---
title: "Get started with gllvmTMB"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Get started with gllvmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  message = FALSE, warning = FALSE
)
```

`gllvmTMB` fits multivariate models for data where each site,
individual, species, or study has several responses. The first
question is usually not mathematical:

> Which responses vary together, and how much variation is shared
> rather than response-specific?

This page does one small, reproducible example. It loads a prepared
morphometrics object, fits the same model from long data and wide
data, checks that the fits agree, and extracts the covariance,
correlation, communality, and ordination summaries a reader normally
wants first.

The example uses the safest public path: Gaussian stacked-trait data
with one observational level, a shared low-rank latent structure, and
trait-specific unique variance. Advanced spatial, phylogenetic,
meta-analytic, mixed-family, and profile-likelihood tutorials remain
under audit until their example objects, diagnostics, and rendered
HTML reviews are ready.

## First fit: Individual × Trait morphology

We have `n` individuals, each measured on `T` morphological traits
across a few sessions. The biological question: do the traits covary
on a small number of latent axes (e.g. a body-size axis, a shape
axis)? With one observational level and modest repetition, both the
shared (`latent()`) and unique (`unique()`) covariance components are
identifiable. No spatial or phylogenetic structure here.

```{r setup}
library(gllvmTMB)
```

For a first fit, load the prepared morphometrics example object. It
contains long data, wide data, the truth table, and the two formulas
used below. The reproducible generator is
`data-raw/examples/make-morphometrics-example.R`, but beginners do not
need to read the generator before fitting the model.

```{r example-data, eval = FALSE}
morph <- readRDS(system.file(
  "extdata", "examples", "morphometrics-example.rds",
  package = "gllvmTMB"
))
df <- morph$data_long
df_wide <- morph$data_wide
truth <- morph$truth
```

```{r example-data-load, include = FALSE}
example_path <- system.file(
  "extdata", "examples", "morphometrics-example.rds",
  package = "gllvmTMB"
)
example_path <- c(
  example_path,
  file.path("inst", "extdata", "examples", "morphometrics-example.rds"),
  file.path("..", "inst", "extdata", "examples", "morphometrics-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "morphometrics-example.rds")
)
example_path <- example_path[nzchar(example_path) & file.exists(example_path)][1]
stopifnot(!is.na(example_path))
morph <- readRDS(example_path)

df <- morph$data_long
df_wide <- morph$data_wide
truth <- morph$truth
trait_names <- truth$trait_names
n_traits <- length(trait_names)
Lambda_true <- truth$Lambda
```

```{r example-data-preview}
morph$story$question
head(df, 6)
```

The example has shared low-rank covariance
$\boldsymbol\Lambda\boldsymbol\Lambda^\top$ plus per-trait unique
variance $\boldsymbol\Psi$. The fit below recovers the individual-level
covariance via `latent()` and `unique()`.

Fit a `d = 2` reduced-rank GLLVM with **`latent() + unique()`** at the
individual level — the shared low-rank structure plus trait-specific
unique variance:

```{r fit-call}
morph$formula_long

fit <- gllvmTMB(
  morph$formula_long,
  data = df,
  trait = morph$fit_args$trait,
  unit = morph$fit_args$unit,
  family = morph$fit_args$family
)
fit
```

### Same model, wide data frame

The same fit reads from a wide data frame (one row per individual, one
column per trait) via the `traits(...)` formula-LHS marker. Internally
the engine pivots to long format, so the fit is byte-identical:

```{r fit-wide}
morph$formula_wide

fit_wide <- gllvmTMB(
  morph$formula_wide,
  data = df_wide,
  unit = morph$fit_args$unit,
  family = morph$fit_args$family
)

# Same fit:
all.equal(as.numeric(logLik(fit)), as.numeric(logLik(fit_wide)))
```

Wide trait tables do not need to be perfectly rectangular. IN under
MIS-21: if an individual is missing one trait measurement, leave that
response cell as `NA`; `gllvmTMB()` drops only the unobserved unit-trait
cell and keeps the other observed traits for that individual. The same
rule applies to long data with `NA` in the response column. OUT: missing
predictors, unit labels, or design-matrix values still error because the
model cannot construct that row.

The recovered loadings (up to column rotation / sign — only
$\boldsymbol{\Lambda}\boldsymbol{\Lambda}^\top$ is identifiable from
the data):

```{r loadings}
Lambda_hat <- fit$report$Lambda_B
rownames(Lambda_hat) <- paste0("trait_", seq_len(n_traits))
colnames(Lambda_hat) <- paste0("LV", seq_len(2))
round(Lambda_hat, 3)
```

The implied trait covariance vs the truth (off-diagonal pattern is
what should agree — diagonals include unique variance):

```{r cov}
list(true_LL  = round(Lambda_true %*% t(Lambda_true), 2),
     fitted_LL = round(Lambda_hat  %*% t(Lambda_hat),  2))
```

The full covariance target is stored in the example object:

```{r sigma-truth}
round(truth$Sigma, 2)
```

Ordination of individuals on the two latent axes:

```{r ord, fig.cap = "Individual scores on the two latent trait axes."}
ord <- extract_ordination(fit, level = "unit")
plot(ord$scores[, 1], ord$scores[, 2], pch = 19, col = "steelblue",
     xlab = "LV 1", ylab = "LV 2",
     main = "Individuals on latent trait axes")
abline(h = 0, v = 0, lty = 2, col = "grey70")
```

That's the whole model — `latent()` for shared trait covariance,
`unique()` for trait-specific residual variance, plus trait-specific
intercepts.

Per-trait communality $c_t^2$ — the proportion of each trait's
variance shared with the others:

```{r communality}
extract_communality(fit, "unit")
```

Pairwise trait correlations with 95% Fisher-z Wald confidence
intervals. `method = "fisher-z"` is the default and keeps bounds in
`[-1, 1]`; `method = "profile"` and `method = "bootstrap"` are slower
alternatives, and `method = "wald"` is kept as an alias of
`"fisher-z"`:

```{r cor}
corr_rows <- extract_correlations(fit, tier = "unit")
corr_rows
```

For reports, plot the same tidy rows instead of hand-indexing a
correlation matrix:

```{r cor-plot, fig.width = 7, fig.height = 3.8, fig.cap = "Pairwise trait correlations from the first fitted model. Points are Fisher-z estimates; horizontal segments are finite 95% interval bounds."}
plot_correlations(corr_rows, sort = "magnitude")
```

The matrix form, if you want to look at the correlation structure as
a whole rather than per pair:

```{r cor-matrix, fig.width = 6.4, fig.height = 4.8, fig.cap = "Trait correlation matrix from the first fitted model. Cells are point estimates; use the pairwise correlation table above when interval bounds matter."}
sigma_corr_rows <- extract_Sigma_table(
  fit,
  level = "unit",
  measure = "correlation",
  entries = "all"
)
plot_Sigma_heatmap(sigma_corr_rows)
```

## Acknowledgements

`gllvmTMB` is a standalone TMB-based package. Its design borrows ideas
from several earlier projects, whose corresponding GPL-3 code is
incorporated and whose authors are credited in `DESCRIPTION` (as
copyright holders) and listed in `inst/COPYRIGHTS`:

* [**sdmTMB**](https://pbs-assess.github.io/sdmTMB/) (Anderson, Ward, English, Barnett, Thorson, et al.): sparse SPDE / GMRF spatial machinery.
* [**glmmTMB**](https://glmmtmb.github.io/glmmTMB/) (Brooks, Bolker, Kristensen, Magnusson, Skaug, Nielsen, Berg, van Bentham, et al.): `rr` / `propto` / `equalto` / `diag` covariance dispatch. The reduced-rank `rr()` machinery in particular is the contribution of [**McGillycuddy, Popovic, Bolker & Warton (2025)**](https://doi.org/10.18637/jss.v112.i01) *J. Stat. Softw.* 112(1).
* [**GALAMM**](https://lcbc-uio.github.io/galamm/) (Sørensen): mixed-response (`gfam`) and confirmatory fixed-loading (`lambda_constraint`) patterns.
* [**MCMCglmm**](https://cran.r-project.org/package=MCMCglmm) (Hadfield): sparse $A^{-1}$ phylogeny representation used by `phylo_latent()`.
* [**TMB**](https://github.com/kaskr/adcomp) (Kristensen et al.) and [**fmesher**](https://inlabru-org.github.io/fmesher/) (Lindgren): autodiff and finite-element mesh.
