---
title: "Get started with gllvmTMB"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Get started with gllvmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  message = FALSE, warning = FALSE
)
```

`gllvmTMB` fits multivariate models for data where each site,
individual, species, or study has several responses. The first
question is usually not mathematical:

> Which responses vary together, and how much variation is shared
> rather than response-specific?

This page does one small, reproducible example. It loads a prepared
morphometrics object, fits the model from a wide trait table first,
checks the equivalent long-data call, and extracts the covariance,
correlation, communality, and ordination summaries a reader normally
wants first.

The example uses the safest public path: Gaussian stacked-trait data
with one observational level, a shared low-rank latent structure, and
trait-specific unique variance. Advanced spatial, phylogenetic,
meta-analytic, mixed-family, and profile-likelihood tutorials remain
under audit until their example objects, diagnostics, and rendered
HTML reviews are ready.

## First fit: wide individual-by-trait morphology

We have `n` individuals, each measured on `T` morphological traits
across a few sessions. The biological question: do the traits covary
on a small number of latent axes (e.g. a body-size axis, a shape
axis)? With one observational level and modest repetition, both the
shared (`latent()`) and unique (`unique()`) covariance components are
identifiable. No spatial or phylogenetic structure here.

```{r setup}
library(gllvmTMB)
```

For a first fit, load the prepared morphometrics example object. It
contains long data, wide data, the truth table, and the two formulas
used below. The reproducible generator is
`data-raw/examples/make-morphometrics-example.R`, but beginners do not
need to read the generator before fitting the model.

```{r example-data, eval = FALSE}
morph <- readRDS(system.file(
  "extdata", "examples", "morphometrics-example.rds",
  package = "gllvmTMB"
))
df <- morph$data_long
df_wide <- morph$data_wide
truth <- morph$truth
```

```{r example-data-load, include = FALSE}
example_path <- system.file(
  "extdata", "examples", "morphometrics-example.rds",
  package = "gllvmTMB"
)
example_path <- c(
  example_path,
  file.path("inst", "extdata", "examples", "morphometrics-example.rds"),
  file.path("..", "inst", "extdata", "examples", "morphometrics-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "morphometrics-example.rds")
)
example_path <- example_path[nzchar(example_path) & file.exists(example_path)][1]
stopifnot(!is.na(example_path))
morph <- readRDS(example_path)

df <- morph$data_long
df_wide <- morph$data_wide
truth <- morph$truth
trait_names <- truth$trait_names
n_traits <- length(trait_names)
Lambda_true <- truth$Lambda
```

```{r example-data-preview}
morph$story$question
head(df, 6)
```

The example has shared low-rank covariance
$\boldsymbol\Lambda\boldsymbol\Lambda^\top$ plus per-trait unique
variance $\boldsymbol\Psi$. The fit below recovers the individual-level
covariance via `latent()` and `unique()`.

Fit a `d = 2` reduced-rank GLLVM with **`latent() + unique()`** at the
individual level — the shared low-rank structure plus trait-specific
unique variance:

```{r fit-call}
morph$formula_wide

fit <- gllvmTMB(
  morph$formula_wide,
  data = df_wide,
  unit = morph$fit_args$unit,
  family = morph$fit_args$family
)
fit
```

Wide trait tables do not need to be perfectly rectangular. IN under
MIS-21: if an individual is missing one trait measurement, leave that
response cell as `NA`; `gllvmTMB()` drops only the unobserved unit-trait
cell and keeps the other observed traits for that individual. The same
rule applies to long data with `NA` in the response column. OUT: missing
predictors, unit labels, or design-matrix values still error because the
model cannot construct that row.

### Same model, long data

If your workflow already stores one row per `(individual, trait)`
observation, use the same entry point with the long formula and explicit
`trait =` argument. Internally the wide formula pivots to this stacked
form, so the fits are byte-identical:

```{r fit-long}
morph$formula_long

fit_long <- gllvmTMB(
  morph$formula_long,
  data = df,
  trait = morph$fit_args$trait,
  unit = morph$fit_args$unit,
  family = morph$fit_args$family
)

# Same fit:
all.equal(as.numeric(logLik(fit)), as.numeric(logLik(fit_long)))
```

## Check the fitted response distribution

Before interpreting `Sigma`, separate two questions. First, ask whether
the numerical fit looks healthy. Then ask whether the fitted response
distribution is visibly mismatched to the observed stacked traits.

The public diagnostic scope is deliberately narrow. IN under DIA-08 /
DIA-10: `check_gllvmTMB()` reports optimizer, gradient, Hessian,
`sdreport()`, restart, boundary, and latent-identifiability status.
IN under DIA-11 / DIA-12: `predictive_check()` and `residuals()` give
fitted-model diagnostic displays with exact randomized-quantile
residuals for Gaussian, Poisson, and NB2 rows. PARTIAL: these checks
are diagnostic displays; they do not calibrate intervals, prove latent
rank, run formal uniformity tests, or draw from a Bayesian posterior.
PLANNED: exact residuals for delta, hurdle, truncated, ordinal, and
mixture families remain future validation work.

```{r fit-health}
health <- check_gllvmTMB(fit)
health[, c("component", "status", "message", "action")]
```

For this Gaussian teaching fit, randomized-quantile residuals are exact
family-CDF residuals. The table keeps every observation row and records
whether the residual could be computed.

```{r residual-table}
rq <- residuals(fit, type = "randomized_quantile", seed = 1)
head(rq[, c("trait", "family", "observed", "residual", "status")], 8)
```

Use the plot as a quick fitted-response check before turning to
covariance interpretation:

```{r residual-qq, fig.width = 6.2, fig.height = 4.2, fig.cap = "Randomized-quantile residual Q-Q check for the fitted Gaussian morphometrics model. The plot is a fitted-model diagnostic display, not interval calibration or a posterior predictive check."}
predictive_check(fit, type = "rq_qq", seed = 1)
```

The recovered loadings (up to column rotation / sign — only
$\boldsymbol{\Lambda}\boldsymbol{\Lambda}^\top$ is identifiable from
the data):

```{r loadings}
Lambda_hat <- fit$report$Lambda_B
rownames(Lambda_hat) <- paste0("trait_", seq_len(n_traits))
colnames(Lambda_hat) <- paste0("LV", seq_len(2))
round(Lambda_hat, 3)
```

The implied trait covariance vs the truth (off-diagonal pattern is
what should agree — diagonals include unique variance):

```{r cov}
list(true_LL  = round(Lambda_true %*% t(Lambda_true), 2),
     fitted_LL = round(Lambda_hat  %*% t(Lambda_hat),  2))
```

The full covariance target is stored in the example object:

```{r sigma-truth}
round(truth$Sigma, 2)
```

Ordination of individuals on the two latent axes:

```{r ord, fig.cap = "Individual scores on the two latent trait axes."}
ord <- extract_ordination(fit, level = "unit")
plot(ord$scores[, 1], ord$scores[, 2], pch = 19, col = "steelblue",
     xlab = "LV 1", ylab = "LV 2",
     main = "Individuals on latent trait axes")
abline(h = 0, v = 0, lty = 2, col = "grey70")
```

That's the whole model — `latent()` for shared trait covariance,
`unique()` for trait-specific residual variance, plus trait-specific
intercepts.

Per-trait communality $c_t^2$ — the proportion of each trait's
variance shared with the others:

```{r communality}
extract_communality(fit, "unit")
```

Pairwise trait correlations with 95% Fisher-z Wald confidence
intervals. `method = "fisher-z"` is the default and keeps bounds in
`[-1, 1]`; `method = "profile"` and `method = "bootstrap"` are slower
alternatives, and `method = "wald"` is kept as an alias of
`"fisher-z"`:

```{r cor}
corr_rows <- extract_correlations(fit, tier = "unit")
corr_rows
```

For reports, plot the same tidy rows instead of hand-indexing a
correlation matrix:

```{r cor-plot, fig.width = 7, fig.height = 3.8, fig.cap = "Pairwise trait correlations from the first fitted model. Points are Fisher-z estimates; horizontal segments are finite 95% interval bounds."}
plot_correlations(corr_rows, sort = "magnitude")
```

The matrix form is IN under EXT-30 as a display over the same rows:
the upper triangle shows point estimates, and the lower triangle shows
the finite Fisher-z interval bounds supplied by
`extract_correlations()`. OUT for EXT-30: the plot does not compute or
calibrate uncertainty itself.

```{r cor-matrix, fig.width = 7.4, fig.height = 5.1, fig.cap = "Trait correlation matrix from the first fitted model. Upper-triangle cells show Fisher-z point estimates; lower-triangle cells show finite 95% interval bounds."}
plot_correlations(
  corr_rows,
  style = "heatmap",
  matrix_layout = "estimate_ci",
  title = "First-model trait correlations"
)
```

## Acknowledgements

`gllvmTMB` is a standalone TMB-based package. Its design borrows ideas
from several earlier projects, whose corresponding GPL-3 code is
incorporated and whose authors are credited in `DESCRIPTION` (as
copyright holders) and listed in `inst/COPYRIGHTS`:

* [**sdmTMB**](https://pbs-assess.github.io/sdmTMB/) (Anderson, Ward, English, Barnett, Thorson, et al.): sparse SPDE / GMRF spatial machinery.
* [**glmmTMB**](https://glmmtmb.github.io/glmmTMB/) (Brooks, Bolker, Kristensen, Magnusson, Skaug, Nielsen, Berg, van Bentham, et al.): `rr` / `propto` / `equalto` / `diag` covariance dispatch. The reduced-rank `rr()` machinery in particular is the contribution of [**McGillycuddy, Popovic, Bolker & Warton (2025)**](https://doi.org/10.18637/jss.v112.i01) *J. Stat. Softw.* 112(1).
* [**GALAMM**](https://lcbc-uio.github.io/galamm/) (Sørensen): mixed-response (`gfam`) and confirmatory fixed-loading (`lambda_constraint`) patterns.
* [**MCMCglmm**](https://cran.r-project.org/package=MCMCglmm) (Hadfield): sparse $A^{-1}$ phylogeny representation used by `phylo_latent()`.
* [**TMB**](https://github.com/kaskr/adcomp) (Kristensen et al.) and [**fmesher**](https://inlabru-org.github.io/fmesher/) (Lindgren): autodiff and finite-element mesh.
