---
title: "Get started with gllvmTMB"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Get started with gllvmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  message = FALSE, warning = FALSE
)
```

`gllvmTMB` fits **generalised linear latent variable models (GLLVMs)**
for unit × trait data — site × species, individual × trait, species
× trait, paper × item, or any other Unit × Trait formulation. The
package is a TMB engine that supports `glmmTMB` / `brms`-style
covariance dispatch via **three covstruct modes** expressed with four
formula keywords:

* **Paired decomposition** — `latent + unique` together gives
  $\boldsymbol\Sigma = \boldsymbol\Lambda\boldsymbol\Lambda^\top + \mathbf S$
  (shared low-rank plus per-trait unique).
* **Marginal only** — `indep` alone gives $\boldsymbol\Sigma = \mathrm{diag}(\sigma^2_t)$
  (per-trait variance, no cross-trait shared structure).
* **Full unstructured** — `dep` alone gives $\boldsymbol\Sigma$ free,
  parameterised via Cholesky.

Built on top: response distributions for binary / count / continuous
/ ordinal / hurdle data, sparse SPDE spatial random fields,
phylogenetic random effects (sparse-$A^{-1}$ representation for big
trees), and multi-tier hierarchies — all turned on **only when the
biological question requires them**. Each topical article in the
side-bar walks through one such extension on a runnable example.

## What a GLLVM estimates

A GLLVM models multivariate data: a matrix of responses
$\mathbf Y = (y_{it})$ with rows $i = 1, \ldots, n$ (units — sites,
individuals, species, …) and columns $t = 1, \ldots, T$ (responses —
species, traits, behaviours, items, …). It combines two familiar
ideas:

* Like a **generalised linear mixed model**, each response gets its
  own distribution and link function $g_t$ so binary, count, and
  continuous responses are all on the same footing.
* Like **factor analysis**, the covariance among the $T$ responses
  is summarised by a small number $d \ll T$ of latent dimensions.

The expected value $\mu_{it} = E(y_{it})$ is linked to a linear
predictor through $g_t(\mu_{it}) = \eta_{it}$, where

\[
\eta_{it} = \mathbf x_i^\top \boldsymbol\beta_t
          + \mathbf z_i^\top \boldsymbol\lambda_t .
\]

* $\mathbf x_i$ are measured predictors with response-specific
  coefficients $\boldsymbol\beta_t$ — the regression part of the
  model.
* $\mathbf z_i$ is a length-$d$ vector of latent scores for unit $i$,
  and $\boldsymbol\lambda_t$ is a length-$d$ loading vector for
  response $t$. Stacking the loadings across responses gives a
  $T \times d$ matrix $\boldsymbol\Lambda$ whose $t$-th row is
  $\boldsymbol\lambda_t$.

In the Gaussian case ($g_t = \mathrm{identity}$), the model can be
written equivalently as

\[
y_{it} \;=\; \eta_{it} + \varepsilon_{it},
\qquad
\varepsilon_{it} \sim \mathcal N(0,\, S_{tt}),
\]

so the leftover $y_{it} - \eta_{it} = \varepsilon_{it}$ — what's
not explained by the regression part and the shared latent dimensions
— has variance $S_{tt}$ specific to response $t$. This is the
**response-specific unique variance** (also called the "specific"
variance in factor analysis). Stacking the $T$ such variances on the
diagonal gives the $T \times T$ **diagonal matrix**

\[
\mathbf S \;=\; \mathrm{diag}(S_{11}, S_{22}, \ldots, S_{TT}) .
\]

The marginal covariance among responses (after measured predictors)
then decomposes as

\[
\boldsymbol\Sigma
\;=\;
\underbrace{\boldsymbol\Lambda\,\boldsymbol\Lambda^\top}_{\text{shared}}
\;+\;
\underbrace{\mathbf S}_{\text{unique}} .
\]

The two pieces are complementary:

* $\boldsymbol\Lambda \boldsymbol\Lambda^\top$ is the $T \times T$
  **shared** covariance among responses, induced through the $d$
  latent dimensions. Because $d \ll T$, the shared component has
  reduced rank — covariance among many responses is captured by a
  few latent dimensions instead of $T(T-1)/2$ free pairwise entries.
  In `gllvmTMB`, $\boldsymbol\Lambda$ is parameterised by the
  formula keyword `latent(0 + trait | unit, d = K)`.
* $\mathbf S$ is the $T \times T$ **diagonal** of response-specific
  unique variances $\mathrm{Var}(\varepsilon_{it}) = S_{tt}$ (one
  entry per response, off-diagonal entries are zero by construction).
  In `gllvmTMB`, $\mathbf S$ is parameterised by
  `unique(0 + trait | unit)`.

A useful per-response summary is **communality**:

\[
c_t^2 \;=\;
\frac{(\boldsymbol\Lambda\boldsymbol\Lambda^\top)_{tt}}
     {(\boldsymbol\Lambda\boldsymbol\Lambda^\top)_{tt} + S_{tt}} .
\]

The numerator is the variance of response $t$ explained by the shared
latent structure; the denominator is the total model-implied residual
variance of response $t$. So $c_t^2$ is the proportion of response
$t$'s variance shared with the others. High $c_t^2$ means response
$t$ is strongly integrated; low $c_t^2$ means response-specific
variation dominates.

A fitted GLLVM therefore returns:

* **Regression coefficients** $\boldsymbol\beta_t$ for measured
  predictors.
* **Latent scores** $\mathbf Z$ placing units in $d$-dimensional
  latent space (model-based ordination).
* **Loadings** $\boldsymbol\Lambda$ describing how strongly each
  response is associated with each latent dimension (biplot arrows).
* **Residual covariance / correlation**
  $\boldsymbol\Sigma = \boldsymbol\Lambda\boldsymbol\Lambda^\top + \mathbf S$
  among responses.
* **Communality** $c_t^2$ per response.

## Example: Individual × Trait morphology

We have `n` individuals, each measured on `T` morphological traits
across a few sessions. The biological question: do the traits covary
on a small number of latent axes (e.g. a body-size axis, a shape
axis)? With one observational level and modest repetition, both the
shared (`latent()`) and unique (`unique()`) covariance components are
identifiable. No spatial or phylogenetic structure here.

```{r setup}
library(gllvmTMB)
```

Simulate `n = 60` individuals × `T = 5` traits × 3 sessions with two
true latent axes plus trait-specific unique variance:

```{r sim}
set.seed(1)
n_indiv   <- 60
n_session <- 3
n_traits  <- 5
trait_names <- paste0("trait_", seq_len(n_traits))

Lambda_true <- matrix(c( 1.0,  0.6, -0.4,  0.2,  0.5,
                         0.0,  0.8,  0.5, -0.3,  0.2),
                      nrow = n_traits, ncol = 2)
S_B_true     <- c(0.20, 0.15, 0.30, 0.20, 0.25)
sigma_eps    <- 0.20
alpha        <- c(0, 0.5, -0.3, 0.2, 0.1)

# Stable per-individual trait values (between-individual rr + unique)
u <- matrix(rnorm(n_indiv * 2), n_indiv, 2)
between_ind <- matrix(alpha, n_indiv, n_traits, byrow = TRUE) +
               u %*% t(Lambda_true) +
               matrix(rnorm(n_indiv * n_traits,
                            sd = sqrt(S_B_true[col(matrix(0, n_indiv, n_traits))])),
                      n_indiv, n_traits)

# Three sessions per individual, each with measurement noise
df <- expand.grid(individual_id = seq_len(n_indiv),
                  session_id    = seq_len(n_session),
                  trait_idx     = seq_len(n_traits))
df$value <- between_ind[cbind(df$individual_id, df$trait_idx)] +
            rnorm(nrow(df), sd = sigma_eps)
df$individual <- factor(df$individual_id)
df$trait      <- factor(trait_names[df$trait_idx], levels = trait_names)
df$session    <- factor(paste(df$individual_id, df$session_id, sep = "_"))
df <- df[, c("individual", "session", "trait", "value")]
head(df, 6)
```

The DGP has three variance sources: shared low-rank
$\boldsymbol\Lambda\boldsymbol\Lambda^\top$, per-trait
between-individual unique $\mathbf S_B$, and per-session noise
$\sigma^2_\varepsilon$. The fit below recovers the first two via
`latent()` and `unique()`; the per-session noise is absorbed into the
Gaussian residual variance $\sigma^2_d$ (one parameter per trait).

Fit a `d = 2` reduced-rank GLLVM with **`latent() + unique()`** at the
individual level — the shared low-rank structure plus trait-specific
unique variance:

```{r fit-call}
fit <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | individual, d = 2) +
          unique(0 + trait | individual),
  data = df,
  unit = "individual"
)
fit
```

The recovered loadings (up to column rotation / sign — only
$\boldsymbol{\Lambda}\boldsymbol{\Lambda}^\top$ is identifiable from
the data):

```{r loadings}
Lambda_hat <- fit$report$Lambda_B
rownames(Lambda_hat) <- paste0("trait_", seq_len(n_traits))
colnames(Lambda_hat) <- paste0("LV", seq_len(2))
round(Lambda_hat, 3)
```

The implied trait covariance vs the truth (off-diagonal pattern is
what should agree — diagonals include unique variance):

```{r cov}
list(true_LL  = round(Lambda_true %*% t(Lambda_true), 2),
     fitted_LL = round(Lambda_hat  %*% t(Lambda_hat),  2))
```

Ordination of individuals on the two latent axes:

```{r ord, fig.cap = "Individual scores on the two latent trait axes."}
ord <- extract_ordination(fit, level = "unit")
plot(ord$scores[, 1], ord$scores[, 2], pch = 19, col = "steelblue",
     xlab = "LV 1", ylab = "LV 2",
     main = "Individuals on latent trait axes")
abline(h = 0, v = 0, lty = 2, col = "grey70")
```

That's the whole model — `latent()` for shared trait covariance,
`unique()` for trait-specific residual variance, plus trait-specific
intercepts.

Per-trait communality $c_t^2$ — the proportion of each trait's
variance shared with the others:

```{r communality}
extract_communality(fit, "unit")
```

Pairwise trait correlations with 95% Wald confidence intervals.
`method = "wald"` is the default and uses a Fisher transform internally
to keep bounds in `[-1, 1]`; `method = "profile"` and
`method = "bootstrap"` are slower alternatives. The old
`method = "fisher-z"` spelling remains as a compatibility alias:

```{r cor}
extract_correlations(fit, tier = "unit")
```

The matrix form, if you want to look at the correlation structure as
a whole rather than per pair:

```{r cor-matrix}
round(cov2cor(extract_Sigma(fit, level = "unit")$Sigma), 2)
```

## Acknowledgements

`gllvmTMB` is a standalone TMB-based package. Its design borrows ideas
from several earlier projects, whose corresponding GPL-3 code is
incorporated and whose authors are credited in `DESCRIPTION` (as
copyright holders) and listed in `inst/COPYRIGHTS`:

* [**sdmTMB**](https://pbs-assess.github.io/sdmTMB/) (Anderson, Ward, English, Barnett, Thorson, et al.): sparse SPDE / GMRF spatial machinery.
* [**glmmTMB**](https://glmmtmb.github.io/glmmTMB/) (Brooks, Bolker, Kristensen, Magnusson, Skaug, Nielsen, Berg, van Bentham, et al.): `rr` / `propto` / `equalto` / `diag` covariance dispatch. The reduced-rank `rr()` machinery in particular is the contribution of [**McGillycuddy, Popovic, Bolker & Warton (2025)**](https://doi.org/10.18637/jss.v112.i01) *J. Stat. Softw.* 112(1).
* [**GALAMM**](https://lcbc-uio.github.io/galamm/) (Sørensen): mixed-response (`gfam`) and confirmatory fixed-loading (`lambda_constraint`) patterns.
* [**MCMCglmm**](https://cran.r-project.org/package=MCMCglmm) (Hadfield): sparse $A^{-1}$ phylogeny representation used by `phylo_latent()`.
* [**TMB**](https://github.com/kaskr/adcomp) (Kristensen et al.) and [**fmesher**](https://inlabru-org.github.io/fmesher/) (Lindgren): autodiff and finite-element mesh.
