---
title: "Get started with gllvmTMB"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Get started with gllvmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  message = FALSE, warning = FALSE
)
```

`gllvmTMB` fits multivariate models for data where each site,
individual, species, or study has several responses. The first
question is usually not mathematical:

> Which responses vary together, and how much variation is shared
> rather than response-specific?

This page does one small, reproducible example. It loads a prepared
morphometrics object, fits the model from a wide trait table first,
checks the equivalent long-data call, and extracts the covariance,
correlation, communality, and ordination summaries a reader normally
wants first.

`gllvmTMB` is experimental and pre-CRAN (version 0.5.0): the formula
grammar, defaults, and reported summaries and intervals may still
change before the API is committed-stable.

The fitted Gaussian model starts with observations, not with the
covariance matrix. For each individual and trait, the teaching model is:

$$
\begin{aligned}
y_{it}
&=
\alpha_t + \boldsymbol{\lambda}_t^{\mathsf T}\mathbf{u}_i + e_{it},\\
\mathbf{u}_i
&\sim \mathcal{N}(\mathbf{0}, \mathbf{I}_d),\\
e_{it}
&\sim \mathcal{N}(0, \psi_t).
\end{aligned}
$$

Here $y_{it}$ is the observed value for individual $i$ and trait $t$.
The intercept $\alpha_t$ is the average for trait $t$. The vector
$\mathbf{u}_i$ is the individual's latent score: where that individual
sits on the unobserved trait axes. The row vector
$\boldsymbol{\lambda}_t$ says how strongly trait $t$ responds to those
axes. The leftover term $e_{it}$ is the part of trait $t$ not shared
with the other traits.

In generalized models, $\eta_{it}$ is the linear predictor and the link
function $g()$ connects it to the mean response:
$g(\mu_{it}) = \eta_{it}$. This first example uses a Gaussian family
with the identity link, so the linear predictor is the mean itself.

The R formula for that model is compact:

```{r first-model-sketch, eval = FALSE}
fit <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~ 1 +
    latent(1 | individual, d = 2),
  data = df_wide,
  unit = "individual",
  family = gaussian()
)
```

After fitting, the covariance among traits is the first report-ready
summary implied by the latent part of the model:

$$
\begin{aligned}
\operatorname{Cov}(\mathbf{y}_i)
&=
\boldsymbol{\Sigma},\\
\boldsymbol{\Sigma}
&=
\boldsymbol{\Lambda}\boldsymbol{\Lambda}^{\mathsf T}
+
\boldsymbol{\Psi},\\
\boldsymbol{\Psi}
&=
\operatorname{diag}(\psi_1,\ldots,\psi_T).
\end{aligned}
$$

In words: total trait covariance = shared multivariate structure +
response-specific variation. `Lambda` collects the trait loadings: one
row per trait and one column per latent variable. The product
`Lambda Lambda^T` is the shared covariance created by those axes. `Psi`
is the diagonal matrix of trait-specific variance left over after the
shared axes. Ordinary `latent()` estimates both pieces by default.

For a first fit, read `Sigma`, correlations, and communality before
reading individual `Lambda` entries. The latent variables can be flipped or
turned without changing the covariance they imply, so raw loading
columns need a rotation or confirmatory constraint before they become a
biological story.

## First fit: wide individual-by-trait morphology

We have `n` individuals, each measured on `T` morphological traits
across a few sessions. The biological question: do the traits covary
on a small number of latent variables (e.g. a body-size axis, a shape
axis)? With one observational level and modest repetition, ordinary
`latent()` estimates both the shared low-rank component and the
trait-specific diagonal `Psi`. No spatial or phylogenetic structure
here.

```{r setup}
library(gllvmTMB)
```

For a first fit, load the prepared morphometrics example object. It
contains long data, wide data, the truth table, and the two formulas
used below. The reproducible generator is
`data-raw/examples/make-morphometrics-example.R`, but beginners do not
need to read the generator before fitting the model.

```{r example-data, eval = FALSE}
morph <- readRDS(system.file(
  "extdata", "examples", "morphometrics-example.rds",
  package = "gllvmTMB"
))
df <- morph$data_long
df_wide <- morph$data_wide
truth <- morph$truth
```

```{r example-data-load, include = FALSE}
example_path <- system.file(
  "extdata", "examples", "morphometrics-example.rds",
  package = "gllvmTMB"
)
example_path <- c(
  file.path("inst", "extdata", "examples", "morphometrics-example.rds"),
  file.path("..", "inst", "extdata", "examples", "morphometrics-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "morphometrics-example.rds"),
  example_path
)
example_path <- example_path[nzchar(example_path) & file.exists(example_path)][1]
stopifnot(!is.na(example_path))
morph <- readRDS(example_path)

df <- morph$data_long
df_wide <- morph$data_wide
truth <- morph$truth
trait_names <- truth$trait_names
n_traits <- length(trait_names)
Lambda_true <- truth$Lambda
```

```{r example-data-preview}
morph$story$question
head(df, 6)
```

The example has shared low-rank covariance
$\boldsymbol\Lambda\boldsymbol\Lambda^\top$ plus per-trait diagonal
variance $\boldsymbol\Psi$. The fit below recovers the individual-level
covariance via ordinary `latent()`.

In wide data, the formula term is
`latent(1 | individual, d = 2)`. In long data, the equivalent term is
`latent(0 + trait | individual, d = 2)`. The `1` in the wide form is
the compact shorthand for the same trait-specific columns used in the
long form.

Fit a `d = 2` reduced-rank GLLVM with ordinary `latent()` at the
individual level — the shared low-rank structure plus trait-specific
unique variance, which ordinary `latent()` carries by default:

```{r fit-call}
morph$formula_wide

fit <- gllvmTMB(
  morph$formula_wide,
  data = df_wide,
  unit = morph$fit_args$unit,
  family = morph$fit_args$family
)

# A finite likelihood confirms that the model fitted; diagnostics come next.
as.numeric(logLik(fit))
```

### Same model, long data

If your workflow already stores one row per `(individual, trait)`
observation, use the same entry point with the long formula and explicit
`trait =` argument. Internally the wide formula pivots to this stacked
form, so the two calls fit the same model — the log-likelihoods agree:

```{r fit-long}
morph$formula_long

fit_long <- gllvmTMB(
  morph$formula_long,
  data = df,
  trait = morph$fit_args$trait,
  unit = morph$fit_args$unit,
  family = morph$fit_args$family
)

# Same fit:
all.equal(as.numeric(logLik(fit)), as.numeric(logLik(fit_long)))
```

### What if some response cells are missing?

Wide trait tables do not need to be perfectly rectangular. If an
individual is missing one trait measurement, leave that response cell as
`NA`; `gllvmTMB()` can treat only that unit-trait cell as missing and
keep the other observed traits for the individual. The same rule applies
to long data with `NA` in the response column. Missing predictors and
missing grouping variables are a separate problem: by default they fail
loudly because the model cannot construct that row.

## Check the fitted response distribution

Before interpreting `Sigma`, separate two questions. First, ask whether
the numerical fit looks healthy. Then ask whether the fitted response
distribution is visibly mismatched to the observed stacked traits. For
this Gaussian example, use `check_gllvmTMB()` first; if a row is not
`PASS`, read its `action` before interpreting covariance. The residual
and predictive-check plots are diagnostic displays here. They do not
prove the latent rank or calibrate uncertainty intervals.

```{r fit-health}
health <- check_gllvmTMB(fit)
health[, c("component", "status", "message", "action")]
```

For this Gaussian teaching fit, randomized-quantile residuals are exact
family-CDF residuals. The table keeps every observation row and records
whether the residual could be computed.

```{r residual-table}
rq <- residuals(fit, type = "randomized_quantile", seed = 1)
head(rq[, c("trait", "family", "observed", "residual", "status")], 8)
```

Use the plot as a quick fitted-response check before turning to
covariance interpretation:

```{r residual-qq, fig.width = 6.2, fig.height = 4.2, fig.cap = "Randomized-quantile residual Q-Q check for the fitted Gaussian morphometrics model. The plot is a fitted-model diagnostic display, not interval calibration or a posterior predictive check."}
predictive_check(fit, type = "rq_qq", seed = 1)
```

## Read the covariance summaries first

Start with the fitted `Sigma` rows. These rows describe the covariance
surface implied by the model, not a separate model:

```{r sigma-table}
sigma_rows <- extract_Sigma_table(fit, level = "unit")
sigma_rows
```

The full covariance target is stored in the example object, so this
teaching example can show what the fitted model is trying to recover:

```{r sigma-truth}
round(truth$Sigma, 2)
```

Ordination of individuals on the two latent axes. These axes come from
the fitted latent model
($\boldsymbol\Sigma = \boldsymbol\Lambda\boldsymbol\Lambda^\top + \boldsymbol\Psi$),
not an eigen-decomposition of the raw data — this is model-based
ordination, not PCA or NMDS:

```{r ord, fig.cap = "Individual scores on the two latent trait axes."}
ord <- extract_ordination(fit, level = "unit")
plot(ord$scores[, 1], ord$scores[, 2], pch = 19, col = "steelblue",
     xlab = "LV 1", ylab = "LV 2",
     main = "Individuals on latent trait axes")
abline(h = 0, v = 0, lty = 2, col = "grey70")
```

That's the whole model — `latent()` for shared trait covariance and the
diagonal trait-specific $\boldsymbol\Psi$ component, plus trait-specific
intercepts.

Per-trait communality $c_t^2$ — the proportion of each trait's
variance shared with the others:

```{r communality}
extract_communality(fit, "unit")
```

Pairwise trait correlations with 95% Fisher-z Wald confidence
intervals. `method = "fisher-z"` is the default and keeps bounds in
`[-1, 1]`; `method = "profile"` and `method = "bootstrap"` are slower
alternatives for follow-up articles, and `method = "wald"` is kept as
an alias of `"fisher-z"`. These are nominal Fisher-z Wald intervals:
they are useful diagnostics, but their empirical coverage is not
calibrated, so read them as recovery-grade, not as
validated coverage.

```{r cor}
corr_rows <- extract_correlations(fit, tier = "unit")
corr_rows
```

For reports, plot the same tidy rows instead of hand-indexing a
correlation matrix:

```{r cor-plot, fig.width = 7, fig.height = 3.8, fig.cap = "Pairwise trait correlations from the first fitted model. Points are Fisher-z estimates; horizontal segments are finite 95% interval bounds."}
plot_correlations(corr_rows, sort = "magnitude")
```

The matrix form is a display over the same rows: the upper triangle
shows point estimates, and the lower triangle shows the finite Fisher-z
interval bounds supplied by `extract_correlations()`. The plot does not
compute or calibrate uncertainty itself.

```{r cor-matrix, fig.width = 7.4, fig.height = 5.1, fig.cap = "Trait correlation matrix from the first fitted model. Upper-triangle cells show Fisher-z point estimates; lower-triangle cells show finite 95% interval bounds."}
plot_correlations(
  corr_rows,
  style = "heatmap",
  matrix_layout = "estimate_ci",
  title = "First-model trait correlations"
)
```

## Then inspect loadings and scores

The recovered loadings are useful, but they are not the first
interpretation target. Only
$\boldsymbol{\Lambda}\boldsymbol{\Lambda}^{\mathsf T}$ is identified
directly from the data; raw loading columns can change sign or rotate.

```{r loadings}
Lambda_hat <- getLoadings(fit, level = "unit")
rownames(Lambda_hat) <- paste0("trait_", seq_len(n_traits))
colnames(Lambda_hat) <- paste0("LV", seq_len(ncol(Lambda_hat)))
round(Lambda_hat, 3)
```

Do not over-interpret the signs of these unrotated loadings. Use them
as a quick look at the latent variables, then use `Sigma`, correlations, and
communality for report-ready interpretation unless you have chosen a
rotation or confirmatory constraint.

The shared covariance from the loadings can still be compared with the
known truth in this simulated teaching object. The off-diagonal pattern
is what should agree; the full `Sigma` diagonal also includes `Psi`.

```{r cov}
list(true_LL  = round(Lambda_true %*% t(Lambda_true), 2),
     fitted_LL = round(Lambda_hat  %*% t(Lambda_hat),  2))
```

That is the whole first model: trait-specific intercepts, individual
latent scores, loadings from those scores to traits, and the diagonal
`Psi` carried by ordinary `latent()`.

## Acknowledgements

`gllvmTMB` is a standalone TMB-based package. Its design borrows ideas
from several earlier projects, whose corresponding GPL-3 code is
incorporated and whose authors are credited in `DESCRIPTION` (as
copyright holders) and listed in `inst/COPYRIGHTS`:

* [**sdmTMB**](https://pbs-assess.github.io/sdmTMB/) (Anderson, Ward, English, Barnett, Thorson, et al.): sparse SPDE / GMRF spatial machinery.
* [**glmmTMB**](https://glmmtmb.github.io/glmmTMB/) (Brooks, Bolker, Kristensen, Magnusson, Skaug, Nielsen, Berg, van Bentham, et al.): `rr` / `propto` / `equalto` / `diag` covariance dispatch. The reduced-rank `rr()` machinery in particular is the contribution of [**McGillycuddy, Popovic, Bolker & Warton (2025)**](https://doi.org/10.18637/jss.v112.i01) *J. Stat. Softw.* 112(1).
* [**GALAMM**](https://lcbc-uio.github.io/galamm/) (Sørensen): mixed-response (`gfam`) and confirmatory fixed-loading (`lambda_constraint`) patterns.
* [**MCMCglmm**](https://cran.r-project.org/package=MCMCglmm) (Hadfield): sparse $A^{-1}$ phylogeny representation used by `phylo_latent()`.
* [**TMB**](https://github.com/kaskr/adcomp) (Kristensen et al.) and [**fmesher**](https://inlabru-org.github.io/fmesher/) (Lindgren): autodiff and finite-element mesh.
