---
title: "Two-stage workflow for species-level trait data (Corvidae-style proxy)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Two-stage workflow for species-level trait data (Corvidae-style proxy)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

> **Note on data.** The original Nakagawa et al. (in prep)
> functional-biogeography manuscript demonstrates this workflow on
> the Corvidae 123-species AVONET-style dataset. To keep this
> article self-contained and avoid data-licensing concerns, we use
> a **simulated proxy** with the same structure (~30 species x 6
> traits, with species-level trait constants and a coalescent
> phylogeny). The numbers in this article are not real Corvidae
> values -- they illustrate the workflow only.

When trait values are **species-level constants** (typical of
AVONET-style trait data: each species has one beak length, one
wing length, etc., not multiple replicates per species), the
full one-shot stacked-trait GLLVM is poorly informed about
between-site covariance -- the data simply do not contain enough
sites-of-the-same-species to estimate global functional
integration directly. The recommended workflow is **two-stage**:

* **Stage 1** -- fit one model per trait,
  `value ~ 0 + site + phylo_scalar(species)`,
  to extract site-level fixed-effect estimates and their standard
  errors (which give the marginal sampling variances).

* **Stage 2** -- fit a multivariate site-level model on the
  stage-1 estimates using `meta_known_V(value, V = V)` so that
  stage-1 sampling uncertainty is propagated into the stage-2
  covariance estimates (rather than treating the stage-1 values
  as error-free data).

Both stages use **direct `gllvmTMB()` calls** -- no helper
functions are needed.

## Setup

```{r setup}
library(gllvmTMB)
library(ape)
```

## Simulate a Corvidae-style proxy

We simulate ~30 species x 6 traits with a known between-site
loading structure: traits 1-3 (beak morphology) load on one
latent variable and traits 4-6 (body size) load on another. A
coalescent phylogeny gives realistic between-species correlation.

```{r sim}
set.seed(123)
n_sp     <- 30
n_traits <- 6

tree <- ape::rcoal(n_sp)
tree$tip.label <- paste0("sp", seq_len(n_sp))
Cphy <- ape::vcv(tree, corr = TRUE)

trait_names <- c("Beak.Length", "Beak.Width", "Beak.Depth",
                 "Tarsus.Length", "Wing.Length", "Mass")

# True between-site reduced-rank loadings (hand-set so traits 1-3
# load on a "beak" axis and traits 4-6 on a "body-size" axis)
Lambda_B_true <- matrix(c(
   1.0,  0.7,  0.6,  0.1,  0.2,  0.0,   # LV 1: beak axis
   0.0,  0.2, -0.1,  0.8,  1.0,  0.9    # LV 2: body axis
), nrow = n_traits, ncol = 2)

sim <- simulate_site_trait(
  n_sites               = 40,
  n_species             = n_sp,
  n_traits              = n_traits,
  mean_species_per_site = 8,
  Lambda_B   = Lambda_B_true,
  psi_B      = rep(0.2, n_traits),
  Cphy       = Cphy,
  sigma2_phy = rep(0.5, n_traits),
  seed       = 123
)
df_long <- sim$data
levels(df_long$species) <- tree$tip.label
levels(df_long$trait)   <- trait_names

dim(df_long)
table(df_long$trait)[1:3]
```

## Stage 1: per-trait gllvmTMB() fits with phylogenetic random effect

For each trait we fit a univariate model with a site fixed effect
and a phylogenetic random effect on species. The formula
`value ~ 0 + site + phylo_scalar(species)` gives one coefficient
per site (the site-level BLUP) and its standard error (which
becomes the per-cell sampling variance at stage 2).

```{r stage1}
trait_levels <- levels(df_long$trait)
stage1_fits <- lapply(trait_levels, function(tl) {
  sub <- droplevels(df_long[df_long$trait == tl, , drop = FALSE])
  sp_here <- levels(sub$species)
  C_sub   <- Cphy[sp_here, sp_here, drop = FALSE]
  gllvmTMB(
    value ~ 0 + site + phylo_scalar(species, vcv = C_sub),
    data  = sub,
    trait = "trait",
    unit  = "site"
  )
})
names(stage1_fits) <- trait_levels
```

Extract site-level estimates and sampling variances from each
fit. The fixed effects are the `site*` coefficients; their
squared standard errors are the per-cell sampling variances that
stage 2 propagates.

```{r stage1-extract}
stage1_list <- lapply(trait_levels, function(tl) {
  td <- tidy(stage1_fits[[tl]], "fixed", conf.int = FALSE)
  site_rows <- grepl("^site", td$term)
  td <- td[site_rows, , drop = FALSE]
  data.frame(
    site         = factor(sub("^site", "", td$term)),
    trait        = factor(tl, levels = trait_levels),
    value        = td$estimate,
    sampling_var = td$std.error^2,
    stringsAsFactors = FALSE,
    row.names    = NULL
  )
})

stage1_summary <- do.call(rbind, stage1_list)
rownames(stage1_summary) <- NULL

# Keep only sites that appear in every trait (required for stacking)
site_counts <- table(stage1_summary$site)
keep_sites  <- names(site_counts)[site_counts == length(trait_levels)]
stage1_summary <- droplevels(
  stage1_summary[as.character(stage1_summary$site) %in% keep_sites, ]
)

head(stage1_summary, 8)
```

Each row of `stage1_summary` is one `(site, trait)` pair with the
site-level estimate and its propagated sampling variance.

## Stage 2: multivariate site-level meta-regression

Stage 2 fits a between-site reduced-rank covariance on the
stage-1 estimates. We call `gllvmTMB()` directly with
`meta_known_V()` to inject the stage-1 sampling variances as a
known-V term:

```{r stage2}
# Build the known-V diagonal matrix (independent rows: one
# estimate per cell).
V <- diag(stage1_summary$sampling_var)

fit2 <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | site, d = 2) +
          unique(0 + trait | site) +
          meta_known_V(value, V = V),
  data    = stage1_summary,
  known_V = V,
  trait   = "trait",
  unit    = "site"
)
fit2$opt$convergence
```

`meta_known_V(value, V = V)` fixes the residual covariance of the
observed stage-1 estimates to the known sampling-variance matrix
`V`. The `latent(0 + trait | site, d = 2)` term recovers the
two-factor between-site structure that generated the data.

### When rows share within-study sampling correlation

If multiple effect sizes come from the same study (e.g. multiple
traits measured on the same individual), the off-diagonal blocks
of `V` are non-zero. Use `block_V()` to build a block-diagonal
`V`:

```{r block-v-example, eval = FALSE}
# Example (not run): 3 effect sizes per study, compound-symmetric
# within-study correlation
V_blk <- block_V(
  study_id     = stage1_summary$site,
  sampling_var = stage1_summary$sampling_var,
  rho_within   = 0.4
)

fit2_blk <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | site, d = 2) +
          meta_known_V(value, V = V_blk),
  data    = stage1_summary,
  known_V = V_blk,
  trait   = "trait",
  unit    = "site"
)
```

## Compare recovered Sigma_B to truth

The headline recovery check: the implied between-site covariance
$\boldsymbol{\Sigma}_B = \boldsymbol\Lambda_B
\boldsymbol\Lambda_B^{\top}$ should match the truth (up to
rotation; the diagonal entries and off-diagonal pattern are
rotation-invariant).

```{r compare-sigma}
Sigma_B_true  <- Lambda_B_true %*% t(Lambda_B_true)
Sigma_B_hat   <- fit2$report$Lambda_B %*% t(fit2$report$Lambda_B)
rownames(Sigma_B_true) <- colnames(Sigma_B_true) <- trait_names
rownames(Sigma_B_hat)  <- colnames(Sigma_B_hat)  <- trait_names

list(true      = round(Sigma_B_true, 2),
     recovered = round(Sigma_B_hat, 2))
```

The off-diagonal pattern of the recovered $\boldsymbol\Sigma_B$
should match the truth: traits 1-3 (beak) positively correlated
with each other and weakly correlated with traits 4-6 (body);
traits 4-6 positively correlated with each other.

```{r plot-sigma, fig.cap = "Recovered Sigma_B entries vs the true values used to simulate the data. Points should lie along the 1:1 line; deviations come from sampling noise (single simulation; the recovery improves with more sites and more replication per species).", eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
df_sigma <- data.frame(
  true   = as.vector(Sigma_B_true),
  fitted = as.vector(Sigma_B_hat)
)
ggplot(df_sigma, aes(true, fitted)) +
  geom_abline(slope = 1, intercept = 0,
              colour = "grey40", linetype = 2) +
  geom_point(colour = "#1f78b4", size = 2) +
  labs(x = "True Sigma_B[i,j]",
       y = "Recovered Sigma_B[i,j]",
       title = "Stage-2 recovery of the between-site trait covariance",
       subtitle = "Single Corvidae-style proxy run") +
  theme_minimal(base_size = 12) +
  coord_equal()
```

## When to use the two-stage workflow

* **Use it** when each species has only one trait value (the
  AVONET-style scenario) and your sites contribute species
  occupancy / abundance information rather than repeated trait
  measurement.
* **Do not use it** when you have replicate trait measurements per
  species per site -- the one-shot stacked-trait GLLVM can fit
  those directly. See
  [Functional biogeography](functional-biogeography.html).
* **The propagation matters**: stage 2 with `meta_known_V()`
  correctly propagates the stage-1 sampling variances. Naively
  feeding stage-1 estimates as if they were error-free data will
  **under-estimate** Sigma_B uncertainty.

## Summary: the direct-call pattern

You do not need helper functions. The full two-stage workflow in
three steps:

```r
# Step 1 -- per-trait gllvmTMB() fits
stage1_fits <- lapply(trait_levels, function(tl) {
  sub   <- droplevels(df_long[df_long$trait == tl, ])
  C_sub <- Cphy[levels(sub$species), levels(sub$species)]
  gllvmTMB(value ~ 0 + site + phylo_scalar(species, vcv = C_sub),
           data = sub)
})
names(stage1_fits) <- trait_levels

# Step 2 -- assemble stage-1 summary (estimates + sampling variances)
stage1_summary <- do.call(rbind, lapply(trait_levels, function(tl) {
  td <- tidy(stage1_fits[[tl]], "fixed")
  td <- td[grepl("^site", td$term), ]
  data.frame(site = factor(sub("^site", "", td$term)),
             trait = factor(tl), value = td$estimate,
             sampling_var = td$std.error^2)
}))

# Step 3 -- stage-2 multivariate fit with meta_known_V()
V    <- diag(stage1_summary$sampling_var)
fit2 <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2) +
                      meta_known_V(value, V = V),
  data    = stage1_summary,
  known_V = V,
  trait   = "trait",
  unit    = "site"
)
```

## See also

* [Functional biogeography](functional-biogeography.html) -- the
  one-shot version when replicate trait data are available.
* [Phylogenetic GLLVM](phylogenetic-gllvm.html) -- the paired
  phylogenetic + non-phylogenetic decomposition used as the
  one-shot alternative for species-level data with replication.
* [Choose your model](choose-your-model.html) -- decision tree for
  which workflow matches your data shape.
