---
title: "Mixed responses: Gaussian + binomial + Poisson rows in one model"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Mixed responses: Gaussian + binomial + Poisson rows in one model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

Some Unit x Trait datasets mix response distributions: one trait is
continuous (e.g. body mass), another is binary (e.g. presence /
absence of a behaviour), a third is a count (e.g. number of
offspring). Standard JSDM software typically forces all responses
into one family and either log-transforms / scales or drops responses
that don't fit. `gllvmTMB` borrows GALAMM's `gfam(list(...))`
mechanism (Sorensen 2024) and lets the user specify a different
response family per row -- all in the same fit, with shared latent
structure.

## API

Pass a **list** of family objects via the `family` argument and add
a `family_var` attribute pointing to the column of `data` that picks
the family per row:

```r
fam <- list(gaussian(), binomial(), poisson())
attr(fam, "family_var") <- "family"   # name of the column on `data`

fit <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2),
  data   = df,
  family = fam
)
```

Each row of `data` whose `family` value is `"g"` (the first level of
the `family` factor, mapped to family code 0 internally) goes through
the Gaussian (identity-link) likelihood; rows with `"b"` go through
the binomial (logit), and `"p"` rows through the Poisson (log).
Latent structure (`latent`, `unique`, `phylo_*`, `spatial_*`, ...) is
shared across families. The implementation is verified by
`tests/testthat/test-stage37-mixed-family.R`.

## Worked example

Three "traits", one per response distribution. We simulate Gaussian
data on the link scale, then transform two of the columns to binary
and to counts so each trait is on its natural scale.

```{r setup}
library(gllvmTMB)
```

```{r sim}
set.seed(2025)
n_traits <- 3
sim <- simulate_site_trait(
  n_sites               = 50,
  n_species             = 10,
  n_traits              = n_traits,
  mean_species_per_site = 5,
  Lambda_B = matrix(c(1.0,  0.7, -0.3,
                      0.3, -0.5,  0.8),
                    nrow = n_traits, ncol = 2),
  psi_B    = rep(0.3, n_traits),
  seed     = 2025
)
df <- sim$data

# Recode trait_1 as Gaussian (raw), trait_2 as binary, trait_3 as count
df$family <- factor(
  with(df, ifelse(trait == "trait_1", "g",
           ifelse(trait == "trait_2", "b", "p"))),
  levels = c("g", "b", "p")
)
df$value[df$family == "b"] <- as.integer(df$value[df$family == "b"] > 0)
df$value[df$family == "p"] <- pmax(0L,
  as.integer(round(df$value[df$family == "p"] + 1))
)

table(family = df$family)
head(df[, c("site", "species", "trait", "family", "value")], 6)
```

## Fit

```{r fit}
fam <- list(gaussian(), binomial(), poisson())
attr(fam, "family_var") <- "family"

fit <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2),
  data   = df,
  family = fam
)
fit
```

The `family_id_vec` data field stores the per-row family code
(0 = Gaussian, 1 = binomial, 2 = Poisson). Every row gets the correct
one:

```{r family-id}
table(family_id_vec = fit$tmb_data$family_id_vec)
```

The wide-format `traits(...)` entry point cannot express per-row
family dispatch directly, because per-row family selection needs a
long-format `family` column (one value per `(unit, trait)` row). For
fully mixed-family fits the canonical recipe is the long-format
formula above. For a Gaussian-only subset of the same traits the
wide form is fine; this article keeps the focus on the mixed-family
mechanic.

## What's shared, what's per-family

* **Shared across families**: the latent scores `Z` (one per site,
  length `d = 2`) and the loadings `Lambda` (`n_traits x d`). This is
  the point of fitting the three trait columns jointly -- they share
  latent structure.
* **Per-family**: the link function and the observation likelihood.
  The Gaussian rows contribute a `dnorm()` term (with a single
  `sigma_eps`), the binomial rows a
  `dbinom(size = 1, prob = plogis(eta))` term, and the Poisson rows a
  `dpois(lambda = exp(eta))` term.

## Currently supported families

The list passed to `family =` can mix any of these 15 families.
Each row of `data` picks the family from the list via the
`family_var` column:

| Family | `family_id_vec` code | Link | Notes |
|---|---|---|---|
| `gaussian()` | 0 | identity | continuous, real-valued |
| `binomial()` | 1 | logit / probit / cloglog | binary 0/1 or multi-trial via `cbind(succ, fail)` / `weights` |
| `poisson()` | 2 | log | non-negative integer counts |
| `lognormal()` | 3 | log | positive continuous; uses `sigma_eps` as the log-scale SD |
| `Gamma(link = "log")` | 4 | log | positive continuous; uses `sigma_eps` as the coefficient of variation |
| `nbinom2()` | 5 | log | over-dispersed counts; per-trait `phi_nbinom2` |
| `tweedie()` | 6 | log | continuous w/ Dirac point mass at 0; per-trait `phi_tweedie`, `p_tweedie` |
| `Beta()` | 7 | logit | continuous on (0, 1); per-trait `phi_beta` |
| `betabinomial()` | 8 | logit | over-dispersed binary / multi-trial; per-trait `phi_betabinom` |
| `student()` | 9 | identity | heavy-tailed continuous; per-trait `sigma_student`, `df_student` |
| `truncated_poisson()` | 10 | log | zero-truncated counts |
| `truncated_nbinom2()` | 11 | log | zero-truncated over-dispersed counts; per-trait `phi_truncnb2` |
| `delta_lognormal()` | 12 | logit / log | hurdle; per-trait `sigma_lognormal_delta` |
| `delta_gamma()` | 13 | logit / log | hurdle; per-trait `phi_gamma_delta` |
| `ordinal_probit()` | 14 | probit | ordered categorical; per-trait cut-points (see [Ordinal-probit](ordinal-probit.html)) |

`sigma_eps` is shared across the "continuous-noise" families
(Gaussian, lognormal, Gamma); for binomial and Poisson rows it is
unused. If you mix Gaussian + lognormal + Gamma, the single
`sigma_eps` is interpreted on each family's natural scale (SD on the
response scale for Gaussian, SD on the log scale for lognormal, CV
for Gamma); their values may not be directly comparable but each is
the correct conditional noise parameter for its family's likelihood.

## Backward compatibility

Passing a scalar `family = gaussian()` (or any single family) goes
through the same code path with `family_id_vec` filled with a single
value -- no change in behaviour relative to single-family fits. This
is verified by the "scalar family still works (backward compatible)"
test in `tests/testthat/test-stage37-mixed-family.R`.

## When to use mixed-response models

The general rule: **only when the latent structure should be shared
across response distributions.** If your binary "behaviour"
measurements and your continuous "body mass" measurements are
genuinely capturing parts of the same multivariate phenomenon (e.g.
"individual quality"), a mixed-response GLLVM is the right tool. If
they're independent, fit them separately.

## See also

* `?gllvmTMB` -- the `family` argument is documented under "Mixed
  responses".
* [Lambda constraints](lambda-constraint.html) -- when you also want
  to **pin** specific loading entries in a mixed-family fit (e.g. a
  CFA pattern where Gaussian items load on one factor and binary
  items on another), combine `family = list(...)` with
  `lambda_constraint = list(B = M)`.
* [Ordinal-probit](ordinal-probit.html) -- the threshold model used
  for the `ordinal_probit()` family in the table above.
* `tests/testthat/test-stage37-mixed-family.R` -- verifies per-row
  dispatch and backward compatibility with scalar `family`.
* **Sorensen, O. (2024)** Multilevel semiparametric latent variable
  modeling in R with galamm. *Multivariate Behavioral Research*
  59(5):1098-1105. <https://doi.org/10.1080/00273171.2024.2385336>
  -- the original `gfam(list(...))` mechanism.
