---
title: "The animal model: heritability and genetic covariance"
tier: 3 # under-audit quantitative-genetic draft; visible as a Developer Note
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{The animal model: heritability and genetic covariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = requireNamespace("MASS", quietly = TRUE),
  message  = FALSE,
  warning  = FALSE
)
```

> **Developer Note -- under audit.** This page is buildable and linked
> from Developer Notes so maintainers can audit the pedigree reader
> path. It is not a first-stop tutorial. Start with
> [Get started](gllvmTMB.html), [First Gaussian morphology model](morphometrics.html),
> and [Covariance and correlation](covariance-correlation.html) before
> using this page for animal-model syntax.

This article is for an evolutionary or ecological reader with
phenotypes measured on individuals whose genealogy is known
(pedigree). The working question is: **how much of the phenotypic
variance is additive genetic, and how do genetic variances and
covariances among traits structure microevolution?**

The animal model is the standard quantitative-genetic tool for that
question (Henderson 1976; Kruuk 2004; Wilson *et al.* 2010). In
`gllvmTMB` it is the `animal_*()` keyword family — a pedigree-
indexed parallel to the phylogenetic `phylo_*()` family, routed
through the same engine path (Hadfield & Nakagawa 2010).

## Variance partition

For a single trait measured on individual $i$, the simple animal
model is

$$
y_i = \mu + a_i + e_i,
$$

where $a_i$ is individual $i$'s **breeding value** (additive
genetic merit) and $e_i$ is a residual. Stacked over all $n$
individuals:

$$
\boldsymbol y = \mu \boldsymbol 1 +
\boldsymbol u + \boldsymbol e,
\qquad
\boldsymbol u \sim
\mathrm{MVN}(\boldsymbol 0,\, \sigma^2_{\mathrm A}\, \boldsymbol A),
\qquad
\boldsymbol e \sim
\mathrm{MVN}(\boldsymbol 0,\, \sigma^2_{\mathrm R}\, \boldsymbol I).
$$

Here $\boldsymbol A$ is the **additive genetic relationship matrix**
derived from the pedigree (Henderson's recursive formula; or
equivalently $A_{ij} = 2\theta_{ij}$ where $\theta_{ij}$ is the
coefficient of coancestry). $\sigma^2_{\mathrm A}$ is the
**additive genetic variance**, $\sigma^2_{\mathrm R}$ is the
**residual variance**, and their sum is the total phenotypic
variance $V_{\mathrm P} = V_{\mathrm A} + V_{\mathrm R}$.
**Narrow-sense heritability** is the proportion explained by
additive genetics:

$$
h^2 = \frac{V_{\mathrm A}}{V_{\mathrm P}}.
$$

For two or more traits, $V_{\mathrm A}$ becomes a $T \times T$
**genetic covariance matrix** $\boldsymbol G$, and the residual a
matrix $\boldsymbol R$. The trait-stacked model reads

$$
\mathrm{vec}(\boldsymbol u) \sim
\mathrm{MVN}\!\bigl(\boldsymbol 0,\, \boldsymbol G \otimes \boldsymbol A\bigr),
\qquad
\mathrm{vec}(\boldsymbol e) \sim
\mathrm{MVN}\!\bigl(\boldsymbol 0,\, \boldsymbol R \otimes \boldsymbol I\bigr).
$$

`gllvmTMB` estimates $\boldsymbol G$ either as a full unstructured
matrix (`animal_dep()`), per-trait independent diagonal entries
(`animal_indep()`; the older `animal_unique()` spelling remains
soft-deprecated compatibility syntax), or as a reduced-rank factor-analytic
parameterisation $\boldsymbol G \approx \boldsymbol\Lambda
\boldsymbol\Lambda^{\!\top} + \boldsymbol\Psi$
(`animal_latent()`; use `unique = FALSE` only for the loadings-only
subset).

Because the `animal_*()` keywords are pure syntactic sugar over the
`phylo_*()` engine path (only the relatedness-matrix input differs),
these intercept structures are **family-general**: the per-trait
`animal_indep()` / `animal_scalar()` decomposition recovers not only
under the Gaussian model used in the tutorials below, but also under
the count, proportion, ordinal, and binomial response families the
engine supports. The worked examples here are Gaussian for clarity;
switch `family =` to fit the same genetic structures on a
non-Gaussian trait.

## Setup

```{r setup}
library(gllvmTMB)
library(MASS)
```

## A small pedigree fixture

We construct a 4-founder, 16-offspring half-sib pedigree (20
individuals total). Half-sib designs are the simplest informative
pedigree because each offspring shares one parent with several
siblings, generating the relatedness structure animal models exploit.

```{r ped}
set.seed(7)
n_founders <- 4L
n_offspring <- 16L
n_ind <- n_founders + n_offspring

ped <- data.frame(
  id   = paste0("i", seq_len(n_ind)),
  sire = c(rep(NA, n_founders),
           rep(c("i1", "i2"), length.out = n_offspring)),
  dam  = c(rep(NA, n_founders),
           rep(c("i3", "i4"), length.out = n_offspring)),
  stringsAsFactors = FALSE
)
head(ped)
```

`gllvmTMB::pedigree_to_A()` builds the relatedness matrix
$\boldsymbol A$ from the pedigree using Henderson's recursive
formula. Founders are assumed unrelated and non-inbred
($A_{\mathrm{founder},\mathrm{founder}} = 1$).

```{r A}
A <- gllvmTMB::pedigree_to_A(ped)
round(A[1:6, 1:6], 3)
```

Diagonal entries are 1 (founders) or $1 + F_i$ (inbred individuals;
none here). Off-diagonals carry the additive relatedness — full
siblings share 0.5, half-siblings 0.25, parents and offspring 0.5.

## Tutorial 1: Heritability of a single trait

We simulate a single trait with known $V_{\mathrm A} = 0.5$ and
$V_{\mathrm R} = 0.5$, so true $h^2 = 0.5$.

```{r sim1}
VA_true <- 0.5
VR_true <- 0.5

u <- as.numeric(MASS::mvrnorm(1, mu = rep(0, n_ind),
                              Sigma = VA_true * A))
e <- rnorm(n_ind, sd = sqrt(VR_true))
y <- u + e

df1 <- data.frame(
  id    = factor(ped$id, levels = ped$id),
  trait = factor("body_size"),
  value = y
)
```

The single-trait animal model uses `animal_scalar()` for the
additive genetic component. Because there is only one trait, the
fixed-effect part is just an intercept (`~ 1`) rather than the
`0 + trait` parameterisation used in the multi-trait sections
below. The `unit` and `cluster` keys are both the individual
identifier.

```{r fit1}
fit1 <- gllvmTMB(
  value ~ 1 + animal_scalar(id, pedigree = ped),
  data = df1,
  trait = "trait",
  unit = "id",
  cluster = "id"
)
fit1$opt$convergence
```

Extract the variance components and heritability. `animal_scalar()`
reports its variance σ²_a as `fit$report$lam_phy` (the
phylogenetic-tier shared variance — recall that `animal_*` and
`phylo_*` share the engine path, hence the `phy` naming). The
residual variance is `sigma_eps²`.

```{r h1}
VA_hat <- as.numeric(fit1$report$lam_phy)
VR_hat <- as.numeric(fit1$report$sigma_eps)^2
h2 <- VA_hat / (VA_hat + VR_hat)
data.frame(
  param = c("V_A", "V_R", "h^2"),
  true  = c(VA_true, VR_true, VA_true / (VA_true + VR_true)),
  est   = signif(c(VA_hat, VR_hat, h2), 3)
)
```

For 20 individuals the estimate is noisy — Kruuk (2004) and Wilson
*et al.* (2010) both note that animal models on wild data typically
need hundreds to thousands of records with informative pedigree
links. This example is for syntax illustration, not power.

## Tutorial 2: Bivariate genetic covariance

We now simulate two correlated traits with $\boldsymbol G =
\bigl(\begin{smallmatrix}0.50 & 0.30\\ 0.30 & 0.45\end{smallmatrix}\bigr)$
and $\boldsymbol R = 0.5\,\boldsymbol I$. The true genetic
correlation is $r_{\mathrm G} = 0.30 / \sqrt{0.50 \cdot 0.45}
\approx 0.632$.

```{r sim2}
G_true <- matrix(c(0.50, 0.30,
                   0.30, 0.45), 2, 2)
R_true <- diag(c(0.50, 0.50))

U <- MASS::mvrnorm(1, mu = rep(0, 2 * n_ind),
                   Sigma = kronecker(G_true, A))
U <- matrix(U, n_ind, 2, byrow = FALSE)
E <- matrix(rnorm(2 * n_ind, sd = sqrt(0.5)), n_ind, 2)
Y <- U + E
colnames(Y) <- c("body_size", "wing_length")
dimnames(G_true) <- list(colnames(Y), colnames(Y))

df2 <- data.frame(
  id    = factor(rep(ped$id, each = 2), levels = ped$id),
  trait = factor(rep(colnames(Y), times = n_ind),
                 levels = colnames(Y)),
  value = as.numeric(t(Y))
)
head(df2)
```

The full unstructured $\boldsymbol G$ is fit with `animal_dep()`.
The factor-analytic alternative is `animal_latent()`, which now carries
the shared factor-analytic part and the trait-specific diagonal
$\boldsymbol\Psi$ by default. At $d = T$, the folded latent form spans
the full $T \times T$ space, equivalent to `animal_dep()`.

```{r fit2}
fit2 <- gllvmTMB(
  value ~ 0 + trait +
    animal_latent(id, d = 1, pedigree = ped) +
    indep(0 + trait | id),
  data = df2,
  trait = "trait",
  unit = "id",
  cluster = "id"
)
fit2$opt$convergence
```

The model has a one-factor genetic structure (`animal_latent(d = 1)`)
with the default per-trait genetic diagonal, plus a non-genetic
trait-level residual covariance (`indep(0 + trait | id)`). The
genetic covariance matrix $\boldsymbol G$ is rotation-invariant and
recovered by combining the shared and diagonal parts:

```{r G2}
G2_compare <- compare_Sigma_table(
  fit2,
  truth = G_true,
  level = "phy",
  measure = "covariance",
  entries = "unique"
)
G2_compare[, c("trait_i", "trait_j", "estimate", "truth", "error")]

R_G_true <- stats::cov2cor(G_true)
R_G_compare <- compare_Sigma_table(
  fit2,
  truth = R_G_true,
  level = "phy",
  measure = "correlation",
  entries = "offdiag"
)
R_G_compare[, c("trait_i", "trait_j", "estimate", "truth", "error")]
```

The bivariate $\boldsymbol G$ encodes both the per-trait variances
(diagonals) and the genetic covariance (off-diagonal). The genetic
correlation $r_{\mathrm G}$ — returned by `extract_Sigma_table(...,
measure = "correlation")`, or independently via
`extract_correlations(fit, tier = "phy")` if you need a CI — is the
standardised version and is the parameter usually reported in
evolutionary studies (Wilson *et al.* 2010, Tutorial 2).

## Tutorial 3: Factor-analytic G matrix

For $T \geq 3$ traits the full $\boldsymbol G$ has $T(T+1)/2$ free
parameters, which is a lot to estimate from a noisy wild pedigree.
**Factor-analytic** $\boldsymbol G$ matrices are the parsimonious
alternative (Smith, Cullis & Gilmour 2001; Kirkpatrick & Meyer 2004;
Runcie & Mukherjee 2013): approximate
$\boldsymbol G \approx \boldsymbol\Lambda \boldsymbol\Lambda^{\!\top}
+ \boldsymbol\Psi$ where $\boldsymbol\Lambda$ is a $T \times d$
loading matrix ($d \ll T$) and $\boldsymbol\Psi$ is diagonal. In
`gllvmTMB` this is `animal_latent(d = K)` — the same folded
decomposition as Tutorial 2, but with $d < T$.

The biological justification (Cheverud 1996; Wagner & Altenberg 1996;
Runcie & Mukherjee 2013) is that organisms develop through a limited
number of **modular** pathways, so the trait covariance is expected
to be sparse and low rank. Runcie & Mukherjee's BSFG (Bayesian
Sparse Factor analysis of G-matrices) scales this idea to thousands
of traits via shrinkage priors on $\boldsymbol\Lambda$. `gllvmTMB`'s
`animal_latent()` is the frequentist (TMB Laplace) counterpart —
faster on moderate-dim problems, applicable to mixed response
families via the GLLVM machinery, no sparsity penalty in v0.2.0.

Below we simulate three traits with a rank-1 shared genetic axis
plus trait-specific genetic and non-genetic noise.

```{r sim3}
n_traits <- 3
Lambda_g_true <- matrix(c(0.70, 0.50, 0.30), n_traits, 1)
psi_g_true <- c(0.10, 0.12, 0.08)
psi_r_true <- c(0.20, 0.18, 0.22)
trait_names <- c("body_size", "wing_length", "tarsus")

g_shared <- as.numeric(MASS::mvrnorm(1, mu = rep(0, n_ind),
                                     Sigma = A))
g_trait <- sapply(seq_len(n_traits), function(t) {
  as.numeric(MASS::mvrnorm(1, mu = rep(0, n_ind),
                           Sigma = psi_g_true[t] * A))
})

Y3 <- g_shared %*% t(Lambda_g_true) + g_trait +
  sweep(matrix(rnorm(n_ind * n_traits), n_ind, n_traits),
        2, sqrt(psi_r_true), "*")
colnames(Y3) <- trait_names

df3 <- data.frame(
  id    = factor(rep(ped$id, each = n_traits), levels = ped$id),
  trait = factor(rep(trait_names, times = n_ind),
                 levels = trait_names),
  value = as.numeric(t(Y3))
)
```

```{r fit3}
fit3 <- gllvmTMB(
  value ~ 0 + trait +
    animal_latent(id, d = 1, pedigree = ped) +
    indep(0 + trait | id),
  data = df3,
  trait = "trait",
  unit = "id",
  cluster = "id"
)
fit3$opt$convergence
```

Extract the implied $\boldsymbol G$, the rank-1 loadings, and the
genetic correlation matrix:

```{r G3}
G3_rows <- extract_Sigma_table(
  fit3,
  level = "phy",
  measure = "covariance",
  entries = "unique"
)
G3_rows[, c("trait_i", "trait_j", "estimate", "interval_status")]

Lambda_hat <- fit3$report$Lambda_phy  # shared loadings (rotation-ambiguous)
round(Lambda_hat, 3)
```

```{r G3-correlation, fig.width = 5.8, fig.height = 4.8}
G3_corr_rows <- extract_Sigma_table(
  fit3,
  level = "phy",
  measure = "correlation",
  entries = "all"
)
plot_Sigma_heatmap(
  G3_corr_rows,
  label_digits = 2,
  title = "Genetic trait correlations",
  subtitle = "Point estimates from the animal-level total covariance."
)
```

The shared $\boldsymbol\Lambda$ is **rotation-invariant only up to
sign and orthogonal rotation** — individual entries are not directly
interpretable without a constraint. The implied $\boldsymbol G$ and
the genetic correlation matrix $\boldsymbol R_{\mathrm G}$ are
rotation-invariant and are the right summaries to report. See
[lambda-constraint](lambda-constraint.html) for confirmatory loadings
when interpretation of $\boldsymbol\Lambda$ entries is required.

## Tutorial 4: Heritable reaction norms (random regression)

The tutorials above partition variance in a trait's **level**. Many
quantitative-genetic questions instead concern the **slope** of a
phenotype on an environmental gradient $x$ — the reaction norm — and
whether that slope is itself heritable (Meyer & Hill 1997; Nussey,
Wilson & Brommer 2007). The random-regression animal model gives each
individual its own slope $\beta_{\mathrm a}(i)$ drawn from a
genetic distribution:

$$
\eta_{io} = \mu_{t(o)} + \beta_{\mathrm a}(i)\, x_{io},
\qquad
\boldsymbol\beta_{\mathrm a} \sim
\mathrm{MVN}\!\bigl(\boldsymbol 0,\, \sigma^2_{\mathrm{slope}}\,\boldsymbol A\bigr),
$$

so that $\sigma^2_{\mathrm{slope}}$ is the additive genetic variance
in the slope and the $\boldsymbol A$ structure makes related
individuals have correlated reaction norms. In `gllvmTMB` this is the
`animal_slope(x | id)` keyword. It carries **one** shared slope
variance (slopes are shared across traits, in the same sense that
`animal_scalar()` shares one intercept variance); the relatedness
matrix is taken from a sibling `animal_*()` term in the same formula.

A 20-individual pedigree is far too small to estimate a slope
variance, so this tutorial builds a larger multi-generation pedigree
(96 individuals over three generations). We keep sires and dams in
disjoint sets so no individual is both.

```{r rr-ped}
make_rr_pedigree <- function(n_sire = 6L, n_dam = 6L, n_gen = 3L,
                             n_off_per_gen = 28L, seed = 11L) {
  set.seed(seed)
  sires0 <- paste0("s", seq_len(n_sire))
  dams0  <- paste0("d", seq_len(n_dam))
  ids  <- c(sires0, dams0)
  sire <- rep(NA_character_, length(ids))
  dam  <- rep(NA_character_, length(ids))
  sire_pool <- sires0
  dam_pool  <- dams0
  k <- 0L
  for (g in seq_len(n_gen)) {
    born <- character(0)
    for (j in seq_len(n_off_per_gen)) {
      k <- k + 1L
      new_id <- paste0("o", k)
      ids  <- c(ids, new_id)
      sire <- c(sire, sample(sire_pool, 1L))
      dam  <- c(dam,  sample(dam_pool, 1L))
      born <- c(born, new_id)
    }
    half <- length(born) %/% 2L
    sire_pool <- c(sire_pool, born[seq_len(half)])
    dam_pool  <- c(dam_pool,  born[(half + 1L):length(born)])
  }
  data.frame(id = ids, sire = sire, dam = dam, stringsAsFactors = FALSE)
}

ped_rr <- make_rr_pedigree()
A_rr   <- gllvmTMB::pedigree_to_A(ped_rr)
n_rr   <- nrow(A_rr)
ids_rr <- rownames(A_rr)
```

Simulate per-individual breeding values for the slope from
$\mathcal N(\boldsymbol 0, \sigma^2_{\mathrm{slope}}\,\boldsymbol A)$
with true $\sigma^2_{\mathrm{slope}} = 0.4$, three traits sharing the
slope, and five replicate measurements per individual so $x$ varies
within individual:

```{r rr-sim}
V_slope_true <- 0.4
L_rr <- t(chol(A_rr + diag(1e-8, n_rr)))
beta_rr <- as.numeric(L_rr %*% rnorm(n_rr, sd = sqrt(V_slope_true)))
names(beta_rr) <- ids_rr

T_rr    <- 3L
nrep_rr <- 5L
df_rr <- expand.grid(
  id    = factor(ids_rr, levels = ids_rr),
  trait = factor(paste0("t", seq_len(T_rr)), levels = paste0("t", seq_len(T_rr))),
  rep   = seq_len(nrep_rr)
)
df_rr$x <- rnorm(nrow(df_rr))
df_rr$value <- c(2, 1.5, 1)[as.integer(df_rr$trait)] +
  beta_rr[as.character(df_rr$id)] * df_rr$x +
  rnorm(nrow(df_rr), sd = 0.5)
```

Fit the random-regression animal model. `animal_slope(x | id)`
supplies the heritable slope; the sibling `animal_scalar(id, pedigree
= ped_rr)` both fits a genetic intercept and provides the relatedness
matrix that `animal_slope()` reuses. (Calling `animal_slope(x | id)`
on its own — with no sibling `animal_*()` term and no relatedness
input — errors, because the engine has no $\boldsymbol A$ to attach
the slopes to.)

```{r rr-fit}
fit_rr <- gllvmTMB(
  value ~ 0 + trait +
    animal_scalar(id, pedigree = ped_rr) +
    animal_slope(x | id),
  data = df_rr,
  trait = "trait",
  unit = "id",
  cluster = "id"
)
fit_rr$opt$convergence
```

The slope variance is reported as `fit$report$sigma_slope` (an SD;
square it for $V_{\mathrm{slope}}$). The per-individual slope
predictions (genetic BLUPs) live in the augmented random-effect block
`b_phy_slope`, indexed by the rows of `Ainv_phy_rr`:

```{r rr-recovery}
V_slope_hat <- as.numeric(fit_rr$report$sigma_slope)^2

b_full  <- fit_rr$tmb_obj$env$last.par.best
blup    <- as.numeric(b_full[names(b_full) == "b_phy_slope"])
aug_id  <- rownames(fit_rr$tmb_data$Ainv_phy_rr)
blup_id <- blup[match(ids_rr, aug_id)]

data.frame(
  quantity = c("V_slope (genetic slope variance)",
               "cor(BLUP slope, true slope)"),
  true     = c(V_slope_true, NA),
  estimate = c(signif(V_slope_hat, 3),
               round(cor(blup_id, beta_rr), 3))
)
```

The genetic slope variance is recovered to within sampling noise and
the predicted per-individual slopes correlate strongly with the
simulated truth. As with the level-heritability example, the point
estimate is noisy at this sample size — reaction-norm variances need
many individuals with repeated measures across the gradient.

**Relationship to `phylo_slope()`.** `animal_slope(x | id)` is exact
syntactic sugar for `phylo_slope(x | id)` with the relatedness matrix
supplied as `vcv = A`: the parser rewrites the former to the latter.
Supplying the *same* $\boldsymbol A$ to both routings gives a
byte-identical fit (here the optimised parameter vectors match to
machine precision):

```{r rr-equiv}
fit_animal <- gllvmTMB(
  value ~ 0 + trait + animal_scalar(id, A = A_rr) + animal_slope(x | id),
  data = df_rr, trait = "trait", unit = "id", cluster = "id"
)
fit_phylo <- gllvmTMB(
  value ~ 0 + trait + phylo_scalar(id, vcv = A_rr) +
    phylo_slope(x | id, vcv = A_rr),
  data = df_rr, trait = "trait", unit = "id", cluster = "id"
)
max(abs(as.numeric(fit_animal$opt$par) - as.numeric(fit_phylo$opt$par)))
```

This single-shared-variance slope is the validated heritable-slope
form. The richer **augmented** reaction-norm models also fit: the
parser now routes an augmented `1 + x | id` left-hand side through the
`animal_*()` keywords, with the soft-deprecated compatibility spelling
`animal_unique(1 + x | id)` driving the correlated $2 \times 2$
(intercept, slope) genetic engine (the
intercept-slope correlation estimated) and `animal_indep(1 + x | id)`
the diagonal variant (correlation pinned to 0). Both are
byte-equivalent to the corresponding soft-deprecated compatibility
`phylo_unique` / first-line `phylo_indep`
augmented fit with the relatedness matrix supplied as `vcv = A`. What
is *not* yet surfaced is the **read-out**: `extract_Sigma()` does not
return the assembled $2 \times 2$ (intercept, slope) genetic covariance
for the correlated `animal_unique(1 + x | id)` fit, so treat it as a
fitting demonstration rather than a covariance-recovery workflow. The
trait-specific and reduced-rank slope structures
(`animal_dep()` / `animal_latent()` and the trait-augmented
`(0 + trait + trait:x | id)` left-hand side) remain a follow-up (see
[What's not yet supported](#whats-not-yet-supported)).

## Adding non-genetic random effects

Wild quantitative-genetic models often need additional random effects
to avoid confounding (Kruuk 2004 §4; Wilson *et al.* 2010 Table 1):

- **Permanent environment** ($V_{\mathrm{PE}}$): non-genetic
  individual effect when individuals are measured multiple times.
- **Maternal environmental** ($V_{\mathrm M}$): shared environment
  among offspring of the same mother (e.g. lactation effects).
- **Common environment** ($V_{\mathrm{CE}}$): shared nest, year, or
  spatial unit.

These all map to ordinary `re_int` random intercepts via the
`(1 | g)` syntax — distinct from `animal_*()` which is reserved for
the pedigree-structured genetic part. A model with $V_{\mathrm A} +
V_{\mathrm M} + V_{\mathrm R}$ would read

```r
gllvmTMB(
  value ~ 0 + trait +
    animal_scalar(id, pedigree = ped) +    # V_A
    (1 | mother) +                          # V_M (environmental)
    ...,
  data = df, trait = "trait", unit = "id", cluster = "id"
)
```

(not fit here for space).

## What's not yet supported {#whats-not-yet-supported}

`animal_*()` in `gllvmTMB` v0.2.0 covers the **single-A-matrix**
animal model — one pedigree-derived relatedness matrix, applied to
the genetic part of the model. Several v0.3.0 follow-ups extend
this:

- **Maternal-genetic effects** ($V_{\mathrm C}$ in Wilson *et al.*
  2010): a second pedigree-indexed random effect with its own
  variance. Estimating $V_{\mathrm A}$, $V_{\mathrm C}$, and their
  covariance requires *two* simultaneous A-matrix terms in the same
  fit. Tracked as register row **ANI-09**.
- **Augmented (structured) reaction norms.** The
  single-shared-variance heritable slope `animal_slope(x | id)` is
  supported and validated ([Tutorial 4](#tutorial-4-heritable-reaction-norms-random-regression),
  register row **ANI-06**). The augmented intercept+slope forms also
  now route through the `animal_*()` keywords and **fit**:
  the soft-deprecated compatibility spelling
  `animal_unique(1 + x | id)` drives the correlated $2 \times 2$
  (intercept, slope) genetic engine and `animal_indep(1 + x | id)` the
  diagonal variant, both byte-equivalent to the corresponding
  soft-deprecated compatibility `phylo_unique` / first-line
  `phylo_indep` augmented fit with `vcv = A`. What is
  *not* yet delivered is, first, the **read-out** -- `extract_Sigma()`
  does not surface the assembled $2 \times 2$ genetic covariance for
  the correlated `animal_unique(1 + x | id)` fit -- and, second, the
  trait-specific and reduced-rank slope structures (`animal_dep()` /
  `animal_latent()` and the trait-augmented
  `(0 + trait + trait:x | id)` left-hand side; Meyer & Hill 1997),
  whose dedicated engine paths remain a v0.3.0 follow-up.

For these post-v0.2.0 needs the workhorse alternatives are
[`MCMCglmm`](https://cran.r-project.org/package=MCMCglmm) (Hadfield
2010, Bayesian), [ASReml-R](https://www.vsni.co.uk/software/asreml-r)
(commercial, REML), and [`WOMBAT`](https://didgeridoo.une.edu.au/km/wombat.php)
(free, REML).

## Citations and further reading

- **Henderson (1976)** *Biometrics* 32:69 — recursive formula for
  $\boldsymbol A^{-1}$ from pedigree.
- **Kruuk (2004)** *Phil. Trans. R. Soc. B* 359:873 — review of the
  animal model in natural populations; the conceptual anchor for
  evolutionary-ecology use.
- **Wilson *et al.* (2010)** *J. Anim. Ecol.* 79:13 — practical
  tutorial for ecologists, with the gryphon worked examples in ASReml,
  MCMCglmm, and WOMBAT.
- **Hadfield & Nakagawa (2010)** *J. Evol. Biol.* 23:494 — sparse
  $\boldsymbol A^{-1}$ representation used by `MCMCglmm` and inherited
  by `gllvmTMB`'s engine path.
- **Smith, Cullis & Gilmour (2001)** *Crop Sci.* 41:1138 —
  factor-analytic $\boldsymbol G$ matrices in plant-breeding
  multi-environment trials (one origin of the parameterisation
  behind `animal_latent()`).
- **Kirkpatrick & Meyer (2004)** *Heredity* 93:512 — reduced-rank
  $\boldsymbol G$ matrices in animal-breeding context.
- **Meyer & Hill (1997)** *Livest. Prod. Sci.* 47:185 —
  random-regression / covariance-function models for heritable
  reaction norms; the parameterisation behind `animal_slope()`.
- **Nussey, Wilson & Brommer (2007)** *J. Evol. Biol.* 20:831 —
  the evolutionary-ecology framing of heritable phenotypic plasticity
  (reaction norms) estimated by random-regression animal models.
- **Runcie & Mukherjee (2013)** *Genetics* 194:753 — Bayesian Sparse
  Factor analysis of $\boldsymbol G$ matrices (BSFG); the canonical
  high-dimensional factor-analytic genetic model. `animal_latent()`
  is the frequentist (TMB Laplace) counterpart, faster on moderate
  $T$ and extensible to mixed response families.
- **Cheverud (1996)** *Am. Zool.* 36:44 and **Wagner & Altenberg
  (1996)** *Evolution* 50:967 — biological argument for modular
  (low-rank, sparse) $\boldsymbol G$.

## See also

- [Phylogenetic trait covariance](phylogenetic-gllvm.html) — the
  cross-species parallel, sharing the engine path.
- [Lambda constraint](lambda-constraint.html) — pinning loading
  entries for confirmatory analysis.
- [GLLVM vocabulary](gllvm-vocabulary.html) — definitions of
  $\boldsymbol\Lambda$, $\boldsymbol\Sigma$, $h^2$, and the rotation
  ambiguity.
- [Choose your model](choose-your-model.html) — decision tree across
  the 4×5 covariance keyword grid.
- [Pitfalls](pitfalls.html) — the **A** (relatedness) vs **V**
  (sampling variance) naming distinction and other gotchas.
