---
title: "The animal model: heritability and genetic covariance"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{The animal model: heritability and genetic covariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = requireNamespace("MASS", quietly = TRUE),
  message  = FALSE,
  warning  = FALSE
)
```

This article is for an evolutionary or ecological reader with
phenotypes measured on individuals whose genealogy is known
(pedigree). The working question is: **how much of the phenotypic
variance is additive genetic, and how do genetic variances and
covariances among traits structure microevolution?**

The animal model is the standard quantitative-genetic tool for that
question (Henderson 1976; Kruuk 2004; Wilson *et al.* 2010). In
`gllvmTMB` it is the `animal_*()` keyword family — a pedigree-
indexed parallel to the phylogenetic `phylo_*()` family, routed
through the same engine path (Hadfield & Nakagawa 2010).

## Variance partition

For a single trait measured on individual $i$, the simple animal
model is

$$
y_i = \mu + a_i + e_i,
$$

where $a_i$ is individual $i$'s **breeding value** (additive
genetic merit) and $e_i$ is a residual. Stacked over all $n$
individuals:

$$
\boldsymbol y = \mu \boldsymbol 1 +
\boldsymbol u + \boldsymbol e,
\qquad
\boldsymbol u \sim
\mathrm{MVN}(\boldsymbol 0,\, \sigma^2_{\mathrm A}\, \boldsymbol A),
\qquad
\boldsymbol e \sim
\mathrm{MVN}(\boldsymbol 0,\, \sigma^2_{\mathrm R}\, \boldsymbol I).
$$

Here $\boldsymbol A$ is the **additive genetic relationship matrix**
derived from the pedigree (Henderson's recursive formula; or
equivalently $A_{ij} = 2\theta_{ij}$ where $\theta_{ij}$ is the
coefficient of coancestry). $\sigma^2_{\mathrm A}$ is the
**additive genetic variance**, $\sigma^2_{\mathrm R}$ is the
**residual variance**, and their sum is the total phenotypic
variance $V_{\mathrm P} = V_{\mathrm A} + V_{\mathrm R}$.
**Narrow-sense heritability** is the proportion explained by
additive genetics:

$$
h^2 = \frac{V_{\mathrm A}}{V_{\mathrm P}}.
$$

For two or more traits, $V_{\mathrm A}$ becomes a $T \times T$
**genetic covariance matrix** $\boldsymbol G$, and the residual a
matrix $\boldsymbol R$. The trait-stacked model reads

$$
\mathrm{vec}(\boldsymbol u) \sim
\mathrm{MVN}\!\bigl(\boldsymbol 0,\, \boldsymbol G \otimes \boldsymbol A\bigr),
\qquad
\mathrm{vec}(\boldsymbol e) \sim
\mathrm{MVN}\!\bigl(\boldsymbol 0,\, \boldsymbol R \otimes \boldsymbol I\bigr).
$$

`gllvmTMB` estimates $\boldsymbol G$ either as a full unstructured
matrix (`animal_dep()`), per-trait independent diagonal entries
(`animal_unique()`), or as a reduced-rank factor-analytic
parameterisation $\boldsymbol G \approx \boldsymbol\Lambda
\boldsymbol\Lambda^{\!\top} + \boldsymbol\Psi$
(`animal_latent()` paired with `animal_unique()`).

## Setup

```{r setup}
library(gllvmTMB)
library(MASS)
```

## A small pedigree fixture

We construct a 4-founder, 16-offspring half-sib pedigree (20
individuals total). Half-sib designs are the simplest informative
pedigree because each offspring shares one parent with several
siblings, generating the relatedness structure animal models exploit.

```{r ped}
set.seed(7)
n_founders <- 4L
n_offspring <- 16L
n_ind <- n_founders + n_offspring

ped <- data.frame(
  id   = paste0("i", seq_len(n_ind)),
  sire = c(rep(NA, n_founders),
           rep(c("i1", "i2"), length.out = n_offspring)),
  dam  = c(rep(NA, n_founders),
           rep(c("i3", "i4"), length.out = n_offspring)),
  stringsAsFactors = FALSE
)
head(ped)
```

`gllvmTMB::pedigree_to_A()` builds the relatedness matrix
$\boldsymbol A$ from the pedigree using Henderson's recursive
formula. Founders are assumed unrelated and non-inbred
($A_{\mathrm{founder},\mathrm{founder}} = 1$).

```{r A}
A <- pedigree_to_A(ped)
round(A[1:6, 1:6], 3)
```

Diagonal entries are 1 (founders) or $1 + F_i$ (inbred individuals;
none here). Off-diagonals carry the additive relatedness — full
siblings share 0.5, half-siblings 0.25, parents and offspring 0.5.

## Tutorial 1: Heritability of a single trait

We simulate a single trait with known $V_{\mathrm A} = 0.5$ and
$V_{\mathrm R} = 0.5$, so true $h^2 = 0.5$.

```{r sim1}
VA_true <- 0.5
VR_true <- 0.5

u <- as.numeric(MASS::mvrnorm(1, mu = rep(0, n_ind),
                              Sigma = VA_true * A))
e <- rnorm(n_ind, sd = sqrt(VR_true))
y <- u + e

df1 <- data.frame(
  id    = factor(ped$id, levels = ped$id),
  trait = factor("body_size"),
  value = y
)
```

The single-trait animal model uses `animal_scalar()` for the
additive genetic component. Because there is only one trait, the
fixed-effect part is just an intercept (`~ 1`) rather than the
`0 + trait` parameterisation used in the multi-trait sections
below. The `unit` and `cluster` keys are both the individual
identifier.

```{r fit1}
fit1 <- gllvmTMB(
  value ~ 1 + animal_scalar(id, pedigree = ped),
  data = df1,
  trait = "trait",
  unit = "id",
  cluster = "id"
)
fit1$opt$convergence
```

Extract the variance components and heritability. `animal_scalar()`
reports its variance σ²_a as `fit$report$lam_phy` (the
phylogenetic-tier shared variance — recall that `animal_*` and
`phylo_*` share the engine path, hence the `phy` naming). The
residual variance is `sigma_eps²`.

```{r h1}
VA_hat <- as.numeric(fit1$report$lam_phy)
VR_hat <- as.numeric(fit1$report$sigma_eps)^2
h2 <- VA_hat / (VA_hat + VR_hat)
data.frame(
  param = c("V_A", "V_R", "h^2"),
  true  = c(VA_true, VR_true, VA_true / (VA_true + VR_true)),
  est   = signif(c(VA_hat, VR_hat, h2), 3)
)
```

For 20 individuals the estimate is noisy — Kruuk (2004) and Wilson
*et al.* (2010) both note that animal models on wild data typically
need hundreds to thousands of records with informative pedigree
links. This example is for syntax illustration, not power.

## Tutorial 2: Bivariate genetic covariance

We now simulate two correlated traits with $\boldsymbol G =
\bigl(\begin{smallmatrix}0.50 & 0.30\\ 0.30 & 0.45\end{smallmatrix}\bigr)$
and $\boldsymbol R = 0.5\,\boldsymbol I$. The true genetic
correlation is $r_{\mathrm G} = 0.30 / \sqrt{0.50 \cdot 0.45}
\approx 0.632$.

```{r sim2}
G_true <- matrix(c(0.50, 0.30,
                   0.30, 0.45), 2, 2)
R_true <- diag(c(0.50, 0.50))

U <- MASS::mvrnorm(1, mu = rep(0, 2 * n_ind),
                   Sigma = kronecker(G_true, A))
U <- matrix(U, n_ind, 2, byrow = FALSE)
E <- matrix(rnorm(2 * n_ind, sd = sqrt(0.5)), n_ind, 2)
Y <- U + E
colnames(Y) <- c("body_size", "wing_length")
dimnames(G_true) <- list(colnames(Y), colnames(Y))

df2 <- data.frame(
  id    = factor(rep(ped$id, each = 2), levels = ped$id),
  trait = factor(rep(colnames(Y), times = n_ind),
                 levels = colnames(Y)),
  value = as.numeric(t(Y))
)
head(df2)
```

The full unstructured $\boldsymbol G$ is fit by pairing
`animal_latent()` (the shared factor-analytic part) with
`animal_unique()` (the trait-specific diagonal). At $d = T$, the
paired form spans the full $T \times T$ space — equivalent to
`animal_dep()` on a full $\boldsymbol G$.

```{r fit2}
fit2 <- gllvmTMB(
  value ~ 0 + trait +
    animal_latent(id, d = 1, pedigree = ped) +
    animal_unique(id, pedigree = ped) +
    unique(0 + trait | id),
  data = df2,
  trait = "trait",
  unit = "id",
  cluster = "id"
)
fit2$opt$convergence
```

The model has a one-factor genetic structure (`animal_latent(d = 1)`)
plus per-trait genetic diagonal (`animal_unique`), plus a non-genetic
trait-level residual covariance (`unique(0 + trait | id)`). The
genetic covariance matrix $\boldsymbol G$ is rotation-invariant and
recovered by combining the shared and diagonal parts:

```{r G2}
G2_compare <- compare_Sigma_table(
  fit2,
  truth = G_true,
  level = "phy",
  measure = "covariance",
  entries = "unique"
)
G2_compare[, c("trait_i", "trait_j", "estimate", "truth", "error")]

R_G_true <- stats::cov2cor(G_true)
R_G_compare <- compare_Sigma_table(
  fit2,
  truth = R_G_true,
  level = "phy",
  measure = "correlation",
  entries = "offdiag"
)
R_G_compare[, c("trait_i", "trait_j", "estimate", "truth", "error")]
```

The bivariate $\boldsymbol G$ encodes both the per-trait variances
(diagonals) and the genetic covariance (off-diagonal). The genetic
correlation $r_{\mathrm G}$ — returned by `extract_Sigma_table(...,
measure = "correlation")`, or independently via
`extract_correlations(fit, tier = "phy")` if you need a CI — is the
standardised version and is the parameter usually reported in
evolutionary studies (Wilson *et al.* 2010, Tutorial 2).

## Tutorial 3: Factor-analytic G matrix

For $T \geq 3$ traits the full $\boldsymbol G$ has $T(T+1)/2$ free
parameters, which is a lot to estimate from a noisy wild pedigree.
**Factor-analytic** $\boldsymbol G$ matrices are the parsimonious
alternative (Smith, Cullis & Gilmour 2001; Kirkpatrick & Meyer 2004;
Runcie & Mukherjee 2013): approximate
$\boldsymbol G \approx \boldsymbol\Lambda \boldsymbol\Lambda^{\!\top}
+ \boldsymbol\Psi$ where $\boldsymbol\Lambda$ is a $T \times d$
loading matrix ($d \ll T$) and $\boldsymbol\Psi$ is diagonal. In
`gllvmTMB` this is `animal_latent(d = K) + animal_unique()` — the
same pairing as Tutorial 2, but with $d < T$.

The biological justification (Cheverud 1996; Wagner & Altenberg 1996;
Runcie & Mukherjee 2013) is that organisms develop through a limited
number of **modular** pathways, so the trait covariance is expected
to be sparse and low rank. Runcie & Mukherjee's BSFG (Bayesian
Sparse Factor analysis of G-matrices) scales this idea to thousands
of traits via shrinkage priors on $\boldsymbol\Lambda$. `gllvmTMB`'s
`animal_latent()` is the frequentist (TMB Laplace) counterpart —
faster on moderate-dim problems, applicable to mixed response
families via the GLLVM machinery, no sparsity penalty in v0.2.0.

Below we simulate three traits with a rank-1 shared genetic axis
plus trait-specific genetic and non-genetic noise.

```{r sim3}
n_traits <- 3
Lambda_g_true <- matrix(c(0.70, 0.50, 0.30), n_traits, 1)
psi_g_true <- c(0.10, 0.12, 0.08)
psi_r_true <- c(0.20, 0.18, 0.22)
trait_names <- c("body_size", "wing_length", "tarsus")

g_shared <- as.numeric(MASS::mvrnorm(1, mu = rep(0, n_ind),
                                     Sigma = A))
g_trait <- sapply(seq_len(n_traits), function(t) {
  as.numeric(MASS::mvrnorm(1, mu = rep(0, n_ind),
                           Sigma = psi_g_true[t] * A))
})

Y3 <- g_shared %*% t(Lambda_g_true) + g_trait +
  sweep(matrix(rnorm(n_ind * n_traits), n_ind, n_traits),
        2, sqrt(psi_r_true), "*")
colnames(Y3) <- trait_names

df3 <- data.frame(
  id    = factor(rep(ped$id, each = n_traits), levels = ped$id),
  trait = factor(rep(trait_names, times = n_ind),
                 levels = trait_names),
  value = as.numeric(t(Y3))
)
```

```{r fit3}
fit3 <- gllvmTMB(
  value ~ 0 + trait +
    animal_latent(id, d = 1, pedigree = ped) +
    animal_unique(id, pedigree = ped) +
    unique(0 + trait | id),
  data = df3,
  trait = "trait",
  unit = "id",
  cluster = "id"
)
fit3$opt$convergence
```

Extract the implied $\boldsymbol G$, the rank-1 loadings, and the
genetic correlation matrix:

```{r G3}
G3_rows <- extract_Sigma_table(
  fit3,
  level = "phy",
  measure = "covariance",
  entries = "unique"
)
G3_rows[, c("trait_i", "trait_j", "estimate", "interval_status")]

Lambda_hat <- fit3$report$Lambda_phy  # shared loadings (rotation-ambiguous)
round(Lambda_hat, 3)
```

```{r G3-correlation, fig.width = 5.8, fig.height = 4.8}
G3_corr_rows <- extract_Sigma_table(
  fit3,
  level = "phy",
  measure = "correlation",
  entries = "all"
)
plot_Sigma_heatmap(
  G3_corr_rows,
  label_digits = 2,
  title = "Genetic trait correlations",
  subtitle = "Point estimates from the animal-level total covariance."
)
```

The shared $\boldsymbol\Lambda$ is **rotation-invariant only up to
sign and orthogonal rotation** — individual entries are not directly
interpretable without a constraint. The implied $\boldsymbol G$ and
the genetic correlation matrix $\boldsymbol R_{\mathrm G}$ are
rotation-invariant and are the right summaries to report. See
[lambda-constraint](lambda-constraint.html) for confirmatory loadings
when interpretation of $\boldsymbol\Lambda$ entries is required.

## Adding non-genetic random effects

Wild quantitative-genetic models often need additional random effects
to avoid confounding (Kruuk 2004 §4; Wilson *et al.* 2010 Table 1):

- **Permanent environment** ($V_{\mathrm{PE}}$): non-genetic
  individual effect when individuals are measured multiple times.
- **Maternal environmental** ($V_{\mathrm M}$): shared environment
  among offspring of the same mother (e.g. lactation effects).
- **Common environment** ($V_{\mathrm{CE}}$): shared nest, year, or
  spatial unit.

These all map to ordinary `re_int` random intercepts via the
`(1 | g)` syntax — distinct from `animal_*()` which is reserved for
the pedigree-structured genetic part. A model with $V_{\mathrm A} +
V_{\mathrm M} + V_{\mathrm R}$ would read

```r
gllvmTMB(
  value ~ 0 + trait +
    animal_scalar(id, pedigree = ped) +    # V_A
    (1 | mother) +                          # V_M (environmental)
    ...,
  data = df, trait = "trait", unit = "id", cluster = "id"
)
```

(not fit here for space).

## What's not yet supported

`animal_*()` in `gllvmTMB` v0.2.0 covers the **single-A-matrix**
animal model — one pedigree-derived relatedness matrix, applied to
the genetic part of the model. Several v0.3.0 follow-ups extend
this:

- **Maternal-genetic effects** ($V_{\mathrm C}$ in Wilson *et al.*
  2010): a second pedigree-indexed random effect with its own
  variance. Estimating $V_{\mathrm A}$, $V_{\mathrm C}$, and their
  covariance requires *two* simultaneous A-matrix terms in the same
  fit. Tracked as register row **ANI-09**.
- **Sparse $\boldsymbol A^{-1}$ direct engine path** (Hadfield &
  Nakagawa 2010). v0.2.0 densifies $\boldsymbol A^{-1}$ internally;
  large pedigrees ($n \gtrsim 500$) will be substantially faster
  when the direct sparse path lands. Tracked as **ANI-08**.
- **Random-regression / heritable slopes**
  (`animal_slope(x | id)`): per-individual slope on a covariate
  with a $\boldsymbol G$-matrix covariance — the canonical
  quantitative-genetic reaction-norm model (Meyer & Hill 1997).
  Parser support is staged; engine support is v0.3.0.

For these post-v0.2.0 needs the workhorse alternatives are
[`MCMCglmm`](https://cran.r-project.org/package=MCMCglmm) (Hadfield
2010, Bayesian), [ASReml-R](https://www.vsni.co.uk/software/asreml-r)
(commercial, REML), and [`WOMBAT`](https://didgeridoo.une.edu.au/km/wombat.php)
(free, REML).

## Citations and further reading

- **Henderson (1976)** *Biometrics* 32:69 — recursive formula for
  $\boldsymbol A^{-1}$ from pedigree.
- **Kruuk (2004)** *Phil. Trans. R. Soc. B* 359:873 — review of the
  animal model in natural populations; the conceptual anchor for
  evolutionary-ecology use.
- **Wilson *et al.* (2010)** *J. Anim. Ecol.* 79:13 — practical
  tutorial for ecologists, with the gryphon worked examples in ASReml,
  MCMCglmm, and WOMBAT.
- **Hadfield & Nakagawa (2010)** *J. Evol. Biol.* 23:494 — sparse
  $\boldsymbol A^{-1}$ representation used by `MCMCglmm` and inherited
  by `gllvmTMB`'s engine path.
- **Smith, Cullis & Gilmour (2001)** *Crop Sci.* 41:1138 —
  factor-analytic $\boldsymbol G$ matrices in plant-breeding
  multi-environment trials (one origin of the parameterisation
  behind `animal_latent()`).
- **Kirkpatrick & Meyer (2004)** *Heredity* 93:512 — reduced-rank
  $\boldsymbol G$ matrices in animal-breeding context.
- **Runcie & Mukherjee (2013)** *Genetics* 194:753 — Bayesian Sparse
  Factor analysis of $\boldsymbol G$ matrices (BSFG); the canonical
  high-dimensional factor-analytic genetic model. `animal_latent()`
  is the frequentist (TMB Laplace) counterpart, faster on moderate
  $T$ and extensible to mixed response families.
- **Cheverud (1996)** *Am. Zool.* 36:44 and **Wagner & Altenberg
  (1996)** *Evolution* 50:967 — biological argument for modular
  (low-rank, sparse) $\boldsymbol G$.

## See also

- [Phylogenetic trait covariance](phylogenetic-gllvm.html) — the
  cross-species parallel, sharing the engine path.
- [Lambda constraint](lambda-constraint.html) — pinning loading
  entries for confirmatory analysis.
- [GLLVM vocabulary](gllvm-vocabulary.html) — definitions of
  $\boldsymbol\Lambda$, $\boldsymbol\Sigma$, $h^2$, and the rotation
  ambiguity.
- [Choose your model](choose-your-model.html) — decision tree across
  the 4×5 covariance keyword grid.
- [Pitfalls](pitfalls.html) — the **A** (relatedness) vs **V**
  (sampling variance) naming distinction and other gotchas.
