---
title: "Formula keyword grid"
description: "Choose a correlation source and covariance mode, then translate the intended model into matching long and traits(...) wide formulas."
tier: 2
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Formula keyword grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This page is the syntax lookup for choosing a covariance keyword. The formula
grammar has two decisions: what relates the grouping levels, and what covariance
among traits is required at that grouping level.

The keywords below are live API, but a valid formula does not guarantee that a
complex covariance is estimable from a particular data set. Fit health,
replication, trait count, and the scientific covariance question still matter.

## Choose in two steps

### 1. What relates the grouping levels?

- **No prefix:** levels are independent, as in an ordinary random effect.
- **`animal_`:** individuals are related through a pedigree, `A`, or `Ainv`.
- **`phylo_`:** species are related through a tree or phylogenetic covariance.
- **`spatial_`:** locations share a mesh-based spatial field, represented with
  a sparse Gaussian Markov random field (GMRF) approximation so nearby
  locations can be correlated.
- **`kernel_`:** levels are related through a supplied dense covariance matrix
  `K` — the generic dense-matrix source (the fifth grid row).

These sources are alternatives based on study design, not a progression from
fine to broad scale. Add a source prefix only when that structured relationship
is part of the estimand.

### 2. What trait covariance is needed?

There are **three fundamental cross-trait structures**, from simplest to richest:

- **`indep`:** a separate marginal variance for each trait and zero cross-trait
  covariance — a diagonal model. Its parsimony modifier `common = TRUE` ties all
  trait variances to **one shared value** (the *scalar* special case).
- **`dep`:** a full unstructured trait covariance with (T(T+1)/2) covariance
  parameters. Use only when trait count and grouping-level replication support
  a saturated covariance.
- **`latent`:** a parsimonious reduced-rank cross-trait covariance
  \(\boldsymbol\Lambda\boldsymbol\Lambda^{\mathsf T}\) represented by (d) latent
  variables. Its modifier `unique = TRUE` adds the trait-diagonal \(\Psi\)
  companion.

**`scalar` and `unique` are modifiers, not separate modes.** `scalar` is
`indep` with the variances tied to one shared value; `unique` is `latent` with
its diagonal-\(\Psi\) companion. This mirrors how the deprecated `unique()`
keyword now lives as `latent(unique = TRUE)`. The named `scalar()`,
`phylo_scalar()`, `animal_scalar()`, `spatial_scalar()`, and `kernel_scalar()`
keywords fit exactly this model but are **soft-deprecated compatibility syntax**
(they emit a one-time warning); write `indep(..., common = TRUE)` in new code.

All fitted keyword covariances are random-effect covariances on the model's
linear-predictor or latent scale. They are not automatically response-scale
covariances for non-Gaussian traits.

## The 5 × 3 grid

The rows are the five correlation **sources** across grouping levels; the
columns are the three fundamental trait-covariance **modes**. Every cell is a
live keyword. The *scalar* and *unique* sub-cases are modifiers, listed under
the table.

<div class="table-responsive">
<table class="table table-striped" style="min-width: 640px;">
<caption>Source × mode covariance keyword grid</caption>
<thead><tr>
<th>Source</th><th>Independent</th><th>Dependent</th><th>Latent</th>
</tr></thead>
<tbody>
<tr><th>none</th><td><code>indep()</code></td>
<td><code>dep()</code></td><td><code>latent()</code></td></tr>
<tr><th>animal</th><td><code>animal_indep()</code></td>
<td><code>animal_dep()</code></td><td><code>animal_latent()</code></td></tr>
<tr><th>phylogenetic</th><td><code>phylo_indep()</code></td>
<td><code>phylo_dep()</code></td><td><code>phylo_latent()</code></td></tr>
<tr><th>spatial</th><td><code>spatial_indep()</code></td>
<td><code>spatial_dep()</code></td><td><code>spatial_latent()</code></td></tr>
<tr><th>kernel</th><td><code>kernel_indep()</code></td>
<td><code>kernel_dep()</code></td><td><code>kernel_latent()</code></td></tr>
</tbody>
</table>
</div>

**Scalar (one shared variance).** Add `common = TRUE` to any `indep` term —
e.g. `indep(0 + trait | g, common = TRUE)`. The soft-deprecated `scalar()` /
`*_scalar()` keywords fit the same model but warn once and are kept only for
compatibility. Omitting a multi-trait covariance term entirely means no such
term is fitted.

**Unique (diagonal-\(\Psi\) companion).** Add `unique = TRUE` to a `latent`
term (see the latent default below). The standalone `unique()` keyword is
soft-deprecated in favour of `indep()` (for a plain diagonal) and
`latent(unique = TRUE)` (for the paired companion).

## What the modes estimate

For the intercept-only grid, let $K_{gg'}$ be an entry of the relationship or
model-implied covariance operator among grouping levels. For the ordinary
no-prefix row, $\mathbf K = \mathbf I$. Animal, phylogenetic, and dense-kernel
inputs supply fixed covariance operators and may have non-unit diagonals.
Spatial terms instead imply $\mathbf K(\boldsymbol\theta)$ through a
mesh-based precision whose range and scale parameters are estimated.

The clearest ordering-free definition is

\[
\operatorname{Cov}(b_{gt}, b_{g't'})
=
\Sigma_{T,tt'} K_{gg'},
\]

where $\boldsymbol\Sigma_T$ is the trait covariance implied by the selected
mode: $\mathrm{diag}(\psi_1,\ldots,\psi_T)$ for `indep`
(collapsing to the shared $\sigma^2\mathbf I_T$ under the `common = TRUE` /
scalar modifier), a free $\boldsymbol\Sigma_T$ for `dep`, and
$\boldsymbol\Lambda\boldsymbol\Lambda^{\mathsf T}$ with an optional
$\boldsymbol\Psi$ companion for `latent`. This elementwise form does not depend
on whether software stacks traits or grouping levels first.

The diagonal in `animal_indep()` or `phylo_indep()` is diagonal **across
traits**, but observations remain correlated **within each trait** through the
pedigree or phylogeny. It is not independent residual noise.

### The important latent default

Here, `unique =` is an argument of a latent term that toggles its diagonal Psi
companion; it is not a separate covariance mode or a call to the deprecated
covariance-function syntax.

The diagonal Psi companion is controlled differently across sources:

- ordinary `latent()` includes
  $\boldsymbol\Lambda\boldsymbol\Lambda^{\mathsf T}+\boldsymbol\Psi$ by
  default; use `unique = FALSE` for loadings only;
- `animal_latent()`, `phylo_latent()`, `spatial_latent()`, and
  `kernel_latent()` are loadings-only by default;
- use `unique = TRUE` on a source-specific or single-kernel latent term when
  that same source should also carry a trait-diagonal Psi component.

For animal, phylogenetic, and kernel models, source-specific Psi is diagonal
across traits but still correlated across individuals, species, or other levels
through the supplied relationship or covariance operator. Spatial Psi is also
diagonal across traits, while each trait-specific field remains correlated
across locations through the mesh-based spatial precision.

These diagonal statements describe intercept-only trait covariance. With an
augmented intercept-and-slope term, `*_indep(1 + x | group)` fixes the
within-trait intercept--slope correlation to zero. The random-slope section
below gives the corresponding blocks.

## Ordinary long and wide syntax

The long and wide forms below express the same covariance choices. The
[Morphometrics](morphometrics.html) article executes and compares both data
shapes.

```{r ordinary-syntax, eval = FALSE}
# Decomposed covariance: Lambda Lambda^T + Psi
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | individual, d = 2),
  data = df_long,
  trait = "trait",
  unit = "individual"
)

fit_wide <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~ 1 +
    latent(1 | individual, d = 2),
  data = df_wide,
  unit = "individual"
)
```

Replace `latent(...)` with `indep(...)` for a diagonal baseline or `dep(...)`
for a full covariance. In wide formulas, `1` inside a bar-style term expands to
the long-form `0 + trait` representation.

## Exact source-specific syntax

The latent animal/phylogenetic helpers take the grouping column directly. The
`indep` and `dep` counterparts use a bar formula, and the one-shared-variance
(scalar) case is the `common = TRUE` modifier on `indep`. Spatial helpers all
use the bar-form orientation.

### Animal

```{r animal-syntax, eval = FALSE}
animal_indep(0 + trait | id, pedigree = ped, common = TRUE)  # one shared variance
animal_indep(0 + trait | id, pedigree = ped)
animal_dep(0 + trait | id, pedigree = ped)
animal_latent(id, d = 2, pedigree = ped)                 # loadings only
animal_latent(id, d = 2, pedigree = ped, unique = TRUE)  # + animal Psi
```

### Phylogenetic

```{r phylo-syntax, eval = FALSE}
phylo_indep(0 + trait | species, tree = tree, common = TRUE)  # one shared variance
phylo_indep(0 + trait | species, tree = tree)
phylo_dep(0 + trait | species, tree = tree)
phylo_latent(species, d = 2, tree = tree)                 # loadings only
phylo_latent(species, d = 2, tree = tree, unique = TRUE)  # + phylo Psi
```

Use `tree =` for the usual phylogenetic workflow. `vcv =`, `A =`, and `Ainv =`
support advanced precomputed covariance/precision inputs; their row and column
names must match the grouping levels.

### Spatial

```{r spatial-syntax, eval = FALSE}
spatial_indep(0 + trait | site, mesh = mesh, common = TRUE)  # one shared variance
spatial_indep(0 + trait | site, mesh = mesh)
spatial_dep(0 + trait | site, mesh = mesh)
spatial_latent(0 + trait | site, d = 2, mesh = mesh)
spatial_latent(0 + trait | site, d = 2, mesh = mesh, unique = TRUE)
```

Spatial models require a mesh built from the model's coordinate columns with
`make_mesh()`. Pass that mesh either inside the spatial keyword or through the
top-level `mesh =` argument to `gllvmTMB()`. A source-valid formula still needs
enough locations and spatial spread to identify range and covariance
parameters.

## Dense kernels (the fifth grid row)

A supplied dense covariance matrix `K` is the generic dense-matrix source. Its
keywords follow the same grid as the other sources, with the tied-variance case
as the `common = TRUE` modifier on `kernel_indep()`.

```{r kernel-syntax, eval = FALSE}
kernel_indep(unit_id, K = K, name = "known", common = TRUE)  # one shared variance x K
kernel_indep(unit_id, K = K, name = "known")
kernel_dep(unit_id, K = K, name = "known")
kernel_latent(unit_id, K = K, d = 2, name = "known")
kernel_latent(
  unit_id, K = K, d = 2, name = "known", unique = TRUE
)
```

For a single named kernel, `unique = TRUE` adds that kernel's diagonal Psi
companion. Multiple named-kernel fits currently use the latent-only route; do
not assume that each component receives its own Psi. When kernels overlap
strongly, the fitted sum may be more stable than attribution to individual
kernel components.

Named kernels should encode a relationship fixed before fitting. The model
does not turn an analyst-supplied association matrix into evidence of a causal
or reciprocal process. When more than one named kernel is fitted, use
`diagnose_kernel_separability()` before interpreting component-specific output;
collapse indistinguishable kernels rather than assigning them separate
biological stories.

## Other formula markers

### Random slopes

`phylo_slope(x | species)` and `animal_slope(x | id)` are live slope-only
structured routes. They are not additional covariance modes and should not be
confused with an augmented intercept-and-slope covariance such as
`animal_dep(1 + x | id, ...)` or `spatial_latent(1 + x | site, ...)`.

An augmented random-slope term describes deviations around a fixed
population-average slope. Include that fixed slope separately:

```{r structured-slope-syntax, eval = FALSE}
# Long form
value ~ 0 + trait + (0 + trait):x +
  phylo_indep(0 + trait + (0 + trait):x | species)

# Equivalent wide form
traits(t1, t2, t3) ~ 1 + x +
  phylo_indep(1 + x | species)
```

The three structured modes do not merely represent increasing computational
complexity. They estimate different covariance objects:

- `phylo_indep(1 + x | species)` and
  `spatial_indep(1 + x | site)` estimate one \(2 \times 2\)
  intercept--slope block with the cross-correlation fixed to zero. Extract it
  at `level = "phy"` or `level = "spatial"`.
- `phylo_latent(1 + x | species, d = K)` and
  `spatial_latent(1 + x | site, d = K)` estimate separate reduced-rank
  cross-trait covariance matrices for the intercept and slope columns. They
  do not estimate intercept--slope covariance. Extract the per-column matrices
  at `level = "phy_slope"` or `level = "spde_slope"`.
- `phylo_dep(1 + x | species)` and
  `spatial_dep(1 + x | site)` estimate a full interleaved
  \(2T \times 2T\) covariance across trait intercepts and slopes. Extract it at
  `level = "phy"` or `level = "spatial"`.

Choose the mode from the scientific covariance assumption, not as an automatic
fallback after a difficult fit. Centre and scale `x`, retain the transformation,
and confirm repeated, overlapping within-unit coverage before fitting a random
slope. The [behavioural reaction-norm guide](random-regression-reaction-norms.html)
shows those design checks for the ordinary Gaussian case. Family-specific
response requirements still apply; changing `family =` is not the only change
needed when moving from continuous traits to counts, proportions, trials, or
ordered categories.

### Known sampling covariance

`meta_V()` is not a grid row. It adds a known observation/sampling covariance
to a meta-analytic model rather than estimating a random-effect trait
covariance. The exact-additive route requires the same row-aligned matrix in
the formula marker and the top-level argument:

```{r known-v-syntax, eval = FALSE}
value ~ 0 + trait + latent(0 + trait | study, d = 1) + meta_V(V = V)
# In gllvmTMB(...), also pass known_V = V.
```

Use `A`/`Ainv` for relatedness and `V` for known sampling covariance; they are
different scientific objects. This syntax lookup does not provide the applied
estimand, diagnostic, and interpretation workflow needed for a full
meta-analytic example. `meta_known_V()` remains a soft-deprecated alias;
new code uses `meta_V()`.

## What not to combine

At one grouping level and correlation source, specify one trait-covariance
story:

- do not combine `dep` with `indep` or `latent` at that source;
- do not combine standalone `indep` with `latent` at that source;
- the same rules apply to ordinary, animal, phylogenetic, spatial, and
  single-kernel terms.

Different grouping levels can carry different covariance terms—for example, a
between-individual `latent()` term and a within-individual `indep()` term—when
the design identifies both.

## If the first choice is too complex

- **Full `dep()` is unstable or imprecise:** try a lower-rank `latent()` model,
  or `indep()` if cross-trait covariance is not essential.
- **An added LV is weak or unstable across refits:** reduce `d` and compare fit
  health and information criteria.
- **A diagonal variance is at zero:** treat it as a boundary estimate; compare
  the simpler model rather than interpreting a tiny component literally.
- **A relationship input fails:** check grouping-level names, matrix dimensions,
  positive-semidefiniteness, tree tips, or mesh/data alignment.
- **You are unsure which source prefix to use:** return to the biological
  sampling design. A more elaborate source is not automatically a better model.

See [Fit diagnostics](fit-diagnostics.html) before interpreting the selected
covariance and [Covariance and correlation](covariance-correlation.html) for the
corresponding estimands.
