---
title: "Formula keyword grid"
tier: 2 # technical reference for the covariance keyword grammar
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Formula keyword grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This reference is for readers who already know they need a
stacked-trait model and want to choose the right covariance keyword.
The keywords are organised as a 4 x 4 grid: the row names the
correlation across units, and the column names the covariance mode
across traits. The four correlation rows go from finest-grained
(individual pedigree) to broadest (geography).

> **Scope boundary.** This page is a technical lookup for the formula
> grammar, not a worked validation article. The long and wide formula
> grammar, `traits(...)` expansion, the ordinary `latent()` path with
> default diagonal $\boldsymbol\Psi$, the `animal_*()` keywords, and the
> phylogenetic and spatial keyword families are validated for their core
> paths. A few surfaces are still under validation at greater recovery
> depth — notably the standalone diagonal, full, and slope forms on the
> no-prefix row, the augmented spatial `latent(unique = TRUE)` fold, and
> the exact single-V meta-analytic path — and proportional `meta_V()` is
> not yet available. Phylogenetic, spatial, animal, and meta-analysis
> worked examples stay hidden until their article-specific gates pass.
> The per-capability validation status lives in the validation-debt
> register.

| correlation \ mode | scalar | indep | dep | latent |
|---|---|---|---|---|
| none | (omit) | `indep()` | `dep()` | `latent()` |
| **animal** | `animal_scalar()` | `animal_indep()` | `animal_dep()` | `animal_latent()` |
| phylo | `phylo_scalar()` | `phylo_indep()` | `phylo_dep()` | `phylo_latent()` |
| spatial | `spatial_scalar()` | `spatial_indep()` | `spatial_dep()` | `spatial_latent()` |

Plus the **random-slope** keywords alongside the grid:
`phylo_slope(x | species)` and `animal_slope(x | id)` — for
per-group random regression slopes on a continuous covariate.

The diagonal `Psi` companion (`Sigma = Lambda Lambda^T + Psi`) is controlled
by the `unique =` **argument** on `latent()` / `*_latent()`, not by a separate
keyword. Ordinary `latent()` carries `Psi` by default; use
`latent(..., unique = FALSE)` for the loadings-only subset, and
`*_latent(..., unique = TRUE)` to fold the `Psi` companion into a
source-specific term. This includes the augmented random-slope form
`*_latent(1 + x | group, unique = TRUE)`, which folds in the intercept–slope
companion for the ordinary, `phylo`, and `animal` sources. The spatial
augmented fold (`spatial_latent(1 + x, unique = TRUE)`) is still under
validation and the kernel augmented fold is future work. For a
standalone marginal-only diagonal, use `indep()` / `*_indep()`.

Status labels during the public-surface reset:

| row | Public status |
|---|---|
| none | covered for long/wide grammar and the ordinary Gaussian `latent()` path with default $\boldsymbol\Psi$; standalone `indep()` and `dep()` are partial outside that core, and the scalar cell has no public no-prefix keyword |
| animal | covered for the five covariance keywords, the A/Ainv helper path, and `animal_slope()`; cross-package agreement remains partial |
| phylo | covered for the main paired decomposition, the sparse A^-1 route, and the scalar, marginal-only, saturated, and slope paths |
| spatial | covered for mesh construction, mode dispatch, orientation, scalar/marginal-only/saturated recovery, and legacy-alias deprecation; the augmented `spatial_latent(unique = TRUE)` total-covariance recovery remains partial |
| meta_V | partial for exact single-V workflows, covered for block-V construction, blocked for proportional mode |

The per-capability validation status lives in the validation-debt register.

Use this page as a grammar lookup. For a worked example, start with
[`morphometrics`](morphometrics.html) for non-phylogenetic
ordinary `latent()` models. Phylogenetic, spatial, animal, and
meta-analysis worked examples are under audit until their validation
and rendered HTML gates pass.

## The Four Modes

The mode controls the trait covariance matrix `Sigma` at one model
level.

| mode | Meaning | Use when |
|---|---|---|
| scalar | one shared variance for all traits | every trait should have the same group-level scale |
| indep | a standalone diagonal covariance with no cross-trait covariance | you want to state marginal-only intent explicitly |
| dep | a full unstructured positive-definite `Sigma` | the number of traits is small enough for a saturated covariance |
| latent | low-rank shared covariance `Lambda Lambda^T`; ordinary no-prefix `latent()` also includes the diagonal `Psi` companion by default | a few latent variables should explain cross-trait association while retaining trait-specific variance |

For ordinary no-prefix models, `latent()` is the default decomposed model:

```text
Sigma = Lambda Lambda^T + Psi
Psi = diag(psi_1, ..., psi_T)
```

The same alignment appears in the worked articles:

| Goal | Long syntax | Wide syntax | Mathematical target |
|---|---|---|---|
| decomposed covariance | `latent(0 + trait | unit, d = K)` | `latent(1 | unit, d = K)` | $\Sigma = \Lambda \Lambda^\top + \Psi$ |
| shared latent variables only | `latent(0 + trait | unit, d = K, unique = FALSE)` | `latent(1 | unit, d = K, unique = FALSE)` | $\Lambda \Lambda^\top$ |
| marginal-only diagonal | `indep(0 + trait | unit)` | `indep(1 | unit)` | diagonal $\Sigma$ with no cross-trait covariance |
| saturated covariance | `dep(0 + trait | unit)` | `dep(1 | unit)` | full unstructured $\Sigma$ |

Ordinary `latent()` fits the shared-plus-diagonal-Psi decomposition;
use `latent(..., unique = FALSE)` for the loadings-only subset, and `indep()`
for a standalone marginal-only diagonal model. Old ordinary `common = TRUE`
formulas should move to `indep(..., common = TRUE)` for standalone diagonal
models, or to `latent(..., common = TRUE)` for ordinary paired decompositions.
`dep()` is the full unstructured baseline and should not be combined with the
matching `latent()` or `indep()` term at the same level.

## The Four Correlation Rows

The row controls how the grouping units are correlated. The rows
are pedagogically ordered from finest-grained (pedigree, one
individual at a time) to broadest (geographic distance).

| row | Unit correlation | Typical grouping | How A is supplied |
|---|---|---|---|
| none | independent groups | sites, individuals, sessions | — |
| **animal** | **the additive-genetic relatedness matrix A** | **individuals in a pedigree** | **`pedigree =` (3-column id/sire/dam) or `A =` or `Ainv =`** |
| phylo | a phylogenetic relationship matrix from `tree` or `vcv` | species or clades | `tree =` or `vcv =` or **`A =`** or **`Ainv =`** (A / Ainv aliases align with `animal_*` per Design 14 §3) |
| spatial | an SPDE/GMRF spatial precision matrix from `mesh` | sites with coordinates | `mesh =` |

**The A vs V naming boundary.** The `animal_*` and `phylo_*` keywords
accept the **A** (or **Ainv**) argument for *relatedness*. The
separate `meta_V(V = V)` keyword accepts **V** for
*sampling variance* in meta-analysis. These are mathematically
distinct (relatedness covariance vs measurement-error variance),
and the argument naming keeps them visually separated.

No-prefix keywords such as `latent()`, `indep()`, and `dep()` use
independent group effects. The animal row uses
`animal_*()` keywords with a pedigree (or pre-computed A / Ainv).
The phylogenetic row uses `phylo_*()` keywords and a tree or
covariance matrix. The spatial row
uses `spatial_*()` keywords and an SPDE mesh, often created with
`make_mesh()`.

## Long And Wide Syntax

Most readers can start from a wide data frame: one row per unit and one
column per response. Use `traits(...)` on the formula left-hand side to
name those response columns and keep the right-hand side compact. If
your data are already long, use one row per `(unit, trait)` observation
and pass `trait =`.

```{r}
# Long format
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | individual, d = 2),
  data = df_long,
  trait = "trait",
  unit = "individual"
)

# Wide data-frame format
fit_wide <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~ 1 +
    latent(1 | individual, d = 2),
  data = df_wide,
  unit = "individual"
)
```

In the wide form, `1` supplies the trait-specific intercepts and
`latent(1 | individual)` expands to `latent(0 + trait | individual)`;
ordinary `latent()` now includes the diagonal Psi companion by default.
The same shorthand applies to `indep()`, `dep()`, bar-style
`phylo_indep()` / `phylo_dep()`, and `spatial_*()` terms.
Species-axis phylogenetic calls such as
`phylo_latent(species, d = K, tree = tree)` and
`phylo_indep(species, tree = tree)` already name their phylogenetic
axis, so they pass through unchanged.

## Per-Cell Examples

Use one cell from a row at a time unless you are deliberately showing
explicit-Psi compatibility syntax. The examples below show the long-form
syntax; the wide-form equivalent replaces `0 + trait` with `1` inside
bar-style terms.

### No Correlation

```{r}
# Explicit marginal-only diagonal covariance
indep(0 + trait | unit)

# Full unstructured trait covariance
dep(0 + trait | unit)

# Reduced-rank shared trait covariance plus default diagonal Psi
latent(0 + trait | unit, d = 2)
```

The scalar no-correlation cell is not a covariance keyword. Ordinary
random intercepts such as `(1 | batch)` pass through as ordinary random
intercepts rather than trait covariance terms.

### Animal

```{r}
# One shared additive-genetic variance
animal_scalar(id, pedigree = ped)

# Explicit marginal-only animal-model diagonal
animal_indep(0 + trait | id, pedigree = ped)

# Full unstructured animal-model trait covariance
animal_dep(0 + trait | id, pedigree = ped)

# Reduced-rank additive-genetic shared covariance plus optional Psi
animal_latent(id, d = 2, pedigree = ped, unique = TRUE)
```

The animal row uses the same four covariance modes as the grid, but
the correlation among individuals comes from a pedigree, a dense
relatedness matrix `A`, or a sparse inverse `Ainv`. Use
`animal_latent(..., unique = TRUE)` for a decomposed additive-genetic model,
`animal_latent(..., unique = FALSE)` for the loadings-only subset,
`animal_indep()` for a marginal-only diagonal animal model, and
`animal_dep()` as the saturated small-trait comparator.

### Phylogenetic

```{r}
# One shared phylogenetic variance
phylo_scalar(species, tree = tree)

# Explicit marginal-only phylogenetic diagonal
phylo_indep(0 + trait | species, tree = tree)

# Full unstructured phylogenetic trait covariance
phylo_dep(0 + trait | species, tree = tree)

# Reduced-rank phylogenetic shared covariance
phylo_latent(species, d = 2, tree = tree)
```

Use `phylo_latent(..., unique = TRUE)` for the folded decomposition of
shared phylogenetic trait axes plus trait-specific phylogenetic
variance. The default `phylo_latent()` is loadings-only.
Use `phylo_dep()` as a saturated comparator when the number of traits
is small enough.

### Spatial

```{r}
# One shared spatial variance
spatial_scalar(0 + trait | site, mesh = mesh)

# Explicit marginal-only spatial fields
spatial_indep(0 + trait | site, mesh = mesh)

# Full unstructured spatial trait covariance
spatial_dep(0 + trait | site, mesh = mesh)

# Reduced-rank shared spatial fields plus optional spatial Psi
spatial_latent(0 + trait | site, d = 2, mesh = mesh, unique = TRUE)
```

The spatial row uses the SPDE/GMRF approximation inherited from
`sdmTMB`. Prefer `spatial_indep()` for a marginal-only spatial
diagonal. The folded `spatial_latent(..., unique = TRUE)`
shared-plus-Psi spatial decomposition is accepted grammar, but its
total-covariance recovery is still under validation.

## Helpers Outside The Grid

Three exported helpers are related to the grid but are not grid
cells.

`phylo_slope(x | species)` adds a phylogenetic random slope for a
covariate `x`. It is an extension hook, not one of the five covariance
modes.

`animal_slope(x | id)` adds the animal-model analogue: a random slope
whose groups are related by a pedigree, `A`, or `Ainv`. Its recovery is
validated against the matching `phylo_slope()` call.

`meta_V(V = V)` adds known sampling covariance for
meta-analytic workflows. It uses a user-supplied covariance matrix
rather than one of the grid's estimated correlation rows. Exact
single-V workflows are still under validation, block-V construction is
covered, and proportional known-V mode is not yet available (it awaits
a post-CRAN validation path).

## What Not To Combine

The parser rejects combinations that tell the same covariance story
twice at the same level:

- `dep()` with `latent()` or `indep()` on the same group;
- `indep()` with `latent()` on the same group;
- `phylo_dep()` with `phylo_latent()` or `phylo_indep()` for the same
  phylogenetic tier;
- `spatial_dep()` with `spatial_latent()` or `spatial_indep()` for the
  same spatial tier.

When in doubt, start with `latent()` for a decomposed model,
`indep()` for a diagonal baseline, and `dep()` for a saturated
baseline. Then use the row prefix only when the grouping units are
phylogenetically or spatially correlated.
