---
title: "Formula keyword grid"
tier: 2 # technical reference for the covariance keyword grammar
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Formula keyword grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This reference is for readers who already know they need a
stacked-trait model and want to choose the right covariance keyword.
The keywords are organised as a 3 x 5 grid: the row names the
correlation across units, and the column names the covariance mode
across traits.

| correlation \ mode | scalar | unique | indep | dep | latent |
|---|---|---|---|---|---|
| none | (omit) | `unique()` | `indep()` | `dep()` | `latent()` |
| phylo | `phylo_scalar()` | `phylo_unique()` | `phylo_indep()` | `phylo_dep()` | `phylo_latent()` |
| spatial | `spatial_scalar()` | `spatial_unique()` | `spatial_indep()` | `spatial_dep()` | `spatial_latent()` |

Use this page as a grammar lookup. For a worked example, start with
[`morphometrics`](morphometrics.html) for non-phylogenetic
`latent() + unique()` models or
[`phylogenetic-gllvm`](phylogenetic-gllvm.html) for phylogenetic
trait covariance.

## The Five Modes

The mode controls the trait covariance matrix `Sigma` at one model
level.

| mode | Meaning | Use when |
|---|---|---|
| scalar | one shared variance for all traits | every trait should have the same group-level scale |
| unique | a diagonal matrix `S` of trait-specific variances | pairing with `latent()` or fitting independent trait variances |
| indep | the same diagonal covariance as standalone `unique()` | you want to state marginal-only intent explicitly |
| dep | a full unstructured positive-definite `Sigma` | the number of traits is small enough for a saturated covariance |
| latent | low-rank shared covariance `Lambda Lambda^T` | a few latent axes should explain cross-trait association |

The most common decomposed model pairs `latent()` and `unique()`:

```text
Sigma = Lambda Lambda^T + S
```

Standalone `latent()` fits the no-residual reduced-rank subset.
Standalone `unique()` and `indep()` both fit independent per-trait
variances, but `indep()` tells the next reader that the marginal-only
interpretation is intentional. `dep()` is the full unstructured
baseline and should not be combined with the matching `latent()`,
`unique()`, or `indep()` term at the same level.

## The Three Correlation Rows

The row controls how the grouping units are correlated.

| row | Unit correlation | Typical grouping |
|---|---|---|
| none | independent groups | sites, individuals, sessions |
| phylo | a phylogenetic relationship matrix from `tree` or `vcv` | species or clades |
| spatial | an SPDE/GMRF spatial precision matrix from `mesh` | sites with coordinates |

No-prefix keywords such as `latent()`, `unique()`, `indep()`, and
`dep()` use independent group effects. The phylogenetic row uses
`phylo_*()` keywords and a tree or covariance matrix. The spatial row
uses `spatial_*()` keywords and an SPDE mesh, often created with
`make_mesh()`.

## Long And Wide Syntax

Long data are canonical: one row per `(unit, trait)` observation. Wide
data frames use `traits(...)` on the formula left-hand side; that lets
the right-hand side use compact syntax.

```{r}
# Long format
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | individual, d = 2) +
    unique(0 + trait | individual),
  data = df_long,
  unit = "individual"
)

# Wide data-frame format
fit_wide <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~ 1 +
    latent(1 | individual, d = 2) +
    unique(1 | individual),
  data = df_wide,
  unit = "individual"
)
```

In the wide form, `1` supplies the trait-specific intercepts and
`latent(1 | individual)` expands to `latent(0 + trait | individual)`.
The same shorthand applies to `unique()`, `indep()`, `dep()`,
bar-style `phylo_indep()` / `phylo_dep()`, and `spatial_*()` terms.
Species-axis phylogenetic calls such as
`phylo_latent(species, d = K, tree = tree)` and
`phylo_unique(species, tree = tree)` already name their phylogenetic
axis, so they pass through unchanged.

## Per-Cell Examples

Use one cell from a row at a time unless you are fitting the
`latent + unique` decomposition. The examples below show the long-form
syntax; the wide-form equivalent replaces `0 + trait` with `1` inside
bar-style terms.

### No Correlation

```{r}
# Diagonal trait variances, often paired with latent()
unique(0 + trait | unit)

# Explicit marginal-only diagonal covariance
indep(0 + trait | unit)

# Full unstructured trait covariance
dep(0 + trait | unit)

# Reduced-rank shared trait covariance
latent(0 + trait | unit, d = 2)
```

The scalar no-correlation cell is not a covariance keyword. Ordinary
random intercepts such as `(1 | batch)` pass through as ordinary random
intercepts rather than trait covariance terms.

### Phylogenetic

```{r}
# One shared phylogenetic variance
phylo_scalar(species, tree = tree)

# Trait-specific phylogenetic diagonal
phylo_unique(species, tree = tree)

# Explicit marginal-only phylogenetic diagonal
phylo_indep(0 + trait | species, tree = tree)

# Full unstructured phylogenetic trait covariance
phylo_dep(0 + trait | species, tree = tree)

# Reduced-rank phylogenetic shared covariance
phylo_latent(species, d = 2, tree = tree)
```

Use `phylo_latent() + phylo_unique()` when the scientific question is
about shared phylogenetic trait axes plus trait-specific phylogenetic
variance. Use `phylo_dep()` as a saturated comparator when the number
of traits is small enough.

### Spatial

```{r}
# One shared spatial variance
spatial_scalar(0 + trait | site, mesh = mesh)

# Trait-specific spatial fields
spatial_unique(0 + trait | site, mesh = mesh)

# Explicit marginal-only spatial fields
spatial_indep(0 + trait | site, mesh = mesh)

# Full unstructured spatial trait covariance
spatial_dep(0 + trait | site, mesh = mesh)

# Reduced-rank shared spatial fields
spatial_latent(0 + trait | site, d = 2, mesh = mesh)
```

The spatial row uses the SPDE/GMRF approximation inherited from
`sdmTMB`. `spatial_unique()` is the canonical replacement for the
legacy `spatial()` alias.

## Helpers Outside The Grid

Two exported helpers are related to the grid but are not grid cells.

`phylo_slope(x | species)` adds a phylogenetic random slope for a
covariate `x`. It is an extension hook, not one of the five covariance
modes.

`meta_known_V(value, V = V)` adds known sampling covariance for
meta-analytic workflows. It uses a user-supplied covariance matrix
rather than one of the grid's estimated correlation rows.

## What Not To Combine

The parser rejects combinations that tell the same covariance story
twice at the same level:

- `dep()` with `latent()`, `unique()`, or `indep()` on the same group;
- `indep()` with `latent()` or `unique()` on the same group;
- `phylo_dep()` with `phylo_latent()`, `phylo_unique()`, or
  `phylo_indep()` for the same phylogenetic tier;
- `spatial_dep()` with `spatial_latent()`, `spatial_unique()`, or
  `spatial_indep()` for the same spatial tier.

When in doubt, start with `latent() + unique()` for a decomposed model,
`indep()` for a diagonal baseline, and `dep()` for a saturated
baseline. Then use the row prefix only when the grouping units are
phylogenetically or spatially correlated.
