---
title: "Formula keyword grid"
tier: 2 # technical reference for the covariance keyword grammar
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Formula keyword grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This reference is for readers who already know they need a
stacked-trait model and want to choose the right covariance keyword.
The keywords are organised as a 4 x 5 grid: the row names the
correlation across units, and the column names the covariance mode
across traits. The four correlation rows go from finest-grained
(individual pedigree) to broadest (geography).

| correlation \ mode | scalar | unique | indep | dep | latent |
|---|---|---|---|---|---|
| none | (omit) | `unique()` | `indep()` | `dep()` | `latent()` |
| **animal** | `animal_scalar()` | `animal_unique()` | `animal_indep()` | `animal_dep()` | `animal_latent()` |
| phylo | `phylo_scalar()` | `phylo_unique()` | `phylo_indep()` | `phylo_dep()` | `phylo_latent()` |
| spatial | `spatial_scalar()` | `spatial_unique()` | `spatial_indep()` | `spatial_dep()` | `spatial_latent()` |

Plus the **random-slope** keywords alongside the grid:
`phylo_slope(x | species)` and `animal_slope(x | id)` — for
per-group random regression slopes on a continuous covariate.

Status labels during the public-surface reset:

| row | Public status | Validation note |
|---|---|---|
| none | covered for long/wide grammar and the core Gaussian `latent() + unique()` path; standalone `indep()` and `dep()` are partial outside that core, and the scalar cell has no public no-prefix keyword | FG-01--FG-06 and FAM-01 covered; FG-07--FG-09 partial |
| animal | covered for the five covariance keywords and A/Ainv helper path; `animal_slope()` and cross-package agreement remain partial | ANI-01--ANI-05, ANI-07, and ANI-08 covered; ANI-06, ANI-09, and ANI-10 partial |
| phylo | covered for the main paired decomposition and sparse A^-1 route; scalar, marginal-only, saturated, and slope paths are partial at the current validation depth | FG-12 and PHY-01--PHY-03 covered; PHY-04--PHY-06 partial |
| spatial | covered for mesh construction, mode dispatch, orientation, and legacy-alias deprecation; covariance recovery depth remains partial | SPA-01 and SPA-05--SPA-07 covered; SPA-02--SPA-04 partial |
| meta_V | partial for exact single-V workflows, covered for block-V construction, blocked for proportional mode | MET-01 partial; MET-02 covered; MET-03 blocked; MET-04 partial |

Use this page as a grammar lookup. For a worked example, start with
[`morphometrics`](morphometrics.html) for non-phylogenetic
`latent() + unique()` models. Phylogenetic, spatial, animal, and
meta-analysis worked examples are under audit until their validation
and rendered HTML gates pass.

## The Five Modes

The mode controls the trait covariance matrix `Sigma` at one model
level.

| mode | Meaning | Use when |
|---|---|---|
| scalar | one shared variance for all traits | every trait should have the same group-level scale |
| unique | a diagonal matrix `Psi` of trait-specific variances `psi` | pairing with `latent()` or fitting independent trait variances |
| indep | the same diagonal covariance as standalone `unique()` | you want to state marginal-only intent explicitly |
| dep | a full unstructured positive-definite `Sigma` | the number of traits is small enough for a saturated covariance |
| latent | low-rank shared covariance `Lambda Lambda^T` | a few latent axes should explain cross-trait association |

The most common decomposed model pairs `latent()` and `unique()`:

```text
Sigma = Lambda Lambda^T + diag(psi)
```

The same alignment appears in the worked articles:

| Goal | Long syntax | Wide syntax | Mathematical target |
|---|---|---|---|
| shared latent axes | `latent(0 + trait | unit, d = K)` | `latent(1 | unit, d = K)` | $\Lambda \Lambda^\top$ |
| trait-specific diagonal | `unique(0 + trait | unit)` | `unique(1 | unit)` | $\Psi = \mathrm{diag}(\psi)$ |
| decomposed covariance | `latent(...) + unique(...)` | `latent(...) + unique(...)` | $\Sigma = \Lambda \Lambda^\top + \Psi$ |
| marginal-only diagonal | `indep(0 + trait | unit)` | `indep(1 | unit)` | diagonal $\Sigma$ with no cross-trait covariance |
| saturated covariance | `dep(0 + trait | unit)` | `dep(1 | unit)` | full unstructured $\Sigma$ |

Standalone `latent()` fits the no-residual reduced-rank subset.
Standalone `unique()` and `indep()` both fit independent per-trait
variances, but `indep()` tells the next reader that the marginal-only
interpretation is intentional. `dep()` is the full unstructured
baseline and should not be combined with the matching `latent()`,
`unique()`, or `indep()` term at the same level.

## The Four Correlation Rows

The row controls how the grouping units are correlated. The rows
are pedagogically ordered from finest-grained (pedigree, one
individual at a time) to broadest (geographic distance).

| row | Unit correlation | Typical grouping | How A is supplied |
|---|---|---|---|
| none | independent groups | sites, individuals, sessions | — |
| **animal** | **the additive-genetic relatedness matrix A** | **individuals in a pedigree** | **`pedigree =` (3-column id/sire/dam) or `A =` or `Ainv =`** |
| phylo | a phylogenetic relationship matrix from `tree` or `vcv` | species or clades | `tree =` or `vcv =` or **`A =`** or **`Ainv =`** (A / Ainv aliases align with `animal_*` per Design 14 §3) |
| spatial | an SPDE/GMRF spatial precision matrix from `mesh` | sites with coordinates | `mesh =` |

**The A vs V naming boundary.** The `animal_*` and `phylo_*` keywords
accept the **A** (or **Ainv**) argument for *relatedness*. The
separate `meta_V(V = V)` keyword accepts **V** for
*sampling variance* in meta-analysis. These are mathematically
distinct (relatedness covariance vs measurement-error variance),
and the argument naming keeps them visually separated. See
[`docs/design/14-known-relatedness-keywords.md`](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/14-known-relatedness-keywords.md)
§3 for the full rule.

No-prefix keywords such as `latent()`, `unique()`, `indep()`, and
`dep()` use independent group effects. The animal row uses
`animal_*()` keywords with a pedigree (or pre-computed A / Ainv).
The phylogenetic row uses `phylo_*()` keywords and a tree or
covariance matrix. The spatial row
uses `spatial_*()` keywords and an SPDE mesh, often created with
`make_mesh()`.

## Long And Wide Syntax

Most readers can start from a wide data frame: one row per unit and one
column per response. Use `traits(...)` on the formula left-hand side to
name those response columns and keep the right-hand side compact. If
your data are already long, use one row per `(unit, trait)` observation
and pass `trait =`.

```{r}
# Long format
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | individual, d = 2) +
    unique(0 + trait | individual),
  data = df_long,
  trait = "trait",
  unit = "individual"
)

# Wide data-frame format
fit_wide <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~ 1 +
    latent(1 | individual, d = 2) +
    unique(1 | individual),
  data = df_wide,
  unit = "individual"
)
```

In the wide form, `1` supplies the trait-specific intercepts and
`latent(1 | individual)` expands to `latent(0 + trait | individual)`.
The same shorthand applies to `unique()`, `indep()`, `dep()`,
bar-style `phylo_indep()` / `phylo_dep()`, and `spatial_*()` terms.
Species-axis phylogenetic calls such as
`phylo_latent(species, d = K, tree = tree)` and
`phylo_unique(species, tree = tree)` already name their phylogenetic
axis, so they pass through unchanged.

## Per-Cell Examples

Use one cell from a row at a time unless you are fitting the
`latent + unique` decomposition. The examples below show the long-form
syntax; the wide-form equivalent replaces `0 + trait` with `1` inside
bar-style terms.

### No Correlation

```{r}
# Diagonal trait variances, often paired with latent()
unique(0 + trait | unit)

# Explicit marginal-only diagonal covariance
indep(0 + trait | unit)

# Full unstructured trait covariance
dep(0 + trait | unit)

# Reduced-rank shared trait covariance
latent(0 + trait | unit, d = 2)
```

The scalar no-correlation cell is not a covariance keyword. Ordinary
random intercepts such as `(1 | batch)` pass through as ordinary random
intercepts rather than trait covariance terms.

### Animal

```{r}
# One shared additive-genetic variance
animal_scalar(id, pedigree = ped)

# Trait-specific additive-genetic diagonal
animal_unique(id, pedigree = ped)

# Explicit marginal-only animal-model diagonal
animal_indep(0 + trait | id, pedigree = ped)

# Full unstructured animal-model trait covariance
animal_dep(0 + trait | id, pedigree = ped)

# Reduced-rank additive-genetic shared covariance
animal_latent(id, d = 2, pedigree = ped)
```

The animal row uses the same five covariance modes as the grid, but
the correlation among individuals comes from a pedigree, a dense
relatedness matrix `A`, or a sparse inverse `Ainv`. Use
`animal_latent() + animal_unique()` for a decomposed additive-genetic
model, `animal_indep()` for a marginal-only diagonal animal model,
and `animal_dep()` as the saturated small-trait comparator.

### Phylogenetic

```{r}
# One shared phylogenetic variance
phylo_scalar(species, tree = tree)

# Trait-specific phylogenetic diagonal
phylo_unique(species, tree = tree)

# Explicit marginal-only phylogenetic diagonal
phylo_indep(0 + trait | species, tree = tree)

# Full unstructured phylogenetic trait covariance
phylo_dep(0 + trait | species, tree = tree)

# Reduced-rank phylogenetic shared covariance
phylo_latent(species, d = 2, tree = tree)
```

Use `phylo_latent() + phylo_unique()` when the scientific question is
about shared phylogenetic trait axes plus trait-specific phylogenetic
variance. Use `phylo_dep()` as a saturated comparator when the number
of traits is small enough.

### Spatial

```{r}
# One shared spatial variance
spatial_scalar(0 + trait | site, mesh = mesh)

# Trait-specific spatial fields
spatial_unique(0 + trait | site, mesh = mesh)

# Explicit marginal-only spatial fields
spatial_indep(0 + trait | site, mesh = mesh)

# Full unstructured spatial trait covariance
spatial_dep(0 + trait | site, mesh = mesh)

# Reduced-rank shared spatial fields
spatial_latent(0 + trait | site, d = 2, mesh = mesh)
```

The spatial row uses the SPDE/GMRF approximation inherited from
`sdmTMB`. `spatial_unique()` is the canonical replacement for the
legacy `spatial()` alias.

## Helpers Outside The Grid

Three exported helpers are related to the grid but are not grid
cells.

`phylo_slope(x | species)` adds a phylogenetic random slope for a
covariate `x`. It is an extension hook, not one of the five covariance
modes.

`animal_slope(x | id)` adds the animal-model analogue: a random slope
whose groups are related by a pedigree, `A`, or `Ainv`. It is parser
accepted, but the validation register still marks recovery depth as
partial (ANI-06).

`meta_V(V = V)` adds known sampling covariance for
meta-analytic workflows. It uses a user-supplied covariance matrix
rather than one of the grid's estimated correlation rows. Exact
single-V workflows are partial (MET-01), block-V construction is
covered (MET-02), and proportional known-V mode is blocked until a
post-CRAN validation path lands (MET-03).

## What Not To Combine

The parser rejects combinations that tell the same covariance story
twice at the same level:

- `dep()` with `latent()`, `unique()`, or `indep()` on the same group;
- `indep()` with `latent()` or `unique()` on the same group;
- `phylo_dep()` with `phylo_latent()`, `phylo_unique()`, or
  `phylo_indep()` for the same phylogenetic tier;
- `spatial_dep()` with `spatial_latent()`, `spatial_unique()`, or
  `spatial_indep()` for the same spatial tier.

When in doubt, start with `latent() + unique()` for a decomposed model,
`indep()` for a diagonal baseline, and `dep()` for a saturated
baseline. Then use the row prefix only when the grouping units are
phylogenetically or spatially correlated.
