---
title: "Behavioural syndromes (2-level GLLVM)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Behavioural syndromes (2-level GLLVM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

## The biological question

Animal **personality** — consistent individual differences in behaviour
across time and context — has become central to ecology and evolution
(Réale et al. 2007). Réale and colleagues organised animal temperament
along five axes: boldness, exploration, activity, aggression, and
sociability. When multiple behaviours correlate consistently across
individuals, they form a **behavioural syndrome** (Sih et al. 2004).
The key empirical question is whether syndrome structure exists
*between* individuals and, separately, whether individuals show
correlated *within-individual* fluctuations across measurement
occasions.

These two sources of covariance are biologically distinct. Between-individual
covariance captures stable personality: a bold individual is also, on
average, more exploratory. Within-individual covariance captures
**behavioural lability** — whether an individual's motivational state
on a given day affects all of its behaviours simultaneously (Westneat
et al. 2015). Bell et al. (2009) synthesised repeatability data for
diverse behaviours and found a median repeatability of approximately
0.37 — meaning that between-individual variance accounts for roughly a
third of total phenotypic variance. Yet the within-individual residual
is far from noise: it carries information about plasticity, context
dependence, and measurement bias, and must be modelled explicitly
rather than discarded (Westneat et al. 2015).

A **two-level GLLVM** with `latent() + unique()` at both the between-
and within-individual tiers is the natural tool here. At the between-
individual tier it estimates a low-rank personality structure (the
syndrome). At the within-individual tier it estimates a lability
structure — shared fluctuations in behaviour across sessions — while
simultaneously delivering per-trait repeatabilities as ICCs. This
article demonstrates the whole workflow on a realistic simulated
dataset modelled after great-tit personality studies (e.g. Dingemanse
et al. 2002).

The [Individual morphometrics](morphometrics.html) article covers the
one-level case (one measurement per individual per trait). The
two-level extension shown here is the next rung on the
[complexity ladder](../index.html#models-by-complexity).

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
library(dplyr)
```

## Simulated data

### Design

We simulate a great-tit-style personality assay: **80 individuals**,
each tested in **5 independent sessions** separated by at least one
week (a standard protocol for estimating repeatability; Bell et al.
2009). At each session, eight behavioural measures are recorded: four
on a **proactive-reactive** axis (boldness, exploration speed, novel-
object approach, latency to emerge from cover) and four on an
**aggression-activity** axis (mirror aggression, dominance score,
open-arena activity, restlessness). The long-format data have
$80 \times 5 \times 8 = 3\,200$ rows.

> **Wide-format input.** Personality data often arrive *wide* (one row
> per `(individual, session)`, one column per trait). The formula-wide
> path names the response columns on the LHS and keeps the RHS compact:
> `traits(boldness, exploration, ..., restlessness) ~ 1 +
> latent(1 | individual, d = 2) + unique(1 | individual) +
> latent(1 | session_id, d = 1) + unique(1 | session_id)`. The long
> form below stays canonical because it makes the two grouping levels
> visible row by row.

### True parameter structure

The simulation has three layers:

1. **Between-individual personality structure** — a $T \times K_B$
   loading matrix $\boldsymbol{\Lambda}_B$ ($T = 8$ traits,
   $K_B = 2$ personality axes) plus per-trait unique variances
   $\mathbf{S}_B$ (diagonal).

2. **Within-individual lability axis** — a $T \times K_W$ loading
   matrix $\boldsymbol{\Lambda}_W$ ($K_W = 1$) capturing a shared
   motivational state that shifts all behaviours similarly on a given
   day, plus per-trait within-individual unique variances
   $\mathbf{S}_W$.

3. **Observation noise** is assumed negligible (Bell et al. 2009);
   for Gaussian responses any residual measurement error is absorbed
   into $\mathbf{S}_W$ and is not separately identifiable from
   $s^2_{Wt}$ at the row level.

The implied per-trait repeatability is:

$$
R_t \;=\; \frac{(\boldsymbol{\Sigma}_B)_{tt}}
               {(\boldsymbol{\Sigma}_B)_{tt} + (\boldsymbol{\Sigma}_W)_{tt}}.
$$

We set parameters so that empirical repeatabilities fall in the range
0.30–0.50, matching the Bell et al. (2009) meta-analytic median of
0.37.

```{r sim}
set.seed(2025)

n_ind     <- 80L
n_session <- 5L
T         <- 8L

trait_names <- c(
  "boldness",       # proactive-reactive axis (1-4)
  "exploration",
  "novel_object",
  "latency_emerge",
  "mirror_aggression",  # aggression-activity axis (5-8)
  "dominance",
  "arena_activity",
  "restlessness"
)

# ---- True between-individual loading matrix (8 x 2) ----------------------
# LV1 = "proactive-reactive" personality axis (boldness, exploration, ...)
# LV2 = "aggression-activity" axis (mirror-aggression, dominance, ...)
# Note: latency_emerge loads negatively on LV1 (short latency = proactive)
# Loadings scaled to give empirical repeatabilities near 0.37 (Bell et al. 2009)
Lambda_B_true <- matrix(c(
  # LV1 (proactive-reactive)  LV2 (aggression-activity)
   0.40,                       0.05,   # boldness
   0.35,                       0.05,   # exploration
   0.35,                       0.05,   # novel_object
  -0.33,                       0.00,   # latency_emerge (negative: short = proactive)
   0.00,                       0.42,   # mirror_aggression
   0.05,                       0.35,   # dominance
   0.05,                       0.38,   # arena_activity
   0.00,                       0.30    # restlessness
), nrow = T, ncol = 2, byrow = TRUE,
   dimnames = list(trait_names, c("proactive_reactive", "aggression_activity")))

# ---- Per-trait between-individual unique variances (S_B) -----------------
# Modest — about 0.06-0.07 on the latent scale. These represent stable
# between-individual differences not captured by the two shared axes.
S_B_true <- c(
  boldness         = 0.07,
  exploration      = 0.07,
  novel_object     = 0.07,
  latency_emerge   = 0.07,
  mirror_aggression = 0.06,
  dominance        = 0.07,
  arena_activity   = 0.07,
  restlessness     = 0.06
)

# ---- True within-individual loading vector (8 x 1) -----------------------
# A single "lability" axis: when motivation is high, all behaviours are
# more extreme (higher boldness, shorter latency, more aggression, etc.)
Lambda_W_true <- matrix(c(
   0.15,   # boldness
   0.14,   # exploration
   0.15,   # novel_object
  -0.12,   # latency_emerge (negative: motivated bird emerges faster)
   0.13,   # mirror_aggression
   0.12,   # dominance
   0.14,   # arena_activity
   0.12    # restlessness
), nrow = T, ncol = 1, byrow = TRUE,
   dimnames = list(trait_names, "lability"))

# ---- Per-trait within-individual unique variances (S_W) ------------------
# Smaller than S_B: within-individual fluctuations are more constrained
S_W_true <- c(
  boldness         = 0.10,
  exploration      = 0.09,
  novel_object     = 0.11,
  latency_emerge   = 0.07,
  mirror_aggression = 0.07,
  dominance        = 0.08,
  arena_activity   = 0.09,
  restlessness     = 0.08
)

# ---- Observation-level noise (independent measurement error) -------------
# Note: for Gaussian responses, sigma_eps^2 is absorbed into s_Wt^2 by the
# engine (both vary at the row level and are not separately identifiable),
# so we fold it into var_W below.  This matches the MS convention.
sigma_eps_true <- 0.48

# ---- Implied per-trait repeatabilities (target ~0.30-0.50) ---------------
var_B <- diag(Lambda_B_true %*% t(Lambda_B_true)) + S_B_true                      # Sigma_B diagonals
var_W <- diag(Lambda_W_true %*% t(Lambda_W_true)) + S_W_true + sigma_eps_true^2   # Sigma_W diagonals (sigma_eps absorbed)
R_true <- var_B / (var_B + var_W)
cat("True per-trait repeatabilities:\n")
round(R_true, 3)

# ---- Simulate data ---------------------------------------------------------
# Between-individual scores (one per individual, stable across sessions)
u_B <- matrix(rnorm(n_ind * 2), n_ind, 2)       # n_ind x K_B

# Per-individual trait means = mu_t + Lambda_B u_i + eps_B_it
mu_trait <- rep(0, T)  # global trait means (trait intercepts, set to 0 for clarity)
eps_B <- matrix(
  rnorm(n_ind * T, sd = sqrt(rep(S_B_true, each = n_ind))),
  n_ind, T, byrow = FALSE
)
between_ind <- matrix(mu_trait, n_ind, T, byrow = TRUE) +
               u_B %*% t(Lambda_B_true) + eps_B   # n_ind x T

# Session-level IDs (one factor level per individual x session combination)
session_id_levels <- paste(
  rep(seq_len(n_ind), each = n_session),
  rep(seq_len(n_session), n_ind),
  sep = "_"
)
n_obs <- n_ind * n_session  # 400 individual-sessions

# Within-individual lability scores (one per session)
u_W <- matrix(rnorm(n_obs, 0, 1), n_obs, 1)     # n_obs x K_W

# Per-session-within-individual deviation
eps_W <- matrix(
  rnorm(n_obs * T, sd = sqrt(rep(S_W_true, each = n_obs))),
  n_obs, T, byrow = FALSE
)
within_obs <- u_W %*% t(Lambda_W_true) + eps_W   # n_obs x T

# Observation noise (independent)
noise <- matrix(rnorm(n_obs * T, sd = sigma_eps_true), n_obs, T)

# Assemble: each session-level observation = individual mean + within dev + noise
ind_idx <- rep(seq_len(n_ind), each = n_session)  # which individual for each obs

# Wide matrix: n_obs rows, T columns
Y_wide <- between_ind[ind_idx, ] + within_obs + noise

# Long format ---------------------------------------------------------------
df <- data.frame(
  individual = factor(rep(rep(seq_len(n_ind), each = n_session), T)),
  session_id = factor(rep(session_id_levels, T)),
  trait      = factor(
    rep(trait_names, each = n_obs),
    levels = trait_names
  ),
  value = as.vector(Y_wide)
)

# Sanity checks
stopifnot(nrow(df) == n_ind * n_session * T)
stopifnot(nlevels(df$individual) == n_ind)
stopifnot(nlevels(df$session_id) == n_ind * n_session)
stopifnot(nlevels(df$trait) == T)

head(df, 8)
```

The `session_id` factor has $80 \times 5 = 400$ levels — one per
individual-session combination. This is the within-individual grouping
factor: all observations from the same individual on the same day
share a session-level random effect.

> **Factor level order matters.** `trait` is created with
> `levels = trait_names` so that the rows of estimated loading matrices
> are always in the same order as `Lambda_B_true`. Without explicit
> levels, R sorts alphabetically and the recovery comparison breaks.

## The model

The two-level GLLVM decomposes trait variance at two nested tiers:

$$
y_{its} \;=\; \mu_t
          \;+\; \boldsymbol{\lambda}_{Bt}^\top \mathbf{u}_{Bi}
          \;+\; e_{Bit}
          \;+\; \boldsymbol{\lambda}_{Wt}^\top \mathbf{u}_{Wi_s}
          \;+\; e_{Wit_s}
          \;+\; \varepsilon_{its},
$$

where:

- $\mu_t$ is the global mean for trait $t$,
- $\mathbf{u}_{Bi} \sim \mathcal{N}(\mathbf{0}, \mathbf{I}_{K_B})$ are
  the $K_B = 2$ **between-individual personality scores** for
  individual $i$ (stable across sessions),
- $e_{Bit} \sim \mathcal{N}(0, s_{Bt}^2)$ is the trait-specific
  between-individual residual (personality variance not captured by the
  two shared axes),
- $\mathbf{u}_{Wi_s} \sim \mathcal{N}(\mathbf{0}, \mathbf{I}_{K_W})$ are
  the $K_W = 1$ **within-individual lability scores** for session $s$
  of individual $i$,
- $e_{Wit_s} \sim \mathcal{N}(0, s_{Wt}^2)$ is the trait-specific
  within-individual residual (session-to-session fluctuation not shared
  with other traits, including any residual measurement error),
- $\varepsilon_{its} \sim \mathcal{N}(0, \sigma_\varepsilon^2)$ is
  independent observation noise. **For Gaussian responses
  $\sigma^2_\varepsilon$ and $s^2_{Wt}$ both vary at the row level
  and are not separately identifiable**; the engine absorbs
  $\sigma^2_\varepsilon$ into $s^2_{Wt}$ (note `unique_W` but no
  `sigma_eps` in the `summary()` output below).

The implied between-individual and within-individual trait covariances are:

$$
\boldsymbol{\Sigma}_B = \boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^\top + \mathbf{S}_B,
\qquad
\boldsymbol{\Sigma}_W = \boldsymbol{\Lambda}_W \boldsymbol{\Lambda}_W^\top + \mathbf{S}_W.
$$

The per-trait repeatability (ICC) is then:

$$
R_t = \frac{(\boldsymbol{\Sigma}_B)_{tt}}{(\boldsymbol{\Sigma}_B)_{tt} + (\boldsymbol{\Sigma}_W)_{tt}}.
$$

In the `gllvmTMB` formula:

- `latent(0 + trait | individual, d = 2)` fits $\boldsymbol{\Lambda}_B$
  and the individual scores $\mathbf{U}_B$.
- `unique(0 + trait | individual)` fits the diagonal $\mathbf{S}_B$.
- `latent(0 + trait | session_id, d = 1)` fits $\boldsymbol{\Lambda}_W$
  and the session scores $\mathbf{U}_W$.
- `unique(0 + trait | session_id)` fits the diagonal $\mathbf{S}_W$.

The `unit = "individual"` argument tells the engine which column
defines the between-individual grouping, and `unit_obs = "session_id"`
identifies the within-individual grouping.

```{r fit}
fit <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | individual, d = 2) +   # between-individual personality axes
          unique(0 + trait | individual) +           # trait-specific between-individual residual
          latent(0 + trait | session_id, d = 1) +   # within-individual lability axis
          unique(0 + trait | session_id),            # trait-specific within-individual residual
  data     = df,
  unit     = "individual",
  unit_obs = "session_id"
)
fit
```

## Overview: `summary(fit)`

```{r summary}
summary(fit)
```

The summary panel reports fixed-effect trait intercepts ($\hat{\mu}_t$),
the estimated between-individual ($\hat{\boldsymbol{\Sigma}}_B$) and
within-individual ($\hat{\boldsymbol{\Sigma}}_W$) covariance structures,
per-trait ICCs (repeatabilities), communalities, and convergence
diagnostics. The maximum gradient element and positive-definite Hessian
flag indicate whether the MLE is reliable.

## Detailed inspection

### Between-individual trait correlation (personality structure)

```{r inspect}
# Between-individual correlation matrix
R_B_hat <- extract_Sigma(fit, level = "unit")$R
cat("Estimated between-individual trait correlation matrix (R_B):\n")
round(R_B_hat, 3)
```

The off-diagonal pattern reveals the syndrome structure. Traits on the
same personality axis (proactive-reactive, or aggression-activity) are
expected to correlate positively; across axes the correlations should
be weak.

### Within-individual trait correlation (lability structure)

```{r inspect-w}
# Within-individual correlation matrix
R_W_hat <- extract_Sigma(fit, level = "unit_obs")$R
cat("Estimated within-individual trait correlation matrix (R_W):\n")
round(R_W_hat, 3)
```

The within-individual correlations reflect shared motivational state
across sessions. Because the true $\boldsymbol{\Lambda}_W$ has all
positive entries (except for `latency_emerge`), we expect most
within-individual correlations to be positive — mirroring the overall
motivational pattern.

### Per-trait repeatabilities

```{r inspect-icc}
# Per-trait repeatabilities (ICC_t = Sigma_B_tt / (Sigma_B_tt + Sigma_W_tt))
# For Gaussian responses, sigma_eps^2 is absorbed into s_Wt^2 (not separately
# identifiable), so it lives inside (Sigma_W)_tt -- this matches R_true above.
rep_df <- extract_repeatability(fit)
cat("\nEstimated repeatabilities (R_t) with 95% CIs:\n")
print(rep_df)
cat("\nTrue repeatabilities:\n")
print(round(R_true, 3))
cat("\nMedian estimated repeatability:", round(median(rep_df$R), 3),
    " (Bell et al. 2009 meta-analytic median ~ 0.37)\n")
```

### Between-individual communalities

```{r inspect-comm-b}
# Communality at B level: proportion of between-individual variance that is
# shared across traits via the two personality axes
comm_B <- extract_communality(fit, "B")
cat("Between-individual communalities (shared via personality axes):\n")
round(comm_B, 3)
```

A communality near 0.6 means 60% of that trait's between-individual
variance is shared with the other traits through the two latent
personality axes; the remaining 40% is trait-specific.

### Within-individual communalities

```{r inspect-comm-w}
# Communality at W level: proportion of within-individual variance that is
# shared across traits via the lability axis
comm_W <- extract_communality(fit, "W")
cat("Within-individual communalities (shared via lability axis):\n")
round(comm_W, 3)
```

Within-individual communalities quantify how much of each trait's
session-to-session fluctuation is correlated with the fluctuations of
all other traits — i.e. driven by the shared lability axis. Lower values
here indicate largely trait-specific plasticity.

## Ordination: individuals in personality space

```{r ord, fig.cap = "Individuals in personality space. Each point is one of the 80 great-tit individuals, plotted on the two between-individual latent axes (LV1 = proactive-reactive, LV2 = aggression-activity). Spread along each axis represents variation in that personality dimension across individuals."}
ord_B <- extract_ordination(fit, "B")
scores_df <- as.data.frame(ord_B$scores)
scores_df$individual <- rownames(scores_df)

ggplot(scores_df, aes(LV1, LV2)) +
  geom_hline(yintercept = 0, colour = "grey70", linetype = 2) +
  geom_vline(xintercept = 0, colour = "grey70", linetype = 2) +
  geom_point(colour = "#2c7bb6", alpha = 0.75, size = 2.2) +
  labs(
    x     = "LV1 (proactive-reactive axis)",
    y     = "LV2 (aggression-activity axis)",
    title = "Individuals in personality space"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

Each point is one individual's position on the two between-individual
latent axes. Individuals with high LV1 scores are more proactive
(bolder, faster to explore, quicker to emerge) and those with high LV2
scores are more aggressive and active. The cloud of points covers the
full personality space, confirming that both axes capture genuine
individual variation.

## Recovery: estimated vs true loadings

Because latent factors are identified only up to an orthogonal
rotation, we use Procrustes alignment (via `compare_loadings()`) before
comparing the estimated $\hat{\boldsymbol{\Lambda}}_B$ to the true
$\boldsymbol{\Lambda}_B$.

```{r recovery, fig.cap = "Estimated vs true between-individual loadings (Lambda_B) after Procrustes alignment. Each point is one (trait, latent-factor) loading entry. Points on or near the 1:1 dashed line indicate clean recovery."}
Lambda_B_hat <- ord_B$loadings  # T x K_B

# Procrustes-align Lambda_B_hat to Lambda_B_true
proc_B <- compare_loadings(Lambda_B_hat, Lambda_B_true)
Lambda_B_aligned <- proc_B$Lambda_a_rot

# Assemble for plotting
df_recovery <- data.frame(
  trait  = rep(trait_names, ncol(Lambda_B_true)),
  factor = rep(colnames(Lambda_B_true), each = T),
  true   = as.vector(Lambda_B_true),
  fitted = as.vector(Lambda_B_aligned)
)

cat("Correlation (estimated vs true) per personality axis after Procrustes alignment:\n")
cat("  LV1:", round(proc_B$cor_per_factor[1], 3), "\n")
cat("  LV2:", round(proc_B$cor_per_factor[2], 3), "\n")

ggplot(df_recovery, aes(true, fitted, colour = factor, shape = factor)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_vline(xintercept = 0, colour = "grey80") +
  geom_point(size = 3, alpha = 0.85) +
  scale_colour_manual(
    values = c(proactive_reactive = "#2c7bb6", aggression_activity = "#d7191c"),
    name   = "Personality axis"
  ) +
  scale_shape_manual(
    values = c(proactive_reactive = 16, aggression_activity = 17),
    name   = "Personality axis"
  ) +
  labs(
    x     = expression("True " * Lambda[B] * " entry"),
    y     = expression("Estimated " * hat(Lambda)[B] * " entry (Procrustes-aligned)"),
    title = "Recovery of between-individual loadings"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

```{r recovery-sigma, fig.cap = "Estimated vs true between-individual trait correlations (off-diagonal entries of Sigma_B). Perfect recovery would put all points on the 1:1 line."}
# Optional: compare Sigma_B off-diagonal entries
Sigma_B_true <- Lambda_B_true %*% t(Lambda_B_true) + diag(S_B_true)
Sigma_B_hat  <- extract_Sigma(fit, level = "unit")$Sigma

# Off-diagonal indices
n_traits <- T
off_diag <- which(lower.tri(matrix(0, n_traits, n_traits)))
true_corr <- cov2cor(Sigma_B_true)[off_diag]
hat_corr  <- cov2cor(Sigma_B_hat)[off_diag]

cat("Correlation between estimated and true Sigma_B off-diagonal entries:",
    round(cor(hat_corr, true_corr), 3), "\n")

df_sigma <- data.frame(true = true_corr, fitted = hat_corr)

ggplot(df_sigma, aes(true, fitted)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(colour = "#2c7bb6", alpha = 0.80, size = 2.5) +
  labs(
    x     = expression("True " * Sigma[B] * " correlation (off-diagonal)"),
    y     = expression("Estimated " * hat(Sigma)[B] * " correlation (off-diagonal)"),
    title = "Recovery of between-individual trait correlations"
  ) +
  xlim(-1, 1) + ylim(-1, 1) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## Summary of recovered parameters

```{r recovery-summary}
# Compact table: repeatability and communality recovery
tab <- data.frame(
  trait        = trait_names,
  R_true       = round(R_true, 3),
  R_hat        = round(rep_df$R, 3),
  R_lower      = round(rep_df$lower, 3),
  R_upper      = round(rep_df$upper, 3),
  comm_B_hat   = round(comm_B, 3),
  comm_W_hat   = round(comm_W, 3)
)
tab
cat("\nMean absolute error in repeatability:", round(mean(abs(tab$R_hat - tab$R_true)), 3), "\n")
```

Estimated repeatabilities fall within a reasonable band of the true
values. The communality estimates quantify the syndrome strength: a
communality of 0.6 on `boldness` means that 60% of the stable
between-individual variance in boldness is shared with other traits
via the two personality axes — the hallmark of a genuine behavioural
syndrome (Bell et al. 2009; Réale et al. 2007).

## What this model reveals about the syndrome

The two-level GLLVM answers three distinct questions simultaneously:

1. **Syndrome strength** — the between-individual trait correlations
   ($\hat{\boldsymbol{R}}_B$) and communalities quantify whether
   behaviours co-vary systematically across individuals.

2. **Personality axes** — the loading matrix $\hat{\boldsymbol{\Lambda}}_B$
   identifies the latent dimensions organising that syndrome, and the
   ordination scores place individuals along those dimensions.

3. **Lability structure** — the within-individual correlations
   ($\hat{\boldsymbol{R}}_W$) reveal whether behavioural fluctuations
   across sessions are coordinated (shared lability axis) or trait-
   specific. Non-trivial within-individual communalities indicate that
   individual motivational state or context — not just measurement
   noise — drives correlated variation across sessions.

Repeatabilities connect both levels: a high ICC means the between-
individual variance dominates, so syndrome conclusions generalise
across contexts; a low ICC implies that the within-individual signal
is large and syndrome inferences are session-specific.

## References

Bell, A. M., Hankison, S. J. & Laskowski, K. L. (2009) The
repeatability of behaviour: a meta-analysis. *Animal Behaviour*
**77**, 771–783. <https://doi.org/10.1016/j.anbehav.2008.12.022>

Dingemanse, N. J. & Wright, J. (eds. 2020) *Animal Personality.*
Springer.

Réale, D., Reader, S. M., Sol, D., McDougall, P. T. & Dingemanse,
N. J. (2007) Integrating animal temperament within ecology and
evolution. *Biological Reviews* **82**, 291–318.
<https://doi.org/10.1111/j.1469-185X.2007.00010.x>

Sih, A., Bell, A. M. & Johnson, J. C. (2004) Behavioral syndromes:
an ecological and evolutionary overview. *Trends in Ecology and
Evolution* **19**, 372–378.

Westneat, D. F., Wright, J. & Dingemanse, N. J. (2015) The biology
hidden inside residual within-individual phenotypic variation.
*Biological Reviews* **90**, 729–743.
<https://doi.org/10.1111/brv.12131>

Nakagawa et al. (in prep). *Functional biogeography using generalised
linear latent variable modelling.* (the conceptual paper motivating
`gllvmTMB`'s unit × trait framing and the two-level decomposition).
