---
title: "Behavioural syndromes in repeated measurements"
description: "Use repeated Gaussian behaviours to separate stable between-individual covariance from within-occasion covariance and point repeatability."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Behavioural syndromes in repeated measurements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.2,
  fig.height = 4.8,
  fig.align = "center",
  cache = FALSE,
  message = FALSE
)
```

> **Evidence boundary.** This article demonstrates point estimation for one
> Gaussian repeated-measures
> model. It does not establish confidence-interval calibration.

A behavioural syndrome is stable covariance among behaviours at the
individual level. Repeated measurements also contain covariance among
behaviours within an individual from one occasion to another. Those are
different biological summaries:

- **Between individuals:** do consistently bolder individuals also tend to be
  more exploratory or active?
- **Within individuals:** when an individual's measurement departs from its
  usual level on one occasion, do several behaviours depart together?

This article separates those two covariance matrices and computes
trait-specific repeatability. The within-occasion component can combine
short-term state, unmeasured context, and measurement error; the model does not
identify which of those mechanisms caused it.

## The model

For individual \(i\), occasion \(s\), and trait \(t\), we fit

$$
y_{ist} = \mu_t + b_{it} + w_{ist}.
$$

$$
\mathbf b_i \sim \operatorname{MVN}(0, \Sigma_B),
\qquad
\mathbf w_{is} \sim \operatorname{MVN}(0, \Sigma_W).
$$

Each covariance uses a low-rank shared component plus a trait-specific
diagonal:

$$
\Sigma_B = \Lambda_B\Lambda_B^\top + \Psi_B,
\qquad
\Sigma_W = \Lambda_W\Lambda_W^\top + \Psi_W.
$$

The per-trait repeatability, or intraclass correlation, is

$$
R_t =
\frac{(\Sigma_B)_{tt}}
     {(\Sigma_B)_{tt} + (\Sigma_W)_{tt}}.
$$

| Target | Formula term | Interpretation |
|---|---|---|
| \(\Sigma_B\) | `latent(... \| individual)` | Stable covariance among individual means |
| \(\Sigma_W\) | `latent(... \| occasion)` | Covariance among occasion-level deviations, including inseparable measurement error |
| \(R_t\) | both tiers | Proportion of fitted variance that lies between individuals for trait \(t\) |

Repeatability is not heritability: stable genetic and environmental
differences both contribute to \(\Sigma_B\). This model is also not a
reaction norm. A reaction norm includes a measured environmental gradient and
asks whether individuals differ in their slopes along it.

## Data requirements

The example uses standardized continuous traits and a Gaussian likelihood.
Before fitting real data:

1. Give every measurement occasion a unique identifier nested within
   individual. Reusing `occasion = 1, 2, ...` across individuals would join
   unrelated observations into the same random-effect level.
2. Confirm that individuals have repeated occasions. Uneven replication is
   allowed, but individuals with little information contribute little to the
   between/within separation.
3. Check that each `(individual, occasion, trait)` cell represents the
   intended observation. Resolve accidental duplicates before fitting.
4. Put continuous traits on interpretable, comparable scales when their raw
   units differ greatly. Counts, proportions, and ordinal scores need
   likelihoods appropriate to their sampling process rather than automatic
   Gaussian treatment.
5. Choose the two ranks deliberately. This teaching simulation plants
   \(d_B=2\) and \(d_W=1\); it does not estimate those ranks.

See [Choosing latent rank](model-selection-latent-rank.html) for rank
sensitivity and [Handling missing data](missing-data.html) when the repeated
trait grid is incomplete.

## Simulate a truth-checked example

```{r setup}
library(gllvmTMB)
```

We simulate six behaviours for 150 individuals measured on six occasions.
The between-individual tier has two anonymous latent axes; the occasion tier
has one. We do not attach biological names to raw latent axes because their
orientation and sign are not identified without an explicit rotation or
constraint.

```{r simulate}
set.seed(2025)
n_individual <- 150L
n_occasion <- 6L
trait_names <- c(
  "boldness", "exploration", "latency_emerge",
  "aggression", "activity", "restlessness"
)
n_traits <- length(trait_names)

Lambda_B_true <- matrix(
  c(
     0.55, 0.00,
     0.45, 0.05,
    -0.40, 0.00,
     0.05, 0.55,
     0.05, 0.45,
     0.00, 0.35
  ),
  n_traits,
  2,
  byrow = TRUE,
  dimnames = list(trait_names, c("B1", "B2"))
)
psi_B_true <- setNames(rep(0.10, n_traits), trait_names)

Lambda_W_true <- matrix(
  c(0.25, 0.20, -0.18, 0.22, 0.20, 0.18),
  n_traits,
  1,
  dimnames = list(trait_names, "W1")
)
# This diagonal includes unshared occasion variation and measurement error.
psi_W_true <- setNames(rep(0.35, n_traits), trait_names)

Sigma_B_true <- tcrossprod(Lambda_B_true) + diag(psi_B_true)
Sigma_W_true <- tcrossprod(Lambda_W_true) + diag(psi_W_true)
R_true <- diag(Sigma_B_true) /
  (diag(Sigma_B_true) + diag(Sigma_W_true))

between <- matrix(rnorm(n_individual * 2), n_individual, 2) %*%
  t(Lambda_B_true) +
  sweep(
    matrix(rnorm(n_individual * n_traits), n_individual, n_traits),
    2,
    sqrt(psi_B_true),
    "*"
  )

n_rows_wide <- n_individual * n_occasion
individual_index <- rep(seq_len(n_individual), each = n_occasion)
within <- matrix(rnorm(n_rows_wide), n_rows_wide, 1) %*%
  t(Lambda_W_true) +
  sweep(
    matrix(rnorm(n_rows_wide * n_traits), n_rows_wide, n_traits),
    2,
    sqrt(psi_W_true),
    "*"
  )
Y <- between[individual_index, ] + within

df_wide <- data.frame(
  individual = factor(individual_index),
  occasion = factor(paste(
    individual_index,
    rep(seq_len(n_occasion), n_individual),
    sep = "_"
  ))
)
df_wide[trait_names] <- as.data.frame(Y)

df_long <- data.frame(
  individual = factor(rep(individual_index, n_traits)),
  occasion = factor(rep(df_wide$occasion, n_traits)),
  trait = factor(
    rep(trait_names, each = n_rows_wide),
    levels = trait_names
  ),
  value = as.vector(Y)
)

stopifnot(
  !anyDuplicated(df_long[c("individual", "occasion", "trait")]),
  all(table(df_long$individual) == n_occasion * n_traits),
  all(table(df_long$occasion) == n_traits)
)
head(df_long)
```

The example puts all unshared occasion-level variation, including measurement
error, inside \(\Psi_W\). That matches what this data resolution can identify;
it does not pretend to estimate a separate measurement-error variance.

## Fit long and wide forms

The long form has one row per individual-occasion-trait observation. The wide
form has one row per individual occasion and one response column per trait.
Both use the same `gllvmTMB()` entry point.

```{r fit}
fit_control <- gllvmTMBcontrol(
  start_method = list(method = "indep"),
  optimizer = "optim",
  optArgs = list(method = "BFGS")
)

fit_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | individual, d = 2, unique = TRUE) +
    latent(0 + trait | occasion, d = 1, unique = TRUE),
  data = df_long,
  trait = "trait",
  unit = "individual",
  unit_obs = "occasion",
  family = gaussian(),
  control = fit_control
)

fit_wide <- gllvmTMB(
  traits(
    boldness, exploration, latency_emerge,
    aggression, activity, restlessness
  ) ~ 1 +
    latent(1 | individual, d = 2, unique = TRUE) +
    latent(1 | occasion, d = 1, unique = TRUE),
  data = df_wide,
  unit = "individual",
  unit_obs = "occasion",
  family = gaussian(),
  control = fit_control
)

c(
  long = as.numeric(logLik(fit_long)),
  wide = as.numeric(logLik(fit_wide)),
  absolute_difference = abs(
    as.numeric(logLik(fit_long)) - as.numeric(logLik(fit_wide))
  )
)
```

The two likelihoods should agree up to numerical precision. This verifies the
data-shape translation, not two independent scientific analyses.

## Inspect every numerical warning

```{r diagnostics}
health <- check_gllvmTMB(fit_long)

data.frame(
  converged = isTRUE(fit_long$fit_health$converged),
  raw_max_gradient = signif(fit_long$fit_health$max_gradient, 4),
  raw_gradient_threshold = 0.01,
  scaled_gradient_descriptive = signif(
    fit_long$fit_health$scaled_gradient,
    4
  )
)

core_components <- c(
  "optimizer_convergence",
  "max_gradient",
  "sdreport",
  "pd_hessian"
)

health[
  health$component %in% core_components,
  c("component", "status", "message")
]

stopifnot(
  isTRUE(fit_long$fit_health$converged),
  all(health$status[health$component %in% core_components] == "PASS")
)

health[
  health$status != "PASS",
  c("component", "status", "value", "threshold", "message", "action")
]
```

The rendered BFGS fixture passes the optimiser and raw-gradient checks, so its
point covariance is numerically usable. The `sdreport()` and Hessian rows also
pass, which supports local Wald calculations. Keep those decisions separate:
an optimiser failure or raw maximum gradient above `0.01` challenges the point
surface, whereas an `sdreport()` or Hessian warning blocks or weakens Wald
inference without automatically invalidating a stable rotation-invariant point
covariance. The objective-scaled gradient is descriptive and cannot override a
failed optimiser or large raw gradient. Matching long/wide likelihoods checks
data-shape translation only; it does not establish interval coverage.

The rotation warning concerns the raw between-individual loading matrix. It
does not invalidate the rotation-invariant covariance \(\Sigma_B\), but it
does mean that unrotated labels such as “LV1 = boldness” are unsafe.

## Recover both covariance tiers

```{r covariance-recovery}
Sigma_B_hat <- extract_Sigma(
  fit_long,
  level = "unit",
  part = "total"
)$Sigma
Sigma_W_hat <- extract_Sigma(
  fit_long,
  level = "unit_obs",
  part = "total"
)$Sigma

relative_error <- c(
  between = norm(Sigma_B_hat - Sigma_B_true, "F") /
    norm(Sigma_B_true, "F"),
  within = norm(Sigma_W_hat - Sigma_W_true, "F") /
    norm(Sigma_W_true, "F")
)
round(relative_error, 3)
```

These are descriptive errors for one simulated dataset. They check that the
worked example recovers its two scientific targets; they are not a
repeated-sampling validation study.

```{r covariance-figure, fig.class="wide-scientific-figure", fig.cap = "Fitted versus true trait correlations at the between-individual and within-individual tiers. The diagonal line marks exact recovery; points are estimates, not confidence intervals.", fig.alt = "Two scatterplot panels compare fitted with true correlations among six behaviours. One panel shows stable between-individual correlations and the other shows within-individual occasion correlations. A diagonal line in each panel marks exact recovery."}
between_comparison <- compare_Sigma_table(
  fit_long,
  truth = Sigma_B_true,
  level = "unit",
  measure = "correlation",
  entries = "upper"
)
between_comparison$comparison <- "Between individuals"

within_comparison <- compare_Sigma_table(
  fit_long,
  truth = Sigma_W_true,
  level = "unit_obs",
  measure = "correlation",
  entries = "upper"
)
within_comparison$comparison <- "Within individuals"

covariance_comparison <- rbind(
  between_comparison,
  within_comparison
)

plot_Sigma_comparison(
  covariance_comparison,
  measure = "correlation",
  facet = "comparison",
  style = "scatter"
)
```

The plotted targets are the total covariance matrices. The simulation also has
shared and diagonal pieces, but this single fit does not justify a biological
interpretation of their exact split. In real data, interpret total
between/within covariance first and treat loading or communality stories as
rank- and rotation-dependent.

## Recover repeatability

`extract_repeatability()` uses the Wald route by default. We display only its
point estimate because this article does not establish interval calibration.
A request for `method = "profile"` stops rather than silently substituting a
Wald interval; use `method = "bootstrap"` only when its refit-based limitations
are appropriate for the analysis.

```{r repeatability}
repeatability <- extract_repeatability(fit_long)

repeatability_comparison <- data.frame(
  trait = repeatability$trait,
  truth = unname(R_true[repeatability$trait]),
  estimate = repeatability$R
)
repeatability_comparison$absolute_error <- abs(
  repeatability_comparison$estimate - repeatability_comparison$truth
)
repeatability_comparison

cat(
  "Mean absolute repeatability error:",
  round(mean(repeatability_comparison$absolute_error), 3),
  "\n"
)
```

A larger \(R_t\) means that more of the fitted variance for trait \(t\) lies
between individuals rather than among repeated occasions. It does not show
that the trait is genetic, invariant across environments, or transferable to
an unmeasured context.

## What the example supports

This article supports a narrow conclusion: for a complete Gaussian
repeated-measures design with supplied ranks, gllvmTMB can separate and report
total between-individual covariance, total within-individual covariance, and
point repeatability. The example does not:

- choose the ranks;
- separate short-term biology from measurement error inside \(\Sigma_W\);
- give rotation-invariant biological names to latent axes;
- calibrate the available Wald or bootstrap intervals; or
- estimate context-dependent slopes.

Move to a random-regression model only when the data contain a measured
environmental gradient and the question concerns individual-specific slopes.
For difficult fits, use [Can I trust this fit?](fit-diagnostics.html) and
[Convergence and start values](convergence-start-values.html).

## References

Bell, A. M., Hankison, S. J., & Laskowski, K. L. (2009). The
repeatability of behaviour: a meta-analysis. *Animal Behaviour*, 77, 771–783.
<https://doi.org/10.1016/j.anbehav.2008.12.022>

Nakagawa, S., & Schielzeth, H. (2010). Repeatability for Gaussian and
non-Gaussian data: a practical guide for biologists. *Biological Reviews*, 85,
935–956. <https://doi.org/10.1111/j.1469-185X.2010.00141.x>

Réale, D., Reader, S. M., Sol, D., McDougall, P. T., & Dingemanse, N. J.
(2007). Integrating animal temperament within ecology and evolution.
*Biological Reviews*, 82, 291–318.
<https://doi.org/10.1111/j.1469-185X.2007.00010.x>

Westneat, D. F., Wright, J., & Dingemanse, N. J. (2015). The biology hidden
inside residual within-individual phenotypic variation. *Biological Reviews*,
90, 729–743. <https://doi.org/10.1111/brv.12131>
