---
title: "How to choose your model: a decision tree"
tier: 3 # under-audit model-choice draft; visible as a Developer Note
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{How to choose your model: a decision tree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

> **Developer Note -- under audit.** This page is buildable and linked
> from Developer Notes so maintainers can audit the model-choice path.
> It is not yet the package's beginner model guide. If you are new to
> `gllvmTMB`, start with [Get started](gllvmTMB.html),
> [First Gaussian morphology model](morphometrics.html), and
> [Formula keyword grid](api-keyword-grid.html) before using this
> decision tree.
>
> **Current status -- validated rungs are runnable here; two
> capabilities are scoped honestly.**
>
> This page is a Tier 3 routing draft while several linked examples remain
> internal. The runnable chunks below exercise named formula rungs on small
> fixtures, but the page is not yet a public decision guide. Two things on this
> page are deliberately bounded by the
> [capability status](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/61-capability-status.md):
>
> - **Random-regression slopes.** Section 7e showcases phylogenetic
>   random slopes (`phylo_slope`, Gaussian and core families) with
>   recovery evidence; the pedigree parallel `animal_slope` is now
>   recovery-validated too (register `ANI-06`, byte-equivalent to
>   `phylo_slope(vcv = A)` to 1e-5). The structured slope modes
>   (`phylo_indep/dep/latent` and their spatial analogues) all ship
>   with built engines and Gaussian recovery; the diagonal
>   (`phylo_indep`) and reduced-rank (`phylo_latent`, `spatial_latent`)
>   modes additionally cover core non-Gaussian families, while the
>   full-unstructured `_dep` modes stay **Gaussian-only** with
>   non-Gaussian reserved (fail-loud) pending identifiability work.
>   Section 5 gives the per-mode status.
> - **Confidence intervals.** Section 8 explains the Wald / profile /
>   bootstrap choice. The *machinery* runs for all three, but
>   **calibrated coverage is an open gate** outside Gaussian
>   single-tier fits (`CI-08`, `CI-10`); the section says so plainly
>   rather than implying the intervals are validated everywhere.

`gllvmTMB` covers a large model space — one, two, or three nested
observational levels, optional phylogenetic structure, optional
spatial structure, optional relatedness-structured random slopes
(reaction norms), Gaussian / binary / Poisson / mixed responses,
reduced-rank or full-rank trait covariance. That generality is
helpful when your data are unusual but unhelpful when you are
picking a formula on day 1.

This draft is the decision tree we want: six questions
about your data, each answered with a small piece of the formula,
that you assemble at the end into the model that matches your
problem. A seventh question — *how should I report uncertainty?* —
is answered in section 8. If you already know the rung you want,
jump to the linked article for that rung, but note that several linked
worked examples are intentionally internal until their rendered/browser and
evidence gates close.

## How to use this article

Walk down the six questions in order. At each step you pick one
option, which gives you a piece of the formula or the call. By
the end of section 6 you have all the pieces; section 7 shows the
canonical assembled formulas for the most common combinations, and
section 8 covers confidence intervals.

The six questions:

1. What does each row represent? (one, two, or three nested levels)
2. What is the response distribution?
3. Are rows phylogenetically related?
4. Are rows spatially organised?
5. Do slopes vary by relatedness (reaction norms)?
6. Do you want a reduced-rank story for trait covariance?

### The complexity ladder, at a glance

Each rung adds one structural piece to the previous. Pick the
highest rung you actually need — extra structure costs
identifiability and fit time.

| Rung | Pieces |
|---|---|
| 0 | Individual × trait — `latent(... \| unit)` |
| 1 | Site × species (JSDM) — `+ family = binomial()` |
| 2 | + phylogeny — `+ phylo_latent(...)` plus ordinary `latent()` when a non-phylogenetic species tier is needed |
| 3 | + spatial — `+ spatial_indep(0 + trait \| coords)` |
| 4 | + relatedness-structured random slope (reaction norm) — `+ phylo_slope(x \| species)` |
| 5 | Two-level (between / within) — `+ latent(... \| unit_obs)` |
| 6 | Three-level (cluster > unit > unit_obs) — `+ indep(0 + trait \| cluster)` |
| 7 | Capstone — phylogeny + space + tiers together |

Below the table, sections 1–6 ask one question per rung and let
you assemble the formula piece by piece. Section 7 shows the
canonical assembled formulas for the most common combinations, and
section 8 covers interval estimation.

## 1. What does each row represent?

`gllvmTMB` accepts data in two shapes through one entry point.
Long-format data have one row per `(unit, trait)` observation
(`gllvmTMB(value ~ ..., data = df_long, trait = "trait", unit = "...")`);
wide-
format data have one row per unit and one column per trait,
marked by `traits(...)` on the formula left-hand side
(`gllvmTMB(traits(t1, t2, t3) ~ ..., data = df_wide,
unit = "...")`). Internally the engine pivots to long; the two
shapes produce byte-identical fits. The first decision is
whether your data have **one** observational level or **two**.

| Your data look like... | Observational levels | Random-effect grouping | Example article |
|---|---|---|---|
| One row per individual / site, with several traits as columns. Use the wide path directly, or reshape to long. | One: `unit` | <code>latent(0 + trait &#124; unit)</code> (long) / <code>latent(1 &#124; unit, d = K)</code> (wide via `traits()`) | [Morphometrics](morphometrics.html) (individuals × continuous traits) |
| Sites that each contain multiple species, with traits measured for each (site, species) combination; or repeated measures per individual; or papers with multiple items each. | Two: `unit` (between) and `unit_obs` (within) | <code>latent(0 + trait &#124; unit)</code> *and* <code>latent(0 + trait &#124; unit_obs)</code> when both tiers need a shared-plus-diagonal decomposition | [Behavioural syndromes](behavioural-syndromes.html) and [Functional biogeography](functional-biogeography.html) |
| The within-unit rows themselves nest inside a higher grouping — e.g. observations within individuals within populations, or items within papers within research groups. | Three: `cluster` > `unit` > `unit_obs` | add <code>indep(0 + trait &#124; cluster)</code> for a standalone diagonal cluster tier | §7f (worked example below) |
| Sites that each contain multiple species, with traits measured for each (site, species) combination; or repeated measures per individual; or papers with multiple items each. | Two: `unit` (between) and `unit_obs` (within) | <code>latent(0 + trait &#124; unit)</code> *and* <code>indep(0 + trait &#124; unit_obs)</code> | [Behavioural syndromes](behavioural-syndromes.html) and [Functional biogeography](functional-biogeography.html) |
| The within-unit rows themselves nest inside a higher grouping — e.g. observations within individuals within populations, or items within papers within research groups. | Three: `cluster` > `unit` > `unit_obs` | add <code>indep(0 + trait &#124; cluster)</code> on top of the two-level pair | §7f (worked example below) |

If you have a single observational level, the `unit` column can be
called anything (`individual`, `site`, `study`, `paper`); pass its
name to `gllvmTMB(..., unit = "your_column")`. The two-level case
needs a between-unit column and a within-unit column; pass them as
`unit = "..."` and `unit_obs = "..."` (the defaults `site` /
`site_species` are filled in by `simulate_site_trait()`). The
three-level case adds a `cluster = "..."` column above the unit —
each tier contributes its own per-trait variance (`sd_q` for the
cluster tier; see §7f), which is the natural layout for nested
ecological or bibliometric data.

This article writes the worked examples in §7 in long format because
the long form makes the multi-component formulas easier to read.
Each example has an equivalent wide form via `traits(...)`; the
[Get Started vignette](../gllvmTMB.html) shows the long/wide
equivalence check (`all.equal(logLik(fit), logLik(fit_wide))`).

## 2. What is the response distribution?

Latent structure is shared across response distributions, but the
likelihood term per row is family-specific. Pick the family that
matches your trait scale.

| Trait values look like... | `family` argument | Notes |
|---|---|---|
| Continuous, real-valued (mass, length, log-abundance) | `family = gaussian()` | Identity link. Default. |
| 0/1 presence-absence, yes-no, item-passed | `family = binomial()` | Logit link. Used for joint species distribution models and binary-item systematic mapping; see [Joint SDM](joint-sdm.html). |
| Counts (offspring, observations) | `family = poisson()` | Log link. |
| Different traits use different families | `family = list(gaussian(), binomial(), poisson())` plus a `family_var` column | One family per row. See [Response families](response-families.html). |

For the mixed-response case the `family` list and the `family_var`
attribute look like this:

```r
fam <- list(gaussian(), binomial(), poisson())
attr(fam, "family_var") <- "family"   # column on `data` selecting the row's family
```

If you are unsure, start with the family that matches your most
abundant trait type and read the response-family reference before
committing to a list.

## 3. Are rows phylogenetically related?

A phylogenetic random effect is appropriate when species (or other
taxa) appear repeatedly in your data and closely related taxa are
likely to share trait values for evolutionary reasons. The
canonical phylogenetic decomposition pairs four components: a
phylogenetic shared low-rank axis and a phylogenetic trait-specific
diagonal (both carried by `phylo_latent(..., unique = TRUE)`), a
non-phylogenetic shared axis (`latent`), and the ordinary diagonal
`Psi` companion included by default in `latent()`. The full
decomposition reads

The
folded phylogenetic decomposition pairs two species-level covariance
matrices: a phylogenetic shared-plus-diagonal tier (`phylo_latent(..., unique = TRUE)`)
and, when needed, a non-phylogenetic shared-plus-diagonal species tier
(`latent`). The full decomposition reads

$$
\Omega = (\Sigma_{\text{phy}} \otimes A) + (\Sigma_{\text{non}} \otimes I), \quad
\Sigma_{\text{phy}} = \Lambda_{\text{phy}} \Lambda_{\text{phy}}^{\top} + \boldsymbol\Psi_{\text{phy}}, \quad
\Sigma_{\text{non}} = \Lambda_{\text{non}} \Lambda_{\text{non}}^{\top} + \boldsymbol\Psi_{\text{non}}.
$$

Without all four components the decomposition is degenerate:
`extract_phylo_signal()` requires both `latent()` and `phylo_*()`
to return a meaningful three-way split. See
[Phylogenetic GLLVM](phylogenetic-gllvm.html) for the worked
example.

| Your situation | Add to formula (in-keyword `tree =` or `vcv =`) |
|---|---|
| Rows are not phylogenetically structured (e.g. individuals from one species, or sites without taxonomic identity). | Nothing. |
| Phylogenetic structure with one shared phylogenetic variance per trait (simplest phylo model). | `+ phylo_scalar(species, vcv = Cphy)` |
| Folded phylogenetic decomposition (when traits can be both phylogenetically and non-phylogenetically correlated). | `+ phylo_latent(species, d = K_phy, tree = tree, unique = TRUE) + latent(0 + trait \| species, d = K_non)` |
| Many species (n > ~200) and you want speed. | (same folded form) with `tree = tree` (sparse $A^{-1}$ trick); much faster than `vcv = Cphy` at this scale. |

The crossover where the sparse $A^{-1}$ path overtakes the dense
path sits around `n_species ≈ 150-200`. Above that, the speedup is
roughly an order of magnitude. The
[Phylogenetic GLLVM article](phylogenetic-gllvm.html) shows the
benchmark, the worked four-component recovery, and the
`extract_communality()` / `extract_phylo_signal()` outputs.

### 3a. Are individuals related by a pedigree (animal model)?

If your rows are **individuals** rather than species, and you
have a **pedigree** (a three-column id/sire/dam table) instead of
a phylogeny, swap `phylo_*()` for the parallel `animal_*()`
family. Mathematically identical to the phylogenetic case — both
use a known relatedness matrix $\mathbf A$ — only the source of
$\mathbf A$ differs (pedigree → Henderson's recursive formula vs
tree → branch-length VCV). The `animal_*` keywords accept three
input forms:

```r
# 1. Three-column pedigree (MCMCglmm-style; columns id/sire/dam
#    or animal/dam/sire — resolved by name with positional fallback):
+ animal_latent(individual, d = K, pedigree = ped, unique = TRUE)

# 2. Pre-computed A matrix:
+ animal_latent(individual, d = K, A = A_matrix, unique = TRUE)

# 3. Sparse precision Ainv (densified for v0.2.0):
+ animal_latent(individual, d = K, Ainv = Ainv_matrix, unique = TRUE)
```

The full 4-keyword grid (`animal_scalar`, `animal_indep`, `animal_dep`,
`animal_latent`) plus the random-
regression `animal_slope(x | individual)` mirror their `phylo_*`
siblings 1-to-1. Use the animal family for quantitative-genetics
animal models: variance-component decompositions, factor-analytic
G-matrix estimation, reaction-norm (random regression) models.
See [Design 14](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/14-known-relatedness-keywords.md)
for the math + the **A-vs-V naming boundary** (A for relatedness;
V is reserved for `meta_V()` sampling variance — do not blur).
Worked example in [The animal model: heritability and genetic
covariance](animal-model.html).

## 4. Are rows spatially organised?

If your sites carry coordinates and you suspect residual spatial
autocorrelation, add a sparse SPDE/GMRF random field. Each trait
gets its own field by default.

| Your situation | Add to formula | Argument to `gllvmTMB()` |
|---|---|---|
| Sites have no spatial structure (or you do not have coordinates). | Nothing. | — |
| Sites have coordinates and each trait should have its own spatial pattern. | <code>+ spatial_indep(0 + trait &#124; coords)</code> | `mesh = make_mesh(df, c("lon", "lat"), cutoff = ...)` |
| Sites have coordinates and you want one shared spatial-variance scale across traits (parsimony). | <code>+ spatial_scalar(0 + trait &#124; coords)</code> | `mesh = make_mesh(df, c("lon", "lat"), cutoff = ...)` |
| You want traits to share a low-rank spatial structure. | <code>+ spatial_latent(0 + trait &#124; coords, d = K)</code> — K shared SPDE fields driven by a T × K loading matrix Λ_spa. | `mesh = make_mesh(df, c("lon", "lat"), cutoff = ...)` (the spatial analogue of `phylo_latent()`). |

The SPDE engine wins on speed when the dataset has more than
roughly 100 sites; below that, the dense `glmmTMB::exp()` covariance
is competitive.

## 5. Do slopes vary by relatedness (reaction norms)?

So far every structural term has been an *intercept* random effect:
each group (species, individual, site) gets its own shift in the
trait mean. A **random-regression** or **reaction-norm** model goes
one step further and lets each group have its own *slope* on a
covariate `x` — how strongly trait values respond to an environmental
gradient, a dose, a time index. When those per-group slopes are
themselves structured by **relatedness** (a phylogeny or a pedigree),
closely related groups are expected to share similar slopes, and the
slope variance is the quantity you want to estimate (Meyer & Hill
1997; de Villemereuil 2018).

| Your situation | Add to formula | Argument to `gllvmTMB()` |
|---|---|---|
| Slopes do not vary by group, or you have no covariate of interest. | Nothing — keep the intercept-only structure from §1–§4. | — |
| **Per-species slope on `x`, structured by the phylogeny** (one shared slope variance, slopes correlated through the tree). | <code>+ phylo_slope(x &#124; species, tree = tree)</code> | `unit = "species"` |
| Per-individual slope on `x`, structured by a **pedigree** (the quantitative-genetic heritable reaction norm). | <code>+ animal_slope(x &#124; id)</code> with a relatedness matrix `A` | `phylo_vcv = A`, `unit = "id"` |

`phylo_slope(x | species)` is the **validated** entry point: it fits
one phylogenetically-structured slope variance and per-species
slopes drawn from $\mathcal N(\mathbf 0,\, \sigma^2_{\text{slope}}\,
\mathbf C_{\text{phy}})$. It composes with the intercept terms from
§3 — you can have phylogenetic random *intercepts*
(`phylo_latent(..., unique = TRUE)`) and a phylogenetic
random *slope* in the same fit. The
worked recovery is in §7e below. `animal_slope(x | id)` is the
pedigree parallel (same engine, relatedness matrix $\mathbf A$ from
the pedigree instead of the tree).

> **Scope (per-mode status).** The slope engine has filled in since
> earlier drafts; the honest bounds now sit at non-Gaussian coverage
> for the saturated modes, not at "implemented vs not". Per the
> [capability status](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/61-capability-status.md)
> and validation-debt register:
>
> - **Correlated intercept–slope reaction norms.** The folded
>   `phylo_latent(1 + x | species, unique = TRUE)` estimates the
>   phylogenetic reaction-norm covariance `Sigma = Lambda Lambda^T + Psi`
>   over the (intercept, slope) coefficients — including the
>   intercept–slope correlation — and surfaces it via
>   `extract_Sigma(fit, level = "phy")` (register `ANI-11`, Gaussian).
>   The pedigree twin `animal_latent(1 + x | id, unique = TRUE)` is the
>   same engine. (§7e shows it run.)
> - **Structured slope modes** `phylo_indep` / `phylo_dep` /
>   `phylo_latent` (and their `spatial_*` analogues) all ship with
>   built engines and Gaussian recovery. The diagonal `phylo_indep`
>   slope is covered across the core families; the reduced-rank
>   `phylo_latent` and `spatial_latent` slopes cover Gaussian plus
>   several non-Gaussian families. The full-unstructured `phylo_dep` /
>   `spatial_dep` slopes are **Gaussian-only**, with non-Gaussian
>   reserved (fail-loud) until the unstructured cross-(intercept,slope)
>   covariance is shown identifiable (`PHY-18`, `SPA-10`).
> - **`animal_slope` recovery** is **validated** (register `ANI-06`):
>   byte-equivalent to `phylo_slope(vcv = A)` to 1e-5 over the recovery
>   grid. Treat its Gaussian point estimates as covered, not
>   provisional.
>
> Two-part (delta / hurdle) response families — `delta_gamma()`,
> `delta_lognormal()` — take **no** random slopes (or any latent /
> tier structure) by design: with two link scales there is no single
> scale on which a species×species correlation is defined (see
> [Design 62](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/62-two-part-family-naming-and-scope.md)).
> They are fixed-effect response distributions only.

## 6. Do you have a reduced-rank story for trait covariance?

Even with no phylogenetic or spatial structure, you have to decide
how the trait covariance itself is parameterised. This is the
choice between *factor analysis* (a few latent dimensions
explaining most cross-trait covariance) and *unstructured
covariance* (every pair gets its own parameter).

| Your situation | Add to formula | Notes |
|---|---|---|
| You expect a small number of latent dimensions (size, shape, life-history pace) to explain most of the cross-trait covariance. T traits, K << T factors. | <code>+ latent(0 + trait &#124; site, d = K)</code> | The default GLLVM choice. K = 1–3 is typical. |
| You want full unstructured between-unit covariance with no rank reduction. | <code>+ dep(0 + trait &#124; site)</code> | Reasonable when T is small (T ≤ 5) and you have many units. |
| You want only a standalone trait-specific diagonal tier. | <code>+ indep(0 + trait &#124; site)</code> | Use this when the model should have no off-diagonal covariance at that tier. |
| Confirmatory: you know which traits load on which factor and want to pin those entries of $\boldsymbol\Lambda$. | <code>+ latent(0 + trait &#124; site, d = K)</code> plus `lambda_constraint = list(unit = M)` | M is a `T × K` matrix with `NA` for free entries and numeric values for fixed ones. See [Lambda constraint](lambda-constraint.html); `?suggest_lambda_constraint` proposes a sensible matrix from a fitted unconstrained model. |

For two-level decomposed models, put `latent()` at each tier that needs
shared axes plus a diagonal `Psi` companion. The full between-unit
piece is now `latent(0 + trait | site, d = K)`, and the within-unit
piece substitutes the within-unit grouping column (e.g. `site_species`
for site × species data, or the value passed via `unit_obs = "..."`
for other domains). Use `latent(..., unique = FALSE)` only when you
want the no-Psi subset, and use `indep()` for standalone diagonal tiers.

| You want full unstructured between-unit covariance with no rank reduction. | <code>+ indep(0 + trait &#124; site)</code> (specific variance) plus a separate term for off-diagonal structure as needed | Reasonable when T is small (T ≤ 5) and you have many units. |
| Confirmatory: you know which traits load on which factor and want to pin those entries of $\boldsymbol\Lambda$. | <code>+ latent(0 + trait &#124; site, d = K)</code> plus `lambda_constraint = list(unit = M)` | M is a `T × K` matrix with `NA` for free entries and numeric values for fixed ones. See [Lambda constraint](lambda-constraint.html); `?suggest_lambda_constraint` proposes a sensible matrix from a fitted unconstrained model. |

For two-level models you typically include ordinary `latent()`
at each level, so the full between-unit piece is
`latent(0 + trait | site, d = K)` and the
within-unit piece substitutes the within-unit grouping column
(e.g. `site_species` for site × species data, or the value passed via
`unit_obs = "..."` for other domains).
Ordinary `latent()` already carries the per-trait residual diagonal
$\boldsymbol\Psi$ by default ($\Sigma = \Lambda\Lambda^{\top} +
\boldsymbol\Psi$), so each level's `latent()` supplies both the shared
factors and each trait's own residual variance; use
`latent(..., unique = FALSE)` for the loadings-only subset. A
standalone `indep(0 + trait | cluster)` adds a further diagonal-only
tier when the data nest one level deeper.

## 7. Final formula assembly

Below are the canonical formulas for the most common combinations.
The chunks below run live to confirm the formulas parse and recover
their simulated truth; each fits in a few seconds on the small
simulations.

```{r setup}
library(gllvmTMB)
library(ape)
```

### 7a. Individuals × continuous traits, no extra structure

The simplest case (rung 0): one level, Gaussian, no phylogeny, no
space. See [Morphometrics](morphometrics.html) for the full
walkthrough.

```{r ex-1, results = "hide"}
set.seed(1)
n_ind <- 30; n_traits <- 4
df_ind <- data.frame(
  individual = factor(rep(seq_len(n_ind), each = n_traits)),
  trait      = factor(rep(paste0("trait_", seq_len(n_traits)), n_ind),
                      levels = paste0("trait_", seq_len(n_traits))),
  value      = rnorm(n_ind * n_traits)
)
fit_ind <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | individual, d = 2),
  data = df_ind,
  trait = "trait",
  unit = "individual"
)
```

### 7b. Sites × species, presence/absence

Rung 1 with a binomial response: a joint species distribution
model. See [Joint SDM](joint-sdm.html).

```{r ex-2, results = "hide"}
set.seed(2)
sim_b <- simulate_site_trait(
  n_sites = 30, n_species = 6, n_traits = 4,
  mean_species_per_site = 4,
  Lambda_B = matrix(rnorm(8, sd = 0.6), 4, 2),
  psi_B      = rep(0.3, 4), seed = 2
)
df_b <- sim_b$data
df_b$value <- as.integer(df_b$value > 0)

fit_jsdm <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | site, d = 2),
  data   = df_b,
  trait  = "trait",
  family = binomial()
)
```

### 7c. Add phylogeny

Rung 2: the canonical folded phylogenetic decomposition combines
`phylo_latent(unique = TRUE)` (the phy side) with
ordinary `latent()` (the non-phy side, including its default diagonal
`Psi`). With a tree in hand, prefer
`tree = tree` (in-keyword) for the sparse $A^{-1}$ path. See
[Phylogenetic GLLVM](phylogenetic-gllvm.html) for the worked
example with simulation recovery and `extract_phylo_signal()`.

Rung 2: the canonical paired phylogenetic decomposition combines
folded `phylo_latent()` (the phy side) with
ordinary `latent()` (the non-phy side). With a tree in hand, prefer
`tree = tree` (in-keyword) for the sparse $A^{-1}$ path. See
[Phylogenetic GLLVM](phylogenetic-gllvm.html) for the worked
example with simulation recovery and `extract_phylo_signal()`.

```{r ex-3, results = "hide"}
set.seed(3)
n_sp <- 8
tree <- ape::rcoal(n_sp); tree$tip.label <- paste0("sp", seq_len(n_sp))

sim_p <- simulate_site_trait(
  n_sites = 30, n_species = n_sp, n_traits = 4,
  mean_species_per_site = 5,
  Lambda_B = matrix(rnorm(8, sd = 0.5), 4, 2),
  psi_B = rep(0.3, 4),
  Cphy = ape::vcv(tree, corr = TRUE), sigma2_phy = rep(0.4, 4),
  seed = 3
)
df_p <- sim_p$data
levels(df_p$species) <- tree$tip.label

fit_phy <- gllvmTMB(
  value ~ 0 + trait +
          phylo_latent(species, d = 1, tree = tree, unique = TRUE) +
          latent(0 + trait | species, d = 1),
  data = df_p,
  trait = "trait",
  unit = "species"
)
```

### 7d. Add space

Rung 3: append `spatial_indep(0 + trait | coords)` and pass a mesh. The
simulator places sites in `[0,1]^2` automatically when
`spatial_range` is supplied. The SPDE engine wins on speed when
the dataset has more than roughly 100 sites; below that the dense
`glmmTMB::exp()` covariance is competitive.

```{r ex-4, results = "hide"}
set.seed(4)
sim_s <- simulate_site_trait(
  n_sites = 40, n_species = 1, n_traits = 3,
  mean_species_per_site = 1,
  Lambda_B = matrix(0.0, 3, 1), psi_B = rep(0.1, 3),
  spatial_range = 0.3, sigma2_spa = rep(0.4, 3),
  seed = 4
)
df_s <- sim_s$data
mesh <- make_mesh(df_s, c("lon", "lat"), cutoff = 0.1)

fit_spa <- gllvmTMB(
  value ~ 0 + trait + spatial_indep(0 + trait | coords, mesh = mesh),
  data = df_s,
  trait = "trait"
)
```

### 7e. Add a phylogenetic random slope (reaction norm)

Rung 4: append `phylo_slope(x | species, tree = tree)` to give each
species its own slope on the covariate `x`, with the per-species
slopes correlated through the phylogeny. Here we simulate true
per-species slopes from $\mathcal N(\mathbf 0,\, \sigma^2_{\text{slope}}\,
\mathbf C_{\text{phy}})$ with $\sigma^2_{\text{slope}} = 0.4$ and check
that the fit recovers both the slope SD and the per-species slopes.

```{r ex-slope, results = "hide"}
set.seed(6)
n_sp  <- 25L
n_trt <- 3L
n_rep <- 4L
tree  <- ape::rcoal(n_sp)
tree$tip.label <- paste0("sp", seq_len(n_sp))
Cphy  <- ape::vcv(tree, corr = TRUE)

## True per-species slopes ~ N(0, sigma2_slope * Cphy)
sigma2_slope <- 0.4
L_chol       <- t(chol(Cphy + diag(1e-8, n_sp)))
slope_true   <- as.numeric(L_chol %*% rnorm(n_sp, sd = sqrt(sigma2_slope)))
names(slope_true) <- tree$tip.label

df_rr <- expand.grid(
  species = factor(tree$tip.label, levels = tree$tip.label),
  trait   = factor(paste0("t", seq_len(n_trt)), levels = paste0("t", seq_len(n_trt))),
  rep     = seq_len(n_rep)
)
df_rr$x     <- rnorm(nrow(df_rr))
mu_trait    <- c(2, 1.5, 1)[as.integer(df_rr$trait)]
df_rr$value <- mu_trait +
  slope_true[as.character(df_rr$species)] * df_rr$x +
  rnorm(nrow(df_rr), sd = 0.5)

fit_rr <- gllvmTMB(
  value ~ 0 + trait + phylo_slope(x | species, tree = tree),
  data = df_rr,
  unit = "species"
)
```

The slope SD lives in `fit$report$sigma_slope`; the per-species
slopes are the `b_phy_slope` random effects, indexed against the
augmented relatedness matrix:

```{r slope-recovery}
sigma_slope_hat <- as.numeric(fit_rr$report$sigma_slope)

b_best    <- fit_rr$tmb_obj$env$last.par.best
b_slope   <- as.numeric(b_best[names(b_best) == "b_phy_slope"])
aug_names <- rownames(fit_rr$tmb_data$Ainv_phy_rr)
b_tips    <- b_slope[match(tree$tip.label, aug_names)]

data.frame(
  quantity = c("slope SD", "per-species slope recovery (cor)"),
  estimate = signif(c(sigma_slope_hat, cor(b_tips, slope_true)), 3),
  truth    = c(round(sqrt(sigma2_slope), 3), NA)
)
```

The recovered slope SD sits close to the simulated $\sqrt{0.4}
\approx 0.63$, and the fitted per-species slopes correlate strongly
with the truth. For a **pedigree** instead of a tree, swap in
`animal_slope(x | id)` and pass the relatedness matrix as
`phylo_vcv = A`; that pedigree path is recovery-validated (register
`ANI-06`, byte-equivalent to `phylo_slope(vcv = A)` to 1e-5).

The richer correlated intercept–slope form fits and surfaces its 2×2
(intercept, slope) covariance through `extract_Sigma(fit, level =
"phy")` (register `ANI-11`, Gaussian). It is shown here **not run**
only to keep this page to a single worked fit:

```{r ex-slope-aug, eval = FALSE}
# Folded phylogenetic reaction-norm covariance: phylo_latent(1 + x | species,
# unique = TRUE) estimates Sigma = Lambda Lambda^T + Psi over the
# (intercept, slope) coefficients; read it back via extract_Sigma(level = "phy").
# animal_latent(1 + x | id, unique = TRUE) is the pedigree twin.
fit_rr_aug <- gllvmTMB(
  value ~ 0 + trait +
    phylo_latent(0 + trait + (0 + trait):x | species, d = 1, unique = TRUE),
  data = df_rr, phylo_tree = tree, unit = "species"
)
extract_Sigma(fit_rr_aug, level = "phy")   # phylo reaction-norm Sigma
```

### 7f. Three-level cluster tier

Rung 6: when the within-unit rows nest inside a higher grouping,
add a third per-trait diagonal tier `indep(0 + trait | cluster)`
and pass all three grouping keys. Each tier gets its own per-trait
variance; the cluster tier's is reported as `fit$report$sd_q`.

```{r ex-cluster, results = "hide"}
set.seed(11)
n_cluster  <- 25L
n_unit_per <- 4L
n_obs_per  <- 3L
trait_lv   <- c("a", "b", "c")
sd_cluster_true <- c(0.5, 0.7, 0.4)

rows <- expand.grid(
  obs       = seq_len(n_obs_per),
  unit_j    = seq_len(n_unit_per),
  clus      = seq_len(n_cluster),
  trait_idx = seq_along(trait_lv)
)
rows$cluster  <- factor(rows$clus)
rows$unit     <- factor(paste(rows$clus, rows$unit_j, sep = "_"))
rows$unit_obs <- factor(paste(rows$clus, rows$unit_j, rows$obs, sep = "_"))
rows$trait    <- factor(trait_lv[rows$trait_idx], levels = trait_lv)

q_cluster  <- sapply(seq_along(trait_lv),
                     function(t) rnorm(n_cluster, sd = sd_cluster_true[t]))
eta_c      <- q_cluster[cbind(as.integer(rows$cluster), rows$trait_idx)]
rows$value <- eta_c + rnorm(nrow(rows), sd = 0.3)

fit_cluster <- gllvmTMB(
  value ~ 0 + trait + indep(0 + trait | cluster),
  data     = rows,
  unit     = "unit",
  unit_obs = "unit_obs",
  cluster  = "cluster"
)
```

```{r cluster-recovery}
data.frame(
  trait   = trait_lv,
  sd_true = sd_cluster_true,
  sd_est  = signif(as.numeric(fit_cluster$report$sd_q), 3)
)
```

The cluster-tier SDs recover their simulated values. The same
keyword works across the core families — pass `family = poisson()`,
`Beta()`, `Gamma(link = "log")`, etc., and add the matching
within-tier terms as needed.

### 7g. Two-level + phylo + space (capstone)

Rung 7: the full Nakagawa et al. functional-biogeography model — two
levels, phylogeny, and space all at once. The formula reads as a sum
of the pieces from sections 1–6. The runnable end-to-end version
lives in [Functional biogeography](functional-biogeography.html);
in this article we just show the formula:

```r
fit_full <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):env_1 + (0 + trait):env_2 +
          spatial_indep(0 + trait | coords, mesh = mesh) +
          latent(0 + trait | site, d = 2) +
          latent(0 + trait | site_species, d = 1) +
          phylo_latent(species, d = 1, tree = tree, unique = TRUE),
  data = df,
  trait = "trait"
)
```

## 8. How should you report uncertainty? (Wald / profile / bootstrap)

Every CI-producing extractor (`extract_repeatability()`,
`extract_communality()`, `extract_correlations()`) and the
`confint()` method accept a `method` argument with three values. The
choice trades speed against how faithfully the interval respects the
skewed, boundary-bounded sampling distributions of variance
components.

| Method | Speed | Behaviour | When to use |
|---|---|---|---|
| `"wald"` | fastest (reads `sd_report` directly) | symmetric on the natural scale; can cross a boundary (e.g. $R < 0$) | a quick check on a well-identified fit |
| `"profile"` | fast (TMB inner optim, warm-started) | respects skewness; one-sided near $\sigma^2 \to 0$ | the default for derived ratio quantities |
| `"bootstrap"` | slowest (refits per replicate) | empirical sampling distribution for the fitted target | non-monotone profiles, joint uncertainty, reportable intervals when the relevant coverage row supports them |

```{r ci-setup, results = "hide"}
s_ci <- simulate_site_trait(
  n_sites = 40, n_species = 6, n_traits = 3, mean_species_per_site = 4,
  Lambda_B = matrix(c(0.9, 0.4, -0.3), 3, 1),
  psi_B = c(0.40, 0.30, 0.50), psi_W = c(0.30, 0.40, 0.30),
  beta = matrix(0, 3, 2), seed = 42
)
fit_ci <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | site, d = 1) +
    indep(0 + trait | site_species),
  data  = s_ci$data,
  trait = "trait"
)
```

```{r ci-methods}
rep_wald <- extract_repeatability(fit_ci, level = 0.95, method = "wald")
rep_boot <- extract_repeatability(fit_ci, level = 0.95, method = "bootstrap",
                                  nsim = 20, seed = 1)
rep_wald
rep_boot
```

The point estimates agree across methods (they share the same MLE);
the **interval widths** differ. The bootstrap interval is the one to
cite when the Wald symmetry assumption is doubtful — near a boundary,
or for a ratio quantity like repeatability that lives in $[0, 1]$.

> **Honest coverage caveat.** The CI *machinery* runs for all three
> methods, but **calibrated coverage has been established only for
> Gaussian single-tier fits at $d = 1$ and $d = 3$.** A production
> coverage study found most other Gaussian cells below the 94% target
> (`CI-08`), and **mixed-family** (e.g. Gaussian + binomial)
> communality/correlation coverage collapses to 0.55–0.82
> (`CI-10`) — a known engine deficiency, not a near-complete feature.
> Treat intervals on mixed-family fits, and on high-dimensional
> Gaussian fits, as **indicative rather than calibrated**; prefer
> bootstrap and report the method you used. See the
> [capability status](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/61-capability-status.md)
> §3 and the
> [profile-likelihood CI article](profile-likelihood-ci.html) for the
> full treatment, including direct access to `TMB::tmbprofile()` and
> `TMB::tmbroot()`.

For a deeper walk through the three-method API — variance components,
correlations, and the boundary behaviour — see
[Profile-likelihood confidence intervals](profile-likelihood-ci.html).

## 9. Diagnostic checklist

After any fit, run through these in order:

| Check | How |
|---|---|
| Optimiser convergence and Hessian PD-ness. | `sanity_multi(fit)` |
| Implied between-unit trait covariance. | `extract_Sigma_table(fit, level = "unit")` (and `extract_Sigma_table(fit, level = "unit_obs")` for two-level fits). |
| Interpretable loadings. | `getLoadings(fit, level = "unit", rotate = "varimax")` |
| Did you fit unconstrained `latent()` with `d > 1`? Loadings are not unique. | Use `rotate = "varimax"` for a quick post-hoc rotation, or call `suggest_lambda_constraint()` to propose a confirmatory `lambda_constraint` matrix and refit. |
| Communalities / ICCs. | `extract_ICC_site(fit)`, `extract_communality(fit, level = "unit")`, `extract_communality(fit, level = "unit_obs")` |

If `sanity_multi()` flags a non-PD Hessian or a near-zero loading
diagonal, the most common causes are an over-rich model
(`d` too large, too many `latent()` terms relative to data) or
factor-level ordering issues in the long-format data. The
[Morphometrics](morphometrics.html) article has a worked example
of the factor-level pitfall.

## See also

* [Morphometrics](morphometrics.html) — rung 0: individuals × continuous traits.
* [Joint SDM](joint-sdm.html) — rung 1: binary multivariate.
* [Behavioural syndromes](behavioural-syndromes.html) — internal two-observational-level worked-example draft.
* [Phylogenetic GLLVM](phylogenetic-gllvm.html) — internal phylogenetic-dependence draft for the canonical paired four-component decomposition.
* [The animal model](animal-model.html) — internal pedigree-structured `animal_*` and heritable reaction-norm draft.
* [Functional biogeography](functional-biogeography.html) — internal capstone draft: two-level + space + phylogeny.
* [Response families](response-families.html) — supported likelihoods and mixed-family dispatch.
* [Lambda constraint](lambda-constraint.html) — confirmatory factor analysis with fixed loadings; see also `?suggest_lambda_constraint`.
* [Profile-likelihood confidence intervals](profile-likelihood-ci.html) — the Wald / profile / bootstrap CI choice (section 8) in depth.
* [Pitfalls](pitfalls.html) — common gotchas in data prep, formula construction, and interpretation.

## References

* Meyer, K. & Hill, W. G. (1997). Estimation of genetic and phenotypic
  covariance functions for longitudinal or 'repeated' records by
  restricted maximum likelihood. *Livestock Production Science* **47**,
  185–200. — random-regression / reaction-norm model.
* de Villemereuil, P. (2018). Quantitative genetic methods depending on
  the nature of the phenotypic trait. *Annals of the New York Academy of
  Sciences* **1422**, 29–47. — reaction norms and relatedness-structured
  slopes.
