---
title: "How to choose your model: a decision tree"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{How to choose your model: a decision tree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

`gllvmTMB` covers a large model space — one or two observational
levels, optional phylogenetic structure, optional spatial structure,
Gaussian / binary / Poisson / mixed responses, reduced-rank or
full-rank trait covariance. That generality is helpful when your
data are unusual but unhelpful when you are picking a formula on
day 1.

This article is the decision tree we wish we had: five questions
about your data, each answered with a small piece of the formula,
that you assemble at the end into the model that matches your
problem. If you already know the rung you want, jump to the linked
article for that rung.

## How to use this article

Walk down the five questions in order. At each step you pick one
option, which gives you a piece of the formula or the call. By
the end of section 5 you have all the pieces; section 6 shows the
canonical assembled formulas for the most common combinations.

The five questions:

1. What does each row represent?
2. What is the response distribution?
3. Are rows phylogenetically related?
4. Are rows spatially organised?
5. Do you want a reduced-rank story for trait covariance?

### The complexity ladder, at a glance

Each rung adds one structural piece to the previous. Pick the
highest rung you actually need — extra structure costs
identifiability and fit time.

| Rung | Pieces |
|---|---|
| 0 | Individual × trait — `latent(... \| unit)` |
| 1 | Site × species (JSDM) — `+ family = binomial()` |
| 2 | + phylogeny — `+ phylo_latent + phylo_unique` (paired with `latent + unique`) |
| 3 | + spatial — `+ spatial_unique(0 + trait \| coords)` |
| 4 | Two-level (between / within) — `+ latent(... \| unit_obs)` |
| 5 | Capstone — all of the above |

Below the table, sections 1–5 ask one question per rung and let
you assemble the formula piece by piece. Section 6 shows the
canonical assembled formulas for the most common combinations.

## 1. What does each row represent?

`gllvmTMB` accepts data in two shapes through one entry point.
Long-format data have one row per `(unit, trait)` observation
(`gllvmTMB(value ~ ..., data = df_long, unit = "...")`); wide-
format data have one row per unit and one column per trait,
marked by `traits(...)` on the formula left-hand side
(`gllvmTMB(traits(t1, t2, t3) ~ ..., data = df_wide,
unit = "...")`). Internally the engine pivots to long; the two
shapes produce byte-identical fits. The first decision is
whether your data have **one** observational level or **two**.

| Your data look like... | Observational levels | Random-effect grouping | Example article |
|---|---|---|---|
| One row per individual / site, with several traits as columns. Use the wide path directly, or reshape to long. | One: `unit` | <code>latent(0 + trait &#124; unit)</code> (long) / <code>latent(1 &#124; unit, d = K)</code> (wide via `traits()`) | [Morphometrics](morphometrics.html) (individuals × continuous traits) |
| Sites that each contain multiple species, with traits measured for each (site, species) combination; or repeated measures per individual; or papers with multiple items each. | Two: `site` (between) and `site_species` (within) | <code>latent(0 + trait &#124; site)</code> *and* <code>latent(0 + trait &#124; site_species)</code> | [Behavioural syndromes](behavioural-syndromes.html) and [Functional biogeography](functional-biogeography.html) |

If you have a single observational level, the `unit` column can be
called anything (`individual`, `site`, `study`, `paper`); pass its
name to `gllvmTMB(..., unit = "your_column")`. The two-level case
needs both `site` and `site_species` columns; the
`simulate_site_trait()` helper shows the canonical layout.

This article writes the examples in §6 in long format because the
long form makes the multi-component formulas easier to read.
Each example has an equivalent wide form via `traits(...)`; the
[Get Started vignette](../gllvmTMB.html) shows the long/wide
equivalence check (`all.equal(logLik(fit), logLik(fit_wide))`).

## 2. What is the response distribution?

Latent structure is shared across response distributions, but the
likelihood term per row is family-specific. Pick the family that
matches your trait scale.

| Trait values look like... | `family` argument | Notes |
|---|---|---|
| Continuous, real-valued (mass, length, log-abundance) | `family = gaussian()` | Identity link. Default. |
| 0/1 presence-absence, yes-no, item-passed | `family = binomial()` | Logit link. Used for joint species distribution models and binary-item systematic mapping; see [Joint SDM](joint-sdm.html). |
| Counts (offspring, observations) | `family = poisson()` | Log link. |
| Different traits use different families | `family = list(gaussian(), binomial(), poisson())` plus a `family_var` column | One family per row. See [Response families](response-families.html). |

For the mixed-response case the `family` list and the `family_var`
attribute look like this:

```r
fam <- list(gaussian(), binomial(), poisson())
attr(fam, "family_var") <- "family"   # column on `data` selecting the row's family
```

If you are unsure, start with the family that matches your most
abundant trait type and read the response-family reference before
committing to a list.

## 3. Are rows phylogenetically related?

A phylogenetic random effect is appropriate when species (or other
taxa) appear repeatedly in your data and closely related taxa are
likely to share trait values for evolutionary reasons. The
canonical phylogenetic decomposition pairs four components: a
phylogenetic shared low-rank axis (`phylo_latent`), a
phylogenetic trait-specific diagonal (`phylo_unique`), a
non-phylogenetic shared axis (`latent`), and a non-phylogenetic
trait-specific diagonal (`unique`). The full decomposition reads

$$
\Omega = (\Sigma_{\text{phy}} \otimes A) + (\Sigma_{\text{non}} \otimes I), \quad
\Sigma_{\text{phy}} = \Lambda_{\text{phy}} \Lambda_{\text{phy}}^{\top} + S_{\text{phy}}, \quad
\Sigma_{\text{non}} = \Lambda_{\text{non}} \Lambda_{\text{non}}^{\top} + S_{\text{non}}.
$$

Without all four components the decomposition is degenerate:
`extract_phylo_signal()` requires both `latent()` and `phylo_*()`
to return a meaningful three-way split. See
[Phylogenetic GLLVM](phylogenetic-gllvm.html) for the worked
example.

| Your situation | Add to formula (in-keyword `tree =` or `vcv =`) |
|---|---|
| Rows are not phylogenetically structured (e.g. individuals from one species, or sites without taxonomic identity). | Nothing. |
| Phylogenetic structure with one shared phylogenetic variance per trait (simplest phylo model). | `+ phylo_scalar(species, vcv = Cphy)` |
| **Canonical paired phylogenetic decomposition** (recommended when traits can be both phylogenetically and non-phylogenetically correlated). | `+ phylo_latent(species, d = K_phy, tree = tree) + phylo_unique(species, tree = tree) + latent(0 + trait \| species, d = K_non) + unique(0 + trait \| species)` |
| Many species (n > ~200) and you want speed. | (same paired form) with `tree = tree` (sparse $A^{-1}$ trick); much faster than `vcv = Cphy` at this scale. |

The crossover where the sparse $A^{-1}$ path overtakes the dense
path sits around `n_species ≈ 150-200`. Above that, the speedup is
roughly an order of magnitude. The
[Phylogenetic GLLVM article](phylogenetic-gllvm.html) shows the
benchmark, the worked four-component recovery, and the
`extract_communality()` / `extract_phylo_signal()` outputs.

## 4. Are rows spatially organised?

If your sites carry coordinates and you suspect residual spatial
autocorrelation, add a sparse SPDE/GMRF random field. Each trait
gets its own field by default.

| Your situation | Add to formula | Argument to `gllvmTMB()` |
|---|---|---|
| Sites have no spatial structure (or you do not have coordinates). | Nothing. | — |
| Sites have coordinates and each trait should have its own spatial pattern. | <code>+ spatial_unique(0 + trait &#124; coords)</code> | `mesh = make_mesh(df, c("lon", "lat"), cutoff = ...)` |
| Sites have coordinates and you want one shared spatial-variance scale across traits (parsimony). | <code>+ spatial_scalar(0 + trait &#124; coords)</code> | `mesh = make_mesh(df, c("lon", "lat"), cutoff = ...)` |
| You want traits to share a low-rank spatial structure. | <code>+ spatial_latent(0 + trait &#124; coords, d = K)</code> — K shared SPDE fields driven by a T × K loading matrix Λ_spa. | `mesh = make_mesh(df, c("lon", "lat"), cutoff = ...)` (the spatial analogue of `phylo_latent()`). |

The SPDE engine wins on speed when the dataset has more than
roughly 100 sites; below that, the dense `glmmTMB::exp()` covariance
is competitive.

## 5. Do you have a reduced-rank story for trait covariance?

Even with no phylogenetic or spatial structure, you have to decide
how the trait covariance itself is parameterised. This is the
choice between *factor analysis* (a few latent dimensions
explaining most cross-trait covariance) and *unstructured
covariance* (every pair gets its own parameter).

| Your situation | Add to formula | Notes |
|---|---|---|
| You expect a small number of latent dimensions (size, shape, life-history pace) to explain most of the cross-trait covariance. T traits, K << T factors. | <code>+ latent(0 + trait &#124; site, d = K)</code> | The default GLLVM choice. K = 1–3 is typical. |
| You want full unstructured between-unit covariance with no rank reduction. | <code>+ unique(0 + trait &#124; site)</code> (specific variance) plus a separate term for off-diagonal structure as needed | Reasonable when T is small (T ≤ 5) and you have many units. |
| Confirmatory: you know which traits load on which factor and want to pin those entries of $\boldsymbol\Lambda$. | <code>+ latent(0 + trait &#124; site, d = K)</code> plus `lambda_constraint = list(B = M)` | M is a `T × K` matrix with `NA` for free entries and numeric values for fixed ones. See [Lambda constraint](lambda-constraint.html); `?suggest_lambda_constraint` proposes a sensible matrix from a fitted unconstrained model. |

For two-level models you typically include both `latent()` and `unique()`
at each level, so the full between-unit piece is
`latent(0 + trait | site, d = K) + unique(0 + trait | site)` and the
within-unit piece substitutes the within-unit grouping column
(e.g. `site_species` for site × species data, or the value passed via
`unit_obs = "..."` for other domains).
The `latent()` term gives the shared latent factors; the `unique()` term
gives each trait its own residual variance on top.

## 6. Final formula assembly

Below are the canonical formulas for the most common combinations.
The chunks below run live to confirm the formulas parse; each fits
in a few seconds on the small simulations.

```{r setup}
library(gllvmTMB)
library(ape)
```

### 6a. Individuals × continuous traits, no extra structure

The simplest case (rung 0): one level, Gaussian, no phylogeny, no
space. See [Morphometrics](morphometrics.html) for the full
walkthrough.

```{r ex-1, results = "hide"}
set.seed(1)
n_ind <- 30; n_traits <- 4
df_ind <- data.frame(
  individual = factor(rep(seq_len(n_ind), each = n_traits)),
  trait      = factor(rep(paste0("trait_", seq_len(n_traits)), n_ind),
                      levels = paste0("trait_", seq_len(n_traits))),
  value      = rnorm(n_ind * n_traits)
)
fit_ind <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | individual, d = 2),
  data = df_ind,
  unit = "individual"
)
```

### 6b. Sites × species, presence/absence

Rung 1 with a binomial response: a joint species distribution
model. See [Joint SDM](joint-sdm.html).

```{r ex-2, results = "hide"}
set.seed(2)
sim_b <- simulate_site_trait(
  n_sites = 30, n_species = 6, n_traits = 4,
  mean_species_per_site = 4,
  Lambda_B = matrix(rnorm(8, sd = 0.6), 4, 2),
  S_B      = rep(0.3, 4), seed = 2
)
df_b <- sim_b$data
df_b$value <- as.integer(df_b$value > 0)

fit_jsdm <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | site, d = 2) +
          unique(0 + trait | site),
  data   = df_b,
  family = binomial()
)
```

### 6c. Add phylogeny

Rung 2: the canonical paired phylogenetic decomposition combines
`phylo_latent + phylo_unique` (the phy side) with
`latent + unique` (the non-phy side). With a tree in hand, prefer
`tree = tree` (in-keyword) for the sparse $A^{-1}$ path. See
[Phylogenetic GLLVM](phylogenetic-gllvm.html) for the worked
example with simulation recovery and `extract_phylo_signal()`.

```{r ex-3, results = "hide"}
set.seed(3)
n_sp <- 8
tree <- ape::rcoal(n_sp); tree$tip.label <- paste0("sp", seq_len(n_sp))

sim_p <- simulate_site_trait(
  n_sites = 30, n_species = n_sp, n_traits = 4,
  mean_species_per_site = 5,
  Lambda_B = matrix(rnorm(8, sd = 0.5), 4, 2),
  S_B = rep(0.3, 4),
  Cphy = ape::vcv(tree, corr = TRUE), sigma2_phy = rep(0.4, 4),
  seed = 3
)
df_p <- sim_p$data
levels(df_p$species) <- tree$tip.label

fit_phy <- gllvmTMB(
  value ~ 0 + trait +
          phylo_latent(species, d = 1, tree = tree) +
          phylo_unique(species, tree = tree) +
          latent(0 + trait | species, d = 1) +
          unique(0 + trait | species),
  data = df_p,
  unit = "species"
)
```

### 6d. Add space

Rung 3: append `spatial_unique(0 + trait | coords)` and pass a mesh. The
simulator places sites in `[0,1]^2` automatically when
`spatial_range` is supplied. The SPDE engine wins on speed when
the dataset has more than roughly 100 sites; below that the dense
`glmmTMB::exp()` covariance is competitive.

```{r ex-4, results = "hide"}
set.seed(4)
sim_s <- simulate_site_trait(
  n_sites = 40, n_species = 1, n_traits = 3,
  mean_species_per_site = 1,
  Lambda_B = matrix(0.0, 3, 1), S_B = rep(0.1, 3),
  spatial_range = 0.3, sigma2_spa = rep(0.4, 3),
  seed = 4
)
df_s <- sim_s$data
mesh <- make_mesh(df_s, c("lon", "lat"), cutoff = 0.1)

fit_spa <- gllvmTMB(
  value ~ 0 + trait + spatial_unique(0 + trait | coords, mesh = mesh),
  data = df_s
)
```

### 6e. Two-level + phylo + space (capstone)

The full Nakagawa et al. functional-biogeography model: two levels,
phylogeny, and space all at once. The formula reads as a sum of
the pieces from sections 1–5. The runnable end-to-end version
lives in [Functional biogeography](functional-biogeography.html);
in this article we just show the formula:

```r
fit_full <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):env_1 + (0 + trait):env_2 +
          spatial_unique(0 + trait | coords, mesh = mesh) +
          latent(0 + trait | site, d = 2) +
          unique(0 + trait | site) +
          latent(0 + trait | site_species, d = 1) +
          unique(0 + trait | site_species) +
          phylo_latent(species, d = 1, tree = tree) +
          phylo_unique(species, tree = tree),
  data = df
)
```

## 7. Diagnostic checklist

After any fit, run through these in order:

| Check | How |
|---|---|
| Optimiser convergence and Hessian PD-ness. | `sanity_multi(fit)` |
| Implied between-unit trait covariance. | `extract_Sigma(fit, level = "unit")` (and `extract_Sigma(fit, level = "unit_obs")` for two-level fits). |
| Interpretable loadings. | `getLoadings(fit, level = "unit", rotate = "varimax")` |
| Did you fit unconstrained `latent()` with `d > 1`? Loadings are not unique. | Use `rotate = "varimax"` for a quick post-hoc rotation, or call `suggest_lambda_constraint()` to propose a confirmatory `lambda_constraint` matrix and refit. |
| Communalities / ICCs. | `extract_ICC_site(fit)`, `extract_communality(fit, level = "unit")`, `extract_communality(fit, level = "unit_obs")` |

If `sanity_multi()` flags a non-PD Hessian or a near-zero loading
diagonal, the most common causes are an over-rich model
(`d` too large, too many `latent()` terms relative to data) or
factor-level ordering issues in the long-format data. The
[Morphometrics](morphometrics.html) article has a worked example
of the factor-level pitfall.

## See also

* [Morphometrics](morphometrics.html) — rung 0: individuals × continuous traits.
* [Joint SDM](joint-sdm.html) — rung 1: binary multivariate.
* [Behavioural syndromes](behavioural-syndromes.html) — two observational levels.
* [Phylogenetic GLLVM](phylogenetic-gllvm.html) — phylogenetic dependence (the canonical paired four-component decomposition).
* [Functional biogeography](functional-biogeography.html) — capstone: two-level + space + phylogeny.
* [Response families](response-families.html) — supported likelihoods and mixed-family dispatch.
* [Lambda constraint](lambda-constraint.html) — confirmatory factor analysis with fixed loadings; see also `?suggest_lambda_constraint`.
* [Pitfalls](pitfalls.html) — common gotchas in data prep, formula construction, and interpretation.
