---
title: "Convergence and hard fits"
description: "Diagnose a numerically difficult fit and compare start or optimiser strategies without turning a small criterion into a validity certificate."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Convergence and hard fits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = TRUE
)
```

Some multivariate latent models are hard because the likelihood has
several plausible basins—a basin is a local solution reached from one set of
starting values—weakly identified variance splits, or a flat curvature
direction near a boundary. The practical question is
therefore not just "did the optimizer print convergence code 0?" but
"which part of the fit is healthy, which part is weak, and what should
I try next?"

This article covers `check_gllvmTMB()`, restart history, protected or
skipped `TMB::sdreport()` status, and `gllvmTMBcontrol(se = FALSE)`.
Bootstrap and profile uncertainty are follow-up paths when Wald standard
errors are unsafe, but they have narrower evidence: bootstrap summaries
depend on successful refits, and profile intervals apply only to supported
targets with stable profiles. The start strategies below are options to try,
not established defaults or universal convergence recommendations.

The most important distinction is simple. `fit$fit_health$converged` is the
package's conservative point-stationarity conjunction: optimiser success, a
finite objective, and a small raw maximum gradient. By contrast, `pd_hessian`
reports TMB's curvature result. `FALSE` means ordinary Wald uncertainty is not
supported; `NA` after `se = FALSE` means the Hessian was not computed. Keep the
point-fit and inference checks separate, and do not use objective scaling to
override a failed optimiser code.

```{r setup}
library(gllvmTMB)
set.seed(20260519)
```

## Fit a small model

The example below uses a two-trait Gaussian model. The scientific
target is the trait covariance
`Sigma = Lambda Lambda^T + Psi`, where `Lambda` contains the shared
latent loadings and `Psi` is the diagonal trait-specific variance.
In current ordinary `latent()` models, that diagonal `Psi` companion is
included by default; `latent(..., unique = FALSE)` requests the older
no-Psi subset.

Use this alignment while reading the diagnostics:

| Symbol | R syntax or object | Plain meaning |
|---|---|---|
| `Lambda` | `latent(0 + trait | site, d = 1)`; `Lambda_B` in the simulator | The shared loading axis that creates cross-trait covariance. |
| `Psi` | default companion inside `latent(0 + trait | site, d = 1)`; `psi_B` in the simulator | Trait-specific variance that is not shared with the other trait. |
| `Sigma` | `extract_Sigma_table(fit, level = "unit", part = "total")` | The full trait covariance, `Lambda Lambda^T + Psi`, used for interpretation. |

```{r simulate}
sim <- simulate_site_trait(
  n_sites = 24, n_species = 1, n_traits = 2,
  mean_species_per_site = 1,
  Lambda_B = matrix(c(0.7, 0.4), nrow = 2, ncol = 1),
  psi_B = c(0.3, 0.3),
  seed = 20260519
)
df_long <- sim$data
```

The long-format call is the canonical interface: one row per
`(unit, trait)` observation, with the `trait` and `unit` columns named
explicitly.

```{r fit-long}
fit <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | site, d = 1),
  data = df_long,
  trait = "trait",
  unit = "site",
  control = gllvmTMBcontrol(n_init = 2, init_jitter = 0.05)
)
```

The same model can be written with the wide data-frame formula
interface. This route is useful when your data start as one row per
unit and one column per trait.

```{r wide-form}
df_wide <- stats::reshape(
  df_long[c("site", "trait", "value")],
  idvar = "site", timevar = "trait", direction = "wide"
)
names(df_wide) <- sub("^value\\.", "", names(df_wide))

fit_wide <- gllvmTMB(
  traits(trait_1, trait_2) ~ 1 +
    latent(1 | site, d = 1),
  data = df_wide,
  unit = "site",
  control = gllvmTMBcontrol(n_init = 2, init_jitter = 0.05)
)

signif(abs(as.numeric(logLik(fit)) - as.numeric(logLik(fit_wide))), 3)
```

The printed difference is exactly zero at the shown precision, so the long and
wide calls reached the same likelihood for this example. That agreement does
not replace the health checks below.

## Read fit health

Start with the conservative point-stationarity verdict. It requires an
optimiser success code, a finite objective, and an unscaled maximum gradient
below `0.01`. The objective-scaled gradient is retained as a descriptive signal,
not a substitute for those checks.

```{r diagnose}
data.frame(
  converged = isTRUE(fit$fit_health$converged),
  optimizer_converged = isTRUE(fit$fit_health$optimizer_converged),
  stationary_by_gradient = isTRUE(fit$fit_health$stationary_by_gradient),
  raw_max_gradient = signif(fit$fit_health$max_gradient, 4),
  raw_gradient_threshold = 0.01,
  scaled_gradient_descriptive = signif(fit$fit_health$scaled_gradient, 4),
  objective = signif(fit$fit_health$objective, 6),
  optimizer_code = fit$fit_health$convergence
)

health <- check_gllvmTMB(fit)
triage_components <- c(
  "optimizer_convergence", "max_gradient", "sdreport", "pd_hessian",
  "hessian_rank", "restart_history", "boundary_flags", "weak_axis_unit",
  "near_zero_psi_unit"
)
health[
  health$component %in% triage_components,
  c("component", "status", "value")
]
```

Read the components in three lanes:

| Lane | Components | What it answers | Typical next action |
|---|---|---|---|
| Convergence | optimizer code, raw maximum gradient, restart history; scaled gradient is descriptive | Is the selected point sufficiently stationary, and did starts find competing objective values? | Try more starts, rescale predictors, or compare optimizers. |
| Hessian-based inference | `sdreport`, `pd_hessian`, Hessian rank, fixed-effect SE | Is local curvature usable for Wald uncertainty? | If unavailable or weak, inspect the target and use a supported profile or bootstrap route. |
| Identifiability and interpretation | weak axes, boundary flags, near-zero `Psi`, rotation convention | Is a covariance component or latent axis weak, redundant, or convention-dependent? | Compare lower rank, simpler covariance, and rotation-invariant `Sigma`; use the advanced identifiability check when appropriate. |

In this baseline fit, the conservative convergence conjunction is `TRUE`, both restarts reach
the same displayed objective, `sdreport` is available, and the reported Hessian
rank is full. However, `pd_hessian` is `FALSE`. That combination says the
selected point is stationary but local curvature still warns against a blanket
Wald-uncertainty claim. It does not by itself prove either model failure or a
benign rotation ridge; inspect the target and its profile or refit behaviour.

This separation matters. A fit can be stationary while Wald uncertainty is
unavailable, or it can have a positive-definite Hessian while an extra latent
axis remains scientifically weak. Do not flatten every warning into model
failure.

After the numerical triage, inspect the fitted response distribution
before leaning on covariance interpretation. `residuals(fit,
type = "randomized_quantile")` and `predictive_check(fit,
type = "rq_qq")` are available for the scoped Gaussian, Poisson, and
NB2 diagnostic paths. They serve the same role as the fit-health
table: useful for surfacing mismatch, not proof of interval
calibration, latent-rank adequacy, or Bayesian posterior prediction.
The worked diagnostic sequence is shown in
[Can I trust this fit?](fit-diagnostics.html).

## Convergence is not identifiability

Convergence asks whether optimization reached a sufficiently stationary point.
Identifiability asks whether the data can distinguish the parameters or latent
structure being interpreted. Raw `Lambda` loadings also depend on a rotation
convention, whereas `Sigma` is rotation invariant and is therefore the safer
cross-start comparison.

Simulation can assess a predeclared rank-selection rule only when the data-
generating rank is fixed independently of the fitted model. Simulating from a
rank-two fit and successfully refitting rank two tests stability under that
fitted scenario; it cannot establish that rank two was required by the original
data. For a study-level validation, predeclare candidate ranks, simulate from
known rank-one and rank-two models over realistic sample sizes, and report how
often the full selection workflow recovers each generating rank.

## Choose a start strategy

Start values are opt-in tools. The actual default is one `nlminb` fit using
historical starts with `se = TRUE`. The strategies below are experimental:
their recovery benefit across
families and ranks is not yet established by simulation, so treat the
table as a menu of things to try on a hard fit, not a proven policy.

| What you observe | Strategy | Control |
|---|---|---|
| No warning or competing objective yet | Use the default first | `gllvmTMBcontrol()` |
| Starts reach different objective values | Add restarts and modest jitter | `gllvmTMBcontrol(n_init = 5, init_jitter = 0.1)` |
| A non-Gaussian reduced-rank fit repeatedly stops poorly | Seed loadings and scores from residual structure | `gllvmTMBcontrol(start_method = list(method = "res", jitter.sd = 0.2))` |
| A Gaussian two-level latent fit is sensitive to its initial covariance | Warm-start from the matching independent model | `gllvmTMBcontrol(start_method = list(method = "indep"))` |
| A scientifically sensible simpler fit is already available | Copy compatible starting parameters | `gllvmTMBcontrol(start_from = simpler_fit)` |
| A fitted `phi` parameter is unstable, especially for NB1/NB2 | Warm up dispersion trait by trait | `gllvmTMBcontrol(init_strategy = "single_trait_warmup")` |
| `nlminb` does not reach a stationary point | Compare `optim` with BFGS | `gllvmTMBcontrol(optimizer = "optim", optArgs = list(method = "BFGS"))` |
| Point estimates are stable but Wald calculation is the fragile step | Deliberately skip standard errors | `gllvmTMBcontrol(se = FALSE)` |

For a difficult model, start with more than one restart and inspect
`fit$restart_history`. This is currently a low-level fitted-object diagnostic
field, so its columns may evolve during the experimental lifecycle.

```{r restart-history}
fit$restart_history[, c("restart", "start_method", "objective",
                        "convergence", "selected")]
```

Lower objective values are preferred. Here the two displayed objectives are
equal at the shown precision and restart 1 is selected. That supports, but does
not prove, that the starts reached the same solution. Do not privilege a start
simply because it converged: compare the selected fit's optimiser code and raw gradient
and rotation-invariant targets such as `Sigma`. Compare raw `Lambda` only after
alignment or rotation handling.

## Use no-SE fits honestly

`gllvmTMBcontrol(se = FALSE)` intentionally skips `TMB::sdreport()`. Use it only
after an ordinary fit has shown that point estimates are stable but
standard-error calculation is the fragile step. The skipped status is `NA`, not evidence
that a computed Hessian was non-positive-definite.

```{r se-false-status}
fit_no_se <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | site, d = 1),
  data = df_long,
  trait = "trait",
  unit = "site",
  control = gllvmTMBcontrol(se = FALSE, n_init = 2, init_jitter = 0.05)
)

fit_health <- check_gllvmTMB(fit_no_se)
fit_health[
  fit_health$component %in% c("sdreport", "pd_hessian", "max_fixed_se"),
  c("component", "status", "value", "message")
]
```

This option is not a way to hide convergence problems. It is a way to
avoid making Hessian-based uncertainty claims when the Hessian path is
the fragile part of the workflow.

## Bootstrap after a one-core smoke test

Bootstrap refits the model many times. It is one practical uncertainty
path when `pdHess` is
false, when `sdreport()` fails, or when the target is nonlinear, but
it is not a cure for an unstable fitted surface. `bootstrap_Sigma()`
exposes `n_cores` so independent refits can run in parallel.

The Gaussian `bootstrap_Sigma()` route has direct simulate-refit contract and
plumbing tests, but this article does not claim uniform interval coverage
across all targets. Non-Gaussian and mixed-family refit plumbing is shallower;
treat those intervals as an audit trail until the relevant family and
target have coverage evidence.

```{r bootstrap, eval = FALSE}
boot_smoke <- bootstrap_Sigma(
  fit,
  n_boot = 25,
  level = "unit",
  what = c("Sigma", "R", "communality"),
  seed = 1,
  n_cores = 1
)

boot_smoke$n_failed
extract_Sigma_table(boot_smoke, level = "unit", entries = "upper")
```

The 25-replicate call is only a plumbing smoke test. Once the model refits
cleanly, increase to at least 200 replicates and optionally use multiple cores.
Even 200 is illustrative for percentile endpoints because only a handful of
draws determine each tail; increase `n_boot` further for final reporting and
check that endpoints are stable. Always record the failure count. A bootstrap
CI with many failed refits is a diagnostic result, not just an
inconvenience; report the failure count and simplify the model before
treating the interval as a stable uncertainty summary.

## Profile when the target is scalar

Profile likelihood is a targeted route for scalar quantities such as
repeatability, communality, selected correlations, or a single variance
share. Use it when the target has a stable profile
and the model is not bouncing between likelihood basins. It is not a
general substitute for bootstrap across every derived table entry.

For full latent covariance models, some derived targets still fall back to
Wald or bootstrap while constrained-profile machinery is expanded.
The output method column should be read literally: if it says
`"wald"`, the interval is Hessian-based even if the requested method
was `"profile"`.

The generic `confint(method = "profile")` and
`confint(method = "bootstrap")` routes are contract-tested, but route
availability is not the same as empirical interval calibration.
Target-specific profile helpers are narrower: check
`profile_targets(fit)` before treating a derived quantity as
profile-ready, and read the returned method and status literally.
For the interval workflow, continue to
[Profile-likelihood confidence intervals](profile-likelihood-ci.html);
for failed or irregular profiles, use its
[troubleshooting decision table](profile-likelihood-ci.html#troubleshoot-the-interval-that-came-back).

## Troubleshooting ladder

Use this order when a hard model does not give clean inference:

1. Read `check_gllvmTMB(fit)` and `fit$restart_history`.
2. Check whether the optimizer, gradient, `pdHess`, or `sdreport`
   component is the first weak link.
3. Use `residuals()` or `predictive_check()` to check fitted-response
   mismatch before interpreting covariance summaries.
4. Inspect boundary flags for near-zero variance, redundant rank, or
   unstable dispersion.
5. Try more starts and modest jitter.
6. Try residual starts for non-Gaussian reduced-rank models.
7. Try independent or simpler-model starts for Gaussian two-level
   models.
8. Try `optimizer = "optim"` with `optArgs = list(method = "BFGS")`, or a
   longer `nlminb` budget such as
   `gllvmTMBcontrol(optArgs = list(control = list(eval.max = 5000, iter.max = 4000)))`.
9. If point estimates are stable but `pdHess` remains false, use
   bootstrap or profile uncertainty.
10. If the target still changes across starts, simplify rank or
   covariance structure before making a scientific claim.

## The bottom line

The carry-forward principle is that hard fits need honest triage. A
weak Hessian should change the uncertainty workflow, not automatically
discard every point estimate the model produced.
