---
title: "Convergence, identifiability, and start values"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Convergence, identifiability, and start values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

> **Scope boundary.** This article covers the robust-modeling
> diagnostics implemented in Design 49: `check_gllvmTMB()`,
> `fit$fit_health`, `fit$restart_history`, `fit$start_provenance`,
> protected or skipped `TMB::sdreport()` status, and
> `gllvmTMBcontrol(se = FALSE)`. The validation-debt register marks
> `check_gllvmTMB()` and the `se = FALSE` point-estimate path as
> covered (DIA-08, DIA-10), while a deterministic in-fit
> `sdreport()` failure fixture remains partial (DIA-09).
> Start-method convergence-rate claims are still M3.3a/M3.4 evidence
> work, not default-policy claims.

Some multivariate latent models are hard because the likelihood has
several plausible basins, weakly identified variance splits, or a
flat curvature direction near a boundary. The practical question is
therefore not just "did the optimizer print convergence code 0?" but
"which part of the fit is healthy, which part is weak, and what should
I try next?"

The most important rule is simple: `pdHess = FALSE` means
Hessian-based standard errors and naive Wald intervals are unsafe. It
does not automatically mean the point estimates, fitted values,
likelihood, or rotation-invariant covariance summaries are useless.
When the point estimate is scientifically useful but the Hessian is
weak, fit with `se = FALSE` or keep the degraded `sdreport()` status
visible, then use bootstrap or profile workflows for uncertainty.

```{r setup}
library(gllvmTMB)
set.seed(20260519)
```

## Fit A Small Model

The example below uses a two-trait Gaussian model. The scientific
target is the trait covariance
`Sigma = Lambda Lambda^T + Psi`, where `Lambda` contains the shared
latent loadings and `Psi` is the diagonal trait-specific variance.

```{r simulate}
sim <- simulate_site_trait(
  n_sites = 24, n_species = 1, n_traits = 2,
  mean_species_per_site = 1,
  Lambda_B = matrix(c(0.7, 0.4), nrow = 2, ncol = 1),
  psi_B = c(0.3, 0.3),
  seed = 20260519
)
df_long <- sim$data
```

The long-format call is the canonical interface: one row per
`(unit, trait)` observation, with the `trait` and `unit` columns named
explicitly.

```{r fit-long}
fit <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | site, d = 1) +
    unique(0 + trait | site),
  data = df_long,
  trait = "trait",
  unit = "site",
  control = gllvmTMBcontrol(se = FALSE, n_init = 2, init_jitter = 0.05)
)
```

The same model can be written with the wide data-frame formula
interface. This route is useful when your data start as one row per
unit and one column per trait.

```{r wide-form, eval = FALSE}
df_wide <- stats::reshape(
  df_long[c("site", "trait", "value")],
  idvar = "site", timevar = "trait", direction = "wide"
)
names(df_wide) <- sub("^value\\.", "", names(df_wide))

fit_wide <- gllvmTMB(
  traits(trait_1, trait_2) ~ 1 +
    latent(1 | site, d = 1) +
    unique(1 | site),
  data = df_wide,
  unit = "site",
  control = gllvmTMBcontrol(se = FALSE, n_init = 2, init_jitter = 0.05)
)
```

## Read Fit Health

`check_gllvmTMB()` returns a stable table. It is designed for users,
tests, and simulation scripts so they do not have to parse optimizer
messages by hand.

```{r diagnose}
check_gllvmTMB(fit)
```

The components have different meanings:

| Component | What it answers | Typical next action |
|---|---|---|
| optimizer convergence | Did the optimizer report a clean stop? | If not, try more starts, a longer optimizer budget, or `optim(BFGS)`. |
| max gradient | Is the objective flat at the reported optimum? | Rescale predictors, simplify the model, or compare optimizers. |
| `pdHess` | Is local curvature positive definite for Hessian-based inference? | If false, avoid naive Wald intervals and inspect boundary or rank flags. |
| `sdreport` | Did `TMB::sdreport()` produce standard errors? | If skipped or failed, keep point estimates but route uncertainty elsewhere. |
| restart history | Did several starts reach the same basin? | Use the objective spread to decide whether starts are stable. |
| boundary flags | Are variance or loading parameters near a boundary? | Try a lower rank, simpler covariance structure, or profile the target. |

This separation matters. A fit can have useful point estimates while
`sdreport` is skipped or while `pdHess` is false. The diagnosis should
tell the user what is degraded, not flatten every warning into model
failure.

## Choose A Start Strategy

Start values are opt-in tools. The default remains conservative until
M3.3a/M3.4 simulations show which strategies improve recovery across
families and ranks.

| Situation | Control |
|---|---|
| Routine fit | `gllvmTMBcontrol()` |
| Suspect local optima | `gllvmTMBcontrol(n_init = 5, init_jitter = 0.1)` |
| Non-Gaussian reduced-rank fit | `gllvmTMBcontrol(start_method = list(method = "res", jitter.sd = 0.2))` |
| Gaussian two-level latent+unique fit | `gllvmTMBcontrol(start_method = list(method = "indep"))` |
| Known simpler model is available | `gllvmTMBcontrol(start_from = simpler_fit)` |
| Count-family dispersion is unstable | `gllvmTMBcontrol(init_strategy = "single_trait_warmup")` |
| `nlminb` reports a suspicious stop | `gllvmTMBcontrol(optimizer = "optim", optArgs = list(method = "BFGS"))` |
| Point estimate is needed but SE calculation is unreliable | `gllvmTMBcontrol(se = FALSE)` |

For a difficult model, start with more than one restart and inspect
`fit$restart_history`:

```{r restart-history}
fit$restart_history[, c("restart", "start_method", "objective",
                        "convergence", "selected")]
```

Do not privilege a start method simply because it converged. Compare
the objective value, the diagnostics, and the scientific summaries.
Reduced-rank likelihoods can converge cleanly in an inferior basin.

## Use No-SE Fits Honestly

`gllvmTMBcontrol(se = FALSE)` intentionally skips `TMB::sdreport()`.
This mirrors the workflow used for difficult TMB-family models: keep
the point estimate, record that Hessian-based standard errors were not
computed, and obtain uncertainty from a refit-based method.

```{r se-false-status}
fit$fit_health$sdreport_ok
fit$sdreport_error
```

This option is not a way to hide convergence problems. It is a way to
avoid making Hessian-based uncertainty claims when the Hessian path is
the fragile part of the workflow.

## Bootstrap With Multiple Cores

Bootstrap is slower than Wald intervals because it refits the model
many times. It is also the natural uncertainty path when `pdHess` is
false, when `sdreport()` fails, or when the target is nonlinear.
Laptops often have enough cores to make bootstrap practical, so
`bootstrap_Sigma()` exposes `n_cores`.

```{r bootstrap, eval = FALSE}
boot <- bootstrap_Sigma(
  fit,
  n_boot = 200,
  level = "unit",
  what = c("Sigma", "R", "communality"),
  seed = 1,
  n_cores = 4
)

boot$n_failed
boot$ci_lower$Sigma_B
boot$ci_upper$Sigma_B
```

Use `n_cores = 1` for the first smoke test. Once the model refits
cleanly, increase `n_cores` and record the failure count. A bootstrap
CI with many failed refits is a diagnostic result, not just an
inconvenience.

## Profile When The Target Is Scalar

Profile likelihood is often faster than bootstrap for scalar targets
such as repeatability, communality, selected correlations, or a
single variance share. Use it when the target has a stable profile
and the model is not bouncing between likelihood basins.

For full latent+unique models, some derived targets still fall back to
Wald or bootstrap while constrained-profile machinery is expanded.
The output method column should be read literally: if it says
`"wald"`, the interval is Hessian-based even if the requested method
was `"profile"`.

## Troubleshooting Ladder

Use this order when a hard model does not give clean inference:

1. Read `check_gllvmTMB(fit)` and `fit$restart_history`.
2. Check whether the optimizer, gradient, `pdHess`, or `sdreport`
   component is the first weak link.
3. Inspect boundary flags for near-zero variance, redundant rank, or
   unstable dispersion.
4. Try more starts and modest jitter.
5. Try residual starts for non-Gaussian reduced-rank models.
6. Try independent or simpler-model starts for Gaussian two-level
   models.
7. Try `optim(BFGS)` or a longer `nlminb` budget.
8. If point estimates are stable but `pdHess` remains false, use
   bootstrap or profile uncertainty.
9. If the target still changes across starts, simplify rank or
   covariance structure before making a scientific claim.

## What Is Next

Implemented now:

- machine-readable fit health via `check_gllvmTMB()` and
  `fit$fit_health`;
- restart history and start provenance;
- protected or skipped `sdreport()` status;
- `gllvmTMBcontrol(se = FALSE)` for point-estimate-first hard fits;
- explicit controls for multistart, jitter, residual starts,
  independent starts, manual starts, single-trait warmup, and
  `optim(BFGS)`.

Partial:

- direct in-fit `sdreport()` failure tests are still hard to trigger
  deterministically in a small test, so DIA-09 remains partial;
- non-Gaussian start-method recommendations are available but not yet
  promoted as defaults.

Planned:

- the M3.3a target-explicit pilot will compare Gaussian, `nbinom2`,
  mixed-family, and two-level Gaussian fits using the new diagnostic
  fields;
- larger stress lanes will decide whether `nbinom2` fragility is a
  model-identifiability problem or a start/diagnostic problem;
- profile and bootstrap workflows will gain stronger target-parallel
  tooling for local multicore and larger compute runs;
- simulation diagnostic plots will show convergence, `pdHess`,
  objective spread, selected-start frequency, coverage, and failed
  refits by family and start strategy.

The carry-forward principle is that hard fits need honest triage. A
weak Hessian should change the uncertainty workflow, not automatically
discard every point estimate the model produced.
