---
title: "Covariance and correlation: when you need `unique()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Covariance and correlation: when you need `unique()`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

Suppose you want between-individual correlations among behaviours such as
boldness, exploration, activity, aggression, and sociability. If you fit only
`latent()`, the recovered correlations can be too large because the model has
no trait-specific variance term. Pair `latent()` with `unique()` when
trait-specific variance should stay in the denominator of the correlation
matrix.

```{r early-fit-pattern, eval = FALSE}
fit_latent_only <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | individual, d = 2),
  data = df, trait = "trait", unit = "individual"
)

fit_latent_unique <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | individual, d = 2) +
    unique(0 + trait | individual),
  data = df, trait = "trait", unit = "individual"
)
```

If your data are wide, with one row per individual and one column per
behaviour, use the same `gllvmTMB()` entry point with `traits(...)` on the
left-hand side:

```{r early-fit-pattern-wide, eval = FALSE}
fit_latent_unique <- gllvmTMB(
  traits(boldness, exploration, activity, aggression, sociability) ~
    1 + latent(1 | individual, d = 2) + unique(1 | individual),
  data = df_wide, unit = "individual"
)
```

## The decomposition

For the `latent() + unique()` model, the implied trait covariance is

$$
\boldsymbol\Sigma_\text{level} \;=\; \underbrace{\boldsymbol\Lambda_\text{level} \boldsymbol\Lambda_\text{level}^{\!\top}}_{\text{shared low-rank}} \;+\; \underbrace{\boldsymbol\Psi_\text{level}}_{\text{trait-specific diagonal}},
$$

where `level` names the covariance tier being extracted, such as `unit`
for between-individual covariance or `unit_obs` for within-individual or
observation-level covariance. $\boldsymbol\Lambda$ is the $T \times K$
loading matrix (from the `latent()` term) and $\boldsymbol\Psi$ is a
$T \times T$ diagonal matrix of trait-specific unique variances (from the
`unique()` term). The correlation matrix is then

$$
\mathbf R \;=\; \mathbf D^{-1/2}\,\boldsymbol\Sigma\,\mathbf D^{-1/2},
\qquad \mathbf D \;=\; \mathrm{diag}(\boldsymbol\Sigma).
$$

Both pieces matter: the diagonal of $\boldsymbol\Sigma$ is
$(\boldsymbol\Lambda \boldsymbol\Lambda^{\!\top})_{tt} + \psi_{tt}$ — the
sum of squared loadings *plus* the trait-specific unique variance.
Drop $\boldsymbol\Psi$ and the diagonal shrinks; the off-diagonals stay the
same; $\mathbf R$ swells.

Read the equation beside the R syntax:

| Symbol | R syntax or extractor | Plain meaning |
|---|---|---|
| `level` | `level = "unit"` or `level = "unit_obs"` | Which covariance tier you are extracting. |
| $\boldsymbol\Lambda_\text{level}$ | `latent(..., d = K)` | The `K` loading axes that create shared cross-trait covariance. |
| $\boldsymbol\Lambda\boldsymbol\Lambda^\top$ | `extract_Sigma_table(fit, level = "unit", part = "shared")` | The covariance explained by the shared latent axes only. |
| $\boldsymbol\Psi_\text{level}$ and $\psi_{tt}$ | `unique(...)`; `extract_Sigma_table(fit, level = "unit", part = "unique")` | Trait-specific variance that stays in the denominator of correlations. |
| $\boldsymbol\Sigma_\text{level}$ | `latent(...) + unique(...)`; `extract_Sigma_table(fit, level = "unit", part = "total")` | The full trait covariance matrix used for reporting. |
| $\mathbf R$ | `extract_correlations(fit, tier = "unit")` or `$R` from `extract_Sigma()` | The correlation matrix after standardising $\boldsymbol\Sigma$. |

## A side-by-side demonstration

We use a prepared Gaussian behavioural-syndrome example object. The generator
is `data-raw/examples/make-covariance-edge-cases-example.R`; the article uses
the shipped object so the first thing the reader sees is the model, not a long
data-generating block.

The truth has both a shared low-rank component and a trait-specific unique
diagonal. Fit the same data two ways:

* `Model A`: `latent(0 + trait | individual, d = 2)` only
* `Model B`: `latent(0 + trait | individual, d = 2) + unique(0 + trait | individual)`

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

```{r load-example, eval = FALSE}
covex <- readRDS(system.file(
  "extdata", "examples", "covariance-edge-cases-example.rds",
  package = "gllvmTMB"
))
df <- covex$data_long
df_wide <- covex$data_wide
truth <- covex$truth
```

```{r load-example-data, include = FALSE}
covex_path <- system.file(
  "extdata", "examples", "covariance-edge-cases-example.rds",
  package = "gllvmTMB"
)
covex_path <- c(
  covex_path,
  file.path("inst", "extdata", "examples", "covariance-edge-cases-example.rds"),
  file.path("..", "inst", "extdata", "examples", "covariance-edge-cases-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "covariance-edge-cases-example.rds")
)
covex_path <- covex_path[nzchar(covex_path) & file.exists(covex_path)][1]
stopifnot(!is.na(covex_path))
covex <- readRDS(covex_path)
df <- covex$data_long
df_wide <- covex$data_wide
truth <- covex$truth
trait_names <- truth$trait_names
Sigma_true <- truth$Sigma
```

```{r show-example}
covex$story$question
head(df, 6)
```

The long-format formulas are:

```{r formulas-long}
covex$edge_cases$latent_only$formula_long
covex$formula_long
```

The wide data-frame formula expresses the same recommended model through the
`traits(...)` left-hand side:

```{r formula-wide}
covex$formula_wide
```

```{r fit}
ctl <- gllvmTMBcontrol(se = FALSE)

fit_A <- gllvmTMB(
  covex$edge_cases$latent_only$formula_long,
  data = df,
  trait = covex$fit_args$trait,
  unit = covex$fit_args$unit,
  family = covex$fit_args$family,
  control = ctl
)

fit_B <- gllvmTMB(
  covex$formula_long,
  data = df,
  trait = covex$fit_args$trait,
  unit = covex$fit_args$unit,
  family = covex$fit_args$family,
  control = ctl
)

fit_B_wide <- gllvmTMB(
  covex$formula_wide,
  data = df_wide,
  unit = covex$fit_args$unit,
  family = covex$fit_args$family,
  control = ctl
)
```

Long and wide fits use different data shapes, but here they represent the same
likelihood:

```{r long-wide-check}
c(
  long = as.numeric(logLik(fit_B)),
  wide = as.numeric(logLik(fit_B_wide))
)
```

Now extract the implied $\hat{\boldsymbol\Sigma}_{\text{unit}}$ and
between-individual correlation matrix from each fit:

```{r extract}
ext_A <- suppressMessages(
  extract_Sigma(fit_A, level = "unit", part = "total")
)
ext_B <- extract_Sigma(fit_B, level = "unit", part = "total")
```

Notice that the call on `fit_A` emits a one-shot **advisory**
explaining that without `unique()` the returned $\hat{\boldsymbol\Sigma}$
is latent-only (we suppressed the message above for clean output but the
message is also stored in `ext_A$note`).

```{r show-note}
ext_A$note  # advisory: diag missing
```

## The correlations side by side

```{r corr-comparison, fig.width = 7.5, fig.height = 5.2, fig.cap = "Trait-pair correlation errors for the latent-only model and the latent + unique model. Positive values mean the fitted correlation is larger than truth; zero means exact recovery for that pair."}
cmp_A <- compare_Sigma_table(
  fit_A,
  truth = Sigma_true,
  level = "unit",
  measure = "correlation",
  entries = "upper"
)
cmp_A$comparison <- "Model A: latent only"

cmp_B <- compare_Sigma_table(
  fit_B,
  truth = Sigma_true,
  level = "unit",
  measure = "correlation",
  entries = "upper"
)
cmp_B$comparison <- "Model B: latent + unique"

corr_comparison <- rbind(cmp_A, cmp_B)
plot_Sigma_comparison(
  corr_comparison,
  measure = "correlation",
  facet = "comparison",
  sort = "trait"
)
```

`Model A` shows correlations that are **uniformly larger** than truth.
`Model B` pulls those errors back toward zero. The mechanism:
Model A's $\hat{\boldsymbol\Sigma}$ has small diagonals
(only $\Lambda \Lambda^{\!\top}$), so when we divide by $\sqrt{D}$ the
off-diagonals get inflated. Model B's $\hat{\boldsymbol\Sigma}$ has
the full $\Lambda \Lambda^{\!\top} + \boldsymbol\Psi$ — diagonals at the right scale
and correlations at the right magnitude.

## Report-ready Sigma rows

For reports, make the row object first. Each covariance target is one
tidy row with trait names, matrix part, estimate, interval placeholders,
and validation status:

```{r sigma-table-rows}
sigma_rows_B <- extract_Sigma_table(
  fit_B,
  level = "unit",
  part = "total",
  measure = "covariance",
  entries = "upper"
)
sigma_rows_B
```

`part = "total"` is the default and is what you almost always want for
reporting. `"shared"` is useful if you specifically want the latent-implied
component (e.g. for ordination interpretation or communality). `"unique"`
gives the diagonal of $\boldsymbol\Psi$ as a named numeric vector.

The same table helper exposes the three decomposition parts without
hand-indexing matrices:

```{r sigma-part-rows}
sigma_part_rows_B <- rbind(
  extract_Sigma_table(fit_B, level = "unit", part = "shared", entries = "diag"),
  extract_Sigma_table(fit_B, level = "unit", part = "unique", entries = "diag"),
  extract_Sigma_table(fit_B, level = "unit", part = "total", entries = "diag")
)

sigma_part_rows_B[c("component", "matrix", "trait_i", "trait_j", "estimate")]
```

If you need a matrix for algebra checks, use `extract_Sigma()` directly:

```{r sigma-matrix-backend}
# Lambda Lambda^T alone (the "shared" component)
extract_Sigma(fit_B, level = "unit", part = "shared")$Sigma |> round(2)

# psi -- trait-specific unique variances (the "unique" component)
extract_Sigma(fit_B, level = "unit", part = "unique")$s |> round(2)

# Sigma_unit = Lambda Lambda^T + Psi (the "total" -- what you usually want)
extract_Sigma(fit_B, level = "unit", part = "total")$Sigma |> round(2)
```

```{r sigma-table-plot, fig.width = 7, fig.height = 3.8, fig.cap = "Upper-triangle off-diagonal entries of Sigma_unit from the latent + unique model. Open points mark point estimates without finite interval bounds; bootstrap-derived rows are the next step for Sigma uncertainty."}
plot_Sigma_table(sigma_rows_B, sort = "magnitude")
```

## Communality and ICC need the full $\Sigma$ too

Two trait-level summaries used in behavioural-syndromes and
phenotypic-integration workflows both depend
on the *full* $\boldsymbol\Sigma = \boldsymbol\Lambda \boldsymbol\Lambda^{\!\top} + \boldsymbol\Psi$
decomposition. If you compute them from a latent-only fit you get the
wrong answer.

### Communality

$$
c_t^2 \;=\; \frac{(\boldsymbol\Lambda \boldsymbol\Lambda^{\!\top})_{tt}}{\boldsymbol\Sigma_{tt}} \;=\; \frac{\sum_{\ell=1}^d \Lambda_{t\ell}^2}{\sum_{\ell=1}^d \Lambda_{t\ell}^2 + \psi_{tt}}.
$$

This is the proportion of trait $t$'s variance that is *shared* with
the other traits via the latent factors. It is bounded between 0 and 1
and is a proportion-of-variance summary, analogous in scale to $h^2$
(heritability) or $R^2$.

For a reader-facing table, the alignment is:

| Symbol | R output | Interpretation |
|---|---|---|
| $c_t^2$ | `extract_communality(fit, level = "unit")` | Fraction of trait `t` variance explained by the shared latent axes. |
| numerator | `extract_Sigma(fit, level = "unit", part = "shared")` diagonal | Shared variance for trait `t`. |
| denominator | `extract_Sigma(fit, level = "unit", part = "total")` diagonal | Shared variance plus trait-specific unique variance. |

**The catch**: with latent-only fits, $\psi_{tt} = 0$ for every trait by
construction, so $c_t^2 = 1$ for every trait — the communality is
identically 1 and tells you nothing. Adding `unique()` gives the
denominator a per-trait $\psi_{tt}$ slot, and communality becomes
informative.

```{r communality}
extract_communality(fit_A, level = "unit")  # latent-only fit: all = 1, useless
extract_communality(fit_B, level = "unit")  # latent + unique: meaningful
```

For the report surface, keep the exact table but compare the rows with
the correlation plot helper. The plot uses the same Fisher-z intervals as
`extract_correlations()`, so it shows uncertainty rather than only a
matrix cell value:

```{r communality-correlation-rows}
corr_B <- extract_correlations(fit_B, tier = "unit")
corr_B
```

```{r communality-correlation-plot, fig.width = 7, fig.height = 3.8, fig.cap = "Pairwise correlations from the latent + unique model. Points are Fisher-z estimates; horizontal segments show finite 95% interval bounds."}
plot_correlations(corr_B, sort = "magnitude")
```

### Site-level / individual-level ICC

For two-level fits the site-level (individual-level) ICC is

$$
R_t \;=\; \frac{(\boldsymbol\Sigma_{\text{unit}})_{tt}}{(\boldsymbol\Sigma_{\text{unit}})_{tt} + (\boldsymbol\Sigma_{\text{unit\_obs}})_{tt}},
$$

i.e. the proportion of total trait variance attributable to between-
unit differences. Each piece needs the full decomposition for its
level — $\Sigma_{\text{unit}} = \Lambda_{\text{unit}} \Lambda_{\text{unit}}^{\!\top} + \boldsymbol\Psi_{\text{unit}}$ and
$\Sigma_{\text{unit\_obs}} = \Lambda_{\text{unit\_obs}} \Lambda_{\text{unit\_obs}}^{\!\top} + \boldsymbol\Psi_{\text{unit\_obs}}$. Both `unique()` terms
are required (one per level) for $R_t$ to be honest.

In gllvmTMB this is

```r
extract_ICC_site(fit)
```

— see the recommended **two-`unique()`** pattern below.

## Binomial responses: the link's implicit residual

Binary outcomes (`family = binomial()`) are a special case. The
link function fixes a latent-scale residual variance: $\pi^2/3$
for `logit`, $1$ for `probit`, $\pi^2/6$ for `cloglog`. This
implicit residual already plays the role of $\boldsymbol\Psi$ on the
latent scale, so an explicit `unique()` term on top of a binary
fit is typically not identified.

`extract_Sigma()` has a `link_residual = "auto"` option that adds
the link-specific implicit residual to
$\mathrm{diag}(\boldsymbol\Sigma)$ for the marginal latent-scale
interpretation:

```{r binomial-demo, eval = FALSE}
# (illustrative — not run; needs a binary fit)
extract_Sigma(fit_binary, level = "unit", part = "total",
              link_residual = "auto")
```

For continuous traits, the recommended starting pattern is the paired
`latent() + unique()` decomposition. For count, mixed-family,
phylogenetic, and spatial fits, use the same idea only after checking
the corresponding validation-row status; those worked examples remain
under audit during the public-surface reset.

## Two-level (between + within) models: two `unique()` terms

For repeated-measures data the recommended pattern is two `latent() + unique()`
pairs (Nakagawa & Schielzeth 2010; Nakagawa, Johnson & Schielzeth 2017):

```r
fit_two_level <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | individual, d = d_B) +
    unique(0 + trait | individual) +
    latent(0 + trait | obs_id, d = d_W) +
    unique(0 + trait | obs_id),
  data = df,
  trait = "trait",
  unit = "individual",
  unit_obs = "obs_id"
)
```

giving
$\boldsymbol\Sigma_{\text{unit}} = \boldsymbol\Lambda_{\text{unit}} \boldsymbol\Lambda_{\text{unit}}^{\!\top} + \boldsymbol\Psi_{\text{unit}}$
(behavioural syndromes — between-individual covariance) and
$\boldsymbol\Sigma_{\text{unit\_obs}} = \boldsymbol\Lambda_{\text{unit\_obs}} \boldsymbol\Lambda_{\text{unit\_obs}}^{\!\top} + \boldsymbol\Psi_{\text{unit\_obs}}$
(integrated plasticity — within-individual covariance).
**Each level has its own `unique()`**.

## Observation-level random effects for non-Gaussian fits

For overdispersed Poisson and similar count families, the latent-scale
"unique component" is recovered by adding an observation-level random
effect (OLRE; Nakagawa & Schielzeth 2010):

$$
\ln \lambda_{ij} = \eta_{ij} + e_{ij}, \qquad e_{ij} \sim \mathcal N(0, \sigma_e^2).
$$

The $\sigma_e^2$ on the latent log scale plays the role of $\boldsymbol\Psi$.
Native OLRE support is available: add an `obs_id` column (one level per
row), include `unique(0 + trait | obs_id)` in the formula, pass
`unit_obs = "obs_id"`, and use `extract_residual_split(fit)` to separate
$\sigma^2_e$ (the estimated OLRE variance) from $\sigma^2_d$ (the
distribution-specific latent residual; Nakagawa & Schielzeth 2010;
Nakagawa, Johnson & Schielzeth 2017).

## Summary

| You want… | You need | Notes |
|---|---|---|
| Cross-trait correlations on a Gaussian / lognormal / Gamma fit | `latent() + unique()` at every level | Otherwise correlations are inflated. |
| Correlations on a binary fit | `latent()` only; `link_residual = "auto"` for marginal scale | Implicit residual depends on link (π²/3, 1, π²/6). |
| Phylogenetic decomposition | `phylo_latent() + phylo_unique()` for the phylogenetic tier; add `latent() + unique()` for the non-phylogenetic species tier | $\Omega = \Sigma_\text{phy} + \Sigma_\text{non}$, with each $\Sigma$ using shared + `diag(psi)` pieces. |
| Communality | `extract_communality(fit)` + the right latent/unique pattern | Communality formula uses Σ; needs full Σ to be meaningful. |
| `Lambda Lambda^T` only | `extract_Sigma(fit, part = "shared")` | Useful for ordination. |
| `Psi` / `psi` only | `extract_Sigma(fit, part = "unique")` | Trait-specific unique variances as a vector. |

## See also

* [Pitfalls](pitfalls.html) — common mistakes including correlation inflation
* [`?unique`](../reference/unique.html) — the formula keyword for the unique diagonal
* [`?extract_Sigma`](../reference/extract_Sigma.html) — the unified extractor
* [`?suggest_lambda_constraint`](../reference/suggest_lambda_constraint.html) — a helper for confirmatory loading structures
* [Formula keyword grid](api-keyword-grid.html) — syntax map for moving from this decomposition to other covariance rows
* **Nakagawa & Schielzeth (2010)** Repeatability for Gaussian and non-Gaussian data: a practical guide for biologists. *Biological Reviews* 85(4):935–956. <https://doi.org/10.1111/j.1469-185X.2010.00141.x>
* **Westneat, Wright & Dingemanse** (2015) The biology hidden inside residual within-individual phenotypic variation. *Biological Reviews* **90**, 729–743. <https://doi.org/10.1111/brv.12131>
