---
title: "Covariance and correlation: when you need `unique()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Covariance and correlation: when you need `unique()`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

> **Preview — Gaussian validated; mixed-family `extract_correlations()` is M1 milestone work.**
>
> The Σ vs correlation decomposition shown here is `covered` on
> Gaussian fits (EXT-01 + EXT-04 Fisher-z/Wald `covered`).
> `extract_correlations()` profile + bootstrap on mixed-family
> fits walks to `covered` after the **M1** milestone closes. See
> the [validation-debt register](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/35-validation-debt-register.md).

Suppose you want between-individual correlations among behaviours such as
boldness, exploration, activity, aggression, and sociability. If you fit only
`latent()`, the recovered correlations can be too large because the model has
no trait-specific variance term. Pair `latent()` with `unique()` when
trait-specific variance should stay in the denominator of the correlation
matrix.

```{r early-fit-pattern, eval = FALSE}
fit_latent_only <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | individual, d = 2),
  data = df, unit = "individual"
)

fit_latent_unique <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | individual, d = 2) +
    unique(0 + trait | individual),
  data = df, unit = "individual"
)
```

If your data are wide, with one row per individual and one column per
behaviour, use the same `gllvmTMB()` entry point with `traits(...)` on the
left-hand side:

```{r early-fit-pattern-wide, eval = FALSE}
fit_latent_unique <- gllvmTMB(
  traits(boldness, exploration, activity, aggression, sociability) ~
    1 + latent(1 | individual, d = 2) + unique(1 | individual),
  data = df_wide, unit = "individual"
)
```

## The decomposition

For the `latent() + unique()` model, the implied trait covariance is

$$
\boldsymbol\Sigma_\text{level} \;=\; \underbrace{\boldsymbol\Lambda_\text{level} \boldsymbol\Lambda_\text{level}^{\!\top}}_{\text{shared low-rank}} \;+\; \underbrace{\boldsymbol\Psi_\text{level}}_{\text{trait-specific diagonal}},
$$

where `level` names the covariance tier being extracted, such as `unit`
for between-individual covariance or `unit_obs` for within-individual or
observation-level covariance. $\boldsymbol\Lambda$ is the $T \times K$
loading matrix (from the `latent()` term) and $\boldsymbol\Psi$ is a
$T \times T$ diagonal matrix of trait-specific unique variances (from the
`unique()` term). The correlation matrix is then

$$
\mathbf R \;=\; \mathbf D^{-1/2}\,\boldsymbol\Sigma\,\mathbf D^{-1/2},
\qquad \mathbf D \;=\; \mathrm{diag}(\boldsymbol\Sigma).
$$

Both pieces matter: the diagonal of $\boldsymbol\Sigma$ is
$(\boldsymbol\Lambda \boldsymbol\Lambda^{\!\top})_{tt} + \psi_{tt}$ — the
sum of squared loadings *plus* the trait-specific unique variance.
Drop $\boldsymbol\Psi$ and the diagonal shrinks; the off-diagonals stay the
same; $\mathbf R$ swells.

## A side-by-side demonstration

Simulate one Gaussian dataset where the true between-individual
covariance has both a shared low-rank component and a trait-specific
unique diagonal, then
fit it two ways:

* `Model A`: `latent(0 + trait | individual, d = 2)` only
* `Model B`: `latent(0 + trait | individual, d = 2) + unique(0 + trait | individual)`

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

```{r sim}
set.seed(2026)
n_ind <- 200
trait_names <- c("boldness", "exploration", "activity",
                 "aggression", "sociability")
T <- length(trait_names)

## True between-individual covariance Sigma_unit = Lambda Lambda^T + diag(psi)
Lambda_true <- matrix(c(
  0.9,  0.1,
  1.0,  0.2,
  0.8, -0.4,
  0.7, -0.5,
  0.6,  0.6
), nrow = T, ncol = 2, byrow = TRUE,
   dimnames = list(trait_names, c("syndrome_1", "syndrome_2")))
S_true     <- c(0.30, 0.20, 0.40, 0.25, 0.35)
Sigma_true <- Lambda_true %*% t(Lambda_true) + diag(S_true)
sigma2_eps_true <- 0.05  # tiny observation-level noise

u <- matrix(rnorm(n_ind * 2), n_ind, 2)
e <- matrix(rnorm(n_ind * T, sd = sqrt(S_true[col(matrix(0, n_ind, T))])),
            n_ind, T)
y <- u %*% t(Lambda_true) + e +
     matrix(rnorm(n_ind * T, sd = sqrt(sigma2_eps_true)), n_ind, T)

df <- data.frame(
  individual = factor(rep(seq_len(n_ind), each = T)),
  trait      = factor(rep(trait_names, n_ind), levels = trait_names),
  value      = as.vector(t(y))
)
```

```{r fit}
fit_A <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | individual, d = 2),
  data = df, unit = "individual"
)

fit_B <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | individual, d = 2) +
          unique(0 + trait | individual),
  data = df, unit = "individual"
)
```

Now extract the implied $\hat{\boldsymbol\Sigma}_{\text{unit}}$ and
between-individual correlation matrix from each fit:

```{r extract}
ext_A <- suppressMessages(
  extract_Sigma(fit_A, level = "unit", part = "total")
)
ext_B <- extract_Sigma(fit_B, level = "unit", part = "total")
```

Notice that the call on `fit_A` emits a one-shot **advisory**
explaining that without `unique()` the returned $\hat{\boldsymbol\Sigma}$
is latent-only (we suppressed the message above for clean output but the
message is also stored in `ext_A$note`).

```{r show-note}
ext_A$note  # advisory: diag missing
```

## The correlations side by side

```{r corr-comparison, fig.width = 8.5, fig.height = 4.5, fig.cap = "True (left) vs. recovered correlations from Model A (latent only) and Model B (latent + unique). Model A's off-diagonals are all systematically larger than truth — that's the inflation unique() prevents."}
make_long <- function(R, label) {
  data.frame(
    trait_i = factor(rep(rownames(R), ncol(R)), levels = rownames(R)),
    trait_j = factor(rep(colnames(R), each = nrow(R)), levels = colnames(R)),
    rho     = as.vector(R),
    panel   = label
  )
}
R_true_named <- cov2cor(Sigma_true)
rownames(R_true_named) <- colnames(R_true_named) <- trait_names
df_corr <- rbind(
  make_long(R_true_named,   "1. Truth"),
  make_long(ext_A$R,        "2. Model A (latent only)"),
  make_long(ext_B$R,        "3. Model B (latent + unique)")
)
df_corr$panel <- factor(df_corr$panel, levels = c(
  "1. Truth", "2. Model A (latent only)", "3. Model B (latent + unique)"))

ggplot(df_corr, aes(trait_j, trait_i, fill = rho)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 3) +
  scale_fill_gradient2(low = "#3b82c6", mid = "white", high = "#d33",
                       midpoint = 0, limits = c(-1, 1)) +
  scale_y_discrete(limits = rev) +
  facet_wrap(~ panel) +
  labs(x = NULL, y = NULL, fill = expression(rho)) +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(),
        strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 35, hjust = 1))
```

The middle panel (`Model A`) shows correlations that are
**uniformly larger** than truth (left panel). The right panel
(`Model B`) recovers the true correlations to within sampling noise.
The mechanism: Model A's $\hat{\boldsymbol\Sigma}$ has small diagonals
(only $\Lambda \Lambda^{\!\top}$), so when we divide by $\sqrt{D}$ the
off-diagonals get inflated. Model B's $\hat{\boldsymbol\Sigma}$ has
the full $\Lambda \Lambda^{\!\top} + \boldsymbol\Psi$ — diagonals at the right scale
and correlations at the right magnitude.

## What `extract_Sigma()` gives you

The unified extractor exposes the three parts of the decomposition
explicitly:

```{r parts}
# Lambda Lambda^T alone (the "shared" component)
extract_Sigma(fit_B, level = "unit", part = "shared")$Sigma |> round(2)

# diag(s_unit) — trait-specific unique variances (the "unique" component)
extract_Sigma(fit_B, level = "unit", part = "unique")$s    |> round(2)

# Sigma_unit = Lambda Lambda^T + diag(s_unit)  (the "total" — what you usually want)
extract_Sigma(fit_B, level = "unit", part = "total")$Sigma |> round(2)
```

`part = "total"` is the default and is what you almost always want for
reporting. `"shared"` is useful if you specifically want the latent-implied
component (e.g. for ordination interpretation or communality). `"unique"`
gives the diagonal of $\boldsymbol\Psi$ as a named numeric vector.

## Communality and ICC need the full $\Sigma$ too

Two trait-level summaries used in behavioural-syndromes and
phenotypic-integration workflows both depend
on the *full* $\boldsymbol\Sigma = \boldsymbol\Lambda \boldsymbol\Lambda^{\!\top} + \boldsymbol\Psi$
decomposition. If you compute them from a latent-only fit you get the
wrong answer.

### Communality

$$
c_t^2 \;=\; \frac{(\boldsymbol\Lambda \boldsymbol\Lambda^{\!\top})_{tt}}{\boldsymbol\Sigma_{tt}} \;=\; \frac{\sum_{\ell=1}^d \Lambda_{t\ell}^2}{\sum_{\ell=1}^d \Lambda_{t\ell}^2 + \psi_{tt}}.
$$

This is the proportion of trait $t$'s variance that is *shared* with
the other traits via the latent factors. It is bounded between 0 and 1
and is a proportion-of-variance summary, analogous in scale to $h^2$
(heritability) or $R^2$.

**The catch**: with latent-only fits, $\psi_{tt} = 0$ for every trait by
construction, so $c_t^2 = 1$ for every trait — the communality is
identically 1 and tells you nothing. Adding `unique()` gives the
denominator a per-trait $\psi_{tt}$ slot, and communality becomes
informative.

```{r communality}
extract_communality(fit_A, level = "unit")  # latent-only fit: all = 1, useless
extract_communality(fit_B, level = "unit")  # latent + unique: meaningful
```

### Site-level / individual-level ICC

For two-level fits the site-level (individual-level) ICC is

$$
R_t \;=\; \frac{(\boldsymbol\Sigma_{\text{unit}})_{tt}}{(\boldsymbol\Sigma_{\text{unit}})_{tt} + (\boldsymbol\Sigma_{\text{unit\_obs}})_{tt}},
$$

i.e. the proportion of total trait variance attributable to between-
unit differences. Each piece needs the full decomposition for its
level — $\Sigma_{\text{unit}} = \Lambda_{\text{unit}} \Lambda_{\text{unit}}^{\!\top} + \boldsymbol\Psi_{\text{unit}}$ and
$\Sigma_{\text{unit\_obs}} = \Lambda_{\text{unit\_obs}} \Lambda_{\text{unit\_obs}}^{\!\top} + \boldsymbol\Psi_{\text{unit\_obs}}$. Both `unique()` terms
are required (one per level) for $R_t$ to be honest.

In gllvmTMB this is

```r
extract_ICC_site(fit)
```

— see the recommended **two-`unique()`** pattern below.

## Binomial responses: the link's implicit residual

Binary outcomes (`family = binomial()`) are a special case. The
link function fixes a latent-scale residual variance: $\pi^2/3$
for `logit`, $1$ for `probit`, $\pi^2/6$ for `cloglog`. This
implicit residual already plays the role of $\boldsymbol\Psi$ on the
latent scale, so an explicit `unique()` term on top of a binary
fit is typically not identified.

`extract_Sigma()` has a `link_residual = "auto"` option that adds
the link-specific implicit residual to
$\mathrm{diag}(\boldsymbol\Sigma)$ for the marginal latent-scale
interpretation:

```{r binomial-demo, eval = FALSE}
# (illustrative — not run; needs a binary fit)
extract_Sigma(fit_binary, level = "unit", part = "total",
              link_residual = "auto")
```

For everything else — continuous traits, count traits, mixed
families — the recommended pattern is the paired `latent() +
unique()` decomposition. The phylogenetic case extends this to a
paired four-component decomposition (`phylo_latent +
phylo_unique` on the phy side, `latent + unique` on the non-phy
side); see the [Phylogenetic GLLVM article](phylogenetic-gllvm.html)
for the worked example.

## Two-level (between + within) models: two `unique()` terms

For repeated-measures data the recommended pattern is two `latent() + unique()`
pairs (Nakagawa et al. *in prep*):

```r
fit_two_level <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | individual, d = d_B) +
    unique(0 + trait | individual) +
    latent(0 + trait | obs_id, d = d_W) +
    unique(0 + trait | obs_id),
  data = df,
  unit = "individual",
  unit_obs = "obs_id"
)
```

giving
$\boldsymbol\Sigma_{\text{unit}} = \boldsymbol\Lambda_{\text{unit}} \boldsymbol\Lambda_{\text{unit}}^{\!\top} + \boldsymbol\Psi_{\text{unit}}$
(behavioural syndromes — between-individual covariance) and
$\boldsymbol\Sigma_{\text{unit\_obs}} = \boldsymbol\Lambda_{\text{unit\_obs}} \boldsymbol\Lambda_{\text{unit\_obs}}^{\!\top} + \boldsymbol\Psi_{\text{unit\_obs}}$
(integrated plasticity — within-individual covariance).
**Each level has its own `unique()`**.

## Observation-level random effects for non-Gaussian fits

For overdispersed Poisson and similar count families, the latent-scale
"unique component" is recovered by adding an observation-level random
effect (OLRE; Nakagawa & Schielzeth 2010):

$$
\ln \lambda_{ij} = \eta_{ij} + e_{ij}, \qquad e_{ij} \sim \mathcal N(0, \sigma_e^2).
$$

The $\sigma_e^2$ on the latent log scale plays the role of $\boldsymbol\Psi$.
Native OLRE support is available: add an `obs_id` column (one level per
row), include `unique(0 + trait | obs_id)` in the formula, pass
`unit_obs = "obs_id"`, and use `extract_residual_split(fit)` to separate
$\sigma^2_e$ (the estimated OLRE variance) from $\sigma^2_d$ (the
distribution-specific latent residual; Nakagawa & Schielzeth 2010;
Nakagawa, Johnson & Schielzeth 2017).

## Summary

| You want… | You need | Notes |
|---|---|---|
| Cross-trait correlations on a Gaussian / lognormal / Gamma fit | `latent() + unique()` at every level | Otherwise correlations are inflated. |
| Correlations on a binary fit | `latent()` only; `link_residual = "auto"` for marginal scale | Implicit residual depends on link (π²/3, 1, π²/6). |
| Phylogenetic decomposition | `phylo_latent() + phylo_unique()` for the phylogenetic tier; add `latent() + unique()` for the non-phylogenetic species tier | $\Omega = \Sigma_\text{phy} + \Sigma_\text{non}$, with each $\Sigma$ using shared + `diag(psi)` pieces. |
| Communality | `extract_communality(fit)` + the right latent/unique pattern | Communality formula uses Σ; needs full Σ to be meaningful. |
| `Lambda Lambda^T` only | `extract_Sigma(fit, part = "shared")` | Useful for ordination. |
| `S` only | `extract_Sigma(fit, part = "unique")` | Trait-specific unique variances as a vector. |

## See also

* [Pitfalls](pitfalls.html) — common mistakes including correlation inflation
* [`?unique`](../reference/unique.html) — the formula keyword for the unique diagonal
* [`?extract_Sigma`](../reference/extract_Sigma.html) — the unified extractor
* [`?suggest_lambda_constraint`](../reference/suggest_lambda_constraint.html) — a helper for confirmatory loading structures
* [Choose your model](choose-your-model.html) — decision tree
* **Nakagawa & Schielzeth (2010)** Repeatability for Gaussian and non-Gaussian data: a practical guide for biologists. *Biological Reviews* 85(4):935–956. <https://doi.org/10.1111/j.1469-185X.2010.00141.x>
* **Nakagawa et al.** (*in prep*) Quantifying between- and within-individual correlations and the degree of trait integration.
