---
title: "Cross-family trait correlations, including nominal responses"
description: "Put a Gaussian, binary, count, ordinal, and nominal trait on one shared latent factor, recover their cross-family correlations, and read the reference-invariant summary for the categorical trait."
tier: 2 # worked demonstration of the multi-family latent factor and the 2C nominal summary
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Cross-family trait correlations, including nominal responses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

Traits do not arrive on one scale. A single organism carries continuous
measurements, presence/absence marks, counts, ordered scores, and unordered
categories at once; a single site records abundance, occupancy, and a dominant
class. The question this article answers is narrow and concrete: can a
**Gaussian, a binary, a count, an ordinal, and a nominal trait** sit on **one
shared latent factor** in a single `gllvmTMB()` fit, and can the model report
the correlations *between* those families — including the categorical one?

The answer is yes, with caveats that matter, and this page shows both. It is a
young capability (0.6), demonstrated on a single powered simulation. Read the
recovered numbers as evidence that the estimator tracks the truth in a favourable
regime, not as calibrated intervals.

## Why put different families on one latent scale

The motivation is the same one that drives joint species distribution models and
phenotypic-integration analyses: the interesting quantity is not each response
on its own but how they **covary**. If a species that is more abundant is also
more likely to be present, and its dominant habitat class shifts with both, that
joint structure is the signal. A shared latent factor $\mathbf u_i$ is the
device that couples the families: every response loads on the same axes, so a
correlation *between families* is well-defined even though the families live on
different observation scales.

The alternative — fit each response separately and correlate the point
predictions — discards the measurement model and gives correlations that are
biased by each family's link and dispersion. Putting the families on one latent
scale keeps the coupling in the model where it belongs.

## The model

For unit $i$ and pseudo-trait $t$, a $d$-dimensional shared factor
$\mathbf u_i \sim \mathcal N(\mathbf 0, \mathbf I_d)$ drives every family through
its own linear predictor and link:

$$
\eta_{it} = \mu_t + \boldsymbol\lambda_t^{\!\top}\mathbf u_i,
\qquad
y_{it} \mid \mathbf u_i \sim \text{family}_t(\,g_t^{-1}(\eta_{it})\,).
$$

Gaussian, binomial, and Poisson traits each occupy a single row of the loading
matrix $\boldsymbol\Lambda$. An `ordinal_probit()` trait occupies one row and
crosses a set of ordered thresholds. A `multinomial()` trait is different: an
unordered response with $K$ categories is not one latent dimension but
$K-1$ of them — the baseline-category contrasts — so it expands into $K-1$
rows of $\boldsymbol\Lambda$, named `"<trait>:<cat>"`. The shared low-rank
covariance among all pseudo-traits is
$\boldsymbol\Lambda\boldsymbol\Lambda^{\!\top}$, and the cross-family
correlations are its standardised off-diagonals — rotation-invariant, and the
quantity to report.

In this demonstration the nominal trait is fitted **loadings-only**
(`latent(..., unique = FALSE)`): the $K-1$ contrast rows share the factor but
carry no per-contrast diagonal $\boldsymbol\Psi$. A unique/Psi tier on a
categorical response is deferred (see the caveats).

## A five-family worked example

We simulate `r 500` units, each observed `r 6` times, with five traits sharing
one 3-dimensional factor:

| trait | family | value column holds |
|---|---|---|
| `g` | `gaussian()` | a real number |
| `b` | `binomial()` | 0 / 1 |
| `p` | `poisson()` | a count |
| `o` | `ordinal_probit()` | an ordered code `1..Ko` |
| `cat` | `multinomial()` | a category code `1..K` |

The truth is a single loading matrix over the six pseudo-traits
`g, b, p, o, cat:2, cat:3` (the nominal trait `cat` contributes the last two
rows). Everything the model must recover is contained in
$\boldsymbol\Lambda\boldsymbol\Lambda^{\!\top}$.

```{r setup}
library(gllvmTMB)
```

### Simulate

```{r simulate}
set.seed(20260718)

N     <- 500L   # units
reps  <- 6L     # observations per unit (replication sharpens recovery)
d     <- 3L     # shared latent dimensions
units <- seq_len(N)

# True loadings for the six pseudo-traits on the three shared axes.
# The nominal trait `cat` (K = 3) contributes the rows cat:2 and cat:3.
Lambda <- rbind(
  g       = c( 1.2,  0.2,  0.0),
  b       = c( 0.9,  0.6,  0.1),
  p       = c( 0.3,  1.0,  0.4),
  o       = c( 0.7,  0.4,  0.9),
  `cat:2` = c( 1.0,  0.5,  0.2),
  `cat:3` = c(-0.5,  0.8,  0.6)
)

# Trait means / baseline intercepts (ordinal handled by thresholds instead).
mu  <- c(g = 0, b = 0, p = log(2), o = 0, `cat:2` = 0, `cat:3` = 0)
tau <- c(-1.6, -0.2, 1.2)   # 3 thresholds -> Ko = 4 ordered categories

# One latent position per unit, shared across that unit's replicates.
U   <- matrix(rnorm(N * d), N, d)
Eta <- U %*% t(Lambda)                 # N x 6 unit-level linear predictors
colnames(Eta) <- rownames(Lambda)

sim_one_rep <- function(rep_id) {
  g_val <- mu["g"] + Eta[, "g"] + rnorm(N)                      # Gaussian
  b_val <- rbinom(N, 1L, plogis(mu["b"] + Eta[, "b"]))         # binary
  p_val <- rpois(N, exp(mu["p"] + Eta[, "p"]))                  # count
  o_val <- findInterval(mu["o"] + Eta[, "o"] + rnorm(N), tau) + 1L  # ordinal
  e2 <- mu["cat:2"] + Eta[, "cat:2"]                            # nominal
  e3 <- mu["cat:3"] + Eta[, "cat:3"]                            #   contrasts
  P  <- cbind(1, exp(e2), exp(e3)); P <- P / rowSums(P)         # softmax
  cat_val <- apply(P, 1L, function(pr) sample.int(3L, 1L, prob = pr))
  rbind(
    data.frame(unit = units, trait = "g",   family = "g", value = g_val),
    data.frame(unit = units, trait = "b",   family = "b", value = b_val),
    data.frame(unit = units, trait = "p",   family = "p", value = p_val),
    data.frame(unit = units, trait = "o",   family = "o", value = o_val),
    data.frame(unit = units, trait = "cat", family = "m", value = cat_val)
  )
}

dat <- do.call(rbind, lapply(seq_len(reps), sim_one_rep))
dat$unit   <- factor(dat$unit)
dat$trait  <- factor(dat$trait,  levels = c("g", "b", "p", "o", "cat"))
dat$family <- factor(dat$family, levels = c("g", "b", "p", "o", "m"))

table(dat$trait)
```

The `family` column maps each row to its response family; the nominal trait is
the single column `cat`, whose values are category codes `1..K`. It expands into
the pseudo-traits `cat:2` and `cat:3` inside the fit, not in the data.

### Fit

One `gllvmTMB()` call carries all five families on the shared factor. The
`family` argument is a named list keyed by the `family` column's levels, and
`attr(family, "family_var")` names that column.

```{r fit}
family <- list(
  g = gaussian(), b = binomial(), p = poisson(),
  o = ordinal_probit(), m = multinomial()
)
attr(family, "family_var") <- "family"

fit <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | unit, d = 3, unique = FALSE),
  data   = dat,
  family = family,
  trait  = "trait",
  unit   = "unit",
  silent = TRUE
)

c(convergence = fit$opt$convergence, pdHess = isTRUE(fit$sd_report$pdHess))
```

### Recover the cross-family correlations

`extract_Sigma()` with `part = "shared"` returns the latent low-rank covariance
$\boldsymbol\Lambda\boldsymbol\Lambda^{\!\top}$ over all six pseudo-traits. We
standardise it and compare against the truth built from `Lambda`.

```{r recover}
Sigma_true <- Lambda %*% t(Lambda)
R_true <- cov2cor(Sigma_true)

Sig_hat <- extract_Sigma(fit, level = "unit", part = "shared", link_residual = "none")
R_hat   <- cov2cor(Sig_hat$Sigma)

# align the recovered matrix to the truth's ordering by name
ord   <- rownames(R_true)
R_hat <- R_hat[ord, ord]

pairs <- rbind(
  c("g", "b"), c("g", "p"), c("g", "o"),
  c("g", "cat:2"), c("g", "cat:3"),
  c("b", "cat:2"), c("p", "cat:3"), c("o", "cat:2")
)

data.frame(
  pair      = paste(pairs[, 1], pairs[, 2], sep = " ~ "),
  true      = round(mapply(function(i, j) R_true[i, j], pairs[, 1], pairs[, 2]), 3),
  recovered = round(mapply(function(i, j) R_hat[i, j], pairs[, 1], pairs[, 2]), 3),
  row.names = NULL
)
```

The recovered correlations track the truth across all five families,
including the two nominal contrasts `cat:2` and `cat:3` and their sign — a
strong positive `g ~ cat:2`, a negative `g ~ cat:3`. The remaining error is
sampling noise at this `N`; larger `N` (or more replicates per unit) shrinks it.
Every off-diagonal pair, not just the eight highlighted here, is available in
`R_hat`.

```{r recover-plot, fig.width = 5.2, fig.height = 5.2, fig.cap = "True versus recovered latent cross-family correlations, all off-diagonal pairs. The line is the identity; points near it indicate faithful recovery in this powered run."}
offd <- upper.tri(R_true)
plot(R_true[offd], R_hat[offd],
     xlab = "true correlation", ylab = "recovered correlation",
     xlim = c(-1, 1), ylim = c(-1, 1), pch = 19, col = "#3366aa")
abline(0, 1, lty = 2, col = "grey40")
```

## Summarising the nominal trait: the 2C multiple correlation

The pairwise table above lists a *separate* correlation for each nominal
contrast (`cat:2`, `cat:3`). Often a reader wants **one number** for "how
strongly is the categorical trait associated with each partner?" that does not
depend on which category was chosen as the reference. That is what
`extract_cross_correlations()` reports (reporting decision 2C): a
**reference-invariant multiple correlation** between the categorical trait and
each single-scale partner, optionally with the underlying $(K-1)$-vector.

```{r cross-cor}
xc <- extract_cross_correlations(fit, level = "unit", link_residual = "auto")
xc
```

The single-number summary is reported on the **observation scale**, via the
softmax link residual. This is the deliberate default, and the reason is in the
next section.

## Honest caveats

These are limits of the capability as it stands at 0.6, not incidental notes.

- **The nominal trait is loadings-only here.** It is fitted with
  `latent(..., unique = FALSE)`, so the $K-1$ contrast rows share the factor but
  carry no per-contrast diagonal $\boldsymbol\Psi$. A unique/Psi tier on a
  categorical response is deferred; the pairwise latent correlations (the matrix)
  are the recovered quantities.

- **The single-number summary belongs on the observation scale.** The
  reference-invariant multiple correlation is computed on the observation scale
  using the softmax's link residual $(\pi^2/6)(\mathbf I + \mathbf J)$ — the
  covariance of the differenced Gumbel errors in the random-utility
  representation of the softmax (McFadden 1974), which reduces to the binomial's
  $\pi^2/3$ when $K = 2$. On the **latent** scale (`link_residual = "none"`) the
  multiple correlation degenerates toward 1 whenever the $K-1$ contrast block
  spans the partner, because a $(K-1)$-dimensional block can align with almost
  any single direction. That is why the observation scale is the sensible default
  for the one-number summary, while the per-pair latent correlations remain the
  right object for the matrix.

- **Recovering a nominal trait's covariance is data-hungry.** A single
  categorical draw carries little information about a $(K-1)$-dimensional
  liability. With one draw per unit the reduced-rank estimate can rail at
  $\pm 1$; this is an information limitation, not a bug. It resolves with
  replication (multiple draws per unit, as here) or many units. The powered
  regime in this article — `r 500` units $\times$ `r 6` replicates — is what
  makes the recovery above possible; do not expect the same at one observation
  per unit.

- **Intervals are not yet calibrated.** This is a 0.6 capability. Coverage of
  the cross-family correlations has not been certified; treat the recovered
  numbers and any interval outputs as recovery-oriented, not as calibrated
  confidence statements.

- **One multinomial trait per fit.** The current implementation supports a single
  `multinomial()` trait in a fit. Two nominal responses on one shared factor is
  not yet supported.

## See also

- [Unordered categories with `multinomial()`](multinomial.html) — the
  baseline-category model, reading contrasts and predicted probabilities, and the
  standalone among-category table in the phylogenetic setting.
- [Covariance and correlation: the model behind Sigma](covariance-correlation.html)
  — $\boldsymbol\Sigma$, $\boldsymbol\Lambda$, $\boldsymbol\Psi$, and the
  `link_residual` conventions used above.
- [`?extract_Sigma`](../reference/extract_Sigma.html) — the unified covariance
  extractor.
- [`?extract_cross_correlations`](../reference/extract_cross_correlations.html) —
  the reference-invariant nominal summary (decision 2C).
- **McFadden (1974)** Conditional logit analysis of qualitative choice behavior.
  In *Frontiers in Econometrics*, ed. P. Zarembka, pp. 105–142. Academic Press.
- **Nakagawa & Schielzeth (2010)** Repeatability for Gaussian and non-Gaussian
  data. *Biological Reviews* 85(4):935–956.
  <https://doi.org/10.1111/j.1469-185X.2010.00141.x>
