---
title: "Cross-lineage coevolution kernels"
tier: 3 # internal methods draft until COE-04 and interval/rho gates close
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Cross-lineage coevolution kernels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Build policy for this article:
# - Light chunks (kernel construction, fixture inspection, the Gamma heatmap)
#   run at build time; they touch only matrices and the portable RDS fixture,
#   never a TMB fit.
# - Every heavy chunk -- the cross-kernel fits, the null fit, the rho grid
#   profile, the sparse-vs-dense sensitivity sweep -- is eval = FALSE and shows
#   the exact shipped syntax with precomputed / illustrative numbers.
#   Coevolution recovery fits are slow and have OOM'd pkgdown before; do not
#   flip these to eval = TRUE in the pkgdown build. The numbers quoted in the
#   prose are the planted truth from the fixture and representative recovery
#   values from tests/testthat/test-coevolution-recovery.R.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7.2, fig.height = 4.4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

This article is a worked methods example for a comparative biologist with
**two interacting lineages** -- hosts and parasites, plants and pollinators,
or any mutualist/antagonist pair. The estimand is the **coevolution matrix**
`Gamma`: a host-trait by partner-trait covariance block that says which host
traits move with which partner traits *after* both phylogenies and the
observed association structure are accounted for.

> **Internal article gate.** This page is a Tier 3 methods draft while the
> Paper 2 coevolution evidence remains fixed-rho, point-estimate, and
> diagnostic-only. It may be used to keep the syntax and estimand coherent, but
> it is not yet a release-ready public worked example.

The model is a stacked-trait phylogenetic GLLVM in which a single
**cross-clade kernel** `K*` carries the coevolution signal in its
off-diagonal block. As far as we have found, the one-matrix stacked-`Y`
PGLLVM with an association-derived off-diagonal block and the estimand
`Gamma = Lambda_H Lambda_P^T` is novel: HMSC (Ovaskainen et al. 2017) is the
closest joint model but does not enforce a cross-clade latent coupling, and
the dyadic phylogenetic mixed models of Rafferty & Ives (2013) and Hadfield &
Nakagawa (2010) model the *interaction strength* as the response rather than
the two lineages' *traits* jointly. We therefore present this as a methods
contribution and are deliberate about what the current engine does and does
not yet certify.

Scope boundary: **IN** (`COE-02`, `KER-02`), this article fits one named dense
kernel tier with `kernel_latent()` and extracts the cross-lineage shared block
with `extract_Gamma()`. **PARTIAL** (`COE-03`, `COE-04`), tests now cover
latent-only two-kernel fixtures and kernel-separability diagnostics, including
a raw-versus-residualized reciprocal-dependence screen for present-day/tip
matching. The article therefore describes the two-component Paper 2 estimand,
but still teaches only a fixed-kernel, point-estimate workflow. The current
Paper 2 first-wave syntax is latent-only; `kernel_unique()` remains
soft-deprecated compatibility syntax for older explicit-Psi fixtures and
equivalence tests. **PLANNED**, in-engine `rho` estimation, calibrated
intervals, module uncertainty, rank calibration, formal null thresholds, and
broader recovery grids remain later Design 65 work.

The target Paper 2 identity separates two cross-lineage covariance sources:

$$
\mathrm{Cov}(\eta_{H,ia}, \eta_{P,jb})
= C_{\mathrm{phy},ij}\Gamma_{\mathrm{phy},ab}
+ C_{\mathrm{tip},ij}\Gamma_{\mathrm{tip},ab}.
$$

`C_phy` is the phylogenetically structured association kernel: it asks whether
related hosts tend to match related partners in a conserved way. `C_tip` is a
present-day/tip-level matching kernel: it asks whether the realised network has
extra trait matching after the phylogenetic cross pattern has been accounted
for. When those two fixed kernels are too similar, the honest answer is one
network-conditioned covariance block, not two biological stories.

## Why coevolution lives in an off-diagonal block

Stack the two lineages into one response matrix. Hosts contribute their
traits in the host rows; partners contribute theirs in the partner rows; the
cross-lineage cells are structurally missing, because a host individual has no
partner traits and vice versa:

$$
Y^* =
\begin{bmatrix}
Y_H & \mathrm{NA} \\
\mathrm{NA} & Y_P
\end{bmatrix}.
$$

The model puts a single between-species covariance kernel on the stacked
rows:

$$
K^* =
\begin{bmatrix}
A_H & K_{HP} \\
K_{HP}^{T} & A_P
\end{bmatrix},
\qquad
\mathrm{Var}\{\mathrm{vec}(\eta^* - M^*)\}
= (\Lambda\Lambda^{T}) \otimes K^*.
$$

Older explicit-Psi compatibility fits add a diagonal `kernel_unique()` term at
the same named tier. The current article teaches the latent-only first wave,
because the Paper 2 multi-kernel path deliberately defers explicit
kernel-level `Psi`.

The two diagonal blocks `A_H` and `A_P` are the within-lineage phylogenetic
covariances: they describe how host species resemble each other, and how
partner species resemble each other, by descent alone. They cannot, by
construction, link a host trait to a partner trait. **All** of the
host-partner coupling is forced through the off-diagonal block `K_HP`. If
`K_HP = 0` the stacked model factorises into two independent single-lineage
PGLLVMs and no coevolution can be expressed.

Writing the latent loadings as host rows stacked over partner rows,
`Lambda = rbind(Lambda_H, Lambda_P)`, the implied covariance between a host
trait `a` (on species `i`) and a partner trait `b` (on species `j`) is

$$
\mathrm{Cov}(\eta_{H,ia}, \eta_{P,jb})
= \Gamma_{ab}\, K_{HP,ij},
\qquad
\Gamma = \Lambda_H \Lambda_P^{T}.
$$

So `K_HP` carries the *species-level* coupling -- which host species is
coupled to which partner species, driven by the observed associations -- and
`Gamma` carries the *trait-level* coupling, the host-trait by partner-trait
matrix we want to read. The two are separable because `K_HP` is supplied (it
is a property of the phylogenies and the association matrix `W`), so the fit
only has to learn the trait loadings.

This is the structural reason classical correlation-of-contrasts methods
struggle with interaction effects (Drury et al. 2016): they have no place to
put `K_HP`, and the within-lineage structure absorbs signal that actually
belongs to the cross block. The explicit `K*` makes the cross-block
contribution identifiable as a separate term.

## Build the cross-lineage kernel

`make_cross_kernel()` assembles `K*` from the host relatedness matrix `A_H`,
the partner relatedness matrix `A_P`, and the association matrix `W`. The
off-diagonal block is built as `K_HP = rho * L_H * W_scaled * L_P^T`, where
`L_H L_H^T = A_H`, `L_P L_P^T = A_P`, and `W` is spectrally scaled so that the
assembled `K*` stays positive semidefinite for `|rho| <= 1`. In this first
workflow `rho` is a **user-supplied** coupling strength rather than an
estimated parameter (see the `rho` profile section below).

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

```{r load-example, eval = FALSE}
coev <- readRDS(system.file(
  "extdata", "examples", "coevolution-kernel-example.rds",
  package = "gllvmTMB"
))
```

```{r load-example-hidden, include = FALSE}
example_path <- system.file(
  "extdata", "examples", "coevolution-kernel-example.rds",
  package = "gllvmTMB"
)
example_path <- c(
  example_path,
  file.path("inst", "extdata", "examples", "coevolution-kernel-example.rds"),
  file.path("..", "inst", "extdata", "examples", "coevolution-kernel-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "coevolution-kernel-example.rds")
)
example_path <- example_path[nzchar(example_path) & file.exists(example_path)][1]
stopifnot(!is.na(example_path))
coev <- readRDS(example_path)
```

The fixture has `r nrow(coev$A_H)` host species and `r nrow(coev$A_P)`
partner species, four traits, and `r coev$truth$n_rep` replicate measurements
per species. The wide table has one row per sampled individual; host rows
carry host traits and `NA` partner traits, and partner rows carry partner
traits and `NA` host traits:

```{r example-preview}
coev$story$question
head(coev$data_wide, 6)
```

The long table makes the block-missing cells explicit as one `(row_id, trait)`
row per observation. Both shapes fit the same model through the single
`gllvmTMB()` entry point:

```{r long-preview}
head(coev$data_long, 8)
```

Now assemble the cross kernel. The light checks confirm that the assembled
matrix is positive semidefinite and matches the kernel stored in the fixture:

```{r build-kernel}
K_star <- make_cross_kernel(
  coev$A_H,
  coev$A_P,
  coev$W,
  rho = coev$truth$rho
)

# PSD floor: the smallest eigenvalue must not be meaningfully negative.
range(eigen(K_star, symmetric = TRUE, only.values = TRUE)$values)

# Off-diagonal energy: this is the cross block that carries coevolution.
max(abs(K_star[rownames(coev$A_H), rownames(coev$A_P)]))

# Agreement with the kernel stored in the fixture.
max(abs(K_star - coev$K_star))
```

The **null** kernel keeps both within-lineage phylogenies but zeroes the
cross block. It is the model in which the same descent structure is allowed,
but no host-partner coupling is permitted -- the correct baseline for asking
whether the off-diagonal block is doing any work:

```{r null-kernel}
K_null <- coev$K_null
max(abs(K_null[rownames(coev$A_H), rownames(coev$A_P)]))
```

## Screen phylogenetic and tip kernels before splitting them

The two-component Paper 2 split only makes sense when the proposed
phylogenetic and tip kernels carry distinguishable cross-lineage information.
For an observed interaction-count matrix, a defensible first tip candidate is
the symmetric reciprocal-dependence surface

$$
W_{\mathrm{recip},ij} =
\sqrt{p(j \mid i)\,p(i \mid j)}.
$$

The current COE-04 gate treats raw `W_recip` as a sensitivity target, not a
promotion target. In the deterministic test fixture, the raw reciprocal tip
kernel has **moderate** overlap with the phylogenetic cross kernel and receives
the `sensitivity_required` recommendation. A residualized-plus-opposed tip
candidate is **near orthogonal** and receives `separable_candidate`.

```{r reciprocal-screen, eval = FALSE}
p_col_given_row <- sweep(counts, 1, rowSums(counts), "/")
p_row_given_col <- sweep(counts, 2, colSums(counts), "/")
W_recip <- sqrt(p_col_given_row * p_row_given_col)

K_phy <- make_cross_kernel(A_H, A_P, W_phy, rho = 0.55)
I_H <- diag(nrow(A_H), dimnames = dimnames(A_H))
I_P <- diag(nrow(A_P), dimnames = dimnames(A_P))
K_tip_raw <- make_cross_kernel(I_H, I_P, W_recip, rho = 0.55)

raw_dx <- diagnose_kernel_separability(
  phy = K_phy,
  tip_raw = K_tip_raw
)
raw_dx$pairs[, c("similarity", "overlap_class", "recommendation")]
#> moderate overlap; recommendation = "sensitivity_required"

W_tip_resid <- matrix(
  resid(lm(as.vector(W_recip) ~ as.vector(W_phy))),
  nrow(W_recip), ncol(W_recip),
  dimnames = dimnames(W_recip)
)
K_tip_resid <- make_cross_kernel(I_H, I_P, W_tip_resid, rho = 0.55)

resid_dx <- diagnose_kernel_separability(
  phy = K_phy,
  tip_resid = K_tip_resid
)
resid_dx$pairs[, c("similarity", "overlap_class", "recommendation")]
#> near_orthogonal; recommendation = "separable_candidate"
```

This screen does not prove causal coevolution, and it does not deliver
intervals. It decides which model is scientifically defensible before a fit:
collapse to one cross-lineage covariance when the kernels overlap too strongly,
run a sensitivity analysis in the moderate zone, and fit separate
`name = "phy"` / `name = "tip"` tiers only when the fixed kernels are
distinguishable enough for component-specific `Gamma` blocks to mean
something.

## The workflow at a glance

The full coevolution analysis is three steps:

1. **Build** `K* <- make_cross_kernel(A_H, A_P, W, rho)`.
2. **Fit** `kernel_latent(species, K = K_star, d = 2, name = "cross")`
   on the block-`NA` stacked `Y*`.
3. **Extract** `extract_Gamma(fit, level = "cross", row_traits, col_traits)`.
4. **Summarise modules** with `extract_coevolution_modules()` when a
   standardized host-axis / partner-axis decomposition is useful.

`kernel_latent()` estimates the shared reduced-rank `Lambda Lambda^T`
covariance attached to `K*`, and that shared component is the source of
`Gamma`. The name `"cross"` is the level that `extract_Sigma()`,
`extract_Gamma()`, and `extract_coevolution_modules()` read back. Explicit
`kernel_unique()` can still reproduce older Gaussian C2 fixtures as
compatibility syntax, but it is no longer the first-line coevolution teaching
path.

The wide `traits(...)` call names the four response columns on the
left-hand side:

```{r fit-wide, eval = FALSE}
ctrl <- gllvmTMBcontrol(se = TRUE)

fit_cross <- gllvmTMB(
  traits(h_size, h_defence, p_size, p_attack) ~ 1 +
    kernel_latent(species, K = K_star, d = 2, name = "cross"),
  data = coev$data_wide,
  unit = "row_id",
  cluster = "species",
  family = gaussian(),
  control = ctrl
)

fit_cross$opt$convergence       #> 0
fit_cross$fit_health$pd_hessian #> TRUE
```

The long-format call is the same model written against the canonical
`value ~ 0 + trait` response grammar. Use it when the data already arrive as
one `(unit, trait)` row per observation; the two fits agree up to optimiser
noise (the C2 fixture test asserts the wide/long `logLik` difference is below
`1e-6`):

```{r fit-long, eval = FALSE}
fit_cross_long <- gllvmTMB(
  value ~ 0 + trait +
    kernel_latent(species, K = K_star, d = 2, name = "cross"),
  data = coev$data_long,
  trait = "trait",
  unit = "row_id",
  cluster = "species",
  family = gaussian(),
  control = ctrl
)

abs(as.numeric(logLik(fit_cross)) -
    as.numeric(logLik(fit_cross_long)))
#> ~ 1e-9
```

## The null comparison: is the cross block improving fit?

The central diagnostic comparison is between the cross-kernel fit and the
block-diagonal null. Both fits use the *same* formula, the *same* rank `d = 2`,
and the *same* data; the only difference is whether the kernel has a
host-partner bridge. Because `rho` is supplied through `K*` rather than
estimated in TMB, treat the likelihood difference as a fixed-kernel
model-comparison diagnostic, not as a calibrated likelihood-ratio test.

```{r fit-null, eval = FALSE}
fit_null <- gllvmTMB(
  traits(h_size, h_defence, p_size, p_attack) ~ 1 +
    kernel_latent(species, K = K_null, d = 2, name = "cross"),
  data = coev$data_wide,
  unit = "row_id",
  cluster = "species",
  family = gaussian(),
  control = ctrl
)

c(
  cross       = as.numeric(logLik(fit_cross)),
  null        = as.numeric(logLik(fit_null)),
  improvement = as.numeric(logLik(fit_cross)) -
                as.numeric(logLik(fit_null))
)
#> cross   null   improvement
#> (the cross fit clears the null by several logLik units on K*-structured data)
```

Two facts make this a useful diagnostic rather than a tautology. First, the
null fit cannot express any host-partner covariance, so its `Gamma` block
collapses to zero (the recovery test asserts `max(abs(Gamma_null)) < 1e-8`).
Second, on data generated *with* a cross block, the cross-kernel fit improves
the likelihood by a margin the test gates at more than one logLik unit. If
your real data showed no improvement, that would be the honest signal that the
association matrix carries little trait coevolution -- the workflow is built
to let the null win when it should.

## Extract and read Gamma

`extract_Gamma()` slices the host-trait by partner-trait block of the fitted
shared covariance, `Gamma = Lambda_H Lambda_P^T`. The rows are host traits and
the columns are partner traits:

```{r extract-gamma, eval = FALSE}
Gamma_hat <- extract_Gamma(
  fit_cross,
  level = "cross",
  row_traits = coev$truth$host_traits,
  col_traits = coev$truth$partner_traits
)

Gamma_null <- extract_Gamma(
  fit_null,
  level = "cross",
  row_traits = coev$truth$host_traits,
  col_traits = coev$truth$partner_traits
)

# Sign-tolerant agreement with the planted truth (the test gates this > 0.9):
abs(cor(as.vector(Gamma_hat), as.vector(coev$truth$Gamma)))
#> ~ 0.97
```

The planted truth in this fixture is below. Because `Gamma` is a covariance
block, both sign and scale carry meaning: a positive cell means the host and
partner traits rise together along the cross-kernel latent axes, and a
negative cell means they move in opposite directions.

```{r truth-gamma}
round(coev$truth$Gamma, 2)
```

A worked reading of these cells:

- **`h_size` x `p_size`** is positive: large-bodied host species are coupled,
  through the association structure, to large-bodied partner species -- a
  size-matching signal.
- **`h_size` x `p_attack`** carries the sign of the size-versus-attack
  coupling for hosts; read it as the covariance the shared latent axes induce
  between host body size and partner attack investment.
- **`h_defence` row** couples host defence investment to both partner traits
  with the pattern set by row 2 of `Lambda_H`; each cell is the covariance
  contributed by the shared axes, not a standalone correlation.

The fitted `Gamma_hat` reproduces this sign-and-magnitude pattern up to the
loading-orientation caveat below; the null `Gamma` block is numerically zero.
The figure uses the fixture's planted truth and an illustrative fitted block
(a small jitter of the truth standing in for a recovered fit) so that the
heatmap renders without a TMB fit at build time:

```{r gamma-plot, fig.cap = "Cross-lineage shared covariance. The cross-kernel fit recovers the planted host-trait by partner-trait Gamma block; the block-diagonal null leaves that block at zero. The fitted panel here is an illustrative stand-in (truth plus small noise) so the figure builds without a heavy fit.", fig.alt = "Three heatmap facets compare the true Gamma block, an illustrative fitted cross-kernel Gamma block, and the block-diagonal null. The cross-kernel panel reproduces the planted sign pattern, while the null panel contains zero-valued cells."}
set.seed(1)
Gamma_true <- coev$truth$Gamma
Gamma_hat_illustrative <- Gamma_true +
  matrix(rnorm(length(Gamma_true), 0, 0.04), nrow(Gamma_true))
dimnames(Gamma_hat_illustrative) <- dimnames(Gamma_true)
Gamma_null_illustrative <- matrix(
  0, nrow(Gamma_true), ncol(Gamma_true),
  dimnames = dimnames(Gamma_true)
)

gamma_frame <- function(M, source) {
  out <- as.data.frame(as.table(M), stringsAsFactors = FALSE)
  names(out) <- c("host_trait", "partner_trait", "estimate")
  out$source <- source
  out
}

gamma_plot_data <- rbind(
  gamma_frame(Gamma_true, "Truth"),
  gamma_frame(Gamma_hat_illustrative, "Cross kernel (illustrative)"),
  gamma_frame(Gamma_null_illustrative, "Block-diagonal null")
)
gamma_plot_data$source <- factor(
  gamma_plot_data$source,
  levels = c("Truth", "Cross kernel (illustrative)", "Block-diagonal null")
)
gamma_plot_data$host_trait <- factor(
  gamma_plot_data$host_trait,
  levels = rev(coev$truth$host_traits)
)
gamma_plot_data$partner_trait <- factor(
  gamma_plot_data$partner_trait,
  levels = coev$truth$partner_traits
)
limit <- max(abs(gamma_plot_data$estimate))

ggplot(gamma_plot_data,
       aes(x = partner_trait, y = host_trait, fill = estimate)) +
  geom_tile(color = "#d9d9d9", linewidth = 0.7) +
  geom_text(
    aes(label = sprintf("%.2f", estimate),
        color = abs(estimate) > 0.55 * limit),
    size = 3.6
  ) +
  facet_wrap(~ source, nrow = 1) +
  scale_fill_gradient2(
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0, limits = c(-limit, limit),
    name = "Shared\ncovariance"
  ) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "white"),
                     guide = "none") +
  scale_x_discrete(labels = function(x) gsub("_", " ", x, fixed = TRUE)) +
  scale_y_discrete(labels = function(x) gsub("_", " ", x, fixed = TRUE)) +
  labs(x = "Partner trait", y = "Host trait") +
  coord_equal() +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 25, hjust = 1),
    legend.position = "right"
  )
```

## Profiling rho: a sensitivity grid, not a fitted parameter

`rho` lives inside `K*`, not in the TMB parameter vector, so it is **not**
estimated by the fit. The honest workflow is a fixed-kernel sensitivity grid:
use `profile_cross_rho()` to rebuild `K*` over plausible `rho` values, refit
the same model at each grid point, and read the likelihood and `Gamma` as a
function of `rho`.

```{r rho-grid, eval = FALSE}
rho_grid <- seq(0, 0.8, by = 0.2)

rho_results <- profile_cross_rho(
  coev$A_H, coev$A_P, coev$W,
  rho = rho_grid,
  refit = function(K, rho) {
    gllvmTMB(
      traits(h_size, h_defence, p_size, p_attack) ~ 1 +
        kernel_latent(species, K = K, d = 2, name = "cross"),
      data = coev$data_wide,
      unit = "row_id",
      cluster = "species",
      family = gaussian(),
      control = gllvmTMBcontrol(se = FALSE)
    )
  },
  metrics = function(fit, K, rho) {
    list(
      gamma_cor = abs(cor(
        as.vector(extract_Gamma(fit, level = "cross",
                                row_traits = coev$truth$host_traits,
                                col_traits = coev$truth$partner_traits)),
        as.vector(coev$truth$Gamma)
      ))
    )
  }
)
rho_results
```

A representative profile from data generated at `rho = 0.65` looks like the
table below (illustrative numbers; `rho = 0` is exactly the block-diagonal
null):

```
  rho    logLik  pd_hess  gamma_cor
  0.0  -1180.4     TRUE       0.04
  0.2  -1131.7     TRUE       0.71
  0.4  -1102.5     TRUE       0.93
  0.6  -1094.1     TRUE       0.97
  0.8  -1098.8     TRUE       0.95
```

Read the grid two ways. The `logLik` column peaks near the data-generating
`rho`, so the fixed-kernel sensitivity curve points to the best-supported
coupling strength in this fixture. The `gamma_cor` column shows that the
*interpretation* of `Gamma`
stabilises across the upper half of the grid: the sign pattern is the same
whether you pick `rho = 0.4` or `rho = 0.6`. Report the coevolution result
only when the biology is stable across the `rho` values you would defend for
your `W`, and quote the `rho` range you used. In-engine estimation of `rho`
is planned later Design 65 work; until then the sensitivity grid is the
supported route.

## Single-W sensitivity: one association matrix is one replicate

The sharpest caveat in this whole workflow is statistical, not computational.
The model learns the cross block from **one** realised association matrix `W`.
A single `W` is one replicate of the coevolutionary process (Boettiger et al.
2012 make the analogous point for phylogenetic signal: power scales with
independent replication, and a single shared structure is not many
replicates). Two species-rich, densely connected lineages can still deliver
only weak information about `Gamma` if the associations are sparse or
concentrated on a few links.

The C2 recovery test makes this concrete with a sparse-versus-dense sweep on
the *same* underlying association surface. Thinning `W` to its top links
(keeping the strongest 1% and zeroing the rest) leaves fewer realised
host-partner connections, and `Gamma` recovery degrades accordingly:

```{r w-sensitivity, eval = FALSE}
# Dense W: the full association surface (the fixture's W).
K_dense  <- make_cross_kernel(coev$A_H, coev$A_P, coev$W, rho = coev$truth$rho)
fit_dense <- gllvmTMB( ... K = K_dense  ... )   # full syntax as above

# Sparse W: keep only the strongest links, zero the rest.
W_sparse <- coev$W
keep <- W_sparse >= quantile(as.vector(W_sparse), probs = 0.99)
W_sparse <- W_sparse * keep
K_sparse <- make_cross_kernel(coev$A_H, coev$A_P, W_sparse, rho = coev$truth$rho)
fit_sparse <- gllvmTMB( ... K = K_sparse ... )

# Recovery is strictly worse under the thinned W:
#   abs(cor(Gamma_dense,  truth))  ~ 0.97   (many realised links)
#   abs(cor(Gamma_sparse, truth))  ~ 0.74   (few realised links)
# In the sparse case, assess uncertainty with bootstrap or a documented
# sensitivity workflow before making interval-scale claims.
```

The practical reading: a `Gamma` estimate is only as trustworthy as the
**association richness** behind it. Treat the number of realised, well-spread
host-partner links as the effective sample size for the cross block, not the
number of species or the number of measurements. When `W` is thin, report
`Gamma` as weak, sign-level evidence rather than a precise coevolution
estimate, and lean on the `rho` profile and the null comparison rather than
on a single point estimate.

## Honest-uncertainty checklist

Before reporting a `Gamma`, walk this list -- it is the data-condition and
identifiability discipline the engine cannot enforce for you:

- **Convergence and the Hessian.** Confirm `fit$opt$convergence == 0` and
  `fit$fit_health$pd_hessian == TRUE`. A `pdHess = FALSE` result means the
  mode is not a well-defined optimum and the loadings -- hence `Gamma` -- are
  not safely interpretable. Sparse `W`, an over-large rank `d`, or too few
  replicates per species are the usual causes; simplify before trusting the
  block.
- **Loading orientation.** `Gamma = Lambda_H Lambda_P^T` is identified only up
  to the loading constraints. gllvmTMB anchors `Lambda_H` lower-triangular
  with a positive leading diagonal (the recovery test checks
  `Lambda[1, 2] == 0` and positive pivots), which fixes the rotation so that
  `Gamma` is comparable across fits. Without that anchor the same fit could
  return a rotated `Gamma` with a different-looking sign pattern -- so quote
  `Gamma` as the *block*, and read individual cells against the orientation
  convention rather than as standalone parameters.
- **No calibrated intervals yet.** `extract_Gamma()` returns a point estimate;
  it does not attach a calibrated standard error or interval. For uncertainty,
  use a bootstrap or profile workflow, and remember the single-`W`
  replication ceiling above caps how precise any interval can honestly be.
- **Data-condition warnings.** Sparse `W`, a near-singular `A_H` or `A_P`, or
  a `rho` near the PSD boundary all widen the effective uncertainty in ways
  the point estimate does not show. `make_cross_kernel()` will refuse a `rho`
  that breaks positive semidefiniteness; treat that error as a modelling
  signal, not an obstacle to engineer around.

Report the cross-kernel result only when convergence is clean, the
block-diagonal null is not competitive, and the biological reading of `Gamma`
is stable across the `rho` values and the association density you would
defend.

## What to read next

- [`make_cross_kernel()`](../reference/make_cross_kernel.html) documents how
  `A_H`, `A_P`, `W`, and `rho` are scaled into `K*`.
- [`kernel_latent()`](../reference/kernel_latent.html) documents the generic
  dense-kernel formula keywords.
- [`extract_Gamma()`](../reference/extract_Gamma.html) documents the
  cross-lineage covariance extractor and its point-estimate scope.
- [`phylogenetic-gllvm`](phylogenetic-gllvm.html) is the ordinary
  single-lineage phylogenetic trait-covariance example; start there when all
  traits live on one tree.
- [`missing-data`](missing-data.html) explains the block-`NA` response-mask
  machinery that lets the stacked `Y*` carry structural gaps without
  listwise deletion.

## References

- Boettiger, C., Coop, G. & Ralph, P. (2012). Is your phylogeny informative?
  Measuring the power of comparative methods. *Evolution* 66:2240-2251.
- Drury, J., Clavel, J., Manceau, M. & Morlon, H. (2016). Estimating the
  effect of competition on trait evolution using maximum likelihood
  inference. *Systematic Biology* 65:700-710.
- Hadfield, J. D. & Nakagawa, S. (2010). General quantitative genetic methods
  for comparative biology. *Journal of Evolutionary Biology* 23:494-508.
- Manceau, M., Lambert, A. & Morlon, H. (2017). A unifying comparative
  phylogenetic framework including traits coevolving across interacting
  lineages. *Systematic Biology* 66:551-568.
- Ovaskainen, O. et al. (2017). How to make more out of community data?
  *Ecology Letters* 20:561-576.
- Rafferty, N. E. & Ives, A. R. (2013). Phylogenetic trait-based analyses of
  ecological networks. *Ecology* 94:2321-2333.
