---
title: "Cross-family correlations and environmental predictors"
description: "Fit one predictor-informed latent model across five response families, recover shared correlations, and interpret rotation-invariant trait effects."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Cross-family correlations and environmental predictors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
options(cli.num_colors = 1, crayon.enabled = FALSE)
```

Can one environmental gradient explain a shared biological axis while the
responses use different distributions? Yes: one ordinary `latent()` block can
couple continuous, binary, count, ordered, and unordered traits, while
`lv = ~ x` associates the shared axes with a unit-level predictor.

The long-data call is the canonical form; the equivalent wide-data formula
uses `traits(...)` through the same `gllvmTMB()` entry point:

```{r first-look, eval=FALSE}
fit_long <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | unit, d = 3, unique = FALSE, lv = ~ x),
  data = dat, family = family, trait = "trait", unit = "unit"
)
fit_wide <- gllvmTMB(
  traits(g, b, p, o, cat) ~ 1 + latent(1 | unit, d = 3, unique = FALSE, lv = ~ x),
  data = dat_wide,
  family = list(g = gaussian(), b = binomial(), p = poisson(),
                o = ordinal_probit(), cat = multinomial()),
  unit = "unit"
)
```

Traits do not arrive on one scale. A single organism carries continuous
measurements, presence/absence marks, counts, ordered scores, and unordered
categories at once; a single site records abundance, occupancy, and a dominant
class. The question this article answers is narrow and concrete: can a
**Gaussian, a binary, a count, an ordinal, and a nominal trait** sit on **one
shared latent model**, can an observed gradient be associated with its axes,
and can the model report the correlations *between* those families?

This page covers only the **ordinary-unit cross-family route**: one
`multinomial()` trait and its non-nominal partners share
`latent(0 + trait | unit, d, lv = ~ x)`. A powered simulation demonstrates
point recovery for the shared covariance and predictor effect. Separate
route-health canaries exercise every registered native family, but they are not
recovery or interval studies. This page does not admit multiple nominal traits,
structured tiers, augmented slopes, fixed `X + X_lv`, missing responses or LV
predictors, REML, or calibrated arbitrary-composition intervals. The five-family latent
covariance is valid, but the one-number observation-scale nominal summary is
shown on a non-ordinal subset because the current automatic residual convention
is not coherent for an ordinal-probit partner.

## Why put different families on one latent scale

The motivation is the same one that drives joint species distribution models and
phenotypic-integration analyses: the interesting quantity is not each response
on its own but how they **covary**. If a species that is more abundant is also
more likely to be present, and its dominant habitat class shifts with both, that
joint structure is the signal. A shared latent factor $\mathbf u_i$ is the
device that couples the families: every response loads on the same axes, so a
correlation *between families* is well-defined even though the families live on
different observation scales.

The alternative — fit each response separately and correlate the point
predictions — discards the measurement model and gives correlations that are
biased by each family's link and dispersion. Putting the families on one latent
scale keeps the coupling in the model where it belongs.

## The model

For unit $i$, predictors $\mathbf M_i$ shift the mean of a $d$-dimensional
latent score, while $\mathbf e_i$ records remaining unit-to-unit innovation:

$$
\mathbf z_i = \mathbf M_i\boldsymbol\alpha + \mathbf e_i,
\qquad \mathbf e_i \sim \mathcal N(\mathbf 0, \mathbf I_d).
$$

Each pseudo-trait $t$ uses its own response family and link:

$$
\eta_{it} = \mu_t + \boldsymbol\lambda_t^{\!\top}\mathbf z_i,
\qquad
y_{it} \mid \mathbf u_i \sim \text{family}_t(\,g_t^{-1}(\eta_{it})\,).
$$

Gaussian, binomial, and Poisson traits each occupy a single row of the loading
matrix $\boldsymbol\Lambda$. An `ordinal_probit()` trait occupies one row and
crosses a set of ordered thresholds. A `multinomial()` trait is different: an
unordered response with $K$ categories is not one latent dimension but
$K-1$ of them — the baseline-category contrasts — so it expands into $K-1$
rows of $\boldsymbol\Lambda$, named `"<trait>:<cat>"`. The shared low-rank
covariance among all pseudo-traits is
$\boldsymbol\Lambda\boldsymbol\Lambda^{\!\top}$, and the cross-family
correlations are its standardised off-diagonals — rotation-invariant, and the
quantity to report.

The predictor effect on each trait's linear-predictor scale is also
rotation-invariant:

$$
\mathbf B_{lv} = \boldsymbol\Lambda\boldsymbol\alpha^\top.
$$

Raw `alpha` and `Lambda` entries depend on the chosen latent-axis orientation;
`B_lv` and `Lambda %*% t(Lambda)` are the cross-fit scientific targets.

`latent()` can carry a **between-unit** trait-specific variance $\boldsymbol\Psi$
(`unique = TRUE`), so the full tier covariance is
$\boldsymbol\Sigma = \boldsymbol\Lambda\boldsymbol\Lambda^\top + \boldsymbol\Psi$
— $\boldsymbol\Psi$ is the variance each trait has among units *beyond* the
shared factor. For a nominal contrast with one categorical draw per unit, the
corresponding between-unit $\boldsymbol\Psi$ is not identified. Replication can
identify that variance in principle, but the current engine conservatively maps
it off even for the replicated design used here. Admitted partner traits retain
fitted diagonal companions only when the free loading and diagonal parameters
fit within the available covariance moments. This rank-3 teaching model instead
requests the identified loadings-only form with `unique = FALSE`; the
cross-family correlations assessed below live in
$\boldsymbol\Lambda\boldsymbol\Lambda^\top$.

## A five-family worked example

We simulate `r 500` units, each observed `r 6` times, with five traits sharing
one 3-dimensional factor. A standardised environmental gradient shifts the
mean of all three axes; unit innovations create the remaining correlation:

| trait | family | value column holds |
|---|---|---|
| `g` | `gaussian()` | a real number |
| `b` | `binomial()` | 0 / 1 |
| `p` | `poisson()` | a count |
| `o` | `ordinal_probit()` | an ordered code `1..Ko` |
| `cat` | `multinomial()` | a category code `1..K` |

The truth is a single loading matrix over the six pseudo-traits
`g, b, p, o, cat:2, cat:3` (the nominal trait `cat` contributes the last two
rows). Everything the model must recover is contained in
$\boldsymbol\Lambda\boldsymbol\Lambda^{\!\top}$.

```{r setup}
library(gllvmTMB)
```

### Simulate

```{r simulate}
set.seed(20260718)

N     <- 500L   # units
reps  <- 6L     # observations per unit (replication sharpens recovery)
d     <- 3L     # shared latent dimensions
units <- seq_len(N)

# True loadings for the six pseudo-traits on the three shared axes.
# The nominal trait `cat` (K = 3) contributes the rows cat:2 and cat:3.
Lambda <- rbind(
  g       = c( 1.2,  0.2,  0.0),
  b       = c( 0.9,  0.6,  0.1),
  p       = c( 0.3,  1.0,  0.4),
  o       = c( 0.7,  0.4,  0.9),
  `cat:2` = c( 1.0,  0.5,  0.2),
  `cat:3` = c(-0.5,  0.8,  0.6)
)

# Trait means / baseline intercepts (ordinal handled by thresholds instead).
mu  <- c(g = 0, b = 0, p = log(2), o = 0, `cat:2` = 0, `cat:3` = 0)
tau <- c(-1.6, -0.2, 1.2)   # 3 thresholds -> Ko = 4 ordered categories

# One predictor-informed latent position per unit, shared across replicates.
x <- as.numeric(scale(seq(-1, 1, length.out = N)))
alpha <- c(0.55, -0.30, 0.20)
E <- matrix(rnorm(N * d), N, d)
U_mean <- outer(x, alpha)
U_total <- U_mean + E
Eta <- U_total %*% t(Lambda)           # N x 6 unit-level predictors
colnames(Eta) <- rownames(Lambda)
B_lv_true <- drop(Lambda %*% alpha)

sim_one_rep <- function(rep_id) {
  g_val <- mu["g"] + Eta[, "g"] + rnorm(N)                      # Gaussian
  b_val <- rbinom(N, 1L, plogis(mu["b"] + Eta[, "b"]))         # binary
  p_val <- rpois(N, exp(mu["p"] + Eta[, "p"]))                  # count
  o_val <- findInterval(mu["o"] + Eta[, "o"] + rnorm(N), tau) + 1L  # ordinal
  e2 <- mu["cat:2"] + Eta[, "cat:2"]                            # nominal
  e3 <- mu["cat:3"] + Eta[, "cat:3"]                            #   contrasts
  P  <- cbind(1, exp(e2), exp(e3)); P <- P / rowSums(P)         # softmax
  cat_val <- apply(P, 1L, function(pr) sample.int(3L, 1L, prob = pr))
  rbind(
    data.frame(unit = units, replicate = rep_id, trait = "g",   family = "g", x = x, value = g_val),
    data.frame(unit = units, replicate = rep_id, trait = "b",   family = "b", x = x, value = b_val),
    data.frame(unit = units, replicate = rep_id, trait = "p",   family = "p", x = x, value = p_val),
    data.frame(unit = units, replicate = rep_id, trait = "o",   family = "o", x = x, value = o_val),
    data.frame(unit = units, replicate = rep_id, trait = "cat", family = "m", x = x, value = cat_val)
  )
}

dat <- do.call(rbind, lapply(seq_len(reps), sim_one_rep))
dat$unit   <- factor(dat$unit)
dat$trait  <- factor(dat$trait,  levels = c("g", "b", "p", "o", "cat"))
dat$family <- factor(dat$family, levels = c("g", "b", "p", "o", "m"))

table(dat$trait)

dat_wide <- reshape(
  dat[c("unit", "replicate", "x", "trait", "value")],
  idvar = c("unit", "replicate", "x"), timevar = "trait", direction = "wide"
)
names(dat_wide) <- sub("^value\\.", "", names(dat_wide))
```

The `family` column maps each row to its response family; the nominal trait is
the single column `cat`, whose values are category codes `1..K`. It expands into
the pseudo-traits `cat:2` and `cat:3` inside the fit, not in the data.

### Fit

One `gllvmTMB()` call carries all five families on the shared factor. The
`family` argument is a named list keyed by the `family` column's levels, and
`attr(family, "family_var")` names that column.

```{r fit}
family <- list(
  g = gaussian(), b = binomial(), p = poisson(),
  o = ordinal_probit(), m = multinomial()
)
attr(family, "family_var") <- "family"

fit <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | unit, d = 3, unique = FALSE, lv = ~ x),
  data   = dat,
  family = family,
  trait  = "trait",
  unit   = "unit",
  silent = TRUE
)

c(convergence = fit$opt$convergence, pdHess = isTRUE(fit$sd_report$pdHess))
```

The explicit `unique = FALSE` requests the loadings-only covariance. At this
rank, adding an automatic diagonal $\Psi$ would introduce more free covariance
parameters than the observed trait covariance can identify, so the parser
refuses that saturated form and recommends a lower rank, a scientifically
appropriate common diagonal variance, or this loadings-only model.

### Recover the predictor effect

`extract_lv_effects(type = "trait_effect")` returns $B_{lv}$ rather than raw
axis coefficients. Each row is an association with `x` on that response's
linear-predictor scale.

```{r recover-predictor}
B_hat <- extract_lv_effects(fit, type = "trait_effect")
B_compare <- data.frame(
  trait = B_hat$trait,
  true = unname(B_lv_true[B_hat$trait]),
  recovered = B_hat$estimate
)
B_compare
```

For Gaussian traits this is an associated mean change. For Poisson traits it
is a log expected-count change; exponentiate it for a multiplicative contrast.
For ordinal and nominal rows it remains a latent linear-predictor association.
The model is associational: changing `x` is not interpreted as an intervention.

### Recover the cross-family correlations

`extract_Sigma()` with `part = "shared"` returns the latent low-rank covariance
$\boldsymbol\Lambda\boldsymbol\Lambda^{\!\top}$ over all six pseudo-traits. We
standardise it and compare against the truth built from `Lambda`.

```{r recover}
Sigma_true <- Lambda %*% t(Lambda)
R_true <- cov2cor(Sigma_true)

Sig_hat <- extract_Sigma(fit, level = "unit", part = "shared", link_residual = "none")
R_hat   <- cov2cor(Sig_hat$Sigma)

# align the recovered matrix to the truth's ordering by name
ord   <- rownames(R_true)
R_hat <- R_hat[ord, ord]

pairs <- rbind(
  c("g", "b"), c("g", "p"), c("g", "o"),
  c("g", "cat:2"), c("g", "cat:3"),
  c("b", "cat:2"), c("p", "cat:3"), c("o", "cat:2")
)

data.frame(
  pair      = paste(pairs[, 1], pairs[, 2], sep = " ~ "),
  true      = round(mapply(function(i, j) R_true[i, j], pairs[, 1], pairs[, 2]), 3),
  recovered = round(mapply(function(i, j) R_hat[i, j], pairs[, 1], pairs[, 2]), 3),
  row.names = NULL
)
```

The recovered correlations track the truth across all five families,
including the two nominal contrasts `cat:2` and `cat:3` and their sign — a
strong positive `g ~ cat:2`, a negative `g ~ cat:3`. The remaining error is
sampling noise at this `N`; larger `N` (or more replicates per unit) shrinks it.
Every off-diagonal pair, not just the eight highlighted here, is available in
`R_hat`.

```{r recover-plot, fig.width = 5.2, fig.height = 5.2, fig.cap = "True versus recovered latent cross-family correlations, all off-diagonal pairs. The line is the identity; points near it indicate faithful recovery in this powered run."}
offd <- upper.tri(R_true)
plot(R_true[offd], R_hat[offd],
     xlab = "true correlation", ylab = "recovered correlation",
     xlim = c(-1, 1), ylim = c(-1, 1), pch = 19, col = "#3366aa")
abline(0, 1, lty = 2, col = "grey40")
```

## Summarising the nominal trait: the reference-invariant multiple correlation

The pairwise table above lists a *separate* correlation for each nominal
contrast (`cat:2`, `cat:3`). Often a reader wants **one number** for "how
strongly is the categorical trait associated with each partner?" that does not
depend on which category was chosen as the reference. That is what
`extract_cross_correlations()` reports: a
**reference-invariant multiple correlation** between the categorical trait and
each single-scale partner, optionally with the underlying $(K-1)$-vector.

The five-family fit contains an ordinal-probit partner. Its threshold model
already fixes the latent residual at one, so an `"auto"` summary would add that
unit residual a second time. `extract_cross_correlations()` therefore refuses
that combination rather than attenuating the reported association. We retain
the valid five-family latent-covariance result above, then refit the same data
without the ordinal trait for the observation-scale one-number summary.

```{r cross-cor}
ordinal_refusal <- tryCatch(
  extract_cross_correlations(fit, level = "unit", link_residual = "auto"),
  error = function(e) cli::ansi_strip(conditionMessage(e))
)
ordinal_refusal

dat_summary <- droplevels(dat[dat$trait != "o", ])
family_summary <- family[c("g", "b", "p", "m")]
attr(family_summary, "family_var") <- "family"

fit_summary <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | unit, d = 3, unique = FALSE, lv = ~ x),
  data   = dat_summary,
  family = family_summary,
  trait  = "trait",
  unit   = "unit",
  silent = TRUE
)

xc <- extract_cross_correlations(
  fit_summary, level = "unit", link_residual = "auto"
)
xc
```

The single-number summary is reported on the **observation scale**. The
`"auto"` route adds the applicable family-specific residual for each admitted
partner and the full softmax residual matrix for the nominal block. This is the
deliberate default, and the reason is in the next section.

## Honest caveats

These are limits of the capability as it stands at 0.6, not incidental notes.

- **Fitted $\boldsymbol\Psi$ and the fixed link residual are different objects.**
  The **between-unit** $\boldsymbol\Psi$ inside `latent()` is an estimated
  among-unit diagonal companion. The extractor's $R_{\mathrm{link}}$ is a
  separate family-specific observation-scale addition; for the nominal block it
  is $(\pi^2/6)(\mathbf I+\mathbf J)$. With one categorical draw per unit the
  nominal contrast $\boldsymbol\Psi$ is not identified. Replication can identify
  it in principle, but the current engine conservatively maps it off even in the
  replicated design here; admitted partner diagonals remain fitted where the
  model identifies them. A replication-aware multinomial refinement is future
  work. Either way the pairwise latent correlations (the matrix) are the
  recovered quantities. See
  [the vocabulary of $\boldsymbol\Psi$](gllvm-vocabulary.html) and
  [covariance and correlation](covariance-correlation.html).

- **The single-number summary belongs on the observation scale.** For an
  admitted non-ordinal partner set, the reference-invariant multiple correlation
  uses each applicable family-specific residual and the softmax residual
  $(\pi^2/6)(\mathbf I + \mathbf J)$ — the
  covariance of the differenced Gumbel errors in the random-utility
  representation of the softmax (McFadden 1974), which reduces to the binomial's
  $\pi^2/3$ when $K = 2$. On the **latent** scale (`link_residual = "none"`) the
  multiple correlation degenerates toward 1 whenever the $K-1$ contrast block
  spans the partner, because a $(K-1)$-dimensional block can align with almost
  any single direction. That is why the observation scale is the sensible default
  for the one-number summary, while the per-pair latent correlations remain the
  right object for the matrix. Ordinal-probit is not admitted on this automatic
  summary route because its residual is already fixed by the threshold model;
  the typed refusal above prevents double counting.

- **Recovering a nominal trait's covariance is data-hungry.** A single
  categorical draw carries little information about a $(K-1)$-dimensional
  liability. With one draw per unit the reduced-rank estimate can rail at
  $\pm 1$; this is an information limitation, not a bug. It resolves with
  replication (multiple draws per unit, as here) or many units. The powered
  regime in this article — `r 500` units $\times$ `r 6` replicates — is what
  makes the recovery above possible; do not expect the same at one observation
  per unit.

- **Intervals are cell-specific, not universal.** Existing evidence calibrates
  selected rank-1 Gaussian-anchor `B_lv` cells. It does not calibrate the
  rank-3 arbitrary composition shown here or cross-family correlation
  intervals. Treat these outputs as point/recovery-oriented.

- **Gaussian and lognormal residual scales are separate when they coexist.**
  Gaussian residual SD uses raw response units; lognormal residual SD uses
  `log(y)`. A joint fit therefore reports two `sigma_eps` slots, shared within
  family rather than across these incomparable scales. Pure fits retain the
  historical one-slot contract.

- **One multinomial trait per fit.** The current implementation supports a single
  `multinomial()` trait in a fit. Two nominal responses on one shared factor is
  not yet supported.

## See also

- [Unordered categories with `multinomial()`](multinomial.html) — the
  baseline-category model, reading contrasts and predicted probabilities, and the
  standalone among-category table in the phylogenetic setting.
- [Covariance and correlation: the model behind Sigma](covariance-correlation.html)
  — $\boldsymbol\Sigma$, $\boldsymbol\Lambda$, $\boldsymbol\Psi$, and the
  `link_residual` conventions used above.
- [`?extract_Sigma`](../reference/extract_Sigma.html) — the unified covariance
  extractor.
- [`?extract_cross_correlations`](../reference/extract_cross_correlations.html) —
  the reference-invariant nominal summary.
- **McFadden (1974)** Conditional logit analysis of qualitative choice behavior.
  In *Frontiers in Econometrics*, ed. P. Zarembka, pp. 105–142. Academic Press.
- **Nakagawa & Schielzeth (2010)** Repeatability for Gaussian and non-Gaussian
  data. *Biological Reviews* 85(4):935–956.
  <https://doi.org/10.1111/j.1469-185X.2010.00141.x>
