---
title: "Cross-lineage coevolution kernels"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Cross-lineage coevolution kernels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7.2, fig.height = 4.4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

This article is for a comparative-methods reader with two interacting
lineages: for example hosts and parasites, plants and pollinators, or
mutualists. The working question is: after accounting for the host
phylogeny, the partner phylogeny, and the observed association matrix
`W`, can we estimate the host-trait by partner-trait covariance block?

Scope boundary: **IN** (`COE-02`, `KER-02`), this article fits one
named dense kernel tier with `kernel_latent()` plus `kernel_unique()`
and extracts the cross-lineage shared block with `extract_Gamma()`.
**PARTIAL**, `rho` is supplied when `K_star` is built, and
`extract_Gamma()` returns point estimates without calibrated
intervals. **PLANNED**, multiple simultaneous kernel tiers and
in-engine `rho` estimation remain later Design 65 work.

The core fit is a `traits(...)` wide-data formula. The article also
fits the same model with the canonical long `value ~ ...` formula
after loading the example fixture.

```{r model-shape, eval = FALSE}
fit_cross <- gllvmTMB(
  traits(h_size, h_defence, p_size, p_attack) ~ 1 +
    kernel_latent(species, K = K_star, d = 2, name = "cross") +
    kernel_unique(species, K = K_star, name = "cross"),
  data = coev$data_wide,
  unit = "row_id",
  cluster = "species",
  family = gaussian()
)
```

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

```{r load-example, eval = FALSE}
coev <- readRDS(system.file(
  "extdata", "examples", "coevolution-kernel-example.rds",
  package = "gllvmTMB"
))
```

```{r load-example-hidden, include = FALSE}
example_path <- system.file(
  "extdata", "examples", "coevolution-kernel-example.rds",
  package = "gllvmTMB"
)
example_path <- c(
  example_path,
  file.path("inst", "extdata", "examples", "coevolution-kernel-example.rds"),
  file.path("..", "inst", "extdata", "examples", "coevolution-kernel-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "coevolution-kernel-example.rds")
)
example_path <- example_path[nzchar(example_path) & file.exists(example_path)][1]
stopifnot(!is.na(example_path))
coev <- readRDS(example_path)
```

```{r example-preview}
coev$story$question
head(coev$data_wide, 6)
```

The wide table has one row per sampled lineage individual. Host rows
carry host traits and `NA` partner traits; partner rows carry partner
traits and `NA` host traits. The long table makes those block-missing
cells explicit as one row per `(row_id, trait)` pair:

```{r long-preview}
head(coev$data_long, 8)
```

## Build the cross-lineage kernel

`make_cross_kernel()` combines the host relatedness matrix `A_H`, the
partner relatedness matrix `A_P`, and the association matrix `W` into
one positive semidefinite kernel:

$$
K^* =
\begin{bmatrix}
A_H & K_{HP} \\
K_{HP}^{T} & A_P
\end{bmatrix}.
$$

The off-diagonal block `K_HP` is where the observed host-partner
associations enter. In this first workflow, `rho` is a user-supplied
coupling strength rather than an estimated parameter.

```{r build-kernel}
K_star <- make_cross_kernel(
  coev$A_H,
  coev$A_P,
  coev$W,
  rho = coev$truth$rho
)

range(eigen(K_star, symmetric = TRUE, only.values = TRUE)$values)
max(abs(K_star - coev$K_star))
```

The null model keeps the two lineage phylogenies but removes the
cross-lineage block. That block-diagonal kernel asks whether the same
within-lineage structure is enough without host-partner covariance.

```{r null-kernel}
K_null <- coev$K_null
max(abs(K_null[rownames(coev$A_H), rownames(coev$A_P)]))
```

## Fit the cross-kernel model

The wide data-frame formula is compact because `traits(...)` names the
four response columns on the left-hand side. The named kernel tier
`"cross"` is the level that `extract_Sigma()` and `extract_Gamma()`
will read later. `kernel_latent()` estimates the shared reduced-rank
`Lambda Lambda^T` covariance attached to `K_star`; `kernel_unique()`
adds trait-specific diagonal `psi` variation at the same tier.

```{r fit-wide}
ctrl <- gllvmTMBcontrol(se = FALSE)

fit_cross <- suppressMessages(gllvmTMB(
  traits(h_size, h_defence, p_size, p_attack) ~ 1 +
    kernel_latent(species, K = K_star, d = 2, name = "cross") +
    kernel_unique(species, K = K_star, name = "cross"),
  data = coev$data_wide,
  unit = "row_id",
  cluster = "species",
  family = gaussian(),
  control = ctrl
))

fit_cross$opt$convergence
```

The long formula is the same model written with the canonical
long-format response column. It is useful when the data already arrive
as one `(unit, trait)` row per observation.

```{r fit-long}
fit_cross_long <- suppressMessages(gllvmTMB(
  value ~ 0 + trait +
    kernel_latent(species, K = K_star, d = 2, name = "cross") +
    kernel_unique(species, K = K_star, name = "cross"),
  data = coev$data_long,
  trait = "trait",
  unit = "row_id",
  cluster = "species",
  family = gaussian(),
  control = ctrl
))

cat("Long/wide logLik difference:",
    signif(abs(as.numeric(logLik(fit_cross)) -
               as.numeric(logLik(fit_cross_long))), 3), "\n")
```

Now fit the block-diagonal null on the same data. It uses the same
formula and rank, but the kernel has no host-partner bridge.

```{r fit-null}
fit_null <- suppressMessages(gllvmTMB(
  traits(h_size, h_defence, p_size, p_attack) ~ 1 +
    kernel_latent(species, K = K_null, d = 2, name = "cross") +
    kernel_unique(species, K = K_null, name = "cross"),
  data = coev$data_wide,
  unit = "row_id",
  cluster = "species",
  family = gaussian(),
  control = ctrl
))

c(
  cross = as.numeric(logLik(fit_cross)),
  null = as.numeric(logLik(fit_null)),
  improvement = as.numeric(logLik(fit_cross)) - as.numeric(logLik(fit_null))
)
```

## Extract Gamma

`extract_Gamma()` slices the shared part of the fitted kernel-tier
`Sigma`. For this article, the rows are host traits and the columns
are partner traits:

$$
\Gamma_{HP} = \Lambda_H \Lambda_P^T.
$$

Here, `Lambda` is the trait-by-latent-axis loading matrix for the
shared kernel tier.

Because this target is a covariance block, its sign and scale matter:
positive cells mean the host and partner traits rise together along
the cross-kernel latent axes; negative cells mean they move in opposite
directions.

```{r extract-gamma}
Gamma_hat <- extract_Gamma(
  fit_cross,
  level = "cross",
  row_traits = coev$truth$host_traits,
  col_traits = coev$truth$partner_traits
)

Gamma_null <- extract_Gamma(
  fit_null,
  level = "cross",
  row_traits = coev$truth$host_traits,
  col_traits = coev$truth$partner_traits
)

round(coev$truth$Gamma, 2)
round(Gamma_hat, 2)
round(Gamma_null, 4)

abs(cor(as.vector(Gamma_hat), as.vector(coev$truth$Gamma)))
```

The null fit has an approximately zero `Gamma` block because its
kernel has no cross-lineage covariance. The cross-kernel fit recovers
the planted sign pattern and improves the likelihood on these data.

```{r gamma-plot, fig.cap = "Cross-lineage shared covariance. The fitted cross-kernel model recovers the planted host-trait by partner-trait Gamma block, while the block-diagonal null leaves that block at zero.", fig.alt = "Three heatmap facets compare the true Gamma block, the fitted cross-kernel Gamma block, and the block-diagonal null. The cross-kernel panel reproduces the planted sign pattern, while the null panel contains zero-valued cells."}
gamma_frame <- function(M, source) {
  out <- as.data.frame(as.table(M), stringsAsFactors = FALSE)
  names(out) <- c("host_trait", "partner_trait", "estimate")
  out$source <- source
  out
}

gamma_plot_data <- rbind(
  gamma_frame(coev$truth$Gamma, "Truth"),
  gamma_frame(Gamma_hat, "Cross kernel"),
  gamma_frame(Gamma_null, "Block-diagonal null")
)
gamma_plot_data$source <- factor(
  gamma_plot_data$source,
  levels = c("Truth", "Cross kernel", "Block-diagonal null")
)
gamma_plot_data$host_trait <- factor(
  gamma_plot_data$host_trait,
  levels = rev(coev$truth$host_traits)
)
gamma_plot_data$partner_trait <- factor(
  gamma_plot_data$partner_trait,
  levels = coev$truth$partner_traits
)
limit <- max(abs(gamma_plot_data$estimate))

ggplot(gamma_plot_data,
       aes(x = partner_trait, y = host_trait, fill = estimate)) +
  geom_tile(color = "#d9d9d9", linewidth = 0.7) +
  geom_text(
    aes(label = sprintf("%.2f", estimate),
        color = abs(estimate) > 0.55 * limit),
    size = 3.6
  ) +
  facet_wrap(~ source, nrow = 1) +
  scale_fill_gradient2(
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0, limits = c(-limit, limit),
    name = "Shared\ncovariance"
  ) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "white"),
                     guide = "none") +
  scale_x_discrete(labels = function(x) gsub("_", " ", x, fixed = TRUE)) +
  scale_y_discrete(labels = function(x) gsub("_", " ", x, fixed = TRUE)) +
  labs(x = "Partner trait", y = "Host trait") +
  coord_equal() +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 25, hjust = 1),
    legend.position = "right"
  )
```

## Read the result cautiously

The model learns cross-lineage covariance from one realised
association matrix `W`. That makes the workflow useful, but it also
means sparse or weakly structured `W` can support much less information
than a dense simulation suggests. The C2 validation test therefore
includes a single-`W` sensitivity check: thinning the same association
surface reduces `Gamma` recovery relative to the dense case.

For an applied analysis, treat `rho` as part of the sensitivity
workflow. Rebuild `K_star` over a grid of plausible `rho` values, refit
the model, and compare `logLik`, `Gamma`, and fit diagnostics:

```{r rho-grid, eval = FALSE}
rho_grid <- seq(0, 0.8, by = 0.2)
rho_results <- lapply(rho_grid, function(rho) {
  K_rho <- make_cross_kernel(coev$A_H, coev$A_P, coev$W, rho = rho)
  fit_rho <- gllvmTMB(
    traits(h_size, h_defence, p_size, p_attack) ~ 1 +
      kernel_latent(species, K = K_rho, d = 2, name = "cross") +
      kernel_unique(species, K = K_rho, name = "cross"),
    data = coev$data_wide,
    unit = "row_id",
    cluster = "species",
    family = gaussian(),
    control = gllvmTMBcontrol(se = FALSE)
  )
  list(rho = rho, logLik = logLik(fit_rho),
       Gamma = extract_Gamma(fit_rho, level = "cross",
                             row_traits = coev$truth$host_traits,
                             col_traits = coev$truth$partner_traits))
})
```

Report the cross-kernel result only when the biological interpretation
is stable across the `rho` values you would defend for `W`, and when
the block-diagonal null is not competitive.

## What to read next

- [`make_cross_kernel()`](../reference/make_cross_kernel.html) documents
  how `A_H`, `A_P`, `W`, and `rho` are scaled into `K_star`.
- [`kernel_latent()`](../reference/kernel_latent.html) documents the
  generic dense-kernel formula keywords.
- [`extract_Gamma()`](../reference/extract_Gamma.html) documents the
  cross-lineage covariance extractor and its point-estimate scope.
- [`phylogenetic-gllvm`](phylogenetic-gllvm.html) is the ordinary
  single-lineage phylogenetic trait-covariance example; start there
  when all traits live on one tree.
