---
title: "Current limitations and boundaries"
description: "Decide which gllvmTMB models and results are ready for your analysis, which remain experimental, and what to check before interpretation."
tier: 1 # Primary Getting Started safety and scope page for first-time users.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Current limitations and boundaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

## Can I use gllvmTMB for this analysis?

Read this page before choosing a model. `gllvmTMB` is experimental, and its
implemented surface is larger than the set of models with deep recovery and
comparison evidence. A formula being accepted tells you that the route is
implemented; it does not, by itself, show that the covariance is identifiable
for your data or that its uncertainty is calibrated.

The safest current starting boundary is an ordinary Gaussian model fitted with
the default Laplace engine using `indep()` or `dep()`. The frozen production
study gave positive point-estimation evidence for those two routes in its
tested three-trait complete-data designs. It did not earn a package-wide
dependable-core claim. The Gaussian latent decomposition remains useful as a
teaching and characterization model, while non-Gaussian latent evidence is
family- and design-specific.

This page reports the evidence boundary for the development version leading to
0.7.0, updated after production confirmation on 9 August 2026. The package
version remains 0.6.0. CRAN availability will not, by itself, be a scientific
certification.

## What the status labels mean

**Dependable core** means the route passed every preregistered first-release
recovery and comparator gate. No candidate receives this label before all of
its release cells pass.

**Experimental/partial** means a useful implementation and some evidence
exist, but the evidence is narrower than the general model description. Read
the linked model guide and do not generalise beyond its stated design.

**Not recommended for this claim** means the package may expose related code,
but the intended scientific interpretation has not earned a public reliability
claim. Choose a simpler route or another package.

## What production confirmation found

The frozen study completed 44,800 fits across 28 pilot-admitted cells. Gaussian
`indep()`, Gaussian `dep()`, Poisson-log rank-1
`latent(unique = TRUE)`, and Binomial(10)-logit rank-1
`latent(unique = TRUE)` passed their prespecified small/large core
point-estimation gates. Gaussian `latent(unique = TRUE)` and NB2-log
`latent(unique = TRUE)` remained characterization-only.

That result is deliberately narrower than a dependable-core label. Not every
eligible cell entered production, every campaign contained failed cells, and
the observable terminal-status detector found only 19 of 80 catastrophic truth
errors. The passing Poisson and binomial pairs are therefore clean-data,
tested-regime point evidence, not family-wide robustness or reliable
silent-failure protection.

Within the package-wide **Experimental/partial** status, the exact-aa evidence
uses three more precise result phrases. **Narrow tested-regime point evidence**
means the named core cells passed only in their stated designs. **Narrow
clean-data point evidence** adds the restriction that adjacent robustness cells
did not all pass. **Characterization-only** means the route can be studied and
reported transparently, but its tested pair did not pass the release gate.

## Current decision table

For the ordinary latent route, `latent(..., unique = TRUE)` is the default and
represents
\(\Sigma = \Lambda\Lambda^\mathsf{T} + \Psi\), where \(\Psi\) is diagonal.
Setting `unique = FALSE` requests the narrower loadings-only subset.

| Analysis or result | Current status | What you may report | What you must not infer |
|---|---|---|---|
| Ordinary Gaussian `indep()` | Experimental/partial | **Narrow tested-regime point evidence:** trait-specific independent variance point estimates in the tested three-trait complete-data regimes (`n = 60` and `n = 240`) after fit-health checks | Calibrated uncertainty, family-wide robustness, or protection across arbitrary sample sizes and boundary regimes |
| Ordinary Gaussian `dep()` | Experimental/partial | **Narrow tested-regime point evidence:** a full trait covariance point estimate in the tested three-trait complete-data regimes (`n = 60` and `n = 240`) after fit-health checks | That a positive-definite Hessian proves every covariance entry is well estimated, or that the result transfers to structured sources |
| Ordinary Gaussian `latent()` | Experimental/partial | **Characterization-only:** rotation-invariant `Sigma`, correlations, communality, and an ordination after diagnostics; the `n = 60` core cell was not admitted and correlation stress cells failed production gates | A dependable-core claim, biological meaning for raw loading signs/orientation, or reliability in weak-variance and boundary regimes |
| Poisson-log rank-1 `latent()` | Experimental/partial | **Narrow clean-data point evidence:** point estimates and rotation-invariant covariance summaries in the tested three-trait designs (`n = 100` and `n = 300`) | Family-wide robustness: the rare-level robustness cell failed its fixed-effect-bias gate; no general interval or detector claim follows |
| NB2-log rank-1 `latent()` | Experimental/partial | **Characterization-only:** point results with the exact sample size and dispersion diagnostics reported; the `n = 300` core cell passed | A dependable point-estimation claim: the `n = 100` and missing-response cells produced catastrophic dispersion failures that diagnostics often missed |
| NB2-log latent model with missing responses | Not recommended for this claim | Nothing as dependable evidence; the tested cell may be used only to study the failure mechanism | That standard optimizer, Hessian, or boundary diagnostics clear the fit: 30 of 41 catastrophic truth errors in the tested missing-response cell were missed |
| Binomial(10)-logit rank-1 `latent()` | Experimental/partial | **Narrow clean-data point evidence:** point estimates and rotation-invariant covariance summaries in the tested three-trait designs (`n = 100` and `n = 300`) | Family-wide robustness, Bernoulli equivalence, separation protection, or reliability in the held/failed small-sample silent-failure cells |
| Other response families | Experimental/partial | Results only within the conditions stated in the response-family and model-specific guides | That family construction or one successful fit establishes broad recovery |
| Ordinal models | Not recommended for this claim | Exploratory point fits only with model-specific validation; there is no calibrated ordinal degeneracy detector | That nominal convergence rules out a degenerate fit, or that binomial detector thresholds transfer to ordinal fits |
| Mixed-family models across traits | Experimental/partial | Point estimates only for explicitly documented family combinations, with one family/link per trait | Direct comparison of correlations defined on incompatible latent or response scales |
| Family/link varying across rows within one trait | Not recommended for this claim | Nothing yet; the fit now fails loudly before optimization | That a shared loading is coherent across Poisson-log and binomial-logit/probit merely because the engine accepts per-row family IDs; the planned Poisson-log/binomial-cloglog integrated route is not yet validated |
| Non-unit likelihood weights | Not recommended for likelihood-based inference | Weighted-objective point estimates only, with the weighting rule reported | Ordinary Hessian/Wald uncertainty, `logLik()`, AIC, BIC, or likelihood-ratio interpretation; no sandwich route is certified |
| Phylogenetic, animal, spatial, or kernel covariance | Experimental/partial | Point estimates for the exact documented route after structure-specific diagnostics | That ordinary-model evidence transfers to a structured source, that spatial `tau` or field SD has validated absolute-scale recovery, or that structured intervals are calibrated |
| Random slopes or several grouping tiers | Experimental/partial | Exact admitted cells only, with the covariance block named explicitly | A grid-wide claim across families, modes, slope counts, or grouping levels |
| Missing response cells | Experimental/partial | Point estimates after confirming the intended cells were omitted from the likelihood; the strongest worked evidence is Gaussian | That every family has equal recovery depth or that missingness is ignorable by default |
| One modelled missing predictor with `mi()` | Experimental/partial | Native-Laplace results for one explicitly modelled predictor in the documented predictor-family routes | VA support, several `mi()` predictors, general MNAR handling, or missing grouping variables |
| Variational approximation | Experimental/partial | Opt-in exploratory point estimates within the admitted VA fence, using the current automatic GH evaluator with `H = 7`, with the returned method metadata and the exact family/rank evidence named | General point recovery: only 1 of 36 family-by-rank cells passed the current campaign's overall point route, and that sole pass was Poisson-log `q = 5`, outside the public `d <= 2` fence; its ELBO is not a log-likelihood, so ordinary `logLik()`, AIC, BIC, and likelihood-ratio comparisons are undefined; do not infer calibrated covariance, Wald intervals, or evidence equivalent to Laplace |
| Adaptive Gauss-Hermite quadrature | Experimental/partial | Opt-in sensitivity analysis in its documented scalar-random-effect regime, stating the ridge setting | Ordinary ML interpretation: AGHQ defaults to a `tau = 2` penalised MAP fit, so ordinary AIC, BIC, and likelihood-ratio interpretations do not apply; the evaluated node/ridge rule is not a generally calibrated estimator |
| EVA | Not recommended for this claim | Nothing yet | That internal design work implies an available estimator |
| Julia engine | Experimental/partial | Optional-bridge results only after checking the route-specific parity evidence | Full R/TMB feature parity or a required dependency for the ordinary workflow |
| Response-column coefficient models (`column_coef()`, `phylo_coef()`, `animal_coef()`, `kernel_coef()`, `spatial_coef()`) | Experimental/partial | Gaussian point estimates for a random intercept, slope, or both, varying across response columns, fitted through long or `traits(...)` wide data. `phylo_coef()` and `kernel_coef()` can fix a numeric mixture strength or estimate one interior value; `animal_coef()` fixes it; `spatial_coef()` fixes it at 1 and estimates the spatial range | Calibrated intervals, non-Gaussian responses, an estimated mixture strength for the animal or spatial routes, or recovery outside the tested Gaussian cells |
| Prediction at new locations for integrated (iSDM) fits (`predict()` on `isdm_sources()` fits) | Experimental/partial | Point predictions, including maps, at new in-hull coordinates for the ordinary intercept-only spatial tier; training-row predictions reproduce the fitted values exactly | Held-out map accuracy — the preregistered accuracy campaign did not pass its gate; a standard error or confidence interval on `newdata` predictions (`se.fit` is refused there); or predictions for spatial slope tiers, which remain unsupported |
| Predictor-informed latent scores (`latent(..., lv = ~ x)`) | Experimental/partial | Point estimates of the rotation-invariant effect of a predictor acting through the shared latent axes, combining traits from more than one response family in one model, in the specific tested construction and evidence cells | Broad recovery or calibration across arbitrary family combinations; support for missing latent-score predictors, combined fixed-and-latent-score predictors, REML, or a structured (phylogenetic/spatial/kernel) source on this term |
| Multinomial structured routes | Experimental/partial | Fixed-effect point estimates and probability predictions for one unordered multinomial response are well tested. Several structured extensions (a phylogenetic route, a shared-latent cross-family route, pedigree- or kernel-sourced routes, and ordinary group random intercepts) exist and each has its own narrower tested evidence | Fitting more than one multinomial response in the same model; augmented (intercept-and-slope) structured forms; or calibrated intervals on any structured multinomial route |
| Estimated source strength (`rho = NULL` on `phylo_*`/`animal_*`/`kernel_*`/`spatial_*` terms) | Experimental/partial | Bounded point-recovery evidence for a native Gaussian model with complete replicated trait vectors, one known relatedness source, retained observation-level residual variance, and no competing ordinary covariance term | General recovery across other source/covariance-mode combinations — most combinations tested so far did not pass; any confidence interval on the estimated strength, which does not yet exist; or recovery for augmented (intercept-and-slope) forms |
| Wald, bootstrap, or profile intervals | Not recommended for this claim | A target-specific returned interval together with its method and status label, subject to the narrow exception below | That method availability establishes repeated-sampling coverage |

## What to interpret first

For latent models, start from quantities that do not change when the latent
axes are rotated:

1. total covariance, `Sigma`;
2. correlations derived from `Sigma`;
3. shared covariance, `Lambda Lambda^T`;
4. the diagonal unique component, `Psi`;
5. communality.

Raw loading columns can change sign, order, or orientation without changing the
fitted covariance. Use them for an ordination only after applying a stated
rotation convention; do not treat an arbitrary raw orientation as biology.

## Current uncertainty limits

The package provides Wald, bootstrap, and profile machinery for named targets,
but these routes do not share one package-wide calibration claim.

`profile_ci_total_variance()` is a penalty-profile approximation, not a
certified exact likelihood-ratio profile. Historical ordinary Gaussian
`n_units = 150`, `d` in `{1, 2}` cells and the newer `n_units = 400`, `d = 2`
cell passed their numerical coverage gates, but the retained endpoints do not
prove constrained-refit convergence and exact target attainment. Every
computed row is therefore `route-only`; the former `certified-0.94` labels are
withdrawn until the mechanism is repaired and recalibrated.

There is one exact loading-interval exception. Standardized symmetric
joint-delta Wald intervals are certified for the structurally free strict-lower
targets in native pinned unrotated ordinary Gaussian three-trait cells
`(n_units=150,d=2)`, `(n_units=400,d=1)`, and `(n_units=400,d=2)`, at 95%.
This is one frozen DGP: trait intercepts `(-0.20, 0.10, 0.25)`, unique standard
deviations `(0.70, 0.80, 0.90)`, and loading vector `(0.80, 0.45, -0.35)` for
`d=1`, with second column `(0, 0.70, 0.40)` for `d=2`. Coverage is conditional on eligible fits
(optimizer convergence, converged fit health, available
`sdreport()`, and a positive-definite Hessian); availability was 98.82%,
93.38%, and 96.18% in the three certified cells. The `(n_units=150,d=1)` cell
failed. No other loading, unique-variance, or intercept regime inherits the
result. Pinned diagnostic rows, Fisher-z Wald,
arbitrary constraints, rotated fits, and every neighbouring family, tier,
rank, sample size, or confidence level remain uncalibrated.

For spatial and `phylo_diag` models, current simulation and bootstrap fallbacks
may condition on fitted random effects. The returned intervals can therefore
be too narrow. Read the warning and do not report those intervals as calibrated
unconditional uncertainty.

VA interval methods remain labelled `calibrated = FALSE`. AGHQ's node and
ridge choices were evaluated for numerical failure and runaway avoidance in a
narrow regime, not for general interval coverage.

## Check every fit before interpretation

At minimum, run:

```{r fit-health}
check_gllvmTMB(fit)
diagnostic_table(fit)
```

Inspect optimizer status, raw and scaled gradients, Hessian status, boundary
flags, weak latent axes, non-finite estimates, restart history, and the
diagnostics specific to the response family. Procedural convergence and a
positive-definite Hessian are useful signals, but neither proves that the
scientific estimand was recovered.

In the frozen production study, the observable terminal-status detector had
sensitivity 0.2375 and specificity 0.9627: it missed 61 of 80 catastrophic
truth errors, all in the tested NB2 cells. Treat diagnostics as triage signals,
not as certificates that a scientifically plausible estimate is accurate.

See [Can I trust this fit?](fit-diagnostics.html) for the first diagnostic
workflow and [Convergence and start values](convergence-start-values.html) when
the fit is weak, unstable, or boundary-pinned.

## When another package is a better choice

| Your main task | Start with |
|---|---|
| A single-response GLMM | `glmmTMB` |
| A single-response spatial model | `sdmTMB` |
| One- or two-response distributional regression | `drmTMB` |
| An established matrix-interface GLLVM workflow using VA, EVA, or Laplace | `gllvm` |
| Bayesian multivariate posterior inference | `MCMCglmm` or `brms` |
| A stacked-trait multivariate likelihood with explicit covariance modes | `gllvmTMB`, within the boundary above |

Next, fit the [simplest Gaussian GLLVM](morphometrics.html), use the
[function map](function-map-cheatsheet.html) to find an operation, or consult
the [response-family guide](response-families.html) before choosing a
non-Gaussian likelihood.
