---
title: "Current limitations and boundaries"
description: "Decide which gllvmTMB models and results are ready for your analysis, which remain experimental, and what to check before interpretation."
tier: 1 # Primary Getting Started safety and scope page for first-time users.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Current limitations and boundaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

## Can I use gllvmTMB for this analysis?

`gllvmTMB` models several responses measured on the same sampling units, such
as species counts at sites or traits measured on individuals. A **generalized
linear latent-variable model (GLLVM)** uses a few unobserved variables to
describe how responses vary together after accounting for measured predictors.
It can help ask which species respond similarly to an environmental gradient,
or how much variation is shared among traits. Shared variation alone
does not identify a cause or demonstrate an interaction between species.

For a first runnable example, follow
[Get started: fit continuous morphological traits](gllvmTMB.html#first-fit-wide-individual-by-trait-morphology).
It loads data included with the package, fits a model, checks the fit, and reads
trait covariances. Both wide and long data are shown. This is a teaching example;
running it successfully does not establish reliability for a new dataset.

`gllvmTMB` is experimental. Its formulas and summaries may change, and support
for a model does not mean that all its estimates and intervals have been shown
to be reliable. Before reporting a result:

- Match the response model to your data. Continuous measurements, counts, and
  presence/absence need different choices; see the
  [response-family guide](response-families.html).
- [Check the fitted model](fit-diagnostics.html). A fit can appear numerically
  healthy and still give poor estimates, particularly for some negative-binomial
  count models and models with missing responses.
- Read the evidence for your model in the table below. Results for a simple
  model do not automatically apply after adding a phylogeny, spatial structure,
  or another source of variation.
- Treat uncertainty separately from point estimates. An available confidence
  interval does not guarantee that it contains the true value as often as its
  stated confidence level implies. See [Current uncertainty limits](#current-uncertainty-limits).

For continuous traits, the strongest starting evidence here is for Gaussian
models with independent trait variation (`indep()`) or a full trait covariance
(`dep()`), in the tested three-trait complete-data examples. The Gaussian latent
model used in the introductory tutorial has not passed all its recovery checks.
For counts and binary responses, use the matching row below; the Gaussian
results do not establish reliability for those data.

## Choose a model within the evidence

This page describes the 0.7.1 source version. It separates three practical
decisions:

- **A safer starting point within the tested conditions:** point-estimate
  checks passed for the model and data design stated. This does not establish
  reliability for other designs or for confidence intervals.
- **Exploratory only:** you can fit and describe the stated model, but its
  accuracy is incomplete or a relevant check failed.
- **Do not use it for this purpose:** the requested conclusion is unsupported,
  or known failures make that interpretation unreliable.

Match the *whole* model to the evidence: its response family, number of
responses, sample size, covariance structure, and missing-data pattern all
matter. Adding a phylogeny, spatial term, or other component does not inherit
the evidence for the simpler model. The technical record behind these checks is
available in [Validation oracles](validation-oracles.html).

## Choose a response model

For the ordinary latent route, `latent(..., unique = TRUE)` is the default and
represents
\(\Sigma = \Lambda\Lambda^\mathsf{T} + \Psi\), where \(\Psi\) is diagonal.
Setting `unique = FALSE` requests the narrower loadings-only subset.

| Your model or question | What you can use | Important limit |
|---|---|---|---|
| Gaussian `indep()` | **Safer starting point:** trait-specific variance point estimates after fit-health checks, for complete three-trait data in the tested `n = 60` and `n = 240` designs | No calibrated uncertainty, broad robustness claim, or guarantee at other sample sizes or boundary conditions |
| Gaussian `dep()` | **Safer starting point:** a full trait-covariance point estimate after fit-health checks, for complete three-trait data in the tested `n = 60` and `n = 240` designs | A positive-definite Hessian does not show that every covariance is accurate, and this evidence does not transfer to structured models |
| Gaussian `latent()` | **Exploratory:** rotation-invariant covariance (`Sigma`), correlations, communality, and an ordination after diagnostics | The `n = 60` central design was not ready for release and correlation-stress designs failed accuracy checks. Do not attach biological meaning to raw loading signs or orientation, or assume reliability near weak-variance or boundary cases |
| Poisson-log rank-1 `latent()` | **Exploratory point estimates:** rotation-invariant covariance summaries for clean three-trait data in the tested `n = 100` and `n = 300` designs | A rare-level design had excessive fixed-effect bias. This does not establish general robustness, interval reliability, or dependable failure detection |
| NB2-log rank-1 `latent()` | **Exploratory:** report point results with their exact sample size and dispersion diagnostics; the `n = 300` central design passed | Do not make a dependable point-estimation claim: `n = 100` and missing-response designs had severe dispersion errors that diagnostics often missed |
| NB2-log latent model with missing responses | **Do not use for dependable results.** The tested model is useful only for studying the failure mechanism | Standard optimizer, Hessian, and boundary diagnostics missed 30 of 41 severe truth errors in the tested missing-response design |
| Binomial(10)-logit rank-1 `latent()` | **Exploratory point estimates:** rotation-invariant covariance summaries for clean three-trait data in the tested `n = 100` and `n = 300` designs | No broad robustness claim, Bernoulli equivalence, separation protection, or reliability in the small-sample designs where failures could be silent |
| Other response families | **Exploratory:** only within the conditions in the response-family and model-specific guides | Constructing a family, or obtaining one successful fit, does not establish broad recovery |
| Ordinal models | **Do not use for a validated conclusion.** Exploratory point fits need model-specific validation | Nominal convergence does not rule out a degenerate fit, and binomial diagnostic thresholds do not apply to ordinal fits |
| Mixed response families across traits | **Exploratory:** documented combinations with one family and link per trait | Do not directly compare correlations that live on incompatible latent or response scales |
| Family or link varying between rows of one trait | **Do not use.** The fit now stops before optimisation | An accepted per-row family identifier does not make a shared loading coherent across Poisson-log and binomial-logit/probit. The planned Poisson-log/binomial-cloglog integrated route is not yet validated |

## Check additional model components

<details>
<summary>More complex structure: covariance, time, groups, and missing data</summary>

| Your model or question | Availability | What you can use | Important limit |
|---|---|---|---|
| Non-unit likelihood weights | Not recommended for likelihood-based inference | Weighted-objective point estimates only, with the weighting rule reported | Ordinary Hessian/Wald uncertainty, `logLik()`, AIC, BIC, or likelihood-ratio interpretation; no sandwich route is certified |
| Phylogenetic, animal, spatial, or kernel covariance | Experimental/partial | Point estimates for the exact documented route after structure-specific diagnostics | That ordinary-model evidence transfers to a structured source, that spatial `tau` or field SD has validated absolute-scale recovery, or that structured intervals are calibrated |
| Gaussian temporal covariance | Experimental/partial | Use the [temporal guide](temporal-ar1.html) only for repeated Gaussian measurements with complete panels, at least three traits, and three occasions per series | General recovery, calibrated intervals, temporal slopes, higher rank, or non-Gaussian support |
| Temporal plus another source | Experimental/partial | A narrow Gaussian AR1 model plus one kernel, phylogenetic, animal, or spatial source; see the temporal guide for the required complete-panel design | Broad recovery or coverage, OU combinations, multiple structured sources, source-by-time interactions, and spatial forecasting |
| Random slopes or several grouping tiers | Experimental/partial | Only the exact family, mode, slope-count, and grouping combinations with documented recovery evidence, with the covariance block named explicitly | A grid-wide claim across families, modes, slope counts, or grouping levels |
| Missing response cells | Experimental/partial | Point estimates after confirming the intended cells were omitted from the likelihood; the strongest worked evidence is Gaussian | That every family has equal recovery depth or that missingness is ignorable by default |
| One modelled missing predictor with `mi()` | Experimental/partial | Native-Laplace results for one explicitly modelled predictor in the documented predictor-family routes | VA support, several `mi()` predictors, general MNAR handling, or missing grouping variables |

</details>

<details>
<summary>Alternative fitting methods and the optional Julia connection</summary>

| Your model or question | Availability | What you can use | Important limit |
|---|---|---|---|
| Variational approximation | Experimental/partial | Opt-in exploratory point estimates for the documented families and ranks, using the current automatic GH evaluator with `H = 7`, and report the method metadata and exact family/rank | General point recovery outside those families and ranks; its ELBO is not a log-likelihood, so ordinary `logLik()`, AIC, BIC, and likelihood-ratio comparisons are undefined; do not infer calibrated covariance, Wald intervals, or evidence equivalent to Laplace |
| Adaptive Gauss-Hermite quadrature | Experimental/partial | Opt-in sensitivity analysis in its documented scalar-random-effect regime, stating the ridge setting | Ordinary ML interpretation: AGHQ defaults to a `tau = 2` penalised MAP fit, so ordinary AIC, BIC, and likelihood-ratio interpretations do not apply; the evaluated node/ridge rule is not a generally calibrated estimator |
| EVA | Not recommended for this claim | No public estimator is available | That an implementation detail implies an available estimator |
| Julia engine | Experimental/partial | Optional-bridge results for the documented model types; start with [Julia bridge setup](../reference/gllvm_julia_setup.html) | Full R/TMB feature parity or a required dependency for the ordinary workflow |

</details>

<details>
<summary>Specialised response-column, integrated, and multinomial routes</summary>

| Your model or question | Availability | What you can use | Important limit |
|---|---|---|---|
| Response-column coefficient models (`column_coef()`, `phylo_coef()`, `animal_coef()`, `kernel_coef()`, `spatial_coef()`) | Experimental/partial | Gaussian point estimates for a random intercept, slope, or both, varying across response columns, fitted through long or `traits(...)` wide data. `phylo_coef()` and `kernel_coef()` can fix a numeric mixture strength or estimate one interior value; `animal_coef()` fixes it; `spatial_coef()` fixes it at 1 and estimates the spatial range | Calibrated intervals, non-Gaussian responses, an estimated mixture strength for the animal or spatial routes, or recovery outside the tested Gaussian cells |
| Prediction at new locations for integrated (iSDM) fits (`predict()` on `isdm_sources()` fits) | Experimental/partial | Point predictions, including maps, at new in-hull coordinates for the ordinary intercept-only spatial tier; training-row predictions reproduce the fitted values exactly | Held-out map accuracy, which has not been demonstrated in the evaluated design; a standard error or confidence interval on `newdata` predictions (`se.fit` is refused there); or predictions for spatial slope tiers, which remain unsupported |
| Predictor-informed latent scores (`latent(..., lv = ~ x)`) | Experimental/partial | Point estimates of the rotation-invariant effect of a predictor acting through the shared latent axes, combining traits from more than one response family in one model, in the specific tested construction and evidence cells | Broad recovery or calibration across arbitrary family combinations; support for missing latent-score predictors, combined fixed-and-latent-score predictors, REML, or a structured (phylogenetic/spatial/kernel) source on this term |
| Multinomial structured routes | Experimental/partial | Fixed-effect point estimates and probability predictions for one unordered multinomial response are well tested. Several structured extensions (a phylogenetic route, a shared-latent cross-family route, pedigree- or kernel-sourced routes, and ordinary group random intercepts) exist and each has its own narrower tested evidence | Fitting more than one multinomial response in the same model; augmented (intercept-and-slope) structured forms; or calibrated intervals on any structured multinomial route |
| Estimated source strength (`rho = NULL` on `phylo_*`/`animal_*`/`kernel_*`/`spatial_*` terms) | Experimental/partial | Bounded point-recovery evidence for a native Gaussian model with complete replicated trait vectors, one known relatedness source, retained observation-level residual variance, and no competing ordinary covariance term | General recovery across other source/covariance-mode combinations — most combinations tested so far did not pass; any confidence interval on the estimated strength, which does not yet exist; or recovery for augmented (intercept-and-slope) forms |

</details>

## What to interpret first

For latent models, start from quantities that do not change when the latent
axes are rotated:

1. total covariance, `Sigma`;
2. correlations derived from `Sigma`;
3. shared covariance, `Lambda Lambda^T`;
4. the diagonal unique component, `Psi`;
5. communality.

Raw loading columns can change sign, order, or orientation without changing the
fitted covariance. Use them for an ordination only after applying a stated
rotation convention; do not treat an arbitrary raw orientation as biology.

## Current uncertainty limits

Wald, bootstrap, and profile methods are available for named targets, but their
availability does not establish repeated-sampling coverage. Report the target,
method, and returned status; calibrated uncertainty is not recommended as a
general claim. The narrow loading-interval exception is described below.

`profile_ci_total_variance()` is a penalty-profile approximation, not a
certified exact likelihood-ratio profile. Historical ordinary Gaussian
`n_units = 150`, `d` in `{1, 2}` cells and the newer `n_units = 400`, `d = 2`
cell met their prespecified numerical coverage criteria, but the retained
endpoints do not
prove constrained-refit convergence and exact target attainment. Every
computed row remains an approximate calculation rather than a certified exact
profile. The earlier 0.94 coverage result does not justify exact-profile
language until the mechanism is repaired and recalibrated.

There is one exact loading-interval exception. Standardized symmetric
joint-delta Wald intervals are certified for the structurally free strict-lower
targets in native pinned unrotated ordinary Gaussian three-trait cells
`(n_units=150,d=2)`, `(n_units=400,d=1)`, and `(n_units=400,d=2)`, at 95%.
This is one frozen DGP: trait intercepts `(-0.20, 0.10, 0.25)`, unique standard
deviations `(0.70, 0.80, 0.90)`, and loading vector `(0.80, 0.45, -0.35)` for
`d=1`, with second column `(0, 0.70, 0.40)` for `d=2`. Coverage is conditional on eligible fits
(optimizer convergence, converged fit health, available
`sdreport()`, and a positive-definite Hessian); availability was 98.82%,
93.38%, and 96.18% in the three certified cells. The `(n_units=150,d=1)` cell
failed. No other loading, unique-variance, or intercept regime inherits the
result. Pinned diagnostic rows, Fisher-z Wald,
arbitrary constraints, rotated fits, and every neighbouring family, tier,
rank, sample size, or confidence level remain uncalibrated.

For spatial and `phylo_diag` models, current simulation and bootstrap fallbacks
may condition on fitted random effects. The returned intervals can therefore
be too narrow. Read the warning and do not report those intervals as calibrated
unconditional uncertainty.

VA interval methods remain labelled `calibrated = FALSE`. AGHQ's node and
ridge choices were evaluated for numerical failure and runaway avoidance in a
narrow regime, not for general interval coverage.

## Check every fit before interpretation

At minimum, run:

```{r fit-health}
check_gllvmTMB(fit)
diagnostic_table(fit)
```

Inspect optimizer status, raw and scaled gradients, Hessian status, boundary
flags, weak latent axes, non-finite estimates, restart history, and the
diagnostics specific to the response family. Procedural convergence and a
positive-definite Hessian are useful signals, but neither proves that the
scientific estimand was recovered.

In the tested NB2 examples, the observable terminal-status detector had
sensitivity 0.2375 and specificity 0.9627: it missed 61 of 80 catastrophic
truth errors. Treat diagnostics as triage signals, not as certificates that a
scientifically plausible estimate is accurate.

See [Can I trust this fit?](fit-diagnostics.html) for the first diagnostic
workflow and [Convergence and start values](convergence-start-values.html) when
the fit is weak, unstable, or boundary-pinned.

## When another package is a better choice

| Your main task | Start with |
|---|---|
| A single-response GLMM | `glmmTMB` |
| A single-response spatial model | `sdmTMB` |
| One- or two-response distributional regression | `drmTMB` |
| An established matrix-interface GLLVM workflow using VA, EVA, or Laplace | `gllvm` |
| Bayesian multivariate posterior inference | `MCMCglmm` or `brms` |
| A stacked-trait multivariate likelihood with explicit covariance modes | `gllvmTMB`, within the boundary above |

Next, fit the [simplest Gaussian GLLVM](morphometrics.html), use the
[function map](function-map-cheatsheet.html) to find an operation, or consult
the [response-family guide](response-families.html) before choosing a
non-Gaussian likelihood.
