---
title: "Data-shape flowchart: which worked example matches your data?"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Data-shape flowchart: which worked example matches your data?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

The `gllvmTMB` worked-example library covers many situations, and
choosing the right starting article is mostly a question of **what
shape your data has**, not which model machinery you want. This page
maps the common data shapes to specific worked examples in one
flowchart, so you can land on the right page without reading
everything.

If you already know the *question* you want to answer (e.g. "is there
phylogenetic signal in these traits?") rather than the data shape,
start with the [Choose your model](choose-your-model.html) page
instead. This page is the data-first companion. New to the
vocabulary? See [gllvmTMB vocabulary](gllvm-vocabulary.html) first.

## The flowchart

```{r flowchart, echo = FALSE, eval = requireNamespace("DiagrammeR", quietly = TRUE)}
DiagrammeR::mermaid("
graph TD
  A[Start: what is each row in your data?]
  A -->|One row per individual, several traits as columns| B[Cross-sectional traits, no grouping]
  A -->|One row per site x species observation| C[Site x species long format]
  A -->|One row per respondent x item| D[Survey / IRT data]
  A -->|One row per species, traits = columns| E[Species-level constants]
  A -->|Pedigree available: id / sire / dam| F[Animal-model quantitative genetics]

  B --> B1[morphometrics<br/>Rank-2 GLLVM, size + shape axes]
  C --> C1{Phylogeny on species?}
  C1 -->|No| C2{Spatial autocorrelation between sites?}
  C1 -->|Yes| C3[phylogenetic-gllvm<br/>Paired phylo + non-phylo decomposition]
  C2 -->|No| C4{Repeated measures on same individual?}
  C2 -->|Yes| C5[joint-sdm<br/>SPDE / mesh, presence-absence]
  C4 -->|No| C6[functional-biogeography<br/>Trait integration, between + within site]
  C4 -->|Yes| C7[behavioural-syndromes<br/>Repeatability, between vs within individual]
  D --> D1{Items mix Likert + binary + counts?}
  D1 -->|Yes| D2[psychometrics-irt<br/>+ mixed-response<br/>CFA + per-row family dispatch]
  D1 -->|No| D3[psychometrics-irt<br/>CFA with one item family]
  E --> E1[corvidae-two-stage<br/>Stage-1 per-trait fits, stage-2 meta-V]
  F --> F1[animal_*() keywords<br/>V_A via pedigree-derived A matrix<br/>Worked example: v0.3.0 ANI-09]
", height = 750)
```

A static fallback in case the JavaScript-rendered diagram doesn't
load:

```
Start: what is each row in your data?

  1) One row per individual, several traits as columns
     -> morphometrics                          (Rank-2 GLLVM)

  2) One row per (site x species) observation
     2a) No phylogeny, no spatial, no repeated measures
         -> functional-biogeography            (between + within site)
     2b) No phylogeny, no spatial, repeated measures on individuals
         -> behavioural-syndromes              (repeatability)
     2c) No phylogeny, spatial autocorrelation between sites
         -> joint-sdm                          (SPDE / mesh)
     2d) Phylogeny on species
         -> phylogenetic-gllvm                 (paired phylo decomposition)
     2e) Mixed response families per row
         -> mixed-response                     (Gaussian + binomial + Poisson)

  3) One row per (respondent x item) -- survey / IRT
     -> psychometrics-irt                      (CFA, optionally mixed-family)

  4) One row per species, traits are columns (AVONET-style)
     -> corvidae-two-stage                     (stage-1 per-trait + stage-2 meta-V)

  5) Pedigree available (3-column id / sire / dam table)
     -> animal_*() keywords                    (V_A via pedigree-derived A;
                                                v0.3.0 worked example, ANI-09)
```

## Notes on the branches

**Cross-sectional traits, no grouping** ([morphometrics](morphometrics.html))
&mdash; The simplest case. One observational level (individuals),
several continuous traits per individual, shared low-rank covariance
between traits. No phylogeny, no spatial structure, no nesting. This
is the foundation for everything else; if your data fits here, start
here even if you eventually need a more complex model.

**Site x species long format, between + within decomposition**
([functional-biogeography](functional-biogeography.html)) &mdash; Each
row is one species's trait value at one site. The "trait integration"
question asks whether trait covariance is mostly **between** sites
(do entire sites have a syndrome) or **within** sites (do co-occurring
species share a syndrome regardless of site). The paired
`latent() + unique()` decomposition fits both layers.

**Site x species with repeated measures**
([behavioural-syndromes](behavioural-syndromes.html)) &mdash; Same data
shape as functional-biogeography, but with repeated observations on
the same individual. The model decomposes variance into
between-individual (the "personality" axis) and within-individual
(the residual / plasticity axis); the canonical behavioural-ecology
summary is per-trait *repeatability* $R_t$.

**Site x species with spatial autocorrelation**
([joint-sdm](joint-sdm.html)) &mdash; When sites have spatial
coordinates and species-presence patterns are spatially clustered, the
SPDE / mesh approximation models the between-site covariance as a
continuous spatial Gaussian field rather than an independent random
effect. This is the spatial-ecology variant of functional-biogeography.

**Site x species with phylogeny on species**
([phylogenetic-gllvm](phylogenetic-gllvm.html)) &mdash; When species
share a phylogenetic tree, the canonical model is the paired
phylogenetic + non-phylogenetic decomposition (four pieces:
$\boldsymbol{\Lambda}_{\rm phy}$, $\boldsymbol{\Psi}_{\rm phy}$,
$\boldsymbol{\Lambda}_{\rm non}$, $\boldsymbol{\Psi}_{\rm non}$).
For small trees, a [three-piece fallback]
(phylogenetic-gllvm.html#three-piece-fallback-when-the-paired-form-is-under-identified)
is sometimes more appropriate. The output of `extract_phylo_signal()`
reports the partition $H^2 + C^2_{\rm non} + \Psi = 1$.

**Mixed-family rows in one fit**
([mixed-response](mixed-response.html)) &mdash; Use when one trait is
continuous, another is binary, another is a count, and you want them
to share latent structure. Pass `family = list(gaussian(),
binomial(), poisson())` with a `family_var` column.

**Survey / IRT data**
([psychometrics-irt](psychometrics-irt.html)) &mdash; One row per
(respondent, item). The article shows a Confirmatory Factor Analysis
(CFA) with the lambda-constraint mechanism: pinning the block-
diagonal off-pattern entries to zero. Mixing item families (Likert
items as Gaussian, yes/no items as binomial) uses the same `family =
list(...)` dispatch as `mixed-response`.

**Species-level constants** (corvidae-two-stage) &mdash; When each
species has one trait value (no within-species replication), the
one-shot stacked-trait GLLVM is poorly informed. The recommended
workflow is two-stage: (1) per-trait univariate fits to extract
site-level estimates and their SEs; (2) a multivariate
meta-regression on the stage-1 estimates using `meta_known_V()` to
propagate the stage-1 uncertainty into the stage-2 covariance.

**Pedigree available — animal-model quantitative genetics** &mdash;
When rows are individuals and you have a three-column pedigree
(id, sire, dam), use the `animal_*()` keyword family. This is the
canonical QG animal model: the random effect's covariance is the
pedigree-derived additive-genetic relatedness matrix
$\mathbf{A}$ (Henderson 1976). Pass the pedigree directly:
`animal_latent(individual, d = K, pedigree = ped)`, or pre-compute
with `pedigree_to_A(ped)` and pass `A =`. The full grid
(`animal_scalar`, `_unique`, `_indep`, `_dep`, `_latent` plus
`animal_slope` for random regression / reaction norms) mirrors
`phylo_*()` 1-to-1. See [Design 14](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/14-known-relatedness-keywords.md)
for the math and the A-vs-V naming boundary (A for relatedness; V
is reserved for `meta_known_V()` sampling variance). A full worked
example with simulation recovery is v0.3.0 work (ANI-09).

## When the branches overlap

Real data often fits more than one branch. Some examples:

- **Site x species + phylogeny + spatial**: use the
  [functional-biogeography](functional-biogeography.html) article as
  the base (between/within decomposition), then add a `phylo_*()`
  term plus a `spatial_*()` term. The keyword grid composes; each
  tier gets its own paired decomposition.
- **Mixed-family + CFA structure**: combine the
  [mixed-response](mixed-response.html) recipe (per-row `family =
  list(...)`) with the
  [lambda-constraint](lambda-constraint.html) recipe (pinned
  off-pattern entries). Both arguments accept their pieces
  simultaneously.
- **IRT + repeated measures**: the
  [behavioural-syndromes](behavioural-syndromes.html) example
  generalises trivially; the "individual" is now "respondent" and
  the "trait" is now "item". The model machinery doesn't change.

## See also

- [Choose your model](choose-your-model.html) -- the question-first
  decision tree (this page is the data-shape-first companion).
- [gllvmTMB vocabulary](gllvm-vocabulary.html) -- plain-English
  glossary for the terms used in each branch.
- [API keyword grid](api-keyword-grid.html) -- the 4x5 covariance
  keyword reference (animal · phylo · spatial · none × scalar ·
  unique · indep · dep · latent).
