---
title: "Explaining Latent Ecological Axes With Predictors"
description: "Relate an observed environmental gradient to a shared latent ecological axis, then interpret the rotation-invariant trait-scale effect."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Explaining Latent Ecological Axes With Predictors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Scientific question

Suppose an ecologist has measured three continuous community traits at each
site. Does a moisture gradient accompany a coordinated change across those
traits, beyond the unexplained site-to-site position on the same shared
ecological axis?

This article fits a **predictor-informed latent variable**: moisture shifts the
mean of a latent ecological score, while a standard-normal innovation retains
unexplained differences among sites. The target for interpretation is the
trait-scale effect $B_{lv}$, not the arbitrary orientation of a fitted latent
axis. The result describes an association between moisture and community
traits; the model and these simulated data do not identify a causal effect.

## Simulate one healthy teaching example

The example has 72 sites, three Gaussian responses, and one latent axis. We
simulate this fixture directly because its known $B_{lv}$ and score components
are the quantities that this article needs to explain. The seed, factor levels,
and predictor values are fixed, so the long and wide data frames contain the
same observations.

```{r setup}
library(gllvmTMB)

set.seed(7301)

n_sites <- 72L
trait_names <- c("leaf_area", "wood_density", "seed_mass")
site_names <- sprintf("site_%02d", seq_len(n_sites))
moisture <- as.numeric(scale(seq(-1.5, 1.5, length.out = n_sites)))

Lambda <- matrix(c(0.70, -0.45, 0.55), ncol = 1,
                 dimnames = list(trait_names, "LV1"))
alpha <- 0.65
trait_intercept <- c(0.10, -0.05, 0.08)
psi <- c(0.18, 0.14, 0.16)

score_innovation <- rnorm(n_sites)
score_mean <- moisture * alpha
score_total <- score_mean + score_innovation

mu <- outer(score_total, Lambda[, 1])
mu <- sweep(mu, 2, trait_intercept, "+")
Y <- mu + matrix(
  rnorm(n_sites * length(trait_names),
        sd = rep(psi, each = n_sites)),
  nrow = n_sites,
  dimnames = list(site_names, trait_names)
)

df_wide <- data.frame(
  site = factor(site_names, levels = site_names),
  moisture = moisture,
  Y,
  check.names = FALSE
)

df_long <- data.frame(
  site = factor(rep(site_names, each = length(trait_names)),
                levels = site_names),
  trait = factor(rep(trait_names, times = n_sites),
                 levels = trait_names),
  moisture = rep(moisture, each = length(trait_names)),
  value = as.vector(t(Y))
)

head(df_long, 6)
head(df_wide, 3)
```

For site $i$, the score model is

\[
z_i = M_i\alpha + e_i, \qquad e_i \sim N(0,1).
\]

Here $M_i$ is the site's numeric moisture value. The loading matrix
$\Lambda$ maps the score to the three response traits. The ordinary native
`latent()` model includes a diagonal unique-variance companion $\Psi$ by
default, so its unit-tier covariance is

\[
\Sigma = \Lambda\Lambda^\top + \Psi.
\]

The moisture association on the trait linear-predictor scale is

\[
B_{lv} = \Lambda\alpha^\top.
\]

Unlike the separate fitted values of $\Lambda$ and $\alpha$, $B_{lv}$ does
not change under an orthogonal rotation of the latent axis. It is therefore
the appropriate quantity for comparing the scientific association across
fits.

## Fit the same model from long or wide data

Both calls use `gllvmTMB()`. The long formula names each response row with
`trait`; the compact wide formula names the three response columns with
`traits()`. Moisture appears only in `lv = ~ moisture`. It is deliberately not
also included as an ordinary fixed effect, because the combined fixed-effect
and latent-score route is outside the supported model taught here.

```{r fit-long}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | site, d = 1, lv = ~ moisture),
  data = df_long,
  unit = "site",
  trait = "trait",
  control = gllvmTMBcontrol(
    se = TRUE,
    optimizer = "optim",
    optArgs = list(method = "BFGS")
  )
)
```

```{r fit-wide}
fit_wide <- gllvmTMB(
  traits(leaf_area, wood_density, seed_mass) ~ 1 +
    latent(1 | site, d = 1, lv = ~ moisture),
  data = df_wide,
  unit = "site",
  control = gllvmTMBcontrol(
    se = TRUE,
    optimizer = "optim",
    optArgs = list(method = "BFGS")
  )
)
```

The remainder of the article uses `fit_long`. The two calls are equivalent
data shapes for the same model; they are not two independent analyses.

## Interpret the trait-scale association

`summary(fit_long)` includes the latent predictor coefficients by axis and
the corresponding trait-scale effects. The axis coefficients depend on the
orientation of the fitted axes; compare trait-scale effects across fits.
For a binary probit model, a trait-scale effect describes a change on the
probit linear-predictor scale, rather than a constant change in probability.

For loading-ridge fits, retained penalised-Hessian metadata supplies local
posterior standard deviations for the axis coefficients and delta-method
uncertainty for trait effects. This approximation is conditional on the
chosen ridge scale; it does not establish repeated-sampling coverage.
The `std.error` column then holds an approximate posterior SD; `lower` and
`upper` are normal-approximation credible bounds, not sampling confidence
intervals.
Older fits without that metadata return LV point estimates with unavailable
standard errors and bounds. Inspect `uncertainty_status` before interpreting
any uncertainty column.
For a categorical predictor, use `lv = ~ group`. Default treatment coding
compares each level with the reference level, which you can choose with
`relevel()` before fitting. For example, `Fox_NativenessNative` represents
Native minus Introduced when Introduced is the reference. The corresponding
trait-scale coefficient is that contrast on each word's link scale.
Explicit `lv = ~ 0 + group` creates full group indicators and is rejected
because they span a constant under the no-LV-intercept contract.

```{r latent-summary}
summary(fit_long)
```

`extract_lv_effects(type = "trait_effect")` returns $B_{lv}$, with one row
per trait and LV predictor. For this native Gaussian route, `method = "wald"`
requests delta-method standard errors and Wald bounds from the fitted model.

```{r trait-effect}
B_hat <- extract_lv_effects(
  fit_long,
  type = "trait_effect",
  method = "wald",
  conf.level = 0.95
)

B_hat[c("trait", "predictor", "estimate", "std.error", "lower", "upper")]
```

With one latent axis, both `ordiplot(fit_long)` and
`plot(fit_long, type = "ordination")` show a one-dimensional display.
There is no second axis. Score uncertainty is currently unavailable for
predictor-informed latent fits, so keep `ellipse = FALSE`.

A loading ridge can be requested with
`control = gllvmTMBcontrol(loading_ridge = 2)`. The value is the prior standard
deviation $\tau$ on free ordinary between-unit loadings: $\tau = 2$ means independent
$N(0, 2^2)$ priors. Smaller positive values apply stronger shrinkage, and
`loading_ridge = Inf` disables the penalty. For unstable binomial fits with ordinary between-unit latent loadings,
an explicit $\tau = 2$ fit is a starting sensitivity analysis. The
ordinary Laplace default remains unpenalised; evidence for this sensitivity
choice does not justify changing the default across response families.
A ridged fit is a penalised estimate. Do not compare it with an unpenalised
fit using ordinary AIC, BIC, or a likelihood-ratio test.

Because Gaussian responses use the identity link, an estimate is the fitted
change in that trait's mean for a one-unit increase in the standardized
moisture gradient through the shared latent axis. A positive estimate means
the trait increases along the moisture-associated component; a negative
estimate means it decreases. The signs and magnitudes belong to the named
traits and remain interpretable even if the fitted latent axis is rotated or
sign-flipped.

These are Wald summaries, not a promise that every fitted model has calibrated
95% coverage. Retained simulation evidence supports the named complete-response
Gaussian regime, with failed Hessian or unavailable-interval attempts counted
separately. Do not transfer that evidence to a different family, rank, missing
data pattern, or covariance tier.

## A categorical predictor: 35 articles and 22 words

The following complete example uses your own long-format CSV with columns
`word_present` (0 or 1), `analysis_word`, `article_id`, and `Fox_Nativeness`.
It is not run when building this article because the input is user data.
Replace the file path with your local CSV. The checks require 35 articles,
22 words, and one nativeness category per article. Keep the raw CSV outside
the package repository.

Use one spelling consistently: the actual response-column name,
`analysis_word`, appears in both the fixed intercepts and the latent loading
block. If you rename the data column to `trait`, use `trait` in both places
and set `trait = "trait"`. The placeholder `trait`
inside `latent()` is equivalent when `trait = "analysis_word"` is supplied.
This equivalence also holds with the LV predictor and loading prior.

Choose Introduced as the reference before fitting. With the default
treatment contrasts, `Fox_NativenessNative` then means Native minus
Introduced. Its trait effect is a difference on each word's probit linear
predictor scale, not a constant probability difference.

```{r fox-word-sensitivity, eval=FALSE}
library(gllvmTMB)
dat <- read.csv("path/to/fox-word-data.csv", check.names = FALSE)
needed <- c("word_present", "analysis_word", "article_id", "Fox_Nativeness")
stopifnot(all(needed %in% names(dat)), !anyNA(dat[needed]))
stopifnot(all(dat$word_present %in% c(0, 1)))
stopifnot(all(vapply(split(dat$Fox_Nativeness, dat$article_id),
                    function(x) length(unique(x)) == 1L, logical(1))))
dat$article_id <- factor(dat$article_id)
dat$analysis_word <- factor(dat$analysis_word)
dat$Fox_Nativeness <- relevel(factor(dat$Fox_Nativeness), ref = "Introduced")
stopifnot(nlevels(dat$article_id) == 35L,
          nlevels(dat$analysis_word) == 22L,
          setequal(levels(dat$Fox_Nativeness), c("Introduced", "Native")))

set.seed(1467)
m1 <- gllvmTMB(
  word_present ~ 0 + analysis_word +
    latent(0 + analysis_word | article_id, d = 1, lv = ~ Fox_Nativeness),
  data = dat, trait = "analysis_word", unit = "article_id",
  family = binomial(link = "probit"),
  control = gllvmTMBcontrol(loading_ridge = Inf)
)
set.seed(1467)
m2 <- gllvmTMB(
  word_present ~ 0 + analysis_word +
    latent(0 + analysis_word | article_id, d = 2, lv = ~ Fox_Nativeness),
  data = dat, trait = "analysis_word", unit = "article_id",
  family = binomial(link = "probit"),
  control = gllvmTMBcontrol(loading_ridge = Inf)
)
set.seed(1467)
m2_ridge <- gllvmTMB(
  word_present ~ 0 + analysis_word +
    latent(0 + analysis_word | article_id, d = 2, lv = ~ Fox_Nativeness),
  data = dat, trait = "analysis_word", unit = "article_id",
  family = binomial(link = "probit"),
  control = gllvmTMBcontrol(loading_ridge = 2)
)

# Each summary includes the separate LV predictor coefficient block.
summary(m1)
summary(m2)
summary(m2_ridge)
fits <- list(d1 = m1, d2 = m2, d2_ridge_tau2 = m2_ridge)
trait_effects <- lapply(fits, extract_lv_effects, type = "trait_effect")
trait_effects
diagnostics <- lapply(fits, gllvmTMB_diagnose)
diagnostics

# One-dimensional scores only; vertical position is a plotting baseline.
ordiplot(m1, axes = 1L, biplot = FALSE, ellipse = FALSE)
```

Inspect the diagnostics before interpreting differences between these fits.
Compare the named-word trait effects across ranks and ridge choices;
raw-axis coefficients can change orientation. The ridge may reduce runaway
loadings, but it does not guarantee a stable fit or accurate effects for
this data set. Any finite ridged LV uncertainty is conditional on $\tau = 2$
and remains a local posterior approximation. Predictor-informed score
uncertainty is unavailable in `ordiplot()`, including for the ridged fit.

## Decompose each fitted score

The total fitted score is the sum of the moisture-informed mean and the
site-specific innovation. `extract_ordination()` exposes all three components
on the same fitted axis.

```{r score-decomposition}
ord_total <- extract_ordination(fit_long, level = "unit", component = "total")
ord_mean <- extract_ordination(fit_long, level = "unit", component = "mean")
ord_innovation <- extract_ordination(
  fit_long,
  level = "unit",
  component = "innovation"
)

score_check <- max(abs(
  ord_total$scores - ord_mean$scores - ord_innovation$scores
))
score_check

head(data.frame(
  site = rownames(ord_total$scores),
  total = ord_total$scores[, 1],
  mean = ord_mean$scores[, 1],
  innovation = ord_innovation$scores[, 1]
))
```

The identity `total = mean + innovation` checks the score decomposition. The
mean component answers how moisture is associated with the shared axis; the
innovation records remaining site-to-site position after that association.
Neither component, by itself, makes moisture a causal exposure.

## Use one latent axis across two response families

A named mixed-response route can place a continuous trait and a count trait on
the same predictor-informed axis. Here temperature is associated with both
responses through that axis. The Gaussian response uses an identity link; the
Poisson response uses a log link. The shared $B_{lv}$ estimates therefore live
on each response's linear-predictor scale, not on one common response scale.

```{r mixed-data}
set.seed(7302)

temperature <- as.numeric(scale(seq(-1.4, 1.4, length.out = n_sites)))
mixed_score <- 0.60 * temperature + rnorm(n_sites)
continuous <- rnorm(n_sites, 0.15 + 0.70 * mixed_score, 0.30)
abundance <- rpois(n_sites, exp(log(4) - 0.55 * mixed_score))

mixed_wide <- data.frame(
  site = factor(site_names, levels = site_names),
  temperature = temperature,
  continuous = continuous,
  abundance = abundance
)

mixed_long <- data.frame(
  site = factor(rep(site_names, each = 2L), levels = site_names),
  trait = factor(
    rep(c("continuous", "abundance"), n_sites),
    levels = c("continuous", "abundance")
  ),
  family = factor(
    rep(c("continuous", "abundance"), n_sites),
    levels = c("continuous", "abundance")
  ),
  temperature = rep(temperature, each = 2L),
  value = as.vector(rbind(continuous, abundance))
)

mixed_families_long <- list(
  continuous = gaussian(),
  abundance = poisson()
)
attr(mixed_families_long, "family_var") <- "family"

mixed_families_wide <- list(
  continuous = gaussian(),
  abundance = poisson()
)
```

The long call uses an explicit `family` selector column. In the `traits()`
call, the names of the family list match the selected response columns, so the
wide-to-long rewrite supplies the same selector internally.

```{r mixed-fit-long}
fit_mixed_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | site, d = 1, unique = FALSE,
           lv = ~ temperature),
  data = mixed_long,
  unit = "site",
  trait = "trait",
  family = mixed_families_long,
  control = gllvmTMBcontrol(se = FALSE)
)
```

```{r mixed-fit-wide}
fit_mixed_wide <- gllvmTMB(
  traits(continuous, abundance) ~ 1 +
    latent(1 | site, d = 1, unique = FALSE,
           lv = ~ temperature),
  data = mixed_wide,
  unit = "site",
  family = mixed_families_wide,
  control = gllvmTMBcontrol(se = FALSE)
)
```

```{r mixed-effects}
B_mixed_long <- extract_lv_effects(
  fit_mixed_long,
  type = "trait_effect"
)
B_mixed_wide <- extract_lv_effects(
  fit_mixed_wide,
  type = "trait_effect"
)

data.frame(
  trait = B_mixed_long$trait,
  predictor = B_mixed_long$predictor,
  long_estimate = B_mixed_long$estimate,
  wide_estimate = B_mixed_wide$estimate
)
```

For `continuous`, the estimate is an associated change in the Gaussian mean
per one-unit increase in standardized temperature through the shared axis. For
`abundance`, it is a change in the log expected count; exponentiating it gives
the corresponding multiplicative change. This named rank-1,
`unique = FALSE`, complete-response route has retained point-recovery evidence.
The example deliberately requests no standard errors so the teaching path does
not turn a family-specific calibration result into a general interval claim.
Separately retained simulation evidence supports target-wise `B_lv` Wald
coverage for this exact Gaussian + Poisson archetype.

> **Which analyses does the evidence support?** This article demonstrates
> complete-response Gaussian models and a Gaussian + Poisson model with one
> latent axis and `unique = FALSE`, all fitted with the native TMB engine at
> the ordinary unit level. Other supported response-family and link
> combinations can be fitted with one numeric unit-level `lv` predictor and
> no ordinary fixed-effect covariates. With `unique = FALSE`, the latent rank
> can range from one to the number of responses before any multinomial
> expansion. The default $\Psi$ component adds further restrictions to avoid
> fitting more mean and covariance parameters than the data can distinguish.
> Small rank-2 and rank-3 examples have fitted successfully across all supported
> families; these checks alone do not establish accurate estimates or reliable
> confidence intervals for every combination.
>
> Simulation evidence supports estimates and Wald intervals in the tested
> Gaussian rank-1 and rank-2 settings. For binomial-only models, interval
> evidence is limited to rank-1 models with multiple trials per observation
> and logit, probit, or cloglog links. Categorical `lv` predictors have
> point-estimate recovery evidence for Gaussian responses, but their interval
> coverage has not been established. Tested mixed-response models have
> point-estimate recovery evidence at rank 1 with `unique = FALSE`. Eight
> tested combinations containing Gaussian responses, including the Gaussian +
> Poisson example here, also have evidence for the coverage of each individual
> $B_{lv}$ Wald interval. This does not establish simultaneous coverage of all
> effects or coverage for arbitrary mixtures of response families.
>
> Beta-only models did not converge reliably enough in the recovery study;
> ordinal-probit-only models did not recover the shared covariance reliably
> enough. Treat those estimates as exploratory. Before interpreting a different
> response combination, check its family-specific evidence and use simulations
> resembling your data to assess estimation and interval coverage.
>
> The Julia bridge supports complete-response Gaussian point estimates with
> `unique = FALSE`; its optional Wald intervals have not been shown to achieve
> their stated coverage. Native ordinary `latent()` instead uses
> $\Lambda\Lambda^\top + \Psi$ by default. The examples here do not establish
> support for REML, ordinary fixed-effect covariates alongside `lv` predictors
> (including the same predictor in both), calibrated Julia intervals, broader
> missing-response patterns, non-canonical links, ranks above the number of
> responses, extra grouping levels, phylogenetic, spatial, animal, or kernel
> covariance, missing `lv` predictors, or profile and bootstrap intervals.

## What to report

For an applied analysis, report the response family and link, latent rank,
unit tier, the exact `lv` predictor, the trait-scale $B_{lv}$ estimates, and
the interval method. State that $B_{lv}$ is rotation-invariant, describe the
score innovation, and keep the conclusion associational unless the study
design supplies a separate causal argument.
