---
title: "Explaining Latent Ecological Axes With Predictors"
description: "Relate an observed environmental gradient to a shared latent ecological axis, then interpret the rotation-invariant trait-scale effect."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Explaining Latent Ecological Axes With Predictors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Scientific question

Suppose an ecologist has measured three continuous community traits at each
site. Does a moisture gradient accompany a coordinated change across those
traits, beyond the unexplained site-to-site position on the same shared
ecological axis?

This article fits a **predictor-informed latent variable**: moisture shifts the
mean of a latent ecological score, while a standard-normal innovation retains
unexplained differences among sites. The target for interpretation is the
trait-scale effect $B_{lv}$, not the arbitrary orientation of a fitted latent
axis. The result describes an association between moisture and community
traits; the model and these simulated data do not identify a causal effect.

## Simulate one healthy teaching example

The example has 72 sites, three Gaussian responses, and one latent axis. We
simulate this fixture directly because its known $B_{lv}$ and score components
are the quantities that this article needs to explain. The seed, factor levels,
and predictor values are fixed, so the long and wide data frames contain the
same observations.

```{r setup}
library(gllvmTMB)

set.seed(7301)

n_sites <- 72L
trait_names <- c("leaf_area", "wood_density", "seed_mass")
site_names <- sprintf("site_%02d", seq_len(n_sites))
moisture <- as.numeric(scale(seq(-1.5, 1.5, length.out = n_sites)))

Lambda <- matrix(c(0.70, -0.45, 0.55), ncol = 1,
                 dimnames = list(trait_names, "LV1"))
alpha <- 0.65
trait_intercept <- c(0.10, -0.05, 0.08)
psi <- c(0.18, 0.14, 0.16)

score_innovation <- rnorm(n_sites)
score_mean <- moisture * alpha
score_total <- score_mean + score_innovation

mu <- outer(score_total, Lambda[, 1])
mu <- sweep(mu, 2, trait_intercept, "+")
Y <- mu + matrix(
  rnorm(n_sites * length(trait_names),
        sd = rep(psi, each = n_sites)),
  nrow = n_sites,
  dimnames = list(site_names, trait_names)
)

df_wide <- data.frame(
  site = factor(site_names, levels = site_names),
  moisture = moisture,
  Y,
  check.names = FALSE
)

df_long <- data.frame(
  site = factor(rep(site_names, each = length(trait_names)),
                levels = site_names),
  trait = factor(rep(trait_names, times = n_sites),
                 levels = trait_names),
  moisture = rep(moisture, each = length(trait_names)),
  value = as.vector(t(Y))
)

head(df_long, 6)
head(df_wide, 3)
```

For site $i$, the score model is

\[
z_i = M_i\alpha + e_i, \qquad e_i \sim N(0,1).
\]

Here $M_i$ is the site's numeric moisture value. The loading matrix
$\Lambda$ maps the score to the three response traits. The ordinary native
`latent()` model includes a diagonal unique-variance companion $\Psi$ by
default, so its unit-tier covariance is

\[
\Sigma = \Lambda\Lambda^\top + \Psi.
\]

The moisture association on the trait linear-predictor scale is

\[
B_{lv} = \Lambda\alpha^\top.
\]

Unlike the separate fitted values of $\Lambda$ and $\alpha$, $B_{lv}$ does
not change under an orthogonal rotation of the latent axis. It is therefore
the appropriate quantity for comparing the scientific association across
fits.

## Fit the same model from long or wide data

Both calls use `gllvmTMB()`. The long formula names each response row with
`trait`; the compact wide formula names the three response columns with
`traits()`. Moisture appears only in `lv = ~ moisture`. It is deliberately not
also included as an ordinary fixed effect, because the combined fixed-effect
and latent-score route is outside the supported model taught here.

```{r fit-long}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | site, d = 1, lv = ~ moisture),
  data = df_long,
  unit = "site",
  trait = "trait",
  control = gllvmTMBcontrol(
    se = TRUE,
    optimizer = "optim",
    optArgs = list(method = "BFGS")
  )
)
```

```{r fit-wide}
fit_wide <- gllvmTMB(
  traits(leaf_area, wood_density, seed_mass) ~ 1 +
    latent(1 | site, d = 1, lv = ~ moisture),
  data = df_wide,
  unit = "site",
  control = gllvmTMBcontrol(
    se = TRUE,
    optimizer = "optim",
    optArgs = list(method = "BFGS")
  )
)
```

The remainder of the article uses `fit_long`. The two calls are equivalent
data shapes for the same model; they are not two independent analyses.

## Interpret the trait-scale association

`extract_lv_effects(type = "trait_effect")` returns $B_{lv}$, with one row
per trait and LV predictor. For this native Gaussian route, `method = "wald"`
requests delta-method standard errors and Wald bounds from the fitted model.

```{r trait-effect}
B_hat <- extract_lv_effects(
  fit_long,
  type = "trait_effect",
  method = "wald",
  conf.level = 0.95
)

B_hat[c("trait", "predictor", "estimate", "std.error", "lower", "upper")]
```

Because Gaussian responses use the identity link, an estimate is the fitted
change in that trait's mean for a one-unit increase in the standardized
moisture gradient through the shared latent axis. A positive estimate means
the trait increases along the moisture-associated component; a negative
estimate means it decreases. The signs and magnitudes belong to the named
traits and remain interpretable even if the fitted latent axis is rotated or
sign-flipped.

These are Wald summaries, not a promise that every fitted model has calibrated
95% coverage. Retained simulation evidence supports the named complete-response
Gaussian regime, with failed Hessian or unavailable-interval attempts counted
separately. Do not transfer that evidence to a different family, rank, missing
data pattern, or covariance tier.

## Decompose each fitted score

The total fitted score is the sum of the moisture-informed mean and the
site-specific innovation. `extract_ordination()` exposes all three components
on the same fitted axis.

```{r score-decomposition}
ord_total <- extract_ordination(fit_long, level = "unit", component = "total")
ord_mean <- extract_ordination(fit_long, level = "unit", component = "mean")
ord_innovation <- extract_ordination(
  fit_long,
  level = "unit",
  component = "innovation"
)

score_check <- max(abs(
  ord_total$scores - ord_mean$scores - ord_innovation$scores
))
score_check

head(data.frame(
  site = rownames(ord_total$scores),
  total = ord_total$scores[, 1],
  mean = ord_mean$scores[, 1],
  innovation = ord_innovation$scores[, 1]
))
```

The identity `total = mean + innovation` checks the score decomposition. The
mean component answers how moisture is associated with the shared axis; the
innovation records remaining site-to-site position after that association.
Neither component, by itself, makes moisture a causal exposure.

## Use one latent axis across two response families

A named mixed-response route can place a continuous trait and a count trait on
the same predictor-informed axis. Here temperature is associated with both
responses through that axis. The Gaussian response uses an identity link; the
Poisson response uses a log link. The shared $B_{lv}$ estimates therefore live
on each response's linear-predictor scale, not on one common response scale.

```{r mixed-data}
set.seed(7302)

temperature <- as.numeric(scale(seq(-1.4, 1.4, length.out = n_sites)))
mixed_score <- 0.60 * temperature + rnorm(n_sites)
continuous <- rnorm(n_sites, 0.15 + 0.70 * mixed_score, 0.30)
abundance <- rpois(n_sites, exp(log(4) - 0.55 * mixed_score))

mixed_wide <- data.frame(
  site = factor(site_names, levels = site_names),
  temperature = temperature,
  continuous = continuous,
  abundance = abundance
)

mixed_long <- data.frame(
  site = factor(rep(site_names, each = 2L), levels = site_names),
  trait = factor(
    rep(c("continuous", "abundance"), n_sites),
    levels = c("continuous", "abundance")
  ),
  family = factor(
    rep(c("continuous", "abundance"), n_sites),
    levels = c("continuous", "abundance")
  ),
  temperature = rep(temperature, each = 2L),
  value = as.vector(rbind(continuous, abundance))
)

mixed_families_long <- list(
  continuous = gaussian(),
  abundance = poisson()
)
attr(mixed_families_long, "family_var") <- "family"

mixed_families_wide <- list(
  continuous = gaussian(),
  abundance = poisson()
)
```

The long call uses an explicit `family` selector column. In the `traits()`
call, the names of the family list match the selected response columns, so the
wide-to-long rewrite supplies the same selector internally.

```{r mixed-fit-long}
fit_mixed_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | site, d = 1, unique = FALSE,
           lv = ~ temperature),
  data = mixed_long,
  unit = "site",
  trait = "trait",
  family = mixed_families_long,
  control = gllvmTMBcontrol(se = FALSE)
)
```

```{r mixed-fit-wide}
fit_mixed_wide <- gllvmTMB(
  traits(continuous, abundance) ~ 1 +
    latent(1 | site, d = 1, unique = FALSE,
           lv = ~ temperature),
  data = mixed_wide,
  unit = "site",
  family = mixed_families_wide,
  control = gllvmTMBcontrol(se = FALSE)
)
```

```{r mixed-effects}
B_mixed_long <- extract_lv_effects(
  fit_mixed_long,
  type = "trait_effect"
)
B_mixed_wide <- extract_lv_effects(
  fit_mixed_wide,
  type = "trait_effect"
)

data.frame(
  trait = B_mixed_long$trait,
  predictor = B_mixed_long$predictor,
  long_estimate = B_mixed_long$estimate,
  wide_estimate = B_mixed_wide$estimate
)
```

For `continuous`, the estimate is an associated change in the Gaussian mean
per one-unit increase in standardized temperature through the shared axis. For
`abundance`, it is a change in the log expected count; exponentiating it gives
the corresponding multiplicative change. This named rank-1,
`unique = FALSE`, complete-response route has retained point-recovery evidence.
The example deliberately requests no standard errors so the teaching path does
not turn a family-specific calibration result into a general interval claim.
Separately retained simulation evidence supports target-wise `B_lv` Wald
coverage for this exact Gaussian + Poisson archetype.

> **Current boundary.** This article teaches the native, ordinary unit-tier,
> complete-response Gaussian model and one named Gaussian + Poisson,
> rank-1, loadings-only mixed example. Registered native family/link rows now
> compose in this ordinary predictor-informed block at rank 1 through the
> number of logical responses, with automatic $\Psi$ or `unique = FALSE`;
> bounded rank-2/rank-3 canaries exercise every registered family. Those
> canaries establish route health, not general recovery or interval
> calibration. Retained native evidence covers named
> Gaussian rank-1 and rank-2 cells; interval evidence for pure binomial models
> is limited to rank-1, multi-trial logit, probit, and cloglog cells.
> Factor-valued LV predictors have point-estimate and recovery evidence, but
> their intervals are not calibrated. The Julia bridge provides
> complete-response, loadings-only (`unique = FALSE`) point estimates with
> optional uncalibrated Wald plumbing; native ordinary `latent()` instead uses
> $\Lambda\Lambda^\top + \Psi$ by default. This article does not support REML,
> an ordinary fixed predictor combined with the same `lv` predictor, Julia
> interval calibration, broader response masks, non-canonical links, ranks
> above the number of logical responses, extra grouping tiers, structured
> phylogenetic, spatial, animal, or
> kernel sources, missing LV predictors, or profile and bootstrap intervals.
> Separately named mixed-response cells have retained rank-1, loadings-only
> point-recovery evidence. Eight preregistered Gaussian-anchor archetypes also
> have target-wise Wald calibration evidence, including the Gaussian + Poisson
> example shown here; this is not a simultaneous all-target or arbitrary-mixture
> interval claim. Pure Beta and pure ordinal-probit retain separate point-
> recovery HOLDs. Consult the family-specific boundary before interpreting a
> different admitted composition.

## What to report

For an applied analysis, report the response family and link, latent rank,
unit tier, the exact `lv` predictor, the trait-scale $B_{lv}$ estimates, and
the interval method. State that $B_{lv}$ is rotation-invariant, describe the
score innovation, and keep the conclusion associational unless the study
design supplies a separate causal argument.
