---
title: "Can I trust this fit?"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Can I trust this fit?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6.2,
  fig.height = 4.2,
  fig.align  = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

Before interpreting a fitted GLLVM, separate two questions. Did the
optimizer find a numerically healthy point estimate? Does the fitted
response distribution show an obvious mismatch to the observed stacked
traits? This article answers those questions for a small count example
before moving on to covariance, ordination, repeatability, or interval
claims.

The example is deliberately modest: three behaviours counted on each
individual, fitted with one shared latent axis. `latent()` estimates a
low-rank axis of between-individual covariance; this is a diagnostic
example, not the full ordinary `latent()` covariance-decomposition
tutorial from [Morphometrics](morphometrics.html).

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

## Fit A Small Model

The data have one row per `(individual, trait)` observation. The
working question is whether the three behaviours show a shared
between-individual axis and whether the fitted Poisson response model
looks reasonable enough to inspect.

```{r simulate}
set.seed(20260609)

n_ind <- 48L
trait_names <- c("forage", "alarm", "explore")
n_traits <- length(trait_names)

z <- rnorm(n_ind, sd = 0.45)
eta <- cbind(
  forage = 0.25 + 0.75 * z,
  alarm = -0.15 - 0.45 * z,
  explore = 0.05 + 0.55 * z
)
y <- matrix(
  rpois(n_ind * n_traits, lambda = exp(as.vector(eta))),
  nrow = n_ind,
  ncol = n_traits
)

df_long <- data.frame(
  individual = factor(rep(seq_len(n_ind), each = n_traits)),
  trait = factor(rep(trait_names, n_ind), levels = trait_names),
  count = as.vector(t(y))
)

head(df_long, 8)
```

The long-format call is canonical. It names the response, the trait
column, and the unit column explicitly.

```{r fit-long}
fit_long <- gllvmTMB(
  count ~ 0 + trait +
    latent(0 + trait | individual, d = 1),
  data = df_long,
  trait = "trait",
  unit = "individual",
  family = poisson(),
  control = gllvmTMBcontrol(n_init = 2, init_jitter = 0.03)
)

fit <- fit_long
fit
```

The same model can be fitted from a wide data frame with the
`traits(...)` left-hand side. The wide call is useful when the data
start as one row per individual and one column per behaviour.

```{r fit-wide}
df_wide <- stats::reshape(
  df_long,
  idvar = "individual",
  timevar = "trait",
  direction = "wide"
)
names(df_wide) <- sub("^count\\.", "", names(df_wide))

fit_wide <- gllvmTMB(
  traits(forage, alarm, explore) ~ 1 +
    latent(1 | individual, d = 1),
  data = df_wide,
  unit = "individual",
  family = poisson(),
  control = gllvmTMBcontrol(n_init = 2, init_jitter = 0.03)
)

signif(
  abs(as.numeric(logLik(fit_long)) - as.numeric(logLik(fit_wide))),
  3
)
```

That last number should be close to zero: the two calls reach the same
stacked-trait likelihood.

## Read Fit Health

`check_gllvmTMB()` returns a machine-readable fit-health table. Start
there before interpreting coefficients, latent axes, or covariance
summaries.

```{r fit-health}
health <- check_gllvmTMB(fit)
key_components <- c(
  "optimizer_convergence",
  "max_gradient",
  "sdreport",
  "pd_hessian",
  "weak_axis_unit",
  "boundary_sigma_eps"
)
health[
  health$component %in% key_components,
  c("component", "status", "message")
]
```

Read the rows literally. Optimizer and gradient rows speak to the
point estimate. `sdreport` and `pd_hessian` speak to Hessian-based
standard errors and Wald-style inference. A weak-axis row says a
latent dimension may be hard to interpret; it does not by itself prove
that the fitted response distribution is wrong.

The same table records intentionally skipped standard errors. A no-SE
fit can still be useful for point estimates or refit-based uncertainty,
but it should not be reported as if Wald standard errors were computed.

```{r no-se-health}
fit_no_se <- gllvmTMB(
  count ~ 0 + trait +
    latent(0 + trait | individual, d = 1),
  data = df_long,
  trait = "trait",
  unit = "individual",
  family = poisson(),
  control = gllvmTMBcontrol(se = FALSE)
)

no_se_health <- check_gllvmTMB(fit_no_se)
no_se_health[
  no_se_health$component %in% c("sdreport", "pd_hessian", "max_fixed_se"),
  c("component", "status", "message")
]
```

For hard fits, move from this table to the
[Convergence and start values](convergence-start-values.html) article
before making scientific claims.

## Check Response Diagnostics

Numerical convergence is not enough. For Gaussian, Poisson, and NB2
rows, `residuals(type = "randomized_quantile")` computes exact
family-CDF residuals. `diagnostic_table()` turns the attached metadata
into ordinary data frames for reports and articles.

```{r residual-table}
rq <- residuals(
  fit,
  type = "randomized_quantile",
  seed = 20260609
)

diagnostic_table(rq, table = "row_status")
diagnostic_table(rq, table = "fit_health_status")

head(
  diagnostic_table(rq, table = "data")[
    , c("trait", "family", "observed", "residual", "status")
  ],
  8
)
```

The `row_status` table tells you whether every model row produced a
usable residual. If a row is marked `unsupported_family`,
`missing_response`, or `nonfinite_observed`, keep that status in the
report rather than silently dropping the row.

```{r residual-qq, fig.cap = "Randomized-quantile residual Q-Q check for the fitted Poisson GLLVM. This is a fitted-model diagnostic display, not interval calibration, latent-rank proof, or a Bayesian posterior predictive check."}
predictive_check(
  fit,
  type = "rq_qq",
  seed = 20260609
)
```

For count models, a rootogram is often easier to read than a Q-Q plot.
Bars near zero mean the observed and fitted-model simulated count
frequencies are similar on the square-root scale.

```{r rootogram, fig.cap = "Rootogram comparing observed behavioural counts with fitted-model simulations. Large systematic departures would point to a response-family, link, covariate, or random-effect mismatch."}
predictive_check(
  fit,
  type = "rootogram",
  ndraws = 40,
  seed = 20260609,
  max_count = 7L
)
```

## What To Do Next

Use the diagnostics as routing information, not as a single pass/fail
stamp.

| What you see | What it means | Next step |
|---|---|---|
| Optimizer, gradient, `sdreport`, and `pd_hessian` all pass | The numerical fit is a reasonable starting point | Inspect `Sigma`, correlations, ordination, or fitted effects. |
| Optimizer or gradient rows warn/fail | The point estimate may be unstable | Try more starts, rescaling, simpler rank, or an alternative optimizer. |
| `sdreport` or `pd_hessian` warns | Standard errors or Wald intervals are degraded | Keep point estimates visible; route uncertainty to profile/bootstrap where supported. |
| Residual rows have non-`ok` status | Some rows were not diagnosed | Inspect missing values, non-finite responses, and family support. |
| Q-Q or rootogram shows systematic structure | The fitted response distribution may be wrong | Revisit family, link, covariates, offsets, or random-effect structure. |

If the fit-health and response checks are acceptable but the latent rank
is still a modelling choice, continue to
[How many latent dimensions should I fit?](model-selection-latent-rank.html).
That article compares one ML candidate set with AIC/BIC and uses REML
only after a Gaussian model has been chosen.

This page covers the public diagnostic path that is currently backed by
the validation register: fit health and no-SE status (DIA-08, DIA-10),
fitted-model predictive diagnostics and residuals for the scoped
Gaussian / Poisson / NB2 rows (DIA-11, DIA-12), and report-ready
metadata tables (DIA-13). It does not calibrate confidence intervals,
prove the latent rank, run formal residual tests, or draw from a
Bayesian posterior. Exact residual support for delta, hurdle,
truncated, ordinal, and mixture-family rows remains outside this
article's current claims.
