---
title: "Can I trust this fit?"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Can I trust this fit?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6.2,
  fig.height = 4.2,
  fig.align  = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

Before interpreting a fitted GLLVM, separate two questions. Did the
optimizer find a numerically healthy point estimate? Does the fitted
response distribution show an obvious mismatch to the three behavioural
responses arranged in long format? This article answers those questions for a small count example
before moving on to covariance, ordination, repeatability, or interval
claims.

`gllvmTMB` 0.5.0 is a pre-CRAN release with an experimental lifecycle.
Use these diagnostics to decide what to inspect or refit next, not as a
certificate that every inferential target is calibrated.

The example is deliberately modest: three behaviours counted on each
individual, fitted with one shared latent variable. `latent()` estimates a
low-rank axis of between-individual covariance; this is a diagnostic
example, not the full ordinary `latent()` covariance-decomposition
tutorial from [Morphometrics](morphometrics.html).

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

## Fit A Small Model

The data have one row per `(individual, trait)` observation. The
working question is whether the three behaviours show a shared
between-individual axis and whether the fitted Poisson response model
looks reasonable enough to inspect.

```{r simulate}
set.seed(20260609)

n_ind <- 48L
trait_names <- c("forage", "alarm", "explore")
n_traits <- length(trait_names)

z <- rnorm(n_ind, sd = 0.45)
eta <- cbind(
  forage = 0.25 + 0.75 * z,
  alarm = -0.15 - 0.45 * z,
  explore = 0.05 + 0.55 * z
)
y <- matrix(
  rpois(n_ind * n_traits, lambda = exp(as.vector(eta))),
  nrow = n_ind,
  ncol = n_traits
)

df_long <- data.frame(
  individual = factor(rep(seq_len(n_ind), each = n_traits)),
  trait = factor(rep(trait_names, n_ind), levels = trait_names),
  count = as.vector(t(y))
)

head(df_long, 8)
```

The long-format call is canonical. It names the response, the trait
column, and the unit column explicitly.

```{r fit-long}
fit_long <- gllvmTMB(
  count ~ 0 + trait +
    latent(0 + trait | individual, d = 1),
  data = df_long,
  trait = "trait",
  unit = "individual",
  family = poisson(),
  control = gllvmTMBcontrol(n_init = 2, init_jitter = 0.03)
)

fit <- fit_long
fit
```

The same model can be fitted from a wide data frame with the
`traits(...)` left-hand side. The wide call is useful when the data
start as one row per individual and one column per behaviour.

```{r fit-wide}
df_wide <- stats::reshape(
  df_long,
  idvar = "individual",
  timevar = "trait",
  direction = "wide"
)
names(df_wide) <- sub("^count\\.", "", names(df_wide))

fit_wide <- gllvmTMB(
  traits(forage, alarm, explore) ~ 1 +
    latent(1 | individual, d = 1),
  data = df_wide,
  unit = "individual",
  family = poisson(),
  control = gllvmTMBcontrol(n_init = 2, init_jitter = 0.03)
)

signif(
  abs(as.numeric(logLik(fit_long)) - as.numeric(logLik(fit_wide))),
  3
)
```

The observed difference is about `2.14e-09`, which is negligible at the
reported precision: the two calls reach the same stacked-trait likelihood.

## Read Fit Health

The authoritative convergence verdict is `fit$fit_health$converged`. It uses
the largest absolute gradient divided by `1 + abs(objective)`, so the gradient
is judged relative to the scale of the fitted objective. Values below `1e-3`
count as a usable optimum. Raw optimizer codes, the unscaled maximum gradient,
and Hessian curvature remain supporting evidence rather than substitutes for
that verdict.

```{r convergence-verdict}
data.frame(
  converged = isTRUE(fit$fit_health$converged),
  scaled_gradient = signif(fit$fit_health$scaled_gradient, 4),
  threshold = 1e-3
)
```

`check_gllvmTMB()` returns the detailed, machine-readable fit-health table used
in simulations, tests, and reports. `gllvmTMB_diagnose()` instead prints a
guided narrative for an interactive first look. Start with either one before
interpreting coefficients, latent variables, or covariance summaries.

```{r fit-health}
health <- check_gllvmTMB(fit)
key_components <- c(
  "optimizer_convergence",
  "max_gradient",
  "sdreport",
  "pd_hessian",
  "weak_axis_unit",
  "boundary_sigma_eps"
)
health[
  health$component %in% key_components,
  c("component", "status", "message")
]
```

These six displayed rows pass for this example. The full table contains more
checks, so always inspect every warning rather than treating a selected subset
as the whole verdict. Here are all non-passing rows, including their thresholds,
meaning, and suggested action.

```{r fit-health-warnings}
health[
  health$status != "PASS",
  c("component", "status", "value", "threshold", "message", "action")
]
```

In this fit, the two warnings describe the same feature: the fitted
trait-specific diagonal standard deviation is almost zero. The simulation
planted a shared latent component without an additional trait-specific
between-individual component, so a boundary estimate is plausible here. It
still matters for interpretation: this example supports a shared axis, but it
does not demonstrate recovery of a separate diagonal variance. “Mapped off”
means a parameter was deliberately fixed or excluded from the fitted model,
rather than estimated near zero.

Read the rows literally. Optimizer and gradient rows speak to the
point estimate. `sdreport` and `pd_hessian` speak to Hessian-based
standard errors and Wald-style inference. A weak-axis row says a
latent variable explains only a small share of the fitted common covariance
and may therefore be hard to interpret; it does not by itself prove
that the fitted response distribution is wrong.

`boundary_sigma_eps` checks the continuous-family residual scale when
that parameter is active. A warning means the estimate is below the
reported threshold, so the residual scale is close to its boundary; it
does not apply when the model or response family maps that parameter
off.

The same table records intentionally skipped standard errors. A no-SE
fit can still be useful for point estimates or refit-based uncertainty, such
as a bootstrap that repeatedly fits the model to new samples, but it should not
be reported as if Wald standard errors were computed.

```{r no-se-health}
fit_no_se <- gllvmTMB(
  count ~ 0 + trait +
    latent(0 + trait | individual, d = 1),
  data = df_long,
  trait = "trait",
  unit = "individual",
  family = poisson(),
  control = gllvmTMBcontrol(se = FALSE)
)

no_se_health <- check_gllvmTMB(fit_no_se)
no_se_health[
  no_se_health$component %in% c("sdreport", "pd_hessian", "max_fixed_se"),
  c("component", "status", "message")
]
```

For hard fits, move from this table to the
[Convergence and start values](convergence-start-values.html) article
before making scientific claims.

## Check Response Diagnostics

Numerical convergence is not enough. For Gaussian, Poisson, and NB2 rows,
`residuals(type = "randomized_quantile")` evaluates each fitted family's
cumulative distribution function at the observed response. These are exact
CDF calculations at the fitted parameter values and fitted random-effect modes,
but they do not propagate parameter, refit, or random-effect uncertainty and
are not a formal calibration test. NB1 uses the same implemented exact-CDF
route and has a display smoke test, but it has not yet reached the same depth of
evidence as Gaussian, Poisson, and NB2. `diagnostic_table()` turns the attached
metadata into ordinary data frames for reports and articles.

```{r residual-table}
rq <- residuals(
  fit,
  type = "randomized_quantile",
  seed = 20260609
)

diagnostic_table(rq, table = "row_status")
diagnostic_table(rq, table = "fit_health_status")

head(
  diagnostic_table(rq, table = "data")[
    , c("trait", "family", "observed", "residual", "status")
  ],
  8
)
```

The `row_status` table tells you whether every model row produced a
usable residual. If a row is marked `unsupported_family`,
`missing_response`, or `nonfinite_observed`, keep that status in the
report rather than silently dropping the row.

The Q-Q points below stay broadly near the reference line through the centre,
but the alarm and forage panels depart in the upper tail. The display is useful
because it does not turn a mostly straight pattern into a blanket pass: those
tails deserve inspection, and the plot is not evidence that the latent rank or
uncertainty estimates are calibrated.

```{r residual-qq, fig.cap = "Randomized-quantile residual Q-Q check for the fitted Poisson GLLVM. This is a fitted-model diagnostic display, not interval calibration, latent-rank proof, or a Bayesian posterior predictive check."}
predictive_check(
  fit,
  type = "rq_qq",
  seed = 20260609
)
```

For Poisson and NB2 count models, a rootogram is often easier to read than a
Q-Q plot. Bars near zero mean the observed and fitted-model simulated count
frequencies are similar on the square-root scale. The simulations default to
`condition_on_RE = TRUE`, so they condition on the fitted random-effect modes;
they check the fitted response pattern, not the marginal random-effect
distribution. The plot below shows noticeable alternating discrepancies,
especially for forage counts 3 and 5. They do not all point in one direction,
but they are large enough to investigate rather than dismiss. Simulation noise
remains, even with the larger draw count used here.

```{r rootogram, fig.cap = "Rootogram comparing observed behavioural counts with simulations conditional on the fitted random-effect modes. Persistent departures can flag a response-family, link, covariate, offset, or fitted-linear-predictor mismatch; this display does not test the marginal random-effect distribution."}
predictive_check(
  fit,
  type = "rootogram",
  ndraws = 200,
  seed = 20260609,
  max_count = 7L
)
```

## What To Do Next

Use the diagnostics as routing information, not as a single pass/fail
stamp.

| What you see | What it means | Next step |
|---|---|---|
| The four numerical rows shown above all pass | The numerical fit is a reasonable starting point, provided the remaining rows contain no unexplained warnings | Inspect every non-PASS row, then inspect `Sigma`, correlations, ordination, or fitted effects. |
| Optimizer or gradient rows warn/fail | The point estimate may be unstable | Try more starts, rescaling, simpler rank, or an alternative optimizer. |
| `sdreport` or `pd_hessian` warns | Wald uncertainty is unavailable or local curvature may be unreliable | Inspect the scientific target and use a supported profile or refit-based bootstrap route where available. |
| Residual rows have non-`ok` status | Some rows were not diagnosed | Inspect missing values, non-finite responses, and family support. |
| Q-Q or rootogram shows systematic structure | The fitted conditional response pattern may be wrong | Revisit family, link, covariates, offsets, or the fitted linear predictor; investigate random effects separately. |

If the fit-health and response checks are acceptable but the latent rank
is still a modelling choice, continue to
[How many latent dimensions should I fit?](model-selection-latent-rank.html).
That article compares one ML candidate set with AIC/BIC and uses REML
only after a Gaussian model has been chosen.

This page covers fit health and no-SE status, fitted-model predictive
diagnostics and residuals for Gaussian, Poisson, and NB2 rows, and report-ready
metadata tables. NB1 uses an implemented exact-CDF route, but its evidence is
currently shallower than for those three families.

The page does not calibrate confidence intervals, prove the latent rank, run
formal residual tests, test the marginal random-effect distribution, or draw
from a Bayesian posterior. Exact residual support for delta, hurdle, truncated,
ordinal, and mixture-family rows remains outside its current claims.
