---
title: "Response-specific predictor effects"
description: "Specify which predictors affect which responses, including intentional zero coefficients, without confusing fixed-effect structure with covariance constraints."
tier: 2
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Response-specific predictor effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

Use `Xcoef_fixed` when a direct predictor effect is fixed at zero for some
responses but estimated for others. This is an exact equality constraint on the
mean model—not shrinkage, variable selection, or evidence that an estimated
effect is null.

You can impose named zero constraints in maximum-likelihood fits. Non-zero
fixed values and `REML = TRUE` are not supported.

## Decide whether zero is structural

For unit (i) and response (t), consider

\[
\eta_{it} = \mathbf x_{it}^{\mathsf T}\boldsymbol\beta_t
             + \boldsymbol\lambda_t^{\mathsf T}\mathbf u_i.
\]

`Xcoef_fixed` pins one named entry of the direct mean coefficient
$\boldsymbol\beta_t$. It does not directly constrain the loadings, latent
scores, Psi, or other covariance parameters—although their fitted estimates can
change after the mean restriction changes.

Use an exact zero only when it is justified before seeing the fitted estimate,
for example by experimental design, temporal ordering, measurement definition,
preregistration, or a defensible structural model.

- **Known by design that the direct coefficient is zero:** use
  `Xcoef_fixed`.
- **Want to learn whether the coefficient is near zero:** leave it free and
  report its uncertainty or compare prespecified restricted and unrestricted
  models.
- **Want shrinkage or data-driven selection:** use a method designed for that
  purpose.
- **Want to remove a response:** change the response set.
- **Want to constrain which traits load on an LV:** use the separate loading-
  constraint interface; that changes $\boldsymbol\Lambda$, not
  $\boldsymbol\beta$.

Pinning a false zero can move the omitted mean pattern into other fixed effects,
residual variation, or covariance components.

## Worked example

Suppose treatment is assigned after a baseline measurement. By design, the
treatment cannot cause the baseline response, while its effects on later growth
and colour responses remain free. We encode that prespecified temporal zero.

```{r setup}
library(gllvmTMB)

set.seed(20260623)
n_unit <- 36
site <- factor(seq_len(n_unit))
treatment <- rnorm(n_unit)

dat <- expand.grid(
  site = site,
  trait = factor(c("baseline", "growth", "colour")),
  KEEP.OUT.ATTRS = FALSE
)
dat$treatment <- treatment[as.integer(dat$site)]
dat$eta <- with(dat, ifelse(
  trait == "baseline", 0.3,
  ifelse(trait == "growth", 0.4 + 0.8 * treatment,
         -0.2 - 0.5 * treatment)
))
dat$y <- dat$eta + rnorm(nrow(dat), sd = 0.2)
```

### Inspect the expanded coefficient names

Constraint names must match the expanded fixed-effect design, not a guess based
on the raw formula. Factor contrasts, interactions, transformations, and factor
level order can all change these names.

```{r names}
fixed_formula <- ~ 0 + trait + (0 + trait):treatment
colnames(model.matrix(fixed_formula, dat))
```

The baseline treatment coefficient is named `traitbaseline:treatment`, so the
constraint is:

```{r fit-long}
fit_long <- gllvmTMB(
  y ~ 0 + trait + (0 + trait):treatment,
  data = dat,
  trait = "trait",
  unit = "site",
  family = gaussian(),
  Xcoef_fixed = c("traitbaseline:treatment" = 0),
  silent = TRUE
)

long_fixed <- tidy(fit_long, "fixed")
long_fixed[grepl(":treatment$", long_fixed$term),
           c("term", "estimate", "std.error", "status")]
```

The zero is imposed, not estimated. Therefore `estimate = 0`,
`std.error = NA`, and `status = "fixed"` are expected. The missing standard
error does not indicate a failed Hessian calculation. Inference for all other
parameters is conditional on the zero restriction being correct, and the model
degrees of freedom decrease by one for each pinned coefficient.

### The same constraint from wide data

The equivalent `traits(...)` call reaches the same stacked design and therefore
uses the same expanded coefficient name.

```{r fit-wide}
dat_wide <- stats::reshape(
  dat[c("site", "trait", "treatment", "y")],
  idvar = c("site", "treatment"),
  timevar = "trait",
  direction = "wide"
)
names(dat_wide) <- sub("^y\\.", "", names(dat_wide))

fit_wide <- gllvmTMB(
  traits(baseline, growth, colour) ~ 1 + treatment,
  data = dat_wide,
  unit = "site",
  family = gaussian(),
  Xcoef_fixed = c("traitbaseline:treatment" = 0),
  silent = TRUE
)

wide_fixed <- tidy(fit_wide, "fixed")
comparison <- merge(
  long_fixed[grepl(":treatment$", long_fixed$term),
             c("term", "estimate", "status")],
  wide_fixed[grepl(":treatment$", wide_fixed$term),
             c("term", "estimate", "status")],
  by = "term",
  suffixes = c("_long", "_wide")
)
comparison
```

Equivalent designs and contrast settings produce the same names and estimates;
arbitrarily different codings need not.

To pin several prespecified direct effects, supply several uniquely named
entries, for example:

```{r multiple-constraints, eval = FALSE}
Xcoef_fixed = c(
  "traitbaseline:treatment" = 0,
  "traitcolour:temperature" = 0
)
```

## Errors and next actions

- **A name is not recognized:** inspect `colnames(model.matrix(...))` for the
  exact expanded design and check factor levels and contrasts.
- **A name is duplicated:** provide each expanded coefficient once.
- **A non-zero fixed value is requested:** only structural zero is implemented;
  leave the coefficient free if zero is not the intended value.
- **`REML = TRUE` errors:** fixed-coefficient pinning is currently ML-only.
- **A fixed row has an unavailable SE or CI:** this is expected because the
  value was imposed rather than estimated.
- **The scientific question is whether the effect equals zero:** do not impose
  the answer with `Xcoef_fixed`; fit an appropriate unrestricted comparison.

## Related tools

`Xcoef_fixed` acts on direct response-specific mean coefficients. It is
different from:

- [Formula keyword grid](api-keyword-grid.html), which chooses covariance
  structure;
- [Pre-fit response screening](pre-fit-response-screening.html), which checks
  data support and formula estimability rather than selecting responses or
  predictors;
- [`gllvmTMB()` reference](../reference/gllvmTMB.html), which documents the
  complete argument contract.
