---
title: "Fix predictor effects at zero"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Fix predictor effects at zero}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

`Xcoef_fixed` controls the fixed-effect part of the model. Use it when the same
stacked-trait formula is convenient, but a predictor should be estimated for
some responses and fixed at zero for others.

For one unit `i` and response `t`, a simple stacked-trait model has a mean
structure like

$$
\eta_{it} = x_{it}^{T}\beta_t + \lambda_t^{T}u_i .
$$

Here `beta` is the fixed-effect coefficient vector for the mean, `lambda` is a
loading row, and `u_i` is the unit score. `Xcoef_fixed` changes the first part,
`x beta`. It does not change `Lambda`, `Psi`, the latent rank, or the response
set.

## Why fix a coefficient?

Sometimes a predictor is meaningful for one response but not another. In a
systematic map, an urban-gradient predictor might be scientifically meaningful
for biodiversity outcomes but not for a metadata indicator. In a joint species
distribution model, a covariate might be pre-specified for some species or
response groups but not others. In a morphometrics example, a treatment might
affect growth traits but not a recording-quality indicator.

You could split the model into several formulas. `Xcoef_fixed` gives a narrower
route: keep one stacked model and pin only the irrelevant fixed-effect
coefficients at structural zero.

## Example

```{r setup}
library(gllvmTMB)
```

Create three responses measured on the same units. The predictor `x` should
affect `growth` and `colour`, but the `metadata` response is included only as a
control indicator, so we do not want `x` to define that response's mean.

```{r data}
set.seed(20260623)
n_unit <- 36
site <- factor(seq_len(n_unit))
x_site <- rnorm(n_unit)

dat <- expand.grid(
  site = site,
  trait = factor(c("growth", "colour", "metadata")),
  KEEP.OUT.ATTRS = FALSE
)
dat$x <- x_site[as.integer(dat$site)]
dat$eta <- with(dat, ifelse(
  trait == "growth", 0.4 + 0.8 * x,
  ifelse(trait == "colour", -0.2 - 0.5 * x, 0.1)
))
dat$y <- dat$eta + rnorm(nrow(dat), sd = 0.2)
```

The coefficient names come from the expanded fixed-effect design, not from the
raw formula text. Check those names before supplying `Xcoef_fixed`.

```{r names}
colnames(model.matrix(~ 0 + trait + (0 + trait):x, dat))
```

Now fix only the `metadata` slope for `x` at zero.

```{r fit}
fit <- gllvmTMB(
  y ~ 0 + trait + (0 + trait):x + latent(0 + trait | site, d = 1, unique = FALSE),
  data = dat,
  unit = "site",
  family = gaussian(),
  Xcoef_fixed = c("traitmetadata:x" = 0),
  silent = TRUE
)

tidy(fit, "fixed")
```

The fixed row remains in the fitted object. Its estimate is `0`, its standard
error is `NA`, and its `status` is `"fixed"`. The other response-specific slopes
are still estimated.

If your data start in wide form, use the same `gllvmTMB()` entry point with the
`traits(...)` left-hand side. The expanded coefficient name is the same because
both forms reach the same stacked fixed-effect design.

```{r wide-fit}
dat_wide <- stats::reshape(
  dat[c("site", "trait", "x", "y")],
  idvar = c("site", "x"),
  timevar = "trait",
  direction = "wide"
)
names(dat_wide) <- sub("^y\\.", "", names(dat_wide))

fit_wide <- gllvmTMB(
  traits(growth, colour, metadata) ~ 1 + x +
    latent(1 | site, d = 1, unique = FALSE),
  data = dat_wide,
  unit = "site",
  family = gaussian(),
  Xcoef_fixed = c("traitmetadata:x" = 0),
  silent = TRUE
)

tidy(fit_wide, "fixed")
```

## Relation to other tools

`Xcoef_fixed` is not variable selection. It should express a pre-specified
model structure: this response does not get this predictor effect. If you are
unsure whether a binary response should enter the model at all, start with
`screen_gllvmTMB()` instead.

`Xcoef_fixed` is also not a loading constraint. Use loading constraints when
you want to control which responses define a latent axis. Use `Xcoef_fixed`
when you want to control which predictors enter each response's mean.

The current implementation supports exact zero constraints in ML fits. Native
TMB fits support the full named `Xcoef_fixed` contract. Admitted
`engine = "julia"` fixed-effect-X rows use the same R argument and pass a
zero-mask to GLLVM.jl. `REML = TRUE`, non-zero fixed values, and unsupported
Julia fixed-effect-X rows remain gated.
