---
title: "Functional biogeography GLLVM: M1 to M2 to extensions"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Functional biogeography GLLVM: M1 to M2 to extensions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

The replicated site--species--trait GLLVM is the canonical
functional-biogeography model of Nakagawa et al. (in prep). This
article fits it as a **two-step model ladder**:

* **§1 -- M1 (functional integration core).** Trait covariation
  decomposed into a between-site component
  $\boldsymbol{\Sigma}_B$ (global functional integration) and a
  within-site component $\boldsymbol{\Sigma}_W$ (local functional
  integration). No space, no phylogeny -- just the syndrome story.
* **§2 -- M2 (controls for space and phylogeny).** Re-fit M1 with
  spatial and phylogenetic adjustment terms. Compare M2's
  $\boldsymbol{\Sigma}_B$ / $\boldsymbol{\Sigma}_W$ to M1's. If the
  syndrome story holds, the controls have done their job.
* **§3 -- Extensions.** Promote space or phylogeny to its own
  covariance decomposition (M3, M4) when the inferential target is
  *spatial* or *phylogenetically conserved* trait syndromes. We
  preview the syntax and point at companion articles; the fits
  themselves live elsewhere.

## A family of models

Before §1, the conceptual scaffolding. The replicated
site--species--trait GLLVM is **one of a family** of trait-covariance
models. The right model depends on the response shape:

| Data / question | Recommended model |
|---|---|
| Species occurrences/abundances; traits explain species responses | Site $\times$ species JSDM / fourth-corner GLLVM ([`joint-sdm`](joint-sdm.html)) |
| One CWM/functional summary per site and trait | Site-summary / CWM site $\times$ trait GLLVM |
| **Species-in-site trait observations**, enough within-site richness | **Replicated site--species--trait GLLVM** (this article) |
| Species have one trait value each (AVONET-style) | Two-stage site-summary model ([`corvidae-two-stage`](corvidae-two-stage.html)) |
| Trait-state counts or functional-gene frequencies | Trait-state abundance/count GLLVM |
| Trait integration across species/evolution | Species $\times$ trait phylogenetic GLLVM ([`phylogenetic-gllvm`](phylogenetic-gllvm.html)) |

This article models direct trait observations $y_{sit}$ on multiple
species at each site. The other rows of the table point at the right
article for the other response shapes; do not force this model onto
the wrong one.

A second conceptual move runs through §1 and §2: the **adjustment vs
partitioning** distinction for spatial and phylogenetic terms.

* **Adjustment models** (M2, this article's main focus). Spatial
  and phylogenetic terms are *controls*, so the primary biological
  targets remain $\boldsymbol{\Sigma}_B$ and $\boldsymbol{\Sigma}_W$.
  Use `spatial_unique()` and `phylo_unique()` -- per-trait variances
  on a shared correlation structure ($\mathbf{K}_{\text{space}}$ and
  $\mathbf{C}_{\text{phy}}$).
* **Partitioning models** (M3, M4 in §3). Spatial and phylogenetic
  signals get their own reduced-rank covariance decomposition
  ($\boldsymbol{\Sigma}_R$, $\boldsymbol{\Sigma}_P$) via
  `spatial_latent + spatial_unique` and
  `phylo_latent + phylo_unique`. The question shifts to: are
  syndromes *spatially structured* or *phylogenetically conserved*?
  These models are more data-demanding; treat them as sensitivity
  analyses.

## Setup and DGP

```{r setup}
library(gllvmTMB)
library(ape)
library(ggplot2)
```

> **Wide-format input.** The data here are naturally long (one row per
> `(site, species, trait)` cell), but field datasets sometimes arrive
> wide (one row per `(site, species)`, one column per trait). The
> formula-wide path keeps that shape readable:
> `traits(trait_1, ..., trait_T) ~ 1 + latent(1 | site_species, d = 2)
> + unique(1 | site_species)`. The long form below stays canonical
> because it keeps `site`, `species`, and `site_species` visible for the
> crossed ecological interpretation.

A 60-site, 30-species, 6-trait fixture with mean 5 species per site
(`r as.character(round(60 * 5))` rows-ish) gives both M1 and M2 enough
information to recover the simulation truth on the largest covariance
entries. The DGP includes:

* a rank-2 between-site truth ($d_B = 2$): two trait syndromes with
  opposite signs across the two trait blocks (1-3 vs 4-6);
* a rank-1 within-site truth ($d_W = 1$): a single shared local axis;
* small phylogenetic and spatial signals so M2 has something to
  control for, but small enough that the M1-vs-M2 comparison is
  honest -- the syndrome story should *not* be an artefact of
  geography or shared ancestry.

```{r sim}
set.seed(42)
n_sites   <- 60
n_species <- 30
n_traits  <- 6

tree <- ape::rcoal(n_species)
tree$tip.label <- paste0("sp", seq_len(n_species))

## Two clear syndromes (rank d_B = 2 truth)
Lambda_B_true <- matrix(c(
   0.9,  0.7,  0.6, -0.2, -0.5, -0.3,    # axis 1 (traits 1-3 high; 4-6 low)
   0.2, -0.1,  0.0,  0.7,  0.7,  0.5     # axis 2 (traits 4-6 high)
), nrow = n_traits, ncol = 2)
Lambda_W_true <- matrix(c(0.5, 0.4, 0.3, 0.4, 0.3, 0.2),
                        nrow = n_traits, ncol = 1)
S_B_true <- rep(0.15, n_traits)
S_W_true <- rep(0.20, n_traits)

sim <- simulate_site_trait(
  n_sites = n_sites, n_species = n_species, n_traits = n_traits,
  mean_species_per_site = 5,
  Lambda_B = Lambda_B_true, Lambda_W = Lambda_W_true,
  S_B = S_B_true, S_W = S_W_true,
  Cphy = ape::vcv(tree, corr = TRUE),
  sigma2_phy    = rep(0.10, n_traits),
  sigma2_sp     = rep(0.05, n_traits),
  spatial_range = 0.4,
  sigma2_spa    = rep(0.10, n_traits),
  sigma2_eps    = 0.10,
  seed          = 42
)
df <- sim$data
levels(df$species) <- tree$tip.label

Sigma_B_true <- Lambda_B_true %*% t(Lambda_B_true) + diag(S_B_true)
Sigma_W_true <- Lambda_W_true %*% t(Lambda_W_true) + diag(S_W_true)
```

# §1 -- M1: functional integration core

The lead-in model. No space, no phylogeny. The mean structure is

$$
\mu_{st} = \alpha_t + \mathbf{x}_s^{\!\top} \boldsymbol{\beta}_t,
$$

a per-trait intercept $\alpha_t$ plus per-trait slopes
$\boldsymbol{\beta}_t$ on two site-level predictors `env_1, env_2`.
The random-effect structure is

$$
y_{sit} = \mu_{st} + u_{st} + e_{sit} + \delta_{sit},
$$

with

* $\mathbf{u}_s \sim \mathcal{N}(\mathbf{0}, \boldsymbol{\Sigma}_B)$,
  $\boldsymbol{\Sigma}_B = \boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^{\!\top} + \boldsymbol{\Psi}_B$
  -- *global functional integration*, between-site trait covariance.
* $\mathbf{e}_{si} \sim \mathcal{N}(\mathbf{0}, \boldsymbol{\Sigma}_W)$,
  $\boldsymbol{\Sigma}_W = \boldsymbol{\Lambda}_W \boldsymbol{\Lambda}_W^{\!\top} + \boldsymbol{\Psi}_W$
  -- *local functional integration*, within-site species deviations.
* $\delta_{sit} \sim \mathcal{N}(0, \sigma^2_\delta)$ -- residual.

`gllvmTMB` syntax:

```{r fit-m1}
t0 <- Sys.time()
fit_M1 <- gllvmTMB(
  value ~ 0 + trait +
          (0 + trait):env_1 + (0 + trait):env_2 +
          latent(0 + trait | site, d = 2) +
          unique(0 + trait | site) +
          latent(0 + trait | site_species, d = 1) +
          unique(0 + trait | site_species),
  data    = df,
  cluster = "species"
)
t_M1 <- round(as.numeric(Sys.time() - t0, units = "secs"), 1)
fit_M1$opt$convergence
```

Wall time: `r t_M1` s. Extract $\boldsymbol{\Sigma}_B$ and
$\boldsymbol{\Sigma}_W$:

```{r m1-sigma}
Sigma_B_M1 <- extract_Sigma(fit_M1, level = "unit",     part = "total")$Sigma
Sigma_W_M1 <- extract_Sigma(fit_M1, level = "unit_obs", part = "total")$Sigma
round(Sigma_B_M1, 2)
round(Sigma_W_M1, 2)
```

The block structure is recovered: traits 1-3 covary positively,
traits 4-6 covary positively, and the two blocks covary negatively --
matching the rank-2 simulation truth. Per-trait communality
$c^2_{B,t} = (\boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^{\!\top})_{tt} / (\boldsymbol{\Sigma}_B)_{tt}$
and the site-level ICC
$R^{(\text{site})}_t = (\boldsymbol{\Sigma}_B)_{tt} / [(\boldsymbol{\Sigma}_B)_{tt} + (\boldsymbol{\Sigma}_W)_{tt}]$:

```{r m1-comm-icc}
round(extract_communality(fit_M1, level = "unit"), 3)
round(extract_ICC_site(fit_M1), 3)
```

Three things to watch for in any M1 fit, each a separate identifiability concern:

> **Note (1) -- $\boldsymbol{\Sigma}_W$ vs $\sigma_\delta$
> identifiability.** With no within-cell replication (one observation
> per (`site`, `species`, `trait`) triple), only the *sum*
> $\boldsymbol{\Sigma}_W^{*} = \boldsymbol{\Sigma}_W + \mathrm{diag}(\sigma^2_\delta)$
> is identified. The engine handles this by auto-suppressing
> `sigma_eps` to a tiny floor whenever a per-row `unique()` is
> present (you'll see a one-shot inform when M1 fits). Replication
> within cell -- multiple individuals of the same species at the
> same site, or repeated visits -- is the design move that splits
> them. See [`pitfalls`](pitfalls.html) for a worked demonstration.

> **Note (2) -- communality is rank-conditional.** $c^2_{B,t}$
> depends on the choice of $d_B$. Refitting M1 at $d_B = 1$ on the
> same data gives different communalities than at $d_B = 2$:

```{r m1-rank-conditional}
fit_M1_d1 <- gllvmTMB(
  value ~ 0 + trait +
          (0 + trait):env_1 + (0 + trait):env_2 +
          latent(0 + trait | site, d = 1) +
          unique(0 + trait | site) +
          latent(0 + trait | site_species, d = 1) +
          unique(0 + trait | site_species),
  data    = df,
  cluster = "species"
)
rbind(
  d_B_1 = round(extract_communality(fit_M1_d1, level = "unit"), 3),
  d_B_2 = round(extract_communality(fit_M1,    level = "unit"), 3)
)
```

> The two communality vectors disagree by 0.05--0.4 across traits.
> The reason is structural: $c^2_{B,t}$ is the diagonal of
> $\boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^{\!\top}$ over the
> diagonal of $\boldsymbol{\Sigma}_B$, and shrinking $d_B$ by one
> takes a column out of $\boldsymbol{\Lambda}_B$. **Always report the
> rank you used.** Pick $d_B$ on a model-comparison criterion (AIC,
> cross-validated likelihood) before reporting communality, not
> after.

> **Note (3) -- the response shape determines the question.** This
> article uses direct trait observations $y_{sit}$. For
> occurrence/abundance data with traits as covariates, fit a JSDM /
> fourth-corner GLLVM ([`joint-sdm`](joint-sdm.html)). For one
> CWM-style summary per site, use the site-summary CWM model. For
> AVONET-style data with one trait value per species, use the
> [two-stage workflow](corvidae-two-stage.html). The replicated
> site--species--trait model below is right for our DGP; do not
> force it onto a different one.

# §2 -- M2: spatial and phylogenetic adjustment

Same equation as §1 but with four extra random-effect terms:

$$
y_{sit} = \mu_{st} + u_{st} + e_{sit} + r_{st} + p_{it} + q_{it} + \delta_{sit},
$$

with

* $r_{\cdot t} \sim \mathcal{N}(\mathbf{0}, \sigma^2_{R,t}\,\mathbf{K}_{\text{space}})$
  -- per-trait Matérn spatial field on the SPDE mesh; *spatial control*.
* $p_{\cdot t} \sim \mathcal{N}(\mathbf{0}, \sigma^2_{P,t}\,\mathbf{C}_{\text{phy}})$
  -- per-trait *phylogenetic* species random effect on the shared
  $\mathbf{C}_{\text{phy}}$; *phylogenetic control*.
* $q_{\cdot t} \sim \mathcal{N}(\mathbf{0}, \sigma^2_{Q,t}\,\mathbf{I})$
  -- per-trait *non-phylogenetic* species random effect (independent
  across species); the residual species effect after phylogeny.

These are the **adjustment-mode** terms: each trait carries its own
variance on a shared correlation structure
($\mathbf{K}_{\text{space}}$, $\mathbf{C}_{\text{phy}}$, $\mathbf{I}$),
without introducing a cross-trait spatial or phylogenetic loading
matrix. This keeps the focal interpretation on $\boldsymbol{\Sigma}_B$
and $\boldsymbol{\Sigma}_W$.

The $p_{it} + q_{it}$ pair is the same Hadfield--Nakagawa (2010)
decomposition used in the
[`phylogenetic-gllvm`](phylogenetic-gllvm.html) article, just placed
at the *cluster* (species) slot of a crossed (site $\times$ species)
fit rather than at the unit slot of a per-species fit. The two
correlation matrices $\mathbf{C}_{\text{phy}}$ and $\mathbf{I}$ are
orthogonal, so $\sigma^2_{P,t}$ and $\sigma^2_{Q,t}$ are jointly
identifiable -- see the note after the fit for the empirical regime.

```{r fit-m2}
mesh <- make_mesh(df, c("lon", "lat"), cutoff = 0.1)

t0 <- Sys.time()
fit_M2 <- gllvmTMB(
  value ~ 0 + trait +
          (0 + trait):env_1 + (0 + trait):env_2 +
          spatial_unique(0 + trait | coords, mesh = mesh) +
          latent(0 + trait | site, d = 2) +
          unique(0 + trait | site) +
          latent(0 + trait | site_species, d = 1) +
          unique(0 + trait | site_species) +
          unique(0 + trait | species) +              # q_it
          phylo_unique(species, tree = tree),         # p_it
  data    = df,
  cluster = "species"
)
t_M2 <- round(as.numeric(Sys.time() - t0, units = "secs"), 1)
fit_M2$opt$convergence
```

Wall time: `r t_M2` s.

A note on identifiability. With $n_{\text{species}} \geq 100$ and a
crossed (site $\times$ species) design, $\sigma^2_{Q,t}$ recovers
within $\sim$10% relative error and $\sigma^2_{P,t}$ within $\sim$50%
(per-trait estimates can be noisy when one trait's phylogenetic
signal is weak; the trait-averaged $\bar{\sigma}^2_P$ is more
stable). At small $n_{\text{species}}$ or weak phylogenetic signal,
the $p_{it}$ and $q_{it}$ pieces cannot be cleanly separated and one
should compare M1 (without `unique(0 + trait | species)`) to M2 (with
it) to confirm both terms contribute on the focal data. The engine
prints a one-shot informational note pointing at this regime when it
detects the pairing.

## The pedagogical moment: do the controls move the syndrome?

Re-extract $\boldsymbol{\Sigma}_B$ and $\boldsymbol{\Sigma}_W$ from
M2 and compare to M1's:

```{r m2-sigma}
Sigma_B_M2 <- extract_Sigma(fit_M2, level = "unit",     part = "total")$Sigma
Sigma_W_M2 <- extract_Sigma(fit_M2, level = "unit_obs", part = "total")$Sigma

R_B_M1 <- extract_Sigma(fit_M1, level = "unit")$R
R_B_M2 <- extract_Sigma(fit_M2, level = "unit")$R
R_W_M1 <- extract_Sigma(fit_M1, level = "unit_obs")$R
R_W_M2 <- extract_Sigma(fit_M2, level = "unit_obs")$R

c(max_abs_diff_R_B = round(max(abs(R_B_M2 - R_B_M1)), 3),
  max_abs_diff_R_W = round(max(abs(R_W_M2 - R_W_M1)), 3))
```

Side-by-side correlation-matrix heatmaps:

```{r heatmap-rb, fig.cap = "Cross-trait correlation matrix at the between-site (Sigma_B) tier under M1 (no controls) and M2 (spatial + phylogenetic adjustment). The block structure (traits 1-3 vs 4-6) is preserved across models.", fig.width = 8, fig.height = 4}
heatmap_df <- function(R, label) {
  rn <- rownames(R) %||% paste0("trait_", seq_len(nrow(R)))
  data.frame(
    model  = label,
    trait_i = factor(rep(rn, ncol(R)), levels = rn),
    trait_j = factor(rep(rn, each = nrow(R)), levels = rn),
    rho     = as.vector(R)
  )
}

R_B_long <- rbind(
  heatmap_df(R_B_M1, "M1 (no controls)"),
  heatmap_df(R_B_M2, "M2 (+ space + phylo)")
)

ggplot(R_B_long, aes(trait_i, trait_j, fill = rho)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 2.8) +
  scale_fill_gradient2(low = "#1f78b4", mid = "white", high = "#b2182b",
                       midpoint = 0, limits = c(-1, 1)) +
  facet_wrap(~ model) +
  coord_equal() +
  labs(x = NULL, y = NULL,
       title = expression("Between-site trait correlations " *
                          R[B] * " = cor(" * Sigma[B] * ")")) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r heatmap-rw, fig.cap = "Cross-trait correlation matrix at the within-site (Sigma_W) tier under M1 and M2. The local-integration story is similarly stable to the addition of controls.", fig.width = 8, fig.height = 4}
R_W_long <- rbind(
  heatmap_df(R_W_M1, "M1 (no controls)"),
  heatmap_df(R_W_M2, "M2 (+ space + phylo)")
)

ggplot(R_W_long, aes(trait_i, trait_j, fill = rho)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 2.8) +
  scale_fill_gradient2(low = "#1f78b4", mid = "white", high = "#b2182b",
                       midpoint = 0, limits = c(-1, 1)) +
  facet_wrap(~ model) +
  coord_equal() +
  labs(x = NULL, y = NULL,
       title = expression("Within-site trait correlations " *
                          R[W] * " = cor(" * Sigma[W] * ")")) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

The block structure of $\mathbf{R}_B$ -- traits 1-3 positively
correlated, traits 4-6 positively correlated, the two blocks
negatively correlated -- is *the same* under M1 and M2. That is the
desired finding. It tells the reader the syndrome was **not** a
geography or ancestry artefact: even after controlling for the
spatial and phylogenetic structure that the data carry, the same two
syndromes survive.

If the comparison instead showed M2's $\mathbf{R}_B$ flattening
(syndrome erased) or shifting sign (syndrome reversed), you would
have a different and equally interesting story: the apparent global
functional integration was largely a function of spatial covariation
in the environmental drivers, or of shared ancestry. Either finding
should be reported -- the M1 vs M2 comparison is *the* diagnostic
for which.

> **Sample size guidance**. M2 is a multi-tier model with five
> random-effect tiers and ~70 free parameters at $T = 6$.
> Identifiability requires enough **groups** at each tier, not just
> enough total observations:
>
> - $n_{\text{sites}} \geq 50$ for stable $\boldsymbol{\Sigma}_B$
> - mean species per site $\geq 5$ for identifiable $\boldsymbol{\Sigma}_W$
>   (with single observation per (site, species, trait) cell, only
>   $\boldsymbol{\Sigma}_W^{\star} = \boldsymbol{\Sigma}_W +
>   \mathrm{diag}(\sigma^2_\delta)$ is identifiable -- see Note 1
>   in §1)
> - $n_{\text{species}} \geq 50$ for the per-trait phylogenetic
>   ($\sigma^2_{P,t}$) and non-phylogenetic species ($\sigma^2_{Q,t}$)
>   variances
> - phylogenetic signal at moderate-or-stronger
>   ($\lambda_{\text{Pagel}} \gtrsim 0.4$) for the phylogenetic
>   decomposition to bite
>
> Below these thresholds, fall back to **M1** -- drop the controls,
> report $\boldsymbol{\Sigma}_B$ and $\boldsymbol{\Sigma}_W$ as
> *partial* (not "fully adjusted") global and local integration.
> Don't add tiers your data can't support: gllvmTMB makes complex
> models *expressible*, but **expressibility is not
> identifiability**. A formal regime map across M1–M4 is in
> preparation (`dev/design/11-identifiability-regime-map.md`); a
> dedicated article will follow once the simulation grid completes.

# §3 -- Extensions: when space and phylogeny become the target

When the question is about space or phylogeny *itself* -- "are
syndromes spatially structured?", "is one of the syndromes
phylogenetically conserved?" -- promote the corresponding signal
from a per-trait control to a full reduced-rank covariance
decomposition. We sketch the syntax for two extension models below
and point at the dedicated articles for the worked fits.

## M3 -- spatial trait syndromes

Add a `spatial_latent` term in addition to `spatial_unique` to
estimate $\boldsymbol{\Sigma}_R = \boldsymbol{\Lambda}_R\boldsymbol{\Lambda}_R^{\!\top} + \boldsymbol{\Psi}_R$,
the spatial trait-covariance decomposition. Loadings on
$\boldsymbol{\Lambda}_R$ tell you which traits hang together
*spatially*.

```r
fit_M3 <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):env_1 + (0 + trait):env_2 +
          spatial_latent(0 + trait | coords, d = 2, mesh = mesh) +
          spatial_unique(0 + trait | coords, mesh = mesh) +
          latent(0 + trait | site, d = 2) +
          unique(0 + trait | site) +
          latent(0 + trait | site_species, d = 1) +
          unique(0 + trait | site_species) +
          phylo_unique(species, tree = tree),
  data = df, cluster = "species"
)
```

The spatial term is the same engine as the phylogenetic
two-piece decomposition documented in
[`phylogenetic-gllvm`](phylogenetic-gllvm.html). M3 needs more
spatial signal than M2 to identify both
$\boldsymbol{\Lambda}_R$ and $\boldsymbol{\Psi}_R$ separately;
treat its loadings as a sensitivity analysis to M2's per-trait
spatial variances. A worked example will appear in the planned
`spatial-trait-syndromes` article.

## M4 -- phylogenetically conserved syndromes

Symmetric to M3, but on the phylogeny side: pair `phylo_latent`
with `phylo_unique` to estimate
$\boldsymbol{\Sigma}_P = \boldsymbol{\Lambda}_P\boldsymbol{\Lambda}_P^{\!\top} + \boldsymbol{\Psi}_P$,
the phylogenetic two-piece decomposition. Loadings on
$\boldsymbol{\Lambda}_P$ tell you which traits cluster together
*along the tree*.

```r
fit_M4 <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):env_1 + (0 + trait):env_2 +
          spatial_unique(0 + trait | coords, mesh = mesh) +
          latent(0 + trait | site, d = 2) +
          unique(0 + trait | site) +
          latent(0 + trait | site_species, d = 1) +
          unique(0 + trait | site_species) +
          phylo_latent(species, d = 2, tree = tree) +
          phylo_unique(species, tree = tree),
  data = df, cluster = "species", unit = "species"  # PGLLVM mode
)
```

The two-piece phylogenetic fit is the focus of the
[`phylogenetic-gllvm`](phylogenetic-gllvm.html) article and its
diagnostic counterpart [`two-U-phylogeny`](two-U-phylogeny.html).
That article also covers the identifiability conditions: $K_P$
($d$ in `phylo_latent`) is identifiable up to rotation, and the
two-piece split between $\boldsymbol{\Lambda}_P\boldsymbol{\Lambda}_P^{\!\top}$
and $\boldsymbol{\Psi}_P$ requires either $K_P < T$ or a
constraint matrix.

## What we did *not* fit

We deliberately do not promote the ladder beyond M4 in this
article. The "kitchen-sink" model that combines latent partitioning
on every tier ($\boldsymbol{\Lambda}_B$, $\boldsymbol{\Lambda}_W$,
$\boldsymbol{\Lambda}_R$, $\boldsymbol{\Lambda}_P$) is conceptually
beautiful but statistically fragile -- the partitioning at one tier
trades off against the others, and the loadings stop being separately
identifiable without strong constraints. M2 is the sweet spot for
publication-grade fits; M3 and M4 are sensitivity analyses;
beyond that, the data have to carry the burden of identifiability
that the model can no longer.

# Diagnostics

```{r diagnostics}
sanity_multi(fit_M2)
```

Convergence, max gradient, Hessian PD-ness, fixed-effect SE, and
near-zero loading diagonals -- the report flags any rank that was
set higher than the data support. The fixed-effect SE check is the
loosest filter: in fits with a phylogenetic random effect the SE
column also picks up some derived sd-report quantities and can flag
WARN even when the underlying fixed-effect estimates are well-
behaved (compare to the `tidy(fit_M2, "fixed")` output). The other
four checks are the ones to act on.

# See also

* [`phylogenetic-gllvm`](phylogenetic-gllvm.html) -- the two-piece
  phylogenetic decomposition (M4 above) at species level.
* [`joint-sdm`](joint-sdm.html) -- non-Gaussian JSDM /
  fourth-corner model: traits explain species responses to environment.
* [`corvidae-two-stage`](corvidae-two-stage.html) -- AVONET-style
  data where each species has one trait value.
* [`pitfalls`](pitfalls.html) -- common gotchas: factor level order,
  $\boldsymbol{\Lambda}$ rotation, the DGP-must-match-the-fit rule.
* [`profile-likelihood-ci`](profile-likelihood-ci.html) -- CIs on
  $\boldsymbol{\Sigma}_B$, $\boldsymbol{\Sigma}_W$ entries via
  profile likelihood, Wald, or parametric bootstrap.
* [`simulation-recovery`](simulation-recovery.html) -- multi-replicate
  recovery study for the full ladder.
