---
title: "Joint ecological intensity from opportunistic records"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4.5
)
set.seed(20260816)
```

```{r setup, message = FALSE}
library(gllvmTMB)
```

Opportunistic occurrence records — GBIF downloads, portal submissions,
atlas compilations — are plentiful exactly where people go and sparse
everywhere else. This article fits a **joint** model to such records
alone: many species, one shared ecological gradient, and an explicit,
named recording-bias term. It is the presence-only opener of the
species-distribution series; the integrated articles that follow show
what changes when a designed survey is added to the same model.

No integration machinery appears here. The model is an ordinary
`gllvmTMB()` call — `poisson()` counts, an effort offset, a bias
covariate, and a `latent()` term. Everything it estimates is
**relative ecological intensity**: where each species' recorded
intensity is higher or lower, after effort and the named bias are
accounted for. It does not estimate abundance, absolute occurrence
probability, or detectability, and no number below should be read as if
it did.

# Why model species jointly when the data are mostly zeros?

Consider a wetland restoration programme comparing six species across a
landscape of cells. A separate single-species model per taxon gives six
maps, but discards a useful biological premise: species can share an
unmeasured ecological gradient while keeping species-specific
departures. Jointly, the sparse species borrow strength from the common
gradient, and the model reports which species co-vary beyond what the
measured environment explains.

For cell $c$ and species $s$, the ecological linear predictor is

$$
  \eta_{cs} = \alpha_s + x_c \beta_s + \lambda_s z_c + e_{cs},
  \qquad z_c \sim N(0, 1), \qquad e_{cs} \sim N(0, \psi_s^2),
$$

so the between-cell species covariance is
$\Sigma = \Lambda \Lambda^\top + \Psi$ with
$\Lambda = (\lambda_1, \ldots, \lambda_S)^\top$ and
$\Psi = \mathrm{diag}(\psi_1^2, \ldots, \psi_S^2)$. The rank-one term is
the shared unmeasured gradient; the free diagonal keeps it from being
forced to explain every species-specific deviation.

The records themselves are counts with a known effort exposure $a_c$
and a measured bias covariate $b_c$:

$$
  N_{cs} \sim \mathrm{Poisson}(\mu_{cs}), \qquad
  \log \mu_{cs} = \log a_c + \eta_{cs} + b_c \gamma_s .
$$

The $\gamma_s$ terms belong to the recording process, not to ecology:
they describe each species' association with the bias covariate
(accessibility, urban proximity, observer density) beyond what the
effort offset already carries. Identifying them is a **design
assumption**, not a package guarantee — the bias covariate must vary
separately enough from the environmental predictors. In the simulation
below the two are independent by construction; in a real dataset you
must argue this from the study design.

# Simulate a six-species wetland community

Three hundred cells, six species, one portal-style record stream.
Effort varies tenfold across the landscape and is treated as known (in
practice: visit counts, checklist counts, or records of a reference
group). The bias covariate is simulated independently of the wetland
gradient — the assumption flagged above, made true here by design.

```{r simulate}
n_cell  <- 300
cells   <- paste0("cell_", seq_len(n_cell))
species <- c("sedge_wren", "rail", "bittern",
             "sundew", "sphagnum", "darner")

x <- as.numeric(scale(runif(n_cell)))             # measured wetland gradient
b <- as.numeric(scale(runif(n_cell)))             # accessibility (bias), independent of x
z <- rnorm(n_cell)                                # unmeasured shared gradient
effort <- exp(log(0.12) + 1.15 * (b - min(b)) / diff(range(b)) * 2)  # ~10x range, tracks access

alpha  <- c(-0.4,  0.1, -0.8,  0.3,  0.6, -0.2)
beta   <- c( 0.9,  0.5,  0.7,  1.1,  0.8, -0.3)   # wetland responses
lambda <- c( 0.8,  0.6,  0.7, -0.5, -0.6,  0.4)   # shared-gradient loadings
psi    <- c( 0.4,  0.3,  0.5,  0.3,  0.4,  0.3)   # species-specific residual sd
gamma  <- c( 0.6,  0.1,  0.0,  0.4,  0.2,  0.9)   # recording-bias associations

dat <- expand.grid(cell_id = cells, trait = species,
                   stringsAsFactors = FALSE)
ci <- match(dat$cell_id, cells)
si <- match(dat$trait, species)
eta <- alpha[si] + x[ci] * beta[si] + z[ci] * lambda[si] +
  rnorm(nrow(dat), 0, psi[si])
dat$value      <- rpois(nrow(dat), effort[ci] * exp(eta + b[ci] * gamma[si]))
dat$trait      <- factor(dat$trait, levels = species)
dat$cell_id    <- factor(dat$cell_id)
dat$env        <- x[ci]
dat$access     <- b[ci]
dat$log_effort <- log(effort[ci])
mean(dat$value == 0)
```

Most cells hold zero records for most species — the familiar shape of
opportunistic data.

# One ordinary fit

Every piece of the model above maps onto the standard long-format
grammar: per-species intercepts and slopes as `trait` interactions, the
known effort as an offset, and the shared gradient as a rank-one
`latent()` term (which carries its diagonal $\Psi$ companion by
default):

```{r fit}
fit <- gllvmTMB(
  value ~ 0 + trait + trait:env + trait:access + offset(log_effort) +
    latent(0 + trait | cell_id, d = 1),
  data   = dat,
  trait  = "trait",
  unit   = "cell_id",
  family = poisson(),
  silent = TRUE
)
fit$opt$convergence
```

`0` means the optimizer reached a point it judges stationary; run the
package's own health check rather than stopping there:

```{r health}
health <- check_gllvmTMB(fit)
health[health$component %in% c("optimizer_convergence", "pd_hessian"),
       c("component", "status")]
```

# Does it recover what it claims to estimate?

Two claims are on the table: the per-species environmental responses
$\beta_s$, and the per-species recording-bias associations $\gamma_s$.
Both are single-draw illustrations — the spread you see is sampling
noise at this design size, not bias.

```{r beta-figure, fig.cap = "Per-species wetland-gradient slopes, simulated truth against estimate, from one fit on one simulated draw."}
est_beta <- unname(fit$opt$par[grep(":env", fit$X_fix_names)])
plot(beta, est_beta, pch = 19, cex = 1.3,
     xlim = range(c(beta, est_beta)) + c(-0.15, 0.15),
     ylim = range(c(beta, est_beta)) + c(-0.15, 0.15),
     xlab = "true wetland-gradient slope",
     ylab = "estimated")
abline(0, 1, lty = 2)
text(beta, est_beta, labels = species, pos = 3, cex = 0.8)
```

```{r gamma-figure, fig.cap = "Per-species recording-bias associations, simulated truth against estimate. These are recording-process quantities, not ecology; their identification rests on the bias covariate varying separately from the environmental predictors."}
est_gamma <- unname(fit$opt$par[grep(":access", fit$X_fix_names)])
plot(gamma, est_gamma, pch = 19, cex = 1.3,
     xlim = range(c(gamma, est_gamma)) + c(-0.15, 0.15),
     ylim = range(c(gamma, est_gamma)) + c(-0.15, 0.15),
     xlab = "true recording-bias association",
     ylab = "estimated")
abline(0, 1, lty = 2)
text(gamma, est_gamma, labels = species, pos = 3, cex = 0.8)
```

The two panels are not equally tight, and the difference is
instructive. The ecological slopes recover closely; the recording-bias
associations scatter more, because they are informed only by how record
counts co-vary with accessibility after effort — a weaker signal at
this record density. That pattern is a reasonable expectation for
models like this, not a guarantee from one draw. Because the bias
associations are estimated *jointly* with the
ecology, a species recorded mostly near roads is still not mistaken for
a species that prefers whatever habitat roads run through — provided,
again, that roads and habitat vary separately.

# The joint part: residual co-occurrence

The quantity a stack of single-species models cannot give you is the
between-cell species correlation after environment, effort, and bias
are accounted for — the rotation-invariant summary of
$\Sigma = \Lambda \Lambda^\top + \Psi$:

```{r correlations}
corr_rows <- extract_correlations(
  fit,
  tier = "unit",
  method = "fisher-z"
)
head(corr_rows[order(-abs(corr_rows$correlation)), ], 5)
```

The `lower`/`upper` bounds and the `interval_status` column are
Fisher-z heuristics, and the fit reports them as exactly that —
`heuristic_unvalidated`. Calibrated coverage for this route has not
been established; read the bounds as descriptive spread, not as a
coverage guarantee.

The point estimates recover the block structure written into
$\Lambda$: the two plants load together (the strongest printed pair),
and every printed bird–plant pair is negative, matching their
opposite-signed loadings. In a real analysis these correlations are
descriptions of residual co-occurrence, not causal interaction claims.

# What this model cannot tell you

Everything above is **relative** intensity, and the boundary is worth
stating precisely:

- **No absolute quantities.** Abundance, occupancy probability, and
  detectability are all inaccessible from opportunistic records alone,
  because the recording process and the ecological process enter the
  intensity multiplicatively and only their product is observed.
- **The effort offset is assumed known.** If $a_c$ is itself estimated
  or mis-specified, its error moves directly into the intercepts and,
  when correlated with covariates, into the slopes.
- **Bias identification is a design property.** When the bias covariate
  tracks the environment, $\beta_s$ and $\gamma_s$ are confounded and
  no fitting choice can separate them.

The way past each of these limits is more data of a different kind, not
more records of the same kind. Adding even a small designed survey —
detection/non-detection with known effort — anchors the recording
levels and is exactly what the integrated route provides:

# See also

- [Integrating opportunistic records with a repeated survey](integrated-two-source-example.html)
  — the two-source integrated model this article's fences point toward.
- [How big does an integrated survey design need to be?](integrated-survey-design.html)
- [More than two sources](integrated-multi-source-example.html)
- [Joint species distribution model](joint-sdm.html) — the binary-occurrence
  JSDM guide, without the point-process framing.

# References

Blanchet, F.G., Cazelles, K. and Gravel, D. (2020). Co-occurrence is not
evidence of ecological interactions. *Ecology Letters* **23**, 1050–1063.

Elith, J., Graham, C.H., Anderson, R.P., et al. (2006). Novel methods improve
prediction of species' distributions from occurrence data. *Ecography* **29**,
129–151.

Renner, I.W., Elith, J., Baddeley, A., Fithian, W., Hastie, T., Phillips,
S.J., Popovic, G. and Warton, D.I. (2015). Point process models for
presence-only analysis. *Methods in Ecology and Evolution* **6**, 366–379.
<https://doi.org/10.1111/2041-210X.12352>

Warton, D.I., Renner, I.W. and Ramp, D. (2013). Model-based control of
observer bias for the analysis of presence-only data in ecology. *PLoS ONE*
**8**, e79168. <https://doi.org/10.1371/journal.pone.0079168>

Warton, D.I., Blanchet, F.G., O'Hara, R.B., Ovaskainen, O., Taskinen, S.,
Walker, S.C. and Hui, F.K.C. (2015). So many variables: joint modeling in
community ecology. *Trends in Ecology & Evolution* **30**, 766–779.

Zurell, D., Pollock, L.J. and Thuiller, W. (2018). Do joint species
distribution models reliably detect interspecific interactions from
co-occurrence data in homogenous environments? *Ecography* **41**, 1812–1819.
