---
title: "gllvmTMB vocabulary"
description: "Look up the model terms, covariance pieces, grouping tiers, identifiability concepts, and uncertainty language used across the guides."
tier: 2
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{gllvmTMB vocabulary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

GLLVMs combine vocabulary from multivariate statistics, ordination, factor
analysis, and mixed models. This glossary translates the terms used in the
public `gllvmTMB` guides and points to the page where each idea is demonstrated.

These definitions describe the current public interface. They do not imply
that every combination of response family, covariance source, and inferential
method has equal evidence.

## Data and formula terms

### Trait and response matrix

A **trait** is one response variable in the multivariate model. Depending on the
application, traits may be species occurrences, morphological measurements,
questionnaire items, or study outcomes. Placing units in rows and traits in
columns gives the **response matrix**.

### Unit and grouping key

A **unit** is the entity carrying the multivariate response: for example, a
site, individual, species, or study. A **grouping key** is a data column that
identifies which observations share a random-effect or covariance component.
In `gllvmTMB()`, `unit` identifies the between-unit level, `unit_obs` identifies
within-unit replicate cells, `cluster` supplies a further ordinary or structured
grouping, and optional `cluster2` supplies an additional plain-diagonal grouping.
These are modelling roles rather than fixed biological meanings: the columns
that fill them depend on the study design.

### Long and wide data

**Stacked long data** have one row per `(unit, trait)` observation, with a
response column such as `value`, a `trait` factor, and the grouping columns.
This is the package's internal representation and the most explicit formula
form.

**Wide data** have one row per unit and one column per trait. The `traits(...)`
left-hand side converts that shape to the same stacked representation. These two
formulas express the same two-LV model. Long calls name both the trait and unit
columns explicitly. Wide calls omit `trait =` because `traits(...)` supplies
the response columns, but still name `unit`; override the default
`unit = "site"` whenever another column carries that role.

```{r long-wide-formulas, eval = FALSE}
# Long
gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | site, d = 2),
  data = dat_long,
  trait = "trait",
  unit = "site"
)

# Wide
gllvmTMB(
  traits(t1, t2, t3) ~ 1 +
    latent(1 | site, d = 2),
  data = dat_wide,
  unit = "site"
)
```

### Family and link

The **family** is the probability distribution used for a response, such as
Gaussian, binomial, or Poisson. The **link function** maps the response mean to
the linear-predictor scale: identity for the usual Gaussian model, logit for a
common binomial model, and log for a common Poisson model. See [Response
families](response-families.html) before combining families in one fit.

## Latent-variable terms

### Latent variable and latent score

A **latent variable (LV)** is an unobserved dimension used to represent shared
variation among traits. `d = 2` asks for two LVs, conventionally labelled `LV1`
and `LV2`.

A **latent score** is a particular unit's fitted position on those LVs. In the
ordinary latent model, unit $i$ has score vector $\mathbf u_i$ with the
identifying convention $\operatorname{Var}(\mathbf u_i)=\mathbf I_d$ before
conditioning on its responses. Fitted scores are conditional estimates from
the likelihood, responses, predictors, and covariance structure; they are not
directly observed measurements. Use `extract_ordination()` to obtain scores and
loadings together.

### Loading and loading matrix

A **loading** $\Lambda_{tk}$ describes how strongly trait $t$ changes with
LV $k$. The $T \times d$ **loading matrix** $\boldsymbol\Lambda$ contains
one row per trait and one column per LV. Large absolute values indicate a strong
relationship with that fitted direction on the linear-predictor scale. Their
magnitudes are not directly comparable across response families or links.
Individual loading signs and axes also depend on orientation, so interpret
rotation-invariant summaries first.

### Shared low-rank covariance

The product


\[
\boldsymbol\Lambda\boldsymbol\Lambda^{\mathsf T}
\]

is the **shared low-rank covariance** generated by the LVs. It is unchanged by
an orthogonal rotation of the loading matrix and is therefore safer to compare
across refits than raw loading columns.

### Psi: tier-specific diagonal variance

$\boldsymbol\Psi = \mathrm{diag}(\psi_1, \ldots, \psi_T)$ is a diagonal
covariance component at a named model tier. It captures trait-specific variance
at that tier that is not shared through the LVs. Ordinary `latent()` includes
this diagonal companion by default; `latent(..., unique = FALSE)` requests the
loadings-only subset.

Psi is not a generic synonym for every residual variance. In particular, it is
distinct from Gaussian observation residual variance, a diagonal component at
another grouping tier, and a distribution-specific link-scale residual used by
some extractors.

### Sigma: model-implied trait covariance

For the ordinary latent decomposition at one tier,

\[
\boldsymbol\Sigma
=
\boldsymbol\Lambda\boldsymbol\Lambda^{\mathsf T}
+
\boldsymbol\Psi.
\]

$\boldsymbol\Sigma$ is the **model-implied trait covariance** at that tier.
Use `extract_Sigma()` for matrices and `extract_Sigma_table()` for report-ready
rows. See [Covariance and correlation](covariance-correlation.html) for the
worked interpretation.

### Communality

**Communality** is the fraction of a trait's variance at a chosen tier and scale
that is attributed to the shared LV component:

\[
c_t^2 =
\frac{(\boldsymbol\Lambda\boldsymbol\Lambda^{\mathsf T})_{tt}}
     {\Sigma_{tt}}.
\]

It is not automatically the fraction of total phenotypic variance across the
whole model. If Psi is excluded by construction, communality can equal one
mechanically; that is not evidence that the traits are perfectly integrated.

### Model-based ordination

**Model-based ordination** displays fitted latent scores and trait loadings from
a probability model. Unlike PCA or NMDS, its mean structure and residual
behavior depend on the chosen response family and link. The displayed axes are
still orientation-dependent, so conclusions should be checked against
rotation-invariant covariance summaries.

## Covariance-structure terms

### Covariance tier

A **covariance tier** is one variance component in the model, attached to a
grouping key and interpreted at that level. Between-unit and within-unit
components are different tiers even when they use the same covariance mode.
Do not use *tier*, *grouping key*, *correlation source*, and *LV axis* as
synonyms.

### Correlation source

The **correlation source** says how levels of a grouping factor are related:
ordinary independent levels, pedigree/animal relatedness, phylogenetic
relatedness, spatial dependence, or a supplied dense kernel. Phylogenetic and
spatial labels describe structured covariance sources or variance-share
categories; they are not peer data columns alongside `unit` and `unit_obs`.

### Covariance mode

The **covariance mode** describes the trait-by-trait covariance within a tier:

- `scalar`: one common variance where that source supplies a scalar form;
- `indep`: separate trait variances with zero cross-trait covariance;
- `dep`: a full unstructured trait covariance;
- `latent`: reduced-rank shared covariance, with the diagonal Psi convention
  determined by the keyword family and its `unique` argument.

The canonical source-by-mode interface crosses ordinary, animal,
phylogenetic, and spatial sources with four taught modes: scalar, independent,
dependent, and latent. A separate dense-kernel interface provides
`kernel_indep()`, `kernel_dep()`, and `kernel_latent()`.
See the [API keyword grid](api-keyword-grid.html) for exact syntax.

## Scale and identification terms

### Fitted covariance and augmented link scale

The covariance fitted for a random-effect tier is not always the same estimand
as a marginal or liability-scale covariance. With `link_residual = "none"`, an
extractor reports the fitted model covariance. With `link_residual = "auto"`,
supported extractors add a family-specific diagonal convention before
standardising or partitioning variance. This is a post-fit scale choice, not an
extra likelihood term, and some family-specific additions are approximations.

### Rotation equivalence and loading constraints

If $\mathbf Q$ is orthogonal, $\boldsymbol\Lambda$ and
$\boldsymbol\Lambda\mathbf Q$ imply the same shared covariance. The identified
object is therefore an equivalence class—or its covariance—not a unique raw
loading orientation. A loading constraint or post-hoc rotation chooses an
orientation; it does not create information that the data lack.

### Weak identification

A parameter is **weakly identified** when the likelihood changes little over a
range of substantively different values. Large standard errors, unstable
refits, flat profiles, or collapsing extra LVs can be warning signs.
Repeated starts, refits, and likelihood profiles can reveal instability under a
fitted scenario, but they do not by themselves prove that the original data
identify a particular rank. Testing a rank-selection procedure requires
simulations with a predeclared generating rank and a denominator that retains
failed or unusable fits.

### Boundary

A **boundary** is the edge of the parameter space, such as a variance equal to
zero or a correlation near $-1$ or $1$. Ordinary quadratic approximations
can be unreliable there. Profile intervals can display asymmetric or one-sided
likelihood information, but their reference distribution and calibration are
target-specific. See [Profile likelihood: routes and
fallbacks](profile-likelihood-ci.html).

## Fit and uncertainty terms

### Convergence, gradient, and Hessian

**Optimizer convergence** means the numerical optimizer stopped according to
its criterion. The **gradient** measures the remaining local slope of the
objective. The conservative package verdict requires an unscaled maximum
gradient below `0.01`; `scaled_gradient` is descriptive and cannot rescue a
failed optimiser or a large raw gradient. The **Hessian** measures local
curvature. A positive-definite Hessian supports local standard-error
calculations, but does not prove identifiability, model adequacy, or interval
calibration. Use [Fit diagnostics](fit-diagnostics.html) before interpreting
estimates.

### Wald, profile, bootstrap, and Fisher-z intervals

A **Wald interval** uses local Hessian/delta-method curvature on a chosen working
scale. A **profile interval** inverts a likelihood profile. A **parametric
bootstrap interval** simulates and refits from the fitted model. A **Fisher-z
interval** transforms a correlation with $\operatorname{atanh}(\rho)$, uses
the approximate standard error $1 / \sqrt{n_{\mathrm{eff}} - 3}$, and
back-transforms the limits to $[-1, 1]$. In mixed and non-Gaussian latent
models, treat this as a sensitivity interval based on an explicit or
tier-derived effective sample size, not as Hessian-based inference or a
coverage guarantee.

None is universally best. The implemented route, fallback, cutoff, and evidence
depend on the target. Inspect method or status fields where the extractor
provides them, together with its messages and warnings, rather than assuming
the requested method was used.

## Missing-data terms

### Response mask

A **response mask** retains the identity of a missing response cell while
giving that cell zero contribution to the response likelihood. With
`miss_control(response = "include")`, `predict_missing()` later returns a
plug-in fitted value for each masked cell; it does not currently return a
prediction interval.

### Modelled missing predictor and conditional mode

`mi(x)` declares a modelled missing predictor and `impute` specifies its
predictor model. The predictor may live at the unit level, a coarser
`mi_group()` level, or a phylogenetic-tip level. For a continuous Gaussian
predictor, `imputed()` returns a fitted **conditional mode** and, when
available, a local Gaussian/Laplace conditional standard error from the joint
Hessian. It is not a conditional mean, a multiple-imputation draw, a Bayesian
posterior summary, or a calibrated interval. See [Handling missing
data](missing-data.html).

## Where to go next

- [Get started](../articles/gllvmTMB.html): fit and interpret the first model.
- [API keyword grid](api-keyword-grid.html): choose a covariance source and
  mode.
- [Covariance and correlation](covariance-correlation.html): interpret
  $\boldsymbol\Lambda$, $\boldsymbol\Psi$, and $\boldsymbol\Sigma$.
- [Fit diagnostics](fit-diagnostics.html): decide whether a fitted model is
  numerically trustworthy.
- [Handling missing data](missing-data.html): choose between response masking
  and a missing-predictor model.
