---
title: "Integrating opportunistic records with a repeated survey: a small worked example"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gllvmTMB)
set.seed(20260815)
```

> **Developer preview.** The integrated two-source route shown here is
> experimental and unexported: the entry point is internal, its interface may
> change without deprecation, and no released capability claim is made. The
> example exists so that early users can see the intended workflow end to end
> on simulated data.

# The question

Opportunistic portal records (GBIF-style) are abundant but arrive where
people go; a designed detection/non-detection survey is small but has known
effort. This example asks a deliberately modest question of a **simulated
three-species grassland songbird guild** on a prairie-inspired landscape:
after giving the portal records their own recording-bias term, do the two
sources agree on one shared habitat gradient?

Both sources see the same ecological state for species $s$ in cell $c$,

$$
\eta_{cs} = \alpha_s + x_c\,\beta_s + u_c\,\lambda_s,
$$

with $x_c$ a measured habitat covariate and $u_c$ an unmeasured shared
spatial gradient carried by a latent field with species loadings
$\lambda_s$. The portal branch adds what only it suffers -- an
accessibility bias with species-specific coefficients $\gamma_s$ and its own
spatial recording field -- while the survey branch is a plain
complementary-log-log detection/non-detection likelihood with known visit
support:

$$
N^{G}_{cs} \sim \text{Poisson}\!\left[a^{G}_c
  \exp\{\eta_{cs} + b_c\gamma_s + h_c\,\delta_s\}\right],
\qquad
Y^{S}_{csv} \sim \text{Bernoulli}\!\left[1 -
  \exp\{-a^{S}_{cv}\exp(\eta_{cs})\}\right].
$$

Everything is **relative intensity**: no absolute abundance, occupancy, or
detectability claim is available from this design.

# Simulate a small two-source data set

One hundred and eight cells, three species, one portal record stream, three
survey visits. The fields here are simple smooth surfaces -- enough to give
the model something to find without pretending to be a validation campaign.

```{r simulate}
n_lon <- 12; n_lat <- 9
grid <- expand.grid(lon = seq(0, 1.2, length.out = n_lon),
                    lat = seq(0, 0.9, length.out = n_lat))
grid <- grid[order(grid$lat, grid$lon), ]
n_cell <- nrow(grid)
cells   <- paste0("cell_", seq_len(n_cell))
species <- c("gralark", "dickcis", "bobolnk")   # simulated guild

## habitat covariate, shared gradient u, and an accessibility index
x <- as.numeric(scale(grid$lon))                        # measured habitat
u <- 0.8 * sin(pi * grid$lat / 0.9) * cos(pi * grid$lon / 1.2)  # unmeasured
b <- as.numeric(scale(-grid$lat + 0.3 * cos(2 * pi * grid$lon)))# access bias
## the DGP here has NO portal-only spatial field (delta = 0): the fitted
## model still estimates one, and treating it as a nuisance is the honest
## reading at this design size (see the closing cautions)

alpha  <- c(-0.3, 0.1, 0.4)
beta   <- c(0.6, -0.4, 0.5)
lambda <- c(0.9, 0.7, -0.6)
gamma  <- c(0.5, -0.3, 0.2)

eta <- outer(u, lambda) +
  sweep(outer(x, beta), 2, alpha, "+")                  # n_cell x 3

## portal effort concentrates in accessible cells; survey support is designed
a_g <- 3 * exp(0.4 * b + rnorm(n_cell, sd = 0.2))
a_s <- rep(0.9, n_cell)

base <- expand.grid(cell_id = cells, trait = species,
                    stringsAsFactors = FALSE)
base <- base[order(match(base$cell_id, cells), match(base$trait, species)), ]
ci <- match(base$cell_id, cells)
si <- match(base$trait, species)

gbif <- transform(base, source = "gbif", survey_event_id = NA_character_,
  branch = "count", support = a_g[ci], lon = grid$lon[ci], lat = grid$lat[ci],
  value = rpois(nrow(base),
    a_g[ci] * exp(eta[cbind(ci, si)] + b[ci] * gamma[si])),
  visit = NA_integer_)

pa <- do.call(rbind, lapply(1:3, function(v) transform(base,
  source = "survey", survey_event_id = paste0("pa_v", v, "_", base$cell_id),
  branch = "pa", support = a_s[ci], lon = grid$lon[ci], lat = grid$lat[ci],
  value = rbinom(nrow(base), 1, -expm1(-a_s[ci] * exp(eta[cbind(ci, si)]))),
  visit = v)))

rows <- rbind(gbif, pa)
X <- matrix(x[match(rows$cell_id, cells)], ncol = 1,
            dimnames = list(NULL, "env"))
B <- matrix(ifelse(rows$source == "gbif",
                   b[match(rows$cell_id, cells)], NA_real_),
            ncol = 1, dimnames = list(NULL, "bias"))
table(rows$source)
```

# Fit the integrated model

The two sources enter one likelihood: shared fixed effects and shared latent
field, a portal-only bias coefficient and portal-only spatial field, and the
survey's cloglog detection branch -- selected internally per row. A mesh over
the cells carries the spatial terms.

```{r fit}
mesh <- make_mesh(rows[, c("lon", "lat")], c("lon", "lat"), cutoff = 0.085)
fit <- gllvmTMB:::.gll_isdm_fit(rows = rows, X = X, B = B, d = 1,
                                mesh = mesh, spatial = TRUE,
                                control = gllvmTMBcontrol(), silent = TRUE)
fit$opt$convergence
```

# What did the two sources agree on?

The portal-bias coefficients are the part of the model only the integration
can give you -- the survey anchors the shared state, so $\gamma$ is
identified rather than silently absorbed. At this deliberately small design
expect useful but noisy estimates:

```{r gamma}
idx <- grep("b_bias", fit$X_fix_names)
round(rbind(truth = gamma,
            estimate = unname(fit$opt$par[idx])), 2)
```

The shared-state loadings and the portal-only field's covariance are
reported on the link scale (both are sign-invariant as outer products):

```{r sigma}
round(fit$report$Sigma_spde_slope_intercept, 3)
round(fit$report$Sigma_spde_slope_slope, 3)
```

# What to check before believing more than this

Three honest cautions, each with a package-level reason.

1. **Design size governs identifiability.** Source-specific spatial
   structure is only weakly identified on small grids: internal known-truth
   campaigns place reliable recovery of a portal-only *field* far beyond a
   design of this size, with identifiability bought by adding cells (at
   maintained per-cell records), not by piling records into the same cells.
   Treat the portal-only field here as a nuisance adjustment, not an
   estimate to interpret.
2. **Relative intensity only.** Nothing in this design identifies absolute
   abundance or detectability; comparisons are within-map, not across
   species in absolute terms.
3. **Experimental surface.** The entry point is internal and unexported;
   check convergence and positive-definiteness on every fit, and expect the
   interface to change.

```{r session, include=FALSE}
## kept minimal deliberately; this article is a workflow illustration only
```
