---
title: "How big does an integrated survey design need to be?"
description: "Audit the per-cell support of a presence-only plus structured-survey design before you fit it, then read a known-truth design curve showing that spatial replication pays only when per-cell information is held fixed."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{How big does an integrated survey design need to be?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  cache = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(gllvmTMB)
set.seed(20260816)
```

> **Experimental.** The model on this page is fitted through the ordinary
> `gllvmTMB()` entry point — there is no separate integrated-model function —
> but the route is experimental: the interface may change without deprecation,
> and no released capability claim is made.

# The question

You have tens of thousands of opportunistic portal records and money for a
limited structured survey. The portal records arrive where people go; the
survey is small but has known effort. Before writing the grant, you need to
answer a question that is not about estimation at all:

**where should the survey effort go, and is the design big enough?**

This page is about judging a design *before* you fit it. Its companion,
[Integrating opportunistic records with a repeated
survey](integrated-two-source-example.html), asks the estimation question —
*do the two sources agree on one shared gradient?* — and answers it with a
single fit. Read that one first if you want the model explained. Read this
one if you are deciding how many cells to lay out.

The short version of the answer, and the whole argument of this page:

> **Spatial replication helps if and only if per-cell information is
> preserved.** Extra cells buy identifiability that extra records piled into
> the same cells do not. That is a mechanism with conditions attached, not a
> rule of thumb — the conditions are set out in
> [What the curve does not tell you](#what-the-curve-does-not-tell-you), and
> they matter more than the headline.

Everything here is **relative intensity**. Presence-only data cannot identify
absolute abundance, occupancy, or detectability (Fithian et al. 2015,
*Methods in Ecology and Evolution* 6:424–438), and nothing on this page
changes that.

# The design levers

Both sources see the same ecological state for species $s$ in cell $c$,

$$
\eta_{cs} = \alpha_s + x_c\,\beta_s + u_c\,\lambda_s ,
$$

with $x_c$ a measured habitat covariate and $u_c$ an unmeasured spatial
gradient carried by a latent field with species loadings $\lambda_s$. The
portal branch adds an accessibility bias with species-specific coefficients
$\gamma_s$; the survey branch is a complementary-log-log
detection/non-detection likelihood with known visit support. The companion
article writes both likelihoods out in full.

Three things are yours to choose, and they are not interchangeable:

- **the number of cells**, $C$ — how far the design reaches, and how many
  independent looks the shared field gets;
- **the per-cell record support** — how much portal effort lands in a typical
  cell;
- **the survey replication** — how many visits each surveyed cell receives.

The rest of this page is about the first two, because that is where the
measured evidence sits.

# Audit your own design first

The audit below needs no model. It is arithmetic on your own data, and it is
the step that decides whether the design curve later on this page is even
relevant to you.

We use a **simulated three-species desert lizard guild** on an arid
south-western landscape; the species names are invented. Portal records
concentrate near roads, so accessibility drives portal effort while the
survey's support is designed.

```{r simulate}
n_lon <- 10; n_lat <- 8
grid <- expand.grid(lon = seq(0, 1.0, length.out = n_lon),
                    lat = seq(0, 0.8, length.out = n_lat))
grid <- grid[order(grid$lat, grid$lon), ]
n_cell  <- nrow(grid)
cells   <- paste0("cell_", seq_len(n_cell))
species <- c("spinytail", "bandedgecko", "dunerunner")   # simulated guild

## measured habitat, an unmeasured shared gradient, and road accessibility
x <- as.numeric(scale(grid$lon))
u <- 0.8 * sin(pi * grid$lat / 0.8) * cos(pi * grid$lon / 1.0)
b <- as.numeric(scale(-grid$lat + 0.3 * cos(2 * pi * grid$lon)))

alpha  <- c(-0.2,  0.2, 0.4)
beta   <- c( 0.6, -0.4, 0.5)
lambda <- c( 0.9,  0.7, -0.6)
gamma  <- c( 0.5, -0.3, 0.2)

eta <- outer(u, lambda) + sweep(outer(x, beta), 2, alpha, "+")

## portal effort follows accessibility; survey support is uniform by design
a_g <- 3 * exp(0.4 * b + rnorm(n_cell, sd = 0.2))
a_s <- rep(0.9, n_cell)

base <- expand.grid(cell_id = cells, trait = species,
                    stringsAsFactors = FALSE)
base <- base[order(match(base$cell_id, cells), match(base$trait, species)), ]
ci <- match(base$cell_id, cells)
si <- match(base$trait, species)

## the contract's source labels are fixed strings, "gbif" and "survey";
## they name the two roles, not a particular data provider
portal <- transform(base, source = "gbif", support = a_g[ci],
  lon = grid$lon[ci], lat = grid$lat[ci],
  value = rpois(nrow(base),
    a_g[ci] * exp(eta[cbind(ci, si)] + b[ci] * gamma[si])))

pa <- do.call(rbind, lapply(1:2, function(v) transform(base,
  source = "survey", support = a_s[ci],
  lon = grid$lon[ci], lat = grid$lat[ci],
  value = rbinom(nrow(base), 1,
    -expm1(-a_s[ci] * exp(eta[cbind(ci, si)]))))))

rows <- rbind(portal, pa)
c(cells = n_cell, portal_rows = nrow(portal), survey_rows = nrow(pa))
```

Now the audit itself. Three numbers decide whether your cell count means what
you think it means.

```{r audit}
recs <- tapply(portal$value, portal$cell_id, sum)
recs <- recs[cells]

audit <- c(
  cells                = length(recs),
  total_records        = sum(recs),
  median_per_cell      = median(recs),
  cells_with_no_record = sum(recs == 0),
  share_in_top_decile  = sum(sort(recs, decreasing = TRUE)[
                              seq_len(ceiling(0.1 * length(recs)))]) / sum(recs)
)
round(audit, 3)
```

`share_in_top_decile` is the one to watch. Perfectly even effort would put
0.10 of the records in the best-sampled tenth of cells; this deliberately mild
simulation puts about 0.31. When that number climbs toward 1, the *effective*
number of well-supported cells is far below your cell count, and a plan to
"add more cells" will in practice spread the same records thinner. The design
curve below was measured under the opposite condition — per-cell support held
fixed while cells were added — and it does not describe the thinning case.

```{r fig-audit, fig.cap = "Design audit of the simulated portal stream. Left: records per cell, showing how far the per-cell support spreads. Right: the cumulative share of all records against the cumulative share of cells, ranked from best- to worst-sampled; the diagonal is even effort. The gap between the curve and the diagonal is the part of your cell count that is not doing the work you think it is.", fig.width = 7, fig.height = 3.6}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 1.0), cex = 0.9)

hist(recs, breaks = 12, col = "grey85", border = "white",
     main = "Records per cell", xlab = "portal records in a cell", ylab = "cells")
abline(v = median(recs), lty = 2)

srt <- sort(recs, decreasing = TRUE)
plot(seq_along(srt) / length(srt), cumsum(srt) / sum(srt), type = "l", lwd = 2,
     xlim = c(0, 1), ylim = c(0, 1), main = "Concentration of effort",
     xlab = "cumulative share of cells", ylab = "cumulative share of records")
abline(0, 1, lty = 3)

par(op)
```

# One fit, to fix the workflow

The design question does not need a big fit, but it does need the workflow to
be concrete. This is the same route the companion article walks through, on a
deliberately small design, so that the rest of the page has something to point
at.

Four columns do the integrating work: a **source gate** so that portal-only
terms are structurally absent from the survey likelihood, a **log support**
offset carrying known effort, a **structurally zero** accessibility covariate
on survey rows, and a per-row **family selector**.

```{r assemble}
dat <- rows
dat$trait   <- factor(dat$trait)
dat$cell_id <- factor(dat$cell_id)

dat$isdm_gbif    <- as.integer(dat$source == "gbif")   # source gate
dat$log_support  <- log(dat$support)                   # known effort
dat$env    <- x[match(as.character(dat$cell_id), cells)]
dat$access <- ifelse(dat$isdm_gbif == 1L,              # portal-only, so a
                     b[match(as.character(dat$cell_id), cells)], 0)  # zero,
                                                       # never an NA
dat$isdm_family <- factor(
  ifelse(dat$source == "gbif", "gbif", "survey_pa"),
  levels = c("gbif", "survey_pa"))
fam <- list(gbif = poisson(), survey_pa = binomial(link = "cloglog"))
attr(fam, "family_var") <- "isdm_family"
```

The route is intrinsically long-format: a per-row family selector has no
wide-data-frame equivalent, because a wide frame has one row per unit and this
model needs one row per (cell, species, source). There is no `traits(...)`
twin for this call.

```{r fit}
mesh <- make_mesh(dat[, c("lon", "lat")], c("lon", "lat"), cutoff = 0.12)
fit <- gllvmTMB(
  value ~ 0 + trait + trait:env + trait:isdm_gbif + trait:access +
    offset(log_support) +
    indep(0 + trait | cell_id) +
    spatial_latent(0 + trait | cell_id, d = 1),
  data   = dat,
  trait  = "trait",
  unit   = "cell_id",
  family = fam,
  mesh   = mesh,
  silent = TRUE
)
fit$opt$convergence
```

A converged optimizer is not an admissible fit, and admissibility is the very
thing this page is about — so run the check here rather than taking the
convergence code at face value:

```{r fit-health}
health <- check_gllvmTMB(fit)
health[health$component %in% c("optimizer_convergence", "pd_hessian"),
       c("component", "status")]
```

The portal-bias coefficients come back in the order `factor()` chose, so
realign the simulated truth before comparing:

```{r gamma}
idx <- grep("access", fit$X_fix_names)
tab <- rbind(truth    = gamma[match(levels(dat$trait), species)],
             estimate = unname(fit$opt$par[idx]))
colnames(tab) <- levels(dat$trait)
round(tab, 2)
```

Eighty cells is roughly a fifth of the smallest design in the evidence below,
and these are single-draw numbers, so read them as a
demonstration that the route runs and the columns line up rather than as a
recovery result. They do carry one design lesson, though, and it is the reason
this page exists: **different quantities in the same model have different
design requirements.** An accessibility coefficient is a fixed effect that
every cell contributes to, so it is comparatively cheap. The amplitude of a
spatial field is not — it is estimated from how the field varies *across*
cells, so it is the quantity that has to be bought with cells. That is what
the next section measures.

# The design curve

The numbers in this section come from an internal known-truth simulation
campaign: **1,600 fits, no errors**, run on a shared compute server. The
domain was grown at **fixed spatial range and fixed cell size**, so the number
of cells rises *without* diluting the sampling in any one cell. That
separation is the point of the design; it is what lets the curve say something
about cells specifically rather than about "more data" in general.

They are hard-coded here, exactly as the campaign recorded them, so that you
can see the evidence rather than a summary of it. Nothing on this page re-runs
the campaign — a single fit at the largest design point took about 107 s.

```{r curve-data}
## per-cell effort level E = 2 (the higher of the campaign's two settings)
design_curve <- data.frame(
  n_cell  = c(  360,   810,  1440,  2250),
  med_rel = c(0.993, 0.985, 0.981, 0.633),  # lower is better
  pd_rate = c(0.330, 0.430, 0.455, 0.555),  # higher is better
  cos95   = c(0.255, 0.375, 0.470, 0.575),  # higher is better
  conv    = c(0.940, 0.985, 0.995, 1.000)   # higher is better
)

## the lower effort level, E = 1: only the two endpoints were reported
e1_endpoints <- data.frame(n_cell  = c(  360,  2250),
                           med_rel = c(0.997, 0.982))

design_curve
```

What the four columns are:

- **`med_rel`** — the median relative error in the recovered amplitude of the
  spatial field. This is the campaign's headline accuracy axis, and its target
  was **0.25 or below**. Lower is better.
- **`conv`** — the proportion of fits that converged.
- **`pd_rate`** — the proportion of fits reaching an admissible solution, in
  the positive-definite sense the package checks on every fit.
- **`cos95`** — the campaign's agreement score between the recovered and the
  true gradient direction. Higher is better.

Every one of the four moves monotonically in the right direction as cells are
added, and `conv` reaches 1.000.

Read `pd_rate` before taking much comfort from that. `conv` reaching 1.000 is
the *optimizer* declaring itself satisfied; `pd_rate` is the share of those
fits whose curvature is actually admissible, and at the largest design tested
— 2,250 cells — it is **0.555**. Barely more than half. It rises from 0.330,
so the trend is real, but it is the axis furthest from solved and its
destination lies outside the measured span. Adding cells at fixed per-cell
support *improves* admissibility; on this evidence it does not yet deliver it.

```{r fig-curve, fig.cap = "What extra cells buy, at fixed spatial range and fixed cell size. Left: median relative amplitude error falls, but does not reach the campaign's 0.25 target within the measured span; the shaded region is beyond any measurement, and the grey open points are the lower effort level (endpoints only). Right: the three higher-is-better axes, all monotone. The horizontal axis is on a log scale in both panels.", fig.width = 7.5, fig.height = 4}
op <- par(mfrow = c(1, 2), mar = c(4.4, 4.4, 2.6, 1.0), cex = 0.9)

## --- amplitude ---
plot(design_curve$n_cell, design_curve$med_rel, type = "n", log = "x",
     xlim = c(320, 20000), ylim = c(0, 1.05),
     xlab = "cells (log scale)", ylab = "median relative amplitude error",
     main = "Amplitude: better, not yet good")
rect(2600, -0.1, 30000, 1.2, col = grey(0.92), border = NA)
text(7600, 0.86, "beyond the\nmeasured span", cex = 0.8, col = grey(0.35))
box()
abline(h = 0.25, lty = 3)
text(430, 0.30, "campaign target 0.25", cex = 0.75, adj = 0)
lines(e1_endpoints$n_cell, e1_endpoints$med_rel, type = "b",
      pch = 1, lty = 2, col = grey(0.45))
lines(design_curve$n_cell, design_curve$med_rel, type = "b", pch = 19, lwd = 2)
legend("bottomleft", bty = "n", cex = 0.75,
       legend = c("higher effort (E = 2)", "lower effort (E = 1), endpoints"),
       pch = c(19, 1), lty = c(1, 2), col = c("black", grey(0.45)))

## --- the higher-is-better axes ---
plot(design_curve$n_cell, design_curve$conv, type = "n", log = "x",
     xlim = c(320, 2600), ylim = c(0, 1.05),
     xlab = "cells (log scale)", ylab = "proportion of fits",
     main = "Admissibility and agreement")
for (j in seq_along(v <- c("conv", "pd_rate", "cos95"))) {
  lines(design_curve$n_cell, design_curve[[v[j]]], type = "b",
        pch = c(15, 17, 18)[j], lty = j, lwd = 2, col = grey(c(0, 0.35, 0.6)[j]))
}
legend("bottomright", bty = "n", cex = 0.75, legend = v,
       pch = c(15, 17, 18), lty = 1:3, col = grey(c(0, 0.35, 0.6)))

par(op)
```

# Why it is cells, and not records

A monotone improvement on its own is weak evidence: almost any axis improves
when you add data. What makes this a statement about *cells* is a second arm
of the same programme, in which the spatial **range was shrunk** instead —
which also raises the number of patches the field spans, but does so by making
each patch smaller relative to the sampling. That arm moved in the **opposite
direction on every one of the four axes**.

Two arms, both increasing spatial replication, opposite outcomes. The
difference between them is whether per-cell information survived the change.
Hence the conditional form:

> Spatial replication helps **if and only if** per-cell information is
> preserved. Cells buy what records cannot.

This is also why the audit comes before the curve. Adding cells to a design
whose records are already concentrated in a tenth of the map is closer to the
range-shrink arm than to the domain-growth arm, and the curve will not
describe it.

# What the curve does not tell you {#what-the-curve-does-not-tell-you}

This is the more important half of the page, and it is deliberately longer
than the claim.

**The target was not reached.** At the campaign's own accuracy target
(`med_rel` at or below 0.25), 2,250 cells is **not enough**. The last doubling
moved the amplitude error from 0.98 to 0.63 — real progress, and still nowhere
near the target. Any figure for how many cells *would* be enough is an
extrapolation past the last measured point; the campaign's own reading is
roughly two to three further doublings, on the order of 10,000 to 20,000
cells, and it labels that as extrapolation rather than measurement. So do we.

**Cost grows with cells.** Fit time rose roughly linearly with cell count —
about 9.6 s at 360 cells and 107 s at 2,250. The next doubling is a designed
compute campaign, not something you run while you wait.

**The regime is narrow.** These numbers were measured at a fixed spatial
range, a fixed cell size, two per-cell effort levels, and one synthetic
data-generating process with known truth. They are not a power analysis for
any real taxon, grid, or question.

Specifically, you **cannot** conclude from this page that:

1. some particular number of cells is right for your study — the curve fixes
   range and cell size, and your field's range relative to your cell size is
   the quantity that actually drives the result;
2. 2,250 cells is sufficient — on the campaign's own target it is not;
3. 10,000 to 20,000 cells is sufficient — that is an extrapolation beyond
   every measured point, and it may be optimistic or pessimistic;
4. adding cells helps in *your* design — if adding cells thins your per-cell
   records, that is the arm that moved the other way;
5. anything at all about absolute abundance, occupancy, or detectability —
   the design is relative-intensity throughout;
6. the same ordering holds for a different taxon, likelihood, bias structure,
   or number of species — none of those were varied;
7. admissibility becomes likely if you simply add cells — at the largest
   design measured, `pd_rate` is 0.555, so roughly one fit in two still fails
   the positive-definiteness check. Growing the domain moves that number in
   the right direction and does not finish the job.

What you *can* take away is the mechanism and the audit: find out whether your
own design would hold per-cell support fixed as it grows, because that
condition is what separates the two arms.

# What this route does and does not cover

**Covered here:** a design audit you can run on your own two-source data
before fitting anything, and a worked assembly-and-fit of the integrated
route — one presence-only count stream and one detection/non-detection survey
sharing an ecological linear predictor, with a portal-only bias term and known
per-row support, through `gllvmTMB()`.

**Works but not certified:** the design guidance itself. It rests on a
synthetic known-truth campaign in a single narrow regime. The direction of the
effect and its conditional form are what the evidence supports; the location
of the accuracy frontier is not established, and the estimates from the small
fit above are illustration, not validation.

**Not available:** a power calculator or a recommended cell count for a real
study; more than two sources; calibrated confidence intervals for this route;
weighting one source's contribution against the other's; and any measured
result outside the fixed-range, fixed-cell-size regime described above. Each
is real work, not an oversight.

# Reference

Fithian, W., Elith, J., Hastie, T. and Keith, D.A. (2015) Bias correction in
species distribution models: pooling survey and collection data for multiple
species. *Methods in Ecology and Evolution* **6**, 424–438.
