---
title: "Integrating opportunistic records with a repeated survey: a small worked example"
description: "Fit a presence-only portal stream and a repeated presence/absence survey in one likelihood through the ordinary gllvmTMB() call, and read what the shared spatial gradient and the portal-only bias term do and do not tell you."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Integrating opportunistic records with a repeated survey: a small worked example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 5,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
library(gllvmTMB)
library(ggplot2)
set.seed(20260815)
```

> **Experimental.** The model here is fitted through the ordinary
> `gllvmTMB()` entry point — there is no separate integrated-model function —
> but the route is experimental: the interface may change without deprecation,
> and no released capability claim is made. The example exists so that early
> users can see the intended workflow end to end on simulated data.

# The question

Opportunistic portal records (GBIF-style) are abundant but arrive where
people go; a designed detection/non-detection survey is small but has known
effort. This example asks a deliberately modest question of a **simulated
three-species grassland songbird guild** on a prairie-inspired landscape:
after giving the portal records their own recording-bias term, do the two
sources agree on one shared habitat gradient?

Both sources see the same ecological state for species $s$ in cell $c$,

$$
\eta_{cs} = \alpha_s + x_c\,\beta_s + u_c\,\lambda_s,
$$

with $x_c$ a measured habitat covariate and $u_c$ an unmeasured shared
spatial gradient carried by a latent field with species loadings
$\lambda_s$. The portal branch adds what only it suffers -- an
accessibility bias with species-specific coefficients $\gamma_s$ and its own
spatial recording field -- while the survey branch is a plain
complementary-log-log detection/non-detection likelihood with known visit
support:

$$
N^{G}_{cs} \sim \text{Poisson}\!\left[a^{G}_c
  \exp\{\eta_{cs} + b_c\gamma_s + h_c\,\delta_s\}\right],
\qquad
Y^{S}_{csv} \sim \text{Bernoulli}\!\left[1 -
  \exp\{-a^{S}_{cv}\exp(\eta_{cs})\}\right].
$$

Everything is **relative intensity**: no absolute abundance, occupancy, or
detectability claim is available from this design.

This is the intrinsic shape of the problem, not a stylistic choice: the
per-row family selector below needs several observation rows per (cell,
species) that carry *different* response families, which is exactly what the
long-format grammar represents and the wide `traits()` grammar cannot -- a
wide row is one (cell, species) cell with one family. There is no wide
version of this example to show alongside it.

# Simulate a small two-source data set

One hundred and eight cells, three species, one portal record stream, three
survey visits. The fields here are simple smooth surfaces -- enough to give
the model something to find without pretending to be a validation campaign.

```{r simulate}
n_lon <- 12; n_lat <- 9
grid <- expand.grid(lon = seq(0, 1.2, length.out = n_lon),
                    lat = seq(0, 0.9, length.out = n_lat))
grid <- grid[order(grid$lat, grid$lon), ]
n_cell <- nrow(grid)
cells   <- paste0("cell_", seq_len(n_cell))
species <- c("gralark", "dickcis", "bobolnk")   # simulated guild

## habitat covariate, shared gradient u, and an accessibility index
x <- as.numeric(scale(grid$lon))                        # measured habitat
u <- 0.8 * sin(pi * grid$lat / 0.9) * cos(pi * grid$lon / 1.2)  # unmeasured
b <- as.numeric(scale(-grid$lat + 0.3 * cos(2 * pi * grid$lon)))# access bias
## the DGP here has NO portal-only spatial field (delta = 0): the fitted
## model still estimates one, and treating it as a nuisance is the honest
## reading at this design size (see the closing cautions)

alpha  <- c(-0.3, 0.1, 0.4)
beta   <- c(0.6, -0.4, 0.5)
lambda <- c(0.9, 0.7, -0.6)
gamma  <- c(0.5, -0.3, 0.2)

eta <- outer(u, lambda) +
  sweep(outer(x, beta), 2, alpha, "+")                  # n_cell x 3

## portal effort concentrates in accessible cells; survey support is designed
a_g <- 3 * exp(0.4 * b + rnorm(n_cell, sd = 0.2))
a_s <- rep(0.9, n_cell)

base <- expand.grid(cell_id = cells, trait = species,
                    stringsAsFactors = FALSE)
base <- base[order(match(base$cell_id, cells), match(base$trait, species)), ]
ci <- match(base$cell_id, cells)
si <- match(base$trait, species)

gbif <- transform(base, source = "gbif", survey_event_id = NA_character_,
  branch = "count", support = a_g[ci], lon = grid$lon[ci], lat = grid$lat[ci],
  value = rpois(nrow(base),
    a_g[ci] * exp(eta[cbind(ci, si)] + b[ci] * gamma[si])),
  visit = NA_integer_)

pa <- do.call(rbind, lapply(1:3, function(v) transform(base,
  source = "survey", survey_event_id = paste0("pa_v", v, "_", base$cell_id),
  branch = "pa", support = a_s[ci], lon = grid$lon[ci], lat = grid$lat[ci],
  value = rbinom(nrow(base), 1, -expm1(-a_s[ci] * exp(eta[cbind(ci, si)]))),
  visit = v)))

rows <- rbind(gbif, pa)
table(rows$source)
```

The shared gradient `u` is what both sources are supposed to see in common.
Mapping it over the grid before fitting anything gives a visual reference
for the "do the two sources agree?" question the rest of the article works
toward:

```{r u-map, fig.cap = "The simulated shared spatial gradient u across the 108-cell grid. Both the portal stream and the survey observe species intensities driven partly by this common field; neither source sees u directly."}
ggplot(grid, aes(lon, lat, fill = u)) +
  geom_raster() +
  scale_fill_gradient2(low = "#8c4d8f", mid = "white", high = "#0a617d") +
  coord_fixed() +
  labs(x = "Longitude (arbitrary units)", y = "Latitude (arbitrary units)",
       fill = "u") +
  theme_minimal()
```

# Assemble the long table

`gllvmTMB()` works on long data: one row per (cell, species, source). Three
columns do the work that makes this an *integrated* model rather than two
models glued together.

```{r assemble}
dat <- rows
dat$trait   <- factor(dat$trait)
dat$cell_id <- factor(dat$cell_id)

## 1. the source gate: 1 on portal rows, 0 on survey rows. Every portal-only
##    term is written as an interaction with this, so it is structurally
##    absent from the survey likelihood rather than merely small.
dat$isdm_gbif <- as.integer(dat$source == "gbif")

## 2. known support, on the log scale: portal effort and survey visit area.
##    This is the change-of-support term, not a parameter.
dat$log_support <- log(dat$support)

## 3. the covariates. `bias` is portal-only, so it is a structural zero on
##    survey rows -- a zero, never an NA, because a survey row genuinely has
##    no accessibility term rather than a missing one.
dat$env  <- x[match(as.character(dat$cell_id), cells)]
dat$bias <- ifelse(dat$isdm_gbif == 1L,
                   b[match(as.character(dat$cell_id), cells)], 0)

## the per-row family selector, and the family list it selects from
dat$isdm_family <- factor(ifelse(dat$source == "gbif", "gbif", "survey_pa"),
                          levels = c("gbif", "survey_pa"))
fam <- list(gbif = poisson(), survey_pa = binomial(link = "cloglog"))
attr(fam, "family_var") <- "isdm_family"
```

The family list is the heart of it. Ordinarily `gllvmTMB()` requires one
family per trait; here a single species is Poisson on its portal rows and
Bernoulli-cloglog on its survey rows. That mix is admitted for exactly this
contract, because cloglog is the link under which a detection/non-detection
observation is consistent with the same underlying intensity that generates
the counts -- and because Poisson and Bernoulli carry no dispersion parameter,
so nothing per-species becomes ambiguous.

**The contract is exact, not approximate.** Mixing families within a trait is
admitted only for `family = list(gbif = poisson(), survey_pa =
binomial(link = "cloglog"))` (Poisson-log paired with binomial-cloglog),
together with a `family_var` attribute naming the per-row selector column and
a matching factor column in the data. Anything short of that -- a missing or
misnamed `family_var` attribute, a `source` present without its matching
`isdm_family` levels, or only one of the two sources actually present in the
data -- falls back to the ordinary one-family-per-trait rule and errors
there, because from the fitter's point of view it no longer looks like an
integrated fit at all. Swapping the survey link for logit or probit is
refused outright: those links do not share the intensity that generates the
portal counts, so the two branches would no longer describe the same
$\eta_{cs}$.

# Fit the integrated model

The two sources enter one likelihood: shared fixed effects and shared latent
field, a portal-only bias coefficient and portal-only spatial field, and the
survey's cloglog detection branch -- selected per row by `isdm_family`. A mesh
over the cells carries the spatial terms.

```{r fit}
mesh <- make_mesh(dat[, c("lon", "lat")], c("lon", "lat"), cutoff = 0.085)
fit <- gllvmTMB(
  value ~ 0 + trait + trait:env + trait:isdm_gbif + trait:bias +
    offset(log_support) +
    indep(0 + trait | cell_id) +
    spatial_latent(1 + isdm_gbif | cell_id, d = 1),
  data   = dat,
  trait  = "trait",
  unit   = "cell_id",
  family = fam,
  mesh   = mesh,
  silent = TRUE
)
```

The mesh triangulates the 108 cells for both the shared and the portal-only
spatial fields:

```{r plot-mesh, fig.cap = "Triangular SPDE mesh built from the 108 grid cells. The mesh carries both the shared spatial field u and the portal-only recording field at its vertices; the observation-to-mesh projection maps each back to the cell rows fitted above."}
plot(mesh)
```

<!--
MAINTAINER DECISION, 2026-08-16 (Shinichi): "keep the WARN fit - it's honest."
This fit reports pd_hessian = WARN and that is DELIBERATE. Do not change the
model, the design size, or the seed to make it pass. The warning is the
teaching content: it is localised by measurement (the same data refitted
non-spatially returns PASS, so the spatial fields are the difficulty, not the
integration), and it is the hinge to the companion design article. Making it
pass would mean dropping the augmented spatial slope -- the exact route the
public admission opened -- or growing the design past what an article can
render. See docs/dev-log/decisions.md, 2026-08-16.
-->

# Did the fit converge, and can you trust its curvature?

```{r convergence}
fit$opt$convergence
```

`0` is the optimizer's own signal that it stopped at a point it judges
stationary -- not proof that the point is a good estimate, and not a
substitute for checking curvature. `check_gllvmTMB()` runs that second,
independent check: it reports whether the Hessian at the optimum is
positive-definite, which is what actually licenses Wald-type standard
errors.

```{r health}
health <- check_gllvmTMB(fit)
health[
  health$component %in% c("optimizer_convergence", "pd_hessian"),
  c("component", "status", "value")
]
```

`optimizer_convergence` reads `PASS`: the optimizer itself is content. But
`pd_hessian` reads `WARN` -- the Hessian at this optimum is *not*
positive-definite. That combination (stationary point, non-PD curvature) is
consistent with the first caution below: source-specific spatial structure
is only weakly identified at this design size, and a weakly identified
spatial covariance is exactly the kind of thing that leaves the Hessian
non-PD even while the optimizer is satisfied.

It is worth being precise about *which* part of the model is responsible,
because it is not the integration. Refitting these same data with the
spatial terms replaced by an ordinary non-spatial `latent()` block -- same
two sources, same shared predictor, same portal-bias coefficient -- returns
`pd_hessian = PASS`. The integrated likelihood is not the difficulty; the
*spatial fields* are, at 108 cells. That is the same conclusion the design
article reaches from a much larger set of fits, and it is why the companion
[How big does an integrated survey design need to be?](integrated-survey-design.html)
exists.

Read this as the honest state of a small illustrative fit, not as a bug --
but do not trust Wald-type standard errors from this particular fit, and do
not skip this check on your own data because a real analysis is unlikely to
be this forgiving. See
[Convergence and start values](convergence-start-values.html) for the fuller
triage (more starts, rescaling, a simpler covariance) when `pd_hessian`
warns on a fit you intend to use.

# What did the two sources agree on?

The portal-bias coefficients are the part of the model only the integration
can give you -- the survey anchors the shared state, so $\gamma$ is
identified rather than silently absorbed. At this deliberately small design
expect useful but noisy estimates:

```{r gamma}
## factor() ordered the species alphabetically, so the fitted coefficients come
## back in that order -- realign the simulated truth before comparing, and name
## the columns so the pairing is visible rather than assumed.
idx <- grep("bias", fit$X_fix_names)
gamma_ordered <- gamma[match(levels(dat$trait), species)]
tab <- rbind(truth = gamma_ordered, estimate = unname(fit$opt$par[idx]))
colnames(tab) <- levels(dat$trait)
round(tab, 2)
```

```{r gamma-plot, fig.width = 5.2, fig.height = 5.2, fig.cap = "True versus fitted portal accessibility-bias coefficients gamma, one point per species. The line marks exact recovery; distance from it is recovery error at this design size."}
gamma_df <- data.frame(
  species  = levels(dat$trait),
  truth    = gamma_ordered,
  estimate = unname(fit$opt$par[idx])
)
lims <- range(c(gamma_df$truth, gamma_df$estimate)) + c(-0.1, 0.1)
ggplot(gamma_df, aes(truth, estimate, label = species)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey55", linewidth = 0.5) +
  geom_point(size = 2.8, colour = "#0a617d") +
  geom_text(vjust = -1, size = 3.3) +
  coord_equal(xlim = lims, ylim = lims) +
  labs(x = "True gamma", y = "Fitted gamma") +
  theme_minimal()
```

All three species land close to the identity line at this size -- the
portal-only bias term is doing what it is meant to, separating recording
effort from the shared ecological signal rather than absorbing it into
$\beta$ or $\lambda$. `bobolnk` (the smallest true $|\gamma|$) is the noisiest
of the three in relative terms, which is the expected pattern: a
near-zero true effect is the hardest one to pin down precisely, not evidence
that the method singles that species out.

The shared-state loadings and the portal-only field's covariance are
reported on the link scale (both are sign-invariant as outer products).
Rather than the raw internal report fields, use the public accessor, which
attaches trait names and validates that the fit actually carries this term:

```{r sigma}
sigma_slope <- extract_Sigma(fit, level = "spde_slope")
round(sigma_slope$intercept, 3)
round(sigma_slope$slope, 3)
```

`sigma_slope$intercept` is the shared spatial field's cross-species
covariance: it is what carries the "one shared gradient" question, and its
sign pattern should echo the simulated loadings `lambda` above (`gralark`
and `dickcis` positively coupled, `bobolnk` moving the other way).
`sigma_slope$slope` is the portal-only field's covariance -- here it is
estimating structure that the data-generating process never put there
(`delta = 0` above), so its magnitude is a measure of how much unclaimed
spatial variance the portal branch is willing to soak up at this design
size, not a recovered effect to interpret species by species. Reading it
alongside the first caution below is deliberate: this is exactly the
quantity that caution says to treat as a nuisance adjustment.

# What to check before believing more than this

Three honest cautions, each with a package-level reason.

1. **Design size governs identifiability.** Source-specific spatial
   structure is only weakly identified on small grids: internal known-truth
   campaigns place reliable recovery of a portal-only *field* far beyond a
   design of this size, with identifiability bought by adding cells (at
   maintained per-cell records), not by piling records into the same cells.
   Treat the portal-only field here as a nuisance adjustment, not an
   estimate to interpret.
2. **Relative intensity only.** Nothing in this design identifies absolute
   abundance or detectability; comparisons are within-map, not across
   species in absolute terms.
3. **Experimental surface.** The route is public but experimental: the data
   contract and the terms above may change without deprecation. Check
   convergence and positive-definiteness on every fit, and do not build a
   pipeline on this interface yet.

## What this route does and does not cover

**Covered here:** two sources — one presence-only count stream and one
detection/non-detection survey — sharing one ecological linear predictor,
with a portal-only bias term and known per-row support, fitted through
`gllvmTMB()`.

**Works but not certified:** the estimates themselves. The route is
contract-tested and the model fits, but recovery accuracy at this or any
particular design size is not established, and the portal-only *spatial*
field in particular rests on development experience rather than a completed
recovery study. Read the numbers above as illustration, not validation.

**Not available:** more than two sources; calibrated confidence intervals for
this route; and weighting one source's contribution against the other's.
Each is real work, not an oversight.

## See also

- [Multivariate spatial models with an SPDE mesh](spatial-models.html) for
  `spatial_latent()` and mesh construction on their own, outside the
  two-source setting.
- [Choose a response family](response-families.html) for the ordinary
  one-family-per-trait rule that this article's per-row contract is an
  exception to.

# References

Fithian, W., Elith, J., Hastie, T. and Keith, D.A. (2015). Bias correction in
species distribution models: pooling survey and collection data for multiple
species. *Methods in Ecology and Evolution* **6**, 424–438.
<https://doi.org/10.1111/2041-210X.12242>

Fletcher, R.J. Jr, Hefley, T.J., Robertson, E.P., Zuckerberg, B., McCleery,
R.A. and Dorazio, R.M. (2019). A practical guide for combining data to model
species distributions. *Ecology* **100**, e02710.
<https://doi.org/10.1002/ecy.2710>

Isaac, N.J.B., Jarzyna, M.A., Keil, P., et al. (2020). Data integration for
large-scale models of species distributions. *Trends in Ecology & Evolution*
**35**, 56–67. <https://doi.org/10.1016/j.tree.2019.08.006>

Koshkina, V., Wang, Y., Gordon, A., Dorazio, R.M., White, M. and Stone, L.
(2017). Integrated species distribution models: combining presence-background
data and site-occupancy data with imperfect detection. *Methods in Ecology
and Evolution* **8**, 420–430. <https://doi.org/10.1111/2041-210X.12738>

Miller, D.A.W., Pacifici, K., Sanderlin, J.S. and Reich, B.J. (2019). The
recent past and promising future for data integration methods to estimate
species' distributions. *Methods in Ecology and Evolution* **10**, 22–37.
<https://doi.org/10.1111/2041-210X.13110>

```{r session, include=FALSE}
## kept minimal deliberately; this article is a workflow illustration only
```
