---
title: "What do repeated survey visits add to an integrated model?"
description: "Separate a poor visit from genuinely low ecological intensity: repeated survey visits with visit-varying conditions, fitted through the ordinary gllvmTMB() call, and the boundary between this route and a separate-detection occupancy model."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{What do repeated survey visits add to an integrated model?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
set.seed(20260817)
library(gllvmTMB)
```

> **Experimental.** The model here is fitted through the ordinary
> `gllvmTMB()` entry point with the `isdm_sources()` declaration — the same
> experimental route as its sibling articles: the interface may change
> without deprecation, and no released capability claim is made. The example
> exists so that early users can see what visit structure adds, end to end,
> on simulated data.

A designed survey earns its keep twice. The
[two-source article](integrated-two-source-example.html) showed the first
half: detection/non-detection with known effort anchors the recording levels
that presence-only records cannot anchor. This article shows the second half:
**repeated visits at the same cells, under conditions that vary between
visits, let the model tell a poor visit apart from genuinely low ecological
intensity.** A single zero is ambiguous — the species may not have been
there, or the visit may have been short, wet, or badly timed. Replication
under varying conditions is what resolves the ambiguity, and it is a property
of the survey *design*, not of any fitting trick.

Everything here stays inside the integrated route's standing boundary:
**relative ecological intensity** — no abundance, no absolute occurrence
probability, and (as the last section makes precise) no free-standing
detectability parameter either.

# How a visit condition enters the admitted model

Every survey visit in the integrated route is a complementary-log-log
thinning of the same shared intensity. For cell $c$, species $s$ and visit
$v$ with known support $a_{cv}$ and a measured visit condition $w_{cv}$
(observer-hours, weather, time of day):

$$
  Y_{csv} \sim \mathrm{Bernoulli}(p_{csv}), \qquad
  p_{csv} = 1 - \exp\!\left[-a_{cv}\,
    \exp(\eta_{cs} + w_{cv}\,\delta_s)\right],
$$

with $\eta_{cs}$ the same ecological predictor every source shares. The
visit-condition effects $\delta_s$ act **on the log-intensity scale**: a
bad visit scales down the effective intensity the visit gets to thin,
exactly as a smaller support $a_{cv}$ would. This is the
"abundance-induced detection" view of Royle & Nichols (2003) — detection
improves with intensity, and visit conditions shift the rate, not a separate
coin-flip.

Two things follow. First, no new interface is needed: $w_{cv}$ is an
ordinary covariate column, set to its measured value on survey rows and to
zero on portal rows, so `trait:cond` terms are survey-only by construction —
the same source-gating device the two-source article uses for the portal's
bias covariate. Second, this is **not** the classical occupancy factorisation
$\Pr(Y=1) = p \cdot q$ with a separate logit detection process (MacKenzie
*et al.* 2002). That model is deliberately outside the admitted set: a logit
detection arm breaks the thinned-Poisson coherence that lets all sources
share one intensity, which is why `isdm_sources()` refuses logit detection
declarations at the door. The last section returns to what that boundary
costs and buys.

# Simulate: a portal stream plus four visits under varying conditions

Six wetland species over 250 cells. The portal stream is the familiar
biased count channel. The survey visits each cell four times; each visit has
its own condition $w_{cv}$, simulated independently of the wetland gradient —
the same design assumption the GBIF-only article flags for its bias
covariate, and it binds here identically.

```{r simulate}
n_cell  <- 250
n_visit <- 4
cells   <- paste0("cell_", seq_len(n_cell))
species <- c("sedge_wren", "rail", "bittern",
             "sundew", "sphagnum", "darner")

x <- as.numeric(scale(runif(n_cell)))    # wetland gradient (ecology)
b <- as.numeric(scale(runif(n_cell)))    # accessibility (portal bias)
z <- rnorm(n_cell)                       # unmeasured shared gradient

alpha  <- c(-0.4,  0.1, -0.8,  0.3,  0.6, -0.2)
beta   <- c( 0.9,  0.5,  0.7,  1.1,  0.8, -0.3)   # ecology
lambda <- c( 0.8,  0.6,  0.7, -0.5, -0.6,  0.4)
gamma  <- c( 0.6,  0.1,  0.0,  0.4,  0.2,  0.9)   # portal bias
delta  <- c( 0.7, -0.5,  0.3,  0.6, -0.4,  0.5)   # visit conditions

a_g <- 2.0 * exp(0.3 * b)                # portal effort tracks access
a_s <- 0.8                               # per-visit survey support

base <- expand.grid(cell_id = cells, trait = species,
                    stringsAsFactors = FALSE)
ci <- match(base$cell_id, cells)
si <- match(base$trait, species)
eta <- alpha[si] + x[ci] * beta[si] + z[ci] * lambda[si]

portal <- transform(base, isdm_source = "gbif", support = a_g[ci],
  value  = rpois(nrow(base), a_g[ci] * exp(eta + b[ci] * gamma[si])),
  env    = x[ci], access = b[ci], cond = 0, visit = NA_integer_)

visits <- do.call(rbind, lapply(seq_len(n_visit), function(v) {
  w <- rnorm(n_cell)                     # this visit's conditions, per cell
  transform(base, isdm_source = "survey", support = a_s,
    value  = rbinom(nrow(base), 1,
      -expm1(-a_s * exp(eta + w[ci] * delta[si]))),
    env    = x[ci], access = 0, cond = w[ci], visit = v)
}))

dat <- rbind(portal, visits)
dat$trait       <- factor(dat$trait, levels = species)
dat$cell_id     <- factor(dat$cell_id)
dat$log_support <- log(dat$support)
table(dat$isdm_source) / length(species)
```

Note the gating columns: `access` is the measured accessibility on portal
rows and zero on survey rows; `cond` is the measured visit condition on
survey rows and zero on portal rows. Each source keeps its own observation
covariates without any new syntax.

# One fit, three kinds of coefficient

```{r fit}
fit <- gllvmTMB(
  value ~ 0 + trait + trait:env + trait:access + trait:cond +
    offset(log_support) + latent(0 + trait | cell_id, d = 1),
  data   = dat,
  trait  = "trait",
  unit   = "cell_id",
  family = isdm_sources(gbif = poisson(), survey = binomial("cloglog")),
  silent = TRUE
)
fit$opt$convergence
```

```{r health}
health <- check_gllvmTMB(fit)
health[health$component %in% c("optimizer_convergence", "pd_hessian"),
       c("component", "status")]
```

Both read `PASS` on this fit: the optimizer settled, and the curvature at
that point is trustworthy (`pd_hessian` — whether the Hessian is
positive-definite; the [two-source article](integrated-two-source-example.html)
walks through what a `WARN` there means).

The formula now carries three interaction families with three different
meanings, and reading them apart is the whole skill of this model class:
`trait:env` is **ecology** (shared by every source), `trait:access` is the
**portal's recording process**, and `trait:cond` is the **survey's visit
process**. One fit, and each coefficient belongs to exactly one of the three.

```{r recovery-figure, fig.width = 8, fig.height = 3.6, fig.cap = "Truth against estimate for the three coefficient families — ecology (shared), portal bias (portal rows only), and visit conditions (survey rows only). One fit on one simulated draw; spread is sampling noise at this design size.", fig.alt = "Three side-by-side scatterplots, each with six labelled species points around a dashed 45-degree identity line on equal axes."}
op <- par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
panel <- function(truth, est, label) {
  lim <- range(c(truth, est)) + c(-0.2, 0.2)
  plot(truth, est, pch = 19, cex = 1.2, asp = 1, xlim = lim, ylim = lim,
       xlab = paste("true", label), ylab = "estimated", main = label)
  abline(0, 1, lty = 2)
  text(truth, est, labels = abbreviate(species, 6),
       pos = rep(c(3, 1), 3), cex = 0.65)
}
panel(beta,  unname(fit$opt$par[grep(":env",    fit$X_fix_names)]), "ecology")
panel(gamma, unname(fit$opt$par[grep(":access", fit$X_fix_names)]), "portal bias")
panel(delta, unname(fit$opt$par[grep(":cond",   fit$X_fix_names)]), "visit conditions")
par(op)
```

All three families recover along the identity line on this draw. The point
is not the precision of any single panel — it is that the three processes
were *separable at all*, and that separation came from the design: the
visit conditions varied between visits at the same cells, and varied
independently of the wetland gradient. The portal-bias panel leans on the
same requirement from the other side — `access` varies independently of
`env` in this simulation, and its recovery depends on that just as it did
in the [presence-only opener](gbif-joint-intensity.html).

# Replication is the ingredient, not the visit count

What actually breaks when replication is removed? Refit the same model
keeping only each cell's **first** visit — same number of survey cells, same
condition covariate, no within-cell replication:

```{r one-visit}
dat1 <- rbind(portal, subset(visits, visit == 1L))
dat1$trait       <- factor(dat1$trait, levels = species)
dat1$cell_id     <- factor(dat1$cell_id)
dat1$log_support <- log(dat1$support)

fit1 <- gllvmTMB(
  value ~ 0 + trait + trait:env + trait:access + trait:cond +
    offset(log_support) + latent(0 + trait | cell_id, d = 1),
  data   = dat1,
  trait  = "trait",
  unit   = "cell_id",
  family = isdm_sources(gbif = poisson(), survey = binomial("cloglog")),
  silent = TRUE
)
round(rbind(
  truth        = delta,
  four_visits  = unname(fit$opt$par[grep(":cond",  fit$X_fix_names)]),
  one_visit    = unname(fit1$opt$par[grep(":cond", fit1$X_fix_names)])
), 2)
```

With one visit per cell the model still runs — nothing refuses it — but on
this draw the visit-condition estimates degrade visibly: each cell now
contributes a single Bernoulli draw whose condition effect must be told
apart from that cell's latent ecological state using cross-cell contrast
alone. The structural argument says this is no accident: **repeated visits
at the same cells, with at least one condition that varies by visit rather
than only by cell,** are what buy the separation, and the visit count
matters only through the replication it provides. The comparison above is a
one-draw illustration of that argument, not a measured design curve — the
[design article](integrated-survey-design.html) shows what a measured curve
looks like for the cell-count question.

# The boundary: detection through intensity, not beside it

The admitted model ties detection to intensity by construction — a species
is seen more often where its intensity is higher, and visit conditions scale
that rate. What it does **not** provide is a free-standing detectability
parameter per species. If a species is simply hard to see — cryptic
colouration, calls that are easy to miss — that shortfall is folded into its
recording-level and intercept terms rather than estimated on its own. And
its ecological map stays accurate only if the shortfall is roughly constant
from cell to cell; if detectability itself varies with habitat, this model
cannot separate that from real ecological change.

If your scientific question is about detectability itself — "what fraction
of occupied cells does one visit find?" — you want the occupancy
factorisation $p \cdot q$ with its own detection submodel (MacKenzie *et
al.* 2002; Guillera-Arroita 2017), which is a different estimand and a
different likelihood, and is refused at the `isdm_sources()` door rather
than half-supported. If your question is where relative intensity is high
and how sources and visits distort your view of it, the route shown here
answers it inside one coherent likelihood.

# What this route does and does not cover

**Covered here:** repeated survey visits as additional cloglog rows sharing
one intensity; per-visit support offsets; visit-condition effects via an
ordinary gated interaction; recovery of ecology, portal-bias, and
visit-condition coefficients illustrated on one known-truth draw.

**Works but not certified:** the estimates themselves — one draw, one
design size, nonspatial, balanced visits.

**Not available:** a separate logit detection process; per-species
detectability at fixed intensity; abundance or absolute occupancy from any
number of visits; and calibrated confidence intervals for this route.

# See also

- [Integrating opportunistic records with a designed survey](integrated-two-source-example.html)
  — the two-source model this article extends with visit structure.
- [How big does an integrated survey design need to be?](integrated-survey-design.html)
- [More than two sources](integrated-multi-source-example.html)
- [Joint ecological intensity from opportunistic records](gbif-joint-intensity.html)
  — the presence-only opener, where the gating device is introduced.

# References

Guillera-Arroita, G. (2017). Modelling of species distributions, range
dynamics and communities under imperfect detection: advances, challenges and
opportunities. *Ecography* **40**, 281–295.
<https://doi.org/10.1111/ecog.02445>

Isaac, N.J.B., Jarzyna, M.A., Keil, P., et al. (2020). Data integration for
large-scale models of species distributions. *Trends in Ecology & Evolution*
**35**, 56–67. <https://doi.org/10.1016/j.tree.2019.08.006>

MacKenzie, D.I., Nichols, J.D., Lachman, G.B., Droege, S., Royle, J.A. and
Langtimm, C.A. (2002). Estimating site occupancy rates when detection
probabilities are less than one. *Ecology* **83**, 2248–2255.

Miller, D.A.W., Pacifici, K., Sanderlin, J.S. and Reich, B.J. (2019). The
recent past and promising future for data integration methods to estimate
species' distributions. *Methods in Ecology and Evolution* **10**, 22–37.
<https://doi.org/10.1111/2041-210X.13110>

Royle, J.A. and Nichols, J.D. (2003). Estimating abundance from repeated
presence–absence data or point counts. *Ecology* **84**, 777–790.
