---
title: "Integrating three data sources at once: portal, atlas, and survey"
description: "Declare several named observation sources with isdm_sources() — portal counts, digitised atlas records, a small detection survey — and fit one shared ecological process across all of them through the ordinary gllvmTMB() call."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Integrating three data sources at once: portal, atlas, and survey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4.5,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
library(gllvmTMB)
set.seed(20260816)
```

> **Experimental.** The multi-source route shown here is fitted through the
> ordinary `gllvmTMB()` entry point with sources declared by `isdm_sources()`.
> It is experimental: the interface may change without deprecation, and no
> released capability claim is made. If you have exactly two sources — one
> portal stream and one survey — start with the companion
> [two-source worked example](integrated-two-source-example.html) instead.

# The question

Real integration problems rarely stop at two sources. For a **simulated guild
of three boreal-wetland dragonflies**, suppose you hold three records of the
same landscape: an opportunistic portal stream (abundant, biased toward
accessible water), a digitised regional atlas (sparser counts, older effort,
its own recording habits), and a small designed detection/non-detection
survey. The question this article walks through: **can all three enter one
likelihood, each keeping its own observation law, while estimating one shared
ecological gradient?**

Every source sees the same ecological state for species $s$ in cell $c$,

$$
\eta_{cs} = \alpha_s + x_c\,\beta_s + u_c\,\lambda_s,
$$

and each source $d$ adds only its own recording level $\gamma_{ds}$ (with
source 1 as the reference, $\gamma_{1s} = 0$) and its own known support
$a_{cd}$:

$$
N^{(d)}_{cs} \sim \text{Poisson}\!\left[a_{cd}\,
  e^{\eta_{cs} + \gamma_{ds}}\right]
\quad\text{(count sources)},
\qquad
Y^{(d)}_{cs} \sim \text{Bernoulli}\!\left[1 -
  e^{-a_{cd}\, e^{\eta_{cs} + \gamma_{ds}}}\right]
\quad\text{(detection source)}.
$$

The complementary log-log link is what makes the detection source consistent
with the count sources: $1 - e^{-a e^{\eta}}$ is exactly the probability that
a Poisson draw with mean $a e^{\eta}$ is non-zero, so every source observes a
thinning of one shared intensity. That argument is per-source, which is why
it does not weaken as sources are added.

Everything here is **relative intensity**: no absolute abundance, occupancy,
or detectability claim is available from data like these, however many
sources you add.

# Simulate three sources over one landscape

One hundred and fifty cells, three species, three sources. The portal stream
is dense, the atlas is thin, the survey is small but designed.

```{r simulate}
n_cell  <- 150
cells   <- paste0("cell_", seq_len(n_cell))
species <- c("emerald", "whiteface", "darner")   # simulated guild
sources <- c("portal", "atlas", "survey")

x <- as.numeric(scale(runif(n_cell)))            # measured wetland gradient
u <- as.numeric(scale(sin(seq_len(n_cell) / 6))) # unmeasured shared gradient

alpha  <- c(-0.2,  0.2,  0.0)
beta   <- c( 0.5, -0.3,  0.4)
lambda <- c( 0.8,  0.5, -0.4)
## per-source recording effects, reference-coded on the portal
gamma  <- rbind(portal = c(0, 0, 0),   # reference source
                atlas  = c(-0.8, -0.5, -1.1),
                survey = c(0, 0, 0))   # zero by simulation choice, not a
                                       # second reference; its level rides
                                       # on its support
support <- c(portal = 2.0, atlas = 0.8, survey = 0.9)

dat <- expand.grid(cell_id = cells, trait = species, isdm_source = sources,
                   stringsAsFactors = FALSE)
ci <- match(dat$cell_id, cells)
si <- match(dat$trait, species)
di <- match(dat$isdm_source, sources)
eta <- alpha[si] + x[ci] * beta[si] + u[ci] * lambda[si] +
  gamma[cbind(di, si)]
dat$support <- support[di]
dat$value <- ifelse(
  dat$isdm_source == "survey",
  rbinom(nrow(dat), 1, -expm1(-dat$support * exp(eta))),
  rpois(nrow(dat), dat$support * exp(eta))
)
dat$trait       <- factor(dat$trait)
dat$cell_id     <- factor(dat$cell_id)
dat$log_support <- log(dat$support)
dat$env         <- x[ci]
dat$src         <- factor(dat$isdm_source, levels = sources)
table(dat$isdm_source, dat$value > 0)
```

Each source shows a different mix of zero and non-zero rows — the atlas is
the sparsest, matching its lower simulated support and negative recording
effects.

# Declare the sources, then fit

The declaration is the whole interface: name every source and state its
observation law. Two laws are admitted — `poisson()` for a count stream and
`binomial("cloglog")` for detection/non-detection — because both observe a
thinning of the same shared intensity. A logit detection model, a
dispersion-carrying family, or an all-detection declaration is refused at
this line, with the reason.

```{r declare}
fam <- isdm_sources(
  portal = poisson(),
  atlas  = poisson(),
  survey = binomial(link = "cloglog")
)
```

The data need an `isdm_source` column naming each row's source (built above),
and every species must be observed by every declared source. The per-source
recording effects enter the formula as an ordinary interaction with the
source factor — because `portal` is the first level, it is the reference and
the fitted terms are `atlas` and `survey` effects *relative to the portal*:

```{r fit}
fit <- gllvmTMB(
  value ~ 0 + trait + trait:env + trait:src + offset(log_support) +
    latent(0 + trait | cell_id, d = 1),
  data   = dat,
  trait  = "trait",
  unit   = "cell_id",
  family = fam,
  silent = TRUE
)
fit$opt$convergence
```

`0` means the optimizer reached a point it judges stationary. That is not the
same as trustworthy curvature, so run the package's own check rather than
stopping here:

```{r health}
health <- check_gllvmTMB(fit)
health[health$component %in% c("optimizer_convergence", "pd_hessian"),
       c("component", "status")]
```

Both pass on this fit (`pd_hessian` checks whether the curvature at the
optimum is trustworthy enough to license standard errors). On the small
spatial design of the two-source article the same check warns — read that
article's discussion of what a `WARN` there means and why it is shown rather
than hidden.

# What the integration bought

The per-source recording effects are the quantities only integration gives
you: with the portal as reference, the fitted `trait:src` terms say how much
each species' atlas and survey records depart from its portal recording
level, on the log scale, after the shared gradient is accounted for.

```{r gamma-figure, fig.cap = "Per-source recording effects, truth against estimate, for the atlas source (the survey's simulated departures are zero by construction). Estimates come from one fit on one simulated draw; the spread you see is sampling noise at this design size, not bias.", fig.alt = "Scatterplot of three labelled species points near a dashed 45-degree identity line, axes on equal scales."}
idx <- grep(":srcatlas", fit$X_fix_names)
est <- unname(fit$opt$par[idx])
truth <- gamma["atlas", match(levels(dat$trait), species)]
plot(truth, est, pch = 19, cex = 1.3,
     xlim = range(c(truth, est)) + c(-0.15, 0.15),
     ylim = range(c(truth, est)) + c(-0.15, 0.15),
     xlab = "true atlas recording effect (log scale)",
     ylab = "estimated",
     main = "")
abline(0, 1, lty = 2)
text(truth, est, labels = levels(dat$trait), pos = 3, cex = 0.85)
```

The estimates track the simulated truths along the identity line. In a real
analysis you would read these as each source's species-specific recording
tendency — which species a source over- or under-records relative to the
portal — and, per the relative-intensity boundary, nothing more absolute
than that.

# How large does a design like this need to be?

Two pieces of measured guidance from this package's internal campaigns, both
with their regimes stated:

1. **Adding sources neither helps nor hurts the sources you already have.**
   In a 1,200-fit known-truth study on exactly this model shape (150 cells,
   3 species, non-spatial, 2–4 fully-crossed sources), each source type's own
   recovery error was flat as sources were added. Integration's cost is not
   interference between arms.
2. **A weak source pays its own price.** Dropping a source's effort to a
   tenth of the reference roughly doubled the error of *that source's*
   recording effects, and left the others essentially untouched.

Both were measured with every source observing every cell — balanced,
fully-crossed designs. Sources with partial, patchy coverage are the harder
and unmeasured case, and the spatial version of this model has its own,
stricter design requirements — see
[How big does an integrated survey design need to be?](integrated-survey-design.html).

# What this route does and does not cover

**Covered here:** any number of named sources declared with `isdm_sources()`,
each a Poisson count stream or a cloglog detection stream, sharing one
ecological process through the ordinary `gllvmTMB()` call, with per-source
recording effects via an ordinary source-factor interaction.

**Works but not certified:** the estimates themselves. The route is
contract-tested and the recovery evidence above is real but nonspatial, at
one design size, on balanced sources. Read fitted numbers as illustration
until your own design has been checked against the design article's
guidance.

**Not available:** an all-detection declaration (refused when you build it —
the detection arm's offset is only admitted alongside a count arm);
per-source bias *covariates* beyond the recording level (expressible by hand
as source-gated interaction columns, as the two-source article shows, but
not yet packaged); weighting one source against another; calibrated
confidence intervals for this route; and anything absolute — abundance,
occupancy, detectability — from any number of presence-only sources.

# See also

- [Integrating opportunistic records with a designed survey](integrated-two-source-example.html)
  — the two-source worked example, including the spatial variant and an
  honest treatment of a non-positive-definite fit.
- [How big does an integrated survey design need to be?](integrated-survey-design.html)
  — the design question, from a 1,600-fit known-truth campaign.
- [Response families](response-families.html) — the admitted families and
  links in general.

# References

Fletcher, R.J. Jr, Hefley, T.J., Robertson, E.P., Zuckerberg, B., McCleery,
R.A. and Dorazio, R.M. (2019). A practical guide for combining data to model
species distributions. *Ecology* **100**, e02710.
<https://doi.org/10.1002/ecy.2710>

Isaac, N.J.B., Jarzyna, M.A., Keil, P., et al. (2020). Data integration for
large-scale models of species distributions. *Trends in Ecology & Evolution*
**35**, 56–67. <https://doi.org/10.1016/j.tree.2019.08.006>

Miller, D.A.W., Pacifici, K., Sanderlin, J.S. and Reich, B.J. (2019). The
recent past and promising future for data integration methods to estimate
species' distributions. *Methods in Ecology and Evolution* **10**, 22–37.
<https://doi.org/10.1111/2041-210X.13110>

```{r session, include=FALSE}
## deliberately minimal; this article is a workflow illustration only
```
