---
title: "When one data source knows where it is and the other does not"
description: "Integrated SDMs fuse a precise presence-only arm with a survey arm whose coordinates are deliberately fuzzed for privacy. That mismatch is covariate measurement error in disguise: it attenuates the environmental slope, and the penalty survives across arm-size designs and a biased opportunistic arm. It is a cost to measure, not a reason to discard a structured arm — once your sampling-bias surrogate carries realistic error, the sign reverses and integration helps."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{When one data source knows where it is and the other does not}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# The situation

An integrated species distribution model fuses data sources with different
sampling protocols: a large, messy, presence-only arm (GBIF, eBird) and a
structured survey arm (a monitoring programme). The whole design rests on the
two arms measuring the same underlying process, so that the structured arm
can discipline the biased one.

They usually differ in one respect nobody puts in the model. **The
presence-only arm knows where it is; the survey arm often does not.**

This is not sloppiness — it is policy. Monitoring programmes routinely
publish coordinates degraded on purpose, to protect site access agreements
and to avoid disclosing the locations of species at risk. The Alberta
Biodiversity Monitoring Institute publishes public survey locations
**fuzzed to roughly 5.5 km**, and coarser still for sensitive species;
precise coordinates require a separate data agreement. Meanwhile an eBird
checklist in the same landscape carries a GPS fix.

So the two arms are fused as though both located their observations equally
well, and one of them did not.

**How large is each arm, in practice?** A presence-only arm is usually much
larger than a structured survey — that is the whole reason to bother fusing
them. This article does not assume one ratio and call it "large" versus
"smaller": it tests **three explicit designs**, precise-arm size against
fuzzed-arm size — 220 / 220 (equal), 400 / 100 (presence-only four times the
survey, the common real-world shape) and 100 / 400 (the survey the larger
arm, a stress test in the opposite direction) — because, as the next section
shows, the arm-size ratio itself turns out to matter.

# Why this is easy to wave away

The intuition that makes this feel harmless goes: *the model has a spatial
random field; positional error is spatial noise; the field will absorb it.*

That intuition is about the **response**. The damage is to the
**covariate**.

Environmental predictors in an SDM are almost never measured at the site.
They are extracted from a raster at the recorded coordinate. If the recorded
coordinate is wrong by 5.5 km, you read temperature, elevation, or canopy
cover **from the wrong place** — and you then regress the species' response,
which happened at the *true* location, on the covariate value from a
different one.

That is textbook errors-in-variables. Its signature is **attenuation**: the
estimated slope is biased toward zero. And it lands on exactly the quantity
an SDM exists to produce — the environmental response curve.

How bad it is depends on one ratio: the positional error relative to the
**correlation length of the environment**, not its size in kilometres. Fuzz
of 5.5 km is nearly harmless across a smooth continental temperature
gradient and severe across a patchy canopy-cover layer. What "correlation
length" means, precisely, and how to compute your own, is given later in
this article, in [Get the ratio right first] — that number is what turns
this whole article from a warning into a decision you can make on your own
data.

# Measuring it

The claim above is checkable, so it should be checked rather than asserted.

The simulation is in `dev/isdm-precision/evidence-arm-weighting.R` and
`dev/isdm-precision/evidence-biased-po.R`; the replicate-level results
plotted below come from
`dev/isdm-precision/generate-arm-weighting-results.R` and
`dev/isdm-precision/generate-biased-po-results.R`, which call the same
simulation and estimator and additionally save every replicate rather than
only the mean.

Two species share a landscape with a single smooth exponential-correlation
environmental surface. Counts are generated from the environment at each
site's **true** location, with a shared slope $\beta = 0.9$. The
presence-only arm records its coordinates exactly; the survey arm's recorded
coordinates are displaced by Gaussian noise with standard deviation `fuzz` ×
the environmental correlation length. The covariate handed to the model is
read at each arm's *recorded* location, which is what an analyst would do.

Three analyses of the identical world, at each of three arm-size designs and
three fuzz levels (0, 0.5, 1.0 × correlation length), **15 replicates per
cell**:

- **`precise`** — fit the presence-only arm alone.
- **`fuzzed`** — fit the survey arm alone.
- **`integrated`** — fit both together, the thing an iSDM actually does.

# What happens

**The headline number is a dose-response, not a single comparison.** Hold
fuzz fixed at 0.5 × correlation length and grow the fuzzed arm's **share** of
the data — 100 rows against 400, then 220 against 220, then 400 against 100 —
and the integration penalty deepens in step:
`integrated − precise` goes **−0.105 → −0.279 → −0.438**. A pure averaging
artefact cannot produce that; it is exactly what attenuation propagating in
proportion to the contaminated arm's share of the data predicts.

```{r arm-weighting-figure, echo = FALSE, fig.width = 9.5, fig.height = 4.0, dpi = 200, out.width = "100%", fig.cap = "The attenuation penalty across three arm-size designs, each n(precise) / n(fuzzed), ordered left to right by the FUZZED arm's share of the data: 400/100 (fuzzed arm one fifth), 220/220 (equal), 100/400 (fuzzed arm four fifths, a stress test). Pale symbols are all 135 replicate fits per design (15 per fuzz level per arm); solid symbols and lines are the design's arm means. The dashed grey line, labelled in the first panel, marks the true slope 0.9. In every design the precise arm sits at truth and the fuzzed arm collapses toward zero; the integrated fit falls between them at a depth that depends on how much of the total data the fuzzed arm supplies.", fig.alt = "Three side-by-side panels, one per arm-size design, each plotting estimated slope against positional fuzz (0, 0.5 and 1.0 times the environmental correlation length) for three analyses: precise, fuzzed and integrated. A dashed grey horizontal reference line, labelled truth, marks the true slope 0.9 in every panel. In all three panels the precise series stays flat near 0.9 and the fuzzed series falls steeply toward 0.2 by fuzz 1.0. The integrated series falls least in the 400-precise/100-fuzzed panel (to about 0.74 at fuzz 1.0), more in the 220/220 panel (to about 0.52), and most in the 100-precise/400-fuzzed panel (to about 0.34): the more of the total data the fuzzed arm supplies, the deeper the integrated series falls."}
f_arm <- "../../dev/isdm-precision/arm-weighting-results.rds"
d_arm <- if (file.exists(f_arm)) readRDS(f_arm) else NULL

arms <- c("precise", "fuzzed", "integrated")
col  <- c(precise = "#0072B2", fuzzed = "#D55E00", integrated = "#009E73")
pchv <- c(precise = 16, fuzzed = 17, integrated = 15)
ltyv <- c(precise = 1, fuzzed = 3, integrated = 2)
xlb  <- "positional fuzz  (× correlation length)"
fz   <- c(0, 0.5, 1.0)
off  <- c(precise = -0.025, fuzzed = 0, integrated = 0.025)
## Panel order = increasing share of the data supplied by the FUZZED arm,
## matching the tables, the prose, and the direction of the claim. Ordered
## any other way the integrated series goes down-up-down and the reader has
## to reassemble the dose-response by hand.
designs <- list(c(400, 100), c(220, 220), c(100, 400))
dtitle <- c("n(precise)=400, n(fuzzed)=100", "n(precise)=220, n(fuzzed)=220",
            "n(precise)=100, n(fuzzed)=400")

## Fallback means, from READER-BLOCKS.md Block 7, so the article always
## builds even if the replicate-level file is absent (dispersion then
## unavailable).
mu_fallback <- list(
  `220_220` = rbind(precise = c(0.898, 0.898, 0.898),
                     fuzzed  = c(0.900, 0.365, 0.205),
                     integrated = c(0.899, 0.619, 0.522)),
  `400_100` = rbind(precise = c(0.893, 0.893, 0.893),
                     fuzzed  = c(0.966, 0.386, 0.190),
                     integrated = c(0.903, 0.788, 0.738)),
  `100_400` = rbind(precise = c(0.912, 0.912, 0.912),
                     fuzzed  = c(0.904, 0.366, 0.208),
                     integrated = c(0.906, 0.474, 0.344))
)

ylim_arm <- if (!is.null(d_arm)) range(c(0, 1.05, d_arm$beta_hat)) else c(0, 1.05)

op <- par(mfrow = c(1, 3), mar = c(3.4, 3.6, 2.2, 0.6),
          oma = c(2.0, 0, 1.6, 0),
          mgp = c(2.1, 0.6, 0), las = 1, cex.axis = 0.78, cex.lab = 0.82,
          tcl = -0.3)
set.seed(1)
for (k in seq_along(designs)) {
  key <- paste(designs[[k]], collapse = "_")
  ## ylim must CONTAIN the replicates, because the caption says it shows all
  ## of them. A fixed limit silently clips whatever a re-run pushes outside.
  plot(NA, xlim = c(-0.06, 1.06), ylim = ylim_arm, xaxt = "n",
       xlab = xlb, ylab = expression("estimated slope  " * hat(beta)))
  axis(1, at = fz, labels = format(fz))
  abline(h = 0.9, col = "grey45", lwd = 1.1, lty = 2)
  if (k == 1) text(1.045, 0.9, "truth", col = "grey35", cex = 0.62,
                   adj = c(1, -0.45))
  if (!is.null(d_arm)) {
    sub <- d_arm[d_arm$n_po == designs[[k]][1] & d_arm$n_survey == designs[[k]][2], ]
    mu <- tapply(sub$beta_hat, list(sub$arm, sub$fuzz), mean)[arms, ]
  } else {
    mu <- mu_fallback[[key]]
  }
  for (a in arms) {
    if (!is.null(d_arm)) {
      s <- sub[sub$arm == a, ]
      points(s$fuzz + off[a] + runif(nrow(s), -0.010, 0.010), s$beta_hat,
             pch = pchv[a], cex = 0.42, col = adjustcolor(col[a], 0.30))
    }
    lines(fz + off[a], mu[a, ], col = col[a], lwd = 2, lty = ltyv[a])
    points(fz + off[a], mu[a, ], pch = pchv[a], cex = 1.0, col = col[a])
  }
  mtext(dtitle[k], side = 3, line = 0.5, cex = 0.72, font = 2)
  if (k == 1) {
    ## seg.len: line type is the channel that carries this in greyscale, so
    ## the key must be long enough to tell dashed from dotted.
    legend("bottomleft", legend = arms, col = col[arms], pch = pchv[arms],
           lty = ltyv[arms], lwd = 1.6, bty = "n", cex = 0.72,
           seg.len = 3, text.col = col[arms])
  }
}
## The fence travels with the PNG, not with the caption: a figure lifted into
## a talk or a thesis loses its caption and keeps its pixels.
mtext("SIMULATED DATA — point estimates only; no interval or coverage claim",
      side = 1, outer = TRUE, line = 0.6, cex = 0.66, col = "grey35")
if (is.null(d_arm))
  mtext("FALLBACK MEANS — replicate file absent; re-run generate-arm-weighting-results.R",
        side = 3, outer = TRUE, line = 0.2, cex = 0.66, col = "#D55E00", font = 2)
par(op)
```

The same numbers, because a figure is read and a table is checked. Mean
estimated slope (true value **0.9**):

```{r arm-weighting-table, echo = FALSE}
if (!is.null(d_arm)) {
  agg <- aggregate(beta_hat ~ n_po + n_survey + fuzz + arm, d_arm, mean)
  tab <- reshape(agg, idvar = c("n_po", "n_survey", "fuzz"),
                 timevar = "arm", direction = "wide")
  names(tab) <- sub("beta_hat\\.", "", names(tab))
  tab <- tab[order(-tab$n_po, tab$fuzz), c("n_po", "n_survey", "fuzz",
                                            "precise", "fuzzed", "integrated")]
  names(tab) <- c("n(precise)", "n(fuzzed)", "fuzz", "precise", "fuzzed",
                   "integrated")
  knitr::kable(tab, digits = 3, row.names = FALSE)
}
```

Three things to read off it.

**The control works.** At `fuzz = 0` the *integrated* estimate sits within
0.011 of 0.9 in all three designs, and eight of the nine arm means are within
0.012. The exception is worth naming rather than rounding away: the fuzzed
arm at 400/100 comes in at **0.966**, 0.066 high — it is the visibly elevated
orange triangle in that panel, and it is the noisiest cell in the grid,
15 replicates of the smallest arm (100 rows) with no fuzz to attenuate it.
The point of the control survives: the bias appears only when positional
error does, not from the integration machinery.

**The precise arm is untouched.** In every design its estimate is fixed
across fuzz levels (bias under 0.02 throughout), because nothing about the
survey arm's positional error touches it. Adding fuzz to the *other* arm
cannot hurt it, which is what makes the integrated column so awkward.

**Attenuation is severe and it survives integration in all three designs.**
At displacement equal to the correlation length (`fuzz = 1.0`) the fuzzed
arm alone retains 21–23% of the true slope in every design. Integration
dilutes that damage but does not remove it, and how much it dilutes depends
entirely on how much of the pooled data the fuzzed arm supplies.

# The midpoint objection

A fair reader's first reaction to a 50/50 arm split is: *of course the
integrated estimate lands between the two single-arm estimates — that is
what an average does.* That was the previous version of this article, and
the objection is correct about that design.

```{r midpoint-table, echo = FALSE}
if (!is.null(d_arm)) {
  ## "Distance from the exact midpoint" is measured WITHIN each replicate
  ## (the midpoint of that replicate's own precise and fuzzed estimates),
  ## then averaged -- not the midpoint of the arm means. The two are not the
  ## same: averaging first can let opposite-signed within-replicate
  ## deviations cancel.
  rows <- list()
  for (des in list(c(220, 220), c(400, 100), c(100, 400))) {
    sub <- d_arm[d_arm$n_po == des[1] & d_arm$n_survey == des[2], ]
    for (fv in c(0.5, 1.0)) {
      s <- sub[sub$fuzz == fv, ]
      w <- reshape(s, idvar = "rep", timevar = "arm", direction = "wide")
      names(w) <- sub("beta_hat\\.", "", names(w))
      mid <- (w$precise + w$fuzzed) / 2
      rows[[length(rows) + 1]] <- data.frame(
        n_po = des[1], n_survey = des[2], fuzz = fv,
        integrated = mean(w$integrated),
        dist_from_midpoint = mean(abs(w$integrated - mid)))
    }
  }
  w <- do.call(rbind, rows)
  w <- w[order(-w$n_po, w$fuzz), ]
  names(w) <- c("n(precise)", "n(fuzzed)", "fuzz", "mean integrated",
                "mean |integrated - own-replicate midpoint|")
  knitr::kable(w, digits = 3, row.names = FALSE)
}
```

**She is right about the design.** At 220/220 the integrated estimate sits
within 0.023–0.041 of the exact midpoint of the two single-arm means — close
enough that the result reads as arithmetic, which is a fair thing to
distrust. The midpoint identity is an artefact **of the balanced design
specifically**: at 400/100 and 100/400 the gap from the midpoint is
0.149–0.215, five to seven times larger, and the identity is gone.

**She is wrong that the claim depends on it.** The number that matters is
not distance from the midpoint — it is whether integrating leaves you
further from the truth than dropping the fuzzed arm, and that holds in
**every** design, including the one most favourable to integration: a
fuzzed arm of only 100 rows against 400 precise rows still loses in **15 of
15** replicates at both non-zero fuzz levels. So does 220/220, and so does
100/400. The `fuzz = 0` control holds in all three (6, 9 and 8 of 15 below
zero — a coin flip, as a control must be).

# Does the arm need to be biased for this to matter?

Everything above shares one property with the article's earlier version: the
opportunistic arm is **unbiased**. It records its own true location and
nothing else is wrong with it. That is not why anyone integrates a
structured survey with an opportunistic one — they integrate because the
opportunistic arm is biased (accessibility, effort, detectability), and the
structured arm is what corrects that bias. A design without sampling bias
removes the entire reason to integrate, and a reader who took the argument
above at face value could wrongly conclude their own survey arm is not worth
keeping.

So the design is repeated with the opportunistic arm made **precise but
sampling-biased**: reporting intensity carries a coefficient of 1.2 on an
accessibility surface, and the accessibility surface is constructed to be
statistically independent of the environmental covariate (realised sample
correlation on this seed: −0.14, i.e. no relationship, in a strongly
autocorrelated field — chance, not confounding). The survey arm is
**unbiased but fuzzed**, as before. The analyst models the bias with an
`access` term, as a competent one would — the accessibility surface is
observable in real analyses (road density, trail networks) even though the
reporting *bias* it induces is not.

```{r biased-po-figure, echo = FALSE, fig.width = 5.2, fig.height = 4.0, dpi = 200, out.width = "70%", fig.cap = "The same penalty, with a precise but sampling-biased opportunistic arm (400 rows, reporting intensity biased by accessibility) against an unbiased but fuzzed survey arm (100 rows). Pale symbols are all 135 replicate fits (15 per fuzz level per arm, three arms, three fuzz levels); solid symbols and lines are means. The bias is modelled with an `access` term in every fit. The integrated estimate still falls below the precise-arm-alone estimate at both non-zero fuzz levels.", fig.alt = "One panel plotting estimated slope against positional fuzz (0, 0.5 and 1.0 times the correlation length) for three analyses: precise (biased but unfuzzed), fuzzed (unbiased but positionally fuzzed), and integrated. A dashed grey reference line, labelled truth, marks the true slope 0.9. The precise series stays flat near 0.90 at every fuzz level. The fuzzed series falls from about 0.92 at zero fuzz to about 0.21 at fuzz 1.0. The integrated series falls from about 0.90 to about 0.80, below the precise series at both non-zero fuzz levels but by a smaller margin than in the unbiased design."}
f_bpo <- "../../dev/isdm-precision/biased-po-results.rds"
d_bpo <- if (file.exists(f_bpo)) readRDS(f_bpo) else NULL
mu_bpo_fallback <- rbind(precise = c(0.903, 0.903, 0.903),
                         fuzzed  = c(0.915, 0.363, 0.206),
                         integrated = c(0.904, 0.835, 0.798))

par(mar = c(5.3, 3.6, 2.2, 0.6), mgp = c(2.1, 0.6, 0), las = 1,
    cex.axis = 0.80, cex.lab = 0.85, tcl = -0.3)
ylim_bpo <- if (!is.null(d_bpo)) range(c(0, 1.05, d_bpo$beta_hat)) else c(0, 1.05)
plot(NA, xlim = c(-0.06, 1.06), ylim = ylim_bpo, xaxt = "n",
     xlab = xlb, ylab = expression("estimated slope  " * hat(beta)))
axis(1, at = fz, labels = format(fz))
abline(h = 0.9, col = "grey45", lwd = 1.1, lty = 2)
text(1.045, 0.9, "truth", col = "grey35", cex = 0.66, adj = c(1, -0.45))
if (!is.null(d_bpo)) {
  mu_bpo <- tapply(d_bpo$beta_hat, list(d_bpo$arm, d_bpo$fuzz), mean)[arms, ]
} else {
  mu_bpo <- mu_bpo_fallback
}
set.seed(3)
for (a in arms) {
  if (!is.null(d_bpo)) {
    s <- d_bpo[d_bpo$arm == a, ]
    points(s$fuzz + off[a] + runif(nrow(s), -0.010, 0.010), s$beta_hat,
           pch = pchv[a], cex = 0.42, col = adjustcolor(col[a], 0.30))
  }
  lines(fz + off[a], mu_bpo[a, ], col = col[a], lwd = 2, lty = ltyv[a])
  points(fz + off[a], mu_bpo[a, ], pch = pchv[a], cex = 1.0, col = col[a])
}
legend("bottomleft", legend = arms, col = col[arms], pch = pchv[arms],
       lty = ltyv[arms], lwd = 1.6, bty = "n", cex = 0.72,
       text.col = col[arms])
mtext("400 precise-but-biased / 100 unbiased-but-fuzzed", side = 3,
      line = 0.5, cex = 0.75, font = 2)
## The fence travels with the PNG, not with the caption.
mtext("SIMULATED DATA — point estimates only; no interval or coverage claim",
      side = 1, line = 3.0, cex = 0.68, col = "grey35")
if (is.null(d_bpo))
  mtext("FALLBACK MEANS — re-run generate-biased-po-results.R",
        side = 1, line = 3.9, cex = 0.68, col = "#D55E00", font = 2)
```

```{r biased-po-table, echo = FALSE}
if (!is.null(d_bpo)) {
  agg3 <- aggregate(beta_hat ~ fuzz + arm, d_bpo, mean)
  w3 <- reshape(agg3, idvar = "fuzz", timevar = "arm", direction = "wide")
  names(w3) <- sub("beta_hat\\.", "", names(w3))
  w3$penalty <- w3$integrated - w3$precise
  w3$hurt <- sapply(w3$fuzz, function(fv) {
    sub <- reshape(d_bpo[d_bpo$fuzz == fv, ], idvar = "rep", timevar = "arm",
                   direction = "wide")
    sum((sub[["beta_hat.integrated"]] - sub[["beta_hat.precise"]]) < 0)
  })
  w3 <- w3[, c("fuzz", "precise", "fuzzed", "integrated", "penalty", "hurt")]
  names(w3) <- c("fuzz", "precise (biased)", "fuzzed (unbiased)", "integrated",
                 "integrated - precise", "replicates hurt (of 15)")
  knitr::kable(w3, digits = 3, row.names = FALSE)
}
```

**The claim survives.** With the motivation for integration actually present
— and correctly modelled — integrating the fuzzed arm still moves the
environmental slope away from truth in 14 of 15 replicates at both non-zero
fuzz levels. The `fuzz = 0` control holds (10 of 15 below zero; a coin
flip).

**Do not compare these magnitudes to the unbiased design above.** This run
fits an `access` term the earlier one does not, so the two are different
models; only the within-run contrast (integrated vs precise, same data) is
valid in each.

**The honest framing here is narrower than it looks, and it is a
trade-off, not a verdict.** This says the penalty falls on the
**environmental slope specifically**. It does not say the survey arm is
worthless — an unbiased structured arm is exactly what identifies and
corrects the reporting bias, and that correction is why it belongs in the
model at all. The uncomfortable finding is that the *same* arm, once its
coordinates are fuzzed, damages the slope while it repairs the bias. Both
are true at once. An article that reported only one of them would be
misleading, and this one is not resolving the trade-off for you — it is
telling you the trade-off exists so you can weigh it against your own data.

## Read this before you apply the result to your own data — the sign reverses

Everything above is conditional on something the simulation quietly grants
itself and **no real analysis has**: the accessibility surface is measured
*exactly*, and modelled. Given that, the presence-only arm's environmental
slope is already unbiased — so integrating a fuzzed arm can only cost, and
the whole article measures a cost with no benefit to weigh it against.

Relax that one assumption and the conclusion flips.
`dev/isdm-precision/evidence-bias-surrogate-error.R` changes one thing at a
time, 12 replicates at `fuzz = 1.0` — the worst case above:

| what the analyst has | precise | integrated | integration **helped** |
|---|---|---|---|
| `access` independent of `env`, measured exactly *(the design above)* | 0.918 | 0.806 | 3 / 12 |
| `access` confounded with `env` (ρ = 0.7), still measured exactly | 0.910 | 0.812 | **0 / 12** |
| confounded, surrogate carries measurement error (sd 0.5) | 1.192 | 1.083 | **11 / 12** |
| confounded, not modelled at all | 1.722 | 1.577 | **12 / 12** |

**The switch is not confounding — it is whether your bias surrogate is
exactly right.** Road density is not accessibility. Distance-to-trail is not
detectability. Once the surrogate carries realistic error, the presence-only
arm is no longer unbiased, and the structured arm — fuzzed coordinates and
all — pulls the estimate back toward truth in 11 or 12 replicates out of 12.

**And note what the last two rows also say: in that regime nothing recovers
the truth.** Precise 1.72, integrated 1.58, against a true 0.9. The *ranking*
flips; both answers are still bad. "Integration helped" there means "less
wrong", not "right".

So the decision this article supports is narrower than its headline:
**positional precision is a cost you should measure, not a reason to discard
a structured arm.** If your bias surrogate is imperfect — and it is — the
structured arm is likely earning its place despite the fuzz.

# The uncomfortable part

Compare the precise-alone and integrated columns in every design above.

At every non-zero fuzz level, in every arm-size design tested, **the
integrated estimate is further from the truth than the presence-only arm
alone**. At `fuzz = 0.5` in the 220/220 design, discarding the survey data
entirely gives 0.898 against a true 0.9; fusing it gives 0.619.

That is a statement about means, and it would be weaker if it were only
that. The replicate scatter in the figures above pairs the two analyses
**within each simulated dataset**, where they saw the same landscape and the
same truth. At `fuzz = 0` integrating helps about as often as it hurts in
every design — a coin flip, which is what a working control looks like. At
every non-zero fuzz it hurts in 15 of 15 replicates in all three arm-size
designs, and in 14 of 15 in the biased-opportunistic-arm design. The penalty
is not an artefact of averaging, and it is not an artefact of an unrealistic
unbiased opportunistic arm; it is what happens each time, across every
design tested here.

That inverts the premise of the method as far as the environmental slope is
concerned. Integrated SDMs are motivated by the structured arm correcting
the opportunistic one. Here the structured arm carries a defect the
opportunistic arm does not have, and integration propagates it — so more
data yields a worse point estimate of the environmental slope, because the
model has no way to know that one arm's covariates are mismeasured.

None of this is an argument against integrated SDMs. It is an argument that
**positional precision belongs in the list of things you check before
fusing two sources**, alongside detection probability, sampling effort, and
sampling bias — and it currently is not on that list.

**And it is a cost to measure, not a verdict to act on.** Every result above
holds the sampling-bias surrogate *exactly right*, which is the one condition
no real analysis meets. Give that surrogate realistic measurement error and
the sign reverses — integration then moves the slope **toward** truth in 11
of 12 replicates at `fuzz = 1.0`. See
[Read this before you apply the result to your own data — the sign reverses].
If you take one thing from this article, take the ratio in
[Get the ratio right first], not a decision to drop your survey arm.

# What to do about it

`gllvmTMB` does not implement a fix, and this article does not claim one.

## Get the ratio right first

The single most useful number in this whole article is a ratio: your
survey arm's positional uncertainty divided by the correlation length of
the covariate you care about. Getting that ratio wrong by even a factor of
three is enough to flip the recommendation, because different software
reports different things under the name "correlation length" or "range".

**The definition used throughout this article, with no convention
involved.** For a covariate with correlation `cor(d) = exp(-d / phi)`, `phi`
is the distance at which the covariate's own spatial correlation falls to
**1/e ≈ 0.368**. That is computable directly from a raster and does not
depend on what any particular package calls "range" — variogram software
commonly reports the *practical range*, conventionally about `3 * phi`, the
distance at which correlation falls to 0.05. Dividing a practical range by 3
before comparing it to positional fuzz is the single most important
arithmetic step in adapting this article to your own data.

```{r cor-length-recipe}
## phi = distance at which the covariate's own correlation falls to 1/e.
## NOTE the n_sub argument: this builds a dense n x n distance matrix, so it
## must be given a SUBSAMPLE of cells, never a whole raster (see below).
cor_length <- function(x, y, z, n_bin = 30, n_sub = 2000, seed = 1) {
  if (length(x) > n_sub) {
    ## Take the subsample reproducibly WITHOUT clobbering the caller's RNG
    ## stream. A bare set.seed() here would silently change every random
    ## draw the user makes afterwards -- the same species of invisible
    ## damage this article is about.
    old <- if (exists(".Random.seed", .GlobalEnv))
      get(".Random.seed", .GlobalEnv) else NULL
    set.seed(seed)
    i <- sample.int(length(x), n_sub)
    if (is.null(old)) rm(".Random.seed", envir = .GlobalEnv) else
      assign(".Random.seed", old, envir = .GlobalEnv)
    x <- x[i]; y <- y[i]; z <- z[i]
  }
  d   <- as.matrix(dist(cbind(x, y)))
  cp  <- outer(z - mean(z), z - mean(z)) / var(z)
  ut  <- upper.tri(d)
  brk <- seq(0, quantile(d[ut], 0.5), length.out = n_bin + 1)
  rho <- tapply(cp[ut], cut(d[ut], brk, include.lowest = TRUE), mean)
  h   <- (brk[-1] + brk[-length(brk)]) / 2
  ok  <- !is.na(rho)
  approx(rho[ok], h[ok], xout = exp(-1))$y
}
```

**Do not hand this a whole raster, and do not coarsen one to make it fit.**
Both halves of that matter:

- It is $O(n^2)$ in memory. 22,500 cells already costs about 25 s and 4 GB;
  a 1 km Alberta raster (~10⁶ cells) would need roughly **8 TB**. The
  `n_sub` argument above takes a random subsample instead, and a subsample
  is enough — 500 to 2,000 cells reproduce the full-raster answer to about
  **±0.007** on the surfaces tested here.
- **`terra::aggregate()` is the wrong fix, and it fails in the dangerous
  direction.** Coarsening the raster inflates `phi_hat` monotonically
  (0.128 → 0.176 at a factor of 6), which *shrinks* your fuzz-to-`phi` ratio
  and makes the situation look safer than it is — the opposite of the
  downward bias described below. Subsample cells; do not average them.

**Verified against known truth**, on simulated exponential-correlation
surfaces where `phi` is set by construction:

```{r cor-length-validate, echo = FALSE}
env_surface <- function(n_side, ell = 0.25, seed = 1) {
  set.seed(seed)
  gx <- seq(0, 1, length.out = n_side)
  g <- expand.grid(lon = gx, lat = gx)
  d <- as.matrix(dist(g)); S <- exp(-d / ell)
  z <- as.numeric(t(chol(S + diag(1e-6, nrow(S)))) %*% rnorm(nrow(g)))
  g$env <- as.numeric(scale(z)); g
}
val <- do.call(rbind, lapply(c(0.10, 0.25, 0.40), function(true_ell) {
  est <- sapply(1:6, function(s) {
    g <- env_surface(30, ell = true_ell, seed = s)
    cor_length(g$lon, g$lat, g$env)
  })
  data.frame(`true phi` = true_ell, `recipe returns` = mean(est),
             `sd over 6 surfaces` = sd(est), check.names = FALSE)
}))
knitr::kable(val, digits = 3, row.names = FALSE)
```

**And its limit, measured.** The bias depends only on `phi` relative to the
**domain side**, and is identical whether the domain is small or large — so
it is a finite-extent artefact, not a units artefact:

```{r cor-length-domain, echo = FALSE}
env_surface_side <- function(n_side, ell, seed, side = 1) {
  set.seed(seed)
  gx <- seq(0, side, length.out = n_side)
  g <- expand.grid(lon = gx, lat = gx)
  d <- as.matrix(dist(g)); S <- exp(-d / ell)
  z <- as.numeric(t(chol(S + diag(1e-6, nrow(S)))) %*% rnorm(nrow(g)))
  g$env <- as.numeric(scale(z)); g
}
dom <- do.call(rbind, lapply(c(0.05, 0.10, 0.25, 0.40), function(r) {
  est1 <- sapply(1:6, function(s) {
    g <- env_surface_side(30, ell = r * 1, seed = s, side = 1)
    cor_length(g$lon, g$lat, g$env)
  })
  data.frame(`phi / domain side` = r,
             `recipe / true phi` = mean(est1) / (r * 1), check.names = FALSE)
}))
knitr::kable(dom, digits = 2, row.names = FALSE)
```

Two things this means for you:

1. The recipe is trustworthy when `phi` is under about a tenth of your
   study region's width. Above that it **underestimates**, because a
   correlation longer than your map cannot be seen on your map.
2. The error is **always downward**, so a computed `fuzz / phi_hat` ratio is
   biased *upward* — the recipe errs toward "check this more carefully",
   which for this article's decision is the safe direction. But it also
   means a large returned `phi` is more trustworthy than a small one.

Also: a single raster gives `phi` to roughly ±20% (sd 0.041 at `phi` =
0.25 above). Do not read the third significant figure.

## Then act on it

1. **Compare the ratio, not a glance at the map.** Use the recipe above on
   your own raster, then compare your survey arm's positional uncertainty
   to it. Fuzz well below the correlation length is benign; by the time
   fuzz reaches half the correlation length, the tables above already show
   severe attenuation and a penalty in every design tested.
2. **Fit the arms separately first.** If the precise-arm-alone and
   integrated slopes disagree materially, that gap is evidence, not noise
   — provided you also know whether your opportunistic arm carries sampling
   bias (see above); in real data the gap can reflect positional error,
   sampling bias, detection differences, or a genuine scale difference all
   at once, and this article's clean simulations cannot tell you which.
3. **Aggregate to the fuzz scale.** If survey coordinates are fuzzed to
   5.5 km, covariates extracted at 5.5 km resolution are read at
   approximately the right place; a 30 m raster read at a fuzzed coordinate
   is not. Matching covariate resolution to positional precision is the
   cheapest available mitigation, and it costs the fine-scale signal.
4. **Ask for the precise coordinates.** Most fuzzing regimes, ABMI's
   included, have a data-agreement route to exact locations. For an
   analysis that turns on an environmental slope, that request is worth
   making before the analysis rather than after.

The general fix — an errors-in-variables layer with a per-arm positional
prior — is a real modelling option and is not implemented here. Treat this
article as a reason to measure your own case, not as a solution to it.

**A related, separate bug worth checking while you are at it.** Positional
error corrupts the covariate at *fit* time. A different, equally invisible
bug corrupts it at *prediction* time: centring and scaling a covariate
separately on the training data and again on a prediction grid — a change
of units the model was never told about. It is demonstrated, with measured
numbers, in the companion article on mapping (`isdm-canada-warbler`); the
short version is that the map's *pattern* survives untouched (within-species
rank correlation exactly 1.000) while predicted abundance moves by up to
5.5× in opposite directions for different species — invisible to every
check a reader would perform on the map itself. It is a different mechanism
from the one measured here, but the lesson is the same: a covariate silently
means something different at two points in the workflow, and nothing warns.

# Scope and limits

Stated plainly, because the tables above are more persuasive than the
evidence behind them warrants if these are omitted:

- **Simulation only, and that is a hard boundary, not a caveat.** No real
  ABMI or GBIF data was fitted anywhere in this article. The 5.5 km figure
  is documented ABMI policy and motivates the design; it is not a
  measurement of bias in any real dataset. A real analysis fusing GBIF and
  ABMI records needs a data agreement for the precise ABMI coordinates (see
  [What to do about it]) and is out of scope for this package's
  documentation — it is a data-access and ethics question, not a modelling
  one, and nothing here substitutes for doing it with real data under that
  agreement.
- **One error model.** Isotropic Gaussian displacement. Administrative
  fuzzing is often uniform-within-a-cell or grid-snapped, which will not
  behave identically.
- **One covariate structure.** A single smooth exponential-correlation
  surface. Patchier or anisotropic covariates will attenuate differently,
  and the fuzz-to-correlation-length ratio is the axis that matters.
- **Fixed effects only.** No latent field is fitted in the arm-weighting or
  biased-arm designs, so this isolates the covariate channel. Positional
  error also corrupts an estimated spatial field; that is not measured
  here.
- **One bias mechanism.** The biased-arm design uses one accessibility
  surface and one bias coefficient. Real reporting bias may be
  multi-causal, non-linear, or itself measured with error.
- **No coverage claim.** Point estimates only. Whether intervals around
  these attenuated slopes retain nominal coverage is a separate question
  and is not answered.
