---
title: "When one data source knows where it is and the other does not"
description: "Integrated SDMs fuse a precise presence-only arm with a survey arm whose coordinates are deliberately fuzzed for privacy. That mismatch is covariate measurement error in disguise: it attenuates the environmental slope by 57-79% in the fuzzed arm, and the pooled fit inherits that attenuation in proportion to the arm's share of the data - the swamping effect Fletcher et al. (2019) describe, here carrying a positional defect. Whether it matters is decided by one ratio, computed here from your own raster. And the conclusion reverses when your sampling-bias surrogate is BOTH imperfect and correlated with the environment - the conjunction, not either alone."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{When one data source knows where it is and the other does not}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

## The replicate-level result files are tracked in the repository and this
## article is built from them. If one is missing the build STOPS: a page that
## quietly substitutes stale hard-coded means for measurements is the exact
## failure mode this article is about.
f_arm  <- "../../dev/isdm-precision/arm-weighting-results.rds"
f_bpo  <- "../../dev/isdm-precision/biased-po-results.rds"
f_rev  <- "../../dev/isdm-precision/surrogate-error-results.rds"
f_grid <- "../../dev/isdm-precision/cor-length-grid.rds"
for (f in c(f_arm, f_bpo, f_rev, f_grid))
  if (!file.exists(f)) stop("missing evidence file: ", f,
    " -- re-run the matching dev/isdm-precision/generate-*.R script")
d_arm  <- readRDS(f_arm)
d_bpo  <- readRDS(f_bpo)
d_rev  <- readRDS(f_rev)
d_grid <- readRDS(f_grid)

## Every load-bearing number in the prose below is pulled from these objects
## by an inline expression, so no sentence can drift from the table above it.
wide <- function(s) {
  w <- reshape(s, idvar = "rep", timevar = "arm", direction = "wide")
  names(w) <- sub("beta_hat\\.", "", names(w)); w
}
cell <- function(np, ns, fz)
  wide(d_arm[d_arm$n_po == np & d_arm$n_survey == ns & d_arm$fuzz == fz, ])
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
```

# The situation

An integrated species distribution model fuses sources with different sampling
protocols: a large, messy, presence-only arm (GBIF, eBird) and a structured
survey arm (a monitoring programme). The design rests on both arms measuring
the same underlying process, so that the structured arm can discipline the
biased one.

They usually differ in one respect nobody puts in the model. **The
presence-only arm knows where it is; the survey arm often does not.**

This is not sloppiness — it is policy. Monitoring programmes routinely publish
coordinates degraded on purpose, to protect site access agreements and to avoid
disclosing the locations of species at risk. The Alberta Biodiversity
Monitoring Institute publishes public survey locations **fuzzed to roughly
5.5 km**, and coarser still for sensitive species; precise coordinates require
a separate data agreement. An eBird checklist in the same landscape carries a
GPS fix. The two arms are then fused as though both located their observations
equally well, and one of them did not.

That deliberate obscuration has analytical consequences is already documented
rather than hypothetical: work on iNaturalist's obscured-geoprivacy setting
reports that obscured locations change the environmental values attached to a
record — elevation, habitat type, extracted covariates — and change the
resulting estimated distributions. Gábor et al. (2022) likewise find that SDM
performance degrades as positional error in occurrence records grows.

The reason this still feels harmless in an *integrated* model is an intuition
about the wrong half of it: *there is a spatial random field, positional error
is spatial noise, the field will absorb it.* That is a statement about the
**response**. The damage is to the **covariate**.

Environmental predictors in an SDM are almost never measured at the site. They
are extracted from a raster at the recorded coordinate. If the recorded
coordinate is wrong by 5.5 km you read temperature, elevation or canopy cover
**from the wrong place**, and then regress the species' response — which
happened at the *true* location — on a covariate value from a different one.

That is classical errors-in-variables, and its signature is long established:
**attenuation**, the estimated slope biased toward zero. Nothing in the
mechanism is new here. What this article measures is what happens to it inside
a two-arm integrated fit, and what decides whether you should care.

Because that last part is not decided by the fuzz in kilometres. It is decided
by the fuzz relative to the **correlation length of the covariate**. Five and a
half kilometres is nearly nothing across a smooth continental temperature
gradient and severe across a patchy canopy-cover layer. So that ratio comes
first, before any simulation.

# The number that decides it

Your survey arm's positional uncertainty, divided by the correlation length of
the covariate you care about. Getting it wrong by a factor of three is enough
to flip the recommendation — and a factor of three is exactly what sits between
the two things software calls "range".

**The definition used here involves no convention.** For a covariate whose
spatial correlation behaves like `exp(-d / phi)`, `phi` is the distance at
which the covariate's own correlation falls to **1/e ≈ 0.368**. That is
computable directly from a raster. Variogram software commonly reports instead
the *practical range*, conventionally about `3 * phi`, the distance at which
correlation falls to 0.05. **Dividing a reported practical range by 3 before
comparing it to your positional fuzz is the single most important arithmetic
step in applying this article to your own data.**

## The recipe

```{r cor-length-recipe}
## phi = the distance at which the covariate's own spatial correlation falls
## to 1/e. NOTE n_sub: this builds a dense n x n distance matrix, so it takes
## a SUBSAMPLE of cells, never a whole raster.
cor_length <- function(x, y, z, n_bin = 30, n_sub = 2000, seed = 1) {
  if (length(x) > n_sub) {
    ## Subsample reproducibly WITHOUT clobbering the caller's RNG stream. A
    ## bare set.seed() here would silently change every random draw the user
    ## makes afterwards -- the same species of invisible damage this article
    ## is about.
    old <- if (exists(".Random.seed", .GlobalEnv))
      get(".Random.seed", .GlobalEnv) else NULL
    set.seed(seed)
    i <- sample.int(length(x), n_sub)
    if (is.null(old)) rm(".Random.seed", envir = .GlobalEnv) else
      assign(".Random.seed", old, envir = .GlobalEnv)
    x <- x[i]; y <- y[i]; z <- z[i]
  }
  d   <- as.matrix(dist(cbind(x, y)))
  cp  <- outer(z - mean(z), z - mean(z)) / var(z)
  ut  <- upper.tri(d)
  brk <- seq(0, quantile(d[ut], 0.5), length.out = n_bin + 1)
  rho <- tapply(cp[ut], cut(d[ut], brk, include.lowest = TRUE), mean)
  h   <- (brk[-1] + brk[-length(brk)]) / 2
  ok  <- !is.na(rho); rho <- rho[ok]; h <- h[ok]

  ## Walk outward to the FIRST crossing of 1/e and interpolate there.
  ## approx(rho, h, xout = 1/e) would sort on the correlations, and an
  ## empirical correlogram is not monotone -- that silently re-pairs
  ## distances with correlations they did not come from.
  below <- which(rho < exp(-1))
  if (!length(below))
    stop("cor_length(): the correlogram never falls to 1/e within half the ",
         "maximum inter-point distance. phi is too long to measure on this ",
         "extent -- your covariate is smoother than your map is wide. ",
         "Use a larger region, and do not substitute the largest distance ",
         "you can see.", call. = FALSE)
  k <- below[1]
  if (k == 1L)
    stop("cor_length(): correlation is already below 1/e in the first ",
         "distance bin, so phi is shorter than the spacing between the cells ",
         "supplied. Supply finer cells; do NOT read the returned bin ",
         "midpoint as phi.", call. = FALSE)
  h[k - 1] + (rho[k - 1] - exp(-1)) / (rho[k - 1] - rho[k]) * (h[k] - h[k - 1])
}
```

**Do not hand it a whole raster, and do not coarsen one to make it fit.** Both
halves matter. It is $O(n^2)$ in memory, and that arithmetic is unforgiving:
`n` cells cost `8 * n^2` bytes per matrix, so 22,500 cells is
`r fmt(8 * 22500^2 / 2^30, 1)` GB and a 1 km Alberta raster (~10⁶ cells) would
need `r fmt(8 * 1e6^2 / 2^40, 0)` TB. The `n_sub` argument takes a random
subsample instead, and a subsample is enough.
**`terra::aggregate()` is the wrong fix and it fails in the dangerous
direction**: coarsening inflates `phi_hat`, which *shrinks* your fuzz-to-`phi`
ratio and makes the situation look safer than it is. Subsample cells; do not
average them. (Coarsening is also not a remedy for the underlying problem — see
[What to do](#what-to-do).)

Working, on a surface whose `phi` is known by construction, and then refusing
to answer on one it cannot measure:

```{r cor-length-demo}
sim_surface <- function(n_side, phi, seed) {
  set.seed(seed)
  gx <- seq(0, 1, length.out = n_side)
  g  <- expand.grid(lon = gx, lat = gx)
  S  <- exp(-as.matrix(dist(g)) / phi)
  g$z <- as.numeric(scale(as.numeric(
    t(chol(S + diag(1e-6, nrow(S)))) %*% rnorm(nrow(g)))))
  g
}

g <- sim_surface(41, phi = 0.05, seed = 1)     # cell spacing 1/40 = 0.025
cor_length(g$lon, g$lat, g$z)                  # true phi = 0.05

g_coarse <- sim_surface(21, phi = 0.02, seed = 1)   # cell spacing 0.05 > phi
try(cor_length(g_coarse$lon, g_coarse$lat, g_coarse$z))
```

That second call is the important one. The correlogram never rises above 1/e,
so there is nothing to interpolate. An earlier version of this recipe returned
`NA` there without comment; it now says which of the two ways it failed,
because a silent `NA` in the one number the whole decision turns on is worse
than no recipe at all.

## What it can and cannot do

Two things can go wrong, and they are different: `phi` can be long relative to
your **map**, or short relative to your **cells**. An earlier experiment here
varied both at once — it scaled the domain and `phi` together on a fixed
lattice — so it could not tell them apart. This one crosses them:

```{r cor-length-grid, echo = FALSE}
## Drift guard: the grid was measured with the copy of the recipe stored in
## dev/isdm-precision/cor-length-recipe.R. If the function shipped above has
## since diverged from it, say so on the page rather than quietly reporting
## one function's accuracy as another's.
rec <- attr(d_grid, "recipe")
if (!is.null(rec) &&
    !identical(paste(deparse(cor_length), collapse = "\n"), rec))
  warning("The recipe shipped above differs from the one used to build this ",
          "table. Re-run dev/isdm-precision/generate-cor-length-grid.R.")

d_grid$ratio <- d_grid$phi_hat / d_grid$phi
gtab <- do.call(rbind, lapply(
  split(d_grid, list(d_grid$n_side, d_grid$phi), drop = TRUE), function(s)
    data.frame(
      `phi / map width`    = s$phi[1],
      `phi / cell spacing` = round(s$phi_over_spacing[1], 1),
      `refused (of 12)`    = sum(is.na(s$phi_hat)),
      `mean phi_hat / phi` = if (all(is.na(s$phi_hat))) NA else
                               mean(s$ratio, na.rm = TRUE),
      `sd over surfaces`   = if (sum(!is.na(s$phi_hat)) < 2) NA else
                               sd(s$ratio, na.rm = TRUE),
      `above truth`        = paste0(sum(s$ratio > 1, na.rm = TRUE), " / ",
                                    sum(!is.na(s$ratio))),
      check.names = FALSE)))
gtab <- gtab[order(gtab$`phi / map width`, gtab$`phi / cell spacing`), ]
knitr::kable(gtab, digits = 3, row.names = FALSE)
```

```{r cor-length-derived, include = FALSE}
## "The region where the recipe is recommended" = the operational advice
## given below: phi under a tenth of the map AND comfortably above the cell
## spacing. Both halves, or the summary silently includes cells the advice
## already tells the reader to stay out of.
benign <- d_grid[d_grid$phi <= 0.10 & d_grid$phi_over_spacing >= 2 &
                 !is.na(d_grid$phi_hat), ]
benign_above <- sum(benign$ratio > 1)
benign_n     <- nrow(benign)
sds <- tapply(benign$ratio, list(benign$n_side, benign$phi),
              function(z) sd(z, na.rm = TRUE))
sd_lo <- min(sds, na.rm = TRUE); sd_hi <- max(sds, na.rm = TRUE)
r040 <- mean(d_grid$ratio[d_grid$phi == 0.40], na.rm = TRUE)
r025 <- mean(d_grid$ratio[d_grid$phi == 0.25], na.rm = TRUE)
## The one cell that answers while barely resolving phi: phi/spacing = 1.2.
r_edge <- mean(d_grid$ratio[abs(d_grid$phi_over_spacing - 1.2) < 1e-8],
               na.rm = TRUE)
```

Read down the table and then across it.

**Down — the bias belongs to the map's width.** As `phi` grows relative to the
map, the recipe reads low: about `r fmt(r025, 2)` × truth at `phi` a quarter of
the map's width, `r fmt(r040, 2)` × at four tenths. Read *across* at a fixed
`phi / map width` and the three lattice resolutions give essentially the same
answer, which is what makes this a finite-extent effect rather than a
resolution one — a correlation longer than your map cannot be seen on your map.
The earlier design could not have shown that; this one does.

**Across — the resolution decides whether you get an answer at all.** Once
`phi` falls to about the spacing between your cells, the recipe refuses on
every surface or nearly every one. The row that answers at
`phi / cell spacing = 1.2` should be read as a warning rather than a success:
it returns a number, and the number is `r fmt(100 * (1 - r_edge), 0)`% low.

**And "the error is always downward" was wrong.** It is a statement about a
mean, applied by readers to a single raster. In the region where the recipe is
recommended — `phi` at or under a tenth of the map and at least twice the cell
spacing — the per-surface standard deviation of `phi_hat / phi` is
`r fmt(sd_lo, 2)`–`r fmt(sd_hi, 2)`, and `r benign_above` of `r benign_n`
individual surfaces (`r fmt(100 * benign_above / benign_n, 0)`%) return a value
**above** truth. So the safety argument that used to be built on "always
downward" does not hold for one raster run once. Use the recipe on several
covariate layers or several subsamples, and do not read the third significant
figure.

**So, operationally.** Trust it when `phi` sits comfortably below a tenth of
your study region's width and comfortably above your cell spacing. Outside that
window it either refuses or reads low, and reading low inflates your computed
`fuzz / phi` ratio — which for this article's decision errs toward looking more
carefully, but is not a licence to treat the number as exact.

# What actually happens

## The design

Two species share a landscape with a single smooth exponential-correlation
surface. Counts are generated from the environment at each site's **true**
location with a shared slope $\beta = 0.9$. The presence-only arm records its
coordinates exactly; the survey arm's recorded coordinates are displaced by
noise with standard deviation `fuzz` × the correlation length. The covariate
handed to the model is read at each arm's *recorded* location, which is what an
analyst would do. Three analyses of the identical world — `precise` (the
presence-only arm alone), `fuzzed` (the survey arm alone) and `integrated`
(both, the thing an iSDM does) — at three fuzz levels and three arm-size
designs, 15 replicates each. Scripts:
`dev/isdm-precision/generate-arm-weighting-results.R` and
`generate-biased-po-results.R`.

Three properties of the design a reader should know before reading the numbers:

- **No spatial field is fitted.** Fixed effects only. This isolates the
  covariate channel cleanly, and it means the article does **not** measure how
  much of the damage a fitted field would absorb — the intuition named at the
  top is set aside here, not tested.
- **The precise arm is generated before `fuzz` is used**, so its estimate is
  bit-identical across the three fuzz levels
  (`r fmt(mean(cell(100, 400, 0)$precise), 8)` at all three in the 100/400
  design). Its flatness in the figure is an identity, not a finding. It gives
  perfect pairing, which is the point.
- **The three designs share their 15 landscapes** (the same replicate seeds).
  They are three designs on one set of landscapes, not 45 independent trials.

## The measurement

```{r arm-weighting-figure, echo = FALSE, fig.width = 9.5, fig.height = 4.0, dpi = 200, out.width = "100%", fig.cap = "The attenuation penalty across three arm-size designs, each n(precise) / n(fuzzed), ordered left to right by the FUZZED arm's share of the data: 400/100 (fuzzed arm one fifth), 220/220 (equal), 100/400 (fuzzed arm four fifths, a stress test). Pale symbols are the 105 distinct fits behind each design - 15 precise (generated before fuzz is applied, so the same 15 are drawn at each fuzz level), 45 fuzzed and 45 integrated. Solid symbols and lines are the design's arm means. The dashed grey line, labelled in the first panel, marks the true slope 0.9. In every design the precise arm sits at truth and the fuzzed arm collapses toward zero; the integrated fit falls between them at a depth set by how much of the pooled data the fuzzed arm supplies.", fig.alt = "Three side-by-side panels, one per arm-size design, each plotting estimated slope against positional fuzz (0, 0.5 and 1.0 times the environmental correlation length) for three analyses: precise, fuzzed and integrated. A dashed grey horizontal reference line, labelled truth, marks the true slope 0.9 in every panel. In all three panels the precise series stays flat near 0.9 and the fuzzed series falls steeply toward 0.2 by fuzz 1.0. The integrated series falls least in the 400-precise/100-fuzzed panel (to about 0.74 at fuzz 1.0), more in the 220/220 panel (to about 0.52), and most in the 100-precise/400-fuzzed panel (to about 0.34): the more of the pooled data the fuzzed arm supplies, the deeper the integrated series falls."}
arms <- c("precise", "fuzzed", "integrated")
col  <- c(precise = "#0072B2", fuzzed = "#D55E00", integrated = "#009E73")
pchv <- c(precise = 16, fuzzed = 17, integrated = 15)
ltyv <- c(precise = 1, fuzzed = 3, integrated = 2)
xlb  <- "positional fuzz  (× correlation length)"
fz   <- c(0, 0.5, 1.0)
off  <- c(precise = -0.025, fuzzed = 0, integrated = 0.025)
## Panel order = increasing share of the data supplied by the FUZZED arm,
## matching the tables, the prose, and the direction of the claim. Ordered
## any other way the integrated series goes down-up-down and the reader has
## to reassemble the dose-response by hand.
designs <- list(c(400, 100), c(220, 220), c(100, 400))
dtitle <- c("n(precise)=400, n(fuzzed)=100", "n(precise)=220, n(fuzzed)=220",
            "n(precise)=100, n(fuzzed)=400")
## ylim must CONTAIN the replicates, because the caption says it shows all of
## them. A fixed limit silently clips whatever a re-run pushes outside.
ylim_arm <- range(c(0, 1.05, d_arm$beta_hat))

op <- par(mfrow = c(1, 3), mar = c(3.4, 3.6, 2.2, 0.6),
          oma = c(2.0, 0, 1.6, 0),
          mgp = c(2.1, 0.6, 0), las = 1, cex.axis = 0.78, cex.lab = 0.82,
          tcl = -0.3)
set.seed(1)
for (k in seq_along(designs)) {
  plot(NA, xlim = c(-0.06, 1.06), ylim = ylim_arm, xaxt = "n",
       xlab = xlb, ylab = expression("estimated slope  " * hat(beta)))
  axis(1, at = fz, labels = format(fz))
  abline(h = 0.9, col = "grey45", lwd = 1.1, lty = 2)
  if (k == 1) text(1.045, 0.9, "truth", col = "grey35", cex = 0.62,
                   adj = c(1, -0.45))
  sub <- d_arm[d_arm$n_po == designs[[k]][1] &
               d_arm$n_survey == designs[[k]][2], ]
  mu <- tapply(sub$beta_hat, list(sub$arm, sub$fuzz), mean)[arms, ]
  for (a in arms) {
    s <- sub[sub$arm == a, ]
    points(s$fuzz + off[a] + runif(nrow(s), -0.010, 0.010), s$beta_hat,
           pch = pchv[a], cex = 0.42, col = adjustcolor(col[a], 0.30))
    lines(fz + off[a], mu[a, ], col = col[a], lwd = 2, lty = ltyv[a])
    points(fz + off[a], mu[a, ], pch = pchv[a], cex = 1.0, col = col[a])
  }
  mtext(dtitle[k], side = 3, line = 0.5, cex = 0.72, font = 2)
  if (k == 1) {
    ## seg.len: line type is the channel that carries this in greyscale, so
    ## the key must be long enough to tell dashed from dotted.
    legend("bottomleft", legend = arms, col = col[arms], pch = pchv[arms],
           lty = ltyv[arms], lwd = 1.6, bty = "n", cex = 0.72,
           seg.len = 3, text.col = col[arms])
  }
}
## The fence travels with the PNG, not with the caption: a figure lifted into
## a talk or a thesis loses its caption and keeps its pixels.
mtext("SIMULATED DATA — point estimates only; no interval or coverage claim",
      side = 1, outer = TRUE, line = 0.6, cex = 0.66, col = "grey35")
par(op)
```

The same numbers, because a figure is read and a table is checked. Mean
estimated slope (true value **0.9**), with the arm-size-weighted average of the
two single-arm means alongside the integrated fit:

```{r arm-weighting-table, echo = FALSE}
agg <- aggregate(beta_hat ~ n_po + n_survey + fuzz + arm, d_arm, mean)
tab <- reshape(agg, idvar = c("n_po", "n_survey", "fuzz"),
               timevar = "arm", direction = "wide")
names(tab) <- sub("beta_hat\\.", "", names(tab))
tab$nwavg <- (tab$n_po * tab$precise + tab$n_survey * tab$fuzzed) /
             (tab$n_po + tab$n_survey)
tab <- tab[order(-tab$n_po, tab$fuzz),
           c("n_po", "n_survey", "fuzz", "precise", "fuzzed", "integrated",
             "nwavg")]
names(tab) <- c("n(precise)", "n(fuzzed)", "fuzz", "precise", "fuzzed",
                "integrated", "n-weighted avg")
knitr::kable(tab, digits = 3, row.names = FALSE)
```

```{r arm-weighting-derived, include = FALSE}
att <- agg[agg$arm == "fuzzed" & agg$fuzz > 0, ]
att_pct <- 100 * (1 - att$beta_hat / 0.9)
att_lo <- min(att_pct); att_hi <- max(att_pct)
att05 <- att_pct[att$fuzz == 0.5]; att10 <- att_pct[att$fuzz == 1]
nw_gap <- sapply(designs, function(de) sapply(c(0.5, 1.0), function(fv) {
  w  <- cell(de[1], de[2], fv)
  nw <- (de[1] * w$precise + de[2] * w$fuzzed) / sum(de)
  mean(abs(w$integrated - nw))
}))
pen <- sapply(designs, function(de) {
  w <- cell(de[1], de[2], 0.5); mean(w$integrated - w$precise) })
pen_ci <- sapply(designs, function(de) {
  w <- cell(de[1], de[2], 0.5); t.test(w$integrated - w$precise)$conf.int })
ctl <- d_arm[d_arm$n_po == 400 & d_arm$n_survey == 100 &
             d_arm$fuzz == 0 & d_arm$arm == "fuzzed", "beta_hat"]
ctl_t <- t.test(ctl, mu = 0.9)
## The zero-fuzz arm means, so "eight of nine within X" is measured, not
## remembered. The ninth is the cell tested above.
z0 <- agg[agg$fuzz == 0, ]
z0_dev <- sort(abs(z0$beta_hat - 0.9))
z0_ok  <- z0_dev[length(z0_dev) - 1]      # the largest of the other eight
sign_cnt <- sapply(designs, function(de) sapply(c(0.5, 1.0), function(fv) {
  w <- cell(de[1], de[2], fv); sum(w$integrated - w$precise < 0) }))
err_cnt <- sapply(designs, function(de) sapply(c(0.5, 1.0), function(fv) {
  w <- cell(de[1], de[2], fv)
  sum(abs(w$integrated - 0.9) > abs(w$precise - 0.9)) }))
```

**The fuzzed arm is gutted.** At half the correlation length it has already
lost `r fmt(min(att05), 0)`–`r fmt(max(att05), 0)`% of the true slope; at a
displacement equal to the correlation length,
`r fmt(min(att10), 0)`–`r fmt(max(att10), 0)`%. Across all six non-zero cells
the attenuation is **`r fmt(att_lo, 0)`–`r fmt(att_hi, 0)`%**. That is the
finding. Everything else follows from it.

**Pooling is an information-weighted average — and saying so is the strongest
form of the argument, not a concession.** A fair reader's first reaction is
that the integrated estimate is bound to land between the two single-arm
estimates, because that is what averaging does. Correct: the last column of the
table is the arm-size-weighted average of the two single-arm means, and it
tracks the integrated fit to within
`r fmt(min(nw_gap), 3)`–`r fmt(max(nw_gap), 3)` **within each replicate**, in
every design including the unbalanced ones. Pooling behaves exactly as pooling
should.

What is not arithmetic is **what is being averaged in**. The fuzzed arm arrives
carrying `r fmt(att_lo, 0)`–`r fmt(att_hi, 0)`% attenuation, and the pooled
estimate inherits it in proportion to that arm's share of the data. Hold fuzz
at 0.5 and grow the fuzzed arm's share — 100 rows against 400, then 220 against
220, then 400 against 100 — and `integrated − precise` deepens in step:
**`r fmt(pen[1])` → `r fmt(pen[2])` → `r fmt(pen[3])`**, with disjoint
confidence intervals
(`r sprintf("[%.3f, %.3f]", pen_ci[1,1], pen_ci[2,1])`,
`r sprintf("[%.3f, %.3f]", pen_ci[1,2], pen_ci[2,2])`,
`r sprintf("[%.3f, %.3f]", pen_ci[1,3], pen_ci[2,3])`).

**This is a known effect, and naming it is the point.** Fletcher et al. (2019)
describe exactly this in integrated SDMs: without explicit weighting, estimates
are largely driven by the abundant presence-only data, and the effect worsens
as that arm grows — the ISDM's parameters shrink toward the presence-only
model's. That is *swamping*, and it is established. This article adds one
narrower thing: **positional error is a defect that gets swamped in**, the
attenuation it causes is severe well before the fuzz reaches the covariate's
correlation length, and the ratio in the previous section is what tells you
whether your case is in that regime. If you already know about swamping, the
new information is what the swamped-in defect is and how to size it.

**The zero-fuzz control mostly works, and where it does not, the mechanism has
a name.** Eight of the nine arm means at `fuzz = 0` sit within
`r fmt(z0_ok)` of truth. The ninth is the fuzzed arm in the 400/100 design, at
`r fmt(mean(ctl))` — `r fmt(ctl_t$statistic, 2)` standard errors above 0.9,
*p* = `r fmt(ctl_t$p.value, 3)`. That is not noise, and calling it noise while
treating small *p*-values elsewhere as findings would be having it both ways.
It is the smallest arm in the grid (100 sites) at log-scale intercepts of −0.3
and 0.1, i.e. counts of order one — the regime where the Poisson
maximum-likelihood slope is upward-biased in small samples. The control's
purpose still holds, since the *large* bias appears only when positional error
does; but this cell is a known small-sample effect, not a fluke.

**Within each replicate, integration loses.** The two analyses saw the same
landscape and the same truth, so they can be paired. Counting by *sign*, the
integrated estimate is lower than the precise-arm estimate in
`r min(sign_cnt)` of 15 replicates in every design at both non-zero fuzz
levels. Counting by *distance from the truth* — the quantity that actually
matters, and the one the prose used to claim while the code counted signs — it
is worse in `r min(err_cnt)` of 15 in the 400/100 design at `fuzz = 0.5` and in
15 of 15 everywhere else. The two counts differ in one cell out of six, and the
smaller one is the one to quote.

## With a biased opportunistic arm

Everything above shares one property that removes the reason anyone integrates:
the opportunistic arm is **unbiased**. Nobody fuses a structured survey with an
opportunistic one for fun — they do it because the opportunistic arm is biased
by accessibility, effort and detectability, and the structured arm is what
corrects that. A reader who took the result above at face value could wrongly
conclude their survey arm is not worth keeping.

So the design is repeated with the opportunistic arm **precise but
sampling-biased**: reporting intensity carries a coefficient of 1.2 on an
accessibility surface drawn independently of the environmental covariate. The
survey arm is **unbiased but fuzzed**, as before. The analyst models the bias
with an `access` term, as a competent one would — accessibility is observable
in real analyses (road density, trail networks) even though the reporting bias
it induces is not.

```{r access-cor, echo = FALSE}
env_surface <- function(ns = 26, ell = 0.25, seed = 1) {
  set.seed(seed)
  gx <- seq(0, 1, length.out = ns); g <- expand.grid(lon = gx, lat = gx)
  S <- exp(-as.matrix(dist(g)) / ell)
  z <- as.numeric(t(chol(S + diag(1e-6, nrow(S)))) %*% rnorm(nrow(g)))
  as.numeric(scale(z))
}
acc_r <- sapply(1:15, function(s)
  cor(env_surface(seed = 3000 + s), env_surface(seed = 3000 + s + 500L)))
```

Independent by construction, but each replicate draws its own pair of surfaces,
so the *realised* correlation varies: across the 15 landscapes it runs from
`r fmt(min(acc_r), 2)` to `r fmt(max(acc_r), 2)` (mean `r fmt(mean(acc_r), 2)`,
sd `r fmt(sd(acc_r), 2)`). Quoting one seed's value would invite the reader to
think the design pins it.

```{r biased-po-figure, echo = FALSE, fig.width = 5.2, fig.height = 4.0, dpi = 200, out.width = "70%", fig.cap = "The same penalty with a precise but sampling-biased opportunistic arm (400 rows, reporting intensity biased by accessibility) against an unbiased but fuzzed survey arm (100 rows). Pale symbols are the 105 distinct fits (15 precise, generated before fuzz is applied and so drawn at each fuzz level; 45 fuzzed; 45 integrated); solid symbols and lines are means. The bias is modelled with an access term in every fit. The integrated estimate still falls below the precise-arm-alone estimate at both non-zero fuzz levels, by a smaller margin than in the unbiased design.", fig.alt = "One panel plotting estimated slope against positional fuzz (0, 0.5 and 1.0 times the correlation length) for three analyses: precise (biased but unfuzzed), fuzzed (unbiased but positionally fuzzed), and integrated. A dashed grey reference line, labelled truth, marks the true slope 0.9. The precise series stays flat near 0.90 at every fuzz level. The fuzzed series falls from about 0.92 at zero fuzz to about 0.21 at fuzz 1.0. The integrated series falls from about 0.90 to about 0.80, below the precise series at both non-zero fuzz levels."}
par(mar = c(5.3, 3.6, 2.2, 0.6), mgp = c(2.1, 0.6, 0), las = 1,
    cex.axis = 0.80, cex.lab = 0.85, tcl = -0.3)
plot(NA, xlim = c(-0.06, 1.06), ylim = range(c(0, 1.05, d_bpo$beta_hat)),
     xaxt = "n", xlab = xlb,
     ylab = expression("estimated slope  " * hat(beta)))
axis(1, at = fz, labels = format(fz))
abline(h = 0.9, col = "grey45", lwd = 1.1, lty = 2)
text(1.045, 0.9, "truth", col = "grey35", cex = 0.66, adj = c(1, -0.45))
mu_bpo <- tapply(d_bpo$beta_hat, list(d_bpo$arm, d_bpo$fuzz), mean)[arms, ]
set.seed(3)
for (a in arms) {
  s <- d_bpo[d_bpo$arm == a, ]
  points(s$fuzz + off[a] + runif(nrow(s), -0.010, 0.010), s$beta_hat,
         pch = pchv[a], cex = 0.42, col = adjustcolor(col[a], 0.30))
  lines(fz + off[a], mu_bpo[a, ], col = col[a], lwd = 2, lty = ltyv[a])
  points(fz + off[a], mu_bpo[a, ], pch = pchv[a], cex = 1.0, col = col[a])
}
legend("bottomleft", legend = arms, col = col[arms], pch = pchv[arms],
       lty = ltyv[arms], lwd = 1.6, bty = "n", cex = 0.72,
       text.col = col[arms])
mtext("400 precise-but-biased / 100 unbiased-but-fuzzed", side = 3,
      line = 0.5, cex = 0.75, font = 2)
## The fence travels with the PNG, not with the caption.
mtext("SIMULATED DATA — point estimates only; no interval or coverage claim",
      side = 1, line = 3.0, cex = 0.68, col = "grey35")
```

```{r biased-po-table, echo = FALSE}
agg3 <- aggregate(beta_hat ~ fuzz + arm, d_bpo, mean)
w3 <- reshape(agg3, idvar = "fuzz", timevar = "arm", direction = "wide")
names(w3) <- sub("beta_hat\\.", "", names(w3))
w3$penalty <- w3$integrated - w3$precise
bpo_cnt <- t(sapply(w3$fuzz, function(fv) {
  s <- wide(d_bpo[d_bpo$fuzz == fv, ])
  c(sum(s$integrated - s$precise < 0),
    sum(abs(s$integrated - 0.9) > abs(s$precise - 0.9)))
}))
w3$lower <- bpo_cnt[, 1]; w3$worse <- bpo_cnt[, 2]
w3 <- w3[, c("fuzz", "precise", "fuzzed", "integrated", "penalty",
             "lower", "worse")]
names(w3) <- c("fuzz", "precise (biased)", "fuzzed (unbiased)", "integrated",
               "integrated - precise", "lower (of 15)",
               "further from truth (of 15)")
knitr::kable(w3, digits = 3, row.names = FALSE)
```

**The penalty survives the realistic design.** With the motivation for
integration present and correctly modelled, integrating the fuzzed arm still
moves the environmental slope further from truth in
`r min(bpo_cnt[w3$fuzz > 0, 2])` of 15 replicates at both non-zero fuzz levels.
It is smaller than in the unbiased design — but **do not compare the magnitudes
across the two runs**: this one fits an `access` term the other does not, so
they are different models, and only the within-run contrast is valid in each.

The finding is narrower than it sounds, though: the penalty falls on the
**environmental slope specifically**. The *same* arm, once its coordinates are
fuzzed, damages the slope while it repairs the reporting bias. Both are true at
once, and which one dominates is the next section — not a rhetorical question,
because the answer changes sign.

# When the conclusion reverses

Everything above grants the simulation something no real analysis has: the
accessibility surface is measured **exactly**, and modelled. Given that, the
presence-only arm's environmental slope is already unbiased — so integrating a
fuzzed arm can only cost, and the whole measurement above is a cost with no
benefit to weigh against it.

Relax that and the ranking flips. `generate-surrogate-error-results.R` changes
one thing at a time, 12 replicates at `fuzz = 1.0`, the worst case above. Note
the second row: it is the cell that identifies the mechanism, and the earlier
version of this table did not have it.

```{r reversal-table, echo = FALSE}
rev_tab <- do.call(rbind, lapply(sort(unique(d_rev$order)), function(o) {
  s <- d_rev[d_rev$order == o, ]
  data.frame(
    `what the analyst has` = s$cond[1],
    `cor(access, env)`     = s$rho[1],
    `surrogate error sd`   = s$err_sd[1],
    precise                = mean(s$precise),
    `precise-arm bias t`   = unname(t.test(s$precise - 0.9)$statistic),
    integrated             = mean(s$integrated),
    `integration helped`   = paste0(
      sum(abs(s$integrated - 0.9) < abs(s$precise - 0.9)), " / 12"),
    check.names = FALSE, row.names = NULL)
}))
knitr::kable(rev_tab, digits = c(0, 1, 1, 3, 2, 3, 0), row.names = FALSE)
```

```{r reversal-derived, include = FALSE}
hlp <- function(o) { s <- d_rev[d_rev$order == o, ]
  sum(abs(s$integrated - 0.9) < abs(s$precise - 0.9)) }
p_flip <- fisher.test(matrix(c(hlp(4), 12 - hlp(4),
                               hlp(3), 12 - hlp(3)), 2))$p.value
p_err  <- fisher.test(matrix(c(hlp(2), 12 - hlp(2),
                               hlp(1), 12 - hlp(1)), 2))$p.value
ci_str <- function(k) { ci <- binom.test(k, 12)$conf.int
  sprintf("%d/12 is [%.3f, %.3f]", k, ci[1], ci[2]) }
```

**The flip needs the conjunction, not either ingredient alone.** Surrogate
error *by itself*, with accessibility independent of the environment, leaves
the presence-only slope unbiased (row 2, bias *t* below 2 in absolute value)
and integration helps in `r hlp(2)` of 12 — indistinguishable from the
`r hlp(1)` of 12 baseline (Fisher *p* = `r fmt(p_err, 2)`). Confounding by
itself, with the surrogate measured exactly, helps in `r hlp(3)` of 12. Only
when the surrogate is **both imperfect and correlated with the environmental
covariate** does the ranking reverse, to `r hlp(4)` of 12
(*p* = `r format.pval(p_flip, digits = 2)` against confounding alone).

That is what errors-in-variables predicts: measurement error in one regressor
spills into another's coefficient only through their covariance. It is also a
much more useful instruction, because the condition is one you can *check*
rather than assume. Compute `cor(env, surrogate)` on your own data. Road
density is correlated with almost every environmental layer, so in practice the
warning usually applies — but now you can tell.

**Read the table as one contrast, not four ordered results.** With 12
replicates the proportions carry wide intervals (`r ci_str(hlp(4))`;
`r ci_str(hlp(1))`). Rows 1 and 2 are not distinguishable from each other, and
rows 4 and 5 are not either. The single informative comparison is
confounding-alone against confounding-plus-error.

**And note what the bottom two rows also say: in that regime nothing recovers
the truth.** Precise `r fmt(rev_tab$precise[4], 2)`, integrated
`r fmt(rev_tab$integrated[4], 2)`, against a true 0.9. The *ranking* flips;
both answers are still bad. "Integration helped" there means "less wrong", not
"right".

So the decision this article supports is narrower than the measurement that
drives it. **Positional precision is a cost to measure, not a reason to discard
a structured arm.** If your bias surrogate is imperfect *and* correlated with
your environment — and for road-density-style surrogates it usually is — the
structured arm is probably earning its place despite the fuzz.

# What to do

`gllvmTMB` does not implement a fix and this article does not claim one. The
general remedy is an errors-in-variables layer with a per-arm positional prior;
that is a real modelling option and it is not implemented here.

1. **Compute the ratio, do not eyeball the map.** Run
   [the recipe](#the-recipe) on your own raster, divide any reported *practical
   range* by 3 first, and compare the result to your survey arm's positional
   uncertainty. Fuzz well below the correlation length is benign; by half the
   correlation length the tables above already show
   `r fmt(min(att05), 0)`%+ attenuation in the affected arm.
2. **Check `cor(env, surrogate)`.** It decides which way the previous section
   points for you. Independent surrogate: the fuzzed arm is a net cost.
   Correlated and imperfect: it is probably a net gain.
3. **Weight the arms, or at least know what the unweighted fit is doing.** The
   penalty tracks the fuzzed arm's share of the pooled data, so an integrated
   fit that lets the larger arm dominate is making that choice implicitly.
   Fletcher et al. (2019) reach the same conclusion from the swamping side.
4. **Fit the arms separately first.** If the precise-arm-alone and integrated
   slopes disagree materially, that gap is evidence rather than noise. In real
   data it can reflect positional error, sampling bias, detection differences,
   or a genuine scale difference all at once, and these clean simulations
   cannot tell you which.
5. **Ask for the precise coordinates.** Most fuzzing regimes, ABMI's included,
   have a data-agreement route to exact locations. For an analysis that turns
   on an environmental slope, that request is worth making before the analysis
   rather than after.

**And do not reach for the obvious mitigation, because it has been tested and
it does not work.** The intuitive fix is to match covariate resolution to
positional precision — if coordinates are fuzzed to 5.5 km, extract covariates
at 5.5 km, so they are read at approximately the right place. An earlier
version of this article recommended it. **Gábor et al. (2022) tested it
directly and the answer is in their title: positional errors in species
distribution modelling are not overcome by the coarser grains of analysis.**
Moudrý et al. (2023) restate the same finding. Coarsening destroys fine-scale
signal and does not buy back what positional error took, so it costs twice.
Measure the ratio, and treat the positional error as a property of the data
rather than something a resolution choice can absorb.

**A related defect worth checking while you are there.** Positional error
corrupts the covariate at *fit* time; scaling it separately on the training
data and on a prediction grid corrupts it at *prediction* time. That error is
affine, so it preserves the map's rank order **exactly** — every hotspot stays
a hotspot, nothing warns — while the predicted values move by an amount that
depends on the two datasets' moments. The companion `isdm-canada-warbler`
article shows the fix; a worked case with measured magnitudes is in
`dev/isdm-precision/READER-BLOCKS.md` (Block 5).

# Scope and limits

- **Simulation only, and that is a hard boundary, not a caveat.** No real ABMI
  or GBIF data was fitted anywhere here. The 5.5 km figure is documented ABMI
  policy and motivates the design; it is not a measurement of bias in any real
  dataset.
```{r truncation, include = FALSE}
## The simulations apply pmin(pmax(coord + rnorm(n, 0, fuzz * ell), 0), 1),
## so displacement is clipped at the domain boundary. Measure what that
## actually does rather than describing the intended mechanism.
set.seed(11); n <- 5e5; ell <- 0.25
trunc <- t(sapply(c(0.5, 1.0), function(fzz) {
  lo <- runif(n); la <- runif(n)
  ex <- rnorm(n, 0, fzz * ell); ey <- rnorm(n, 0, fzz * ell)
  lf <- pmin(pmax(lo + ex, 0), 1); af <- pmin(pmax(la + ey, 0), 1)
  c(clipped  = mean(lo + ex < 0 | lo + ex > 1 | la + ey < 0 | la + ey > 1),
    realised = mean(sqrt((lf - lo)^2 + (af - la)^2)) /
               mean(sqrt(ex^2 + ey^2)))
}))
rownames(trunc) <- c("0.5", "1.0")
```

- **The displacement is a truncated Gaussian, not the isotropic Gaussian the
  fuzz axis names.** Coordinates are clipped to the unit domain, so at
  `fuzz = 1.0` about `r fmt(100 * trunc["1.0", "clipped"], 0)`% of survey rows
  are clipped on at least one axis and the realised mean displacement is about
  `r fmt(100 * trunc["1.0", "realised"], 0)`% of nominal
  (`r fmt(100 * trunc["0.5", "clipped"], 0)`% and
  `r fmt(100 * trunc["0.5", "realised"], 0)`% at `fuzz = 0.5`). The fuzz axis
  is therefore mislabelled by roughly 10–15%, in the conservative direction —
  the fuzzed arm is if anything less damaged than the label implies.
  Administrative fuzzing is often uniform-within-a-cell or grid-snapped, which
  will behave differently again.
- **One covariate structure.** A single smooth exponential-correlation
  surface. Patchier or anisotropic covariates attenuate differently.
- **The simulated landscape sits where the recipe reads low.** Its `phi` is
  0.25 of the domain side, so a reader applying this article's own recipe to
  this article's own landscape would get about `r fmt(0.25 * r025, 2)` rather
  than 0.25, and would read the `fuzz = 1.0` cell as a ratio near
  `r fmt(1 / r025, 1)` rather than 1.0. That is exactly the arithmetic the
  article asks readers to perform, so it is stated rather than left to be
  discovered.
- **Fixed effects only.** No latent field is fitted in either design, which
  isolates the covariate channel and means the field-absorption intuition is
  set aside, not tested. Positional error also corrupts an estimated spatial
  field; that is not measured here.
- **One bias mechanism.** One accessibility surface, one bias coefficient.
  Real reporting bias may be multi-causal, non-linear, or itself measured with
  error.
- **Fifteen replicates, and three designs on one set of landscapes.** The
  effects reported are large relative to their confidence intervals, but the
  three designs are not three independent confirmations, and no result here is
  a precision estimate.
- **No coverage claim.** Point estimates only. Whether intervals around these
  attenuated slopes retain nominal coverage is a separate question and is not
  answered.

## References

- Fletcher, R. J. Jr., et al. (2019). *A practical guide for combining data to
  model species distributions.* Establishes that unweighted integrated models
  are driven by the abundant presence-only arm, increasingly so as that arm
  grows.
- Gábor, L., Jetz, W., Lu, M., et al. (2022). *Positional errors in species
  distribution modelling are not overcome by the coarser grains of analysis.*
  Also reports SDM performance declining as positional error grows.
- Moudrý, V., et al. (2023). Restates the Gábor et al. result in a review of
  spatial-resolution choices in species distribution modelling.
