---
title: "An integrated SDM for twelve species, end to end"
description: "How to combine eBird-style opportunistic counts with a structured detection survey for a twelve-species songbird community when the two sources share no locations, and how to make a map out of the result. Covers the two decisions you will actually face -- how many shared spatial gradients to fit, and what to do about overdispersion -- and is honest about what the map does not carry."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{An integrated SDM for twelve species, end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

## ---------------------------------------------------------------------------
## Evidence files. Every replicated number in this article is READ from one of
## these, not recomputed at render time: they are campaigns of tens of fits at
## about a minute each, and a documentation build must not run them. If one is
## missing the build STOPS rather than falling back to a hard-coded number --
## a page that quietly substitutes a remembered figure for a measurement is
## the failure mode this article spends its length guarding against.
## ---------------------------------------------------------------------------
D <- "../../dev/isdm-precision"
need <- c(probe = "cawa12-probe.rds", aic = "cawa12-aic.rds",
          uniq  = "cawa12-unique.rds", quad = "cawa12-quadrature.rds",
          mapd  = "cawa12-mapdata.rds", gate = "precision-sim-gate.rds")
paths <- stats::setNames(file.path(D, need), names(need))
absent <- paths[!file.exists(paths)]
if (length(absent))
  stop("missing evidence file(s): ", paste(absent, collapse = ", "),
       " -- re-run the matching dev/isdm-precision/generate-*.R script")
pr <- readRDS(paths[["probe"]]); ai <- readRDS(paths[["aic"]])
un <- readRDS(paths[["uniq"]]);  qd <- readRDS(paths[["quad"]])
md <- readRDS(paths[["mapd"]]);  gt <- readRDS(paths[["gate"]])
us <- un$summary

f2 <- function(x, d = 2) formatC(x, format = "f", digits = d)

## Derived objects are built ONCE, here, so that a section can be moved
## without breaking an inline reference somewhere else on the page.
sw  <- pr$sweep              # count-ceiling sweep, 3 seeds per rung
ss  <- pr$sweep_summary      # recovery envelope, 12 seeds per rung
ff  <- ai$fits; sel <- ai$selection; ts <- ai$tell_summary
so  <- us$slope_overall
dl  <- us$dloglik
v   <- us$vmr[order(us$vmr$sigma_od), ]
ps  <- us$psi[order(us$psi$sigma_od), ]
hi  <- ps$sigma_od >= 0.55
psi_ratio_hi <- ps$psi_median[hi] / ps$sigma2[hi]
rp  <- qd$replicates
sd_ <- qd$sweep_decoupled$summary
npp <- qd$nodes_per_presence
mc  <- mean(qd$replicates$sd)
big_beta <- abs(rp$truth) > 0.4
nll_gain <- t(sapply(split(ff, ff$seed), function(z) c(
  `2->3` = z$objective[z$k == 2] - z$objective[z$k == 3],
  `3->4` = z$objective[z$k == 3] - z$objective[z$k == 4])))
k3l <- ai$loadings[ai$loadings$k == 3, ]
smallest_col <- sapply(split(k3l, k3l$seed),
                       function(z) z$column[which.min(z$col_rms)])
nz3 <- ai$near_zero_tell[ai$near_zero_tell$k == 3, ]
nz4 <- ai$near_zero_tell[ai$near_zero_tell$k == 4, ]
sv3_nonzero <- nz3$sv_rel_3[nz3$sv_rel_3 > 1e-6]

## One aggregation of the count-ceiling sweep, used at two densities: a short
## excerpt in the main path, the whole thing in the record at the end.
short_reason <- c("convergence != 0" = "no convergence",
                  "iterations <= 1 (#1167 silent failure)" = "1 iteration",
                  "error: All 1 restarts failed." = "optimiser error")
agg <- do.call(rbind, lapply(split(sw, sw$kappa), function(z) data.frame(
  kappa      = z$kappa[1],
  mean_ct    = round(mean(z$mean_count), 2),
  max_ct     = paste(range(z$max_count), collapse = "-"),
  iters      = if (all(is.na(z$iterations))) "-"
               else paste(range(z$iterations, na.rm = TRUE), collapse = "-"),
  objective  = if (all(is.na(z$objective))) "-"
               else format(mean(z$objective), digits = 2, scientific = TRUE),
  ok         = paste0(sum(z$ok), "/", nrow(z)),
  verdict    = if (all(z$ok)) "healthy"
               else unname(short_reason[unique(z$reason[!z$ok])[1]]))))
rownames(agg) <- NULL

## Capture a condition without letting its ABSENCE break the build. A vignette
## that calls conditionMessage() on whatever tryCatch() hands back will die
## with an unrelated error the day the package stops signalling -- a change
## worth noticing, but not one that should take the whole website down.
warn_of <- function(expr) {
  w <- tryCatch({ force(expr); NULL }, warning = function(w) w)
  if (is.null(w)) "(behaviour has changed: no warning was emitted)"
  else paste0("<", class(w)[1], ">\n", conditionMessage(w))
}
err_of <- function(expr) {
  e <- tryCatch({ force(expr); NULL }, error = function(e) e)
  if (is.null(e)) "(behaviour has changed: no error was raised)"
  else paste0("<", class(e)[1], ">\n", conditionMessage(e))
}
msg_of <- function(expr) {
  m <- tryCatch({ force(expr); NULL }, message = function(m) m)
  if (is.null(m)) "(behaviour has changed: no message was emitted)"
  else conditionMessage(m)
}
```

# What this shows

You have two kinds of data on the same birds. One is **opportunistic** --
[eBird](https://ebird.org) checklists, [GBIF](https://www.gbif.org) records --
with counts, wildly uneven effort and no sampling design. (If those names are
new: eBird is a global bird-recording platform built on checklists, and GBIF,
the Global Biodiversity Information Facility, is the portal where species
records from museums, surveys and citizen science are aggregated and
downloaded.) The other is a **structured survey**: a protocol, fixed
stations, detection or non-detection, and far fewer sites. An **integrated
species distribution model** fits both at once.

The motivation, in one line, is that the designed sample should discipline the
messy one -- though note up front that this article does not *test* that: it
never compares an opportunistic-only fit, a survey-only fit and the integrated
one. What it does is run one integrated model end to end for **twelve boreal
songbirds**, with the two sources at locations that share nothing at all, and
finish with the thing the exercise is for: a map of relative intensity for all
twelve species.

Twelve species rather than two, because twelve makes two decisions real that
you will meet on your own data.

1. **How many shared spatial gradients should the model fit?** The species
   are tied together by unmeasured surfaces they all respond to, and you have
   to say how many. [Decision 1: how many shared gradients?] settles it with
   AIC, and shows why the rule of thumb most people reach for instead points
   one gradient too high.
2. **What do you do about overdispersion?** The count arm is Poisson, and
   Poisson has no dispersion parameter, so extra noise has to go somewhere.
   [Decision 2: overdispersion in an arm with no dispersion parameter]
   measures the option this package offers.

**Getting the package.** gllvmTMB is not on CRAN; install it from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("itchyshin/gllvmTMB")
library(gllvmTMB)
```

**And note: you need the repository to run this page's code, not just the
installed package.** The simulator and the saved campaign results live in
`dev/isdm-precision/` in the
[gllvmTMB source tree](https://github.com/itchyshin/gllvmTMB); copying chunks
into a fresh session will not work on its own. Everything you need for *your
own* data — `isdm_sources()`, `make_mesh()`, `gllvmTMB()`, `predict()` — is in
the installed package.

Two neighbouring questions have articles of their own rather than a paragraph
here: how much survey you need
([*How big does an integrated survey design need to be?*](integrated-survey-design.html))
and what repeat visits buy you
([*What do repeated survey visits add to an integrated model?*](integrated-repeated-visits.html)).

## The data are simulated

**No real GBIF or [ABMI](https://abmi.ca) (Alberta Biodiversity Monitoring
Institute) data were analysed for this article.** Everything below
describes a simulated boreal songbird community shaped on a real system --
Canada Warbler (*Cardellina canadensis*) and eleven co-occurring species in
Alberta, an eBird-style opportunistic arm and an ABMI-style structured detection arm
(a standardised provincial survey programme) -- and no number here is a measurement about any real taxon.

Three reasons, all of which you would meet on the way to the real analysis:

1. **The structured arm is not freely available at usable precision.** ABMI
   publishes public survey locations fuzzed to roughly 5.5 km; precise
   coordinates need a **separate data agreement**, and fitting the fuzzed ones
   would measure the fuzz as much as the birds -- the subject of the companion
   article
   [*When one data source knows where it is and the other does not*](isdm-spatial-precision.html).
2. **A vignette must not depend on the network.** [rgbif](https://docs.ropensci.org/rgbif/) will fetch the
   opportunistic arm, but a documentation build that phones an API is a
   documentation build that breaks.
3. **Real integration needs cleaning and matching work** that would swamp
   everything this page is trying to show.

Simulating also buys the one thing real data cannot: a known truth, against
which *did the model recover it* is a question with an answer.

**This simulation is the easy version of the problem**, in two specific ways
that matter for whether any of it transfers: the opportunistic records are
scattered at random rather than clustered along roads, and the environmental
covariate is uncorrelated with the unmeasured spatial structure by
construction. Neither holds on real data.
[Before you use this on real data] spells out what that leaves untested.

# The two arms, and the one field they share

Everything here rests on a single idea: the two data sources are two ways of
observing **the same underlying abundance**. Write the log of that abundance
for species $j$ at location $i$ as

$$\eta_{ij} \;=\; \alpha_j \;+\; \beta_j\,\mathrm{env}_i
  \;+\; \lambda_{1j} f_{1i} \;+\; \lambda_{2j} f_{2i}
  \;+\; \log(\text{effort}_i)
  \;+\; \kappa\,\mathbf{1}\{\text{arm}_i = \text{opportunistic}\},$$

and read it one piece at a time.

**$\alpha_j + \beta_j\,\mathrm{env}_i$ -- the habitat part.** Each species
has its own intercept and its own slope on the environmental covariate,
exactly as it would in a per-species GLM. These slopes are what most people
came for.

**$\lambda_{1j} f_{1i} + \lambda_{2j} f_{2i}$ -- the shared unmeasured
gradients.** A **latent spatial field** is everything spatial you did not
measure, estimated as a smooth surface: soil, moisture, disturbance history,
whatever else organises the community. There are two of them here, estimated
rather than supplied, and each species carries its own pair of **loadings** on
them. A species with a large positive loading on the first surface is common
where that surface runs high, and species with similar loadings co-occur --
which is how a community model differs from twelve separate SDMs.

**$\log(\text{effort}_i)$ -- the offset.** A known multiplier with no fitted
coefficient: doubling the effort doubles the expected count, and the model is
told that rather than asked to discover it. Any other *known* adjustment goes
in the same place. If you carry [QPAD](https://github.com/borealbirds/QPAD)-style
detectability offsets for time of day, date and distance (Sólymos et al.
2013), add them into this term rather than modelling them
separately.

**$\kappa$ -- the arm indicator.** The two sources report on different overall
scales, and one intercept per source absorbs that difference so the shared
parts do not have to.

Counts and detections then come off that same $\eta$ by different routes:
$y_{ij}\sim\text{Poisson}(e^{\eta_{ij}})$ on the opportunistic arm, and
$y_{ij}\sim\text{Bernoulli}(1-e^{-e^{\eta_{ij}}})$ -- a **complementary
log-log** link -- on the structured one. The cloglog link is not a taste: it
is exactly the probability that a Poisson count with that same mean is at
least one, so "at least one bird detected here" and "this many birds counted
there" become statements about one abundance rather than two. `isdm_sources()`
is what tells the model which rows follow which process.

**The mesh, and why your two sources need not share a single site.** A
**mesh** is a triangulated grid laid over the study area. The latent field is
estimated at the mesh nodes, and any location -- from either arm -- takes its
value as a weighted average of the nodes around it. That is the whole
mechanism by which two data sources sharing no row, no coordinate and no
label share a field: **the field lives on the mesh, not on the data.**

That is the question that stops most people before they start. You do not need
to match an eBird checklist to an ABMI station, grid both onto a common
raster, or throw away the source with fewer sites. Build the mesh over the
union of both, and project every row onto it.

The simulator draws the two arms **independently** over one landscape, so
nothing here quietly assumes a convenience the real data will not have:

```{r data}
local({
  article_dir <- getwd()
  on.exit(setwd(article_dir), add = TRUE)
  setwd(normalizePath("../.."))
  suppressMessages(source("dev/isdm-precision/generate-cawa12.R"))
})

dat   <- sim_cawa12(seed = 1L)
truth <- attr(dat, "truth")
cells <- attr(dat, "cells")

str(dat[, c("cell_id", "trait", "isdm_source", "env", "log_effort", "value")])
```

**This `str()` is the shape your own data has to reach, so read it as a
target.** One row per species per location. `cell_id` labels the location
(give every location its own label — it is bookkeeping here, and
[the reference section](#reference-four-things-that-do-not-do-what-you-would-guess)
shows it never enters the prediction). `trait` is the species. `isdm_source`
says which dataset the row came from. `value` is the count on the
opportunistic rows and detected/not-detected (1/0) on the survey rows.
`log_effort` is the log of whatever measures how hard that row was looked for
— checklist duration, point-count minutes — and any known detectability
corrections you already compute (QPAD-style offsets for time of day, date,
distance) are added into it. Building this frame from an eBird download plus
your own survey table — stacking the two, agreeing on the covariate columns,
constructing `log_effort` for each arm — is real work this page does not show;
it is ordinary `rbind()` and `dplyr` work, but budget time for it.

**What counts as a "site"?** Whatever unit carries one location and one
effort. In this simulation a site is a randomly placed point — despite the
`cell_id` name, **nothing here is a grid cell, and no gridding is required**:
the model works on exact coordinates, the spatial field is a continuous
surface, and the mesh is a computational device for estimating it, not a grid
your data must live on. In your own data an eBird checklist is a natural site
(one place, one duration); pooling checklists into grid cells and summing
their effort is a *choice* you may make, never a requirement. A survey
station is a site as it stands. The whole question — sites, cells,
checklists, and why the mesh is none of them — has its own short article:
[*What is a site? The unit of analysis*](unit-of-analysis.html). And the table is a **complete crossing**:
every site has a row for every species, *zeros included*. Those zeros are
data — "we looked here and did not find this species" — and they exist in
eBird only for **complete checklists**, where the observer reported everything
detected. A raw presence-only download has no zeros to give, which is exactly
why the [presence-only section](#if-your-opportunistic-arm-has-no-counts)
treats it as a different problem.

**If your survey records counts rather than detections, keep them.** Declare
both arms Poisson — `isdm_sources(ebird = poisson(), survey = poisson())` is
accepted — and do not binarise your abundances.

```{r arms}
tab <- table(cells$arm)
cat("opportunistic cells:", tab[["ebird"]],
    " structured stations:", tab[["abmi"]],
    "\nspecies:", length(truth$code), " rows:", nrow(dat), "\n")

## Disjoint by construction: no cell id, and no coordinate, is shared.
po <- cells[cells$arm == "ebird", ]; sv <- cells[cells$arm == "abmi", ]
cat("shared cell ids:", length(intersect(po$cell_id, sv$cell_id)),
    " shared (X, Y):",
    length(intersect(paste(po$X, po$Y), paste(sv$X, sv$Y))), "\n")

## Effort varies WITHIN each arm, so the offset is not a relabelling of arm.
cat("sd(log effort) within arm:", f2(sd(po$log_effort), 3), "/",
    f2(sd(sv$log_effort), 3),
    " cor(log effort, arm):",
    f2(cor(cells$log_effort, cells$arm == "ebird"), 3), "\n")

cat("largest opportunistic count:", max(dat$value[dat$isdm_source == "ebird"]),
    " detection rate, structured arm:",
    f2(mean(dat$value[dat$isdm_source == "abmi"]), 3), "\n")
```

One thing to do before you mesh anything: **project your coordinates**. A
degree of longitude and a degree of latitude are different physical distances
everywhere but the equator, and a mesh built on degrees treats them as equal.
These coordinates are in UTM kilometres inside a single zone. The
[For the record] section returns to this, because the obvious way of checking
such a choice -- fit both and compare -- does not work.

# Fit it, and check that it is really a fit

```{r fit}
t0 <- proc.time()[["elapsed"]]
f  <- cawa12_fit(dat)
elapsed <- proc.time()[["elapsed"]] - t0
fit <- f$fit

cat("model: "); print(cawa12_formula()[[3]])
cat("\naccepted:", f$ok, " convergence:", f$convergence,
    " iterations:", f$iterations,
    "\nobjective:", f2(f$objective, 3), " seconds:", f2(elapsed, 1),
    " mesh nodes:", cawa12_mesh(dat)$mesh$n, "\n")
```

`spatial_latent(0 + trait | coords, d = 2)` is the shared field, with
per-species loadings on it; `d = 2` is the number of gradients, the subject of
the next section. The mesh is built from `dat` -- the union of both arms'
locations -- and every row from either arm is projected onto it. The
`cutoff` in `make_mesh()` -- an interface inherited from
[sdmTMB](https://pbs-assess.github.io/sdmTMB/), the single-species spatial
package these helpers descend from -- is the minimum spacing between mesh
nodes, **in the units of your coordinates** -- kilometres here, because the coordinates were
projected first. This page uses `cutoff = 40` on a domain roughly 225 by
443 km, giving `r cawa12_mesh(dat)$mesh$n` nodes: about a tenth of the
domain's shorter side is a reasonable starting point. A smaller cutoff means
more nodes, a finer field, and a slower fit; if your conclusions change when
you halve it, the mesh was too coarse.

This is a **small example**: `r nrow(cells)` sampled locations, a
`r cawa12_mesh(dat)$mesh$n`-node mesh, about a minute per fit. Nothing on this
page measures how any of it scales to tens of thousands of checklists or a
mesh an order of magnitude larger.

**`convergence == 0` is not the check.** This package can return
`convergence == 0` after a **single** iteration, at an astronomically large
objective, with every spatial parameter still sitting at its starting value:
a fit that never happened, reporting success. So look at `iterations` as
well. Counts in the tens or hundreds are the healthy shape; **1 means the
optimiser never moved**. `cawa12_fit()` accepts a fit only when all three of
`convergence == 0`, `iterations > 1` and a finite objective hold.

**Keep the counts moderate.** That failure arrives when the counts get large,
and *large* is not a feeling, so it was measured: the opportunistic arm's
overall intensity -- the same $\kappa$ the model formula above carries as that
arm's intercept, here set in the simulation -- was raised step by step, three
datasets per step. `kappa` in the table below is the value it was set to.

```{r ceiling-short, echo = FALSE}
excerpt <- agg[agg$kappa %in% c(1, 2.3, 2.4, 3),
               c("kappa", "mean_ct", "max_ct", "iters", "ok", "verdict")]
rownames(excerpt) <- NULL
excerpt
```

Two things to take from that. The cliff is **not far off** -- fits are healthy
up to a mean cell count near `r f2(max(sw$mean_count[sw$kappa == 2.3]), 1)`
and all three seeds fail one step later -- and a cap on the largest single
cell is a **poor early-warning sign**, because the largest count overlaps across the cliff
(`r pr$envelope$max_cell_count_healthy` in a healthy fit against
`r pr$envelope$min_cell_count_failing` in a rejected one); what controls it is
the *mean* cell intensity. So *keep your largest cell count under about
`r pr$envelope$safe_max_cell_count_recommended`* is a recommendation with
margin, not a measured threshold.

Accuracy starts decaying two steps **before** the optimiser breaks, while
every fit still reports itself healthy. The full sweep, and the recovery
numbers behind that sentence, are in [For the record].

## What this one fit recovered

```{r recovery-one}
est <- cawa12_env_slopes(fit)
cmp <- data.frame(species = truth$code, truth = round(truth$beta, 3),
                  estimate = round(unname(est), 3))
cmp$error <- round(cmp$estimate - cmp$truth, 3)
cmp
cat("cor(estimate, truth):", f2(cor(cmp$estimate, cmp$truth), 3),
    " mean |error|:", f2(mean(abs(cmp$error)), 3),
    " mean signed error:", f2(mean(cmp$error), 3), "\n")
```

That is **one** dataset, and a single replicate scatters. Read it against the
12-seed envelope instead: correlation with truth
`r f2(ss$mean_cor[ss$kappa == 1], 3)`, sd
`r f2(ss$sd_cor[ss$kappa == 1], 3)`. Judging an estimator from one replicate
is how people conclude it is broken when it is not.

**The slopes do come back shrunk**, here and in every replicate: the mean
signed error across the campaign is
`r f2(mean(ss$mean_signed_err), 3)` against a true slope SD of
`r f2(sd(truth$beta), 3)`. This article makes no claim that these estimates
are unbiased.

# Decision 1: how many shared gradients?

**Verdict: fit `d = 1, 2, 3, 4`, and take the lowest AIC.** On ten replicate
datasets whose true number of gradients is 2, AIC picked 2 in
**`r sum(sel$aic_pick == 2)` of `r nrow(sel)`**. The rule of thumb you would
otherwise reach for -- fit a generous number, then stop where a loading column
goes near zero -- got it right in **1 of 10**, and no threshold rescues it.

`d` is the number of independent unmeasured gradients the community responds
to. `d = 1` says one hidden surface organises all twelve species; `d = 4` says
four do. The model will fit whichever number you write, without complaint,
which is why this is a decision and not a setting.

```{r aic-table, echo = FALSE}
by_k <- do.call(rbind, lapply(split(ff, ff$k), function(z) data.frame(
  d = z$k[1], df = z$df[1],
  mean_AIC       = round(mean(z$AIC), 1),
  AIC_range      = paste(round(range(z$AIC), 1), collapse = " - "),
  median_secs    = round(median(z$secs), 1))))
by_k$mean_delta_AIC <- round(c(mean(sel$delta_aic_1), mean(sel$delta_aic_2),
                               mean(sel$delta_aic_3), mean(sel$delta_aic_4)), 2)
rownames(by_k) <- NULL
by_k
cat("AIC picks d = 2 in", sum(sel$aic_pick == 2), "of", nrow(sel),
    "replicates; BIC picks d = 2 in", sum(sel$bic_pick == 2), "\n")
```

Fitting too few gradients is expensive -- `d = 1` costs
`r f2(mean(ff$objective[ff$k == 1]) - mean(ff$objective[ff$k == 2]), 1)`
units of log-likelihood on average -- while each gradient past the truth buys
almost nothing and pays about 20 AIC for the privilege. That asymmetry is why
AIC lands on the right answer so consistently here. One caveat on the
comparison itself: at a superfluous rank (a number of gradients `d` above the truth) the extra
loading column sits on the
boundary of its parameter space, so AIC's nominal penalty is conservative in
the direction that flatters this 10-of-10 -- the same boundary problem noted
for `Psi` in [For the record].

```{r rank-fig, echo = FALSE, fig.width = 6.6, fig.height = 3.6, dpi = 200, out.width = "100%", fig.cap = "Choosing the number of shared gradients on ten replicate datasets whose true number is 2. Panel A plots each replicate's AIC relative to its own best value: the minimum falls at d = 2 in every replicate, fitting too few at d = 1 costs 114 to 175 AIC units, and each gradient beyond the truth costs about 20 AIC for almost no gain in fit. Panel B tests the rule of thumb AIC is competing against. Each replicate was fitted with one gradient too many, and the three loading columns' root-mean-square values are plotted relative to the largest, the digit giving the column number. A reader stopping when a column falls below a near-zero cut reads the answer correctly in one replicate of ten: the smallest of the three columns runs from 0.00 to 0.34 of the largest, with only one replicate below a 0.05 cut, and the SMALLEST column is column 2 rather than the last in five of the ten replicates, because the fitted columns get mixed among themselves. No uncertainty is shown anywhere in this figure.", fig.alt = "Two panels. Panel A: ten grey lines, one per replicate dataset, plotting AIC minus the best AIC against the number of gradients fitted, 1 to 4. All ten start high at 1, between about 114 and 175, drop to exactly zero at 2, marked by a red point labelled true rank 2, then rise gently to roughly 15 to 20 at 3 and 32 to 38 at 4. Panel B: ten horizontal rows, one per replicate seed, each carrying three bold digits 1, 2 and 3 placed along a horizontal axis from 0 to 1 giving each loading column's root-mean-square relative to the largest column. Digit 1 sits at 1.0 in every row by construction. The other two digits scatter between about 0.02 and 0.8, and in half the rows digit 3 lies to the right of digit 2. A red dashed vertical line at 0.05 marks a near-zero cut; only one row has any digit to the left of it."}
op <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(3.6, 3.6, 2.2, 0.6), oma = c(2.6, 0, 1.2, 0),
    mgp = c(2.1, 0.55, 0), las = 1, cex.axis = 0.72, cex.lab = 0.8, tcl = -0.3)

DD <- as.matrix(sel[, paste0("delta_aic_", 1:4)])
plot(NA, xlim = c(0.8, 4.2), ylim = c(-14, max(DD) * 1.06), xaxt = "n",
     xlab = "gradients fitted (d)", ylab = "AIC - AIC(best d)")
axis(1, at = 1:4, cex.axis = 0.72); abline(h = 0, col = "grey85")
for (i in seq_len(nrow(DD)))
  lines(1:4, DD[i, ], col = adjustcolor("grey35", 0.5), lwd = 1)
points(rep(1:4, each = nrow(DD)), as.vector(DD), pch = 16, cex = 0.5,
       col = adjustcolor("grey20", 0.6))
points(rep(2, nrow(DD)), DD[, 2], pch = 21, bg = "#B2182B", col = "white",
       cex = 1, lwd = 0.6)
text(2, -9, "true rank 2", cex = 0.68, col = "#B2182B", font = 2)
text(4.15, max(DD) * 0.62, "extra gradients cost\n~20 AIC each and\nbuy almost no fit",
     cex = 0.6, col = "grey30", adj = 1)
mtext("A. AIC picks the true number, 10 of 10", side = 3, line = 0.4,
      cex = 0.78, font = 2, adj = 0)

k3    <- ai$loadings[ai$loadings$k == 3, ]
seeds <- sort(unique(k3$seed))
rel   <- t(sapply(seeds, function(s) {
  v <- k3$col_rms[k3$seed == s][order(k3$column[k3$seed == s])]; v / max(v) }))
plot(NA, xlim = c(-0.03, 1.05), ylim = c(0.4, length(seeds) + 1.1), yaxt = "n",
     xlab = "loading-column rms / largest column", ylab = "")
axis(2, at = seq_along(seeds), labels = seeds, cex.axis = 0.58)
mtext("replicate (seed)", side = 2, line = 2.4, cex = 0.7, las = 0)
abline(v = 0.05, lty = 2, col = "#B2182B")
for (i in seq_along(seeds)) {
  segments(min(rel[i, ]), i, 1, i, col = "grey85")
  text(rel[i, ], rep(i, 3), labels = 1:3, cex = 0.6, font = 2,
       col = c("grey15", "#2166AC", "#B2182B"))
}
text(0.075, length(seeds) + 0.85, "a 0.05 \"near zero\" cut", cex = 0.6,
     col = "#B2182B", adj = 0)
mtext("B. the near-zero column: 1 of 10", side = 3, line = 0.4, cex = 0.78,
      font = 2, adj = 0)

mtext("Ten replicate datasets; true number of gradients 2. B is fitted with three; digits are column numbers.",
      side = 1, outer = TRUE, line = 0.3, cex = 0.6, col = "grey30")
mtext("The smallest column is column 2, not the last, in 5 of 10 replicates.",
      side = 1, outer = TRUE, line = 1.2, cex = 0.6, col = "grey30")
mtext("SIMULATED DATA -- no uncertainty shown", side = 3, outer = TRUE,
      line = -0.2, cex = 0.66, col = "#B2182B", font = 2)
par(op)
```

## Why the near-zero-loading rule fails

It fails for a reason worth knowing, because it is not specific to this
package. The fitted loading columns are **not pinned to particular
gradients**: any rotation of them describes the same model equally well, so
the columns come out mixed among themselves. "The extra column" is therefore
not the last column. Fitted with three gradients, the *smallest* column is
column 2 rather than column 3 in `r sum(smallest_col == 2)` of
`r length(smallest_col)` replicates. A reader following the recipe inspects
the third column, finds a substantial value, and concludes the third gradient
is real. Panel B above is that fact drawn.

Even ignoring the ordering, the gap is not there to find: an unneeded
direction does not always collapse to zero, and nothing inside a single fit
separates "a tenth of the leading gradient" from a weak real one. **So compare
by AIC.** If you want a loading-based sanity check anyway, use the singular
values of the loading matrix rather than its columns -- they are at least
immune to the mixing -- quote your threshold, and treat the answer as a hint.
Here is that matrix for the live fit:

```{r loadings-live}
L <- fit$report$Lambda_spde          # 12 x d -- use this, not the packed vector
dim(L)
round(sqrt(colMeans(L^2)), 4)        # per-column rms
round(svd(L)$d / max(svd(L)$d), 4)   # singular values, relative to the largest
```

Use `fit$report$Lambda_spde` rather than the packed parameter vector: the
packed version holds `P*d - d(d-1)/2` values rather than `P` by `d`, so naming
it with your species vector mislabels silently.

## What it costs

```{r rank-cost, echo = FALSE}
data.frame(d = by_k$d, median_secs_per_fit = by_k$median_secs,
           fits_in_this_campaign = as.integer(table(ff$k)))
```

Fitting four values of `d` on one dataset is a few minutes at this size. You
do not have to repeat the ten-dataset campaign -- fit your own `d = 1..4`
once and read the AIC.

# Decision 2: overdispersion in an arm with no dispersion parameter

**Verdict.** In this simulation, `spatial_latent(..., unique = TRUE)`
substantially improved the **point estimates** -- RMSE in the environmental
slopes fell from `r f2(so$rmse_base, 3)` to `r f2(so$rmse_uniq, 3)` -- but by
a mechanism that looks more like **regularisation than like estimating the
overdispersion**: how far a slope moved tracked the base model's own error
(`r f2(so$cor_move_vs_base_error, 2)`) far more closely than it tracked the
species' simulated overdispersion (`r f2(so$cor_move_vs_sigma2, 2)`). Do not
read `Psi` as a measurement of how overdispersed your counts are, and do not
expect your intervals to improve. It costs about
`r f2(us$timing$uniq_median_secs / us$timing$base_median_secs, 1)` times the
runtime and makes fits less stable. And you cannot tell from a
variance-to-mean ratio whether you need it.

Here is the problem in terms you already have. With an overdispersed Poisson
GLMM you would add an **observation-level random effect** and let it soak up
the extra noise. This model has no such tier by default, so extra-Poisson
variation goes to the term with the most freedom to absorb it: the shared
spatial field. A field busy mopping up observation noise is not estimating
space, and your slopes are estimated conditionally on it -- which is how
overdispersion you ignored comes back as bias in the slopes you care about.

`spatial_latent(..., unique = TRUE)` adds a per-species tier of its own,
called `Psi`, alongside the shared field, to play a role **similar to** the
observation-level random effect in a GLMM -- similar, not identical: `Psi` is
itself spatially structured rather than independent noise per row, which is
part of why its behaviour below is odd, and the record section says more. It
has to be asked for:

```{r unique-formula, eval = FALSE}
value ~ 0 + trait + trait:env + isdm_source + offset(log_effort) +
  spatial_latent(0 + trait | coords, d = 2, unique = TRUE)
```

It was measured rather than argued: overdispersion was added to the
opportunistic arm only, the twelve species carrying a ladder of severity from
`r f2(min(un$design$sigma_od), 2)` (none at all) to
`r f2(max(un$design$sigma_od), 2)`, and both models were fitted to identical
data on an identical mesh.

## Your counts will look overdispersed even when they are not

The obvious pre-fit check is the variance-to-mean ratio, and it misleads. The
table puts the observed VMR beside a **matched counterfactual** -- same
landscape, same effort, same species, simulated with no overdispersion at all.

```{r vmr, echo = FALSE}
data.frame(species = v$trait, sigma = v$sigma_od,
           VMR = round(v$vmr_mean, 2),
           VMR_no_overdisp = round(v$vmr_equi_mean, 2),
           frac_above = round(v$frac_gt_equi, 2))
```

**That counterfactual column runs from `r f2(min(v$vmr_equi_mean), 1)` to
`r f2(max(v$vmr_equi_mean), 1)` with no overdispersion at all**: the shared
spatial field, the environmental gradient and the varying effort inflate it on
their own. So "VMR far above 1" does not demonstrate overdispersion in a model
like this one -- and on real data you have no counterfactual to compare it
against. Only at the top of the ladder is the contrast clear; at
$\sigma \le 0.45$ the observed VMR beats its own counterfactual in
`r f2(min(v$frac_gt_equi[v$sigma_od <= 0.45]), 2)` to
`r f2(max(v$frac_gt_equi[v$sigma_od <= 0.45]), 2)` of replicates, which is a
coin flip dressed up as a diagnostic.

## What turning it on actually did

```{r unique-slopes, echo = FALSE}
data.frame(
  quantity = c("median |change| in env slope, in base SEs",
               "median SE ratio, unique / base",
               "95% coverage of the true slope (est ± 1.96 SE), base",
               "95% coverage of the true slope (est ± 1.96 SE), unique",
               "bias, base", "bias, unique", "RMSE, base", "RMSE, unique"),
  value = round(c(so$median_abs_d_est_in_se, so$se_ratio_median,
                  so$coverage_base, so$coverage_uniq,
                  so$bias_base, so$bias_uniq, so$rmse_base, so$rmse_uniq), 3))
```

The expected story is that ignoring overdispersion makes your intervals too
narrow and the fix widens them. **That is the wrong way round here.** The
slopes move -- about `r f2(so$median_abs_d_est_in_se, 2)` of a standard error
at the median -- and the standard errors move **down**, not up (median ratio
`r f2(so$se_ratio_median, 3)`). The plausible reading is that the base model's
coverage failure was **bias** rather than understated uncertainty: freeing the
shared field from having to absorb observation noise de-attenuates the slopes
and tightens them at once. Take that as the likely direction rather than a
settled result -- the coverage figures come from `r us$n_pairs_healthy`
replicate pairs, twelve species inside each sharing one estimated field, and
carry no uncertainty of their own. Coverage rises from
`r f2(so$coverage_base, 3)` to `r f2(so$coverage_uniq, 3)`, **overshooting**
the nominal 0.95, so this is no route to well-calibrated intervals either.

`Psi` itself is a **threshold detector, not a measurement**: below about
$\sigma = 0.5$ it sits at zero, and above it, it fires in most replicates but
recovers only `r f2(100 * min(psi_ratio_hi), 0)` to
`r f2(100 * max(psi_ratio_hi), 0)` per cent of the true variance.

**And it costs stability.** The unique model failed to produce a usable
curvature matrix -- the thing standard errors are computed from -- in
`r us$robustness$pdHess$n_total - us$robustness$pdHess$n_pdHess_true` of the
`r us$n_pairs_se_complete` pairs where both models returned standard errors.
Variance components pinned at zero are where that happens.

```{r unique-selection, echo = FALSE}
cat("over", dl$n, "pairs -- median delta logLik (unique - base):",
    f2(dl$median, 2), "\nextra parameters:", dl$extra_params,
    " median delta AIC:", f2(dl$dAIC_median, 2),
    "\nreplicates favouring unique:", dl$n_favouring_unique, "of", dl$n, "\n")
```

By likelihood it fits clearly better; by AIC it barely pays, and only in about
half the replicates. Restricting the comparison to the pairs with a
well-behaved curvature matrix reproduces every number above, so none of it is
an artefact of that subset -- details, and one caveat about testing the
log-likelihood difference, in [For the record].

# The map

Three things have to be right before a surface means anything.

**The spatial field must be in the prediction.** It is, from gllvmTMB 0.7.0
onwards. On an older install, `predict(newdata = )` silently dropped it, and a
map drawn that way shows a field-free surface while claiming the field. Check
your version before trusting a map made on an old install.

**The scale must be effort-free.** `type = "response"` includes the row's
offset, so it returns an expected count *at that effort* -- right for a
training row, wrong for a map. Setting `log_effort` to zero on the grid gives
relative intensity, exactly rather than approximately, because the offset is
re-evaluated against `newdata`.

**The covariate must be on the scale it was fitted on.** `env` was
standardised once, on the training data, and the grid must reuse those same
two numbers. Re-standardise the grid on its own mean and SD and the map's
*pattern* survives almost untouched while its *values* do not -- so no visual
check will catch it.

```{r map-grid}
grid <- md$grid[, c("X", "Y", "env")]     # already on the TRAINING scale
newd <- do.call(rbind, lapply(levels(dat$trait), function(sp)
  transform(grid, trait = sp)))
newd$trait       <- factor(newd$trait, levels = levels(dat$trait))
newd$cell_id     <- factor("po001", levels = levels(dat$cell_id))
newd$isdm_source <- factor("ebird", levels = levels(dat$isdm_source))
newd$log_effort  <- 0                     # relative intensity, NOT a count
newd$value       <- 0                     # the design matrix evaluates the LHS

p <- suppressWarnings(suppressMessages(
  predict(fit, newdata = newd, type = "response")))
E <- matrix(p$est, ncol = nlevels(dat$trait),
            dimnames = list(NULL, levels(dat$trait)))
cat("grid cells:", nrow(grid), " species:", ncol(E),
    " relative intensity:", f2(min(E), 3), "to", f2(max(E), 2), "\n")

## The saved surfaces the figure is drawn from came from this same fit.
agree <- max(abs(E - md$est))
cat("max |live prediction - saved surface|:", signif(agree, 3), "\n")
```

The figure below is drawn from those saved surfaces. The line above is the
check that the fit you have just run reproduces them: the largest disagreement
anywhere on the grid is `r signif(agree, 2)`, against surface values running
from `r f2(min(E), 3)` to `r f2(max(E), 2)`.

## What the map does not carry -- read this before the picture

**No uncertainty, of any kind.** There is no interval behind any value in the
figure below. `se.fit` is refused outright with `newdata`, and the refusal is
deliberate rather than an oversight:

```{r se-fit-refusal}
cat(err_of(predict(fit, newdata = newd, type = "link", se.fit = TRUE)))
```

The standard error that *is* available at training rows ignores the error in
reconstructing the latent field, which is exactly what dominates on a map.
Map-scale uncertainty needs machinery this package does not have yet.

**Nothing outside the mesh.** The grid deliberately extends 6 per cent beyond
the sampled extent, the way a real prediction map would, and `predict()` says
so:

```{r hull-warning}
cat(warn_of(suppressMessages(predict(fit, newdata = newd, type = "response"))))
```

Out there the mesh basis is all zeros, so the field is **exactly zero** --
indistinguishable, on a coloured surface, from a region where the field is
genuinely near zero. A blank corner reads as a cold one.
`r sum(md$out_of_hull)` of `r nrow(md$grid)` cells are in that state here, and
the figure hatches them rather than colouring them. Clip your grid, or extend
your mesh; do not suppress the warning, which is what the chunk above did
purely to keep the figure inline.

**A little extrapolation in the covariate.**
`r md$env_grid_cells_outside_training_range` of `r nrow(md$grid)` grid cells
carry an `env` value outside the training range, a consequence of that same
padding.

**No claim that the slopes behind it are unbiased.** They come back shrunk, as
the recovery table above showed.

```{r map-plot, echo = FALSE, fig.width = 6.6, fig.height = 7.5, dpi = 200, out.width = "100%", fig.cap = "Relative intensity for all twelve species from one integrated fit, on a shared logarithmic colour scale. Canada Warbler is drawn large, carrying the axes in projected UTM kilometres, the sampled locations and the labelled scale; the other eleven are thumbnails ordered by the correlation of each surface with the flagship's, so the grid reads as a gradient from most like it to opposite rather than as an arbitrary sequence. Colour is never the only channel: isolines at the same levels as the colour-bar ticks are drawn on every panel, and the rail under each panel places that species' 5th to 95th percentile and median on the shared scale, because a shared ramp is exactly what compresses amplitude away. Cells outside the mesh hull are hatched rather than coloured: there the spatial field is exactly zero because the mesh basis is empty, not because it was estimated to be small. The surfaces carry NO uncertainty of any kind. The simulated-data fence is drawn inside the canvas because a PNG lifted into a talk keeps its pixels and loses its caption.", fig.alt = "A large map panel for Canada Warbler beside a four-by-three grid of eleven smaller maps, one per species, all on a shared viridis colour scale for relative intensity on a log scale. The flagship panel carries UTM easting and northing axes in kilometres, white isolines, and small white circles marking all 360 sampled locations. The broad pattern is a bright high-intensity region in the north and a dark low-intensity region in the south-west. The eleven thumbnails show the same landscape expressed through different loadings, running from surfaces closely resembling the flagship to one close to its inverse. A short horizontal rail under each panel marks that species' fifth to ninety-fifth percentile with a notch at the median. A few cross-hatched cells sit in the corners, outside the mesh hull. Red text at the foot reads SIMULATED DATA, no uncertainty shown, no real observations."}
source("../../dev/isdm-precision/cawa12-map.R")
invisible(cawa12_map_plot(md))
```

The map is drawn from one dataset. Across `r length(md$ok_seeds)` replicates
on the same landscape the drawn quantity replicates well -- median
within-species correlation of the log surfaces
`r f2(median(md$stability$median_cor), 3)`, lowest single pairing
`r f2(min(md$stability$min_cor), 3)`. That is a statement about the surface's
**stability**, not about its uncertainty.

# If your opportunistic arm has no counts

eBird checklists carry counts. Many opportunistic downloads do not: a raw GBIF
export is often **presence-only** -- a location, a species, and nothing to put
in `value`.

The standard route (Warton & Shepherd 2010; Renner et al. 2015) is to fit an
inhomogeneous Poisson process as a **weighted Poisson GLM**: each presence
record becomes a row with a tiny weight `w` and response `y = 1/w`, and a grid
of background *quadrature* nodes covering the study area becomes rows with
weight equal to their cell area and response `y = 0`. Fitted that way, the
GLM's likelihood is the point-process likelihood. `gllvmTMB()` accepts exactly
that through `weights =`:

```{r po-recipe, eval = FALSE}
po <- rbind(
  data.frame(trait = sp, env = env_at_presences, y = 1e6, w = 1e-6),
  data.frame(trait = sp, env = env_at_nodes,     y = 0,   w = cell_area)
)
gllvmTMB(y ~ 0 + trait + trait:env, data = po, trait = "trait",
         unit = "obs_id", family = poisson(), weights = po$w)
```

The device recovers the true environmental slopes in this setting: over
`r nrow(qd$replicates)` species and 20 replicate datasets no species differed
detectably from its true value, and a pooled test for uniform attenuation gave
`r f2(qd$pooled_power$attenuation_mean, 4)` against a true value of 1. What
that test could and could not have detected, and how dense the quadrature grid
needs to be, are in [For the record].

The package warns about such a fit, in its own words:

```{r po-warning, echo = FALSE, comment = ""}
cat(qd$fence$weights_warning)
```

**It does not combine with the two-arm model.** Passing weights into an
`isdm_sources()` fit is refused:

```{r po-refusal, echo = FALSE, comment = ""}
cat(qd$fence$isdm_refusal_unit$msg)
```

If your opportunistic arm is genuinely presence-only, fit it with the
quadrature device **on its own**, the way the standalone
[*Joint ecological intensity from opportunistic records*](gbif-joint-intensity.html)
article does.

# Do you need grid cells? No — and here is what they would cost

Nothing on this page was gridded. Every observation sits at its exact
coordinates, the spatial field is a continuous surface, and the mesh is the
computer's scaffolding for that surface, not a grid the data live on. What a
"site" is, why a count needs an observation event rather than a map square,
and why the mesh is not a grid are untangled properly in
[*What is a site? The unit of analysis*](unit-of-analysis.html).

But gridding is what many workflows are used to, so the cost of doing it
anyway was measured rather than argued: the same twelve-species datasets were
refitted with every location snapped to the centre of its grid cell and the
habitat value re-read there — which is what gridding does to your data. Six
datasets per cell size:

```{r gridding}
gr <- readRDS("../../dev/isdm-precision/cawa12-gridding.rds")$summary
data.frame(cell = c("exact points", "6 x 11 km", "12 x 22 km", "25 x 44 km"),
           mean_cor = gr$cor.1, sd_cor = gr$cor.2,
           mean_abs_err = gr$mae.1, sd = gr$mae.2)
```

Two things to read off it.

**Cells of typical atlas sizes cost nothing here.** At 6 × 11 km and even
12 × 22 km, slope recovery is indistinguishable from exact coordinates.

**The damage starts when displacement approaches the habitat's own scale —
and it looks like shrinkage, not scrambling.** At 25 × 44 km cells a record
can move ~25 km from where it was, the typical slope error rises from
`r gr$mae.1[1]` to `r gr$mae.1[4]` — about
`r round(100 * (gr$mae.1[4] / gr$mae.1[1] - 1))`% worse — while the species
*ranking* barely moves (correlation `r gr$cor.1[4]`). Gridding too coarsely
does not make your results look wrong; it quietly pulls the habitat effects
toward zero. That is the same positional-error mechanism the companion
article
[*When one data source knows where it is and the other does not*](isdm-spatial-precision.html)
measures in depth, and the same number decides it: **displacement relative to
the distance over which your habitat layer changes**. This landscape's
habitat changes over ~100 km, so small cells are harmless; on a patchy
10-km-scale habitat layer the free zone would shrink accordingly — compute
the ratio for your own raster with that article's recipe rather than reusing
these cell sizes.

One honest limit: this measured only the *positional* cost of gridding — the
records moved, but each site kept its own row. Pooling many checklists into
one cell row (summing counts and effort) changes the data in additional ways
not measured here. And the choice of cell size is a genuine model choice:
analyses of the same data on different grids can disagree with each other,
which is a documented problem in spatial statistics, not a rounding
curiosity.

# Before you use this on real data

**This simulation lives in the friendly regime, and that is the biggest
caveat on the page.** Two things are true here by construction and are false
almost everywhere else. First, the latent surfaces are residualised against
`env`, so the environmental covariate and the unmeasured spatial structure are
uncorrelated; on real data habitat and unmeasured spatial structure are
correlated in essentially every dataset, and the price of getting that wrong
is not small -- in this same simulator, removing the residualisation drives
the estimated slopes to about +21 against a true mean near +1.3. How the model
behaves under realistic covariate-field correlation is measured nowhere on
this page, and it is the first thing to establish before trusting a slope from
real data. Second, the opportunistic arm's bias is a single constant, so its
sampling is spatially uniform: **preferential sampling** -- birders on roads,
near towns and trailheads -- is untested here, and it is the failure mode this
literature worries about most.

**Nothing here shows that integrating helped.** This page never fits the
opportunistic arm alone, the survey alone, and the two together and compares
them. That comparison is the standard one in this field, and it is the obvious
next thing to run on your own data.

**The slopes come back shrunk.** Every recovery table on this page shows a
consistent downward pull. Do not describe estimates from this machinery as
unbiased, and do not build an effect-size claim on their magnitude without
running your own recovery check.

**The map carries no uncertainty at all.** Not a confidence interval, not a
standard error, not a coefficient of variation. A smooth colourful surface
implies precision it does not have, and this one has none to report.

**Your survey coordinates are probably fuzzed, and that matters.** Real
structured-survey data frequently does not know exactly where it is -- public
ABMI locations are displaced by kilometres -- and the consequences are not
small. Read
[*When one data source knows where it is and the other does not*](isdm-spatial-precision.html)
before fusing real fuzzed data. In particular, do **not** expect coarsening
your raster to fix it: Gábor et al. (2022) is the direct test of that
intuition, and the intuition does not survive it.

**The specific numbers are properties of this design.** AIC's 10 of 10, the
loading rule's 1 of 10, the overdispersion results: all measured at twelve
species, one landscape, one true number of gradients, moderate counts, and a
small mesh. The *mechanisms* generalise; the hit rates do not.

# Recap

`isdm_sources()` names the two observation processes. The arms sit at disjoint
locations and still share one field, because the field lives on a mesh built
over the union of both and every row is projected onto it. Project your
coordinates before meshing. Check `iterations`, not just `convergence`, and
keep your counts well under the ceiling. Choose the number of shared gradients
by AIC, not by hunting for a small loading column. Consider `unique = TRUE`
when a count arm may be overdispersed -- in this simulation it clearly helped
the point estimates, by a mechanism that looks more like regularisation than
like estimating the overdispersion, so do not read `Psi` as a measurement and
do not expect your intervals to improve. Zero the offset on the prediction
grid, keep the covariate on its training scale, and hatch whatever falls
outside the mesh.

Point estimates and a surface. No intervals -- and this page says so, rather
than letting a smooth picture imply otherwise.

**A first-time reader can stop here.** What follows is reference material:
four behaviours to know about on the day something does not work, and the
audit trail behind every number above.

# Reference: four things that do not do what you would guess

Each of these is an **algebraic identity**, not a tendency measured on a
sample: it holds exactly, for the reason given, and no number of replicates
would make it more or less true. They are collected here because each has cost
somebody an afternoon.

They are demonstrated on a deliberately small model -- three of the twelve
species -- because the size of a fit cannot change an identity.

```{r small}
sub <- droplevels(dat[dat$trait %in% c("CAWA", "OVEN", "TEWA"), ])
fam <- isdm_sources(ebird = poisson(), abmi = binomial(link = "cloglog"))
msh <- make_mesh(sub, c("X", "Y"), cutoff = 40)
FIT <- function(f, ...) suppressMessages(suppressWarnings(gllvmTMB(
  f, data = sub, trait = "trait", unit = "cell_id", family = fam,
  silent = TRUE, ...)))
```

## 1. An offset that never varies within an arm does nothing

An `offset()` is a known multiplier with no estimated coefficient -- data, not
a parameter -- so it cannot compete with anything for fit. If it never varies
*within* an arm, the arm's own intercept reproduces it exactly, and adding it
changes nothing:

```{r offset-absorbed}
sub$const_off <- ifelse(sub$isdm_source == "ebird", 1.0, 0)   # constant per arm
o_none  <- FIT(value ~ 0 + trait + trait:env + isdm_source)$opt$objective
o_const <- FIT(value ~ 0 + trait + trait:env + isdm_source +
                 offset(const_off))$opt$objective
o_vary  <- FIT(value ~ 0 + trait + trait:env + isdm_source +
                 offset(log_effort))$opt$objective
c(no_offset = o_none, constant_offset = o_const, varying_offset = o_vary)
cat("relative difference, constant offset vs none:",
    format((o_const - o_none) / o_none, digits = 3), "\n")
```

The first two are the same model to well inside the optimiser's own tolerance.
The real `log_effort` offset varies *within* each arm, which is why removing
it does change the objective. If your own effort is one number per source,
your offset is a relabelling of your source indicator, and the offset idiom is
not doing anything for you.

## 2. The unit column of a prediction grid, and when it stops being inert

The design matrix requires *some* valid factor level in the unit column. In
this model the only random effect is the spatial field, keyed on `(X, Y)`, so
the choice is arbitrary -- provably so:

```{r unit-column}
alt <- newd; alt$cell_id <- factor("sv060", levels = levels(dat$cell_id))
p_alt <- suppressWarnings(suppressMessages(
  predict(fit, newdata = alt, type = "response")))
cat("max |difference| between the two unit levels:",
    max(abs(p$est - p_alt$est)),
    "\nover a prediction range of", f2(diff(range(p$est)), 2), "\n")
```

Exactly zero -- an opportunistic cell's label and a survey station's give the
same answer. **This stops being true the moment your model has a genuine
unit-level random effect.** Then the label you wrote on the grid is used:
`predict()` re-adds the matching random effect and tells you it has done so.

```{r unit-re-message}
set.seed(3)
n <- 150; ev <- rnorm(n); site_eff <- rnorm(n, 0, 0.8)
toy <- do.call(rbind, lapply(1:2, function(j) data.frame(
  site = factor(seq_len(n)), env = ev, trait = c("a", "b")[j],
  value = rpois(n, exp(0.5 + c(0.9, -0.4)[j] * ev + site_eff)))))
toy$trait <- factor(toy$trait)
toy_fit <- suppressWarnings(suppressMessages(gllvmTMB(
  value ~ 0 + trait + trait:env + (1 | site), data = toy, trait = "trait",
  unit = "site", family = poisson(), silent = TRUE)))
toy_grid <- expand.grid(env = seq(-2, 2, length.out = 10),
                        trait = levels(toy$trait))
toy_grid$value <- 0
g1 <- transform(toy_grid, site = factor("1",  levels = levels(toy$site)))
g2 <- transform(toy_grid, site = factor("77", levels = levels(toy$site)))
cat(msg_of(predict(toy_fit, newdata = g1, type = "link")))

q1 <- suppressMessages(predict(toy_fit, newdata = g1, type = "link"))
q2 <- suppressMessages(predict(toy_fit, newdata = g2, type = "link"))
cat("\nsite \"1\" vs site \"77\":", f2(max(abs(q1$est - q2$est)), 3),
    "on the link scale\n")
```

So pick that level deliberately, or ask for a prediction with no site effect
in it at all using `re_form = ~0`. **Neither is what this article's map uses**,
and on purpose: `re_form = ~0` would zero the spatial field too, and that
field is the entire point of the map.

## 3. The name after the `|` in a spatial term is ignored

There is no column called `coords` in the data, and there does not need to be.
The geometry comes entirely from `mesh =`; the token to the right of the `|`
is never read.

```{r token-inert}
sp_obj <- function(tok) FIT(stats::as.formula(paste0(
  "value ~ 0 + trait + trait:env + isdm_source + offset(log_effort) + ",
  "spatial_latent(0 + trait | ", tok, ", d = 1)")), mesh = msh)$opt$objective
o <- c(coords = sp_obj("coords"), banana = sp_obj("banana"),
       cell_id = sp_obj("cell_id"))
o
cat("largest difference:", max(o) - min(o), "\n")
```

`cell_id` is a real column with a real meaning, and it makes no difference
whatsoever -- which is the direction that matters. Writing
`spatial_latent(0 + trait | site)` with a real `site` column and a real
intention will *also* be ignored, and nothing will say so.

## 4. A mesh passed without a spatial term does nothing

`mesh =` is consumed by the `spatial_*()` term and by nothing else. With no
such term in the formula, a mesh built with `make_mesh()` changes nothing --
the fit runs, converges, and is entirely non-spatial:

```{r mesh-ignored}
m <- c(no_mesh   = FIT(value ~ 0 + trait + trait:env + isdm_source +
                         offset(log_effort))$opt$objective,
       with_mesh = FIT(value ~ 0 + trait + trait:env + isdm_source +
                         offset(log_effort), mesh = msh)$opt$objective)
m
cat("difference:", abs(diff(m)), "\n")
```

The package gained a warning for this case (issue
[#1165](https://github.com/itchyshin/gllvmTMB/issues/1165)). Whether it reaches
you depends on the path -- when this page was built, the integrated fit above
produced no warning:

```{r mesh-ignored-warn}
cat(warn_of(gllvmTMB(
  value ~ 0 + trait + trait:env + isdm_source + offset(log_effort),
  data = sub, trait = "trait", unit = "cell_id", family = fam,
  silent = TRUE, mesh = msh)))
```

So check that the `spatial_*()` term is in your formula rather than trusting a
complaint to arrive. The objectives above are the reliable evidence: identical
to the last decimal means the mesh did nothing.

The *other* raw-mesh case is loud and correct -- a raw `fm_mesh_2d()` object
*with* a spatial term is refused, which matters because most `fmesher` and
INLA tutorials teach you to build one by hand:

```{r raw-mesh-build}
raw_bnd  <- fmesher::fm_extensions(cbind(sub$X, sub$Y), convex = c(30, 60))
raw_mesh <- fmesher::fm_mesh_2d(loc = cbind(sub$X, sub$Y), boundary = raw_bnd,
                                max.edge = c(40, 120))
```

```{r raw-mesh-error}
cat(err_of(FIT(value ~ 0 + trait + trait:env + isdm_source + offset(log_effort) +
                 spatial_latent(0 + trait | coords, d = 1), mesh = raw_mesh)))
```

# For the record

This section is the audit trail: the campaigns behind the numbers above, the
measurements that did not make the main path, and the places where the
evidence is weaker than a plain sentence would suggest. It is here so the
claims can be checked, not because a first-time reader needs it.

## Where the numbers come from

One fit runs live on this page. Everything replicated is read from a saved
campaign:

```{r provenance, echo = FALSE}
data.frame(
  campaign = c("intensity ceiling + recovery envelope", "rank selection",
               "overdispersion (unique = TRUE)", "presence-only quadrature",
               "map surfaces"),
  fits = c(nrow(pr$sweep) + nrow(pr$timing) + sum(pr$sweep_summary$n_rep),
           nrow(ai$fits), 2L * us$n_seeds_fitted,
           nrow(qd$replicate_fit_health) +
             nrow(qd$sweep_decoupled$long) / nrow(qd$replicates) +
             nrow(qd$sweep_coincident$long) / nrow(qd$replicates),
           length(md$ok_seeds)),
  file = c("cawa12-probe.rds", "cawa12-aic.rds", "cawa12-unique.rds",
           "cawa12-quadrature.rds", "cawa12-mapdata.rds"),
  row.names = NULL)
```

Each has its generating script beside it in `dev/isdm-precision/`
(`generate-<name>.R`), and each ran against gllvmTMB `r ai$package_version`.

## How the simulated field was built

Both latent surfaces are residualised against `env` at the sampled locations,
and the second against the first, so that the covariate and the latent field
are not two spatial terms fighting over one pool of variance. That step is not
a subtlety: skipping it drives the estimated slopes to about +21 against a
true mean near +1.3. It is also the assumption that makes this simulation
friendlier than real data, as [Before you use this on real data] says.

## The full count-ceiling sweep

Fourteen rungs of the opportunistic arm's overall intensity, three seeds each,
every rung unanimous across its three seeds:

```{r ceiling-full, echo = FALSE}
agg
```

At $\kappa = `r f2(pr$envelope$last_healthy_kappa, 1)`$ all three seeds fit
healthily; at `r f2(pr$envelope$first_failing_kappa, 1)` all three fail after
about a dozen iterations at an objective near 1e14. At
`r f2(pr$envelope$silent_1iter_kappa, 1)` the failure goes **silent** -- one
iteration, objective near 1e24, `convergence == 0`. Beyond that the optimiser
errors outright.

Recovery of the environmental slopes, four rungs, 12 replicate seeds each, all
accepted:

```{r recovery-envelope, echo = FALSE}
data.frame(kappa = ss$kappa,
           mean_cell_count = round(sapply(ss$kappa, function(k)
             mean(sw$mean_count[sw$kappa == k])), 2),
           accepted        = paste0(ss$n_ok, "/", ss$n_rep),
           cor_est_truth   = paste0(f2(ss$mean_cor, 3),
                                    " (sd ", f2(ss$sd_cor, 3), ")"),
           mean_abs_err    = f2(ss$mean_abs_err, 3),
           mean_signed_err = f2(ss$mean_signed_err, 3))
```

Accuracy is already decaying two rungs before the optimiser breaks, while
every fit still reports itself healthy. The numerical check tells you when to
stop trusting the optimiser; nothing tells you when to stop trusting the
estimate except a recovery check of your own.
$\kappa = `r f2(pr$envelope$chosen_kappa, 1)`$ was chosen on this evidence, at
a median `r f2(pr$envelope$secs_per_fit_median_serial, 1)` s per fit
(`r f2(pr$envelope$secs_per_fit_range_serial[1], 1)`--`r f2(pr$envelope$secs_per_fit_range_serial[2], 1)` s
across five serial fits on a shared machine).

## The near-zero-loading rule, in full

Made falsifiable as *"a column is spent when its root-mean-square falls below
`thr` times the largest column's"*, and applied to the same ten replicates at
two fitted ranks:

```{r tell-sweep, echo = FALSE}
data.frame(fitted_rank = ts$fitted_k, threshold = ts$thr,
           by_columns = paste0(ts$n_correct_colrms, "/", ts$n),
           by_singular_values = paste0(ts$n_correct_sv, "/", ts$n),
           modal_rank = ts$modal_rank_colrms)
```

Reading the columns gives the right answer in 1 of 10 at every threshold from
0.005 to 0.10, with a modal answer of 3: the heuristic over-counts by one, and
loosening the threshold does not rescue it. The rotation-invariant version --
threshold the singular values instead -- reaches
`r ts$n_correct_sv[ts$thr == 0.10 & ts$fitted_k == 3][1]` of 10 at a 0.10 cut
and 1 of 10 at every tighter cut.

Fitted with four gradients, the smallest singular value drops below a
millionth of the leading one in `r sum(nz4$sv_rel_4 < 1e-6)` of
`r nrow(nz4)` replicates -- so one direction does collapse -- but with three
gradients the third singular value still carries
`r f2(100 * min(sv3_nonzero), 0)` to `r f2(100 * max(sv3_nonzero), 0)` per
cent of the leading one in `r length(sv3_nonzero)` of `r nrow(nz3)`
replicates.

Degrees of freedom rise by `r ff$df[ff$k == 2][1] - ff$df[ff$k == 1][1]`,
`r ff$df[ff$k == 3][1] - ff$df[ff$k == 2][1]` and
`r ff$df[ff$k == 4][1] - ff$df[ff$k == 3][1]` across `d = 1 -> 2 -> 3 -> 4`.
Past the true number the likelihood is essentially flat: across the ten
replicates the third column buys at most `r f2(max(nll_gain[, 1]), 2)` units
of log-likelihood against a 20-AIC charge, and the fourth at most
`r f2(max(nll_gain[, 2]), 3)` against an 18-AIC charge. Two caveats on the
comparison. Refitting one dataset at `d = 2` and `d = 3` under eight different
RNG states gave an AIC range of exactly
`r f2(ai$optimiser_noise_summary$aic_range[1], 3)` -- which establishes that
these fits do not depend on the RNG state, and nothing more; the starts are a
deterministic function of the data, so it does not rule out all ten replicates
sharing one local optimum. And at a superfluous rank the extra loading column
sits on the boundary of its parameter space, where AIC's nominal penalty is
conservative -- in the direction that flatters the 10-of-10 result.

## The overdispersion campaign

Overdispersion was added to the opportunistic arm only, as a species-specific
observation-level effect $u_{ij}\sim N(-\sigma_j^2/2,\ \sigma_j^2)$ on the log
scale -- mean-centred on the intensity scale, so that $\sigma_j$ alone is
manipulated and not abundance.

```{r unique-design, echo = FALSE}
h <- us$health
data.frame(
  quantity = c("seeds offered", "screened out on the count cap",
               "seeds fitted", "base model accepted", "unique model accepted",
               "healthy pairs", "pairs with SEs on both models",
               "pairs with a usable curvature matrix (unique)",
               "median seconds, base", "median seconds, unique"),
  value = c(as.character(c(length(un$design$seeds_offered),
            length(un$design$seeds_screened_out),
            us$n_seeds_fitted, sum(h$base_ok), sum(h$uniq_ok),
            us$n_pairs_healthy, us$n_pairs_se_complete,
            us$robustness$pdHess$n_pdHess_true)),
            f2(us$timing$base_median_secs, 1),
            f2(us$timing$uniq_median_secs, 1)))
```

Any seed whose largest opportunistic count reached `r un$design$count_cap` was
rejected **before** fitting, because that is the ceiling from the first
section. Iteration counts on the unique model run in the hundreds, so no
silent failure occurred -- the rejected fits failed loudly.

Restricting the comparison to the
`r us$robustness$pdHess$n_pdHess_true` pairs with a well-behaved curvature
matrix reproduces everything: SE ratio
`r f2(us$robustness$pdHess$se_ratio_median, 3)`, coverage
`r f2(us$robustness$pdHess$coverage_base, 3)` to
`r f2(us$robustness$pdHess$coverage_uniq, 3)`, RMSE
`r f2(us$robustness$pdHess$rmse_base, 3)` to
`r f2(us$robustness$pdHess$rmse_uniq, 3)`.

Three caveats on the analysis itself.

**Do not test that log-likelihood difference against a $\chi^2$.** Under the
null the extra variance parameters sit on the boundary of their range (they
cannot be negative), and the usual reference distribution does not apply
there; the correct one is a mixture. The AIC comparison in the main path
avoids the question.

**`Psi` is not an i.i.d. observation-level tier.** It is a per-species
*spatially structured* field sharing the base precision of the shared field
(`src/gllvmTMB.cpp`), so it can only partly represent i.i.d. overdispersion --
which is the likeliest explanation for the attenuation above. Whether a
genuine observation-level tier, shaped like
`latent(0 + trait | cell_id, unique = TRUE)`, would recover $\sigma^2$ at the
right magnitude is the honest next measurement, and this campaign did not run
it.

**Nothing here was tested above the count screen**, which is exactly where
detection of `Psi` would improve. The count ceiling forbids looking.

## Presence-only: power, and how dense the grid must be

```{r po-recovery, echo = FALSE}
data.frame(species = rp$species, truth = round(rp$truth, 3),
           mean_est = round(rp$mean, 3), bias = round(rp$bias, 4),
           t = round(rp$t, 2), p = round(rp$p, 3))
```

The smallest *p* is `r f2(min(rp$p), 3)`, uncorrected, and every 95 per cent
interval contains the truth. The pooled test -- a per-seed regression through
the origin of the twelve estimates on the twelve truths, which equals 1 under
no uniform bias -- gives `r f2(qd$pooled_power$attenuation_mean, 5)`, 95 per
cent CI [`r f2(qd$pooled_power$ci[1], 4)`, `r f2(qd$pooled_power$ci[2], 4)`],
*p* = `r f2(qd$pooled_power$p, 3)`. That design would have detected a uniform
multiplicative distortion of `r f2(qd$pooled_power$mde_80_pct, 1)` per cent or
more, and detected nothing.

Two things it does not establish. First, **it certifies the device and the
optimiser, not the quadrature**: at these settings the quadrature grid
coincides with the lattice the point pattern was thinned from, so the
objective there is the exact likelihood of the generating model -- verified
against a direct `glm()` on the lattice counts to a maximum absolute
difference of
`r format(max(sapply(qd$exactness, function(z) z$max_abs_diff)), digits = 2)`.
Second, **the smallest bias this design could have detected** is
`r f2(min(rp$mde_80), 3)` to `r f2(max(rp$mde_80), 3)` on the slope scale --
`r f2(min(rp$mde_80_pct[big_beta]), 1)` to
`r f2(max(rp$mde_80_pct[big_beta]), 1)` per cent of truth for the
`r sum(big_beta)` species with `|beta| > 0.4`, but
`r f2(min(rp$mde_80_pct[!big_beta]), 0)` to
`r f2(max(rp$mde_80_pct[!big_beta]), 0)` per cent for the `r sum(!big_beta)`
whose true slope sits near zero, which is to say no power there at all. Quote
absolute numbers for those species, not percentages.

Grid density, measured with the grids **decoupled** from the generating
lattice so that what is measured is genuine quadrature error; 12 seeds per
rung:

```{r po-density, echo = FALSE}
data.frame(nodes = sd_$n_nodes,
           nodes_per_presence = paste0(round(npp$min, 1), "-", round(npp$max, 1)),
           mean_bias = round(sd_$mean_bias, 4),
           shift_vs_finest = round(sd_$mean_abs_shift_vs_finest, 4),
           pct_of_MC_sd = paste0(round(100 * sd_$mean_abs_shift_vs_finest / mc), "%"))
```

```{r quad-exponent}
## Fit the shift against node count on the log scale, and report the interval
## rather than the point -- with six rungs the slope is not well determined.
ex <- lm(log(sd_$mean_abs_shift_vs_finest[-nrow(sd_)]) ~ log(sd_$n_nodes[-nrow(sd_)]))
c(exponent = round(unname(coef(ex)[2]), 3),
  round(setNames(confint(ex)[2, ], c("lo", "hi")), 3))
```

Estimates approach the truth from below, monotonically. How fast is not
something six rungs can pin down -- the interval above includes -1 -- so do
not plan around a rate. Judged against the only scale that matters to a reader
(the Monte Carlo scatter of a single estimate, `r f2(mc, 4)`) they are stable
for practical purposes from about
`r sd_$n_nodes[which(sd_$mean_abs_shift_vs_finest / mc < 0.10)[1]]` nodes.
"Nodes per presence" is the right currency and it varies across species --
from `r f2(npp$min[npp$n_nodes == 3600], 1)` to
`r f2(npp$max[npp$n_nodes == 3600], 1)` per presence at 3,600 nodes here,
because rare species get more nodes per presence than common ones. And the
sweep never flattens: "increase the density until the estimates stop moving"
really means *until they move less than your Monte Carlo noise*.

## Projection: the likelihood has an opinion, and it is not the one you want

At this latitude a degree of latitude is about 1.7 times a degree of longitude
in kilometres, so a mesh built on `lon`/`lat` imposes a correlation structure
stretched by that factor in one direction. The instinct is to check the choice
by fitting both and comparing. That does not work.

```{r projection-null}
pj <- readRDS("../../dev/isdm-precision/cawa12-projection.rds")
data.frame(seed        = pj$per_seed$seed,
           d_objective = round(pj$per_seed$d_obj, 2),
           err_lonlat  = round(pj$per_seed$err_lonlat, 4),
           err_utm     = round(pj$per_seed$err_utm, 4))
c(mean_d_obj = round(pj$mean_d_obj, 2), se = round(pj$se_d_obj, 2),
  sign_flips = pj$sign_flips, utm_better = pj$utm_better_n, seeds = pj$n_seeds)
```

Six seeds, the same twelve-species design as everything else on this page, and
the two meshes matched at `r pj$per_seed$nodes_lonlat[1]` and
`r pj$per_seed$nodes_utm[1]` nodes so neither is handed extra latent
dimensions. The likelihood has a strong, consistent opinion -- the unprojected
mesh wins by `r abs(round(pj$mean_d_obj, 1))` objective units, SE
`r round(pj$se_d_obj, 1)`, same sign in all `r pj$n_seeds` seeds -- while on
accuracy six seeds cannot tell: the projected mesh recovers the slopes better
in `r pj$utm_better_n` of `r pj$n_seeds` seeds, mean absolute errors
`r round(pj$mean_err_lonlat, 3)` and `r round(pj$mean_err_utm, 3)` (paired
*t* = 0.78, *p* = 0.47). That comparison could only have detected a gap of
about 0.065, some 30 per cent of the error level itself, so it is underpowered
rather than a demonstrated tie.

And here the worked example undercuts the lesson, which is worth saying rather
than hiding. The latent surfaces in this simulation are built as functions of
**longitude and latitude**, not of projected kilometres, and `f1`'s
east-to-north stretch measures about 1.0 in degrees and 1.7 in kilometres. So
for *this* generating process the field really is closer to isotropic in
degrees, the unprojected mesh really is the better-specified model, and the
likelihood is not being fooled -- it is preferring the model that matches the
truth.

- **What is demonstrated.** The likelihood tracks the geometry of the truth.
  Holding both meshes fixed and varying only how stretched the simulated field
  is in kilometres, the gap moves with it: about -35 objective units at a
  km-stretch of 1.81, -25 at 1.37, and only -8 at 1.18, where the projected
  mesh starts winning seeds. Monotone, *r* = -0.93. With the meshes identical
  across those runs, nothing about the discretisation can explain it.
- **What is not demonstrated.** That the likelihood is blind to a projection
  error. It is not blind here; it is right.
- **Why you still cannot use it to choose.** The likelihood answers *which
  mesh matches my field's stretch* -- not *which coordinate system is
  physically correct*. Those coincide only when you already know the stretch,
  which is the thing you are trying to model.

Project first because degrees are not a distance -- a degree of longitude is
`r round(111*cos(56*pi/180))` km at this latitude against 111 km for a degree
of latitude. The likelihood will express a firm view about your mesh, and that
view is informative about your field's shape rather than about your
projection. Do not read one as the other.

`add_utm_columns()` warns rather than fails when a domain straddles a UTM zone
boundary, a real risk for any study area wider than six degrees:

```{r zone-warning}
straddling <- data.frame(lon = c(-114.5, -112.5), lat = c(53.6, 55.4))
cat(warn_of(add_utm_columns(straddling, ll_names = c("lon", "lat"))))
cat("\nUTM zones spanned by this article's data:",
    length(unique(floor((cells$lon + 180) / 6) + 1)), "\n")
```

## The bias question, and a correction

A separate 30-replicate, 12-species recovery campaign against this same
machinery reported a positive mean error at *t* = 4.17, and an earlier version
of this article called it "small but statistically real". It is not: that test
treated 360 errors as independent when all twelve species within a replicate
share one estimated field.

```{r bias-clustering}
err      <- gt$results$err                  # 30 replicates x 12 species
rep_mean <- tapply(err, gt$results$rep, mean)
naive <- t.test(err); clustered <- t.test(rep_mean)
data.frame(
  treating = c("360 errors as independent", "30 replicate means"),
  mean = c(f2(mean(err), 4), f2(mean(rep_mean), 4)),
  t    = c(f2(naive$statistic, 2), f2(clustered$statistic, 2)),
  p    = c(format(naive$p.value, digits = 2), f2(clustered$p.value, 3)))
cat("SD of replicate means:", f2(sd(rep_mean), 4),
    "  if species were independent:", f2(sd(err) / sqrt(12), 4), "\n")
```

The correct test does not reject. The signed bias is also a property of
**one** landscape and one draw of species truths, both fixed across those
replicates, so it has not been shown to be a mechanism rather than noise. What
remains true, and is visible in this article's own recovery table, is that the
slopes come back shrunk.

## Known issues behind this page

- The one-iteration silent failure at large counts:
  [#1167](https://github.com/itchyshin/gllvmTMB/issues/1167).
- The inert grouping token on a `spatial_*()` term:
  [#1163](https://github.com/itchyshin/gllvmTMB/issues/1163).
- A mesh silently ignored when no `spatial_*()` term is present:
  [#1165](https://github.com/itchyshin/gllvmTMB/issues/1165) — **fixed**; the
  package now warns, as shown in section 4 above.

## References

Gábor, L., Jetz, W., Lu, M., Rocchini, D., Cord, A. F., Malavasi, M.,
Zarzo-Arias, A., Barták, V., & Moudrý, V. (2022). Positional errors in species
distribution modelling are not overcome by the coarser grains of analysis.
*Methods in Ecology and Evolution*, **13**, 2289--2302.

Renner, I. W., Elith, J., Baddeley, A., Fithian, W., Hastie, T., Phillips,
S. J., Popovic, G., & Warton, D. I. (2015). Point process models for
presence-only analysis. *Methods in Ecology and Evolution*, **6**, 366--379.

Sólymos, P., Matsuoka, S. M., Bayne, E. M., Lele, S. R., Fontaine, P.,
Cumming, S. G., Stralberg, D., Schmiegelow, F. K. A., & Song, S. J. (2013).
Calibrating indices of avian density from non-standardized survey data: making
the most of a messy situation. *Methods in Ecology and Evolution*, **4**,
1047--1058.

Warton, D. I., & Shepherd, L. C. (2010). Poisson point process models solve
the "pseudo-absence problem" for presence-only data in ecology. *Annals of
Applied Statistics*, **4**, 1383--1402.
