---
title: "An integrated SDM end to end, and the map at the end of it"
description: "A two-arm integrated SDM for a boreal warbler: opportunistic counts plus a structured detection survey, a shared spatial field, and a relative-intensity map. Shows what the map does and does not carry, and where predict() will warn at you."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{An integrated SDM end to end, and the map at the end of it}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# What this shows

An integrated species distribution model fuses an opportunistic
presence-only arm with a structured survey arm, letting the structured arm
discipline the biased one. This article runs that end to end for a boreal
songbird system and finishes with the thing the whole exercise is for: a
relative-intensity map.

The map step is new. Until recently `predict(newdata = )` silently dropped
the spatial field, so a map drawn from it showed a field-free surface while
claiming the field. That is fixed, and this article is partly a
demonstration that it is.

## About the data — read this before reading anything else

**The data below are simulated.** They are shaped on a real study system —
Canada Warbler (*Cardellina canadensis*) and a co-occurring boreal warbler
in Alberta, an eBird-style opportunistic arm and an ABMI-style structured
detection arm — but no real records are fitted here.

Two reasons, both worth knowing if you plan the real analysis:

1. **The structured arm is not freely available at usable precision.** ABMI
   publishes public survey locations fuzzed to roughly 5.5 km, coarser for
   species at risk; precise coordinates need a separate data agreement. Why
   that matters enormously for an integrated model is the subject of a
   companion article — the short version is that it attenuates the
   environmental slope, and integrating a fuzzed arm can leave you further
   from the truth than discarding it.
2. **A vignette must not depend on the network.** The opportunistic arm
   *is* freely available (GBIF holds ~4,200 coordinate-bearing Alberta
   records for Canada Warbler, 98% of them from eBird), and `rgbif` will
   fetch them — but a documentation build that phones an API is a
   documentation build that breaks.

So: simulated, to the real system's dimensions. Every number below is
reproducible from this page.

# The two arms

```{r data}
suppressMessages(library(gllvmTMB))
set.seed(11)

n_cell <- 120L
cells <- paste0("c", seq_len(n_cell))
lon <- runif(n_cell, -120, -110)
lat <- runif(n_cell, 49, 60)

## A smooth habitat covariate, and a latent spatial field the covariate
## does not explain -- the thing the SPDE tier is there to capture.
env <- as.numeric(scale(sin(lon / 2) + cos(lat / 3)))
u <- as.numeric(scale(sin(lon) * cos(lat)))

spp <- c("cawa", "oven")
a0 <- c(-1.0, -0.6)     # species intercepts
bt <- c(0.8, 0.4)       # environmental slopes -- the estimand of interest
lm2 <- c(0.9, 0.5)      # loading on the shared spatial field

mk <- function(src, kind, effort) {
  do.call(rbind, lapply(seq_along(spp), function(j) {
    d <- data.frame(cell_id = cells, trait = spp[j], lon = lon, lat = lat,
                    env = env, isdm_source = src, effort = effort,
                    stringsAsFactors = FALSE)
    eta <- a0[j] + bt[j] * env + lm2[j] * u + log(effort)
    d$value <- if (kind == "count") {
      rpois(n_cell, exp(eta))
    } else {
      rbinom(n_cell, 1, -expm1(-exp(eta)))
    }
    d
  }))
}

## Opportunistic arm: counts, high effort. Structured arm: detection /
## non-detection, lower effort. Both see the SAME underlying intensity.
dat <- rbind(mk("ebird", "count", 3.0), mk("survey", "pa", 0.8))
dat$trait <- factor(dat$trait)
dat$cell_id <- factor(dat$cell_id)
dat$isdm_source <- factor(dat$isdm_source, levels = c("ebird", "survey"))
dat$log_effort <- log(dat$effort)
str(dat[, c("cell_id", "trait", "isdm_source", "value", "env")])
```

`isdm_sources()` is what tells the model the two arms are different
observation processes on one shared intensity: a Poisson likelihood for the
counts, a complementary-log-log Bernoulli for detection/non-detection.

```{r fit}
mesh <- make_mesh(dat, c("lon", "lat"), cutoff = 0.8)
fam <- isdm_sources(ebird = poisson(), survey = binomial(link = "cloglog"))

fit <- suppressMessages(suppressWarnings(gllvmTMB(
  value ~ 0 + trait + trait:env + isdm_source + offset(log_effort) +
    spatial_scalar(0 + trait | coords),
  data = dat, trait = "trait", unit = "cell_id",
  family = fam, mesh = mesh, silent = TRUE
)))
fit$opt$convergence
```

# The fitted output tells you which arm each row came from

```{r arm-column}
head(predict(fit), 4)
```

The `isdm_source` column matters more than it looks. On the response scale
`est` **mixes scales by construction** — the count arm returns an expected
count, the detection arm returns a probability — so without the arm label
the column is two different quantities interleaved, distinguishable only by
memory. `fitted()` carries it too.

# The map

Three things have to be right for a map to mean anything.

**The spatial field must actually be in the prediction.** It is; this is the
part that used to be silently dropped.

**The scale must be effort-free.** `type = "response"` includes the row's
offset, so it returns an expected count *at that effort* — fine for a
training row, wrong for a map. Setting the offset to zero in `newdata` gives
relative intensity, and it is exact rather than approximate because the
offset is re-evaluated against `newdata`.

**The grid must lie inside the mesh.** More on that below.

```{r map}
gx <- seq(min(lon), max(lon), length.out = 30)
gy <- seq(min(lat), max(lat), length.out = 30)
g0 <- expand.grid(lon = gx, lat = gy)
g0$env <- as.numeric(scale(sin(g0$lon / 2) + cos(g0$lat / 3)))

grid <- do.call(rbind, lapply(spp, function(sp) transform(g0, trait = sp)))
grid$trait <- factor(grid$trait, levels = levels(dat$trait))
grid$cell_id <- factor(cells[1], levels = levels(dat$cell_id))
grid$isdm_source <- factor("ebird", levels = levels(dat$isdm_source))
grid$log_effort <- 0     # relative intensity, NOT an expected count
grid$value <- 0          # placeholder: the design matrix evaluates the LHS

map <- suppressWarnings(suppressMessages(
  predict(fit, newdata = grid, type = "response")
))
range(map$est)
```

```{r map-plot, fig.cap = "Relative intensity for one species across the prediction grid, from the integrated fit. The offset is set to zero, so this is a relative intensity surface rather than an expected count at any particular effort. It carries NO uncertainty: there is no interval construction behind these values, and none should be inferred from the smoothness of the surface.", fig.alt = "A smoothly varying colour surface over a rectangular longitude-latitude grid, with higher values concentrated in a broad diagonal band."}
i <- map$trait == "cawa"
z <- matrix(map$est[i], nrow = length(gx), ncol = length(gy))
image(gx, gy, z, xlab = "longitude", ylab = "latitude",
      main = "Canada Warbler: relative intensity")
points(lon, lat, pch = 3, cex = 0.5)
```

## `predict()` will warn if your grid leaves the mesh

The chunk above is wrapped in `suppressWarnings()`, which is a cheat worth
exposing. Run it without:

```{r hull-warning}
w <- tryCatch(
  suppressMessages(predict(fit, newdata = grid, type = "response")),
  warning = function(w) class(w)[1]
)
w
```

A rectangular grid over a scattered set of survey points inevitably pokes
outside the mesh hull. Out there `fmesher` returns an all-zero basis row, so
the spatial field is **exactly zero** — indistinguishable, on the map, from
a region where the field is genuinely near zero. A blank corner reads as a
cold one.

`make_mesh()` rejects such rows at *fit* time, so the honest thing is for
`predict()` not to be quietly more permissive. Clip the grid to the meshed
domain, or extend the mesh to cover it — do not suppress the warning, which
is what this article did purely to keep the figure inline.

# What this map does not carry

Stated because a smooth colour surface is persuasive out of all proportion
to the evidence behind it:

- **No uncertainty, of any kind.** There is no interval on these values.
  `se.fit` is deliberately refused with `newdata`, and that refusal is load
  bearing: the available training-row standard error is conditional and
  fixed-effect-only, and a 1,600-fit measurement put its coverage of the
  true linear predictor at **0.23–0.82**, falling as the grid grows, because
  the latent-field reconstruction error it ignores is exactly what dominates
  on a map. Map-scale uncertainty needs an RE-aware construction that does
  not exist yet.
- **No claim about the environmental slopes' calibration beyond one
  measured grid.** The Wald intervals on `trait:env` were measured as
  calibrated (48/48 cell x species verdicts inside a [0.92, 0.98] band, at
  n = 2000 replicates) — but on the simulation grid that campaign ran, not
  here.
- **No positional-error correction.** Everything above assumes both arms
  know where they are. Real structured-survey data frequently does not, and
  the consequences are not small — see the companion article on
  spatial-precision mismatch before fusing real fuzzed data.
- **No claim that this taxon is settled.** The real Canada Warbler /
  ABMI pairing was assessed and came back **conditional**, pending exactly
  the coordinate-precision question above.

# Recap

`isdm_sources()` for the observation processes, a `spatial_*()` term for the
shared field, an offset for effort, and `predict(newdata = )` with the
offset zeroed for a relative-intensity surface. The arm column tells you
which scale each row is on; the hull warning tells you where the map stopped
being informed by data.

Point estimates and a surface — no intervals, and the article says so
rather than letting the picture imply otherwise.
