---
title: "When one data source knows where it is and the other does not"
description: "Monitoring programmes often publish survey coordinates deliberately blurred to protect sensitive sites, while an eBird or GBIF record carries a GPS fix. If you fuse the two, you are reading your habitat raster at the wrong place for one of the arms - and that drags the estimated habitat effect toward zero, by 57-79% in the blurred arm here. The combined fit inherits that in proportion to how much of the data the blurred arm supplies. Whether it matters for you comes down to one ratio you can compute from your own raster. And the answer flips - the blurred arm becomes worth keeping - when the surrogate you use to model where people record is both imperfectly measured and correlated with habitat, which is the usual situation with real data."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{When one data source knows where it is and the other does not}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

## The replicate-level result files are tracked in the repository and this
## article is built from them. If one is missing the build STOPS: a page that
## quietly substitutes stale hard-coded means for measurements is the exact
## failure mode this article is about.
f_arm  <- "../../dev/isdm-precision/arm-weighting-results.rds"
f_bpo  <- "../../dev/isdm-precision/biased-po-results.rds"
f_rev  <- "../../dev/isdm-precision/surrogate-error-results.rds"
f_grid <- "../../dev/isdm-precision/cor-length-grid.rds"
for (f in c(f_arm, f_bpo, f_rev, f_grid))
  if (!file.exists(f)) stop("missing evidence file: ", f,
    " -- re-run the matching dev/isdm-precision/generate-*.R script")
d_arm  <- readRDS(f_arm)
d_bpo  <- readRDS(f_bpo)
d_rev  <- readRDS(f_rev)
d_grid <- readRDS(f_grid)

## Every load-bearing number in the prose below is pulled from these objects
## by an inline expression, so no sentence can drift from the table above it.
wide <- function(s) {
  w <- reshape(s, idvar = "rep", timevar = "arm", direction = "wide")
  names(w) <- sub("beta_hat\\.", "", names(w)); w
}
cell <- function(np, ns, fz)
  wide(d_arm[d_arm$n_po == np & d_arm$n_survey == ns & d_arm$fuzz == fz, ])
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
```

# The situation

You have two kinds of data on the same species. One is a large, messy pile of
opportunistic records — [eBird](https://ebird.org) checklists,
[GBIF](https://www.gbif.org) (Global Biodiversity Information Facility)
occurrences — with no survey
design behind it. The other is a structured monitoring programme: fewer sites,
but a protocol. An integrated species distribution model (iSDM) fits both at
once, on the reasoning that the structured arm can discipline the biased one
(each data source inside an integrated model is called an **arm** -- the word
appears throughout these articles).

The two arms usually differ in one respect that nobody puts in the model. **The
opportunistic arm knows where it is; the survey arm often does not.**

This is not sloppiness — it is policy. Monitoring programmes routinely publish
coordinates degraded on purpose, to protect site access agreements and to avoid
disclosing the locations of species at risk. [ABMI](https://abmi.ca), the
Alberta Biodiversity Monitoring Institute, publishes public survey locations
**blurred to roughly 5.5 km**, and coarser still for sensitive species; precise coordinates require
a separate data agreement. An eBird checklist in the same landscape carries a
GPS fix. The two arms are then fused as though both located their observations
equally well, and one of them did not.

Throughout this article that deliberate blurring is called **fuzz**, and it is
measured not in kilometres but relative to the habitat layer — for reasons that
become the whole point below.

This is a documented problem, not a hypothetical one. Work on [iNaturalist](https://www.inaturalist.org)'s
obscured-geoprivacy setting reports that obscured locations change the
environmental values attached to a record — elevation, habitat type, extracted
covariates — and change the resulting estimated distributions (Koo et al.
2025). Gábor et al.
(2022) likewise find that SDM performance degrades as positional error in
occurrence records grows.

## Why the spatial random field does not rescue you

The reason fuzzed coordinates still feel harmless in an integrated model is a
sensible-sounding intuition: *there is a spatial random field in the model,
positional error is spatial noise, so the field will absorb it.* That intuition
is about the **response**. The damage is to the **covariate**.

Environmental predictors in an SDM are almost never measured at the site. They
are extracted from a raster at the recorded coordinate. If the recorded
coordinate is wrong by 5.5 km you read temperature, elevation or canopy cover
**from the wrong place**, and then regress the species' response — which
happened at the *true* location — on a habitat value from a different one.

The consequence is well known and has a name. When a predictor is measured with
error, the estimated slope on that predictor is dragged toward zero: the habitat
effect looks weaker than it really is. Statisticians call the situation
*errors-in-variables* and the symptom *attenuation*; both terms are used below,
and both mean what the previous sentence says. Nothing about the mechanism is
new here. What this article measures is what the mechanism does inside a
two-arm integrated fit, and what decides whether you should care.

That last part is not decided by the fuzz in kilometres. It is decided by the
fuzz **relative to how far you have to move before the habitat layer stops
looking like itself**. Five and a half kilometres is nearly nothing across a
smooth continental temperature gradient, and severe across a patchy
canopy-cover layer. So that ratio comes first, before any simulation.

# The number that decides it

The number is your survey arm's positional uncertainty divided by the
**correlation length** of the habitat layer you care about — the distance over
which values in that layer stay similar to each other.

Getting that number wrong by a factor of three is enough to flip the
recommendation, and a factor of three is exactly what sits between the two
quantities that software calls "range". So it is worth being exact about which
one is meant.

For a covariate whose spatial correlation behaves like `exp(-d / phi)`, `phi` is
the distance at which the covariate's correlation with itself falls to
**1/e ≈ 0.368**. That is what "correlation length" means here, and it is
computable directly from a raster. Software that fits a **variogram** -- the
standard geostatistical plot of how similarity between raster values decays
with distance -- commonly reports instead the *practical range*, conventionally about `3 * phi`, the distance at which
correlation falls to 0.05. **Dividing a reported practical range by 3 before
comparing it to your positional fuzz is the single most important arithmetic
step in applying this article to your own data.**

## The recipe

The function below measures `phi` from a set of cells. It works by building a
**correlogram** — the average correlation between pairs of cells, binned by how
far apart they are — and then reading off the distance at which that curve
crosses 1/e.

```{r cor-length-recipe}
## phi = the distance at which the covariate's own spatial correlation falls
## to 1/e. NOTE n_sub: this builds a dense n x n distance matrix, so it takes
## a SUBSAMPLE of cells, never a whole raster.
cor_length <- function(x, y, z, n_bin = 30, n_sub = 2000, seed = 1) {
  if (length(x) > n_sub) {
    ## Subsample reproducibly WITHOUT clobbering the caller's RNG stream: a
    ## bare set.seed() here would change every random draw you make
    ## afterwards, invisibly.
    old <- if (exists(".Random.seed", .GlobalEnv))
      get(".Random.seed", .GlobalEnv) else NULL
    set.seed(seed)
    i <- sample.int(length(x), n_sub)
    if (is.null(old)) rm(".Random.seed", envir = .GlobalEnv) else
      assign(".Random.seed", old, envir = .GlobalEnv)
    x <- x[i]; y <- y[i]; z <- z[i]
  }
  d   <- as.matrix(dist(cbind(x, y)))
  cp  <- outer(z - mean(z), z - mean(z)) / var(z)
  ut  <- upper.tri(d)
  brk <- seq(0, quantile(d[ut], 0.5), length.out = n_bin + 1)
  rho <- tapply(cp[ut], cut(d[ut], brk, include.lowest = TRUE), mean)
  h   <- (brk[-1] + brk[-length(brk)]) / 2
  ok  <- !is.na(rho); rho <- rho[ok]; h <- h[ok]

  ## Walk outward to the FIRST crossing of 1/e and interpolate there.
  ## approx(rho, h, xout = 1/e) would sort on the correlations, and a measured
  ## correlogram does not fall monotonically -- sorting silently re-pairs
  ## distances with correlations they did not come from.
  below <- which(rho < exp(-1))
  if (!length(below))
    stop("cor_length(): the correlogram never falls to 1/e within half the ",
         "maximum inter-point distance. phi is too long to measure on this ",
         "extent -- your covariate is smoother than your map is wide. ",
         "Use a larger region, and do not substitute the largest distance ",
         "you can see.", call. = FALSE)
  k <- below[1]
  if (k == 1L)
    stop("cor_length(): correlation is already below 1/e in the first ",
         "distance bin, so phi is shorter than the spacing between the cells ",
         "supplied. Supply finer cells; do NOT read the returned bin ",
         "midpoint as phi.", call. = FALSE)
  h[k - 1] + (rho[k - 1] - exp(-1)) / (rho[k - 1] - rho[k]) * (h[k] - h[k - 1])
}
```

**Do not hand it a whole raster, and do not coarsen one to make it fit.** Both
halves matter. The distance matrix grows with the square of the number of
cells, and that arithmetic is unforgiving: `n` cells cost `8 * n^2` bytes per
matrix, so 22,500 cells is `r fmt(8 * 22500^2 / 2^30, 1)` GB and a 1 km Alberta
raster (~10⁶ cells) would need `r fmt(8 * 1e6^2 / 2^40, 0)` TB. The `n_sub`
argument takes a random subsample instead, and a subsample is enough.
**`terra::aggregate()` is the wrong fix, and it fails in the dangerous
direction**: coarsening inflates the estimated `phi`, which *shrinks* your
fuzz-to-`phi` ratio and makes your situation look safer than it is. Subsample
cells; do not average them. (Coarsening is also not a remedy for the underlying
problem — see [What to do](#what-to-do).)

Here it is working on a surface whose `phi` is known by construction, and then
refusing to answer on one it cannot measure:

```{r cor-length-demo}
sim_surface <- function(n_side, phi, seed) {
  set.seed(seed)
  gx <- seq(0, 1, length.out = n_side)
  g  <- expand.grid(lon = gx, lat = gx)
  S  <- exp(-as.matrix(dist(g)) / phi)
  g$z <- as.numeric(scale(as.numeric(
    t(chol(S + diag(1e-6, nrow(S)))) %*% rnorm(nrow(g)))))
  g
}

g <- sim_surface(41, phi = 0.05, seed = 1)     # cell spacing 1/40 = 0.025
cor_length(g$lon, g$lat, g$z)                  # true phi = 0.05

g_coarse <- sim_surface(21, phi = 0.02, seed = 1)   # cell spacing 0.05 > phi
try(cor_length(g_coarse$lon, g_coarse$lat, g_coarse$z))
```

The second call is the important one. The cells are further apart than the
correlation length, so the correlogram is already below 1/e in its first bin and
there is nothing to interpolate. The function stops and says which of the two
ways it failed, rather than returning a number you would have no reason to
distrust.

## Where the recipe can be trusted

Two things can go wrong, and they are different: `phi` can be long relative to
your **map**, or short relative to your **cells**. The grid below crosses those
two so they can be told apart — twelve simulated surfaces per cell, each with a
`phi` that is known exactly.

```{r cor-length-grid, echo = FALSE}
## Drift guard: the grid was measured with the copy of the recipe stored in
## dev/isdm-precision/cor-length-recipe.R. If the function shipped above has
## since diverged from it, say so on the page rather than quietly reporting
## one function's accuracy as another's.
rec <- attr(d_grid, "recipe")
if (!is.null(rec) &&
    !identical(paste(deparse(cor_length), collapse = "\n"), rec))
  warning("The recipe shipped above differs from the one used to build this ",
          "table. Re-run dev/isdm-precision/generate-cor-length-grid.R.")

d_grid$ratio <- d_grid$phi_hat / d_grid$phi
gtab <- do.call(rbind, lapply(
  split(d_grid, list(d_grid$n_side, d_grid$phi), drop = TRUE), function(s)
    data.frame(
      `phi / map width`    = s$phi[1],
      `phi / cell spacing` = round(s$phi_over_spacing[1], 1),
      `refused (of 12)`    = sum(is.na(s$phi_hat)),
      `mean phi_hat / phi` = if (all(is.na(s$phi_hat))) NA else
                               mean(s$ratio, na.rm = TRUE),
      `sd over surfaces`   = if (sum(!is.na(s$phi_hat)) < 2) NA else
                               sd(s$ratio, na.rm = TRUE),
      `above truth`        = paste0(sum(s$ratio > 1, na.rm = TRUE), " / ",
                                    sum(!is.na(s$ratio))),
      check.names = FALSE)))
gtab <- gtab[order(gtab$`phi / map width`, gtab$`phi / cell spacing`), ]
knitr::kable(gtab, digits = 3, row.names = FALSE)
```

```{r cor-length-derived, include = FALSE}
## "The region where the recipe is recommended" = the operational advice
## given below: phi under a tenth of the map AND comfortably above the cell
## spacing. Both halves, or the summary silently includes cells the advice
## already tells the reader to stay out of.
benign <- d_grid[d_grid$phi <= 0.10 & d_grid$phi_over_spacing >= 2 &
                 !is.na(d_grid$phi_hat), ]
benign_above <- sum(benign$ratio > 1)
benign_n     <- nrow(benign)
sds <- tapply(benign$ratio, list(benign$n_side, benign$phi),
              function(z) sd(z, na.rm = TRUE))
sd_lo <- min(sds, na.rm = TRUE); sd_hi <- max(sds, na.rm = TRUE)
r040 <- mean(d_grid$ratio[d_grid$phi == 0.40], na.rm = TRUE)
r025 <- mean(d_grid$ratio[d_grid$phi == 0.25], na.rm = TRUE)
## The one cell that answers while barely resolving phi: phi/spacing = 1.2.
r_edge <- mean(d_grid$ratio[abs(d_grid$phi_over_spacing - 1.2) < 1e-8],
               na.rm = TRUE)
```

`mean phi_hat / phi` is the recipe's answer divided by the truth, so 1.00 is
perfect and anything below 1 means the recipe read short. Read down the table
and then across it.

**Down — the bias belongs to the map's width.** As `phi` grows relative to the
map, the recipe reads low: about `r fmt(r025, 2)` × truth at `phi` a quarter of
the map's width, `r fmt(r040, 2)` × at four tenths. Read *across* at a fixed
`phi / map width` and the three lattice resolutions give essentially the same
answer. That is what makes it a map-size effect and not a resolution one: a
correlation longer than your map cannot be seen on your map.

**Across — the resolution decides whether you get an answer at all.** Once
`phi` falls to about the spacing between your cells, the recipe refuses on
every surface or nearly every one. The row that answers at
`phi / cell spacing = 1.2` should be read as a warning rather than a success:
it returns a number, and the number is `r fmt(100 * (1 - r_edge), 0)`% low.

**The error is not reliably downward on a single raster.** The column means
above are means over twelve surfaces, and it is tempting to read "reads low" as
a guarantee for your one raster. It is not. Even in the region where this
recipe is recommended — `phi` at or under a tenth of the map and at least twice
the cell spacing — the surface-to-surface standard deviation of
`phi_hat / phi` is `r fmt(sd_lo, 2)`–`r fmt(sd_hi, 2)`, and
`r benign_above` of `r benign_n` individual surfaces
(`r fmt(100 * benign_above / benign_n, 0)`%) come out **above** truth. Run the
recipe on several covariate layers or several subsamples, and do not read the
third significant figure.

**So, operationally.** Trust it when `phi` sits comfortably below a tenth of
your study region's width and comfortably above your cell spacing. Outside that
window it either refuses or reads low. Reading low inflates your computed
`fuzz / phi` ratio, which for this article's decision errs toward looking more
carefully — but that is not a licence to treat the number as exact.

# What actually happens

## The design

Two species share a landscape with a single smooth habitat surface. Counts are
generated from the habitat value at each site's **true** location, with a shared
slope $\beta = 0.9$ — that is the number every estimate below is trying to
recover. The presence-only arm records its coordinates exactly. The survey arm's
recorded coordinates are displaced by noise whose standard deviation is `fuzz` ×
the correlation length, so `fuzz = 0.5` means "displaced by about half the
distance over which the habitat layer stays similar to itself".

The covariate handed to the model is then read at each arm's *recorded*
location, which is what an analyst would do. Three analyses of the identical
world are compared — `precise` (the presence-only arm alone), `fuzzed` (the
survey arm alone) and `integrated` (both together, the thing an iSDM does) — at
three fuzz levels and three arm-size designs, 15 replicate landscapes each.

Three features of the design worth knowing before reading the numbers:

- **No spatial field is fitted.** Fixed effects only. This isolates the
  covariate channel cleanly, and it means the article does **not** measure how
  much of the damage a fitted field would absorb — the intuition named at the
  top is set aside here, not tested.
- **The precise arm is generated before `fuzz` is used**, so its estimate is
  identical to the last digit across the three fuzz levels
  (`r fmt(mean(cell(100, 400, 0)$precise), 8)` at all three in the 100/400
  design). Its flat line in the figure is therefore built in, not a result. It
  buys perfect pairing between the three analyses, which is the point.
- **The three designs share their 15 landscapes** (the same replicate seeds).
  They are three designs on one set of landscapes, not 45 independent trials.

## The measurement

```{r arm-weighting-figure, echo = FALSE, fig.width = 9.5, fig.height = 4.0, dpi = 200, out.width = "100%", fig.cap = "The attenuation penalty across three arm-size designs, each n(precise) / n(fuzzed), ordered left to right by the FUZZED arm's share of the data: 400/100 (fuzzed arm one fifth), 220/220 (equal), 100/400 (fuzzed arm four fifths, a stress test). Pale symbols are the 105 distinct fits behind each design - 15 precise (generated before fuzz is applied, so the same 15 are drawn at each fuzz level), 45 fuzzed and 45 integrated. Solid symbols and lines are the design's arm means. The dashed grey line, labelled in the first panel, marks the true slope 0.9. In every design the precise arm sits at truth and the fuzzed arm collapses toward zero; the integrated fit falls between them at a depth set by how much of the pooled data the fuzzed arm supplies.", fig.alt = "Three side-by-side panels, one per arm-size design, each plotting estimated slope against positional fuzz (0, 0.5 and 1.0 times the environmental correlation length) for three analyses: precise, fuzzed and integrated. A dashed grey horizontal reference line, labelled truth, marks the true slope 0.9 in every panel. In all three panels the precise series stays flat near 0.9 and the fuzzed series falls steeply toward 0.2 by fuzz 1.0. The integrated series falls least in the 400-precise/100-fuzzed panel (to about 0.74 at fuzz 1.0), more in the 220/220 panel (to about 0.52), and most in the 100-precise/400-fuzzed panel (to about 0.34): the more of the pooled data the fuzzed arm supplies, the deeper the integrated series falls."}
arms <- c("precise", "fuzzed", "integrated")
col  <- c(precise = "#0072B2", fuzzed = "#D55E00", integrated = "#009E73")
pchv <- c(precise = 16, fuzzed = 17, integrated = 15)
ltyv <- c(precise = 1, fuzzed = 3, integrated = 2)
xlb  <- "positional fuzz  (× correlation length)"
fz   <- c(0, 0.5, 1.0)
off  <- c(precise = -0.025, fuzzed = 0, integrated = 0.025)
## Panel order = increasing share of the data supplied by the FUZZED arm,
## matching the tables, the prose, and the direction of the claim. Ordered
## any other way the integrated series goes down-up-down and the reader has
## to reassemble the dose-response by hand.
designs <- list(c(400, 100), c(220, 220), c(100, 400))
dtitle <- c("n(precise)=400, n(fuzzed)=100", "n(precise)=220, n(fuzzed)=220",
            "n(precise)=100, n(fuzzed)=400")
## ylim must CONTAIN the replicates, because the caption says it shows all of
## them. A fixed limit silently clips whatever a re-run pushes outside.
ylim_arm <- range(c(0, 1.05, d_arm$beta_hat))

op <- par(mfrow = c(1, 3), mar = c(3.4, 3.6, 2.2, 0.6),
          oma = c(2.0, 0, 1.6, 0),
          mgp = c(2.1, 0.6, 0), las = 1, cex.axis = 0.78, cex.lab = 0.82,
          tcl = -0.3)
set.seed(1)
for (k in seq_along(designs)) {
  plot(NA, xlim = c(-0.06, 1.06), ylim = ylim_arm, xaxt = "n",
       xlab = xlb, ylab = expression("estimated slope  " * hat(beta)))
  axis(1, at = fz, labels = format(fz))
  abline(h = 0.9, col = "grey45", lwd = 1.1, lty = 2)
  if (k == 1) text(1.045, 0.9, "truth", col = "grey35", cex = 0.62,
                   adj = c(1, -0.45))
  sub <- d_arm[d_arm$n_po == designs[[k]][1] &
               d_arm$n_survey == designs[[k]][2], ]
  mu <- tapply(sub$beta_hat, list(sub$arm, sub$fuzz), mean)[arms, ]
  for (a in arms) {
    s <- sub[sub$arm == a, ]
    points(s$fuzz + off[a] + runif(nrow(s), -0.010, 0.010), s$beta_hat,
           pch = pchv[a], cex = 0.42, col = adjustcolor(col[a], 0.30))
    lines(fz + off[a], mu[a, ], col = col[a], lwd = 2, lty = ltyv[a])
    points(fz + off[a], mu[a, ], pch = pchv[a], cex = 1.0, col = col[a])
  }
  mtext(dtitle[k], side = 3, line = 0.5, cex = 0.72, font = 2)
  if (k == 1) {
    ## seg.len: line type is the channel that carries this in greyscale, so
    ## the key must be long enough to tell dashed from dotted.
    legend("bottomleft", legend = arms, col = col[arms], pch = pchv[arms],
           lty = ltyv[arms], lwd = 1.6, bty = "n", cex = 0.72,
           seg.len = 3, text.col = col[arms])
  }
}
## The fence travels with the PNG, not with the caption: a figure lifted into
## a talk or a thesis loses its caption and keeps its pixels.
mtext("SIMULATED DATA — point estimates only; no interval or coverage claim",
      side = 1, outer = TRUE, line = 0.6, cex = 0.66, col = "grey35")
par(op)
```

The same numbers as a table, since a figure is read and a table is checked.
Mean estimated slope (true value **0.9**), with the arm-size-weighted average
of the two single-arm means alongside the integrated fit:

```{r arm-weighting-table, echo = FALSE}
agg <- aggregate(beta_hat ~ n_po + n_survey + fuzz + arm, d_arm, mean)
tab <- reshape(agg, idvar = c("n_po", "n_survey", "fuzz"),
               timevar = "arm", direction = "wide")
names(tab) <- sub("beta_hat\\.", "", names(tab))
tab$nwavg <- (tab$n_po * tab$precise + tab$n_survey * tab$fuzzed) /
             (tab$n_po + tab$n_survey)
tab <- tab[order(-tab$n_po, tab$fuzz),
           c("n_po", "n_survey", "fuzz", "precise", "fuzzed", "integrated",
             "nwavg")]
names(tab) <- c("n(precise)", "n(fuzzed)", "fuzz", "precise", "fuzzed",
                "integrated", "n-weighted avg")
knitr::kable(tab, digits = 3, row.names = FALSE)
```

```{r arm-weighting-derived, include = FALSE}
att <- agg[agg$arm == "fuzzed" & agg$fuzz > 0, ]
att_pct <- 100 * (1 - att$beta_hat / 0.9)
att_lo <- min(att_pct); att_hi <- max(att_pct)
att05 <- att_pct[att$fuzz == 0.5]; att10 <- att_pct[att$fuzz == 1]
nw_gap <- sapply(designs, function(de) sapply(c(0.5, 1.0), function(fv) {
  w  <- cell(de[1], de[2], fv)
  nw <- (de[1] * w$precise + de[2] * w$fuzzed) / sum(de)
  mean(abs(w$integrated - nw))
}))
pen <- sapply(designs, function(de) {
  w <- cell(de[1], de[2], 0.5); mean(w$integrated - w$precise) })
pen_ci <- sapply(designs, function(de) {
  w <- cell(de[1], de[2], 0.5); t.test(w$integrated - w$precise)$conf.int })
ctl <- d_arm[d_arm$n_po == 400 & d_arm$n_survey == 100 &
             d_arm$fuzz == 0 & d_arm$arm == "fuzzed", "beta_hat"]
ctl_t <- t.test(ctl, mu = 0.9)
## The zero-fuzz arm means, so "eight of nine within X" is measured, not
## remembered. The ninth is the cell tested above.
z0 <- agg[agg$fuzz == 0, ]
z0_dev <- sort(abs(z0$beta_hat - 0.9))
z0_ok  <- z0_dev[length(z0_dev) - 1]      # the largest of the other eight
sign_cnt <- sapply(designs, function(de) sapply(c(0.5, 1.0), function(fv) {
  w <- cell(de[1], de[2], fv); sum(w$integrated - w$precise < 0) }))
err_cnt <- sapply(designs, function(de) sapply(c(0.5, 1.0), function(fv) {
  w <- cell(de[1], de[2], fv)
  sum(abs(w$integrated - 0.9) > abs(w$precise - 0.9)) }))
```

**The fuzzed arm is gutted.** Displaced by half the correlation length, it has
already lost `r fmt(min(att05), 0)`–`r fmt(max(att05), 0)`% of the true slope;
displaced by a full correlation length,
`r fmt(min(att10), 0)`–`r fmt(max(att10), 0)`%. Across all six non-zero cells
the loss is **`r fmt(att_lo, 0)`–`r fmt(att_hi, 0)`%**. That is the finding.
Everything else follows from it.

**Pooling behaves exactly as pooling should.** The integrated estimate lands
between the two single-arm estimates, which is what averaging does — the last
column of the table is the arm-size-weighted average of the two single-arm
means, and it tracks the integrated fit to within
`r fmt(min(nw_gap), 3)`–`r fmt(max(nw_gap), 3)` **within each replicate**, in
every design including the unbalanced ones. There is nothing pathological in the
fitting.

The problem is **what is being averaged in**. The fuzzed arm arrives carrying
`r fmt(att_lo, 0)`–`r fmt(att_hi, 0)`% attenuation, and the pooled estimate
inherits it in proportion to that arm's share of the data. Hold fuzz at 0.5 and
grow the fuzzed arm's share — 100 rows against 400, then 220 against 220, then
400 against 100 — and the gap `integrated − precise` deepens in step:
**`r fmt(pen[1])` → `r fmt(pen[2])` → `r fmt(pen[3])`**, with non-overlapping
confidence intervals
(`r sprintf("[%.3f, %.3f]", pen_ci[1,1], pen_ci[2,1])`,
`r sprintf("[%.3f, %.3f]", pen_ci[1,2], pen_ci[2,2])`,
`r sprintf("[%.3f, %.3f]", pen_ci[1,3], pen_ci[2,3])`).

**The general mechanism is already documented.** Fletcher et al. (2019) show
that in integrated SDMs without explicit weighting, estimates are largely driven
by the abundant presence-only data, increasingly so as that arm grows — the
integrated model's parameters shrink toward the presence-only model's. That is
*swamping*. What this article adds is narrower: positional error is one of the
defects that gets swamped in, the attenuation it causes is severe well before
the fuzz reaches the covariate's correlation length, and the ratio in the
previous section is what tells you whether your case is in that regime.

**The zero-fuzz control mostly works, and where it does not, the cause is
identifiable.** Eight of the nine arm means at `fuzz = 0` sit within
`r fmt(z0_ok)` of truth. The ninth is the fuzzed arm in the 400/100 design, at
`r fmt(mean(ctl))` — `r fmt(ctl_t$statistic, 2)` standard errors above 0.9,
*p* = `r fmt(ctl_t$p.value, 3)`, so it is not noise. It is the smallest arm in
the grid (100 sites) at log-scale intercepts of −0.3 and 0.1, i.e. counts of
order one, which is the regime where the Poisson maximum-likelihood slope is
biased upward in small samples. The control still does its job — the *large*
bias appears only when positional error does — but this one cell is a known
small-sample effect rather than a fluke.

**Landscape by landscape, integration loses.** The precise-arm and integrated
analyses saw the same landscape and the same truth, so they can be paired.
Counting how often the integrated estimate is simply *lower*, it is lower in
`r min(sign_cnt)` of 15 replicates in every design at both non-zero fuzz levels.
Counting how often it is *further from the truth* — the quantity that actually
matters — it is worse in `r min(err_cnt)` of 15 in the 400/100 design at
`fuzz = 0.5` and in 15 of 15 everywhere else. The two counts differ in one cell
out of six, and the smaller one is the one to quote.

## With a biased opportunistic arm

Everything above gives the opportunistic arm a property that removes the whole
reason anyone integrates: it is **unbiased**. Nobody fuses a structured survey
with an opportunistic pile for fun. They do it because the opportunistic arm is
biased by accessibility, effort and detectability, and the structured arm is
what corrects that. Taken at face value, the result above could leave you
concluding that your survey arm is not worth keeping — so the design is repeated
with that motivation put back in.

In this second run the opportunistic arm is **precise but sampling-biased**:
reporting intensity carries a coefficient of 1.2 on an accessibility surface
drawn independently of the habitat covariate. The survey arm is **unbiased but
fuzzed**, as before. The analyst models the bias with an `access` term, as a
competent one would — accessibility is observable in real analyses (road
density, trail networks) even though the reporting bias it induces is not.

```{r access-cor, echo = FALSE}
env_surface <- function(ns = 26, ell = 0.25, seed = 1) {
  set.seed(seed)
  gx <- seq(0, 1, length.out = ns); g <- expand.grid(lon = gx, lat = gx)
  S <- exp(-as.matrix(dist(g)) / ell)
  z <- as.numeric(t(chol(S + diag(1e-6, nrow(S)))) %*% rnorm(nrow(g)))
  as.numeric(scale(z))
}
acc_r <- sapply(1:15, function(s)
  cor(env_surface(seed = 3000 + s), env_surface(seed = 3000 + s + 500L)))
```

The two surfaces are independent by construction, but each replicate draws its
own pair, so the correlation actually realised in any one landscape varies:
across the 15 landscapes it runs from `r fmt(min(acc_r), 2)` to
`r fmt(max(acc_r), 2)` (mean `r fmt(mean(acc_r), 2)`, sd
`r fmt(sd(acc_r), 2)`). The whole range is quoted because a single seed's value
would suggest the design pins it at zero, and it does not.

```{r biased-po-figure, echo = FALSE, fig.width = 5.2, fig.height = 4.0, dpi = 200, out.width = "70%", fig.cap = "The same penalty with a precise but sampling-biased opportunistic arm (400 rows, reporting intensity biased by accessibility) against an unbiased but fuzzed survey arm (100 rows). Pale symbols are the 105 distinct fits (15 precise, generated before fuzz is applied and so drawn at each fuzz level; 45 fuzzed; 45 integrated); solid symbols and lines are means. The bias is modelled with an access term in every fit. The integrated estimate still falls below the precise-arm-alone estimate at both non-zero fuzz levels, by a smaller margin than in the unbiased design.", fig.alt = "One panel plotting estimated slope against positional fuzz (0, 0.5 and 1.0 times the correlation length) for three analyses: precise (biased but unfuzzed), fuzzed (unbiased but positionally fuzzed), and integrated. A dashed grey reference line, labelled truth, marks the true slope 0.9. The precise series stays flat near 0.90 at every fuzz level. The fuzzed series falls from about 0.92 at zero fuzz to about 0.21 at fuzz 1.0. The integrated series falls from about 0.90 to about 0.80, below the precise series at both non-zero fuzz levels."}
par(mar = c(5.3, 3.6, 2.2, 0.6), mgp = c(2.1, 0.6, 0), las = 1,
    cex.axis = 0.80, cex.lab = 0.85, tcl = -0.3)
plot(NA, xlim = c(-0.06, 1.06), ylim = range(c(0, 1.05, d_bpo$beta_hat)),
     xaxt = "n", xlab = xlb,
     ylab = expression("estimated slope  " * hat(beta)))
axis(1, at = fz, labels = format(fz))
abline(h = 0.9, col = "grey45", lwd = 1.1, lty = 2)
text(1.045, 0.9, "truth", col = "grey35", cex = 0.66, adj = c(1, -0.45))
mu_bpo <- tapply(d_bpo$beta_hat, list(d_bpo$arm, d_bpo$fuzz), mean)[arms, ]
set.seed(3)
for (a in arms) {
  s <- d_bpo[d_bpo$arm == a, ]
  points(s$fuzz + off[a] + runif(nrow(s), -0.010, 0.010), s$beta_hat,
         pch = pchv[a], cex = 0.42, col = adjustcolor(col[a], 0.30))
  lines(fz + off[a], mu_bpo[a, ], col = col[a], lwd = 2, lty = ltyv[a])
  points(fz + off[a], mu_bpo[a, ], pch = pchv[a], cex = 1.0, col = col[a])
}
legend("bottomleft", legend = arms, col = col[arms], pch = pchv[arms],
       lty = ltyv[arms], lwd = 1.6, bty = "n", cex = 0.72,
       text.col = col[arms])
mtext("400 precise-but-biased / 100 unbiased-but-fuzzed", side = 3,
      line = 0.5, cex = 0.75, font = 2)
## The fence travels with the PNG, not with the caption.
mtext("SIMULATED DATA — point estimates only; no interval or coverage claim",
      side = 1, line = 3.0, cex = 0.68, col = "grey35")
```

```{r biased-po-table, echo = FALSE}
agg3 <- aggregate(beta_hat ~ fuzz + arm, d_bpo, mean)
w3 <- reshape(agg3, idvar = "fuzz", timevar = "arm", direction = "wide")
names(w3) <- sub("beta_hat\\.", "", names(w3))
w3$penalty <- w3$integrated - w3$precise
bpo_cnt <- t(sapply(w3$fuzz, function(fv) {
  s <- wide(d_bpo[d_bpo$fuzz == fv, ])
  c(sum(s$integrated - s$precise < 0),
    sum(abs(s$integrated - 0.9) > abs(s$precise - 0.9)))
}))
w3$lower <- bpo_cnt[, 1]; w3$worse <- bpo_cnt[, 2]
w3 <- w3[, c("fuzz", "precise", "fuzzed", "integrated", "penalty",
             "lower", "worse")]
names(w3) <- c("fuzz", "precise (biased)", "fuzzed (unbiased)", "integrated",
               "integrated - precise", "lower (of 15)",
               "further from truth (of 15)")
knitr::kable(w3, digits = 3, row.names = FALSE)
```

**The penalty survives the realistic design.** With the motivation for
integration present and correctly modelled, integrating the fuzzed arm still
moves the habitat slope further from truth in
`r min(bpo_cnt[w3$fuzz > 0, 2])` of 15 replicates at both non-zero fuzz levels.
The gap is smaller than in the unbiased design — but **do not compare the
magnitudes across the two runs**: this one fits an `access` term the other does
not, so they are different models, and only the within-run contrast is valid in
each.

The finding is also narrower than it may sound. The penalty falls on the
**habitat slope specifically**. The *same* survey arm, once its coordinates are
fuzzed, damages that slope while it repairs the reporting bias. Both are true at
once, and which one wins is the next section. It matters, because the answer
changes sign.

```{r reversal-derived, include = FALSE}
hlp <- function(o) { s <- d_rev[d_rev$order == o, ]
  sum(abs(s$integrated - 0.9) < abs(s$precise - 0.9)) }
p_flip <- fisher.test(matrix(c(hlp(4), 12 - hlp(4),
                               hlp(3), 12 - hlp(3)), 2))$p.value
p_err  <- fisher.test(matrix(c(hlp(2), 12 - hlp(2),
                               hlp(1), 12 - hlp(1)), 2))$p.value
ci_str <- function(k) { ci <- binom.test(k, 12)$conf.int
  sprintf("%d/12 is [%.3f, %.3f]", k, ci[1], ci[2]) }
```

# When the conclusion reverses

**Read this section if you read nothing else.** Everything so far has given the
simulation something no real analysis has: the accessibility surface — the thing
that explains where people happened to record — is measured **exactly**, and put
in the model. Given that, the presence-only arm's habitat slope is already
unbiased, so adding a fuzzed arm can only cost you. Every number above is a cost
with no benefit to weigh against it.

Real data never gives you that. You proxy "where people record" with road
density, trail networks, effort — surrogates that are (a) measured imperfectly
and (b) correlated with your habitat layer, because roads follow terrain and
land cover. **When both of those are true at once, the ranking flips: the
fuzzed survey arm earns its place.** Integrating it moved the habitat slope
*closer* to truth in `r hlp(4)` of 12 landscapes, against `r hlp(3)` of 12 when
the same surrogate was measured exactly. Drop the survey arm in that situation
and you will probably do worse, fuzz and all.

Two things to hold on to before taking that as licence. **It needs both
problems together, not either one alone** — an imperfect surrogate that happens
to be unrelated to habitat does not flip anything. And in the regime where the
flip happens, **neither answer is close to the truth**; "integration helped"
means "less wrong", not "right". Both points are measured below.

## What was varied

One thing changes at a time, 12 replicate landscapes each, all at `fuzz = 1.0` —
the worst case measured above. The five rows are:

- **measured exactly** — the baseline used in the previous section: the
  accessibility surface is unrelated to habitat and known without error.
- **surrogate error only** — the surrogate is measured with error
  (sd 0.5) but is still unrelated to habitat.
- **confounding only** — the surrogate is correlated with habitat
  (`cor = 0.7`) but is measured exactly.
- **confounding + surrogate error** — both at once. This is the row that
  resembles real data.
- **confounded, not modelled** — the surrogate is correlated with habitat and
  the analyst leaves it out of the model altogether.

```{r reversal-table, echo = FALSE}
rev_tab <- do.call(rbind, lapply(sort(unique(d_rev$order)), function(o) {
  s <- d_rev[d_rev$order == o, ]
  data.frame(
    `what the analyst has` = s$cond[1],
    `cor(access, env)`     = s$rho[1],
    `surrogate error sd`   = s$err_sd[1],
    precise                = mean(s$precise),
    `precise-arm bias t`   = unname(t.test(s$precise - 0.9)$statistic),
    integrated             = mean(s$integrated),
    `integration helped`   = paste0(
      sum(abs(s$integrated - 0.9) < abs(s$precise - 0.9)), " / 12"),
    check.names = FALSE, row.names = NULL)
}))
knitr::kable(rev_tab, digits = c(0, 1, 1, 3, 2, 3, 0), row.names = FALSE)
```

`precise-arm bias t` is how many standard errors the presence-only-alone slope
sits away from the true 0.9; roughly, values beyond about 2 in absolute value
say that arm is biased on its own. `integration helped` counts the landscapes
in which the integrated estimate was closer to 0.9 than the presence-only arm
alone.

## Reading the table

**Neither problem flips it alone.** An imperfect surrogate by itself, with
accessibility unrelated to habitat, leaves the presence-only slope unbiased
(row 2, bias *t* below 2 in absolute value) and integration helps in
`r hlp(2)` of 12 — no different from the `r hlp(1)` of 12 baseline (Fisher
*p* = `r fmt(p_err, 2)`). Correlation with habitat by itself, with the surrogate
measured exactly, helps in `r hlp(3)` of 12. Only when the surrogate is **both
imperfect and correlated with habitat** does the ranking reverse, to `r hlp(4)`
of 12 (*p* = `r format.pval(p_flip, digits = 2)` against correlation alone).

That is exactly what measurement error in a regressor is expected to do: error
in one predictor spills into another predictor's coefficient only through the
covariance between them. It is also a more useful instruction than a general
warning, because the condition is one you can *check* rather than assume.
Compute `cor(env, surrogate)` on your own data. Road density is correlated with
almost every environmental layer, so in practice the warning usually applies —
but now you can tell.

**Treat the table as one comparison, not four ranked results.** Twelve
landscapes per row is not many, and the proportions are correspondingly
imprecise: `r ci_str(hlp(4))` and `r ci_str(hlp(1))`. With counts this small,
rows 1 and 2 cannot be told apart from each other, and neither can rows 4 and 5.
The one comparison the design can actually resolve is correlation-alone against
correlation-plus-error.

**And the bottom two rows say something else that matters: in that regime
nothing recovers the truth.** Precise `r fmt(rev_tab$precise[4], 2)`,
integrated `r fmt(rev_tab$integrated[4], 2)`, against a true 0.9. The *ranking*
flips; both answers are still bad.

So the decision this article supports is narrower than the measurement driving
it. **Positional precision is a cost to measure, not a reason to discard a
structured arm.** If your bias surrogate is imperfect *and* correlated with your
habitat layer — and for road-density-style surrogates it usually is — the
structured arm is probably earning its place despite the fuzz.

# What to do

`gllvmTMB` does not implement a fix and this article does not claim one. The
general remedy is an errors-in-variables layer with a per-arm positional prior:
a real modelling option, and not implemented here.

1. **Compute the ratio, do not eyeball the map.** Run
   [the recipe](#the-recipe) on your own raster, divide any reported *practical
   range* by 3 first, and compare the result to your survey arm's positional
   uncertainty. Fuzz well below the correlation length is benign; by half the
   correlation length the tables above already show
   `r fmt(min(att05), 0)`%+ attenuation in the affected arm.
2. **Check `cor(env, surrogate)`.** It decides which way the previous section
   points for you. Surrogate unrelated to habitat: the fuzzed arm is a net cost.
   Correlated with habitat and imperfectly measured: it is probably a net gain.
3. **Weight the arms, or at least know what the unweighted fit is doing.** The
   penalty tracks the fuzzed arm's share of the pooled data, so an integrated
   fit that lets the larger arm dominate is making that choice implicitly.
   Fletcher et al. (2019) reach the same conclusion from the swamping side.
4. **Fit the arms separately first.** If the precise-arm-alone and integrated
   slopes disagree materially, that gap is evidence rather than noise. In real
   data it can reflect positional error, sampling bias, detection differences,
   or a genuine scale difference all at once, and these clean simulations
   cannot tell you which.
5. **Ask for the precise coordinates.** Most fuzzing regimes, ABMI's included,
   have a data-agreement route to exact locations. For an analysis that turns
   on a habitat slope, that request is worth making before the analysis rather
   than after.

**Do not reach for the obvious mitigation, because it has been tested and it
does not work.** The intuitive fix is to match covariate resolution to
positional precision — if coordinates are fuzzed to 5.5 km, extract covariates
at 5.5 km, so they are read at approximately the right place. **Gábor et al.
(2022) tested that directly, and the answer is in their title: positional errors
in species distribution modelling are not overcome by the coarser grains of
analysis.** Moudrý et al. (2023) restate the same finding. Coarsening destroys
fine-scale signal and does not buy back what positional error took, so it costs
twice. Measure the ratio, and treat the positional error as a property of the
data rather than something a resolution choice can absorb.

**A related defect worth checking while you are there.** Positional error
corrupts the covariate at *fit* time. Scaling that covariate separately on the
training data and on a prediction grid corrupts it at *prediction* time — a
different error with the same flavour. That one is affine, so it preserves the
map's rank order **exactly**: every hotspot stays a hotspot and nothing warns
you, while the predicted values themselves shift by an amount that depends on
the two datasets' means and standard deviations. The companion
`isdm-canada-warbler` article shows the fix.

# Scope and limits

- **Everything here is simulated, and that is a hard boundary rather than a
  caveat.** No real ABMI or GBIF data was fitted anywhere in this article. The
  5.5 km figure is documented ABMI policy and motivates the design; it is not a
  measurement of bias in any real dataset.
```{r truncation, include = FALSE}
## The simulations apply pmin(pmax(coord + rnorm(n, 0, fuzz * ell), 0), 1),
## so displacement is clipped at the domain boundary. Measure what that
## actually does rather than describing the intended mechanism.
set.seed(11); n <- 5e5; ell <- 0.25
trunc <- t(sapply(c(0.5, 1.0), function(fzz) {
  lo <- runif(n); la <- runif(n)
  ex <- rnorm(n, 0, fzz * ell); ey <- rnorm(n, 0, fzz * ell)
  lf <- pmin(pmax(lo + ex, 0), 1); af <- pmin(pmax(la + ey, 0), 1)
  c(clipped  = mean(lo + ex < 0 | lo + ex > 1 | la + ey < 0 | la + ey > 1),
    realised = mean(sqrt((lf - lo)^2 + (af - la)^2)) /
               mean(sqrt(ex^2 + ey^2)))
}))
rownames(trunc) <- c("0.5", "1.0")
```

- **The displacement is clipped at the edge of the map, so the fuzz axis is
  slightly mislabelled.** Displaced coordinates are pushed back inside the unit
  domain, so at `fuzz = 1.0` about `r fmt(100 * trunc["1.0", "clipped"], 0)`%
  of survey rows are clipped on at least one axis and the displacement actually
  realised is about `r fmt(100 * trunc["1.0", "realised"], 0)`% of nominal
  (`r fmt(100 * trunc["0.5", "clipped"], 0)`% and
  `r fmt(100 * trunc["0.5", "realised"], 0)`% at `fuzz = 0.5`). The axis labels
  are therefore off by roughly 10–15%, in the conservative direction — the
  fuzzed arm is if anything less damaged than the label implies. Real
  administrative fuzzing is often uniform-within-a-cell or grid-snapped, which
  will behave differently again.
- **One covariate structure.** A single smooth surface with exponential
  correlation. Patchier or directional covariates will attenuate differently.
- **The simulated landscape sits where the recipe reads low.** Its `phi` is
  0.25 of the domain side, so a reader applying this article's own recipe to
  this article's own landscape would get about `r fmt(0.25 * r025, 2)` rather
  than 0.25, and would read the `fuzz = 1.0` cell as a ratio near
  `r fmt(1 / r025, 1)` rather than 1.0. That is exactly the arithmetic the
  article asks you to perform, so it is stated here rather than left to be
  discovered.
- **Fixed effects only.** No spatial field is fitted in either design, which
  isolates the covariate channel and means the field-absorption intuition from
  the opening is set aside, not tested. Positional error also corrupts an
  estimated spatial field; that is not measured here.
- **One bias mechanism.** One accessibility surface, one bias coefficient. Real
  reporting bias may be multi-causal, non-linear, or itself measured with error.
- **Fifteen replicates, and three designs on one set of landscapes.** The
  effects reported are large relative to their confidence intervals, but the
  three designs are not three independent confirmations, and no number here is
  a precision estimate.
- **No coverage claim.** Point estimates only. Whether confidence intervals
  around these attenuated slopes retain their nominal coverage is a separate
  question, and it is not answered here.

## References

- Fletcher, R. J. Jr., Hefley, T. J., Robertson, E. P., Zuckerberg, B.,
  McCleery, R. A., & Dorazio, R. M. (2019). *A practical guide for combining
  data to model species distributions.* Ecology, **100**(6), e02710. Establishes that unweighted integrated models
  are driven by the abundant presence-only arm, increasingly so as that arm
  grows.
- Koo, K. S., Lee, K.-H., Lee, D., & Jang, Y. (2025). *Impact of obscured
  data on species distribution models.* Conservation Biology, **39**, e70050.
  The direct test of iNaturalist's obscured-geoprivacy setting: obscuring
  changed the environmental values attached to records and the estimated
  distributions of three endangered species.
- Gábor, L., Jetz, W., Lu, M., et al. (2022). *Positional errors in species
  distribution modelling are not overcome by the coarser grains of analysis.*
  Methods in Ecology and Evolution, **13**, 2289--2302. Also reports SDM performance declining as positional error grows.
- Moudrý, V., Keil, P., Gábor, L., et al. (2023). *Scale mismatches between
  predictor and response variables in species distribution modelling: A review
  of practices for appropriate grain selection.* Progress in Physical
  Geography: Earth and Environment, **47**, 467--482. Restates the Gábor et al.
  result in a review of spatial-resolution choices in species distribution
  modelling.

## For the record

Everything on this page is built from replicate-level result files tracked in
the repository, so no sentence above can drift from the numbers beneath it. For
anyone reproducing or extending the work:

- **Simulation scripts and per-replicate results:** `dev/isdm-precision/`.
  `generate-arm-weighting-results.R` produces the three-design run,
  `generate-biased-po-results.R` the biased-opportunistic-arm run,
  `generate-surrogate-error-results.R` the reversal run, and
  `generate-cor-length-grid.R` the recipe-accuracy grid. Each writes the
  matching `.rds` this article reads.
- **A worked case of the training/prediction scaling defect**, with measured
  magnitudes, is in `dev/isdm-precision/READER-BLOCKS.md` (Block 5).
- **Changes since earlier versions of this page**, recorded so that anyone
  holding an older copy can see what moved: the correlation-length recipe used
  to return `NA` silently when it could not measure `phi`, and now stops with a
  message naming which of the two failure modes occurred; the recipe-accuracy
  experiment used to vary map width and cell spacing together and so could not
  separate them, and now crosses them; the claim that the recipe's error is
  "always downward" was a statement about a mean over surfaces and has been
  replaced with the per-surface spread reported above; the paired-comparison
  count is now reported by distance from the truth as well as by sign, and the
  smaller count is the one quoted; the reversal table gained the
  surrogate-error-only row that isolates the mechanism; and an earlier version
  of this article recommended matching covariate resolution to positional
  precision, which Gábor et al. (2022) tested and found does not work.
