---
title: "An integrated SDM for twelve species, end to end"
description: "Twelve boreal songbirds, an opportunistic count arm and a structured detection arm at disjoint locations, and one shared spatial field. Twelve species make two decisions real that two species cannot: how many latent spatial axes to fit (AIC does it; the near-zero-loading rule of thumb does not), and what to do about overdispersion in a Poisson arm that has no dispersion parameter. Ends in a relative-intensity map, and in what the map does not carry."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{An integrated SDM for twelve species, end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

## ---------------------------------------------------------------------------
## Evidence files. Every replicated number in this article is READ from one of
## these, not recomputed at render time: they are campaigns of tens of fits at
## about a minute each, and a documentation build must not run them. If one is
## missing the build STOPS rather than falling back to a hard-coded number --
## a page that quietly substitutes a remembered figure for a measurement is
## the failure mode this article spends its length guarding against.
## ---------------------------------------------------------------------------
D <- "../../dev/isdm-precision"
need <- c(probe = "cawa12-probe.rds", aic = "cawa12-aic.rds",
          uniq  = "cawa12-unique.rds", quad = "cawa12-quadrature.rds",
          mapd  = "cawa12-mapdata.rds", gate = "precision-sim-gate.rds")
paths <- stats::setNames(file.path(D, need), names(need))
absent <- paths[!file.exists(paths)]
if (length(absent))
  stop("missing evidence file(s): ", paste(absent, collapse = ", "),
       " -- re-run the matching dev/isdm-precision/generate-*.R script")
pr <- readRDS(paths[["probe"]]); ai <- readRDS(paths[["aic"]])
un <- readRDS(paths[["uniq"]]);  qd <- readRDS(paths[["quad"]])
md <- readRDS(paths[["mapd"]]);  gt <- readRDS(paths[["gate"]])
us <- un$summary

f2 <- function(x, d = 2) formatC(x, format = "f", digits = d)

## Capture a warning without letting its ABSENCE break the build. A vignette
## that calls conditionMessage() on whatever tryCatch() hands back will die
## with an unrelated error the day the package stops warning -- a change worth
## noticing, but not one that should take the whole website down with it.
warn_of <- function(expr) {
  w <- tryCatch({ force(expr); NULL }, warning = function(w) w)
  if (is.null(w)) "(no warning was emitted -- behaviour has changed)"
  else paste0("<", class(w)[1], ">\n", conditionMessage(w))
}
err_of <- function(expr) {
  e <- tryCatch({ force(expr); NULL }, error = function(e) e)
  if (is.null(e)) "(no error was raised -- behaviour has changed)"
  else paste0("<", class(e)[1], ">\n", conditionMessage(e))
}
```

# What this article is

An **integrated species distribution model** fuses an opportunistic arm --
eBird checklists, GBIF records -- with a structured survey arm, so the
designed sample can discipline the messy one. This article runs one end to
end for a community of **twelve boreal songbirds**, with the two arms at
locations that share nothing, and finishes with the thing the exercise is
for: a relative-intensity map for all twelve.

Twelve species is the point, not decoration. With two species there is one
sensible latent structure and no dispersion question worth asking. With
twelve, two decisions become real, and neither has an obvious answer:

1. **How many latent spatial axes?** The shared field has a rank `d`, and
   the model will fit any `d` you name. Section
   [Decision 1: the rank of the shared field] measures what happens when you
   ask AIC (it works, in 10 replicate datasets out of 10) and what happens
   when you use the rule of thumb everyone reaches for instead -- *stop when
   a loading column goes near zero* -- which gets the answer right in 1 of
   10, for two separate reasons.
2. **What absorbs overdispersion in a Poisson arm?** Poisson has no
   dispersion parameter. Section
   [Decision 2: overdispersion in an arm with no dispersion parameter]
   measures `spatial_latent(..., unique = TRUE)` as the mechanism, and the
   answer is the opposite of the one most people would predict.

Everything else here -- the two arms, the mesh, the offset, the map, the
presence-only route -- is machinery in service of those two.

## Where the numbers come from

One fit runs live on this page. Everything replicated is read from a saved
campaign, listed here so each claim can be weighed against the design behind
it:

```{r provenance, echo = FALSE}
data.frame(
  campaign = c("intensity ceiling + recovery envelope", "rank selection",
               "overdispersion (unique = TRUE)", "presence-only quadrature",
               "map surfaces"),
  fits = c(nrow(pr$sweep) + nrow(pr$timing) + sum(pr$sweep_summary$n_rep),
           nrow(ai$fits), 2L * us$n_seeds_fitted,
           nrow(qd$replicate_fit_health) +
             nrow(qd$sweep_decoupled$long) / nrow(qd$replicates) +
             nrow(qd$sweep_coincident$long) / nrow(qd$replicates),
           length(md$ok_seeds)),
  file = c("cawa12-probe.rds", "cawa12-aic.rds", "cawa12-unique.rds",
           "cawa12-quadrature.rds", "cawa12-mapdata.rds"),
  row.names = NULL)
```

Each has its generating script beside it (`generate-<name>.R`), and each ran
against gllvmTMB `r ai$package_version`.

## The data are simulated, and that is a scope decision

**No real GBIF or ABMI data were analysed for this article.** Everything
below describes a simulated boreal songbird community shaped on a real
system -- Canada Warbler (*Cardellina canadensis*) and eleven co-occurring
species in Alberta, an eBird-style opportunistic arm and an ABMI-style
structured detection arm -- and no number here is a measurement about any
real taxon.

Three reasons, all of which you would meet on the way to the real analysis:

1. **The structured arm is not freely available at usable precision.** ABMI
   publishes public survey locations fuzzed to roughly 5.5 km, coarser for
   species at risk; precise coordinates require a **separate data
   agreement**. Fitting real ABMI locations here would need that agreement,
   and fitting the fuzzed public ones would measure the fuzz as much as the
   birds -- the subject of the companion article
   [*When one data source knows where it is and the other does not*](isdm-spatial-precision.html)
   (a pkgdown article, not an installed vignette, so `vignette()` will not
   find it).
2. **A vignette must not depend on the network.** The opportunistic arm *is*
   freely available and `rgbif` will fetch it, but a documentation build that
   phones an API is a documentation build that breaks.
3. **Real integration needs cleaning and matching work** that would swamp
   everything this page is trying to show.

Simulating also buys the one thing real data cannot: a known truth, against
which *did the model recover it* is a question with an answer.

# The two arms

Real opportunistic and structured arms almost never share a location, a grid
cell, or a unit label -- an eBird checklist and an ABMI station land wherever
their own protocol put them. The simulator draws them **independently** over
one landscape, so nothing here quietly assumes a convenience the real data
will not have.

```{r data}
suppressMessages(source("../../dev/isdm-precision/generate-cawa12.R"))

dat   <- sim_cawa12(seed = 1L)
truth <- attr(dat, "truth")
cells <- attr(dat, "cells")

str(dat[, c("cell_id", "trait", "isdm_source", "env", "log_effort", "value")])
```

The simulator is one file, sourced rather than retyped, so the fit below is
the same object every campaign in this article measured. Its data-generating
process is

$$\eta_{ij} \;=\; \alpha_j \;+\; \beta_j\,\mathrm{env}_i
  \;+\; \lambda_{1j} f_{1i} \;+\; \lambda_{2j} f_{2i}
  \;+\; \log(\text{effort}_i)
  \;+\; \kappa\,\mathbf{1}\{\text{arm}_i = \text{opportunistic}\},$$

with counts $y_{ij}\sim\text{Poisson}(e^{\eta_{ij}})$ on the opportunistic
arm and detections $y_{ij}\sim\text{Bernoulli}(1-e^{-e^{\eta_{ij}}})$ --
cloglog -- on the structured one. **The truth is rank 2**: two latent
spatial axes, each species carrying its own loading on each. Both axes are
residualised against `env` at the sampled locations, and $f_2$ against
$f_1$, so the covariate and the latent field are not two competing spatial
terms fighting over one pool of variance. Skipping that step is not a
subtlety: it drives the estimated slopes to about +21 against a true mean
near +1.3.

```{r arms}
tab <- table(cells$arm)
cat("opportunistic cells:", tab[["ebird"]],
    " structured stations:", tab[["abmi"]],
    "\nspecies:", length(truth$code), " rows:", nrow(dat), "\n")

## Disjoint by construction: no cell id, and no coordinate, is shared.
po <- cells[cells$arm == "ebird", ]; sv <- cells[cells$arm == "abmi", ]
cat("shared cell ids:", length(intersect(po$cell_id, sv$cell_id)),
    " shared (X, Y):",
    length(intersect(paste(po$X, po$Y), paste(sv$X, sv$Y))), "\n")

## Effort varies WITHIN each arm, so the offset is not a relabelling of arm.
cat("sd(log effort) within arm:", f2(sd(po$log_effort), 3), "/",
    f2(sd(sv$log_effort), 3),
    " cor(log effort, arm):",
    f2(cor(cells$log_effort, cells$arm == "ebird"), 3), "\n")

## Projected to UTM km BEFORE anything is meshed, inside a single zone.
cat("UTM zones spanned:", length(unique(floor((cells$lon + 180) / 6) + 1)),
    " X range (km):", paste(round(range(cells$X)), collapse = "-"),
    " Y range (km):", paste(round(range(cells$Y)), collapse = "-"), "\n")

cat("largest opportunistic count:", max(dat$value[dat$isdm_source == "ebird"]),
    " mean:", f2(mean(dat$value[dat$isdm_source == "ebird"]), 2),
    " detection rate, structured arm:",
    f2(mean(dat$value[dat$isdm_source == "abmi"]), 3), "\n")
```

Coordinates are projected to UTM kilometres before anything is meshed,
because a degree of longitude and a degree of latitude are different physical
distances everywhere but the equator, and an isotropic SPDE built on degrees
treats them as equal. Section [The one thing the likelihood cannot tell you]
returns to that, because the usual way of checking such a choice does not
work.

# One fit, and the check that it is a fit

```{r fit}
t0 <- proc.time()[["elapsed"]]
f  <- cawa12_fit(dat)
elapsed <- proc.time()[["elapsed"]] - t0
fit <- f$fit

cat("model: "); print(cawa12_formula()[[3]])
cat("\naccepted:", f$ok, " convergence:", f$convergence,
    " iterations:", f$iterations,
    "\nobjective:", f2(f$objective, 3), " seconds:", f2(elapsed, 1),
    " mesh nodes:", cawa12_mesh(dat)$mesh$n, "\n")
```

`isdm_sources()` is what tells the model the two arms are different
observation processes on one shared intensity: a Poisson likelihood for the
counts, a complementary-log-log Bernoulli for detection / non-detection.
`spatial_latent(0 + trait | coords, d = 2)` is the shared field, with
per-species loadings on it. The mesh is built from `dat` -- the union of both
arms' locations -- and every row from either arm is projected onto it. That
is the whole mechanism by which two arms sharing no row, coordinate or label
share a latent field: the field lives on the mesh, not on the data.

**`convergence == 0` is not the check.** This package has a silent failure
([issue #1167](https://github.com/itchyshin/gllvmTMB/issues/1167)) in which
an SPDE fit at large counts returns `convergence == 0` after a **single**
iteration at an astronomically large objective, every spatial parameter still
at its start value -- the sweep below shows the magnitudes. `cawa12_fit()` therefore rejects a fit unless **all three** hold:
`convergence == 0`, `iterations > 1`, and a finite objective. Iteration
counts in the dozens or hundreds are the healthy shape.

## How large can the counts get?

Measured, because *large* is not a feeling. The opportunistic arm's intensity
offset $\kappa$ was swept, three seeds per rung, every rung unanimous across
its three seeds:

```{r envelope, echo = FALSE}
sw  <- pr$sweep
agg <- do.call(rbind, lapply(split(sw, sw$kappa), function(z) data.frame(
  kappa           = z$kappa[1],
  mean_cell_count = round(mean(z$mean_count), 2),
  max_cell_count  = paste(range(z$max_count), collapse = "-"),
  iterations      = if (all(is.na(z$iterations))) "-"
                    else paste(range(z$iterations, na.rm = TRUE), collapse = "-"),
  mean_objective  = if (all(is.na(z$objective))) "-"
                    else format(mean(z$objective), digits = 3, scientific = TRUE),
  accepted        = paste0(sum(z$ok), "/", nrow(z)),
  verdict         = if (all(z$ok)) "healthy" else unique(z$reason[!z$ok])[1])))
rownames(agg) <- NULL
agg
```

Three things in that table are worth taking away rather than skimming.

**The cliff is real and it is not far off.** At
$\kappa = `r f2(pr$envelope$last_healthy_kappa, 1)`$ all three seeds fit
healthily; at `r f2(pr$envelope$first_failing_kappa, 1)` all three fail,
stopping after about a dozen iterations at an objective near 1e14. At
`r f2(pr$envelope$silent_1iter_kappa, 1)` the failure goes **silent** -- one
iteration, objective near 1e24, `convergence == 0` -- the exact #1167
signature. Beyond that the optimiser errors outright.

**A per-cell cap is a weak gate.** The largest single-cell count *overlaps*
across the cliff: `r pr$envelope$max_cell_count_healthy` in a healthy fit
against `r pr$envelope$min_cell_count_failing` in a rejected one. The
controlling quantity is the **mean** cell intensity, not the maximum. So the
figure worth carrying -- *keep your largest cell count under about
`r pr$envelope$safe_max_cell_count_recommended`* -- is a recommendation with
margin, not a measured threshold. The measured facts are the two overlapping
numbers.

**The statistical envelope is tighter than the numerical one.** Recovery of
the environmental slopes, four rungs, 12 replicate seeds each, all accepted:

```{r recovery-envelope, echo = FALSE}
ss <- pr$sweep_summary
data.frame(kappa = ss$kappa,
           mean_cell_count = round(sapply(ss$kappa, function(k)
             mean(sw$mean_count[sw$kappa == k])), 2),
           accepted        = paste0(ss$n_ok, "/", ss$n_rep),
           cor_est_truth   = paste0(f2(ss$mean_cor, 3),
                                    " (sd ", f2(ss$sd_cor, 3), ")"),
           mean_abs_err    = f2(ss$mean_abs_err, 3),
           mean_signed_err = f2(ss$mean_signed_err, 3))
```

Accuracy is already decaying two rungs before the optimiser breaks, while
every fit still reports itself healthy. That is the more useful of the two
warnings: the numerical gate tells you when to stop trusting the optimiser,
and nothing tells you when to stop trusting the estimate except a recovery
check of your own. $\kappa = `r f2(pr$envelope$chosen_kappa, 1)`$ was chosen
on this evidence -- best mean correlation, tightest spread -- at a median
`r f2(pr$envelope$secs_per_fit_median_serial, 1)` s per fit
(`r f2(pr$envelope$secs_per_fit_range_serial[1], 1)`--`r f2(pr$envelope$secs_per_fit_range_serial[2], 1)` s
across five serial fits on a shared machine).

**And there is a bias in that table that was not removed.** The signed error
is negative at every rung, averaging
`r f2(mean(ss$mean_signed_err), 3)` against a true slope SD of
`r f2(sd(truth$beta), 3)`. Slopes come back shrunk. This article makes no
claim that these estimates are unbiased, here or anywhere below.

## What this one fit recovered

```{r recovery-one}
est <- cawa12_env_slopes(fit)
cmp <- data.frame(species = truth$code, truth = round(truth$beta, 3),
                  estimate = round(unname(est), 3))
cmp$error <- round(cmp$estimate - cmp$truth, 3)
cmp
cat("cor(estimate, truth):", f2(cor(cmp$estimate, cmp$truth), 3),
    " mean |error|:", f2(mean(abs(cmp$error)), 3),
    " mean signed error:", f2(mean(cmp$error), 3), "\n")
```

One seed. Read it against the 12-seed envelope above
(`cor` `r f2(ss$mean_cor[ss$kappa == 1], 3)`, sd
`r f2(ss$sd_cor[ss$kappa == 1], 3)`), not on its own: a single replicate
scatters, and judging an estimator from one is how people conclude it is
broken when it is not.

# Decision 1: the rank of the shared field

`d` is the **rank** of the shared spatial structure -- the number of
independent latent surfaces the community draws on, each species carrying its
own loading on each. `d = 1` says one unmeasured gradient organises the whole
community; `d = 4` says four do. With two species the question barely exists.
With twelve it is the first real modelling decision, and the model will fit
whichever `d` you write.

## AIC decides it

Ten replicate datasets, each fitted at `d = 1, 2, 3, 4`, true rank 2:

```{r aic-table, echo = FALSE}
ff  <- ai$fits
sel <- ai$selection
by_k <- do.call(rbind, lapply(split(ff, ff$k), function(z) data.frame(
  d = z$k[1], df = z$df[1],
  mean_AIC       = round(mean(z$AIC), 1),
  AIC_range      = paste(round(range(z$AIC), 1), collapse = " - "),
  median_secs    = round(median(z$secs), 1))))
by_k$mean_delta_AIC <- round(c(mean(sel$delta_aic_1), mean(sel$delta_aic_2),
                               mean(sel$delta_aic_3), mean(sel$delta_aic_4)), 2)
by_k$delta_AIC_range <- c(
  paste(round(range(sel$delta_aic_1), 1), collapse = " - "), "0 - 0",
  paste(round(range(sel$delta_aic_3), 1), collapse = " - "),
  paste(round(range(sel$delta_aic_4), 1), collapse = " - "))
rownames(by_k) <- NULL
by_k
cat("AIC picks d = 2 in", sum(sel$aic_pick == 2), "of", nrow(sel),
    "replicates; BIC picks d = 2 in", sum(sel$bic_pick == 2), "of",
    nrow(sel), "\n")
## per-replicate gain in negative log-likelihood bought by each extra column
nll_gain <- t(sapply(split(ff, ff$seed), function(z) c(
  `2->3` = z$objective[z$k == 2] - z$objective[z$k == 3],
  `3->4` = z$objective[z$k == 3] - z$objective[z$k == 4])))
```

```{r rank-fig, echo = FALSE, fig.width = 6.6, fig.height = 3.6, dpi = 200, out.width = "100%", fig.cap = "Choosing the rank of the shared spatial field on ten replicate datasets whose true rank is 2. Panel A plots each replicate's AIC relative to its own best rank: the minimum falls at d = 2 in every replicate, under-fitting at d = 1 costs 114 to 175 AIC units, and each axis beyond the truth costs about 20 AIC for almost no gain in fit. Panel B tests the rule of thumb AIC is competing against. Each replicate was fitted at rank 3, one axis too many, and the three loading columns' root-mean-square values are plotted relative to the largest, the digit giving the column number. A reader stopping when a column falls below a near-zero cut reads the rank correctly in one replicate of ten: the smallest of the three columns runs from 0.00 to 0.34 of the largest, with only one replicate below a 0.05 cut, and the SMALLEST column is column 2 rather than the last in five of the ten replicates, because the loading matrix is identified only up to rotation. No uncertainty is shown anywhere in this figure.", fig.alt = "Two panels. Panel A: ten grey lines, one per replicate dataset, plotting AIC minus the best AIC against fitted rank 1 to 4. All ten start high at rank 1, between about 114 and 175, drop to exactly zero at rank 2, marked by a red point labelled true rank 2, then rise gently to roughly 15 to 20 at rank 3 and 32 to 38 at rank 4. Panel B: ten horizontal rows, one per replicate seed, each carrying three bold digits 1, 2 and 3 placed along a horizontal axis from 0 to 1 giving each loading column's root-mean-square relative to the largest column. Digit 1 sits at 1.0 in every row by construction. The other two digits scatter between about 0.02 and 0.8, and in half the rows digit 3 lies to the right of digit 2. A red dashed vertical line at 0.05 marks a near-zero cut; only one row has any digit to the left of it."}
op <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(3.6, 3.6, 2.2, 0.6), oma = c(2.6, 0, 1.2, 0),
    mgp = c(2.1, 0.55, 0), las = 1, cex.axis = 0.72, cex.lab = 0.8, tcl = -0.3)

DD <- as.matrix(sel[, paste0("delta_aic_", 1:4)])
plot(NA, xlim = c(0.8, 4.2), ylim = c(-14, max(DD) * 1.06), xaxt = "n",
     xlab = "rank d fitted", ylab = "AIC - AIC(best rank)")
axis(1, at = 1:4, cex.axis = 0.72); abline(h = 0, col = "grey85")
for (i in seq_len(nrow(DD)))
  lines(1:4, DD[i, ], col = adjustcolor("grey35", 0.5), lwd = 1)
points(rep(1:4, each = nrow(DD)), as.vector(DD), pch = 16, cex = 0.5,
       col = adjustcolor("grey20", 0.6))
points(rep(2, nrow(DD)), DD[, 2], pch = 21, bg = "#B2182B", col = "white",
       cex = 1, lwd = 0.6)
text(2, -9, "true rank 2", cex = 0.68, col = "#B2182B", font = 2)
text(4.15, max(DD) * 0.62, "extra axes cost\n~20 AIC each and\nbuy almost no fit",
     cex = 0.6, col = "grey30", adj = 1)
mtext("A. AIC picks the true rank, 10 of 10", side = 3, line = 0.4,
      cex = 0.78, font = 2, adj = 0)

k3    <- ai$loadings[ai$loadings$k == 3, ]
seeds <- sort(unique(k3$seed))
rel   <- t(sapply(seeds, function(s) {
  v <- k3$col_rms[k3$seed == s][order(k3$column[k3$seed == s])]; v / max(v) }))
plot(NA, xlim = c(-0.03, 1.05), ylim = c(0.4, length(seeds) + 1.1), yaxt = "n",
     xlab = "loading-column rms / largest column", ylab = "")
axis(2, at = seq_along(seeds), labels = seeds, cex.axis = 0.58)
mtext("replicate (seed)", side = 2, line = 2.4, cex = 0.7, las = 0)
abline(v = 0.05, lty = 2, col = "#B2182B")
for (i in seq_along(seeds)) {
  segments(min(rel[i, ]), i, 1, i, col = "grey85")
  text(rel[i, ], rep(i, 3), labels = 1:3, cex = 0.6, font = 2,
       col = c("grey15", "#2166AC", "#B2182B"))
}
text(0.075, length(seeds) + 0.85, "a 0.05 \"near zero\" cut", cex = 0.6,
     col = "#B2182B", adj = 0)
mtext("B. the near-zero column: 1 of 10", side = 3, line = 0.4, cex = 0.78,
      font = 2, adj = 0)

mtext("Ten replicate datasets; true rank 2. B is fitted at rank 3; digits are column numbers.",
      side = 1, outer = TRUE, line = 0.3, cex = 0.6, col = "grey30")
mtext("The smallest column is column 2, not the last, in 5 of 10 replicates.",
      side = 1, outer = TRUE, line = 1.2, cex = 0.6, col = "grey30")
mtext("SIMULATED DATA -- no uncertainty shown", side = 3, outer = TRUE,
      line = -0.2, cex = 0.66, col = "#B2182B", font = 2)
par(op)
```

The mechanism is worth understanding, because it is what makes the comparison
trustworthy rather than lucky. Degrees of freedom rise by
`r ff$df[ff$k == 2][1] - ff$df[ff$k == 1][1]`,
`r ff$df[ff$k == 3][1] - ff$df[ff$k == 2][1]` and
`r ff$df[ff$k == 4][1] - ff$df[ff$k == 3][1]` across `d = 1 -> 2 -> 3 -> 4`.
Under-fitting at `d = 1` costs
`r f2(mean(ff$objective[ff$k == 1]) - mean(ff$objective[ff$k == 2]), 1)` units
of negative log-likelihood on average, so rank 2 wins by a wide margin.
**Past the true rank the likelihood is essentially flat**: across the ten
replicates the third column buys at most
`r f2(max(nll_gain[, 1]), 2)` nll units against a 20-AIC charge, and the
fourth at most `r f2(max(nll_gain[, 2]), 3)` against an 18-AIC charge -- which is why `delta`-AIC(4) minus `delta`-AIC(3) sits close
to the bare df penalty in nearly every replicate.

None of this is start-value luck. Refitting one dataset at `d = 2` and
`d = 3` under eight different RNG states gave an AIC range of exactly
`r f2(ai$optimiser_noise_summary$aic_range[1], 3)`: given the data, these
fits are deterministic.

## The near-zero-loading rule of thumb does not work at twelve species

The natural alternative is to skip the model comparison -- fit a generous
rank, look at the loading matrix, stop where a column goes near zero. Made
falsifiable as *"a column is spent when its rms falls below `thr` times the
largest column's"*, and applied to the same ten replicates:

```{r tell-sweep, echo = FALSE}
ts <- ai$tell_summary
data.frame(fitted_rank = ts$fitted_k, threshold = ts$thr,
           by_columns = paste0(ts$n_correct_colrms, "/", ts$n),
           by_singular_values = paste0(ts$n_correct_sv, "/", ts$n),
           modal_rank = ts$modal_rank_colrms)
nz3 <- ai$near_zero_tell[ai$near_zero_tell$k == 3, ]
nz4 <- ai$near_zero_tell[ai$near_zero_tell$k == 4, ]
sv3_nonzero  <- nz3$sv_rel_3[nz3$sv_rel_3 > 1e-6]
k3l <- ai$loadings[ai$loadings$k == 3, ]
smallest_col <- sapply(split(k3l, k3l$seed),
                       function(z) z$column[which.min(z$col_rms)])
```

Reading the columns -- the version a reader actually applies -- gives the
right rank in **1 of 10** replicates at every threshold from 0.005 to 0.10.
The modal answer is 3, not 2: the heuristic **over-counts by one**. Loosening
the threshold does not rescue it, so this is not a tuning problem.

Two distinct mechanisms are at work, and they are worth separating.

**Column order is not magnitude order.** The packed lower-triangular loading
matrix is identified only up to rotation, so "the extra column" is not the
last column. Fitted at rank 3, the *smallest* column is column 2 rather than
column 3 in `r sum(smallest_col == 2)` of the `r length(smallest_col)`
replicates. A reader following the recipe inspects the third column, sees
a substantial value, and concludes the third axis is real. Panel B above is
that fact drawn.

**Even as an unordered set, the gap is not there.** Fitted at rank 4, the
smallest singular value drops below a millionth of the leading one in
`r sum(nz4$sv_rel_4 < 1e-6)` of `r nrow(nz4)` replicates -- so one axis does
collapse -- but the *third* singular value still carries
`r f2(100 * min(sv3_nonzero), 0)` to `r f2(100 * max(sv3_nonzero), 0)` per
cent of the leading one in `r length(sv3_nonzero)` of `r nrow(nz3)`
replicates. Nothing inside a single fit separates "a tenth of the leading
axis" from a weak real axis without a reference distribution, and a
reference distribution is precisely what AIC supplies and eyeballing does
not.

The rotation-invariant version -- threshold the **singular values** of the
loading matrix instead of its columns -- is better and still not good:
`r ts$n_correct_sv[ts$thr == 0.10 & ts$fitted_k == 3][1]` of 10 at a 0.10
cut, and 1 of 10 at every tighter cut in the table.

**So compare ranks by AIC.** If you want a loading-based sanity check, take
singular values rather than columns, quote your threshold, and treat the
answer as a hint. Here is that matrix on the live fit -- worth seeing for how
little it settles on its own:

```{r loadings-live}
L <- fit$report$Lambda_spde          # 12 x d -- use this, not the packed vector
dim(L)
round(sqrt(colMeans(L^2)), 4)        # per-column rms
round(svd(L)$d / max(svd(L)$d), 4)   # singular values, relative to the largest
```

Use `fit$report$Lambda_spde` rather than the packed parameter vector: a
rank-`d` loading matrix for `P` traits is stored as `P*d - d(d-1)/2` free
values, so at 12 species and `d = 2` the vector holds 23 entries against a
12-by-2 matrix, and naming it with the species vector mislabels silently.

## What it costs

```{r rank-cost, echo = FALSE}
data.frame(d = by_k$d, median_secs_per_fit = by_k$median_secs,
           fits_in_this_campaign = as.integer(table(ff$k)))
```

Fitting four ranks on one dataset is a few minutes at this size. Fitting them
on ten datasets is the campaign that produced the table above -- and you do
not have to repeat it. Fit your own `d = 1..4` once and read the AIC.

# Decision 2: overdispersion in an arm with no dispersion parameter

The opportunistic arm is Poisson, and **Poisson has no dispersion
parameter**. Counts more variable than Poisson have nowhere to go except into
the terms that do carry parameters, and in this model the term with the most
freedom is the shared spatial field. A field busy absorbing extra-Poisson
noise is not estimating space -- and the environmental slopes are estimated
conditionally on it.

`spatial_latent(..., unique = TRUE)` adds a per-trait tier, `Psi`, alongside
the shared rank-`d` field:

```{r unique-formula, eval = FALSE}
value ~ 0 + trait + trait:env + isdm_source + offset(log_effort) +
  spatial_latent(0 + trait | coords, d = 2, unique = TRUE)
```

It is **not** the default for a source-tier latent term; it has to be asked
for. Whether it earns its keep was measured rather than argued.

## How it was measured

Genuine overdispersion was added to the opportunistic arm only, as a
species-specific observation-level effect
$u_{ij}\sim N(-\sigma_j^2/2,\ \sigma_j^2)$ on the log scale -- mean-centred
on the intensity scale, so $\sigma_j$ alone is manipulated and not abundance.
The twelve species carry a ladder of $\sigma$ from
`r f2(min(un$design$sigma_od), 2)` (an exact null cell) to
`r f2(max(un$design$sigma_od), 2)`. Both models, with and without
`unique = TRUE`, were fitted to identical data on an identical mesh.

```{r unique-design, echo = FALSE}
h <- us$health
data.frame(
  quantity = c("seeds offered", "screened out on the count cap",
               "seeds fitted", "base model accepted", "unique model accepted",
               "healthy pairs", "pairs with SEs on both models",
               "pairs with a PD Hessian on the unique model",
               "median seconds, base", "median seconds, unique"),
  value = c(as.character(c(length(un$design$seeds_offered),
            length(un$design$seeds_screened_out),
            us$n_seeds_fitted, sum(h$base_ok), sum(h$uniq_ok),
            us$n_pairs_healthy, us$n_pairs_se_complete,
            us$robustness$pdHess$n_pdHess_true)),
            f2(us$timing$base_median_secs, 1),
            f2(us$timing$uniq_median_secs, 1)))
```

The screen matters: any seed whose largest opportunistic count reached
`r un$design$count_cap` was rejected **before** fitting, because that is the
#1167 boundary from the first section. Iteration counts on the unique model
run in the hundreds, so no silent failure occurred -- the rejected fits
failed loudly, on `convergence != 0`.

## Can you see the overdispersion before fitting?

Only by contrast, and this is where the obvious diagnostic misleads. The
variance-to-mean ratio of the raw counts was computed alongside a matched
**equidispersed counterfactual**: same linear predictor, $\sigma = 0$, same
seed.

```{r vmr, echo = FALSE}
v <- us$vmr[order(us$vmr$sigma_od), ]
data.frame(species = v$trait, sigma = v$sigma_od,
           mean = round(v$mean_count, 2),
           VMR = round(v$vmr_mean, 2),
           VMR_equi = round(v$vmr_equi_mean, 2),
           frac_above = round(v$frac_gt_equi, 2))
cat("cor(VMR, sigma^2): observed", f2(us$vmr_cor_sigma2[["observed"]], 3),
    "-- equidispersed counterfactual",
    f2(us$vmr_cor_sigma2[["equidispersed_counterfactual"]], 3), "\n")
```

**The counterfactual column runs from `r f2(min(v$vmr_equi_mean), 1)` to
`r f2(max(v$vmr_equi_mean), 1)` with no overdispersion at all.** The shared
rank-2 field, the environmental gradient and the varying effort all inflate
it. So "VMR much greater than 1" does not demonstrate overdispersion in a
model like this one; only the contrast against a counterfactual you have to
simulate does, and on real data you do not have one. At the top of the ladder
the signal is unambiguous -- the observed VMR exceeds its own counterfactual
in every replicate at $\sigma \ge 0.75$ -- while at
$\sigma \le 0.45$ that fraction runs
`r f2(min(v$frac_gt_equi[v$sigma_od <= 0.45]), 2)` to
`r f2(max(v$frac_gt_equi[v$sigma_od <= 0.45]), 2)`, which is a coin flip
dressed up as a diagnostic.

## What `Psi` recovers

`Psi` is reported as $1/\tau^2$ in the base precision matrix's units; the
table below converts it to the variance scale numerically.

```{r psi, echo = FALSE}
ps <- us$psi[order(us$psi$sigma_od), ]
data.frame(species = ps$trait, sigma = ps$sigma_od,
           true_var = round(ps$sigma2, 3),
           psi_median = signif(ps$psi_median, 2),
           frac_nonzero = round(ps$frac_nonzero, 2))
hi <- ps$sigma_od >= 0.55
psi_ratio_hi <- ps$psi_median[hi] / ps$sigma2[hi]
cat("species-replicates with Psi > 1e-3:",
    f2(us$psi_detection$frac_nonzero_low_sigma, 3), "at",
    us$psi_detection$low_sigma_defined, "vs",
    f2(us$psi_detection$frac_nonzero_high_sigma, 3), "at",
    us$psi_detection$high_sigma_defined, "\n")
```

`Psi` is a **threshold detector, not a measurement**. Below
$\sigma \approx 0.5$ it sits numerically on the boundary at zero; above it,
it fires in most replicates, but its median lands at only
`r f2(100 * min(psi_ratio_hi), 0)` to `r f2(100 * max(psi_ratio_hi), 0)` per
cent of the true variance.

The attenuation is a mechanism mismatch rather than a tuning failure, and it
is visible in the source: `spatial_latent(unique = TRUE)`'s `Psi` is a
per-trait **spatially structured** field on the same base precision as the
shared field (`src/gllvmTMB.cpp`, the `SCALE(GMRF(Q_base), 1/tau)` prior),
not an i.i.d. observation-level nugget. It can only partly represent i.i.d.
overdispersion. Whether a genuine observation-level tier -- something shaped
like `latent(0 + trait | cell_id, unique = TRUE)` -- would recover
$\sigma^2$ at the right magnitude is the honest next measurement, and this
campaign did not run it.

The collapse at the low end has a price. The unique model's Hessian was not
positive-definite in
`r us$robustness$pdHess$n_total - us$robustness$pdHess$n_pdHess_true` of the
`r us$n_pairs_se_complete` SE-complete pairs, and
`r us$n_pairs_healthy - us$n_pairs_se_complete` of the
`r us$n_pairs_healthy` healthy pairs returned no fixed-effect standard errors
at all. Boundary estimates are where variance components misbehave.

## The result is the opposite of the expected one

The expected story is: ignoring overdispersion makes your intervals too
narrow, and the fix widens them. **That is the wrong way round here.**

```{r unique-slopes, echo = FALSE}
so <- us$slope_overall
data.frame(
  quantity = c("median |change| in env slope, in base SEs",
               "median SE ratio, unique / base",
               "95% Wald coverage of the true slope, base",
               "95% Wald coverage of the true slope, unique",
               "bias, base", "bias, unique", "RMSE, base", "RMSE, unique",
               "cor(slope movement, simulated sigma^2)",
               "cor(slope movement, base model's own error)"),
  value = round(c(so$median_abs_d_est_in_se, so$se_ratio_median,
                  so$coverage_base, so$coverage_uniq,
                  so$bias_base, so$bias_uniq, so$rmse_base, so$rmse_uniq,
                  so$cor_move_vs_sigma2, so$cor_move_vs_base_error), 3))
```

The environmental slopes **move** -- about
`r f2(so$median_abs_d_est_in_se, 2)` of a base standard error at the median
-- and the standard errors move **down**, not up: median ratio
`r f2(so$se_ratio_median, 3)`, and much the same at high $\sigma$
(`r f2(so$se_ratio_median_high_sigma, 3)`) as at low
(`r f2(so$se_ratio_median_low_sigma, 3)`).

So the base model's coverage failure is **not** understated uncertainty; it
is **bias**. Freeing the shared field from having to absorb extra-Poisson
noise de-attenuates the slopes *and* tightens them. Coverage rises from
`r f2(so$coverage_base, 3)` to `r f2(so$coverage_uniq, 3)`, which
**overshoots** nominal -- so `unique = TRUE` is not a calibration fix either.
In both models the mean standard error exceeds the across-replicate SD of the
estimate, so the standard errors were already conservative.

And the correction is not targeted. Movement correlates only weakly with the
simulated $\sigma^2$ (`r f2(so$cor_move_vs_sigma2, 2)`) and strongly and
negatively with the base model's own error
(`r f2(so$cor_move_vs_base_error, 2)`): `unique = TRUE` pulls whichever
slopes the base model got wrong, including species with almost no simulated
overdispersion at all.

## Whether to turn it on

```{r unique-selection, echo = FALSE}
dl <- us$dloglik
cat("delta logLik (unique - base) over", dl$n, "pairs: median",
    f2(dl$median, 2), " mean", f2(dl$mean, 2),
    " range", f2(dl$range[1], 2), "to", f2(dl$range[2], 2),
    "\nextra parameters:", dl$extra_params,
    "  median delta AIC:", f2(dl$dAIC_median, 2),
    "  replicates favouring unique:", dl$n_favouring_unique, "of", dl$n, "\n")
```

By likelihood it clearly fits better; by AIC it barely pays, and only in
about half the replicates. Do not read a plain $\chi^2_{12}$ off that delta
log-likelihood: the null puts variance parameters on a boundary, so the
reference distribution is a mixture, not a chi-square.

**The practical reading.** `unique = TRUE` here is worth having for the
**point estimates** -- it cuts RMSE from `r f2(so$rmse_base, 3)` to
`r f2(so$rmse_uniq, 3)` and removes most of the attenuation -- and is not a
route to better-calibrated intervals. It costs about
`r f2(us$timing$uniq_median_secs / us$timing$base_median_secs, 1)` times the
runtime and a materially worse-behaved Hessian. Restricting to the
`r us$robustness$pdHess$n_pdHess_true` pairs with a positive-definite Hessian
reproduces every number above -- SE ratio
`r f2(us$robustness$pdHess$se_ratio_median, 3)`, coverage
`r f2(us$robustness$pdHess$coverage_base, 3)` to
`r f2(us$robustness$pdHess$coverage_uniq, 3)`, RMSE
`r f2(us$robustness$pdHess$rmse_base, 3)` to
`r f2(us$robustness$pdHess$rmse_uniq, 3)` -- so none of it is an artefact of
the selected subset.

One thing this campaign could **not** test: whether any of it holds at counts
above the `r un$design$count_cap` screen. Detection of `Psi` should improve
with counts, and the #1167 ceiling is exactly what forbids looking.

# The map

Three things have to be right before a surface means anything.

**The spatial field must be in the prediction.** It is. Before gllvmTMB 0.7.0
(development), `predict(newdata = )` silently dropped it, so a map drawn that
way showed a field-free surface while claiming the field. On an older install,
check your version before trusting a map made this way.

**The scale must be effort-free.** `type = "response"` includes the row's
offset, so it returns an expected count *at that effort* -- right for a
training row, wrong for a map. Setting `log_effort` to zero on the grid gives
relative intensity, exactly rather than approximately, because the offset is
re-evaluated against `newdata`.

**The covariate must be on the scale it was fitted on.** `env` was
standardised once, on the training data, and the grid reuses those same two
numbers. Re-standardising the grid on its own mean and SD silently changes
what the covariate means: the map's *pattern* survives almost untouched while
its *values* do not, so no visual check catches it.

```{r map-grid}
grid <- md$grid[, c("X", "Y", "env")]     # already on the TRAINING scale
newd <- do.call(rbind, lapply(levels(dat$trait), function(sp)
  transform(grid, trait = sp)))
newd$trait       <- factor(newd$trait, levels = levels(dat$trait))
newd$cell_id     <- factor("po001", levels = levels(dat$cell_id))
newd$isdm_source <- factor("ebird", levels = levels(dat$isdm_source))
newd$log_effort  <- 0                     # relative intensity, NOT a count
newd$value       <- 0                     # the design matrix evaluates the LHS

p <- suppressWarnings(suppressMessages(
  predict(fit, newdata = newd, type = "response")))
E <- matrix(p$est, ncol = nlevels(dat$trait),
            dimnames = list(NULL, levels(dat$trait)))
cat("grid cells:", nrow(grid), " species:", ncol(E),
    " relative intensity:", f2(min(E), 3), "to", f2(max(E), 2), "\n")

## The saved map surfaces came from this same fit on this same grid.
cat("max |live prediction - saved surface|:", max(abs(E - md$est)), "\n")
```

Exactly zero: the figure below is this fit, not a stored approximation of it.

```{r map-plot, echo = FALSE, fig.width = 6.6, fig.height = 7.5, dpi = 200, out.width = "100%", fig.cap = "Relative intensity for all twelve species from one integrated fit, on a shared logarithmic colour scale. Canada Warbler is drawn large, carrying the axes in projected UTM kilometres, the sampled locations and the labelled scale; the other eleven are thumbnails ordered by the correlation of each surface with the flagship's, so the grid reads as a gradient from most like it to opposite rather than as an arbitrary sequence. Colour is never the only channel: isolines at the same levels as the colour-bar ticks are drawn on every panel, and the rail under each panel places that species' 5th to 95th percentile and median on the shared scale, because a shared ramp is exactly what compresses amplitude away. Cells outside the mesh hull are hatched rather than coloured: there the spatial field is exactly zero because fmesher returns an empty basis row, not because it was estimated to be small. The surfaces carry NO uncertainty of any kind. The simulated-data fence is drawn inside the canvas because a PNG lifted into a talk keeps its pixels and loses its caption.", fig.alt = "A large map panel for Canada Warbler beside a four-by-three grid of eleven smaller maps, one per species, all on a shared viridis colour scale for relative intensity on a log scale. The flagship panel carries UTM easting and northing axes in kilometres, white isolines, and small white circles marking all 360 sampled locations. The broad pattern is a bright high-intensity region in the north and a dark low-intensity region in the south-west. The eleven thumbnails show the same landscape expressed through different loadings, running from surfaces closely resembling the flagship to one close to its inverse. A short horizontal rail under each panel marks that species' fifth to ninety-fifth percentile with a notch at the median. A few cross-hatched cells sit in the corners, outside the mesh hull. Red text at the foot reads SIMULATED DATA, no uncertainty shown, no real observations."}
source("../../dev/isdm-precision/cawa12-map.R")
invisible(cawa12_map_plot(md))
```

The map is drawn from one seed. Across `r length(md$ok_seeds)` replicate
datasets on the same fixed landscape the drawn quantity replicates well --
median within-species correlation of the log surfaces
`r f2(median(md$stability$median_cor), 3)`, lowest single pairing
`r f2(min(md$stability$min_cor), 3)` -- which is a statement about the
surface's **stability**, not about its uncertainty.

## What the map does not carry

**No uncertainty, of any kind.** There is no interval behind any value in
the figure. `se.fit` is refused outright with `newdata`, and the refusal is
load-bearing rather than an oversight:

```{r se-fit-refusal}
cat(err_of(predict(fit, newdata = newd, type = "link", se.fit = TRUE)))
```

The training-row standard error that *is* available is conditional and
fixed-effect-only, and the latent-field reconstruction error it ignores is
exactly what dominates on a map. Map-scale uncertainty needs a
random-effect-aware construction that this package does not have yet.

**Nothing outside the mesh.** The grid deliberately extends 6 per cent
beyond the sampled extent, the way a real prediction map would, and
`predict()` says so:

```{r hull-warning}
cat(warn_of(suppressMessages(predict(fit, newdata = newd, type = "response"))))
```

Out there `fmesher` returns an all-zero basis row, so the field is **exactly
zero** -- indistinguishable, on a coloured surface, from a region where the
field is genuinely near zero. A blank corner reads as a cold one.
`r sum(md$out_of_hull)` of `r nrow(md$grid)` cells are in that state here,
and the figure hatches them instead of colouring them. `make_mesh()` rejects
such rows at *fit* time, so the honest thing is for `predict()` not to be
quietly more permissive. Clip the grid, or extend the mesh -- do not suppress
the warning, which is what the chunk above did purely to keep the figure
inline.

**A little extrapolation in the covariate.**
`r md$env_grid_cells_outside_training_range` of `r nrow(md$grid)` grid cells
carry an `env` value outside the training range, a consequence of that same 6
per cent padding.

**No claim that the slopes behind it are unbiased.** The recovery envelope
earlier shows a consistent downward bias, and
[What this article does not establish] carries that caveat with the right
statistics attached.

# Four things that look like modelling choices and are not

Each of the following is an **algebraic identity**, not a tendency measured
on a sample. It holds exactly, for the reason given, and no number of
replicates would make it more or less true. They are collected here because
each has cost somebody an afternoon.

They are demonstrated on a deliberately small model -- three of the twelve
species -- because the size of a fit cannot change an identity, and a faster
page is a page more people read.

```{r small}
sub <- droplevels(dat[dat$trait %in% c("CAWA", "OVEN", "TEWA"), ])
fam <- isdm_sources(ebird = poisson(), abmi = binomial(link = "cloglog"))
msh <- make_mesh(sub, c("X", "Y"), cutoff = 40)
FIT <- function(f, ...) suppressMessages(suppressWarnings(gllvmTMB(
  f, data = sub, trait = "trait", unit = "cell_id", family = fam,
  silent = TRUE, ...)))
```

## 1. An offset constant within an arm is absorbed by the source intercept

An `offset()` is a known multiplier with no estimated coefficient -- data,
not a parameter -- so it cannot compete with anything for fit. But if it
never varies within an arm, the arm's own intercept reproduces it exactly,
and adding it changes nothing:

```{r offset-absorbed}
sub$const_off <- ifelse(sub$isdm_source == "ebird", 1.0, 0)   # constant per arm
o_none  <- FIT(value ~ 0 + trait + trait:env + isdm_source)$opt$objective
o_const <- FIT(value ~ 0 + trait + trait:env + isdm_source +
                 offset(const_off))$opt$objective
o_vary  <- FIT(value ~ 0 + trait + trait:env + isdm_source +
                 offset(log_effort))$opt$objective
c(no_offset = o_none, constant_offset = o_const, varying_offset = o_vary)
cat("relative difference, constant offset vs none:",
    format((o_const - o_none) / o_none, digits = 3), "\n")
```

Identical to well inside the optimiser's own tolerance -- the two models are
the same model. The real `log_effort` offset varies *within* each arm, which
is why the simulation was built that way and why removing it changes the
objective. If your own effort is one number per source, your offset is a
relabelling of your source indicator, and an article teaching the offset
idiom is teaching you something you are not using.

## 2. The unit column of a prediction grid does not enter the prediction

The design matrix requires *some* valid factor level there; this model's only
random effect is the spatial field, projected through `(X, Y)`. So the choice
is arbitrary, and provably so:

```{r unit-column}
alt <- newd; alt$cell_id <- factor("sv060", levels = levels(dat$cell_id))
p_alt <- suppressWarnings(suppressMessages(
  predict(fit, newdata = alt, type = "response")))
cat("max |difference| between two different unit levels:",
    max(abs(p$est - p_alt$est)),
    " over a prediction range of", f2(diff(range(p$est)), 2), "\n")
```

Exactly zero -- an opportunistic cell's label and a survey station's, same
answer. **This stops being true the moment your model has a genuine
unit-level random effect**, and then the package tells you rather than
guessing:

```{r unit-re-warning}
set.seed(3)
n <- 150; ev <- rnorm(n); site_eff <- rnorm(n, 0, 0.8)
toy <- do.call(rbind, lapply(1:2, function(j) data.frame(
  site = factor(seq_len(n)), env = ev, trait = c("a", "b")[j],
  value = rpois(n, exp(0.5 + c(0.9, -0.4)[j] * ev + site_eff)))))
toy$trait <- factor(toy$trait)
toy_fit <- suppressWarnings(suppressMessages(gllvmTMB(
  value ~ 0 + trait + trait:env + (1 | site), data = toy, trait = "trait",
  unit = "site", family = poisson(), silent = TRUE)))
toy_grid <- expand.grid(env = seq(-2, 2, length.out = 10),
                        trait = levels(toy$trait))
toy_grid$value <- 0
toy_grid$site  <- factor(levels(toy$site)[1], levels = levels(toy$site))
cat(warn_of(predict(toy_fit, newdata = toy_grid, type = "link")))
```

It names the tier it dropped and offers two correct alternatives:
`newdata = NULL` for the full conditional predictor, or `re_form = ~0` for
fixed effects only. **Neither is what this article's map uses**, deliberately:
`re_form = ~0` would zero the spatial field too, and that field is the entire
point of the map.

## 3. The grouping token on a `spatial_*()` term is inert

There is no column called `coords` in the data, and there does not need to
be. The geometry comes entirely from `mesh =`; the token to the right of the
`|` is never read.

```{r token-inert}
sp_obj <- function(tok) FIT(stats::as.formula(paste0(
  "value ~ 0 + trait + trait:env + isdm_source + offset(log_effort) + ",
  "spatial_latent(0 + trait | ", tok, ", d = 1)")), mesh = msh)$opt$objective
o <- c(coords = sp_obj("coords"), banana = sp_obj("banana"),
       cell_id = sp_obj("cell_id"))
o
cat("largest difference:", max(o) - min(o), "\n")
```

`cell_id` is a real column with a real meaning, and it makes no difference
whatsoever -- which is the direction that matters. Writing
`spatial_latent(0 + trait | site)` with a real `site` column and a real
intention will *also* be ignored, and nothing will say so. Tracked as
[issue #1163](https://github.com/itchyshin/gllvmTMB/issues/1163).

## 4. A mesh passed without a spatial term is ignored

`mesh =` is consumed by the `spatial_*()` term and by nothing else. With no
such term in the formula, passing a mesh -- of either kind -- changes
nothing:

```{r mesh-ignored}
raw_bnd  <- fmesher::fm_extensions(cbind(sub$X, sub$Y), convex = c(30, 60))
raw_mesh <- fmesher::fm_mesh_2d(loc = cbind(sub$X, sub$Y), boundary = raw_bnd,
                                max.edge = c(40, 120))
m <- c(no_mesh   = FIT(value ~ 0 + trait + trait:env + isdm_source +
                         offset(log_effort))$opt$objective,
       raw_mesh  = FIT(value ~ 0 + trait + trait:env + isdm_source +
                         offset(log_effort), mesh = raw_mesh)$opt$objective,
       make_mesh = FIT(value ~ 0 + trait + trait:env + isdm_source +
                         offset(log_effort), mesh = msh)$opt$objective)
m
cat("largest difference:", max(m) - min(m), "\n")
```

A reader who builds a mesh and then forgets to add the `spatial_*()` term
gets a converged, ordinary-looking, entirely non-spatial fit, with no error
and no warning. Tracked as
[issue #1165](https://github.com/itchyshin/gllvmTMB/issues/1165). Check that
a `spatial_*()` term is actually in your formula rather than trusting the
absence of a complaint.

The *other* raw-mesh case is loud and correct -- a raw `fm_mesh_2d()` object
*with* a spatial term is refused. That matters because most `fmesher` and
INLA tutorials teach you to build one by hand:

```{r raw-mesh-error}
cat(err_of(FIT(value ~ 0 + trait + trait:env + isdm_source + offset(log_effort) +
                 spatial_latent(0 + trait | coords, d = 1), mesh = raw_mesh)))
```

# The one thing the likelihood cannot tell you

Those four are things the model absorbs silently and correctly. Here is one
it absorbs silently and **wrongly**.

The mesh must be built on projected coordinates. At this domain's latitude a
degree of latitude is about 1.7 times a degree of longitude in kilometres, so
an isotropic SPDE fitted on `lon`/`lat` imposes a correlation structure
stretched by that factor in one direction -- an artefact of the coordinate
system, not a property of the birds.

The instinct is to check this the way one checks any modelling choice: fit
both and compare. **That does not work.**

```{r projection-null}
pj <- readRDS("../../dev/isdm-precision/cawa12-projection.rds")
data.frame(seed        = pj$per_seed$seed,
           d_objective = round(pj$per_seed$d_obj, 2),
           err_lonlat  = round(pj$per_seed$err_lonlat, 4),
           err_utm     = round(pj$per_seed$err_utm, 4))
c(mean_d_obj = round(pj$mean_d_obj, 2), se = round(pj$se_d_obj, 2),
  sign_flips = pj$sign_flips, utm_better = pj$utm_better_n, seeds = pj$n_seeds)
```

Six seeds, the same twelve-species design as everything else on this page, and
the two meshes matched at **`r pj$per_seed$nodes_lonlat[1]`** and
**`r pj$per_seed$nodes_utm[1]`** nodes so neither is handed extra latent
dimensions.

Read those two rows together, because separately each is misleading.

**The likelihood has a strong and consistent opinion.** The unprojected mesh
wins by `r abs(round(pj$mean_d_obj, 1))` objective units, SE
`r round(pj$se_d_obj, 1)`, with the sign the same in **all
`r pj$n_seeds` seeds**. This is not a close call and it is not noise.

**Its opinion does not track accuracy.** On the quantity the model exists to
estimate, the two meshes are indistinguishable: the projected mesh recovers the
slopes better in `r pj$utm_better_n` of `r pj$n_seeds` seeds, with mean
absolute errors of `r round(pj$mean_err_lonlat, 3)` and
`r round(pj$mean_err_utm, 3)`. A coin flip.

So the warning is not that the likelihood is uninformative. **It is that the
likelihood is confident about a choice it has no business deciding.** Two
meshes are two discretisations of the same continuous field; they carry
different latent dimensions and are not nested, so their objectives are not on
a common scale. A 35-unit gap that never changes sign is the signature of a
number measuring the discretisation rather than the fit to the data.

Comparing log-likelihoods or AIC across meshes will therefore not catch a
projection mistake — and worse, it will answer confidently. Project first
because the geometry is otherwise wrong, not because the fit will tell you.

**An honest disagreement between designs, left unresolved.** The companion
precision article measures a clear recovery advantage for the projected mesh
(five seeds of five, on a three-species design whose field is isotropic in true
distance). This twelve-species design does not reproduce that advantage. Both
measurements are reported as run; which one resembles your data depends on
whether your latent structure is isotropic in kilometres, and neither article
resolves it for you.

`add_utm_columns()` warns rather than fails when a domain straddles a UTM zone
boundary, which is a real risk for any study area wider than six degrees:

```{r zone-warning}
straddling <- data.frame(lon = c(-114.5, -112.5), lat = c(53.6, 55.4))
cat(warn_of(add_utm_columns(straddling, ll_names = c("lon", "lat"))))
```

This domain stays on the zone-11 side of that line, which is why the zone
count printed above is 1.

# If your opportunistic arm has no counts

The arm above models counts, which eBird checklists do carry. Many
opportunistic downloads do not: a raw GBIF export is often **presence-only**
-- a location, a species, and nothing to put in `value`. There is no natural
count to invent, and fitting one as if it were a count is not an option.

The standard route (Warton & Shepherd 2010; Renner et al. 2015) is the
Berman-Turner quadrature device: an inhomogeneous Poisson process fitted as a
**weighted Poisson GLM**, in which presence rows carry a tiny weight `w` with
response `y = 1/w`, and a grid of background quadrature nodes carries weight
equal to its cell area with response `y = 0`. `gllvmTMB()` accepts exactly
that through `weights =`:

```{r po-recipe, eval = FALSE}
po <- rbind(
  data.frame(trait = sp, env = env_at_presences, y = 1e6, w = 1e-6),
  data.frame(trait = sp, env = env_at_nodes,     y = 0,   w = cell_area)
)
gllvmTMB(y ~ 0 + trait + trait:env, data = po, trait = "trait",
         unit = "obs_id", family = poisson(), weights = po$w)
```

The package's own fence on such a fit, captured at runtime rather than
paraphrased:

```{r po-warning, echo = FALSE, comment = ""}
cat(qd$fence$weights_warning)
```

## Does the device recover the truth?

Twenty replicate datasets at these settings, `r nrow(qd$replicates)` species,
all `r sum(qd$replicate_fit_health$ok)` fits accepted under the #1167 rule:

```{r po-recovery, echo = FALSE}
rp <- qd$replicates
data.frame(species = rp$species, truth = round(rp$truth, 3),
           mean_est = round(rp$mean, 3), bias = round(rp$bias, 4),
           mc_sd_of_one = round(rp$sd, 4),
           t = round(rp$t, 2), p = round(rp$p, 3))
big_beta <- abs(rp$truth) > 0.4
```

No species differs detectably from its true value: the smallest *p* is
`r f2(min(rp$p), 3)`, uncorrected, and every 95 per cent interval contains
the truth. The pooled test is far more powerful, and it is the right test for
the failure people actually fear -- a per-seed regression through the origin
of the twelve estimates on the twelve truths, which equals 1 under no uniform
bias: `r f2(qd$pooled_power$attenuation_mean, 5)`, 95 per cent CI
[`r f2(qd$pooled_power$ci[1], 4)`, `r f2(qd$pooled_power$ci[2], 4)`],
*p* = `r f2(qd$pooled_power$p, 3)`. That design would have detected a uniform
multiplicative distortion of `r f2(qd$pooled_power$mde_80_pct, 1)` per cent
or more. It detected nothing.

Two things it does **not** establish, both worth stating plainly.

**It certifies the device and the optimiser, not the quadrature.** At these
settings the quadrature grid *coincides* with the lattice the point pattern
was thinned from, so the Berman-Turner objective there is the exact likelihood
of the generating model -- verified against a direct `glm()` on the lattice
counts to a maximum absolute difference of
`r format(max(sapply(qd$exactness, function(z) z$max_abs_diff)), digits = 2)`.
There is no quadrature error there to find. The sweep below is what measures
quadrature error, on grids deliberately decoupled from that lattice.

**Relative-bias power is worthless where the true slope is near zero.** The
minimum bias this design could detect is `r f2(min(rp$mde_80), 3)` to
`r f2(max(rp$mde_80), 3)` on the slope scale. For the `r sum(big_beta)`
species with `|beta| > 0.4` that is
`r f2(min(rp$mde_80_pct[big_beta]), 1)` to
`r f2(max(rp$mde_80_pct[big_beta]), 1)` per cent of truth; for the
`r sum(!big_beta)` whose true slope sits near zero it is
`r f2(min(rp$mde_80_pct[!big_beta]), 0)` to
`r f2(max(rp$mde_80_pct[!big_beta]), 0)` per cent, which is another way of
saying this design has no relative-bias power there at all. Quote absolute
numbers for those species, not percentages.

## How dense should the quadrature grid be?

Measured with the grids **decoupled** from the generating lattice, so that
what is measured is genuine quadrature error; 12 seeds per rung:

```{r po-density, echo = FALSE}
sd_ <- qd$sweep_decoupled$summary
npp <- qd$nodes_per_presence
mc  <- mean(qd$replicates$sd)
data.frame(nodes = sd_$n_nodes,
           nodes_per_presence = paste0(round(npp$min, 1), "-", round(npp$max, 1)),
           mean_bias = round(sd_$mean_bias, 4),
           mean_shift_vs_finest = round(sd_$mean_abs_shift_vs_finest, 4),
           pct_of_MC_sd = paste0(round(100 * sd_$mean_abs_shift_vs_finest / mc), "%"))
```

Estimates approach the truth **from below, monotonically**, and sub-linearly in
the number of quadrature nodes: fitting the shift against node count on this
sweep gives an exponent of about **-0.58**, not the **-1** that "the error
halves when you double the nodes" would imply. Doubling the nodes buys roughly
a third off the shift, not a half — so budget accordingly. Judged against the
only scale that matters to a reader -- the Monte
Carlo SD of a single estimate, `r f2(mc, 4)` -- they are stable for practical
purposes from about
`r sd_$n_nodes[which(sd_$mean_abs_shift_vs_finest / mc < 0.10)[1]]` nodes,
and further movement past about
`r sd_$n_nodes[which(sd_$mean_abs_shift_vs_finest / mc < 0.05)[1]]` nodes is
under 4 per cent of one estimate's noise.

Two caveats rather than one rule. **"Nodes per presence" is the right currency
and it varies across species** -- from
`r f2(npp$min[npp$n_nodes == 3600], 1)` to
`r f2(npp$max[npp$n_nodes == 3600], 1)` per presence at 3,600 nodes here,
because rare species get more nodes per presence than common ones. And **the
sweep never flattens**: it converges asymptotically, so "increase the density
until the estimates stop moving" really means *"until they move less than your
Monte Carlo noise"*.

## It does not combine with the two-arm fit

```{r po-refusal, echo = FALSE, comment = ""}
cat(qd$fence$isdm_refusal_unit$msg)
```

Confirmed at runtime on this exact model, with unit weights and with genuinely
non-unit ones -- both abort, so it is not a special case for one of them. If
your opportunistic arm is genuinely presence-only, fit it with the quadrature
device **on its own**, the way the standalone
[`gbif-joint-intensity`](gbif-joint-intensity.html) article does. Combining a
quadrature-weighted arm with a structured arm inside `isdm_sources()` is not
supported.

# What this article does not establish

**No real data.** No GBIF and no ABMI records were fitted here. The structured
arm at usable coordinate precision needs a data agreement, and this page is
not a stand-in for that analysis.

**No positional-error correction.** Everything above assumes both arms know
where they are. Real structured-survey data frequently does not, and the
consequences are not small -- see the companion article
[*When one data source knows where it is and the other does not*](isdm-spatial-precision.html)
before fusing real fuzzed data. In particular, do not expect coarsening your
raster to fix it: Gabor et al. (2022), *Positional errors in species
distribution modelling are not overcome by the coarser grains of analysis*,
is the direct test of that intuition and it does not survive.

**No unbiasedness claim -- and the previous version of this caveat was
wrong.** A separate 30-replicate, 12-species recovery campaign against this
same machinery reported a positive mean error at *t* = 4.17, and an earlier
version of this article called it "small but statistically real". It is not
statistically real: that test treated 360 errors as independent when all
twelve species within a replicate share one estimated field.

```{r bias-clustering}
err      <- gt$results$err                  # 30 replicates x 12 species
rep_mean <- tapply(err, gt$results$rep, mean)
naive <- t.test(err); clustered <- t.test(rep_mean)
cat("as 360 independent errors:  mean", f2(mean(err), 4),
    " t", f2(naive$statistic, 2), " p", format(naive$p.value, digits = 2),
    "\nby replicate (the resampling unit):  mean", f2(mean(rep_mean), 4),
    " t", f2(clustered$statistic, 2), " p", f2(clustered$p.value, 3), "\n")
cat("between-replicate SD of the mean error:", f2(sd(rep_mean), 4),
    "-- were species independent within a replicate it would be",
    f2(sd(err) / sqrt(12), 4), "\n")
```

The correct test does not reject. The signed bias is also a property of
**one** landscape and one draw of species truths, both fixed across those
replicates, so it has not been shown to be a mechanism rather than noise.
What remains true, and is visible in this article's own recovery table, is
that the slopes come back **shrunk**; nothing here entitles anyone to call
them unbiased.

**No interval calibration for anything on the map.** Coverage was measured for
`unique = TRUE` against the base model on the *training* rows above, where it
overshoots. Nothing here measures coverage of a mapped surface, and `se.fit`
is refused with `newdata` for that reason.

**The rank result is measured at twelve species, on one design.** AIC's 10/10
and the loading heuristic's 1/10 are properties of this landscape, this
rank-2 truth and this species count. The mechanism behind the heuristic's
failure -- rotational non-identifiability of the loading columns -- is
general; the hit rates are not.

**The overdispersion result is capped by issue #1167.** It could not be tested
at count magnitudes above the screen, which is exactly where detection of
`Psi` would improve.

# Recap

`isdm_sources()` names the two observation processes. The arms sit at disjoint
locations and still share one field, because the field lives on a mesh built
over the union of both and every row is projected onto it. Project the
coordinates before meshing, and do not expect the likelihood to tell you
whether you did. Check `iterations`, not just `convergence`, and keep the
counts well under the ceiling. Choose the rank of the shared field by AIC, not
by hunting for a small loading column. Reach for `unique = TRUE` when an arm
is Poisson and plainly overdispersed -- for the point estimates rather than
the intervals, and expect it to cost you Hessians. Zero the offset on the
prediction grid, keep the covariate on its training scale, and hatch whatever
falls outside the mesh.

Point estimates and a surface. No intervals -- and the article says so, rather
than letting a smooth picture imply otherwise.

## References

Gabor, L., et al. (2022). Positional errors in species distribution modelling
are not overcome by the coarser grains of analysis.

Renner, I. W., Elith, J., Baddeley, A., Fithian, W., Hastie, T., Phillips,
S. J., Popovic, G., & Warton, D. I. (2015). Point process models for
presence-only analysis. *Methods in Ecology and Evolution.*

Warton, D. I., & Shepherd, L. C. (2010). Poisson point process models solve
the "pseudo-absence problem" for presence-only data in ecology. *Annals of
Applied Statistics.*
