---
title: "Joint species distribution model (binary GLLVM)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Joint species distribution model (binary GLLVM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

A reduced-rank binary GLLVM compresses a many-column 0/1 occurrence
matrix into a few latent factors. The classic ecology version is a
**joint species distribution model**: rows are sites, columns are
species, the entries are presence / absence, and the latent factors
capture residual co-occurrence beyond what's explained by
environmental predictors. The same `gllvmTMB(family = binomial())`
call also fits any binary `unit × trait` data — IRT-style item
responses, presence/absence of attributes, etc.

## The model

For binary $y_{it} \in \{0, 1\}$ with logit link,

$$
\mathrm{logit}\bigl(P(y_{it} = 1)\bigr) \;=\; \mathbf x_i^\top
\boldsymbol\beta_t \;+\; \boldsymbol\lambda_t^{\!\top} \mathbf u_i,
\qquad \mathbf u_i \sim \mathcal N(\mathbf 0, \mathbf I_d).
$$

Two distinct covariance objects matter here, and **conflating them is
the most common notation trap in JSDM writing**.

### 1. Between-unit shared covariance (estimated by the GLLVM)

The covariance among traits / species induced by the latent factors
$\mathbf u_i$ is

$$
\boldsymbol\Sigma_B \;=\; \boldsymbol\Lambda \boldsymbol\Lambda^\top.
$$

That's the part the GLLVM **estimates** from cross-trait
co-occurrence on the latent scale. With reduced rank $d \ll T$ it is
identified by the off-diagonal (cross-trait) co-occurrence pattern;
the diagonal is then constrained by the rank.

### 2. Total latent-liability covariance (used for correlations / ICC)

To convert $\boldsymbol\Sigma_B$ into something on a unified
**latent-liability** scale (so we can compute latent-scale
correlations, ICC, communality, $R^2$, …), we use the threshold
representation:

$$
y^*_{it} \;=\; \eta_{it} + \varepsilon_{it},
\qquad \varepsilon_{it} \sim \text{Logistic}(0, 1),
\qquad \mathrm{Var}(\varepsilon) = \pi^2/3.
$$

with $y_{it} = \mathbf 1\{y^*_{it} > 0\}$. The total latent-liability
covariance is then

$$
\boldsymbol\Sigma_{\text{latent}}
\;=\;
\underbrace{\boldsymbol\Sigma_B}_{\text{estimated}}
\;+\;
\underbrace{(\pi^2 / 3)\, \mathbf I}_{\text{fixed by link}}.
$$

The $(\pi^2/3) \mathbf I$ term is **not estimated** — it is the
distribution-specific variance of the standard logistic
($\pi^2/3 \approx 3.29$). For probit it's $1$; for cloglog it's
$\pi^2/6$. It enters whenever you want to put $\boldsymbol\Sigma_B$
on the conventional latent-liability scale (Niku et al. 2017;
Nakagawa & Schielzeth 2010).

In `gllvmTMB`:

* `extract_Sigma(fit, level = "unit", part = "shared")` returns
  $\boldsymbol\Sigma_B = \boldsymbol\Lambda\boldsymbol\Lambda^\top$.
* `extract_Sigma(fit, level = "unit", part = "total")` returns
  $\boldsymbol\Sigma_{\text{latent}} = \boldsymbol\Sigma_B + (\pi^2/3) \mathbf I$.

`extract_correlations()` uses `part = "total"` by default, so the
correlations come out on the latent-liability scale.

### Why this example omits `unique()`

For Gaussian fits, the canonical decomposition pairs
`latent + unique`:

$$\boldsymbol\Sigma_B = \boldsymbol\Lambda\boldsymbol\Lambda^\top + \boldsymbol\Psi$$

where $\boldsymbol\Psi = \mathrm{diag}(\psi_{tt})$ is the per-trait
unique variance estimated from residual variation. A single-trial
Bernoulli response is different: the link already supplies the
latent-scale residual variance ($\pi^2/3$ for logit, 1 for probit).
There is no separate observation-level $\psi$ term for the data to
identify.

That is why the worked example below uses only
`latent(0 + trait | site, d = 2)`. The `latent + unique` binary path
is still useful as a diagnostic or development check, but the current
validation register marks that paired binary regime as RE-09
`partial`, not `covered`. For a public JSDM example, the honest
default is therefore:

| Goal | Recommended term |
|---|---|
| Model residual species co-occurrence in 0/1 data | `latent(0 + trait | site, d = K)` |
| Add a per-trait residual variance to Gaussian or other continuous traits | `latent(...) + unique(...)` |
| Diagnose a boundary or over-parameterised binary fit | try `unique(...)` as a sensitivity check, then report the boundary status |

Bottom line: for **pure single-trial binary** JSDMs, start with
`latent()` alone and treat `unique()` as a diagnostic path.

### What about `dep()` and `indep()` for binary?

The `dep()` and `indep()` keywords are part of the general covariance
grammar, but this article does not use them as binary-JSDM teaching
defaults. The validation-debt register currently marks the `dep()`
and `indep()` rows for the unit tier as FG-08 and FG-07 `partial`
because their stronger recovery evidence is Gaussian-only.

For binary species-occurrence data, choose the term by the biological
question:

| Question | Term to start with |
|---|---|
| Which species share residual co-occurrence after the environment is accounted for? | `latent(0 + trait | site, d = K)` |
| Do species vary independently, with no residual co-occurrence target? | `indep()` can express that model, but it is not the JSDM focus here |
| Do I need a saturated covariance for a small species set? | `dep()` is a possible sensitivity fit, but it is not the validated public path for this binary example |

So the recommendation here is narrow: use `latent(d = K)` for the
worked binary JSDM and treat `dep()` / `indep()` comparisons as
sensitivity checks until the binary-specific validation rows move
from `partial` to `covered`.

### How `extract_Sigma()` handles non-Gaussian fits

`extract_Sigma(fit, part = "total")` looks up each trait's family /
link, and adds the **per-trait latent-scale residual** $\sigma^2_d$ to
the diagonal of $\boldsymbol\Sigma$. The internal helper is
`gllvmTMB:::link_residual_per_trait(fit)`, which uses these formulas
from Nakagawa & Schielzeth (2010) + Nakagawa, Johnson & Schielzeth
(2017):

| Family / link | $\sigma^2_d$ added per trait |
|---|---|
| `gaussian` (identity) | $0$ |
| `binomial(logit)` | $\pi^2/3 \approx 3.290$ |
| `binomial(probit)` | $1$ |
| `binomial(cloglog)` | $\pi^2/6 \approx 1.645$ |
| `poisson(log)` | $\log(1 + 1/\hat\mu_t)$  (lognormal-Poisson) |
| `nbinom2(log)` | $\psi'(\hat\phi_t)$ (trigamma; Stoklosa et al. 2022) |
| `tweedie(log)` | $\log(1 + \hat\phi_t \hat\mu_t^{p-2})$ (Tweedie 1984) |
| `Beta(logit)` | $\psi'(\mu\phi) + \psi'((1-\mu)\phi)$ (Smithson & Verkuilen 2006) |
| `betabinomial(logit)` | $\psi'(\mu\phi) + \psi'((1-\mu)\phi) + \pi^2/3$ |
| `delta_lognormal` | $\hat\sigma^2_{\text{logn}} + \pi^2/3$ |

Each trait gets the residual implied by its family; the diagonal of
$\boldsymbol\Sigma$ is therefore on a unified latent scale. Pass
`link_residual = "none"` to suppress this addition and recover the
bare $\boldsymbol\Lambda\boldsymbol\Lambda^\top + \boldsymbol\Psi$ form.

The article body below fits a **pure binomial** JSDM. End-to-end
mixed-family rigour (validated extractors on `family = list(...)`
fits across the 15-family matrix) is M1 milestone work — see
[validation-debt register MIX-03..MIX-08](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/35-validation-debt-register.md#section-7--mixed-family-fits).

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

## Load and fit

The article uses a shipped teaching fixture rather than building the
data inside the vignette. It contains a complete `Site x Species`
occurrence matrix: 120 sites, 8 species, 2 latent residual
co-occurrence axes, and an environmental covariate `env_1` with
species-specific responses. The true loadings put the species into
four niche quadrants (dry-low, dry-high, wet-low, wet-high), with two
species per quadrant.

```{r jsdm-data}
jsdm_path <- system.file(
  "extdata", "examples", "joint-sdm-example.rds",
  package = "gllvmTMB"
)

source_candidates <- file.path(
  c(".", "..", "../.."),
  "inst", "extdata", "examples", "joint-sdm-example.rds"
)
input_file <- knitr::current_input(dir = TRUE)
if (!is.null(input_file)) {
  source_candidates <- c(
    source_candidates,
    file.path(
      dirname(dirname(dirname(input_file))),
      "inst", "extdata", "examples", "joint-sdm-example.rds"
    )
  )
}
candidate_paths <- c(jsdm_path, source_candidates)
candidate_paths <- candidate_paths[nzchar(candidate_paths)]
jsdm_path <- candidate_paths[file.exists(candidate_paths)][1]
if (is.na(jsdm_path)) {
  stop("Could not find joint-sdm-example.rds.")
}
jsdm <- readRDS(jsdm_path)

df <- jsdm$data_long
df_wide <- jsdm$data_wide
```

The same JSDM fits in both long format and the `traits(...)` wide
format. The long form is shown first; the wide form is the
formula-API equivalent on a pivoted data frame. Both reach the
same engine.

```{r jsdm-fit}
# Long format -- one row per (site, species) observation
fit_jsdm <- gllvmTMB(
  jsdm$formula_long,
  data   = df,
  trait  = jsdm$fit_args$trait,
  unit   = jsdm$fit_args$unit,
  family = jsdm$fit_args$family
)
fit_jsdm
```

```{r jsdm-fit-wide}
# Wide format -- one row per site, one column per species. The
# `traits(...)` LHS names the species columns; the compact RHS
# shorthand uses `1 + env_1`, which expands internally to
# `0 + trait + (0 + trait):env_1`.
fit_jsdm_wide <- gllvmTMB(
  jsdm$formula_wide,
  data   = df_wide,
  unit   = jsdm$fit_args$unit,
  family = jsdm$fit_args$family
)
all.equal(logLik(fit_jsdm), logLik(fit_jsdm_wide), tolerance = 1e-8)
```

### Long or wide data shape

JSDM is the package's natural long-format use case because each row
already carries `(site, species)` and the response factor `trait`
indexes species. When the data already live as one row per site and
one binary column per species, the `traits(...)` wide form is the same
model through the same `gllvmTMB()` entry point. In either shape, true
absences should be explicit zeros. Missing response cells are not a
safe shortcut for absences, because missingness can also mean the site
or species was not surveyed.

### Where loading constraints fit

The loading matrix $\boldsymbol\Lambda$ is usually exploratory in a
JSDM: its rotation is arbitrary, while
$\boldsymbol\Lambda\boldsymbol\Lambda^\top$ is the interpretable
co-occurrence target. If you have an a-priori hypothesis that a
species group defines an axis, or that a species should have zero
loading on an axis, that is a **confirmatory loading** question rather
than the default species-distribution workflow.

`gllvmTMB` supports that confirmatory path with
`lambda_constraint = list(unit = M)`, where `NA` means "estimate this
loading" and a number means "pin this loading during fitting." Binary
IRT-style loading constraints are covered by the LAM-03 and LAM-04
validation rows, but an ecology-first JSDM article should introduce
them only as an optional follow-up after the reader understands
correlations, $\boldsymbol\Sigma$, and ordination.

## Latent-scale residual species correlations

Pairwise species correlations on the latent-liability scale. The
default `method = "fisher-z"` gives Fisher-transform Wald CIs that
are fast (seconds) and bounded inside $[-1, 1]$ by construction.
`method = "wald"` is an accepted alias for the same Fisher-z path.
Two heavier alternatives are also available: `method = "bootstrap"`
re-fits the model under a parametric bootstrap for a full
sampling-distribution CI, and `method = "profile"` inverts the
profile likelihood pair-by-pair (cost scales as $O(T(T-1)/2)$
constrained refits).

One scale caveat is worth stating up front, because it trips people
up on binary fits. The default `link_residual = "auto"` puts the
correlation on the **latent-liability** scale — it adds the fixed
logistic residual $\pi^2/3$ to the diagonal of
$\boldsymbol\Sigma_B$ before forming the correlation, which is the
interpretable co-occurrence quantity for a JSDM. `fisher-z`,
`wald`, and `bootstrap` all honour that argument. The current
`profile` path, however, always reports the correlation of the
**bare** $\boldsymbol\Sigma_B = \boldsymbol\Lambda\boldsymbol\Lambda^\top$
(it ignores `link_residual`), so its point estimate is the
undiluted latent-factor correlation, not the latent-liability one.
The worked comparison below makes the difference explicit and shows
how to line the methods up like-for-like.

**What is and isn't certified.** The point machinery and the three
CI methods are validated for *behaviour* — they run and return the
intervals their definitions imply. **Calibrated coverage is a
separate, still-open question.** Even for Gaussian fits the
coverage study clears the $\geq 94\%$ target only at $d = 1$ and
$d = 3$ (most cells of the grid still fall short), and **mixed-family
fits — `family = list(...)` combinations such as Gaussian + binomial
— have measured coverage of only 0.55–0.82, well below target.**
Treat the intervals in this article as point-correct but *not*
coverage-certified, and do **not** read the mixed-family case as
ready: that is a known engine deficiency, not a near-complete
feature. See the
[validation-debt register](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/35-validation-debt-register.md)
(CI-08, CI-10) for the live status.

```{r jsdm-cor}
corr_rows <- extract_correlations(
  fit_jsdm,
  tier = "unit",
  method = "fisher-z",
  link_residual = "auto"
)
corr_rows
```

The matrix view below shows the same point estimates as a
species-by-species map. For a binary JSDM, the useful first picture is
the **latent-liability correlation** pattern after adding the fixed
logistic link residual to the denominator; raw
$\boldsymbol\Sigma_B$ entries can be dominated by loading scale and are
harder to interpret. The Fisher-z interval bounds stay in `corr_rows`
rather than being printed inside every matrix cell.

```{r jsdm-correlation-heatmap, fig.width = 6.8, fig.height = 5.8, fig.cap = "Residual species correlations from the fitted binary JSDM. Cells show Fisher-z point estimates on the latent-liability scale after accounting for env_1 and the fixed logistic link residual; interval bounds are available in the preceding table.", fig.alt = "Heatmap of pairwise residual species correlations for eight species. Blue cells indicate positive residual co-occurrence, red cells indicate negative residual co-occurrence, and diagonal self-correlations are one."}
plot_correlations(
  corr_rows,
  style = "heatmap",
  matrix_layout = "by_level",
  label_type = "estimate",
  label_digits = 2,
  include_diagonal = TRUE,
  title = "Residual species correlations",
  subtitle = "Latent-liability correlations after accounting for env_1."
) +
  ggplot2::labs(caption = NULL)
```

Use these correlations, rather than the signs of individual loadings,
for the rotation-invariant co-occurrence story.

## Confidence intervals on the binary latent scale

This section works through the three CI methods on two estimands —
a pairwise species **correlation** and a per-species **ICC** — to
show what each method returns and where it currently stops short.
The bootstrap chunks re-fit the model dozens of times; they are run
here at a deliberately small `nsim` for speed and are slower than
the Fisher-z default. Read the caveat box above first: the point
estimates are trustworthy, the *coverage* of these intervals is
not yet certified, and the mixed-family case is failing.

### Three CI methods for one correlation

Take a single species pair (`sp_1`, `sp_2`) and ask for the same
correlation three ways. Keeping `link_residual = "auto"` for the
Fisher-z and bootstrap calls holds them on the latent-liability
scale, so their point estimates agree and only the interval differs:

```{r jsdm-cor-methods}
pair_ij <- c("sp_1", "sp_2")

cor_fisher <- extract_correlations(
  fit_jsdm, tier = "unit", pair = pair_ij,
  method = "fisher-z", link_residual = "auto"
)
cor_boot <- extract_correlations(
  fit_jsdm, tier = "unit", pair = pair_ij,
  method = "bootstrap", nsim = 20, seed = 1, link_residual = "auto"
)
rbind(cor_fisher, cor_boot)[, c("trait_i", "trait_j",
                                "correlation", "lower", "upper", "method")]
```

The Fisher-z and bootstrap point estimates match because both add
the $\pi^2/3$ logistic residual to the diagonal; the bootstrap
interval is the empirical 2.5/97.5 percentile across the refits.

The `profile` method is the one that reports a *different* number,
because it currently works on the bare $\boldsymbol\Sigma_B$ scale.
To compare it fairly, request the Fisher-z interval with
`link_residual = "none"` so both sit on the same (undiluted) scale:

```{r jsdm-cor-profile}
cor_profile <- extract_correlations(
  fit_jsdm, tier = "unit", pair = pair_ij,
  method = "profile"
)
cor_fisher_bare <- extract_correlations(
  fit_jsdm, tier = "unit", pair = pair_ij,
  method = "fisher-z", link_residual = "none"
)
rbind(cor_profile, cor_fisher_bare)[, c("trait_i", "trait_j",
                                        "correlation", "lower", "upper", "method")]
```

Now the two point estimates agree: the profile correlation matches
the bare-scale Fisher-z value. The profile *interval* is a
likelihood-ratio inversion around that point — but note it can come
back as `NA` when the constrained refit fails to bracket the bound
(a near-boundary correlation here is a common cause), which is the
same not-yet-calibrated CI story flagged above. The take-home is
*not* that one method is wrong, but that **`profile` and the default
latent-liability correlations live on different scales** until the
profile path learns to honour `link_residual`. For the interpretable
JSDM co-occurrence story, prefer the latent-liability scale
(`auto`), and use Fisher-z for speed or the bootstrap when you want
an empirical sampling-distribution interval.

### ICC (communality) on the latent scale

For a single-trial binary JSDM, the natural intra-class quantity is
the share of each species' latent-liability variance that is shared
co-occurrence rather than the fixed logistic residual:

$$
\mathrm{ICC}_t \;=\;
\frac{(\boldsymbol\Lambda\boldsymbol\Lambda^\top)_{tt}}
     {(\boldsymbol\Lambda\boldsymbol\Lambda^\top)_{tt} + \pi^2/3}.
$$

With `latent()` and no `unique()` at the unit level there is no
estimated $\boldsymbol\Psi$, so `extract_communality(level = "unit",
link_residual = "auto")` returns exactly this ratio — the
proportion of the latent-liability variance that the shared factors
explain. (This is *not* the between-vs-within `extract_repeatability()`
ICC, which needs a paired `unique(0 + trait | site) + unique(0 +
trait | site_species)` design; a single-trial occurrence matrix has
no within-site replication for that, so `extract_repeatability()`
and `extract_ICC_site()` are not applicable to this fit.)

```{r jsdm-icc-point}
icc_point <- extract_communality(
  fit_jsdm, level = "unit", link_residual = "auto"
)
round(icc_point, 3)
```

A parametric bootstrap puts an interval on each species' ICC:

```{r jsdm-icc-boot}
icc_boot <- extract_communality(
  fit_jsdm, level = "unit", link_residual = "auto",
  ci = TRUE, method = "bootstrap", nsim = 20, seed = 1
)
icc_boot[, c("trait", "c2", "lower", "upper", "method")]
```

The intervals are wide — eight species and 120 sites is modest
information for a per-species variance share, and the upper bounds
for the more strongly-loading species run close to 1.

Two honest limitations on the ICC CI in the current build:

* **Wald is not implemented** for communality (it is a non-linear
  function of several parameters); requesting `method = "wald"`
  emits a note and falls back to the bootstrap.
* **Profile bounds are not yet delivered** for this proportion on a
  single-trial binary fit — the constrained refit fails to bracket
  and the interval returns `NA`. The point estimate is still the
  correct latent-scale ICC; only the profile *interval* is missing.

These gaps are exactly the open-coverage story flagged above: the
estimands compute, but calibrated interval machinery on the binary
and mixed-family scales is still in progress (CI-08, CI-10). Do not
present these intervals as finished, and do not extend them to
mixed-family `family = list(...)` fits, where the measured coverage
is failing.

## Ordination biplot

Site scores on the two latent axes, with species loadings shown as
arrows. Axes are residual co-occurrence after `env_1` has been
accounted for; species loading in the same direction tend to
co-occur (after accounting for the environment).

```{r jsdm-biplot, fig.width = 6.5, fig.height = 5, fig.cap = "Ordination biplot from the fitted binary JSDM, using varimax-rotated and sign-anchored site scores and species loadings. Loading orientation and sign are not identifiable; the model-implied Sigma and correlations remain the rotation-invariant interpretation target.", fig.alt = "Two-dimensional ordination biplot showing site scores as points and species loading arrows after accounting for the environmental predictor. Species arrows in similar directions indicate similar residual co-occurrence patterns, but loading sign and rotation are arbitrary."}
plot(
  fit_jsdm,
  type = "ordination",
  level = "unit",
  rotation = "varimax",
  sign_anchor = "auto",
  standardize_loadings = TRUE
) +
  ggplot2::labs(
    title = "Residual species ordination",
    subtitle = "Varimax-rotated site scores and species loading directions.",
    caption = NULL
  )
```

The fixture's truth places the eight species into **four designed
association groups**: `sp_1` / `sp_2`, `sp_3` / `sp_4`, `sp_5` /
`sp_6`, and `sp_7` / `sp_8`. The fitted arrows should be read for
broad grouping, not axis labels. The exact orientation is arbitrary —
only $\boldsymbol\Lambda\boldsymbol\Lambda^\top$ is identifiable.
Species arrows pointing in the same direction co-occur after `env_1`
is accounted for; arrows pointing in opposite directions tend to
mutually exclude.

## See also

* [`vignette("functional-biogeography")`](functional-biogeography.html) — Gaussian stacked-trait analogue.
* [`vignette("response-families")`](response-families.html) — full table of supported families and the per-family $\sigma^2_d$.

## References

* Niku, J., Warton, D.I., Hui, F.K.C., & Taskinen, S. (2017).
  Generalized linear latent variable models for multivariate count
  and biomass data in ecology. *Journal of Agricultural, Biological,
  and Environmental Statistics* **22**, 498–522.
  <https://link.springer.com/article/10.1007/s13253-017-0304-7>
* Niku, J., Hui, F.K.C., Taskinen, S., & Warton, D.I. (2019).
  gllvm: Fast analysis of multivariate abundance data with
  generalized linear latent variable models in r. *Methods in
  Ecology and Evolution* **10**, 2173–2182.
  <https://doi.org/10.1111/2041-210X.13303>
* Nakagawa, S., & Schielzeth, H. (2010). Repeatability for Gaussian
  and non-Gaussian data: a practical guide for biologists.
  *Biological Reviews* **85**, 935–956.
  <https://doi.org/10.1111/j.1469-185X.2010.00141.x>
* Nakagawa, S., Johnson, P.C.D., & Schielzeth, H. (2017). The
  coefficient of determination R² and intra-class correlation
  coefficient from generalized linear mixed-effects models revisited
  and expanded. *Journal of the Royal Society Interface* **14**,
  20170213. <https://doi.org/10.1098/rsif.2017.0213>
* Warton, D.I., Blanchet, F.G., O'Hara, R.B., Ovaskainen, O.,
  Taskinen, S., Walker, S.C., & Hui, F.K.C. (2015). So many
  variables: joint modeling in community ecology. *Trends in
  Ecology & Evolution* **30**, 766–779.
  <https://doi.org/10.1016/j.tree.2015.09.007>
