---
title: "Joint species distribution model (binary GLLVM)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Joint species distribution model (binary GLLVM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

A reduced-rank binary GLLVM compresses a many-column 0/1 occurrence
matrix into a few latent factors. The classic ecology version is a
**joint species distribution model**: rows are sites, columns are
species, the entries are presence / absence, and the latent factors
capture residual co-occurrence beyond what's explained by
environmental predictors. The same `gllvmTMB(family = binomial())`
call also fits any binary `unit ├Ч trait` data тАФ IRT-style item
responses, presence/absence of attributes, etc.

## The model

For binary $y_{it} \in \{0, 1\}$ with logit link,

$$
\mathrm{logit}\bigl(P(y_{it} = 1)\bigr) \;=\; \mathbf x_i^\top
\boldsymbol\beta_t \;+\; \boldsymbol\lambda_t^{\!\top} \mathbf u_i,
\qquad \mathbf u_i \sim \mathcal N(\mathbf 0, \mathbf I_d).
$$

Two distinct covariance objects matter here, and **conflating them is
the most common notation trap in JSDM writing**.

### 1. Between-unit shared covariance (estimated by the GLLVM)

The covariance among traits / species induced by the latent factors
$\mathbf u_i$ is

$$
\boldsymbol\Sigma_B \;=\; \boldsymbol\Lambda \boldsymbol\Lambda^\top.
$$

That's the part the GLLVM **estimates** from cross-trait
co-occurrence on the latent scale. With reduced rank $d \ll T$ it is
identified by the off-diagonal (cross-trait) co-occurrence pattern;
the diagonal is then constrained by the rank.

### 2. Total latent-liability covariance (used for correlations / ICC)

To convert $\boldsymbol\Sigma_B$ into something on a unified
**latent-liability** scale (so we can compute latent-scale
correlations, ICC, communality, $R^2$, тАж), we use the threshold
representation:

$$
y^*_{it} \;=\; \eta_{it} + \varepsilon_{it},
\qquad \varepsilon_{it} \sim \text{Logistic}(0, 1),
\qquad \mathrm{Var}(\varepsilon) = \pi^2/3.
$$

with $y_{it} = \mathbf 1\{y^*_{it} > 0\}$. The total latent-liability
covariance is then

$$
\boldsymbol\Sigma_{\text{latent}}
\;=\;
\underbrace{\boldsymbol\Sigma_B}_{\text{estimated}}
\;+\;
\underbrace{(\pi^2 / 3)\, \mathbf I}_{\text{fixed by link}}.
$$

The $(\pi^2/3) \mathbf I$ term is **not estimated** тАФ it is the
distribution-specific variance of the standard logistic
($\pi^2/3 \approx 3.29$). For probit it's $1$; for cloglog it's
$\pi^2/6$. It enters whenever you want to put $\boldsymbol\Sigma_B$
on the conventional latent-liability scale (Niku et al. 2017;
Nakagawa & Schielzeth 2010).

In `gllvmTMB`:

* `extract_Sigma(fit, level = "unit", part = "shared")` returns
  $\boldsymbol\Sigma_B = \boldsymbol\Lambda\boldsymbol\Lambda^\top$.
* `extract_Sigma(fit, level = "unit", part = "total")` returns
  $\boldsymbol\Sigma_{\text{latent}} = \boldsymbol\Sigma_B + (\pi^2/3) \mathbf I$.

`extract_correlations()` uses `part = "total"` by default, so the
correlations come out on the latent-liability scale.

### Why we drop `unique()` for pure binary

For Gaussian fits, the canonical decomposition pairs
`latent + unique`:

$$\boldsymbol\Sigma_B = \boldsymbol\Lambda\boldsymbol\Lambda^\top + \mathbf S$$

where $\mathbf S = \mathrm{diag}(S_{tt})$ is the per-trait unique
variance тАФ *estimated* from per-row residual variability. For
**pure binary 1-obs-per-cell** fits, $S_{tt}$ carries no information:
binary data do not constrain observation-level variability beyond
the link residual. Some Bayesian implementations (e.g. Hadfield's
MCMCglmm) add a small `╧Г┬▓_units` term for MCMC mixing reasons, but
it isn't estimated from data; `gllvmTMB` is frequentist and simply
omits it.

What happens if you add `unique()` anyway? It runs, but is mostly
wasted effort:

| Question | Answer |
|---|---|
| Engine accepts `unique()` in a pure binary fit? | тЬЕ Yes тАФ runs without error |
| Estimates per-trait `sd_B[t]`? | тЬЕ Yes тАФ but hits the boundary at ~ 0 (unidentifiable from binary 1-obs-per-cell) |
| Issues a warning? | тЭМ Not yet тАФ Phase D follow-up will add a one-shot info message and `map`-suppress those parameters |
| Affects $\boldsymbol\Sigma_B = \boldsymbol\Lambda\boldsymbol\Lambda^\top$? | тЭМ No тАФ $\boldsymbol\Lambda$ is unchanged |
| Affects $\boldsymbol\Sigma_{\text{latent}}$? | Marginal тАФ diagonal becomes $\boldsymbol\Lambda\boldsymbol\Lambda^\top + \mathrm{diag}(\hat{sd}_B^2) + (\pi^2/3) \mathbf I$, but $\hat{sd}_B^2 \approx 0$ at the boundary so it's essentially the same as without `unique()` |
| Affects fit time / Hessian | Slightly: extra $T$ parameters pinned at the boundary, Hessian can be ill-conditioned. `sanity_multi(fit)` may flag it |

Bottom line: drop `unique()` for **pure** binary fits. Keep it for
**mixed-family** fits where it's freely estimated for the non-binary
traits.

### What about `dep` and `indep` for binary?

The other two covstruct modes тАФ `dep` (full unstructured ╬г) and
`indep` (diagonal-only ╬г) тАФ also have identifiability issues with
binary 1-obs-per-cell. Empirical comparison on simulated data
(`T = 6` species, `n = 150` sites, rank-2 truth):

| Mode | logL | ╬г_B diag (estimated) | Off-diag agreement vs truth |
|---|---|---|---|
| **`latent(d = 2)`** (recommended) | **тИТ593.65** | 1.07тАУ3.31 | $r = 0.80$ тЬУ |
| `latent(d = T)` (= `dep`) | тИТ593.45 | 1.11тАУ3.41 | $r = 0.79$ тЬУ |
| `dep` (full unstructured) | тИТ593.45 | 1.11тАУ3.41 | $r = 0.79$ тЬУ |
| **`indep`** (diagonal-only) | **тИТ616.85** ЁЯФ┤ | **0, 0, 0, 0, 0, 0** ЁЯФ┤ | NA (no off-diagonals) |

(True diagonal: 1.49тАУ1.85; true off-diagonal range: [тИТ1.77, +1.66].)

What this confirms:

- **`dep` is mathematically equivalent to `latent(d = T)`** тАФ identical
  logL, identical ╬г.
- **Reduced-rank `latent(d < T)` matches the saturated logL within
  0.2 nats** тАФ the *off-diagonals* (the identifiable part) are
  captured equally well by `latent(d = 2)` as by `dep`. The rank
  constraint additionally regularises the *diagonals*, which would
  otherwise be unidentifiable individually for binary, giving
  estimates ~10тАУ20% closer to truth.
- **`indep` is catastrophic for binary**: logL is ~23 nats worse,
  every per-trait variance pinned at exactly zero, no off-diagonals
  modelled. Avoid `indep` for any binary fit where cross-species
  / cross-trait correlations matter тАФ which is essentially every
  realistic JSDM.

Recommended for binary: **`latent(d = K)` with `K < T`**. Skip `dep`
(no advantage), skip `indep` (broken).

### Mixed-family fits

When the response includes a *mix* of families (some traits binary,
some Gaussian, some Poisson, тАж), the picture changes:

- For **non-binary traits**, the unique component $S_{tt}$ IS
  identifiable. Including `unique(0 + trait | unit)` adds a
  per-trait random intercept that's freely estimated.
- For **binary traits in a mixed fit**, $S_{tt}$ is still
  unidentifiable; the optimiser pins it near zero (boundary).

So `unique()` is fine in mixed-family fits тАФ you just need to know
the per-trait `sd_B` will report effectively-zero values for binary
rows. The Phase D follow-up will add per-family-aware
`map`-suppression so the optimiser doesn't waste iterations on
those boundary parameters; current behaviour is correct, just
slightly inefficient. The `extract_Sigma()` machinery handles each
trait's link residual separately by family тАФ table below.

### How `extract_Sigma()` handles non-Gaussian and mixed-family fits

`extract_Sigma(fit, part = "total")` looks up each trait's family /
link, and adds the **per-trait latent-scale residual** $\sigma^2_d$ to
the diagonal of $\boldsymbol\Sigma$. The internal helper is
`gllvmTMB:::link_residual_per_trait(fit)`, which uses these formulas
from Nakagawa & Schielzeth (2010) + Nakagawa, Johnson & Schielzeth
(2017):

| Family / link | $\sigma^2_d$ added per trait |
|---|---|
| `gaussian` (identity) | $0$ |
| `binomial(logit)` | $\pi^2/3 \approx 3.290$ |
| `binomial(probit)` | $1$ |
| `binomial(cloglog)` | $\pi^2/6 \approx 1.645$ |
| `poisson(log)` | $\log(1 + 1/\hat\mu_t)$  (lognormal-Poisson) |
| `nbinom2(log)` | $\psi'(\hat\phi_t)$ (trigamma; Stoklosa et al. 2022) |
| `tweedie(log)` | $\log(1 + \hat\phi_t \hat\mu_t^{p-2})$ (Tweedie 1984) |
| `Beta(logit)` | $\psi'(\mu\phi) + \psi'((1-\mu)\phi)$ (Smithson & Verkuilen 2006) |
| `betabinomial(logit)` | $\psi'(\mu\phi) + \psi'((1-\mu)\phi) + \pi^2/3$ |
| `delta_lognormal` | $\hat\sigma^2_{\text{logn}} + \pi^2/3$ |

For mixed-family fits, each trait gets the residual implied by **its
own** family; the diagonal of $\boldsymbol\Sigma$ is therefore on a
unified latent scale. Pass `link_residual = "none"` to suppress this
addition and recover the bare $\boldsymbol\Lambda\boldsymbol\Lambda^\top
+ \mathbf S$ form.

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

## Simulate and fit

A simulated `Site ├Ч Species` occurrence dataset тАФ 100 sites, 8
species, 2 underlying latent factors, plus an environmental
covariate `env_1` with species-specific responses. The true loadings
are designed so each species falls into one of **four niche
quadrants** of the latent space (e.g. dry-low, dry-high, wet-low,
wet-high), with two species per quadrant. The biplot below should
recover those four clusters.

```{r jsdm-sim}
set.seed(2025)
T       <- 8           # 8 species
n_sites <- 100         # 100 sites

## Designed loadings: 8 species in 4 niche quadrants
##   sp1, sp2  : (- F1, - F2)  тАФ e.g. "dry, lowland"
##   sp3, sp4  : (- F1, + F2)  тАФ e.g. "dry, highland"
##   sp5, sp6  : (+ F1, - F2)  тАФ e.g. "wet, lowland"
##   sp7, sp8  : (+ F1, + F2)  тАФ e.g. "wet, highland"
Lam_true <- matrix(c(
  ## Column 1: F1 (e.g. moisture; positive = wet)
  -1.2, -1.0, -1.1, -0.9,  +1.0, +1.2, +1.1, +0.9,
  ## Column 2: F2 (e.g. elevation; positive = highland)
  -1.0, -1.2, +1.1, +0.9,  -1.0, -1.1, +1.0, +1.2
), nrow = T, ncol = 2)

sim <- simulate_site_trait(
  n_sites = n_sites, n_species = 6, n_traits = T,
  mean_species_per_site = 5,
  Lambda_B = Lam_true, S_B = rep(0.4, T),
  sigma2_eps = 0.01, seed = 2025
)
df <- sim$data
df$value <- as.integer(df$value > 0)   # binarise
```

```{r jsdm-fit}
fit_jsdm <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):env_1 +
          latent(0 + trait | site, d = 2),
  data   = df,
  family = binomial()
)
fit_jsdm
```

## Latent-scale residual species correlations (with 95% CIs)

Pairwise species correlations on the latent-liability scale. The
default `method = "wald"` gives approximate Wald CIs that use a
Fisher transform internally, so they are fast (seconds) and bounded
inside $[-1, 1]$ by construction.
Profile-likelihood (`method = "profile"`) is more accurate for
skewed sampling distributions but scales as $O(T(T-1)/2)$
constrained refits тАФ use it for final, publication-grade CIs once
Phase K's warm-started `TMB::tmbprofile()` accelerates the path.
`method = "bootstrap"` is also available for full sampling-
distribution CIs.

```{r jsdm-cor}
extract_correlations(fit_jsdm, tier = "unit")
```

The two $\boldsymbol\Sigma$ matrices side-by-side: the **shared**
component $\boldsymbol\Sigma_B = \boldsymbol\Lambda\boldsymbol\Lambda^\top$
that the GLLVM estimates, and the **total latent-liability**
covariance $\boldsymbol\Sigma_{\text{latent}} = \boldsymbol\Sigma_B + (\pi^2/3) \mathbf I$
that adds the fixed link residual on the diagonal:

```{r jsdm-sigma}
Sigma_shared <- extract_Sigma(fit_jsdm, level = "unit", part = "shared")$Sigma
Sigma_total  <- extract_Sigma(fit_jsdm, level = "unit", part = "total")$Sigma
list(Sigma_shared = round(Sigma_shared, 2),
     Sigma_total  = round(Sigma_total,  2))
```

Notice the off-diagonals are identical (the link residual is
diagonal) and the total has $+ \pi^2/3 \approx 3.29$ added to each
diagonal entry.

## Ordination biplot

Site scores on the two latent axes, with species loadings shown as
arrows. Axes are residual co-occurrence after `env_1` has been
accounted for; species loading in the same direction tend to
co-occur (after accounting for the environment).

```{r jsdm-biplot, fig.width = 6.5, fig.height = 5}
## Site scores
ord <- extract_ordination(fit_jsdm, level = "unit")
scores <- data.frame(
  LV1 = ord$scores[, 1],
  LV2 = ord$scores[, 2]
)

## Species loadings (Lambda_B)
Lambda_hat <- fit_jsdm$report$Lambda_B
loadings <- data.frame(
  trait = paste0("sp", seq_len(nrow(Lambda_hat))),
  LV1   = Lambda_hat[, 1],
  LV2   = Lambda_hat[, 2]
)

## Scale arrows so they're comparable to score cloud
sc <- 0.7 * max(abs(scores)) / max(abs(loadings[, c("LV1", "LV2")]))
loadings$LV1 <- loadings$LV1 * sc
loadings$LV2 <- loadings$LV2 * sc

ggplot() +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.4) +
  geom_vline(xintercept = 0, colour = "grey80", linewidth = 0.4) +
  geom_point(data = scores, aes(LV1, LV2),
             colour = "steelblue", alpha = 0.45, size = 1.8) +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = LV1, yend = LV2),
               arrow = arrow(length = unit(0.22, "cm")),
               colour = "tomato", linewidth = 0.5) +
  geom_text(data = loadings,
            aes(LV1, LV2, label = trait),
            colour = "tomato", size = 3.6,
            nudge_x = 0.15, nudge_y = 0.05) +
  coord_equal() +
  labs(x = "Latent variable 1", y = "Latent variable 2") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

The species arrows recover the **four design quadrants**: `sp1, sp2`
in the lower-left, `sp3, sp4` upper-left, `sp5, sp6` lower-right,
`sp7, sp8` upper-right (up to a sign flip / rotation of the latent
axes тАФ only $\boldsymbol\Lambda\boldsymbol\Lambda^\top$ is
identifiable). Species arrows pointing in the same direction
co-occur after `env_1` is accounted for; arrows pointing in opposite
directions tend to mutually exclude.

## See also

* [`vignette("functional-biogeography")`](functional-biogeography.html) тАФ Gaussian stacked-trait analogue.
* [`vignette("cross-package-validation")`](cross-package-validation.html) тАФ `gllvm` log-likelihood comparison on a binary fit.
* [`vignette("response-families")`](response-families.html) тАФ full table of supported families and the per-family $\sigma^2_d$.

## References

* Niku, J., Warton, D.I., Hui, F.K.C., & Taskinen, S. (2017).
  Generalized linear latent variable models for multivariate count
  and biomass data in ecology. *Journal of Agricultural, Biological,
  and Environmental Statistics* **22**, 498тАУ522.
  <https://link.springer.com/article/10.1007/s13253-017-0304-7>
* Niku, J., Hui, F.K.C., Taskinen, S., & Warton, D.I. (2019).
  gllvm: Fast analysis of multivariate abundance data with
  generalized linear latent variable models in r. *Methods in
  Ecology and Evolution* **10**, 2173тАУ2182.
  <https://doi.org/10.1111/2041-210X.13303>
* Nakagawa, S., & Schielzeth, H. (2010). Repeatability for Gaussian
  and non-Gaussian data: a practical guide for biologists.
  *Biological Reviews* **85**, 935тАУ956.
  <https://doi.org/10.1111/j.1469-185X.2010.00141.x>
* Nakagawa, S., Johnson, P.C.D., & Schielzeth, H. (2017). The
  coefficient of determination R┬▓ and intra-class correlation
  coefficient from generalized linear mixed-effects models revisited
  and expanded. *Journal of the Royal Society Interface* **14**,
  20170213. <https://doi.org/10.1098/rsif.2017.0213>
* Warton, D.I., Blanchet, F.G., O'Hara, R.B., Ovaskainen, O.,
  Taskinen, S., Walker, S.C., & Hui, F.K.C. (2015). So many
  variables: joint modeling in community ecology. *Trends in
  Ecology & Evolution* **30**, 766тАУ779.
  <https://doi.org/10.1016/j.tree.2015.09.007>
