---
title: "Joint species distribution models for binary occurrence data"
description: "Fit a binary latent-factor JSDM in long or wide form and interpret residual species co-occurrence on a declared liability scale."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Joint species distribution models for binary occurrence data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

After an environmental gradient is in the model, which species still
tend to occur together, and which species avoid each other? That is
the first joint species distribution model (JSDM) question this page
answers.

The data are binary occurrence records: one row per site, one column
or trait per species, and a 0/1 response for absence/presence. The
model uses `latent()` to estimate a low-rank residual co-occurrence
structure across species after `env_1` is accounted for.

This example supports both long data and the wide `traits(...)` formula,
using a binomial-logit model with `latent(0 + trait | site, d = K)` for
residual co-occurrence. The Fisher-z intervals below are model-based
uncertainty diagnostics; their repeated-sampling coverage has not been
established for this model class. Single-species spatial SDMs belong in
`sdmTMB`, and multivariate spatial fields via `spatial_*()` are beyond the
scope of this introductory JSDM example.

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

## Load and fit

The article uses a shipped teaching fixture rather than building the
data inside the vignette. It contains a complete `Site x Species`
occurrence matrix: 120 sites, 8 species, 2 latent residual
co-occurrence axes, and an environmental covariate `env_1` with
species-specific responses. The true loadings put the species into
four niche quadrants (dry-low, dry-high, wet-low, wet-high), with two
species per quadrant.

```{r jsdm-data}
jsdm_path <- system.file(
  "extdata", "examples", "joint-sdm-example.rds",
  package = "gllvmTMB"
)

source_candidates <- file.path(
  c(".", "..", "../.."),
  "inst", "extdata", "examples", "joint-sdm-example.rds"
)
input_file <- knitr::current_input(dir = TRUE)
if (!is.null(input_file)) {
  source_candidates <- c(
    source_candidates,
    file.path(
      dirname(dirname(dirname(input_file))),
      "inst", "extdata", "examples", "joint-sdm-example.rds"
    )
  )
}
candidate_paths <- c(jsdm_path, source_candidates)
candidate_paths <- candidate_paths[nzchar(candidate_paths)]
jsdm_path <- candidate_paths[file.exists(candidate_paths)][1]
if (is.na(jsdm_path)) {
  stop("Could not find joint-sdm-example.rds.")
}
jsdm <- readRDS(jsdm_path)

df <- jsdm$data_long
df_wide <- jsdm$data_wide
```

The same JSDM fits in both long format and the `traits(...)` wide
format. The long form is shown first; the wide form is the
formula-API equivalent on a pivoted data frame. Both reach the
same engine.

```{r jsdm-fit}
# Long format -- one row per (site, species) observation
fit_jsdm <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):env_1 +
    latent(0 + trait | site, d = 2),
  data   = df,
  trait  = "trait",
  unit   = "site",
  family = binomial()
)
fit_jsdm
```

```{r jsdm-fit-wide}
# Wide format -- one row per site, one column per species. The
# `traits(...)` LHS names the species columns; the compact RHS
# shorthand uses `1 + env_1`, which expands internally to
# `0 + trait + (0 + trait):env_1`.
fit_jsdm_wide <- gllvmTMB(
  traits(sp_1, sp_2, sp_3, sp_4, sp_5, sp_6, sp_7, sp_8) ~
    1 + env_1 + latent(1 | site, d = 2),
  data   = df_wide,
  unit   = "site",
  family = binomial()
)
all.equal(logLik(fit_jsdm), logLik(fit_jsdm_wide), tolerance = 1e-8)
```

### Long or wide data shape

JSDM is the package's natural long-format use case because each row
already carries `(site, species)` and the response factor `trait`
indexes species. When the data already live as one row per site and
one binary column per species, the `traits(...)` wide form is the same
model through the same `gllvmTMB()` entry point. In either shape, true
absences should be explicit zeros. Missing response cells are not a
safe shortcut for absences, because missingness can also mean the site
or species was not surveyed.

## What `latent()` estimates

For binary occurrence data with a logit link, the fitted linear
predictor is

$$
\mathrm{logit}\{P(y_{it} = 1)\} =
\mathbf x_i^\top \boldsymbol\beta_t +
\boldsymbol\lambda_t^\top \mathbf u_i,
\qquad
\mathbf u_i \sim \mathcal N(\mathbf 0, \mathbf I_d).
$$

The site score $\mathbf u_i$ is the unobserved residual state of site
$i$, and the species loading $\boldsymbol\lambda_t$ says how species
$t$ responds to those residual axes. The rotation of
$\boldsymbol\Lambda$ is arbitrary, so the stable co-occurrence target
is the species covariance implied by the loadings. For this single-trial
Bernoulli fixture, each species-site cell contributes one binary observation.
The default between-site diagonal companion is therefore not separately
identified from the link scale and the engine maps it off:

$$
\boldsymbol\Sigma_B =
\boldsymbol\Lambda\boldsymbol\Lambda^\top.
$$

For a binary logit JSDM, correlations are easier to interpret on the
latent-liability scale. `extract_correlations(link_residual = "auto")`
adds the fixed logistic residual variance $\pi^2 / 3$ to the diagonal:

$$
\boldsymbol\Sigma_{\mathrm{latent}} =
\boldsymbol\Lambda\boldsymbol\Lambda^\top +
(\pi^2 / 3)\mathbf I.
$$

The logistic residual term is fixed by the binomial logit link, not estimated
from the data. Ordinary `latent()` normally carries a diagonal
$\boldsymbol\Psi_B$ companion, but the fit message explains why that default
component is mapped off for these single-trial cells. Multi-trial binomial data
or genuine repeated observations can identify additional variation; they are
different designs and should not be inferred from this occurrence matrix.

### Where loading constraints fit

The loading matrix $\boldsymbol\Lambda$ is usually exploratory in a
JSDM: its rotation is arbitrary, while
$\boldsymbol\Lambda\boldsymbol\Lambda^\top$ is the interpretable
co-occurrence target. If you have an a-priori hypothesis that a
species group defines an axis, or that a species should have zero
loading on an axis, that is a **confirmatory loading** question rather
than the default species-distribution workflow.

`gllvmTMB` supports that confirmatory path with
`lambda_constraint = list(unit = M)`, where `NA` means "estimate this
loading" and a number means "pin this loading during fitting." Numeric
pins are exact model restrictions: because the latent scores have unit
variance, a loading pinned to `1` is assumed to equal one; it is not a
cosmetic sign convention. The current route acts on the packed
lower-triangular loading matrix, so it is not a general free-form
confirmatory-factor interface.

Specify a biological loading hypothesis before inspecting this fitted
loading matrix, verify that the matrix rows follow the response order,
and stop if the constrained fit is unhealthy. A matrix suggested from
the same data can provide a reproducible descriptive orientation, but
it remains exploratory and does not turn selected zeros into biological
facts. This guide therefore keeps correlations and
$\boldsymbol\Sigma$ as its interpretation targets.

## Latent-scale residual species correlations

Pairwise species correlations on the latent-liability scale. The
opt-in `method = "fisher-z"` gives Fisher-transform heuristic bounds that
are fast (seconds) and bounded inside $[-1, 1]$ by construction.
`method = "wald"` is an accepted alias for the same Fisher-z path.
Two heavier alternatives are also available: `method = "bootstrap"`
re-fits the model under a parametric bootstrap for a full
sampling-distribution CI, and `method = "profile"` inverts the
profile likelihood pair-by-pair (cost scales as $O(T(T-1)/2)$
constrained refits).

One scale caveat is worth stating up front, because it trips people
up on binary fits. The default `link_residual = "auto"` puts the
correlation on the **latent-liability** scale — it adds the fixed
logistic residual $\pi^2/3$ to the diagonal of
$\boldsymbol\Sigma_B$ before forming the correlation, which is the
interpretable co-occurrence quantity for a JSDM. `fisher-z`,
`wald`, and `bootstrap` all honour that argument. The current
`profile` path, however, always reports the correlation of the
**bare** $\boldsymbol\Sigma_B = \boldsymbol\Lambda\boldsymbol\Lambda^\top$
(it ignores `link_residual`), so its point estimate is the
undiluted latent-factor correlation, not the latent-liability one.
The optional comparison snippets below make the difference explicit
for readers who want to audit the scale.

What is and is not certified: the point estimates and Fisher-z
bounds are useful diagnostics for this fitted example, but the
package-wide calibrated-coverage work is still open. Treat intervals
here as model-based uncertainty summaries, not as a claim that every
binary or mixed-family covariance interval has reached nominal
coverage.

```{r jsdm-cor}
corr_rows <- extract_correlations(
  fit_jsdm,
  tier = "unit",
  method = "fisher-z",
  link_residual = "auto"
)
corr_rows
```

The matrix view below shows the same point estimates as a
species-by-species map. For a binary JSDM, the useful first picture is
the **latent-liability correlation** pattern after adding the fixed
logistic link residual to the denominator; raw
$\boldsymbol\Sigma_B$ entries can be dominated by loading scale and are
harder to interpret. The Fisher-z interval bounds stay in `corr_rows`
rather than being printed inside every matrix cell.

```{r jsdm-correlation-heatmap, fig.width = 6.8, fig.height = 5.8, out.extra='class="wide-scientific-figure"', fig.cap = "Residual species correlations from the fitted binary JSDM. Cells show Fisher-z point estimates on the latent-liability scale after accounting for env_1 and the fixed logistic link residual; interval bounds are available in the preceding table.", fig.alt = "Heatmap of pairwise residual species correlations for eight species. Blue cells indicate positive residual co-occurrence, red cells indicate negative residual co-occurrence, and diagonal self-correlations are one."}
plot_correlations(
  corr_rows,
  style = "heatmap",
  matrix_layout = "by_level",
  label_type = "estimate",
  label_digits = 2,
  include_diagonal = TRUE,
  title = "Residual species correlations",
  subtitle = "Latent-liability correlations after accounting for env_1."
) +
  ggplot2::labs(caption = NULL)
```

Use these correlations, rather than the signs of individual loadings,
for the rotation-invariant co-occurrence story.

## Confidence intervals on the binary latent scale

This section works through the three CI methods on two estimands —
a pairwise species **correlation** and a per-species **ICC** — to
show what each method returns and where it currently stops short.
These chunks are optional because the bootstrap and profile paths
re-fit the model; run them when the interval method itself is the
question, not for a first pass through the article.

### Three CI methods for one correlation

Take a single species pair (`sp_1`, `sp_2`) and ask for the same
correlation three ways. Keeping `link_residual = "auto"` for the
Fisher-z and bootstrap calls holds them on the latent-liability
scale, so their point estimates agree and only the interval differs:

```{r jsdm-cor-methods, eval = FALSE}
pair_ij <- c("sp_1", "sp_2")

cor_fisher <- extract_correlations(
  fit_jsdm, tier = "unit", pair = pair_ij,
  method = "fisher-z", link_residual = "auto"
)
cor_boot <- extract_correlations(
  fit_jsdm, tier = "unit", pair = pair_ij,
  method = "bootstrap", nsim = 20, seed = 1, link_residual = "auto"
)
rbind(cor_fisher, cor_boot)[, c("trait_i", "trait_j",
                                "correlation", "lower", "upper", "method")]
```

The Fisher-z and bootstrap point estimates match because both add
the $\pi^2/3$ logistic residual to the diagonal; the bootstrap
interval is the empirical 2.5/97.5 percentile across the refits.

The `profile` method is the one that reports a *different* number,
because it currently works on the bare $\boldsymbol\Sigma_B$ scale.
To compare it fairly, request the Fisher-z interval with
`link_residual = "none"` so both sit on the same (undiluted) scale:

```{r jsdm-cor-profile, eval = FALSE}
cor_profile <- extract_correlations(
  fit_jsdm, tier = "unit", pair = pair_ij,
  method = "profile"
)
cor_fisher_bare <- extract_correlations(
  fit_jsdm, tier = "unit", pair = pair_ij,
  method = "fisher-z", link_residual = "none"
)
rbind(cor_profile, cor_fisher_bare)[, c("trait_i", "trait_j",
                                        "correlation", "lower", "upper", "method")]
```

Now the two point estimates agree: the profile correlation matches
the bare-scale Fisher-z value. The profile *interval* is a
likelihood-ratio inversion around that point — but note it can come
back as `NA` when the constrained refit fails to bracket the bound
(a near-boundary correlation here is a common cause), which is the
same not-yet-calibrated CI story flagged above. The take-home is
*not* that one method is wrong, but that **`profile` and the default
latent-liability correlations live on different scales** until the
profile path learns to honour `link_residual`. For the interpretable
JSDM co-occurrence story, prefer the latent-liability scale
(`auto`), and use Fisher-z for speed or the bootstrap when you want
an empirical sampling-distribution interval.

### ICC (communality) on the latent scale

For a single-trial binary JSDM, the natural communality quantity is the
share of each species' latent-liability variance represented by the shared
factors rather than the fixed logistic residual:

$$
\mathrm{ICC}_t \;=\;
\frac{(\boldsymbol\Lambda\boldsymbol\Lambda^\top)_{tt}}
     {(\boldsymbol\Lambda\boldsymbol\Lambda^\top)_{tt} + \pi^2/3}.
$$

The unit-level `latent()` term estimates the shared loading covariance, and
`link_residual = "auto"` adds the fixed logistic residual variance $\pi^2/3$,
so `extract_communality(level = "unit",
link_residual = "auto")` returns exactly this ratio — the
proportion of the latent-liability variance that the shared factors
explain. (This is *not* the between-vs-within `extract_repeatability()`
ICC, which needs paired between- and within-unit diagonal tiers such as
`indep(0 + trait | site) + indep(0 + trait | site_species)`; a single-trial occurrence matrix has
no within-site replication for that, so `extract_repeatability()`
and `extract_ICC_site()` are not applicable to this fit.)

```{r jsdm-icc-point}
icc_point <- extract_communality(
  fit_jsdm, level = "unit", link_residual = "auto"
)
round(icc_point, 3)
```

A parametric bootstrap puts an interval on each species' ICC:

```{r jsdm-icc-boot, eval = FALSE}
icc_boot <- extract_communality(
  fit_jsdm, level = "unit", link_residual = "auto",
  ci = TRUE, method = "bootstrap", nsim = 20, seed = 1
)
icc_boot[, c("trait", "c2", "lower", "upper", "method")]
```

The intervals are wide — eight species and 120 sites is modest
information for a per-species variance share, and the upper bounds
for the more strongly-loading species run close to 1.

Two limitations apply to the ICC interval in this example:

* **Wald is not implemented** for communality (it is a non-linear
  function of several parameters); requesting `method = "wald"`
  emits a note and falls back to the bootstrap.
* **Profile bounds are not yet delivered** for this proportion on a
  single-trial binary fit — the constrained refit fails to bracket
  and the interval returns `NA`. The point estimate is still the
  correct latent-scale ICC; only the profile *interval* is missing.

The point estimates are available, but interval coverage on binary and
mixed-family scales has not been established. Report the point estimate and
the interval method used; do not present these intervals as
coverage-certified.

## Ordination biplot

Site scores on the two latent variables (LV1, LV2), with species
loadings shown as arrows. This is *model-based* ordination: LV1 and
LV2 are the latent variables of a fitted GLLVM, estimated jointly with
the `env_1` regression — not a variance-maximising PCA, a
distance-based NMDS embedding, or a covariate-constrained CCA/RDA of
the raw occurrence matrix. The plotted site scores are predicted
(empirical-Bayes) conditional modes of the residual site states
$u_i$: shrunk toward zero, and carrying prediction uncertainty the
point display does not show. The rotation-invariant interpretation
target is the model-implied residual co-occurrence
$\boldsymbol\Sigma_B$ (defined above), not the axis coordinates
themselves — species loading in the same direction tend to co-occur
after `env_1` is accounted for.

```{r jsdm-biplot, fig.width = 6.5, fig.height = 5, fig.cap = "Ordination biplot from the fitted binary JSDM, using varimax-rotated and sign-anchored site scores and species loadings. Loading orientation and sign are not identifiable; the model-implied Sigma and correlations remain the rotation-invariant interpretation target.", fig.alt = "Two-dimensional ordination biplot showing site scores as points and species loading arrows after accounting for the environmental predictor. Species arrows in similar directions indicate similar residual co-occurrence patterns, but loading sign and rotation are arbitrary."}
plot(
  fit_jsdm,
  type = "ordination",
  level = "unit",
  rotation = "varimax",
  sign_anchor = "auto",
  standardize_loadings = TRUE
) +
  ggplot2::labs(
    title = "Residual species ordination",
    subtitle = "Varimax-rotated site scores and species loading directions.",
    caption = NULL
  )
```

The fixture's truth places the eight species into **four designed
association groups**: `sp_1` / `sp_2`, `sp_3` / `sp_4`, `sp_5` /
`sp_6`, and `sp_7` / `sp_8`. The fitted arrows should be read for
broad grouping, not axis labels. The exact orientation is arbitrary —
only $\boldsymbol\Lambda\boldsymbol\Lambda^\top$ is identifiable.
Species arrows pointing in the same direction co-occur after `env_1`
is accounted for; arrows pointing in opposite directions tend to
mutually exclude.

## See also

* [Morphometrics](morphometrics.html) — Gaussian stacked-trait analogue with ordinary `latent()`.
* [Choosing latent rank](model-selection-latent-rank.html) — how to compare candidate values of `d`.
* Loading-constraint helpers — `?confirmatory_lambda` and
  `?suggest_lambda_constraint` document the advanced API contracts.
* [`vignette("response-families")`](response-families.html) — full table of supported families and the per-family $\sigma^2_d$.

## References

* Niku, J., Warton, D.I., Hui, F.K.C., & Taskinen, S. (2017).
  Generalized linear latent variable models for multivariate count
  and biomass data in ecology. *Journal of Agricultural, Biological,
  and Environmental Statistics* **22**, 498–522.
  <https://link.springer.com/article/10.1007/s13253-017-0304-7>
* Niku, J., Hui, F.K.C., Taskinen, S., & Warton, D.I. (2019).
  gllvm: Fast analysis of multivariate abundance data with
  generalized linear latent variable models in r. *Methods in
  Ecology and Evolution* **10**, 2173–2182.
  <https://doi.org/10.1111/2041-210X.13303>
* Nakagawa, S., & Schielzeth, H. (2010). Repeatability for Gaussian
  and non-Gaussian data: a practical guide for biologists.
  *Biological Reviews* **85**, 935–956.
  <https://doi.org/10.1111/j.1469-185X.2010.00141.x>
* Nakagawa, S., Johnson, P.C.D., & Schielzeth, H. (2017). The
  coefficient of determination R² and intra-class correlation
  coefficient from generalized linear mixed-effects models revisited
  and expanded. *Journal of the Royal Society Interface* **14**,
  20170213. <https://doi.org/10.1098/rsif.2017.0213>
* Warton, D.I., Blanchet, F.G., O'Hara, R.B., Ovaskainen, O.,
  Taskinen, S., Walker, S.C., & Hui, F.K.C. (2015). So many
  variables: joint modeling in community ecology. *Trends in
  Ecology & Evolution* **30**, 766–779.
  <https://doi.org/10.1016/j.tree.2015.09.007>
