---
title: "Joint species distribution models for binary occurrence data"
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Joint species distribution models for binary occurrence data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

After an environmental gradient is in the model, which species still
tend to occur together, and which species avoid each other? That is
the first joint species distribution model (JSDM) question this page
answers.

The data are binary occurrence records: one row per site, one column
or trait per species, and a 0/1 response for absence/presence. The
model uses `latent()` to estimate a low-rank residual co-occurrence
structure across species after `env_1` is accounted for.

Scope boundary. IN: the long and `traits(...)` wide formula paths
(FG-02, FG-03), standalone `latent(0 + trait | site, d = K)` (FG-04),
and binomial-logit response fits (FAM-02), with the package fixture
checked by `test-example-joint-sdm.R` and
`test-joint-sdm-binary-long-wide.R`. PARTIAL: calibrated interval
coverage for covariance-derived quantities remains under CI-08 /
CI-10, so this article treats Fisher-z intervals as useful diagnostics
but not final coverage claims. OUT OF SCOPE: single-species spatial
SDMs belong in `sdmTMB`; multivariate spatial fields via
`spatial_*()` are a later advanced article, not a prerequisite for
this JSDM example.

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

## Load and fit

The article uses a shipped teaching fixture rather than building the
data inside the vignette. It contains a complete `Site x Species`
occurrence matrix: 120 sites, 8 species, 2 latent residual
co-occurrence axes, and an environmental covariate `env_1` with
species-specific responses. The true loadings put the species into
four niche quadrants (dry-low, dry-high, wet-low, wet-high), with two
species per quadrant.

```{r jsdm-data}
jsdm_path <- system.file(
  "extdata", "examples", "joint-sdm-example.rds",
  package = "gllvmTMB"
)

source_candidates <- file.path(
  c(".", "..", "../.."),
  "inst", "extdata", "examples", "joint-sdm-example.rds"
)
input_file <- knitr::current_input(dir = TRUE)
if (!is.null(input_file)) {
  source_candidates <- c(
    source_candidates,
    file.path(
      dirname(dirname(dirname(input_file))),
      "inst", "extdata", "examples", "joint-sdm-example.rds"
    )
  )
}
candidate_paths <- c(jsdm_path, source_candidates)
candidate_paths <- candidate_paths[nzchar(candidate_paths)]
jsdm_path <- candidate_paths[file.exists(candidate_paths)][1]
if (is.na(jsdm_path)) {
  stop("Could not find joint-sdm-example.rds.")
}
jsdm <- readRDS(jsdm_path)

df <- jsdm$data_long
df_wide <- jsdm$data_wide
```

The same JSDM fits in both long format and the `traits(...)` wide
format. The long form is shown first; the wide form is the
formula-API equivalent on a pivoted data frame. Both reach the
same engine.

```{r jsdm-fit}
# Long format -- one row per (site, species) observation
fit_jsdm <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):env_1 +
    latent(0 + trait | site, d = 2),
  data   = df,
  trait  = "trait",
  unit   = "site",
  family = binomial()
)
fit_jsdm
```

```{r jsdm-fit-wide}
# Wide format -- one row per site, one column per species. The
# `traits(...)` LHS names the species columns; the compact RHS
# shorthand uses `1 + env_1`, which expands internally to
# `0 + trait + (0 + trait):env_1`.
fit_jsdm_wide <- gllvmTMB(
  traits(sp_1, sp_2, sp_3, sp_4, sp_5, sp_6, sp_7, sp_8) ~
    1 + env_1 + latent(1 | site, d = 2),
  data   = df_wide,
  unit   = "site",
  family = binomial()
)
all.equal(logLik(fit_jsdm), logLik(fit_jsdm_wide), tolerance = 1e-8)
```

### Long or wide data shape

JSDM is the package's natural long-format use case because each row
already carries `(site, species)` and the response factor `trait`
indexes species. When the data already live as one row per site and
one binary column per species, the `traits(...)` wide form is the same
model through the same `gllvmTMB()` entry point. In either shape, true
absences should be explicit zeros. Missing response cells are not a
safe shortcut for absences, because missingness can also mean the site
or species was not surveyed.

## What `latent()` estimates

For binary occurrence data with a logit link, the fitted linear
predictor is

$$
\mathrm{logit}\{P(y_{it} = 1)\} =
\mathbf x_i^\top \boldsymbol\beta_t +
\boldsymbol\lambda_t^\top \mathbf u_i,
\qquad
\mathbf u_i \sim \mathcal N(\mathbf 0, \mathbf I_d).
$$

The site score $\mathbf u_i$ is the unobserved residual state of site
$i$, and the species loading $\boldsymbol\lambda_t$ says how species
$t$ responds to those residual axes. The rotation of
$\boldsymbol\Lambda$ is arbitrary, so the stable co-occurrence target
is the species covariance implied by the loadings:

$$
\boldsymbol\Sigma_B = \boldsymbol\Lambda\boldsymbol\Lambda^\top.
$$

For a binary logit JSDM, correlations are easier to interpret on the
latent-liability scale. `extract_correlations(link_residual = "auto")`
adds the fixed logistic residual variance $\pi^2 / 3$ to the diagonal
before forming the correlation matrix:

$$
\boldsymbol\Sigma_{\mathrm{latent}} =
\boldsymbol\Sigma_B + (\pi^2 / 3)\mathbf I.
$$

That residual term is fixed by the binomial logit link, not estimated
from the data. This is why the first binary JSDM model uses
`latent()` alone. The Gaussian `latent() + unique()` decomposition
estimates a separate diagonal $\boldsymbol\Psi$, but a single-trial
Bernoulli response already has its latent-scale residual variance
set by the link. For the same reason, `dep()` and `indep()` are useful
sensitivity forms, but they are not the recommended starting point
for this public binary JSDM path.

### Where loading constraints fit

The loading matrix $\boldsymbol\Lambda$ is usually exploratory in a
JSDM: its rotation is arbitrary, while
$\boldsymbol\Lambda\boldsymbol\Lambda^\top$ is the interpretable
co-occurrence target. If you have an a-priori hypothesis that a
species group defines an axis, or that a species should have zero
loading on an axis, that is a **confirmatory loading** question rather
than the default species-distribution workflow.

`gllvmTMB` supports that confirmatory path with
`lambda_constraint = list(unit = M)`, where `NA` means "estimate this
loading" and a number means "pin this loading during fitting." Binary
IRT-style loading constraints are covered by the LAM-03 and LAM-04
validation rows, but an ecology-first JSDM article should introduce
them only as an optional follow-up after the reader understands
correlations, $\boldsymbol\Sigma$, and ordination.

## Latent-scale residual species correlations

Pairwise species correlations on the latent-liability scale. The
default `method = "fisher-z"` gives Fisher-transform Wald CIs that
are fast (seconds) and bounded inside $[-1, 1]$ by construction.
`method = "wald"` is an accepted alias for the same Fisher-z path.
Two heavier alternatives are also available: `method = "bootstrap"`
re-fits the model under a parametric bootstrap for a full
sampling-distribution CI, and `method = "profile"` inverts the
profile likelihood pair-by-pair (cost scales as $O(T(T-1)/2)$
constrained refits).

One scale caveat is worth stating up front, because it trips people
up on binary fits. The default `link_residual = "auto"` puts the
correlation on the **latent-liability** scale — it adds the fixed
logistic residual $\pi^2/3$ to the diagonal of
$\boldsymbol\Sigma_B$ before forming the correlation, which is the
interpretable co-occurrence quantity for a JSDM. `fisher-z`,
`wald`, and `bootstrap` all honour that argument. The current
`profile` path, however, always reports the correlation of the
**bare** $\boldsymbol\Sigma_B = \boldsymbol\Lambda\boldsymbol\Lambda^\top$
(it ignores `link_residual`), so its point estimate is the
undiluted latent-factor correlation, not the latent-liability one.
The optional comparison snippets below make the difference explicit
for readers who want to audit the scale.

What is and is not certified: the point estimates and Fisher-z
intervals are useful diagnostics for this fitted example, but the
package-wide calibrated-coverage work is still open. Treat intervals
here as model-based uncertainty summaries, not as a claim that every
binary or mixed-family covariance interval has reached nominal
coverage. See the
[validation-debt register](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/35-validation-debt-register.md)
(CI-08, CI-10) for the live status.

```{r jsdm-cor}
corr_rows <- extract_correlations(
  fit_jsdm,
  tier = "unit",
  method = "fisher-z",
  link_residual = "auto"
)
corr_rows
```

The matrix view below shows the same point estimates as a
species-by-species map. For a binary JSDM, the useful first picture is
the **latent-liability correlation** pattern after adding the fixed
logistic link residual to the denominator; raw
$\boldsymbol\Sigma_B$ entries can be dominated by loading scale and are
harder to interpret. The Fisher-z interval bounds stay in `corr_rows`
rather than being printed inside every matrix cell.

```{r jsdm-correlation-heatmap, fig.width = 6.8, fig.height = 5.8, fig.cap = "Residual species correlations from the fitted binary JSDM. Cells show Fisher-z point estimates on the latent-liability scale after accounting for env_1 and the fixed logistic link residual; interval bounds are available in the preceding table.", fig.alt = "Heatmap of pairwise residual species correlations for eight species. Blue cells indicate positive residual co-occurrence, red cells indicate negative residual co-occurrence, and diagonal self-correlations are one."}
plot_correlations(
  corr_rows,
  style = "heatmap",
  matrix_layout = "by_level",
  label_type = "estimate",
  label_digits = 2,
  include_diagonal = TRUE,
  title = "Residual species correlations",
  subtitle = "Latent-liability correlations after accounting for env_1."
) +
  ggplot2::labs(caption = NULL)
```

Use these correlations, rather than the signs of individual loadings,
for the rotation-invariant co-occurrence story.

## Confidence intervals on the binary latent scale

This section works through the three CI methods on two estimands —
a pairwise species **correlation** and a per-species **ICC** — to
show what each method returns and where it currently stops short.
These chunks are optional because the bootstrap and profile paths
re-fit the model; run them when the interval method itself is the
question, not for a first pass through the article.

### Three CI methods for one correlation

Take a single species pair (`sp_1`, `sp_2`) and ask for the same
correlation three ways. Keeping `link_residual = "auto"` for the
Fisher-z and bootstrap calls holds them on the latent-liability
scale, so their point estimates agree and only the interval differs:

```{r jsdm-cor-methods, eval = FALSE}
pair_ij <- c("sp_1", "sp_2")

cor_fisher <- extract_correlations(
  fit_jsdm, tier = "unit", pair = pair_ij,
  method = "fisher-z", link_residual = "auto"
)
cor_boot <- extract_correlations(
  fit_jsdm, tier = "unit", pair = pair_ij,
  method = "bootstrap", nsim = 20, seed = 1, link_residual = "auto"
)
rbind(cor_fisher, cor_boot)[, c("trait_i", "trait_j",
                                "correlation", "lower", "upper", "method")]
```

The Fisher-z and bootstrap point estimates match because both add
the $\pi^2/3$ logistic residual to the diagonal; the bootstrap
interval is the empirical 2.5/97.5 percentile across the refits.

The `profile` method is the one that reports a *different* number,
because it currently works on the bare $\boldsymbol\Sigma_B$ scale.
To compare it fairly, request the Fisher-z interval with
`link_residual = "none"` so both sit on the same (undiluted) scale:

```{r jsdm-cor-profile, eval = FALSE}
cor_profile <- extract_correlations(
  fit_jsdm, tier = "unit", pair = pair_ij,
  method = "profile"
)
cor_fisher_bare <- extract_correlations(
  fit_jsdm, tier = "unit", pair = pair_ij,
  method = "fisher-z", link_residual = "none"
)
rbind(cor_profile, cor_fisher_bare)[, c("trait_i", "trait_j",
                                        "correlation", "lower", "upper", "method")]
```

Now the two point estimates agree: the profile correlation matches
the bare-scale Fisher-z value. The profile *interval* is a
likelihood-ratio inversion around that point — but note it can come
back as `NA` when the constrained refit fails to bracket the bound
(a near-boundary correlation here is a common cause), which is the
same not-yet-calibrated CI story flagged above. The take-home is
*not* that one method is wrong, but that **`profile` and the default
latent-liability correlations live on different scales** until the
profile path learns to honour `link_residual`. For the interpretable
JSDM co-occurrence story, prefer the latent-liability scale
(`auto`), and use Fisher-z for speed or the bootstrap when you want
an empirical sampling-distribution interval.

### ICC (communality) on the latent scale

For a single-trial binary JSDM, the natural intra-class quantity is
the share of each species' latent-liability variance that is shared
co-occurrence rather than the fixed logistic residual:

$$
\mathrm{ICC}_t \;=\;
\frac{(\boldsymbol\Lambda\boldsymbol\Lambda^\top)_{tt}}
     {(\boldsymbol\Lambda\boldsymbol\Lambda^\top)_{tt} + \pi^2/3}.
$$

With `latent()` and no `unique()` at the unit level there is no
estimated $\boldsymbol\Psi$, so `extract_communality(level = "unit",
link_residual = "auto")` returns exactly this ratio — the
proportion of the latent-liability variance that the shared factors
explain. (This is *not* the between-vs-within `extract_repeatability()`
ICC, which needs a paired `unique(0 + trait | site) + unique(0 +
trait | site_species)` design; a single-trial occurrence matrix has
no within-site replication for that, so `extract_repeatability()`
and `extract_ICC_site()` are not applicable to this fit.)

```{r jsdm-icc-point}
icc_point <- extract_communality(
  fit_jsdm, level = "unit", link_residual = "auto"
)
round(icc_point, 3)
```

A parametric bootstrap puts an interval on each species' ICC:

```{r jsdm-icc-boot, eval = FALSE}
icc_boot <- extract_communality(
  fit_jsdm, level = "unit", link_residual = "auto",
  ci = TRUE, method = "bootstrap", nsim = 20, seed = 1
)
icc_boot[, c("trait", "c2", "lower", "upper", "method")]
```

The intervals are wide — eight species and 120 sites is modest
information for a per-species variance share, and the upper bounds
for the more strongly-loading species run close to 1.

Two honest limitations on the ICC CI in the current build:

* **Wald is not implemented** for communality (it is a non-linear
  function of several parameters); requesting `method = "wald"`
  emits a note and falls back to the bootstrap.
* **Profile bounds are not yet delivered** for this proportion on a
  single-trial binary fit — the constrained refit fails to bracket
  and the interval returns `NA`. The point estimate is still the
  correct latent-scale ICC; only the profile *interval* is missing.

These gaps are the open-coverage story flagged above: the estimands
compute, but calibrated interval machinery on the binary and
mixed-family scales is still in progress (CI-08, CI-10). Report the
point estimate and the method used; avoid presenting these intervals
as coverage-certified.

## Ordination biplot

Site scores on the two latent axes, with species loadings shown as
arrows. Axes are residual co-occurrence after `env_1` has been
accounted for; species loading in the same direction tend to
co-occur (after accounting for the environment).

```{r jsdm-biplot, fig.width = 6.5, fig.height = 5, fig.cap = "Ordination biplot from the fitted binary JSDM, using varimax-rotated and sign-anchored site scores and species loadings. Loading orientation and sign are not identifiable; the model-implied Sigma and correlations remain the rotation-invariant interpretation target.", fig.alt = "Two-dimensional ordination biplot showing site scores as points and species loading arrows after accounting for the environmental predictor. Species arrows in similar directions indicate similar residual co-occurrence patterns, but loading sign and rotation are arbitrary."}
plot(
  fit_jsdm,
  type = "ordination",
  level = "unit",
  rotation = "varimax",
  sign_anchor = "auto",
  standardize_loadings = TRUE
) +
  ggplot2::labs(
    title = "Residual species ordination",
    subtitle = "Varimax-rotated site scores and species loading directions.",
    caption = NULL
  )
```

The fixture's truth places the eight species into **four designed
association groups**: `sp_1` / `sp_2`, `sp_3` / `sp_4`, `sp_5` /
`sp_6`, and `sp_7` / `sp_8`. The fitted arrows should be read for
broad grouping, not axis labels. The exact orientation is arbitrary —
only $\boldsymbol\Lambda\boldsymbol\Lambda^\top$ is identifiable.
Species arrows pointing in the same direction co-occur after `env_1`
is accounted for; arrows pointing in opposite directions tend to
mutually exclude.

## Check loadings before naming axes

The ordination plot is useful, but it is not enough to name an axis.
An exploratory loading matrix is identified only up to rotation, so
per-species loading intervals are not well-defined until the loading
orientation has been pinned. If you have a biological hypothesis, use
`confirmatory_lambda()` to encode it. If you do not, use
`suggest_lambda_constraint()` as a statistical fallback, refit, and
then inspect uncertainty.

Here the fallback should not be just one hard cutoff. The helper below
compares the cheap point-threshold and Wald suggestions with the
slower profile-retention suggestion:

* `varimax_threshold` rotates the exploratory loadings and pins small
  point estimates.
* `wald_retention` uses the same salience threshold, but asks whether
  the loading is retained after accounting for Wald uncertainty on the
  standardised loading scale.
* `profile_retention` refits the model one loading at a time and pins
  entries whose likelihood-ratio test does not reject zero.

For a binary logit JSDM, the Wald retention calculation uses the fixed
logistic residual variance, `sigma_d2 = pi^2 / 3`. This is still a
statistical fallback rather than a biological hypothesis; it is a
reproducible orientation for checking which loadings are stable enough
to discuss.

```{r jsdm-suggest-loadings}
loading_suggestions <- suggest_lambda_constraints(
  fit_jsdm,
  methods = c("varimax_threshold", "wald_retention", "profile_retention"),
  threshold = 0.30,
  retention_prob = 0.90,
  sigma_d2 = pi^2 / 3
)

loading_suggestions$summary[, c(
  "method", "decision_basis", "cost", "n_pins", "n_free"
)]

loading_suggestion <- loading_suggestions$recommended
loading_suggestions$recommended_method
loading_suggestion$constraint
loading_suggestion$note
```

These three suggestions can disagree. The profile path costs more, but
it gives the cleanest loading-diagnostics display for this binary
fixture. The cheaper Wald suggestion over-prunes this example and the
refit has a non-positive-definite Hessian, so it is not a good public
figure even though it remains a useful warning about over-aggressive
pinning.

Refit the same binary JSDM with the suggested loading constraint. The
likelihood may change slightly because the constrained fit is no
longer just a rotated copy of the exploratory model; it is a
reporting-oriented sensitivity fit for the loading display.

```{r jsdm-constrained-loadings}
fit_jsdm_constrained <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):env_1 +
    latent(0 + trait | site, d = 2),
  data   = df,
  trait  = "trait",
  unit   = "site",
  family = binomial(),
  lambda_constraint = list(unit = loading_suggestion$constraint)
)
fit_jsdm_constrained
```

`loading_ci()` gives Wald intervals for the pinned-or-free loading
entries. `flag_unreliable_loadings()` adds a practical reliability
flag: by default here, a loading is unreliable if its interval overlaps
the small null region `[-0.1, 0.1]`.

```{r jsdm-loading-ci}
loading_flags <- flag_unreliable_loadings(
  fit_jsdm_constrained,
  level = "unit",
  method = "wald",
  null_region = c(-0.1, 0.1)
)

loading_flags[, c(
  "trait", "axis", "estimate", "lower", "upper", "pinned", "unreliable"
)]
```

The Confidence Eye plot puts that table into the same visual grammar as
the rest of the package: hollow point for the estimate, eye-shaped
region for the interval, grey for pinned loadings, green when the
interval excludes the null region, and red when it overlaps it.

```{r jsdm-loading-confidence-eye, fig.width = 8, fig.height = 4.8, fig.cap = "Confidence Eye plot for the profile-retention constrained binary JSDM loadings. Hollow points are loading estimates, eye-shaped regions are Wald intervals on the profile-retention refit, grey points are entries pinned by the suggested constraint, green intervals exclude the small null region, and red intervals overlap it.", fig.alt = "Two-panel confidence-eye plot of species loadings for LV1 and LV2 after the profile-retention constraint. Pinned entries are grey points at zero; free entries have hollow estimates and eye-shaped Wald intervals coloured by whether the interval overlaps the small null region."}
plot_loadings_confidence_eye(
  fit_jsdm_constrained,
  level = "unit",
  method = "wald",
  null_region = c(-0.1, 0.1),
  sort_by = "magnitude"
) +
  ggplot2::scale_colour_manual(
    values = c(
      "pinned" = "grey50",
      "CI overlaps null" = "#d6604d",
      "CI excludes null" = "#1b7837"
    ),
    breaks = c("pinned", "CI overlaps null", "CI excludes null"),
    name = NULL
  ) +
  ggplot2::labs(
    y = "Loading estimate",
    subtitle = "Profile-retention constraint; pale lens = 95% Wald CI; grey points are pinned to zero."
  )
```

Read this plot as a loading-diagnostics check, not as a replacement for
the residual correlation matrix. Correlations and
$\boldsymbol\Sigma_B$ are rotation-invariant; loadings become
reportable only after the orientation choice is explicit.

## See also

* [Morphometrics](morphometrics.html) — Gaussian stacked-trait analogue with `latent() + unique()`.
* [Choosing latent rank](model-selection-latent-rank.html) — how to compare candidate values of `d`.
* Loading-constraint helpers — `?confirmatory_lambda` and `?suggest_lambda_constraint` are available for scoped loading-diagnostics work; the full public loading-constraint teaching path is still under article-council review.
* [`vignette("response-families")`](response-families.html) — full table of supported families and the per-family $\sigma^2_d$.

## References

* Niku, J., Warton, D.I., Hui, F.K.C., & Taskinen, S. (2017).
  Generalized linear latent variable models for multivariate count
  and biomass data in ecology. *Journal of Agricultural, Biological,
  and Environmental Statistics* **22**, 498–522.
  <https://link.springer.com/article/10.1007/s13253-017-0304-7>
* Niku, J., Hui, F.K.C., Taskinen, S., & Warton, D.I. (2019).
  gllvm: Fast analysis of multivariate abundance data with
  generalized linear latent variable models in r. *Methods in
  Ecology and Evolution* **10**, 2173–2182.
  <https://doi.org/10.1111/2041-210X.13303>
* Nakagawa, S., & Schielzeth, H. (2010). Repeatability for Gaussian
  and non-Gaussian data: a practical guide for biologists.
  *Biological Reviews* **85**, 935–956.
  <https://doi.org/10.1111/j.1469-185X.2010.00141.x>
* Nakagawa, S., Johnson, P.C.D., & Schielzeth, H. (2017). The
  coefficient of determination R² and intra-class correlation
  coefficient from generalized linear mixed-effects models revisited
  and expanded. *Journal of the Royal Society Interface* **14**,
  20170213. <https://doi.org/10.1098/rsif.2017.0213>
* Warton, D.I., Blanchet, F.G., O'Hara, R.B., Ovaskainen, O.,
  Taskinen, S., Walker, S.C., & Hui, F.K.C. (2015). So many
  variables: joint modeling in community ecology. *Trends in
  Ecology & Evolution* **30**, 766–779.
  <https://doi.org/10.1016/j.tree.2015.09.007>
