---
title: "Explore loading constraints with suggest_lambda_constraint()"
description: "Explore pre-specified and data-derived loading constraints without turning selected pins into biological hypotheses."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Explore loading constraints with suggest_lambda_constraint()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  cache = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE
)
```

When a reduced-rank model has more than one latent dimension, the columns of
its loading matrix $\boldsymbol\Lambda$ can be rotated without changing the
shared covariance $\boldsymbol\Lambda\boldsymbol\Lambda^\top$.
`suggest_lambda_constraint()` supplies a reproducible *identification
scaffold*: a matrix that selects one orientation before fitting. It does not
discover a biological factor structure or test whether a selected loading is
exactly zero.

This page is for an applied user who needs a stable loading orientation for a
plot or a reproducible workflow. It uses the binary JSDM teaching fixture, but
the point is general. The ecological example below is an orientation
demonstration, not evidence for a confirmatory ecological loading structure.

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

## The model and the invariant target

The shipped JSDM fixture records presence or absence for eight species at 120
sites, with a measured environmental predictor. We fit two residual latent
dimensions after allowing each species its own environmental response.

```{r data}
jsdm_path <- system.file(
  "extdata", "examples", "joint-sdm-example.rds",
  package = "gllvmTMB"
)
if (!nzchar(jsdm_path)) {
  jsdm_path <- file.path(
    "inst", "extdata", "examples", "joint-sdm-example.rds"
  )
}
jsdm <- readRDS(jsdm_path)
df_long <- jsdm$data_long
```

For a binomial-logit JSDM, the model is

$$
\operatorname{logit}\{P(y_{it}=1)\} =
\mathbf{x}_i^\top\boldsymbol\beta_t +
\boldsymbol\lambda_t^\top\mathbf{u}_i,
\qquad \mathbf{u}_i \sim \mathcal{N}(\mathbf{0}, \mathbf{I}_d).
$$

Here $\boldsymbol\lambda_t$ is the row of loadings for species $t$.
The individual columns of $\boldsymbol\Lambda$ depend on the chosen
orientation; the implied covariance is the more stable scientific target.
See [Joint species distribution models](joint-sdm.html) for a full analysis
and [Covariance and correlation](covariance-correlation.html) for the
interpretation of $\boldsymbol\Sigma$.

## Choose an identification convention before fitting

`suggest_lambda_constraint()` returns a matrix with `NA` for an estimated
entry and a number for an exact pin. Its default `"lower_triangular"`
convention fixes only the strict upper triangle to zero. For $d = 2$, that is
one zero; for $d = K$, it is $K(K - 1)/2$ zeros. These zeros identify an
orientation. They are not selected from the JSDM's fitted loadings.

```{r constraint}
model_formula <- value ~ 0 + trait + (0 + trait):env_1 +
  latent(0 + trait | site, d = 2)

sug <- suggest_lambda_constraint(
  model_formula,
  data = df_long,
  trait = "trait",
  unit = "site",
  convention = "lower_triangular"
)

sug$constraint
sug$usage_hint
```

The figure makes the contract visible. The single blue cell is fixed exactly
at zero; every white cell is estimated. The first two rows carry the minimal
lower-triangular scaffold, while the remaining rows are unconstrained.

```{r constraint-figure, fig.alt="A species-by-latent-dimension constraint matrix. White cells labelled estimate are free loadings. One dark blue cell in the first row and second latent dimension is fixed at zero, which supplies the lower-triangular orientation convention."}
constraint_plot <- as.data.frame(as.table(sug$constraint), stringsAsFactors = FALSE)
names(constraint_plot) <- c("species", "dimension", "value")
constraint_plot$status <- ifelse(
  is.na(constraint_plot$value), "estimated", "fixed at 0"
)

ggplot(constraint_plot, aes(x = dimension, y = species, fill = status)) +
  geom_tile(colour = "white", linewidth = 0.7) +
  geom_text(aes(label = ifelse(is.na(value), "estimate", "0")), size = 3) +
  scale_fill_manual(values = c("estimated" = "grey95", "fixed at 0" = "#0A617D")) +
  labs(
    x = "Latent dimension", y = "Species", fill = NULL,
    title = "A lower-triangular loading orientation",
    subtitle = "`NA` means estimate; numeric entries are exact pins"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), legend.position = "top")
```

## Fit and check the oriented model

Pass the returned matrix to `gllvmTMB()` through `lambda_constraint`. The
constraint changes the loading orientation, not the purpose of the model:
we still interpret residual co-occurrence through
$\boldsymbol\Lambda\boldsymbol\Lambda^\top$ and the corresponding
latent-scale correlations.

```{r fit}
fit_oriented <- gllvmTMB(
  model_formula,
  data = df_long,
  trait = "trait",
  unit = "site",
  family = binomial(),
  lambda_constraint = list(unit = sug$constraint)
)

health <- check_gllvmTMB(fit_oriented)
health[health$component %in% c("optimizer_convergence", "max_gradient", "pd_hessian"), ]
```

Do not interpret loading labels if these health rows fail. A constraint can
choose a coordinate system; it cannot repair a weakly identified model or add
information that the data do not contain.

```{r covariance}
Sigma_oriented <- extract_Sigma(fit_oriented, level = "unit", part = "shared")
round(Sigma_oriented$Sigma, 2)
```

## Compare exploratory retention conventions

The lower-triangular convention above is fixed before fitting. The other
conventions start from an exploratory fit and propose exact zero pins for a
new, data-derived constrained refit. They answer different screening
questions, so their disagreement is useful to inspect rather than something
to hide. Because an exact zero changes the fitted model, its implied covariance
can differ from the exploratory fit; this is not merely a coordinate rotation.

```{r exploratory-fit}
fit_exploratory <- gllvmTMB(
  model_formula,
  data = df_long,
  trait = "trait",
  unit = "site",
  family = binomial()
)

comparison_fast <- suggest_lambda_constraints(
  fit_exploratory,
  methods = c("varimax_threshold", "wald_retention"),
  threshold = 0.30,
  retention_prob = 0.90
)
comparison_table <- data.frame(
  convention = c("Varimax threshold", "Wald retention"),
  decision = c(
    "Rotate first, then propose zeros for small standardised loadings.",
    "Propose zeros when the fitted uncertainty does not support a loading above the chosen threshold."
  ),
  proposed_zero_pins = vapply(
    comparison_fast$suggestions,
    function(x) x$n_pins,
    integer(1)
  ),
  free_loadings = vapply(
    comparison_fast$suggestions,
    function(x) length(x$constraint) - x$n_pins,
    integer(1)
  )
)
knitr::kable(comparison_table)
```

`varimax_threshold` first rotates the loading axes so that species associate
more clearly with one dimension, then proposes zero pins for standardised
loadings below 0.30. Treat 0.30 as a display choice, not a biological cutoff.
`wald_retention` retains a loading only when the fitted model assigns at least
a 90% probability to its standardised magnitude exceeding 0.30; otherwise it
proposes a zero pin. Neither convention is a hypothesis test for a biological
zero. The figure compares the resulting pin matrices: blue cells are exact
zero restrictions in the proposed constrained refit; white cells remain
estimated.

```{r fast-comparison-figure, fig.alt="Two species-by-latent-dimension matrices compare varimax-threshold and Wald-retention orientation suggestions. Blue cells are proposed exact zero pins and white cells remain estimated. The patterns can differ because one method uses point estimates and the other also uses baseline standard errors."}
make_constraint_frame <- function(suggestion, method) {
  out <- as.data.frame(as.table(suggestion$constraint), stringsAsFactors = FALSE)
  names(out) <- c("species", "dimension", "value")
  out$method <- method
  out$status <- ifelse(is.na(out$value), "estimated", "proposed zero")
  out
}

fast_constraints <- do.call(
  rbind,
  Map(
    make_constraint_frame,
    comparison_fast$suggestions,
    names(comparison_fast$suggestions)
  )
)

ggplot(fast_constraints, aes(x = dimension, y = species, fill = status)) +
  geom_tile(colour = "white", linewidth = 0.7) +
  geom_text(aes(label = ifelse(is.na(value), "estimate", "0")), size = 3) +
  facet_wrap(~ method) +
  scale_fill_manual(values = c("estimated" = "grey95", "proposed zero" = "#0A617D")) +
  labs(
    x = "Latent dimension", y = "Species", fill = NULL,
    title = "Exploratory retention conventions can propose different pins",
    subtitle = "Each blue cell would become an exact zero in a constrained refit"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), legend.position = "top")
```

### The slower profile-retention convention

`profile_retention` replaces the Wald screen with a one-entry-at-a-time
likelihood-ratio refit. It is useful when you want to see how a
likelihood-based screen differs from the fast conventions, but it is not a
confirmation procedure: the same exploratory data proposed both the loading
and the test. It is also expensive—one refit for each testable loading—and
leaves an entry free when a test refit is unhealthy. Run it deliberately after
the exploratory fit is healthy.

```{r profile-retention, eval=FALSE}
comparison_profile <- suggest_lambda_constraints(
  fit_exploratory,
  methods = "profile_retention",
  retention_prob = 0.90
)
comparison_profile$suggestions$profile_retention$constraint

# This is a new, data-derived constrained refit, not a harmless rotation of
# the exploratory model. Screen it before describing its covariance summaries.
fit_profile_orientation <- gllvmTMB(
  model_formula,
  data = df_long,
  trait = "trait",
  unit = "site",
  family = binomial(),
  lambda_constraint = list(unit = comparison_profile$recommended$constraint)
)
check_gllvmTMB(fit_profile_orientation)
```

The code is not run during routine article rendering because it performs many
refits. It is intentionally written out in full so that the cost and the
post-fit health check are visible rather than hidden. If the exploratory or
constrained fit is unhealthy, simplify the model or collect more informative
data; do not treat the profile screen as a repair.

## What this helper does not establish

The other conventions in `suggest_lambda_constraint()` include
`"varimax_threshold"`, `"wald_retention"`, and `"profile_retention"`.
They can propose an exploratory, data-derived constrained refit after
inspecting a fit. They do not convert the selected zeros into prior biological
hypotheses, and fitting the selected constraint to the same data does not
confirm those zeros. Therefore this article does not compare their refits with
AIC, BIC, or likelihood-ratio tests, and it does not report loading confidence
intervals after selection.

If you have a biological hypothesis before inspecting the data, construct its
matrix explicitly with `confirmatory_lambda()` and state the exact numerical
pins as model assumptions. If you need to choose the number of latent
dimensions, compare a pre-specified set of candidate ranks with the separate
[Choosing latent rank](model-selection-latent-rank.html) workflow. That page
uses `logLik()`, AIC, BIC, and `check_gllvmTMB()` for its demonstrated
Gaussian candidate set; it does not validate information criteria as a
universal rank-selection rule.

## Next steps

* `?suggest_lambda_constraint` documents all conventions and arguments.
* [Joint species distribution models](joint-sdm.html) shows the full binary
  occurrence workflow and its rotation-invariant covariance interpretation.
* [Can I trust this fit?](fit-diagnostics.html) explains the health table
  shown above and what to do when a fit is weak.
