---
title: "Confirmatory loadings for binary species distributions"
tier: 3 # Hidden while the binary loading-constraint reader path is under article-council review.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Confirmatory loadings for binary species distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

> **Internal article gate:** This page is kept buildable for
> maintainers while the binary loading-constraint teaching path is under
> article-council review. The current public binary worked example is
> [`vignette("joint-sdm")`](joint-sdm.html); do not treat this page as a
> first-stop tutorial until its rendered article review is recorded.

## Why this article exists

Joint species distribution models (JSDMs) factor a multi-species
community into shared latent variables plus species-specific loadings on
those axes (Pollock *et al.* 2014; Warton *et al.* 2015; Niku *et al.*
2017, 2019). The fitted loading matrix $\boldsymbol{\Lambda}$ is the
biological story: which species respond to which latent gradient and
how strongly.

But $\boldsymbol{\Lambda}$ is **identified only up to rotation**.
Exploratory fits give a $\boldsymbol{\Lambda}$ that is mathematically
correct but visually arbitrary -- two different optimisation starts can
return rotated versions of the same fit, and the orientation of the
axes is not the orientation suggested by your biology (Niku *et al.*
2019; Tobler *et al.* 2019).

When you have **prior knowledge** about which species should (or
should not) load on which latent variables, you can encode that knowledge
directly as zero pins on $\boldsymbol{\Lambda}$. The model becomes
*confirmatory* rather than *exploratory*, and the axes inherit a fixed
biological meaning (Norberg *et al.* 2019 discuss confirmatory JSDM
comparisons across 33 datasets). This article walks the
`lambda_constraint` mechanism on a binary species distribution dataset
with prior functional-group membership.

A complementary integration of prior knowledge is **trait mediation**,
as in HMSC (Ovaskainen *et al.* 2017; Tikhonov *et al.* 2019), where
loadings are continuous functions of measured species traits. Use
HMSC's trait machinery when traits are continuous and many; use
`lambda_constraint` when prior knowledge is discrete and precise
(known structural zeros, functional-group membership, biological
hypotheses to test).

## When you need `lambda_constraint`

Reach for `lambda_constraint` when **any** of the following is true:

- You have a categorical hypothesis about a species -- "shade-tolerant
  species should not respond to the light gradient" (a structural zero).
- You have **functional-group** classifications and want each group to
  load on a specific axis.
- You want to **test** a hypothesis: fit an exploratory model and a
  constrained model, then compare by likelihood-ratio test or AIC.
- You need the rotation pinned so two analysts working on the same data
  get the same axes.

You probably don't need `lambda_constraint` when:

- You have rich continuous trait data -- use HMSC's $T r$ matrix instead
  (Ovaskainen *et al.* 2017; Tikhonov *et al.* 2019).
- You have only one latent dimension ($d = 1$) -- rotation
  indeterminacy doesn't exist, and the identification pin is automatic.
- Your goal is purely predictive, not interpretive -- the exploratory
  fit gives the same likelihood as any rotated equivalent.

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)

set.seed(20260527)
```

## A 20-species forest understory community

We simulate a community of 20 forest understory plants at 40 sites
(the moderate-$n$ regime is on purpose -- at very large $n$ the
exploratory fit's leakage on the structural-zero entries becomes
invisible, and the article would have nothing visual to compare).
The community has **three functional groups**:

- **Group A -- shade specialists** (5 species): respond positively to
  the shade-tolerance axis (LV1); do **not** respond to the
  drought-tolerance axis (LV2).
- **Group B -- drought specialists** (5 species): respond positively
  to LV2; do **not** respond to LV1.
- **Group C -- no prior hypothesis** (10 species): loadings are free
  on both axes. These are species the ecologist has not pre-classified
  or chooses not to constrain.

Group C is a single category, not two. (An earlier draft separated
"generalists" from "unknown" species -- the distinction was cosmetic
because both classes carry `NA` in the constraint matrix, so we
collapse them into one C group.)

We use a **binary probit link** because (i) presence/absence is the
canonical JSDM response type, and (ii) probit's latent residual
variance is $\sigma^2_d = 1$ exactly -- the cleanest non-Gaussian
identifiability story (see the `response-families` vignette).

### True loadings

```{r dgp-truth}
n_species <- 20L
n_sites   <- 40L
d         <- 2L

species_names <- c(
  paste0("A_shade_",   1:5),
  paste0("B_drought_", 1:5),
  paste0("C_free_",    1:10)
)

# True loadings: structural zeros encode the functional-group hypothesis.
# We deliberately use moderate (not huge) magnitudes so the exploratory
# fit's sampling-noise leakage on the structural-zero entries is visible.
Lambda_true <- matrix(0, nrow = n_species, ncol = d)
rownames(Lambda_true) <- species_names
colnames(Lambda_true) <- c("LV1_shade", "LV2_drought")

# Group A (rows 1-5):  load on LV1 only.
Lambda_true[1:5,   1] <- stats::runif(5, 0.5, 1.0)
Lambda_true[1:5,   2] <- 0
# Group B (rows 6-10): load on LV2 only.
Lambda_true[6:10,  1] <- 0
Lambda_true[6:10,  2] <- stats::runif(5, 0.5, 1.0)
# Group C (rows 11-20): no prior, signed on both axes.
Lambda_true[11:20, 1] <- stats::runif(10, -1.0, 1.0)
Lambda_true[11:20, 2] <- stats::runif(10, -1.0, 1.0)

round(Lambda_true, 2)
```

### Simulating the binary observations

```{r dgp-fit-data}
# Latent factors: one row per site.
U <- matrix(stats::rnorm(n_sites * d), nrow = n_sites, ncol = d)

# Species intercepts (modest variation; binary probit calibrated near 0.5).
alpha_true <- stats::rnorm(n_species, mean = 0, sd = 0.3)

# Linear predictor (probit scale): n_sites x n_species.
eta <- matrix(alpha_true, nrow = n_sites, ncol = n_species, byrow = TRUE) +
       U %*% t(Lambda_true)

# Binary presence/absence.
prob   <- stats::pnorm(eta)
y_wide <- matrix(stats::rbinom(length(prob), 1L, prob),
                 nrow = n_sites, ncol = n_species)
colnames(y_wide) <- species_names
```

### Both data shapes: long and wide

`gllvmTMB()` accepts the **same fit** in two equivalent shapes
(see the `gllvm-vocabulary` vignette for full details):

- **wide** -- one row per site, one column per species. Natural shape
  for a site-by-species table from a vegetation survey.
- **long** -- one row per (site, species) pair. Natural shape after a
  `tidyr::pivot_longer()`.

```{r data-shapes}
# Wide: 40 sites x 20 species, plus a site identifier.
df_wide <- as.data.frame(y_wide)
df_wide$site <- factor(seq_len(n_sites))

dim(df_wide)
df_wide[1:3, c("site", "A_shade_1", "B_drought_1", "C_free_1")]

# Long: 40 * 20 = 800 rows.
df_long <- data.frame(
  site  = factor(rep(seq_len(n_sites), times = n_species)),
  trait = factor(rep(species_names, each = n_sites),
                 levels = species_names),
  value = as.integer(c(y_wide))
)

dim(df_long)
head(df_long, 3)
```

### Prevalence diagnostic

Before fitting, sanity-check that the simulated data are not saturated
at 0 or 1 -- that would make the fit ill-conditioned regardless of
constraint structure.

```{r prevalence, fig.width = 6, fig.height = 3, fig.cap = "Per-species prevalences (proportion of sites at which each species is present) for all 20 simulated species, coloured by functional group. The community is well-conditioned: no species is at 0% or 100% prevalence, so each species contributes information to the latent-axis estimation.", fig.alt = "Bar chart of 20 species prevalences ranging roughly between 0.15 and 0.85. Bars are coloured by functional group: A_shade (orange), B_drought (teal), C_free (grey)."}
df_prev <- data.frame(
  species = factor(species_names, levels = species_names),
  prev    = colMeans(y_wide),
  group   = factor(c(rep("A_shade", 5), rep("B_drought", 5),
                     rep("C_free", 10)),
                   levels = c("A_shade", "B_drought", "C_free"))
)

ggplot(df_prev, aes(x = species, y = prev, fill = group)) +
  geom_col() +
  scale_fill_manual(values = c("A_shade" = "#e08214",
                               "B_drought" = "#2c7a7b",
                               "C_free" = "grey60")) +
  geom_hline(yintercept = c(0.05, 0.95),
             linetype = "dashed", colour = "grey40") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7)) +
  labs(x = NULL, y = "Prevalence", fill = NULL,
       title = "Species prevalences (target: away from 0 and 1)")
```

## The exploratory fit (rotation-arbitrary baseline)

A pure exploratory fit identifies $\boldsymbol{\Lambda}$ only up to
rotation. To make even the *exploratory* model identified, the
package adopts a **lower-triangular** convention internally:
$\Lambda_{i, j} = 0$ for $j > i$ when $i \le d$, and the diagonal is
constrained to be positive. This is documented in
`?lambda_constraint`.

We fit it in **long** form. (The wide interface gives the identical
likelihood from the same data in `traits(all_of(species_names)) ~ ...`
form; the `gllvm-vocabulary` vignette demonstrates the long/wide
equivalence in detail, so we don't repeat the duplicate fit here.)

```{r fit-exploratory-long}
fit_exp_long <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2L),
  data   = df_long,
  family = stats::binomial(link = "probit")
)

cat("Long exploratory  -- conv:", fit_exp_long$opt$convergence,
    "  logLik:", round(as.numeric(logLik(fit_exp_long)), 2), "\n")
```

## The biological constraint matrix

The biology says: group A loads on LV1 (shade) only, group B loads on
LV2 (drought) only, group C is free on both axes. The package gives
us a one-call builder for that exact pattern:

```{r constraint-matrix}
M_bio <- confirmatory_lambda(
  species  = species_names,
  group    = c(rep("A_shade",   5),
               rep("B_drought", 5),
               rep("C_free",   10)),
  d        = 2L,
  loads_on = list(A_shade = 1L, B_drought = 2L),
  axis_labels = c("LV1_shade", "LV2_drought")
)
M_bio
```

`confirmatory_lambda()` takes care of all the identification
bookkeeping for you:

- For each named group it writes zeros on every axis the group does
  **not** load on (group A gets `M[1:5, 2] = 0`; group B gets
  `M[6:10, 1] = 0`).
- It picks one anchor species per axis (the first species in each
  loaded group) and pins it at `+1` to fix the axis scale and sign
  (here `A_shade_1` anchors LV1 and `B_drought_1` anchors LV2).
- Group C is absent from `loads_on`, so it remains `NA` (free) on
  both axes.

Counting pins:

- **2 anchor pins** -- $\Lambda_{1,1} = 1$ and $\Lambda_{6,2} = 1$.
- **5 group-A structural zeros** at $\Lambda_{1{:}5,\, 2} = 0$.
- **5 group-B structural zeros** at $\Lambda_{6{:}10,\, 1} = 0$.

That is **12 pins total**, leaving $2 \times 20 - 12 = 28$ free
loading entries. Compared to the exploratory baseline (which carries
one upper-triangular pin internally), the constrained model adds
**11 net pins**, so the LRT below uses df = 11.

If your hypothesis doesn't map cleanly onto "this group loads on that
axis" -- for example a free-form CFA pattern, idiosyncratic
species-by-species pins, or a confirmatory design from outside
ecology -- build the matrix by hand using `NA` for free entries and
numeric values for pinned entries. If you have **no** prior
hypothesis at all, see the *fallback* section at the end of this
article that introduces `suggest_lambda_constraint()`.

### Visualising what gets pinned

```{r constraint-tile, fig.width = 4.5, fig.height = 5, fig.cap = "The constraint matrix M_bio. Red cells are the ten biological structural zeros (group A on LV2, group B on LV1). Green cells are the two positive anchors (rows 1 and 6) that fix the scale and sign of each latent axis. White cells (group C and the non-anchor diagonal of A and B) are free and estimated from the data.", fig.alt = "Tile plot of the 20-species by 2-axis constraint matrix. Ten cells are red (pinned to zero), two cells are green (pinned to +1), and 28 cells are white (free)."}
df_pin <- data.frame(
  species = factor(rep(species_names, times = d), levels = rev(species_names)),
  axis    = factor(rep(c("LV1_shade", "LV2_drought"), each = n_species),
                   levels = c("LV1_shade", "LV2_drought")),
  pin     = as.numeric(M_bio)
)
df_pin$kind <- factor(
  ifelse(is.na(df_pin$pin),     "free",
  ifelse(df_pin$pin == 0,       "zero pin",
                                "anchor (+1)")),
  levels = c("zero pin", "anchor (+1)", "free")
)

ggplot(df_pin, aes(x = axis, y = species, fill = kind)) +
  geom_tile(colour = "grey80") +
  scale_fill_manual(values = c("zero pin"    = "#d6604d",
                               "anchor (+1)" = "#1b7837",
                               "free"        = "white"),
                    name = NULL) +
  theme_minimal(base_size = 9) +
  theme(panel.grid = element_blank()) +
  labs(x = NULL, y = NULL,
       title = "Constraint matrix M_bio",
       subtitle = "Red = pinned to 0; Green = pinned to +1; White = free")
```

## Fitting the constrained model

We fit the constrained model in long form (the wide interface gives
the identical likelihood, exactly as for the exploratory fit above):

```{r fit-con-long}
fit_con_long <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2L),
  data              = df_long,
  family            = stats::binomial(link = "probit"),
  lambda_constraint = list(unit = M_bio)
)

cat("Long constrained -- conv:", fit_con_long$opt$convergence,
    "  logLik:", round(as.numeric(logLik(fit_con_long)), 2), "\n")
```

For the rest of the article we abbreviate the long-form fits as
`fit_con` and `fit_exp`, since the diagnostic helpers below expect the
long shape.

```{r fit-con-alias}
fit_con <- fit_con_long
fit_exp <- fit_exp_long
```

## Does the constraint improve the model?

There are three things to check, and they tell complementary stories:

1. **LRT** -- Is the constraint *inconsistent* with the data?
   (A small $p$ would say yes; large $p$ means the constraint is
   compatible.)
2. **AIC** -- Does the constraint *improve* expected predictive
   accuracy after paying the parameter-parsimony price?
3. **BIC** -- Does the constraint *improve* with a stronger penalty
   for parameters?

The constrained model is **nested inside** the exploratory model
(11 net pins added), so the LRT statistic
$-2(\ell_{\text{con}} - \ell_{\text{exp}})$ is approximately
$\chi^2_{11}$ under the null that the biological structure is correct.

```{r model-compare}
ll_exp <- as.numeric(logLik(fit_exp))
ll_con <- as.numeric(logLik(fit_con))
df_exp <- attr(logLik(fit_exp), "df")
df_con <- attr(logLik(fit_con), "df")

lrt_stat <- 2 * (ll_exp - ll_con)
lrt_df   <- df_exp - df_con
lrt_p    <- stats::pchisq(lrt_stat, df = lrt_df, lower.tail = FALSE)

aic_exp <- AIC(fit_exp); aic_con <- AIC(fit_con)
bic_exp <- BIC(fit_exp); bic_con <- BIC(fit_con)

data.frame(
  model        = c("exploratory", "constrained"),
  free_params  = c(df_exp, df_con),
  logLik       = round(c(ll_exp, ll_con), 2),
  AIC          = round(c(aic_exp, aic_con), 1),
  BIC          = round(c(bic_exp, bic_con), 1)
)
```

```{r lrt-summary}
data.frame(
  LR_statistic = round(lrt_stat, 2),
  df           = lrt_df,
  p_value      = signif(lrt_p, 3),
  delta_AIC    = round(aic_con - aic_exp, 1),
  delta_BIC    = round(bic_con - bic_exp, 1)
)
```

Reading these together:

- **LRT** -- the constrained likelihood is only slightly below the
  exploratory likelihood, and the difference is well within what
  $\chi^2_{11}$ expects under the null. A large $p$ means we fail to
  reject the structural hypothesis -- the constraint is *compatible*
  with the data. (Failing to reject is not the same as proving the
  hypothesis true; it is the standard inferential reading.)
- **AIC** -- negative $\Delta \text{AIC}_{\text{con}-\text{exp}}$ means
  the constrained model is preferred after the 11-parameter parsimony
  penalty. The biology buys us a more efficient description of the
  same data.
- **BIC** -- BIC penalises parameters more aggressively than AIC, so
  $\Delta \text{BIC}$ favours the constrained model by an even larger
  margin. By BIC the structural hypothesis is **strongly preferred**.

Because the truth here was simulated with the structural zeros in
place, all three diagnostics point the same way. If you supplied an
incorrect constraint, the LRT $p$ would be small and $\Delta \text{AIC}$
would be positive -- both would push you back to the exploratory model.

## Inspecting the constrained loadings

We pull the fitted loadings through the package's public helper
`getLoadings()` rather than poking into the fit object directly:

```{r get-loadings}
Lambda_exp <- getLoadings(fit_exp, level = "unit")
Lambda_con <- getLoadings(fit_con, level = "unit")

# Friendlier axis labels (the helper returns "LV1", "LV2"; we rename
# in-place to remind the reader which axis is which biology).
colnames(Lambda_exp) <- colnames(Lambda_con) <-
  c("LV1_shade", "LV2_drought")
```

### Structural-zero leakage: a one-line summary

If the biology says "Group A does not respond to drought (LV2)" and
the truth is consistent with that hypothesis, the constrained fit
forces those entries to zero by construction. The exploratory fit
does not -- sampling noise leaks into the entries that *should* be
zero. We can put a single number on the leakage:

```{r leakage}
zero_idx <- rbind(
  cbind(row = 1:5,  col = 2L),   # Group A on LV2 (should be zero)
  cbind(row = 6:10, col = 1L)    # Group B on LV1 (should be zero)
)

leakage <- data.frame(
  model    = c("exploratory", "constrained"),
  mean_abs = c(
    mean(abs(Lambda_exp[zero_idx])),
    mean(abs(Lambda_con[zero_idx]))
  )
)
leakage$mean_abs <- round(leakage$mean_abs, 3)
leakage
```

The exploratory mean-absolute leakage is a real, sampling-driven
number that *should* be zero under the biological hypothesis; the
constrained value is exactly zero by construction. The model-comparison
table above showed that paying for those structural zeros buys
$\Delta \text{AIC} < 0$ and $\Delta \text{BIC} \ll 0$ -- i.e., the
constraint is not a cosmetic improvement, it is a measurable one.

### Exploratory vs constrained -- side by side

```{r loadings-compare, fig.width = 8, fig.height = 5, fig.cap = "Estimated loadings from the exploratory fit (top row) and the constrained fit (bottom row), separated by latent axis (LV1 left, LV2 right). In the constrained fit, group A is exactly zero on LV2 and group B is exactly zero on LV1 -- the imposed biological pins. In the exploratory fit, those same entries show sampling-noise leakage (small but visible bars) because the rotation convention does not know about the biology.", fig.alt = "Four-panel bar chart, 2 by 2 grid. Rows are model (exploratory, constrained); columns are axis (LV1, LV2). Group A bars are non-zero on both panels in the exploratory row but exactly zero on LV2 in the constrained row. Group B bars are non-zero on both panels in the exploratory row but exactly zero on LV1 in the constrained row. Group C bars are scattered in sign and magnitude in all four panels."}
make_df <- function(Lambda, label) {
  data.frame(
    species = factor(rep(species_names, times = d), levels = species_names),
    axis    = factor(rep(c("LV1_shade", "LV2_drought"), each = n_species),
                     levels = c("LV1_shade", "LV2_drought")),
    loading = as.numeric(Lambda),
    group   = factor(rep(rep(c("A_shade", "B_drought", "C_free"),
                              times = c(5L, 5L, 10L)), times = d),
                     levels = c("A_shade", "B_drought", "C_free")),
    model   = label
  )
}

df_compare <- rbind(
  make_df(Lambda_exp, "exploratory"),
  make_df(Lambda_con, "constrained")
)
df_compare$model <- factor(df_compare$model,
                           levels = c("exploratory", "constrained"))

ggplot(df_compare, aes(x = species, y = loading, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey40") +
  facet_grid(model ~ axis) +
  scale_fill_manual(values = c("A_shade"   = "#e08214",
                               "B_drought" = "#2c7a7b",
                               "C_free"    = "grey60")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 6),
        strip.text  = element_text(face = "bold")) +
  labs(x = NULL, y = expression(hat(Lambda)), fill = NULL,
       title = "Exploratory vs constrained loadings",
       subtitle = "Top: sampling-noise leakage on the biological zeros. Bottom: zeros enforced.")
```

## Error-aware loading interpretation

A point estimate of $\hat\Lambda_{i,k}$ is half the story. Two
loadings of the same magnitude can carry wildly different
uncertainty; a "dominant" loading whose 95% CI brackets zero is not
evidence that the species responds to the gradient. This is one of
the few cases where the JSDM/GLLVM literature lags the psychometric
literature, where Wald, profile, and bootstrap CIs on loadings have
been routine for decades (Cudeck & O'Dell 1994; Zhang *et al.* 2010).
On the ecology side it is essentially absent from the ML/TMB JSDM
toolchain -- `gllvm`, `sjSDM`, `ecoCopula` all return point estimates
of $\Lambda$ without per-entry CIs in their default summaries -- and
the only published ecology-side warning that GLLVM Wald SEs can
*overstate* confidence is ter Braak & van der Veen (2025).

The package gives us three helpers for the confirmatory case.

### Per-entry confidence intervals

`loading_ci()` returns a Wald CI per loading via the delta method on
a numerical Jacobian of $\Lambda$ with respect to the packed
reduced-rank parameters. It is **gated to confirmatory fits**
because per-entry CIs are not well-defined under rotation
indeterminacy (Mansolf & Reise 2016) -- the helper errors on an
exploratory fit and points you at `confirmatory_lambda()` or at the
rotation-invariant alternatives (`extract_communality(ci = TRUE)`,
`extract_Sigma()`).

```{r loading-ci}
ci <- loading_ci(fit_con, level = "unit")
head(ci, 10)
```

When the fitted Hessian is positive-definite, pinned entries have
`se = 0` exactly (they're not random variables -- the constraint
nailed them down) and free entries inherit the SE from TMB's
`sdreport` covariance restricted to the loading parameters. In this
fixture the Hessian is non-positive-definite (see the Confidence Eye
section), so `loading_ci()` returns `se`/`lower`/`upper` as `NA` with
`ci_status = "not_available_non_positive_definite_hessian"`; the table
above illustrates the output shape, not usable Wald intervals.

### Reliability classification

Once you have CIs, you can ask the basic decision question: which
loadings are *not* consistent with the null hypothesis that the
species responds trivially to the axis? The reference design here
is Zientek & Thompson (2007), who proposed "retain a loading only if
the CI excludes a biologically negligible region." `flag_unreliable_loadings()`
implements that contract with a user-supplied `null_region`:

```{r flag-unreliable}
fl <- flag_unreliable_loadings(fit_con, null_region = c(-0.1, 0.1))

# Tabulate reliability classes
table(
  pinned       = fl$pinned,
  reliability  = ifelse(is.na(fl$unreliable), "pinned",
                 ifelse(fl$unreliable,        "CI overlaps null",
                                              "CI excludes null"))
)
```

Read this table as: pinned entries carry no inference, and each free
entry is classified reliable/unreliable by whether its CI excludes the
null region -- *when* Wald intervals are available. In this
non-positive-definite fixture every free entry's CI is `NA`, so the
classification collapses to the pinned bucket; a genuine reliability
split needs a positive-definite fit, and per-entry interval calibration
itself remains an open gate (`CI-08`/`CI-10`).

### Symmetric vs asymmetric Wald

The default `loading_ci(method = "wald")` gives symmetric CIs on the
$\Lambda$ scale. That is *first-order* asymptotically correct but
ignores the fact that the standardised loading
$\rho = \Lambda / \sqrt{\Lambda^2 + \sigma_d^2}$ is bounded in
$(-1, 1)$. For large $|\hat\Lambda|$ that boundedness pushes the
sampling distribution of $\hat\Lambda$ rightward, so symmetric Wald
under-states the upper bound. `method = "wald_asym"` corrects this by
running Wald on the Fisher-z scale and back-transforming:

```{r symmetric-vs-asymmetric}
ci_sym  <- loading_ci(fit_con, level = "unit", method = "wald")
ci_asym <- loading_ci(fit_con, level = "unit", method = "wald_asym")

cmp <- data.frame(
  trait    = ci_sym$trait,
  axis     = ci_sym$axis,
  estimate = round(ci_sym$estimate, 2),
  sym_CI   = sprintf("[%.2f, %.2f]", ci_sym$lower,  ci_sym$upper),
  asym_CI  = sprintf("[%.2f, %.2f]", ci_asym$lower, ci_asym$upper),
  skew     = round((ci_asym$upper - ci_asym$estimate) -
                   (ci_asym$estimate - ci_asym$lower), 2)
)
# Show the largest-magnitude free entries -- where asymmetry matters most.
free_big <- !ci_sym$pinned & abs(ci_sym$estimate) > 0.6
cmp[free_big, ][order(-abs(cmp$estimate[free_big])), ][1:6, ]
```

For loadings near zero, symmetric and asymmetric agree. For loadings
where $|\hat\Lambda|$ is large, asymmetric Wald widens the
upper bound and tightens the lower bound -- the qualitatively right
shape, matching what profile likelihood and parametric bootstrap
would give up to higher-order corrections (see the discussion of
Mansolf & Reise 2016 and ter Braak & van der Veen 2025 in the
references).

### Confidence Eye plot

`plot_loadings_confidence_eye()` is the natural visual companion.
Closest published precedent is Hoegh & Roberts (2020), whose
`UncertainOrd` plots site-score uncertainty on ordination biplots; to
our knowledge no equivalent per-species $\Lambda$ uncertainty plot
exists in the ML/TMB JSDM lineage. The "Confidence Eye" contract is:
pale CI region with thick semi-transparent line, hollow point
estimate, optional shaded "negligible" band so the reader can tell at
a glance which CIs sit outside it.

```{r confidence-eye, fig.width = 8, fig.height = 4.5, fig.cap = "Diagnostic Confidence Eye attempt for the constrained loadings. In this fixture the Wald loading-CI path is unavailable because the fitted Hessian is non-positive-definite, so the rendered plot shows hollow point estimates only rather than confidence-eye polygons. This is kept as an internal review signal, not as a public-ready loading-uncertainty figure.", fig.alt = "Two-panel chart with 20 species on the x-axis, separated by latent axis (LV1, LV2). Hollow point estimates are shown, but no confidence-eye polygons appear because the fitted Hessian is non-positive-definite and Wald loading intervals are unavailable for this fit."}
plot_loadings_confidence_eye(
  fit_con,
  level       = "unit",
  null_region = c(-0.1, 0.1)
)
```

The current rendered output is a review blocker, not a finished teaching
figure. It shows where the Confidence Eye should appear, but the fitted
Hessian is non-positive-definite and the Wald loading-CI helper correctly
refuses to draw uncertainty polygons. Before this article returns to the
public navigation, this section needs either a fixture with positive-definite
loading intervals or a profile/bootstrap loading-interval path that can carry
the intended uncertainty display.

## Communality and uniqueness

Loadings are easier to interpret when normalised. Two complementary
views:

- **Total communality** (per species): the share of the species'
  total variance explained by the latent factors together,
  $h_i = \sum_k \Lambda_{i, k}^2 / \big(\sum_k \Lambda_{i, k}^2 + \sigma^2_d\big)$,
  with $\sigma^2_d = 1$ for the probit link.
  The package's `extract_communality()` returns this directly.
- **Per-axis share**: the partition of the latent variance between
  LV1 and LV2, $\Lambda_{i, k}^2 / \sum_{k'} \Lambda_{i, k'}^2$.

```{r communality-total}
# Total communality per species (latent share of total variance).
total_comm <- extract_communality(fit_con, level = "unit")
round(total_comm, 2)
```

```{r communality-per-axis-numeric}
# Per-axis share of latent variance.
row_var     <- rowSums(Lambda_con^2)
communality <- Lambda_con^2 / row_var
round(communality, 2)
```

A stacked bar makes the per-axis partition immediate:

```{r communality-plot, fig.width = 7, fig.height = 4, fig.cap = "Per-axis share of latent variance in the constrained fit. Each species' latent variance is partitioned between LV1 (shade) and LV2 (drought). Group A species are 100 percent on LV1 by construction; group B species are 100 percent on LV2 by construction; group C species partition the variance freely.", fig.alt = "Stacked bar chart, one bar per species, normalised to 1. Group A bars are fully orange (100% LV1). Group B bars are fully teal (100% LV2). Group C bars have mixed orange-and-teal proportions."}
df_comm <- data.frame(
  species = factor(rep(species_names, times = d), levels = species_names),
  axis    = factor(rep(c("LV1_shade", "LV2_drought"), each = n_species),
                   levels = c("LV1_shade", "LV2_drought")),
  share   = as.numeric(communality)
)

ggplot(df_comm, aes(x = species, y = share, fill = axis)) +
  geom_col() +
  scale_fill_manual(values = c("LV1_shade"   = "#e08214",
                               "LV2_drought" = "#2c7a7b")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7)) +
  labs(x = NULL, y = "Per-axis share", fill = NULL,
       title = "Per-axis latent-variance share (constrained fit)")
```

## Variance partition with confidence intervals

The communality plot above gives the per-axis split of the *latent*
variance. The complementary view is the per-species variance
*decomposition*: how much of each species' total liability variance
comes from the shared latent factors versus the link-implicit residual?
For the probit fit this partition has two components:

- **`shared_unit`** --
  $(\boldsymbol\Lambda\boldsymbol\Lambda^\top)_{tt} / \boldsymbol\Sigma_{tt}$,
  the share of species $t$'s liability variance attributable to the
  shared latent factors at the unit (site) tier. Identical numerator
  to the total communality above, but expressed as a proportion of
  the **total** liability variance.
- **`link_residual`** --
  $\sigma^2_d / \boldsymbol\Sigma_{tt}$, the probit noise floor.
  Fixed at $\sigma^2_d = 1$ for probit, so it equals
  $1 - \texttt{shared\_unit}$ by construction.

`extract_proportions()` returns the point estimates:

```{r vp-point}
vp_point <- extract_proportions(fit_con, format = "wide")
# Round only the numeric proportion columns; keep the trait label intact.
num_cols <- vapply(vp_point, is.numeric, logical(1))
vp_point[, num_cols] <- lapply(vp_point[, num_cols, drop = FALSE],
                               round, digits = 2)
vp_point
```

Profile-likelihood CIs on `shared_unit` come from
`profile_ci_proportions()`, or equivalently via the unified
`confint()` surface. We restrict to three representative species (one
from each functional group) so the demo renders in well under a minute
rather than half an hour; widen `demo_idx` to profile more species:

```{r vp-profile, cache = TRUE}
demo_idx <- c(1,    # Group A: shade specialist
              6,    # Group B: drought specialist
              11)   # Group C: free
demo_species <- species_names[demo_idx]

vp_ci <- profile_ci_proportions(fit_con,
                                components = "shared_unit",
                                trait_idx  = demo_idx,
                                level      = 0.95)
vp_show <- vp_ci[, c("trait", "proportion", "lower", "upper")]
vp_show[, c("proportion", "lower", "upper")] <-
  lapply(vp_show[, c("proportion", "lower", "upper")], round, digits = 2)
vp_show
```

The same numbers are reachable through the unified `confint()` surface
(`confint(fit_con, parm = "proportion:shared_unit:<trait>", method = "profile")`),
which is the recommended user-facing entry point: it places variance
partitions on the same `parm`/`method` API as Lambda entries,
correlations, ICC, communality, and fixed effects. See
`?confint.gllvmTMB_multi` for the full grammar.

Two implementation notes worth flagging:

- `profile_ci_proportions()` is a constrained Lagrange refit, like
  `profile_ci_communality()`. The difference is that the denominator
  $\boldsymbol\Sigma_{tt}$ floats across **all** variance components
  present in the fit, so the CI is on the proportion of the *total*
  liability variance, not the proportion of the tier-internal
  variance. For this single-tier fit the two collapse to the same
  number; on a multi-tier fit (e.g. with `indep()` or
  source-specific `*_indep()` terms) they would diverge.
- `link_residual` is flagged as `method = "(unavailable)"` here -- for
  probit it is a known constant ($\sigma^2_d = 1$), so there is no
  profile to invert and the bounds collapse to the point estimate.

A stacked bar with the profile CI overlaid on the boundary between
the two components makes the partition immediate:

```{r vp-plot, fig.width = 7.5, fig.height = 4, fig.cap = "Variance partition for three representative species from the constrained probit fit. The orange segment is the shared-latent contribution `shared_unit`; the grey segment is the probit link residual. The vertical error bars at the boundary between the segments are the 95 percent profile-likelihood CI on `shared_unit`; their asymmetry is the bounded-ratio property profile likelihood captures and symmetric Wald would miss.", fig.alt = "Stacked bar chart, one bar per species, normalised to 1. Lower (orange) segment is the shared-latent proportion; upper (grey) segment is the link-residual proportion. A short vertical error bar sits at the boundary between segments, asymmetric, never crossing 0 or 1."}
vp_long <- data.frame(
  species   = factor(vp_ci$trait, levels = demo_species),
  component = "shared_unit",
  prop      = vp_ci$proportion,
  lower     = vp_ci$lower,
  upper     = vp_ci$upper
)
vp_res <- vp_long
vp_res$component <- "link_residual"
vp_res$prop      <- 1 - vp_long$prop
vp_res$lower     <- NA_real_
vp_res$upper     <- NA_real_

vp_stack <- rbind(vp_long, vp_res)
vp_stack$component <- factor(vp_stack$component,
                             levels = c("link_residual", "shared_unit"))

ggplot() +
  geom_col(data = vp_stack,
           aes(x = species, y = prop, fill = component)) +
  geom_errorbar(data = vp_long,
                aes(x = species, ymin = lower, ymax = upper),
                width = 0.30, colour = "grey20") +
  scale_fill_manual(
    values = c(shared_unit   = "#e08214",
               link_residual = "grey85"),
    labels = c(shared_unit   = "shared_unit (latent)",
               link_residual = "link_residual (probit floor)")
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(x = NULL, y = "Proportion of liability variance",
       fill = NULL,
       title = "Variance partition with 95% profile CI on shared_unit")
```

How to read this:

- Some of the three demo species have **two-sided CIs** --
  the profile likelihood crosses the $\chi^2$ cutoff on both sides
  of the point estimate. These are the cleanest summaries.
- Others have **one-sided CIs** -- the profile crosses the
  cutoff on one side but not the other within the grid extent. The
  unreached side is reported as `NA` (the profile didn't bracket the
  cutoff at all) or as the proportion boundary `0`/`1` (the profile
  ran into the natural support boundary first). This is the profile's
  honest answer when the data don't constrain that side of the bound
  sharply: profile_ci_proportions() refuses to extrapolate. Symmetric
  Wald on the raw proportion would have invented a number instead.
- Group C species (signed on both LV1 and LV2) tend to have smaller
  `shared_unit` proportions because splitting variance across two
  axes leaves smaller per-axis magnitudes; the total latent share is
  correspondingly lower.
- Where they're available, CI **widths** are wider for species with
  `shared_unit` near 0.5 and narrower at the extremes -- the standard
  bounded-proportion pattern. The asymmetry (wider reach toward the
  far boundary) is what profile likelihood captures and symmetric
  Wald on the raw proportion would miss.

For species with one-sided CIs, the companion `profile_proportions()`
function (same fit, same components) returns the LR curve itself --
you can inspect where the profile crosses the $\chi^2$ cutoff and
decide whether the available bound is informative for your purpose.
The bracket-and-bisect inversion in `profile_ci_proportions()` keeps
its answer honest by reporting `NA` rather than extrapolating beyond
where the data constrain.

The same machinery extends to multi-tier fits (with `indep()` or
source-specific `*_indep()` terms) where the partition has three or more
components and `link_residual` is the smallest piece rather than half
the bar. The `parm` token grammar is uniform --
`"proportion:<component>[:trait]"` -- so the demo above generalises
without learning new function names.

## Reliability regime: what we know empirically

`lambda_constraint` on binary fits is validated in
`tests/testthat/test-m2-3-lambda-constraint-binary.R`
(parameter recovery at `n_items` $\in \{20, 50\} \times d \in \{1, 2\}$)
plus cross-package agreement with `mirt` and `galamm`
(Sorensen 2024) in
`tests/testthat/test-m2-3-mirt-cross-check.R` and
`tests/testthat/test-m2-3-galamm-cross-check.R`.
`suggest_lambda_constraint()` is validated in
`tests/testthat/test-m2-4-suggest-lambda-constraint-binary.R`.

The validation-debt rows `LAM-01`, `LAM-03`, `LAM-04` are at
`covered` for binary
(see [`docs/design/35-validation-debt-register.md`](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/35-validation-debt-register.md)).

**Practical regime**:

- Recovery is reliable at `n_items` $\ge 20$, $d \in \{1, 2\}$.
- $d = 3$ at `n_items = 10` is the documented boundary -- the
  suggester may not always stabilise; check `fit_health$pd_hessian`.
- This article's fixture (`n_species = 20`, `d = 2`, `n_sites = 40`)
  matches the validated `(n_items, d)` grid for point recovery
  (`LAM-03`: `n_items` $\in \{20, 50\} \times d \in \{1, 2\}$, validated
  at `n_resp` $\approx 400$--$500$). It sits well below that sample
  size (`n_sites = 40`, chosen to make leakage visible), and the
  validated regime does not certify fit health or interval calibration
  -- consistent with this fixture's non-positive-definite Hessian (see
  the Confidence Eye section).

## When *not* to use `lambda_constraint`

Three honest caveats:

1. **You only have $d = 1$**. There is nothing to rotate, no
   biological zeros to pin (except trivial sign anchors). The
   identification pin is automatic; you don't need an explicit
   `lambda_constraint` argument.
2. **Your prior knowledge is continuous and trait-mediated**. If
   "species with bigger seeds tend to load more on LV1," that's a
   smooth-function-of-traits hypothesis. HMSC's $T r$ matrix
   (Tikhonov *et al.* 2019) is built for it; `lambda_constraint` would
   discretise the prior into a coarse pin pattern and lose information.
3. **You don't actually have prior knowledge**. Confirmatory analysis
   without genuine *a priori* hypotheses is just exploratory analysis
   with extra steps. Run exploratory first; bring `lambda_constraint`
   when you have a hypothesis to test.

## Fallback: no hypothesis? Use `suggest_lambda_constraint()`

When you have no biological hypothesis to encode, the package can still
propose a minimal, purely *statistical* set of pins that make
$\boldsymbol{\Lambda}$ identified -- via `suggest_lambda_constraint()`,
including data-driven conventions (`varimax_threshold`,
`profile_retention`) that read the exploratory fit and pin entries the
data leave small or unreliable. That fallback path -- the suggester
conventions, the data-driven refits, their model comparison, and the
non-positive-definite vs positive-definite identifiability lesson -- is
covered in its own companion article,
[`vignette("lambda-constraint-suggest")`](lambda-constraint-suggest.html),
so this confirmatory walkthrough stays focused on the hypothesis-driven
case.

## See also

- [`vignette("joint-sdm")`](joint-sdm.html) -- exploratory binary JSDM
  without constraint; this article's companion.
- [`vignette("lambda-constraint-suggest")`](lambda-constraint-suggest.html)
  -- the `suggest_lambda_constraint()` fallback when you have no
  hypothesis: data-driven retention conventions and the non-PD vs PD
  identifiability lesson.
- [`vignette("psychometrics-irt")`](psychometrics-irt.html) -- mixed
  Likert + binary confirmatory factor analysis using the same
  `lambda_constraint` mechanism.
- [`vignette("response-families")`](response-families.html) -- the
  full per-family table including $\sigma^2_d$.
- `?confirmatory_lambda` -- biology-aware constraint builder used in
  this article.
- `?loading_ci`, `?flag_unreliable_loadings`,
  `?plot_loadings_confidence_eye` -- error-aware loading interpretation.
- `?lambda_constraint` and `?suggest_lambda_constraint` -- argument
  reference for the underlying mechanism and the statistical-scaffold
  fallback.
- `?extract_communality` (with `ci = TRUE`) and `?extract_Sigma` --
  rotation-invariant alternatives when the fit is exploratory rather
  than confirmatory.

## How `lambda_constraint` is implemented

Internally, `lambda_constraint = list(unit = M)` is translated to a TMB
`map` factor that fixes the corresponding entries of the
packed-`theta` parameter vector. Diagonal pins use the engine's
positive-diagonal parameterisation; off-diagonal pins set the
corresponding `theta` entries to fixed values and remove them from the
optimisation. The bookkeeping lives in
[`R/lambda-constraint.R`](https://github.com/itchyshin/gllvmTMB/blob/main/R/lambda-constraint.R)
and is verified by
[`tests/testthat/test-lambda-constraint.R`](https://github.com/itchyshin/gllvmTMB/blob/main/tests/testthat/test-lambda-constraint.R).

## References

- Comrey, A.L., & Lee, H.B. (1992). *A First Course in Factor
  Analysis* (2nd ed.). Hillsdale, NJ: Lawrence Erlbaum. (Source of
  the 0.30 / 0.40 / 0.50 standardised-loading retention thresholds
  used by `convention = "varimax_threshold"`.)
- Cudeck, R., & O'Dell, L.L. (1994). Applications of standard error
  estimates in unrestricted factor analysis: significance tests for
  factor loadings and correlations. *Psychological Bulletin* **115**,
  475--487. <https://doi.org/10.1037/0033-2909.115.3.475>
- Hoegh, A., & Roberts, D.W. (2020). Evaluating and presenting
  uncertainty in model-based unconstrained ordination. *Ecology and
  Evolution* **10**, 59--69.
  <https://doi.org/10.1002/ece3.5752>
- Mansolf, M., & Reise, S.P. (2016). Exploratory bifactor analysis:
  the Schmid-Leiman orthogonalization and Jennrich-Bentler analytic
  rotations. *Multivariate Behavioral Research* **51**, 698--717.
  <https://doi.org/10.1080/00273171.2016.1215898>
- Niku, J., Warton, D.I., Hui, F.K.C., & Taskinen, S. (2017).
  Generalized linear latent variable models for multivariate count
  and biomass data in ecology. *Journal of Agricultural, Biological,
  and Environmental Statistics* **22**, 498--522.
  <https://doi.org/10.1007/s13253-017-0304-7>
- Niku, J., Hui, F.K.C., Taskinen, S., & Warton, D.I. (2019).
  gllvm: Fast analysis of multivariate abundance data with
  generalized linear latent variable models in R. *Methods in
  Ecology and Evolution* **10**, 2173--2182.
  <https://doi.org/10.1111/2041-210X.13303>
- Norberg, A., Abrego, N., Blanchet, F.G., Adler, F.R., Anderson,
  B.J., Anttila, J., *et al.* (2019). A comprehensive evaluation of
  predictive performance of 33 species distribution models at
  species and community levels. *Ecological Monographs* **89**(3),
  e01370. <https://doi.org/10.1002/ecm.1370>
- McGillycuddy, M., Warton, D.I., Popovic, G., & Bolker, B.M. (2025).
  Parsimoniously fitting large multivariate random effects in
  `glmmTMB`. *Journal of Statistical Software* **112**(1), 1--19.
  <https://doi.org/10.18637/jss.v112.i01>
- Ovaskainen, O., Tikhonov, G., Norberg, A., Blanchet, F.G., Duan, L.,
  Dunson, D., Roslin, T., & Abrego, N. (2017). How to make more out
  of community data? A conceptual framework and its implementation as
  models and software. *Ecology Letters* **20**, 561--576.
  <https://doi.org/10.1111/ele.12757>
- Pollock, L.J., Tingley, R., Morris, W.K., Golding, N., O'Hara, R.B.,
  Parris, K.M., Vesk, P.A., & McCarthy, M.A. (2014). Understanding
  co-occurrence by modelling species simultaneously with a joint
  species distribution model (JSDM). *Methods in Ecology and Evolution*
  **5**, 397--406.
  <https://doi.org/10.1111/2041-210X.12180>
- Sorensen, O. (2024). Multilevel semiparametric latent variable
  modeling in R with `galamm`. *Multivariate Behavioral Research*
  **59**(5), 1098--1105.
  <https://doi.org/10.1080/00273171.2024.2385336>
- ter Braak, C.J.F., & van der Veen, B. (2025). Generalized linear
  latent variable models can overstate trait--environment
  confidence. *Environmental and Ecological Statistics* (in press).
  <https://doi.org/10.1007/s10651-025-00696-0>
- Tikhonov, G., Opedal, O.H., Abrego, N., Lehikoinen, A., de Jonge,
  M.M.J., Oksanen, J., & Ovaskainen, O. (2019). Joint species
  distribution modelling with the R package `Hmsc`. *Methods in
  Ecology and Evolution* **11**, 442--447.
  <https://doi.org/10.1111/2041-210X.13345>
- Tobler, M.W., Kery, M., Hui, F.K.C., Guillera-Arroita, G.,
  Knaus, P., & Sattler, T. (2019). Joint species distribution
  models with species correlations and imperfect detection.
  *Ecology* **100**(8), e02754.
  <https://doi.org/10.1002/ecy.2754>
- Warton, D.I., Blanchet, F.G., O'Hara, R.B., Ovaskainen, O.,
  Taskinen, S., Walker, S.C., & Hui, F.K.C. (2015). So many
  variables: joint modelling in community ecology. *Trends in
  Ecology & Evolution* **30**, 766--779.
  <https://doi.org/10.1016/j.tree.2015.09.007>
- Zhang, G., Preacher, K.J., & Luo, S. (2010). Bootstrap confidence
  intervals for ordinary least squares factor loadings and
  correlations in exploratory factor analysis. *Multivariate
  Behavioral Research* **45**, 104--134.
  <https://doi.org/10.1080/00273170903504836>
- Zientek, L.R., & Thompson, B. (2007). Applying the bootstrap to the
  multivariate case: bootstrap component / factor analysis. *Behavior
  Research Methods* **39**, 318--325.
  <https://doi.org/10.3758/BF03193163>
