---
title: "Confirmatory loadings: pinning Lambda entries with `lambda_constraint`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Confirmatory loadings: pinning Lambda entries with `lambda_constraint`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

> **Preview — Gaussian path validated; binary IRT is M2.3 milestone work.**
>
> The Gaussian-path `lambda_constraint` workflow (LAM-01 + LAM-02
> `covered` per PR-0B.3 row #18 walk) is the production surface
> today. Binary IRT confirmatory-loadings (LAM-03) walks to
> `covered` after the **M2.3** milestone closes. See the
> [validation-debt register](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/35-validation-debt-register.md).

By default, `gllvmTMB`'s reduced-rank loading matrix
$\boldsymbol{\Lambda}$ is exploratory: lower-triangular with positive
diagonal, identifiable up to *trait permutation* and *column rotation*.
This is fine for ordination, but **for confirmatory hypotheses you
often want to fix specific entries during MLE** -- e.g.

* "Trait $t_1$ defines factor 1 (set $\Lambda_{1,1} = 1$, all other
  loadings on factor 1 free)" -- the GALAMM (Sorensen 2024) /
  classical IRT convention.
* "Trait $t_3$ does NOT load on factor 2 (set $\Lambda_{3,2} = 0$)" --
  testing a structural zero implied by theory.

`gllvmTMB` supports both via the `lambda_constraint` argument, which
takes an `n_traits x d` matrix where `NA` means "estimate freely" and
a numeric value means "pin to that value during fit". The mechanism
follows GALAMM (Sorensen 2024) and is verified by recovery tests in
`tests/testthat/test-lambda-constraint.R`.

## Setup

```{r setup}
library(gllvmTMB)
```

## When do you actually *need* `lambda_constraint`?

The honest answer: **less often than you might think**. The
implied trait covariance
$\boldsymbol{\Sigma}_{\text{unit}} = \boldsymbol\Lambda \boldsymbol\Lambda^{\!\top} + \boldsymbol{\Psi}$
is *always* identifiable, regardless of any rotational ambiguity in
$\boldsymbol\Lambda$ itself. A short decision tree:

| Your goal | Do you need `lambda_constraint`? |
|---|---|
| Estimate $\boldsymbol\Sigma_{\text{unit}}$ (correlations between traits, communalities) | **No.** $\boldsymbol\Sigma_{\text{unit}}$ is rotation-invariant. Just fit and call `extract_Sigma(level = "unit")`. |
| Plot an ordination (`ordiplot()`) | **No.** Axes are rotation-invariant up to reflection / ordering. |
| Compare loadings across two `gllvmTMB` fits on the same data | **No.** Use `compare_loadings()` (Procrustes alignment) instead. |
| Compare loadings against another package (`gllvm`, `glmmTMB::rr()`, ...) | **No.** Use Procrustes alignment again -- both packages have their own arbitrary rotation. |
| Test "does trait $t$ load on factor $k$?" (confirmatory) | **Yes.** Pin $\Lambda_{t,k} = 0$ and likelihood-ratio test against the unconstrained fit. |
| Anchor factor 1 to a specific trait (GALAMM / IRT convention) | **Yes.** Pin $\Lambda_{1,1} = 1$. |
| Get a *unique* numerical $\boldsymbol\Lambda$ for one-shot reporting | **Yes.** Pin to lower-triangular (call `suggest_lambda_constraint()` below). |

If your goal isn't in the bottom three rows, fitting without any
constraint and using rotation-invariant accessors (`extract_Sigma`,
`extract_communality`, `extract_ordination`) is usually the right
move.

## Pre-check: ask the package what it would suggest

If you decide you do want a constraint, the package can construct a
sensible default for you. `suggest_lambda_constraint()` takes either
a fitted model or a `formula + data` and returns the matrix to pass
to `lambda_constraint = list(B = ...)`.

```{r suggest-pre-check}
sim_pre <- simulate_site_trait(
  n_sites = 60, n_species = 1, n_traits = 4, mean_species_per_site = 1,
  Lambda_B = matrix(0.0, 4, 2), psi_B = rep(0, 4),
  beta = matrix(0, 4, 2), seed = 1
)
sug_lt <- suggest_lambda_constraint(
  value ~ 0 + trait + latent(0 + trait | site, d = 2) +
                      unique(0 + trait | site),
  data = sim_pre$data,
  convention = "lower_triangular"
)
sug_lt$constraint
sug_lt$note
sug_lt$usage_hint
```

The three available conventions:

| Convention | What it pins | When to use |
|---|---|---|
| `"lower_triangular"` *(default)* | Upper triangle of $\boldsymbol\Lambda$ pinned to 0 -- `K(K-1)/2` pins | Standard rotational fix; gives a unique $\hat{\boldsymbol\Lambda}$ |
| `"pin_top_one"` | $\Lambda_{1,1} = 1$ only | Single-anchor convention (GALAMM / IRT) -- sets scale, does *not* remove rotation |
| `"none"` | All `NA` (no pins) | If you plan to use post-hoc rotation (varimax) instead |

For $K = 1$, `lower_triangular` returns all-`NA` because there is no
rotational ambiguity to fix -- the function knows this and tells you
in `note`.

## Hands-on: four `lambda_constraint` patterns

This article demonstrates four `lambda_constraint` patterns on the
same simulated data:

1. **Exploratory** -- no constraint (default).
2. **Diagonal pinned to 1** -- the GALAMM / IRT convention.
3. **Off-diagonal pinned to 0** -- structural zero.
4. **Off-diagonal pinned to a non-zero target** -- confirmatory test.

## Simulate a known $\boldsymbol{\Lambda}_B$

Four traits, two latent factors, 60 sites -- same simulator as the
package's recovery tests.

```{r sim}
set.seed(1)
n_traits <- 4
d        <- 2
Lambda_B_true <- matrix(0, nrow = n_traits, ncol = d)
Lambda_B_true[lower.tri(Lambda_B_true, diag = TRUE)] <-
  c(1.0,  0.4, -0.3,  0.2,    # column 1: diag + below
    1.0,  0.6, -0.2)           # column 2: diag + below
Lambda_B_true

sim <- simulate_site_trait(
  n_sites               = 60,
  n_species             = 10,
  n_traits              = n_traits,
  mean_species_per_site = 5,
  Lambda_B              = Lambda_B_true,
  psi_B                 = rep(0.3, n_traits),
  seed                  = 1
)
df_long <- sim$data
head(df_long, 4)
```

The simulator returns long-format data (one row per `(site, species, trait)`
observation). The same simulated data can be reshaped to one of the
two wide forms accepted by `gllvmTMB`:

```{r build-wide}
library(tidyr)
df_wide <- tidyr::pivot_wider(
  df_long,
  id_cols     = c(site, site_species),
  names_from  = trait,
  values_from = value
)
head(df_wide, 4)
```

## 1. Exploratory fit (no constraint)

The same model can be fitted from either data shape. We show both for
the exploratory fit; the constrained fits in the next sections re-use
the long-form recipe.

```{r fit-exploratory}
# Long format (canonical)
fit0_long <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2) +
                      unique(0 + trait | site),
  data = df_long
)

# Wide data-frame format (same model with compact formula syntax)
fit0_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3, trait_4) ~ 1 +
          latent(1 | site, d = 2) +
          unique(1 | site),
  data = df_wide
)

# Verify the two routes give the same fit.
cat("Long/wide logLik difference:",
    signif(abs(as.numeric(logLik(fit0_long)) -
               as.numeric(logLik(fit0_wide))), 3),
    "\n")

# The rest of the article uses the canonical long-format fit.
fit0 <- fit0_long
L0 <- fit0$report$Lambda_B
rownames(L0) <- paste0("trait_", seq_len(n_traits))
colnames(L0) <- paste0("LV", seq_len(d))
round(L0, 3)
```

The lower-triangular convention forces $\Lambda_{1,2} = 0$ and
$\Lambda_{1,1}, \Lambda_{2,2} > 0$ for identifiability. The remaining
five entries are free.

## 2. Diagonal pinned to 1 (GALAMM / IRT convention)

```{r fit-galamm}
cnst_diag <- matrix(NA_real_, nrow = n_traits, ncol = d)
diag(cnst_diag) <- 1
cnst_diag

fit_diag <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2) +
                      unique(0 + trait | site),
  data              = df_long,
  lambda_constraint = list(B = cnst_diag)
)
L_diag <- fit_diag$report$Lambda_B
rownames(L_diag) <- paste0("trait_", seq_len(n_traits))
colnames(L_diag) <- paste0("LV", seq_len(d))
round(L_diag, 3)
```

The first two diagonal entries are exactly 1 (verified by the
`test-lambda-constraint.R` recovery tests). The remaining
lower-triangle entries are free and the fit finds their MLE under
that constraint.

## 3. Off-diagonal pin to zero (structural zero)

Hypothesis: trait 4 should not load on factor 2 -- there is a
biological reason to believe its variance is fully captured by factor
1 plus its unique variance. Set $\Lambda_{4,2} = 0$:

```{r fit-zero-pin}
cnst_zero <- matrix(NA_real_, nrow = n_traits, ncol = d)
diag(cnst_zero) <- 1
cnst_zero[4, 2] <- 0    # trait 4 does NOT load on LV 2
cnst_zero

fit_zero <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2) +
                      unique(0 + trait | site),
  data              = df_long,
  lambda_constraint = list(B = cnst_zero)
)
L_zero <- fit_zero$report$Lambda_B
rownames(L_zero) <- paste0("trait_", seq_len(n_traits))
colnames(L_zero) <- paste0("LV", seq_len(d))
round(L_zero, 3)
```

The pinned entry is exactly 0 (machine precision). A likelihood-ratio
test against the unconstrained model tests whether the structural
zero is supported by the data:

```{r lrt-zero}
ll_diag <- as.numeric(stats::logLik(fit_diag))
ll_zero <- as.numeric(stats::logLik(fit_zero))
LR      <- 2 * (ll_diag - ll_zero)
df_diff <- 1
p_value <- pchisq(LR, df = df_diff, lower.tail = FALSE)
data.frame(
  ll_unconstrained = ll_diag,
  ll_zero_pinned   = ll_zero,
  LR_stat          = LR,
  df               = df_diff,
  p_value          = p_value
)
```

If `p_value` is large, the data are compatible with $\Lambda_{4,2} = 0$;
if small, the data reject the structural zero. (Remember
likelihood-ratio tests on boundary parameters can have
non-standard reference distributions; treat as a heuristic.)

## 4. Off-diagonal pin to a non-zero target

The previous case used 0; you can pin to any value. For example, fix
$\Lambda_{3,1}$ at $-0.5$ and let everything else float:

```{r fit-nonzero-pin}
cnst_05 <- matrix(NA_real_, nrow = n_traits, ncol = d)
diag(cnst_05) <- 1
cnst_05[3, 1] <- -0.5
cnst_05

fit_05 <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2) +
                      unique(0 + trait | site),
  data              = df_long,
  lambda_constraint = list(B = cnst_05)
)
L_05 <- fit_05$report$Lambda_B
rownames(L_05) <- paste0("trait_", seq_len(n_traits))
colnames(L_05) <- paste0("LV", seq_len(d))
round(L_05, 3)
```

`L_05[3, 1]` is exactly $-0.5$ (verified by the recovery tests).

## Comparison across the four fits

```{r side-by-side}
side_by_side <- list(
  exploratory       = round(L0,     3),
  diag_pinned_to_1  = round(L_diag, 3),
  zero_pin_at_42    = round(L_zero, 3),
  nonzero_pin_at_31 = round(L_05,   3)
)
side_by_side
```

A faceted heatmap makes the structure obvious -- each panel shows
$\hat{\boldsymbol\Lambda}_B$ from one of the four fits, with red /
blue indicating the sign of each loading. Pinned cells are outlined
in heavy black:

```{r side-by-side-heatmap, fig.width = 7.5, fig.height = 4, eval = requireNamespace("ggplot2", quietly = TRUE), fig.cap = "Recovered Lambda_B from the four fits. Each panel is a 4 (trait) x 2 (latent factor) heatmap. Heavy black outlines mark cells whose value was pinned by `lambda_constraint`."}
library(ggplot2)
make_long <- function(L, fit_name, pinned_idx = matrix(integer(0), 0, 2)) {
  df_l <- data.frame(
    trait  = factor(rep(rownames(L), ncol(L)), levels = rownames(L)),
    factor = factor(rep(colnames(L), each = nrow(L)), levels = colnames(L)),
    value  = as.vector(L),
    pinned = FALSE,
    fit    = fit_name
  )
  for (k in seq_len(nrow(pinned_idx))) {
    df_l$pinned[df_l$trait  == rownames(L)[pinned_idx[k, 1]] &
                df_l$factor == colnames(L)[pinned_idx[k, 2]]] <- TRUE
  }
  df_l
}
pinned_diag <- rbind(c(1, 1), c(2, 2))
pinned_zero <- rbind(c(1, 1), c(2, 2), c(4, 2))
pinned_05   <- rbind(c(1, 1), c(2, 2), c(3, 1))
df_heat <- rbind(
  make_long(L0,     "1. exploratory"),
  make_long(L_diag, "2. diag pinned to 1",        pinned_diag),
  make_long(L_zero, "3. + zero pin at L[4,2]",    pinned_zero),
  make_long(L_05,   "4. + nonzero pin at L[3,1]", pinned_05)
)
df_heat$fit <- factor(df_heat$fit, levels = c(
  "1. exploratory", "2. diag pinned to 1",
  "3. + zero pin at L[4,2]", "4. + nonzero pin at L[3,1]"))

ggplot(df_heat, aes(factor, trait, fill = value)) +
  geom_tile(colour = "grey90") +
  geom_tile(data = subset(df_heat, pinned),
            colour = "black", linewidth = 1, fill = NA) +
  geom_text(aes(label = sprintf("%.2f", value)),
            colour = "black", size = 3) +
  scale_fill_gradient2(low = "#3b82c6", mid = "white", high = "#d33",
                       midpoint = 0,
                       limits = c(-1, 1) * max(abs(df_heat$value)) * 1.1) +
  facet_wrap(~ fit, ncol = 4) +
  labs(x = "latent factor", y = "trait",
       fill = expression(hat(Lambda))) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(),
        legend.position = "right",
        strip.text = element_text(face = "bold"))
```

The two off-pattern pins ($\Lambda_{4,2}$ in panel 3, $\Lambda_{3,1}$
in panel 4) are held *exactly* at their target values; everything
else floats and adjusts. Panel 1's exploratory loadings are
rotation-only-identified -- a different optimisation start could give
a numerically different $\hat{\boldsymbol\Lambda}$ but the same
implied $\boldsymbol\Sigma_{\text{unit}}$.

## How `lambda_constraint` is implemented

Internally, `lambda_constraint = list(B = M)` is translated to a TMB
`map` factor that fixes the corresponding entries of the packed-theta
parameter vector. The bookkeeping (which packed-theta index
corresponds to which $\Lambda_{i,j}$) lives in
`R/lambda-constraint.R` and is verified by
`tests/testthat/test-lambda-constraint.R` (pinning diagonals,
off-diagonals to zero, off-diagonals to non-zero, upper-triangle
pins silently ignored, dimension-mismatch errors, `W`-level
constraints, combined `B + W` constraints).

## See also

* `?gllvmTMB` -- the `lambda_constraint` argument is documented under
  "Confirmatory loadings".
* [`vignette("functional-biogeography")`](functional-biogeography.html)
  -- the full Nakagawa et al. model also accepts `lambda_constraint`
  for $\boldsymbol\Lambda_B$ and $\boldsymbol\Lambda_W$ simultaneously.
* **Sorensen, O. (2024)** Multilevel semiparametric latent variable
  modeling in R with galamm. *Multivariate Behavioral Research*
  59(5):1098-1105. <https://doi.org/10.1080/00273171.2024.2385336> --
  the original `lambda` mechanism in galamm.
