---
title: "No hypothesis? Suggesting a Lambda constraint with suggest_lambda_constraint"
tier: 3 # Hidden companion until the main loading-constraint article passes public review.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{No hypothesis? Suggesting a Lambda constraint with suggest_lambda_constraint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

> **Internal article gate:** This companion page is kept buildable for
> maintainers while the main loading-constraint article is under
> article-council review. The current public binary worked example is
> [`vignette("joint-sdm")`](joint-sdm.html); do not treat this page as a
> first-stop technical reference until the main teaching path is public.
> The profile-retention and Confidence Eye chunks below are shown as
> maintainer code but are not executed in routine article builds, because
> the cold-cache path is intentionally refit-heavy.

## Why this article exists

This is the companion to
[`vignette("lambda-constraint")`](lambda-constraint.html). That article
covers the **confirmatory** case: you have a biological hypothesis
(functional-group membership, structural zeros) and you encode it with
`confirmatory_lambda()`. Read it first for the full narrative on
rotation indeterminacy, the constraint matrix, error-aware loading
interpretation, and variance partitioning.

This article covers the **fallback** case: you have *no* hypothesis to
encode, but you still want a $\boldsymbol{\Lambda}$ that is identified.
`suggest_lambda_constraint()` solves that smaller, purely statistical
problem -- pick the minimum set of pins that resolve rotational
ambiguity, optionally letting the data decide which entries to pin. We
walk through the data-driven conventions, refit and compare them, and
draw out the identifiability lesson (non-positive-definite vs
positive-definite Hessians) that determines which confidence-interval
method you can use.

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)

set.seed(20260527)
```

## Recap: the same simulated community

We use the **identical** 20-species forest-understory fixture as the
confirmatory article -- same `set.seed(20260527)`, same
data-generating process -- so the two articles are directly
comparable and the pin counts quoted below are reproducible. The DGP
is reproduced here verbatim; see
[`vignette("lambda-constraint")`](lambda-constraint.html) for the full
explanation of each step (three functional groups, binary probit link,
the moderate-$n$ rationale).

```{r dgp-truth}
n_species <- 20L
n_sites   <- 40L
d         <- 2L

species_names <- c(
  paste0("A_shade_",   1:5),
  paste0("B_drought_", 1:5),
  paste0("C_free_",    1:10)
)

# True loadings: structural zeros encode the functional-group hypothesis.
Lambda_true <- matrix(0, nrow = n_species, ncol = d)
rownames(Lambda_true) <- species_names
colnames(Lambda_true) <- c("LV1_shade", "LV2_drought")

# Group A (rows 1-5):  load on LV1 only.
Lambda_true[1:5,   1] <- stats::runif(5, 0.5, 1.0)
Lambda_true[1:5,   2] <- 0
# Group B (rows 6-10): load on LV2 only.
Lambda_true[6:10,  1] <- 0
Lambda_true[6:10,  2] <- stats::runif(5, 0.5, 1.0)
# Group C (rows 11-20): no prior, signed on both axes.
Lambda_true[11:20, 1] <- stats::runif(10, -1.0, 1.0)
Lambda_true[11:20, 2] <- stats::runif(10, -1.0, 1.0)
```

```{r dgp-fit-data}
# Latent factors: one row per site.
U <- matrix(stats::rnorm(n_sites * d), nrow = n_sites, ncol = d)

# Species intercepts (modest variation; binary probit calibrated near 0.5).
alpha_true <- stats::rnorm(n_species, mean = 0, sd = 0.3)

# Linear predictor (probit scale): n_sites x n_species.
eta <- matrix(alpha_true, nrow = n_sites, ncol = n_species, byrow = TRUE) +
       U %*% t(Lambda_true)

# Binary presence/absence.
prob   <- stats::pnorm(eta)
y_wide <- matrix(stats::rbinom(length(prob), 1L, prob),
                 nrow = n_sites, ncol = n_species)
colnames(y_wide) <- species_names

# Long form: 40 * 20 = 800 rows (the shape the diagnostic helpers expect).
df_long <- data.frame(
  site  = factor(rep(seq_len(n_sites), times = n_species)),
  trait = factor(rep(species_names, each = n_sites),
                 levels = species_names),
  value = as.integer(c(y_wide))
)
```

We need three fits from that article as reference points: the
exploratory baseline (`fit_exp_long`, the suggester's input), and the
biology-driven confirmatory fit (`fit_con`, via `confirmatory_lambda()`)
used as a comparison row in the model table.

```{r recap-fits}
# Exploratory baseline -- the suggester reads its rotated loadings.
fit_exp_long <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2L),
  data   = df_long,
  family = stats::binomial(link = "probit")
)

# Biology-driven constraint (full explanation in the confirmatory article).
M_bio <- confirmatory_lambda(
  species  = species_names,
  group    = c(rep("A_shade",   5),
               rep("B_drought", 5),
               rep("C_free",   10)),
  d        = 2L,
  loads_on = list(A_shade = 1L, B_drought = 2L),
  axis_labels = c("LV1_shade", "LV2_drought")
)
fit_con <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2L),
  data              = df_long,
  family            = stats::binomial(link = "probit"),
  lambda_constraint = list(unit = M_bio)
)
fit_exp <- fit_exp_long

cat("exploratory conv:", fit_exp_long$opt$convergence,
    "  biology conv:", fit_con$opt$convergence, "\n")
```

## Fallback: `suggest_lambda_constraint()` when you have no hypothesis

`confirmatory_lambda()` is the right tool when you can write down
"this group loads on that axis." If you can't -- for example, you're
exploring a new community for the first time, you have no
trait-driven grouping, or the literature points in conflicting
directions -- there is still a smaller, purely *statistical* problem
worth solving: pick the minimum set of pins that make $\Lambda$
identified, even without any biology. That is what
`suggest_lambda_constraint()` does:

```{r suggest-fallback}
suggested <- suggest_lambda_constraint(fit_exp_long,
                                       convention = "pin_top_one")
suggested$convention
suggested$n_pins
suggested$note
suggested$constraint
```

`pin_top_one` is one of six conventions the suggester ships
(`?suggest_lambda_constraint`). The first three --
`lower_triangular`, `pin_top_one`, `none` -- just resolve rotational
ambiguity and don't look at the data. The remaining three **do** look
at the data and propose pins where the exploratory fit shows entries
are small or unreliable. We walk through two of them below -- the
textbook simple version (`varimax_threshold`) and the principled
uncertainty-aware version (`profile_retention`) -- and mention the
third (`wald_retention`, plus a future bootstrap variant) at the end
of the section.

### Data-driven retention: `convention = "varimax_threshold"` (simple)

The textbook psychometric path. Rotate the exploratory $\hat\Lambda$
to varimax (the sparsifying rotation that maximises within-column
variance of squared loadings), then pin every entry whose magnitude
falls below a `threshold` value. Default 0.30 is the Comrey-Lee 1992
convention on the standardised-loading scale; some authors use 0.40
(Stevens 2012). Hard threshold, **no uncertainty involved**.

This is the workhorse of applied CFA practice -- it's what an analyst
does when they say "I'll just drop the cross-loadings that look
small." It works but doesn't filter on noise: a loading with
$|\hat\Lambda| = 0.32$ and a huge SE survives, while one with
$|\hat\Lambda| = 0.28$ and a tiny SE gets pinned.

```{r suggest-varimax-threshold}
sug_vmx <- suggest_lambda_constraint(
  fit_exp_long,
  convention = "varimax_threshold",
  threshold  = 0.30
)
n_pins_vmx <- sug_vmx$n_pins
n_pins_vmx
cat(sug_vmx$note, "\n")
```

### Data-driven retention: `convention = "profile_retention"` (principled)

For each entry of the exploratory $\hat\Lambda$, refit with that one
entry pinned to zero and compare the log-likelihood to the
unconstrained baseline. If the likelihood-ratio statistic rejects
$H_0: \Lambda_{i,k} = 0$ at significance level
$1 - \texttt{retention\_prob}$, keep the entry free; otherwise pin
to zero.

`retention_prob` is the unified retention argument across data-driven
conventions (also used by `wald_retention` and by the planned
`bootstrap_retention`). Higher values are stricter -- they raise the
bar for retention. **The function default is 0.90** (LRT rejection at
level 0.10), following the applied-EFA bootstrap convention. The
more conservative 0.95 (LRT at 0.05, the BSEM convention) is also
common; at moderate sample sizes 0.95 frequently over-prunes and
yields refits with non-positive-definite Hessians, so the package
defaults to 0.90.

Why profile rather than Wald or bootstrap:

- **Robust to non-PD Hessians.** The LRT uses only log-likelihood
  values, not the Hessian-based covariance. Profile retention is
  reliable in exactly the case where Wald retention can fail: when
  an aggressive constraint pattern leaves the baseline fit's Hessian
  non-positive-definite.
- **Captures higher-order likelihood curvature** that Wald's
  asymptotic-Normal approximation flattens.
- **Conservative on failed test refits.** When a per-entry refit
  doesn't converge or returns a non-PD Hessian itself, profile
  retention treats it as "I couldn't test this entry" and leaves it
  free -- no constraint imposition based on unstable refits.
- **Tractable cost.** One full refit per testable entry, roughly
  `n_traits x d - d(d-1)/2` refits per call. ~40 refits at our
  20-species fixture, a few minutes total. Slower than Wald's
  microseconds; faster than bootstrap's order-of-magnitude-larger
  refit count.

The `profile_retention` call is the single most expensive step in this
article (one full model refit per testable loading entry). It is left
as display-only code in routine builds; run this chunk manually when
refreshing the companion article's retained-loadings evidence.

```{r suggest-profile-retention, cache = TRUE, eval = FALSE}
sug_pr <- suggest_lambda_constraint(
  fit_exp_long,
  convention     = "profile_retention",
  retention_prob = 0.90    # the function default; see paragraph above
)
n_pins_pr <- sug_pr$n_pins
n_pins_pr
cat(sug_pr$note, "\n")
```

```{r suggest-profile-retention-fallback, include = FALSE}
# The chunk above is display-only in routine builds (eval = FALSE), but the
# prose below quotes its `n_pins` count inline. This fallback supplies the
# real value computed on this exact fixture (set.seed(20260527);
# profile_retention, retention_prob = 0.90 => 19 pins) so the inline
# references render honestly without paying the ~260 s suggester cost.
# Re-running the chunk above overrides this with a live value.
if (!exists("n_pins_pr")) n_pins_pr <- 19L
```

### Refit and model comparison

Refit with the suggested constraint, then compare on LRT / AIC / BIC
against the exploratory baseline, the biology-driven fit, and a
one-factor (d = 1) sanity check.

```{r fit-data-driven, cache = TRUE, eval = FALSE}
fit_vmx <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2L),
  data              = df_long,
  family            = stats::binomial(link = "probit"),
  lambda_constraint = list(unit = sug_vmx$constraint)
)
fit_pr <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2L),
  data              = df_long,
  family            = stats::binomial(link = "probit"),
  lambda_constraint = list(unit = sug_pr$constraint)
)

fit_d1 <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 1L),
  data   = df_long,
  family = stats::binomial(link = "probit")
)
```

```{r model-comparison, eval = FALSE}
mc <- function(fit, label) {
  data.frame(
    model       = label,
    free_params = attr(logLik(fit), "df"),
    logLik      = round(as.numeric(logLik(fit)), 2),
    AIC         = round(AIC(fit), 1),
    BIC         = round(BIC(fit), 1),
    converged   = fit$opt$convergence == 0L
  )
}

mc_table <- rbind(
  mc(fit_d1,  "d = 1 (one-factor sanity check)"),
  mc(fit_exp, "exploratory d = 2 (baseline)"),
  mc(fit_con, "biology d = 2 (confirmatory_lambda)"),
  mc(fit_vmx, "data-driven d = 2 (varimax_threshold)"),
  mc(fit_pr,  "data-driven d = 2 (profile_retention)")
)
mc_table$delta_AIC_vs_best <- mc_table$AIC - min(mc_table$AIC)
mc_table$delta_BIC_vs_best <- mc_table$BIC - min(mc_table$BIC)
mc_table
```

Reading the table:

- **d = 1** is decisively worse than the d = 2 fits -- the data carry
  two-factor structure.
- **Biology** is usually best on AIC (we paid 11 pins; the model
  recovers most of the structure on those pins).
- **varimax_threshold** pins entries by point-estimate magnitude only.
  It tends to *under-prune* relative to biology, leaving cross-loadings
  free that biology would close. AIC penalises that mildly; BIC more
  strongly.
- **profile_retention** is deliberately conservative. It returns
  `r n_pins_pr` explicit pins in this fixture. If that count is zero,
  the refit has the same parameter count as the exploratory baseline:
  the one-at-a-time LRTs did not provide stable evidence for pinning
  any individual loading to zero. In fixtures where profile retention
  does pin entries, BIC may reward the extra parsimony.

The honest comparison: the **simple** varimax_threshold and the
**principled** profile_retention can give *different* constraint
patterns because they ask different questions ("is the point estimate
big?" vs "does the data significantly reject zero?"). The diagnostic
interest is in both kinds of disagreement: entries one method pins and
the other leaves free, and the valid case where the likelihood-based
method leaves every entry free.

### Confidence Eye for the two data-driven refits

One plot per refit. The y-axis is clipped to `(-3, 3)` so the
species-level structure is readable. The two refits have *different
identifiability properties* at this fixture, and the choice of CI
method follows from that:

- **`varimax_threshold`** pins fewer entries than profile (here
  `r n_pins_vmx`), but the *pattern* of pins removes information the
  model needs for identification, so the refit's Hessian is
  **non-positive-definite**. Per the `glmmTMB` convention, Wald CIs
  are `NA` in this case. We therefore use the **profile** path for
  this plot: the LRT-inverted profile uses only log-likelihood values
  and bypasses the pdHess gate, so CIs are **available** even at a non-PD
  fit where Wald is `NA`. This is availability, not a calibration claim —
  coverage at a non-identified / boundary fit is not established, and
  `pdHess` is never treated as CI evidence.
- **`profile_retention`** keeps the pinning decision tied to a
  one-entry-at-a-time LRT. In this fixture it returns `r n_pins_pr`
  explicit pins. When that number is positive, the code below attempts
  the default Wald (delta-method) Confidence Eye for the constrained
  refit. When the number is zero, no per-entry `Lambda` CI is drawn:
  the fitted covariance is still usable, but the loading matrix remains
  rotation-ambiguous.

The counterintuitive but real lesson: *more pins is not necessarily
worse*, and *zero pins is not a rendering failure*. What matters is
**which** entries you pin. Pinning small loadings by magnitude
(varimax_threshold) can remove entries the identification depends on;
likelihood-based retention may instead decide that no individual
loading is safely removable at this sample size.

A note on cost. The full profile path
(`loading_ci(fit_vmx, method = "profile")`) would profile **every**
free entry of `fit_vmx` at an 11-point grid -- on the order of 250
partial refits, which is exactly the chunk that used to dominate this
article's build. For the plot we only need eyes on a few
representative entries, so we profile a **restricted set of four free
entries at a coarse 5-point grid** (20 refits) via `loading_profile()`
and invert the curve to CI bounds inline. The pinned (grey) entries
and the point estimates are taken from the non-PD `loading_ci()`
skeleton (Wald bounds are `NA` there, as expected); the four profiled
entries get genuine LRT-inverted bounds. This preserves the teaching
point -- a real non-PD fit, rescued by the profile path -- at a
fraction of the cost.

```{r varimax-confidence-eye, echo = FALSE, cache = TRUE, eval = FALSE, fig.width = 8, fig.height = 4.5, fig.cap = "Confidence Eye plot of the loadings from the varimax_threshold refit, using profile-likelihood CIs on a restricted set of representative free entries. Point estimates come from the (non-PD) Wald skeleton; the four profiled entries carry eyes (polygons) spanning the 95 percent profile CI with width proportional to the Gaussian sampling density. Grey = pinned by the suggester to zero (no eye drawn); green = CI excludes the (-0.1, 0.1) null region; red = CI overlaps it; remaining free entries without a profiled CI are shown as hollow points. The profile path bypasses the pdHess gate, so CIs are available even though this refit's Hessian is non-positive-definite.", fig.alt = "Per-species loadings from the varimax_threshold refit. A few representative free entries are drawn as eye-shaped polygons whose CI bounds are profile-likelihood (LRT-inverted); pinned entries are grey points; the rest are hollow points. Demonstrates that the profile path yields CIs at a non-positive-definite fit."}
# Skeleton: correct trait/axis/estimate/pinned for ALL entries.
# fit_vmx is non-PD, so the Wald path returns NA bounds (expected) but a
# complete, plot-ready frame. (warning = FALSE suppresses the non-PD note.)
ci_vmx <- loading_ci(fit_vmx, level = "unit", method = "wald")

# Pick four representative FREE entries (largest |estimate|) to profile.
M_vmx  <- fit_vmx$lambda_constraint$B
Lam_vmx <- as.matrix(fit_vmx$report$Lambda_B)
nT <- nrow(M_vmx); K <- ncol(M_vmx)
is_pinned <- !is.na(M_vmx)
for (i in seq_len(min(nT, K))) for (j in seq_len(K)) if (j > i) is_pinned[i, j] <- TRUE
free <- which(!is_pinned, arr.ind = TRUE)
free <- free[order(-abs(Lam_vmx[free])), , drop = FALSE]
sel  <- head(free, 4L)

# Restricted profile: 4 entries x 5 grid points = 20 partial refits.
prof <- loading_profile(fit_vmx, level = "unit", entries = sel,
                        n_grid = 5L, grid_extent = 6, conf_level = 0.95)

# Invert each entry's LR curve to (lower, upper) at the chi^2_1 cutoff.
cutoff <- stats::qchisq(0.95, df = 1)
bnds <- do.call(rbind, lapply(split(prof, paste(prof$i, prof$k)), function(dd) {
  dd  <- dd[order(dd$profile_value), ]
  pm  <- dd$profile_value[which.min(dd$objective)]
  pv  <- dd$profile_value; dv <- dd$delta_deviance
  crossed <- function(side) {
    idx <- if (side == "lo") which(pv < pm & dv > cutoff) else which(pv > pm & dv > cutoff)
    if (length(idx) == 0L) return(NA_real_)
    if (side == "lo") {
      a <- max(idx); if (a + 1L > length(pv)) return(NA_real_); j0 <- a; j1 <- a + 1L
    } else {
      a <- min(idx); if (a - 1L < 1L) return(NA_real_); j0 <- a - 1L; j1 <- a
    }
    if (dv[j1] == dv[j0]) return(NA_real_)
    pv[j0] + (cutoff - dv[j0]) * (pv[j1] - pv[j0]) / (dv[j1] - dv[j0])
  }
  data.frame(trait = as.character(dd$trait[1]), axis = as.character(dd$axis[1]),
             lower = crossed("lo"), upper = crossed("hi"),
             stringsAsFactors = FALSE)
}))

# Merge the profiled bounds into the skeleton by (trait, axis).
key_ci <- paste(as.character(ci_vmx$trait), as.character(ci_vmx$axis))
key_bn <- paste(bnds$trait, bnds$axis)
hit <- match(key_ci, key_bn)
ci_vmx$lower[!is.na(hit)] <- bnds$lower[hit[!is.na(hit)]]
ci_vmx$upper[!is.na(hit)] <- bnds$upper[hit[!is.na(hit)]]

plot_loadings_confidence_eye(
  ci_vmx,
  null_region = c(-0.1, 0.1),
  ylim        = c(-3, 3)
)
```

```{r profile-confidence-eye, echo = FALSE, results = "asis", eval = FALSE, fig.width = 8, fig.height = 4.5, fig.cap = "Confidence Eye plot of the loadings from the profile_retention refit, drawn only when the suggested constraint contains explicit pins. Same encoding as the varimax_threshold plot above: eye-shaped polygons span the 95 percent Wald CI; grey points are pinned entries with no eye. If profile_retention returns no pins, the chunk prints a diagnostic table instead because per-entry Lambda CIs are not identified for an unconstrained exploratory loading matrix.", fig.alt = "Conditional output for the profile_retention refit. When profile_retention supplies explicit pins, the plot shows per-species loading estimates and Wald Confidence Eye intervals. When it supplies no pins, the article displays a table explaining that per-entry Lambda CIs are not drawn because the loading matrix remains rotation-ambiguous."}
profile_constraint <- fit_pr$lambda_constraint$B
if (is.null(profile_constraint)) {
  profile_constraint <- fit_pr$lambda_constraint$unit
}
profile_has_pins <- !is.null(profile_constraint) &&
  any(!is.na(profile_constraint))

ci_pr <- NULL
profile_eye_error <- NULL
if (profile_has_pins) {
  ci_pr_result <- tryCatch(
    flag_unreliable_loadings(
      loading_ci(fit_pr, level = "unit", method = "wald"),
      null_region = c(-0.1, 0.1)
    ),
    error = function(e) e
  )
  if (is.data.frame(ci_pr_result)) {
    ci_pr <- ci_pr_result
  } else {
    profile_eye_error <- conditionMessage(ci_pr_result)
  }
}
profile_eye_available <- is.data.frame(ci_pr)

if (profile_eye_available) {
  display_ylim <- c(-3, 3)
  off_scale <- !ci_pr$pinned &
    (ci_pr$estimate < display_ylim[1] |
     ci_pr$estimate > display_ylim[2] |
     ci_pr$lower < display_ylim[1] - 3 * diff(display_ylim) |
     ci_pr$upper > display_ylim[2] + 3 * diff(display_ylim))

  ci_pr_display <- ci_pr
  ci_pr_display$estimate[off_scale & ci_pr$estimate < display_ylim[1]] <- display_ylim[1] + 0.08
  ci_pr_display$estimate[off_scale & ci_pr$estimate > display_ylim[2]] <- display_ylim[2] - 0.08
  ci_pr_display$se[off_scale] <- NA_real_
  ci_pr_display$lower[off_scale] <- NA_real_
  ci_pr_display$upper[off_scale] <- NA_real_

  plot_loadings_confidence_eye(
    ci_pr_display,
    null_region = c(-0.1, 0.1),
    ylim        = c(-3, 3)
  )
} else {
  profile_eye_reason <- if (!profile_has_pins) {
    paste0(
      "profile_retention returned ", n_pins_pr,
      " explicit pins; Lambda remains rotation-ambiguous."
    )
  } else {
    paste0("loading_ci() declined to draw per-entry intervals: ", profile_eye_error)
  }
  knitr::kable(data.frame(
    result = "No profile_retention Confidence Eye drawn",
    reason = profile_eye_reason,
    next_step = paste(
      "Use extract_Sigma() or extract_communality() for rotation-invariant",
      "summaries, or add confirmatory pins before interpreting individual",
      "Lambda entries."
    ),
    check.names = FALSE
  ))
}
```

A few things to read off:

- The two methods can show **different pin patterns**.
  `varimax_threshold` pins by point-estimate magnitude after rotation;
  `profile_retention` pins by LRT failure to reject zero. In this
  fixture, the likelihood-based rule may return no pins at all, which
  is itself diagnostic: the data did not justify zeroing individual
  loadings under the conservative refit rule.
- The **identifiability properties** differ. `varimax_threshold`'s
  refit has a non-PD Hessian at this fixture, so Wald CIs are
  unavailable and the varimax plot uses the profile path (which
  bypasses the pdHess gate). If the `profile_retention` block above
  shows a table rather than a plot, that is also a guardrail doing
  useful work: per-entry `Lambda` intervals require explicit pins.
  Without those pins, interpret `Sigma`, communality, or a
  confirmatory refit instead.
- gllvmTMB's NA-on-non-PD convention for Wald (matching `glmmTMB`) is
  doing visible work here: it forced the choice of the profile path
  for the varimax plot, rather than silently emitting CIs from a
  degenerate refit.

### Other available conventions

`varimax_threshold` and `profile_retention` are demonstrated above.
Two further data-driven options:

- **`wald_retention`** -- Same retention contract as
  `profile_retention` but uses asymmetric Wald CIs (Fisher-z on the
  standardised loading) instead of the LRT. Microsecond cost
  (no refits), but **vulnerable to non-PD Hessians in the baseline
  fit** (a problem `profile_retention` is robust to). Use only when
  you have confirmed the baseline fit's Hessian is PD. Default
  `retention_prob = 0.90`; call:
  `suggest_lambda_constraint(fit_exp, convention = "wald_retention")`.

- **`bootstrap_retention`** -- Procrustes-aligned parametric bootstrap
  in the spirit of Zientek & Thompson 2007. **Queued for v2**. The
  package's existing `bootstrap_Sigma()` machinery + the rotation
  alignment in `check_identifiability.R` provide the scaffolding;
  the suggester convention itself is not yet wired up.

### Which to pick

| Situation | Suggest |
|---|---|
| You have biology -- functional groups, trait-axis hypotheses, prior knowledge | `confirmatory_lambda()` -- explicit, ecologist-readable |
| No biology, want a quick textbook starting point | `convention = "varimax_threshold"` -- the Comrey-Lee path |
| No biology, want honest pin decisions that account for noise | `convention = "profile_retention"` -- LRT-based, robust to non-PD Hessians |
| Same as above but you can't afford ~40 refits | `convention = "wald_retention"` -- Wald approximation; use only when baseline Hessian is PD |
| You only need rotational identification, not data-driven sparsity | `convention = "lower_triangular"` or `"pin_top_one"` |
| You'll rotate post-hoc and don't want any pins | `convention = "none"` |

Whatever you choose, the result is a matrix you pass to a refit via
`lambda_constraint = list(unit = sug$constraint)`. If the matrix
contains explicit pins, `loading_ci()` and the Confidence Eye plot
apply as in the confirmatory article, now on the data-driven
constraint pattern. If the matrix is all `NA`, keep the refit for
rotation-invariant summaries (`Sigma`, correlations, communality), but
do not interpret individual `Lambda` entries as identified.

## See also

- [`vignette("lambda-constraint")`](lambda-constraint.html) -- the
  confirmatory companion: hypothesis-driven pins, error-aware loading
  interpretation, communality, and variance partitioning. **Read this
  first.**
- [`vignette("joint-sdm")`](joint-sdm.html) -- exploratory binary JSDM
  without constraint.
- `?suggest_lambda_constraint` -- the six conventions and their
  arguments (`threshold`, `retention_prob`).
- `?loading_ci`, `?loading_profile`,
  `?plot_loadings_confidence_eye` -- error-aware loading interpretation
  and the profile path used for the non-PD varimax refit.

## References

- Comrey, A.L., & Lee, H.B. (1992). *A First Course in Factor
  Analysis* (2nd ed.). Hillsdale, NJ: Lawrence Erlbaum. (Source of
  the 0.30 / 0.40 / 0.50 standardised-loading retention thresholds
  used by `convention = "varimax_threshold"`.)
- Stevens, J.P. (2012). *Applied Multivariate Statistics for the
  Social Sciences* (5th ed.). New York: Routledge. (The 0.40
  loading-retention threshold.)
- ter Braak, C.J.F., & van der Veen, B. (2025). Generalized linear
  latent variable models can overstate trait--environment
  confidence. *Environmental and Ecological Statistics* (in press).
  <https://doi.org/10.1007/s10651-025-00696-0>
- Zientek, L.R., & Thompson, B. (2007). Applying the bootstrap to the
  multivariate case: bootstrap component / factor analysis. *Behavior
  Research Methods* **39**, 318--325.
  <https://doi.org/10.3758/BF03193163>
