---
title: "Handling missing data"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Handling missing data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# SKELETON STAGE: all chunks are eval=FALSE so the article renders instantly.
# As each implementation phase ships, chunks are re-enabled one section at a
# time and driven by tiny toy fixtures (n <= 40, no latent variables) to keep
# pkgdown build time under 30 s.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

Missing data in `gllvmTMB` has two meanings.
A **response cell** can be missing: an individual was measured but one trait
was not recorded, leaving a gap in the multivariate trait matrix.
A **predictor** can be missing: a covariate needed by the response model was
not observed for some rows.
The current implementation treats those two cases differently, and both differ
from the third case--measurement error--which is explicitly out of scope for
this layer (see Section 8).

The starting point is what *not* to do.
Listwise deletion discards every row with any gap, which in a multivariate
trait matrix can mean throwing away most of the data and biasing estimates
whenever missingness is not completely at random.
Mean-fill and single imputation invent values and then treat them as if they
were observed, understating uncertainty and distorting the covariance
structure that an ordination or a phylogenetic model is trying to estimate.
For structured ecological data--shared species means, phylogenetic covariance,
spatial fields, latent axes--these ad-hoc fixes are particularly damaging
because they sever exactly the correlations the model exists to exploit.
`gllvmTMB` instead keeps the missingness *inside* the likelihood: nothing is
deleted and nothing is invented, and the uncertainty that missingness creates
propagates through the same joint Hessian as every other parameter.

For missing response cells, `gllvmTMB` can retain the original rows and
evaluate the observed-cell likelihood by masking the unobserved entries.
Missing cells contribute no direct response likelihood, but the fitted object
still knows where those cells were.
For missing predictors, `gllvmTMB` requires an explicit predictor model
declared through `mi()` in the response formula and `impute = list(...)` in
the fit call.
A missing predictor is a latent random effect integrated out by TMB's Laplace
approximation; this is frequentist FIML, not multiple imputation.

This is likelihood-based missing-data handling.
It is not Bayesian posterior imputation, not multiple imputation, and not an
EM interface.
For cases where those approaches are preferred--rich GNN phylogenetic
imputation, mixed trait types, tree uncertainty--the `pigauto` package
provides the complementary preprocessing path (Section 6).

### Status: what is shipped versus planned

This article documents the missing-data layer ahead of its implementation, so
the reader can see the intended design as a whole.
Read the status tags literally:

- **SHIPPED today.** Only the conservative default: rows with a missing
  *response* cell are dropped before the likelihood is built, and a missing
  *predictor* is an error. This is ordinary complete-case fitting.
- **PLANNED (Phase 1, the response mask).** `miss_control(response =
  "include")`, the `is_y_observed` mask, original-row accounting in
  `fit$missing_data`, and `predict_missing()` for reconstructed response cells.
  This is the design's "easy win" but is **not yet in the package**.
- **PLANNED (Phase 2 onward, missing predictors).** The `mi()` formula token,
  the `impute = list(...)` covariate-model list, `impute_model(family = )`,
  `imputed()`, and the continuous/discrete latent-predictor machinery.

Every code chunk below is therefore illustrative (`eval = FALSE`): it shows the
intended syntax, not output from a current fit. Each section labels the phase
its API belongs to.

## 1. What is fitted now

The default--and, today, the only behaviour--is conservative complete-case
fitting: a missing response cell drops its row, and a missing predictor is an
error. The arguments below describe the planned routes out of that default.

```{r what-is-fitted-default}
# FILL: Phase 1 -- run miss_control() and show its defaults once the API lands
# miss_control()
```

Once shipped, `missing = miss_control(response = "include")` will retain rows
that have at least one observed response cell, and `missing =
miss_control(predictor = "model")` will be used only when the response formula
contains at least one supported `mi()` term and `impute` supplies the
corresponding predictor model.

The table below maps each missing-data case to its intended route and the
boundary that route stops at. The phase tag in the final column says when that
route is expected to land.

| Missing-data case | Intended route | Boundary / phase |
| --- | --- | --- |
| Missing trait cell (any family) | `missing = miss_control(response = "include")` | All predictors, grouping variables, offsets, weights, and structured-effect inputs must be complete for that row (Phase 1). |
| One missing continuous predictor, fixed covariate model | `traits(y1 + y2 ~ mi(body_mass) + z)` with `impute = list(body_mass = body_mass ~ z)` | The `mi()` variable must be numeric; one Gaussian covariate model; no multiple missing predictors yet (Phase 2a). |
| One missing continuous predictor, grouped covariate model | As above with `impute = list(body_mass = body_mass ~ z + (1 | site))` | One grouped random intercept in the predictor model; independent of response random effects (Phase 2b). |
| Species-level missing trait (broadcast to obs) | `impute = list(body_mass = body_mass ~ 1 + phylo(1 | species, tree = tree))` | Species-level value broadcast to observation rows; phylogenetic covariate model (Phase 3). |
| Non-Gaussian / categorical missing predictor | `impute_model(family = )` with a non-Gaussian family | Lognormal, Gamma, beta, Poisson, ordinal, factor -- later work (Phase 5). |

The `gllvmTMB` response matrix is **multivariate**: a single fit can have many
trait columns and many species.
Missing cells are indexed at the (row, trait) level, not at the response-row
level as in a univariate model.

## 2. Missing response cells in `traits()`

For a multivariate response model, each (row, trait) cell is either observed or
missing.
The observed-cell likelihood for row `i` and trait `j` is:

```
is_y_observed[i, j] = 1  -->  full log p(y_{ij} | eta_{ij})
is_y_observed[i, j] = 0  -->  0  (cell masked; row kept for random-effect
                                   borrowing across traits and species)
```

A row with at least one observed trait cell is retained.
A row where every trait is missing is also retained for row accounting but
contributes zero direct response likelihood.

This example creates a tiny trait matrix with deliberate missing cells:

```{r missing-response-example}
# FILL: Phase 1 -- enable with tiny toy fixture once miss_control(response="include")
# lands (no model fit needed here -- show data setup + formula syntax only).
#
# set.seed(1)
# n <- 20; p <- 3
# dat <- data.frame(
#   species = rep(paste0("sp", 1:n), 1),
#   x       = rnorm(n)
# )
# Y <- matrix(rnorm(n * p), n, p,
#             dimnames = list(NULL, c("trait1", "trait2", "trait3")))
# Y[3, 1] <- NA; Y[7, 2] <- NA; Y[12, 1] <- NA; Y[12, 3] <- NA  # deliberate gaps
#
# fit_resp <- gllvmTMB(
#   traits(trait1 + trait2 + trait3 ~ x),
#   family  = gaussian(),
#   data    = cbind(dat, Y),
#   missing = miss_control(response = "include")
# )
```

Cross-trait borrowing still operates on the retained rows because the latent
structure is shared across all cells in a row.
A row where trait 1 is missing but traits 2 and 3 are observed still
contributes to the species means and latent axes for all traits.

The analogy with `drmTMB` partial bivariate pairs (where one of `y1`/`y2` can
be observed) generalises here: an observed cell informs its own trait
mean/loadings, but cross-trait species correlation is identified only from
rows where both cells are co-observed.

## 3. Inspecting `fit$missing_data` and `nobs()`

Every fitted object that used `miss_control(response = "include")` or
`miss_control(predictor = "model")` carries a `fit$missing_data` slot with
row-accounting and diagnostics.

```{r inspect-metadata}
# FILL: Phase 1 -- enable once fit_resp above works.
#
# nobs(fit_resp)                     # likelihood-contributing cells
# fit_resp$missing_data$counts       # retained rows / observed cells / missing cells
# fit_resp$missing_data$version      # implementation slice, e.g. "MD1"
# head(fit_resp$missing_data$original_row)
# fit_resp$missing_data$observed_y   # n x p logical mask
```

The most useful fields are:

| Field | Meaning |
| --- | --- |
| `version` | Implementation slice tag, e.g. `"MD1"`, `"MD2"`, `"MD3a"`. |
| `response_policy` | Whether response rows/cells were dropped or masked. |
| `predictor_policy` | Whether missing predictors fail or are modelled. |
| `original_row` | Row indices in the user's original data frame. |
| `model_row` | Row indices in the retained model frame. |
| `observed_y` | Logical matrix mask (n x p) for multivariate fits. |
| `counts` | Named integer vector: retained rows, observed cells, missing cells, likelihood-contributing cells. |
| `predictors` | Metadata and fitted values for explicitly modelled `mi()` predictors. |

For ordinary user work, prefer `nobs()`, `fitted()`, `residuals()`, and
`predict_missing()` over hand-parsing the slot.
The slot exists so row accounting stays auditable after fitting.

A co-observed cell count diagnostic is planned for the multivariate Gaussian
case: if very few rows have both traits observed, the cross-trait residual
correlation is weakly identified and `gllvmTMB` will warn.

## 4. `predict_missing()` for response cells

Missing response cells have predicted values from the fitted model because
their predictors and random effects are known (or estimated).
Use `predict_missing()` to retrieve those reconstructed values alongside their
standard errors.

```{r predict-missing}
# FILL: Phase 1 -- enable once predict_missing() is exported.
#
# pm <- predict_missing(fit_resp)
# head(pm)
# # Columns: original_row, trait, fitted, std_error
# # std_error comes from the joint Hessian (sdreport block), not a
# # posterior interval and not a multiple-imputation draw.
```

`predict_missing()` is for **response** cells only.
It is distinct from `imputed()`, which is reserved for missing **predictors**
(Section 5).
The terminology follows the shared contract with `drmTMB`: predicted/
reconstructed responses vs EBLUP conditional modes for latent predictors.

Residuals for missing cells are `NA`; fitted values are not.
This is deliberate: a residual requires an observed value to compute, but a
fitted value does not.

## 5. Missing predictors: `mi()` and `impute_model(family=)`

Missing predictors are not handled by row deletion or sentinel imputation.
The response model needs the predictor value, so `gllvmTMB` asks for an
explicit model for that predictor.
The missing values become latent random effects integrated out by the Laplace
approximation, jointly with the response-model random effects.

The joint likelihood factorises as:

```
L(theta) = integral
  p(y_obs | x_obs, x_mis, b, theta_y)
  * p(x_obs, x_mis | b_x, theta_x)
  * p(b, b_x | theta_b)
  d x_mis db db_x
```

In `gllvmTMB`, `x_mis` is declared in the TMB `random` set and integrated out
by the Laplace approximation together with all other random effects.
`sdreport` then returns the conditional mode (EBLUP) and its standard error
from the inverse joint Hessian.

How the integral is taken depends on the type of the missing predictor, and
the two cases are handled differently:

- **Continuous predictors (Laplace).** A missing continuous covariate is a
  real-valued latent quantity. The Laplace approximation replaces its integral
  with the joint mode plus a Gaussian curvature term, exactly as for any other
  random effect. The reported value is its conditional mode (EBLUP). This is
  the Phase 2-4 path and the one all the worked sketches below use.
- **Discrete predictors (exact finite-state summation).** A missing factor or
  ordinal covariate has finitely many possible values, so the "integral" is a
  finite sum: the likelihood is evaluated at each admissible level and weighted
  by the covariate model's category probabilities. No Gaussian approximation is
  used for the discrete dimension. This is planned for Phase 5 and is the
  reason categorical predictors are scheduled last.

### 5a. Continuous predictor, fixed covariate model (Phase 2a)

```{r missing-predictor-fixed}
# FILL: Phase 2a -- enable once mi() + impute list + x_mis latent are wired.
# Tiny fixture: n = 24, p = 2 traits, 4 missing body_mass values.
#
# set.seed(2)
# n <- 24
# dat2 <- data.frame(
#   moisture   = rnorm(n),
#   body_mass  = rnorm(n),
#   species    = paste0("sp", 1:n)
# )
# Y2 <- cbind(
#   trait1 = 0.5 + 0.8 * dat2$body_mass + rnorm(n, sd = 0.2),
#   trait2 = -0.2 - 0.4 * dat2$moisture + rnorm(n, sd = 0.25)
# )
# dat2$body_mass[c(3, 11, 18)] <- NA  # three missing predictor values
#
# fit_pred <- gllvmTMB(
#   traits(trait1 + trait2 ~ mi(body_mass) + moisture),
#   family  = gaussian(),
#   data    = cbind(dat2, Y2),
#   impute  = list(body_mass = body_mass ~ moisture),
#   missing = miss_control(predictor = "model")
# )
#
# imputed(fit_pred)   # EBLUP conditional modes + SEs for x_mis
```

Read `imputed(fit_pred)` as a fitted missing-predictor summary.
The `estimate` column is the conditional mode (EBLUP) used in the fitted
likelihood.
The `std_error` column is the likelihood-based conditional standard error from
the joint Hessian.
It is not a posterior interval and is not a set of multiple-imputation draws.

### 5b. Predictor with grouped covariate model (Phase 2b)

The same `mi()` route can include one grouped random intercept in the predictor
model when missingness is tied to a grouping level:

```{r missing-predictor-grouped}
# FILL: Phase 2b
#
# fit_grp <- gllvmTMB(
#   traits(trait1 + trait2 ~ mi(body_mass) + moisture),
#   family  = gaussian(),
#   data    = cbind(dat2, Y2),
#   impute  = list(body_mass = body_mass ~ moisture + (1 | site)),
#   missing = miss_control(predictor = "model")
# )
```

The grouped effect is a covariate-model random intercept, not a response-model
random effect.
It is kept independent of the response random effects by default (Level-1
independent default; Level-2 joint field is opt-in via
`correlate_with = "response"`, deferred to Phase 4).

### 5c. Species-level predictor with phylogenetic covariate model (Phase 3)

The flagship case for ecology: a missing species-level trait used as a
predictor, with its imputation model inheriting phylogenetic structure:

```{r missing-predictor-phylo}
# FILL: Phase 3 -- enable once phylo covariate model is wired.
# Tiny fixture: n_species = 12, a minimal Brownian-motion tree, 3 missing
# body_mass values at species level broadcast to observation rows.
#
# fit_phy <- gllvmTMB(
#   traits(trait1 + trait2 ~ mi(body_mass) + moisture),
#   family  = gaussian(),
#   data    = cbind(dat2, Y2),
#   impute  = list(
#     body_mass = impute_model(
#       body_mass ~ 1 + phylo(1 | species, tree = tiny_tree),
#       family = gaussian()
#     )
#   ),
#   missing = miss_control(predictor = "model")
# )
#
# imputed(fit_phy)
```

Phylogenetic imputation adds noise when phylogenetic signal is weak
(Penone et al. 2014; Molina-Venegas 2024).
A phylo-signal gate is planned as a Phase 3 verification check; it will warn
when the estimated signal is too weak to trust the imputed values.
Trait databases are typically MNAR (missing not at random): rare species with
few measurements are systematically under-represented.
The current implementation does not correct for MNAR; a delta-adjustment
sensitivity analysis is a Phase 3+ gate, not a v1 argument.

### 5d. Multi-source and spatial covariate models (Phase 4)

Phase 4 adds spatial (`spatial()`), relatedness-matrix (`relmat()`), and
animal-model (`animal()`) covariate structures, and enables the Level-2 joint
field where the predictor model explicitly shares a structured field with the
response model (with eigenvector-fixed-effect orthogonalisation to guard
against structured confounding; Dupont, Marques and Kneib 2023).

```{r missing-predictor-phase4}
# FILL: Phase 4 -- spatial, relmat, animal, multi-source, Level-2 joint field.
```

### 5e. Non-Gaussian and categorical predictors (Phase 5)

```{r missing-predictor-nongaussian}
# FILL: Phase 5 -- lognormal/Gamma, beta/logit-normal, Poisson/NB,
# ordinal cumulative-logit, unordered categorical (factor).
# Mirrors drmTMB MD5 non-Gaussian extension.
```

## 6. FIML versus multiple imputation

The missing-data layer described above is a frequentist **full-information
maximum-likelihood (FIML)** approach: one joint model, fitted once, with the
missingness handled inside the likelihood. It is deliberately an
**alternative** to multiple imputation (MI), not a wrapper around it. The two
strategies answer the same question by different routes.

Multiple imputation runs in three steps. An imputation model **draws** several
plausible completed datasets; the **analysis** model is fitted to each; and the
results are **pooled** by Rubin's rules, which combine the within- and
between-imputation variance into a single standard error. The completed
datasets are an explicit ensemble, and the between-imputation spread is how MI
represents the uncertainty that missingness adds.

FIML never forms that ensemble. The missing values are estimated jointly with
every other parameter in a single fit: missing responses are masked and
reconstructed afterward, and missing predictors are latent quantities
integrated out by the Laplace approximation. The uncertainty that MI recovers
from between-imputation spread is recovered here from the curvature of the same
joint likelihood--the inverse Hessian that `sdreport` already returns. Under a
correctly specified model and data missing at random, FIML and proper MI are
asymptotically equivalent (Schafer and Graham 2002; Schenker and Taylor 1996);
FIML simply reaches the answer without a completed-dataset ensemble and without
a pooling step.

| | Multiple imputation (incl. `pigauto`) | FIML (this layer) |
| --- | --- | --- |
| Workflow | draw -> analyse -> pool (Rubin's rules) | one joint model, fitted once |
| Completed datasets | an explicit ensemble (e.g. `m = 50`) | none; nothing is filled in |
| Uncertainty | within- + between-imputation variance | inverse joint Hessian (`sdreport`) |
| Missing predictors | requires covariates complete before imputing | latent, integrated out inside the fit |
| Imputation model | separate from the analysis model | *is* the analysis model's structure |

### Congeniality is automatic here

A recurring difficulty in MI is **congeniality** (Meng 1994): the imputation
model and the analysis model are specified separately, and if the imputation
model omits structure the analysis model relies on--an interaction, a
nonlinearity, a phylogenetic covariance--the pooled inference is biased even
when each step is internally correct. In the FIML layer this mismatch cannot
arise by construction: there is only one model. The "imputation model" for a
missing predictor is the covariate model declared in `impute`, and it lives in
the same likelihood as the response model, sharing its parameters and its
structured fields. The imputation and analysis models are congenial because
they are the same model.

This is also why the covariate model is **level-aware** rather than a generic
default. A missing species-level trait is imputed with a phylogenetic
covariate model (`phylo()`); a missing site-level covariate is imputed with a
spatial one (`spatial()`); an animal-model or relatedness structure
(`animal()`, `relmat()`) or the shared latent axes can play the same role. The
imputation model inherits the biological covariance of the analysis it serves--
but it is declared explicitly per predictor, not silently copied from the
response model, so the inheritance is visible and auditable.

### The `pigauto` sister path

Multiple imputation remains available, and for some problems it is the better
tool. `pigauto`
([itchyshin.github.io/pigauto](https://itchyshin.github.io/pigauto/index.html))
is the standalone MI path: a phylogenetic trait-imputation engine--a
Brownian-motion baseline blended with a graph-neural-network component--that
produces completed datasets and propagates uncertainty by multiple imputation
and Rubin's rules (`impute()` -> `multi_impute()` -> `with_imputations()` ->
`pool_mi()`), with conformal intervals and tree-uncertainty support. It imputes
missing **traits** but requires the covariates to be fully observed, and it is
agnostic to the downstream model.

The two packages are complementary rather than competing:

- Prefer **`pigauto` (MI)** when you want rich GNN-based phylogenetic
  imputation, a mix of trait types in one imputation step, explicit
  tree-uncertainty propagation, or a completed dataset you can hand to any
  downstream model.
- Prefer **`gllvmTMB` (FIML)** when you want a single transparent joint model
  with no preprocessing step, when the missingness is in **predictors** (which
  `pigauto` does not handle), or when you want the missing-data uncertainty to
  propagate through the very same likelihood as the rest of the fit.

The interfaces are kept separate on purpose: there is no `engine = "mi"` inside
`miss_control()`, because `mi()` means a latent missing predictor *inside* the
model while MI means multiple imputation *outside* it. A documented
`pigauto`-then-`gllvmTMB` workflow (impute with `pigauto`, fit `gllvmTMB` on
each completed dataset, pool) is the bridge when both are wanted.

## 7. Current boundaries

The design deliberately keeps several features out of scope for the planned
Phase 1 through Phase 3, independent of which phases have shipped:

- EM, profile, and REML missing-data engines are not implemented.
  `engine = "laplace"` is the only v1 option; `"em"` and `"profile"` are
  reserved names that are not yet accepted.
- Multiple missing predictors (`mi(x1)` and `mi(x2)` in the same formula) are
  not implemented.
- Non-Gaussian, bounded, or categorical missing predictors (lognormal, Gamma,
  beta, Poisson, NB, ordinal, factor) are Phase 5+ work.
- Structured covariate models for non-Gaussian predictors (grouped or phylo
  binary/ordinal) are not implemented.
- Interactions inside `mi()`, splines inside `mi()`, and predictor-model
  random slopes are not implemented.
- `imputed()` does not summarize missing **responses**; use `predict_missing()`
  for that.
- `simulate_imputed()` is deferred to Phase 5: if shipped, it will mean
  Gaussian-approximation or parametric-bootstrap draws, not posterior draws.
- Simulated imputations, credible intervals, and multiple-imputation pooling
  are not implemented in this layer; see `pigauto` for the MI sister path.
- Dense known sampling-covariance matrices with partial multivariate response
  rows are not implemented.
- MNAR sensitivity, bootstrap-SE cross-checks, and phylo-signal reliability
  diagnostics are Phase 3+ verification gates, not v1 API arguments.

When you encounter one of these boundaries, treat the error as a design limit
rather than a failed fit.
For now, use the supported likelihood-based route, fall back to complete-case
fitting, or handle the unsupported missingness outside `gllvmTMB` with a
method whose assumptions you can state explicitly.

## 8. Non-goal: measurement error

Missing predictors in this layer treat the **observed** values of `x` as
exact.
Only `x_mis` (the unobserved values) is latent.

**Measurement error** -- where the true `x` is latent even when nominally
observed, requiring an observation model `x_obs | x_true` with known standard
errors, replicates, or a validation subsample -- is **out of scope** for
Phase 1 through Phase 3.
The missing-predictor registry and `fit$missing_data` slot are designed so a
later measurement-error lane can reuse the latent-covariate machinery (a
possible future `me(x, se = se_x)` argument); but v1 exposes only `mi(x)` for
missing predictors and the two interfaces will remain distinct.

## References

Core references for the statistical foundations:

- Rubin (1976) *Biometrika* 63:581--592. MAR validity.
- Schafer & Graham (2002) *Psychological Methods* 7:147--177. FIML as
  alternative to MI.
- Schenker & Taylor (1996) *Computational Statistics & Data Analysis*
  22:425--446. Asymptotic equivalence of FIML and proper MI.
- Meng (1994) *Statistical Science* 9:538--558. Congeniality of the imputation
  and analysis models.
- Ibrahim, Chen, Lipsitz & Herring (2005) *JASA* 100:332--346. Missing
  covariates in GLMs.
- Kristensen et al. (2016) *JSS* 70(5). TMB Laplace + automatic
  differentiation.
- Robinson (1991) *Statistical Science* 6:15--32. EBLUP and best prediction.
- Penone et al. (2014) *Methods Ecol. Evol.* 5:961--970. Phylogenetic
  imputation adds noise when signal is weak.
- Molina-Venegas (2024) *Methods Ecol. Evol.* Limits of phylogenetic
  imputation.
- Dupont, Marques & Kneib (2023) arXiv:2309.16861. Structured confounding in
  spatial models.
- Wang, Edge, Schraiber & Pennell (2025 preprint, PMC -- not peer-reviewed).
  Phylogenetic analogue of structured confounding.
