---
title: "Handling missing data"
description: "Choose among omitted, retained, or explicitly modelled missing values and verify how each choice changes the estimand and fitted rows."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Handling missing data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

Missing responses and missing predictors require different models. This guide
shows the two current workflows and the two extractors that go with them.

**A response cell is missing**, such as one unmeasured trait. Fit with
`miss_control(response = "include")`, then inspect the masked cells with
`predict_missing()`.

**One unit-level predictor is missing.** Fit with `mi(x)`, an `impute` model,
and `miss_control(predictor = "model")`, then inspect it with `imputed()`.

The examples below exercise Gaussian routes that have direct fit-equivalence
and recovery tests. Binary,
ordered, and unordered missing-predictor models are also implemented, but they
have different output contracts and are summarized rather than expanded into
more long examples here.

## Before fitting: state the assumption

These workflows do not make missingness harmless. Likelihood-based inference
requires an ignorable missingness process: after conditioning on variables in
the model, the probability that a value is missing must not depend on its
unobserved value. The response and predictor models must also be adequate.
`gllvmTMB` does not model a missing-not-at-random process.

By default, `miss_control()` drops rows with missing responses and rejects
missing predictors. The likelihood-based routes in this article are explicit
choices:

```{r controls}
library(gllvmTMB)

miss_control()
miss_control(response = "include")
miss_control(predictor = "model")
```

Measurement error is a different problem. If an observed predictor is noisy,
its true value is latent even where the recorded value is present. The `mi()`
workflow below handles absent predictor values, not measurement error.

## Missing response cells

Suppose four trait measurements are absent from a two-trait data set. With
`response = "include"`, the cells remain identifiable in the stacked data but
contribute zero to the response likelihood. The fitted model can then return a
plug-in fitted value for each masked cell.

```{r response-data}
set.seed(42)
n_site <- 24

dat <- data.frame(
  site = factor(seq_len(n_site)),
  z = rnorm(n_site)
)
dat$t1 <- 0.4 + 0.7 * dat$z + rnorm(n_site, sd = 0.4)
dat$t2 <- -0.2 + 0.3 * dat$z + rnorm(n_site, sd = 0.5)
dat$t1[c(3, 12)] <- NA
dat$t2[c(7, 19)] <- NA

colSums(is.na(dat[c("t1", "t2")]))
```

### Wide and long calls

The wide call uses `traits(...)`. The equivalent long call spells out the
trait-specific intercepts and slopes. Both enter through `gllvmTMB()`.

```{r response-wide-fit}
fit_response_wide <- gllvmTMB(
  traits(t1, t2) ~ 1 + z,
  data = dat,
  unit = "site",
  family = gaussian(),
  missing = miss_control(response = "include"),
  control = gllvmTMBcontrol(se = FALSE)
)
```

```{r response-long-fit}
dat_long <- rbind(
  data.frame(site = dat$site, trait = "t1", value = dat$t1, z = dat$z),
  data.frame(site = dat$site, trait = "t2", value = dat$t2, z = dat$z)
)
dat_long$trait <- factor(dat_long$trait, levels = c("t1", "t2"))

fit_response_long <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):z,
  data = dat_long,
  trait = "trait",
  unit = "site",
  family = gaussian(),
  missing = miss_control(response = "include"),
  control = gllvmTMBcontrol(se = FALSE)
)
```

The two forms maximize the same observed-data likelihood in this example.

```{r response-parity}
data.frame(
  format = c("wide", "long"),
  log_likelihood = c(logLik(fit_response_wide), logLik(fit_response_long)),
  likelihood_rows = c(nobs(fit_response_wide), nobs(fit_response_long))
)
```

### Reconstruct the masked responses

`predict_missing()` returns one row per missing response cell. Its `est` column
is a fitted linear predictor by default; use `type = "response"` for the
inverse-link mean. These are plug-in fitted values based on the fitted
parameters and conditional random-effect estimates. They do not currently
carry reconstruction standard errors or prediction intervals.

```{r response-predict}
missing_response <- predict_missing(fit_response_wide, type = "response")
missing_response
```

For wide input, `original_row` is the row of `dat`, while `model_row` is the
internally stacked cell row. Join on `site` plus `trait` when identifying a
cell; those columns remain stable if the stacking order changes.

`nobs(fit_response_wide)` is 44 because 48 cells were supplied and four were
masked. Retaining a cell's identity is not the same as adding information: the
missing response itself contributes nothing to the likelihood. Estimating
cross-trait dependence also requires adequate overlap among observed traits
within the relevant units or covariance groups.

## One missing continuous predictor

A missing predictor must be declared in both parts of the joint model:

- `mi(x)` places the unit-level predictor in the response formula;
- `impute = list(x = x ~ z)` specifies a model for the predictor;
- `miss_control(predictor = "model")` activates the route.

The following wide-format example models one continuous predictor. The
predictor is observed once per site and is shared by both trait responses.

```{r predictor-data}
set.seed(47)
n_site <- 30

pred_dat <- data.frame(
  site = factor(seq_len(n_site)),
  z = rnorm(n_site)
)
pred_dat$x <- 0.3 + 0.8 * pred_dat$z + rnorm(n_site, sd = 0.5)
pred_dat$t1 <- 0.5 + 1.1 * pred_dat$x + rnorm(n_site, sd = 0.4)
pred_dat$t2 <- -0.2 + 1.1 * pred_dat$x + rnorm(n_site, sd = 0.4)
pred_dat$x[c(4, 13, 21)] <- NA
```

```{r predictor-fit}
fit_predictor <- gllvmTMB(
  traits(t1, t2) ~ 1 + mi(x),
  data = pred_dat,
  unit = "site",
  family = gaussian(),
  impute = list(x = x ~ z),
  missing = miss_control(predictor = "model"),
  control = gllvmTMBcontrol(se = TRUE)
)

imputed_wide <- imputed(fit_predictor)
imputed_wide
```

For a continuous Gaussian predictor, `estimate` is a conditional mode and
`std_error` is a local Gaussian/Laplace conditional standard error when the
joint Hessian is available. It is not a multiple-imputation draw, a Bayesian
posterior interval, or evidence of calibrated interval coverage. The `level`
and `level_id` columns are the safe join key. Here they identify `site`; the
row ordinals refer to predictor-model levels and are not general data-row keys.

The same model can start from long data. The shared `mi(x)` coefficient applies
to both traits, while `0 + trait` gives each trait its own intercept.

```{r predictor-long-fit}
pred_long <- rbind(
  data.frame(site = pred_dat$site, trait = "t1", value = pred_dat$t1,
             x = pred_dat$x, z = pred_dat$z),
  data.frame(site = pred_dat$site, trait = "t2", value = pred_dat$t2,
             x = pred_dat$x, z = pred_dat$z)
)
pred_long$trait <- factor(pred_long$trait, levels = c("t1", "t2"))

fit_predictor_long <- gllvmTMB(
  value ~ 0 + trait + mi(x),
  data = pred_long,
  trait = "trait",
  unit = "site",
  family = gaussian(),
  impute = list(x = x ~ z),
  missing = miss_control(predictor = "model"),
  control = gllvmTMBcontrol(se = TRUE)
)

imputed_long <- imputed(fit_predictor_long)
merge(
  imputed_wide[c("level", "level_id", "estimate")],
  imputed_long[c("level", "level_id", "estimate")],
  by = c("level", "level_id"),
  suffixes = c("_wide", "_long"),
  sort = FALSE
)
```

## Other implemented predictor types

The response formula still uses `mi(x)`. The predictor family is set inside
`impute_model()`.

- **Continuous:** `impute_model(x ~ z, family = gaussian())`, or the shorthand
  `x ~ z`, returns a conditional mode and, when available, a conditional SE.
- **Binary:** `impute_model(x ~ z, family = binomial())` returns the fitted
  conditional probability of level 1, without a parameter-uncertainty SE.
- **Ordered:** `impute_model(x ~ z, family = cumulative_logit())` returns a
  fitted conditional expected numeric score, without an SE. The public summary
  does not return the per-level probability vector.
- **Unordered:** `impute_model(x ~ z, family = categorical())` returns the
  numeric code of the fitted modal category, without an SE. The public summary
  does not currently return its category label or per-level probabilities.

The engine marginalizes over discrete missing states under the estimated joint
model. `imputed()` exposes a probability for the binary case and a one-number
summary for the ordered and unordered cases; it does not expose parameter
uncertainty for any of them.

Gaussian predictor models support fixed effects, one grouped random intercept,
one coarser level declared with `mi_group()`, or an intercept-only phylogenetic
effect. The discrete predictor models are fixed-effect models. See
[`impute_model()`](../reference/impute_model.html) for the current call shapes.
Do not extrapolate those routes to spatial, animal, count, bounded, or multiple
simultaneous missing predictors.

## Choose another route when needed

- **More than one predictor is missing:** use a suitable external
  multiple-imputation or joint-model method.
- **Missingness may depend on the unobserved value:** perform a sensitivity
  analysis with a method that models the missingness mechanism.
- **An observed predictor has measurement error:** use a measurement-error
  model; `mi()` is not one.
- **You need uncertainty intervals for reconstructed responses:** treat
  `predict_missing()` as point output and use a method with
  predictive-uncertainty support.
- **You need completed data sets for several downstream analyses:** use a
  multiple-imputation workflow rather than treating `imputed()` as completed
  data.

`gllvmTMB` estimates its supported missing-predictor component jointly with the
response model. Joint fitting makes the assumptions explicit, but it does not
guarantee that the predictor model is correctly specified or automatically
congenial with every scientific analysis. Multiple imputation remains a
different and often preferable workflow when the goal is a set of completed
data sets, pooling across downstream models, or broader multivariable
missingness.

## What the current evidence supports

The evidence is strongest for Gaussian response masking and for selected
continuous, binary, ordered, unordered, grouped, and phylogenetic predictor
models under known generating processes. It does not establish nominal
interval coverage across missingness mechanisms, response families, covariance
structures, or missingness rates. In particular, this article is not an MNAR
robustness study.

## Next decision

Proceed to covariance interpretation only after the retained-row counts,
missingness assumption, predictor model (when used), and the output from
`check_gllvmTMB()` are all reported. Use [Can I trust this
fit?](fit-diagnostics.html) for the
post-fit checks. If the missingness mechanism or imputation model cannot be
defended, stop and compare a sensitivity analysis or an external multiple-
imputation workflow rather than treating one completed fit as definitive.

## References

- Rubin, D. B. (1976). Inference and missing data. *Biometrika*, 63, 581--592.
- Nakagawa, S. & Freckleton, R. P. (2008). Missing inaction: the dangers of
  ignoring missing data. *Trends in Ecology & Evolution*, 23, 592--596.
- Schafer, J. L. & Graham, J. W. (2002). Missing data: our view of the state
  of the art. *Psychological Methods*, 7, 147--177.
- Ibrahim, J. G., Chen, M.-H., Lipsitz, S. R. & Herring, A. H. (2005).
  Missing-data methods for generalized linear models: a comparative review.
  *Journal of the American Statistical Association*, 100, 332--346.
- Kristensen, K., Nielsen, A., Berg, C. W., Skaug, H. & Bell, B. M. (2016).
  TMB: automatic differentiation and Laplace approximation. *Journal of
  Statistical Software*, 70(5), 1--21.
