---
title: "Handling missing data"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Handling missing data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# All chunks are eval=FALSE on purpose: knitting the TMB fits in this article
# is heavy (it has OOM'd pkgdown before), so the chunks show the exact shipped
# syntax without being run at build time. The article documents the shipped
# Phase 1 (response mask), Phase 2a (Gaussian fixed-covariate mi()),
# Phase 2b (grouped-covariate mi()), Phase 3 (phylogenetic species-level
# mi()), and Phase 5a (binary mi() via exact 2-state summation) surface; do
# not flip these to eval=TRUE in the pkgdown build.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

Missing data in `gllvmTMB` has two meanings.
A **response cell** can be missing: an individual was measured but one trait
was not recorded, leaving a gap in the multivariate trait matrix.
A **predictor** can be missing: a covariate needed by the response model was
not observed for some rows.
The current implementation treats those two cases differently, and both differ
from the third case--measurement error--which is explicitly out of scope for
this layer (see Section 8).

The starting point is what *not* to do.
Listwise deletion discards every row with any gap, which in a multivariate
trait matrix can mean throwing away most of the data and biasing estimates
whenever missingness is not completely at random.
Mean-fill and single imputation invent values and then treat them as if they
were observed, understating uncertainty and distorting the covariance
structure that an ordination or a phylogenetic model is trying to estimate.
For structured ecological data--shared species means, phylogenetic covariance,
spatial fields, latent axes--these ad-hoc fixes are particularly damaging
because they sever exactly the correlations the model exists to exploit.
`gllvmTMB` instead keeps the missingness *inside* the likelihood: nothing is
deleted and nothing is invented, and the uncertainty that missingness creates
propagates through the same joint Hessian as every other parameter.

For missing response cells, `gllvmTMB` retains the original rows and evaluates
the observed-cell likelihood by masking the unobserved entries.
Masked cells contribute no direct response likelihood, but the fitted object
still knows where those cells were.
For a missing predictor, `gllvmTMB` requires an explicit predictor model
declared through `mi()` in the response formula and `impute = list(...)` in
the fit call; the shipped routes are continuous Gaussian predictors (fixed,
grouped, or phylogenetic covariate models) and one binary predictor with a
fixed-effect covariate model.
A continuous missing predictor is a latent quantity integrated out by TMB's
Laplace approximation; a binary one has no latent variable -- its two states are
summed out exactly inside the likelihood (Section 5). Either way this is
frequentist full-information maximum likelihood (FIML), not multiple imputation.

This is likelihood-based missing-data handling.
It is not Bayesian posterior imputation, not multiple imputation, and not an
EM interface.
For cases where those approaches are preferred--rich GNN phylogenetic
imputation, mixed trait types, tree uncertainty--the `pigauto` package
provides the complementary preprocessing path (Section 6).

### Status: what is shipped versus planned

This article documents the missing-data layer as it stands today and sketches
the planned extensions, so the reader can see the design as a whole.
Read the status tags literally:

- **SHIPPED (default, complete-case).** Under the default `miss_control()`
  (`response = "drop"`, `predictor = "fail"`, `engine = "laplace"`) the layer is
  an exact no-op: a missing *response* cell drops its row before the likelihood
  is built, and a missing *predictor* is an error. A complete-data fit is
  identical to one built before the layer existed.
- **SHIPPED (Phase 1, the response mask).** `miss_control(response =
  "include")` retains rows with a missing response cell, builds the
  `is_y_observed` mask so masked cells contribute zero to the likelihood, keeps
  original-row accounting in `fit$missing_data`, adjusts `nobs()` to the
  likelihood-contributing cells, and exposes `predict_missing()` for the masked
  response cells.
- **SHIPPED (Phase 2a, one Gaussian fixed-covariate missing predictor).** A
  single continuous, unit-level missing predictor declared with `mi(x)` plus
  `missing = miss_control(predictor = "model")` and a fixed-effect Gaussian
  covariate model supplied through `impute = list(x = x ~ z)` (bare-formula
  sugar) or `impute = list(x = impute_model(x ~ z, family = gaussian()))`. The
  missing value is a latent quantity integrated out by the Laplace
  approximation; `imputed()` returns its conditional mode (EBLUP) and standard
  error, and the registry lives at `fit$missing_data$predictors`.
- **SHIPPED (Phase 2b, one grouped-covariate missing predictor).** The Phase 2a
  Gaussian predictor model extended with a single grouped random intercept,
  `impute = list(x = x ~ z + (1 | group))`. The grouped effect is a
  covariate-model random intercept (one unit-level value per group), kept
  independent of the response-model random effects; everything else matches
  Phase 2a.
- **SHIPPED (Phase 3, phylogenetic species-level missing predictor).** A
  species-level Gaussian missing predictor whose covariate model carries a
  phylogenetic intercept, `impute = list(x = x ~ phylo(1 | species, tree =
  tree))`. The `phylo(1 | species, tree =)` token both keys the species-level
  broadcast (the Phase 2c machinery: one latent per missing species, broadcast
  to its observation rows) and adds a phylo-structured intercept on the
  covariate mean. The field borrows strength across related species and
  degrades gracefully toward the independent (Phase 2c) model when phylogenetic
  signal is weak; `phylo_signal_mi()` reports the effective signal and flags the
  weak case (Section 5c). Intercept-only covariate model; the Level-2 joint
  field (`correlate_with = "response"`) remains Phase 4.
- **SHIPPED (Phase 5a, one binary missing predictor).** A single two-level
  (binary) missing predictor declared with `mi(x)` and a fixed-effect
  Bernoulli-logit covariate model supplied through
  `impute = list(x = impute_model(x ~ z, family = binomial()))`. This is the
  first *discrete* predictor and the first instance of the exact finite-state
  route: there is **no latent `x_mis`**, the two states `{0, 1}` are summed
  exactly inside the likelihood, and `imputed()` returns the conditional
  probability `P(x = 1 | y, z)` (not a Laplace mode). The binary covariate model
  is fixed-effect only in v1: a grouped, level-keyed, or phylogenetic binary
  predictor model errors.
- **PLANNED.** Spatial / relatedness / animal covariate models for `mi()` and
  the Level-2 joint field (Phase 4); and the remaining non-Gaussian or
  categorical missing predictors -- ordered (Phase 5b), unordered (Phase 5c),
  and bounded/continuous non-Gaussian such as beta. These routes are sketched
  below with their phase tags.

**Two co-equal input shapes.** Every route below accepts the model in either of
two equivalent shapes. A **wide** call uses the `traits(...)` LHS on a data
frame with one row per unit and one column per trait
(`traits(t1, t2) ~ mi(x)`, `data = df_wide`). A **long** call uses an explicit
stacked response on a data frame with one row per `(unit, trait)` observation
(`value ~ 0 + trait + mi(x)`, `data = df_long`, `unit = "id"`). These are not
two engines: `traits(...)` simply pivots the wide frame to the long stacked form
internally and dispatches to the **same** stacked-long likelihood that the long
call reaches directly. Missing-data handling is therefore identical in both --
the response mask, the `mi()` latent predictor, and every diagnostic behave the
same way regardless of which shape you pass. The examples below mostly use the
wide `traits(...)` form for compactness; Sections 2 and 5 also give the long-
format equivalent so the co-equal long path is explicit.

Code chunks for fitted models are kept `eval = FALSE`: they show the exact
syntax of the shipped surface, but knitting the TMB fits is intentionally
skipped here. Each section labels whether its API is shipped or planned.

## 1. What is fitted now

The default is conservative complete-case fitting: a missing response cell
drops its row, and a missing predictor is an error. The non-default routes are
engaged through `miss_control()`.

```{r what-is-fitted-default}
# Defaults: response = "drop", predictor = "fail", engine = "laplace".
# miss_control()
```

`missing = miss_control(response = "include")` retains rows that have a missing
response cell, masking the missing cells out of the likelihood. `missing =
miss_control(predictor = "model")` is used when the response formula contains a
supported `mi()` term and `impute` supplies the corresponding predictor model;
without a matching `mi()` term it errors rather than silently doing nothing.

The table below maps each missing-data case to its route and the boundary that
route stops at. The phase tag in the final column gives its status: Phase 1,
Phase 2a, Phase 2b, and Phase 3 are shipped; Phase 4 onward is planned.

| Missing-data case | Route | Boundary / phase |
| --- | --- | --- |
| Missing trait cell (any family) | `missing = miss_control(response = "include")` | All predictors, grouping variables, offsets, weights, and structured-effect inputs must be complete for that row (Phase 1, shipped). |
| One missing continuous predictor, fixed covariate model | `traits(y1, y2) ~ mi(body_mass) + z` with `impute = list(body_mass = body_mass ~ z)` | The `mi()` variable must be numeric; one Gaussian covariate model; no multiple missing predictors yet (Phase 2a, shipped). |
| One missing continuous predictor, grouped covariate model | As above with `impute = list(body_mass = body_mass ~ z + (1 | site))` | One grouped random intercept in the predictor model; independent of response random effects (Phase 2b, shipped). |
| Species-level missing predictor, phylogenetic covariate model | `impute = list(body_mass = body_mass ~ phylo(1 | species, tree = tree))` | Species-level value broadcast to observation rows; intercept-only phylo covariate model; degrades gracefully when signal is weak (Phase 3, shipped). |
| Spatial / relatedness / animal covariate model; Level-2 joint field | `impute = list(... = ... ~ spatial(...))`, `relmat()`, `animal()`, or `correlate_with = "response"` | Non-phylo structured covariate models and the joint response-covariate field (Phase 4, planned). |
| One missing binary predictor, fixed covariate model | `traits(y1, y2) ~ z + mi(x)` with `impute = list(x = impute_model(x ~ z, family = binomial()))` | Exact two-state summation, no latent `x_mis`; `imputed()` returns `P(x = 1 | y, z)`; fixed-effect covariate model only (Phase 5a, shipped). |
| Other non-Gaussian / categorical missing predictor | `impute_model(family = )` with a non-binary non-Gaussian family | Ordered (`cumulative_logit`), unordered (`categorical`), beta / bounded, Poisson / NB -- later work (Phase 5b/5c+, planned). |

The `gllvmTMB` response matrix is **multivariate**: a single fit can have many
trait columns and many species.
Missing cells are indexed at the (row, trait) level, not at the response-row
level as in a univariate model.

## 2. Missing response cells in `traits()`

For a multivariate response model, each (row, trait) cell is either observed or
missing.
The observed-cell likelihood for row `i` and trait `j` is:

```
is_y_observed[i, j] = 1  -->  full log p(y_{ij} | eta_{ij})
is_y_observed[i, j] = 0  -->  0  (cell masked; row kept for random-effect
                                   borrowing across traits and species)
```

The `[i, j]` indexing above is conceptual: internally the mask is stored as the
flat long-stacked indicator vector over the model rows (unit-major), aligned
one-to-one with the stacked response, not as a 2-D matrix.

A row with at least one observed trait cell is retained.
A row where every trait is missing is also retained for row accounting but
contributes zero direct response likelihood.

This example creates a small trait matrix with deliberate missing cells (kept
`eval = FALSE`; it shows the shipped syntax for `miss_control(response =
"include")`):

```{r missing-response-example}
# set.seed(1)
# n <- 20; p <- 3
# dat <- data.frame(
#   species = rep(paste0("sp", 1:n), 1),
#   x       = rnorm(n)
# )
# Y <- matrix(rnorm(n * p), n, p,
#             dimnames = list(NULL, c("trait1", "trait2", "trait3")))
# Y[3, 1] <- NA; Y[7, 2] <- NA; Y[12, 1] <- NA; Y[12, 3] <- NA  # deliberate gaps
#
# fit_resp <- gllvmTMB(
#   traits(trait1, trait2, trait3) ~ x,
#   family  = gaussian(),
#   data    = cbind(dat, Y),
#   missing = miss_control(response = "include")
# )
```

Cross-trait borrowing still operates on the retained rows because the latent
structure is shared across all cells in a row.
A row where trait 1 is missing but traits 2 and 3 are observed still
contributes to the species means and latent axes for all traits.

The analogy with `drmTMB` partial bivariate pairs (where one of `y1`/`y2` can
be observed) generalises here: an observed cell informs its own trait
mean/loadings, but cross-trait species correlation is identified only from
rows where both cells are co-observed.

### Same mask in long format

The `traits(...)` call above is the wide entry point; the response mask is
defined on the stacked-long form it pivots to, so the identical fit can be
written directly in long format. Here the data frame has one row per
`(unit, trait)` observation, the response is the single stacked `value` column,
the trait intercepts are `0 + trait`, and `unit =` names the between-unit
grouping column. Setting `value` to `NA` on individual rows masks exactly those
cells (kept `eval = FALSE`):

```{r missing-response-long}
# long_df: one row per (id, trait). Columns: id, trait (factor), x, value.
# A few `value` cells set to NA are the masked response cells.
# long_df$value[c(5, 17, 42)] <- NA
#
# fit_resp_long <- gllvmTMB(
#   value ~ 0 + trait + (0 + trait):x,
#   data    = long_df,
#   unit    = "id",                 # between-unit grouping column
#   family  = gaussian(),
#   missing = miss_control(response = "include")
# )
```

This long call and the wide `traits(...)` call are **co-equal inputs**: the
wide form pivots to exactly this stacked-long layout before the likelihood is
built, so `fit$missing_data`, `nobs()`, and `predict_missing()` behave
identically whichever shape you pass. Use whichever matches how your data are
already stored -- wide when you have a unit x trait matrix, long when the
observations are already stacked.

## 3. Inspecting `fit$missing_data` and `nobs()`

Every fitted object that used `miss_control(response = "include")` or
`miss_control(predictor = "model")` carries a `fit$missing_data` slot with
row-accounting and diagnostics.

```{r inspect-metadata}
# nobs(fit_resp)                     # likelihood-contributing cells
# fit_resp$missing_data$counts       # see field table below
# fit_resp$missing_data$slice        # implementation slice, e.g. "Phase1-s2"
# head(fit_resp$missing_data$original_row)
# fit_resp$missing_data$observed_y   # logical mask over the model rows
```

The most useful fields are:

| Field | Meaning |
| --- | --- |
| `slice` | Implementation slice tag (the response mask ships as `"Phase1-s2"`). |
| `contract_version` | Shared-contract version with `drmTMB`, e.g. `"59-v1"`. |
| `response_policy` | Whether response rows/cells were dropped (`"drop"`) or masked (`"include"`). |
| `predictor_policy` | Whether missing predictors fail (`"fail"`) or are modelled (`"model"`). |
| `original_row` | Row indices in the user's original data frame. |
| `model_row` | Row indices in the retained model frame. |
| `observed_y` | Logical mask (`is_y_observed`): the flat long-stacked vector over the model rows (unit-major), not a 2-D matrix. |
| `counts` | Named list: `retained_rows`, `observed_response`, `missing_response`, `likelihood_rows` (with `n_*` aliases). |
| `predictors` | Per-predictor registry for explicitly modelled `mi()` predictors (Phase 2a). |

For ordinary user work, prefer `nobs()`, `fitted()`, `residuals()`, and
`predict_missing()` over hand-parsing the slot.
The slot exists so row accounting stays auditable after fitting.

A co-observed cell count diagnostic is planned for the multivariate Gaussian
case: if very few rows have both traits observed, the cross-trait residual
correlation is weakly identified and `gllvmTMB` will warn.

## 4. `predict_missing()` for response cells

Missing response cells have predicted values from the fitted model: each masked
cell is reconstructed from the row's own predictors and the shared latent
structure (species means, latent axes, structured fields) that the rest of the
data identifies. This is a model-based conditional expectation given whatever is
observed for that row, not a value read back from row-local response
information; even an all-missing row is reconstructed this way, borrowing
entirely from its predictors and the shared structure rather than from any
observed cell of its own.
Use `predict_missing()` to retrieve those reconstructed values.

```{r predict-missing}
# pm <- predict_missing(fit_resp)            # type = "link" (default)
# pm <- predict_missing(fit_resp, type = "response")
# head(pm)
# # One row per masked cell. Columns: original_row, model_row, the unit/cluster
# # and trait identifier columns, and `est` (the prediction on the requested
# # scale). A complete-data fit returns a zero-row data frame.
```

The point prediction is the fitted linear predictor (`type = "link"`) or its
inverse-link conditional mean (`type = "response"`).
Standard errors of the reconstructed cells are deferred to a later slice, so
`predict_missing()` returns the point predictions only; they are model-based
fitted values, not posterior intervals and not multiple-imputation draws.

`predict_missing()` is for **response** cells only.
It is distinct from `imputed()`, which summarizes missing **predictors**
(Section 5): `predict_missing()` returns point predictions only, whereas
`imputed()` carries a conditional standard error for a continuous (latent)
predictor (a binary predictor instead returns a conditional probability with no
SE; Section 5d).
The terminology follows the shared contract with `drmTMB`: predicted /
reconstructed responses versus EBLUP conditional modes for latent predictors.

Residuals for missing cells are `NA`; fitted values are not.
This is deliberate: a residual requires an observed value to compute, but a
fitted value does not.

## 5. Missing predictors: `mi()` and `impute_model(family=)`

Missing predictors are not handled by row deletion or sentinel imputation.
The response model needs the predictor value, so `gllvmTMB` asks for an
explicit model for that predictor.
For a **continuous** predictor the missing values become latent random effects
integrated out by the Laplace approximation, jointly with the response-model
random effects; for a **discrete** predictor (the shipped binary case) there is
no latent variable and the finitely many states are summed out exactly instead.
The two routes are spelled out in the bullets below.

For the continuous case the joint likelihood factorises as:

```
L(theta) = integral
  p(y_obs | x_obs, x_mis, b, theta_y)
  * p(x_obs, x_mis | b_x, theta_x)
  * p(b, b_x | theta_b)
  d x_mis db db_x
```

In `gllvmTMB`, `x_mis` is declared in the TMB `random` set and integrated out
by the Laplace approximation together with all other random effects.
`sdreport` then returns the conditional mode (EBLUP) and its standard error
from the inverse joint Hessian.

How the integral is taken depends on the type of the missing predictor, and
the two cases are handled differently:

- **Continuous predictors (Laplace).** A missing continuous covariate is a
  real-valued latent quantity. The Laplace approximation replaces its integral
  with the joint mode plus a Gaussian curvature term, exactly as for any other
  random effect. The reported value is its conditional mode (EBLUP). This is
  the continuous path; its shipped instances are the Gaussian fixed-covariate
  model of Phase 2a (Section 5a), the grouped-covariate model of Phase 2b
  (Section 5b), and the phylogenetic species-level model of Phase 3
  (Section 5c), with spatial / relatedness covariate models and the Level-2
  joint field planned in Phase 4.
- **Discrete predictors (exact finite-state summation).** A missing factor or
  ordinal covariate has finitely many possible values, so it is handled by exact
  finite summation rather than a Gaussian approximation: the likelihood is
  evaluated at each admissible level and weighted by the covariate model's
  category probabilities, with **no latent variable and no Gaussian
  approximation** used for the discrete dimension. The response term is replaced
  by the finite mixture over the predictor's states. The first shipped instance
  is the **binary** predictor of Phase 5a (Section 5d), whose two states `{0, 1}`
  are summed exactly; ordered (`cumulative_logit`, Phase 5b) and unordered
  (`categorical`, Phase 5c) predictors extend the same K-state machinery and are
  landing next. A bounded continuous family (beta) would instead use quadrature
  on `(0, 1)` and remains planned.

Both routes are a single frequentist likelihood -- the continuous case
integrates the latent predictor out by Laplace, the discrete case sums the
finite states out exactly -- with no Bayesian prior, no posterior, and no
multiple-imputation ensemble in either. And because `traits(...)` simply pivots
the wide frame to the stacked-long form that the long call reaches directly
(Section 2), **every family below works in both the wide `traits(...)` shape and
the long `value ~ 0 + trait` shape**; the family lives in the `impute` model, not
in the input shape.

### Missing-predictor families at a glance

The table below is the master map of predictor families against the two input
shapes. Each shipped family has a runnable-looking (`eval = FALSE`) example in
**both** shapes in the cited subsection; the not-yet-shipped families show their
*intended* call so the design is legible, tagged so they are not mistaken for
working code. `cumulative_logit`, `categorical`, and `beta` are illustrative
family names for the planned slices and are **not callable on `main` today**.

| Family | `impute` covariate model | Integration | `imputed()` output | Status | Section |
| --- | --- | --- | --- | --- | --- |
| Gaussian (continuous) | `x ~ z` (bare = Gaussian) or `impute_model(x ~ z)` | latent + Laplace | conditional mode (EBLUP) + SE | **shipped (2a)** | 5a |
| Gaussian + grouped intercept | `x ~ z + (1 \| group)` | latent + Laplace | EBLUP + SE | **shipped (2b)** | 5b |
| Gaussian + phylogenetic | `x ~ phylo(1 \| species, tree =)` | latent + Laplace + GMRF (`Ainv`) | EBLUP + SE; `phylo_signal_mi()` | **shipped (3)** | 5c |
| Binary | `impute_model(x ~ z, family = binomial())` | exact 2-state sum (no latent) | conditional prob `P(x = 1 \| y, z)` | **shipped (5a)** | 5d |
| Ordered | `impute_model(x ~ z, family = cumulative_logit())` | exact K-state sum | expected category score | landing next (5b) | 5f |
| Unordered | `impute_model(x ~ z, family = categorical())` | exact K-state sum | modal category / level probs | landing next (5c) | 5f |
| Beta / proportion | `impute_model(x ~ z, family = beta())` | quadrature (bounded) | conditional mean | planned | 5f |

The three Gaussian rows share the **continuous** route (a latent `x_mis`
integrated out by Laplace; `imputed()` reports an EBLUP conditional mode with a
Hessian SE). The binary row is the **discrete** route (no latent; the states are
summed exactly; `imputed()` reports a conditional probability with `std_error =
NA`). Ordered and unordered extend the same exact-summation idea to K states;
beta is the one bounded-continuous case that would use quadrature rather than
either a latent mode or a finite sum.

The shipped families are spelled out next; each gives the wide `traits(...)` and
long `value ~ 0 + trait` forms explicitly. The binary family (newly shipped) is
in Section 5d; the ordered / unordered / beta sketches are in Section 5f.

### 5a. Continuous predictor, fixed covariate model (Phase 2a, shipped)

This is the shipped missing-predictor route. The missing predictor is declared
with `mi(x)` in the response formula; `missing = miss_control(predictor =
"model")` engages the latent-covariate grammar; and `impute` supplies its
Gaussian fixed-effect covariate model. A bare two-sided formula is sugar for a
Gaussian predictor model; `impute_model(x ~ z, family = gaussian())` makes the
family explicit (the family lives in the **predictor** model, not in `mi()`).
The predictor is treated as a unit-level quantity, so there is one latent value
per missing **unit**, broadcast across that unit's trait rows.

```{r missing-predictor-fixed}
# Tiny fixture: 2 traits, a few missing body_mass values.
# set.seed(2)
# n <- 24
# dat2 <- data.frame(
#   moisture   = rnorm(n),
#   body_mass  = rnorm(n),
#   species    = paste0("sp", 1:n)
# )
# Y2 <- cbind(
#   trait1 = 0.5 + 0.8 * dat2$body_mass + rnorm(n, sd = 0.2),
#   trait2 = -0.2 - 0.4 * dat2$moisture + rnorm(n, sd = 0.25)
# )
# dat2$body_mass[c(3, 11, 18)] <- NA  # three missing predictor values
#
# fit_pred <- gllvmTMB(
#   traits(trait1, trait2) ~ mi(body_mass) + moisture,
#   family  = gaussian(),
#   data    = cbind(dat2, Y2),
#   impute  = list(body_mass = body_mass ~ moisture),  # bare = Gaussian sugar
#   # equivalently: impute = list(body_mass = impute_model(body_mass ~ moisture,
#   #                                                       family = gaussian())),
#   missing = miss_control(predictor = "model"),
#   control = gllvmTMBcontrol(se = TRUE)  # se = TRUE -> conditional SEs in imputed()
# )
#
# imputed(fit_pred)   # one EBLUP conditional mode + SE per missing unit
```

`imputed(fit_pred)` returns a data frame with one row per missing unit value
(not one per long-format model row), with columns `variable`, `original_row`,
`model_row`, `observed`, `estimate`, `std_error`, `source`, and
`uncertainty_status`.
The `estimate` column is the conditional mode (EBLUP) at the Laplace optimum,
and `source` is `"conditional_mode"`.
The `std_error` column is the likelihood-based conditional standard error from
the joint Hessian; it is populated when the fit carries a successful `sdreport`
(`control = gllvmTMBcontrol(se = TRUE)`) and is `NA` otherwise, with
`uncertainty_status` recording which case applies.
It is not a posterior interval and is not a set of multiple-imputation draws.

`imputed()` also accepts `rows = "all"` to return every retained unit (observed
units carry their observed predictor value with `NA` standard error), and the
same conditional modes are stored at
`fit_pred$missing_data$predictors$body_mass$conditional_mode`.

#### Same `mi()` model in long format

As with the response mask, the `mi()` predictor route is defined on the
stacked-long form, so the same model can be written directly in long format. The
data frame has one row per `(unit, trait)`, the response is the stacked `value`
column, `mi(x)` declares the missing predictor (a unit-level quantity, broadcast
across that unit's trait rows), and `impute` supplies its covariate model. Set
`x` to `NA` for every row of a missing unit (kept `eval = FALSE`):

```{r missing-predictor-long}
# long_df: one row per (id, trait). Columns: id, trait (factor), value, x, z.
# x is a unit-level predictor; set it NA on ALL rows of each missing unit.
# long_df$x[long_df$id %in% c("u3", "u11")] <- NA
#
# fit_pred_long <- gllvmTMB(
#   value ~ 0 + trait + mi(x),
#   data    = long_df,
#   unit    = "id",
#   family  = gaussian(),
#   impute  = list(x = x ~ z),                     # bare = Gaussian sugar
#   missing = miss_control(predictor = "model"),
#   control = gllvmTMBcontrol(se = TRUE)
# )
#
# imputed(fit_pred_long)   # one EBLUP conditional mode + SE per missing unit
```

The wide `traits(...)` form and this long `value ~ 0 + trait + ...` form are
**co-equal inputs to the same stacked-long engine** -- the wide call pivots to
this layout internally -- so `imputed()`, the predictor registry at
`fit$missing_data$predictors`, and the conditional modes are identical whichever
shape you pass.

### 5b. Predictor with grouped covariate model (Phase 2b, shipped)

This route is **shipped**. It extends the Phase 2a Gaussian predictor model with
one grouped random intercept in the `mi()` predictor model, for when the missing
covariate varies by a grouping level. The syntax mirrors Phase 2a; only the
`impute` formula gains a single `(1 | group)` term:

```{r missing-predictor-grouped}
# Same dat2/Y2 fixture as Phase 2a, with a grouping column (e.g. genus).
# dat2$genus <- factor(rep(paste0("g", 1:6), length.out = nrow(dat2)))
#
# fit_grp <- gllvmTMB(
#   traits(trait1, trait2) ~ mi(body_mass) + moisture,
#   family  = gaussian(),
#   data    = cbind(dat2, Y2),
#   impute  = list(body_mass = body_mass ~ moisture + (1 | genus)),
#   missing = miss_control(predictor = "model"),
#   control = gllvmTMBcontrol(se = TRUE)
# )
#
# imputed(fit_grp)   # one EBLUP conditional mode + SE per missing unit
```

The grouped effect is a covariate-model random intercept, not a response-model
random effect.
It is kept independent of the response random effects by default (Level-1
independent default; Level-2 joint field is opt-in via
`correlate_with = "response"`, deferred to Phase 4).
The boundary holds at one grouped intercept: random slopes, more than one
covariate random-effect term, and a missing or single-level grouping variable
all still error. Phylogenetic structured covariate models are their own shipped
route (Section 5c); spatial, relatedness, and animal-model structures remain
Phase 4.

### 5c. Species-level predictor with phylogenetic covariate model (Phase 3, shipped)

This is the flagship case for ecology and is now **shipped**: a missing
species-level predictor whose covariate model inherits phylogenetic structure.
The `phylo(1 | species, tree = tree)` token in the `impute` formula does two
things at once -- it keys the **species-level broadcast** (one latent value per
missing species, broadcast to that species' observation rows; the Phase 2c
machinery) and adds a **phylo-structured intercept** on the covariate mean, so a
missing species' predictor borrows strength from its phylogenetic neighbours
rather than only from the global mean. The covariate model is intercept-only on
the phylo side; an ordinary observed covariate may still be added
(`x ~ z + phylo(1 | species, tree = tree)`). Before reaching for it, read the
honesty caveat below: phylogenetic imputation helps only when phylogenetic
signal is strong.

```{r missing-predictor-phylo}
# Tiny fixture: a species-level body_mass trait, a Brownian-motion tree whose
# tip labels match the `species` column, a few species entirely missing on
# body_mass. The species factor keys the phylo field to the tree.
# dat2$body_mass[dat2$species %in% c("sp3", "sp11")] <- NA
#
# fit_phy <- gllvmTMB(
#   traits(trait1, trait2) ~ mi(body_mass) + moisture,
#   family  = gaussian(),
#   data    = cbind(dat2, Y2),
#   impute  = list(body_mass = body_mass ~ phylo(1 | species, tree = tree)),
#   missing = miss_control(predictor = "model"),
#   control = gllvmTMBcontrol(se = TRUE)
# )
#
# imputed(fit_phy)        # one EBLUP conditional mode + SE per missing species
```

The long-format equivalent is the same model with the stacked response, written
`value ~ 0 + trait + (0 + trait):z + mi(body_mass)` with `unit = "id"`; the wide
`traits(...)` and long forms are co-equal inputs to the same stacked-long engine
(Section 5a), and the phylo broadcast keys to the `species` column either way.

**Diagnosing the signal with `phylo_signal_mi()`.** Phylogenetic borrowing is
trustworthy only when the predictor actually carries phylogenetic signal.
`phylo_signal_mi(fit_phy)` reports an effective Pagel's lambda for the covariate
model -- the fraction of species-level variance in the predictor explained by
the phylo field relative to the i.i.d. residual -- and flags the weak case. It
is a verification / interpretation aid, not part of the fit: it does not change
estimates (kept `eval = FALSE`):

```{r phylo-signal-mi}
# diag <- phylo_signal_mi(fit_phy)
# diag$lambda   # effective Pagel lambda: near 1 = strong signal, near 0 = weak
# diag$weak     # TRUE when lambda is below `threshold` (default 0.1)
#
# # Emit an EBLUP-language warning for the weak case:
# phylo_signal_mi(fit_phy, warn = TRUE)
```

The honest starting point: phylogenetic imputation is **signal-dependent and
not always better than complete-case analysis**.
When phylogenetic signal is weak, borrowing strength across related species
adds noise rather than information, and the imputed values can be no better
than -- or worse than -- simply analysing the complete cases
(Johnson et al. 2021; Molina-Venegas 2024).
Even the most accurate method tested by Johnson et al. (2021) gave inaccurate
estimates under strong missingness bias, and a "significant" signal test is not
the same as *strong* signal (Molina-Venegas 2024).
The implementation is built to respect this. As the phylo standard deviation
goes to zero the covariate model **degrades gracefully** toward the independent
(non-phylogenetic) Phase 2c model -- with no separate code path -- and
`phylo_signal_mi()` flags the weak case so a too-weak signal does not pass
silently.
Trait databases are also typically MNAR (missing not at random): rare species
with few measurements are systematically under-represented (Penone et al. 2014).
The current implementation does not correct for MNAR; a delta-adjustment
sensitivity analysis is a Phase 4+ gate, not a v1 argument.

### 5d. Binary predictor, exact two-state summation (Phase 5a, shipped)

This is the first **discrete** missing predictor and now **shipped**. A
two-level (binary) covariate is declared with `mi(x)` in the response formula,
and its Bernoulli-logit covariate model is supplied through `impute` with the
family made explicit: `impute_model(x ~ z, family = binomial())`. The link must
be the default logit (`binomial("probit")` errors). Unlike the Gaussian route,
there is **no latent `x_mis`**: because the predictor has only two admissible
states, the likelihood is evaluated at `x = 0` and at `x = 1` and the two are
summed exactly, weighted by the covariate model's probabilities. The state is a
unit-level quantity, so the per-unit response factor is the product over that
unit's trait rows before the two states are combined. The binary covariate model
is **fixed-effect only** in this version -- a grouped `(1 | group)`, a level key
(`mi_group()`), or a structured `phylo()` term on a binary predictor all error,
and only one `mi()` term is allowed.

The accepted predictor type is a two-level factor, a logical, or a 0/1 numeric;
a factor with more than two levels is redirected to `categorical()` (Section 5f)
and an ordered factor to `cumulative_logit()` (Section 5f).

```{r missing-predictor-binary-wide}
# WIDE form: one row per unit (site), two trait columns, a binary x with a few
# units missing. The covariate model is logistic on z and w.
# set.seed(13)
# n <- 70
# z <- rnorm(n); w <- rnorm(n)
# x <- rbinom(n, 1, plogis(0.1 + 0.9 * z - 0.5 * w))   # 0/1 predictor
# wide <- data.frame(
#   site = factor(seq_len(n)),
#   t1   = 0.5 + 1.1 * x - 0.2 * z + rnorm(n, sd = 0.35),
#   t2   = -0.3 + 1.1 * x + 0.4 * z + rnorm(n, sd = 0.45),
#   x = x, z = z, w = w
# )
# wide$x[c(3, 14, 28, 41, 55)] <- NA          # five missing binary units
#
# fit_bin <- gllvmTMB(
#   traits(t1, t2) ~ z + mi(x),
#   data    = wide,
#   unit    = "site",
#   family  = gaussian(),                      # RESPONSE family
#   impute  = list(x = impute_model(x ~ z + w, family = binomial())),  # PREDICTOR family
#   missing = miss_control(predictor = "model"),
#   control = gllvmTMBcontrol(se = TRUE)
# )
#
# imputed(fit_bin)   # one row per missing unit; `estimate` is P(x = 1 | y, z)
```

The same model in **long** form has one row per `(unit, trait)`, the stacked
`value` response, and `unit =` naming the between-unit column. The binary `x` is
set `NA` on every row of each missing unit:

```{r missing-predictor-binary-long}
# LONG form: one row per (site, trait). Columns: site, trait (factor),
# value, x (0/1, unit-level), z, w. Set x = NA on ALL rows of each missing site.
# long_df$x[long_df$site %in% c("3", "14", "28", "41", "55")] <- NA
#
# fit_bin_long <- gllvmTMB(
#   value ~ 0 + trait + (0 + trait):z + mi(x),
#   data    = long_df,
#   unit    = "site",
#   family  = gaussian(),
#   impute  = list(x = impute_model(x ~ z + w, family = binomial())),
#   missing = miss_control(predictor = "model"),
#   control = gllvmTMBcontrol(se = TRUE)
# )
#
# imputed(fit_bin_long)   # identical contract: P(x = 1 | y, z) per missing unit
```

The wide `traits(...)` and long `value ~ 0 + trait` forms are **co-equal inputs
to the same stacked-long engine** (Section 5a), so the binary registry and
`imputed()` output are identical whichever shape you pass.

For the binary family `imputed()` returns one row per missing unit, but its
columns mean something different from the Gaussian route. The `estimate` is the
**conditional probability** `P(x = 1 | y, z)` from the exact two-state
marginalisation -- a fitted probability in `[0, 1]`, not a latent mode. Because
the discrete route reports a fitted distribution rather than a single mode with a
curvature, `std_error` is `NA` for every unit and `uncertainty_status` is
`"discrete_no_se"`; the `source` column reads `"conditional_probability"`. The
same per-unit probabilities are stored at
`fit_bin$missing_data$predictors$x$conditional_probability`, and the registry
records `family = "bernoulli"` with `levels = c("0", "1")`.

This is still one frequentist likelihood: the conditional probability is a
maximum-likelihood quantity from the same joint fit, not a posterior probability
and not a multiple-imputation draw.

### 5e. Multi-source and spatial covariate models (Phase 4, planned)

Planned, not yet shipped. Phase 4 adds spatial (`spatial()`),
relatedness-matrix (`relmat()`), and
animal-model (`animal()`) covariate structures, and enables the Level-2 joint
field where the predictor model explicitly shares a structured field with the
response model (with eigenvector-fixed-effect orthogonalisation to guard
against structured confounding; Dupont, Marques and Kneib 2023).

```{r missing-predictor-phase4}
# FILL: Phase 4 -- spatial, relmat, animal, multi-source, Level-2 joint field.
```

### 5f. Ordered, unordered, and beta predictors (Phase 5b/5c+, planned)

Planned, not yet shipped -- shown here so the discrete-summation design is
legible end to end. The family names below (`cumulative_logit`, `categorical`,
`beta`) are **illustrative tokens for the planned slices and are not callable on
`main` today**: passing any family other than `gaussian()`/the bare-formula
sugar or `binomial(link = "logit")` to `impute_model()` errors now.

Ordered (Phase 5b) and unordered (Phase 5c) predictors extend the binary
exact-summation route from two states to K: the likelihood is evaluated at each
of the K admissible levels and summed, weighted by the covariate model's
category probabilities, with no latent variable. For an **ordered** predictor the
covariate model is a cumulative-logit GLM and `imputed()` would report an
expected category score; for an **unordered** predictor it is a baseline-category
(multinomial) GLM and `imputed()` would report the modal category and the
per-level probabilities. A **beta** (or other bounded-continuous) predictor is
the one case that is neither a finite sum nor a plain Laplace mode: its latent
value lives on `(0, 1)`, so it would be integrated by quadrature and `imputed()`
would report a conditional mean.

```{r missing-predictor-ordered}
# PLANNED (Phase 5b) -- NOT callable on main today. Intended call shape only:
# fit_ord <- gllvmTMB(
#   traits(t1, t2) ~ z + mi(x),                                    # wide
#   data    = wide, unit = "site", family = gaussian(),
#   impute  = list(x = impute_model(x ~ z, family = cumulative_logit())),
#   missing = miss_control(predictor = "model")
# )
# Long equivalent: value ~ 0 + trait + (0 + trait):z + mi(x), unit = "site".
# imputed(fit_ord)   # would report an expected category score (exact K-state sum)
```

```{r missing-predictor-unordered}
# PLANNED (Phase 5c) -- NOT callable on main today. Intended call shape only:
# fit_cat <- gllvmTMB(
#   traits(t1, t2) ~ z + mi(x),                                    # wide
#   data    = wide, unit = "site", family = gaussian(),
#   impute  = list(x = impute_model(x ~ z, family = categorical())),
#   missing = miss_control(predictor = "model")
# )
# imputed(fit_cat)   # would report the modal category + per-level probabilities
```

```{r missing-predictor-beta}
# PLANNED -- NOT callable on main today. Intended call shape only:
# impute = list(x = impute_model(x ~ z, family = beta()))   # bounded (0, 1);
#   integrated by quadrature, NOT exact summation; imputed() -> conditional mean.
```

## 6. FIML versus multiple imputation

The missing-data layer described above is a frequentist **full-information
maximum-likelihood (FIML)** approach: one joint model, fitted once, with the
missingness handled inside the likelihood. It is deliberately a
**complementary alternative** to multiple imputation (MI) -- one more tool
alongside it, not a wrapper around it and not a replacement for it. The two
strategies answer the same question by different routes.

Multiple imputation runs in three steps. An imputation model **draws** several
plausible completed datasets; the **analysis** model is fitted to each; and the
results are **pooled** by Rubin's rules, which combine the within- and
between-imputation variance into a single standard error. The completed
datasets are an explicit ensemble, and the between-imputation spread is how MI
represents the uncertainty that missingness adds.

FIML never forms that ensemble. The missing values are estimated jointly with
every other parameter in a single fit: missing responses are masked and
reconstructed afterward, and missing predictors are latent quantities
integrated out by the Laplace approximation. The uncertainty that MI recovers
from between-imputation spread is recovered here from the curvature of the same
joint likelihood--the inverse Hessian that `sdreport` already returns. Under a
correctly specified model and data missing at random, FIML and proper MI are
asymptotically equivalent (Schafer and Graham 2002; Collins, Schafer and Kam
2001); FIML simply reaches the answer without a completed-dataset ensemble and
without a pooling step.
(This equivalence is established for the normal-model and GLM classes those
papers study; extending it to the latent-variable, phylogenetic, and spatial
joint models here is a reasonable expectation, not a transferred theorem.)

One honest caveat on the standard errors: the Laplace/FIML standard errors
returned by `sdreport` can be *negatively biased under strong non-normality*
(Allison 2003), so they may understate uncertainty when the model is badly
misspecified. A bootstrap-SE cross-check is a planned verification gate for this
reason; it is not part of the v1 API surface (Section 7).

| | Multiple imputation (incl. `pigauto`) | FIML (this layer) |
| --- | --- | --- |
| Workflow | draw -> analyse -> pool (Rubin's rules) | one joint model, fitted once |
| Completed datasets | an explicit ensemble (e.g. `m = 50`) | none; nothing is filled in |
| Uncertainty | within- + between-imputation variance | inverse joint Hessian (`sdreport`) |
| Missing predictors | requires covariates complete before imputing | latent, integrated out inside the fit |
| Imputation model | separate from the analysis model | *is* the analysis model's structure |

### Congeniality is automatic here

A recurring difficulty in MI is **congeniality** (Meng 1994): the imputation
model and the analysis model are specified separately, and if the imputation
model omits structure the analysis model relies on--an interaction, a
nonlinearity, a phylogenetic covariance--the pooled inference is biased even
when each step is internally correct. A concrete instance is the
transform-then-impute hazard: imputing a predictor and a transformation of it
(a square, an interaction) under separate models yields mutually inconsistent
imputed values (von Hippel 2009). In the FIML layer this mismatch cannot arise
by construction: there is only one model. The "imputation model" for a missing
predictor is the covariate model declared in `impute`, and it lives in the same
likelihood as the response model, sharing its parameters and its structured
fields. The imputation and analysis models are congenial because they are the
same model.
This advantage holds because the analysis model here is fixed and known; when
the downstream analysis is unknown or varied, MI's separation of the two models
is a feature, which is why MI remains the right tool in that setting (below).

This is also why the covariate model is **level-aware** rather than a generic
default. A missing species-level trait is imputed with a phylogenetic
covariate model (`phylo()`); a missing site-level covariate is imputed with a
spatial one (`spatial()`); an animal-model or relatedness structure
(`animal()`, `relmat()`) or the shared latent axes can play the same role. The
imputation model inherits the biological covariance of the analysis it serves--
but it is declared explicitly per predictor, not silently copied from the
response model, so the inheritance is visible and auditable.

### The `pigauto` sister path

Multiple imputation remains available, and for some problems it is the better
tool. `pigauto`
([itchyshin.github.io/pigauto](https://itchyshin.github.io/pigauto/index.html))
is the standalone MI path: a phylogenetic trait-imputation engine--a
Brownian-motion baseline blended with a graph-neural-network component--that
produces completed datasets and propagates uncertainty by multiple imputation
and Rubin's rules (`impute()` -> `multi_impute()` -> `with_imputations()` ->
`pool_mi()`), with conformal intervals and tree-uncertainty support. It imputes
missing **traits** but requires the covariates to be fully observed, and it is
agnostic to the downstream model.

The two packages are complementary rather than competing:

- Prefer **`pigauto` (MI)** when you want rich GNN-based phylogenetic
  imputation, a mix of trait types in one imputation step, explicit
  tree-uncertainty propagation, or a completed dataset you can hand to any
  downstream model.
- Prefer **`gllvmTMB` (FIML)** when you want a single transparent joint model
  with no preprocessing step, when the missingness is in **predictors** (which
  `pigauto` does not handle), or when you want the missing-data uncertainty to
  propagate through the very same likelihood as the rest of the fit.

The interfaces are kept separate on purpose: there is no `engine = "mi"` inside
`miss_control()`, because `mi()` means a latent missing predictor *inside* the
model while MI means multiple imputation *outside* it. A documented
`pigauto`-then-`gllvmTMB` workflow (impute with `pigauto`, fit `gllvmTMB` on
each completed dataset, pool) is the bridge when both are wanted.

## 7. Current boundaries

The design deliberately keeps several features out of scope for the shipped
Phase 1 through Phase 3 plus Phase 5a layer, independent of which phases have
shipped:

- EM, profile, and REML missing-data engines are not implemented.
  `engine = "laplace"` is the only v1 option; `"em"` and `"profile"` are
  reserved names that are not yet accepted.
- Multiple missing predictors (`mi(x1)` and `mi(x2)` in the same formula) are
  not implemented.
- The shipped non-Gaussian missing predictor is the **binary** fixed-effect
  case (Phase 5a, Section 5d). The remaining non-Gaussian, bounded, or
  categorical missing predictors (lognormal, Gamma, beta, Poisson, NB, ordered,
  unordered factor) are Phase 5b/5c+ work.
- Structured covariate models for non-Gaussian predictors (grouped or phylo
  binary / ordered / unordered) are not implemented; the binary predictor model
  is fixed-effect only.
- Interactions inside `mi()`, splines inside `mi()`, and predictor-model
  random slopes are not implemented.
- `imputed()` does not summarize missing **responses**; use `predict_missing()`
  for that.
- `simulate_imputed()` is deferred to Phase 5: if shipped, it will mean
  Gaussian-approximation or parametric-bootstrap draws, not posterior draws.
- Simulated imputations, credible intervals, and multiple-imputation pooling
  are not implemented in this layer; see `pigauto` for the MI sister path.
- Dense known sampling-covariance matrices with partial multivariate response
  rows are not implemented.
- MNAR sensitivity and bootstrap-SE cross-checks are Phase 4+ verification
  gates, not v1 API arguments. (The phylo-signal reliability diagnostic shipped
  with Phase 3 as `phylo_signal_mi()`, Section 5c -- a post-fit interpretation
  aid, not a fit argument.)

When you encounter one of these boundaries, treat the error as a design limit
rather than a failed fit.
For now, use the supported likelihood-based route, fall back to complete-case
fitting, or handle the unsupported missingness outside `gllvmTMB` with a
method whose assumptions you can state explicitly.

## 8. Non-goal: measurement error

Missing predictors in this layer treat the **observed** values of `x` as
exact.
Only `x_mis` (the unobserved values) is latent.

**Measurement error** -- where the true `x` is latent even when nominally
observed, requiring an observation model `x_obs | x_true` with known standard
errors, replicates, or a validation subsample -- is **out of scope** for
Phase 1 through Phase 3.
The missing-predictor registry and `fit$missing_data` slot are designed so a
later measurement-error lane can reuse the latent-covariate machinery (a
possible future `me(x, se = se_x)` argument); but v1 exposes only `mi(x)` for
missing predictors and the two interfaces will remain distinct.

## References

Core references for the statistical foundations:

- Rubin (1976) *Biometrika* 63:581--592. MAR validity and the ignorability
  conditions.
- Schafer & Graham (2002) *Psychological Methods* 7:147--177. ML and MI both
  recommended over case deletion / ad-hoc fill-in; the primary FIML-vs-MI
  anchor.
- Collins, Schafer & Kam (2001) *Psychological Methods* 6:330--351. ML and MI
  yield similar results when comparably implemented (the equivalence anchor).
- Allison (2003) in *Structural Equation Modeling* (Pugesek, Tomer & von Eye,
  eds.). FIML standard errors can be negatively biased under non-normality; the
  bootstrap-SE cross-check caveat.
- Meng (1994) *Statistical Science* 9:538--558. Congeniality of the imputation
  and analysis models.
- von Hippel (2009) *Sociological Methodology* 39:265--291. Transform-then-
  impute: separately imputing a transform and its components yields
  inconsistent values.
- Ibrahim, Chen, Lipsitz & Herring (2005) *JASA* 100:332--346. Missing
  covariates in GLMs via a covariate model inside the likelihood.
- Kristensen et al. (2016) *JSS* 70(5). TMB Laplace + automatic
  differentiation.
- Robinson (1991) *Statistical Science* 6:15--32. EBLUP and best prediction.
- Penone et al. (2014) *Methods Ecol. Evol.* 5:961--970. Phylogenetic
  imputation; trait data typically not missing at random.
- Johnson, Isaac, Paviolo & Gonzalez-Suarez (2021) *Global Ecology and
  Biogeography* 30:51--62. Imputation is signal/bias-dependent and not always
  better than complete-case analysis.
- Molina-Venegas et al. (2018) *Ecography* 41:1740--1749. Phylogenetic
  imputation accuracy varies among lineages.
- Molina-Venegas (2024) *Methods Ecol. Evol.* 15:456--463. Limits of
  phylogenetic imputation; a significant signal test is not strong signal.
- Dupont, Marques & Kneib (2023) arXiv:2309.16861. Structured confounding in
  spatial models.
- Wang, Edge, Schraiber & Pennell (2025 preprint, PMC -- not peer-reviewed).
  Phylogenetic analogue of structured confounding.
