---
title: "Handling missing data"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Handling missing data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# All chunks are eval=FALSE on purpose: knitting the TMB fits in this article
# is heavy (it has OOM'd pkgdown before), so the chunks show the exact shipped
# syntax without being run at build time. The article documents the shipped
# Phase 1 (response mask) and Phase 2a (Gaussian fixed-covariate mi()) surface;
# do not flip these to eval=TRUE in the pkgdown build.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

Missing data in `gllvmTMB` has two meanings.
A **response cell** can be missing: an individual was measured but one trait
was not recorded, leaving a gap in the multivariate trait matrix.
A **predictor** can be missing: a covariate needed by the response model was
not observed for some rows.
The current implementation treats those two cases differently, and both differ
from the third case--measurement error--which is explicitly out of scope for
this layer (see Section 8).

The starting point is what *not* to do.
Listwise deletion discards every row with any gap, which in a multivariate
trait matrix can mean throwing away most of the data and biasing estimates
whenever missingness is not completely at random.
Mean-fill and single imputation invent values and then treat them as if they
were observed, understating uncertainty and distorting the covariance
structure that an ordination or a phylogenetic model is trying to estimate.
For structured ecological data--shared species means, phylogenetic covariance,
spatial fields, latent axes--these ad-hoc fixes are particularly damaging
because they sever exactly the correlations the model exists to exploit.
`gllvmTMB` instead keeps the missingness *inside* the likelihood: nothing is
deleted and nothing is invented, and the uncertainty that missingness creates
propagates through the same joint Hessian as every other parameter.

For missing response cells, `gllvmTMB` retains the original rows and evaluates
the observed-cell likelihood by masking the unobserved entries.
Masked cells contribute no direct response likelihood, but the fitted object
still knows where those cells were.
For a missing predictor, `gllvmTMB` requires an explicit predictor model
declared through `mi()` in the response formula and `impute = list(...)` in
the fit call; the first shipped route is one continuous Gaussian predictor with
a fixed-effect covariate model.
A missing predictor is a latent quantity integrated out by TMB's Laplace
approximation; this is frequentist full-information maximum likelihood (FIML),
not multiple imputation.

This is likelihood-based missing-data handling.
It is not Bayesian posterior imputation, not multiple imputation, and not an
EM interface.
For cases where those approaches are preferred--rich GNN phylogenetic
imputation, mixed trait types, tree uncertainty--the `pigauto` package
provides the complementary preprocessing path (Section 6).

### Status: what is shipped versus planned

This article documents the missing-data layer as it stands today and sketches
the planned extensions, so the reader can see the design as a whole.
Read the status tags literally:

- **SHIPPED (default, complete-case).** Under the default `miss_control()`
  (`response = "drop"`, `predictor = "fail"`, `engine = "laplace"`) the layer is
  an exact no-op: a missing *response* cell drops its row before the likelihood
  is built, and a missing *predictor* is an error. A complete-data fit is
  identical to one built before the layer existed.
- **SHIPPED (Phase 1, the response mask).** `miss_control(response =
  "include")` retains rows with a missing response cell, builds the
  `is_y_observed` mask so masked cells contribute zero to the likelihood, keeps
  original-row accounting in `fit$missing_data`, adjusts `nobs()` to the
  likelihood-contributing cells, and exposes `predict_missing()` for the masked
  response cells.
- **SHIPPED (Phase 2a, one Gaussian fixed-covariate missing predictor).** A
  single continuous, unit-level missing predictor declared with `mi(x)` plus
  `missing = miss_control(predictor = "model")` and a fixed-effect Gaussian
  covariate model supplied through `impute = list(x = x ~ z)` (bare-formula
  sugar) or `impute = list(x = impute_model(x ~ z, family = gaussian()))`. The
  missing value is a latent quantity integrated out by the Laplace
  approximation; `imputed()` returns its conditional mode (EBLUP) and standard
  error, and the registry lives at `fit$missing_data$predictors`.
- **PLANNED.** A grouped covariate random intercept in the predictor model
  (Phase 2b, in progress); species- or group-level broadcast of an imputed
  value (Phase 2c); a phylogenetic / spatial / relatedness covariate model for
  `mi()` (Phase 3); and non-Gaussian or categorical missing predictors
  (Phase 5). These routes are sketched below with their phase tags.

Code chunks for fitted models are kept `eval = FALSE`: they show the exact
syntax of the shipped surface, but knitting the TMB fits is intentionally
skipped here. Each section labels whether its API is shipped or planned.

## 1. What is fitted now

The default is conservative complete-case fitting: a missing response cell
drops its row, and a missing predictor is an error. The non-default routes are
engaged through `miss_control()`.

```{r what-is-fitted-default}
# Defaults: response = "drop", predictor = "fail", engine = "laplace".
# miss_control()
```

`missing = miss_control(response = "include")` retains rows that have a missing
response cell, masking the missing cells out of the likelihood. `missing =
miss_control(predictor = "model")` is used when the response formula contains a
supported `mi()` term and `impute` supplies the corresponding predictor model;
without a matching `mi()` term it errors rather than silently doing nothing.

The table below maps each missing-data case to its route and the boundary that
route stops at. The phase tag in the final column gives its status: Phase 1 and
Phase 2a are shipped; Phase 2b onward is planned.

| Missing-data case | Route | Boundary / phase |
| --- | --- | --- |
| Missing trait cell (any family) | `missing = miss_control(response = "include")` | All predictors, grouping variables, offsets, weights, and structured-effect inputs must be complete for that row (Phase 1, shipped). |
| One missing continuous predictor, fixed covariate model | `traits(y1 + y2 ~ mi(body_mass) + z)` with `impute = list(body_mass = body_mass ~ z)` | The `mi()` variable must be numeric; one Gaussian covariate model; no multiple missing predictors yet (Phase 2a, shipped). |
| One missing continuous predictor, grouped covariate model | As above with `impute = list(body_mass = body_mass ~ z + (1 | site))` | One grouped random intercept in the predictor model; independent of response random effects (Phase 2b, planned). |
| Species-level missing trait (broadcast to obs) | `impute = list(body_mass = body_mass ~ 1 + phylo(1 | species, tree = tree))` | Species-level value broadcast to observation rows; phylogenetic covariate model (Phase 2c / Phase 3, planned). |
| Non-Gaussian / categorical missing predictor | `impute_model(family = )` with a non-Gaussian family | Lognormal, Gamma, beta, Poisson, ordinal, factor -- later work (Phase 5, planned). |

The `gllvmTMB` response matrix is **multivariate**: a single fit can have many
trait columns and many species.
Missing cells are indexed at the (row, trait) level, not at the response-row
level as in a univariate model.

## 2. Missing response cells in `traits()`

For a multivariate response model, each (row, trait) cell is either observed or
missing.
The observed-cell likelihood for row `i` and trait `j` is:

```
is_y_observed[i, j] = 1  -->  full log p(y_{ij} | eta_{ij})
is_y_observed[i, j] = 0  -->  0  (cell masked; row kept for random-effect
                                   borrowing across traits and species)
```

A row with at least one observed trait cell is retained.
A row where every trait is missing is also retained for row accounting but
contributes zero direct response likelihood.

This example creates a small trait matrix with deliberate missing cells (kept
`eval = FALSE`; it shows the shipped syntax for `miss_control(response =
"include")`):

```{r missing-response-example}
# set.seed(1)
# n <- 20; p <- 3
# dat <- data.frame(
#   species = rep(paste0("sp", 1:n), 1),
#   x       = rnorm(n)
# )
# Y <- matrix(rnorm(n * p), n, p,
#             dimnames = list(NULL, c("trait1", "trait2", "trait3")))
# Y[3, 1] <- NA; Y[7, 2] <- NA; Y[12, 1] <- NA; Y[12, 3] <- NA  # deliberate gaps
#
# fit_resp <- gllvmTMB(
#   traits(trait1 + trait2 + trait3 ~ x),
#   family  = gaussian(),
#   data    = cbind(dat, Y),
#   missing = miss_control(response = "include")
# )
```

Cross-trait borrowing still operates on the retained rows because the latent
structure is shared across all cells in a row.
A row where trait 1 is missing but traits 2 and 3 are observed still
contributes to the species means and latent axes for all traits.

The analogy with `drmTMB` partial bivariate pairs (where one of `y1`/`y2` can
be observed) generalises here: an observed cell informs its own trait
mean/loadings, but cross-trait species correlation is identified only from
rows where both cells are co-observed.

## 3. Inspecting `fit$missing_data` and `nobs()`

Every fitted object that used `miss_control(response = "include")` or
`miss_control(predictor = "model")` carries a `fit$missing_data` slot with
row-accounting and diagnostics.

```{r inspect-metadata}
# nobs(fit_resp)                     # likelihood-contributing cells
# fit_resp$missing_data$counts       # see field table below
# fit_resp$missing_data$slice        # implementation slice, e.g. "Phase1-s2"
# head(fit_resp$missing_data$original_row)
# fit_resp$missing_data$observed_y   # logical mask over the model rows
```

The most useful fields are:

| Field | Meaning |
| --- | --- |
| `slice` | Implementation slice tag (the response mask ships as `"Phase1-s2"`). |
| `contract_version` | Shared-contract version with `drmTMB`, e.g. `"59-v1"`. |
| `response_policy` | Whether response rows/cells were dropped (`"drop"`) or masked (`"include"`). |
| `predictor_policy` | Whether missing predictors fail (`"fail"`) or are modelled (`"model"`). |
| `original_row` | Row indices in the user's original data frame. |
| `model_row` | Row indices in the retained model frame. |
| `observed_y` | Logical mask (`is_y_observed`) over the model rows. |
| `counts` | Named list: `retained_rows`, `observed_response`, `missing_response`, `likelihood_rows` (with `n_*` aliases). |
| `predictors` | Per-predictor registry for explicitly modelled `mi()` predictors (Phase 2a). |

For ordinary user work, prefer `nobs()`, `fitted()`, `residuals()`, and
`predict_missing()` over hand-parsing the slot.
The slot exists so row accounting stays auditable after fitting.

A co-observed cell count diagnostic is planned for the multivariate Gaussian
case: if very few rows have both traits observed, the cross-trait residual
correlation is weakly identified and `gllvmTMB` will warn.

## 4. `predict_missing()` for response cells

Missing response cells have predicted values from the fitted model because
their predictors and random effects are known (or estimated).
Use `predict_missing()` to retrieve those reconstructed values.

```{r predict-missing}
# pm <- predict_missing(fit_resp)            # type = "link" (default)
# pm <- predict_missing(fit_resp, type = "response")
# head(pm)
# # One row per masked cell. Columns: original_row, model_row, the unit/cluster
# # and trait identifier columns, and `est` (the prediction on the requested
# # scale). A complete-data fit returns a zero-row data frame.
```

The point prediction is the fitted linear predictor (`type = "link"`) or its
inverse-link conditional mean (`type = "response"`).
Standard errors of the reconstructed cells are deferred to a later slice, so
`predict_missing()` returns the point predictions only; they are model-based
fitted values, not posterior intervals and not multiple-imputation draws.

`predict_missing()` is for **response** cells only.
It is distinct from `imputed()`, which summarizes missing **predictors**
(Section 5).
The terminology follows the shared contract with `drmTMB`: predicted /
reconstructed responses versus EBLUP conditional modes for latent predictors.

Residuals for missing cells are `NA`; fitted values are not.
This is deliberate: a residual requires an observed value to compute, but a
fitted value does not.

## 5. Missing predictors: `mi()` and `impute_model(family=)`

Missing predictors are not handled by row deletion or sentinel imputation.
The response model needs the predictor value, so `gllvmTMB` asks for an
explicit model for that predictor.
The missing values become latent random effects integrated out by the Laplace
approximation, jointly with the response-model random effects.

The joint likelihood factorises as:

```
L(theta) = integral
  p(y_obs | x_obs, x_mis, b, theta_y)
  * p(x_obs, x_mis | b_x, theta_x)
  * p(b, b_x | theta_b)
  d x_mis db db_x
```

In `gllvmTMB`, `x_mis` is declared in the TMB `random` set and integrated out
by the Laplace approximation together with all other random effects.
`sdreport` then returns the conditional mode (EBLUP) and its standard error
from the inverse joint Hessian.

How the integral is taken depends on the type of the missing predictor, and
the two cases are handled differently:

- **Continuous predictors (Laplace).** A missing continuous covariate is a
  real-valued latent quantity. The Laplace approximation replaces its integral
  with the joint mode plus a Gaussian curvature term, exactly as for any other
  random effect. The reported value is its conditional mode (EBLUP). This is
  the continuous path; its shipped instance is the Gaussian fixed-covariate
  model of Phase 2a (Section 5a), with structured covariate models planned in
  Phases 2b-4.
- **Discrete predictors (exact finite-state summation).** A missing factor or
  ordinal covariate has finitely many possible values, so the "integral" is a
  finite sum: the likelihood is evaluated at each admissible level and weighted
  by the covariate model's category probabilities. No Gaussian approximation is
  used for the discrete dimension. This is planned for Phase 5 and is the
  reason categorical predictors are scheduled last.

### 5a. Continuous predictor, fixed covariate model (Phase 2a, shipped)

This is the shipped missing-predictor route. The missing predictor is declared
with `mi(x)` in the response formula; `missing = miss_control(predictor =
"model")` engages the latent-covariate grammar; and `impute` supplies its
Gaussian fixed-effect covariate model. A bare two-sided formula is sugar for a
Gaussian predictor model; `impute_model(x ~ z, family = gaussian())` makes the
family explicit (the family lives in the **predictor** model, not in `mi()`).
The predictor is treated as a unit-level quantity, so there is one latent value
per missing **unit**, broadcast across that unit's trait rows.

```{r missing-predictor-fixed}
# Tiny fixture: 2 traits, a few missing body_mass values.
# set.seed(2)
# n <- 24
# dat2 <- data.frame(
#   moisture   = rnorm(n),
#   body_mass  = rnorm(n),
#   species    = paste0("sp", 1:n)
# )
# Y2 <- cbind(
#   trait1 = 0.5 + 0.8 * dat2$body_mass + rnorm(n, sd = 0.2),
#   trait2 = -0.2 - 0.4 * dat2$moisture + rnorm(n, sd = 0.25)
# )
# dat2$body_mass[c(3, 11, 18)] <- NA  # three missing predictor values
#
# fit_pred <- gllvmTMB(
#   traits(trait1 + trait2 ~ mi(body_mass) + moisture),
#   family  = gaussian(),
#   data    = cbind(dat2, Y2),
#   impute  = list(body_mass = body_mass ~ moisture),  # bare = Gaussian sugar
#   # equivalently: impute = list(body_mass = impute_model(body_mass ~ moisture,
#   #                                                       family = gaussian())),
#   missing = miss_control(predictor = "model"),
#   control = gllvmTMBcontrol(se = TRUE)  # se = TRUE -> conditional SEs in imputed()
# )
#
# imputed(fit_pred)   # one EBLUP conditional mode + SE per missing unit
```

`imputed(fit_pred)` returns a data frame with one row per missing unit value
(not one per long-format model row), with columns `variable`, `original_row`,
`model_row`, `observed`, `estimate`, `std_error`, `source`, and
`uncertainty_status`.
The `estimate` column is the conditional mode (EBLUP) used in the fitted
likelihood, and `source` is `"conditional_mode"`.
The `std_error` column is the likelihood-based conditional standard error from
the joint Hessian; it is populated when the fit carries a successful `sdreport`
(`control = gllvmTMBcontrol(se = TRUE)`) and is `NA` otherwise, with
`uncertainty_status` recording which case applies.
It is not a posterior interval and is not a set of multiple-imputation draws.

`imputed()` also accepts `rows = "all"` to return every retained unit (observed
units carry their observed predictor value with `NA` standard error), and the
same conditional modes are stored at
`fit_pred$missing_data$predictors$body_mass$conditional_mode`.

### 5b. Predictor with grouped covariate model (Phase 2b, planned)

This route is **planned, not yet shipped**. A grouped covariate random
intercept in the predictor model is rejected today with a clear error. The
intended syntax adds one grouped random intercept to the `mi()` predictor model
when missingness is tied to a grouping level:

```{r missing-predictor-grouped}
# FILL: Phase 2b
#
# fit_grp <- gllvmTMB(
#   traits(trait1 + trait2 ~ mi(body_mass) + moisture),
#   family  = gaussian(),
#   data    = cbind(dat2, Y2),
#   impute  = list(body_mass = body_mass ~ moisture + (1 | site)),
#   missing = miss_control(predictor = "model")
# )
```

The grouped effect is a covariate-model random intercept, not a response-model
random effect.
It is kept independent of the response random effects by default (Level-1
independent default; Level-2 joint field is opt-in via
`correlate_with = "response"`, deferred to Phase 4).

### 5c. Species-level predictor with phylogenetic covariate model (Phase 3, planned)

This is the flagship case for ecology and is **planned, not yet shipped**: a
missing species-level trait used as a predictor, with its imputation model
inheriting phylogenetic structure (structured covariate models are rejected
today). It also requires the species-level broadcast (Phase 2c). Before
reaching for it, read the honesty caveat below: phylogenetic imputation helps
only when phylogenetic signal is strong.

```{r missing-predictor-phylo}
# FILL: Phase 3 -- enable once phylo covariate model is wired.
# Tiny fixture: n_species = 12, a minimal Brownian-motion tree, 3 missing
# body_mass values at species level broadcast to observation rows.
#
# fit_phy <- gllvmTMB(
#   traits(trait1 + trait2 ~ mi(body_mass) + moisture),
#   family  = gaussian(),
#   data    = cbind(dat2, Y2),
#   impute  = list(
#     body_mass = impute_model(
#       body_mass ~ 1 + phylo(1 | species, tree = tiny_tree),
#       family = gaussian()
#     )
#   ),
#   missing = miss_control(predictor = "model")
# )
#
# imputed(fit_phy)
```

The honest starting point: phylogenetic imputation is **signal-dependent and
not always better than complete-case analysis**.
When phylogenetic signal is weak, borrowing strength across related species
adds noise rather than information, and the imputed values can be no better
than -- or worse than -- simply analysing the complete cases
(Johnson et al. 2021; Molina-Venegas 2024).
Even the most accurate method tested by Johnson et al. (2021) gave inaccurate
estimates under strong missingness bias, and a "significant" signal test is not
the same as *strong* signal (Molina-Venegas 2024).
This is exactly why a phylo-signal gate is planned as a Phase 3 verification
check: it will warn when the estimated signal is too weak to trust the imputed
values, and the design's intended behaviour is to degrade gracefully toward an
independent (non-phylogenetic) covariate model when signal is weak.
Trait databases are also typically MNAR (missing not at random): rare species
with few measurements are systematically under-represented (Penone et al. 2014).
The current implementation does not correct for MNAR; a delta-adjustment
sensitivity analysis is a Phase 3+ gate, not a v1 argument.

### 5d. Multi-source and spatial covariate models (Phase 4, planned)

Planned, not yet shipped. Phase 4 adds spatial (`spatial()`),
relatedness-matrix (`relmat()`), and
animal-model (`animal()`) covariate structures, and enables the Level-2 joint
field where the predictor model explicitly shares a structured field with the
response model (with eigenvector-fixed-effect orthogonalisation to guard
against structured confounding; Dupont, Marques and Kneib 2023).

```{r missing-predictor-phase4}
# FILL: Phase 4 -- spatial, relmat, animal, multi-source, Level-2 joint field.
```

### 5e. Non-Gaussian and categorical predictors (Phase 5, planned)

Planned, not yet shipped. Passing a non-Gaussian family to `impute_model()`
errors today (only `gaussian()` and the bare-formula sugar are accepted).

```{r missing-predictor-nongaussian}
# FILL: Phase 5 -- lognormal/Gamma, beta/logit-normal, Poisson/NB,
# ordinal cumulative-logit, unordered categorical (factor).
# Mirrors drmTMB MD5 non-Gaussian extension.
```

## 6. FIML versus multiple imputation

The missing-data layer described above is a frequentist **full-information
maximum-likelihood (FIML)** approach: one joint model, fitted once, with the
missingness handled inside the likelihood. It is deliberately an
**alternative** to multiple imputation (MI), not a wrapper around it. The two
strategies answer the same question by different routes.

Multiple imputation runs in three steps. An imputation model **draws** several
plausible completed datasets; the **analysis** model is fitted to each; and the
results are **pooled** by Rubin's rules, which combine the within- and
between-imputation variance into a single standard error. The completed
datasets are an explicit ensemble, and the between-imputation spread is how MI
represents the uncertainty that missingness adds.

FIML never forms that ensemble. The missing values are estimated jointly with
every other parameter in a single fit: missing responses are masked and
reconstructed afterward, and missing predictors are latent quantities
integrated out by the Laplace approximation. The uncertainty that MI recovers
from between-imputation spread is recovered here from the curvature of the same
joint likelihood--the inverse Hessian that `sdreport` already returns. Under a
correctly specified model and data missing at random, FIML and proper MI are
asymptotically equivalent (Schafer and Graham 2002; Collins, Schafer and Kam
2001); FIML simply reaches the answer without a completed-dataset ensemble and
without a pooling step.
(This equivalence is established for the normal-model and GLM classes those
papers study; extending it to the latent-variable, phylogenetic, and spatial
joint models here is a reasonable expectation, not a transferred theorem.)

| | Multiple imputation (incl. `pigauto`) | FIML (this layer) |
| --- | --- | --- |
| Workflow | draw -> analyse -> pool (Rubin's rules) | one joint model, fitted once |
| Completed datasets | an explicit ensemble (e.g. `m = 50`) | none; nothing is filled in |
| Uncertainty | within- + between-imputation variance | inverse joint Hessian (`sdreport`) |
| Missing predictors | requires covariates complete before imputing | latent, integrated out inside the fit |
| Imputation model | separate from the analysis model | *is* the analysis model's structure |

### Congeniality is automatic here

A recurring difficulty in MI is **congeniality** (Meng 1994): the imputation
model and the analysis model are specified separately, and if the imputation
model omits structure the analysis model relies on--an interaction, a
nonlinearity, a phylogenetic covariance--the pooled inference is biased even
when each step is internally correct. A concrete instance is the
transform-then-impute hazard: imputing a predictor and a transformation of it
(a square, an interaction) under separate models yields mutually inconsistent
imputed values (von Hippel 2009). In the FIML layer this mismatch cannot arise
by construction: there is only one model. The "imputation model" for a missing
predictor is the covariate model declared in `impute`, and it lives in the same
likelihood as the response model, sharing its parameters and its structured
fields. The imputation and analysis models are congenial because they are the
same model.
This advantage holds because the analysis model here is fixed and known; when
the downstream analysis is unknown or varied, MI's separation of the two models
is a feature, which is why MI remains the right tool in that setting (below).

This is also why the covariate model is **level-aware** rather than a generic
default. A missing species-level trait is imputed with a phylogenetic
covariate model (`phylo()`); a missing site-level covariate is imputed with a
spatial one (`spatial()`); an animal-model or relatedness structure
(`animal()`, `relmat()`) or the shared latent axes can play the same role. The
imputation model inherits the biological covariance of the analysis it serves--
but it is declared explicitly per predictor, not silently copied from the
response model, so the inheritance is visible and auditable.

### The `pigauto` sister path

Multiple imputation remains available, and for some problems it is the better
tool. `pigauto`
([itchyshin.github.io/pigauto](https://itchyshin.github.io/pigauto/index.html))
is the standalone MI path: a phylogenetic trait-imputation engine--a
Brownian-motion baseline blended with a graph-neural-network component--that
produces completed datasets and propagates uncertainty by multiple imputation
and Rubin's rules (`impute()` -> `multi_impute()` -> `with_imputations()` ->
`pool_mi()`), with conformal intervals and tree-uncertainty support. It imputes
missing **traits** but requires the covariates to be fully observed, and it is
agnostic to the downstream model.

The two packages are complementary rather than competing:

- Prefer **`pigauto` (MI)** when you want rich GNN-based phylogenetic
  imputation, a mix of trait types in one imputation step, explicit
  tree-uncertainty propagation, or a completed dataset you can hand to any
  downstream model.
- Prefer **`gllvmTMB` (FIML)** when you want a single transparent joint model
  with no preprocessing step, when the missingness is in **predictors** (which
  `pigauto` does not handle), or when you want the missing-data uncertainty to
  propagate through the very same likelihood as the rest of the fit.

The interfaces are kept separate on purpose: there is no `engine = "mi"` inside
`miss_control()`, because `mi()` means a latent missing predictor *inside* the
model while MI means multiple imputation *outside* it. A documented
`pigauto`-then-`gllvmTMB` workflow (impute with `pigauto`, fit `gllvmTMB` on
each completed dataset, pool) is the bridge when both are wanted.

## 7. Current boundaries

The design deliberately keeps several features out of scope for the planned
Phase 1 through Phase 3, independent of which phases have shipped:

- EM, profile, and REML missing-data engines are not implemented.
  `engine = "laplace"` is the only v1 option; `"em"` and `"profile"` are
  reserved names that are not yet accepted.
- Multiple missing predictors (`mi(x1)` and `mi(x2)` in the same formula) are
  not implemented.
- Non-Gaussian, bounded, or categorical missing predictors (lognormal, Gamma,
  beta, Poisson, NB, ordinal, factor) are Phase 5+ work.
- Structured covariate models for non-Gaussian predictors (grouped or phylo
  binary/ordinal) are not implemented.
- Interactions inside `mi()`, splines inside `mi()`, and predictor-model
  random slopes are not implemented.
- `imputed()` does not summarize missing **responses**; use `predict_missing()`
  for that.
- `simulate_imputed()` is deferred to Phase 5: if shipped, it will mean
  Gaussian-approximation or parametric-bootstrap draws, not posterior draws.
- Simulated imputations, credible intervals, and multiple-imputation pooling
  are not implemented in this layer; see `pigauto` for the MI sister path.
- Dense known sampling-covariance matrices with partial multivariate response
  rows are not implemented.
- MNAR sensitivity, bootstrap-SE cross-checks, and phylo-signal reliability
  diagnostics are Phase 3+ verification gates, not v1 API arguments.

When you encounter one of these boundaries, treat the error as a design limit
rather than a failed fit.
For now, use the supported likelihood-based route, fall back to complete-case
fitting, or handle the unsupported missingness outside `gllvmTMB` with a
method whose assumptions you can state explicitly.

## 8. Non-goal: measurement error

Missing predictors in this layer treat the **observed** values of `x` as
exact.
Only `x_mis` (the unobserved values) is latent.

**Measurement error** -- where the true `x` is latent even when nominally
observed, requiring an observation model `x_obs | x_true` with known standard
errors, replicates, or a validation subsample -- is **out of scope** for
Phase 1 through Phase 3.
The missing-predictor registry and `fit$missing_data` slot are designed so a
later measurement-error lane can reuse the latent-covariate machinery (a
possible future `me(x, se = se_x)` argument); but v1 exposes only `mi(x)` for
missing predictors and the two interfaces will remain distinct.

## References

Core references for the statistical foundations:

- Rubin (1976) *Biometrika* 63:581--592. MAR validity and the ignorability
  conditions.
- Schafer & Graham (2002) *Psychological Methods* 7:147--177. ML and MI both
  recommended over case deletion / ad-hoc fill-in; the primary FIML-vs-MI
  anchor.
- Collins, Schafer & Kam (2001) *Psychological Methods* 6:330--351. ML and MI
  yield similar results when comparably implemented (the equivalence anchor).
- Meng (1994) *Statistical Science* 9:538--558. Congeniality of the imputation
  and analysis models.
- von Hippel (2009) *Sociological Methodology* 39:265--291. Transform-then-
  impute: separately imputing a transform and its components yields
  inconsistent values.
- Ibrahim, Chen, Lipsitz & Herring (2005) *JASA* 100:332--346. Missing
  covariates in GLMs via a covariate model inside the likelihood.
- Kristensen et al. (2016) *JSS* 70(5). TMB Laplace + automatic
  differentiation.
- Robinson (1991) *Statistical Science* 6:15--32. EBLUP and best prediction.
- Penone et al. (2014) *Methods Ecol. Evol.* 5:961--970. Phylogenetic
  imputation; trait data typically not missing at random.
- Johnson, Isaac, Paviolo & Gonzalez-Suarez (2021) *Global Ecology and
  Biogeography* 30:51--62. Imputation is signal/bias-dependent and not always
  better than complete-case analysis.
- Molina-Venegas et al. (2018) *Ecography* 41:1740--1749. Phylogenetic
  imputation accuracy varies among lineages.
- Molina-Venegas (2024) *Methods Ecol. Evol.* 15:456--463. Limits of
  phylogenetic imputation; a significant signal test is not strong signal.
- Dupont, Marques & Kneib (2023) arXiv:2309.16861. Structured confounding in
  spatial models.
- Wang, Edge, Schraiber & Pennell (2025 preprint, PMC -- not peer-reviewed).
  Phylogenetic analogue of structured confounding.
