---
title: "Mixed-family fits: latent-scale covariance and correlations"
tier: 3 # under-audit mixed-family extractor draft; visible as a Developer Note
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Mixed-family fits: latent-scale covariance and correlations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

> **Developer Note -- under audit.** This page is buildable and linked
> from Developer Notes so maintainers can audit the mixed-family
> extractor path. It is not a first-stop tutorial. Start with
> [Get started](gllvmTMB.html), [Response families](response-families.html),
> and [Covariance and correlation](covariance-correlation.html) before
> using this page.
>
> This page stays internal until the mixed-family article covers the
> broader response story: Gaussian, binomial, Poisson/NB,
> beta/proportion, and delta/hurdle cases (designed, route-only, and
> pending a convergence fix; see below). It is not the
> loading-constraint lane, and it should not be promoted before the
> binary lambda/JSDM article plan lands.

`gllvmTMB` fits **mixed-family** GLLVMs where each trait can have a
different response distribution within a single fit. The compact
fixture here uses Gaussian, binomial, and Poisson rows. The broader
mixed-response article still needs NB, beta/proportion, and explicit
delta/hurdle examples before this page becomes public. The
package's headline differentiator is that the latent-factor structure
is shared *across* traits even when the response distributions differ.
This article walks through fitting a mixed-family model and using the
extractors to recover the trait-covariance structure on a common,
comparable latent scale.

Scope boundary: **IN** are long-format mixed-family fits whose
families have a defined latent residual scale (MIX-01..MIX-09) and
report-ready extraction from fitted objects (DIA-13). **PARTIAL** are
profile and bootstrap uncertainty summaries on mixed-family targets
(CI-10 / EXT-04), and, as of the 2026-07-05 design decision, delta /
hurdle mixed-family latent-scale correlations (MIX-10): these are
now designed and route-only under a constraint (latent structure on
the positive/continuous submodel only; the occurrence submodel is
fixed-effects-only), but the live fit is not yet available pending a
convergence fix.

The broader family map is:

| Family group | This page does | Status for public teaching |
|---|---|---|
| Gaussian, binomial, Poisson | Fits all three together in the runnable fixture and extracts latent-scale Sigma, correlations, communality, and bootstrap Sigma examples | Internal worked extractor draft; rows `MIX-01`--`MIX-09` supply the current evidence |
| Negative binomial and beta/proportion | Shares the same one-part latent-residual logic, but this page does not yet include a runnable NB/beta fixture | Needs a broader mixed-response teaching pass before public promotion |
| Delta and hurdle families | Names the route-only boundary only | `MIX-10` is designed and route-only (positive-part latent scale, occurrence fixed-effects-only) but not yet shippable: a probe found latent delta-lognormal recovery landing on a convergence boundary |
| Mixture families | Not taught here | Remain outside this article until their validation rows move |

This article is long-format only because the teaching fixture uses
per-row `family_var` dispatch. A wide-data companion needs a separate
family-by-trait teaching fixture before it can be shown honestly.

## Reader path and scope

This internal draft answers one extractor question: after fitting a
mixed-family GLLVM, which summaries live on a common latent scale?

| Reader question | Model object | Code section | Readout |
|---|---|---|---|
| How does per-row family dispatch enter one fit? | `family = fixture$family_list` with `family_var = "family"` | `fixture`, `fixture-data`, `fit` | Gaussian, binomial, and Poisson rows in one long-format model |
| Is the fit in a usable diagnostic state? | randomized-quantile residual metadata | `diagnostic-status` | row-status counts and compact fit-health PASS/WARN counts |
| What is the mixed-family covariance target? | `extract_Sigma_table(..., link_residual = "auto")` | `sigma` | total latent-scale covariance rows with interval status |
| What correlations can a reader plot? | `plot_Sigma_heatmap()` | `corr` | point-estimate correlation heatmap including family-specific link residuals |
| Which interval paths are available but partial? | `extract_correlations()` and `bootstrap_Sigma()` | `corr-fisher`, `corr-bootstrap`, `bootstrap` | Fisher-z and bootstrap examples with CI-10 boundary |

Scope boundary by row: `MIX-01`--`MIX-09` cover the long-format engine and
extractor surfaces used here; `DIA-13` covers diagnostic metadata tables;
`EXT-01`, `EXT-05`, `EXT-06`, `EXT-13`, `EXT-18`, and `EXT-27` cover the
extractor and point-plot helpers this page demonstrates. `CI-10` remains
partial, so the interval examples are API demonstrations, not calibrated
mixed-family coverage proof. `MIX-10` is `partial`: delta and hurdle
mixed-family latent-scale correlations are now designed and route-only
(latent structure constrained to the positive/continuous submodel, with
the occurrence submodel fixed-effects-only), but the live fit is not yet
available -- a probe found latent delta-lognormal recovery landing on a
convergence boundary, so this remains a design decision pending a fix,
not a shipped capability.

## The setup

The package ships a small reproducible fixture for testing and
illustration: three traits, three families.

```{r setup}
library(gllvmTMB)
```

```{r fixture}
fixture <- gllvmTMB:::load_mixed_family_fixture(n_families = 3L)
str(fixture, max.level = 1)
```

The data is in long format — one row per `(site, trait)` pair —
with a `family` column that tells the engine which response
distribution applies to each row:

```{r fixture-data}
head(fixture$data[, c("site", "trait", "family", "value")], 6)
table(fixture$data$family)
```

trait_1 is Gaussian, trait_2 is binomial (binary 0/1), trait_3 is
Poisson (non-negative integers). The `family_list` attribute
`family_var = "family"` tells `gllvmTMB()` to dispatch per-row by
the `family` column.

## Fitting

A single `gllvmTMB()` call handles the compact mixed-family case. The
formula has a single shared rank-1 latent factor. Ordinary `latent()`
carries the per-trait $\boldsymbol\Psi$ residual by default, so the
implied unit-level covariance is the full ordinary decomposition
$\boldsymbol\Lambda\boldsymbol\Lambda^\top + \boldsymbol\Psi$
rather than the shared low-rank axis alone.

```{r fit, results='hide'}
fit <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 1),
  data   = fixture$data,
  trait  = "trait",
  family = fixture$family_list   # list(gaussian(), binomial(), poisson())
)
```

```{r fit-summary}
fit$opt$convergence            # 0 = converged
as.numeric(logLik(fit))
```

Each row of the engine evaluates the likelihood appropriate for its
own family — Gaussian rows use the Gaussian density, binomial rows
use the Bernoulli, Poisson rows use the Poisson — while sharing a
single rank-1 latent factor across traits.

Before interpreting the covariance summaries, check that the fit and
the residual rows are in the expected diagnostic state. The helper
`diagnostic_table()` exposes diagnostic metadata as plain data frames;
it does not compute a new test.

```{r diagnostic-status}
rq_resid <- residuals(fit, type = "randomized_quantile", seed = 1)
gllvmTMB::diagnostic_table(rq_resid, table = "row_status")
gllvmTMB::diagnostic_table(rq_resid, table = "fit_health_status")
```

The `unsupported_family` rows are the Gaussian rows, where randomized
quantile residuals are not needed because ordinary residuals are
already continuous. The `fit_health_status` table is the compact
PASS / WARN / FAIL count from `check_gllvmTMB(fit)`.

## Sigma on the latent scale

`extract_Sigma()` returns the trait-covariance matrix on the latent
scale. For this mixed-family fit, the fitted unit-level matrix is the
full ordinary decomposition that ordinary `latent()` carries by
default ($\boldsymbol\Lambda\boldsymbol\Lambda^\top + \boldsymbol\Psi$);
`link_residual = "auto"` adds the per-family **link residual** to the
diagonal when a total latent-scale correlation is requested. The link
residual is the variance of the
link function's underlying distribution (π²/3 for binomial-logit,
$\log(1 + 1/\mu)$ for Poisson-log, etc.). This puts all traits on a
common scale so the off-diagonal correlations are interpretable:

```{r sigma}
Sigma_rows <- extract_Sigma_table(
  fit,
  level = "unit",
  part = "total",
  measure = "covariance",
  entries = "unique",
  link_residual = "auto"
)
Sigma_rows[, c("trait_i", "trait_j", "estimate", "interval_status")]
```

The corresponding heatmap shows the implied correlation matrix. Cells
are point estimates on the same link-residual-adjusted scale:

```{r corr, fig.width = 5.8, fig.height = 4.8, fig.cap = "Mixed-family latent-scale trait correlations. Cells show point estimates from the total unit-level Sigma matrix after adding the family-specific link residual to each diagonal.", fig.alt = "Three by three heatmap of latent-scale correlations among one Gaussian trait, one binomial trait, and one Poisson trait. Trait 1 and trait 2 are moderately positively correlated, trait 1 and trait 3 are moderately negatively correlated, and trait 2 and trait 3 are near zero."}
R_rows <- extract_Sigma_table(
  fit,
  level = "unit",
  part = "total",
  measure = "correlation",
  entries = "all",
  link_residual = "auto"
)
plot_Sigma_heatmap(
  R_rows,
  label_digits = 2,
  title = "Mixed-family trait correlations",
  subtitle = "Point estimates include family-specific link residuals."
) +
  ggplot2::labs(caption = NULL)
```

By default `link_residual = "auto"` adds the per-family residual to
the diagonal. Passing `link_residual = "none"` returns the structural
latent-scale matrix
$\boldsymbol\Lambda\boldsymbol\Lambda^\top + \boldsymbol\Psi$
that ordinary `latent()` carries -- useful for comparing factor
structure across fits but not for interpreting correlations between
traits of different families.

## Correlations with uncertainty

`extract_correlations()` returns a tidy data frame of pairwise
correlations + confidence intervals. This article shows the two
routes that target the total mixed-family $\Sigma$ surface:

```{r corr-fisher}
extract_correlations(fit, tier = "unit", method = "fisher-z",
                     link_residual = "auto")
```

The default Fisher-z interval is fast and bounded in $[-1, 1]$.
Bootstrap intervals are slower, but they preserve the per-row family
dispatch in simulated refits:

```{r corr-bootstrap, results='hide'}
boot_ci <- extract_correlations(fit, tier = "unit", method = "bootstrap",
                                nsim = 30L, seed = 42L,
                                link_residual = "auto")
```

```{r corr-bootstrap-show}
boot_ci
```

Bootstrap and Fisher-z CIs target the same surface
(correlation on $\Sigma_\text{total}$). The
`method = "profile"` path operates on a different (rotation-
invariant) target — the rank-$d$ factor structure
$\Lambda\Lambda^\top$ — and addresses the latent factor
pattern rather than the observable-scale correlation; like the
other interval routes on this page it is a route-only
demonstration whose mixed-family coverage is uncalibrated
(CI-10, currently failing the gate), not yet a calibrated
inference route.

## Communality: how much variance is explained by the latent factor?

`extract_communality()` returns the proportion of each trait's
latent-scale variance attributable to the shared rank-$d$ factor --
the part of the ordinary decomposition not absorbed by the per-trait
$\boldsymbol\Psi$ residual that `latent()` carries by default:

```{r communality}
extract_communality(fit, level = "unit", link_residual = "auto")
```

For mixed-family fits, the denominator includes the per-family
link residual, so the partition is honest: non-Gaussian traits
have larger denominators (from the link residual) and therefore
typically smaller $H^2$.

## Bootstrap inference on the full Σ matrix

`bootstrap_Sigma()` re-fits the model on parametric-bootstrap
replicates of the data and returns percentile CIs on every cell
of $\Sigma$. The simulate step is family-aware: binomial rows
get fresh 0/1 draws, Poisson rows get fresh non-negative-integer
draws, etc.

```{r bootstrap, results='hide'}
boot <- bootstrap_Sigma(
  fit, n_boot = 20L, level = "unit", what = "Sigma",
  conf = 0.95, seed = 42L, progress = FALSE
)
```

```{r bootstrap-show}
extract_Sigma_table(boot, level = "unit", entries = "upper")
```

`boot$n_failed` records how many refits failed to converge; with
a well-identified fixture this should be small (a handful at most).

## What's covered today

This article uses the M1 validated machinery:

| Capability | Status (validation-debt register) |
|---|---|
| Mixed-family engine acceptance (`family = list(...)`) | covered (MIX-01, MIX-02) |
| `extract_Sigma()` on mixed-family | covered (MIX-03) |
| `extract_correlations()` Fisher-z + Wald | covered (MIX-04) |
| `extract_communality()` on mixed-family | covered (MIX-05) |
| `extract_repeatability()` on mixed-family | covered (MIX-06; M1.6 added the per-family `sigma2_d` to vW) |
| `extract_Omega()` cross-tier | covered (MIX-07) |
| `bootstrap_Sigma()` per-row family preserved | covered (MIX-08) |
| `simulate.gllvmTMB_multi()` family-aware draws | covered (MIS-05; 6 families) |
| Delta / hurdle mixed-family latent-scale correlations | partial (MIX-10; route-only design accepted, live fit pending a convergence fix) |

See the
[validation-debt register](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/35-validation-debt-register.md)
for the full row-by-row evidence.

## See also

- [Covariance and correlation](https://itchyshin.github.io/gllvmTMB/articles/covariance-correlation.html) — the conceptual primer on Σ vs correlation + when an `indep()` diagonal is needed.
- [Profile-likelihood CIs](https://itchyshin.github.io/gllvmTMB/articles/profile-likelihood-ci.html) — inference machinery for derived quantities.
- [Choose your model](https://itchyshin.github.io/gllvmTMB/articles/choose-your-model.html) — decision tree mapping data shapes to articles.

## References

- Nakagawa S, Schielzeth H (2010). *Repeatability for Gaussian and non-Gaussian data: a practical guide for biologists.* Biol. Rev. **85**: 935–956.
- Nakagawa S, Johnson PCD, Schielzeth H (2017). *The coefficient of determination $R^2$ and intra-class correlation coefficient from generalized linear mixed-effects models revisited and expanded.* J. R. Soc. Interface **14**: 20170213.
