---
title: "Mixed-family fits: latent-scale covariance and correlations"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Mixed-family fits: latent-scale covariance and correlations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

`gllvmTMB` fits **mixed-family** GLLVMs where each trait can have a
different response distribution (Gaussian + binomial + Poisson + …)
within a single fit. The package's headline differentiator is that
the latent-factor structure is shared *across* traits even when the
response distributions differ. This article walks through fitting a
mixed-family model and using the extractors to recover the
trait-covariance structure on a common, comparable latent scale.

## The setup

The package ships a small reproducible fixture for testing and
illustration: three traits, three families.

```{r setup}
library(gllvmTMB)
```

```{r fixture}
fixture <- gllvmTMB:::load_mixed_family_fixture(n_families = 3L)
str(fixture, max.level = 1)
```

The data is in long format — one row per `(site, trait)` pair —
with a `family` column that tells the engine which response
distribution applies to each row:

```{r fixture-data}
head(fixture$data[, c("site", "trait", "family", "value")], 6)
table(fixture$data$family)
```

trait_1 is Gaussian, trait_2 is binomial (binary 0/1), trait_3 is
Poisson (non-negative integers). The `family_list` attribute
`family_var = "family"` tells `gllvmTMB()` to dispatch per-row by
the `family` column.

## Fitting

A single `gllvmTMB()` call handles the mixed-family case:

```{r fit, results='hide'}
fit <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 1),
  data   = fixture$data,
  family = fixture$family_list   # list(gaussian(), binomial(), poisson())
)
```

```{r fit-summary}
fit$opt$convergence            # 0 = converged
as.numeric(logLik(fit))
```

Each row of the engine evaluates the likelihood appropriate for its
own family — Gaussian rows use the Gaussian density, binomial rows
use the Bernoulli, Poisson rows use the Poisson — while sharing a
single rank-1 latent factor across traits.

## Sigma on the latent scale

`extract_Sigma()` returns the trait-covariance matrix on the latent
scale. For a mixed-family fit, the diagonal includes the per-family
**link residual** — the variance of the link function's underlying
distribution (π²/3 for binomial-logit, $\log(1 + 1/\mu)$ for
Poisson-log, etc.). This puts all traits on a common scale so the
off-diagonal correlations are interpretable:

```{r sigma}
Sigma <- extract_Sigma(fit, level = "unit", part = "total",
                       link_residual = "auto")
Sigma$Sigma
```

The `$R` element gives the implied correlation matrix:

```{r corr}
round(Sigma$R, 3)
```

By default `link_residual = "auto"` adds the per-family residual to
the diagonal. Passing `link_residual = "none"` returns the bare
$\Lambda\Lambda^\top + \Psi$ matrix — useful for comparing factor
structure across fits but not for interpreting correlations between
traits of different families.

## Correlations with uncertainty

`extract_correlations()` returns a tidy data frame of pairwise
correlations + confidence intervals. Three CI methods are
available:

```{r corr-fisher}
extract_correlations(fit, tier = "unit", method = "fisher-z",
                     link_residual = "auto")
```

The default (Fisher-z) is fast + bounded in $[-1, 1]$. For
variance-component-style boundary cases, profile-likelihood CIs
are tighter:

```{r corr-bootstrap, results='hide'}
boot_ci <- extract_correlations(fit, tier = "unit", method = "bootstrap",
                                nsim = 30L, seed = 42L,
                                link_residual = "auto")
```

```{r corr-bootstrap-show}
boot_ci
```

Bootstrap and Fisher-z CIs target the same surface
(correlation on $\Sigma_\text{total}$). The
`method = "profile"` path operates on a different (rotation-
invariant) target — the rank-$d$ factor structure
$\Lambda\Lambda^\top$ — and is recommended when you need
inference on the latent factor pattern rather than on the
observable-scale correlation.

## Communality: how much variance is explained by the latent factor?

`extract_communality()` returns the proportion of each trait's
latent-scale variance attributable to the shared rank-$d$
factor:

```{r communality}
extract_communality(fit, level = "unit", link_residual = "auto")
```

For mixed-family fits, the denominator includes the per-family
link residual, so the partition is honest: non-Gaussian traits
have larger denominators (from the link residual) and therefore
typically smaller $H^2$.

## Bootstrap inference on the full Σ matrix

`bootstrap_Sigma()` re-fits the model on parametric-bootstrap
replicates of the data and returns percentile CIs on every cell
of $\Sigma$. The simulate step is family-aware: binomial rows
get fresh 0/1 draws, Poisson rows get fresh non-negative-integer
draws, etc.

```{r bootstrap, results='hide'}
boot <- bootstrap_Sigma(
  fit, n_boot = 20L, level = "unit", what = "Sigma",
  conf = 0.95, seed = 42L, progress = FALSE
)
```

```{r bootstrap-show}
boot$point_est$Sigma_unit
boot$ci_lower$Sigma_unit
boot$ci_upper$Sigma_unit
```

`boot$n_failed` records how many refits failed to converge; with
a well-identified fixture this should be small (a handful at most).

## What's covered today

This article uses the M1 validated machinery:

| Capability | Status (validation-debt register) |
|---|---|
| Mixed-family engine acceptance (`family = list(...)`) | covered (MIX-01, MIX-02) |
| `extract_Sigma()` on mixed-family | covered (MIX-03) |
| `extract_correlations()` Fisher-z + Wald | covered (MIX-04) |
| `extract_communality()` on mixed-family | covered (MIX-05) |
| `extract_repeatability()` on mixed-family | covered (MIX-06; M1.6 added the per-family `sigma2_d` to vW) |
| `extract_Omega()` cross-tier | covered (MIX-07) |
| `bootstrap_Sigma()` per-row family preserved | covered (MIX-08) |
| `simulate.gllvmTMB_multi()` family-aware draws | covered (MIS-05; 6 families) |

See the
[validation-debt register](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/35-validation-debt-register.md)
for the full row-by-row evidence.

## See also

- [Covariance and correlation](https://itchyshin.github.io/gllvmTMB/articles/covariance-correlation.html) — the conceptual primer on Σ vs correlation + when `unique()` is needed.
- [Profile-likelihood CIs](https://itchyshin.github.io/gllvmTMB/articles/profile-likelihood-ci.html) — inference machinery for derived quantities.
- [Choose your model](https://itchyshin.github.io/gllvmTMB/articles/choose-your-model.html) — decision tree mapping data shapes to articles.

## References

- Nakagawa S, Schielzeth H (2010). *Repeatability for Gaussian and non-Gaussian data: a practical guide for biologists.* Biol. Rev. **85**: 935–956.
- Nakagawa S, Johnson PCD, Schielzeth H (2017). *The coefficient of determination $R^2$ and intra-class correlation coefficient from generalized linear mixed-effects models revisited and expanded.* J. R. Soc. Interface **14**: 20170213.
