---
title: "How many latent dimensions should I fit?"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{How many latent dimensions should I fit?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6.4,
  fig.height = 4.2,
  fig.align  = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

Before interpreting a latent axis, choose the rank you are willing to
defend. This article asks a practical question: when the same
multi-trait data can be fitted with one latent dimension, two latent
dimensions, or a richer rank, how do AIC, BIC, fit health, and
biological interpretability help decide which model to report?

The example simulates five continuous traits with two planted latent
dimensions. We fit candidate ranks with `gllvmTMB()`, compare their
information criteria, check numerical health, and interpret the chosen
model as a rank-conditional summary rather than proof that nature has
exactly two hidden axes.

A safe reporting order is: first check numerical and response
diagnostics; next compare one ML candidate set with AIC/BIC; then
interpret rotation-invariant summaries for the chosen rank; finally,
for Gaussian fits only, refit that chosen model with REML for covariance
summaries. The REML log-likelihood is not part of the ML AIC/BIC table.

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

## Start With One Data Problem

The fixture has one row per `(individual, trait)` observation. The
traits are continuous measurements with a planted *size* axis and a
planted *shape* axis. The true between-individual covariance is

$$
\Sigma = \Lambda \Lambda^\top + \Psi,
$$

where `Lambda` contains the shared latent loadings and `Psi` is the
trait-specific diagonal from `unique()`.

```{r example-data, eval = FALSE}
rank_ex <- readRDS(system.file(
  "extdata", "examples", "model-selection-rank-example.rds",
  package = "gllvmTMB"
))
df <- rank_ex$data_long
df_wide <- rank_ex$data_wide
truth <- rank_ex$truth
```

```{r example-data-load, include = FALSE}
rank_path <- system.file(
  "extdata", "examples", "model-selection-rank-example.rds",
  package = "gllvmTMB"
)
rank_path <- c(
  rank_path,
  file.path("inst", "extdata", "examples", "model-selection-rank-example.rds"),
  file.path("..", "inst", "extdata", "examples", "model-selection-rank-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "model-selection-rank-example.rds")
)
rank_path <- rank_path[nzchar(rank_path) & file.exists(rank_path)][1]
stopifnot(!is.na(rank_path))
rank_ex <- readRDS(rank_path)

df <- rank_ex$data_long
df_wide <- rank_ex$data_wide
truth <- rank_ex$truth
trait_names <- truth$trait_names
```

```{r data-preview}
rank_ex$story$question
head(df, 8)
round(truth$Lambda, 2)
```

The planted `d = 2` truth is only a teaching fixture. In real data, the
rank is unknown; the candidate set is a set of defensible summaries, not
a direct view into biological machinery.

The long and wide versions of the selected-rank formula are:

```{r formula-preview}
rank_ex$formula_long
rank_ex$formula_wide
```

## Fit Candidate Ranks

The diagonal baseline has no shared latent axis, so it uses `indep()`
to make the standalone marginal-only intent explicit. The `d >= 1`
candidates use the same fixed effects and the same explicit `Psi`
component through `unique()`, changing only the rank of the `latent()`
term.

```{r rank-helpers}
candidate_formula <- function(d) {
  if (identical(d, 0L)) {
    return(value ~ 0 + trait + indep(0 + trait | individual))
  }

  stats::as.formula(paste0(
    "value ~ 0 + trait + ",
    "latent(0 + trait | individual, d = ", d, ") + ",
    "unique(0 + trait | individual)"
  ))
}

fit_candidate <- function(d, data, control) {
  suppressMessages(suppressWarnings(gllvmTMB(
    candidate_formula(d),
    data = data,
    trait = "trait",
    unit = "individual",
    family = gaussian(),
    control = control
  )))
}

summarise_candidate <- function(fit, d) {
  ll <- logLik(fit)
  health <- check_gllvmTMB(fit)
  weak_axis <- health[health$component == "weak_axis_unit", , drop = FALSE]
  health_status <- if (any(health$status == "FAIL")) {
    "FAIL"
  } else if (any(health$status == "WARN")) {
    "WARN"
  } else {
    "PASS"
  }

  data.frame(
    d = d,
    model = if (d == 0L) "diagonal indep" else paste0("latent d = ", d),
    logLik = as.numeric(ll),
    df = attr(ll, "df"),
    nobs = attr(ll, "nobs"),
    AIC = AIC(fit),
    BIC = BIC(fit),
    convergence = fit$opt$convergence,
    pd_hessian = isTRUE(fit$fit_health$pd_hessian),
    max_gradient = fit$fit_health$max_gradient,
    weak_axis = if (nrow(weak_axis)) weak_axis$status[1] else "not fitted",
    health = health_status,
    stringsAsFactors = FALSE
  )
}
```

```{r fit-candidates}
candidate_d <- 0:3
rank_control <- gllvmTMBcontrol(n_init = 2, init_jitter = 0.02)

fits <- setNames(
  lapply(candidate_d, fit_candidate, data = df, control = rank_control),
  paste0("d", candidate_d)
)

rank_table <- do.call(
  rbind,
  lapply(candidate_d, function(d) summarise_candidate(fits[[paste0("d", d)]], d))
)
rank_table$delta_AIC <- rank_table$AIC - min(rank_table$AIC)
rank_table$delta_BIC <- rank_table$BIC - min(rank_table$BIC)

rank_table_display <- rank_table[
  , c(
    "model", "logLik", "df", "nobs", "AIC", "BIC",
    "delta_AIC", "delta_BIC", "pd_hessian", "weak_axis", "health"
  )
]
rank_table_display[, c("logLik", "AIC", "BIC", "delta_AIC", "delta_BIC")] <-
  round(rank_table_display[, c("logLik", "AIC", "BIC", "delta_AIC", "delta_BIC")], 1)
rank_table_display

selected_d <- rank_table$d[which.min(rank_table$BIC)]
fit_selected <- fits[[paste0("d", selected_d)]]
```

Here both criteria support `d = 2`, the planted rank. The `d = 3`
model gains a tiny amount of log-likelihood, but it pays for extra
parameters and its fit-health rows warn about a weak latent axis. That
is the common rank-selection pattern: a richer model can fit slightly
better while adding an axis that is not stable enough to interpret.

## Check The Wide Formula Too

The same `d = 2` model can be written from a wide data frame with
`traits(...)` on the left-hand side. This is the compact form for users
who already have one row per individual and one column per trait.

```{r wide-fit}
fit_d2_wide <- gllvmTMB(
  rank_ex$formula_wide,
  data = df_wide,
  unit = rank_ex$fit_args$unit,
  family = rank_ex$fit_args$family,
  control = rank_control
)

signif(
  abs(as.numeric(logLik(fits$d2)) - as.numeric(logLik(fit_d2_wide))),
  3
)
```

That number should be close to zero. The long and wide calls describe
the same stacked-trait likelihood; only the data shape differs.

## Read AIC And BIC Beside Diagnostics

`AIC()` and `BIC()` work through `logLik.gllvmTMB_multi()`. The
log-likelihood is the TMB Laplace approximation to the marginal
likelihood, `df` is the number of free fitted parameters after TMB
mapping, and `nobs` is the number of observed response cells
contributing to the likelihood. Compare AIC/BIC only among fits with
the same response cells, family/link target, weights, and likelihood
approximation.

```{r ic-plot, fig.cap = "AIC and BIC differences for candidate latent ranks. Both criteria support the planted d = 2 model in this deterministic Gaussian fixture; the overfit d = 3 candidate improves log-likelihood only slightly and is penalised."}
ic_plot_data <- rbind(
  data.frame(d = rank_table$d, criterion = "AIC", delta = rank_table$delta_AIC),
  data.frame(d = rank_table$d, criterion = "BIC", delta = rank_table$delta_BIC)
)
ic_plot_data$rank_label <- factor(
  ifelse(ic_plot_data$d == 0L, "diagonal", paste0("d=", ic_plot_data$d)),
  levels = c("diagonal", "d=1", "d=2", "d=3")
)

ggplot(ic_plot_data, aes(rank_label, delta, group = criterion,
                         colour = criterion)) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey55") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c(AIC = "#0a617d", BIC = "#b0582b")) +
  labs(
    x = "Candidate rank",
    y = "Difference from best value",
    colour = "Criterion"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

Use the criteria as routing information after fit health, not instead
of fit health. A model with the lowest AIC/BIC but failed convergence,
a non-positive-definite Hessian, or a weak added axis is not ready for
biological interpretation.

```{r overfit-health}
health_d3 <- check_gllvmTMB(fits$d3)
health_cols <- intersect(
  c("component", "status", "value", "message", "action"),
  names(health_d3)
)
health_d3[
  health_d3$component %in% c(
    "optimizer_convergence",
    "max_gradient",
    "pd_hessian",
    "hessian_rank",
    "weak_axis_unit",
    "cross_loading_structure_unit"
  ),
  health_cols,
  drop = FALSE
]
```

If AIC and BIC disagree, treat the disagreement as a prompt to inspect
the fitted covariance, weak-axis rows, biological plausibility, and
prediction/residual diagnostics. Do not promote a higher rank only
because it shaves a small number of AIC units off a numerically weaker
fit. Once those checks point to a defensible Gaussian model, use a REML
refit only for the final covariance summary, not for the ML candidate
table.

## Interpret The Selected Model

After choosing a rank, interpret invariant summaries first. `Sigma`,
correlations, and communalities do not depend on how the latent axes
are rotated. Raw loading columns are useful for exploratory labels, but
with `d > 1` their signs and rotations need care.

```{r sigma-recovery}
Sigma_hat <- extract_Sigma(fit_selected, level = "unit", part = "total")$Sigma
Sigma_true <- truth$Sigma

ij <- which(upper.tri(Sigma_true, diag = TRUE), arr.ind = TRUE)
sigma_compare <- data.frame(
  entry = paste(rownames(Sigma_true)[ij[, 1]], colnames(Sigma_true)[ij[, 2]],
                sep = " - "),
  type = ifelse(ij[, 1] == ij[, 2], "variance", "covariance"),
  truth = Sigma_true[ij],
  estimate = Sigma_hat[ij],
  stringsAsFactors = FALSE
)

off <- upper.tri(Sigma_true)
data.frame(
  selected_d = selected_d,
  relative_Frobenius_error =
    norm(Sigma_hat - Sigma_true, "F") / norm(Sigma_true, "F"),
  off_diagonal_correlation = stats::cor(Sigma_hat[off], Sigma_true[off])
)
```

```{r sigma-scatter, fig.cap = "Truth-versus-fit covariance entries for the selected latent-rank model. This rotation-invariant comparison is safer than comparing raw loading columns before rotation or constraints."}
lims <- range(c(sigma_compare$truth, sigma_compare$estimate))

ggplot(sigma_compare, aes(truth, estimate, shape = type, colour = type)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey55", linewidth = 0.5) +
  geom_point(size = 2.6, alpha = 0.9) +
  coord_equal(xlim = lims, ylim = lims) +
  scale_colour_manual(values = c(variance = "#0a617d", covariance = "#8c4d8f")) +
  labs(
    x = "True Sigma entry",
    y = "Fitted Sigma entry",
    colour = "Entry",
    shape = "Entry"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

The selected rank recovers the planted covariance pattern in this
fixture. That statement is narrower than "AIC/BIC proved the true
rank." It means the selected fitted model is a useful summary of this
candidate set and this simulated data-generating mechanism.

## Refit The Selected Gaussian Model With REML

The rank comparison above uses maximum likelihood (ML), which is the
estimator used by `gllvmTMB()` unless `REML = TRUE` is supplied. After
choosing a Gaussian model to report, a restricted maximum-likelihood
(REML) refit is often the better variance-component summary because it
integrates over the fixed-effect coefficients before estimating the
covariance parameters.

```{r reml-selected}
fit_reml_long <- gllvmTMB(
  candidate_formula(selected_d),
  data = df,
  trait = "trait",
  unit = "individual",
  family = gaussian(),
  REML = TRUE,
  control = rank_control
)

fit_reml_wide <- gllvmTMB(
  rank_ex$formula_wide,
  data = df_wide,
  unit = rank_ex$fit_args$unit,
  family = rank_ex$fit_args$family,
  REML = TRUE,
  control = rank_control
)

data.frame(
  estimator = c("ML", "REML"),
  logLik = round(c(
    as.numeric(logLik(fit_selected)),
    as.numeric(logLik(fit_reml_long))
  ), 2),
  df = c(
    attr(logLik(fit_selected), "df"),
    attr(logLik(fit_reml_long), "df")
  ),
  nobs = c(
    attr(logLik(fit_selected), "nobs"),
    attr(logLik(fit_reml_long), "nobs")
  )
)

signif(
  abs(as.numeric(logLik(fit_reml_long)) -
        as.numeric(logLik(fit_reml_wide))),
  3
)
```

The ML and REML log-likelihoods answer different questions, so do not
mix them in the AIC/BIC table. Use one estimator consistently for a
candidate set, and report which estimator produced the final
covariance summaries. In the current release, `REML = TRUE` is guarded
to Gaussian fits without observation weights, retained missing response
cells, or `mi()` predictor models. Fixed-effect profile confidence
intervals are also ML-only; use Wald intervals or refit with
`REML = FALSE` before profiling fixed effects.

## What To Report

Report the candidate set and the checks that constrained it.

| Report item | Why it matters |
|---|---|
| Candidate ranks and formulas | AIC/BIC only compare the models you fit. |
| `logLik`, `df`, `nobs`, AIC, BIC, and deltas | Readers can see the likelihood gain and the penalty. |
| Estimator for candidate comparison and final covariance summaries | Use one ML candidate table; use REML only for the chosen Gaussian covariance summary. |
| `check_gllvmTMB()` rows | A low criterion value is not enough when Hessian, gradient, or weak-axis rows warn. |
| Rotation-invariant summaries | `Sigma`, correlations, and communalities are safer than raw loading labels. |
| Biological interpretation | A rank is useful only if the resulting summaries answer the study question. |

This article covers a model-selection workflow for Gaussian
`latent() + unique()` candidate ranks using base `AIC()` / `BIC()`,
`logLik()`, and `check_gllvmTMB()` (FG-04, FG-06, DIA-03, DIA-08,
DIA-10), then shows a Gaussian-only `REML = TRUE` refit for the chosen
rank (MIS-33).
It does not validate AIC/BIC as universally correct latent-rank
selectors, calibrate confidence intervals, compare likelihoods from
different fitting approximations, or advertise REML for non-Gaussian,
weighted, retained-missing-response, or `mi()` predictor fits.
For deeper Gaussian simulate-refit rank checks, use
`check_identifiability()`; for fitted response diagnostics, start with
[Can I trust this fit?](fit-diagnostics.html).

## References

* Akaike, H. (1974). A new look at the statistical model identification.
  *IEEE Transactions on Automatic Control* **19**, 716-723.
* Schwarz, G. (1978). Estimating the dimension of a model. *The Annals
  of Statistics* **6**, 461-464.
