---
title: "Individual morphometrics: the simplest GLLVM"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Individual morphometrics: the simplest GLLVM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

This is the simplest case the package handles, and it is the
foundation for everything else on the
[complexity ladder](../index.html#models-by-complexity). One
observational level (individuals), several continuous traits per
individual, and shared low-rank covariance between traits. No
phylogeny, no spatial structure, no nesting. This article walks
through that model end-to-end.
The working question is simple: if five body measurements are all
partly driven by size and shape, can a rank-2 GLLVM recover that
shared structure and still leave trait-specific variation on the
diagonal?

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
library(dplyr)
```

## Simulate

We pretend to measure $n = 150$ sparrows on five body traits. The
true loading matrix has a *size* axis (factor 1, all positive: bigger
birds tend to be bigger on every measurement) and a *shape* axis
(factor 2, mixed signs: long-and-skinny vs. short-and-stocky). Each
trait also has a small trait-specific unique variance $\psi_t^2$:

```{r sim}
set.seed(2025)
n_ind <- 150
trait_names <- c("length", "mass", "wing", "tarsus", "bill")
T <- length(trait_names)

# True loadings: factor 1 = "size", factor 2 = "shape"
Lambda_true <- matrix(c(
  0.9,  0.1,
  1.0,  0.2,
  0.8, -0.4,
  0.7, -0.5,
  0.6,  0.6
), nrow = T, ncol = 2, byrow = TRUE,
   dimnames = list(trait_names, c("size", "shape")))

# True trait-specific unique (residual) variances
psi2_true <- c(length = 0.15, mass = 0.10, wing = 0.20, tarsus = 0.12, bill = 0.18)

# Latent factors + trait-specific residuals (no global sigma_eps in truth)
u <- matrix(rnorm(n_ind * 2), n_ind, 2)
eps <- sapply(seq_len(T), function(t) rnorm(n_ind, sd = sqrt(psi2_true[t])))
Y <- u %*% t(Lambda_true) + eps
rownames(Y) <- as.character(seq_len(n_ind))
colnames(Y) <- trait_names

# Wide-format data: one row per individual, one column per trait
df_wide <- data.frame(
  individual = factor(seq_len(n_ind)),
  Y,
  check.names = FALSE
)

# Long-format data: one row per (individual, trait)
df <- data.frame(
  individual = factor(rep(seq_len(n_ind), each = T)),
  trait      = factor(rep(trait_names, n_ind), levels = trait_names),
  value      = as.vector(t(Y))
)
head(df, 8)
```

```{r sim-truth}
Sigma_true <- Lambda_true %*% t(Lambda_true) + diag(psi2_true)  # full between-individual covariance
round(Sigma_true, 2)
```

> **Pitfall: factor level order.** The line
> `factor(rep(trait_names, n_ind), levels = trait_names)` is critical.
> Without the explicit `levels = trait_names`, R sorts levels
> alphabetically (`bill, length, mass, tarsus, wing`), and the rows
> of fitted $\hat{\boldsymbol\Sigma}_B$ will not match the rows of
> your `Lambda_true`. The model is correct; the comparison is not.
> This is a subtle data-prep gotcha worth flagging because it makes
> "recovery" look broken when only the row labels are misaligned.

## Fit

The same model can be fitted from long or wide data. The long form is
canonical because it makes the `(individual, trait)` rows explicit;
the wide forms are the convenient equivalents for readers who already
have one row per individual and one column per trait.

```{r fit}
# Long format (canonical)
fit_long <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | individual, d = 2) +
          unique(0 + trait | individual),
  data = df,
  unit = "individual"      # tell the engine which column groups observations
)

# Wide data-frame format (same model with compact formula syntax)
fit_wide_formula <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~ 1 +
          latent(1 | individual, d = 2) +
          unique(1 | individual),
  data = df_wide,
  unit = "individual"
)

# Wide matrix format (same model for matrix-first workflows)
fit_wide_matrix <- gllvmTMB_wide(Y, d = 2, family = gaussian())

# The rest of the article uses the canonical long-format fit.
fit <- fit_long
fit

cat("Long/formula-wide logLik difference:",
    signif(abs(as.numeric(logLik(fit_long)) -
               as.numeric(logLik(fit_wide_formula))), 3),
    "\n")
cat("Long/matrix-wide logLik difference:",
    signif(abs(as.numeric(logLik(fit_long)) -
               as.numeric(logLik(fit_wide_matrix))), 3),
    "\n")
```

gllvmTMB defaults to `unit = "site"`, meaning it looks for a column
literally named `site` in the data. This morphometric data set has no
`site` column — the grouping factor is `individual`. Passing
`unit = "individual"` redirects the engine to the correct column;
everything else is standard GLLVM machinery.

`traits(length, mass, wing, tarsus, bill)` is the formula-level wide
data-frame marker. Because the response traits are named on the LHS,
the RHS can use the compact shorthand: `1` becomes the trait-specific
intercepts `0 + trait`, and `latent(1 | individual)` becomes
`latent(0 + trait | individual)`. `gllvmTMB_wide(Y, ...)` is the
matrix-in entry point for the same model when the response already
lives in a numeric matrix.

## What model did we fit?

For each individual $i = 1, \dots, n$ we observe a vector of trait
measurements $\mathbf{y}_i = (y_{i1}, \dots, y_{iT})^\top$ — for
example, body length, mass, wing chord, tarsus, bill depth in a
sparrow study; tail length, fin span, gill area in a fish study; or
six items on a psychometric scale in a behavioural study. The
reduced-rank GLLVM is

$$
y_{it} \;=\; \mu_t \;+\; \boldsymbol{\lambda}_t^{\!\top} \mathbf{u}_i \;+\;
\varepsilon_{it},
\qquad
\mathbf{u}_i \sim \mathcal{N}(\mathbf{0}, \mathbf{I}_d), \quad
\varepsilon_{it} \sim \mathcal{N}(0, \psi_{tt}),
$$

where $\boldsymbol{\lambda}_t \in \mathbb{R}^d$ is the loading vector
for trait $t$, $\mathbf{u}_i$ is a $d$-dimensional latent factor for
individual $i$, and $\varepsilon_{it}$ is the trait-specific residual
with variance $\psi_{tt}$. The `unique()` term collects those residual
variances on the diagonal of $\boldsymbol\Psi$. The implied
between-individual trait covariance is

$$
\boldsymbol{\Sigma}_B
\;=\;
\underbrace{\boldsymbol{\Lambda} \boldsymbol{\Lambda}^{\!\top}}_{\text{shared}}
\;+\;
\underbrace{\boldsymbol\Psi}_{\text{unique}},
\qquad \boldsymbol{\Lambda} \in \mathbb{R}^{T \times d}.
$$

For $T = 5$ traits and $d = 2$ factors this is still a small example;
the practical gain is interpretability, not a dramatic parameter
saving. As $T$ grows, the low-rank part scales with $T \cdot d$
loadings rather than the full $T(T + 1) / 2$ covariance surface. That
low-rank structure is the point of GLLVMs: shared biological axes
first, trait-specific variation second.

> **`+ unique(0 + trait | individual)` with cross-sectional data.**
> With cross-sectional data (one observation per individual per
> trait), the `(trait, individual)` pair uniquely identifies every
> row. gllvmTMB detects this automatically: it fixes
> $\sigma_\varepsilon$ to a negligibly small constant (and prints an
> `i Auto-suppressing sigma_eps: ...` message), so the `unique()` term
> absorbs the full observation-level residual and becomes the
> trait-specific residual variance $\psi_t^2$. The confounding that would
> arise if both $\sigma_\varepsilon$ and $\psi_t^2$ were estimated freely
> is therefore resolved by construction — you do not need to choose
> between them. With *longitudinal* data (multiple sessions per
> individual), both terms are separately identified and can be
> estimated without any auto-suppression; the `unique()` term then
> captures genuine between-individual trait-specific variance on top of
> the latent structure. See the
> [Covariance & correlation article](covariance-correlation.html)
> for the side-by-side comparison.

```{r fit-unique-s}
# Trait-specific unique standard deviations (psi_t)
psi_hat <- extract_Sigma(fit, level = "unit", part = "unique")$s
cat("Estimated psi_t:", round(psi_hat, 3), "\n")
cat("True      psi_t:", round(sqrt(psi2_true), 3), "\n")
```

## Recovered $\hat{\boldsymbol\Sigma}_B$ vs truth

```{r recover-once}
Sigma_hat <- extract_Sigma(fit, level = "unit")$Sigma
round(Sigma_hat, 2)

cat("Frobenius |Sigma_hat - Sigma_true| / |Sigma_true| =",
    round(norm(Sigma_hat - Sigma_true, "F") / norm(Sigma_true, "F"), 3), "\n")
```

A side-by-side correlation heatmap shows whether the recovered
trait correlations agree with truth at the entry level:

```{r corr-heatmap, fig.width = 7.5, fig.height = 3.5, fig.cap = "True (left) vs fitted (right) trait-pair correlations from Sigma_B = Lambda Lambda^T. The 1:1 agreement of the colours and labels is the visual signature of clean recovery."}
suppressPackageStartupMessages(library(ggplot2))
make_corr_long <- function(M, label) {
  R <- cov2cor(M)
  data.frame(
    trait_i = factor(rep(rownames(R), ncol(R)), levels = rownames(R)),
    trait_j = factor(rep(colnames(R), each = nrow(R)), levels = colnames(R)),
    rho     = as.vector(R),
    panel   = label
  )
}
rownames(Sigma_true) <- colnames(Sigma_true) <- trait_names
rownames(Sigma_hat)  <- colnames(Sigma_hat)  <- trait_names
df_corr <- rbind(make_corr_long(Sigma_true, "True"),
                 make_corr_long(Sigma_hat,  "Fitted"))
df_corr$panel <- factor(df_corr$panel, levels = c("True", "Fitted"))

ggplot(df_corr, aes(trait_j, trait_i, fill = rho)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 3.2) +
  scale_fill_gradient2(low = "#3b82c6", mid = "white", high = "#d33",
                       midpoint = 0, limits = c(-1, 1)) +
  facet_wrap(~ panel) +
  scale_y_discrete(limits = rev) +
  labs(x = NULL, y = NULL, fill = expression(rho)) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(),
        strip.text = element_text(face = "bold"))
```

> **What `extract_Sigma(level = "unit")$Sigma` returns.** The *total*
> between-unit covariance
> $\boldsymbol\Sigma_B = \boldsymbol\Lambda \boldsymbol\Lambda^{\!\top} + \boldsymbol\Psi$
> — the sum of the reduced-rank shared component and the diagonal
> unique component. If you fit without `+ unique()`, $\boldsymbol\Psi = 0$
> and you get only $\boldsymbol\Lambda \boldsymbol\Lambda^{\!\top}$,
> which understates the diagonal. Adding `+ unique()` (as in the fit
> above) fills in the diagonal correctly. Use `part = "shared"` /
> `part = "unique"` to pull just one component.

## Pairwise correlations and communality

For inference on individual cross-trait correlations or per-trait
communality, the canonical extractors return tidy data frames.
`extract_correlations()` defaults to `method = "fisher-z"` (fast
Fisher-transform Wald CI; bounds guaranteed in $[-1, 1]$).
Communality point estimates are enough for this first worked example.
When interval estimates matter, prefer profile or Wald-style intervals
where the target supports them. Bootstrap is a slower, deliberate option
for a separate cached check, not part of this rendered article.

```{r ci-correlations}
extract_correlations(fit, tier = "unit")
```

```{r ci-communality}
extract_communality(fit, level = "unit")
```

Communality $c_t^2 = (\boldsymbol\Lambda\boldsymbol\Lambda^{\!\top})_{tt} / \boldsymbol\Sigma_{B,tt}$
is the fraction of trait $t$'s variance shared with the others via
the latent factors; high $c_t^2$ means strongly integrated, low
$c_t^2$ means trait-specific variation dominates.

## Recovery across replicates

A single fit cannot tell you whether the model is recovering truth
or just one lucky draw. Repeat over `R = 20` independently simulated
datasets and look at the distribution of the bias. The loop uses the
same long/wide formula pair as above; it checks the wide equivalent on
the first replicate only, so the article does not double the cost of
every recovery fit.

```{r recover-replicates, fig.height = 4}
run_one <- function(seed) {
  set.seed(seed)
  u   <- matrix(rnorm(n_ind * 2), n_ind, 2)
  eps <- sapply(seq_len(T), function(t) rnorm(n_ind, sd = sqrt(psi2_true[t])))
  Y   <- u %*% t(Lambda_true) + eps
  colnames(Y) <- trait_names
  df_rep_wide <- data.frame(
    individual = factor(seq_len(n_ind)),
    Y,
    check.names = FALSE
  )
  df_rep <- data.frame(
    individual = factor(rep(seq_len(n_ind), each = T)),
    trait      = factor(rep(trait_names, n_ind), levels = trait_names),
    value      = as.vector(t(Y))
  )

  fit_rep_long <- suppressMessages(suppressWarnings(gllvmTMB(
    value ~ 0 + trait +
            latent(0 + trait | individual, d = 2) +
            unique(0 + trait | individual),
    data = df_rep, unit = "individual"
  )))
  if (identical(seed, 1L)) {
    fit_rep_wide <- suppressMessages(suppressWarnings(gllvmTMB(
      traits(length, mass, wing, tarsus, bill) ~ 1 +
              latent(1 | individual, d = 2) +
              unique(1 | individual),
      data = df_rep_wide, unit = "individual"
    )))
    stopifnot(
      abs(as.numeric(logLik(fit_rep_long)) -
            as.numeric(logLik(fit_rep_wide))) < 1e-6
    )
  }
  Sigma_hat_rep <- extract_Sigma(fit_rep_long, level = "unit")$Sigma
  data.frame(
    seed = seed,
    entry = paste0("[", rep(1:T, each = T), ",", rep(1:T, T), "]"),
    true = as.vector(Sigma_true),
    fitted = as.vector(Sigma_hat_rep)
  )
}
rep_df <- do.call(rbind, lapply(1:20, run_one))

ggplot(rep_df, aes(true, fitted)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey40", linetype = 2) +
  geom_point(alpha = 0.4, colour = "#1f78b4", size = 1.6) +
  labs(x = expression("True"~Sigma[B]~"entry"),
       y = expression("Fitted"~hat(Sigma)[B]~"entry"),
       title = "Recovery on 20 replicate datasets, n = 150 individuals") +
  theme_minimal(base_size = 11)
```

```{r recover-summary}
rep_df |>
  mutate(off_diag = sub("\\[(.),(.)\\]", "\\1\\2", entry),
         off_diag = substr(off_diag, 1, 1) != substr(off_diag, 2, 2)) |>
  group_by(off_diag) |>
  summarise(
    bias = mean(fitted - true),
    rmse = sqrt(mean((fitted - true)^2)),
    .groups = "drop"
  )
```

The off-diagonal entries (the cross-trait covariances — the *biology*)
are recovered with bias close to zero. Diagonal entries are slightly
under at $n = 150$, which is the expected $O(n^{-1/2})$ finite-sample
behaviour and shrinks as $n$ grows.

## Why `latent + unique`? `dep` and `indep` on the same data

The morphometric workflow pairs `latent(d = K)` with `unique()` to
estimate $\boldsymbol\Sigma_B = \boldsymbol\Lambda\boldsymbol\Lambda^\top + \boldsymbol{\Psi}$.
Two alternative covstruct modes — `dep` (full unstructured $\boldsymbol\Sigma_B$
with $T(T+1)/2$ free entries) and `indep` (diagonal-only
$\boldsymbol\Sigma_B = \mathrm{diag}(\sigma^2_1, \dots, \sigma^2_T)$, no
cross-trait modelling) — are also valid keywords. Refit the same data
under each to see what the user actually loses or gains.

```{r dep-indep-compare, message = FALSE, warning = FALSE}
fit_full_long <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | individual, d = T),
  data = df, unit = "individual"
)
fit_full_wide <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~ 1 +
          latent(1 | individual, d = T),
  data = df_wide, unit = "individual"
)

fit_dep_long <- gllvmTMB(
  value ~ 0 + trait +
          dep(0 + trait | individual),
  data = df, unit = "individual"
)
fit_dep_wide <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~ 1 +
          dep(1 | individual),
  data = df_wide, unit = "individual"
)

fit_indep_long <- gllvmTMB(
  value ~ 0 + trait +
          indep(0 + trait | individual),
  data = df, unit = "individual"
)
fit_indep_wide <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~ 1 +
          indep(1 | individual),
  data = df_wide, unit = "individual"
)

fit_full  <- fit_full_long
fit_dep   <- fit_dep_long
fit_indep <- fit_indep_long

cat("Max long/wide logLik difference across comparison fits:",
    signif(max(abs(c(
      as.numeric(logLik(fit_full_long)) - as.numeric(logLik(fit_full_wide)),
      as.numeric(logLik(fit_dep_long)) - as.numeric(logLik(fit_dep_wide)),
      as.numeric(logLik(fit_indep_long)) - as.numeric(logLik(fit_indep_wide))
    ))), 3),
    "\n")

S1 <- extract_Sigma(fit,       level = "unit")$Sigma
S2 <- extract_Sigma(fit_full,  level = "unit")$Sigma
S3 <- extract_Sigma(fit_dep,   level = "unit")$Sigma
S4 <- extract_Sigma(fit_indep, level = "unit")$Sigma

off       <- function(M) M[upper.tri(M)]
truth_off <- off(Sigma_true)
ll        <- function(f) round(as.numeric(logLik(f)), 2)

data.frame(
  mode               = c("latent(d=2) + unique", "latent(d=T) only",
                         "dep", "indep"),
  logL               = c(ll(fit), ll(fit_full), ll(fit_dep), ll(fit_indep)),
  off_diag_corr_truth = c(round(cor(off(S1), truth_off), 3),
                          round(cor(off(S2), truth_off), 3),
                          round(cor(off(S3), truth_off), 3),
                          NA),
  diag_min_max        = c(sprintf("%.2f - %.2f", min(diag(S1)), max(diag(S1))),
                          sprintf("%.2f - %.2f", min(diag(S2)), max(diag(S2))),
                          sprintf("%.2f - %.2f", min(diag(S3)), max(diag(S3))),
                          sprintf("%.2f - %.2f", min(diag(S4)), max(diag(S4))))
)
```

What this shows:

- **`dep` is mathematically identical to `latent(d = T)` standalone** —
  same logL, same $\boldsymbol\Sigma_B$ to numerical precision. The two
  keywords are aliases of the saturated unstructured fit.
- **Reduced-rank `latent(d = 2) + unique` matches the saturated fit
  within 0.1 nats of logL.** With $T = 5$ traits and a
  rank-2 truth, the rank constraint loses essentially no information
  and gives the same off-diagonal correlation pattern as the saturated
  fit ($r$ vs truth ≈ 0.95 for both).
- **`indep` recovers the diagonals fine** — Gaussian gives per-trait
  identifiability for the marginal variance even with no cross-trait
  modelling — but **loses every off-diagonal**. The off-diagonal
  correlation with truth is `NA` because the fitted off-diagonals are
  exactly zero by construction. logL is materially worse.

This is the **Gaussian counterpart** to the binary comparison in
[Joint species distribution](joint-sdm.html). The qualitative ordering
is the same — `latent + unique` ≥ `dep` ≫ `indep` in modelling
flexibility — but the quantitative gap between `indep` and the others
is much smaller for Gaussian (a few hundred nats here, vs `indep` being
*pinned at zero diagonals* for binary 1-obs-per-cell). Pick `latent + unique`
when you have a strong rank-$K$ hypothesis (size + shape, here);
fall back to `dep` if you suspect $K \approx T$ and want the saturated
fit; reach for `indep` only when you genuinely believe traits are
mutually independent — a rare default for ecological / morphological
data.

## What the latent factors mean

Every factor model has a rotational ambiguity: $\boldsymbol\Lambda$ and
$\boldsymbol\Lambda \mathbf{Q}$ produce the same $\boldsymbol\Sigma_B$
for any orthogonal $\mathbf{Q}$. To compare loadings between fits or
between packages, rotate to a standard convention (varimax for
interpretability, or lower-triangular for a unique fit):

```{r rotate}
L_hat <- extract_ordination(fit, level = "unit")$loadings
rot   <- rotate_loadings(fit, level = "unit", method = "varimax")
round(rot$Lambda, 2)
```

```{r ordi, fig.height = 5}
ord <- extract_ordination(fit, level = "unit")
head(ord$scores)
plot(fit, type = "ordination", level = "unit")
```

The biplot shows the trait loadings on the two latent axes. In a
real morphometric study, the first axis usually picks up *overall
size*, with all traits loading positively — the famous "general
factor" of body size. The second axis picks up shape contrasts.

## What this article does *not* cover

This is the bottom rung of the ladder. The next rungs add structure:

| Rung | New piece | Article |
|---|---|---|
| 1 | Discrete (binary) responses, individuals as sites | [Joint species distribution](joint-sdm.html) |
| 2 | Phylogenetic dependence between rows | [Phylogenetic GLLVM](phylogenetic-gllvm.html) |
| 3 | Spatial dependence between rows | `spatial_*()` keywords (see reference index); article in preparation |
| 4 | Two observational levels (between- and within-unit) | [Behavioural syndromes](behavioural-syndromes.html) |
| 5 | All of the above at once | [Functional biogeography](functional-biogeography.html) |

Read the rungs in order if you are new to GLLVMs. Jump straight to
the rung that matches your data if you are not.

## See also

* [Lambda constraints](lambda-constraint.html) — pin specific entries
  of $\boldsymbol\Lambda$ when you have prior knowledge of the
  factor structure (for example, a confirmatory factor analysis).
