---
title: "Individual morphometrics: the simplest GLLVM"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Individual morphometrics: the simplest GLLVM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

This is the simplest case the package handles, and it is the
foundation for everything else on the
[complexity ladder](../index.html#models-by-complexity). One
observational level (individuals), several continuous traits per
individual, and shared low-rank covariance between traits. No
phylogeny, no spatial structure, no nesting. This article walks
through that model end-to-end.

`gllvmTMB` is experimental and pre-CRAN (version 0.5.0): the surface may
change, and the recovery numbers and intervals below are illustrative,
not validated coverage.

The working question is simple: if five body measurements are all
partly driven by size and shape, can a rank-2 GLLVM recover that
shared structure and still leave trait-specific variation on the
diagonal?

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
library(dplyr)
```

## Example Data

We use a prepared example object rather than asking you to read the
data-generating script first. The script is still in the repository
(`data-raw/examples/make-morphometrics-example.R`) so the fixture is
reproducible, but the article starts where a user starts: with data,
truth, formulas, and one biological question.

```{r example-data, eval = FALSE}
morph <- readRDS(system.file(
  "extdata", "examples", "morphometrics-example.rds",
  package = "gllvmTMB"
))
df <- morph$data_long
df_wide <- morph$data_wide
truth <- morph$truth
estimands <- morph$estimands
```

```{r example-data-load, include = FALSE}
example_path <- system.file(
  "extdata", "examples", "morphometrics-example.rds",
  package = "gllvmTMB"
)
example_path <- c(
  file.path("inst", "extdata", "examples", "morphometrics-example.rds"),
  file.path("..", "inst", "extdata", "examples", "morphometrics-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "morphometrics-example.rds"),
  example_path
)
example_path <- example_path[nzchar(example_path) & file.exists(example_path)][1]
stopifnot(!is.na(example_path))
morph <- readRDS(example_path)

boot_r_path <- system.file(
  "extdata", "examples", "morphometrics-bootstrap-r.rds",
  package = "gllvmTMB"
)
boot_r_path <- c(
  file.path("inst", "extdata", "examples", "morphometrics-bootstrap-r.rds"),
  file.path("..", "inst", "extdata", "examples", "morphometrics-bootstrap-r.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "morphometrics-bootstrap-r.rds"),
  boot_r_path
)
boot_r_path <- boot_r_path[nzchar(boot_r_path) & file.exists(boot_r_path)][1]
stopifnot(!is.na(boot_r_path))
morph_boot_R <- readRDS(boot_r_path)

df <- morph$data_long
df_wide <- morph$data_wide
truth <- morph$truth
estimands <- morph$estimands

n_ind <- truth$n_individuals
trait_names <- truth$trait_names
T <- length(trait_names)
Lambda_true <- truth$Lambda
psi_true <- truth$psi
Sigma_true <- truth$Sigma
```

```{r example-data-preview}
morph$story$question
head(df, 8)
```

The object represents `r n_ind` individual birds measured on five
traits: body length, mass, wing chord, tarsus length, and bill depth.
The true loading matrix has a *size* axis (factor 1, all positive:
bigger birds tend to be bigger on every measurement) and a *shape*
axis (factor 2, mixed signs: long-and-skinny vs. short-and-stocky).
Each trait also has a trait-specific diagonal variance $\psi_t$.

```{r sim-truth}
round(Sigma_true, 2)
estimands
```

> **Pitfall: factor level order.** The example object pins
> `levels(df$trait)` to `truth$trait_names`. Without explicit levels,
> R may sort levels alphabetically (`bill, length, mass, tarsus,
> wing`), and the rows of fitted $\hat{\boldsymbol\Sigma}_B$ will not
> match the rows of the truth table. The model is correct; the
> comparison is not.

## Fit

The same model can be fitted from long or wide data. The long form is
canonical because it makes the `(individual, trait)` rows explicit;
the wide forms are the convenient equivalents for readers who already
have one row per individual and one column per trait.

```{r fit}
# The object stores the exact long and wide formulas used below:
morph$formula_long
morph$formula_wide

# Long format (canonical)
fit_long <- gllvmTMB(
  morph$formula_long,
  data = df,
  trait = morph$fit_args$trait,
  unit = morph$fit_args$unit,
  family = morph$fit_args$family
)

# Wide data-frame format (same model with compact formula syntax)
fit_wide_formula <- gllvmTMB(
  morph$formula_wide,
  data = df_wide,
  unit = morph$fit_args$unit,
  family = morph$fit_args$family
)

# The rest of the article uses the canonical long-format fit.
fit <- fit_long
as.numeric(logLik(fit))

cat("Long/formula-wide logLik difference:",
    signif(abs(as.numeric(logLik(fit_long)) -
               as.numeric(logLik(fit_wide_formula))), 3),
    "\n")
```

gllvmTMB defaults to `unit = "site"`, meaning it looks for a column
literally named `site` in the data. This morphometric data set has no
`site` column — the grouping factor is `individual`. Passing
`unit = "individual"` redirects the engine to the correct column;
everything else is standard GLLVM machinery.

`traits(length, mass, wing, tarsus, bill)` is the formula-level wide
data-frame marker. Because the response traits are named on the LHS,
the RHS can use the compact shorthand: `1` becomes the trait-specific
intercepts `0 + trait`, and `latent(1 | individual)` becomes
`latent(0 + trait | individual)`.

## What model did we fit?

For each individual $i = 1, \dots, n$ we observe a vector of trait
measurements $\mathbf{y}_i = (y_{i1}, \dots, y_{iT})^\top$ — for
example, body length, mass, wing chord, tarsus, bill depth in a
sparrow study; tail length, fin span, gill area in a fish study; or
six items on a psychometric scale in a behavioural study. The
reduced-rank GLLVM is

$$
y_{it} \;=\; \mu_t \;+\; \boldsymbol{\lambda}_t^{\!\top} \mathbf{u}_i \;+\;
\varepsilon_{it},
\qquad
\mathbf{u}_i \sim \mathcal{N}(\mathbf{0}, \mathbf{I}_d), \quad
\varepsilon_{it} \sim \mathcal{N}(0, \psi_{tt}),
$$

where $\boldsymbol{\lambda}_t \in \mathbb{R}^d$ is the loading vector
for trait $t$, $\mathbf{u}_i$ is a $d$-dimensional latent factor for
individual $i$, and $\varepsilon_{it}$ is the trait-specific residual
with variance $\psi_{tt}$. Ordinary `latent()` carries those
residual variances on the diagonal of $\boldsymbol\Psi$ by default. The
implied
between-individual trait covariance is

$$
\boldsymbol{\Sigma}_B
\;=\;
\underbrace{\boldsymbol{\Lambda} \boldsymbol{\Lambda}^{\!\top}}_{\text{shared}}
\;+\;
\underbrace{\boldsymbol\Psi}_{\text{unique}},
\qquad \boldsymbol{\Lambda} \in \mathbb{R}^{T \times d}.
$$

For $T = 5$ traits and $d = 2$ factors this is still a small example;
the practical gain is interpretability, not a dramatic parameter
saving. As $T$ grows, the low-rank part scales with $T \cdot d$
loadings rather than the full $T(T + 1) / 2$ covariance surface. That
low-rank structure is the point of GLLVMs: shared biological axes
first, trait-specific variation second.

> **Diagonal $\boldsymbol\Psi$ with cross-sectional data.**
> With cross-sectional data (one observation per individual per
> trait), the `(trait, individual)` pair uniquely identifies every
> row. gllvmTMB detects this automatically: it fixes
> $\sigma_\varepsilon$ to a negligibly small constant (and prints an
> `i Auto-suppressing sigma_eps: ...` message), so the diagonal
> $\boldsymbol\Psi$ term absorbs the full observation-level residual and
> becomes the trait-specific residual variance $\psi_t$. The confounding
> that would arise if both $\sigma_\varepsilon$ and $\psi_t$ were estimated
> freely is therefore resolved by construction — you do not need to add
> a separate diagonal term for this default latent model. With *longitudinal* data
> (multiple sessions per individual), both terms are separately identified
> and can be estimated without any auto-suppression; the diagonal
> $\boldsymbol\Psi$ term then captures genuine between-individual
> trait-specific variance on top of the latent structure. See the
> [Covariance & correlation article](covariance-correlation.html)
> for the side-by-side comparison.

```{r fit-psi-s}
# Trait-specific unique variances (psi_t)
psi_hat <- extract_Sigma(fit, level = "unit", part = "unique")$s
cat("Estimated psi_t:", round(psi_hat, 3), "\n")
cat("True      psi_t:", round(psi_true, 3), "\n")
```

The extractor stores the diagonal entries in a vector named `s`. In
the text we call the same entries `psi_t`, and the diagonal matrix they
form is `Psi`.

## Recovered $\hat{\boldsymbol\Sigma}_B$ vs truth

The numbers below describe this one simulated draw of `r n_ind`
individuals — descriptive recovery on a single dataset, not repeated
sampling (see the scope note further down).

```{r recover-once}
Sigma_hat <- extract_Sigma(fit, level = "unit")$Sigma
round(Sigma_hat, 2)

cat("Frobenius |Sigma_hat - Sigma_true| / |Sigma_true| =",
    round(norm(Sigma_hat - Sigma_true, "F") / norm(Sigma_true, "F"), 3), "\n")
```

A row-level comparison shows whether the recovered trait correlations
agree with truth. Zero means the fitted correlation equals the known
simulation truth for that trait pair:

```{r corr-comparison, fig.width = 7, fig.height = 4.8, fig.cap = "Trait-pair correlation errors from the total between-unit covariance Sigma_B = Lambda Lambda^T + Psi. Points show fitted minus true correlation; zero means exact recovery for that pair."}
rownames(Sigma_true) <- colnames(Sigma_true) <- trait_names
rownames(Sigma_hat)  <- colnames(Sigma_hat)  <- trait_names
corr_compare <- compare_Sigma_table(
  fit,
  truth = Sigma_true,
  level = "unit",
  measure = "correlation",
  entries = "upper"
)
plot_Sigma_comparison(corr_compare, measure = "correlation")
```

> **What `extract_Sigma(level = "unit")$Sigma` returns.** The *total*
> between-unit covariance
> $\boldsymbol\Sigma_B = \boldsymbol\Lambda \boldsymbol\Lambda^{\!\top} + \boldsymbol\Psi$
> — the sum of the reduced-rank shared component and the diagonal
> unique component. If you request `latent(..., unique = FALSE)`,
> $\boldsymbol\Psi = 0$ and you get only
> $\boldsymbol\Lambda \boldsymbol\Lambda^{\!\top}$, which understates the
> diagonal for this data-generating process. The default `latent()` fit
> above includes $\boldsymbol\Psi$. Use `part = "shared"` / `part = "unique"`
> to pull just one component.

## Pairwise correlations and communality

For inference on individual cross-trait correlations or per-trait
communality, the canonical extractors return tidy data frames.
`extract_correlations()` defaults to `method = "fisher-z"` (fast
Fisher-transform Wald CI; bounds guaranteed in $[-1, 1]$). These are
nominal intervals — their empirical coverage is not calibrated for this
model, so read them as recovery-grade illustration, not validated
coverage.
The table is useful for reporting exact numbers; the plot shows which
trait pairs have the strongest evidence of positive or negative
association. Communality point estimates are enough for this first
worked example. When interval estimates matter, prefer profile or
Wald-style intervals where the target supports them. Bootstrap is a
slower, deliberate option for a separate cached check. This article
loads a small cached `bootstrap_Sigma()` object for the correlation
figure so pkgdown does not refit bootstrap replicates while rendering.

```{r ci-correlations}
corr_rows <- extract_correlations(fit, tier = "unit")
corr_rows
```

```{r ci-correlation-eye, fig.width = 7, fig.height = 4.3, fig.cap = "Pairwise between-individual trait correlations from a cached morphometrics bootstrap fixture. Confidence eyes reconstruct frequentist compatibility from percentile intervals; hollow circles mark estimates. Use a larger bootstrap run before making a final interval-calibration claim."}
plot_correlations(morph_boot_R, tier = "unit", style = "eye", sort = "trait")
```

In this teaching fit, all unit-tier trait correlations are positive, so
the figure mainly ranks their strength along the shared size axis. Tight
confidence eyes near 1 are supplied by the cached bootstrap object
(n_boot = `r morph_boot_R$n_boot`, failed refits =
`r morph_boot_R$n_failed`). Treat this as a rendered plotting example;
final inference should use a bootstrap size chosen for the study rather
than the article's lightweight fixture.

The same cached intervals can be shown as a matrix-style ellipse plot.
Here the shape and fill show the sign and strength of each correlation;
black borders and stars mark pairs whose supplied bootstrap interval does
not cross zero. The stars are a display convention on the cached fixture's
percentile intervals, not a calibrated significance test — choose a
bootstrap size for the study before drawing inferential conclusions.

```{r ci-correlation-ellipse, fig.width = 6.6, fig.height = 5.2, fig.cap = "Ellipse view of the same cached between-individual correlation intervals. Ellipse tilt and colour show the fitted correlation; black borders and stars mark bootstrap percentile intervals that do not cross zero."}
plot(fit, type = "correlation_ellipse", level = "unit", boot = morph_boot_R)
```

```{r ci-communality}
extract_communality(fit, level = "unit")
```

Communality $c_t^2 = (\boldsymbol\Lambda\boldsymbol\Lambda^{\!\top})_{tt} / \boldsymbol\Sigma_{B,tt}$
is the fraction of trait $t$'s variance shared with the others via
the latent factors; high $c_t^2$ means strongly integrated, low
$c_t^2$ means trait-specific variation dominates.

This single generated data set is a teaching example, not a full
simulation study. The test suite refits the long and wide formulas from
the shipped object and checks that the off-diagonal covariance pattern
recovers the known truth. Broader coverage and interval-calibration
claims belong in the simulation-grid articles, which remain under audit.

## Why default `latent()`? `dep` and `indep` on the same data

The morphometric workflow uses ordinary `latent(d = K)` to
estimate $\boldsymbol\Sigma_B = \boldsymbol\Lambda\boldsymbol\Lambda^\top + \boldsymbol{\Psi}$.
Two alternative covstruct modes — `dep` (full unstructured $\boldsymbol\Sigma_B$
with $T(T+1)/2$ free entries) and `indep` (diagonal-only
$\boldsymbol\Sigma_B = \mathrm{diag}(\sigma^2_1, \dots, \sigma^2_T)$, no
cross-trait modelling) — are also valid keywords. Refit the same data
under each to see what the user actually loses or gains.
These are ordinary unit-tier keywords; source-specific forms such as
`phylo_dep()` and `spatial_dep()` affect phylogenetic or spatial
unit-tier covariance instead.

```{r dep-indep-compare, message = FALSE, warning = FALSE}
fit_full_long <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | individual, d = T),
  data = df, trait = "trait", unit = "individual"
)
fit_full_wide <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~ 1 +
          latent(1 | individual, d = T),
  data = df_wide, unit = "individual"
)

fit_dep_long <- gllvmTMB(
  value ~ 0 + trait +
          dep(0 + trait | individual),
  data = df, trait = "trait", unit = "individual"
)
fit_dep_wide <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~ 1 +
          dep(1 | individual),
  data = df_wide, unit = "individual"
)

fit_indep_long <- gllvmTMB(
  value ~ 0 + trait +
          indep(0 + trait | individual),
  data = df, trait = "trait", unit = "individual"
)
fit_indep_wide <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~ 1 +
          indep(1 | individual),
  data = df_wide, unit = "individual"
)

fit_full  <- fit_full_long
fit_dep   <- fit_dep_long
fit_indep <- fit_indep_long

cat("Max long/wide logLik difference across comparison fits:",
    signif(max(abs(c(
      as.numeric(logLik(fit_full_long)) - as.numeric(logLik(fit_full_wide)),
      as.numeric(logLik(fit_dep_long)) - as.numeric(logLik(fit_dep_wide)),
      as.numeric(logLik(fit_indep_long)) - as.numeric(logLik(fit_indep_wide))
    ))), 3),
    "\n")

S1 <- extract_Sigma(fit,       level = "unit")$Sigma
S2 <- extract_Sigma(fit_full,  level = "unit")$Sigma
S3 <- extract_Sigma(fit_dep,   level = "unit")$Sigma
S4 <- extract_Sigma(fit_indep, level = "unit")$Sigma

off       <- function(M) M[upper.tri(M)]
truth_off <- off(Sigma_true)
ll        <- function(f) round(as.numeric(logLik(f)), 2)

data.frame(
  mode               = c("latent(d=2)", "latent(d=T)",
                         "dep", "indep"),
  logL               = c(ll(fit), ll(fit_full), ll(fit_dep), ll(fit_indep)),
  off_diag_corr_truth = c(round(cor(off(S1), truth_off), 3),
                          round(cor(off(S2), truth_off), 3),
                          round(cor(off(S3), truth_off), 3),
                          NA),
  diag_min_max        = c(sprintf("%.2f - %.2f", min(diag(S1)), max(diag(S1))),
                          sprintf("%.2f - %.2f", min(diag(S2)), max(diag(S2))),
                          sprintf("%.2f - %.2f", min(diag(S3)), max(diag(S3))),
                          sprintf("%.2f - %.2f", min(diag(S4)), max(diag(S4))))
)
```

What this shows:

- **`dep` and `latent(d = T)` both recover the saturated covariance
  surface** — the comparison is a useful check that the dense fit and the
  full-rank latent parameterisation agree on the covariance pattern.
  `dep` and `latent(d = T)` give the same saturated fit here — numerically
  equal logL and $\boldsymbol\Sigma_B$ on this fixture.
- **Reduced-rank `latent(d = 2)` with default $\boldsymbol\Psi$ matches the saturated fit
  within 0.1 nats of logL.** With $T = 5$ traits and a
  rank-2 truth, the rank constraint loses essentially no information
  and gives the same off-diagonal correlation pattern as the saturated
  fit ($r$ vs truth ≈ 0.95 for both).
- **`indep` recovers the diagonals fine** — Gaussian gives per-trait
  identifiability for the marginal variance even with no cross-trait
  modelling — but **loses every off-diagonal**. The off-diagonal
  correlation with truth is `NA` because the fitted off-diagonals are
  exactly zero by construction. logL is materially worse.

For this simulated rank-2 Gaussian truth, default `latent(d = 2)` and the
saturated `dep` fit recover nearly the same covariance pattern, while
`indep` cannot recover off-diagonal association because it sets those
entries to zero by construction. Pick `latent(d = K)` when you have a
strong rank-$K$ hypothesis (size + shape, here); fall back to `dep` if
you suspect $K \approx T$ and want the saturated fit; reach for `indep`
only when you genuinely believe traits are mutually independent.

## What the latent factors mean

Every factor model has a rotational ambiguity: $\boldsymbol\Lambda$ and
$\boldsymbol\Lambda \mathbf{Q}$ produce the same $\boldsymbol\Sigma_B$
for any orthogonal $\mathbf{Q}$. To compare loadings between fits or
between packages, rotate to a standard convention (varimax for
interpretability, or lower-triangular for a unique fit):

```{r rotate}
L_hat <- extract_ordination(fit, level = "unit")$loadings
rot   <- rotate_loadings(
  fit,
  level = "unit",
  method = "varimax",
  anchor_traits = c("mass", "wing")
)
round(rot$Lambda, 2)
```

The same convention is available as tidy rows when you want a table
beside the figure:

```{r rotated-loading-table}
loading_table <- extract_rotated_loadings_table(
  fit,
  level = "unit",
  method = "varimax",
  anchor_traits = c("mass", "wing"),
  loading_scale = "standardized"
)

loading_table <- loading_table[
  order(loading_table$axis, -loading_table$abs_loading),
  c("trait", "axis", "loading", "axis_share", "anchor_trait")
]
head(loading_table, 8)
```

```{r ordi, fig.height = 5.6}
ord <- extract_ordination(fit, level = "unit")
head(ord$scores)
plot(
  fit,
  type = "ordination",
  level = "unit",
  anchor_traits = c("mass", "wing"),
  standardize_loadings = TRUE
) +
  ggplot2::labs(
    caption = paste(
      "Grey points are latent scores; arrows are standardized loadings.",
      "Axes use varimax rotation with sign anchors on mass and wing.",
      "Use Sigma and correlation summaries for rotation-invariant interpretation.",
      sep = "\n"
    )
  )
```

This ordination comes from the fitted covariance model
($\boldsymbol\Sigma = \boldsymbol\Lambda\boldsymbol\Lambda^\top + \boldsymbol\Psi$):
unlike PCA or NMDS, the axes are model parameters with uncertainty, not an
eigen- or distance-decomposition of the raw data, and the grey points are
*predicted* latent scores (empirical-Bayes conditional modes of
$\mathbf{u}_i$), not observed coordinates.

The biplot shows the trait loadings on the two latent axes. In a
real morphometric study, the first axis usually picks up *overall
size*, with all traits loading positively — the famous "general
factor" of body size. The second axis picks up shape contrasts. The
sign anchor is a reporting convention: here positive LV1 is anchored to
`mass` and positive LV2 to `wing`; changing signs would not change the
fitted covariance.

## What this article does *not* cover

This is the simplest case on the complexity ladder. Richer models —
binary / [joint species distribution](joint-sdm.html) responses,
phylogenetic and spatial dependence between rows, and two-level
between/within designs — build on this same `latent()` foundation and are
covered in their own guides as they become public.

## See also

* [Covariance and correlation](covariance-correlation.html) — why
  diagonal $\boldsymbol\Psi$ matters for interpreting trait correlations.
* [Formula keyword grid](api-keyword-grid.html) — the syntax map for
  moving from this example to other covariance rows and modes.
