---
title: "Rare items and runaway estimates in Paper × Items evidence synthesis"
description: "Ordinate a papers-by-items evidence table with a probit GLLVM, then meet the two ways such fits break — runaway loadings and separated rare items — and the matched remedy for each: an opt-in loading ridge for one, the opt-in MSPL estimator for the other, never both in one fit."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Rare items and runaway estimates in Paper × Items evidence synthesis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

Evidence synthesis teams increasingly work with a **papers-by-items table**: each
row a published study, each column a binary indicator — did the paper report an
effect size, share its data, preregister, use blinded scoring? The scientific
questions are ordination questions: *which practices travel together across
papers, and does a methods gradient explain them?* That is a job for a joint
latent-variable model of binary responses — the same machinery as a
[joint species distribution model](joint-sdm.html), with papers in place of
sites and reporting items in place of species.

This article fits that model, then walks into the two distinct ways such fits
break, on purpose, and shows the matched remedy for each:

1. **Runaway loadings.** With a strong latent structure and only a few dozen
   papers, plain maximum likelihood can report convergence while the latent
   loadings drift to absurd magnitudes. The matched remedy is an opt-in
   **loading ridge** (`gllvmTMBcontrol(loading_ridge = 2)`).
2. **A separated rare item.** When a rare item is recorded *only* by papers at
   one end of a measured gradient, its fixed-effect estimate has no finite
   maximum. The matched remedy is the opt-in, experimental **MSPL estimator**
   (`estimator = "mspl"`), after the separation is certified by
   `screen_gllvmTMB()`.

Each remedy treats only its own disease: they do not substitute for one
another — the middle of the article demonstrates that too — and they are
never combined in one fit. Ordinary `estimator = "ml"` with no penalty
remains the default throughout the package; nothing here changes any default.

# A papers-by-items corpus

Sixty papers scored on eight binary items, with one measured covariate: a
`design_score` summarising each paper's methodological rigour. Two latent
axes generate residual dependence between items — think of them as unmeasured
"methods culture" gradients across labs. The loadings are deliberately strong,
because that is the regime where the first failure mode lives.

```{r simulate}
library(gllvmTMB)

set.seed(202)
n_paper <- 60
items <- c("defines_urban", "reports_effect_size", "shares_data",
           "preregistered", "blinded_scoring", "reports_power",
           "maps_species", "escalation_plan")
design <- as.numeric(scale(seq_len(n_paper)))
z   <- matrix(rnorm(n_paper * 2), n_paper, 2)     # two latent methods axes
Lam <- matrix(rnorm(16, 0, 3), 8, 2)              # strong true loadings
alpha <- c(0.3, -0.2, -0.5, -0.8, 0.0, -0.4, 0.6, -2.6)
beta  <- c(0.4,  0.6,  0.5,  0.7, 0.2,  0.5, 0.1,  1.2)

eta <- sweep(z %*% t(Lam), 2, alpha, "+") + outer(design, beta)
Y <- matrix(rbinom(n_paper * 8, 1, pnorm(eta)), n_paper, 8,
            dimnames = list(NULL, items))
papers <- data.frame(paper = factor(seq_len(n_paper)),
                     design_score = design, Y)
round(colMeans(Y), 2)
```

Every item is assessed for every paper, and each cell is a single 0/1 record —
a zero means "assessed and absent", not "not reported". That completeness is a
real requirement: missing paper-by-item cells are not supported on the routes
this article uses.

The model is the wide-format call an evidence table naturally invites — probit
link, two latent dimensions, loadings-only residual structure
(`unique = FALSE`, because a free per-item variance is not identifiable from
single 0/1 trials):

```{r form}
form <- traits(defines_urban, reports_effect_size, shares_data,
               preregistered, blinded_scoring, reports_power,
               maps_species, escalation_plan) ~ 1 + design_score +
  latent(1 | paper, d = 2, unique = FALSE)
```

# Failure mode 1: convergence with runaway loadings

Fit it exactly as you would any other model:

```{r ml-plain}
fit_ml <- gllvmTMB(
  form, data = papers, unit = "paper",
  family = binomial("probit"), estimator = "ml",
  control = gllvmTMBcontrol(n_init = 1, se = FALSE, warn_runaway = FALSE)
)
fit_ml$opt$convergence
max(abs(fit_ml$report$Lambda))
```

The optimizer reports `0` — it found a point it judges stationary. But look at
the loadings. On the probit scale a loading *is* a latent standard deviation in
link units: the true values in this simulation top out near
`r round(max(abs(Lam)), 1)`, and probabilities saturate within a few units of
zero. An estimated magnitude in the tens is not a strong effect; it is the
likelihood drifting along a flat direction while the fitted probabilities for
those cells pin at 0 and 1. Convergence codes do not catch this — the flat
direction *looks* stationary.

## The matched remedy: an opt-in loading ridge

`gllvmTMBcontrol(loading_ridge = 2)` adds a Gaussian penalty on the loadings —
a MAP-style shrinkage toward zero with scale `tau = 2` — and nothing else:

```{r ml-ridge}
fit_ridge <- gllvmTMB(
  form, data = papers, unit = "paper",
  family = binomial("probit"), estimator = "ml",
  control = gllvmTMBcontrol(n_init = 1, se = FALSE, warn_runaway = FALSE,
                            loading_ridge = 2)
)
fit_ridge$opt$convergence
max(abs(fit_ridge$report$Lambda))
```

The loadings return to the scale of the truth, and convergence is clean. This
is not a lucky draw: in this package's internal calibration on the
binomial-probit grid, the same penalty took the runaway rate at a comparable
design size from roughly one fit in two to almost none. Two honest boundaries
travel with that record. First, the measured failure corner of the ridge
itself needs the **logit** link and few items together — on logit with small
item counts and large paper counts the penalised fit can still run away, so
the protective record quoted here is the probit record. Second, shrinkage is
not free: the penalty biases every loading toward zero a little to prevent a
few from escaping to infinity. The ridge is an opt-in stabiliser with a
measured track record in this regime, not a default and not a certificate.

The same penalty composes with the package's opt-in adaptive quadrature
engine (`gllvmTMBcontrol(aghq = 3, aghq_ridge = 2)` — `aghq_ridge` is the same
penalty under its original name). AGHQ replaces the Laplace approximation of
the latent integral with an adaptive quadrature; its measured accuracy gains
are for binomial models at unit counts far larger than sixty papers, so at
this corpus size it is the ridge, not the quadrature, doing the visible work.

# Failure mode 2: a separated rare item

Now the second, entirely different pathology. Suppose the escalation-plan item
had been recorded only by the handful of papers at the very top of the design
gradient — a realistic pattern for a practice that only the most rigorous
papers adopt:

```{r separate}
papers2 <- papers
papers2$escalation_plan <- as.integer(design > quantile(design, 0.93))
mean(papers2$escalation_plan)
```

A straight line on `design_score` can now split every presence from every
absence. That is **complete separation**, and it is a property of the fixed
design, not of the latent structure: the maximum-likelihood slope for this
item is infinite, no matter how the latent side is handled.

## First certify it

`screen_gllvmTMB()` runs the formal separation test item by item (it uses the
optional `detectseparation` package):

```{r screen, eval = requireNamespace("detectseparation", quietly = TRUE)}
scr <- screen_gllvmTMB(
  traits(defines_urban, reports_effect_size, shares_data,
         preregistered, blinded_scoring, reports_power,
         maps_species, escalation_plan) ~ 1 + design_score,
  data = papers2, unit = "paper",
  family = binomial("probit"),
  control = screen_control(separation = "fixed")
)
tab <- screen_table(scr, "separation")
tab[, c("traits", "status", "severity", "separated", "infinite_terms")]
```

```{r screen-missing, eval = !requireNamespace("detectseparation", quietly = TRUE), echo = FALSE}
cat("Install the optional detectseparation package to run this certificate.\n")
```

Seven items pass with `overlap`; the escalation item is flagged `complete`,
and the `infinite_terms` column names the exact model terms whose estimates
are infinite. The screen only diagnoses; it never changes the fit.

## What each route does with it

Fit the separated corpus three ways and read the escalation item's slope:

```{r three-routes}
routes <- list(
  ml    = gllvmTMBcontrol(n_init = 1, se = FALSE, warn_runaway = FALSE),
  ridge = gllvmTMBcontrol(n_init = 1, se = FALSE, warn_runaway = FALSE,
                          loading_ridge = 2)
)
slopes <- sapply(names(routes), function(r) {
  f <- suppressWarnings(gllvmTMB(form, data = papers2, unit = "paper",
    family = binomial("probit"), estimator = "ml", control = routes[[r]]))
  co <- coef(f)
  unname(co[grep("escalation_plan:design_score", names(co))])
})
fit_mspl <- suppressWarnings(gllvmTMB(
  form, data = papers2, unit = "paper",
  family = binomial("probit"), estimator = "mspl",
  control = gllvmTMBcontrol(n_init = 1, se = FALSE)
))
co <- coef(fit_mspl)
slopes <- c(slopes,
            mspl = unname(co[grep("escalation_plan:design_score", names(co))]))
round(slopes, 1)
```

The pattern is the whole lesson. Plain ML runs away, as the certificate
predicted. **The loading ridge does not help** — it penalises loadings, and
this pathology lives in the fixed effects; the two failure modes are genuinely
different diseases. MSPL returns a finite slope, because its penalty — a
Jeffreys-prior term on the fixed design, following Sterzinger & Kosmidis
(2023) — targets exactly this boundary. (Supplying `loading_ridge` together
with `estimator = "mspl"` is refused with an error: that hybrid has no
supporting theory.)

The finite MSPL point is a *penalised* estimate, and its fences are strict by
design: no `logLik()`, no AIC/BIC or likelihood-ratio comparison against the
ML fit, and no standard errors or intervals while repeated-sampling
calibration remains in progress — each of those fails with a typed error
rather than returning an uncalibrated number. What you get is a finite,
reportable description of the item's association with the gradient, instead
of a silently deleted item or an unusable coefficient in the hundreds.

# Which remedy, when

| Symptom | Diagnosis | Matched remedy |
|---|---|---|
| `convergence = 0` but loadings at absurd link-scale magnitudes | Flat likelihood direction in the latent structure | `gllvmTMBcontrol(loading_ridge = 2)` — opt-in shrinkage, probit-grid track record |
| `screen_gllvmTMB()` flags an item `complete` / `quasi_complete` | Fixed-design separation; the ML slope is infinite | `estimator = "mspl"` — opt-in, experimental penalised point estimate with explicit fences |
| Both symptoms in the same fit | Two diseases at once — but still one fit | Still one outer penalty: treat the separation with `estimator = "mspl"`. The ridge+MSPL combination is refused, so a co-occurring loading runaway has no packaged remedy — simplify the latent structure or collect more information |

Report, in either case: the certificate table, which route was fitted and why,
the penalty scale (`summary(fit)$estimation` stores `c_n` for MSPL), and the
convergence and gradient diagnostics — not just the finite number.

# See also

- [Joint species distribution model](joint-sdm.html) — the same latent
  machinery on sites × species occurrence data.
- [Screen binary traits before fitting](pre-fit-response-screening.html) —
  the wider pre-fit screening workflow around `screen_gllvmTMB()`.
- [Can I trust this fit?](fit-diagnostics.html) — the general diagnostics this
  article's convergence caveats point into.

# References

- Kosmidis I, Firth D (2021). Jeffreys-prior penalty, finiteness and shrinkage
  in binomial-response generalized linear models. *Biometrika* 108:71–82.
  <https://doi.org/10.1093/biomet/asaa052>
- Sterzinger P, Kosmidis I (2023). Maximum softly-penalized likelihood for
  mixed effects logistic regression. *Statistics and Computing* 33:53.
  <https://doi.org/10.1007/s11222-023-10217-3>
- Sterzinger P, Kosmidis I, Moustaki I (2026). Maximum softly penalized
  likelihood in factor analysis. *Psychometrika* 91:494–507.
  <https://doi.org/10.1017/psy.2026.10092> (adjacent theory for the factor
  extension; not proof for this implementation)
- Warton DI, Blanchet FG, O'Hara RB, Ovaskainen O, Taskinen S, Walker SC,
  Hui FKC (2015). So many variables: joint modeling in community ecology.
  *Trends in Ecology & Evolution* 30:766–779.
- van der Veen B, Hui FKC, Hovstad KA, O'Hara RB (2023). Concurrent
  ordination: simultaneous unconstrained and constrained latent variable
  modelling. *Methods in Ecology and Evolution* 14:683–695.
  <https://doi.org/10.1111/2041-210X.14035>
