---
title: "Keep rare binary items in Paper x Items evidence synthesis with LA-MSPL"
description: "Detect fixed-effect separation in paper-by-item evidence, fit the opt-in softly penalised Laplace estimator, and distinguish it from loading ridge and likelihood inference."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Keep rare binary items in Paper x Items evidence synthesis with LA-MSPL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

A rare outcome may be reported only by papers at the high end of a design or
exposure gradient. Do we silently remove that item, accept a runaway
maximum-likelihood coefficient, or keep it with an estimator whose boundary
behaviour is explicit?

This page uses two separate opt-in tools. `screen_gllvmTMB()` checks whether
the fixed-effect design proves complete or quasi-complete separation.
`estimator = "mspl"` then fits a softly penalised Laplace point estimator. The
screen never chooses the estimator, and MSPL never deletes an item.

The current experimental estimator admits **complete Bernoulli Paper × Item**
data: every item is assessed for every paper and each cell contains one 0/1
trial. A recorded zero means that the item was assessed and absent; an
unreported or unassessed item is missing data and remains outside this release.
MSPL supports logit, probit, and complementary log-log (`cloglog`) links. This
worked example uses `cloglog`, which is often natural when an item records at
least one event over an unequal underlying opportunity. It fits
`latent(..., unique = FALSE)`, so residual dependence between items is
\(\Sigma = \Lambda\Lambda^\top\), with no free diagonal \(\Psi\). A free
single-trial Bernoulli \(\Psi\) is not identifiable and is not admitted here.

The resolved fixed-effect design must have full column rank, and offsets must
be exactly zero. If either condition fails, the fit stops before optimisation;
MSPL does not silently drop columns or reinterpret exposure adjustments.

Missing paper-by-item cells require the later FIML-MSPL extension. Grouped binomial
data--multiple trials in one row, supplied through `cbind(successes, failures)`
or binomial trial weights--also remain deferred because the present
construction and validation campaign cover complete single-trial cells.
Mixed families, rank above two, additional random-effect tiers, and general
inference remain outside this surface. These extensions are deferred, not
claimed to be impossible. `estimator = "ml"` is unchanged and remains the
default.

## A small Paper × Items data set

```{r setup}
library(gllvmTMB)

set.seed(70808)
n_paper <- 36
papers <- data.frame(
  paper = factor(seq_len(n_paper)),
  design_score = seq(-2, 2, length.out = n_paper)
)
papers$item_supports <- rbinom(n_paper, 1, plogis(-0.2 - 0.8 * papers$design_score))
papers$item_harm <- rbinom(n_paper, 1, plogis(0.1 + 0.7 * papers$design_score))
papers$rare_escalation <- as.integer(seq_len(n_paper) %in% c(n_paper - 1L, n_paper))

long <- data.frame(
  paper = rep(papers$paper, times = 3L),
  item = factor(rep(c("item_supports", "item_harm", "rare_escalation"), each = n_paper)),
  design_score = rep(papers$design_score, times = 3L),
  value = c(papers$item_supports, papers$item_harm, papers$rare_escalation)
)
```

Two reported presences do not automatically imply separation. Here their
position on `design_score` is what matters: a linear boundary can isolate both
from every absence. That is a fixed-design property, not a prevalence rule.

## Certify fixed-design separation

The formal detector is optional because it adds the `detectseparation`
dependency only when requested.

```{r separation-screen, eval = requireNamespace("detectseparation", quietly = TRUE)}
screen <- screen_gllvmTMB(
  traits(item_supports, item_harm, rare_escalation) ~ 1 + design_score,
  data = papers,
  unit = "paper",
  family = binomial("cloglog"),
  control = screen_control(separation = "fixed")
)

screen_table(screen, "separation")
```

```{r separation-screen-missing, eval = !requireNamespace("detectseparation", quietly = TRUE), echo = FALSE}
cat("Install the optional detectseparation package to run the fixed-design certificate.\n")
```

The certificate is deliberately conservative about scope. It sees the fixed
design after coefficient maps and structural zeros, but it does not inspect
latent-loading or spatial-covariance divergence. `overlap` therefore means
"this fixed block has a finite GLM estimate," not "every GLLVM parameter is
identified."

Use the result as a decision rule. After fixed-design `overlap`, retain ML
unless another scientific reason calls for regularisation. When complete or
quasi-complete separation affects a scientifically necessary item and the
model lies inside the admitted surface, consider opt-in MSPL. Do not use MSPL
to rescue a constant response: it contains no response variation to estimate.
For missing, grouped, mixed-family, or otherwise unsupported structures,
redesign or filter the analysis transparently, or wait for the corresponding
extension.

## Fit ML and LA-MSPL explicitly

The model asks whether residual item co-occurrence remains after the paper-level
design responses are fitted. Both fits integrate the latent variables with Laplace
approximation. ML optimises the resulting unpenalised approximate likelihood;
MSPL optimises that same approximation plus its explicit soft penalties.

```{r paired-fits}
form_long <- value ~ 0 + item + item:design_score +
  latent(1 | paper, d = 1, unique = FALSE)

ctrl <- gllvmTMBcontrol(
  n_init = 1,
  se = FALSE,
  warn_runaway = FALSE
)

fit_ml_result <- tryCatch(
  suppressWarnings(gllvmTMB(
    form_long,
    data = long,
    unit = "paper",
    trait = "item",
    family = binomial("cloglog"),
    estimator = "ml",
    control = ctrl
  )),
  error = identity
)

fit_ml <- if (inherits(fit_ml_result, "error")) NULL else fit_ml_result
if (is.null(fit_ml)) {
  cat("ML did not return a fitted point:", conditionMessage(fit_ml_result), "\n")
}

fit_mspl <- suppressWarnings(gllvmTMB(
  form_long,
  data = long,
  unit = "paper",
  trait = "item",
  family = binomial("cloglog"),
  estimator = "mspl",
  control = ctrl
))

fit_mspl
```

The finite MSPL coefficient is not evidence that the original ML separation
disappeared. The unpenalised likelihood still has the same boundary direction;
MSPL selects a finite point by adding a sample-size-dependent fixed-effect
Jeffreys term and smooth radial penalties for the latent structure.

```{r coefficient-comparison}
comparison <- data.frame(
  term = names(coef(fit_mspl)),
  ML = if (is.null(fit_ml)) NA_real_ else unname(coef(fit_ml)),
  MSPL = unname(coef(fit_mspl))
)
comparison
```

For the rare item, the fitted point describes reporting increasing sharply
towards the high end of `design_score`. The supported `predict()` route can
display that curve on the response scale while excluding the fitted paper-level latent
scores:

```{r rare-item-prediction}
pred_mspl <- predict(fit_mspl, type = "response", re_form = ~ 0)
rare_curve <- data.frame(
  design_score = long$design_score,
  item = long$item,
  item_probability = pred_mspl$est
)
rare_curve <- rare_curve[rare_curve$item == "rare_escalation", ]
rare_curve[c(1, 9, 18, 27, 36), c("design_score", "item_probability")]
```

This finite curve is the biological interpretation of the penalised point
estimate; it is not evidence that ordinary ML became identifiable, and it does
not carry calibrated uncertainty.

Do not compare these fits with AIC, BIC, a likelihood-ratio test, or
`logLik()`. The MSPL optimum is not the unpenalised maximum-likelihood optimum,
and those methods fail with a typed error. Standard errors and confidence
intervals are also withheld while repeated-sampling calibration is in
progress. Point estimates and predictions are implemented, but the release
ordinary release claim remains experimental until the frozen ordinary promotion
gates complete. Spatial evidence has a separate, explicitly experimental gate.

## The same fit from wide data

Evidence-synthesis teams may begin with a paper-by-item table rather than a
stacked table. The wide `traits(...)` call uses the same `gllvmTMB()` entry point.

```{r wide-fit}
fit_mspl_wide <- suppressWarnings(gllvmTMB(
  traits(item_supports, item_harm, rare_escalation) ~ 1 + design_score +
    latent(1 | paper, d = 1, unique = FALSE),
  data = papers,
  unit = "paper",
  family = binomial("cloglog"),
  estimator = "mspl",
  control = ctrl
))

c(long = fit_mspl$opt$objective, wide = fit_mspl_wide$opt$objective)
```

Long data are canonical when items have item-specific predictors or observation
metadata. Wide data are often the cleanest starting point for an evidence table.
The two calls describe the same scientific model.

## MSPL is not loading ridge

`gllvmTMBcontrol(loading_ridge = 2)` requests the existing Gaussian-prior MAP
penalty on loadings. `aghq_ridge` remains its compatibility spelling; the
penalty does not depend on whether the latent integral is evaluated by Laplace
or AGHQ. Ridge may have lower prediction or estimation risk in some regimes,
so it remains an empirical comparator.

LA-MSPL has a different construction and purpose. It combines the fixed-design
Jeffreys term with radial loading or spatial-covariance penalties, scaled by
`2 * sqrt(p_free / N_eff)`. Supplying a loading ridge together with
`estimator = "mspl"` is a public error because that hybrid has no promoted
theory. The planned simulation campaign includes that hybrid as an internal
ablation.

## Spatial extension

Paper × Items synthesis usually needs the ordinary model above: papers are the
independent sampling units and `latent()` captures residual dependence between
items. MSPL also has an explicitly experimental spatial implementation, but
spatial promotion needs separate mesh, range, marginal-scale, rank, and
prediction evidence. It is not required for, and is not implied by, this
ordinary evidence-synthesis workflow.

## What to report

Report the response support, the fixed-design certificate, the link, the
latent/spatial structure and rank, the estimator label, the penalty scale
`c_n`, convergence and gradient diagnostics, and whether any item was
filtered before analysis. A promotable result is not simply "the optimizer
returned a finite number." It requires a finite stationary point inside a
prespecified regime, retained failures, and a separate assessment of latent
rank collapse.

The fit stores those diagnostics rather than asking users to reconstruct them:

```{r mspl-diagnostics}
mspl_summary <- summary(fit_mspl)
mspl_summary$estimation[c("c_n", "p_free", "N_eff", "scope")]
fit_mspl$fit_health[c(
  "convergence", "max_gradient", "scaled_gradient", "boundary_flags"
)]
```

In this deliberately small example, `near_zero_B_loading` means that MSPL found
a finite point estimate but the latent covariance is still close to a rank
boundary. Use `gllvmTMB_diagnose(fit_mspl)` and simplify, refit, or collect more
information before interpreting that covariance as clean residual
co-occurrence evidence. The example demonstrates boundary-stabilised point
estimation for the separated item; it does not demonstrate a well-resolved
latent rank.

## References

- Kosmidis I, Firth D (2021). Jeffreys-prior penalty, finiteness and shrinkage
  in binomial-response generalized linear models. *Biometrika* 108:71--82.
  <https://doi.org/10.1093/biomet/asaa052>.
- Sterzinger P, Kosmidis I (2023). Maximum softly-penalized likelihood for
  mixed effects logistic regression. *Statistics and Computing* 33:53.
  <https://doi.org/10.1007/s11222-023-10217-3>.
- Sterzinger P, Kosmidis I, Moustaki I (2026). Maximum Softly Penalized
  Likelihood in Factor Analysis. *Psychometrika* 91:494--507.
  <https://doi.org/10.1017/psy.2026.10092>. The paper identifies logistic
  factor analysis as future work; it is adjacent theory, not proof for this
  implementation.
- Sterzinger P, Kosmidis I (2026). Diaconis-Ylvisaker prior penalized
  likelihood for \(p/n \to \kappa \in (0,1)\) logistic regression.
  arXiv:2311.07419v3. <https://doi.org/10.48550/arXiv.2311.07419>. This is a
  high-dimensional logistic-regression result, not GLMM or GLLVM evidence; it
  motivates retaining ridge as an empirical comparator.
- van der Veen B, Hui FKC, Hovstad KA, O'Hara RB (2023). Concurrent
  ordination: Simultaneous unconstrained and constrained latent variable
  modelling. *Methods in Ecology and Evolution* 14:683--695.
  <https://doi.org/10.1111/2041-210X.14035>.
