---
title: "Unordered categories with multinomial()"
description: "Model an unordered categorical response as baseline-category log-odds, read the per-category contrasts and predicted probabilities, and stop at the fixed-effects-only boundary."
tier: 2
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Unordered categories with multinomial()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

Some responses are a single choice among several categories that have **no
natural order**: a bird's diet guild (frugivore, insectivore, granivore), a
plant's pollination syndrome, a patch's dominant land cover. `multinomial()`
models such a response as a set of baseline-category log-odds and recovers how a
predictor moves the odds of each category against a common reference.

This article shows when to reach for it, how to read its coefficients and
predicted probabilities, and — just as important — where it stops in the current
release.

## When to use `multinomial()`

Match the family to how the categories relate, not just to the fact that the
response is a factor.

| Response | Family | Why |
|---|---|---|
| Unordered categories, `K >= 3` | `multinomial()` | No category is "between" the others; each gets its own contrast against a reference. |
| **Ordered** categories, `K >= 3` | `ordinal_probit()` | The order is information. A single latent axis crosses ordered thresholds; see [Choose a response family](response-families.html). |
| Two categories | `binomial(link = "logit")` | One log-odds is enough. `multinomial()` on a 2-level response errors and points you here. |

Two cautions before you start:

- `multinomial()` is a **response** family. It is not the same as
  `categorical()`, which is a constructor for imputing a missing categorical
  *predictor*. The two are unrelated.
- "Ordered-looking" numeric codes are not ordinal by themselves, and genuinely
  ordered categories are not multinomial. Diet guilds are unordered; abundance
  classes (`low < medium < high`) are ordered and belong to `ordinal_probit()`.

## The model: baseline-category log-odds

Pick a reference category (the first factor level by default). For a response
with categories \(1, \dots, K\) and reference \(1\),

\[
\eta_{1} \equiv 0, \qquad
\eta_{k} = \beta_{0k} + x^{\top}\beta_{k}, \quad k = 2, \dots, K,
\]

\[
\Pr(y = k) = \frac{\exp(\eta_{k})}{\sum_{j=1}^{K} \exp(\eta_{j})}
\quad\text{(softmax)} .
\]

The reference is pinned at \(\eta_1 = 0\) so the model is identified, and each
non-reference category carries its own intercept and slopes. A coefficient is a
**contrast against the reference**: \(\beta_{0k}\) is the log-odds of category
\(k\) versus the reference at \(x = 0\), and \(\beta_{k}\) is the change in that
log-odds per unit of \(x\). A single categorical trait therefore occupies
\(K-1\) linear predictors.

Which category is the reference is only a labelling choice: the fitted
probabilities and the maximised likelihood are identical whichever you pick, and
only the coefficient labels change. You can set it with
`multinomial(baseline = ...)` (shown below).

## A worked example: diet guild

We simulate the diet guild of `r 250` bird species — frugivore, insectivore, or
granivore — as a function of one scaled morphological axis, `bill` (larger values
= deeper bills). The data-generating odds make deeper-billed species more likely
to be granivores and finer-billed species more likely to be insectivores,
relative to frugivores.

```{r simulate}
library(gllvmTMB)

set.seed(20260716)
n_species <- 250L
bill <- as.numeric(scale(rnorm(n_species)))

# True baseline-category log-odds, reference = frugivore.
eta <- cbind(
  frugivore   = 0,
  insectivore = 0.3 - 0.9 * bill,
  granivore   = -0.2 + 1.1 * bill
)
prob <- exp(eta - apply(eta, 1L, max))
prob <- prob / rowSums(prob)

guild <- c("frugivore", "insectivore", "granivore")[
  apply(prob, 1L, function(p) sample.int(3L, 1L, prob = p))
]

diet <- data.frame(
  species = factor(seq_len(n_species)),
  trait   = factor("diet"),
  guild   = factor(guild, levels = c("frugivore", "insectivore", "granivore")),
  bill    = bill
)

table(diet$guild)
```

A multinomial response is a **single** categorical column, so the model is fit
from long data only; there is no wide `traits(...)` twin, because that shorthand
is for stacking several response columns into a shared covariance — which a
fixed-effects-only multinomial trait does not use.

```{r fit}
fit <- gllvmTMB(
  guild ~ 0 + trait + (0 + trait):bill,
  data   = diet,
  trait  = "trait",
  unit   = "species",
  family = multinomial(),
  silent = TRUE
)

c(convergence = fit$opt$convergence, pdHess = isTRUE(fit$sd_report$pdHess))
```

## Reading the contrasts

The fit holds \(K - 1 = 2\) contrasts (insectivore and granivore, each versus the
frugivore reference), each with an intercept and a `bill` slope. We pull them
into a readable table.

```{r contrasts}
sdf <- summary(fit$sd_report, "fixed")
est <- sdf[grepl("b_fix", rownames(sdf)), "Estimate"]
names(est) <- fit$X_fix_names

pick <- function(cat, slope = FALSE) {
  key <- if (slope) paste0(cat, ":bill") else paste0(cat, "$")
  unname(est[grepl(key, names(est))])
}

data.frame(
  contrast   = c("insectivore vs frugivore", "granivore vs frugivore"),
  intercept  = round(c(pick("insectivore"), pick("granivore")), 2),
  bill_slope = round(c(pick("insectivore", TRUE), pick("granivore", TRUE)), 2),
  truth_slope = c(-0.9, 1.1)
)
```

Read the `bill` slopes as log-odds changes against the frugivore reference. The
granivore slope is positive: deeper-billed species are more likely to be
granivores than frugivores. The insectivore slope is negative: deeper-billed
species are *less* likely to be insectivores than frugivores. Both recover the
signs and rough magnitudes of the truth. An intercept is the log-odds of that
guild versus frugivory at the mean bill (`bill = 0`).

Coefficients are contrasts, not category-level effects: a positive granivore
slope does not by itself mean the *probability* of granivory rises everywhere,
because every category shares the denominator. For probabilities, predict.

## Predicted category probabilities

`predict(type = "response")` returns a long data frame with one row per
observation and category; the `est` column is the fitted probability, and the
probabilities sum to one within each observation.

```{r predict}
p_hat <- predict(fit, type = "response")
head(p_hat)

# probabilities sum to 1 within each observation
head(tapply(p_hat$est, p_hat$species, sum), 3)
```

To see the trend, evaluate the fitted softmax across a grid of `bill` values
directly from the estimated contrasts (the reference stays at \(\eta = 0\)):

```{r softmax-curve}
bill_grid <- seq(-2, 2, by = 1)
eta_hat <- cbind(
  frugivore   = 0,
  insectivore = pick("insectivore") + pick("insectivore", TRUE) * bill_grid,
  granivore   = pick("granivore")   + pick("granivore",   TRUE) * bill_grid
)
p_grid <- exp(eta_hat) / rowSums(exp(eta_hat))
cbind(bill = bill_grid, round(p_grid, 2))
```

Fitted probability of granivory rises with bill depth while insectivory falls —
the diet composition the data were built from.

Prediction is limited to the fitted data in this release: `predict()` with a
`newdata` argument on a multinomial fit is not supported and fails loud rather
than returning a silently wrong answer. Build the softmax grid from the
coefficients (as above) when you need probabilities at new predictor values.

## Changing the reference category

Any category can be the reference. Refitting with a different baseline relabels
the contrasts but leaves the fit unchanged — same log-likelihood, same fitted
probabilities.

```{r baseline}
fit_gran <- gllvmTMB(
  guild ~ 0 + trait + (0 + trait):bill,
  data   = diet,
  trait  = "trait",
  unit   = "species",
  family = multinomial(baseline = "granivore"),
  silent = TRUE
)

c(default_logLik  = as.numeric(logLik(fit)),
  granivore_logLik = as.numeric(logLik(fit_gran)))
```

The two log-likelihoods match; only whether coefficients read "versus frugivore"
or "versus granivore" changes. Choose the reference that makes your contrasts
easiest to interpret.

## What this release does not do

`multinomial()` is **fixed-effects-only** in this release, and that boundary is
deliberate. An unordered response over \(K\) categories is not one latent
dimension but \(K - 1\) of them (the contrasts), so it does not fit the
single-latent-scale contract that the package's cross-trait covariance machinery
depends on. Concretely:

- Latent, random-effect, or structured terms on a multinomial trait
  (`latent()`, `indep()`, `phylo_*()`, `spatial_*()`, random slopes) are **not
  supported** and fail loud.
- There is **no trait correlation** for a multinomial trait:
  `extract_correlations()` and `extract_Sigma()` refuse a fit containing one.
  The \((K-1) \times (K-1)\) covariance among the category contrasts — the
  natural "correlation table" for an unordered response — is a separate, deferred
  extension.
- Coefficient recovery is validated (see the package's recovery tests), but the
  family is not part of the interval-coverage certification; treat its intervals
  as recovery-oriented.

Within those limits, `multinomial()` gives you an honest, identified
baseline-category model for a single unordered categorical response.

## See also

- [Choose a response family](response-families.html) — matching every response
  type to a family, including the ordered-category route with `ordinal_probit()`.
- `?multinomial` — the family constructor and its `baseline` argument.
- `?ordinal_probit` — for ordered categories.
- `?categorical` — the unrelated missing-*predictor* imputation family.
