---
title: "Unordered categories with multinomial()"
description: "Model an unordered categorical response as baseline-category log-odds, read its fixed-effect contrasts and probabilities, and distinguish the covered core from two partial covariance routes."
tier: 2
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Unordered categories with multinomial()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

Some responses are a single choice among several categories that have **no
natural order**: a bird's diet guild (frugivore, insectivore, granivore), a
plant's pollination syndrome, a patch's dominant land cover. `multinomial()`
models such a response as a set of baseline-category log-odds and recovers how a
predictor moves the odds of each category against a common reference.

This article shows when to reach for it, how to read its coefficients and
predicted probabilities, and — just as important — where it stops in the current
release.

## When to use `multinomial()`

Match the family to how the categories relate, not just to the fact that the
response is a factor.

| Response | Family | Why |
|---|---|---|
| Unordered categories, `K >= 3` | `multinomial()` | No category is "between" the others; each gets its own contrast against a reference. |
| **Ordered** categories, `K >= 3` | `ordinal_probit()` | The order is information. A single latent axis crosses ordered thresholds; see [Choose a response family](response-families.html). |
| Two categories | `binomial(link = "logit")` | One log-odds is enough. `multinomial()` on a 2-level response errors and points you here. |

Two cautions before you start:

- `multinomial()` is a **response** family. It is not the same as
  `categorical()`, which is a constructor for imputing a missing categorical
  *predictor*. The two are unrelated.
- "Ordered-looking" numeric codes are not ordinal by themselves, and genuinely
  ordered categories are not multinomial. Diet guilds are unordered; abundance
  classes (`low < medium < high`) are ordered and belong to `ordinal_probit()`.

## The model: baseline-category log-odds

Pick a reference category (the first factor level by default). For a response
with categories \(1, \dots, K\) and reference \(1\),

\[
\eta_{1} \equiv 0, \qquad
\eta_{k} = \beta_{0k} + x^{\top}\beta_{k}, \quad k = 2, \dots, K,
\]

\[
\Pr(y = k) = \frac{\exp(\eta_{k})}{\sum_{j=1}^{K} \exp(\eta_{j})}
\quad\text{(softmax)} .
\]

The reference is pinned at \(\eta_1 = 0\) so the model is identified, and each
non-reference category carries its own intercept and slopes. A coefficient is a
**contrast against the reference**: \(\beta_{0k}\) is the log-odds of category
\(k\) versus the reference at \(x = 0\), and \(\beta_{k}\) is the change in that
log-odds per unit of \(x\). A single categorical trait therefore occupies
\(K-1\) linear predictors.

Which category is the reference is only a labelling choice: the fitted
probabilities and the maximised likelihood are identical whichever you pick, and
only the coefficient labels change. You can set it with
`multinomial(baseline = ...)` (shown below).

## A worked example: diet guild

We simulate the diet guild of `r 250` bird species — frugivore, insectivore, or
granivore — as a function of one scaled morphological axis, `bill` (larger values
= deeper bills). The data-generating odds make deeper-billed species more likely
to be granivores and finer-billed species more likely to be insectivores,
relative to frugivores.

```{r simulate}
library(gllvmTMB)

set.seed(20260716)
n_species <- 250L
bill <- as.numeric(scale(rnorm(n_species)))

# True baseline-category log-odds, reference = frugivore.
eta <- cbind(
  frugivore   = 0,
  insectivore = 0.3 - 0.9 * bill,
  granivore   = -0.2 + 1.1 * bill
)
prob <- exp(eta - apply(eta, 1L, max))
prob <- prob / rowSums(prob)

guild <- c("frugivore", "insectivore", "granivore")[
  apply(prob, 1L, function(p) sample.int(3L, 1L, prob = p))
]

diet <- data.frame(
  species = factor(seq_len(n_species)),
  trait   = factor("diet"),
  guild   = factor(guild, levels = c("frugivore", "insectivore", "granivore")),
  bill    = bill
)

table(diet$guild)
```

This worked example has one categorical response column, so it uses the long
form directly; there is no useful wide `traits(...)` twin for this isolated
fixed-effect fit. The separate partial cross-family route uses a shared
long-format mixed-family model and is demonstrated in
[Cross-family trait correlations, including nominal responses](cross-family-correlations.html).

```{r fit}
fit <- gllvmTMB(
  guild ~ 0 + trait + (0 + trait):bill,
  data   = diet,
  trait  = "trait",
  unit   = "species",
  family = multinomial(),
  silent = TRUE
)

c(convergence = fit$opt$convergence, pdHess = isTRUE(fit$sd_report$pdHess))
```

## Reading the contrasts

The fit holds \(K - 1 = 2\) contrasts (insectivore and granivore, each versus the
frugivore reference), each with an intercept and a `bill` slope. We pull them
into a readable table.

```{r contrasts}
sdf <- summary(fit$sd_report, "fixed")
est <- sdf[grepl("b_fix", rownames(sdf)), "Estimate"]
names(est) <- fit$X_fix_names

pick <- function(cat, slope = FALSE) {
  key <- if (slope) paste0(cat, ":bill") else paste0(cat, "$")
  unname(est[grepl(key, names(est))])
}

data.frame(
  contrast   = c("insectivore vs frugivore", "granivore vs frugivore"),
  intercept  = round(c(pick("insectivore"), pick("granivore")), 2),
  bill_slope = round(c(pick("insectivore", TRUE), pick("granivore", TRUE)), 2),
  truth_slope = c(-0.9, 1.1)
)
```

Read the `bill` slopes as log-odds changes against the frugivore reference. The
granivore slope is positive: deeper-billed species are more likely to be
granivores than frugivores. The insectivore slope is negative: deeper-billed
species are *less* likely to be insectivores than frugivores. Both recover the
signs and rough magnitudes of the truth. An intercept is the log-odds of that
guild versus frugivory at the mean bill (`bill = 0`).

Coefficients are contrasts, not category-level effects: a positive granivore
slope does not by itself mean the *probability* of granivory rises everywhere,
because every category shares the denominator. For probabilities, predict.

## Predicted category probabilities

`predict(type = "response")` returns a long data frame with one row per
observation and category; the `est` column is the fitted probability, and the
probabilities sum to one within each observation.

```{r predict}
p_hat <- predict(fit, type = "response")
head(p_hat)

# probabilities sum to 1 within each observation
head(tapply(p_hat$est, p_hat$species, sum), 3)
```

To see the trend, evaluate the fitted softmax across a grid of `bill` values
directly from the estimated contrasts (the reference stays at \(\eta = 0\)):

```{r softmax-curve}
bill_grid <- seq(-2, 2, by = 1)
eta_hat <- cbind(
  frugivore   = 0,
  insectivore = pick("insectivore") + pick("insectivore", TRUE) * bill_grid,
  granivore   = pick("granivore")   + pick("granivore",   TRUE) * bill_grid
)
p_grid <- exp(eta_hat) / rowSums(exp(eta_hat))
cbind(bill = bill_grid, round(p_grid, 2))
```

Fitted probability of granivory rises with bill depth while insectivory falls —
the diet composition the data were built from.

Prediction is limited to the fitted data in this release: `predict()` with a
`newdata` argument on a multinomial fit is not supported and fails loud rather
than returning a silently wrong answer. Build the softmax grid from the
coefficients (as above) when you need probabilities at new predictor values.

## Changing the reference category

Any category can be the reference. Refitting with a different baseline relabels
the contrasts but leaves the fit unchanged — same log-likelihood, same fitted
probabilities.

```{r baseline}
fit_gran <- gllvmTMB(
  guild ~ 0 + trait + (0 + trait):bill,
  data   = diet,
  trait  = "trait",
  unit   = "species",
  family = multinomial(baseline = "granivore"),
  silent = TRUE
)

c(default_logLik  = as.numeric(logLik(fit)),
  granivore_logLik = as.numeric(logLik(fit_gran)))
```

The two log-likelihoods match; only whether coefficients read "versus frugivore"
or "versus granivore" changes. Choose the reference that makes your contrasts
easiest to interpret.

## Current scope boundary

The fixed-effect route demonstrated above is **covered** (FAM-20). Two later
extensions are deliberately **partial**, not universal categorical covariance
support:

- FAM-20A admits one `phylo_latent()` term and reports the standalone
  \((K-1) \times (K-1)\) among-category phylogenetic covariance \(V\) described
  below. Its recovery is data-hungry.
- FAM-20B permits one multinomial trait to share an ordinary `latent()` block
  with other response families. The nominal trait remains an explicit \(K-1\)
  baseline-contrast block; `extract_cross_correlations()` supplies the
  reference-invariant partner summary. See
  [Cross-family trait correlations, including nominal responses](cross-family-correlations.html).
- A fixed-effect-only fit still has no latent covariance, so
  `extract_correlations()` and `extract_Sigma()` refuse it rather than fabricate
  a matrix. On the admitted FAM-20B route, point extraction is the default;
  Wald/bootstrap summaries are target-specific and uncalibrated, and nonlinear
  profile requests are withdrawn.
- Multiple multinomial traits, augmented slopes, explicit multinomial
  `unique()`/`indep()`, and unlisted animal, spatial, kernel, or phylogenetic
  tiers remain blocked and fail loud.

Thus coefficient recovery is covered only for FAM-20, while FAM-20A/FAM-20B
retain their narrower `partial` labels. None of these rows establishes interval
coverage.

### The phylogenetic route: a standalone among-category table

The standalone \((K-1) \times (K-1)\) table is most natural in a *phylogenetic*
setting, and gllvmTMB now reports it there (design 84), following the phylogenetic
multinomial GLMM of Mizuno, Drobniak, Williams, Lagisz & Nakagawa (2025, *Journal
of Evolutionary Biology*, doi:10.1093/jeb/voaf116). A `phylo_latent()` factor on
the \(K-1\) category contrasts estimates the among-category covariance \(V\) — how
the category liabilities coevolve across the tree — read out with
`extract_Sigma(fit, level = "phy", part = "shared", link_residual = "none")`.
That explicit call returns the fitted \(V_{\mathrm{phy}}\); the default total
extraction with `link_residual = "auto"` adds the fixed softmax residual
\((\pi^2/6)(\mathbf I + \mathbf J)\). A categorical trait already occupies
\(K-1\) latent dimensions, so a low-rank factor on those contrasts is exactly the
right object; no new likelihood is needed beyond the softmax you have here.

The honest caveat is that this table is **data-hungry**. A single categorical
draw carries little information about the \((K-1)\)-dimensional liability, so
recovering \(V\) needs either per-species replication or a large number of
species; one observation per species is too weakly informative to pin the
among-category correlation reliably (the point estimate is high-variance and can
reach the \(\pm 1\) boundary). Read it as recovery-oriented, and prefer replicated
or large samples — the same discipline the coefficient recovery above is held to.

Two scales are in play, and it helps to name them. \(V\) above is the covariance
of the phylogenetic factor on the latent (log-odds) scale. To place a categorical
trait on the same footing as ordinary single-scale traits you also need its
*observation-scale* residual, and for the softmax that residual is a matrix, not a
scalar: each category contrast carries \(\pi^2/3\) — the same distribution-specific
link variance a `binomial` logit uses (Nakagawa & Schielzeth 2010, *Biological
Reviews* 85, 935–956; Nakagawa, Johnson & Schielzeth 2017, *J. R. Soc. Interface*
14, 20170213) — with a \(\pi^2/6\) covariance between contrasts induced by their
shared baseline category, i.e. \((\pi^2/6)(\mathbf{I} + \mathbf{J})\). That is the
covariance of the differenced Gumbel errors in the softmax's random-utility
representation (McFadden 1974, in *Frontiers in Econometrics*, ed. P. Zarembka,
pp. 105–142); it reduces to the binomial's \(\pi^2/3\) when \(K = 2\), and it is
the scale on which a categorical trait can eventually be made commensurable with
single-scale traits.

Within those limits, `multinomial()` gives you an honest, identified
baseline-category model for a single unordered categorical response.

## See also

- [Choose a response family](response-families.html) — matching every response
  type to a family, including the ordered-category route with `ordinal_probit()`.
- `?multinomial` — the family constructor and its `baseline` argument.
- `?ordinal_probit` — for ordered categories.
- `?categorical` — the unrelated missing-*predictor* imputation family.
