---
title: "Ordinal-probit threshold traits"
tier: 3 # internal threshold-trait technical draft until FAM-14 validation closes
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Ordinal-probit threshold traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE
)
```

> **Preview — `ordinal_probit` family validation is covered for the current
> family-recovery row, but this article stays internal.**
>
> The `ordinal_probit()` family has covered family-recovery evidence
> (FAM-14 in the
> [validation-debt register](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/35-validation-debt-register.md)).
> This page is still not a public worked example because cutpoint extractor
> depth, interval examples, and reader/browser review remain open.

**Internal article gate.** This page is a Tier 3 technical draft while
the ordinal article surface is under audit. It now contains a compact
runnable long/wide example, cutpoint extraction, latent-scale Sigma extraction,
and fit-health diagnostics. It is not a public worked example, an ordinal
interval-coverage claim, or a claim that ordinal randomized-quantile residuals
are available for exact-CDF diagnostics.

Use `ordinal_probit()` when a trait is ordered categorical with at
least three categories and the scientific target is a latent continuous
scale: behavioural intensity scores, ordered parasite-load scores,
ranked breeding states, or threshold traits in phylogenetic comparative
analyses. Use `binomial(link = "probit")` instead when the response has
only two categories.

The reason to prefer the probit threshold model is scale. The latent
variable is

```text
y* = eta + epsilon,    epsilon ~ Normal(0, 1)
```

and the observed category is determined by cutpoints:

```text
y = k when tau[k - 1] < y* <= tau[k]
```

Because `epsilon` has variance 1 by construction, the
distribution-specific latent residual is exactly `sigma_d^2 = 1`.
There is no trigamma or logistic `pi^2 / 3` approximation to add
before comparing variance components to a continuous Gaussian trait.

## Reader path and scope

This internal draft answers five threshold-trait reference questions:

| Reader question | Model object | Code section | Readout |
|---|---|---|---|
| When is `ordinal_probit()` the right family? | ordered categorical trait with `K >= 3` | opening and `Guardrails` | probit threshold model versus binary probit |
| How do long and wide formulas spell the same family? | long `value ~ ...` and wide `traits(...) ~ ...` examples | `Fit The Family` | same compact fixture, same log-likelihood |
| How are cutpoints identified? | `extract_cutpoints()` | `Cutpoints` | `tau_1 = 0`, then `cutpoint_2`, ... rows |
| How do variance components stay comparable? | `extract_Sigma(..., link_residual = "auto")` | `Variance Components` | fixed ordinal residual `sigma_d^2 = 1` |
| Is the tiny fit numerically usable? | `check_gllvmTMB()` | `Diagnostics` | optimizer, gradient, sdreport, and Hessian rows |
| What should readers avoid? | family and residual-scale guardrails | `Guardrails` | no `K = 2` threshold model, no mixed family within one ordinal trait, no OLRE residual scale |

Scope boundary: IN evidence includes `FAM-14` family-recovery coverage and
the runnable local article fixture below. PARTIAL rows include ordinal
cutpoint extractor depth (`EXT-10`), ordinal bridge CI endpoints, and any
profile/bootstrap interval claims for ordinal cutpoints or variance
components. PLANNED or guarded work includes broader applied examples,
ordinal exact-CDF residual diagnostics, interval calibration, and final
article placement review. Until those land, this page stays internal.

## Fit The Family

This small fixture has two ordinal traits scored as `1`, `2`, or `3`.
The latent responses are simulated on a probit scale with one fixed
threshold at `tau_1 = 0` and one free threshold at `tau_2 = 0.8`.

```{r simulate-ordinal}
set.seed(20260619)

n_unit <- 30L
traits <- c("display_score", "parasite_score")
unit_id <- factor(seq_len(n_unit))

unit_effect <- matrix(
  rnorm(n_unit * length(traits), sd = 0.55),
  nrow = n_unit,
  dimnames = list(NULL, traits)
)

eta <- cbind(
  display_score = 0.2 + unit_effect[, "display_score"],
  parasite_score = -0.1 + unit_effect[, "parasite_score"]
)

y_star <- eta + matrix(rnorm(n_unit * length(traits)), nrow = n_unit)
ordinal_y <- apply(y_star, 2, function(x) {
  as.integer(cut(x, breaks = c(-Inf, 0, 0.8, Inf), labels = FALSE))
})

df_wide <- data.frame(
  unit = unit_id,
  display_score = ordinal_y[, "display_score"],
  parasite_score = ordinal_y[, "parasite_score"]
)

df_long <- data.frame(
  unit = rep(unit_id, times = length(traits)),
  trait = factor(rep(traits, each = n_unit), levels = traits),
  value = as.integer(c(
    ordinal_y[, "display_score"],
    ordinal_y[, "parasite_score"]
  ))
)

with(df_long, table(trait, value))
```

The long form is the canonical stacked-trait spelling: one row per
`(unit, trait)` observation, one response column, and a `trait` column
that names the response.

```{r}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    indep(0 + trait | unit),
  data = df_long,
  trait = "trait",
  unit = "unit",
  family = ordinal_probit()
)
```

The wide data-frame form uses `traits(...)` on the left-hand side.
The compact wide syntax expands to the same stacked-trait formula
before fitting.

```{r}
fit_wide <- gllvmTMB(
  traits(display_score, parasite_score) ~ 1 +
    indep(1 | unit),
  data = df_wide,
  unit = "unit",
  family = ordinal_probit()
)

as.numeric(logLik(fit_long) - logLik(fit_wide))
```

The example uses `indep()` because it is fast and stable for a small
article fixture. For phylogenetic ordinal traits, replace it with
`phylo_indep(species, tree = tree)` or another source-specific covariance
keyword after checking the relevant validation row. Source-specific
`*_unique()` syntax remains compatibility spelling for explicit `Psi` in
paired decompositions, not the primary standalone diagonal teaching form.
The family does not change the 4 x 5 covariance keyword grammar.

## Cutpoints

For a `K`-category trait, `gllvmTMB` follows Hadfield's threshold-trait
convention: `tau_1 = 0` is fixed for identifiability, so the model
estimates `K - 2` free cutpoints per ordinal trait:

```text
tau_2, ..., tau_{K - 1}
```

Recover them with `extract_cutpoints()`:

```{r}
cuts <- extract_cutpoints(fit_long)
cuts
```

The returned rows are one row per `(trait, cutpoint)` pair. A
four-category trait has two reported cutpoints, `cutpoint_2` and
`cutpoint_3`; the first threshold `tau_1 = 0` is fixed and is not
reported as an estimated parameter.

This differs from `brms::cumulative()`, which reports `K - 1`
cutpoints under a different location convention. The two conventions
describe the same threshold model up to a shift absorbed by the linear
predictor, but the reported cutpoint count is not the same.

## Variance Components

The main payoff is that ordinal variance components are already on the
latent continuous scale. In the small standalone diagonal example,
`extract_Sigma(fit, level = "unit")` returns the fitted unit-level trait
covariance, and `extract_Sigma(..., link_residual = "auto")` adds the
exact ordinal residual `sigma_d^2 = 1` on the diagonal when needed.

```{r}
Sigma_unit <- extract_Sigma(fit_long, level = "unit")$Sigma
Sigma_total <- extract_Sigma(
  fit_long,
  level = "unit",
  link_residual = "auto"
)$Sigma

Sigma_unit
Sigma_total
```

For a one-trait threshold heritability calculation, the familiar
Dempster-Lerner denominator is direct:

```text
H^2 = sigma_phy^2 / (sigma_phy^2 + 1)
```

The `+ 1` is the fixed latent residual variance. That is the practical
advantage over logit-scale ordinal or binary models when the analysis
needs variance components rather than only fixed-effect contrasts.

## Diagnostics

Run fit-health diagnostics before interpreting the cutpoints or covariance
components. This tiny fixture is useful because all core fit-health rows pass
while still leaving ordinal residual and interval examples outside this page.

```{r diagnostics}
check_gllvmTMB(fit_long)
```

## Guardrails

Use `ordinal_probit()` for ordered categories with `K >= 3`. If a trait
has only two categories, use `binomial(link = "probit")`; the `K = 2`
threshold model reduces exactly to that binomial probit likelihood
(Hadfield 2015, eqn 10).

Do not mix `ordinal_probit()` with another family within the same
trait. Cutpoints are trait-level parameters, so all rows for an
ordinal trait must belong to the ordinal-probit family.

Observation-level `indep()` residuals are structurally unidentifiable
for ordinal-probit traits. The threshold model fixes the residual
scale at 1, and an extra observation-level scale factor can be absorbed
by the cutpoints. The engine maps that observation-level component off
for ordinal-probit traits and reports an informational message.

Correlations from `extract_correlations()` are fitted latent-scale
trait correlations. They are not correlations of the raw integer
category labels. That latent-scale interpretation is usually what you
want for threshold-trait variance components, but it should be named
explicitly in prose and figure captions.

## References

Dempster, E. R. and Lerner, I. M. (1950). Heritability of threshold
characters. *Genetics* 35:212-236.

Falconer, D. S. and Mackay, T. F. C. (1996). *Introduction to
Quantitative Genetics*, 4th ed. Longman.

Felsenstein, J. (2005). Using the quantitative genetic threshold model
for inferences between and within species. *Phil. Trans. R. Soc. B*
360:1427-1434.

Felsenstein, J. (2012). A comparative method for both discrete and
continuous characters using the threshold model. *Am. Nat.*
179:145-156.

Hadfield, J. D. (2015). Increasing the efficiency of MCMC for
hierarchical phylogenetic models of categorical traits using reduced
mixed models. *Methods Ecol. Evol.* 6:706-714.

Mizuno, A. *et al.* (2025). Phylogenetic comparative methods for
threshold traits. *J. Evol. Biol.* 38(12):1699-1712.
