---
title: "Ordinal-probit threshold traits"
tier: 2 # technical reference for the ordinal_probit() family
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Ordinal-probit threshold traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

Use `ordinal_probit()` when a trait is ordered categorical with at
least three categories and the scientific target is a latent continuous
scale: behavioural intensity scores, ordered parasite-load scores,
ranked breeding states, or threshold traits in phylogenetic comparative
analyses. Use `binomial(link = "probit")` instead when the response has
only two categories.

The reason to prefer the probit threshold model is scale. The latent
variable is

```text
y* = eta + epsilon,    epsilon ~ Normal(0, 1)
```

and the observed category is determined by cutpoints:

```text
y = k when tau[k - 1] < y* <= tau[k]
```

Because `epsilon` has variance 1 by construction, the
distribution-specific latent residual is exactly `sigma_d^2 = 1`.
There is no trigamma or logistic `pi^2 / 3` approximation to add
before comparing variance components to a continuous Gaussian trait.

## Fit The Family

Long data are canonical. The response should be integer-coded
`1, ..., K` within each ordinal trait.

```{r}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    phylo_unique(species, tree = tree),
  data = df_long,
  unit = "individual",
  cluster = "species",
  family = ordinal_probit()
)
```

The wide data-frame form uses `traits(...)` on the left-hand side.
The compact wide syntax expands to the same stacked-trait formula
before fitting.

```{r}
fit_wide <- gllvmTMB(
  traits(display_score, parasite_score, rank_score) ~ 1 +
    phylo_unique(species, tree = tree),
  data = df_wide,
  unit = "individual",
  cluster = "species",
  family = ordinal_probit()
)
```

The species-axis phylogenetic keyword already names the grouping axis,
so it is unchanged between long and wide forms. For non-phylogenetic
ordinal traits, replace `phylo_unique()` with the ordinary covariance
keyword you need, such as `latent()` or `unique()`. The family does
not change the 3 x 5 covariance keyword grammar.

## Cutpoints

For a `K`-category trait, `gllvmTMB` follows Hadfield's threshold-trait
convention: `tau_1 = 0` is fixed for identifiability, so the model
estimates `K - 2` free cutpoints per ordinal trait:

```text
tau_2, ..., tau_{K - 1}
```

Recover them with `extract_cutpoints()`:

```{r}
cuts <- extract_cutpoints(fit_long)
cuts
```

The returned rows are one row per `(trait, cutpoint)` pair. A
four-category trait has two reported cutpoints, `cutpoint_2` and
`cutpoint_3`; the first threshold `tau_1 = 0` is fixed and is not
reported as an estimated parameter.

This differs from `brms::cumulative()`, which reports `K - 1`
cutpoints under a different location convention. The two conventions
describe the same threshold model up to a shift absorbed by the linear
predictor, but the reported cutpoint count is not the same.

## Variance Components

The main payoff is that ordinal variance components are already on the
latent continuous scale. In a phylogenetic threshold-trait model,
`extract_Sigma(fit, level = "phy")` returns the phylogenetic trait
covariance, and `extract_Sigma(..., link_residual = "auto")` adds the
exact ordinal residual `sigma_d^2 = 1` on the diagonal when needed.

```{r}
Sigma_phy <- extract_Sigma(fit_long, level = "phy")$Sigma
cuts <- extract_cutpoints(fit_long)
```

For a one-trait threshold heritability calculation, the familiar
Dempster-Lerner denominator is direct:

```text
H^2 = sigma_phy^2 / (sigma_phy^2 + 1)
```

The `+ 1` is the fixed latent residual variance. That is the practical
advantage over logit-scale ordinal or binary models when the analysis
needs variance components rather than only fixed-effect contrasts.

## Guardrails

Use `ordinal_probit()` for ordered categories with `K >= 3`. If a trait
has only two categories, use `binomial(link = "probit")`; the `K = 2`
threshold model reduces exactly to that binomial probit likelihood
(Hadfield 2015, eqn 10).

Do not mix `ordinal_probit()` with another family within the same
trait. Cutpoints are trait-level parameters, so all rows for an
ordinal trait must belong to the ordinal-probit family.

Observation-level `unique()` residuals are structurally unidentifiable
for ordinal-probit traits. The threshold model fixes the residual
scale at 1, and an extra observation-level scale factor can be absorbed
by the cutpoints. The engine maps that observation-level component off
for ordinal-probit traits and reports an informational message.

Correlations from `extract_correlations()` are fitted latent-scale
trait correlations. They are not correlations of the raw integer
category labels. That latent-scale interpretation is usually what you
want for threshold-trait variance components, but it should be named
explicitly in prose and figure captions.

## References

Dempster, E. R. and Lerner, I. M. (1950). Heritability of threshold
characters. *Genetics* 35:212-236.

Falconer, D. S. and Mackay, T. F. C. (1996). *Introduction to
Quantitative Genetics*, 4th ed. Longman.

Felsenstein, J. (2005). Using the quantitative genetic threshold model
for inferences between and within species. *Phil. Trans. R. Soc. B*
360:1427-1434.

Felsenstein, J. (2012). A comparative method for both discrete and
continuous characters using the threshold model. *Am. Nat.*
179:145-156.

Hadfield, J. D. (2015). Increasing the efficiency of MCMC for
hierarchical phylogenetic models of categorical traits using reduced
mixed models. *Methods Ecol. Evol.* 6:706-714.

Mizuno, A. *et al.* (2025). Phylogenetic comparative methods for
threshold traits. *J. Evol. Biol.* 38(12):1699-1712.
