---
title: "A unified phylogenetic + spatial meta-analysis: gllvmTMB-style"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{A unified phylogenetic + spatial meta-analysis: gllvmTMB-style}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

## 1. The Mizuno et al. framework

**Mizuno, Williams, Lagisz, Senior & Nakagawa (2026)** *A unified
framework for phylogenetic and spatial meta-analysis*, EcoEvoRxiv,
<https://doi.org/10.32942/X2WT1B> (companion tutorial code at
<https://ayumi-495.github.io/phylo_spatial_tutorial/>) shows that a
**phylogenetic meta-analysis** and a **spatial meta-analysis** are
the *same* statistical object -- a correlated random-effect model
on a distance metric. The only thing that changes between them is
**what metric defines correlation**: shared evolutionary time for
the phylogenetic case, geographic distance for the spatial case.
Effect sizes $g_k$ with known sampling variances $v_k$ are modelled
as

\[
 g_k \;=\; \mu \;+\; p_{\,i(k)} \;+\; r_{\,s(k)} \;+\; e_k,
 \qquad
 \mathbf{p} \sim \mathcal{N}(\mathbf{0},\,\sigma^{2}_{\text{phy}}\,\mathbf{C}_{\text{phy}}),
 \quad
 \mathbf{r} \sim \mathcal{N}(\mathbf{0},\,\sigma^{2}_{\text{spa}}\,\mathbf{K}_{\text{spa}}),
 \quad
 e_k \sim \mathcal{N}(0, v_k),
\]

with $\mathbf{C}_{\text{phy}}$ a phylogenetic correlation matrix
and $\mathbf{K}_{\text{spa}}$ a Matérn correlation built from
geographic distance. The Mizuno et al. paper covers single-trait
univariate fits in `metafor`, `glmmTMB`, and `brms`. It explicitly
flags **two open gaps**:

1. **Combining phylogeny *and* space in one model** is rarely done
   in practice, even though both terms can be present in the same
   dataset.
2. **Multi-trait outcomes** (e.g. "growth response" *and* "survival
   response" jointly) require an extra trait-covariance layer that
   their three reference packages do not supply directly.

This article fills both gaps using `gllvmTMB`. The same engine that
handles reduced-rank GLLVMs has the building blocks needed:
`phylo_scalar()` for the phylogenetic random effect,
`spatial_unique()` for the sparse-precision spatial Matérn
($\nu = 1$; one independent field per outcome; see
`?spatial_unique` and the
[SPDE benchmark article](spde-vs-glmmTMB.html)),
`meta_known_V()` to plug in known sampling variances, and
`latent()` to add the trait-covariance layer when there is more
than one outcome.

## 2. Simulate a meta-analytic dataset

```{r setup}
library(gllvmTMB)
library(ape)
```

Two outcomes -- *growth response* and *survival response* --
measured across 30 species sharing a phylogeny and 30 study sites
scattered in a unit square. Each outcome contributes 100 effect
sizes with known sampling variances; that is 200 effect sizes in
total. The phylogenetic species effect is shared across outcomes
(one variance); the spatial study effect is independent across
outcomes.

```{r sim}
set.seed(2025)

n_species <- 30
tree <- ape::rcoal(n_species)
tree$tip.label <- paste0("sp", seq_len(n_species))
Cphy <- ape::vcv(tree, corr = TRUE)

n_studies <- 30
study_lon <- runif(n_studies)
study_lat <- runif(n_studies)
D <- as.matrix(stats::dist(cbind(study_lon, study_lat)))
range_sim <- 0.3
K_spa <- exp(-D / range_sim)

sigma2_phy_true <- 0.30
sigma2_spa_true <- 0.20

n_eff_per_outcome <- 100
n_eff <- 2 * n_eff_per_outcome

species_idx <- sample.int(n_species, n_eff, replace = TRUE)
study_idx   <- sample.int(n_studies, n_eff, replace = TRUE)
outcome_lab <- factor(rep(c("growth", "survival"), each = n_eff_per_outcome),
                      levels = c("growth", "survival"))

mu_outcome <- c(growth = 0.20, survival = 0.05)

phy_eff <- as.numeric(t(chol(sigma2_phy_true * Cphy + 1e-8 * diag(n_species))) %*%
                      rnorm(n_species))
spa_eff_growth   <- as.numeric(t(chol(sigma2_spa_true * K_spa + 1e-8 * diag(n_studies))) %*%
                                rnorm(n_studies))
spa_eff_survival <- as.numeric(t(chol(sigma2_spa_true * K_spa + 1e-8 * diag(n_studies))) %*%
                                rnorm(n_studies))

vi <- runif(n_eff, 0.04, 0.16)             # known sampling variances
e  <- rnorm(n_eff, sd = sqrt(vi))           # sampling error per effect
spa_vec <- ifelse(outcome_lab == "growth",
                  spa_eff_growth[study_idx],
                  spa_eff_survival[study_idx])

g <- mu_outcome[as.character(outcome_lab)] +
     phy_eff[species_idx] + spa_vec + e

df <- data.frame(
  study   = factor(paste0("s", study_idx),
                   levels = paste0("s", seq_len(n_studies))),
  species = factor(paste0("sp", species_idx),
                   levels = paste0("sp", seq_len(n_species))),
  lon     = study_lon[study_idx],
  lat     = study_lat[study_idx],
  outcome = outcome_lab,
  value   = g,
  vi      = vi
)
head(df, 4)
```

The `value` column is the Hedges'-$g$-style estimate; `vi` is its
known sampling variance. The columns `species`, `study`, `outcome`
are the three correlation-structure dimensions (phylogenetic,
spatial, trait-covariance).

## 3. Single-trait phylogenetic meta-analysis

Following the Mizuno et al. equation 3 template, slice to the
*growth* outcome and fit a phylogenetic-only meta-analysis:

```{r df1}
df1 <- df[df$outcome == "growth", , drop = FALSE]
df1$outcome <- droplevels(df1$outcome)

## meta_known_V() desugars to equalto(0 + obs | grp_V, V).
## The data must carry an `obs` factor (one level per row) and a
## `grp_V` factor (single level), and known_V must hold the
## diagonal V.
df1$obs   <- factor(seq_len(nrow(df1)))
df1$grp_V <- factor(rep("a", nrow(df1)))
V1 <- diag(df1$vi)
```

The fixed effect is the meta-analytic intercept. We use `~ 1`
rather than `~ 0 + outcome` because the filtered subset has only
one outcome level (`0 + factor` with a single level fails the
base R contrast check).

```{r fit-phy}
fit_phy <- gllvmTMB(
  value ~ 1 + phylo_scalar(species, vcv = Cphy) +
          meta_known_V(value, sampling_var = vi),
  data    = df1,
  trait   = "outcome",
  unit    = "study",
  known_V = V1
)
fit_phy$opt$convergence
fit_phy$report$lam_phy   # phylogenetic variance
```

`lam_phy` is the scale on $\mathbf{C}_{\text{phy}}$, i.e.
$\hat\sigma^{2}_{\text{phy}}$. It should be near the true value of
0.30. This matches the structure of Mizuno et al. equation 3: a
Gaussian outcome with a phylogenetic random effect and a fixed
sampling-error term.

## 4. Single-trait spatial meta-analysis

Same data, but use a **spatial** correlation in place of the
phylogenetic one (Mizuno et al. equation 4). Build a triangulated
mesh on the study coordinates and call `spatial_unique()`:

```{r fit-spa}
mesh1 <- make_mesh(df1, c("lon", "lat"), cutoff = 0.07)
mesh1$mesh$n      # number of mesh nodes

fit_spa <- gllvmTMB(
  value ~ 1 + spatial_unique(0 + outcome | coords) +
              meta_known_V(value, sampling_var = vi),
  data    = df1,
  mesh    = mesh1,
  trait   = "outcome",
  unit    = "study",
  known_V = V1
)
fit_spa$opt$convergence
as.numeric(fit_spa$report$kappa)               # inverse-range
sqrt(8) / as.numeric(fit_spa$report$kappa)     # practical range
```

The estimated practical range -- the distance at which spatial
correlation drops to $\sim 0.13$ -- should sit near the simulated
correlation length 0.3. Note that the simulator drew an
*exponential* field (Matérn $\nu = 1/2$) while `spatial_unique()`
fits Matérn $\nu = 1$, so the two ranges are not numerically
identical; the *shape* of the spatial decay is what is being
recovered.

## 5. Combined phylogenetic + spatial meta-analysis (gap 1)

Mizuno et al. note that combining both correlation sources in one
model is rarely done in practice. With `gllvmTMB` the syntax is
just the union of sections 3 and 4:

```{r fit-both}
fit_both <- gllvmTMB(
  value ~ 1 + phylo_scalar(species, vcv = Cphy) +
              spatial_unique(0 + outcome | coords, mesh = mesh1) +
              meta_known_V(value, sampling_var = vi),
  data    = df1,
  trait   = "outcome",
  unit    = "study",
  known_V = V1
)
fit_both$opt$convergence
```

```{r both-vc}
data.frame(
  source            = c("single phylo", "single spatial",
                        "combined phylo", "combined spatial"),
  variance_or_kappa = c(
    fit_phy$report$lam_phy,
    as.numeric(fit_spa$report$kappa),
    fit_both$report$lam_phy,
    as.numeric(fit_both$report$kappa)
  )
)
sqrt(8) / as.numeric(fit_both$report$kappa)    # spatial practical range, combined
```

A few practical notes on identifiability:

* When phylogenetic and spatial signals overlap in *space* -- i.e.
  closely related species tend to be sampled at nearby sites --
  the two random-effect terms compete for variance and the
  individual variance estimates can shrink relative to the
  single-source fits. In our simulated data the species and study
  indices were drawn independently, so the two variances are well
  separated.
* The phylogenetic variance scalar (`lam_phy`) is shared across
  outcomes when more than one outcome is present
  (`phylo_scalar()` desugars to `propto(0 + species | trait)` with
  a single $\sigma^2_{\text{phy}}$ parameter).
* The SPDE field has its own per-trait precision parameter
  (`tau`) and a *shared* range parameter (`kappa`).

## 6. Multi-trait meta-analysis (gap 2)

Now use **both** outcomes jointly. The extra layer is
`latent(0 + outcome | study, d = 1)`, a rank-1 reduced-rank
between-study term that captures *correlation between outcomes at
the study level* -- a quantity neither `metafor`, `glmmTMB`, nor
`brms` exposes as cleanly when paired with phylogeny and a
sparse-precision spatial field.

```{r df-both}
df_both <- df
df_both$obs   <- factor(seq_len(nrow(df_both)))
df_both$grp_V <- factor(rep("a", nrow(df_both)))
V_both <- diag(df_both$vi)
mesh_both <- make_mesh(df_both, c("lon", "lat"), cutoff = 0.07)
mesh_both$mesh$n
```

```{r fit-multi}
fit_multi <- gllvmTMB(
  value ~ 0 + outcome +
              latent(0 + outcome | study, d = 1) +
              phylo_scalar(species, vcv = Cphy) +
              spatial_unique(0 + outcome | coords, mesh = mesh_both) +
              meta_known_V(value, sampling_var = vi),
  data    = df_both,
  trait   = "outcome",
  unit    = "study",
  known_V = V_both
)
fit_multi$opt$convergence
fit_multi$report$lam_phy                          # phylogenetic variance
sqrt(8) / as.numeric(fit_multi$report$kappa)      # spatial practical range
```

The `latent(0 + outcome | study, d = 1)` term gives a rank-1 trait
covariance $\boldsymbol{\Sigma}_{\text{unit}} = \boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^{\top}$.
This is where `gllvmTMB` extends beyond the Mizuno et al. trio:
combining phylogeny *and* a sparse-precision spatial field *and* a
trait-covariance layer in one fit, all under a single TMB
negative log-likelihood.

## 7. Block-diagonal V: when effect sizes within a study are correlated

Sections 3--6 use **diagonal** `V` (`known_V = diag(vi)`), which
assumes the sampling errors of different rows are independent.
That holds when every row comes from a distinct study, but a
meta-analyst often has *several effect sizes from the same study*
(multiple outcomes, multiple time points, or multiple
sub-experiments) -- and within-study sampling errors are then
correlated.

`gllvmTMB::block_V()` builds the standard compound-symmetric
block-diagonal `V` for this case:
$\mathbf{V}_s = \mathbf{D}_s^{1/2} \mathbf{R}_s \mathbf{D}_s^{1/2}$
within each study, with $\mathbf{R}_s$ having 1 on the diagonal
and `rho_within` on the off-diagonal.

```{r block-V-demo}
## Build a block-diagonal V from the simulated multi-trait data,
## treating the two outcomes from the same study as correlated
## (rho_within = 0.5).
V_blk <- block_V(study_id     = df_both$study,
                 sampling_var = df_both$vi,
                 rho_within   = 0.5)

## Sanity-check: same diagonal as before, plus within-study
## off-diagonals
all.equal(diag(V_blk), df_both$vi)
range(V_blk[upper.tri(V_blk)])     # > 0 within studies, 0 across
```

Refit the multi-trait model with the block-diagonal `V`:

```{r fit-blockV}
fit_multi_block <- gllvmTMB(
  value ~ 0 + outcome +
          latent(0 + outcome | study, d = 1) +
          phylo_scalar(species, vcv = Cphy) +
          spatial_unique(0 + outcome | coords, mesh = mesh_both) +
          meta_known_V(value, sampling_var = vi),
  data    = df_both,
  trait   = "outcome",
  unit    = "study",
  known_V = V_blk
)

c(lam_phy_blk = as.numeric(fit_multi_block$report$lam_phy),
  range_blk   = sqrt(8) / as.numeric(fit_multi_block$report$kappa))
```

Compared to the diagonal-V fit in section 6, the block-diagonal
fit typically *deflates* the residual between-study variance --
because the within-study correlation that the diagonal model
attributed to "genuine between-study heterogeneity" is now
correctly attributed to known sampling correlation. In a real
meta-analysis where you have genuine reason to believe
within-study sampling correlation, this is the correct
specification.

The `meta_known_V(value, V = V)` term in the formula handles the
known-V propagation; pass `known_V = V` to `gllvmTMB()` alongside
it.

## 8. Diagnostics and interpretation

```{r diagnose}
gllvmTMB_diagnose(fit_multi, verbose = TRUE)
```

The implied between-study trait covariance and correlation
matrices:

```{r sigma-b}
SB <- extract_Sigma(fit_multi, level = "unit")
round(SB$Sigma, 3)
round(SB$R,     3)
```

Interpretation for the meta-analyst:

* **$\boldsymbol{\Sigma}_{\text{unit}}$ diagonals** are
  between-study residual variances of each outcome *after*
  phylogeny, geography, and within-study sampling error have been
  removed. They tell you how much study-to-study heterogeneity
  remains in the trait-covariance layer.
* **$\mathbf{R}_{\text{unit}}$ off-diagonal** is the correlation
  between *growth* and *survival* responses across studies. A
  strongly positive value means studies that report large
  positive effects on growth also tend to report large positive
  effects on survival, after the other structures are accounted
  for.
* `lam_phy` and the SPDE practical range ($\sqrt{8}/\hat\kappa$)
  are the same quantities reported in sections 3--5, now estimated
  jointly with the trait-covariance layer.
* Warnings from `gllvmTMB_diagnose()` on small simulated data are
  normal: meta-analytic intercepts are often weakly identified
  when most of the variance is absorbed by the random-effect
  structure, which inflates fixed-effect standard errors. In a
  real application, check the diagnostic on the full dataset
  rather than on a 200-row simulation.

## 9. See also

* [Choose your model](choose-your-model.html) -- decision tree.
* [Stacked-trait GLLVM](stacked-trait-gllvm.html) -- the full
  six-piece decomposition with replicated trait data
  (the alternative to two-stage when data shape allows).
* [Corvidae two-stage workflow](corvidae-two-stage.html) -- a
  different `meta_known_V()` use case: species-level trait
  constants via stage-1 / stage-2 meta-analysis.
* [Phylogenetic GLLVM](phylogenetic-gllvm.html) -- the multi-trait
  reduced-rank phylogenetic case.
* [SPDE benchmark](spde-vs-glmmTMB.html) -- why the spatial term
  is a Matérn $\nu = 1$ sparse-precision field rather than the
  exponential dense one used by `metafor::SPEXP` and
  `glmmTMB::exp()`.
* [Cross-package validation](cross-package-validation.html) --
  agreement of `meta_known_V()` against `glmmTMB::equalto()`.
* [Simulation-based recovery](simulation-recovery.html) -- recovery
  of `phylo_scalar()` and `spatial_unique()` parameters on
  simulated data.
* [Pitfalls](pitfalls.html) -- identifiability and common mistakes
  when stacking covariance structures.
* **Mizuno, Williams, Lagisz, Senior & Nakagawa (2026)** *A unified
  framework for phylogenetic and spatial meta-analysis*,
  EcoEvoRxiv, <https://doi.org/10.32942/X2WT1B> -- the conceptual
  paper; this article extends its single-trait univariate
  framework into the combined-and-multivariate case.
