---
title: "Phylogenetic trait covariance"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Phylogenetic trait covariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = requireNamespace("ape", quietly = TRUE),
  message  = FALSE,
  warning  = FALSE
)
```

This article is for a comparative-methods reader with several traits
measured once per species. The working question is: can we separate
phylogenetically conserved trait covariance from non-phylogenetic
species-level covariance, while still keeping trait-specific variation
on the diagonal?

The model has two species-level trait-covariance matrices. The
phylogenetic component uses the tree correlation matrix `A`; the
non-phylogenetic component uses independent species effects:

$$
g_{\mathrm{phy}} \sim
\mathrm{MVN}(0, \Sigma_{\mathrm{phy}} \otimes A),
\qquad
g_{\mathrm{non}} \sim
\mathrm{MVN}(0, \Sigma_{\mathrm{non}} \otimes I).
$$

Each covariance can use the same `latent() + unique()` decomposition:

$$
\Sigma_{\mathrm{phy}} =
\Lambda_{\mathrm{phy}}\Lambda_{\mathrm{phy}}^T + \boldsymbol\Psi_{\mathrm{phy}},
\qquad
\Sigma_{\mathrm{non}} =
\Lambda_{\mathrm{non}}\Lambda_{\mathrm{non}}^T + \boldsymbol\Psi_{\mathrm{non}}.
$$

Here `Lambda` is the shared low-rank structure and `Psi` is the
trait-specific diagonal variance. The "two-Psi" nickname refers to
this paired phylogenetic and non-phylogenetic decomposition (two
`Psi` diagonals, one at each tier).

## Setup

```{r setup}
library(gllvmTMB)
library(ape)
```

## Simulate

We simulate three traits on 30 species. The truth has four
components, matching the paired decomposition in the theory section:
a phylogenetic shared axis (`Lambda_phy`), a phylogenetic diagonal
component (`psi_phy`), a non-phylogenetic shared axis (`Lambda_non`),
and a non-phylogenetic diagonal component (`psi_non`). This small
example is meant to show syntax and extraction, not to be a full
simulation study.

```{r sim}
set.seed(13)
n_sp <- 30
n_traits <- 3
trait_names <- paste0("trait_", seq_len(n_traits))

tree <- ape::rcoal(n_sp)
tree$tip.label <- paste0("sp", seq_len(n_sp))
Cphy <- ape::vcv(tree, corr = TRUE)
Lphy <- t(chol(Cphy + 1e-8 * diag(n_sp)))

Lambda_phy_true <- matrix(c(0.70, 0.45, 0.25), n_traits, 1)
Lambda_non_true <- matrix(c(0.50, 0.30, 0.20), n_traits, 1)
s_phy_true <- c(0.20, 0.15, 0.12)
s_non_true <- c(0.18, 0.20, 0.16)

g_phy <- Lphy %*% rnorm(n_sp)
g_non <- matrix(rnorm(n_sp), n_sp, 1)
e_phy <- Lphy %*% matrix(rnorm(n_sp * n_traits), n_sp, n_traits)
e_non <- matrix(rnorm(n_sp * n_traits), n_sp, n_traits)

Y <- g_phy %*% t(Lambda_phy_true) +
  sweep(e_phy, 2, sqrt(s_phy_true), "*") +
  g_non %*% t(Lambda_non_true) +
  sweep(e_non, 2, sqrt(s_non_true), "*")
rownames(Y) <- tree$tip.label
colnames(Y) <- trait_names

df <- data.frame(
  species = factor(rep(tree$tip.label, each = n_traits),
                   levels = tree$tip.label),
  trait = factor(rep(trait_names, times = n_sp),
                 levels = trait_names),
  value = as.numeric(t(Y))
)

df_wide <- data.frame(
  species = factor(tree$tip.label, levels = tree$tip.label),
  Y,
  check.names = FALSE
)
head(df)
```

## Fit

The long form is canonical: one row per `(species, trait)` observation.
The wide data-frame form is the same model written from one row per
species and one column per trait. Because `traits(...)` names the
response traits on the left-hand side, the right-hand side can use the
compact covariance syntax.

```{r fit}
# Long format (canonical)
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    phylo_latent(species, d = 1, tree = tree) +
    phylo_unique(species, tree = tree) +
    latent(0 + trait | species, d = 1) +
    unique(0 + trait | species),
  data = df,
  unit = "species",
  cluster = "species"
)

# Wide data-frame format (same model with compact formula syntax)
fit_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~ 1 +
    phylo_latent(species, d = 1, tree = tree) +
    phylo_unique(species, tree = tree) +
    latent(1 | species, d = 1) +
    unique(1 | species),
  data = df_wide,
  unit = "species",
  cluster = "species"
)

fit <- fit_long
fit$opt$convergence

cat("Long/wide logLik difference:",
    signif(abs(as.numeric(logLik(fit_long)) -
               as.numeric(logLik(fit_wide))), 3), "\n")
```

The two forms dispatch to the same long-format engine. In this model,
`phylo_latent(species, d = 1, tree = tree)` estimates the shared
phylogenetic axis, `phylo_unique(species, tree = tree)` estimates the
trait-specific phylogenetic diagonal, `latent(1 | species, d = 1)` in
the wide formula (expanding to `latent(0 + trait | species, d = 1)`
in the long form) estimates the shared non-phylogenetic axis, and
`unique(1 | species)` (expanding to `unique(0 + trait | species)`)
estimates the trait-specific non-phylogenetic diagonal.

`gllvmTMB_wide()` is not the right shortcut for this row-phylogeny
example. That matrix entry point treats matrix columns as the trait
axis and is most natural for site-by-species response matrices, where
the phylogeny belongs to the response columns. Here the phylogeny
belongs to the row unit `species`, so the formula-wide `traits(...)`
path is the readable wide form.

## Recover the phylogenetic covariance

`extract_Sigma(level = "phy")` returns the phylogenetic trait
covariance. The shared part is
`Lambda_phy %*% t(Lambda_phy)`; the unique part is the diagonal
`psi_phy`; the total is their sum.

```{r extract-phy}
Sigma_phy_shared <- extract_Sigma(fit, level = "phy", part = "shared")$Sigma
Sigma_phy_unique <- extract_Sigma(fit, level = "phy", part = "unique")$s
Sigma_phy_total <- extract_Sigma(fit, level = "phy", part = "total")$Sigma

round(Sigma_phy_shared, 2)
round(Sigma_phy_unique, 2)
round(Sigma_phy_total, 2)
```

The non-phylogenetic species-level covariance has the same paired
decomposition as the phylogenetic side:

```{r extract-non}
Sigma_non_shared <- extract_Sigma(fit, level = "unit", part = "shared")$Sigma
Sigma_non_unique <- extract_Sigma(fit, level = "unit", part = "unique")$s
Sigma_non_total  <- extract_Sigma(fit, level = "unit", part = "total")$Sigma

round(Sigma_non_shared, 2)
round(Sigma_non_unique, 2)
round(Sigma_non_total, 2)
```

For a small tree, do not over-interpret either split (between
`Lambda_phy Lambda_phy^T` and `Psi_phy`, or between
`Lambda_non Lambda_non^T` and `Psi_non`). The more stable biological
targets are often the totals `Sigma_phy` and `Sigma_non`, especially
when the tree is small or the ranks are exploratory.

## Communality (within each tier)

Communality `c_t^2` is the fraction of a trait's within-tier variance
explained by the shared latent axis. It is defined only when both a
`latent()` and a `unique()` term are fit at that tier; that is what
the four-component decomposition gives us on both the phylogenetic and
non-phylogenetic sides.

```{r communality}
# Non-phylogenetic communality is one extractor call.
c2_non <- extract_communality(fit, level = "unit")
round(c2_non, 2)

# Phylogenetic communality is the same ratio, computed from the
# Sigma_phy decomposition above.
c2_phy <- diag(Sigma_phy_shared) / diag(Sigma_phy_total)
round(c2_phy, 2)
```

A value near 1 means the trait's within-tier variance is dominated by
the shared axis (high "integration" with the other traits at that
tier). A value near 0 means the trait is mostly idiosyncratic at that
tier.

## Phylogenetic heritability

Phylogenetic heritability `H^2` is a between-tier ratio: the fraction
of total trait variance attributable to the phylogenetic component
(both shared and unique parts of `Sigma_phy`) versus the
non-phylogenetic component (both shared and unique parts of
`Sigma_non`). It is distinct from the within-tier communalities
above.

The convenience extractor returns the full three-way decomposition
`H^2 + C^2_non + Psi = 1` per trait, where:

- `H^2` = phylogenetic share = `diag(Sigma_phy_total) / V_eta`
- `C^2_non` = non-phylogenetic *shared* share =
  `diag(Lambda_non Lambda_non^T) / V_eta`
- `Psi` = non-phylogenetic *unique* share = `psi_non / V_eta`

with `V_eta = diag(Sigma_phy_total) + diag(Lambda_non Lambda_non^T) + psi_non`.

```{r phylo-heritability}
extract_phylo_signal(fit)
```

`C^2_non` is structurally zero when the formula omits the non-phylogenetic
`latent()` term, so the three-way split collapses to `H^2 + Psi`. Adding
`latent(0 + trait | species, d = K)` (long) or `latent(1 | species, d = K)`
(wide), as this article does, is what makes `C^2_non` a genuine
non-phylogenetic shared-axis component rather than a placeholder.

## Compare against simpler and saturated baselines

Three nested covariance intents bracket the four-component model:

- **Independent** (`phylo_indep + indep`): per-trait diagonal only at
  each tier. No shared axis, no cross-trait covariance. The simplest
  model that respects the phylogenetic structure.
- **Two-U** (`phylo_latent + phylo_unique + latent + unique`): the
  four-component decomposition fit in this article. A shared
  low-rank axis plus a trait-specific diagonal at each tier.
- **Dependent** (`phylo_dep + dep`): a full unstructured covariance
  at each tier. Saturated; no shared/unique split is enforced.

The two-Psi decomposition sits between independent (too restrictive
when traits are correlated) and dependent (too many parameters when
the number of traits grows). The cross-check functions refit the
data under the independent and dependent baselines and compare each
component's implied total covariance against the two-Psi fit.

```{r compare-indep}
diag_indep <- compare_indep_vs_two_psi(fit)
diag_indep$agreement
diag_indep$flag
```

```{r compare-dep}
diag_dep <- compare_dep_vs_two_psi(fit)
diag_dep$agreement
diag_dep$flag
```

Each `$agreement` row shows the Frobenius relative disagreement
between the two-Psi fit and the baseline (`rmse / baseline_mag`).
`flag = TRUE` means relative disagreement exceeds `0.10`. A flag is
not a failed model; it tells you the data do not strongly support
the chosen decomposition at the requested rank.

Use `compare_indep_vs_two_psi()` first when the number of traits is
large -- it only refits the diagonal and stays cheap. Use
`compare_dep_vs_two_psi()` when the number of traits is modest and the
saturated refit is affordable -- it is the more informative
benchmark for off-diagonal cross-trait structure.

## When to simplify

The paired `phylo_latent() + phylo_unique() + latent() + unique()`
model is useful when the question is about phylogenetically conserved
trait syndromes alongside non-phylogenetic shared-axis structure.
Simplify when the data do not identify the splits the model asks for:

```{r fallback, eval = FALSE}
fit_phylo_dep <- gllvmTMB(
  value ~ 0 + trait + phylo_dep(0 + trait | species, tree = tree) +
                      dep(0 + trait | species),
  data = df,
  unit = "species",
  cluster = "species"
)
```

`phylo_dep() + dep()` estimates full unstructured trait covariances
at each tier without asking the data to separate a low-rank shared
axis from the diagonal. It is the natural fallback when the two-Psi
diagnostics flag.

## What to read next

- [`compare_dep_vs_two_psi()`](../reference/compare_dep_vs_two_psi.html)
  and [`compare_indep_vs_two_psi()`](../reference/compare_indep_vs_two_psi.html)
  are the likelihood-based cross-checks for this model.
- [`functional-biogeography`](functional-biogeography.html) shows how
  phylogenetic terms behave as controls in a larger site--species--trait
  model.
- [`morphometrics`](morphometrics.html) is the simplest non-phylogenetic
  `latent() + unique()` worked example.

References: Hadfield and Nakagawa (2010) for the phylogenetic mixed
model representation; Meyer and Kirkpatrick (2008) for reduced-rank
covariance estimation; Williams et al. (2025) for fast phylogenetic
GLMMs in the `glmmTMB` ecosystem.
