---
title: "Phylogenetic covariance among traits"
description: "Use replicated Gaussian species traits and a supplied tree to estimate phylogenetically structured covariance with explicit label and health checks."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Phylogenetic covariance among traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  fig.align = "center",
  cache = FALSE,
  eval = requireNamespace("ape", quietly = TRUE),
  message = FALSE
)
```

> **Evidence boundary.** This article uses the package's public interfaces.
> The estimates below are conditional on the supplied tree and fitted
> Gaussian model.

Closely related species can resemble one another across several traits. The
question in this article is: **after allowing for measurement-level noise,
which trait covariances are associated with a supplied phylogeny?**

The first example estimates that phylogenetic covariance directly. A second
example then asks the complementary question: **how much species-level trait
covariance follows the tree, and how much remains non-phylogenetic?**

We use three standardized plant traits measured on ten individuals from each
of 150 species. Replication within species matters here: it gives the model
information to distinguish persistent species differences from
individual-level noise. If you have only one species mean per trait, those
sources are harder to separate and the resulting covariance should be
interpreted accordingly.

## The model

For individual \(i\), species \(s\), and trait \(t\), the example fits

$$
y_{ist} = \mu_t + u_{st} + \varepsilon_{ist},
\qquad
\operatorname{vec}(U) \sim
\operatorname{MVN}(0, \Sigma_{\mathrm{phy,total}} \otimes A),
\qquad
\Sigma_{\mathrm{phy,total}} =
\underbrace{\Lambda_{\mathrm{phy}}\Lambda_{\mathrm{phy}}^{\mathsf T}}_
  {\Sigma_{\mathrm{phy,shared}}} + \Psi_{\mathrm{phy}}.
$$

The supplied matrix \(A\) is the Brownian correlation matrix implied by the
tree. The rank-one matrix
\(\Sigma_{\mathrm{phy,shared}}\) describes one shared phylogenetic trait axis,
while diagonal \(\Psi_{\mathrm{phy}}\) allows additional trait-specific
phylogenetic variance. Their sum is the total phylogenetic covariance.
Independent Gaussian record-level errors \(\varepsilon_{ist}\) account for
variation among individual measurements.

This model does **not** estimate the tree, Pagel's lambda, tree uncertainty, or
genetic heritability. It estimates covariance attributed to one fixed
relationship matrix under the fitted model. The main route throughout this
article is `phylo_latent(..., unique = TRUE)`, which exposes the shared,
diagonal, and total matrices through the public extractor. Other covariance
modes are listed in the
[Formula keyword grid](api-keyword-grid.html#phylogenetic).

::: {.table-responsive}
| Symbol in prose | Keyword / covstruct | DGP draw | Recovery extractor | Planted truth |
|---|---|---|---|---|
| \(\Sigma_{\mathrm{phy,shared}} = \Lambda_{\mathrm{phy}}\Lambda_{\mathrm{phy}}^{\mathsf T}\) | `phylo_latent(..., d = 1, unique = TRUE)` | `phy_shared` with row covariance `A` | `extract_Sigma(level = "phy", part = "shared")` | `Sigma_phy_shared_true` |
| \(\Psi_{\mathrm{phy}}\) | diagonal companion in the same term | `phy_unique` with row covariance `A` | `diag(extract_Sigma(level = "phy", part = "unique")$s)` | `Psi_phy_true` |
| \(\Sigma_{\mathrm{phy,total}} = \Sigma_{\mathrm{phy,shared}} + \Psi_{\mathrm{phy}}\) | same `phylo_latent(..., unique = TRUE)` term | `phy_shared + phy_unique` | `extract_Sigma(level = "phy", part = "total")` | `Sigma_phy_total_true` |
:::

## Simulate one truth-checked example

```{r setup}
library(gllvmTMB)
library(ape)
```

The simulation uses biologically named, standardized traits so that the
correlations have a concrete interpretation. It is one reproducible teaching
dataset, not a repeated-sampling validation study.

```{r simulate}
set.seed(7311)
n_species <- 150L
n_per_species <- 10L
trait_names <- c("leaf_area", "wood_density", "seed_mass")
n_traits <- length(trait_names)

tree <- ape::rcoal(n_species)
tree$tip.label <- paste0("sp", seq_len(n_species))
A <- ape::vcv(tree, corr = TRUE)

Lambda_phy_true <- matrix(c(0.75, 0.48, 0.35), n_traits, 1)
Sigma_phy_shared_true <- Lambda_phy_true %*% t(Lambda_phy_true)
Psi_phy_true <- diag(c(0.25, 0.30, 0.25))
Sigma_phy_total_true <- Sigma_phy_shared_true + Psi_phy_true
for (nm in c(
  "Sigma_phy_shared_true", "Psi_phy_true", "Sigma_phy_total_true"
)) {
  x <- get(nm)
  dimnames(x) <- list(trait_names, trait_names)
  assign(nm, x)
}

L_A <- t(chol(A + 1e-9 * diag(n_species)))
phy_shared <- L_A %*%
  matrix(rnorm(n_species), n_species, 1) %*%
  t(Lambda_phy_true)
phy_unique <- L_A %*%
  matrix(rnorm(n_species * n_traits), n_species, n_traits) %*%
  chol(Psi_phy_true)
species_effect <- phy_shared + phy_unique

trait_means <- c(leaf_area = 0.5, wood_density = 0, seed_mass = -0.5)
residual_sd <- 0.35

df_wide <- expand.grid(
  replicate = seq_len(n_per_species),
  species = tree$tip.label,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
df_wide$species <- factor(df_wide$species, levels = tree$tip.label)
df_wide$individual <- factor(
  paste(df_wide$species, df_wide$replicate, sep = "_")
)

for (j in seq_along(trait_names)) {
  nm <- trait_names[[j]]
  df_wide[[nm]] <- trait_means[[nm]] +
    species_effect[as.integer(df_wide$species), j] +
    rnorm(nrow(df_wide), sd = residual_sd)
}

df_long <- do.call(
  rbind,
  lapply(seq_along(trait_names), function(j) {
    data.frame(
      individual = df_wide$individual,
      species = df_wide$species,
      trait = trait_names[[j]],
      value = df_wide[[trait_names[[j]]]]
    )
  })
)
df_long$species <- factor(df_long$species, levels = tree$tip.label)
df_long$trait <- factor(df_long$trait, levels = trait_names)

head(df_long)
```

## Match the data to the tree

Tree labels and the grouping column must name the same species. Check this
before fitting; silently relying on factor order is unsafe.

```{r tree-data-check}
data_species <- unique(as.character(df_long$species))

alignment <- list(
  data_not_in_tree = setdiff(data_species, tree$tip.label),
  tree_not_in_data = setdiff(tree$tip.label, data_species),
  duplicated_tree_labels = tree$tip.label[duplicated(tree$tip.label)],
  non_positive_branch_lengths = which(tree$edge.length <= 0)
)
alignment

stopifnot(
  lengths(alignment) == 0L,
  identical(levels(df_long$species), tree$tip.label)
)
```

With real data, prune unused tips deliberately and then rebuild the species
factor in the retained tip order. Do not continue when data species are absent
from the tree, tip labels are duplicated, or the covariance derived from the
tree is invalid. Those are data problems, not optimizer problems.

## Fit the same model in long and wide form

The long form is canonical: one row per individual-trait observation. The wide
form is useful when each individual occupies one row and traits occupy
columns. Both calls use the same `gllvmTMB()` entry point.

```{r fit}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    phylo_latent(
      0 + trait | species,
      d = 1,
      tree = tree,
      unique = TRUE
    ),
  data = df_long,
  trait = "trait",
  unit = "individual",
  cluster = "species",
  family = gaussian()
)

fit_wide <- gllvmTMB(
  traits(leaf_area, wood_density, seed_mass) ~ 1 +
    phylo_latent(
      1 | species,
      d = 1,
      tree = tree,
      unique = TRUE
    ),
  data = df_wide,
  unit = "individual",
  cluster = "species",
  family = gaussian()
)

c(
  long = as.numeric(logLik(fit_long)),
  wide = as.numeric(logLik(fit_wide)),
  absolute_difference = abs(
    as.numeric(logLik(fit_long)) - as.numeric(logLik(fit_wide))
  )
)
```

The likelihoods agree up to numerical precision. This checks the data-shape
translation; it does not provide two independent scientific fits.

## Check numerical health before interpretation

`check_gllvmTMB()` is the detailed numerical-health table. Inspect every row;
do not infer component recovery from optimizer or curvature checks.

```{r diagnostics}
health <- check_gllvmTMB(fit_long)
health[
  ,
  c("component", "status", "value", "threshold", "message", "action")
]
```

All 13 rows pass for this fit. The raw maximum gradient is `0.0009637`, below
`0.01`; `sdreport()` is available and the Hessian is positive definite with
reported rank `10/10`.

A passing optimizer and raw maximum gradient below `0.01` support the fitted
point; usable `sdreport()` and a positive-definite Hessian support local Wald
calculations. Neither result proves that the shared and diagonal phylogenetic
components are scientifically well recovered. In a simulation, compare each
component with its known truth; in real data, examine sensitivity to rank,
tree, influential species, and sampling design.

## Compare the fitted covariance with truth

The public extractor returns the shared, diagonal, and total covariance
attributed to the supplied phylogenetic matrix.

```{r covariance-comparison}
Sigma_phy_shared_hat <- extract_Sigma(
  fit_long,
  level = "phy",
  part = "shared"
)$Sigma
Psi_phy_hat <- diag(extract_Sigma(
  fit_long,
  level = "phy",
  part = "unique"
)$s)
Sigma_phy_total_hat <- extract_Sigma(
  fit_long,
  level = "phy",
  part = "total"
)$Sigma

relative_error_first <- function(estimate, truth) {
  norm(estimate - truth, "F") / norm(truth, "F")
}
recovery_first <- data.frame(
  component = c("Sigma_phy_shared", "Psi_phy", "Sigma_phy_total"),
  relative_error = round(c(
    relative_error_first(Sigma_phy_shared_hat, Sigma_phy_shared_true),
    relative_error_first(Psi_phy_hat, Psi_phy_true),
    relative_error_first(Sigma_phy_total_hat, Sigma_phy_total_true)
  ), 3)
)
recovery_first

round(Sigma_phy_shared_hat, 3)
round(Psi_phy_hat, 3)
round(Sigma_phy_total_hat, 3)

sigma_comparison <- compare_Sigma_table(
  fit_long,
  truth = Sigma_phy_total_true,
  level = "phy",
  part = "total",
  measure = "covariance",
  entries = "unique"
)
sigma_comparison[
  ,
  c("trait_i", "trait_j", "estimate", "truth", "error")
]
```

The estimates need not equal truth in one simulated dataset. Read the three
relative errors separately. They are `0.173` for the shared covariance,
`0.228` for phylogenetic \(\Psi\), and `0.096` for their total in this seeded
fit. Better recovery of the total does not imply that the shared axis and
diagonal companion are individually well separated. Numerical convergence is
necessary, but the scientific targets still have finite-sample uncertainty.

The same comparison is often easier to read on the correlation scale.

```{r correlation-comparison, fig.cap = "Estimated minus true phylogenetic correlation for each trait pair. Segments are comparison errors, not confidence intervals.", fig.alt = "A horizontal comparison plot for leaf area, wood density, and seed mass. Each point shows the fitted phylogenetic correlation minus the known simulated correlation, with zero indicating exact recovery."}
correlation_comparison <- compare_Sigma_table(
  fit_long,
  truth = Sigma_phy_total_true,
  level = "phy",
  part = "total",
  measure = "correlation",
  entries = "upper"
)
plot_Sigma_comparison(
  correlation_comparison,
  measure = "correlation"
)
```

This page intentionally reports point estimates only. It does not claim
calibrated confidence-interval coverage for source-specific covariance or
correlation summaries.

## Add a non-phylogenetic species component

Species can share trait combinations for reasons that are not represented by
the supplied tree. To separate the two sources, write the species effect as

$$
b_s = p_s + q_s,
\qquad
\operatorname{vec}(P) \sim
\operatorname{MVN}(0, \Sigma_{\mathrm{phy}} \otimes A),
\qquad
\operatorname{vec}(Q) \sim
\operatorname{MVN}(0, \Sigma_{\mathrm{non}} \otimes I).
$$

Each source is decomposed into a rank-one shared covariance and a diagonal
trait-specific companion:

$$
\Sigma_{\mathrm{phy,total}} =
\underbrace{\Lambda_{\mathrm{phy}}\Lambda_{\mathrm{phy}}^{\mathsf T}}_
  {\Sigma_{\mathrm{phy,shared}}} + \Psi_{\mathrm{phy}},
\qquad
\Sigma_{\mathrm{non,total}} =
\underbrace{\Lambda_{\mathrm{non}}\Lambda_{\mathrm{non}}^{\mathsf T}}_
  {\Sigma_{\mathrm{non,shared}}} + \Psi_{\mathrm{non}}.
$$

The species-level covariance is therefore
\(\Omega_{\mathrm{species}} = \Sigma_{\mathrm{phy,total}} +
\Sigma_{\mathrm{non,total}}\). The alignment table below fixes the meaning of
every claimed covariance before we simulate or fit it.

::: {.table-responsive}
| Symbol in prose | Keyword / covstruct | DGP draw | Recovery extractor | Planted truth |
|---|---|---|---|---|
| \(\Sigma_{\mathrm{phy,shared}} = \Lambda_{\mathrm{phy}}\Lambda_{\mathrm{phy}}^{\mathsf T}\) | `phylo_latent(0 + trait \| species, d = 1, tree = tree_split, unique = TRUE)` | `phy_shared_split` with row covariance `A_split` | `extract_Sigma(level = "phy", part = "shared")` | `Sigma_phy_shared_split_true` |
| \(\Psi_{\mathrm{phy}}\) | diagonal companion in the same `phylo_latent(..., unique = TRUE)` term | `phy_unique_split` with row covariance `A_split` | `diag(extract_Sigma(level = "phy", part = "unique")$s)` | `Psi_phy_split_true` |
| \(\Sigma_{\mathrm{phy,total}} = \Sigma_{\mathrm{phy,shared}} + \Psi_{\mathrm{phy}}\) | same `phylo_latent(..., unique = TRUE)` term | `phy_shared_split + phy_unique_split` | `extract_Sigma(level = "phy", part = "total")` | `Sigma_phy_total_split_true` |
| \(\Sigma_{\mathrm{non,shared}} = \Lambda_{\mathrm{non}}\Lambda_{\mathrm{non}}^{\mathsf T}\) | `latent(0 + trait \| species, d = 1, unique = TRUE)` | `non_shared_split` with independent species rows | `extract_Sigma(level = "unit", part = "shared")` | `Sigma_non_shared_split_true` |
| \(\Psi_{\mathrm{non}}\) | diagonal companion in the same `latent(..., unique = TRUE)` term | `non_unique_split` with independent species rows | `diag(extract_Sigma(level = "unit", part = "unique")$s)` | `Psi_non_split_true` |
| \(\Sigma_{\mathrm{non,total}} = \Sigma_{\mathrm{non,shared}} + \Psi_{\mathrm{non}}\) | same `latent(..., unique = TRUE)` term | `non_shared_split + non_unique_split` | `extract_Sigma(level = "unit", part = "total")` | `Sigma_non_total_split_true` |
| \(\Omega_{\mathrm{species}} = \Sigma_{\mathrm{phy,total}} + \Sigma_{\mathrm{non,total}}\) | sum of both fitted terms | `phy_effect_split + non_effect_split` | sum of both `part = "total"` matrices | `Sigma_phy_total_split_true + Sigma_non_total_split_true` |
:::

Both terms now estimate one shared axis and a diagonal, trait-specific
\(\Psi\) companion. The phylogenetic shared and diagonal draws both follow
`A_split`; their non-phylogenetic counterparts are independent among species.
The `unique =` argument is part of the current latent-term interfaces; it is
distinct from the deprecated standalone `unique()` function.

### Why this example uses 500 species

There is no universal species-count threshold for separating these matrices.
Tree shape, covariance strength, trait count, and within-species replication
all change the information in the data. In a scaled pilot using this same
three-trait model, recovery did not improve monotonically across separately
generated trees as the number of species increased. We use a seeded
500-species dataset to make the seven planted targets visible in one worked
example, not as a recommended minimum for real studies.

The simulation retains three individuals per species. Without within-species
replication, individual noise and independent species-level variation are much
harder to distinguish; a clean optimizer cannot solve that design problem.

```{r simulate-phylo-nonphylo}
set.seed(8622)
n_species_split <- 500L
n_per_species_split <- 3L
trait_names_split <- c("leaf_area", "wood_density", "seed_mass")
n_traits_split <- length(trait_names_split)

tree_split <- ape::rcoal(n_species_split)
tree_split$tip.label <- paste0("sp", seq_len(n_species_split))
A_split <- ape::vcv(tree_split, corr = TRUE)

Lambda_phy_split_true <- matrix(c(0.75, 0.48, 0.35), n_traits_split, 1)
Sigma_phy_shared_split_true <-
  Lambda_phy_split_true %*% t(Lambda_phy_split_true)
Psi_phy_split_true <- diag(c(0.25, 0.30, 0.25))
Sigma_phy_total_split_true <-
  Sigma_phy_shared_split_true + Psi_phy_split_true
Lambda_non_split_true <- matrix(c(0.50, -0.36, 0.28), n_traits_split, 1)
Sigma_non_shared_split_true <-
  Lambda_non_split_true %*% t(Lambda_non_split_true)
Psi_non_split_true <- diag(c(0.20, 0.18, 0.16))
Sigma_non_total_split_true <-
  Sigma_non_shared_split_true + Psi_non_split_true

for (nm in c(
  "Sigma_phy_shared_split_true",
  "Psi_phy_split_true",
  "Sigma_phy_total_split_true",
  "Sigma_non_shared_split_true",
  "Psi_non_split_true",
  "Sigma_non_total_split_true"
)) {
  x <- get(nm)
  dimnames(x) <- list(trait_names_split, trait_names_split)
  assign(nm, x)
}

L_A_split <- t(chol(A_split + 1e-9 * diag(n_species_split)))
phy_shared_split <- L_A_split %*%
  matrix(rnorm(n_species_split), n_species_split, 1) %*%
  t(Lambda_phy_split_true)
phy_unique_split <- L_A_split %*% matrix(
  rnorm(n_species_split * n_traits_split),
  n_species_split,
  n_traits_split
) %*% chol(Psi_phy_split_true)
phy_effect_split <- phy_shared_split + phy_unique_split

non_shared_split <-
  matrix(rnorm(n_species_split), n_species_split, 1) %*%
  t(Lambda_non_split_true)
non_unique_split <- matrix(
  rnorm(n_species_split * n_traits_split),
  n_species_split,
  n_traits_split
) %*% chol(Psi_non_split_true)
non_effect_split <- non_shared_split + non_unique_split

df_split_wide <- expand.grid(
  replicate = seq_len(n_per_species_split),
  species = tree_split$tip.label,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
df_split_wide$species <- factor(
  df_split_wide$species,
  levels = tree_split$tip.label
)
df_split_wide$individual <- factor(
  paste(df_split_wide$species, df_split_wide$replicate, sep = "_")
)

means_split <- c(leaf_area = 0.4, wood_density = 0, seed_mass = -0.4)
for (j in seq_along(trait_names_split)) {
  nm <- trait_names_split[[j]]
  df_split_wide[[nm]] <- means_split[[nm]] +
    phy_effect_split[as.integer(df_split_wide$species), j] +
    non_effect_split[as.integer(df_split_wide$species), j] +
    rnorm(nrow(df_split_wide), sd = 0.25)
}

df_split_long <- do.call(
  rbind,
  lapply(seq_along(trait_names_split), function(j) {
    data.frame(
      individual = df_split_wide$individual,
      species = df_split_wide$species,
      trait = trait_names_split[[j]],
      value = df_split_wide[[trait_names_split[[j]]]]
    )
  })
)
df_split_long$species <- factor(
  df_split_long$species,
  levels = tree_split$tip.label
)
df_split_long$trait <- factor(
  df_split_long$trait,
  levels = trait_names_split
)
```

The long and wide formulas below are the same model. We set `unit = "species"`
so the ordinary `latent()` term is a non-phylogenetic species effect, and
`unit_obs = "individual"` records the replicated observational unit. The fit
chunk displays any messages or warnings instead of hiding them.

```{r fit-phylo-nonphylo, message = TRUE, warning = TRUE}
fit_split_long <- gllvmTMB(
  value ~ 0 + trait +
    phylo_latent(
      0 + trait | species,
      d = 1,
      tree = tree_split,
      unique = TRUE
    ) +
    latent(0 + trait | species, d = 1, unique = TRUE),
  data = df_split_long,
  trait = "trait",
  unit = "species",
  unit_obs = "individual",
  cluster = "species",
  family = gaussian()
)

fit_split_wide <- gllvmTMB(
  traits(leaf_area, wood_density, seed_mass) ~ 1 +
    phylo_latent(
      1 | species,
      d = 1,
      tree = tree_split,
      unique = TRUE
    ) +
    latent(1 | species, d = 1, unique = TRUE),
  data = df_split_wide,
  unit = "species",
  unit_obs = "individual",
  cluster = "species",
  family = gaussian()
)

c(
  long = as.numeric(logLik(fit_split_long)),
  wide = as.numeric(logLik(fit_split_wide)),
  absolute_difference = abs(
    as.numeric(logLik(fit_split_long)) -
      as.numeric(logLik(fit_split_wide))
  )
)
```

Now inspect every fit-health row before reading the split. Passing rows show
that this fitted point is numerically usable and that local Wald calculations
are available; they do not prove that the scientific decomposition is
identified in another dataset.

```{r diagnose-phylo-nonphylo}
health_split <- check_gllvmTMB(fit_split_long)
health_split[, c(
  "component", "status", "value", "threshold", "message", "action"
)]
```

All 16 rows pass for this fit. The raw maximum gradient is `0.0002954`, below
the `0.01` point-stationarity threshold; `sdreport()` is available and the
Hessian is positive definite with reported rank `16/16`. These are numerical
and local-curvature checks, not recovery guarantees.

The six extractor calls recover each shared, diagonal, and total matrix named
in the alignment table. The sum of the two total matrices is the persistent
species covariance; it excludes the independent record-level Gaussian noise
added in the simulation.

```{r recover-phylo-nonphylo}
Sigma_phy_shared_split_hat <- extract_Sigma(
  fit_split_long,
  level = "phy",
  part = "shared"
)$Sigma
Psi_phy_split_hat <- diag(extract_Sigma(
  fit_split_long,
  level = "phy",
  part = "unique"
)$s)
Sigma_phy_total_split_hat <- extract_Sigma(
  fit_split_long,
  level = "phy",
  part = "total"
)$Sigma
Sigma_non_shared_split_hat <- extract_Sigma(
  fit_split_long,
  level = "unit",
  part = "shared"
)$Sigma
Psi_non_split_hat <- diag(extract_Sigma(
  fit_split_long,
  level = "unit",
  part = "unique"
)$s)
Sigma_non_total_split_hat <- extract_Sigma(
  fit_split_long,
  level = "unit",
  part = "total"
)$Sigma

relative_error <- function(estimate, truth) {
  norm(estimate - truth, "F") / norm(truth, "F")
}

recovery_split <- data.frame(
  component = c(
    "Sigma_phy_shared",
    "Psi_phy",
    "Sigma_phy_total",
    "Sigma_non_shared",
    "Psi_non",
    "Sigma_non_total",
    "species total"
  ),
  relative_error = c(
    relative_error(
      Sigma_phy_shared_split_hat,
      Sigma_phy_shared_split_true
    ),
    relative_error(Psi_phy_split_hat, Psi_phy_split_true),
    relative_error(Sigma_phy_total_split_hat, Sigma_phy_total_split_true),
    relative_error(
      Sigma_non_shared_split_hat,
      Sigma_non_shared_split_true
    ),
    relative_error(Psi_non_split_hat, Psi_non_split_true),
    relative_error(Sigma_non_total_split_hat, Sigma_non_total_split_true),
    relative_error(
      Sigma_phy_total_split_hat + Sigma_non_total_split_hat,
      Sigma_phy_total_split_true + Sigma_non_total_split_true
    )
  )
)
recovery_split$relative_error <- round(recovery_split$relative_error, 3)
recovery_split

round(Sigma_phy_shared_split_hat, 3)
round(Psi_phy_split_hat, 3)
round(Sigma_phy_total_split_hat, 3)
round(Sigma_non_shared_split_hat, 3)
round(Psi_non_split_hat, 3)
round(Sigma_non_total_split_hat, 3)
```

The relative errors are descriptive results from this one seeded fit. They
are `0.732`, `0.862`, and `0.340` for the phylogenetic shared, diagonal, and
total matrices; `0.179`, `0.114`, and `0.137` for the corresponding
non-phylogenetic matrices; and `0.197` for the species total. The healthy
optimizer and Hessian therefore do **not** establish close separation of the
phylogenetic shared axis from its diagonal companion in this realization.
These errors are not standard errors or repeated-sampling summaries. This fit
demonstrates the shared/diagonal/total API and an honest truth comparison; it
is not a coverage study or evidence that 500 species will identify the split
for every tree, rank, and covariance regime.

## Interpretation boundaries

The fitted off-diagonal entries answer a conditional model question: how much
cross-trait covariance is associated with the Brownian relationship matrix
constructed from this tree? They do not, by themselves, establish a causal
evolutionary mechanism.

Keep four distinctions clear:

- `Sigma_phy` is a covariance matrix; its correlation matrix standardizes
  entries by fitted phylogenetic variances.
- A phylogenetic covariance is not genetic covariance or heritability.
- The supplied tree and Brownian covariance are treated as fixed. Alternative
  trees or evolutionary kernels can change the answer.
- Replication helps separate species effects from individual noise. With one
  mean per species and trait, independent species variation, measurement
  error, and other tip-level variation can be difficult to distinguish.

This article uses the low-rank `phylo_latent(..., unique = TRUE)` route
throughout. If that decomposition is not supported by the data, simplify or
change covariance mode according to the scientific target rather than adding
complexity automatically. The
[Formula keyword grid](api-keyword-grid.html#phylogenetic) lists the diagonal
and full-covariance alternatives and their exact call shapes.

## What to read next

- [Covariance and correlation](covariance-correlation.html) explains the
  distinction between covariance, correlation, shared structure, and
  trait-specific variance.
- [Can I trust this fit?](fit-diagnostics.html) gives the full numerical and
  response-diagnostic workflow.
- [Convergence and start values](convergence-start-values.html) gives
  symptom-specific recovery steps for difficult fits.
- [Formula keyword grid](api-keyword-grid.html#phylogenetic)
  compares the available phylogenetic covariance modes.

## References

Felsenstein, J. (1985). Phylogenies and the comparative method.
*The American Naturalist*, 125, 1–15.

Hadfield, J. D., & Nakagawa, S. (2010). General quantitative genetic methods
for comparative biology: phylogenies, taxonomies and multi-trait models for
continuous and categorical characters. *Journal of Evolutionary Biology*, 23,
494–508. <https://doi.org/10.1111/j.1420-9101.2009.01915.x>
