---
title: "Phylogenetic ordinal and nominal PGLMMs"
description: "Fit phylogenetic threshold and unordered-category models with ordinal_probit() and multinomial(), following Mizuno et al. (2025) J. Evol. Biol."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Phylogenetic ordinal and nominal PGLMMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

Mizuno, Drobniak, Williams, Lagisz & Nakagawa (2025, *J. Evol. Biol.*
38:1699-1715, doi:10.1093/jeb/voaf116) is a tutorial on phylogenetic
generalized linear mixed models (PGLMMs) for discrete traits — ordinal
(threshold) and nominal (unordered) categories — implemented with
MCMCglmm and brms. This article works through the paper's two worked
examples in `gllvmTMB`: a fast frequentist route to the same models,
using the paper's own archived data
(<https://github.com/Ayumi-495/PGLMM_tutorial>, Zenodo
10.5281/zenodo.17038830, CC BY 4.0).

The recovery numbers below describe these two real bird datasets, fitted
once. They are worked examples, not a calibration study — no interval
here is certified to cover at any nominal rate (see "What this article
does *not* claim" at the end).

## Setup and data

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

The data layer fetches the paper's archived CSVs and Nexus trees to a
local cache (never into the package or the git history — the trees are
Jetz/BirdTree-derived with no explicit licence) and validates the
schema after every fetch: expected columns, the three documented
`Migration` levels, and exact tree-tip/species matching. If a fetch
fails and no cache exists, the loader returns `NULL` with an
explanatory message instead of stale or partial data, and the examples
below are skipped visibly rather than silently.

```{r fetch-data, include = FALSE}
fetch_script <- c(
  file.path("dev", "mizuno-vignette", "fetch-mizuno-data.R"),
  file.path("..", "..", "dev", "mizuno-vignette", "fetch-mizuno-data.R"),
  file.path("..", "dev", "mizuno-vignette", "fetch-mizuno-data.R")
)
fetch_script <- fetch_script[file.exists(fetch_script)][1]
if (!is.na(fetch_script)) {
  source(fetch_script)
  ex1 <- tryCatch(mizuno_load_ordinal(), error = function(e) NULL)
  ex2 <- tryCatch(mizuno_load_nominal(), error = function(e) NULL)
} else {
  ex1 <- NULL
  ex2 <- NULL
}
have_ex1 <- !is.null(ex1)
have_ex2 <- !is.null(ex2)
```

```{r data-unavailable-note, echo = FALSE, results = "asis", eval = !(have_ex1 && have_ex2)}
cat(
  "> **Data unavailable in this build.** This render had no network",
  "access to the archive (or the archive's schema moved) and no local",
  "cache. The affected worked example below is skipped rather than",
  "shown with stale or partial data. Re-run with network access, or see",
  "<https://github.com/Ayumi-495/PGLMM_tutorial>.\n"
)
```

## Example 1 — a phylogenetic ordinal (threshold) PGLMM

The paper's ordinal example (their section 3) regresses a three-level
migratory-behaviour trait on body mass in accipitrid raptors
(Accipitridae, `r if (have_ex1) nrow(ex1$data)` species, one row per
species, a Jetz/BirdTree phylogeny). `gllvmTMB` is a **multivariate**
engine: every covariance-structure keyword (`phylo_latent()`,
`phylo_indep()`, `phylo_dep()`, ...) builds its design matrix through a
`trait` factor, and that factor needs at least two levels — a genuinely
single-trait (univariate) fit of this kind is out of `gllvmTMB`'s scope
and belongs to `glmmTMB` or the paper's own `MCMCglmm`/`brms` route (see
"What this article does *not* claim").

So this article lifts the example to a genuine **bivariate** ordinal
PGLMM, pairing `Migration_ordered` with a second real trait from the
same table, `Habitat.Density` (open/semi-open/dense, also three
levels). The pairing has a real biological question behind it, not just
a workaround: migratory raptors are disproportionately birds of open
country, so a shared phylogenetic axis between migratory tendency and
habitat openness is a reasonable hypothesis, testable directly from the
fitted phylogenetic correlation below. Body mass (log-scale) stays in
the model as a covariate on both traits, as in the paper.

```{r ex1-data, eval = have_ex1}
d1 <- ex1$data
df_long <- rbind(
  data.frame(species = d1$species, trait = "migration",
             value = as.integer(d1$Migration_ordered), logMass = d1$logMass),
  data.frame(species = d1$species, trait = "habitat_density",
             value = as.integer(d1$Habitat.Density), logMass = d1$logMass)
)
df_long$trait <- factor(df_long$trait, levels = c("migration", "habitat_density"))

df_wide <- data.frame(
  species          = d1$species,
  migration        = as.integer(d1$Migration_ordered),
  habitat_density  = as.integer(d1$Habitat.Density),
  logMass          = d1$logMass
)
head(df_long, 4)
head(df_wide, 4)
```

### Fit: long and wide give the same model

```{r ex1-fit, eval = have_ex1}
t0 <- Sys.time()
fit1_long <- gllvmTMB(
  value ~ 0 + trait + trait:logMass +
    phylo_dep(0 + trait | species, tree = ex1$tree),
  data = df_long, trait = "trait", unit = "species", cluster = "species",
  family = ordinal_probit()
)
fit1_wall <- as.numeric(difftime(Sys.time(), t0, units = "secs"))

fit1_wide <- gllvmTMB(
  traits(migration, habitat_density) ~ 1 + logMass +
    phylo_dep(1 | species, tree = ex1$tree),
  data = df_wide, unit = "species", cluster = "species",
  family = ordinal_probit()
)

fit1 <- fit1_long
cat("Wall clock (long fit):", round(fit1_wall, 2), "s\n")
cat("Long/wide logLik difference:",
    signif(abs(as.numeric(logLik(fit1_long)) - as.numeric(logLik(fit1_wide))), 3), "\n")
fit1$fit_health[c("convergence", "pd_hessian", "converged", "sdreport_ok")]
```

`traits(migration, habitat_density) ~ 1 + logMass + phylo_dep(1 | species, tree = tree)`
is the wide equivalent of the long call above. `1` expands to `0 + trait`
(per-trait intercepts) and, because a bare covariate follows it,
`logMass` expands to `trait:logMass` (a **separate** slope per trait,
not one shared slope) — write the long form as `0 + trait + trait:logMass`
to match; `0 + trait + logMass` (one shared slope) is a genuinely
different, non-equivalent model and gives a different log-likelihood.
`phylo_dep(1 | species, tree = tree)` desugars to
`phylo_dep(0 + trait | species, tree = tree)`, the full unstructured
`2 x 2` phylogenetic covariance across the two traits (paper eq 6-11's
Kronecker `Sigma_a %x% A`, applied here to two threshold traits rather
than two continuous ones).

### Cutpoints (Box 2: which convention?)

```{r ex1-cutpoints, eval = have_ex1}
extract_cutpoints(fit1)
```

The paper's Box 2 distinguishes two threshold-model parameterisations:
`MCMCglmm`'s (free intercept, first cutpoint fixed at 0) and `brms`'s
(intercept fixed at 0, all cutpoints free). `gllvmTMB`'s `ordinal_probit()`
follows Hadfield (2015)'s own convention — the same one `MCMCglmm` uses:
`tau_1 = 0` is fixed and the free cutpoints are returned as
`cutpoint_2`, `cutpoint_3`, .... Reading a fit off `MCMCglmm::MCMCglmm()`
into `gllvmTMB` needs no re-anchoring: its free intercept maps directly
onto the `gllvmTMB` trait intercept, and its second cutpoint onto
`cutpoint_2`. Reading a `brms` fit across needs a shift: `brms`'s
`Intercept[1]` becomes the `gllvmTMB` intercept plus `tau_1 = 0`, and
every later `brms` cutpoint shifts by that same constant to land on
`cutpoint_2`, `cutpoint_3`, .... The two parameterisations differ by a
location shift only, never a distributional one, and neither package
performs the shift automatically.

### Phylogenetic heritability (paper eq 4, 18)

```{r ex1-h2, eval = have_ex1}
ps1 <- extract_phylo_signal(fit1, link_residual = "auto")
ps1
```

`link_residual = "auto"` puts the ordinal liability's fixed residual
variance (`sigma^2_d = 1`, the probit-threshold convention) into the
denominator, matching the paper's eq 18
`H^2 = sigma_a^2 / (sigma_a^2 + 1)` exactly. Migratory tendency shows
moderate phylogenetic signal (H^2 ~= `r if (have_ex1) round(ps1$H2[ps1$trait == "migration"], 2)`)
and habitat density shows strong signal
(H^2 ~= `r if (have_ex1) round(ps1$H2[ps1$trait == "habitat_density"], 2)`)
on this fitted tree.

### Do the two traits covary phylogenetically?

```{r ex1-corr, eval = have_ex1}
extract_correlations(fit1, tier = "phy", link_residual = "auto")
```

The estimated phylogenetic correlation between migratory tendency and
habitat density is positive — consistent with the "open-country
migrant" hypothesis above — but this is a **single fitted point
estimate on one dataset with no calibrated interval**, not a tested
claim; see the limits section.

## Example 2 — a phylogenetic nominal (multinomial) PGLMM

The paper's nominal example (their section 4) models AVONET's
`Primary.Lifestyle` locomotory-niche category in thrushes (Turdidae,
`r if (have_ex2) nrow(ex2$data)` species; this subset carries
`r if (have_ex2) nlevels(ex2$data$Primary.Lifestyle)` observed
categories). Unlike Example 1, this genuinely is a single categorical
trait, and it fits `gllvmTMB`'s scope directly: `multinomial()`
internally expands one `K`-category trait into `K - 1` baseline-contrast
pseudo-traits before any design matrix is built, so the single-trait
limitation above never applies here.

```{r ex2-data, eval = have_ex2}
d2 <- ex2$data
K <- nlevels(d2$Primary.Lifestyle)

df2_long <- data.frame(species = d2$species, trait = factor("lifestyle"),
                        value = d2$Primary.Lifestyle)
df2_wide <- data.frame(species = d2$species, lifestyle = d2$Primary.Lifestyle)
table(d2$Primary.Lifestyle)
```

```{r ex2-fit, eval = have_ex2}
t0 <- Sys.time()
fit2_long <- gllvmTMB(
  value ~ 0 + trait + phylo_latent(species, tree = ex2$tree, d = K - 1),
  data = df2_long, trait = "trait", unit = "species", cluster = "species",
  family = multinomial()
)
fit2_wall <- as.numeric(difftime(Sys.time(), t0, units = "secs"))

fit2_wide <- gllvmTMB(
  traits(lifestyle) ~ 1 + phylo_latent(1 | species, tree = ex2$tree, d = K - 1),
  data = df2_wide, unit = "species", cluster = "species",
  family = multinomial()
)

fit2 <- fit2_long
cat("Wall clock (long fit):", round(fit2_wall, 2), "s\n")
cat("Long/wide logLik difference:",
    signif(abs(as.numeric(logLik(fit2_long)) - as.numeric(logLik(fit2_wide))), 3), "\n")
fit2$fit_health[c("convergence", "pd_hessian", "converged", "sdreport_ok")]
```

`phylo_latent(species, tree = tree, d = K - 1)` is the loadings-only
(rank-`K - 1`) phylogenetic ordination over the baseline-category
contrasts (paper eq 33-37); the long and wide calls fit the identical
model (logLik matches to numerical precision).

### Per-contrast phylogenetic heritability (paper eq 4, 19)

```{r ex2-h2, eval = have_ex2}
ps2 <- extract_phylo_signal(fit2, link_residual = "auto")
ps2
```

For an unordered response there is no single H^2 — each `K - 1` contrast
against the reference category (here `Generalist`) gets its own value,
using the fixed nominal-softmax residual `pi^2 / 3` per contrast (eq 19),
never collapsed to one scalar. **The reported correlation between
contrasts is not shown here deliberately** — see the caveat immediately
below.

```{r ex2-h2-plot, eval = have_ex2, fig.width = 5, fig.height = 3.2, fig.cap = "Per-contrast phylogenetic heritability, baseline category Generalist."}
ggplot(ps2, aes(x = trait, y = H2)) +
  geom_col(fill = "#0a617d", width = 0.5) +
  ylim(0, 1) +
  labs(x = NULL, y = expression(H^2), title = "Turdidae Primary.Lifestyle") +
  theme_minimal()
```

> **One categorical draw per species — read the correlation with
> caution, or not at all.** This dataset has exactly one lifestyle
> observation per species, the regime the package's own recovery
> campaign (documented for this exact call shape) found unreliable for
> the **among-category correlation**: variance/H^2 components separate
> and recover correctly, but the correlation estimate rails toward
> +/-1 in a substantial fraction of simulated fits under one draw per
> species, even with a positive-definite Hessian. Five or more
> independent categorical draws per species resolves it in simulation;
> a real comparative dataset with one row per species — like this one —
> cannot supply that, so this article reports the per-contrast H^2
> values above and does not report or interpret a phylogenetic
> correlation between lifestyle categories.

## Comparison with the paper's route

The paper fits both examples with `MCMCglmm` (and `brms`); we do not
have access to its published posterior numbers and do not reproduce
them here — inventing or approximating a number attributed to a
published paper would be worse than not comparing at all.

For Example 1, we instead ran our own independent `MCMCglmm` comparator
on the *same* data, as a second opinion rather than a validation of the
paper's own posterior:

```{r mcmcglmm-comparator, eval = have_ex1 && requireNamespace("MCMCglmm", quietly = TRUE)}
d1$value_f <- factor(as.integer(d1$Migration_ordered), ordered = TRUE)
Ainv <- MCMCglmm::inverseA(ex1$tree, nodes = "TIPS")$Ainv
prior <- list(R = list(V = 1, fix = 1),
              G = list(G1 = list(V = 1, nu = 1, alpha.mu = 0, alpha.V = 1000)))
t0 <- Sys.time()
m <- MCMCglmm::MCMCglmm(
  value_f ~ logMass, random = ~ species, ginverse = list(species = Ainv),
  data = d1, family = "ordinal", prior = prior,
  nitt = 33000, burnin = 3000, thin = 30, verbose = FALSE
)
mcmc_wall <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
h2_mcmc <- mean(m$VCV[, "species"] / (m$VCV[, "species"] + m$VCV[, "units"] + 1))
cat("MCMCglmm wall clock:", round(mcmc_wall, 1), "s\n")
cat("gllvmTMB  H2(migration), joint bivariate fit:",
    round(ps1$H2[ps1$trait == "migration"], 3), "\n")
cat("MCMCglmm  H2(migration), univariate fit:     ", round(h2_mcmc, 3), "\n")
```

The two numbers are not a like-for-like comparison — `MCMCglmm` here
fits `Migration_ordered` **alone**, while `gllvmTMB`'s number comes from
the **joint** bivariate fit with `Habitat.Density` — so read this as a
qualitative sanity check, not an equivalence test. They land close
(within ~0.02 on this dataset), which is reassuring but is a single
comparison on one dataset, not a calibration result.

For Example 2 (nominal), no comparator is run here: `MCMCglmm`'s
categorical family needs a from-scratch multi-trait link-function setup
distinct from the ordinal case above, and building that correctly is
outside this article's scope. The comparison for Example 2 is
**qualitative only** — read the paper directly for its own numbers.

## The paper-equation to `gllvmTMB`-call map

The rows below are reproduced from `gllvmTMB`'s own internal design
record for this alignment (not re-derived here), restricted to the
models this article uses.

| Paper eq | Model | `gllvmTMB` call | Status |
|---|---|---|---|
| 1-5 | Univariate continuous PMM | `phylo_indep(0 + trait \| species, tree = tree)`, `family = gaussian()` | supported for `T >= 2` traits; a genuinely univariate fit is out of scope (see above) |
| 6-11 | Multivariate PMM, Kronecker `Sigma_a %x% A` | `phylo_dep(0 + trait \| species, tree = tree)` | supported; used in Example 1 above |
| 27-32 | Ordinal PGLMM with phylogenetic source | `phylo_latent()` / `phylo_indep()` / `phylo_dep()` + `family = ordinal_probit()` | supported; used in Example 1 above |
| 33-37 | Nominal PGLMM, per-contrast phylogenetic variances/correlations | `phylo_latent(species, tree = tree, d = K - 1)` + `family = multinomial()` | supported for variance/H^2 components; the among-category correlation needs `n_rep >= 5` draws per species, which single-observation comparative data does not have (Example 2 above) |
| 4, 18, 19 | Phylogenetic heritability `H^2 = sigma_a^2 / (sigma_a^2 + sigma_e^2)` | `extract_phylo_signal(fit, link_residual = "auto")` | supported; `sigma_e^2 = 1` for ordinal (eq 18), `pi^2/3` per contrast for nominal (eq 19), never collapsed to a scalar |
| 8 | Phylogenetic correlation between traits/contrasts | `extract_correlations(fit, tier = "phy")` | supported for admitted phylogenetic tiers, including ordinal and multinomial |
| A.1-A.9 | Contrast-matrix reparameterisation of the softmax | n/a | no action needed — `gllvmTMB` already fits the equivalent baseline-category-logit softmax directly |

The full table (20 rows, every mode/source combination, with the
recovery-campaign evidence behind each verdict) is in
`docs/design/123-multinomial-structured-surface.md` inside the package
repository, alongside the register rows it feeds.

## What this article does *not* claim

- **No calibrated interval.** Cutpoint standard errors above are Wald
  approximations reported descriptively; no coverage study backs any
  interval in this article.
- **`gllvmTMB` is a multivariate engine.** A strictly univariate model —
  one categorical trait, no second trait to stack against it — is
  outside `gllvmTMB`'s scope; use `glmmTMB` or the paper's own
  `MCMCglmm`/`brms` route for that case directly.
- **The Example 2 phylogenetic correlation is withheld, not just
  under-powered.** With one categorical draw per species, the
  correlation estimate is known to be unreliable even when the fit
  reports full convergence; do not read the H^2 values above as
  implying a trustworthy correlation exists alongside them.
- **This is one dataset each, fitted once.** Point estimates describe
  these specific fits; they are not simulation-based recovery evidence.
- **No ancestral-state reconstruction.** The paper's Bayesian route
  supports this; `gllvmTMB` does not attempt it.

## See also

* [Phylogenetic covariance among traits](phylogenetic-gllvm.html) —
  the continuous-trait phylogenetic GLLVM this article's ordinal/nominal
  models build on.
* [Unordered categories with `multinomial()`](multinomial.html) — the
  general nominal-response guide.
* [Choose a response family](response-families.html) — when to reach
  for `ordinal_probit()` vs `multinomial()` vs `binomial()`.
