---
title: "Phylogenetic trait covariance"
tier: 3 # internal biological worked-example draft until reader/evidence review passes
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Phylogenetic trait covariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = requireNamespace("ape", quietly = TRUE),
  message  = FALSE,
  warning  = FALSE
)
```

> **Internal article gate.** This is a candidate Tier 1 comparative-methods
> worked example, but it is not ready for the public article dropdown. The page
> stays internal until the phylogenetic/non-phylogenetic reader path, validation
> row bridge, figures, diagnostics, and rendered HTML pass article-council
> review. Start with [Morphometrics](morphometrics.html),
> [Covariance and correlation](covariance-correlation.html), or
> [Formula keyword grid](api-keyword-grid.html) for the current public path.

This article is for a comparative-methods reader with several traits
measured once per species. The working question is: can we separate
phylogenetically conserved trait covariance from non-phylogenetic
species-level covariance, while still keeping trait-specific variation
on the diagonal?

The model has two species-level trait-covariance matrices. The
phylogenetic component uses the tree correlation matrix `A`; the
non-phylogenetic component uses independent species effects:

$$
g_{\mathrm{phy}} \sim
\mathrm{MVN}(0, \Sigma_{\mathrm{phy}} \otimes A),
\qquad
g_{\mathrm{non}} \sim
\mathrm{MVN}(0, \Sigma_{\mathrm{non}} \otimes I).
$$

The phylogenetic covariance uses an explicit source-specific pair,
`phylo_latent() + phylo_unique()`, while the ordinary non-phylogenetic
covariance uses default `latent()` syntax:

$$
\Sigma_{\mathrm{phy}} =
\Lambda_{\mathrm{phy}}\Lambda_{\mathrm{phy}}^T + \boldsymbol\Psi_{\mathrm{phy}},
\qquad
\Sigma_{\mathrm{non}} =
\Lambda_{\mathrm{non}}\Lambda_{\mathrm{non}}^T + \boldsymbol\Psi_{\mathrm{non}}.
$$

Here `Lambda` is the shared low-rank structure and `Psi` is the
trait-specific diagonal variance. The "two-Psi" nickname refers to
this paired phylogenetic and non-phylogenetic decomposition (two
`Psi` diagonals, one at each tier). Use `phylo_unique()` in this
paired sense when the data can identify a separate phylogenetic
diagonal; ordinary `latent()` already includes the non-phylogenetic
diagonal `Psi` companion by default. For a standalone marginal
phylogenetic diagonal model, use `phylo_indep()`; for a standalone
non-phylogenetic diagonal tier, use `indep()` (`FG-05`--`FG-07`).

## Reader path and scope

This draft follows one comparative question through five checks:

| Question | Model object | Code section | Readout |
|---|---|---|---|
| Is trait covariance phylogenetically conserved? | `phylo_latent()` + `phylo_unique()` | `fit`, `extract-phy` | `Sigma_phy` shared, diagonal, and total rows |
| Is there residual species-level covariance not explained by the tree? | ordinary `latent()` at the species tier | `fit`, `extract-non` | `Sigma_non` shared, diagonal, and total rows |
| Do long and wide data give the same likelihood? | long `value ~ ...` and wide `traits(...) ~ ...` calls | `fit` | long/wide log-likelihood difference |
| Is the fitted model numerically usable? | `check_gllvmTMB()` through `diagnostic_table()` | `diagnostic-table` | optimizer, gradient, `sdreport`, and Hessian rows |
| Are the phylogenetic and non-phylogenetic totals interpretable? | `extract_Sigma_table()` + `plot_Sigma_heatmap()` | `extract-total-correlations` | side-by-side total-correlation heatmaps |
| When should the model be simplified? | three-piece fallback and two-Psi cross-checks | `fallback`, `compare-indep`, `compare-dep` | diagnostics and baseline disagreement flags |

Scope boundary: IN rows are the paired phylogenetic decomposition
(`PHY-02`), three-piece fallback (`PHY-03`), phylogenetic independent and
dependent baselines (`PHY-05`), Gaussian phylogenetic slope demonstration
(`PHY-06`), `extract_phylo_signal()` and communality summaries (`PHY-07`,
`PHY-08`, `EXT-05`, `EXT-07`), report-ready Sigma tables and heatmaps
(`EXT-18`, `EXT-27`), and fit-health table support (`DIA-08`). PARTIAL
rows are interval/calibration displays for broader covariance summaries and
profile/bootstrap coverage; point estimates in this article should not be read
as calibrated uncertainty. PLANNED or guarded rows include richer
trait-specific reaction-norm LHS forms and broad non-Gaussian uncertainty
claims.

## Setup

```{r setup}
library(gllvmTMB)
library(ape)
```

## Simulate

We simulate three traits on 30 species. The truth has four
components, matching the paired decomposition in the theory section:
a phylogenetic shared axis (`Lambda_phy`), a phylogenetic diagonal
component (`psi_phy`), a non-phylogenetic shared axis (`Lambda_non`),
and a non-phylogenetic diagonal component (`psi_non`). This small
example is meant to show syntax and extraction, not to be a full
simulation study.

```{r sim}
set.seed(13)
n_sp <- 30
n_traits <- 3
trait_names <- paste0("trait_", seq_len(n_traits))

tree <- ape::rcoal(n_sp)
tree$tip.label <- paste0("sp", seq_len(n_sp))
Cphy <- ape::vcv(tree, corr = TRUE)
Lphy <- t(chol(Cphy + 1e-8 * diag(n_sp)))

Lambda_phy_true <- matrix(c(0.70, 0.45, 0.25), n_traits, 1)
Lambda_non_true <- matrix(c(0.50, 0.30, 0.20), n_traits, 1)
s_phy_true <- c(0.20, 0.15, 0.12)
s_non_true <- c(0.18, 0.20, 0.16)

g_phy <- Lphy %*% rnorm(n_sp)
g_non <- matrix(rnorm(n_sp), n_sp, 1)
e_phy <- Lphy %*% matrix(rnorm(n_sp * n_traits), n_sp, n_traits)
e_non <- matrix(rnorm(n_sp * n_traits), n_sp, n_traits)

Y <- g_phy %*% t(Lambda_phy_true) +
  sweep(e_phy, 2, sqrt(s_phy_true), "*") +
  g_non %*% t(Lambda_non_true) +
  sweep(e_non, 2, sqrt(s_non_true), "*")
rownames(Y) <- tree$tip.label
colnames(Y) <- trait_names

df <- data.frame(
  species = factor(rep(tree$tip.label, each = n_traits),
                   levels = tree$tip.label),
  trait = factor(rep(trait_names, times = n_sp),
                 levels = trait_names),
  value = as.numeric(t(Y))
)

df_wide <- data.frame(
  species = factor(tree$tip.label, levels = tree$tip.label),
  Y,
  check.names = FALSE
)
head(df)
```

## Fit

The long form is canonical: one row per `(species, trait)` observation.
The wide data-frame form is the same model written from one row per
species and one column per trait. Because `traits(...)` names the
response traits on the left-hand side, the right-hand side can use the
compact covariance syntax.

```{r fit}
# Long format (canonical)
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    phylo_latent(species, d = 1, tree = tree) +
    phylo_unique(species, tree = tree) +
    latent(0 + trait | species, d = 1),
  data = df,
  trait = "trait",
  unit = "species",
  cluster = "species"
)

# Wide data-frame format (same model with compact formula syntax)
fit_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~ 1 +
    phylo_latent(species, d = 1, tree = tree) +
    phylo_unique(species, tree = tree) +
    latent(1 | species, d = 1),
  data = df_wide,
  unit = "species",
  cluster = "species"
)

fit <- fit_long
fit$opt$convergence

cat("Long/wide logLik difference:",
    signif(abs(as.numeric(logLik(fit_long)) -
               as.numeric(logLik(fit_wide))), 3), "\n")
```

The two forms dispatch to the same long-format engine. In this model,
`phylo_latent(species, d = 1, tree = tree)` estimates the shared
phylogenetic axis, `phylo_unique(species, tree = tree)` estimates the
trait-specific phylogenetic Psi diagonal, `latent(1 | species, d = 1)` in
the wide formula (expanding to `latent(0 + trait | species, d = 1)`
in the long form) estimates the shared non-phylogenetic axis, and
ordinary `latent()` adds the trait-specific non-phylogenetic Psi
diagonal by default.

## Fit-health diagnostics

Read the diagnostics before interpreting the phylogenetic split. The table
below uses randomized-quantile residual objects as the carrier for the
`check_gllvmTMB()` fit-health rows, then exposes those rows through
`diagnostic_table()`.

```{r diagnostic-table}
rq_long <- residuals(
  fit_long,
  type = "randomized_quantile",
  seed = 20260619
)
rq_wide <- residuals(
  fit_wide,
  type = "randomized_quantile",
  seed = 20260619
)

diagnostic_components <- c(
  "optimizer_convergence",
  "max_gradient",
  "sdreport",
  "pd_hessian"
)

diagnostic_rows <- rbind(
  data.frame(
    layout = "long",
    diagnostic_table(rq_long, table = "check_gllvmTMB")
  ),
  data.frame(
    layout = "wide",
    diagnostic_table(rq_wide, table = "check_gllvmTMB")
  )
)

diagnostic_rows[
  diagnostic_rows$component %in% diagnostic_components,
  c("layout", "component", "status", "message")
]
```

Treat any `WARN` or `FAIL` row as a public-promotion blocker. A clean
diagnostic table is not proof that the phylogenetic split is biologically
identified; it only says the fitted object is numerically usable enough for
the point-estimate readout below.

## Recover the phylogenetic covariance

`extract_Sigma(level = "phy")` returns the phylogenetic trait
covariance. The shared part is
`Lambda_phy %*% t(Lambda_phy)`; the unique part is the diagonal
`psi_phy`; the total is their sum.

```{r extract-phy}
Sigma_phy_rows <- rbind(
  extract_Sigma_table(fit, level = "phy", part = "shared",
                      entries = "unique"),
  extract_Sigma_table(fit, level = "phy", part = "unique",
                      entries = "diag"),
  extract_Sigma_table(fit, level = "phy", part = "total",
                      entries = "unique")
)
Sigma_phy_rows[, c("component", "trait_i", "trait_j",
                   "estimate", "interval_status")]
```

The non-phylogenetic species-level covariance has the same paired
decomposition as the phylogenetic side:

```{r extract-non}
Sigma_non_rows <- rbind(
  extract_Sigma_table(fit, level = "unit", part = "shared",
                      entries = "unique"),
  extract_Sigma_table(fit, level = "unit", part = "unique",
                      entries = "diag"),
  extract_Sigma_table(fit, level = "unit", part = "total",
                      entries = "unique")
)
Sigma_non_rows[, c("component", "trait_i", "trait_j",
                   "estimate", "interval_status")]
```

On the correlation scale, compare the two total covariance targets side
by side:

```{r extract-total-correlations, fig.width = 7.4, fig.height = 4.6, fig.cap = "Phylogenetic and non-phylogenetic trait correlations estimated from the two-tier model.", fig.alt = "Two heatmap facets compare total trait correlations at the phylogenetic and non-phylogenetic unit tiers. Cell labels show estimated correlations for the three simulated traits."}
Sigma_corr_rows <- extract_Sigma_table(
  fit,
  level = c("phy", "unit"),
  part = "total",
  measure = "correlation",
  entries = "all"
)
plot_Sigma_heatmap(
  Sigma_corr_rows,
  label_digits = 2,
  title = "Phylogenetic and non-phylogenetic correlations",
  subtitle = "Facets compare total Sigma at the phy and unit tiers."
)
```

For a small tree, do not over-interpret either split (between
`Lambda_phy Lambda_phy^T` and `Psi_phy`, or between
`Lambda_non Lambda_non^T` and `Psi_non`). The more stable biological
targets are often the totals `Sigma_phy` and `Sigma_non`, especially
when the tree is small or the ranks are exploratory.

### Three-piece fallback when the paired form is under-identified

The paired decomposition fits **four pieces**:
`Lambda_phy`, `Psi_phy`, `Lambda_non`, `Psi_non`. For small trees
(say `n_species < 30`) or near-zero phylogenetic signal, the two
`Psi` diagonals tend to trade off against each other -- only their
sum is well-determined by the data, and `sanity_multi()` will often
flag the fit as boundary-pinned or rank-deficient.

The **three-piece fallback** drops one `Psi` and keeps a single
trait-specific diagonal that is not assigned to either tier:

$$
\boldsymbol{\Omega} =
\boldsymbol{\Lambda}_{\mathrm{phy}}\boldsymbol{\Lambda}_{\mathrm{phy}}^{\top} +
\boldsymbol{\Lambda}_{\mathrm{non}}\boldsymbol{\Lambda}_{\mathrm{non}}^{\top} +
\boldsymbol{\Psi}.
$$

To fit it, drop `phylo_unique()` from the formula and let ordinary
`latent()` carry the single non-phylogenetic diagonal `Psi` companion:

```r
fit_3piece <- gllvmTMB(
  value ~ 0 + trait +
    phylo_latent(species, d = 1, tree = tree) +    # Lambda_phy
    latent(0 + trait | species, d = 1),            # Lambda_non + single Psi
  data = df,
  trait = "trait", unit = "species"
)
```

When you read the output of `extract_phylo_signal()` on a three-piece
fit, `Psi` is the **single** shared diagonal (not the tier-specific
`psi_non` used by the paired form), and the three-way decomposition
`H^2 + C^2_non + Psi = 1` is interpreted with the entire `Psi`
attributed to "trait-specific variance not separated by tier" rather
than to a specifically non-phylogenetic source. The paired form is
preferable when the data support it (it gives a cleaner H^2 reading);
the three-piece fallback is the conservative choice when `sanity_multi()`
or visual inspection flag the paired form as over-parameterised.

## Communality (within each tier)

Communality `c_t^2` is the fraction of a trait's within-tier variance
explained by the shared latent axis. It is defined only when both a
`latent()` and a `unique()` term are fit at that tier; that is what
the four-component decomposition gives us on both the phylogenetic and
non-phylogenetic sides.

```{r communality}
# Non-phylogenetic communality is one extractor call.
c2_non <- extract_communality(fit, level = "unit")
round(c2_non, 2)

# Phylogenetic communality is the same ratio, computed from the
# Sigma_phy decomposition table above.
phy_shared_diag <- subset(Sigma_phy_rows,
                          component == "shared" & diagonal)
phy_total_diag <- subset(Sigma_phy_rows,
                         component == "total" & diagonal)
c2_phy <- phy_shared_diag$estimate / phy_total_diag$estimate
names(c2_phy) <- phy_shared_diag$trait_i
round(c2_phy, 2)
```

A value near 1 means the trait's within-tier variance is dominated by
the shared axis (high "integration" with the other traits at that
tier). A value near 0 means the trait is mostly idiosyncratic at that
tier.

## Phylogenetic heritability

Phylogenetic heritability `H^2` is a between-tier ratio: the fraction
of total trait variance attributable to the phylogenetic component
(both shared and unique parts of `Sigma_phy`) versus the
non-phylogenetic component (both shared and unique parts of
`Sigma_non`). It is distinct from the within-tier communalities
above.

The convenience extractor returns the full three-way decomposition
`H^2 + C^2_non + Psi = 1` per trait, where:

- `H^2` = phylogenetic share = the diagonal of the total phylogenetic
  Sigma divided by `V_eta`
- `C^2_non` = non-phylogenetic *shared* share =
  the diagonal of `Lambda_non Lambda_non^T` divided by `V_eta`
- `Psi` = non-phylogenetic *unique* share = `psi_non / V_eta`

Here `V_eta` is the sum of those three diagonal contributions for each trait.

```{r phylo-heritability}
extract_phylo_signal(fit)
```

`C^2_non` is structurally zero when the formula omits the non-phylogenetic
`latent()` term, so the three-way split collapses to `H^2 + Psi`. Adding
`latent(0 + trait | species, d = K)` (long) or `latent(1 | species, d = K)`
(wide), as this article does, is what makes `C^2_non` a genuine
non-phylogenetic shared-axis component rather than a placeholder.

## Reaction norms: a phylogenetic random slope

The covariance terms above all act on trait *intercepts* -- the
species-level mean of each trait. A different comparative question is
whether species differ in how a trait *responds to an environmental
gradient*, and whether that response is itself phylogenetically
structured. This is a phylogenetic **reaction norm**: a per-species
slope on a continuous covariate, with the slopes drawn from a
phylogenetically correlated distribution

$$
\beta_{\mathrm{slope}} \sim
\mathrm{N}\!\left(0,\; \sigma^2_{\mathrm{slope}}\, A\right),
$$

so closely related species tend to have similar environmental
sensitivities. The keyword is `phylo_slope(x | species)`, where `x` is
the continuous covariate (the environmental gradient).

This is the **one** random-slope mode that is validated for recovery in
this package, and only on a Gaussian response. The honest scope is
spelled out in the caveat below; here we show the worked Gaussian case.

We simulate per-species slopes on a single environmental gradient
`env`, shared across the three traits, then recover the slope variance
and the per-species slopes. As elsewhere in this article, this is a
syntax-and-recovery demonstration on one data set, not a full
simulation study.

```{r reaction-norm-sim}
set.seed(2026)
n_sp_rn <- 25
rn_traits <- paste0("trait_", seq_len(3))

tree_rn <- ape::rcoal(n_sp_rn)
tree_rn$tip.label <- paste0("sp", seq_len(n_sp_rn))
Lphy_rn <- t(chol(ape::vcv(tree_rn, corr = TRUE) + 1e-8 * diag(n_sp_rn)))

sigma_slope_true <- 0.6
slope_sp <- as.numeric(Lphy_rn %*% rnorm(n_sp_rn, sd = sigma_slope_true))
names(slope_sp) <- tree_rn$tip.label

rn <- expand.grid(
  species = factor(tree_rn$tip.label, levels = tree_rn$tip.label),
  trait   = factor(rn_traits, levels = rn_traits),
  rep     = seq_len(5)
)
rn$env <- rnorm(nrow(rn))
mu_trait_rn <- c(2.0, 1.0, 0.0)[as.integer(rn$trait)]
rn$value <- mu_trait_rn +
  slope_sp[as.character(rn$species)] * rn$env +
  rnorm(nrow(rn), sd = 0.5)
head(rn)
```

The slope keyword goes on the right-hand side alongside the fixed
trait means. Pass the tree *inside* the keyword (`tree = tree_rn`);
this is the recommended form and keeps the species index aligned with
the tree.

```{r reaction-norm-fit}
fit_rn <- gllvmTMB(
  value ~ 0 + trait + phylo_slope(env | species, tree = tree_rn),
  data = rn,
  unit = "species"
)
fit_rn$opt$convergence
```

`phylo_slope` estimates a single shared slope variance
`sigma_slope^2` (the slopes are shared across traits in this initial
release). It is reported on the standard-deviation scale as
`log_sigma_slope`:

```{r reaction-norm-sigma}
sigma_slope_hat <- exp(
  fit_rn$opt$par[grepl("log_sigma_slope", names(fit_rn$opt$par))]
)
c(true = sigma_slope_true, estimate = unname(round(sigma_slope_hat, 3)))
```

The per-species slopes themselves are the random effects
`b_phy_slope`, indexed against the augmented `A` rows. Aligning them
to the tip labels recovers the simulated slopes closely:

```{r reaction-norm-slopes}
b_all     <- fit_rn$tmb_obj$env$last.par.best
b_slope   <- as.numeric(b_all[names(b_all) == "b_phy_slope"])
aug_names <- rownames(fit_rn$tmb_data$Ainv_phy_rr)
tip_slope <- b_slope[match(tree_rn$tip.label, aug_names)]

round(cor(tip_slope, slope_sp), 2)   # fitted vs simulated per-species slopes
```

The slope term composes with the intercept-covariance terms from the
rest of this article. For example, a reduced-rank phylogenetic axis on
the intercepts and a phylogenetic reaction-norm slope can be fit
together:

```{r reaction-norm-combined}
fit_rn_latent <- gllvmTMB(
  value ~ 0 + trait +
    phylo_latent(species, d = 1, tree = tree_rn) +
    phylo_slope(env | species, tree = tree_rn),
  data = rn,
  unit = "species"
)
fit_rn_latent$opt$convergence
```

### Scope of the slope machinery (read before extending)

`phylo_slope` is deliberately narrow, and the package is honest about
what is *not* yet available. The validated capability is exactly:

- **`phylo_slope(x | species)` on a Gaussian response**: a single
  shared slope variance on one continuous covariate, slopes shared
  across traits. Recovery of `sigma_slope` and the per-species slopes
  is validated on a Gaussian data-generating process. The parallel
  pedigree keyword `animal_slope(x | id)` uses the same machinery on a
  relatedness matrix `A`.

What is **not** implemented (the fit will stop, or the richer
structure is simply unavailable -- it is not silently approximated):

- **Trait-specific / structured phylogenetic slopes.** Beyond the
  single-variance `phylo_slope` path, the augmented diagonal slope
  `phylo_indep(1 + x | species)` now fits and recovers across the core
  families (Gaussian; binomial probit/logit; Poisson; negative
  binomial; Gamma; Beta; ordinal -- register `PHY-06`, `PHY-11`..`PHY-16`).
  The reduced-rank `phylo_latent(1 + x | species)` slope is likewise
  validated across that family set (`PHY-17`). The richer
  *trait-specific* spelling `phylo_indep(0 + trait + trait:x | species)`
  is recognised by the parser but **reserved for a future release**: it
  **fails loud** (the parser reports that an "LHS richer than
  `0 + trait` is not yet supported"), so you cannot accidentally get an
  unvalidated fit. The full-unstructured `phylo_dep(1 + x | species)`
  augmented slope is validated for `gaussian()` only in this release;
  on a non-Gaussian family it fails loud rather than silently
  approximate (`PHY-18` -- the unstructured cross-(intercept,slope)
  covariance is not yet identifiable off Gaussian). For a correlated
  (intercept, slope) augmented phylogenetic random regression, the
  closed-form `phylo_unique(1 + x | species)` additionally surfaces its
  2x2 via `extract_Sigma(level = "phy")`.
- **Calibrated intervals on the slope variance.** The point estimate
  of `sigma_slope` is what this section demonstrates; calibrated
  confidence-interval coverage for slope variances has not been
  established and should not be reported as such.
- **Non-Gaussian slope recovery on the full-unstructured `phylo_dep`
  path.** The diagonal `phylo_indep` and reduced-rank `phylo_latent`
  augmented slopes now recover across the core families (above); the
  remaining reserve is the full-unstructured `phylo_dep(1 + x |
  species)` slope, which is Gaussian-only and fails loud off Gaussian
  (`PHY-18`). Treat non-Gaussian `phylo_dep` slopes as not yet
  delivered.

If you need a correlated (intercept, slope) covariance, reach for
`phylo_unique(1 + x | species)`, which fits family-generally and
surfaces the 2x2 via `extract_Sigma(level = "phy")`. If your question
genuinely needs trait-specific reaction-norm LHS spellings or a
non-Gaussian full-unstructured `phylo_dep` slope, those remain
reserved (fail-loud) rather than features to reach for now.

## Trait covariance across nested tiers

The phylogenetic and non-phylogenetic split above is one example of a
more general pattern: default `latent()` covariance decompositions and
standalone diagonal `indep()` tiers can be placed at **several nested
grouping tiers** at once. The three grouping slots are named by the
`unit`, `unit_obs`, and `cluster` arguments:

- `unit` -- the primary between-group level whose covariance you most
  want to read (here, the individual);
- `unit_obs` -- the within-`unit` replication level (here, the
  repeated measurement session), which carries the residual;
- `cluster` -- a third grouping slot (here, the population the
  individual belongs to).

This is the layout behind a repeated-measures "behavioural syndrome"
design: traits measured several times per individual, individuals
nested in populations. We decompose the among-individual trait
covariance into a shared axis plus a trait-specific diagonal, and add
a separate population-level diagonal.

```{r tiers-sim}
set.seed(2026)
n_pop     <- 6
n_ind_per <- 10
n_sess    <- 4
behav     <- paste0("behav_", seq_len(3))

pop  <- factor(rep(seq_len(n_pop), each = n_ind_per * n_sess))
ind  <- factor(rep(seq_len(n_pop * n_ind_per), each = n_sess))
sess <- factor(seq_along(ind))

tiers <- data.frame(
  population = rep(pop,  3),
  individual = rep(ind,  3),
  session_id = rep(sess, 3),
  trait      = factor(rep(behav, each = length(sess)), levels = behav)
)

# Truth: a shared among-individual axis + trait-specific diagonals at
# the individual and population tiers, plus residual session noise.
n_ind_tot  <- nlevels(ind)
Lambda_ind <- matrix(c(0.9, 0.6, 0.3), 3, 1)
psi_ind    <- c(0.20, 0.25, 0.30)
psi_pop    <- c(0.40, 0.30, 0.35)
psi_resid  <- c(0.15, 0.15, 0.15)

f_ind      <- rnorm(n_ind_tot)
ind_shared <- f_ind %o% as.numeric(Lambda_ind)
ind_uniq   <- sapply(psi_ind, function(s) rnorm(n_ind_tot, sd = sqrt(s)))
pop_uniq   <- sapply(psi_pop, function(s) rnorm(n_pop,     sd = sqrt(s)))

mu_behav <- c(0.5, 0.0, -0.5)
ti <- as.integer(tiers$trait)
ii <- as.integer(tiers$individual)
pp <- as.integer(tiers$population)
tiers$value <- mu_behav[ti] +
  ind_shared[cbind(ii, ti)] + ind_uniq[cbind(ii, ti)] +
  pop_uniq[cbind(pp, ti)] +
  rnorm(nrow(tiers), sd = sqrt(psi_resid[ti]))
```

The formula places default `latent()` at the individual tier and an
`indep()` diagonal at the population tier. The three grouping columns
are routed by the `unit` / `unit_obs` / `cluster` arguments:

```{r tiers-fit}
fit_tier <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | individual, d = 1) +
    indep(0 + trait | population),
  data     = tiers,
  unit     = "individual",
  unit_obs = "session_id",
  cluster  = "population"
)
fit_tier$opt$convergence
```

`extract_Sigma()` reads the covariance at whichever tier you name.
The `unit` (individual) tier returns the full trait covariance and its
correlation; the `cluster` (population) tier returns the `unique()`
diagonal it carries:

```{r tiers-extract}
# Among-individual total trait covariance (the syndrome of interest).
Sigma_ind <- extract_Sigma(fit_tier, level = "unit", part = "total")
round(Sigma_ind$R, 2)   # correlation among individual-level trait values

# Population-tier trait-specific variances (a diagonal at the cluster slot).
Sigma_pop <- extract_Sigma(fit_tier, level = "cluster", part = "unique")
round(Sigma_pop$s, 3)
```

The table extractor gives the same content per tier in long form, with
an `interval_status` column so you can see which entries carry an
interval:

```{r tiers-table}
ind_total <- extract_Sigma_table(
  fit_tier, level = "unit", part = "total",
  measure = "correlation", entries = "all"
)
ind_total[, c("component", "trait_i", "trait_j",
              "estimate", "interval_status")]
```

Because default `latent()` includes both a shared axis and a diagonal
`Psi` companion, communality is defined there -- the fraction of each
trait's among-individual variance explained by the shared syndrome axis:

```{r tiers-communality}
round(extract_communality(fit_tier, level = "unit"), 2)
```

A practical caveat: the **cluster** tier here has only `r n_pop`
populations, so its diagonal (`Sigma_pop$s` above) is estimated from
few groups and should be read as indicative, not precise -- the same
small-group caution that applies to small trees applies to a
small number of clusters. The among-individual covariance, estimated
from `r n_ind_tot` individuals, is the better-determined target in
this design.

## Compare against simpler and saturated baselines

Three nested covariance intents bracket the four-component model:

- **Independent** (`phylo_indep + indep`): per-trait diagonal only at
  each tier. No shared axis, no cross-trait covariance. The simplest
  model that respects the phylogenetic structure.
- **Two-Psi** (`phylo_latent + phylo_unique + latent`): the
  four-component decomposition fit in this article. A shared
  low-rank axis plus a trait-specific diagonal at each tier.
- **Dependent** (`phylo_dep + dep`): a full unstructured covariance
  at each tier. Saturated; no shared/unique split is enforced.

The two-Psi decomposition sits between independent (too restrictive
when traits are correlated) and dependent (too many parameters when
the number of traits grows). The cross-check functions refit the
data under the independent and dependent baselines and compare each
component's implied total covariance against the two-Psi fit.

```{r compare-indep}
diag_indep <- compare_indep_vs_two_psi(fit)
diag_indep$agreement
diag_indep$flag
```

```{r compare-dep}
diag_dep <- compare_dep_vs_two_psi(fit)
diag_dep$agreement
diag_dep$flag
```

Each `$agreement` row shows the Frobenius relative disagreement
between the two-Psi fit and the baseline (`rmse / baseline_mag`).
`flag = TRUE` means relative disagreement exceeds `0.10`. A flag is
not a failed model; it tells you the data do not strongly support
the chosen decomposition at the requested rank.

Use `compare_indep_vs_two_psi()` first when the number of traits is
large -- it only refits the diagonal and stays cheap. Use
`compare_dep_vs_two_psi()` when the number of traits is modest and the
saturated refit is affordable -- it is the more informative
benchmark for off-diagonal cross-trait structure.

## When to simplify

The paired `phylo_latent() + phylo_unique() + latent()` model is useful
when the question is about phylogenetically conserved trait syndromes
alongside non-phylogenetic shared-axis structure.
Simplify when the data do not identify the splits the model asks for:

```{r fallback, eval = FALSE}
fit_phylo_dep <- gllvmTMB(
  value ~ 0 + trait + phylo_dep(0 + trait | species, tree = tree) +
                      dep(0 + trait | species),
  data = df,
  trait = "trait",
  unit = "species",
  cluster = "species"
)
```

`phylo_dep() + dep()` estimates full unstructured trait covariances
at each tier without asking the data to separate a low-rank shared
axis from the diagonal. It is the natural fallback when the two-Psi
diagnostics flag.

## What to read next

- [`compare_dep_vs_two_psi()`](../reference/compare_dep_vs_two_psi.html)
  and [`compare_indep_vs_two_psi()`](../reference/compare_indep_vs_two_psi.html)
  are the likelihood-based cross-checks for this model.
- [`functional-biogeography`](functional-biogeography.html) shows how
  phylogenetic terms behave as controls in a larger site--species--trait
  model.
- [`cross-lineage-coevolution`](cross-lineage-coevolution.html) extends
  the same dense-relatedness idea to two interacting lineages linked by
  an association matrix `W`.
- [`morphometrics`](morphometrics.html) is the simplest non-phylogenetic
  default-`latent()` worked example.
- **Animal models with individual pedigrees.** If your rows are
  *individuals* with a known pedigree rather than *species* with a
  phylogeny, swap `phylo_*()` for the parallel `animal_*()` family
  ([`animal_scalar`](../reference/animal_scalar.html),
  [`animal_unique`](../reference/animal_unique.html),
  [`animal_indep`](../reference/animal_indep.html),
  [`animal_dep`](../reference/animal_dep.html),
  [`animal_latent`](../reference/animal_latent.html), and
  [`animal_slope`](../reference/animal_slope.html) for
  random-regression / reaction-norm models). The math is
  identical — both keyword families use a known relatedness
  matrix $\mathbf{A}$ — only the source differs (pedigree via
  [`pedigree_to_A()`](../reference/pedigree_to_A.html) vs tree
  via `ape::vcv()`). The pedigree-side worked example with
  heritability, bivariate $\mathbf{G}$, and factor-analytic
  $\mathbf{G}$ is [animal-model](animal-model.html). See
  [Design 14](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/14-known-relatedness-keywords.md)
  for the unified math and the A-vs-V naming boundary.

References: Hadfield and Nakagawa (2010) for the phylogenetic mixed
model representation; Meyer and Kirkpatrick (2008) for reduced-rank
covariance estimation; Williams et al. (2025) for fast phylogenetic
GLMMs in the `glmmTMB` ecosystem.
