---
title: "Common pitfalls and how to avoid them"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Common pitfalls and how to avoid them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

Six recurring mistakes that look like bugs but aren't, plus the
one-line fix for each. Each pitfall starts with the symptom, then the
diagnosis, a short reproducible chunk, and a rule of thumb.

```{r setup}
library(gllvmTMB)
```

## 1. Factor level order on `trait`

**Symptom.** Fitted $\hat{\boldsymbol\Sigma}_{\text{unit}}$ rows don't line up
with the rows of your true `Lambda`.

**Diagnosis.** `factor(rep(traits, n))` defaults to sorting levels
alphabetically. If your `traits` vector is `c("length", "mass", "wing")`,
the model sees levels `c("length", "mass", "wing")`; if it is
`c("wing", "mass", "length")`, the model still sees alphabetical
order. The model is correct; the row labels are out of sync with
your truth matrix.

```{r pitfall-1}
traits <- c("length", "mass", "wing", "tarsus", "bill")

# WRONG: alphabetical levels (bill, length, mass, tarsus, wing)
levels(factor(rep(traits, 3)))

# RIGHT: pin levels to your declared order
levels(factor(rep(traits, 3), levels = traits))
```

**Rule of thumb.** *If you defined a true `Lambda` with named rows,
always pass `levels = <your trait vector>` when building the long-format
factor. Otherwise the recovery comparison is meaningless even when
the fit is right.*

## 2. What `extract_Sigma(level = "unit")` actually returns

**Symptom.** Diagonals of `extract_Sigma(fit, level = "unit")$Sigma`
look "underestimated" by exactly $\sigma^2_\varepsilon$.

**Diagnosis.** `extract_Sigma(level = "unit")` returns the
`latent()`-implied between-unit covariance
$\boldsymbol\Lambda \boldsymbol\Lambda^{\!\top}$ (plus the optional
`unique(0 + trait | site)` variance), **not** the marginal observational
variance $\mathrm{Var}(y_{it}) = \boldsymbol\Lambda \boldsymbol\Lambda^{\!\top} + \sigma^2_\varepsilon$.
The "missing" amount is real residual noise, not bias.

```{r pitfall-2}
set.seed(1)
sim <- simulate_site_trait(
  n_sites = 60, n_species = 1, n_traits = 3,
  mean_species_per_site = 1,
  Lambda_B = matrix(c(1.0, 0.4, -0.3,
                      0.0, 0.8,  0.5), 3, 2),
  S_B  = rep(0, 3), beta = matrix(0, 3, 2), seed = 1
)
fit <- gllvmTMB(value ~ 0 + trait + latent(0 + trait | site, d = 2),
                data = sim$data)

S_hat    <- extract_Sigma(fit, level = "unit")$Sigma
LLt_true <- tcrossprod(sim$truth$Lambda_B)
# Compare to LL^T (the right target) - not to LL^T + sigma^2 I
round(rbind(diag_LLt_true  = diag(LLt_true),
            diag_Sigma_hat = diag(S_hat)), 2)
```

**Rule of thumb.** *Compare `extract_Sigma(level = "unit")` to
$\boldsymbol\Lambda \boldsymbol\Lambda^{\!\top}$. If you want the
marginal Var(y) including residual noise, add
`+ unique(0 + trait | site)` to the formula and inspect
`extract_Sigma(level = "unit", part = "total")$Sigma` — the residual
`sigma_eps` is reported separately in `fit$report$sigma_eps`.*

## 3. The DGP must match the fitted model

**Symptom.** Simulation recovery shows large positive bias on
$\hat{\boldsymbol\Sigma}_{\text{unit}}$ diagonals, even with hundreds of sites.

**Diagnosis.** `simulate_site_trait()` ships with sensible defaults
that include random environmental fixed effects
($\beta_{tk} \sim \mathcal{N}(0, 0.5^2)$) and a per-trait diagonal
$S_B$. If your fitted model has no environmental term and no
`unique(0 + trait | site)`, that variance has nowhere to go and gets
absorbed into $\hat{\boldsymbol\Lambda}_B$. This was a real harness
bug caught during simulation work — see the
[simulation-recovery article](simulation-recovery.html).

```{r pitfall-3}
# CLEAN latent()-only DGP: explicitly switch off beta and S_B
sim_clean <- simulate_site_trait(
  n_sites = 50, n_species = 1, n_traits = 3,
  mean_species_per_site = 1,
  Lambda_B = matrix(c(1.0, 0.4, -0.3,
                      0.0, 0.8,  0.5), 3, 2),
  beta = matrix(0, 3, 2),   # <-- no env effects
  S_B  = rep(0, 3),         # <-- no diagonal trait variance
  seed = 1
)
```

**Rule of thumb.** *When simulating to test recovery of a latent()-only
fit, set `beta = matrix(0, n_traits, n_predictors)` and
`S_B = rep(0, n_traits)`. If you want those components in the DGP,
include them in the formula too: `+ (0 + trait):env_1` and/or
`+ unique(0 + trait | site)`.*

## 4. Lambda is not unique; Sigma is

**Symptom.** Two `gllvmTMB` fits on identical data return different
$\hat{\boldsymbol\Lambda}$ matrices.

**Diagnosis.** Reduced-rank factor models have a rotational
ambiguity: $\boldsymbol\Lambda$ and $\boldsymbol\Lambda \mathbf{Q}$
yield the same $\boldsymbol\Sigma_{\text{unit}}$ for any orthogonal $\mathbf{Q}$.
The implied covariance is identifiable; the loadings alone are not.
`getLoadings()` will surface a one-shot hint when you access raw
$\boldsymbol\Lambda$ on an unconstrained rank-$>1$ fit.

```{r pitfall-4}
# Sigma_unit is rotation-invariant
S_hat <- extract_Sigma(fit, level = "unit")$Sigma
round(S_hat, 2)

# Varimax gives a unique post-hoc rotation for interpretation
rot <- rotate_loadings(fit, level = "unit", method = "varimax")
round(rot$Lambda, 2)

# Or ask the package to suggest a constraint matrix to refit with
sug <- suggest_lambda_constraint(
  value ~ 0 + trait + latent(0 + trait | site, d = 2),
  data = sim$data, convention = "lower_triangular"
)
sug$constraint
```

**Rule of thumb.** *Use `extract_Sigma(level = "unit")`
(rotation-invariant) for covariances and correlations. Use
`rotate_loadings(fit, method = "varimax")` for a unique post-hoc
rotation. Use `suggest_lambda_constraint()` plus a refit only if you
need a specific identifiable $\boldsymbol\Lambda$ for confirmatory
testing — see the [lambda-constraint article](lambda-constraint.html)
for the full decision tree.*

## 5. `phylo_scalar()` vs `phylo_latent()`

**Symptom.** `phylo_scalar()` recovery returns a `lam_phy` that looks
several times smaller than the true per-trait phylogenetic variance
you simulated.

**Diagnosis.** `phylo_scalar()` (the propto-style term) fits a **single
shared scalar** $\lambda$ multiplying a free trait correlation
matrix. Only the product $\lambda \cdot \rho_{tt'}$ is identifiable
— if your data have trait-specific phylogenetic variances, the
fitted scalar can absorb part of that into the correlation matrix
and look small on its own. `phylo_latent(species, d = K)` instead
estimates a $T \times K$ phylogenetic loading matrix, giving
per-trait phylogenetic effects.

```{r pitfall-5, eval = requireNamespace("ape", quietly = TRUE)}
# Tiny tree-based example (only the formulae matter here)
set.seed(1)
tree <- ape::rcoal(15)
tree$tip.label <- paste0("sp", seq_len(15))

# Use phylo_latent (recommended) when traits have their own phylo signal:
#   value ~ 0 + trait + phylo_latent(species, d = 2)
# Use phylo_scalar() (single-scalar) only when one shared phylo factor is
# enough and trait correlations are the inferential target:
#   value ~ 0 + trait + phylo_scalar(species)
```

**Rule of thumb.** *Use `phylo_latent(species, d = K)` (with $K$ up to
`n_traits`) whenever you expect trait-specific phylogenetic effects.
Reserve `phylo_scalar()` for the one-scalar case where the trait
correlation matrix is the target. See
[phylogenetic-gllvm.html](phylogenetic-gllvm.html) for full examples.*

## 6. The `unit = "..."` argument

**Symptom.** Error: `Column site not found in data`.

**Diagnosis.** `gllvmTMB()` defaults to `unit = "site"`, but in
morphometric studies the grouping is often `individual`; in
meta-analyses it might be `study_id`. The engine errors when the
literal column `site` is not present.

```{r pitfall-6}
# Rename to mimic a morphometric dataset
df <- sim$data
df$individual <- df$site
df$site <- NULL

# WRONG (uncomment to see the error):
# gllvmTMB(value ~ 0 + trait + latent(0 + trait | individual, d = 2), data = df)
# #> Error: Column site not found in data

# RIGHT: tell the engine which column groups observations
fit_ind <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | individual, d = 2),
  data = df, unit = "individual"
)
fit_ind$unit_col
```

**Rule of thumb.** *If your grouping column is not literally named
`site`, pass `unit = "<your_column>"` to `gllvmTMB()`. The
[morphometrics article](morphometrics.html) shows this in the
canonical individual × trait setup.*

## See also

* [morphometrics.html](morphometrics.html) — single-level GLLVM with
  `unit = "individual"`.
* [simulation-recovery.html](simulation-recovery.html) — the harness
  bug behind pitfall 3, plus the broader recovery study.
* [lambda-constraint.html](lambda-constraint.html) — the full
  decision tree for whether you actually need a constraint.
* [phylogenetic-gllvm.html](phylogenetic-gllvm.html) — `phylo_latent()`
  vs `phylo_scalar()` head-to-head.
* `?suggest_lambda_constraint`, `?extract_Sigma`,
  `?simulate_site_trait` — the function-level help pages.
