---
title: "Common pitfalls and how to avoid them"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Common pitfalls and how to avoid them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

Seven checks for when a multivariate fit looks wrong. The first step
is to ask whether the data coding, target summary, formula, or
covariance structure is mismatched before treating the result as an
optimizer bug. Each pitfall starts with the symptom, states the
general diagnostic principle, gives one concrete `gllvmTMB()` example,
and ends with a rule of thumb.

The pitfalls below use the long-format API
(`gllvmTMB(value ~ ..., data = df_long, trait = "trait")`). Many
examples have an equivalent wide-data-frame form via the
`traits(...)` LHS marker, but this page shows long form so the
diagnostic target is explicit. The point of each section is not the
particular example model; it is the more general check that applies
to the same model written in long or wide form. See the Get Started
vignette for the long/wide equivalence check.

> **Scope boundary.** This page is a troubleshooting reference for
> already-advertised workflows, not a new capability announcement.
> The examples illustrate general diagnostic checks; they are not a
> claim that these are the only models where the checks matter.
> IN: explicit long-format calls (`FG-02`), the wide `traits(...)`
> interface referenced from Get Started (`FG-03`), Gaussian
> `latent()` / `unique()` covariance examples (`FAM-01`, `FG-04`,
> `FG-05`, `FG-06`), `extract_Sigma()` (`EXT-01`), loading rotation
> and constraint helpers (`EXT-14`, `EXT-15`, `LAM-04`), paired and
> three-piece phylogenetic decompositions (`PHY-02`, `PHY-03`,
> `PHY-07`), and the A/Ainv relatedness path for animal-style
> examples (`ANI-01`, `ANI-07`, `ANI-08`). PARTIAL: `phylo_scalar()`
> and exact single-V `meta_V(V = V)` are shown as caveats rather than
> fully validated workflow recommendations (`PHY-04`, `MET-01`).
> PLANNED/BLOCKED: proportional known-V mode is not presented as a
> current workflow (`MET-03`).

```{r setup}
library(gllvmTMB)
```

```{r load-example-data, include = FALSE}
covex_path <- system.file(
  "extdata", "examples", "covariance-edge-cases-example.rds",
  package = "gllvmTMB"
)
covex_path <- c(
  covex_path,
  file.path("inst", "extdata", "examples", "covariance-edge-cases-example.rds"),
  file.path("..", "inst", "extdata", "examples", "covariance-edge-cases-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "covariance-edge-cases-example.rds")
)
covex_path <- covex_path[nzchar(covex_path) & file.exists(covex_path)][1]
stopifnot(!is.na(covex_path))
covex <- readRDS(covex_path)
```

## 1. Keep trait labels and matrix rows in the same order

**Symptom.** A fitted trait covariance, correlation, or loading matrix
has the right labels but the rows do not line up with the reference
matrix you expected.

**Diagnosis.** `factor(rep(traits, n))` defaults to sorting levels
alphabetically. If your `traits` vector is
`c("boldness", "exploration", "activity")`, the model sees levels
`c("activity", "boldness", "exploration")` unless you pin the order.
The fitted model can be correct while your comparison is wrong: the
row labels are out of sync with the matrix you are using as a
reference.

```{r pitfall-1}
traits <- covex$truth$trait_names

# MISMATCHED: alphabetical levels (activity, aggression, boldness, ...)
levels(factor(rep(traits, 3)))[1:3]

# MATCHED: pin levels to your declared order
levels(factor(rep(traits, 3), levels = traits))[1:3]
```

**Rule of thumb.** *Whenever you compare model output with a known
matrix or a manually ordered display, pin the trait factor with
`levels = <your trait vector>`. Otherwise the comparison can be
meaningless even when the fit is right.*

## 2. Compare the estimate with the estimand you asked for

**Symptom.** A variance or covariance summary looks too small or too
large by a familiar amount, such as a residual variance or a diagonal
trait-specific component.

**Diagnosis.** First check what the extractor is designed to return.
In this example, the formula requests the no-Psi subset with
`latent(..., unique = FALSE)`, so `extract_Sigma(level = "unit")`
returns the `latent()`-implied between-unit covariance
$\boldsymbol\Lambda \boldsymbol\Lambda^{\!\top}$, **not** the marginal
observational variance
$\mathrm{Var}(y_{it}) = \boldsymbol\Lambda \boldsymbol\Lambda^{\!\top} + \sigma^2_\varepsilon$.
The "missing" amount is real residual noise, not bias in the
between-unit covariance.

```{r pitfall-2}
set.seed(1)
sim <- simulate_site_trait(
  n_sites = 60, n_species = 1, n_traits = 3,
  mean_species_per_site = 1,
  Lambda_B = matrix(c(1.0, 0.4, -0.3,
                      0.0, 0.8,  0.5), 3, 2),
  psi_B  = rep(0, 3), beta = matrix(0, 3, 2), seed = 1
)
fit <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2, unique = FALSE),
  data = sim$data, trait = "trait"
)

S_hat    <- extract_Sigma(fit, level = "unit")$Sigma
LLt_true <- tcrossprod(sim$truth$Lambda_B)
# Compare to LL^T (the right target) - not to LL^T + sigma^2 I
round(rbind(diag_LLt_true  = diag(LLt_true),
            diag_Sigma_hat = diag(S_hat)), 2)
```

**Rule of thumb.** *Name the estimand before judging the estimate.
For this example, compare `extract_Sigma(level = "unit")` to
$\boldsymbol\Lambda \boldsymbol\Lambda^{\!\top}$ because the formula requested
`unique = FALSE`. If the scientific target includes trait-specific
diagonal $\boldsymbol\Psi$, use the default `latent()` fit and inspect
`extract_Sigma(level = "unit", part = "total")$Sigma`. The residual
`sigma_eps` is reported separately in `fit$report$sigma_eps`; do not silently
mix it into a between-unit covariance target.*

## 3. Make the data-generating story match the formula

**Symptom.** A simulation-recovery check or a fitted analysis shows
extra variance in a component that should have been clean.

**Diagnosis.** If the data include a component that the formula does
not contain, the fit has to put that variation somewhere. In this
example, `simulate_site_trait()` ships with sensible defaults that
include random environmental fixed effects
($\beta_{tk} \sim \mathcal{N}(0, 0.5^2)$) and a per-trait diagonal
$psi_B$. If your fitted model has no environmental term and no
diagonal $\boldsymbol\Psi$ component, that variance has nowhere to go and gets
absorbed into $\hat{\boldsymbol\Lambda}_B$. This is the general
failure mode: the fitted model can answer a different question from
the data-generating story even when the optimizer behaved correctly.

```{r pitfall-3}
# CLEAN latent()-only DGP: explicitly switch off beta and psi_B
sim_clean <- simulate_site_trait(
  n_sites = 50, n_species = 1, n_traits = 3,
  mean_species_per_site = 1,
  Lambda_B = matrix(c(1.0, 0.4, -0.3,
                      0.0, 0.8,  0.5), 3, 2),
  beta = matrix(0, 3, 2),   # <-- no env effects
  psi_B  = rep(0, 3),         # <-- no diagonal trait variance
  seed = 1
)
```

**Rule of thumb.** *For simulation checks, either remove a component
from the data-generating process or include the matching term in the
formula. For this latent-only example, set
`beta = matrix(0, n_traits, n_predictors)` and
`psi_B = rep(0, n_traits)`. If you want those components in the data,
include them in the formula too: `+ (0 + trait):env_1` and ordinary
default `latent(0 + trait | site, d = K)` for the diagonal
$\boldsymbol\Psi$ companion.*

## 4. Interpret identifiable summaries, not arbitrary coordinates

**Symptom.** Two fits on identical data return different internal
coordinates, such as different $\hat{\boldsymbol\Lambda}$ matrices.

**Diagnosis.** Reduced-rank factor models have a rotational
ambiguity: $\boldsymbol\Lambda$ and $\boldsymbol\Lambda \mathbf{Q}$
yield the same $\boldsymbol\Sigma_{\text{unit}}$ for any orthogonal $\mathbf{Q}$.
The implied covariance is identifiable; the loadings alone are not.
`getLoadings()` will surface a one-shot hint when you access raw
$\boldsymbol\Lambda$ on an unconstrained rank-$>1$ fit.

```{r pitfall-4}
# Sigma_unit is rotation-invariant
S_hat <- extract_Sigma(fit, level = "unit")$Sigma
round(S_hat, 2)

# Varimax gives a unique post-hoc rotation for interpretation
rot <- rotate_loadings(fit, level = "unit", method = "varimax")
round(rot$Lambda, 2)

# Or ask the package to suggest a constraint matrix to refit with
sug <- suggest_lambda_constraint(
  value ~ 0 + trait + latent(0 + trait | site, d = 2),
  data = sim$data, convention = "lower_triangular"
)
sug$constraint
```

**Rule of thumb.** *Report the summary that is identifiable for the
question. In this example, use `extract_Sigma(level = "unit")`
(rotation-invariant) for covariances and correlations. Use
`rotate_loadings(fit, method = "varimax")` for a unique post-hoc
rotation. Use `suggest_lambda_constraint()` plus a refit only if you
need a specific identifiable $\boldsymbol\Lambda$ for confirmatory
testing. The confirmatory-loading article is under audit during the
public-surface reset.*

## 5. Choose the decomposition your design can identify

**Symptom.** A structured covariance term captures the broad shared
pattern, but trait-specific variance lands in the wrong place or is
not separately recoverable.

**Diagnosis.** The general issue is identifiability: do not ask the
data to separate more variance pieces than the design can support.
For any `latent()`, `unique()`, `indep()`, `dep()`, or `*_scalar()`
choice, first name the covariance tier, decide which diagonal variance
is separately identifiable, and use scalar shortcuts only when a shared
variance is the estimand. The phylogenetic example below is one
instance. A bare
`phylo_latent(species, d = K)` is the no-residual reduced-rank subset
-- it captures a shared phylogenetic axis but has no diagonal for
trait-specific phylogenetic variance. A bare `phylo_scalar()` fits a
single shared scalar $\lambda$ multiplying a free trait correlation
matrix, so only the product $\lambda \cdot \rho_{tt'}$ is
identifiable. Neither bare form is a per-trait phylogenetic-variance
recovery model.

The **fully paired phylogenetic decomposition** pairs four
components:

$$
\Sigma_{\text{phy}} = \Lambda_{\text{phy}}\Lambda_{\text{phy}}^{\!\top} + \boldsymbol\Psi_{\text{phy}}, \qquad
\Sigma_{\text{non}} = \Lambda_{\text{non}}\Lambda_{\text{non}}^{\!\top} + \boldsymbol\Psi_{\text{non}}.
$$

```{r pitfall-5, eval = requireNamespace("ape", quietly = TRUE)}
# Tiny tree-based example (only the formulae matter here)
set.seed(1)
tree <- ape::rcoal(15)
tree$tip.label <- paste0("sp", seq_len(15))

# Fully paired phylogenetic decomposition:
#   value ~ 0 + trait +
#     phylo_latent(species, d = K_phy, tree = tree) +
#     phylo_unique(species, tree = tree) +
#     latent(0 + trait | species, d = K_non)

# Single-scalar simplification (only when trait correlations are
# the inferential target and you do not need per-trait variances):
#   value ~ 0 + trait + phylo_scalar(species, tree = tree)
```

**Rule of thumb.** *Start from the scientific target and the design.
Use the paired four-piece decomposition
(`phylo_latent + phylo_unique + latent`) only when **both
$\boldsymbol{\Psi}$ diagonals are identifiable**, typically with
crossed replication, enough species, and strong phylogenetic signal.
Reserve `phylo_scalar()` for the simpler case where a single shared
phylogenetic variance and trait correlations are the inferential
target. The four-piece form is what `extract_phylo_signal()`
decomposes into $H^2 + C^2_{\text{non}} + \psi^2 = 1$. The worked
phylogenetic article is under audit until its helper, validation
rows, and rendered HTML review pass.*

*When the phylogenetic uniqueness component $\boldsymbol\Psi_{\text{phy}}$ is
not separately identifiable from $\Lambda_{\text{phy}}
\Lambda_{\text{phy}}^{\!\top}$ -- common in small `n_species`,
weak phylogenetic signal, or single-replicate-per-tip data --
drop the `phylo_unique()` term and let the species-level
ordinary `latent()` carry the only $\boldsymbol{\Psi}$ in the fit. The model then has
three pieces:*

$$
\Omega = \Lambda_{\text{phy}} \Lambda_{\text{phy}}^{\!\top} +
\Lambda_{\text{non}} \Lambda_{\text{non}}^{\!\top} + \boldsymbol\Psi
$$

*-- $\Omega$ is simply the **total trait covariance**, the sum
of every variance component in whichever model you fit (four
pieces in the paired form above, three pieces here, more if you
add `spatial_*` or extra latent tiers). The number of pieces
follows from the keywords in the formula. `gllvmTMB(... +
phylo_latent + latent, ...)` is a three-piece fallback for
this case. The validation evidence for this fallback is
tracked under `PHY-03`; future worked phylogenetic examples should
present this as one concrete case of the broader identifiability
check, not as a special rule for one model family.*

## 6. Declare the observation-grouping column

**Symptom.** Error: `Column site not found in data`.

**Diagnosis.** Every model needs to know which column identifies the
observational unit for stacked trait rows. `gllvmTMB()` defaults to
`unit = "site"`, but in morphometric studies the grouping is often
`individual`; in meta-analyses it might be `study_id`. The engine
errors when the literal column `site` is not present.

```{r pitfall-6}
# Rename to mimic a morphometric dataset
df <- sim$data
df$individual <- df$site
df$site <- NULL

# MISMATCHED (uncomment to see the error):
# gllvmTMB(value ~ 0 + trait + latent(0 + trait | individual, d = 2),
#          data = df, trait = "trait")
# #> Error: Column site not found in data

# MATCHED: tell the engine which column groups observations
fit_ind <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | individual, d = 2),
  data = df, trait = "trait", unit = "individual"
)
fit_ind$unit_col
```

**Rule of thumb.** *If your grouping column is not literally named
`site`, pass `unit = "<your_column>"` to `gllvmTMB()`. The
[morphometrics article](morphometrics.html) shows this in the
canonical individual × trait setup.*

## 7. Separate relatedness covariance from sampling variance

**Symptom.** A known matrix is available, but it is unclear whether
the matrix belongs in a relatedness keyword or a measurement-error
keyword. For example, you are setting up an animal-model fit, see
that `meta_V()` accepts a `V =` argument, and try to pass your
pedigree-derived relatedness matrix as `V =`. The fit either errors
with a shape mismatch or estimates a variance component with an
unintended interpretation.

**Diagnosis.** First name what the matrix indexes and which variance
term it represents. **A** and **V** mean different things in gllvmTMB:

| Symbol | Meaning | Keyword(s) | Shape |
|--------|---------|-----------|-------|
| **A** / **Ainv** | Known *relatedness* covariance (or its inverse) | `animal_*(..., A = A)`, `animal_*(..., Ainv = Ainv)`, `phylo_*(..., A = A)`, `phylo_*(..., Ainv = Ainv)` | $n \times n$ symmetric PSD |
| **V** | Known *sampling variance* (measurement-error variance in meta-analysis) | `meta_V(V = V)` | scalar per row OR a block-V matrix |

The math is genuinely distinct: $\mathbf{A}$ is the
between-individual relatedness matrix from a pedigree or
phylogeny; $\mathbf{V}$ is the between-study measurement-error
variance from a meta-analysis. Conflating them produces fits whose
variance components answer a different question because the engine
expects each matrix at a different place in the decomposition.

```r
# MISMATCHED: passing relatedness matrix as V to meta_V
fit <- gllvmTMB(value ~ 0 + trait + meta_V(V = A_pedigree),
                data = df, trait = "trait")
# MATCHED: relatedness goes to animal_* / phylo_*, sampling variance to meta_V
fit <- gllvmTMB(value ~ 0 + trait +
                  animal_scalar(individual, A = A_pedigree) +
                  meta_V(V = sampling_variance),
                data = df, trait = "trait")
```

**Rule of thumb.** *Use **A** / **Ainv** for relatedness, use **V**
for sampling variance. Keep the two separate. See
[Design 14 §3](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/14-known-relatedness-keywords.md)
for the full convention, and the [Vocabulary glossary](gllvm-vocabulary.html)
for plain-language definitions.*

## See also

* [morphometrics.html](morphometrics.html) — single-level GLLVM with
  `unit = "individual"`.
* [Formula keyword grid](api-keyword-grid.html) — syntax map for
  covariance keywords and status labels.
* [Response families](response-families.html) — status-labelled family
  lookup.
* Under audit: lambda constraints, phylogenetic GLLVMs, and animal-model
  examples will return after their example and validation gates pass.
* `?suggest_lambda_constraint`, `?extract_Sigma`,
  `?simulate_site_trait` — the function-level help pages.
