---
title: "Common pitfalls and how to avoid them"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Common pitfalls and how to avoid them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

Seven recurring mistakes that look like bugs but aren't, plus the
one-line fix for each. Each pitfall starts with the symptom, then the
diagnosis, a short reproducible chunk, and a rule of thumb.

The pitfalls below use the long-format API
(`gllvmTMB(value ~ ..., data = df_long, trait = "trait")`). Many
examples have an equivalent wide-data-frame form via the
`traits(...)` LHS marker, but this page shows long form so the
diagnostic target is explicit. See the Get Started vignette for the
long/wide equivalence check.

> **Scope boundary.** This page is a troubleshooting reference for
> already-advertised workflows, not a new capability announcement.
> IN: explicit long-format calls (`FG-02`), the wide `traits(...)`
> interface referenced from Get Started (`FG-03`), Gaussian
> `latent()` / `unique()` covariance examples (`FAM-01`, `FG-04`,
> `FG-05`, `FG-06`), `extract_Sigma()` (`EXT-01`), loading rotation
> and constraint helpers (`EXT-14`, `EXT-15`, `LAM-04`), paired and
> three-piece phylogenetic decompositions (`PHY-02`, `PHY-03`,
> `PHY-07`), and the A/Ainv relatedness path for animal-style
> examples (`ANI-01`, `ANI-07`, `ANI-08`). PARTIAL: `phylo_scalar()`
> and exact single-V `meta_V(V = V)` are shown as caveats rather than
> fully validated workflow recommendations (`PHY-04`, `MET-01`).
> PLANNED/BLOCKED: proportional known-V mode is not presented as a
> current workflow (`MET-03`).

```{r setup}
library(gllvmTMB)
```

```{r load-example-data, include = FALSE}
covex_path <- system.file(
  "extdata", "examples", "covariance-edge-cases-example.rds",
  package = "gllvmTMB"
)
covex_path <- c(
  covex_path,
  file.path("inst", "extdata", "examples", "covariance-edge-cases-example.rds"),
  file.path("..", "inst", "extdata", "examples", "covariance-edge-cases-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "covariance-edge-cases-example.rds")
)
covex_path <- covex_path[nzchar(covex_path) & file.exists(covex_path)][1]
stopifnot(!is.na(covex_path))
covex <- readRDS(covex_path)
```

## 1. Factor level order on `trait`

**Symptom.** Fitted $\hat{\boldsymbol\Sigma}_{\text{unit}}$ rows don't line up
with the rows of your true `Lambda`.

**Diagnosis.** `factor(rep(traits, n))` defaults to sorting levels
alphabetically. If your `traits` vector is
`c("boldness", "exploration", "activity")`, the model sees levels
`c("activity", "boldness", "exploration")` unless you pin the order.
The model is correct; the row labels are out of sync with your truth
matrix.

```{r pitfall-1}
traits <- covex$truth$trait_names

# WRONG: alphabetical levels (activity, aggression, boldness, ...)
levels(factor(rep(traits, 3)))[1:3]

# RIGHT: pin levels to your declared order
levels(factor(rep(traits, 3), levels = traits))[1:3]
```

**Rule of thumb.** *If you defined a true `Lambda` with named rows,
always pass `levels = <your trait vector>` when building the long-format
factor. Otherwise the recovery comparison is meaningless even when
the fit is right.*

## 2. What `extract_Sigma(level = "unit")` actually returns

**Symptom.** Diagonals of `extract_Sigma(fit, level = "unit")$Sigma`
look "underestimated" by exactly $\sigma^2_\varepsilon$.

**Diagnosis.** `extract_Sigma(level = "unit")` returns the
`latent()`-implied between-unit covariance
$\boldsymbol\Lambda \boldsymbol\Lambda^{\!\top}$ (plus the optional
`unique(0 + trait | site)` variance), **not** the marginal observational
variance $\mathrm{Var}(y_{it}) = \boldsymbol\Lambda \boldsymbol\Lambda^{\!\top} + \sigma^2_\varepsilon$.
The "missing" amount is real residual noise, not bias.

```{r pitfall-2}
set.seed(1)
sim <- simulate_site_trait(
  n_sites = 60, n_species = 1, n_traits = 3,
  mean_species_per_site = 1,
  Lambda_B = matrix(c(1.0, 0.4, -0.3,
                      0.0, 0.8,  0.5), 3, 2),
  psi_B  = rep(0, 3), beta = matrix(0, 3, 2), seed = 1
)
fit <- gllvmTMB(value ~ 0 + trait + latent(0 + trait | site, d = 2),
                data = sim$data, trait = "trait")

S_hat    <- extract_Sigma(fit, level = "unit")$Sigma
LLt_true <- tcrossprod(sim$truth$Lambda_B)
# Compare to LL^T (the right target) - not to LL^T + sigma^2 I
round(rbind(diag_LLt_true  = diag(LLt_true),
            diag_Sigma_hat = diag(S_hat)), 2)
```

**Rule of thumb.** *Compare `extract_Sigma(level = "unit")` to
$\boldsymbol\Lambda \boldsymbol\Lambda^{\!\top}$. If you want the
marginal Var(y) including residual noise, add
`+ unique(0 + trait | site)` to the formula and inspect
`extract_Sigma(level = "unit", part = "total")$Sigma` — the residual
`sigma_eps` is reported separately in `fit$report$sigma_eps`.*

## 3. The DGP must match the fitted model

**Symptom.** Simulation recovery shows large positive bias on
$\hat{\boldsymbol\Sigma}_{\text{unit}}$ diagonals, even with hundreds of sites.

**Diagnosis.** `simulate_site_trait()` ships with sensible defaults
that include random environmental fixed effects
($\beta_{tk} \sim \mathcal{N}(0, 0.5^2)$) and a per-trait diagonal
$psi_B$. If your fitted model has no environmental term and no
`unique(0 + trait | site)`, that variance has nowhere to go and gets
absorbed into $\hat{\boldsymbol\Lambda}_B$. This was a real harness
bug caught during simulation work.

```{r pitfall-3}
# CLEAN latent()-only DGP: explicitly switch off beta and psi_B
sim_clean <- simulate_site_trait(
  n_sites = 50, n_species = 1, n_traits = 3,
  mean_species_per_site = 1,
  Lambda_B = matrix(c(1.0, 0.4, -0.3,
                      0.0, 0.8,  0.5), 3, 2),
  beta = matrix(0, 3, 2),   # <-- no env effects
  psi_B  = rep(0, 3),         # <-- no diagonal trait variance
  seed = 1
)
```

**Rule of thumb.** *When simulating to test recovery of a latent()-only
fit, set `beta = matrix(0, n_traits, n_predictors)` and
`psi_B = rep(0, n_traits)`. If you want those components in the DGP,
include them in the formula too: `+ (0 + trait):env_1` and/or
`+ unique(0 + trait | site)`.*

## 4. Lambda is not unique; Sigma is

**Symptom.** Two `gllvmTMB` fits on identical data return different
$\hat{\boldsymbol\Lambda}$ matrices.

**Diagnosis.** Reduced-rank factor models have a rotational
ambiguity: $\boldsymbol\Lambda$ and $\boldsymbol\Lambda \mathbf{Q}$
yield the same $\boldsymbol\Sigma_{\text{unit}}$ for any orthogonal $\mathbf{Q}$.
The implied covariance is identifiable; the loadings alone are not.
`getLoadings()` will surface a one-shot hint when you access raw
$\boldsymbol\Lambda$ on an unconstrained rank-$>1$ fit.

```{r pitfall-4}
# Sigma_unit is rotation-invariant
S_hat <- extract_Sigma(fit, level = "unit")$Sigma
round(S_hat, 2)

# Varimax gives a unique post-hoc rotation for interpretation
rot <- rotate_loadings(fit, level = "unit", method = "varimax")
round(rot$Lambda, 2)

# Or ask the package to suggest a constraint matrix to refit with
sug <- suggest_lambda_constraint(
  value ~ 0 + trait + latent(0 + trait | site, d = 2),
  data = sim$data, convention = "lower_triangular"
)
sug$constraint
```

**Rule of thumb.** *Use `extract_Sigma(level = "unit")`
(rotation-invariant) for covariances and correlations. Use
`rotate_loadings(fit, method = "varimax")` for a unique post-hoc
rotation. Use `suggest_lambda_constraint()` plus a refit only if you
need a specific identifiable $\boldsymbol\Lambda$ for confirmatory
testing. The confirmatory-loading article is under audit during the
public-surface reset.*

## 5. Phylogenetic decomposition: paired or scalar?

**Symptom.** Phylogenetic per-trait variances look several times
smaller than the values you simulated, OR the phylogenetic shared
axis absorbs trait-specific variance that should sit on the
diagonal.

**Diagnosis.** A bare `phylo_latent(species, d = K)` is the
no-residual reduced-rank subset — it captures a shared phylogenetic
axis but has no diagonal for trait-specific phylogenetic variance.
A bare `phylo_scalar()` fits a single shared scalar $\lambda$
multiplying a free trait correlation matrix, so only the product
$\lambda \cdot \rho_{tt'}$ is identifiable. Neither bare form
recovers per-trait phylogenetic variances reliably.

The **canonical paired phylogenetic decomposition** pairs four
components:

$$
\Sigma_{\text{phy}} = \Lambda_{\text{phy}}\Lambda_{\text{phy}}^{\!\top} + \boldsymbol\Psi_{\text{phy}}, \qquad
\Sigma_{\text{non}} = \Lambda_{\text{non}}\Lambda_{\text{non}}^{\!\top} + \boldsymbol\Psi_{\text{non}}.
$$

```{r pitfall-5, eval = requireNamespace("ape", quietly = TRUE)}
# Tiny tree-based example (only the formulae matter here)
set.seed(1)
tree <- ape::rcoal(15)
tree$tip.label <- paste0("sp", seq_len(15))

# Canonical paired phylogenetic decomposition:
#   value ~ 0 + trait +
#     phylo_latent(species, d = K_phy, tree = tree) +
#     phylo_unique(species, tree = tree) +
#     latent(0 + trait | species, d = K_non) +
#     unique(0 + trait | species)

# Single-scalar simplification (only when trait correlations are
# the inferential target and you do not need per-trait variances):
#   value ~ 0 + trait + phylo_scalar(species, tree = tree)
```

**Rule of thumb.** *Use the paired four-component decomposition
(`phylo_latent + phylo_unique + latent + unique`) whenever **both
$\boldsymbol{\Psi}$ diagonals are identifiable** -- typically a crossed
(site × species) design with `n_species` around 100 or more and
strong phylogenetic signal. Reserve `phylo_scalar()` for the
simpler case where a single shared phylogenetic variance per trait
is enough and trait correlations are the inferential target. The
four-component form is what `extract_phylo_signal()` decomposes
into $H^2 + C^2_{\text{non}} + \psi^2 = 1$. The worked phylogenetic
article is under audit until its helper, validation rows, and rendered
HTML review pass.*

*When the phylogenetic uniqueness component $\boldsymbol\Psi_{\text{phy}}$ is
not separately identifiable from $\Lambda_{\text{phy}}
\Lambda_{\text{phy}}^{\!\top}$ -- common in small `n_species`,
weak phylogenetic signal, or single-replicate-per-tip data --
drop the `phylo_unique()` term and let the species-level
`unique()` carry the only $\boldsymbol{\Psi}$ in the fit. The model then has
three pieces:*

$$
\Omega = \Lambda_{\text{phy}} \Lambda_{\text{phy}}^{\!\top} +
\Lambda_{\text{non}} \Lambda_{\text{non}}^{\!\top} + \boldsymbol\Psi
$$

*-- $\Omega$ is simply the **total trait covariance**, the sum
of every variance component in whichever model you fit (four
pieces in the paired form above, three pieces here, more if you
add `spatial_*` or extra latent tiers). The number of pieces
follows from the keywords in the formula. `gllvmTMB(... +
phylo_latent + latent + unique, ...)` is the canonical form for
this three-piece case. The validation evidence for this fallback is
tracked under `PHY-03`; the future functional-biogeography capstone
will turn the same $p_{it} + q_{it}$ identifiability case into a
worked biological example after its component helpers and article
review pass.*

## 6. The `unit = "..."` argument

**Symptom.** Error: `Column site not found in data`.

**Diagnosis.** `gllvmTMB()` defaults to `unit = "site"`, but in
morphometric studies the grouping is often `individual`; in
meta-analyses it might be `study_id`. The engine errors when the
literal column `site` is not present.

```{r pitfall-6}
# Rename to mimic a morphometric dataset
df <- sim$data
df$individual <- df$site
df$site <- NULL

# WRONG (uncomment to see the error):
# gllvmTMB(value ~ 0 + trait + latent(0 + trait | individual, d = 2),
#          data = df, trait = "trait")
# #> Error: Column site not found in data

# RIGHT: tell the engine which column groups observations
fit_ind <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | individual, d = 2),
  data = df, trait = "trait", unit = "individual"
)
fit_ind$unit_col
```

**Rule of thumb.** *If your grouping column is not literally named
`site`, pass `unit = "<your_column>"` to `gllvmTMB()`. The
[morphometrics article](morphometrics.html) shows this in the
canonical individual × trait setup.*

## 7. A vs V: relatedness covariance vs sampling variance

**Symptom.** You're setting up an animal-model fit, see that
`meta_V()` accepts a `V =` argument, and try to pass your
pedigree-derived relatedness matrix as `V =`. The fit either
errors with a shape mismatch or produces nonsense estimates.

**Diagnosis.** **A** and **V** mean different things in gllvmTMB:

| Symbol | Meaning | Keyword(s) | Shape |
|--------|---------|-----------|-------|
| **A** / **Ainv** | Known *relatedness* covariance (or its inverse) | `animal_*(..., A = A)`, `animal_*(..., Ainv = Ainv)`, `phylo_*(..., A = A)`, `phylo_*(..., Ainv = Ainv)` | $n \times n$ symmetric PSD |
| **V** | Known *sampling variance* (measurement-error variance in meta-analysis) | `meta_V(V = V)` | scalar per row OR a block-V matrix |

The math is genuinely distinct: $\mathbf{A}$ is the
between-individual relatedness matrix from a pedigree or
phylogeny; $\mathbf{V}$ is the between-study measurement-error
variance from a meta-analysis. Conflating them produces
nonsensical fits because the engine expects each at a different
place in the variance decomposition.

```r
# WRONG: passing relatedness matrix as V to meta_V
fit <- gllvmTMB(value ~ 0 + trait + meta_V(V = A_pedigree),
                data = df, trait = "trait")
# RIGHT: relatedness goes to animal_* / phylo_*, sampling variance to meta_V
fit <- gllvmTMB(value ~ 0 + trait +
                  animal_scalar(individual, A = A_pedigree) +
                  meta_V(V = sampling_variance),
                data = df, trait = "trait")
```

**Rule of thumb.** *Use **A** / **Ainv** for relatedness, use **V**
for sampling variance. Never blur the two. See
[Design 14 §3](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/14-known-relatedness-keywords.md)
for the full convention. The public vocabulary glossary is under audit.*

## See also

* [morphometrics.html](morphometrics.html) — single-level GLLVM with
  `unit = "individual"`.
* [Formula keyword grid](api-keyword-grid.html) — syntax map for
  covariance keywords and status labels.
* [Response families](response-families.html) — status-labelled family
  lookup.
* Under audit: lambda constraints, phylogenetic GLLVMs, and animal-model
  examples will return after their example and validation gates pass.
* `?suggest_lambda_constraint`, `?extract_Sigma`,
  `?simulate_site_trait` — the function-level help pages.
