---
title: "Common pitfalls and how to avoid them"
description: "Troubleshoot six common data, formula, covariance, and interpretation mistakes with concrete long-format checks and examples."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Common pitfalls and how to avoid them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = TRUE
)
```

Six checks for when a multivariate fit looks wrong. First ask whether the data
coding, target summary, formula, or covariance structure is mismatched before
treating the result as an optimizer bug. Each pitfall starts with a symptom,
gives one concrete check, and ends with a rule of thumb.

Use this page to diagnose data and model mismatches after confirming that the
fitted point surface is stable; it is not a substitute for convergence or
uncertainty checks.

The examples use the long-format API because it makes each diagnostic target
explicit. The same checks apply when the model is written in wide data-frame
form with `traits(...)`. See [Get started](../articles/gllvmTMB.html) for the
side-by-side long and wide calls.

> **Before diagnosing a pitfall.** Check `fit$fit_health$converged`,
> `fit$fit_health$optimizer_converged`, and the raw
> `fit$fit_health$max_gradient`, then run `check_gllvmTMB(fit)`. The
> objective-scaled `fit$fit_health$scaled_gradient` is a secondary descriptive
> signal; it cannot override a failed optimiser code or a large raw gradient. A failed
> Hessian or standard-error calculation warns against Wald inference; it does
> not by itself prove that the point estimate failed. Conversely, profiles or
> bootstraps cannot rescue an unstable point surface. See
> [Fit diagnostics](fit-diagnostics.html) and
> [Convergence and start values](convergence-start-values.html) for the full
> decision path.

```{r setup}
library(gllvmTMB)
```

```{r load-example-data, include = FALSE}
covex_path <- system.file(
  "extdata", "examples", "covariance-edge-cases-example.rds",
  package = "gllvmTMB"
)
covex_path <- c(
  covex_path,
  file.path("inst", "extdata", "examples", "covariance-edge-cases-example.rds"),
  file.path("..", "inst", "extdata", "examples", "covariance-edge-cases-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "covariance-edge-cases-example.rds")
)
covex_path <- covex_path[nzchar(covex_path) & file.exists(covex_path)][1]
stopifnot(!is.na(covex_path))
covex <- readRDS(covex_path)
```

## 1. Keep trait labels and matrix rows in the same order

**Symptom.** A fitted trait covariance, correlation, or loading matrix has the
right labels but the rows do not line up with the reference matrix.

**Diagnosis.** `factor(rep(traits, n))` sorts levels alphabetically by default.
If your declared order is `boldness`, `exploration`, `activity`, `aggression`,
`sociability`, the model sees a different order unless you pin the levels. The
fit can be correct while the comparison is wrong.

```{r pitfall-1}
traits <- covex$truth$trait_names

# MISMATCHED: alphabetical levels
levels(factor(rep(traits, 3)))

# MATCHED: pin levels to the declared order
levels(factor(rep(traits, 3), levels = traits))
```

**Rule of thumb.** *Whenever you compare model output with a known matrix or a
manually ordered display, pin the trait factor with `levels = <your trait
vector>`. Otherwise the comparison can be meaningless even when the fit is
right.*

## 2. Compare the estimate with the estimand you asked for

**Symptom.** A variance or covariance summary looks too small or too large by
a familiar amount, such as a residual variance or a diagonal trait-specific
component.

**Diagnosis.** First check what the extractor returns. Here the formula asks
for the loadings-only subset with `latent(..., unique = FALSE)`, so
`extract_Sigma(level = "unit")` returns the between-unit covariance
$\boldsymbol\Lambda\boldsymbol\Lambda^{\!\top}$. It does not return the
marginal observation covariance

$$
\operatorname{Var}(\mathbf y_i) =
\boldsymbol\Lambda\boldsymbol\Lambda^{\!\top} +
\sigma_\varepsilon^2\mathbf I.
$$

The apparently missing amount can therefore be real observation noise rather
than bias in the between-unit covariance.

```{r pitfall-2}
set.seed(1)
sim <- simulate_site_trait(
  n_sites = 60, n_species = 1, n_traits = 3,
  mean_species_per_site = 1,
  Lambda_B = matrix(c(1.0, 0.4, -0.3,
                      0.0, 0.8,  0.5), 3, 2),
  psi_B = NULL, beta = matrix(0, 3, 2), seed = 1
)
fit <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2, unique = FALSE),
  data = sim$data,
  trait = "trait",
  unit = "site"
)

S_hat    <- extract_Sigma(fit, level = "unit")$Sigma
LLt_true <- tcrossprod(sim$truth$Lambda_B)
round(rbind(diag_LLt_true   = diag(LLt_true),
            diag_Sigma_hat = diag(S_hat)), 2)
```

**Rule of thumb.** *Name the estimand before judging the estimate. If the
scientific target includes trait-specific diagonal $\boldsymbol\Psi$, use the
default `latent()` fit, or add `indep(0 + trait | site)` to a loadings-only
formula, and inspect the total between-unit covariance. Observation residual
variance is a separate component; do not silently mix it into that target.
Small finite-sample differences are expected even when the estimand matches.*

## 3. Make the data-generating story match the formula

**Symptom.** A simulation-recovery check shows extra variation in a component
that should have been clean.

**Diagnosis.** `simulate_site_trait()` generates nonzero environmental
coefficients when `beta = NULL`, whereas `psi_B = NULL` means that no
between-unit diagonal component is simulated. Observation residual variance
is generated separately through `sigma2_eps`. If the simulator contains an
environmental or diagonal component that the fitted formula omits, that
variation has to be absorbed elsewhere. The optimizer can behave correctly
while the fitted model answers a different question from the simulation.

```{r pitfall-3}
# A clean latent-only data-generating process
sim_clean <- simulate_site_trait(
  n_sites = 50, n_species = 1, n_traits = 3,
  mean_species_per_site = 1,
  Lambda_B = matrix(c(1.0, 0.4, -0.3,
                      0.0, 0.8,  0.5), 3, 2),
  beta = matrix(0, 3, 2),
  psi_B = NULL,
  seed = 1
)
```

**Rule of thumb.** *For a simulation check, either remove a component from the
data-generating process or include its matching term in the formula. For this
latent-only example, set `beta = matrix(0, n_traits, n_predictors)` and leave
`psi_B = NULL`. If the simulation includes environmental effects, include the
corresponding trait-specific slopes in the fit.*

## 4. Interpret rotation-invariant summaries, not arbitrary coordinates

**Symptom.** Two fits on identical data return different loading coordinates.

**Diagnosis.** Reduced-rank factor models have rotational ambiguity:
$\boldsymbol\Lambda$ and $\boldsymbol\Lambda\mathbf Q$ imply the same shared
covariance for any orthogonal $\mathbf Q$. The covariance is
rotation-invariant; the loading coordinates are not. Rotation invariance does
not guarantee precise estimation, so check fit health and uncertainty
separately.

```{r pitfall-4}
# Sigma is rotation-invariant
round(extract_Sigma(fit, level = "unit")$Sigma, 2)

# Varimax provides one reproducible presentation convention
rot <- rotate_loadings(fit, level = "unit", method = "varimax")
round(rot$Lambda, 2)

# Or obtain an identification scaffold for a constrained refit
sug <- suggest_lambda_constraint(
  value ~ 0 + trait + latent(0 + trait | site, d = 2),
  data = sim$data, convention = "lower_triangular"
)
sug$constraint
```

**Rule of thumb.** *Use `extract_Sigma()` for rotation-invariant covariance
questions. Use `rotate_loadings(..., method = "varimax")` for one ordered,
sign-anchored presentation of the axes, not a uniquely true biological
solution. `suggest_lambda_constraint()` supplies an identification convention;
confirmatory claims require constraints specified from the scientific
hypothesis rather than chosen from the fitted pattern.*

## 5. Declare the observation-grouping column

**Symptom.** Error: `Column site not found in data`.

**Diagnosis.** Every stacked-trait model needs a column identifying the
observational unit. `gllvmTMB()` defaults to `unit = "site"`, but the unit may
instead be an individual, species, or study. A grouping term in the formula
does not replace `unit =`: the argument tells the data-stacking and prediction
machinery which rows belong to the same observational unit.

```{r pitfall-5}
df <- sim$data
df$individual <- df$site
df$site <- NULL

# MISMATCHED (uncomment to see the error):
# gllvmTMB(value ~ 0 + trait + latent(0 + trait | individual, d = 2),
#          data = df, trait = "trait")

# MATCHED: name the observational unit explicitly
fit_ind <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | individual, d = 2),
  data = df, trait = "trait", unit = "individual"
)
fit_ind$fit_health$converged
```

**Rule of thumb.** *If the observational-unit column is not literally named
`site`, pass `unit = "<your_column>"`. The
[morphometrics article](morphometrics.html) shows the canonical individual ×
trait setup.*

## 6. Separate relatedness covariance from sampling-error covariance

**Symptom.** You have a known matrix but are unsure whether it belongs in a
relatedness term or a measurement-error term.

**Diagnosis.** First name what the matrix indexes and what variation it
represents. **A**, **Ainv**, and **V** are not interchangeable.

| Symbol | Meaning | Model route | Required alignment |
|:-------|:--------|:------------|:-------------------|
| **A** | Known relatedness covariance among individuals | `A = A` inside an `animal_*()` or `phylo_*()` term | Symmetric covariance aligned to individual labels |
| **Ainv** | Compatible sparse relatedness precision; it may include unobserved pedigree ancestors | `Ainv = Ainv` inside an `animal_*()` or `phylo_*()` term | Labelled precision aligned to all indexed pedigree nodes |
| **V** | Known sampling-error covariance among observation rows | `meta_V(V = V)` plus top-level `known_V = V` | $n_{obs} \times n_{obs}$ covariance aligned to data rows |

For independent sampling errors, first assemble the row-aligned covariance
matrix, for example `V <- diag(sampling_variance)`. Supply that matrix in the
formula as `meta_V(V = V)` and to `gllvmTMB()` as `known_V = V`. A pedigree
covariance instead belongs in `A =` (or its compatible precision in `Ainv =`)
inside an `animal_*()` term.

**Rule of thumb.** *Use **A** or **Ainv** for relatedness and **V** for
sampling-error covariance. Name what each matrix indexes before choosing a
keyword. See the [Vocabulary glossary](gllvm-vocabulary.html) for
plain-language definitions.*

## See also

* [Pre-fit response screening](pre-fit-response-screening.html) — check
  response support and duplicated Bernoulli patterns before fitting.
* [Fit diagnostics](fit-diagnostics.html) and
  [Convergence and start values](convergence-start-values.html) — decide
  whether a point fit and its Wald inference are usable.
* [Morphometrics](morphometrics.html) — an individual × trait worked example.
* [Formula keyword grid](api-keyword-grid.html) — the covariance syntax map.
* [Response families](response-families.html) — the family and link lookup.
* `?suggest_lambda_constraint`, `?extract_Sigma`, and
  `?simulate_site_trait` — function-level help.
