---
title: "Cross-lineage covariance in a plant--bumblebee example"
description: "Build one fixed cross-lineage kernel and read its plant-by-bumblebee trait covariance block, with an explicit point-estimate boundary."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Cross-lineage covariance in a plant--bumblebee example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4.2,
  fig.align = "center",
  cache = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE
)
```

This article shows the smallest cross-lineage phylogenetic workflow: flower
traits from one lineage and bumblebee traits from another share a supplied,
association-derived kernel. The output is a plant-trait by bumblebee-trait
covariance block, not an interaction model.

> **Synthetic teaching data.** Every species name, trait value, phylogeny, and
> interaction weight below was simulated for this article. The names make the
> workflow concrete; they are not observations from a real plant--bumblebee
> network. A suitable public empirical source is the CC0 data record of Liang
> (2020, [Dryad](https://doi.org/10.5061/dryad.sf7m0cg34)); substituting it
> requires deliberate species-name and trait checks before fitting.

> **Evidence boundary.** This is a fixed-`rho`, Gaussian point-estimate
> example. `rho` is supplied when the kernel is built, not estimated by the
> model. The covariance block has no calibrated interval here, and it is not
> evidence that traits causally coevolved. Sparse association data, a different
> defensible `rho`, or a different interaction matrix can change the result.

> **Scope boundary.** **In:** one named, dense `kernel_latent()` tier and its
> shared plant-by-bumblebee covariance block. **Partial:** the fixed-`rho`
> point estimate demonstrated here; it is sensitive to the supplied kernel and
> association matrix. **Not provided:** estimated `rho`, calibrated intervals,
> a causal coevolution test, or a separately interpreted phylogenetic-versus-tip
> decomposition.

## The model and its estimand

Let (A_F) and (A_B) be fixed plant and bumblebee phylogenetic correlation
matrices, and let (W) be the observed or assumed plant-by-bumblebee
association matrix. `make_cross_kernel()` builds a positive-semidefinite
stacked kernel

$$
K_{\mathrm{cross}} =
\begin{bmatrix}
A_F & C_{FB} \\
C_{FB}^{\mathsf T} & A_B
\end{bmatrix},
$$

where the off-diagonal block (C_{FB}) is a spectrally scaled function of
(A_F), (A_B), (W), and the fixed bridge strength (ho). A latent
Gaussian field (G \sim \operatorname{MVN}(0, K_{\mathrm{cross}})) then gives

$$
Y = M + G\Lambda^{\mathsf T} + E,
\qquad
\Gamma_{FB} = \Lambda_F\Lambda_B^{\mathsf T}.
$$

`Gamma` is the block whose rows are flower traits and whose columns are
bumblebee traits. It summarizes the fitted shared covariance shape conditional
on the supplied kernel. It does not identify a mechanism, establish the
direction of evolutionary change, or turn the interaction matrix into a
response variable.

::: {.table-responsive}
| Symbol in prose | Keyword or helper | Teaching-data draw | Extractor | Known teaching value |
|---|---|---|---|---|
| (K_{\mathrm{cross}}) | `make_cross_kernel()` | fixed from (A_F, A_B, W, \rho) | eigenvalue and name checks | `example$K_cross` |
| (G) | `kernel_latent(..., d = 2, name = "cross")` | (\operatorname{MVN}(0, K_{\mathrm{cross}})) | `extract_Sigma(..., part = "shared")` | `Lambda %*% t(Lambda)` |
| (Gamma_{FB}) | same named kernel tier | `Lambda_plant %*% t(Lambda_bumblebee)` | `extract_Gamma()` | `example$truth$Gamma` |
:::

## Load the synthetic plant--bumblebee fixture

```{r setup}
library(gllvmTMB)

example_path <- system.file(
  "extdata", "examples", "plant-bumblebee-coevolution-example.rds",
  package = "gllvmTMB"
)
example_path <- c(
  example_path,
  file.path("inst", "extdata", "examples", "plant-bumblebee-coevolution-example.rds"),
  file.path("..", "inst", "extdata", "examples", "plant-bumblebee-coevolution-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "plant-bumblebee-coevolution-example.rds")
)
example_path <- example_path[nzchar(example_path) & file.exists(example_path)][1]
stopifnot(!is.na(example_path))
example <- readRDS(example_path)

c(
  plants = example$truth$n_plant,
  bumblebees = example$truth$n_bumblebee,
  repeated_measurements = example$truth$n_replicate
)
head(example$data_wide)
```

Each plant row has two observed flower traits and missing bumblebee traits;
each bumblebee row has the complementary pattern. That block-missing layout is
intentional. The model learns covariance across the two lineages through the
supplied cross-kernel, not by pretending a plant has a tongue length.

## Build and check the fixed cross-lineage kernel

```{r kernel}
K_cross <- make_cross_kernel(
  example$A_plant,
  example$A_bumblebee,
  example$W,
  rho = example$truth$rho
)

kernel_check <- list(
  reconstructed_exactly = isTRUE(all.equal(K_cross, example$K_cross)),
  minimum_eigenvalue = min(eigen(K_cross, symmetric = TRUE, only.values = TRUE)$values),
  plant_names_match = identical(rownames(example$A_plant), rownames(example$W)),
  bumblebee_names_match = identical(colnames(example$A_bumblebee), colnames(example$W))
)
kernel_check
```

For real data, stop here if a species is absent from either tree, if names are
duplicated across the two lineages, or if the interaction matrix is ordered
differently from the trees. Prune and reorder deliberately, then rebuild the
kernel. Do not repair those problems by relying on factor order.

## Fit in wide form

The wide form has one observation per row and the four traits in columns. It
is compact when the two sets of traits are already stored separately.

```{r fit-wide}
fit <- gllvmTMB(
  traits(flower_tube_depth, nectar_volume, tongue_length, body_size) ~ 1 +
    kernel_latent(species, K = K_cross, d = 2, name = "cross"),
  data = example$data_wide,
  unit = "observation",
  cluster = "species",
  family = gaussian(),
  control = gllvmTMBcontrol(se = FALSE)
)

c(
  convergence = fit$opt$convergence,
  log_likelihood = as.numeric(logLik(fit))
)
```

The long form is equivalent in meaning and is often easier to construct from
field data. It uses the same `gllvmTMB()` entry point:

```{r fit-long, eval = FALSE}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    kernel_latent(species, K = K_cross, d = 2, name = "cross"),
  data = example$data_long,
  trait = "trait",
  unit = "observation",
  cluster = "species",
  family = gaussian(),
  control = gllvmTMBcontrol(se = FALSE)
)
```

Here the optimizer reached its normal convergence code. That is a numerical
health check, not validation of an evolutionary conclusion; inspect
`check_gllvmTMB(fit)` before interpreting a real fit.

## Read the plant-by-bumblebee covariance block

```{r gamma}
Gamma_shape <- extract_Gamma(
  fit,
  level = "cross",
  row_traits = example$truth$plant_traits,
  col_traits = example$truth$bumblebee_traits,
  scale = "shape"
)
round(Gamma_shape, 3)
```

The signs and relative sizes describe the fitted shared covariance shape for
this synthetic draw, conditional on the chosen kernel. They are not estimates
of a universal flower--pollinator matching rule. `scale = "effect"` would
multiply this shape by the **fixed** `rho`; it would still not estimate `rho`.

## What to do with empirical data next

For an empirical analysis, first make the biological unit explicit: normally
one row per measured organism or species-level replicate, with a documented
link from each row to a tree tip. Then retain the four checks this small
example makes visible: tree labels, trait availability, interaction-matrix
ordering, and the smallest eigenvalue of the assembled kernel.

You can use `profile_cross_rho()` to refit a short, pre-specified grid of
fixed `rho` values as a sensitivity analysis. That helper compares supplied
kernels; it does not estimate `rho`, produce a profile confidence interval, or
validate a bootstrap. A bootstrap or additional model comparison is therefore
a separate analysis plan, not a conclusion licensed by this article.
