---
title: "Cross-lineage covariance in a plant--bumblebee example"
description: "Build one fixed cross-lineage kernel and read its plant-by-bumblebee trait covariance block, with an explicit point-estimate boundary."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Cross-lineage covariance in a plant--bumblebee example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.0,
  fig.height = 4.8,
  fig.align = "center",
  cache = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE
)
```

This article shows the smallest cross-lineage phylogenetic workflow: flower
traits from one lineage and bumblebee traits from another share a supplied,
association-derived kernel. The output is a plant-trait by bumblebee-trait
covariance block, not an interaction model.

> **Synthetic teaching data.** Every species name, trait value, phylogeny, and
> interaction weight below was simulated for this article. The names make the
> workflow concrete; they are not observations from a real plant--bumblebee
> network. A suitable public empirical source is the CC0 data record of Liang
> (2020, [Dryad](https://doi.org/10.5061/dryad.sf7m0cg34)); substituting it
> requires deliberate species-name and trait checks before fitting.

> **Evidence boundary.** This is a fixed-`rho`, Gaussian point-estimate
> example. `rho` is supplied when the kernel is built, not estimated by the
> model. The covariance block has no calibrated interval here, and it is not
> evidence that traits causally coevolved. Sparse association data, a different
> defensible `rho`, or a different interaction matrix can change the result.

> **Scope boundary.** **In:** one named, dense `kernel_latent()` tier and its
> shared plant-by-bumblebee covariance block. **Partial:** the fixed-`rho`
> point estimate demonstrated here; it is sensitive to the supplied kernel and
> association matrix. **Not provided:** estimated `rho`, calibrated intervals,
> a causal coevolution test, or a separately interpreted phylogenetic-versus-tip
> decomposition.

> **Engine note.** Run this example with `engine = "tmb"` (the default). The
> current `engine = "julia"` bridge deliberately does not route
> `kernel_latent()` or other structured covariance terms. GLLVM.jl has a
> Gaussian dense-covariance building block, but the R bridge has not yet been
> validated to represent this kernel-only model or return its named covariance
> component.

## The model and its estimand

Let (A_F) and (A_B) be fixed plant and bumblebee phylogenetic correlation
matrices, and let (W) be the observed or assumed plant-by-bumblebee
association matrix. `make_cross_kernel()` builds a positive-semidefinite
stacked kernel

$$
K_{\mathrm{cross}} =
\begin{bmatrix}
A_F & C_{FB} \\
C_{FB}^{\mathsf T} & A_B
\end{bmatrix},
$$

where the off-diagonal block (C_{FB}) is a spectrally scaled function of
(A_F), (A_B), (W), and the fixed bridge strength (ho). A latent
Gaussian field (G \sim \operatorname{MVN}(0, K_{\mathrm{cross}})) then gives

$$
Y = M + G\Lambda^{\mathsf T} + E,
\qquad
\Gamma_{FB} = \Lambda_F\Lambda_B^{\mathsf T}.
$$

`Gamma` is the block whose rows are flower traits and whose columns are
bumblebee traits. It summarizes the fitted shared covariance shape conditional
on the supplied kernel. It does not identify a mechanism, establish the
direction of evolutionary change, or turn the interaction matrix into a
response variable.

::: {.table-responsive}
| Symbol in prose | Keyword or helper | Teaching-data draw | Extractor | Known teaching value |
|---|---|---|---|---|
| (K_{\mathrm{cross}}) | `make_cross_kernel()` | fixed from (A_F, A_B, W, \rho) | eigenvalue and name checks | `example$K_cross` |
| (G) | `kernel_latent(..., d = 2, name = "cross")` | (\operatorname{MVN}(0, K_{\mathrm{cross}})) | `extract_Sigma(..., part = "shared")` | `Lambda %*% t(Lambda)` |
| (Gamma_{FB}) | same named kernel tier | `Lambda_plant %*% t(Lambda_bumblebee)` | `extract_Gamma()` | `example$truth$Gamma` |
:::

## Load the synthetic plant--bumblebee fixture

```{r setup}
library(gllvmTMB)
library(ggplot2)

example_path <- system.file(
  "extdata", "examples", "plant-bumblebee-coevolution-example.rds",
  package = "gllvmTMB"
)
example_path <- c(
  example_path,
  file.path("inst", "extdata", "examples", "plant-bumblebee-coevolution-example.rds"),
  file.path("..", "inst", "extdata", "examples", "plant-bumblebee-coevolution-example.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "plant-bumblebee-coevolution-example.rds")
)
example_path <- example_path[nzchar(example_path) & file.exists(example_path)][1]
stopifnot(!is.na(example_path))
example <- readRDS(example_path)

theme_article <- function() {
  theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "#D9E2EC", linewidth = 0.28),
      plot.title = element_text(face = "bold", colour = "#102A43"),
      plot.subtitle = element_text(colour = "#486581"),
      plot.caption = element_text(colour = "#627D98", hjust = 0),
      axis.title = element_text(face = "bold", colour = "#243B53"),
      axis.text = element_text(colour = "#243B53"),
      legend.title = element_text(face = "bold", colour = "#243B53"),
      legend.position = "bottom"
    )
}

matrix_long <- function(x, row_name, col_name, value_name) {
  out <- as.data.frame(as.table(x), stringsAsFactors = FALSE)
  names(out) <- c(row_name, col_name, value_name)
  out
}

c(
  plants = example$truth$n_plant,
  bumblebees = example$truth$n_bumblebee,
  repeated_measurements = example$truth$n_replicate
)
head(example$data_wide)
```

Each plant row has two observed flower traits and missing bumblebee traits;
each bumblebee row has the complementary pattern. That block-missing layout is
intentional. The model learns covariance across the two lineages through the
supplied cross-kernel, not by pretending a plant has a tongue length.

## Example 1: start with an association map, then build one kernel

The association matrix is an input that says which plant--bumblebee links are
more prominent in this teaching system. It is not a response, a causal graph,
or a discovery made by the fitted model. In an empirical analysis it should be
constructed and justified before fitting, rather than tuned until a preferred
trait pattern appears.

```{r association-map, fig.cap="Synthetic association map used as the fixed input W. Darker tiles denote larger supplied association weights; the colours are not fitted probabilities.", fig.alt="A 12 by 8 heatmap of simulated association weights between fictional plant and bumblebee species. Strong values lie near a diagonal band, with weaker values away from it."}
association_long <- matrix_long(
  example$W,
  row_name = "plant_species",
  col_name = "bumblebee_species",
  value_name = "association_weight"
)

ggplot(association_long, aes(bumblebee_species, plant_species,
  fill = association_weight)) +
  geom_tile(colour = "white", linewidth = 0.35) +
  scale_fill_viridis_c(
    option = "C", end = 0.9,
    name = "Supplied\nassociation weight"
  ) +
  labs(
    title = "The association map is declared before model fitting",
    subtitle = "Fictional plant--bumblebee links in the synthetic teaching fixture",
    x = "Bumblebee species", y = "Plant species"
  ) +
  theme_article() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

## Build and check the fixed cross-lineage kernel

```{r kernel}
K_cross <- make_cross_kernel(
  example$A_plant,
  example$A_bumblebee,
  example$W,
  rho = example$truth$rho
)

kernel_check <- list(
  reconstructed_exactly = isTRUE(all.equal(K_cross, example$K_cross)),
  minimum_eigenvalue = min(eigen(K_cross, symmetric = TRUE, only.values = TRUE)$values),
  plant_names_match = identical(rownames(example$A_plant), rownames(example$W)),
  bumblebee_names_match = identical(colnames(example$A_bumblebee), colnames(example$W))
)
kernel_check
```

The association map is transformed into a cross-lineage bridge only after the
tree labels and matrix order agree. The bridge below is the plant-by-bumblebee
off-diagonal block of the resulting kernel; its sign and scale are determined
by the declared kernel construction and fixed `rho`.

```{r kernel-bridge, fig.cap="The plant-by-bumblebee bridge in the fixed cross-lineage kernel. This is an input covariance structure, not a fitted trait covariance block.", fig.alt="A 12 by 8 heatmap of positive kernel bridge values between fictional plant and bumblebee species, with a band of strongest values."}
n_plant <- example$truth$n_plant
kernel_bridge <- K_cross[
  seq_len(n_plant),
  n_plant + seq_len(example$truth$n_bumblebee),
  drop = FALSE
]
rownames(kernel_bridge) <- rownames(example$A_plant)
colnames(kernel_bridge) <- rownames(example$A_bumblebee)
kernel_bridge_long <- matrix_long(
  kernel_bridge,
  row_name = "plant_species",
  col_name = "bumblebee_species",
  value_name = "kernel_bridge"
)

ggplot(kernel_bridge_long, aes(bumblebee_species, plant_species,
  fill = kernel_bridge)) +
  geom_tile(colour = "white", linewidth = 0.35) +
  scale_fill_viridis_c(
    option = "B", end = 0.88,
    name = "Fixed kernel\nbridge"
  ) +
  labs(
    title = "The supplied association map becomes a fixed covariance bridge",
    subtitle = paste0("Kernel built with fixed rho = ", example$truth$rho),
    x = "Bumblebee species", y = "Plant species"
  ) +
  theme_article() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

For real data, stop here if a species is absent from either tree, if names are
duplicated across the two lineages, or if the interaction matrix is ordered
differently from the trees. Prune and reorder deliberately, then rebuild the
kernel. Do not repair those problems by relying on factor order.

## Example 2: fit the shared plant--bumblebee trait block

The wide form has one observation per row and the four traits in columns. It
is compact when the two sets of traits are already stored separately.

```{r fit-wide}
fit <- gllvmTMB(
  traits(flower_tube_depth, nectar_volume, tongue_length, body_size) ~ 1 +
    kernel_latent(species, K = K_cross, d = 2, name = "cross"),
  data = example$data_wide,
  unit = "observation",
  cluster = "species",
  family = gaussian(),
  control = gllvmTMBcontrol(se = FALSE)
)

c(
  convergence = fit$opt$convergence,
  log_likelihood = as.numeric(logLik(fit))
)
```

The long form is equivalent in meaning and is often easier to construct from
field data. It uses the same `gllvmTMB()` entry point:

```{r fit-long, eval = FALSE}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    kernel_latent(species, K = K_cross, d = 2, name = "cross"),
  data = example$data_long,
  trait = "trait",
  unit = "observation",
  cluster = "species",
  family = gaussian(),
  control = gllvmTMBcontrol(se = FALSE)
)
```

Here the optimizer reached its normal convergence code. That is a numerical
health check, not validation of an evolutionary conclusion; inspect
`check_gllvmTMB(fit)` before interpreting a real fit.

## Read the plant-by-bumblebee covariance block

```{r gamma}
Gamma_shape <- extract_Gamma(
  fit,
  level = "cross",
  row_traits = example$truth$plant_traits,
  col_traits = example$truth$bumblebee_traits,
  scale = "shape"
)
round(Gamma_shape, 3)
```

The signs and relative sizes describe the fitted shared covariance shape for
this synthetic draw, conditional on the chosen kernel. They are not estimates
of a universal flower--pollinator matching rule. `scale = "effect"` would
multiply this shape by the **fixed** `rho`; it would still not estimate `rho`.

```{r gamma-block, fig.cap="Fitted Gamma block for the synthetic example. Each tile is a point estimate of shared covariance shape between a flower trait and a bumblebee trait, conditional on the fixed kernel; no interval is shown because this workflow has no calibrated Gamma uncertainty.", fig.alt="A two by two diverging-colour heatmap of fitted shared covariance shape. Flower tube depth is positive with tongue length and negative with body size; nectar volume is positive with tongue length and negative with body size."}
gamma_long <- matrix_long(
  Gamma_shape,
  row_name = "flower_trait",
  col_name = "bumblebee_trait",
  value_name = "covariance_shape"
)
gamma_limit <- max(abs(gamma_long$covariance_shape))
gamma_long$label_colour <- ifelse(
  abs(gamma_long$covariance_shape) > 0.45 * gamma_limit,
  "white", "#102A43"
)
trait_labels <- c(
  flower_tube_depth = "Flower tube depth",
  nectar_volume = "Nectar volume",
  tongue_length = "Tongue length",
  body_size = "Body size"
)
gamma_long$flower_trait <- unname(trait_labels[gamma_long$flower_trait])
gamma_long$bumblebee_trait <- unname(trait_labels[gamma_long$bumblebee_trait])

ggplot(gamma_long, aes(bumblebee_trait, flower_trait,
  fill = covariance_shape)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = sprintf("%.2f", covariance_shape),
    colour = label_colour), fontface = "bold", size = 4.4) +
  scale_colour_identity() +
  scale_fill_gradient2(
    low = "#2166AC", mid = "#F7F7F7", high = "#B2182B",
    midpoint = 0, limits = c(-gamma_limit, gamma_limit),
    name = expression(Gamma["shape"])
  ) +
  labs(
    title = "Fitted shared covariance shape is descriptive",
    subtitle = "Synthetic point estimates conditional on the declared kernel",
    x = "Bumblebee trait", y = "Flower trait"
  ) +
  theme_article()
```

## Example 3: sensitivity to a small set of fixed bridge strengths

The bridge strength `rho` is a modelling choice in this workflow. A useful
check is to refit a short grid of values chosen before looking at the fitted
Gamma block. The curve below compares model fits at four fixed values; it is
not a confidence interval and does not estimate `rho` in the engine.

In this synthetic fixture, a zero bridge fits much worse, whereas the three
positive bridge strengths are almost tied. This is the useful conclusion from
the display: the data separate "no bridge" from "some positive bridge" here,
but do not choose a precise positive value. That is exactly why the plot is a
sensitivity display rather than fitted-rho inference.

```{r rho-sensitivity}
refit_cross <- function(K, rho) {
  gllvmTMB(
    traits(flower_tube_depth, nectar_volume, tongue_length, body_size) ~ 1 +
      kernel_latent(species, K = K, d = 2, name = "cross"),
    data = example$data_wide,
    unit = "observation",
    cluster = "species",
    family = gaussian(),
    control = gllvmTMBcontrol(se = FALSE)
  )
}

rho_profile <- profile_cross_rho(
  example$A_plant,
  example$A_bumblebee,
  example$W,
  rho = c(0, 0.25, example$truth$rho, 0.8),
  refit = refit_cross
)
rho_profile
```

```{r rho-sensitivity-plot, fig.cap="Fixed-rho sensitivity curve for the synthetic example. Relative log likelihood is shown only for four pre-specified, refitted kernels. The vertical rule marks the data-generating rho used for this teaching fixture; it is not an estimated parameter or confidence limit.", fig.alt="A line plot of relative log likelihood over four fixed rho values from zero to 0.8. A vertical dashed line marks the fixture's supplied rho of 0.55."}
rho_ok <- subset(rho_profile, status == "ok")

ggplot(rho_ok, aes(rho, relative_logLik)) +
  geom_hline(yintercept = 0, colour = "#9FB3C8", linewidth = 0.45) +
  geom_vline(
    xintercept = example$truth$rho, linetype = "dashed",
    colour = "#D64545", linewidth = 0.65
  ) +
  geom_line(colour = "#1261A0", linewidth = 0.8) +
  geom_point(colour = "#1261A0", fill = "white", shape = 21, size = 3,
    stroke = 0.9) +
  scale_x_continuous(breaks = rho_ok$rho, limits = c(0, 0.85)) +
  labs(
    title = "The data reject no bridge but do not identify one positive rho",
    subtitle = "The dashed rule marks the supplied teaching value; the positive plateau is nearly flat",
    x = expression("Fixed bridge strength " * rho),
    y = "Relative log likelihood (best fixed value = 0)",
    caption = "No confidence interval is implied: rho is rebuilt into the kernel for each refit."
  ) +
  theme_article()
```

## What to do with empirical data next

For an empirical analysis, first make the biological unit explicit: normally
one row per measured organism or species-level replicate, with a documented
link from each row to a tree tip. Then retain the four checks this small
example makes visible: tree labels, trait availability, interaction-matrix
ordering, and the smallest eigenvalue of the assembled kernel.

You can use `profile_cross_rho()` to refit a short, pre-specified grid of
fixed `rho` values as a sensitivity analysis. The separate
`profile_cross_rho_ci()` helper is screening-grade only: its fixed-grid
coverage has not been calibrated, so do not present its output as a validated
confidence interval. A bootstrap, a null calibration, or a broader model
comparison is therefore a separate analysis plan, not a conclusion licensed by
this article.
