---
title: "Screen binary traits before fitting"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Screen binary traits before fitting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

`screen_gllvmTMB()` is a pre-fit check for candidate responses. It answers a
small question before you fit a stacked-trait GLLVM:

> Are these traits, items, or indicators risky to put into this exact model?

It does not decide which responses to use, remove indicators, choose a latent
rank, prove identifiability, or guarantee convergence. It builds the same long response
shape that `gllvmTMB()` would use, then reports data and formula conditions that
are easy to miss before fitting.

The first screen is for binary and binomial responses. This is the place where
complete or quasi-complete separation is most visible: a constant, rare,
near-universal, duplicated, or nearly deterministic indicator can produce a
very large fitted loading later. The screen makes those risks visible before
the loading matrix is interpreted.

## Setup

```{r setup}
library(gllvmTMB)
```

The example uses a systematic-map style data set. Each row is a review. Each
binary column records whether a review has a topic, region, taxonomic group, or
outcome indicator. In this setting the word `trait` means "response column";
scientifically, it may be a content item or coding indicator.

```{r data}
n <- 40
reviews <- data.frame(
  review = factor(seq_len(n)),
  outcome_richness = rep(c(0, 1), length.out = n),
  outcome_diversity = rep(c(0, 1), length.out = n),
  outcome_no_change = 1 - rep(c(0, 1), length.out = n),
  region_north_america = c(rep(1, 37), rep(0, 3)),
  level_ecosystem = 0
)
```

This candidate set contains three common problems.

`outcome_diversity` duplicates `outcome_richness`. `outcome_no_change` is its
exact complement. `region_north_america` is near-universal, with only three
minority outcomes. `level_ecosystem` is constant.

## Wide-data screen

Most systematic-map data start wide: one row per review and one column per
indicator. Use the same `traits(...)` left-hand side that you would use in a
wide `gllvmTMB()` call.

```{r screen-wide}
scr <- suppressWarnings(screen_gllvmTMB(
  traits(
    outcome_richness,
    outcome_diversity,
    outcome_no_change,
    region_north_america,
    level_ecosystem
  ) ~ 1 + latent(1 | review, d = 2),
  data = reviews,
  unit = "review",
  family = binomial("probit")
))

scr
```

The trait table reports the denominators behind each warning. A 5% prevalence
with 20 observations is not the same evidence as a 5% prevalence with 100000
observations, so the screen always reports both prevalence and minority count.

```{r traits}
screen_table(scr, "traits")
```

The pair table catches exact duplicates, exact complements, and near-duplicate
binary indicators on paired review rows. These are pre-fit data conditions, not
fitted-model diagnostics.

```{r pairs}
screen_table(scr, "pairs")
```

Recommendations are written as actions to inspect, not automatic decisions.
For example, a near-universal indicator may be scientifically central. In that
case, report why it stays in the first latent block and run a sensitivity model.
If it is only a coding artefact, recode it, collapse it with a related item, or
keep it descriptive instead of making it define a latent axis.

```{r recommendations}
screen_table(scr, "recommendations")
```

## Long-data screen

The same screen works after the indicators have already been stacked long. The
long form is useful when your workflow stores one row per `(review, indicator)`
cell.

```{r long}
long <- stats::reshape(
  reviews,
  varying = list(names(reviews)[names(reviews) != "review"]),
  v.names = "present",
  timevar = "trait",
  times = names(reviews)[names(reviews) != "review"],
  idvar = "review",
  direction = "long"
)
long$trait <- factor(long$trait, levels = names(reviews)[names(reviews) != "review"])
row.names(long) <- NULL

scr_long <- suppressWarnings(screen_gllvmTMB(
  present ~ 1 + latent(1 | review, d = 2),
  data = long,
  unit = "review",
  trait = "trait",
  family = binomial("probit")
))

screen_table(scr_long, "traits")
```

The wide and long calls reach the same stacked-trait screen. Use whichever data
shape matches your workflow.

## What to do with warnings

`FAIL` means the candidate response cannot be interpreted as a valid binomial
trait in the screened data, or an exact duplicate/complement is present in the
candidate latent-response block. Fix the data or decide which indicator should
represent that information before the first latent GLLVM fit.

`WARN` means the trait or pair has weak support. Inspect the coding, missingness,
and scientific purpose. Do not lower the latent rank and assume the problem is
gone; a near-constant binary item can dominate a one-axis model too.

`INFO` means the screen saw an imbalanced condition but enough denominator
support that a hard warning would overstate the evidence. This distinction is
why the table reports minority counts.

After fitting, still run `check_gllvmTMB()`. Pre-fit screening and fitted-model
diagnostics answer different questions. `screen_gllvmTMB()` catches candidate
indicator risks before fitting; `check_gllvmTMB()` catches large fitted
loadings, weak axes, optimizer status, and uncertainty-health symptoms after
fitting.

## Evidence base

The screen uses evidence-informed defaults, not universal deletion rules.
Complete and quasi-complete separation in logistic/probit models is a
well-studied maximum-likelihood failure mode (Albert and Anderson 1984). Small
minority counts matter, but the old "10 events per variable" rule is a
simulation-based heuristic whose strictness depends on the modelling context
(Peduzzi et al. 1996; Vittinghoff and McCulloch 2007). Psychometric item
analysis similarly treats item support, redundancy, and interpretability as
diagnostic evidence rather than removal rules (Chalmers 2012).

Post-fit loading thresholds, such as a 0.30 loading-salience rule after
rotation, belong to the identification and interpretation stage. They are not
the same as pre-fit response screening.

## References

- Albert A, Anderson JA (1984). On the existence of maximum likelihood
  estimates in logistic regression models. *Biometrika* 71:1--10.
- Peduzzi P, Concato J, Kemper E, Holford TR, Feinstein AR (1996). A simulation
  study of the number of events per variable in logistic regression analysis.
  *Journal of Clinical Epidemiology* 49:1373--1379.
- Vittinghoff E, McCulloch CE (2007). Relaxing the rule of ten events per
  variable in logistic and Cox regression. *American Journal of Epidemiology*
  165:710--718.
- Chalmers RP (2012). mirt: A multidimensional item response theory package for
  the R environment. *Journal of Statistical Software* 48(6):1--29.
