---
title: "Screen binary traits before fitting"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Screen binary traits before fitting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

`screen_gllvmTMB()` is a pre-fit check for candidate responses. It answers a
small question before you fit a stacked-trait GLLVM:

> Are these traits, items, or indicators risky to put into this exact model?

It does not decide which responses to use, remove indicators, choose a latent
rank, prove identifiability, or guarantee convergence. It builds the same long response
shape that `gllvmTMB()` would use, then reports data and formula conditions that
are easy to miss before fitting.

Scope boundary: the current screen covers binary and binomial
candidate responses before fitting. It is advisory only. Non-binary modules,
optional `detectseparation` / `mirt` comparator checks, and high-dimensional
systematic-map benchmarks remain planned (see the validation-debt register).

The first screen is for binary and binomial responses. This is the place where
complete or quasi-complete separation is most visible: a constant, rare,
near-universal, duplicated, or nearly deterministic indicator can produce a
very large fitted loading later. The screen makes those risks visible before
the loading matrix is interpreted.

## Setup

```{r setup}
library(gllvmTMB)
```

The example uses a systematic-map style data set. Each row is a review. Each
binary column records whether a review has a topic, region, taxonomic group, or
outcome indicator. In this setting the word `trait` means "response column";
scientifically, it may be a content item or coding indicator.

For a binary joint species distribution model (JSDM), read `review` as the
site, plot, survey, or other sampling unit and read each indicator column as
one species' presence-absence response. The screen asks whether those response
columns are well supported for the latent block you are about to fit.

```{r data}
n <- 40
reviews <- data.frame(
  review = factor(seq_len(n)),
  outcome_richness = rep(c(0, 1), length.out = n),
  outcome_diversity = rep(c(0, 1), length.out = n),
  outcome_no_change = 1 - rep(c(0, 1), length.out = n),
  region_north_america = c(rep(1, 37), rep(0, 3)),
  level_ecosystem = 0
)
```

This candidate set contains three common problems.

`outcome_diversity` duplicates `outcome_richness`. `outcome_no_change` is its
exact complement. `region_north_america` is near-universal, with only three
minority outcomes. `level_ecosystem` is constant.

## Wide-data screen

Most systematic-map data start wide: one row per review and one column per
indicator. Use the same `traits(...)` left-hand side that you would use in a
wide `gllvmTMB()` call.

Here `latent(1 | review, d = 2)` requests two ordinary unit-tier latent variables
for the review or site units. The screen uses that requested rank when checking
whether the candidate response block is too small or too redundant; it is not
checking phylogenetic, spatial, or meta-analytic covariance tiers.

The example wraps the call in `suppressWarnings()` only to keep the rendered
article focused on the screen output. In an interactive analysis, read warnings
and then inspect `screen_table(scr, "recommendations")`.

```{r screen-wide}
scr <- suppressWarnings(screen_gllvmTMB(
  traits(
    outcome_richness,
    outcome_diversity,
    outcome_no_change,
    region_north_america,
    level_ecosystem
  ) ~ 1 + latent(1 | review, d = 2),
  data = reviews,
  unit = "review",
  family = binomial("probit")
))

scr
```

The trait table reports the denominators behind each warning. A 5% prevalence
with 20 observations is not the same evidence as a 5% prevalence with 100000
observations, so the screen always reports both prevalence and minority count.

```{r traits}
screen_table(scr, "traits")
```

The pair table catches exact duplicates, exact complements, and near-duplicate
binary indicators on paired review rows. These are pre-fit data conditions, not
fitted-model diagnostics.

```{r pairs}
screen_table(scr, "pairs")
```

Recommendations are written as actions to inspect, not automatic decisions.
For example, a near-universal indicator may be scientifically central. In that
case, report why it stays in the first latent block and run a sensitivity model.
If it is only a coding artefact, recode it, collapse it with a related item, or
keep it descriptive instead of making it define a latent variable.

```{r recommendations}
screen_table(scr, "recommendations")
```

## Long-data screen

The same screen works after the indicators have already been stacked long. The
long form is useful when your workflow stores one row per `(review, indicator)`
cell.

```{r long}
long <- stats::reshape(
  reviews,
  varying = list(names(reviews)[names(reviews) != "review"]),
  v.names = "present",
  timevar = "trait",
  times = names(reviews)[names(reviews) != "review"],
  idvar = "review",
  direction = "long"
)
long$trait <- factor(long$trait, levels = names(reviews)[names(reviews) != "review"])
row.names(long) <- NULL

scr_long <- suppressWarnings(screen_gllvmTMB(
  present ~ 1 + latent(1 | review, d = 2),
  data = long,
  unit = "review",
  trait = "trait",
  family = binomial("probit")
))

screen_table(scr_long, "traits")
```

The wide and long calls reach the same stacked-trait screen. Use whichever data
shape matches your workflow.

## What to do with warnings

`FAIL` means the candidate response cannot be interpreted as a valid binomial
trait in the screened data, or that two candidate responses carry exactly the
same or exactly opposite information in the rows being fitted. Do not treat this
as an automatic species-removal rule. First check coding, taxonomy, sampling
design, and the scientific role of the response; then either justify keeping
both, combine or recode them, or run a sensitivity fit with one representative
response.

`WARN` means the trait or pair has weak support. Inspect the coding, missingness,
and scientific purpose. Do not lower the latent rank and assume the problem is
gone; a near-constant binary item can dominate a one-axis model too.

`INFO` means the screen saw an imbalanced condition but enough denominator
support that a hard warning would overstate the evidence. This distinction is
why the table reports minority counts.

After fitting, still run `check_gllvmTMB()`. Pre-fit screening and fitted-model
diagnostics answer different questions. `screen_gllvmTMB()` catches candidate
indicator risks before fitting; `check_gllvmTMB()` catches large fitted
loadings, weak axes, optimizer status, and uncertainty-health symptoms after
fitting.

## Evidence base

The screen uses evidence-informed defaults, not universal deletion rules.
Complete and quasi-complete separation in logistic/probit models is a
well-studied maximum-likelihood failure mode (Albert and Anderson 1984). Small
minority counts matter, but the old "10 events per variable" rule is a
simulation-based heuristic whose strictness depends on the modelling context
(Peduzzi et al. 1996; Vittinghoff and McCulloch 2007). Separation is also tied
to sparse outcomes, rare exposures, highly correlated predictors, and strong
effects, and can produce infinite maximum-likelihood estimates if left
unchecked (Mansournia et al. 2018). The UCLA OARC separation FAQ is a useful
teaching reference for the practical complete/quasi-complete distinction.

In JSDM work, rare species are not automatically discarded. They are often one
reason to use a joint model, because hierarchical, latent-variable, trait, or
phylogenetic structure can let weakly observed species borrow information from
the rest of the community (Ovaskainen and Soininen 2011; Warton et al. 2015).
At the same time, very sparse or constant responses can make species-level
parameters hard to estimate, slow computation, worsen convergence, or make a
binomial likelihood uninformative (Norberg et al. 2019; Zhang et al. 2020).
Some applied HMSC/JSDM analyses therefore set documented prevalence thresholds,
but those thresholds are analysis choices, not general rules (Cranston et al.
2024). `screen_gllvmTMB()` follows that convention: it flags
all-zero, all-one, near-constant, and very rare candidate species for
inspection, reporting, and possible sensitivity analysis. It does not decide
that a biologically important rare species should be removed.

Psychometric item analysis similarly treats item support, redundancy, response
frequencies, and interpretability as diagnostic evidence rather than removal
rules. The `mirt::itemstats()` interface is a useful precedent because it
computes item summary statistics without first fitting an IRT model (Chalmers
2012). The SAS IRT overview is another applied reference for thinking of
binary responses as items that measure latent traits. Predictor-screening tools
such as `recipes::step_nzv()` and `recipes::step_corr()` are another precedent,
but `screen_gllvmTMB()` is more conservative: it reports candidate response
risks for this formula instead of silently deleting columns. This caution
matters because item-deletion rules can reduce criterion validity even when a
reliability statistic improves (Raykov 2008).

Post-fit loading thresholds, such as a 0.30 loading-salience rule after
rotation, belong to the identification and interpretation stage. They are not
the same as pre-fit response screening.

## References

- Albert A, Anderson JA (1984). On the existence of maximum likelihood
  estimates in logistic regression models. *Biometrika* 71:1--10.
- Peduzzi P, Concato J, Kemper E, Holford TR, Feinstein AR (1996). A simulation
  study of the number of events per variable in logistic regression analysis.
  *Journal of Clinical Epidemiology* 49:1373--1379.
- Vittinghoff E, McCulloch CE (2007). Relaxing the rule of ten events per
  variable in logistic and Cox regression. *American Journal of Epidemiology*
  165:710--718.
- Mansournia MA, Geroldinger A, Greenland S, Heinze G (2018). Separation in
  logistic regression: causes, consequences, and control. *American Journal of
  Epidemiology* 187:864--870.
- UCLA OARC. Complete or quasi-complete separation in logistic/probit
  regression. <https://stats.oarc.ucla.edu/other/mult-pkg/faq/general/faqwhat-is-complete-or-quasi-complete-separation-in-logisticprobit-regression-and-how-do-we-deal-with-them/>
- Chalmers RP (2012). mirt: A multidimensional item response theory package for
  the R environment. *Journal of Statistical Software* 48(6):1--29.
- Chalmers RP. `mirt::itemstats()` documentation: generic item summary
  statistics without prior IRT fitting. <https://rdrr.io/cran/mirt/man/itemstats.html>
- Cranston A, Cooper N, Bro-Jorgensen J (2024). Using joint species distribution
  modelling to identify climatic and non-climatic drivers of Afrotropical
  ungulate distributions. *Ecography* 2024:e07209. doi:10.1111/ecog.07209.
- Norberg A, Abrego N, Blanchet FG, Adler FR, Anderson BJ, Anttila J, Araujo MB,
  Dallas T, Dunson D, Elith J, Foster SD, Fox R, Franklin J, Godsoe W, Guisan A,
  O'Hara B, Hill NA, Holt RD, Hui FKC, Husby M, Kalas JA, Lehikoinen A,
  Luoto M, Mod HK, Newell G, Renner I, Roslin T, Soininen J, Thuiller W,
  Vanhatalo J, Warton D, White M, Zimmermann NE, Gravel D, Ovaskainen O (2019).
  A comprehensive evaluation of predictive performance of 33 species distribution
  models at species and community levels. *Ecological Monographs* 89:e01370.
  doi:10.1002/ecm.1370.
- Ovaskainen O, Soininen J (2011). Making more out of sparse data: hierarchical
  modeling of species communities. *Ecology* 92:289--295. doi:10.1890/10-1251.1.
- An X, Yung Y-F (2014). Item Response Theory: what it is and how you can use
  the IRT procedure to apply it. SAS Global Forum paper SAS364-2014.
  <https://support.sas.com/resources/papers/proceedings14/SAS364-2014.pdf>
- Kosmidis I, Schumacher D. `detectseparation`: detect and check for separation
  and infinite maximum-likelihood estimates in binomial-response GLMs.
  <https://cran.r-project.org/package=detectseparation>
- Kuhn M, Wickham H. `recipes::step_nzv()` and `recipes::step_corr()`
  documentation. <https://recipes.tidymodels.org/reference/step_nzv.html> and
  <https://recipes.tidymodels.org/reference/step_corr.html>
- Raykov T (2008). Alpha if item deleted: a note on loss of criterion validity
  in scale development if maximizing coefficient alpha. *British Journal of
  Mathematical and Statistical Psychology* 61:275--285.
- Warton DI, Blanchet FG, O'Hara RB, Ovaskainen O, Taskinen S, Walker SC,
  Hui FKC (2015). So many variables: joint modeling in community ecology.
  *Trends in Ecology & Evolution* 30:766--779. doi:10.1016/j.tree.2015.09.007.
- Zhang C, Chen Y, Xu B, Xue Y, Ren Y (2020). Improving prediction of rare
  species' distribution from community data. *Scientific Reports* 10:12230.
  doi:10.1038/s41598-020-69157-x.
