---
title: "Screen binary traits before fitting"
description: "Check prevalence, sparse outcomes, duplicated cells, and usable denominators before fitting a binary multivariate model."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Screen binary traits before fitting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = TRUE
)
```

`screen_gllvmTMB()` is a pre-fit check for candidate responses. It answers a
small question before you fit a stacked-trait GLLVM:

> Are these traits, items, or indicators risky to put into this exact model?

It does not decide which responses to use, remove indicators, choose a latent
rank, prove identifiability, or guarantee convergence. It uses the same
wide-to-long stacking route as `gllvmTMB()`, then reports data and formula
conditions that are easy to miss before fitting. It is not a complete dry run
of every parser or engine check.

The current screen covers one `binomial()` family using the logit, probit, or
cloglog link. Trait-level support works for Bernoulli and multi-trial binomial
responses; pairwise duplicate/complement screening is Bernoulli-only. Other
families return `NOT_CHECKED` rather than pretending they were screened.

The screen has three separate jobs:

1. **Marginal response support:** invalid values, unusable denominators,
   constants, rare outcomes, and strong imbalance.
2. **Response-pair redundancy:** exact or near duplicates, complements, and
   unusually strong pairwise association on jointly observed units.
3. **Formula shape:** fixed-effect rank, grouping-level support, and whether the
   requested latent dimension is smaller than the raw number of traits.

Formal complete or quasi-complete separation depends on the response–predictor
design. Rare or constant outcomes can motivate a formal separation check, but
duplicate or complementary response columns are redundancy conditions, not
classical separation. A clean `screen_gllvmTMB()` result therefore does not
prove that separation cannot occur.

## Setup

```{r setup}
library(gllvmTMB)
```

The example uses a systematic-map style data set. Each row is a review. Each
binary column records whether a review has a topic, region, taxonomic group, or
outcome indicator. In this setting the word `trait` means "response column";
scientifically, it may be a content item or coding indicator.

For a binary joint species distribution model (JSDM), read `review` as the
site, plot, survey, or other sampling unit and read each indicator column as
one species' presence-absence response. The screen asks whether those response
columns are well supported for the latent block you are about to fit.

```{r data}
n <- 40
reviews <- data.frame(
  review = factor(seq_len(n)),
  outcome_richness = rep(c(0, 1), length.out = n),
  outcome_diversity = rep(c(0, 1), length.out = n),
  outcome_no_change = 1 - rep(c(0, 1), length.out = n),
  region_north_america = c(rep(1, 37), rep(0, 3)),
  level_ecosystem = 0
)
```

This candidate set contains three common problems.

`outcome_diversity` duplicates `outcome_richness`. `outcome_no_change` is its
exact complement. `region_north_america` is near-universal, with only three
minority outcomes. `level_ecosystem` is constant.

## Wide-data screen

Most systematic-map data start wide: one row per review and one column per
indicator. Use the same `traits(...)` left-hand side that you would use in a
wide `gllvmTMB()` call.

Here `latent(1 | review, d = 2)` requests two ordinary unit-tier latent variables
for the review or site units. The screen warns when `d` is not smaller than the
raw number of traits. It reports pair redundancy separately and does not
estimate an effective rank after collapsing duplicate responses. This page
does not screen phylogenetic, spatial, or meta-analytic covariance tiers.

The thresholds are advisory and configurable. This example stores the defaults
explicitly so the active rules remain visible.

```{r thresholds}
ctl <- screen_control()
data.frame(
  rule = c(
    "minority count: stronger warning",
    "minority count: warning",
    "prevalence: stronger information flag",
    "prevalence: information flag",
    "pair discordance: stronger warning",
    "pair discordance: warning",
    "pair Hamming rate: warning",
    "absolute phi/Jaccard: stronger warning",
    "absolute phi/Jaccard: warning"
  ),
  default = c(
    paste0("< ", ctl$rare_strong_n),
    paste0("< ", ctl$rare_warn_n),
    paste(ctl$prevalence_strong, collapse = " to "),
    paste(ctl$prevalence_warn, collapse = " to "),
    paste0("< ", ctl$discordant_strong_n),
    paste0("< ", ctl$discordant_warn_n),
    paste0("<= ", ctl$hamming_rate_warn),
    paste0(">= ", ctl$phi_strong),
    paste0(">= ", ctl$phi_warn)
  )
)
```

Prevalence outside the displayed interval is flagged; it is not a deletion
cutoff. Minority counts take priority, which prevents a large data set with a
low prevalence from receiving the same warning as a tiny data set with only a
few minority outcomes.

```{r screen-wide}
scr <- screen_gllvmTMB(
  traits(
    outcome_richness,
    outcome_diversity,
    outcome_no_change,
    region_north_america,
    level_ecosystem
  ) ~ 1 + latent(1 | review, d = 2),
  data = reviews,
  unit = "review",
  family = binomial("probit"),
  control = ctl
)

scr
```

The summary counts checks, not traits: `FAIL 4` combines one trait-level and
three pair-level failures, `WARN 1` is one trait warning, and `PASS 13` includes
trait, pair, and formula-shape checks. Statuses are scope-specific. A pair-level
`PASS` cannot cancel a trait-level `FAIL`, and a trait-level `PASS` cannot cancel
an exact redundant-pair finding.

- `PASS` means no condition was flagged at the current thresholds; it is not a
  guarantee of a safe fitted model.
- `WARN` and `INFO` are advisory. `severity` grades the evidence within that
  status, so `WARN` with `severity = "strong"` is not contradictory.
- `FAIL` means invalid binomial input, no usable denominator, a constant
  response that is uninformative for this latent block, or exact pair
  redundancy on the observed overlap.
- `NOT_CHECKED` means the requested screen is unavailable for that row or
  family.

The trait table reports the denominators behind each warning. A 5% prevalence
with 20 Bernoulli trials is not the same evidence as a 5% prevalence with
100000 trials. `n_valid` counts usable rows, while `total_trials` includes the
multi-trial denominator.

```{r traits}
trait_screen <- screen_table(scr, "traits")
trait_screen[, c(
  "trait", "status", "severity", "n_valid", "total_trials",
  "prevalence", "minority_count", "action"
)]
```

The pair table catches exact duplicates, exact complements, and near-duplicate
Bernoulli indicators on jointly observed review rows. `n_pair` is the overlap
denominator; `discordant_n` counts different outcomes, and `hamming_rate` is
that count divided by `n_pair`. “Exact” therefore means exact among the paired
rows shown, not necessarily across rows where one response is missing.

```{r pairs}
pairs <- screen_table(scr, "pairs")
pairs[, c(
  "trait_i", "trait_j", "status", "severity",
  "n_pair", "discordant_n", "hamming_rate", "action"
)]
```

Recommendations are written as actions to inspect, not automatic decisions.
For example, a near-universal indicator may be scientifically central. In that
case, report why it stays in the first latent block and run a sensitivity model.
If it is only a coding artefact, recode it, collapse it with a related item, or
keep it descriptive instead of making it define a latent variable.

```{r recommendations}
recommendations <- screen_table(scr, "recommendations")
recommendations[, c("scope", "status", "action", "trait", "evidence")]
```

The formula and grouping checks are separate tables. Here all 40 units have
both outcome sides across the screened block, and the requested rank is smaller
than the raw trait count.

```{r design-units}
screen_table(scr, "design")[, c(
  "component", "status", "value", "threshold", "message"
)]

as.data.frame(table(
  status = screen_table(scr, "units")$status
))
```

## Long-data screen

The same screen works after the indicators have already been stacked long. The
long form is useful when your workflow stores one row per `(review, indicator)`
cell.

```{r long}
long <- stats::reshape(
  reviews,
  varying = list(names(reviews)[names(reviews) != "review"]),
  v.names = "present",
  timevar = "trait",
  times = names(reviews)[names(reviews) != "review"],
  idvar = "review",
  direction = "long"
)
long$trait <- factor(long$trait, levels = names(reviews)[names(reviews) != "review"])
row.names(long) <- NULL

scr_long <- screen_gllvmTMB(
  present ~ 0 + trait + latent(0 + trait | review, d = 2),
  data = long,
  unit = "review",
  trait = "trait",
  family = binomial("probit"),
  control = ctl
)

long_traits <- screen_table(scr_long, "traits")
long_traits[, c(
  "trait", "status", "n_valid", "prevalence", "minority_count"
)]

identical(
  trait_screen[, c("trait", "status", "n_success", "n_failure")],
  long_traits[, c("trait", "status", "n_success", "n_failure")]
)
```

The printed comparison verifies that the wide and canonical long calls produce
the same trait-support summaries in this example. It does not claim that every
future table is identical under every formula. Use whichever data shape matches
your workflow.

## Apply the screen before fitting

The constant response cannot inform this latent block, and the duplicate and
complementary outcomes should not enter the same block unchanged. The next
candidate keeps one representative response, removes the constant, and retains
the near-universal response because we decide it is scientifically important.
That retained warning must be documented and revisited in a sensitivity fit.

```{r corrected-screen}
candidate <- reviews[c(
  "review",
  "outcome_richness",
  "region_north_america"
)]

scr_candidate <- screen_gllvmTMB(
  traits(outcome_richness, region_north_america) ~
    1 + latent(1 | review, d = 1),
  data = candidate,
  unit = "review",
  family = binomial("probit"),
  control = ctl
)

screen_table(scr_candidate, "recommendations")[, c(
  "scope", "status", "action", "trait", "evidence"
)]
```

This rerun contains only the deliberately retained rarity warning. If the rare
response is not scientifically essential, keep it descriptive or fit a
sensitivity model without it. The screen supports that decision; it does not
make the decision.

## What to do with warnings

Read `FAIL` according to its scope. Invalid values or denominators require data
repair. A constant response is valid binomial data but has no minority support
for the proposed latent block; exclude it from that block or collect informative
variation. Exact duplicate or complementary pairs should be selected, combined,
or recoded before fitting the same latent block unchanged. Always check the
reported `n_pair` first, especially when responses have different missingness.

`WARN` means limited marginal support or unusually strong pairwise redundancy
or association. Inspect the coding, missingness, overlap denominator, and
scientific purpose. Do not lower the latent rank and assume the problem is gone;
a near-constant binary item can dominate a one-axis model too.

`INFO` means the screen saw an imbalanced condition but enough denominator
support that a hard warning would overstate the evidence. This distinction is
why the table reports minority counts.

After fitting, still run `check_gllvmTMB()`. Pre-fit screening and fitted-model
diagnostics answer different questions. `screen_gllvmTMB()` catches candidate
indicator risks before fitting; `check_gllvmTMB()` catches large fitted
loadings, weak axes, optimizer status, and uncertainty-health symptoms after
fitting. The full sequence is shown in
[Can I trust this fit?](fit-diagnostics.html).

## Evidence base

The defaults are evidence-informed heuristics, not universal deletion rules.
Complete and quasi-complete separation is a well-studied maximum-likelihood
failure mode (Albert and Anderson 1984; Mansournia et al. 2018), but its presence
depends on the response–predictor design. Minority-count rules such as “10
events per variable” are context-dependent heuristics rather than hard laws
(Peduzzi et al. 1996; Vittinghoff and McCulloch 2007). This screen reports
conditions that deserve inspection; it does not replace a formal separation
analysis.

Rare species are not automatically discarded in JSDMs. Shared community
structure can make sparse responses scientifically valuable, while very sparse
or constant responses can still make species-specific parameters hard to
estimate (Ovaskainen and Soininen 2011; Warton et al. 2015; Norberg et al.
2019). Documented prevalence thresholds are analysis choices, not general
rules. Retain an important rare species only with a stated reason and a
sensitivity analysis.

Psychometric item analysis uses the same principle: item support, redundancy,
response frequencies, and interpretability are evidence for review rather than
automatic deletion. `mirt::itemstats()` is a useful precedent for inspecting
items before fitting an IRT model (Chalmers 2012). `screen_gllvmTMB()` likewise
reports response risks without silently changing the formula.

Post-fit loading salience, rotation, and latent-rank interpretation belong after
fitting. They are not substitutes for this pre-fit response screen.

## Next decision

Fit only after every `WARN` row has a documented keep, recode, combine, or omit
decision. Then run [Can I trust this fit?](fit-diagnostics.html); a clean
pre-fit screen does not make the fitted optimiser, Hessian, or weak-axis checks
pass automatically. If a scientifically essential response remains constant or
has too little denominator support, report that limitation and defer its
species-specific effect rather than forcing it into the model.

## References

- Albert A, Anderson JA (1984). On the existence of maximum likelihood
  estimates in logistic regression models. *Biometrika* 71:1--10.
- Peduzzi P, Concato J, Kemper E, Holford TR, Feinstein AR (1996). A simulation
  study of the number of events per variable in logistic regression analysis.
  *Journal of Clinical Epidemiology* 49:1373--1379.
- Vittinghoff E, McCulloch CE (2007). Relaxing the rule of ten events per
  variable in logistic and Cox regression. *American Journal of Epidemiology*
  165:710--718.
- Mansournia MA, Geroldinger A, Greenland S, Heinze G (2018). Separation in
  logistic regression: causes, consequences, and control. *American Journal of
  Epidemiology* 187:864--870.
- Chalmers RP (2012). mirt: A multidimensional item response theory package for
  the R environment. *Journal of Statistical Software* 48(6):1--29.
- Chalmers RP. `mirt::itemstats()` documentation: generic item summary
  statistics without prior IRT fitting. <https://rdrr.io/cran/mirt/man/itemstats.html>
- Norberg A, Abrego N, Blanchet FG, Adler FR, Anderson BJ, Anttila J, Araujo MB,
  Dallas T, Dunson D, Elith J, Foster SD, Fox R, Franklin J, Godsoe W, Guisan A,
  O'Hara B, Hill NA, Holt RD, Hui FKC, Husby M, Kalas JA, Lehikoinen A,
  Luoto M, Mod HK, Newell G, Renner I, Roslin T, Soininen J, Thuiller W,
  Vanhatalo J, Warton D, White M, Zimmermann NE, Gravel D, Ovaskainen O (2019).
  A comprehensive evaluation of predictive performance of 33 species distribution
  models at species and community levels. *Ecological Monographs* 89:e01370.
  <https://doi.org/10.1002/ecm.1370>.
- Ovaskainen O, Soininen J (2011). Making more out of sparse data: hierarchical
  modeling of species communities. *Ecology* 92:289--295.
  <https://doi.org/10.1890/10-1251.1>.
- Warton DI, Blanchet FG, O'Hara RB, Ovaskainen O, Taskinen S, Walker SC,
  Hui FKC (2015). So many variables: joint modeling in community ecology.
  *Trends in Ecology & Evolution* 30:766--779.
  <https://doi.org/10.1016/j.tree.2015.09.007>.
