---
title: "Profile likelihood: routes and fallbacks"
description: "Choose and inspect target-specific profile, Wald, or bootstrap routes while keeping returned bounds separate from coverage claims."
tier: 2 # Target-specific lookup page for readers who already have a fitted model.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Profile likelihood: routes and fallbacks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = TRUE
)
```

This reference answers one practical question: when you request a profile
interval, does `gllvmTMB` actually compute one for your target?

Direct parameters and simple linear contrasts can use likelihood profiles.
Nonlinear derived quantities do not currently have profile intervals because
the constrained fits cannot yet be checked reliably in every case. Empirical
coverage has also not been established across all supported models, so this
page does not call any interval method universally best or publication-ready.

The key rule is simple: **choose a route that is actually available for the
target, then inspect the returned bounds and any method or status columns.** A
documented numerical-Wald fallback must retain its label, while a withdrawn
nonlinear route must stop rather than being presented as a profile result.

```{r setup}
library(gllvmTMB)
set.seed(42)
```

## Current route map

The interval API is target-specific. The table below shows what a reader can
request now; it is not a claim about empirical coverage.

| Target | Start here | Current behavior |
|:-------|:-----------|:-----------------|
| Direct fixed effect, dispersion, or variance component | `profile_targets(fit)` followed by `confint(fit, parm = ..., method = "profile")` | Direct likelihood profile when the target is marked ready. |
| Repeatability | `extract_repeatability()` | Wald is the default; requesting profile stops with a typed withdrawal because no full-covariance profile route is available. |
| Trait correlation | `extract_correlations()` | Point-only is the default. Fisher-z and bootstrap bounds are opt-in; a profile interval is not currently available. |
| Communality | `extract_communality()` | Point estimates are the default; Wald and bootstrap bounds are opt-in; a profile interval is not currently available. |
| Phylogenetic signal | `extract_phylo_signal()` | A simple two-component decomposition can use a profile contrast. Richer decompositions return labelled numerical-Wald bounds; a general profile constraint is not implemented. |
| Predictor-informed latent effect $B_{lv}$ | `extract_lv_effects()` | Point estimates and experimental delta-method Wald summaries are available; profile and bootstrap intervals are not currently available. |
| Full covariance matrix | `confint(fit, parm = "Sigma_unit", ...)` | Route depends on the fitted structure. Read the returned method and any fallback message rather than assuming every matrix element was directly profiled. |

Bootstrap is not an automatic remedy for an unusable profile. It relies on
simulation and successful refits, and every active random-effect tier must be
redrawn rather than silently conditioned on its fitted mode. Use enough
replicates to estimate the requested tail probabilities, report failed refits,
and treat small runs only as software smoke tests.

## Fit one model from long or wide data

The same Gaussian model can be fitted through the canonical long interface or
the compact wide data-frame interface. The rest of the article uses the long
fit. Here `unit = "site"` names the between-site tier and
`unit_obs = "site_species"` names the within-site sampling cell used by the
second covariance term. No third grouping role is present, so this model does
not pass `cluster` or `cluster2`.

```{r sim}
s <- simulate_site_trait(
  n_sites               = 50,
  n_species             = 6,
  n_traits              = 3,
  mean_species_per_site = 4,
  Lambda_B              = matrix(c(0.9, 0.4, -0.3), 3, 1),
  psi_B                 = c(0.40, 0.30, 0.50),
  psi_W                 = c(0.30, 0.40, 0.30),
  beta                  = matrix(0, 3, 2),
  seed                  = 42
)
df_long <- s$data

library(tidyr)
df_wide <- tidyr::pivot_wider(
  df_long,
  id_cols = c(site, site_species),
  names_from = trait,
  values_from = value
)
```

```{r fit}
# Long format (canonical)
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | site, d = 1, unique = TRUE) +
    indep(0 + trait | site_species),
  data = df_long,
  trait = "trait",
  unit = "site",
  unit_obs = "site_species"
)

# Wide data-frame format
fit_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~ 1 +
    latent(1 | site, d = 1, unique = TRUE) +
    indep(1 | site_species),
  data = df_wide,
  unit = "site",
  unit_obs = "site_species"
)

data.frame(
  fit = c("long", "wide"),
  logLik = c(as.numeric(logLik(fit_long)), as.numeric(logLik(fit_wide))),
  converged = c(fit_long$fit_health$converged,
                fit_wide$fit_health$converged)
)

fit <- fit_long
```

Do not interpret an interval until the point fit is stable. If
`fit$fit_health$converged` is false, repair the point surface first. A failed
Hessian blocks or weakens Wald inference, but it does not turn an unstable fit
into a valid profile problem. See [Fit diagnostics](fit-diagnostics.html) and
[Convergence and start values](convergence-start-values.html).

## Direct targets: inventory first, then profile

`profile_targets()` lists the fitted quantities that can be sent directly to
`confint(..., method = "profile")`. This public inventory avoids making readers
guess internal parameter names.

```{r targets}
targets <- profile_targets(fit, ready_only = TRUE)
direct_summary <- subset(
  targets,
  target_class %in% c("fixed_effect", "variance")
)
head(direct_summary[c("parm", "target_class", "estimate", "profile_note")], 12)
```

The first between-unit diagonal standard deviation is near its natural
boundary in this fixture. Request its profile by the public `parm` label:

```{r direct-profile}
ci_sd <- confint(
  fit,
  parm = "sd_B[1]",
  method = "profile",
  level = 0.95
)
ci_sd
```

The lower endpoint is `0`, not `NA`: the profile reached the natural variance
boundary. Current profile helpers use these conventions:

* a finite number is an estimated threshold crossing;
* a natural boundary such as `0`, `1`, or `-1` is returned when that is the
  corresponding endpoint;
* an infinite endpoint means the threshold was not crossed before the
  parameter limit;
* `NA` means the profile failed or did not provide enough usable evaluations.

The direct wrapper uses the standard one-degree-of-freedom likelihood-ratio
cutoff. Variance-component boundaries are non-regular, and this implementation
has not been broadly calibrated for every boundary regime.

### Why this page does not offer a t-based profile cutoff

The released direct-profile routes use the standard one-degree-of-freedom
likelihood-ratio reference. A t-based cutoff would require defensible,
target-specific degrees of freedom and repeated-sampling evidence. The number
of units minus the latent rank is not a generally justified substitute for a
Satterthwaite or Kenward--Roger derivation. `gllvmTMB` therefore does not teach
or expose an automatic t-based profile correction in this release.

This is an evidence boundary, not a claim that t-based inference is unimportant.
A future route needs to define the target-specific degrees of freedom, show how
they are estimated, and demonstrate coverage across the intended sample-size
regimes before it becomes reader-facing.

## Derived targets: read the method that came back

### Repeatability uses Wald; profile is withdrawn

For the full covariance definition of repeatability, the default is a
delta-method Wald interval. `method = "profile"` does not perform a constrained
likelihood profile and deliberately stops with
`gllvmTMB_repeatability_profile_withdrawn`; the former profile token targeted
only a diagonal-companion ratio and omitted shared latent variance. Use the
default Wald route or the parametric-bootstrap route, and report its returned
method and limitations.

```{r repeatability}
rep_wald <- extract_repeatability(fit, level = 0.95)
rep_wald
```

The bounds are formed on a transformed scale and back-transformed to
$[0,1]$; they are not symmetric on the natural repeatability scale. They are
Wald bounds, not profile intervals, and this article does not establish their
repeated-sampling calibration.

### Correlation intervals are opt-in

```{r correlations}
cors <- extract_correlations(
  fit,
  tier = "unit",
  level = 0.95,
  method = "fisher-z"
)
cors
```

Fisher-z keeps correlation bounds inside $[-1,1]$, but its effective-sample
size approximation can still be optimistic for complicated latent or
non-Gaussian models. A profile interval is not currently available because the
required constrained refits cannot yet be checked reliably. Here
`interval_status = "heuristic_unvalidated"` makes the Fisher-z limitation explicit:
the bounds use the requested nominal level, but repeated-sampling coverage has
not been established for this model class.

### Phylogenetic signal is decomposition-specific

For a simple two-component decomposition, the package can profile a contrast
of the two log-standard deviations. For richer phylogenetic plus
non-phylogenetic decompositions, `extract_phylo_signal()` currently uses
labelled numerical-Wald bounds. A general profile constraint and a working
general bootstrap interval are not available for those richer decompositions.
Do not relabel the fallback.

## What profile likelihood means

For a scalar target $\theta$ and nuisance parameters $\eta$, the profile
log-likelihood is

$$
\ell_p(\theta_0) = \max_{\eta}\ell(\theta_0,\eta).
$$

The implemented direct route finds the values where the likelihood-ratio
deviance crosses the standard $\chi^2_1$ threshold. Derived quantities such as
correlations or variance proportions require a different constrained-refit
problem because they are nonlinear functions of several fitted parameters.
The package withholds those nonlinear penalty-profile routes until it can enforce
the constraint tightly, verify every constrained optimizer result, and expose
failure diagnostics. That distinction explains why one target may profile
directly while another currently offers only point, Wald, or bootstrap routes.

## Troubleshoot the interval that came back

Before diagnosing a profile, confirm that a profile actually ran. Read the
returned `method` and status first. For withdrawn nonlinear targets such as
repeatability and cross-family correlations, a profile request stops with a
typed error instead of silently substituting a Wald result.

| Observed result | First check | Interpretation and safe next action |
|:----------------|:------------|:------------------------------------|
| Endpoint equals a natural limit such as `0`, `1`, or `-1` | Confirm the target transformation and returned method. | This can be a valid one-sided result. Report the boundary rather than replacing it with `NA`. |
| `NA` endpoint | Check fit health, route messages, and whether finite profile evaluations exist on that side. | The endpoint is unavailable or the numerical profile failed. Do not reinterpret it as a natural boundary. |
| `Inf` or `-Inf` | Inspect the parameter scale, search range, and profile shape. | The threshold may not have been crossed, the transformed natural limit may be infinite, or the search may have failed. Rule out numerical truncation before claiming the parameter is substantively unbounded. |
| Equal lower and upper bounds | Inspect the status for a pinned value, structural zero, or unavailable route. | This need not be an interval estimate. Report why the value is fixed. |
| Profile much wider than Wald | Inspect curve shape and fit health. | Disagreement is a diagnostic signal, not proof that either method is correct. Flatness can reflect weak information; jaggedness can reflect failed constrained refits. |
| Finite bootstrap percentile after an unbounded or failed profile | Check unconditional simulation support, failed-refit count, successful-draw distribution, and Monte Carlo resolution. | A finite percentile from finite successful draws does not prove identification or repair a broken generative simulation. |

For a direct target, `confint_inspect()` returns the profile curve, endpoint
comparison, and diagnostic flags without requiring an internal engine parameter
name:

```{r inspect-direct-profile}
inspection <- confint_inspect(fit, parm = "sd_B[1]")
inspection$bounds
inspection$diagnostics
```

`confint_inspect()` is for direct targets listed by `profile_targets()`. Derived
correlations, communalities, proportions, and latent effects do not currently
have public nonlinear profile curves; a raw direct-parameter profile does not
diagnose a nonlinear derived-target constraint.

When covariance components remain confounded, simplify the model or collect
more informative data. Loading constraints can choose an orientation or encode
a prespecified scientific structure, but they do not create information that
separates redundant variance components. Bootstrap likewise depends on a
credible fitted generative model and successful refits; it is not a universal
rescue for multimodality, weak identification, or an unsupported simulation
route.

## References

* Pawitan, Y. (2001). *In All Likelihood: Statistical Modelling and
  Inference Using Likelihood*. Oxford University Press, chapter 9.
* Venzon, D. J. & Moolgavkar, S. H. (1988). A method for computing
  profile-likelihood-based confidence intervals. *Applied Statistics*
  **37**, 87–94. <https://doi.org/10.2307/2347496>
* Self, S. G. & Liang, K.-Y. (1987). Asymptotic properties of maximum
  likelihood estimators and likelihood-ratio tests under nonstandard
  conditions. *Journal of the American Statistical Association* **82**,
  605–610. <https://doi.org/10.2307/2289471>
* Kristensen, K., Nielsen, A., Berg, C. W., Skaug, H., & Bell, B. M.
  (2016). TMB: Automatic Differentiation and Laplace Approximation.
  *Journal of Statistical Software* **70**(5).
  <https://doi.org/10.18637/jss.v070.i05>
