---
title: "Profile-likelihood confidence intervals"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Profile-likelihood confidence intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

> **Preview — Gaussian validated; mixed-family extension is M3 milestone work.**
>
> Profile, Wald, and bootstrap CIs are end-to-end-validated for
> Gaussian fits (CI-02..CI-07 `covered` in the
> [validation-debt register](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/35-validation-debt-register.md)).
> Extension to non-Gaussian and mixed-family fits is **M3**
> milestone work.

`gllvmTMB` uses **profile-likelihood confidence intervals** as the
default for the ratio-quantity extractors (`extract_repeatability()`,
`extract_communality()`, `extract_phylo_signal()`). Profile likelihood
is fast (TMB's C++ inner optim warm-started from the joint MLE),
respects the skewness of variance-component sampling distributions
(Wald CIs do not), and gives accurate coverage near boundaries (where
Wald often breaks down).

**Note** for cross-trait correlations specifically:
[`extract_correlations()`](../reference/extract_correlations.html)
defaults to `method = "fisher-z"`. Correlations are bounded in
$[-1, 1]$ (not $[0, \infty)$ like variance components), and Fisher's
$z$-transform Wald CI with $\tanh$ back-transform is closed-form and
bounded by construction -- both faster and better-behaved than profile
for correlations specifically. Pass `method = "profile"` explicitly
to get profile-likelihood CIs on correlations.

This article walks through the three-method API, shows how the
methods agree on a simulated DGP, and discusses the speed and
coverage trade-offs. It is also the place to point readers at the
underlying TMB tools (`TMB::tmbprofile()`, `TMB::tmbroot()`,
`TMB::checkConsistency()`) that gllvmTMB's wrappers build on.

```{r setup}
library(gllvmTMB)
set.seed(42)
```

## The three-method API

Every CI-producing function in `gllvmTMB` accepts a `method` argument
with three values:

| Method | Speed | Coverage | When |
|---|---|---|---|
| `"wald"` | fastest | poor near boundary, assumes symmetry | small-N quick check |
| `"profile"` *(default for derived ratio quantities)* | fast | accurate, respects skewness | most cases |
| `"bootstrap"` | slowest | accurate, full sampling distribution | non-monotone profile, full-$\boldsymbol\Sigma$ sampling |

The default `method = "profile"` is based on the maintainer's
observation that parametric bootstrap was ~10-50x slower than the
profile likelihood for the same coverage on routine fits. Bootstrap
is preserved as a fallback for non-monotone cases or when full
sampling distributions are needed (e.g. to draw scatterplots of
joint parameter uncertainty).

### What is profile likelihood?

For a single-parameter null $H_0:\, \theta_t = \theta_0$ where
$\theta_t$ is one component of the parameter vector, the profile
log-likelihood is

$$
\ell_p(\theta_0) = \max_{\theta_{-t}} \ell(\theta_0, \theta_{-t}),
$$

where $\theta_{-t}$ denotes all parameters other than $\theta_t$. By
Wilks's theorem,

$$
2\,(\hat\ell - \ell_p(\theta_0)) \xrightarrow{d} \chi^2_1
$$

asymptotically under $H_0$. The 95% CI for $\theta_t$ is the set of
$\theta_0$ values where the deviance crosses $\chi^2_1(0.95) \approx 3.84$.
[Pawitan (2001)](https://academic.oup.com/book/9268) gives the
canonical exposition; [Venzon & Moolgavkar
(1988)](https://doi.org/10.2307/2347496) was the first uniroot-based
numerical algorithm.

For a derived quantity $q = g(\theta)$, the profile is

$$
\ell_p(q_0) = \max_{\theta:\, g(\theta) = q_0} \ell(\theta).
$$

The constraint $g(\theta) = q_0$ is enforced via Lagrange-style
fix-and-refit, warm-started in TMB's C++ inner optim from the joint
MLE -- typically order-of-magnitude faster than refitting in pure R.

## Worked example: variance components, repeatability, correlations

Simulate a 3-trait, 50-site fit with both between- and within-site
structure (the canonical "behavioural syndromes" set-up):

```{r sim}
s <- simulate_site_trait(
  n_sites               = 50, n_species = 6, n_traits = 3,
  mean_species_per_site = 4,
  Lambda_B              = matrix(c(0.9, 0.4, -0.3), 3, 1),
  psi_B                 = c(0.40, 0.30, 0.50),
  psi_W                 = c(0.30, 0.40, 0.30),
  beta                  = matrix(0, 3, 2), seed = 42
)
df_long <- s$data
```

The same model can be fitted from either the long-format data above
or its wide-format equivalent. We show both for the variance-component
fit; the rest of the article re-uses the long-form fit.

```{r build-wide}
library(tidyr)
df_wide <- tidyr::pivot_wider(
  df_long, id_cols = c(site, site_species),
  names_from = trait, values_from = value
)
```

```{r fit}
# Long format (canonical)
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | site, d = 1) + unique(0 + trait | site) +
    unique(0 + trait | site_species),
  data = df_long
)

# Wide data-frame format (same model with compact formula syntax)
fit_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~ 1 +
    latent(1 | site, d = 1) + unique(1 | site) +
    unique(1 | site_species),
  data = df_wide
)

cat("Long/wide logLik difference:",
    signif(abs(as.numeric(logLik(fit_long)) -
               as.numeric(logLik(fit_wide))), 3), "\n")

fit <- fit_long
```

### Variance components: `tmbprofile_wrapper()`

For direct profile-likelihood CIs on individual variance components,
`tmbprofile_wrapper()` exposes a single-parameter or single-contrast
profile:

```{r tmbprof}
## Profile CI for the third trait's between-site log-SD
ci_log <- tmbprofile_wrapper(
  fit, name = "theta_diag_B", which = 3, level = 0.95
)
ci_log

## Same profile, transformed to the natural variance scale
ci_var <- tmbprofile_wrapper(
  fit, name = "theta_diag_B", which = 3, level = 0.95,
  transform = function(x) exp(2 * x)
)
ci_var
```

If the lower bound is `NA`, the profile became one-sided at the
boundary $\sigma^2 \to 0$. This is the standard near-boundary
behaviour for variance components (the constraint $\sigma^2 \geq 0$
makes the asymptotic $\chi^2_1$ approximation conservative; see e.g.
[Self & Liang (1987)](https://doi.org/10.2307/2289471) on boundary
cases). The upper bound is finite and well-behaved either way.

### Repeatability with three CI methods

Per-trait repeatability $R_t = \sigma^2_{B,t} / (\sigma^2_{B,t} +
\sigma^2_{W,t})$ is the canonical behavioural-ecology summary. The
profile CI uses the linear contrast $2(\theta_{B,t} - \theta_{W,t})$
in `TMB::tmbprofile()`'s `lincomb` argument, then transforms via
$\mathrm{plogis}$ to map back to the unit interval.

> **Honest fallback**: when the fit has a full-$\boldsymbol\Sigma$
> structure (i.e. `latent()` factor model on top of `unique()`),
> proper Lagrange-style profile-likelihood CIs for the ratio quantity
> require a constrained refit per parameter that is not yet
> implemented. `extract_repeatability(method = "profile")` currently
> **falls back to Wald (delta-method)** in this case and emits a
> one-shot inform message. The output's `method` column then reports
> `"wald"` so the actual computation is honest. Use
> `method = "bootstrap"` for an empirical CI in the meantime.

```{r rep}
rep_p <- extract_repeatability(fit, level = 0.95, method = "profile")
rep_w <- extract_repeatability(fit, level = 0.95, method = "wald")
rep_b <- extract_repeatability(fit, level = 0.95, method = "bootstrap",
                               nsim = 30, seed = 1)
rep_p   # method column reads "wald" when profile path falls back
rep_w   # explicit Wald (identical to rep_p above on full-Sigma fits)
rep_b   # genuine bootstrap distribution
```

For this fit, `rep_p` and `rep_w` are typically **identical** because
the profile path falls back to Wald (see callout). `rep_b` shows the
genuine sampling distribution and is what you would cite for
publication-grade CIs today. The Wald CI is symmetric on the natural
scale (which is why it can occasionally violate $R \in [0, 1]$ near
the boundary); a proper profile CI would be symmetric on the
contrast scale (keeping $R$ inside the natural support). Use
`bootstrap` until the contrast-scale profile lands.

### Cross-trait correlations: `extract_correlations()`

`extract_correlations()` returns the implied trait-correlation matrix
at one or more tiers as a tidy data frame. For $T$ traits there are
$T(T-1)/2$ unique correlations per tier.

```{r cors}
cors_w <- extract_correlations(fit, tier = "unit", level = 0.95,
                               method = "wald")
cors_w
```

Bootstrap CIs use the existing `bootstrap_Sigma()` machinery:

```{r cors_boot}
cors_b <- extract_correlations(fit, tier = "unit", level = 0.95,
                               method = "bootstrap", nsim = 30, seed = 1)
cors_b
```

### Phylogenetic signal $H^2$

When the fit has both `phylo_unique(species)` and
`unique(0 + trait | species)`, $H^2 = \sigma^2_{\rm phy} /
(\sigma^2_{\rm phy} + \sigma^2_{\rm non})$ is profileable via a
single linear contrast (analogous to repeatability). For richer
3-component decompositions (PGLLVM with
`phylo_latent + latent + unique`), the extractor returns the point
estimate with a `method = "wald(approx)"` flag in the output's
method column. The constrained-refit profile for the 3-component
case is a known gap; bootstrap is the recommended interim.

## Speed comparison

For a fit with $T = 3$ traits, the three CI methods on per-trait
repeatability take roughly:

| Method | Cost | What it computes |
|---|---|---|
| `"wald"` | free given the fit | transforms `sd_report` directly |
| `"profile"` | ~1-3 s per trait | ~5-10 inner optim evaluations per CI bound, warm-started from the joint MLE |
| `"bootstrap"` *(100 reps)* | ~10-30 s | linear in `nsim`; the bootstrap-Sigma machinery refits the model on each replicate |

For a full 6-trait fit with all derived quantities (~75 correlations
+ 6 $H^2$ + 6 repeatabilities + 30 communalities = ~120 derived
parameters), the profile path completes in seconds whereas a
1000-replicate bootstrap takes minutes to hours.

## When to fall back to bootstrap

Profile likelihood assumes the constrained optimum is unique
(monotone profile). It can fail when:

1. **Multi-modal likelihood**: latent-factor models with weakly
   identified rotation can have multiple equivalent optima.
2. **Boundary cases**: variance components pinned at zero produce
   one-sided profiles. Lower bound is `NA`; report half-CIs
   explicitly.
3. **Joint constraints**: when two derived quantities co-move (e.g.
   $\sigma^2_B + \sigma^2_W$ at fixed total), profile CIs on each
   separately understate joint uncertainty. Bootstrap (or a Bayesian
   fit via `tmbstan::tmbstan(fit$tmb_obj)`) gives the joint
   distribution.

For these cases, switch to `method = "bootstrap"`.

## Related TMB tools

`gllvmTMB`'s profile wrappers are thin layers on top of TMB's own
profile-likelihood machinery. For direct access:

* **`TMB::tmbprofile(fit$tmb_obj, name = ...)`** traces the
  profile-likelihood curve for a single parameter or a linear
  contrast. `tmbprofile_wrapper()` above wraps this with the right
  defaults and a transform argument.
* **`TMB::tmbroot(fit$tmb_obj, name = ...)`** is a root-finding
  variant that returns just the two CI bounds rather than tracing the
  full curve. Faster when you only need bounds; doesn't show the
  shape of the profile.
* **`TMB::checkConsistency(fit$tmb_obj, n = 100)`** simulates from the
  fitted model and checks whether the approximate marginal score is
  centred. A non-centred score is a sign that the Laplace
  approximation is unreliable for that fit (the random-effects
  posterior is far from Gaussian, or the data don't constrain the
  random effects well). Use it during development; it's slower than
  `sanity_multi()` but gives a complementary signal.

## References

* Pawitan, Y. (2001). *In All Likelihood: Statistical Modelling and
  Inference Using Likelihood*. Oxford University Press, ch. 9.
* Venzon, D. J. & Moolgavkar, S. H. (1988). A method for computing
  profile-likelihood-based confidence intervals. *Applied Statistics*
  **37**, 87-94.
* Nakagawa, S. & Schielzeth, H. (2010). Repeatability for Gaussian
  and non-Gaussian data. *Biological Reviews* **85**, 935-956.
* Self, S. G. & Liang, K.-Y. (1987). Asymptotic properties of
  maximum likelihood estimators and likelihood ratio tests under
  nonstandard conditions. *JASA* **82**, 605-610.
* Bates, D. *et al.* `lme4::profile.merMod()` source -- the canonical
  profile-likelihood implementation in R for mixed models.
* Viechtbauer, W. (2010). `metafor::confint.rma.uni()` source --
  the canonical profile-CI implementation for meta-analysis variance
  components.
* Kristensen, K., Nielsen, A., Berg, C. W., Skaug, H., & Bell, B. M.
  (2016). TMB: Automatic Differentiation and Laplace Approximation.
  *Journal of Statistical Software* **70**(5).
  <https://doi.org/10.18637/jss.v070.i05>
