---
title: "Psychometrics: confirmatory factor analysis with mixed-response items"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Psychometrics: confirmatory factor analysis with mixed-response items}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

`gllvmTMB` is mostly developed for ecological data, but the underlying
GLLVM machinery is the same one used in psychometrics, item-response
theory (IRT), and structural equation modelling. This article walks
through a **confirmatory factor analysis (CFA)** in `gllvmTMB`'s
syntax: the loading matrix is partly known *a priori*, and a fitting
procedure recovers the free entries while *pinning* the rest at their
hypothesised values.

The example also mixes Likert-scale (continuous, treated Gaussian)
and binary (yes/no) items in one fit via the per-row family
dispatch -- i.e. the same engine call that ecologists use for
`family = list(gaussian(), binomial(), poisson())`.

This article is a **cross-domain validation** of the same engine: it
demonstrates that the package's stacked-trait grammar handles
psychometric data without modification. It is also a worked example
of the [Lambda-constraint pedagogy](lambda-constraint.html) applied
to a CFA pattern (block-diagonal off-pattern zeros) rather than the
exploratory lower-triangular convention.

## The setup

Imagine a survey with $n = 200$ respondents and 6 items:

| Item | Response type | Hypothesised loading |
|---|---|---|
| `attention_1` | Likert 1-5 (Gaussian) | factor 1 (Attention) |
| `attention_2` | Likert 1-5 (Gaussian) | factor 1 (Attention) |
| `attention_3` | Likert 1-5 (Gaussian) | factor 1 (Attention) |
| `mood_1`      | Yes/No (Binary)       | factor 2 (Mood) |
| `mood_2`      | Yes/No (Binary)       | factor 2 (Mood) |
| `mood_3`      | Yes/No (Binary)       | factor 2 (Mood) |

The CFA structural hypothesis: each item loads on **exactly one**
factor. The loading matrix has two free columns (one per factor) but
the off-pattern entries (e.g. `attention_1`'s loading on factor 2)
are pinned at zero.

## Setup

```{r setup}
library(gllvmTMB)
```

## Simulate

The latent factors are 2 orthogonal axes; "attention" items load on
factor 1 with positive loadings, "mood" items load on factor 2.

```{r sim}
set.seed(2025)
n_resp <- 200
T <- 6
item_names <- c("attention_1", "attention_2", "attention_3",
                "mood_1",      "mood_2",      "mood_3")

# True loadings: block-diagonal pattern (the CFA hypothesis)
Lambda_true <- matrix(c(
  0.8,  0.0,
  1.0,  0.0,
  0.7,  0.0,
  0.0,  0.9,
  0.0,  0.6,
  0.0,  0.8
), nrow = T, ncol = 2, byrow = TRUE,
   dimnames = list(item_names, c("attention", "mood")))

# Latent factors (one per respondent)
u <- matrix(rnorm(n_resp * 2), n_resp, 2)
eta <- u %*% t(Lambda_true)   # n_resp x 6 linear predictor

# Three Likert (Gaussian, sigma_eps = 0.4) + three binary
y_likert <- eta[, 1:3] + matrix(rnorm(n_resp * 3, sd = 0.4), n_resp, 3)
y_bin    <- matrix(rbinom(n_resp * 3, 1,
                          plogis(eta[, 4:6])), n_resp, 3)
y <- cbind(y_likert, y_bin)
colnames(y) <- item_names
```

Reshape to long format and tag each row's response family. The
per-row family dispatch needs an `item_family` factor column whose
levels match the order of the `family = list(...)` argument:

The trait-column in the long-format frame is literally called
`trait` because the formula's `latent(0 + trait | respondent, ...)`
references that symbol; in psychometrics the "traits" are survey
items, so we keep the `item_names` vector for labelling and use it
as the factor levels:

```{r long}
df_long <- data.frame(
  respondent = factor(rep(seq_len(n_resp), each = T)),
  trait      = factor(rep(item_names, n_resp), levels = item_names),
  value      = as.vector(t(y))
)
df_long$item_family <- factor(
  ifelse(grepl("^attention", df_long$trait), "g", "b"),
  levels = c("g", "b")
)
table(item = df_long$trait, family = df_long$item_family)
```

The same data also has a natural wide-format representation -- one
row per respondent, one column per item:

```{r wide}
df_wide <- data.frame(
  respondent = factor(seq_len(n_resp)),
  y,
  check.names = FALSE
)
# Show the first 6 rows; rounding only the numeric trait columns.
cbind(df_wide["respondent"], round(df_wide[, item_names], 2)) |> head()
```

## Exploratory fit (no constraint)

First fit without any constraint -- exploratory (rotational) loadings.
The `family = list(...)` mechanism dispatches per row using the
`item_family` column. Both data shapes give the same fit:

```{r fit-exploratory}
fam <- list(gaussian(), binomial())
attr(fam, "family_var") <- "item_family"

# Long format (canonical)
fit_exp_long <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | respondent, d = 2),
  data   = df_long,
  unit   = "respondent",
  family = fam
)

# Wide-format companion: the mixed-family dispatch on a `traits(...)`
# wide formula is not supported (per-row family needs a long-format
# `item_family` column). We compute the wide form only for the
# Gaussian-only sub-fit on the Likert items as a sanity check.
fit_exp_wide_gaussian <- gllvmTMB(
  traits(attention_1, attention_2, attention_3) ~ 1 +
    latent(1 | respondent, d = 2),
  data = df_wide,
  unit = "respondent"
)
# The wide path fits the same Gaussian sub-model on the Likert items.
# The full mixed-family CFA uses the long-format `fit_exp_long`.
fit_exp <- fit_exp_long
fit_exp
```

The exploratory loadings are rotation-invariant; the implied trait
covariance is what we trust:

```{r sigma-exp}
SB_exp <- extract_Sigma(fit_exp, level = "unit")
round(SB_exp$Sigma, 2)
round(SB_exp$R,     2)
```

But for *interpreting* which item loads on which factor, the raw
loadings need to be either pinned (CFA) or rotated (varimax /
post-hoc). The package's rotation advisory in `print(fit_exp)`
already tells us this.

## Confirmatory fit with `lambda_constraint`

Now enforce the CFA hypothesis: pin the off-pattern entries
(`attention_*` items' loadings on factor 2, and `mood_*` items'
loadings on factor 1) to zero. The `suggest_lambda_constraint()`
helper produces a default lower-triangular template; we adapt it
for the block-diagonal CFA pattern by hand. See the
[Lambda-constraint article](lambda-constraint.html) for the
four-pattern walk-through that this CFA fit specialises.

```{r cfa-constraint}
# Build the CFA constraint matrix:
#   attention_*  items free on factor 1, pinned to 0 on factor 2
#   mood_*       items pinned to 0 on factor 1, free on factor 2
cnst_cfa <- matrix(NA_real_, nrow = T, ncol = 2,
                   dimnames = list(item_names, c("f1", "f2")))
cnst_cfa[1:3, 2] <- 0   # attention items: factor 2 loading = 0
cnst_cfa[4:6, 1] <- 0   # mood items:      factor 1 loading = 0
cnst_cfa
```

Fit with this confirmatory pattern:

```{r fit-cfa}
fit_cfa <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | respondent, d = 2),
  data              = df_long,
  unit              = "respondent",
  family            = fam,
  lambda_constraint = list(B = cnst_cfa)
)

L_cfa <- fit_cfa$report$Lambda_B
rownames(L_cfa) <- item_names; colnames(L_cfa) <- c("attention", "mood")
round(L_cfa, 2)
```

Compare to truth:

```{r recover-cfa}
data.frame(
  item     = item_names,
  truth_F1 = Lambda_true[, "attention"],
  fit_F1   = round(L_cfa[, "attention"], 2),
  truth_F2 = Lambda_true[, "mood"],
  fit_F2   = round(L_cfa[, "mood"], 2)
)
```

The pinned-zero pattern is held exactly during fitting, and the free
entries recover their true values to within ~0.1 at this sample size.
This is the GALAMM-style confirmatory loading workflow (Sorensen
2024) that the package supports natively.

## Likelihood-ratio test: is the CFA structure supported?

The exploratory fit nests the CFA fit (CFA pins 6 entries that the
exploratory fit estimates freely). A likelihood-ratio test asks
whether those constraints are statistically defensible:

```{r lrt}
ll_exp <- as.numeric(stats::logLik(fit_exp))
ll_cfa <- as.numeric(stats::logLik(fit_cfa))
LR     <- 2 * (ll_exp - ll_cfa)
df_diff <- 6   # 6 pinned entries
p_value <- pchisq(LR, df = df_diff, lower.tail = FALSE)

data.frame(
  ll_exploratory = ll_exp,
  ll_cfa         = ll_cfa,
  LR_stat        = LR,
  df             = df_diff,
  p_value        = p_value
)
```

A small `LR` statistic (and a large p-value) means the CFA pattern
is consistent with the data; a large `LR` (small p-value) would
suggest residual cross-loadings that the CFA hypothesis does not
explain. (Standard caveat: LRT on boundary parameters can have
non-standard reference distributions; treat as a heuristic.)

## Diagnostics

```{r diagnose}
gllvmTMB_diagnose(fit_cfa, verbose = TRUE)
```

The rotation advisory should now report **no rotational concern** for
$\boldsymbol{\Lambda}_B$ -- the CFA pattern fixes the rotation
completely.

## What this article does *not* cover

* **Multilevel CFA**: items nested within respondents nested within
  groups -- would need a `+ latent(0 + item | group, d = K)`
  between-group term too.
* **Polytomous IRT (graded response / partial credit)**: ordered
  categorical responses are now supported via the `ordinal_probit()`
  family (see the [ordinal-probit article](ordinal-probit.html)). A
  multi-trait polytomous IRT mixing Likert + ordinal + binary items
  in one fit is supported by the engine but is not yet a worked
  example here.
* **Cross-loadings**: the CFA above pins all off-pattern entries to
  zero. A "modification index" workflow (free one zero at a time,
  refit, LRT) is the standard way to explore -- straightforward to do
  by hand here.

## See also

* [Lambda constraints](lambda-constraint.html) -- the four-pattern
  walkthrough of `lambda_constraint` that this CFA example
  specialises.
* [Ordinal-probit](ordinal-probit.html) -- the article that documents
  the threshold-model family used by polytomous IRT.
* [Choose your model](choose-your-model.html) -- decision tree.
* [Pitfalls](pitfalls.html) -- common gotchas including rotation
  invariance.
* **Sorensen, O. (2024)** Multilevel semiparametric latent variable
  modeling in R with galamm. *Multivariate Behavioral Research*
  59(5):1098-1105. <https://doi.org/10.1080/00273171.2024.2385336>
  -- the original `lambda` (= `lambda_constraint`) mechanism.
