---
title: "Behavioural reaction norms with random slopes"
description: "Use repeated Gaussian behaviours across an observed environment to estimate individual intercept-slope covariance and trait-specific reaction norms."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Behavioural reaction norms with random slopes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.2,
  fig.height = 4.8,
  fig.align = "center",
  cache = FALSE,
  message = FALSE
)
```

> **Evidence boundary.** This article demonstrates point estimation for one
> Gaussian random-slope model.
> It does not establish calibrated intervals or a general rank-selection
> procedure for reaction norms.

Repeated measurements can ask two different questions. A behavioural-syndrome
model asks whether individuals differ consistently in their average
behaviour. A reaction-norm model additionally asks whether individuals differ
in their response to a measured context, such as temperature.

This article estimates three kinds of covariance across boldness,
exploration, and activity:

- covariance among individual intercepts at a declared reference temperature;
- covariance among individual temperature slopes; and
- covariance between the intercepts and slopes.

Use the simpler [behavioural-syndrome model](behavioural-syndromes.html) when
stable between- and within-individual covariance is the target. Add random
slopes only when individual differences in response to a continuous context
are scientifically important and each individual has repeated coverage of
that context.

## Start with the design

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

We use a reproducible simulated example with 60 individuals, five sessions
per individual, and three Gaussian response traits. The simulation generator
is shipped as
[`data-raw/examples/make-behavioural-reaction-norm-example.R`](https://github.com/itchyshin/gllvmTMB/blob/main/data-raw/examples/make-behavioural-reaction-norm-example.R).

```{r example-data, eval = FALSE}
rr <- readRDS(system.file(
  "extdata", "examples", "behavioural-reaction-norm-example.rds",
  package = "gllvmTMB"
))
df <- rr$data_long
df_wide <- rr$data_wide
truth <- rr$truth
```

```{r example-data-load, include = FALSE}
example_path <- c(
  file.path(
    "inst", "extdata", "examples",
    "behavioural-reaction-norm-example.rds"
  ),
  file.path(
    "..", "inst", "extdata", "examples",
    "behavioural-reaction-norm-example.rds"
  ),
  file.path(
    "..", "..", "inst", "extdata", "examples",
    "behavioural-reaction-norm-example.rds"
  ),
  system.file(
    "extdata",
    "examples",
    "behavioural-reaction-norm-example.rds",
    package = "gllvmTMB"
  )
)
example_path <- example_path[nzchar(example_path) & file.exists(example_path)][1]
stopifnot(!is.na(example_path))

rr <- readRDS(example_path)
df <- rr$data_long
df_wide <- rr$data_wide
truth <- rr$truth
trait_names <- truth$trait_names
intercept_names <- paste0("intercept.", trait_names)
slope_names <- paste0("slope.temperature.", trait_names)
```

Temperature is stored both in degrees Celsius and as the centred, scaled
predictor used by the model. In a real analysis, save the centring value and
scale so that the intercept and slope remain interpretable:

```{r centring}
temperature_center_C <- truth$temperature_center_C
temperature_scale_C <- truth$temperature_scale_C

c(
  centre_C = temperature_center_C,
  scale_C = temperature_scale_C
)

head(df[c(
  "individual", "session_id", "temperature_C",
  "temperature", "trait", "value"
)], 6)
```

Here `temperature = 0` means the overall mean temperature. A shift in that
origin changes each random intercept from (u) to (u + cb), where (c) is
the shift and (b) is the random slope. Intercept variance and
intercept--slope covariance therefore depend on the reported reference
temperature; slope variance does not.

The essential design check is within-individual coverage. Every simulated
individual traverses the planned temperature range, rather than different
individuals being observed in different narrow temperature bands:

```{r design-check}
design <- unique(df_wide[c(
  "individual", "session", "temperature_C", "temperature"
)])
by_individual <- split(design, design$individual)

design_summary <- data.frame(
  quantity = c(
    "minimum sessions per individual",
    "minimum within-individual range (degrees C)",
    "SD of individual mean temperatures (degrees C)"
  ),
  value = c(
    min(vapply(by_individual, nrow, integer(1))),
    min(vapply(
      by_individual,
      function(x) diff(range(x$temperature_C)),
      numeric(1)
    )),
    sd(vapply(
      by_individual,
      function(x) mean(x$temperature_C),
      numeric(1)
    ))
  )
)
knitr::kable(design_summary, digits = 2)
```

```{r design-plot, fig.height = 4.2, fig.cap = "Observed temperatures for 12 example individuals. Every line spans all five sessions and approximately the same temperature range."}
shown_individuals <- levels(design$individual)[1:12]
ggplot(
  design[design$individual %in% shown_individuals, ],
  aes(session, temperature_C, group = individual)
) +
  geom_line(colour = "#2b6f8a", alpha = 0.55, linewidth = 0.7) +
  geom_point(colour = "#2b6f8a", alpha = 0.75, size = 1.7) +
  scale_x_continuous(breaks = seq_len(truth$n_sessions_per_individual)) +
  labs(
    x = "Session",
    y = "Temperature (degrees C)",
    title = "Repeated context coverage comes before random slopes"
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank())
```

If individuals experience systematically different mean environments in an
observational study, a single globally centred predictor can mix
within-individual response with between-individual exposure differences. In
that setting, consider separating the within-individual deviation from the
individual's mean context. Also check confounding with session order, age,
season, acclimation, and selective missingness. Without design or experimental
support, the fitted temperature slope is an association, not necessarily a
causal plastic response.

## Long and wide forms of the same model

The long data have one row per `(individual, session_id, trait)` observation.
The wide data have one row per session and one column per response trait:

```{r data-shapes}
head(df, 6)
head(df_wide, 3)
```

The long form makes the trait-specific fixed slopes explicit. The
`traits(...)` form is the equivalent compact syntax for an already-wide data
frame:

```{r formulas}
rr$formula_long
rr$formula_wide
```

```{r fit}
ctl <- gllvmTMBcontrol(se = TRUE)

fit_long <- gllvmTMB(
  rr$formula_long,
  data = df,
  trait = rr$fit_args$trait,
  unit = rr$fit_args$unit,
  unit_obs = rr$fit_args$unit_obs,
  family = rr$fit_args$family,
  control = ctl
)

fit_wide <- gllvmTMB(
  rr$formula_wide,
  data = df_wide,
  unit = rr$fit_args$unit,
  unit_obs = rr$fit_args$unit_obs,
  family = rr$fit_args$family,
  control = ctl
)

cat(
  "Absolute long/wide log-likelihood difference:",
  signif(abs(as.numeric(logLik(fit_long)) - as.numeric(logLik(fit_wide))), 3),
  "\n"
)
fit <- fit_long
```

The individual term contains trait-specific random intercepts and slopes. The
session term contains covariance among traits within a repeated occasion:

```{r formula-pieces, eval = FALSE}
# Long form
latent(
  0 + trait + (0 + trait):temperature | individual,
  d = 1
)
latent(0 + trait | session_id, d = 1)

# Wide shorthand
latent(1 + temperature | individual, d = 1)
latent(1 | session_id, d = 1)
```

The simulation plants rank one at each tier, so this example supplies
`d = 1`; it does not select rank one from the data. Use
[rank sensitivity](model-selection-latent-rank.html) before interpreting a
rank in a new study. We interpret covariance matrices, not named raw latent
axes, because an unconstrained loading axis has arbitrary sign and orientation.

## Inspect the complete fit health

```{r diagnostics}
fit_health <- check_gllvmTMB(fit)
knitr::kable(
  fit_health[c("component", "status", "value")],
  row.names = FALSE
)
```

The audited fixture has optimizer convergence and a raw maximum gradient below
`0.01`, so its point surface is numerically usable. Standard-error calculations
are available, the Hessian is positive definite and full rank, and no boundary
flags appear. The final row records that the separate Gaussian residual scale
is mapped off. That is expected here because the per-session diagonal Ψ already
represents row-level, trait-specific variation. The objective-scaled gradient
is descriptive; it is not the convergence cutoff.

Treat warnings according to the quantity they diagnose. An optimizer failure
or raw maximum gradient above `0.01` challenges the point covariance and should
be repaired before interpretation. An `sdreport()` or Hessian warning makes
local Wald uncertainty unavailable or unreliable, but does not automatically
invalidate a stable point covariance; use a supported profile or refit-based
bootstrap route when appropriate. A boundary warning requires inspecting the
named component and scientific target rather than discarding every summary.

## Population-average temperature slopes

The fixed temperature slopes describe the population-average response. They
should be inspected before variation among individual slopes:

```{r fixed-slopes}
fixed <- tidy(fit, effects = "fixed")
slope_rows <- grepl(":temperature$", fixed$term)

fixed_slope_table <- data.frame(
  trait = sub(
    ":temperature$",
    "",
    sub("^trait", "", fixed$term[slope_rows])
  ),
  estimate = fixed$estimate[slope_rows],
  SE = fixed$std.error[slope_rows],
  truth = unname(truth$beta),
  per_degree_C = fixed$estimate[slope_rows] / temperature_scale_C
)
knitr::kable(fixed_slope_table, digits = 3)
```

In this simulated sample, boldness and exploration increase with temperature,
whereas activity decreases. The per-degree column translates the fitted slope
back from the standardized predictor. In observational data, describe these as
conditional associations unless the design justifies a causal interpretation.

## The augmented covariance

For trait (t), individual (i), and session (s), the fitted model is

$$
y_{ist} = μ_t + β_t x_{is} + u_{it} + b_{it}x_{is} + w_{ist}.
$$

The individual coefficients are stacked as intercept, slope, intercept,
slope, and so on across traits:

$$
\mathbf a_i =
\left(u_{i1}, b_{i1}, u_{i2}, b_{i2}, u_{i3}, b_{i3}\right)^\top,
\qquad
\mathbf a_i \sim \operatorname{MVN}(0, \Sigma_{B,\mathrm{aug}}).
$$

The individual covariance is a rank-one shared component plus a diagonal:

$$
\Sigma_{B,\mathrm{aug}}
= \Lambda_{B,\mathrm{aug}}\Lambda_{B,\mathrm{aug}}^\top
+ \Psi_{B,\mathrm{aug}}.
$$

At the session tier,

$$
\mathbf w_{is} \sim \operatorname{MVN}(0, \Sigma_W),
\qquad
\Sigma_W = \Lambda_W\Lambda_W^\top + \Psi_W.
$$

The simulation generated a shared session effect plus independent Gaussian
noise. With one row per trait and session, those two diagonal sources cannot
be estimated separately. The fitted Ψ at the session tier absorbs that
noise, so `extract_Sigma(level = "unit_obs", part = "total")` is the honest
total within-session covariance. The model assumes the individual and session
effects are independent.

```{r extract-sigma}
Sigma_shared <- extract_Sigma(
  fit,
  level = "unit_slope",
  part = "shared"
)
Sigma_unique <- extract_Sigma(
  fit,
  level = "unit_slope",
  part = "unique"
)
Sigma_total <- extract_Sigma(
  fit,
  level = "unit_slope",
  part = "total"
)
Sigma_W_total <- extract_Sigma(
  fit,
  level = "unit_obs",
  part = "total"
)

round(Sigma_total$Sigma[intercept_names, intercept_names], 2)
round(Sigma_total$Sigma[slope_names, slope_names], 3)
round(Sigma_total$Sigma[intercept_names, slope_names], 3)
```

The intercept--intercept block describes stable behavioural covariance at the
reference temperature. The slope--slope block describes covariance among
individual differences in temperature response. The cross block asks whether
individuals above the population mean at the reference temperature tend to
have steeper or shallower slopes. That last interpretation changes when the
reference temperature changes.

The public extractor also verifies the decomposition:

```{r composition}
max(abs(
  Sigma_total$Sigma -
    (Sigma_shared$Sigma + diag(
      Sigma_unique$s,
      nrow = length(Sigma_unique$s)
    ))
))
```

## Check recovery block by block

One overall error can hide poor recovery of the smaller slope block. We
therefore report each block separately. These are checks of one known
simulation, not empirical coverage or a guarantee for other designs.

```{r recovery-data}
unit_slope_comparison <- function(estimate, truth) {
  idx <- which(lower.tri(truth, diag = TRUE), arr.ind = TRUE)
  row_name <- rownames(truth)[idx[, "row"]]
  col_name <- colnames(truth)[idx[, "col"]]
  row_is_slope <- grepl("^slope\\.", row_name)
  col_is_slope <- grepl("^slope\\.", col_name)

  data.frame(
    estimate = estimate[idx],
    truth = truth[idx],
    block = ifelse(
      row_is_slope & col_is_slope,
      "slope-slope",
      ifelse(
        !row_is_slope & !col_is_slope,
        "intercept-intercept",
        "intercept-slope"
      )
    )
  )
}

cov_compare <- unit_slope_comparison(
  Sigma_total$Sigma,
  truth$Sigma_unit_slope
)

block_error <- do.call(rbind, lapply(
  split(cov_compare, cov_compare$block),
  function(x) data.frame(
    block = x$block[1],
    mean_absolute_error = mean(abs(x$estimate - x$truth)),
    relative_error = sqrt(sum((x$estimate - x$truth)^2)) /
      sqrt(sum(x$truth^2))
  )
))
knitr::kable(block_error, digits = 3, row.names = FALSE)
```

```{r recovery-plot, fig.width = 9, fig.height = 6.2, fig.class="wide-scientific-figure", fig.cap = "Estimated versus true augmented individual covariance entries. Colours separate the intercept, slope, and cross blocks; the dashed line is perfect recovery.", fig.alt = "Scatterplot of fitted against known covariance entries. Green points are intercept-intercept entries, purple points are intercept-slope entries, and orange points are slope-slope entries. A dashed diagonal marks exact recovery."}
ggplot(cov_compare, aes(truth, estimate, colour = block)) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = 2,
    colour = "grey45"
  ) +
  geom_point(size = 2.6, alpha = 0.88) +
  coord_equal() +
  scale_colour_manual(
    values = c(
      "intercept-intercept" = "#1b9e77",
      "intercept-slope" = "#7570b3",
      "slope-slope" = "#d95f02"
    ),
    name = "Covariance block"
  ) +
  labs(
    x = "Known simulation truth",
    y = "Fitted estimate",
    title = "Recovery by covariance block"
  ) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## Pointwise repeatability across temperature

For one trait at standardized temperature (x), the fitted
among-individual variance is

$$
V_{B,t}(x) =
\operatorname{Var}(u_{it})
+ 2x\operatorname{Cov}(u_{it}, b_{it})
+ x^2\operatorname{Var}(b_{it}).
$$

The pointwise repeatability is

$$
R_t(x) = \frac{V_{B,t}(x)}{V_{B,t}(x) + (\Sigma_W)_{tt}}.
$$

This is repeatability at one context value. It is not the correlation between
the same individual measured at two different temperatures.

We restrict the curve to the temperature interval observed for every
simulated individual. The fitted denominator uses only public covariance
output; it does not read an internal residual field.

```{r repeatability-data}
temperature_support <- split(
  design$temperature,
  design$individual
)
x_grid <- seq(
  max(vapply(temperature_support, min, numeric(1))),
  min(vapply(temperature_support, max, numeric(1))),
  length.out = 80
)

reaction_repeatability <- function(
  Sigma_aug,
  Sigma_W,
  x,
  traits,
  source
) {
  rows <- vector("list", length(traits))
  for (k in seq_along(traits)) {
    trait <- traits[k]
    intercept <- paste0("intercept.", trait)
    slope <- paste0("slope.temperature.", trait)
    between <-
      Sigma_aug[intercept, intercept] +
      2 * x * Sigma_aug[intercept, slope] +
      x^2 * Sigma_aug[slope, slope]
    rows[[k]] <- data.frame(
      trait = trait,
      temperature_C = x * temperature_scale_C + temperature_center_C,
      repeatability = between / (between + Sigma_W[trait, trait]),
      source = source
    )
  }
  do.call(rbind, rows)
}

Sigma_W_truth_total <- truth$Sigma_unit_obs +
  diag(truth$sigma_eps^2, nrow = length(trait_names))
dimnames(Sigma_W_truth_total) <- list(trait_names, trait_names)

repeatability_rows <- rbind(
  reaction_repeatability(
    Sigma_total$Sigma,
    Sigma_W_total$Sigma,
    x_grid,
    trait_names,
    "fitted"
  ),
  reaction_repeatability(
    truth$Sigma_unit_slope,
    Sigma_W_truth_total,
    x_grid,
    trait_names,
    "truth"
  )
)

repeatability_wide <- reshape(
  repeatability_rows,
  idvar = c("trait", "temperature_C"),
  timevar = "source",
  direction = "wide"
)
cat(
  "Mean absolute repeatability error:",
  round(mean(abs(
    repeatability_wide$repeatability.fitted -
      repeatability_wide$repeatability.truth
  )), 3),
  "\n"
)
```

```{r repeatability-plot, fig.cap = "Pointwise repeatability over the temperature interval shared by every simulated individual. Solid lines are fitted point estimates and dashed lines are simulation truth; no uncertainty intervals are claimed."}
ggplot(
  repeatability_rows,
  aes(
    temperature_C,
    repeatability,
    colour = trait,
    linetype = source
  )
) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(
    values = c(
      boldness = "#1b9e77",
      exploration = "#d95f02",
      activity = "#7570b3"
    ),
    name = "Trait"
  ) +
  scale_linetype_manual(
    values = c(fitted = "solid", truth = "22"),
    name = NULL
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.25)
  ) +
  labs(
    x = "Temperature (degrees C)",
    y = "Pointwise repeatability",
    title = "Random slopes make repeatability context-dependent",
    subtitle = "Point estimates within common observed support"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

The curves compare fitted point estimates with one known truth. They do not
provide confidence bands or validate interval coverage. In real data, avoid
extending the curve beyond well-replicated within-individual support.

## What to report

Report the context units, centring value, scaling value, and the actual
within-individual context coverage. Then report:

- population-average slopes from `tidy(fit, effects = "fixed")`;
- the rank supplied at each covariance tier;
- the total augmented covariance from
  `extract_Sigma(level = "unit_slope", part = "total")`;
- the reference temperature used to interpret intercept and cross blocks;
- the total within-session covariance from
  `extract_Sigma(level = "unit_obs", part = "total")`; and
- every non-passing row from `check_gllvmTMB()` before interpreting results.

This worked example is limited to an ordinary Gaussian reaction norm with a
rank supplied in advance. It does not establish rank selection, calibrated
intervals for variance or repeatability curves, non-Gaussian random-slope
inference, or a causal temperature effect. Structured phylogenetic, animal,
and spatial random slopes require a separate scientific model and are not
evidence for this ordinary individual-level workflow.

## Next decision

Interpret the reaction-norm covariance only when individuals cover the
environmental gradient, the core health rows pass, and the fitted curve remains
inside the observed within-individual range. If the point fit is unhealthy, go
to [Convergence and hard fits](convergence-start-values.html). If the question
does not require individual slopes, return to
[Behavioural syndromes](behavioural-syndromes.html) and use the simpler
between/within covariance model.

## References

Dingemanse, N. J. and Dochtermann, N. A. (2013). Quantifying individual
variation in behaviour: mixed-effect modelling approaches. *Journal of Animal
Ecology* 82, 39--54.

O'Dea, R. E., Noble, D. W. A., and Nakagawa, S. (2022). Unifying individual
differences in personality, predictability and plasticity: a practical guide.
*Methods in Ecology and Evolution* 13, 278--293.

Stamps, J. A. (2016). Individual differences in behavioural plasticities.
*Biological Reviews* 91, 534--567.
