---
title: "Random regression and reaction-norm slopes"
tier: 3 # internal implementation article; public behavioural example still pending
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Random regression and reaction-norm slopes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  message  = FALSE,
  warning  = FALSE
)
```

This is the internal article for **ordinary individual-level random
regression**. For Gaussian responses, the augmented
`latent(1 + x | individual, d = K) + unique(1 + x | individual)` path now fits
the behavioural reaction-norm target at the ID tier and extracts the full
`Lambda_aug Lambda_aug^T + Psi_B,aug` decomposition.

The article should become public only when the worked example can show the
behavioural-syndrome data story, rendered long and wide examples, diagnostics,
figures, and the non-Gaussian boundary without using deprecated
single-variance slope keywords or structured phylogenetic/spatial dependence
as a substitute.

## The biological question

In behavioural-syndrome data, the unit is the **individual**. Individuals are
measured repeatedly across occasions, and each occasion may carry a continuous
context such as temperature, food availability, social density, or perceived
risk. A random-regression model asks:

> Do individuals differ not only in their average behaviour, but also in how
> their behaviour changes along the context gradient?

Following the behavioural-syndrome appendix, among-individual differences in
average trait expression are **personality**. Among-individual differences in
responsiveness to the context gradient are **plasticity**. The random slope is
therefore an ID-level effect: it belongs at `unit = "individual"`, not at the
within-individual occasion tier.

## Data shape

The long data need one row per `(individual, session_id, trait)` observation:

```{r data-shape, eval = FALSE}
head(behaviour_long)
#>   individual session_id temperature       trait value
#> 1          1        1_1       -1.25    boldness  ...
#> 2          1        1_1       -1.25 exploration  ...
#> 3          1        1_1       -1.25     activity  ...
#> 4          1        1_2        0.10    boldness  ...
```

Use `unit = "individual"` for the between-individual tier and
`unit_obs = "session_id"` for the within-individual occasion tier. The
`session_id` column has one level per individual-occasion combination, so all
traits measured on the same individual at the same occasion share the same
within-individual random effect.

## Current Gaussian random-regression model

For trait `t`, individual `i`, and occasion `j`, with centred/scaled
temperature `temperature_ij`, the target model is

$$
y_{ijt} =
\mu_t + \beta_{t,\mathrm{temp}}\,\mathrm{temperature}_{ij}
+ u_{it} + b_{it}\,\mathrm{temperature}_{ij} + e_{ijt}.
$$

Here `u_it` is the individual-specific random intercept for trait `t`
(personality), `b_it` is the individual-specific random slope for trait `t`
(plasticity), and `e_ijt` is the within-individual occasion-level deviation.
The within-individual component is still a random-intercept/covariance tier
across traits; the random slope lives only at the ID tier.

The implemented long-form syntax pairs the shared `latent()` component with
the augmented diagonal `unique()` component:

```{r target-long-gaussian, eval = FALSE}
fit_rr_gaussian <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):temperature +
    latent(0 + trait + (0 + trait):temperature | individual, d = d_B) +
    unique(0 + trait + (0 + trait):temperature | individual) +
    latent(0 + trait | session_id, d = d_W) +
    unique(0 + trait | session_id),
  data     = behaviour_long,
  trait    = "trait",
  unit     = "individual",
  unit_obs = "session_id"
)
```

The matching wide-data formula is:

```{r target-wide-gaussian, eval = FALSE}
fit_rr_gaussian_wide <- gllvmTMB(
  traits(boldness, exploration, activity) ~ 1 + temperature +
    latent(1 + temperature | individual, d = d_B) +
    unique(1 + temperature | individual) +
    latent(1 | session_id, d = d_W) +
    unique(1 | session_id),
  data     = behaviour_wide,
  unit     = "individual",
  unit_obs = "session_id"
)
```

The augmented between-individual covariance is extracted as shared, unique, or
total:

```{r extract-unit-slope, eval = FALSE}
Sigma_shared <- extract_Sigma(fit_rr_gaussian, level = "unit_slope",
                              part = "shared")
Sigma_unique <- extract_Sigma(fit_rr_gaussian, level = "unit_slope",
                              part = "unique")
Sigma_total  <- extract_Sigma(fit_rr_gaussian, level = "unit_slope",
                              part = "total")
Sigma_total$Sigma
```

This returns the `2T x 2T` covariance with rows labelled
`intercept.<trait>` and `slope.temperature.<trait>`. `part = "shared"` gives
`Lambda_aug Lambda_aug^T`, `part = "unique"` gives the augmented diagonal
`Psi_B,aug`, and `part = "total"` gives their sum.

## What the covariance means

The between-individual component stacks all trait intercepts and slopes for
individual `i`:

$$
\mathbf{b}_{B,i} =
\begin{pmatrix}
\mathbf{u}_i \\
\mathbf{b}_i
\end{pmatrix}
\in \mathbb{R}^{2T}.
$$

The paired decomposition is

$$
\mathbf{b}_{B,i}
= \Lambda_{B,\mathrm{aug}} \mathbf{z}_{B,i} + \boldsymbol\epsilon_{B,i},
\qquad
\mathbf{z}_{B,i} \sim \mathcal{N}(\mathbf{0}, I_{d_B}),
\qquad
\boldsymbol\epsilon_{B,i} \sim \mathcal{N}(\mathbf{0}, \Psi_{B,\mathrm{aug}}),
$$

so that

$$
\Sigma_B^{(\mathrm{aug})}
= \Lambda_{B,\mathrm{aug}}\Lambda_{B,\mathrm{aug}}^\top
+ \Psi_{B,\mathrm{aug}}.
$$

The current Gaussian implementation covers both the low-rank shared component
`Lambda_B,aug Lambda_B,aug^T` and the diagonal augmented `Psi_B,aug`
component. Non-Gaussian augmented `unique()` random slopes remain guarded
until they have their own validation evidence.

Partitioning the augmented covariance into `T x T` blocks gives:

$$
\Sigma_B^{(\mathrm{aug})}
=
\begin{pmatrix}
\Sigma_B^{(u)} & \Sigma_B^{(u,b)} \\
\Sigma_B^{(b,u)} & \Sigma_B^{(b)}
\end{pmatrix}.
$$

These blocks answer different biological questions:

- `Sigma_B^(u)` gives intercept-intercept correlations: personality or
  behavioural-syndrome correlations at the reference context.
- `Sigma_B^(b)` gives slope-slope correlations: plasticity-syndrome
  correlations, asking whether individuals responsive in one trait are also
  responsive in another.
- `Sigma_B^(u,b)` gives intercept-slope correlations: personality-plasticity
  associations, asking whether high-average individuals also respond more or
  less strongly to the context.

Centre the context variable at a meaningful reference point. The intercept is
the expected trait value when `temperature = 0`, so the intercept-intercept and
intercept-slope blocks depend on that choice.

## Repeatability along the gradient

With random slopes, between-individual variance is context-dependent. For a
single trait,

$$
\mathrm{Var}(u_{it} + b_{it} x)
= \mathrm{Var}(u_{it})
+ 2x\,\mathrm{Cov}(u_{it}, b_{it})
+ x^2\,\mathrm{Var}(b_{it}).
$$

The temperature-specific repeatability is therefore

$$
R_t(x) =
\frac{\mathrm{Var}(u_{it} + b_{it} x)}
{\mathrm{Var}(u_{it} + b_{it} x) + (\Sigma_W)_{tt}}.
$$

This is the main reason the article should be about individual reaction norms,
not phylogenetic or spatial slope fields. The applied reader wants to know how
repeatability and behavioural integration change along an observed
within-individual context gradient.

## Current fallback without individual slopes

If the scientific question needs the established full `latent + unique`
decomposition rather than individual-specific slopes, use the two-level
behavioural-syndrome model with average trait-specific context effects:

```{r current-fallback, eval = FALSE}
fit_fixed_context <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):temperature +
    latent(0 + trait | individual, d = d_B) +
    unique(0 + trait | individual) +
    latent(0 + trait | session_id, d = d_W) +
    unique(0 + trait | session_id),
  data     = behaviour_long,
  trait    = "trait",
  unit     = "individual",
  unit_obs = "session_id"
)
```

This fallback estimates the population-average reaction norm for each trait
and the usual between- and within-individual covariance matrices. It does
**not** estimate individual-specific random slopes, plasticity syndromes, or
temperature-specific repeatability.

## Implementation gate

This article returns to the public Model guide only when the implemented
ordinary unit-tier random-regression path is complete enough for a worked
example:

- `latent(0 + trait + (0 + trait):x | individual, d = d_B)` is implemented
  and covered by the focused Gaussian / Poisson tests in
  `test-ordinary-latent-random-regression.R`;
- `unique(0 + trait + (0 + trait):x | individual)` is implemented for
  Gaussian responses and supplies the diagonal `Psi_B,aug` component;
- the equivalent `traits(...)` wide shorthand is implemented for the paired
  `latent + unique` path;
- `extract_Sigma(level = "unit_slope", part = "shared" / "unique" / "total")`
  returns the augmented covariance pieces with labelled intercept and slope
  blocks;
- focused deterministic Gaussian recovery covers the intercept-intercept,
  slope-slope, and intercept-slope blocks; and
- a reader-facing behavioural example with `unit = "individual"`,
  `unit_obs = "session_id"`, diagnostics, and figures still needs to be
  built before public restoration.

Until then, public docs should route users to the current
[Behavioural syndromes](behavioural-syndromes.html) two-level article for
between/within covariance, and to
[Structured random slopes for non-Gaussian traits](random-slopes-nongaussian.html)
only when the scientific model genuinely involves phylogenetic or spatial
dependence.
