---
title: "Random regression and reaction-norm slopes"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Random regression and reaction-norm slopes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = requireNamespace("ape", quietly = TRUE),
  message  = FALSE,
  warning  = FALSE
)
```

This article is for a reader who wants the **slope** of a response on a
continuous covariate to vary across the units of a structured group --
species along a phylogeny, or individuals in a pedigree -- with the
unit-level slopes correlated by a known relatedness matrix
$\mathbf{A}$. In quantitative genetics this is the **random-regression
/ reaction-norm** model (Meyer & Hill 1997; Nussey *et al.* 2007): a
per-unit slope on an environmental gradient, with the slopes treated as
a heritable (or phylogenetically conserved) trait.

## What this article covers, and what it does not

`gllvmTMB` v0.2.0 ships **one** validated random-slope mode: a single
shared slope variance whose unit-level slopes are correlated by
$\mathbf{A}$. It is reached two ways:

- [`phylo_slope(x | species)`](../reference/phylo_slope.html) -- slopes
  correlated by a phylogeny (supply the tree).
- [`animal_slope(x | id)`](../reference/animal_slope.html) -- slopes
  correlated by a pedigree-derived additive-genetic matrix
  $\mathbf{A}$. This is the same engine path:
  `animal_slope(x | id)` desugars to `phylo_slope(x | id, vcv = A)`,
  and the two recover **byte-identically** (shown below).

This single-variance mode is the worked tutorial here. Beyond it, the
package also *accepts and fits* a richer **correlated intercept + slope**
form, [`phylo_unique(1 + x | species)`](../reference/phylo_unique.html),
which is shown in the final section as a fitting demonstration.

Several related forms are **not yet available** and are flagged where
they would naturally appear, so the scope is explicit rather than
implied:

| Form | Status in v0.2.0 |
|---|---|
| `phylo_slope` / `animal_slope` (single shared slope variance) | **Validated** (this article) |
| `phylo_unique(1 + x \| sp)` correlated intercept+slope | Fits; 2x2 (intercept, slope) covariance **reporting** is not surfaced in this build |
| `phylo_indep(1 + x \| sp)` independent (diagonal) intercept+slope | **Forthcoming** -- parser does not yet accept the richer LHS |
| Structured `*_dep` / `*_latent` slopes (phylo or spatial) | **Forthcoming** -- engine in progress |
| Calibrated confidence intervals on slope variances | **Not established** -- point estimates only (see the CI note) |

## The model

For a single continuous covariate $x$, the slope is allowed to vary by
unit $i$ (species or individual). Writing $\eta_{it}$ for the linear
predictor of trait $t$ on unit $i$ at covariate value $x_{io}$,

$$
\eta_{it} \;\mathrel{{+}{=}}\; \beta(i)\, x_{io},
\qquad
\boldsymbol\beta \sim
\mathcal{N}\!\bigl(\mathbf{0},\; \sigma^2_{\text{slope}}\,\mathbf{A}\bigr).
$$

The slope vector $\boldsymbol\beta$ has one entry per unit
($n_{\text{species}}$ or $n_{\text{individuals}}$). It is **shared
across traits** -- the same $\beta(i)$ multiplies $x_{io}$ for every
trait of unit $i$ -- and **one shared variance**
$\sigma^2_{\text{slope}}$ governs how far slopes spread, while
$\mathbf{A}$ (the phylogenetic correlation matrix, or the pedigree
additive-genetic matrix) sets how the slopes covary between related
units. This is the initial-release scope documented on
[`phylo_slope()`](../reference/phylo_slope.html): one covariate, one
slope variance, slopes shared across traits.

The estimand is $\sigma^2_{\text{slope}}$ (how much heritable /
phylogenetic variation there is in the slope), together with the
per-unit slopes $\beta(i)$ themselves.

## Setup

```{r setup}
library(gllvmTMB)
library(ape)
```

## Simulate a phylogenetic reaction norm

We simulate per-species slopes drawn from
$\mathcal{N}(\mathbf 0,\, \sigma^2_{\text{slope}}\,\mathbf{C}_{\text{phy}})$,
with true $\sigma^2_{\text{slope}} = 0.4$, then build a long data frame
with several traits and repeated observations per species at varying
$x$. This is a syntax-and-recovery illustration on one data set, not a
full simulation study.

```{r sim}
set.seed(2026)
n_sp <- 80
n_traits <- 4
tree <- ape::rcoal(n_sp)
tree$tip.label <- paste0("sp", seq_len(n_sp))
Cphy <- ape::vcv(tree, corr = TRUE)

sigma2_slope_true <- 0.4
Lphy <- t(chol(Cphy + diag(1e-8, n_sp)))
beta_slope_true <- as.numeric(
  Lphy %*% rnorm(n_sp, sd = sqrt(sigma2_slope_true))
)
names(beta_slope_true) <- tree$tip.label

n_rep <- 5
trait_lvls <- paste0("t", seq_len(n_traits))
df <- expand.grid(
  species = factor(tree$tip.label, levels = tree$tip.label),
  trait   = factor(trait_lvls, levels = trait_lvls),
  rep     = seq_len(n_rep)
)
df$x <- rnorm(nrow(df))
mu_t  <- c(2, 1.5, 1, 0.5)[as.integer(df$trait)]
df$value <- mu_t +
  beta_slope_true[as.character(df$species)] * df$x +
  rnorm(nrow(df), sd = 0.5)
head(df)
```

## Fit the phylogenetic random slope

The slope term is `phylo_slope(x | species)`; pass the tree inside the
keyword with `tree =`. The fixed part `0 + trait + (0 + trait):x` gives
each trait its own intercept and its own *mean* slope on $x$; the random
`phylo_slope` term adds the per-species deviation around that mean.

```{r fit}
fit <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):x + phylo_slope(x | species, tree = tree),
  data  = df,
  trait = "trait",
  unit  = "species"
)
fit$opt$convergence
```

### Recover the slope variance

The single shared slope standard deviation is reported on the log scale
as `log_sigma_slope` in the optimised parameter vector. Exponentiate to
get $\sigma_{\text{slope}}$:

```{r sigma-slope}
sigma_slope_hat <- exp(
  fit$opt$par[grepl("log_sigma_slope", names(fit$opt$par))]
)
data.frame(
  param = "sigma_slope",
  true  = sqrt(sigma2_slope_true),
  est   = signif(unname(sigma_slope_hat), 3)
)
```

### Recover the per-species slopes

The fitted per-species slopes are the `b_phy_slope` random effects.
They live on the tree-augmented index, so align them to the tip labels
through the rownames of the augmented precision matrix:

```{r per-species-slopes, fig.width = 5.4, fig.height = 5.4}
b_full <- fit$tmb_obj$env$last.par.best
b_slope <- as.numeric(b_full[names(b_full) == "b_phy_slope"])

aug_names <- rownames(fit$tmb_data$Ainv_phy_rr)
tip_idx   <- match(tree$tip.label, aug_names)
b_tips    <- b_slope[tip_idx]

cor_fitted_true <- cor(b_tips, beta_slope_true)
round(cor_fitted_true, 3)

plot(beta_slope_true, b_tips,
     xlab = "True per-species slope",
     ylab = "Fitted per-species slope (BLUP)",
     pch = 19, col = "#3366aa",
     main = "Reaction-norm slopes recover their truth")
abline(0, 1, lty = 2)
```

The shared variance is recovered close to its true value, and the
per-species slopes track the simulated truth with a high correlation.
For a single small data set these are point estimates only -- see the
note on confidence intervals below.

## The pedigree twin: `animal_slope`

When the rows are *individuals* with a known pedigree rather than
*species* with a phylogeny, the identical model is written with
[`animal_slope(x | id)`](../reference/animal_slope.html). The only
change is the source of $\mathbf{A}$: a pedigree additive-genetic matrix
instead of a tree correlation matrix. This is the canonical
quantitative-genetic random-regression model -- **heritable variation
in a reaction-norm slope** (Meyer & Hill 1997; Nussey *et al.* 2007).

To show the equivalence concretely, we treat the same relatedness
matrix two ways on the same data: once through `animal_slope(x | id,
A = A)`, and once through `phylo_slope(x | id, vcv = A)`. Because
`animal_slope` desugars to exactly the latter, the two fits agree to
numerical precision. (When $\mathbf{A}$ is supplied directly via `A =`
/ `vcv =`, give the relatedness key as `cluster =`.)

```{r animal-twin}
set.seed(11)
n_id <- 60
tree_id <- ape::rcoal(n_id)
tree_id$tip.label <- paste0("i", seq_len(n_id))
A <- ape::vcv(tree_id, corr = TRUE)   # stand-in relatedness matrix, rownames = id

s2 <- 0.4
LA <- t(chol(A + diag(1e-8, n_id)))
beta_id_true <- as.numeric(LA %*% rnorm(n_id, sd = sqrt(s2)))
names(beta_id_true) <- tree_id$tip.label

dat <- expand.grid(
  id    = factor(tree_id$tip.label, levels = tree_id$tip.label),
  trait = factor(c("t1", "t2"), levels = c("t1", "t2")),
  rep   = seq_len(5)
)
dat$x <- rnorm(nrow(dat))
dat$value <- c(1, 0.5)[as.integer(dat$trait)] +
  beta_id_true[as.character(dat$id)] * dat$x +
  rnorm(nrow(dat), sd = 0.5)

fit_animal <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):x + animal_slope(x | id, A = A),
  data = dat, trait = "trait", unit = "id", cluster = "id"
)
fit_phylo <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):x + phylo_slope(x | id, vcv = A),
  data = dat, trait = "trait", unit = "id", cluster = "id"
)

ss_animal <- exp(fit_animal$opt$par[grepl("log_sigma_slope",
                                          names(fit_animal$opt$par))])
ss_phylo  <- exp(fit_phylo$opt$par[grepl("log_sigma_slope",
                                         names(fit_phylo$opt$par))])
data.frame(
  routing      = c("animal_slope(x | id, A = A)", "phylo_slope(x | id, vcv = A)"),
  sigma_slope  = signif(c(unname(ss_animal), unname(ss_phylo)), 6)
)

# Byte-equivalence of the full optimised parameter vector
cat("max |opt$par difference| :",
    signif(max(abs(as.numeric(fit_animal$opt$par) -
                   as.numeric(fit_phylo$opt$par))), 3), "\n")
```

The two routings produce the same $\sigma_{\text{slope}}$ and the same
optimised parameter vector. `animal_slope(x | id)` is therefore the
pedigree-facing name for the same random-regression machinery; use it
when your relatedness comes from a pedigree (build $\mathbf{A}$ with
[`pedigree_to_A()`](../reference/pedigree_to_A.html); see the
[animal-model](animal-model.html) article for the pedigree workflow).

## Non-Gaussian families

The random-slope term is not restricted to Gaussian responses -- it
rides on the same GLLVM machinery as the rest of the package, so it
composes with the count, proportion, and ordinal families. As a quick
demonstration, the same phylogenetic reaction norm under a Poisson
log-link:

```{r poisson-slope}
set.seed(5640)
n_sp2 <- 50
tree2 <- ape::rcoal(n_sp2)
tree2$tip.label <- paste0("sp", seq_len(n_sp2))
C2 <- ape::vcv(tree2, corr = TRUE)
L2 <- t(chol(C2 + diag(1e-8, n_sp2)))
beta2_true <- as.numeric(L2 %*% rnorm(n_sp2, sd = sqrt(0.3)))
names(beta2_true) <- tree2$tip.label

dfp <- expand.grid(
  species = factor(tree2$tip.label, levels = tree2$tip.label),
  trait   = factor(paste0("t", 1:3), levels = paste0("t", 1:3)),
  rep     = seq_len(5)
)
dfp$x <- rnorm(nrow(dfp))
eta <- c(2, 2.1, 1.9)[as.integer(dfp$trait)] +
  beta2_true[as.character(dfp$species)] * dfp$x
dfp$value <- rpois(nrow(dfp), exp(eta))

fit_pois <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):x + phylo_slope(x | species, tree = tree2),
  data   = dfp,
  trait  = "trait",
  unit   = "species",
  family = poisson(link = "log")
)
data.frame(
  param = "sigma_slope (poisson)",
  true  = signif(sqrt(0.3), 3),
  est   = signif(unname(exp(fit_pois$opt$par[
    grepl("log_sigma_slope", names(fit_pois$opt$par))])), 3)
)
```

The slope variance recovers under the count family as it does under
Gaussian. The same pattern holds for the other core response families
(negative binomial, gamma, beta, ordinal, binomial) -- swap the
`family =` argument and keep the formula. See
[response-families](response-families.html) for the family catalogue.

## Correlated intercept and slope

The single-variance model above varies only the slope. A fuller
random-regression model lets each unit have **both** a random intercept
and a random slope, with the two correlated -- the standard
$2 \times 2$ $(\text{intercept}, \text{slope})$ covariance of
quantitative-genetic reaction norms. In `gllvmTMB` this is the
augmented form of
[`phylo_unique()`](../reference/phylo_unique.html), written
`phylo_unique(1 + x | species)` on a wide
[`traits()`](../reference/traits.html) surface, or equivalently
`phylo_unique(0 + trait + (0 + trait):x | species)` in long form. Both
spellings dispatch to the same engine.

The model places the per-species intercept and slope under a joint
prior,

$$
\begin{pmatrix} \alpha_i \\ \beta_i \end{pmatrix}
\sim
\mathcal{N}\!\left(\mathbf{0},\;
\boldsymbol\Sigma_{b} \otimes \mathbf{A}\right),
\qquad
\boldsymbol\Sigma_{b} =
\begin{pmatrix}
\sigma^2_{\text{int}} & \rho\,\sigma_{\text{int}}\sigma_{\text{slope}} \\
\rho\,\sigma_{\text{int}}\sigma_{\text{slope}} & \sigma^2_{\text{slope}}
\end{pmatrix}.
$$

The fit below converges and is byte-identical between the long and wide
spellings:

```{r augmented, eval = requireNamespace("tidyr", quietly = TRUE)}
library(tidyr)
set.seed(5640)
n_sp3 <- 60
tree3 <- ape::rcoal(n_sp3)
tree3$tip.label <- paste0("sp", seq_len(n_sp3))
C3 <- ape::vcv(tree3, corr = TRUE)
L3 <- t(chol(C3 + diag(1e-8, n_sp3)))

s2_int <- 0.4; s2_slp <- 0.3; rho <- 0.5
cov_ab <- rho * sqrt(s2_int * s2_slp)
Sigma_b <- matrix(c(s2_int, cov_ab, cov_ab, s2_slp), 2, 2)
ab <- (L3 %*% matrix(rnorm(n_sp3 * 2), n_sp3, 2)) %*% chol(Sigma_b)
colnames(ab) <- c("alpha", "beta")
rownames(ab) <- tree3$tip.label

sr <- expand.grid(
  species = factor(tree3$tip.label, levels = tree3$tip.label),
  rep = seq_len(4)
)
sr$x <- rnorm(nrow(sr))
tl3 <- paste0("t", 1:3)
dl <- merge(sr, data.frame(trait = factor(tl3, levels = tl3)), all = TRUE)
dl <- dl[order(dl$species, dl$rep, dl$trait), ]
dl$value <- c(2, 1, 0.5)[as.integer(dl$trait)] +
  ab[as.character(dl$species), "alpha"] +
  ab[as.character(dl$species), "beta"] * dl$x +
  rnorm(nrow(dl), sd = 0.3)

dw <- as.data.frame(tidyr::pivot_wider(
  dl, id_cols = c(species, rep, x),
  names_from = trait, values_from = value
))

fit_long <- gllvmTMB(
  value ~ 0 + trait + phylo_unique(0 + trait + (0 + trait):x | species,
                                   tree = tree3),
  data = dl, trait = "trait", unit = "species"
)
fit_wide <- gllvmTMB(
  traits(t1, t2, t3) ~ 1 + phylo_unique(1 + x | species, tree = tree3),
  data = dw, unit = "species"
)

c(convergence = fit_long$opt$convergence,
  pd_hessian  = isTRUE(fit_long$fit_health$pd_hessian))
cat("long vs wide logLik difference:",
    signif(abs(as.numeric(logLik(fit_long)) -
               as.numeric(logLik(fit_wide))), 3), "\n")
```

The correlated form **fits** and the two surfaces are the same
likelihood problem. Reading the $2 \times 2$
$(\text{intercept}, \text{slope})$ covariance back out -- the separate
$\sigma^2_{\text{int}}$, $\sigma^2_{\text{slope}}$, and their
correlation $\rho$ -- is the natural next extractor, but that reporting
is **not surfaced in this build**; the single-variance
`phylo_slope` route above is the one with a documented variance
read-out. Treat the augmented form here as a fitting demonstration, not
as a covariance-recovery workflow.

```{r augmented-indep, eval = FALSE}
# not yet: phylo_indep(1 + x | sp) (independent / diagonal intercept+slope,
# correlation pinned to 0) is parser-rejected in v0.2.0 -- the richer LHS is
# "not yet supported". Use the correlated phylo_unique(1 + x | sp) above, or
# the single-variance phylo_slope(x | sp), until the diagonal variant lands.
fit_indep <- gllvmTMB(
  traits(t1, t2, t3) ~ 1 + phylo_indep(1 + x | species, tree = tree3),
  data = dw, unit = "species"
)
```

## A note on confidence intervals

Everything above reports **point** estimates. Calibrated confidence
intervals on slope variances are **not** established in v0.2.0. The
package's interval machinery (profile / Wald / bootstrap) is validated
for *point* behaviour, but the coverage study clears the nominal gate
only for narrow Gaussian cases and fails elsewhere, including for
mixed-family targets. Do not read the point estimates here as if they
came with calibrated intervals; for inference on a slope variance,
treat interval width with caution and prefer external validation (e.g.
a Bayesian refit in `MCMCglmm` / `brms`) until the coverage gate is
met. See [profile-likelihood-ci](profile-likelihood-ci.html) for what
the interval tools currently do and do not guarantee.

## What is forthcoming

The random-slope surface is deliberately narrow in v0.2.0. The
following are recognised but **not yet delivered**, and are named here
so the boundary is explicit:

- **Independent (diagonal) intercept + slope**,
  `phylo_indep(1 + x | species)` -- the variance-only counterpart of
  the correlated `phylo_unique(1 + x | species)`, with the
  intercept-slope correlation pinned to zero. The parser does not yet
  accept the richer LHS.
- **Structured slopes** -- `phylo_dep` / `phylo_latent` and the
  `spatial_*` slope modes (`spatial_indep`, `spatial_dep`,
  `spatial_latent`). These need new engine machinery and are in
  progress; only the single-variance and (fit-only) correlated phylo
  forms exist today.
- **Multi-covariate, per-trait, and reduced-rank** phylogenetic slopes
  -- the initial release fixes one covariate, one slope variance,
  slopes shared across traits.

For these, the established external tools for random-regression /
reaction-norm models are
[`MCMCglmm`](https://cran.r-project.org/package=MCMCglmm) (Hadfield
2010, Bayesian), [ASReml-R](https://www.vsni.co.uk/software/asreml-r)
(REML), and [`brms`](https://cran.r-project.org/package=brms).

## References and further reading

- **Meyer & Hill (1997)** *Livest. Prod. Sci.* 47:185 -- estimating
  covariance functions for random-regression (reaction-norm) models.
- **Nussey, Wilson & Brommer (2007)** *J. Evol. Biol.* 20:831 -- the
  evolutionary-ecology framing of reaction norms as random regressions.
- **Hadfield & Nakagawa (2010)** *J. Evol. Biol.* 23:494 -- the sparse
  $\mathbf{A}^{-1}$ representation shared by the `phylo_*` / `animal_*`
  engine path.
- **Kruuk (2004)** *Phil. Trans. R. Soc. B* 359:873 -- the animal model
  in natural populations.

## See also

- [The animal model](animal-model.html) -- pedigree workflow,
  heritability, and the genetic covariance matrix $\mathbf{G}$.
- [Phylogenetic trait covariance](phylogenetic-gllvm.html) -- the
  intercept-level phylogenetic decomposition that `phylo_slope`
  extends.
- [Choose your model](choose-your-model.html) -- the decision tree
  across the covariance-keyword grid.
- [Response families](response-families.html) -- the family catalogue
  the slope term composes with.
- [GLLVM vocabulary](gllvm-vocabulary.html) -- definitions of
  $\mathbf{A}$, $\boldsymbol\Sigma$, and the rotation conventions.
