---
title: "Structured random slopes for non-Gaussian traits"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Structured random slopes for non-Gaussian traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Chunk policy for this article:
#   * The intro / fixed-data chunks (DGP setup, tiny illustrative Gaussian and
#     Poisson fits on SMALL n) run at build time (eval = TRUE) -- each is a few
#     seconds.
#   * The heavy structured-grid fits at validation n are shown eval = FALSE,
#     with precomputed numbers in prose. Rendering the full phylo_dep / spatial
#     grid at the validation fixtures (n_sp up to 300) would OOM pkgdown, so
#     those chunks document the EXACT shipped syntax without being run here.
#     This mirrors missing-data.Rmd (eval = FALSE syntax) and joint-sdm.Rmd
#     (small real fits) per the article gate (#230 / #347).
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = requireNamespace("ape", quietly = TRUE),
  message  = FALSE,
  warning  = FALSE
)
```

A reaction norm is a slope: how a trait responds to an environmental
gradient. The ordinary individual-level behavioural reaction-norm article is
now the public Gaussian RE-12 worked example for
`latent(1 + x | individual, d = K) + unique(1 + x | individual)`, while plain
non-structured bare-bar `(1 + x | individual)` random slopes remain reserved.
This article covers a different, internal technical surface: **structured**
random slopes, where the slope varies across species along a phylogeny or
across sites in space. Those are random regressions with an explicit dependence
structure.

This article covers two directions that matter for ecological and evolutionary
data:

1. **Non-Gaussian responses.** Counts, proportions, bounded scores, and
   ordinal categories are the rule, not the exception, in trait data. The
   structured random-slope grid now covers them.
2. **Structured intercept + slope covariance.** Beyond a single shared
   slope variance, the augmented forms `phylo_dep(1 + x | species)` and
   `spatial_indep(1 + x | coords)` estimate a per-unit random *intercept*
   and random *slope* together, and let you read the
   intercept-slope covariance back out.

The full grid -- every core family (`poisson`, `nbinom1`, `nbinom2`,
`Gamma`, `Beta`, `binomial`, `ordinal_probit`) crossed with the
phylogenetic and spatial sources, across the `indep`, `latent`, and
`dep` correlation modes -- is now validated (gllvmTMB issues
[#388](https://github.com/itchyshin/gllvmTMB/issues/388),
[#392](https://github.com/itchyshin/gllvmTMB/issues/392),
[#422](https://github.com/itchyshin/gllvmTMB/issues/422),
[#424](https://github.com/itchyshin/gllvmTMB/issues/424),
[#427](https://github.com/itchyshin/gllvmTMB/issues/427),
[#429](https://github.com/itchyshin/gllvmTMB/issues/429); `nbinom1`
added in [#441](https://github.com/itchyshin/gllvmTMB/issues/441)). This
article is the worked tour of that grid.

The scope is deliberately single-slope:

| Claim | Status |
|---|---|
| Structured random regression with one covariate, `1 + x` (`s = 1`) | **Covered** for the core phylogenetic and spatial grid (PHY-11..PHY-18, SPA-08..SPA-10, ANI-11..ANI-12) |
| Gaussian `phylo_dep(1 + x1 + x2 | species)` (`s = 2`) | **Covered** under RE-03 for the structured phylogenetic dependent path |
| Non-Gaussian `phylo_dep(..., s >= 2)` | **Partial**; fail-loud guarded while RE-03 diagnostics continue |
| Delta, hurdle, and two-stage zero-inflated families | **Blocked / out of scope** for latent-scale slope covariance (FAM-17, MIX-10) |
| Confidence intervals on slope variances | **Not calibrated here**; this article reports point estimates and recovery bands, not coverage guarantees |

## The augmented-slope grammar

The random-slope term sits inside the RHS of the formula and follows the
same dual-format grammar as every other covariance keyword in the
package.

| Spelling | Meaning |
|---|---|
| <code>phylo_indep(1 + x | species)</code> | Per-species intercept and slope, each correlated by the phylogeny $\mathbf{A}$, **diagonal** between intercept and slope ($\rho$ pinned to 0). |
| <code>phylo_latent(1 + x | species)</code> | Reduced-rank block over the augmented (intercept, slope) columns. |
| <code>phylo_dep(1 + x | species)</code> | **Full unstructured** $2T \times 2T$ covariance over the per-trait (intercept, slope) columns. |
| <code>spatial_indep(1 + x | coords)</code> | Two independent SPDE fields: one for the intercept surface, one for the slope surface. |
| <code>spatial_dep(1 + x | coords)</code> | Full cross-field covariance between the intercept and slope fields. |

In **long format**, the wide shorthand `(1 + x | species)` expands to
the explicit stacked-trait form `0 + trait + (0 + trait):x | species`;
the two are byte-identical (the recovery tests assert this). In **wide
format**, `traits(t1, t2, ...)` on the LHS lets the RHS use the compact
`(1 + x | ...)` shorthand directly.

```{r setup}
library(gllvmTMB)
library(ape)
```

## A small worked fit: Gaussian, both shapes

We start with the lightest possible illustrative fit so the dual-format
machinery is concrete before the heavier grid. Per-species intercepts and
slopes are drawn jointly from
$\mathcal{N}(\mathbf 0,\, \boldsymbol\Sigma_b \otimes \mathbf{A})$ on a
small tree, with a positive intercept-slope correlation
($\rho = 0.5$). This is a syntax-and-recovery illustration on one small
data set, not a simulation study.

```{r sim-small}
set.seed(2026)
n_sp <- 30
tree <- ape::rcoal(n_sp)
tree$tip.label <- paste0("sp", seq_len(n_sp))
Cphy <- ape::vcv(tree, corr = TRUE)
Lphy <- t(chol(Cphy + diag(1e-8, n_sp)))

# True 2x2 (intercept, slope) covariance, shared across traits
s2_int <- 0.4; s2_slope <- 0.3; rho <- 0.5
cov_is <- rho * sqrt(s2_int * s2_slope)
Sigma_b <- matrix(c(s2_int, cov_is, cov_is, s2_slope), 2, 2)

ab <- (Lphy %*% matrix(rnorm(n_sp * 2), n_sp, 2)) %*% chol(Sigma_b)
colnames(ab) <- c("alpha", "beta")
rownames(ab) <- tree$tip.label

# Long-format frame: trait x species x rep, x shared within a (species, rep)
trait_lvls <- paste0("t", 1:3)
sr <- expand.grid(
  species = factor(tree$tip.label, levels = tree$tip.label),
  rep = seq_len(4)
)
sr$x <- rnorm(nrow(sr))
long <- merge(sr, data.frame(trait = factor(trait_lvls, levels = trait_lvls)),
              all = TRUE)
long <- long[order(long$species, long$rep, long$trait), ]
mu_t <- c(2, 1, 0.5)[as.integer(long$trait)]
long$value <- mu_t +
  ab[as.character(long$species), "alpha"] +
  ab[as.character(long$species), "beta"] * long$x +
  rnorm(nrow(long), sd = 0.3)
head(long)
```

The **long** call uses the explicit stacked-trait grammar; the **wide**
call uses the `traits()` LHS shorthand. The tree is passed once at the
top level via `phylo_tree =`. We use the diagonal `phylo_indep` form
here because it is the lightest structured cell that still estimates both
an intercept variance and a slope variance.

```{r fit-small-long}
fit_long <- gllvmTMB(
  value ~ 0 + trait + phylo_indep(1 + x | species),
  data = long, trait = "trait", phylo_tree = tree, unit = "species"
)
```

The wide form is the same model on one-row-per-`(species, rep)` data; we
reshape with `tidyr::pivot_wider` and confirm the two surfaces agree.

```{r fit-small-wide, eval = requireNamespace("tidyr", quietly = TRUE)}
library(tidyr)
wide <- as.data.frame(tidyr::pivot_wider(
  long, id_cols = c(species, rep, x),
  names_from = trait, values_from = value
))

fit_wide <- gllvmTMB(
  traits(t1, t2, t3) ~ 1 + phylo_indep(1 + x | species),
  data = wide, phylo_tree = tree, unit = "species"
)

cat("long vs wide logLik difference:",
    signif(abs(as.numeric(logLik(fit_long)) -
               as.numeric(logLik(fit_wide))), 3), "\n")
```

The two surfaces are the same likelihood problem; the difference is
numerical zero. The rest of the article uses whichever shape reads more
clearly for the point being made, and always shows both spellings for
each structured cell.

### Reading the intercept-slope covariance

The augmented intercept + slope covariance comes back through
`extract_Sigma()`. For the diagonal `phylo_indep` form the off-diagonal
is pinned to zero by construction, so it reports the two marginal
variances; the correlated `phylo_dep` form (below) reports the full
block. The fitted slope standard deviation for this small fit:

```{r recover-small}
# sd_b carries (sd_intercept, sd_slope); cor_b carries rho (0 here for indep)
sd_b <- as.numeric(fit_long$report$sd_b)
data.frame(
  component = c("sd_intercept", "sd_slope"),
  true      = signif(sqrt(c(s2_int, s2_slope)), 3),
  estimated = signif(sd_b, 3)
)
```

### Always check the fit, do not just trust convergence

A non-positive-definite Hessian (`pdHess = FALSE`) is an **uncertainty
warning**, not model death: the point estimates can still be sensible,
but the standard errors around them are not trustworthy. `gllvmTMB`
surfaces this in `fit$fit_health`, and `check_gllvmTMB()` rolls the
convergence, gradient, and Hessian checks into one table you should read
before interpreting any structured-slope fit.

```{r check-small}
isTRUE(fit_long$fit_health$pd_hessian)
check_gllvmTMB(fit_long)
```

If `pd_hessian` is `FALSE`, the usual remedies are more units (more
species, more sites), more replication per unit, a different seed for the
starting values, or stepping down to a lighter correlation mode
(`dep` $\to$ `latent` $\to$ `indep`). Report the point estimate with the
caveat; do not silently drop the fit.

## A non-Gaussian reaction norm: Poisson counts

Count traits are where random regression earns its keep in ecology --
fecundity, abundance, and visit counts all vary in their environmental
response across species. The slope term composes with the Poisson family
exactly as it does with Gaussian: the augmented (intercept, slope)
contribution enters the linear predictor *before* the family dispatch, so
swapping `family =` is the only change.

Here is a small illustrative Poisson fit that runs at build time. We keep
$n$ small and use the diagonal `phylo_indep` cell so it converges in a few
seconds.

```{r poisson-small}
set.seed(5640)
n_sp2 <- 25
tree2 <- ape::rcoal(n_sp2)
tree2$tip.label <- paste0("sp", seq_len(n_sp2))
C2 <- ape::vcv(tree2, corr = TRUE)
L2 <- t(chol(C2 + diag(1e-8, n_sp2)))

# Independent intercept and slope fields (rho = 0), sd_int = 0.6, sd_slope = 0.5
ab2 <- L2 %*% cbind(rnorm(n_sp2, sd = 0.6), rnorm(n_sp2, sd = 0.5))
rownames(ab2) <- tree2$tip.label

dfp <- merge(
  data.frame(species = factor(tree2$tip.label, levels = tree2$tip.label),
             x = rnorm(n_sp2)),
  data.frame(trait = factor(paste0("t", 1:3), levels = paste0("t", 1:3))),
  all = TRUE
)
# replicate to give the counts something to estimate from
dfp <- dfp[rep(seq_len(nrow(dfp)), each = 4), ]
dfp$x <- dfp$x + rnorm(nrow(dfp), sd = 0.2)
eta <- c(1.4, 1.2, 1.0)[as.integer(dfp$trait)] +
  ab2[as.character(dfp$species), 1] +
  ab2[as.character(dfp$species), 2] * dfp$x
dfp$value <- rpois(nrow(dfp), exp(eta))

fit_pois <- gllvmTMB(
  value ~ 0 + trait + phylo_indep(1 + x | species),
  data = dfp, trait = "trait", phylo_tree = tree2, unit = "species",
  family = poisson(link = "log")
)

c(convergence = fit_pois$opt$convergence,
  pd_hessian  = isTRUE(fit_pois$fit_health$pd_hessian))
sd_b_pois <- as.numeric(fit_pois$report$sd_b)
data.frame(
  component = c("sd_intercept", "sd_slope"),
  true      = c(0.6, 0.5),
  estimated = signif(sd_b_pois, 3)
)
```

The slope variance recovers under the count family. The same pattern
holds for the other core families (`nbinom1`, `nbinom2`, `Gamma`,
`Beta`, `binomial`, `ordinal_probit`) -- swap `family =` and keep the
formula. The
[response-families](response-families.html) article is the family
catalogue these slope terms compose with.

## The correlated form: `phylo_dep(1 + x | species)`

The diagonal `phylo_indep` cell above pins the intercept-slope
correlation to zero. The **full unstructured** `phylo_dep` cell estimates
it. This is the canonical evolutionary reaction-norm question: do species
that start high (large intercept) also respond more steeply (large
slope)? The estimand is the $2T \times 2T$ covariance
$\boldsymbol\Sigma_b$ over the per-trait (intercept, slope) columns, and
its intercept-slope blocks carry that correlation.

This cell is heavy at the validation $n$ ($n_{\text{species}}$ in the
low hundreds for the count families). The chunk below shows the **exact
shipped syntax** for both shapes and the extractor call, with
`eval = FALSE`; the recovery numbers quoted afterward come from the
validation tests, not from a build-time fit.

```{r dep-poisson, eval = FALSE}
# Long format: explicit stacked-trait grammar
fit_dep_long <- gllvmTMB(
  value ~ 0 + trait + phylo_dep(1 + x | species),
  data = long_counts, trait = "trait", phylo_tree = tree, unit = "species",
  family = poisson(link = "log")
)

# Wide format: traits() LHS shorthand, same model
fit_dep_wide <- gllvmTMB(
  traits(t1, t2) ~ 1 + phylo_dep(1 + x | species),
  data = wide_counts, phylo_tree = tree, unit = "species",
  family = poisson(link = "log")
)

# Check before interpreting
check_gllvmTMB(fit_dep_long)

# The full augmented (intercept, slope) covariance, interleaved per trait:
#   intercept.t1, slope.t1, intercept.t2, slope.t2
Sigma_aug <- extract_Sigma(fit_dep_long, level = "phy")$Sigma
round(Sigma_aug, 3)

# Slope variances sit on the interleaved diagonal positions 2, 4, ...;
# the intercept-slope correlation is the (1,2) / (3,4) block off-diagonal.
slope_var <- diag(Sigma_aug)[c(2, 4)]
```

In the validation cell (Poisson, `phylo_dep`, $n_{\text{species}} =
150$, 2 traits, 10 reps), the engine converges with a positive-definite
Hessian and recovers the per-trait slope variances inside the
mean-dependent-family band (within a factor of about 4 of truth -- the
honest tolerance for a count family at this fixture). The
`extract_Sigma(level = "phy")$Sigma` matrix carries interleaved dimnames
`intercept.<t>`, `slope.<t>`, so the intercept-slope correlation reads
directly off the $2 \times 2$ block for each trait.

> **Why the band, not a point claim?** Mean-dependent families (Poisson,
> negative binomial, Gamma, Beta) carry an implicit, mean-varying
> latent-scale residual, so the recoverable precision on a slope variance
> is looser than for a Gaussian or fixed-scale family. The validation
> suite uses a 3x-4x variance band for these families and a tighter
> (2.5x-3x) band for the fixed-scale families (binomial, ordinal probit).
> These are recovery tolerances, **not** calibrated confidence intervals
> -- see the CI note below.

## The spatial twin: `spatial_indep(1 + x | coords)`

When the structured units are *locations* rather than *species*, the same
random-regression idea runs over a spatial field. The slope on $x$ varies
smoothly across space, modelled as a second SPDE field alongside the
intercept field. The `spatial_indep` form treats the two fields as
independent; `spatial_dep` estimates their cross-covariance.

This needs a mesh (`make_mesh()`), which makes it heavier than the phylo
cells, so it is shown `eval = FALSE`. The marginal field standard
deviations come back through `extract_Sigma(level = "spatial")`.

```{r spatial-poisson, eval = FALSE}
library(fmesher)

# A mesh over the site coordinates
mesh <- make_mesh(sites, c("lon", "lat"), cutoff = 0.1)

# Long format
fit_sp_long <- gllvmTMB(
  value ~ 0 + trait + spatial_indep(1 + x | coords),
  data = sites_long, trait = "trait", mesh = mesh,
  family = poisson(link = "log")
)

# Wide format
fit_sp_wide <- gllvmTMB(
  traits(t1, t2, t3) ~ 1 + spatial_indep(1 + x | coords),
  data = sites_wide, mesh = mesh,
  family = poisson(link = "log")
)

check_gllvmTMB(fit_sp_long)

# Marginal field SDs for the intercept and slope surfaces (2x2 cross-field)
S_sp <- extract_Sigma(fit_sp_long, level = "spatial")$Sigma
sqrt(diag(S_sp))   # intercept-field SD, slope-field SD
```

In the validation cell (Poisson, `spatial_indep`, 300 sites), the fit
converges with a positive-definite Hessian, the diagonal `indep` contract
holds (the cross-field correlation is pinned to exactly 0), and both the
intercept-field and slope-field BLUPs correlate with their simulated
truth above 0.8. The marginal field standard deviations recover inside
the family band via `extract_Sigma(level = "spatial")`. The
[functional-biogeography](functional-biogeography.html) article develops
the spatial machinery (mesh construction, range and field-SD
interpretation) in full.

## Where to simulate from

For a Gaussian DGP without phylogeny or space, the shipped
simulators give you a faithful stacked-trait long frame to attach a slope
to. [`simulate_unit_trait()`](../reference/simulate_unit_trait.html)
builds the generic `(unit, observation, trait)` cube and
[`simulate_site_trait()`](../reference/simulate_site_trait.html) the
domain-specific `(site, species, trait)` cube with phylogenetic and
spatial structure baked in. For the **slope** term specifically you add
the per-unit slope draw and the `x` covariate by hand, exactly as the
worked chunks above do -- the simulators give you the intercept-level
scaffold, and the reaction-norm slope is the piece you layer on top.

## The full grid at a glance

Every cell below now fits and recovers within its honest band across the
core families. "Indep" pins the intercept-slope (or cross-field)
correlation to zero; "latent" is the reduced-rank block; "dep" is the
full unstructured covariance.

| Source | `indep` | `latent` | `dep` |
|---|---|---|---|
| <code>phylo_*(1 + x | species)</code> | validated | validated | validated |
| <code>spatial_*(1 + x | coords)</code> | validated | validated | validated |

Families covered for each cell: `gaussian`, `poisson`, `nbinom1`,
`nbinom2`, `Gamma`, `Beta`, `binomial` (incl. multi-trial), and
`ordinal_probit`. The deprecated single-variance slope keywords are not used
in new public examples.

This table covers one random-slope covariate. Gaussian
`phylo_dep(1 + x1 + x2 | species)` has its own RE-03 evidence, but the
non-Gaussian `s >= 2` path remains guarded until that separate gate
clears.

## A note on confidence intervals

Everything here reports **point** estimates and **recovery bands**, not
calibrated confidence intervals. The coverage study clears the nominal
gate only for narrow Gaussian cases; for the non-Gaussian structured
slopes it does not yet. Treat the recovery bands above as evidence that
the *point* estimate lands near truth, not as a coverage guarantee. For
inference on a slope variance, prefer external validation (a Bayesian
refit in [`MCMCglmm`](https://cran.r-project.org/package=MCMCglmm) or
[`brms`](https://cran.r-project.org/package=brms)) until the coverage
gate is met. See
[profile-likelihood-ci](profile-likelihood-ci.html) for what the interval
tools currently do and do not guarantee.

## See also

- [Formula keyword grid](api-keyword-grid.html) -- the covariance keyword
  grammar and current scope labels.
- [Response families](response-families.html) -- the family catalogue the
  slope term composes with.
- [Convergence and starting values](convergence-start-values.html) -- fit
  diagnostics before interpreting random-slope covariance.
- [Formula keyword grid](api-keyword-grid.html) -- the full syntax map
  across the covariance-keyword grid.
