---
title: "Behavioural reaction norms with random slopes"
tier: 3 # internal draft until the reader path is plain-language and fully reviewed
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Behavioural reaction norms with random slopes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6.5,
  fig.height = 4.4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

> **Internal draft.** This page is not yet in the public article menu. It is a
> working behavioural-syndrome example for Gaussian individual-level reaction
> norms, and still needs a plain-language reader pass before promotion.

This draft fits an individual-level random-regression GLLVM for a behavioural
syndrome question. The unit is the **individual**, the repeated occasion is
`session_id`, and the continuous context is centred/scaled `temperature`. The
model asks whether individuals differ in both average behaviour and
temperature responsiveness, and whether those reaction-norm coefficients
covary across behaviours.

Here `latent()` estimates the between-individual shared axes and their
diagonal `Psi` companion for trait-specific variation left after those axes.
The current draft covers the Gaussian version only. It does not yet teach
non-Gaussian reaction norms, calibrated intervals for slope summaries, or the
broader model-choice path for deciding when this random-regression model is
appropriate.

## Setup

```{r setup}
library(gllvmTMB)
library(ggplot2)
library(dplyr)
```

## Example data

We use a prepared example object so the article starts with the data,
truth, formulas, and biological question rather than a long simulation
script. The generator is shipped in the repository as
`data-raw/examples/make-behavioural-reaction-norm-example.R`.

```{r example-data, eval = FALSE}
rr <- readRDS(system.file(
  "extdata", "examples", "behavioural-reaction-norm-example.rds",
  package = "gllvmTMB"
))
df <- rr$data_long
df_wide <- rr$data_wide
truth <- rr$truth
```

```{r example-data-load, include = FALSE}
example_path <- system.file(
  "extdata",
  "examples",
  "behavioural-reaction-norm-example.rds",
  package = "gllvmTMB"
)
example_path <- c(
  example_path,
  file.path(
    "inst",
    "extdata",
    "examples",
    "behavioural-reaction-norm-example.rds"
  ),
  file.path(
    "..",
    "inst",
    "extdata",
    "examples",
    "behavioural-reaction-norm-example.rds"
  ),
  file.path(
    "..",
    "..",
    "inst",
    "extdata",
    "examples",
    "behavioural-reaction-norm-example.rds"
  )
)
example_path <- example_path[nzchar(example_path) & file.exists(example_path)][1]
stopifnot(!is.na(example_path))

rr <- readRDS(example_path)
df <- rr$data_long
df_wide <- rr$data_wide
truth <- rr$truth
trait_names <- truth$trait_names
intercept_names <- paste0("intercept.", trait_names)
slope_names <- paste0("slope.temperature.", trait_names)
```

```{r preview}
rr$story$question
head(df, 9)
```

The long data have one row per `(individual, session_id, trait)`
observation. The matching wide data have one row per session, one column
per response trait, and the same `individual`, `session_id`, and
`temperature` columns:

```{r wide-preview}
head(df_wide, 4)
```

Here the three response traits are `boldness`, `exploration`, and
`activity`. The example is deliberately small but not single-level:
`individual` indexes stable among-individual differences, while
`session_id` indexes occasion-level deviations shared by the traits
within the same repeated measurement.

## Fit the model

The same model can be fitted from long or wide data. The long form is
canonical because it makes the `(individual, session_id, trait)` rows
explicit. The wide `traits(...)` form is the compact equivalent for
readers whose data already have one row per repeated session.

```{r fit}
rr$formula_long
rr$formula_wide

ctl <- gllvmTMBcontrol(
  se = FALSE,
  optimizer = "optim",
  optArgs = list(method = "BFGS")
)

fit_long <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):temperature +
    latent(0 + trait + (0 + trait):temperature | individual, d = 1) +
    latent(0 + trait | session_id, d = 1),
  data = df,
  trait = rr$fit_args$trait,
  unit = rr$fit_args$unit,
  unit_obs = rr$fit_args$unit_obs,
  family = rr$fit_args$family,
  control = ctl
)

fit_wide <- gllvmTMB(
  traits(boldness, exploration, activity) ~ 1 + temperature +
    latent(1 + temperature | individual, d = 1) +
    latent(1 | session_id, d = 1),
  data = df_wide,
  unit = rr$fit_args$unit,
  unit_obs = rr$fit_args$unit_obs,
  family = rr$fit_args$family,
  control = ctl
)

fit <- fit_long

cat(
  "Long/wide logLik difference:",
  signif(abs(as.numeric(logLik(fit_long)) - as.numeric(logLik(fit_wide))), 3),
  "\n"
)
```

The `unit = "individual"` argument defines the between-individual tier.
The `unit_obs = "session_id"` argument defines the occasion tier. The
augmented random slope lives at the individual tier, not the session
tier:

```{r formula-pieces, eval = FALSE}
# Long format
latent(0 + trait + (0 + trait):temperature | individual, d = 1)

# Wide data-frame shorthand
latent(1 + temperature | individual, d = 1)
```

The article uses `se = FALSE` so the vignette renders quickly. That
means the diagnostic table below is a point-estimate fit check, not a
curvature-based interval claim.

```{r diagnostics}
diag_tbl <- check_gllvmTMB(fit)
diag_tbl[
  diag_tbl$component %in% c(
    "optimizer_convergence",
    "max_gradient",
    "boundary_flags",
    "boundary_sigma_eps",
    "sdreport",
    "pd_hessian"
  ),
  c("component", "status", "value", "message")
]
```

## What model did we fit?

For trait `t`, individual `i`, and session `j`,

$$
y_{ijt} =
\mu_t + \beta_t\,\mathrm{temperature}_{ij}
+ u_{it} + b_{it}\,\mathrm{temperature}_{ij}
+ e_{ijt} + \varepsilon_{ijt}.
$$

The random intercept `u_it` is the individual's average deviation for
trait `t` at the reference temperature. In animal-personality language,
that is the personality component. The random slope `b_it` is the
individual's deviation in temperature responsiveness. That is the
reaction-norm or plasticity component.

gllvmTMB stacks all trait-specific intercepts and slopes into one
augmented between-individual vector:

$$
\mathbf{b}_{B,i} =
\begin{pmatrix}
u_{i,\mathrm{boldness}} \\
b_{i,\mathrm{boldness}} \\
u_{i,\mathrm{exploration}} \\
b_{i,\mathrm{exploration}} \\
u_{i,\mathrm{activity}} \\
b_{i,\mathrm{activity}}
\end{pmatrix}.
$$

The fitted decomposition is

$$
\mathbf{b}_{B,i}
= \Lambda_{B,\mathrm{aug}} \mathbf{z}_{B,i}
+ \mathbf{q}_{B,i},
\qquad
\mathbf{q}_{B,i} \sim
\mathcal{N}(\mathbf{0}, \mathrm{diag}(\psi_{B,\mathrm{aug}})).
$$

Therefore the augmented reaction-norm covariance is

$$
\Sigma_{B,\mathrm{aug}}
= \Lambda_{B,\mathrm{aug}}\Lambda_{B,\mathrm{aug}}^\top
+ \mathrm{diag}(\psi_{B,\mathrm{aug}}).
$$

This `2T x 2T` covariance is the main estimand in the article.

## Extract the augmented covariance

```{r extract-sigma}
Sigma_shared <- extract_Sigma(
  fit,
  level = "unit_slope",
  part = "shared"
)
Sigma_unique <- extract_Sigma(
  fit,
  level = "unit_slope",
  part = "unique"
)
Sigma_total <- extract_Sigma(
  fit,
  level = "unit_slope",
  part = "total"
)

round(Sigma_total$Sigma[intercept_names, intercept_names], 2)
round(Sigma_total$Sigma[slope_names, slope_names], 3)
round(Sigma_total$Sigma[intercept_names, slope_names], 3)
```

The three blocks answer different biological questions:

- The intercept-intercept block is the personality or behavioural
  syndrome covariance at `temperature = 0`.
- The slope-slope block is the plasticity-syndrome covariance: whether
  individuals that are temperature-responsive in one behaviour are also
  responsive in another.
- The intercept-slope block is the personality-plasticity association:
  whether high-average individuals also respond more strongly or weakly
  to temperature.

The shared and unique pieces compose the total augmented covariance.
`Sigma_unique$s` is the fitted diagonal `Psi` contribution that ordinary
`latent()` carries by default, ordered to match the rows and columns of
`Sigma_total$Sigma`:

```{r composition}
max(abs(
  Sigma_total$Sigma -
    (Sigma_shared$Sigma + diag(Sigma_unique$s, nrow = length(Sigma_unique$s)))
))
```

## Recovery check

Because this is a generated teaching example, we can compare fitted and
true augmented covariance entries. The plot is a diagnostic for this
single fixture, not a calibrated simulation study.

```{r recovery-data}
unit_slope_comparison <- function(estimate, truth) {
  idx <- which(lower.tri(truth, diag = TRUE), arr.ind = TRUE)
  row_name <- rownames(truth)[idx[, "row"]]
  col_name <- colnames(truth)[idx[, "col"]]
  row_is_slope <- grepl("^slope\\.", row_name)
  col_is_slope <- grepl("^slope\\.", col_name)

  data.frame(
    row = row_name,
    column = col_name,
    estimate = estimate[idx],
    truth = truth[idx],
    block = ifelse(
      row_is_slope & col_is_slope,
      "slope-slope",
      ifelse(
        !row_is_slope & !col_is_slope,
        "intercept-intercept",
        "intercept-slope"
      )
    )
  )
}

cov_compare <- unit_slope_comparison(
  estimate = Sigma_total$Sigma,
  truth = truth$Sigma_unit_slope
)

cat(
  "Relative Frobenius error:",
  round(
    norm(Sigma_total$Sigma - truth$Sigma_unit_slope, "F") /
      norm(truth$Sigma_unit_slope, "F"),
    3
  ),
  "\n"
)
```

```{r recovery-plot, fig.width = 6.5, fig.height = 4.8, fig.cap = "Estimated versus true augmented individual-level covariance entries for the behavioural reaction-norm fixture. The dashed line is perfect recovery; colours identify intercept-intercept, slope-slope, and intercept-slope blocks."}
ggplot(cov_compare, aes(truth, estimate, colour = block)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey45") +
  geom_point(size = 2.6, alpha = 0.88) +
  coord_equal() +
  scale_colour_manual(
    values = c(
      "intercept-intercept" = "#1b9e77",
      "intercept-slope" = "#7570b3",
      "slope-slope" = "#d95f02"
    ),
    name = "Block"
  ) +
  labs(
    x = "Known simulation truth",
    y = "Fitted estimate",
    title = "Recovery of the reaction-norm covariance",
    subtitle = "Point estimates from Sigma_B,aug"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## Repeatability along the temperature gradient

With random slopes, among-individual variance is not constant. For one
trait at a focal temperature value `x`,

$$
\mathrm{Var}(u_{it} + b_{it}x)
= \mathrm{Var}(u_{it})
+ 2x\,\mathrm{Cov}(u_{it}, b_{it})
+ x^2\,\mathrm{Var}(b_{it}).
$$

The temperature-specific repeatability is this among-individual
variance divided by among-individual plus occasion-level and Gaussian
residual variance.

```{r repeatability-data}
Sigma_W_hat <- extract_Sigma(fit, level = "unit_obs")$Sigma
sigma_eps2_hat <- as.numeric(fit$report$sigma_eps[1])^2

reaction_repeatability <- function(Sigma_aug, Sigma_W, sigma_eps2,
                                   traits, source) {
  x_grid <- seq(-1.5, 1.5, length.out = 80)
  out <- vector("list", length(traits))
  for (k in seq_along(traits)) {
    trait <- traits[k]
    intercept <- paste0("intercept.", trait)
    slope <- paste0("slope.temperature.", trait)
    between <- Sigma_aug[intercept, intercept] +
      2 * x_grid * Sigma_aug[intercept, slope] +
      x_grid^2 * Sigma_aug[slope, slope]
    total <- between + Sigma_W[trait, trait] + sigma_eps2
    out[[k]] <- data.frame(
      trait = trait,
      temperature = x_grid,
      repeatability = between / total,
      source = source
    )
  }
  do.call(rbind, out)
}

repeatability_rows <- rbind(
  reaction_repeatability(
    Sigma_total$Sigma,
    Sigma_W_hat,
    sigma_eps2_hat,
    trait_names,
    "estimate"
  ),
  reaction_repeatability(
    truth$Sigma_unit_slope,
    truth$Sigma_unit_obs,
    truth$sigma_eps^2,
    trait_names,
    "truth"
  )
)
```

```{r repeatability-plot, fig.width = 7.0, fig.height = 4.8, fig.cap = "Trait-specific repeatability across the temperature gradient. Solid lines are fitted point estimates; dashed lines are the known simulation truth."}
ggplot(
  repeatability_rows,
  aes(temperature, repeatability, colour = trait, linetype = source)
) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c(
      boldness = "#1b9e77",
      exploration = "#d95f02",
      activity = "#7570b3"
    ),
    name = "Trait"
  ) +
  scale_linetype_manual(
    values = c(estimate = "solid", truth = "22"),
    name = NULL
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.25)
  ) +
  labs(
    x = "Temperature (centred and scaled)",
    y = "Repeatability",
    title = "Repeatability changes along the reaction norm",
    subtitle = "Random slopes make among-individual variance context-dependent"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

The repeatability curves are often the most useful biological summary.
They translate the augmented covariance into the familiar question:
how much of the observed trait variation is due to stable individual
differences at each temperature value?

## Current non-Gaussian boundary

The ordinary augmented `latent()` random-regression path in
this article is Gaussian. The low-rank augmented `latent()` path can
currently fit ordinary non-Gaussian random slopes in a smoke-test
regime, but adding the augmented diagonal `unique()` component is still
guarded for non-Gaussian families.

```{r nongaussian-boundary, eval = FALSE}
# This intentionally errors today.
gllvmTMB(
  value ~ 0 + trait + (0 + trait):temperature +
    latent(0 + trait + (0 + trait):temperature | individual, d = 1) +
    unique(0 + trait + (0 + trait):temperature | individual),
  data = count_data_long,
  trait = "trait",
  unit = "individual",
  unit_obs = "session_id",
  family = poisson()
)
#> Error: Augmented ordinary `unique()` random-regression slopes are currently
#> implemented for Gaussian responses only.
```

For a non-Gaussian reaction-norm analysis today, either fit the
augmented `latent()` part without `unique()` and treat it as early
evidence, or stay with the established non-slope ordinary `latent()`
between/within model. Do not switch to `phylo_*()` or `spatial_*()`
syntax unless the scientific model genuinely contains phylogenetic or
spatial dependence.

## What to report

For this model, the primary reportable pieces are:

- the fixed trait intercepts and temperature slopes, which describe
  the population-average reaction norms;
- `extract_Sigma(level = "unit_slope", part = "total")`, which gives
  the full covariance among individual intercepts and slopes;
- the intercept-intercept, slope-slope, and intercept-slope blocks,
  interpreted as personality syndromes, plasticity syndromes, and
  personality-plasticity associations;
- repeatability curves across a biologically meaningful temperature
  range; and
- diagnostics from `check_gllvmTMB()` before interpreting covariance or
  repeatability.

This draft demonstrates the Gaussian random-regression model only. It does not
establish calibrated confidence intervals for slope variances, slope
correlations, intercept-slope correlations, repeatability curves, or
non-Gaussian augmented `unique()` support.

## References

O'Dea, R. E., Noble, D. W. A., and Nakagawa, S. (2022). Unifying
individual differences in personality, predictability and plasticity: a
practical guide. *Methods in Ecology and Evolution* 13, 278-293.

Stamps, J. A. (2016). Individual differences in behavioural
plasticities. *Biological Reviews* 91, 534-567.

Dingemanse, N. J. and Dochtermann, N. A. (2013). Quantifying individual
variation in behaviour: mixed-effect modelling approaches. *Journal of
Animal Ecology* 82, 39-54.
