---
title: "When a rare species breaks the JSDM"
description: "A rare specialist recorded only at one end of a gradient can make a joint species distribution model unfittable. Certify the problem with screen_gllvmTMB(), keep the species with the opt-in MSPL estimator, and know exactly what the penalised fit does and does not give you."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{When a rare species breaks the JSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4.2,
  message = FALSE,
  warning = FALSE
)
```

The species that break a [joint species distribution model](joint-sdm.html)
are usually the ones you care about most. A rare specialist — recorded only
at the warmest sites, only above the treeline, only in the oldest forest —
gives the fixed-effect design a pattern a straight line can split perfectly,
and at that point its maximum-likelihood coefficient has no finite value.
The bad responses are to delete the species quietly, or to publish whatever
number the optimizer happened to stop at. This page shows the packaged
response: **certify the problem, then keep the species with an estimator
whose behaviour at the boundary is explicit.**

The estimator (`estimator = "mspl"`) is opt-in and experimental; ordinary
maximum likelihood stays the default, and nothing switches automatically.
This page is the species-facing door to that machinery — the
[evidence-synthesis article](mspl-binary-jsdm.html) walks the same tools
through their full paces (including the *other* failure mode, runaway latent
loadings, and its different remedy).

# A butterfly guild with one heat specialist

Ninety sites along a temperature gradient, six butterflies. Five are
ordinary; the sixth is present only at the very warmest sites — the
threshold pattern real specialists actually show.

```{r simulate}
library(gllvmTMB)

set.seed(88)
n_site <- 90
species <- c("meadow_brown", "common_blue", "small_heath",
             "grayling", "wall_brown", "silver_studded")
temp <- as.numeric(scale(seq_len(n_site)))
z   <- matrix(rnorm(n_site * 2), n_site, 2)   # two latent site axes
Lam <- matrix(rnorm(12, 0, 0.9), 6, 2)
alpha <- c(0.6, 0.2, 0.4, -0.5, 0.0, -2.4)
beta  <- c(0.3, -0.4, 0.5, 0.8, -0.2, 1.0)

eta <- sweep(z %*% t(Lam), 2, alpha, "+") + outer(temp, beta)
Y <- matrix(rbinom(n_site * 6, 1, plogis(eta)), n_site, 6,
            dimnames = list(NULL, species))
Y[, "silver_studded"] <- as.integer(temp > quantile(temp, 0.92))
sites <- data.frame(site = factor(seq_len(n_site)), temp = temp, Y)
round(colMeans(Y), 2)
```

Note what makes this dangerous: the problem is not rarity itself. A species
at 8% prevalence scattered across the gradient fits fine. This one's
presences all sit past a temperature threshold, so a line on `temp` can
separate every presence from every absence — and that is a property of the
fixed design, diagnosable before any model is fitted.

# Certify the problem before fitting

`screen_gllvmTMB()` runs the formal separation test species by species
(it uses the optional `detectseparation` package):

```{r screen, eval = requireNamespace("detectseparation", quietly = TRUE)}
scr <- screen_gllvmTMB(
  traits(meadow_brown, common_blue, small_heath,
         grayling, wall_brown, silver_studded) ~ 1 + temp,
  data = sites, unit = "site", family = binomial(),
  control = screen_control(separation = "fixed")
)
screen_table(scr, "separation")[, c("traits", "status", "severity",
                                    "separated", "infinite_terms")]
```

```{r screen-missing, eval = !requireNamespace("detectseparation", quietly = TRUE), echo = FALSE}
cat("Install the optional detectseparation package to run this certificate.\n")
```

Five species pass with `overlap`; the silver-studded is flagged `complete`,
and `infinite_terms` names exactly which of its coefficients have no finite
maximum-likelihood estimate. The screen only diagnoses — it never chooses an
estimator for you.

# What happens if you fit anyway

The long-format JSDM call is the same one the
[JSDM guide](joint-sdm.html) uses; `traits()` wide form fits the identical
model.

```{r long}
long <- data.frame(
  site  = rep(sites$site, times = 6),
  trait = factor(rep(species, each = n_site), levels = species),
  temp  = rep(sites$temp, times = 6),
  value = as.vector(Y)
)

form <- value ~ 0 + trait + trait:temp +
  latent(1 | site, d = 2, unique = FALSE)
```

```{r ml-attempt}
fit_ml <- tryCatch(
  suppressWarnings(gllvmTMB(
    form, data = long, unit = "site", trait = "trait",
    family = binomial(), estimator = "ml",
    control = gllvmTMBcontrol(n_init = 1, se = FALSE, warn_runaway = FALSE)
  )),
  error = identity
)
if (inherits(fit_ml, "error")) {
  cat("ML did not return a fitted point:\n", conditionMessage(fit_ml), "\n")
} else {
  co <- coef(fit_ml)
  cat("ML slope for the separated species:",
      round(unname(co[grep("silver_studded:temp", names(co))]), 1), "\n")
}
```

On this draw ordinary maximum likelihood fails to return a usable fit at
all; on other draws the same disease shows up differently — the optimizer
"converges" with an absurd slope instead. Either way, the certificate above
already told you no finite ML estimate exists for those terms; whatever
number does come back for them is an artefact of where the optimizer
stopped, not an estimate.

# Keep the species with MSPL

```{r mspl-fit}
fit_mspl <- suppressWarnings(gllvmTMB(
  form, data = long, unit = "site", trait = "trait",
  family = binomial(), estimator = "mspl",
  control = gllvmTMBcontrol(n_init = 1, se = FALSE)
))
fit_mspl$opt$convergence
```

MSPL adds a sample-size-scaled Jeffreys-prior penalty on the fixed design
(Kosmidis & Firth 2021; Sterzinger & Kosmidis 2023), which guarantees the
separated coefficients a finite maximum. The species stays in the model, and
its fitted relationship to temperature becomes a curve you can actually
plot — steep, as the data demand, but finite:

```{r rare-curve, fig.cap = "Fitted occurrence probability for the separated heat specialist under MSPL, on the response scale with the site-level latent scores excluded. The curve is a penalised point estimate: finite and reportable, with no uncertainty band because interval methods are deliberately withheld for this estimator.", fig.alt = "A single steep S-shaped curve of occurrence probability against temperature, rising from near zero to near one at the warm end of the gradient."}
pred <- predict(fit_mspl, type = "response", re_form = ~ 0)
idx <- long$trait == "silver_studded"
ord <- order(long$temp[idx])
plot(long$temp[idx][ord], pred$est[idx][ord], type = "l", lwd = 2,
     xlab = "temperature (scaled)", ylab = "occurrence probability",
     ylim = c(0, 1))
points(long$temp[idx], long$value[idx], pch = 3, cex = 0.7)
```

# Why not the loading ridge?

The package's other opt-in stabiliser, `gllvmTMBcontrol(loading_ridge = 2)`,
is sometimes reached for here — and it is the wrong tool for this disease.
It penalises the *latent loadings*; separation lives in the *fixed effects*:

```{r ridge-attempt}
fit_ridge <- tryCatch(
  suppressWarnings(gllvmTMB(
    form, data = long, unit = "site", trait = "trait",
    family = binomial(), estimator = "ml",
    control = gllvmTMBcontrol(n_init = 1, se = FALSE, warn_runaway = FALSE,
                              loading_ridge = 2)
  )),
  error = identity
)
if (inherits(fit_ridge, "error")) {
  cat("Ridge fit did not return a fitted point.\n")
} else {
  co <- coef(fit_ridge)
  cat("convergence:", fit_ridge$opt$convergence,
      "| slope for the separated species:",
      round(unname(co[grep("silver_studded:temp", names(co))]), 1), "\n")
}
```

The separated species' slope is still runaway — the ridge changed nothing
about the boundary the certificate diagnosed. (Combining the ridge with
`estimator = "mspl"` is refused with an error; the two remedies target
different diseases and do not compose.) The ridge's own job — taming runaway
*loadings* — is demonstrated in the
[evidence-synthesis article](mspl-binary-jsdm.html).

# What the penalised fit does and does not give you

**It gives you** a finite, reportable point estimate for every coefficient,
a species kept in the joint model rather than silently dropped, and a
defensible sentence for the paper: *"the species' occurrence rises sharply
past the temperature threshold; its coefficient is a penalised point
estimate because the design shows complete separation."*

**It does not give you** ordinary inference. `logLik()`, AIC/BIC,
likelihood-ratio comparisons against the ML fit, standard errors, and
confidence intervals all fail with a typed error rather than returning
uncalibrated numbers — repeated-sampling calibration for this estimator is
still in progress. And the penalised point is not evidence that the
separation disappeared: the certificate's verdict about the *data* still
stands. Report the certificate together with the estimate.

If your fit shows the *other* classic pathology — convergence with latent
loadings at absurd magnitudes — that is a different disease with a different
remedy (an opt-in loading ridge), and the
[evidence-synthesis article](mspl-binary-jsdm.html) demonstrates the two
side by side, including why they do not substitute for one another.

# See also

- [Joint species distribution model](joint-sdm.html) — the JSDM this page's
  workflow protects.
- [Screen binary traits before fitting](pre-fit-response-screening.html) —
  the wider pre-fit screening workflow.
- [Rare items and runaway estimates in Paper × Items evidence synthesis](mspl-binary-jsdm.html)
  — the same estimator in full, with the ridge comparison and all fences.

# References

- Kosmidis I, Firth D (2021). Jeffreys-prior penalty, finiteness and
  shrinkage in binomial-response generalized linear models. *Biometrika*
  108:71–82. <https://doi.org/10.1093/biomet/asaa052>
- Sterzinger P, Kosmidis I (2023). Maximum softly-penalized likelihood for
  mixed effects logistic regression. *Statistics and Computing* 33:53.
  <https://doi.org/10.1007/s11222-023-10217-3>
