---
title: "Response families"
tier: 2 # technical reference for the multivariate family surface
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Response families}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This reference is for readers who already know the covariance
structure they want and need to choose the response likelihood. The
family controls the link function, the distributional variance, and
the link-scale residual term that `extract_Sigma()` and
`extract_Omega()` use when reporting non-Gaussian covariance.

> **Scope boundary.** This page is a technical lookup for families that
> the multivariate engine maps through `family_to_id()` and for exported
> constructors that are intentionally not advertised as current
> multivariate examples. Gaussian, binomial links, Poisson, and NB2 have
> covered validation rows (FAM-01--FAM-04, FAM-06, FAM-08). The
> beta-binomial, Gamma, beta, lognormal, Student-t, Tweedie, ordinal
> probit, and truncated-count rows are partial at the depth shown here
> (FAM-05, FAM-09--FAM-15). Standard `delta_lognormal()` and
> `delta_gamma()` hurdle fits are engine-mapped, but two-part
> response-scale correlations remain outside this article's advertised
> interpretation boundary (FAM-17), and mixed-family delta/hurdle
> latent-scale correlations (MIX-10) are designed and route-only --
> uncalibrated (CI-08/CI-10) -- rather than an interpretation readers can
> use today. Exported constructors that are not mapped by
> `family_to_id()` remain outside the public multivariate engine surface
> until their validation rows move (FAM-07, FAM-16, FAM-18, FAM-19).

The current multivariate engine recognises 15 family entries. The
source of truth is the `family_to_id()` mapping in `R/fit-multi.R`;
families not mapped there should be treated as unsupported in
`gllvmTMB()` multivariate fits until a later implementation and
simulation check say otherwise.

## Quick Lookup

| Response looks like | Family | Link(s) currently accepted by the multivariate engine | Validation / interpretation status | Notes |
|---|---|---|---|---|
| continuous, unbounded | `gaussian()` | identity | covered (FAM-01) | Default family. |
| binary or binomial trials | `binomial()` | logit, probit, cloglog | covered (FAM-02--FAM-04) | Use `cbind(success, failure)` or `weights` for multi-trial binomial data. |
| counts | `poisson()` | log | covered (FAM-06) | No dispersion parameter; add an observation-level `indep()` diagonal term only when that is identifiable. |
| positive continuous, right-skewed | `lognormal()` | log | partial (FAM-11) | Use only when the response is strictly positive. |
| positive continuous, constant CV | `Gamma(link = "log")` | log | partial (FAM-09) | Base R `Gamma()`, with log link in the engine. |
| overdispersed counts | `nbinom2()` | log | covered (FAM-08) | NB2 variance grows quadratically with the mean. |
| non-negative continuous with zeros | `tweedie()` | log | partial (FAM-13) | Compound Poisson-Gamma style mean-variance relationship. |
| proportions in `(0, 1)` | `Beta()` | logit | partial (FAM-10) | Boundary 0 and 1 values need separate handling before fitting. |
| overdispersed binomial trials | `betabinomial()` | logit | partial (FAM-05) | For k-of-n responses with extra-binomial variation. |
| continuous with heavy tails | `student()` | identity | partial (FAM-12) | Degrees of freedom can be estimated or fixed. |
| positive counts, no zeros | `truncated_poisson()` | log | partial (FAM-15) | Requires integer responses `y >= 1`. |
| overdispersed positive counts, no zeros | `truncated_nbinom2()` | log | partial (FAM-15) | Requires integer responses `y >= 1`. |
| non-negative continuous with exact zeros | `delta_lognormal()` | logit + log | undefined for response-scale interpretation (FAM-17); mixed-family latent-scale correlation is designed and route-only but uncalibrated (MIX-10, `partial`) | Standard hurdle form only; report model/link-scale correlations unless a later response-scale estimand is defined. |
| non-negative continuous with exact zeros | `delta_gamma()` | logit + log | undefined for response-scale interpretation (FAM-17); mixed-family latent-scale correlation is designed and route-only but uncalibrated (MIX-10, `partial`) | Standard hurdle form only; report model/link-scale correlations unless a later response-scale estimand is defined. |
| ordered categories with `K >= 3` | `ordinal_probit()` | probit | covered (FAM-14); article still internal | Threshold model with latent residual variance fixed at 1; the ordinal worked article is under audit for extractor depth, intervals, and public placement. |

Use this table as a likelihood lookup. For covariance keywords, see
[Formula keyword grid](api-keyword-grid.html). In that grid, `indep()` is the
documentary standalone diagonal spelling (`FG-07`). Binary species,
mixed-family, and other non-Gaussian worked examples are under audit until
their example objects, diagnostics, and validation caveats are ready.

## Single-Family Fits

A single `family = ...` value applies the same response family to all
rows. Long data remain the canonical form:

```{r}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | individual, d = 2),
  data = df_long,
  trait = "trait",
  unit = "individual",
  family = gaussian()
)
```

The wide data-frame path reaches the same engine through
`traits(...)`. Use the compact wide syntax when your data already
have one row per unit and one column per trait:

```{r}
fit_wide <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~
    1 +
    latent(1 | individual, d = 2),
  data = df_wide,
  unit = "individual",
  family = gaussian()
)
```

Only the `family` argument changes when you move from Gaussian to
another supported likelihood. The covariance keyword grammar stays
the same, but the diagonal keyword choice should still match the
model interpretation: ordinary `latent()` already includes the diagonal
Psi component by default; use `indep()` for standalone marginal diagonal tiers
and identifiable observation-level random effects.

## Mixed-Family Fits

Mixed-family fits use long data with one row per observation and a
column that selects the family for that row. Pass a list of family
objects and set `attr(fam, "family_var")` to the selector column.

```{r}
df_long$family <- factor(
  df_long$family,
  levels = c("continuous", "presence", "count")
)

fam <- list(gaussian(), binomial(), poisson())
attr(fam, "family_var") <- "family"

fit_mixed <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2),
  data = df_long,
  trait = "trait",
  unit = "site",
  family = fam
)
```

The list order must match the factor levels in the selector column.
If the selector is a character column, the engine sorts the unique
values before matching them, so an explicit factor is safer.

This page is a family lookup, not the mixed-response worked example.
The separate mixed-response article can show a runnable Gaussian +
binomial + Poisson analysis with simulated data and interpretation.
Use non-delta, one-link families for public mixed-family examples for
now. Delta and hurdle families are deliberately not used in the
mixed-family example path here: MIX-10 is designed and route-only
(latent structure constrained to the positive/continuous submodel,
with the occurrence submodel fixed-effects-only), but it is not
advertised: the positive-part latent-scale correlation is route-only
and its coverage is uncalibrated (CI-08/CI-10).

## Scale And Extraction

`extract_Sigma()` reports trait covariance on the latent or link
scale. For Gaussian responses, there is no distribution-specific
latent residual term beyond the modelled covariance. For binomial,
Poisson, Tweedie, beta-binomial, delta, and ordinal responses, the
family determines the residual term that makes the reported
covariance interpretable on the link scale.

That is why family choice is not cosmetic. A Poisson count model and
a Gaussian model can share the same ordinary `latent()` formula but
will not imply the same residual scale. Use `extract_Omega()` and
`extract_Sigma()` after fitting, and read the extractor output on the
scale implied by the family.

Correlations need one extra sentence of caution. `extract_correlations()`
reports correlations in the fitted covariance tier, not correlations
of the raw observed responses. For single-link families this is usually
the quantity readers want: a latent-liability or link-scale trait
correlation. For two-part families such as `delta_lognormal()` and
`delta_gamma()`, there are several possible scientific targets:
presence/absence correlation, positive-response correlation, and total
observed-response correlation. The current engine fits those families
with one shared linear predictor, so the covariance tier is still
well-defined, but the link-residual adjustment used by
`extract_Sigma(link_residual = "auto")` is an approximate diagonal
scale correction, not a full observed-scale two-part correlation
calculus.

The practical rule is to report two-part family correlations as
latent/link-scale model correlations unless a future article or
extractor explicitly defines a response-scale estimand. Do not describe
them as raw biomass, abundance, or total-response correlations without
that extra definition.

## Exported But Not Engine-Mapped

`R/families.R` also contains exported constructor helpers that are not
currently mapped by `family_to_id()` for multivariate `gllvmTMB()`
fits. Do not present these as supported response families in public
examples until the engine, tests, and documentation are updated
together.

| Constructor | Register status | Current status in multivariate `gllvmTMB()` |
|---|---|---|
| `nbinom1()` | partial (FAM-07) | exported constructor, not mapped in `family_to_id()` |
| `gengamma()` | blocked (FAM-19) | exported constructor, not mapped in `family_to_id()` |
| `gamma_mix()` | blocked (FAM-18) | exported constructor, not mapped in `family_to_id()` |
| `lognormal_mix()` | blocked (FAM-18) | exported constructor, not mapped in `family_to_id()` |
| `nbinom2_mix()` | blocked (FAM-18) | exported constructor, not mapped in `family_to_id()` |
| `truncated_nbinom1()` | partial (FAM-15) | exported constructor, not mapped in `family_to_id()` |
| `censored_poisson()` | partial (FAM-16) | exported constructor, not mapped in `family_to_id()` |
| `delta_beta()` | blocked (FAM-17) | exported constructor, not mapped in `family_to_id()` |
| `delta_gengamma()` | blocked (FAM-17) | exported constructor, not mapped in `family_to_id()` |
| `delta_gamma_mix()` | blocked (FAM-17, FAM-18) | exported constructor, not mapped in `family_to_id()` |
| `delta_lognormal_mix()` | blocked (FAM-17, FAM-18) | exported constructor, not mapped in `family_to_id()` |
| `delta_truncated_nbinom1()` | blocked (FAM-17) | exported constructor, not mapped in `family_to_id()` |
| `delta_truncated_nbinom2()` | blocked (FAM-17) | exported constructor, not mapped in `family_to_id()` |
| `delta_poisson_link_gamma()` | blocked compatibility path (FAM-17) | deprecated compatibility helper |
| `delta_poisson_link_lognormal()` | blocked compatibility path (FAM-17) | deprecated compatibility helper |

The practical rule is simple: if a family is in the quick-lookup table
above, it is a documented multivariate engine family. If it appears
only in this unsupported table, treat it as roadmap or compatibility
surface for now.
