---
title: "Response families"
tier: 2 # technical reference for the multivariate family surface
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Response families}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This reference is for readers who already know the covariance
structure they want and need to choose the response likelihood. The
family controls the link function, the distributional variance, and
the link-scale residual term that `extract_Sigma()` and
`extract_Omega()` use when reporting non-Gaussian covariance.

The current multivariate engine recognises 15 family entries. The
source of truth is the `family_to_id()` mapping in `R/fit-multi.R`;
families not mapped there should be treated as unsupported in
`gllvmTMB()` multivariate fits until a later implementation and
simulation check say otherwise.

## Quick Lookup

| Response looks like | Family | Link(s) currently accepted by the multivariate engine | Notes |
|---|---|---|---|
| continuous, unbounded | `gaussian()` | identity | Default family. |
| binary or binomial trials | `binomial()` | logit, probit, cloglog | Use `cbind(success, failure)` or `weights` for multi-trial binomial data. |
| counts | `poisson()` | log | No dispersion parameter; add an observation-level `unique()` term when that is identifiable. |
| positive continuous, right-skewed | `lognormal()` | log | Use only when the response is strictly positive. |
| positive continuous, constant CV | `Gamma(link = "log")` | log | Base R `Gamma()`, with log link in the engine. |
| overdispersed counts | `nbinom2()` | log | NB2 variance grows quadratically with the mean. |
| non-negative continuous with zeros | `tweedie()` | log | Compound Poisson-Gamma style mean-variance relationship. |
| proportions in `(0, 1)` | `Beta()` | logit | Boundary 0 and 1 values need separate handling before fitting. |
| overdispersed binomial trials | `betabinomial()` | logit | For k-of-n responses with extra-binomial variation. |
| continuous with heavy tails | `student()` | identity | Degrees of freedom can be estimated or fixed. |
| positive counts, no zeros | `truncated_poisson()` | log | Requires integer responses `y >= 1`. |
| overdispersed positive counts, no zeros | `truncated_nbinom2()` | log | Requires integer responses `y >= 1`. |
| non-negative continuous with exact zeros | `delta_lognormal()` | logit + log | Standard hurdle form only. |
| non-negative continuous with exact zeros | `delta_gamma()` | logit + log | Standard hurdle form only. |
| ordered categories with `K >= 3` | `ordinal_probit()` | probit | Threshold model with latent residual variance fixed at 1; see [Ordinal-probit threshold traits](ordinal-probit.html). |

Use this table as a likelihood lookup. For covariance keywords, see
[Formula keyword grid](api-keyword-grid.html). For binary species
data and non-Gaussian covariance interpretation, see
[`joint-sdm`](joint-sdm.html).

## Single-Family Fits

A single `family = ...` value applies the same response family to all
rows. Long data remain the canonical form:

```{r}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    latent(0 + trait | individual, d = 2) +
    unique(0 + trait | individual),
  data = df_long,
  unit = "individual",
  family = gaussian()
)
```

The wide data-frame path reaches the same engine through
`traits(...)`. Use the compact wide syntax when your data already
have one row per unit and one column per trait:

```{r}
fit_wide <- gllvmTMB(
  traits(length, mass, wing, tarsus, bill) ~
    1 +
    latent(1 | individual, d = 2) +
    unique(1 | individual),
  data = df_wide,
  unit = "individual",
  family = gaussian()
)
```

Only the `family` argument changes when you move from Gaussian to
another supported likelihood. The covariance keyword grammar stays
the same.

## Mixed-Family Fits

Mixed-family fits use long data with one row per observation and a
column that selects the family for that row. Pass a list of family
objects and set `attr(fam, "family_var")` to the selector column.

```{r}
df_long$family <- factor(
  df_long$family,
  levels = c("continuous", "presence", "count")
)

fam <- list(gaussian(), binomial(), poisson())
attr(fam, "family_var") <- "family"

fit_mixed <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | site, d = 2),
  data = df_long,
  unit = "site",
  family = fam
)
```

The list order must match the factor levels in the selector column.
If the selector is a character column, the engine sorts the unique
values before matching them, so an explicit factor is safer.

This page is a family lookup, not the mixed-response worked example.
The separate mixed-response article can show a runnable Gaussian +
binomial + Poisson analysis with simulated data and interpretation.

## Scale And Extraction

`extract_Sigma()` reports trait covariance on the latent or link
scale. For Gaussian responses, there is no distribution-specific
latent residual term beyond the modelled covariance. For binomial,
Poisson, Tweedie, beta-binomial, delta, and ordinal responses, the
family determines the residual term that makes the reported
covariance interpretable on the link scale.

That is why family choice is not cosmetic. A Poisson count model and
a Gaussian model can share the same `latent() + unique()` formula but
will not imply the same residual scale. Use `extract_Omega()` and
`extract_Sigma()` after fitting, and read the extractor output on the
scale implied by the family.

Correlations need one extra sentence of caution. `extract_correlations()`
reports correlations in the fitted covariance tier, not correlations
of the raw observed responses. For single-link families this is usually
the quantity readers want: a latent-liability or link-scale trait
correlation. For two-part families such as `delta_lognormal()` and
`delta_gamma()`, there are several possible scientific targets:
presence/absence correlation, positive-response correlation, and total
observed-response correlation. The current engine fits those families
with one shared linear predictor, so the covariance tier is still
well-defined, but the link-residual adjustment used by
`extract_Sigma(link_residual = "auto")` is an approximate diagonal
scale correction, not a full observed-scale two-part correlation
calculus.

The practical rule is to report two-part family correlations as
latent/link-scale model correlations unless a future article or
extractor explicitly defines a response-scale estimand. Do not describe
them as raw biomass, abundance, or total-response correlations without
that extra definition.

## Exported But Not Engine-Mapped

`R/families.R` also contains exported constructor helpers that are not
currently mapped by `family_to_id()` for multivariate `gllvmTMB()`
fits. Do not present these as supported response families in public
examples until the engine, tests, and documentation are updated
together.

| Constructor | Current status in multivariate `gllvmTMB()` |
|---|---|
| `nbinom1()` | exported constructor, not mapped in `family_to_id()` |
| `gengamma()` | exported constructor, not mapped in `family_to_id()` |
| `gamma_mix()` | exported constructor, not mapped in `family_to_id()` |
| `lognormal_mix()` | exported constructor, not mapped in `family_to_id()` |
| `nbinom2_mix()` | exported constructor, not mapped in `family_to_id()` |
| `truncated_nbinom1()` | exported constructor, not mapped in `family_to_id()` |
| `censored_poisson()` | exported constructor, not mapped in `family_to_id()` |
| `delta_beta()` | exported constructor, not mapped in `family_to_id()` |
| `delta_gengamma()` | exported constructor, not mapped in `family_to_id()` |
| `delta_gamma_mix()` | exported constructor, not mapped in `family_to_id()` |
| `delta_lognormal_mix()` | exported constructor, not mapped in `family_to_id()` |
| `delta_truncated_nbinom1()` | exported constructor, not mapped in `family_to_id()` |
| `delta_truncated_nbinom2()` | exported constructor, not mapped in `family_to_id()` |
| `delta_poisson_link_gamma()` | deprecated compatibility helper |
| `delta_poisson_link_lognormal()` | deprecated compatibility helper |

The practical rule is simple: if a family is in the quick-lookup table
above, it is a documented multivariate engine family. If it appears
only in this unsupported table, treat it as roadmap or compatibility
surface for now.
