---
title: "Choose a response family"
description: "Match the response support and sampling process to a family before adding multivariate covariance, and stop at the documented specialist boundaries."
tier: 2
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Choose a response family}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

<style type="text/css">
.family-lookup {
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.family-lookup table {
  min-width: 680px;
}
</style>

Choose a family from the values the response can take **and** the process that
generated those values. The family sets the likelihood and link; covariance
terms such as `latent()` or `indep()` describe dependence among responses.
Changing the family can therefore require different response coding,
diagnostics, and interpretation—not merely a different argument value.

Start with the simplest scientifically defensible family, inspect fit and
predictive diagnostics, and change it when the observed mean--variance or zero
pattern contradicts its assumptions.

## Start from the response

### Continuous responses

<div class="table-responsive family-lookup">

| Observed values and sampling process | Family and link | Main assumption or warning |
|---|---|---|
| Any real value; roughly symmetric conditional errors | `gaussian()`; identity | Residual variation is Gaussian on the response scale. |
| Any real value; occasional genuine heavy-tailed observations | `student()`; identity | `df = NULL` estimates degrees of freedom; `student(df = 3)` fixes them. The fitted scale is not the response SD, and the variance is infinite when `df <= 2`. |
| Strictly positive; multiplicative errors are plausible | `lognormal()`; log | The linear predictor describes the mean of `log(y)`, so `exp(eta)` is the conditional median, not the arithmetic mean. |
| Strictly positive; variance grows approximately with the square of the mean | `Gamma(link = "log")`; log | Uses a mean--shape parameterization with constant conditional coefficient of variation within a trait. |
| Non-negative continuous values with exact zeros from one compound process | `tweedie()`; log | Assumes `Var(Y) = phi * mu^p`, with `1 < p < 2`. It is a one-likelihood alternative to an explicit occurrence/positive decomposition. |
| Non-negative continuous values with a scientifically meaningful occurrence process and positive part | `delta_lognormal()` or `delta_gamma()`; logit + log | The current standard hurdle model uses the **same** linear predictor for occurrence and positive mean. See the specialist boundary below before using covariance terms. |

</div>

`Gamma()` and `lognormal()` can both fit right-skewed positive data. Choose from
the sampling mechanism and residual pattern rather than the histogram alone:
lognormal errors are additive after taking logs, whereas Gamma variation is
described through a conditional coefficient of variation.

For exact zeros, ask whether one compound process is plausible or whether zero
versus positive observations represent distinct biological events. That
scientific distinction is more important than which likelihood gives the
smallest AIC in one dataset.

### Binary, trial, proportion, and ordinal responses

<div class="table-responsive family-lookup">

| Observed values and sampling process | Family and link | Main assumption or warning |
|---|---|---|
| A 0/1 event or successes out of known trials | `binomial()`; logit, probit, or cloglog | Use `cbind(successes, failures)` for trials. A flat response plus `weights = n_trials` is also accepted. |
| Successes out of known trials with extra-binomial variation | `betabinomial()`; logit | Uses the same trial encoding as binomial data and estimates a precision parameter for each trait. |
| A continuous proportion strictly inside `(0, 1)` | `Beta()`; logit | Genuine zeros or ones are not Beta observations; do not silently nudge them inward without a scientific measurement model. |
| Ordered categories coded consecutively as `1, ..., K`, with `K >= 3` | `ordinal_probit()`; probit | Category order is part of the model. For two categories, use `binomial(link = "probit")`. Recover estimated thresholds with `extract_cutpoints()`. |

</div>

`ordinal_probit()` uses an underlying liability

\[
y^* = \eta + \varepsilon, \qquad \varepsilon \sim N(0, 1),
\]

and records category \(k\) when \(\tau_{k-1} < y^* \leq \tau_k\).
`gllvmTMB` fixes \(\tau_1 = 0\) as its location convention and estimates the
remaining \(K-2\) free boundaries. `extract_cutpoints(fit)` returns those
boundaries from `cutpoint_2` onward. A cutpoint is a boundary between adjacent
categories on the latent probit scale, not a regression effect or an observed
category-unit change. A predictor shifts the same latent liability across all
boundaries, which is the proportional-threshold assumption.

Prepare ordinal responses deliberately. Use meaningful ordered categories
coded as consecutive integers `1, ..., K`; a numeric code does not make a
nominal category ordinal. The current engine infers \(K\) from the largest
observed code for each trait, so the highest intended category must occur in
the fitted data and gaps must be resolved before fitting. Two-category input is
accepted as the probit special case, but new binary analyses should use
`binomial(link = "probit")` directly.

This lookup stops at family choice and cutpoint interpretation. Ordinal
variance, correlation, or heritability requires a replicated or externally
structured design that identifies the relevant variance tier; a converged fit
with one random effect per observation is not evidence that such a component
has been recovered.

For multi-trial binomial and beta-binomial data, `weights` means trial counts.
For other families, `weights` are likelihood multipliers. Prefer the explicit
`cbind(successes, failures)` form when possible because it makes the denominator
visible in the formula.

### Counts

<div class="table-responsive family-lookup">

| Sampling process | Family and link | Conditional variance |
|---|---|---|
| Counts with no remaining dispersion after the mean structure and random effects | `poisson()`; log | `Var(Y) = mu` |
| Overdispersion that grows approximately linearly with the mean | `nbinom1()`; log | `Var(Y) = mu * (1 + phi)` |
| Overdispersion that grows approximately quadratically with the mean | `nbinom2()`; log | `Var(Y) = mu + mu^2 / phi` |
| Positive counts because zero cannot be observed by design | `truncated_poisson()` or `truncated_nbinom2()`; log | Conditions on `Y >= 1`; every response must be a positive integer. |

</div>

Do not choose a zero-truncated family merely because zeros were deleted during
cleaning. It is appropriate only when the observation process could not produce
or record a zero. Likewise, an observation-level random effect is not an
automatic repair for Poisson overdispersion: first check whether the missing
structure is better represented by NB1/NB2, a predictor, a grouping term, or a
different sampling process.

## One family in long and wide data

A single family applies to every response in the fit. The long and
`traits(...)` wide calls below fit the same Poisson model; only the data shape
and formula shorthand differ.

```{r count-data}
library(gllvmTMB)

set.seed(20260711)
n_unit <- 40L
trait_names <- c("beetles", "moths", "spiders")
unit <- factor(seq_len(n_unit))
z <- rnorm(n_unit, sd = 0.45)

eta <- cbind(
  beetles = 0.2 + 0.8 * z,
  moths = 0.6 - 0.4 * z,
  spiders = 0.1 + 0.5 * z
)
Y <- matrix(
  rpois(length(eta), lambda = exp(eta)),
  nrow = n_unit,
  dimnames = list(NULL, trait_names)
)

counts_wide <- data.frame(unit = unit, Y, check.names = FALSE)
counts_long <- data.frame(
  unit = rep(unit, each = length(trait_names)),
  trait = factor(rep(trait_names, times = n_unit), levels = trait_names),
  count = as.vector(t(Y))
)
```

```{r count-fits}
fit_long <- gllvmTMB(
  count ~ 0 + trait + latent(0 + trait | unit, d = 1),
  data = counts_long,
  trait = "trait",
  unit = "unit",
  family = poisson(),
  silent = TRUE
)

fit_wide <- gllvmTMB(
  traits(beetles, moths, spiders) ~ 1 + latent(1 | unit, d = 1),
  data = counts_wide,
  unit = "unit",
  family = poisson(),
  silent = TRUE
)

data.frame(
  input = c("long", "wide"),
  logLik = c(as.numeric(logLik(fit_long)), as.numeric(logLik(fit_wide)))
)
```

Equivalent long and wide designs should give the same likelihood. This check
does not show that Poisson is the right family; that requires diagnostics and a
scientific mean--variance argument.

## Overdispersed counts and the NB2 dispersion

When counts vary more than a Poisson mean allows even after the mean structure
and covariance terms, `nbinom2()` adds one dispersion parameter per trait. Its
conditional variance is

\[
\operatorname{Var}(Y) = \mu + \frac{\mu^2}{\phi},
\]

so \(\phi\) controls the quadratic extra-Poisson variance: smaller \(\phi\)
means stronger overdispersion, and as \(\phi \to \infty\) the family collapses
back to Poisson. This \(\phi\) is the negative binomial `size` parameter on the
natural scale—the value you would pass to `rnbinom(mu = , size = )`—not a
variance, a standard deviation, or a log-scale quantity.

The simulation below draws three insect counts per site from that exact
mean--variance relationship, sharing one latent axis across traits:

```{r nb2-data}
set.seed(20260817)
n_unit_nb <- 200L
nb_trait_names <- c("aphids", "thrips", "weevils")
nb_unit <- factor(seq_len(n_unit_nb))
z_nb <- rnorm(n_unit_nb, sd = 0.45)

eta_nb <- cbind(
  aphids = 1.0 + 0.8 * z_nb,
  thrips = 1.4 - 0.4 * z_nb,
  weevils = 0.8 + 0.5 * z_nb
)
phi_true <- c(aphids = 1.5, thrips = 2, weevils = 1.2)
Y_nb <- matrix(
  rnbinom(length(eta_nb),
          mu = exp(eta_nb),
          size = rep(phi_true, each = n_unit_nb)),
  nrow = n_unit_nb,
  dimnames = list(NULL, nb_trait_names)
)

nb_wide <- data.frame(unit = nb_unit, Y_nb, check.names = FALSE)
nb_long <- data.frame(
  unit = rep(nb_unit, each = length(nb_trait_names)),
  trait = factor(rep(nb_trait_names, times = n_unit_nb),
                 levels = nb_trait_names),
  count = as.vector(t(Y_nb))
)
```

The long and wide calls change only `family = nbinom2()` relative to the
Poisson fits above:

```{r nb2-fits}
fit_nb_long <- gllvmTMB(
  count ~ 0 + trait + latent(0 + trait | unit, d = 1),
  data = nb_long,
  trait = "trait",
  unit = "unit",
  family = nbinom2(),
  silent = TRUE
)

fit_nb_wide <- gllvmTMB(
  traits(aphids, thrips, weevils) ~ 1 + latent(1 | unit, d = 1),
  data = nb_wide,
  unit = "unit",
  family = nbinom2(),
  silent = TRUE
)

data.frame(
  input = c("long", "wide"),
  logLik = c(as.numeric(logLik(fit_nb_long)), as.numeric(logLik(fit_nb_wide)))
)
```

The fitted dispersions live in `fit$report$phi_nbinom2`, one value per trait in
trait factor-level order, already on the natural \(\phi\) scale:

```{r nb2-dispersion}
data.frame(
  trait = nb_trait_names,
  phi_true = as.numeric(phi_true),
  phi_hat = round(as.numeric(fit_nb_long$report$phi_nbinom2), 2)
)
```

Read these against the variance formula, not as effect sizes: `aphids` with
\(\hat\phi \approx 2\) has extra variance \(\mu^2 / 2\) at mean \(\mu\), which
at \(\mu = 5\) already exceeds the Poisson variance. The estimates sit somewhat
above the simulated values because the `latent()` tier—whose default per-trait
`Psi` acts like an observation-level variance here—absorbs part of the
extra-Poisson variation, leaving less for \(\phi\). The two components compete,
and that competition has a failure mode worth recognizing: when the data carry
little quadratic overdispersion beyond what the covariance terms explain,
\(\phi\) is weakly identified and can run away to a huge value (`1e6` or more).
A runaway \(\hat\phi\) means "this trait looks Poisson after the random
effects", not a precise estimate; refit with `poisson()` and compare rather
than reporting the number.

Exact randomized-quantile residuals are available for NB2 rows and return
standard-normal residuals when the family, link, and mean model are adequate:

```{r nb2-residuals}
r_nb <- residuals(fit_nb_long, type = "randomized_quantile", seed = 1)
round(c(mean = mean(r_nb$residual), sd = sd(r_nb$residual)), 2)
```

The `residual` column should look like draws from \(N(0, 1)\); a mean near 0
and a standard deviation near 1, with no trend against fitted means, is
consistent with the NB2 mean--variance assumption. Systematic departure—for
example residual variance still growing with the mean—points back to the
family table above.

This example demonstrates point estimation and fitted diagnostics only.
Interval coverage for negative binomial fits has not been certified, and
NB2 latent models at small sample sizes have documented dispersion failures;
see [Current limitations and boundaries](current-limits.html) before reporting
uncertainty statements from a fit like this one.

## Gamma's `phi_gamma` is a shape, not a dispersion

Reach for `Gamma(link = "log")` when a positive response has a roughly
constant coefficient of variation as its mean changes. The trap:
`fit$report$phi_gamma` is the **shape** of the mean-shape parameterization,
not a variance or an SD, and a *larger* value means *less* relative
variability. Convert with `scale = mu / phi_gamma` and
`CV(y) = 1 / sqrt(phi_gamma)`.

```{r two-trait-helper}
# Shared plumbing for the compact fits below: two traits per unit in long
# format, the `value ~ 0 + trait + latent(0 + trait | unit)` shape used
# throughout this article.
two_trait_long <- function(unit, trait_names, y_matrix, response = "value") {
  out <- data.frame(
    unit = rep(unit, each = 2),
    trait = factor(rep(trait_names, times = length(unit)), levels = trait_names)
  )
  out[[response]] <- as.vector(t(y_matrix))
  out
}
```

```{r gamma-fit}
set.seed(20260818)
n_gam <- 150L
gam_traits <- c("leaf_area", "root_mass")
z_gam <- rnorm(n_gam, sd = 0.4)
eta_gam <- cbind(leaf_area = 1.5 + 0.6 * z_gam, root_mass = 1.0 - 0.3 * z_gam)
shape_true_gam <- c(leaf_area = 6, root_mass = 3)
Y_gam <- matrix(
  rgamma(length(eta_gam), shape = rep(shape_true_gam, each = n_gam),
         scale = exp(eta_gam) / rep(shape_true_gam, each = n_gam)),
  nrow = n_gam, dimnames = list(NULL, gam_traits)
)
gam_long <- two_trait_long(factor(seq_len(n_gam)), gam_traits, Y_gam)

fit_gam <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | unit, d = 1),
  data = gam_long, trait = "trait", unit = "unit",
  family = Gamma(link = "log"), silent = TRUE
)
phi_gam_hat <- fit_gam$report$phi_gamma
data.frame(trait = gam_traits, shape_true = shape_true_gam,
           phi_hat = round(phi_gam_hat, 2),
           cv_hat = round(1 / sqrt(phi_gam_hat), 2))
```

`leaf_area`'s larger `phi_hat` (about 7.8, shape simulated at 6) corresponds
to the *smaller* CV (about 0.36); `root_mass`'s smaller `phi_hat` (about 3.2)
has the *larger* CV (about 0.56). Report the CV, not `phi_hat` itself,
whenever "more dispersed" is the claim you want to make.

## Student-t's scale is not the response SD

Reach for `student()` when a real-valued response has occasional genuine
heavy-tailed observations that a Gaussian residual would under-predict. The
trap: `fit$report$sigma_student` is the **scale** of the underlying
\(t\)-distribution, not the response SD, and the two diverge as `df`
shrinks: \(\mathrm{SD}(y) = \sigma \sqrt{df / (df - 2)}\), undefined for
\(df \leq 2\). Leave `df = NULL` to estimate degrees of freedom, or fix them
with `student(df = 3)`.

```{r student-fit}
set.seed(20260818)
n_stu <- 150L
stu_traits <- c("bill_len", "wing_len")
z_stu <- rnorm(n_stu, sd = 0.4)
eta_stu <- cbind(bill_len = 10 + 1.5 * z_stu, wing_len = 15 - 1.0 * z_stu)
sigma_true_stu <- c(bill_len = 0.8, wing_len = 0.5)
df_fix <- 5
Y_stu <- matrix(
  eta_stu + rt(length(eta_stu), df = df_fix) * rep(sigma_true_stu, each = n_stu),
  nrow = n_stu, dimnames = list(NULL, stu_traits)
)
stu_long <- two_trait_long(factor(seq_len(n_stu)), stu_traits, Y_stu)

fit_stu <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | unit, d = 1),
  data = stu_long, trait = "trait", unit = "unit",
  family = student(df = df_fix), silent = TRUE
)
sigma_stu_hat <- fit_stu$report$sigma_student
data.frame(trait = stu_traits, sigma_true = sigma_true_stu,
           sigma_hat = round(sigma_stu_hat, 2),
           sd_hat = round(sigma_stu_hat * sqrt(df_fix / (df_fix - 2)), 2))
```

At `df = 5`, `bill_len`'s recovered scale (about 0.80) converts to an SD of
about 1.03—29% larger than the scale itself, and the gap widens as `df`
approaches 2 from above.

## Beta's `phi_beta` is a precision, and boundary values are not Beta data

Reach for `Beta()` when a response is a continuous proportion strictly
inside `(0, 1)`. `fit$report$phi_beta` is a **precision**, not a variance:
larger values mean *less* spread, via `shape1 = mu * phi_beta`,
`shape2 = (1 - mu) * phi_beta` with `mu = plogis(eta)`. The other trap is at
the boundary: genuine 0s or 1s are not Beta observations, and nudging them
inward to make the likelihood accept them silently changes the measurement
model rather than fixing a data-entry problem.

```{r beta-fit}
set.seed(20260818)
n_bet <- 150L
bet_traits <- c("cover_a", "cover_b")
z_bet <- rnorm(n_bet, sd = 0.4)
eta_bet <- cbind(cover_a = 0.2 + 0.8 * z_bet, cover_b = -0.3 - 0.5 * z_bet)
mu_bet <- plogis(eta_bet)
phi_true_bet <- c(cover_a = 12, cover_b = 20)
Y_bet <- matrix(
  rbeta(length(eta_bet), shape1 = mu_bet * rep(phi_true_bet, each = n_bet),
        shape2 = (1 - mu_bet) * rep(phi_true_bet, each = n_bet)),
  nrow = n_bet, dimnames = list(NULL, bet_traits)
)
bet_long <- two_trait_long(factor(seq_len(n_bet)), bet_traits, Y_bet)

fit_bet <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | unit, d = 1),
  data = bet_long, trait = "trait", unit = "unit",
  family = Beta(), silent = TRUE
)
data.frame(trait = bet_traits, phi_true = phi_true_bet,
           phi_hat = round(fit_bet$report$phi_beta, 2))
```

Both precisions recover within simulation noise of their targets (about
10.9 against 12, and 22.2 against 20). If your data legitimately contain 0
or 1, model the boundary process explicitly rather than reporting a Beta fit
against nudged values.

## Multi-trial binomial: `cbind()` vs `weights`, and a `simulate()` gotcha

Reach for `binomial()` with a trial count whenever the response is
successes out of known trials rather than a single 0/1 event. Two encodings
fit *identically*: `cbind(successes, failures) ~ ...`, or a flat response of
**raw success counts** with `weights = n_trials`—`weights` is the trial
count, not a proportion, and passing `successes / n_trials` silently fits
the wrong likelihood.

```{r binomial-trial-fit}
set.seed(20260818)
n_trl <- 150L
trl_traits <- c("infected", "damaged")
n_trials <- 10L
z_trl <- rnorm(n_trl, sd = 0.4)
eta_trl <- cbind(infected = 0.1 + 0.7 * z_trl, damaged = -0.4 - 0.5 * z_trl)
succ_trl <- matrix(
  rbinom(length(eta_trl), size = n_trials, prob = plogis(eta_trl)),
  nrow = n_trl, dimnames = list(NULL, trl_traits)
)
trl_long <- two_trait_long(factor(seq_len(n_trl)), trl_traits, succ_trl,
                            "successes")
trl_long$failures <- n_trials - trl_long$successes

fit_trl_cbind <- gllvmTMB(
  cbind(successes, failures) ~ 0 + trait + latent(0 + trait | unit, d = 1),
  data = trl_long, trait = "trait", unit = "unit",
  family = binomial(), silent = TRUE
)
fit_trl_weights <- gllvmTMB(
  successes ~ 0 + trait + latent(0 + trait | unit, d = 1),
  data = trl_long, trait = "trait", unit = "unit",
  weights = rep(n_trials, nrow(trl_long)), family = binomial(), silent = TRUE
)
c(cbind_logLik = as.numeric(logLik(fit_trl_cbind)),
  weights_logLik = as.numeric(logLik(fit_trl_weights)))
```

The two log-likelihoods match exactly (about -583.04). Separately:
`simulate()` treated every multi-trial binomial response as Bernoulli
regardless of trial count until it was fixed on 2026-08-17; any saved
simulation output for a multi-trial binomial fit produced before that fix is
suspect and should be regenerated.

## Ordinal cutpoints depend on which categories appear

Reach for `ordinal_probit()` when categories are ordered and coded as
consecutive integers `1, ..., K`. `gllvmTMB` fixes \(\tau_1 = 0\) as its
location convention, and `extract_cutpoints()` returns only
\(\tau_2, \ldots, \tau_{K-1}\). The hazard: `K` is inferred **per trait**
from the largest observed category, not declared, so a category that
happens not to occur in the fitted data silently shrinks that trait's model.

```{r ordinal-fit}
set.seed(20260818)
n_ord <- 150L
ord_traits <- c("condition", "vigor")
z_ord <- rnorm(n_ord, sd = 0.4)
taus_true <- c(0, 1.0, 2.2) # tau_1 = 0, tau_2 = 1.0, tau_3 = 2.2 -> K = 4
eta_ord <- cbind(condition = 0.5 + 0.9 * z_ord, vigor = 0.2 - 0.6 * z_ord)
draw_ordinal <- function(eta, taus) {
  ystar <- eta + rnorm(length(eta))
  as.integer(cut(ystar, breaks = c(-Inf, taus, Inf), labels = FALSE))
}
Y_ord <- matrix(draw_ordinal(as.vector(eta_ord), taus_true), nrow = n_ord,
                 dimnames = list(NULL, ord_traits))
Y_ord[, "vigor"] <- pmin(Y_ord[, "vigor"], 3L) # category 4 never occurs here
ord_long <- two_trait_long(factor(seq_len(n_ord)), ord_traits, Y_ord)

fit_ord <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | unit, d = 1),
  data = ord_long, trait = "trait", unit = "unit",
  family = ordinal_probit(), silent = TRUE
)
extract_cutpoints(fit_ord, quiet = TRUE)[, c("trait", "cutpoint_label",
                                              "tau_estimate")]
```

`condition` keeps all four simulated categories and returns both
`cutpoint_2` and `cutpoint_3` (about 1.04 and 2.28, against 1.0 and 2.2).
`vigor` was simulated the same way but its top category never appears after
recoding; the fitted model has only three categories and `cutpoint_3` does
not exist for it. `K` is a property of the observed data, not a declared
setting.

## Truncated counts condition on Y >= 1

Reach for `truncated_poisson()` or `truncated_nbinom2()` only when the
sampling process itself cannot produce or record a zero—not because zeros
were removed during cleaning. The likelihood conditions on \(Y \geq 1\),
renormalizing the count CDF by \(1 - F(0)\); every response must be a
positive integer. `truncated_nbinom2()`'s dispersion lives in
`fit$report$phi_truncnb2`, a vector kept **separate** from `phi_nbinom2`—
fitting the untruncated family on the same data and reading its
`phi_nbinom2` would be a different, wrong, parameter.

```{r truncated-fit}
set.seed(20260818)
rtnbinom2 <- function(n, mu, size) {
  p0 <- dnbinom(0, mu = mu, size = size)
  qnbinom(runif(n, p0, 1), mu = mu, size = size)
}
n_trn <- 300L
trn_traits <- c("visits", "clutch")
z_trn <- rnorm(n_trn, sd = 0.4)
eta_trn <- cbind(visits = 1.2 + 0.6 * z_trn, clutch = 1.4 - 0.4 * z_trn)
phi_true_trn <- c(visits = 3, clutch = 2)
Y_trn <- matrix(
  rtnbinom2(length(eta_trn), mu = exp(eta_trn),
            size = rep(phi_true_trn, each = n_trn)),
  nrow = n_trn, dimnames = list(NULL, trn_traits)
)
trn_long <- two_trait_long(factor(seq_len(n_trn)), trn_traits, Y_trn)

fit_trn <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | unit, d = 1),
  data = trn_long, trait = "trait", unit = "unit",
  family = truncated_nbinom2(), silent = TRUE
)
data.frame(trait = trn_traits, phi_true = phi_true_trn,
           phi_hat = round(fit_trn$report$phi_truncnb2, 2))
```

Both traits recover their dispersion in the right neighborhood (about 2.3
against 3, and 1.7 against 2), and every simulated value is at least 1—no
zero ever enters the truncated likelihood, by construction of the process.

## Different families in one fit

A mixed-family model needs long data because the family selector belongs to
each stacked response row. Use a factor with explicit levels and a **named**
family list; names are matched to selector levels, so accidental list ordering
cannot silently swap likelihoods.

```{r mixed-syntax, eval = FALSE}
set.seed(20260711)
n_mixed <- 60L
mixed_unit <- factor(seq_len(n_mixed))
z_mixed <- rnorm(n_mixed, sd = 0.45)

mixed_long <- rbind(
  data.frame(unit = mixed_unit, trait = "size", family = "continuous",
             value = rnorm(n_mixed, 0.3 + 0.7 * z_mixed)),
  data.frame(unit = mixed_unit, trait = "present", family = "presence",
             value = rbinom(n_mixed, 1, plogis(0.2 + 0.9 * z_mixed))),
  data.frame(unit = mixed_unit, trait = "abundance", family = "count",
             value = rpois(n_mixed, exp(0.4 + 0.5 * z_mixed)))
)
mixed_long$trait <- factor(mixed_long$trait)

mixed_long$family <- factor(
  mixed_long$family,
  levels = c("continuous", "presence", "count")
)

fam <- list(
  continuous = gaussian(),
  presence = binomial(),
  count = poisson()
)
attr(fam, "family_var") <- "family"

fit_mixed <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | unit, d = 1),
  data = mixed_long,
  trait = "trait",
  unit = "unit",
  family = fam
)
```

An unnamed list is still accepted, but its order must match the selector factor
levels exactly. A named list is safer. Keep one family within each trait; in
particular, an ordinal trait must own all of its rows because its thresholds are
estimated per trait.

Check that rule explicitly before fitting:

```{r mixed-family-check, eval = FALSE}
families_per_trait <- tapply(
  as.character(mixed_long$family),
  mixed_long$trait,
  function(x) length(unique(x))
)
stopifnot(all(families_per_trait == 1L))
```

Mixed-family fitting does not make every response directly comparable on its
raw observed scale. Define the scale and scientific target before interpreting
cross-trait covariance.

## Covariance scale and uncertainty

The fitted covariance tier and the observed responses are different objects.
For `extract_Sigma()`:

- `link_residual = "none"` returns the covariance components fitted in the
  selected tier;
- `link_residual = "auto"` adds a family-specific diagonal convention before
  converting to correlations.

The automatic addition is exact for fixed-scale links such as probit and an
approximation for several count, proportion, and two-part families. Neither
setting generally gives the correlation between raw observed responses. Choose
between them by stating the estimand, not by treating `"auto"` as universally
better.

For a fitted mixed-family model, keep the two point-estimate targets separate:

```{r mixed-family-point-estimates, eval = FALSE}
# Covariance fitted at the selected random-effect tier. Entries retain the
# units of their trait-specific linear predictors.
structural <- extract_Sigma(
  fit_mixed,
  level = "unit",
  part = "total",
  link_residual = "none"
)

# Pairwise family/link-scale correlations after adding each trait's
# link-residual convention to its diagonal.
adjusted <- extract_Sigma(
  fit_mixed,
  level = "unit",
  part = "total",
  link_residual = "auto"
)

structural$Sigma
adjusted$R
adjusted$note
```

Raw covariance magnitudes are not directly comparable when one predictor is in
identity-link units, another in logit units, and another in log units. The
correlations in `adjusted$R` are dimensionless, but they remain conditional on
the family, link, fitted mean trajectory, and residual convention for each
trait. For example, the Poisson addition uses a trait-level approximation based
on its fitted mean. These are pairwise model/link-scale summaries—not a single
universal latent scale and not observed-response correlations.

For mixed-family fits, point estimates can describe dependence on the selected
latent or link scale, but broad interval coverage has not been established for
this target. Keep the public workflow point-estimate-only; do not use mixed-family
correlation intervals as calibrated uncertainty.

### Specialist boundary for hurdle families

`delta_lognormal()` and `delta_gamma()` currently use one shared predictor:

\[
\Pr(Y > 0) = \operatorname{logit}^{-1}(\eta), \qquad
E(Y \mid Y > 0) = \exp(\eta).
\]

Use these constructors only when that shared-predictor assumption is
scientifically defensible. The current reader-facing route is the standard
fixed-effect hurdle fit. The package does not yet supply a response-scale
covariance or correlation interpretation for hurdle fits, so do
not describe fitted-tier correlations as correlations of total biomass,
abundance, or another observed two-part response.

`delta_gamma()`'s dispersion trap sits next to `Gamma()`'s and points the
opposite way: `fit$report$phi_gamma_delta` is the positive part's
coefficient of variation **directly**, not a shape—its near-identical
neighbor `phi_gamma` is the inverse-square of the CV, so the two report
names read alike but move oppositely for the same underlying spread.
`delta_lognormal()`'s positive-part dispersion, `sigma_lognormal_delta`, is
already a scale in the usual sense.

```{r delta-gamma-fit}
set.seed(20260818)
n_hur <- 150L
hur_traits <- c("biomass_a", "biomass_b")
z_hur <- rnorm(n_hur, sd = 0.4)
eta_hur <- cbind(biomass_a = 0.3 + 0.8 * z_hur, biomass_b = -0.2 - 0.5 * z_hur)
cv_true_hur <- c(biomass_a = 0.5, biomass_b = 0.7)
draw_delta_gamma <- function(p, mu, cv) {
  occ <- rbinom(length(p), 1, p)
  ifelse(occ == 1, rgamma(length(p), shape = 1 / cv^2, scale = mu * cv^2), 0)
}
Y_hur <- cbind(
  biomass_a = draw_delta_gamma(plogis(eta_hur[, 1]), exp(eta_hur[, 1]), cv_true_hur[1]),
  biomass_b = draw_delta_gamma(plogis(eta_hur[, 2]), exp(eta_hur[, 2]), cv_true_hur[2])
)
hur_long <- two_trait_long(factor(seq_len(n_hur)), hur_traits, Y_hur)

fit_hur <- gllvmTMB(
  value ~ 0 + trait + latent(0 + trait | unit, d = 1),
  data = hur_long, trait = "trait", unit = "unit",
  family = delta_gamma(), silent = TRUE
)
data.frame(trait = hur_traits, cv_true = cv_true_hur,
           cv_hat = round(fit_hur$report$phi_gamma_delta, 2))
```

The recovered CVs (about 0.39 against 0.5, and 0.61 against 0.7) are read
off directly; there is no shape-to-CV conversion here, unlike `Gamma()`
above.

## Diagnose the fitted choice

Family choice is provisional until the fitted model behaves sensibly.

```{r diagnostics, eval = FALSE}
check_gllvmTMB(fit_long)

# Exact randomized-quantile residuals are currently available for Gaussian,
# binomial, Poisson, lognormal, Gamma, NB1, NB2, Beta, betabinomial, student,
# truncated Poisson, truncated NB2, and ordinal_probit rows -- every family
# above except tweedie, delta_lognormal, delta_gamma, and multinomial.
residuals(fit_long, type = "randomized_quantile", seed = 1)

# Rootograms currently support Poisson, NB1, and NB2 count rows.
predictive_check(fit_long, type = "rootogram", ndraws = 100, seed = 1)
```

Systematic residual or rootogram structure can indicate a wrong family, link,
mean model, offset, or random-effect structure. It does not identify which one
automatically. Compare alternatives that represent credible sampling processes,
then recheck convergence and prediction rather than choosing solely by AIC.

Common input failures have direct remedies:

- **Gamma or lognormal data contain zero:** revisit the measurement process;
  use a family that allows zero rather than adding an arbitrary constant.
- **Counts are negative or non-integer:** correct the response definition; a
  count likelihood is not appropriate for transformed or continuous values.
- **Beta data include 0 or 1:** model the boundary process explicitly or choose
  another response representation.
- **Ordinal categories are not consecutive integers starting at 1:** recode an
  ordered factor with `as.integer()` and verify the level order.
- **A link is rejected:** use the accepted link shown above; the constructor may
  allow a base-R link that the multivariate engine does not.
- **Mixed-family names do not match selector levels:** name every family-list
  element exactly as the factor level.
- **A constructor is absent from this page:** it is not a supported
  multivariate response family; choose an available likelihood rather than
  relying on an exported experimental constructor.

For the full diagnostic workflow, continue to
[Fit diagnostics](fit-diagnostics.html). For covariance syntax, see the
[Formula keyword grid](api-keyword-grid.html).
