---
title: "Start here: which species distribution model do your data allow?"
description: "The front door to the species distribution articles. Every model on this site describes the same thing — a surface of expected abundance across a landscape — and your data type is a window onto it. This page shows the four windows, gives a decision tree and a routing table from what you have to what to read, and says plainly where a different package is the better tool."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Start here: which species distribution model do your data allow?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

You have some species records and a question about where things live. The
species distribution literature will offer you a dozen model names, and it is
not obvious which of them your data entitle you to. This page is the shortest
route from *what is in your spreadsheet* to *which article to open*.

It is a hallway, not a room. Nothing is fitted here. Everything below runs on
base R alone, in a couple of seconds, so you can paste it into a console and
see the pictures for yourself.

# One surface, four windows

Every model on this site — and its close neighbours, including occupancy
models — is describing the same underlying object: a **surface of expected
abundance**, one for each species, spread across the landscape. High where the
species thrives, low where it does not. That surface is what you actually want
to know about. You never observe it.

What you observe is a **window** onto it, and the shape of the window is
decided by how the data were collected:

- **Counts with effort** — a checklist, a trap-night, a timed transect — see
  the surface almost directly, scaled by how hard you looked. Twice the effort,
  roughly twice the count.
- **Detection / non-detection** sees only whether the surface was above zero
  at the places you visited. A rich site and a merely adequate one both come
  back as a 1.
- **Presence-only records** — one photo, one point, uploaded by a stranger —
  see the surface through a second filter: *where people go*. A road, a
  car park, and a popular reserve all look like good habitat.
- **Grid cells** see the surface averaged over squares. Averaging flattens
  peaks; a hotspot smaller than a cell stops existing.

```{r windows, ref.label = "windows-code", echo = FALSE, fig.width = 7.2, fig.height = 3.9, dpi = 150, out.width = "100%", fig.cap = "One simulated landscape, seen four ways. The centre panel is the true surface of expected abundance per unit of effort, which no data type observes directly; the flanking panels show what each kind of survey would have recorded from it. (a) counts at twenty places, printed as numbers, with text size marking how much effort each place received. (b) detection or non-detection at thirty-four places, one visit each: filled circles found the species, open circles did not. (c) eighty-five presence-only records, proposed where people are and then thinned by the surface, so they pile up both on the real hotspot and in the accessible bottom-right corner. (d) the same surface averaged into twenty-five coarse cells on the same colour scale, which visibly drains the bright peak. This is a drawn illustration on simulated data, not a fitted model.", fig.alt = "A five-panel figure. The large central panel is a smooth colour surface on a viridis scale with white contour lines, showing a large bright peak in the upper left and a smaller peak in the lower right against a dark background. Four smaller white panels flank it. Panel a shows scattered printed numbers, larger and higher in the upper-left region. Panel b shows scattered circles, mostly filled in the upper left and more often open in the lower right. Panel c shows scattered plus signs forming two clumps, one on the upper-left peak and a denser one in the lower-right corner. Panel d shows a five-by-five grid of coloured squares on the same scale as the central panel, with the peak reduced to a single mid-bright square."}
```

The consequence is the single most useful idea on this site: **the model you
need is not chosen by your species or your region — it is chosen by your
window.** Two people studying the same warbler in the same valley need
different models if one of them has checklists and the other has photographs.

And it is why combining sources is worth the trouble. Two windows onto one
surface are strictly more informative than either alone, provided the model
knows they are windows and lets each keep its own distortion. That is what the
integrated models on this site do.

# Two questions decide your route

**What do you have?** (the window) and **what do you want to know?** (the
question). The tree answers the first; the table underneath adds the second.

```{r tree, ref.label = "tree-code", echo = FALSE, fig.width = 7.2, fig.height = 6.4, dpi = 150, out.width = "100%", fig.cap = "Routing by data type. Read downwards: the top row is what your data look like, the diamonds are the two follow-up questions where the answer sends you off this site, and the bottom row is a destination. Solid boxes are articles here; dashed boxes are honest exits to a better tool elsewhere, distinguished by border style rather than by colour so the distinction survives greyscale printing. The band along the foot applies whenever you hold more than one of the four data types at once.", fig.alt = "A flow chart reading top to bottom. A grey box at the top reads: what do your data look like? Four arrows lead to four boxes: presence-only records; counts with effort; detection or non-detection with one visit per place; and repeat-visit histories. The presence-only box leads to a diamond asking whether the question is a trend over years, whose yes branch reaches a dashed box reading grid cells plus list-based occupancy models, and whose no branch reaches a solid box reading joint intensity from records, the point-process route. Counts with effort leads straight to a solid box reading joint intensity from records, count arm, no recording-bias term. Detection or non-detection leads straight to a solid box reading binary joint SDM, with one visit detection cannot be separated from occurrence. Repeat-visit histories leads to a diamond asking whether presence must be split from detection, whose yes branch reaches a dashed box reading spOccupancy carries the detection submodel, and whose no branch reaches a solid box reading what repeat visits add to one shared intensity. Dotted lines gather the four solid boxes into a wide band at the foot reading: have two or more of these at once? That is the integrated model, one shared surface with each source keeping its own window; this site's centrepiece and the only route that uses all of your data."}
```

The tree routes on data alone. Your question splits the destinations further:

| you have ↓ / you want → | **a map, or a habitat response** | **community structure** — which species covary | **"was it really there?"** — detection separated |
|---|---|---|---|
| **counts with effort** <br> checklists, traps, transects | [joint intensity from records](gbif-joint-intensity.html), count arm · [add space](spatial-models.html) | the same fit — all species on shared latent gradients | not from counts alone; you need repeat visits |
| **detection / non-detection** <br> one visit per place | [binary joint SDM](joint-sdm.html) · or its [survey arm](integrated-two-source-example.html) | [binary joint SDM](joint-sdm.html) · [if a species is too rare](rare-species-jsdm.html) | not possible from one visit — the occupancy model collapses into this one |
| **detection histories** <br> repeat visits | [what repeat visits add](integrated-repeated-visits.html) — but read the exit first | possible, same fit | **spOccupancy territory** — see [the honest exits](#the-honest-exits) |
| **presence-only** <br> one photo, one point | [joint intensity from records](gbif-joint-intensity.html) — the point-process route | the same fit | **the occupancy-trend tradition** — see [the honest exits](#the-honest-exits) |
| **mixed sources** <br> [eBird](https://ebird.org) plus your own survey | **[the integrated model, end to end](isdm-canada-warbler.html)** · [start smaller](integrated-two-source-example.html) · [go wider](integrated-multi-source-example.html) | yes, and it is the same fit | only as far as the survey arm's visits allow |

Two honesty notes about that table. There is no article here devoted to
counts from a *single designed* survey; the machinery for it lives inside the
presence-only article, because a designed count is the same likelihood with
the bias term switched off. And every route in the table returns **relative**
intensity, not absolute abundance or occupancy probability — the individual
articles are specific about what that rules out.

# The honest exits

Some questions are better answered elsewhere. Saying so is part of the routing.

**Occupancy models.** These are in the *same family* as the binary joint SDM
above, with one extra layer: a **detection submodel**, a second process that
says how often you find the species given that it is there. That extra layer
is only estimable because **repeat visits** give you the information to split
the two. If you have repeat visits *and* your question genuinely needs "was it
there" separated from "did we find it" — absolute occupancy, or a long-term
trend from opportunistic data — use
[spOccupancy](https://www.jeffdoser.com/files/spoccupancy-web/), which is built
for exactly that. With a single visit the two models collapse into the same
thing, and that collapsed model is the one this package fits. So this is not a
rival framework; it is one layer further along the same road, and the layer
needs data you may not have.

**Trends from one-photo records**, the way insect recording schemes work. The
established route is grid cells plus occupancy modelling of *list-based visits*
— using the other species recorded on the same list to reconstruct the visit
and detection structure a raw photo stream lacks (van Strien, van Swaay and
Termaat 2013, *Opportunistic citizen science data of animal species produce
reliable estimates of distribution trends if analysed with occupancy models*).
Not this package. This package's presence-only route is aimed at habitat
responses and intensity maps, not at trends.

**Uncertainty on a map.** The prediction maps produced here currently carry no
intervals of any kind — `predict()` refuses `se.fit` with new data rather than
returning a number that ignores the dominant source of error. If your thesis
figure needs a credible interval on every pixel, that is a real limit of this
package today, and you should know it before you build a chapter on it. The
[Canada Warbler article](isdm-canada-warbler.html) states the point in full,
and [Current limitations](current-limits.html) keeps the running list.

# Where to start reading

A short path, in order.

1. **[What is a site? The unit of analysis](unit-of-analysis.html)** — before
   anything else. It settles the data-shape decisions that every route below
   inherits, and it will probably talk you out of gridding your data.
2. **Your cell of the table above.** One article. Fit the toy example in it
   before you touch your own data.
3. **The two warning articles, before real data.**
   [When one data source knows where it is and the other does not](isdm-spatial-precision.html)
   is about location precision, which decides whether combining sources helps
   or hurts. And the *"Before you use this on real data"* section of the
   [Canada Warbler article](isdm-canada-warbler.html) is the checklist of
   things that go wrong first.

If a word on any of those pages is unfamiliar, the
[plain-English vocabulary](gllvm-vocabulary.html) page defines the package's
terms in one line each. If you are sizing a survey rather than analysing one,
[How much survey do you need?](integrated-survey-design.html) is the page for
that.

The integrated route on this site is **experimental**: the interface is
settled enough to use and the estimates are illustrated on known-truth
simulations, but confidence intervals for it are not certified. Each article
says where its own evidence stops.

# How the figures were drawn

Both figures use only base R and take about a second. Nothing in them is
fitted; they are drawings of a definition.

```{r windows-code, eval = FALSE}
set.seed(7)

## One smooth surface: expected abundance per unit of effort on a unit square.
## Nothing here is a fit -- this picture is a definition, not a result.
lambda <- function(x, y) {
  6.0 * exp(-((x - 0.30)^2 + (y - 0.72)^2) / 0.075) +
  3.2 * exp(-((x - 0.78)^2 + (y - 0.24)^2) / 0.045) + 0.35
}
gx <- seq(0, 1, length.out = 120); gy <- gx
Z  <- outer(gx, gy, lambda)

pal  <- hcl.colors(64, "Viridis")       # monotone luminance: survives greyscale
brks <- seq(0, max(Z) * 1.001, length.out = 65)
ins  <- function(n) 0.08 + 0.84 * runif(n)   # keep marks off the panel border

## (a) counts with effort: where you looked, and how hard.
n_a <- 20; ax <- ins(n_a); ay <- ins(n_a)
eff <- sample(c(0.4, 1, 2), n_a, replace = TRUE)
cnt <- rpois(n_a, lambda(ax, ay) * eff)

## (b) detection / non-detection at one visit per place.
n_b <- 34; bx <- ins(n_b); by <- ins(n_b)
det <- rbinom(n_b, 1, 1 - exp(-lambda(bx, by)))

## (c) presence-only: proposed where PEOPLE go, then thinned by the surface.
n_c  <- 85
cand <- cbind(ins(4000), ins(4000))
wt   <- exp(-sqrt((cand[, 1] - 0.94)^2 + (cand[, 2] - 0.06)^2) / 0.26) *
        lambda(cand[, 1], cand[, 2])
kp   <- sample(nrow(cand), n_c, prob = wt)
cx   <- cand[kp, 1]; cy <- cand[kp, 2]

## (d) the same surface averaged over 5 x 5 coarse cells.
K <- 5; ex <- seq(0, 1, length.out = K + 1)
Zc <- outer(seq_len(K), seq_len(K), Vectorize(function(i, j)
  mean(Z[gx >= ex[i] & gx < ex[i + 1], gy >= ex[j] & gy < ex[j + 1]])))

blank <- function() {
  plot.new(); plot.window(c(0, 1), c(0, 1), xaxs = "i", yaxs = "i")
  rect(0, 0, 1, 1, col = "white", border = "grey30")
}
tag <- function(lab, sub) {
  mtext(lab, side = 3, line = 0.10, adj = 0, cex = 0.68, font = 2)
  mtext(sub, side = 1, line = 0.30, cex = 0.60)
}

op <- par(no.readonly = TRUE)
layout(matrix(c(2, 1, 1, 3,
                4, 1, 1, 5), nrow = 2, byrow = TRUE),
       widths = c(1, 0.95, 0.95, 1))
par(mar = c(1.9, 0.5, 1.4, 0.5))

## The centre: the surface itself.
image(gx, gy, Z, col = pal, breaks = brks, axes = FALSE, xlab = "", ylab = "",
      xaxs = "i", yaxs = "i")
contour(gx, gy, Z, add = TRUE, drawlabels = FALSE, col = "white", lwd = 0.6,
        levels = pretty(range(Z), 6))
box(col = "grey20", lwd = 1.4)
mtext("What every model here is after", side = 3, line = 0.25, cex = 0.82, font = 2)
mtext("expected abundance per unit of effort, everywhere", side = 1,
      line = 0.30, cex = 0.66)

blank(); text(ax, ay, cnt, cex = 0.60 + 0.20 * log2(eff * 2.5), col = "grey10")
tag("(a) counts with effort", "number = count, size = effort")

blank(); points(bx, by, pch = ifelse(det == 1, 19, 1), cex = 0.82,
                col = "grey10", lwd = 1.1)
tag("(b) detected or not", "filled = found, open = not")

blank(); points(cx, cy, pch = 3, cex = 0.52, col = "grey10", lwd = 0.9)
tag("(c) presence-only", "clumped by PEOPLE too")

image(ex, ex, Zc, col = pal, breaks = brks, axes = FALSE, xlab = "", ylab = "",
      xaxs = "i", yaxs = "i")
abline(v = ex, h = ex, col = "white", lwd = 0.7); box(col = "grey30")
tag("(d) grid cells", "averaged into squares")
par(op)
```

```{r tree-code, eval = FALSE}
op <- par(no.readonly = TRUE)
par(mar = c(0.2, 0.2, 0.2, 0.2))
plot.new(); plot.window(xlim = c(0, 100), ylim = c(0, 100))

INK <- "grey15"

## A box. `kind` sets the border STYLE as well as the fill, so the distinction
## between "an article on this site" and "a better tool elsewhere" survives
## printing in greyscale: destinations are solid, honest exits are dashed.
bx <- function(x, y, w, h, txt, kind = "branch", gap = 3.4) {
  sty <- switch(kind,
    start  = list(fill = "grey86", lty = 1, lwd = 1.4, font = 2, cex = 0.82),
    branch = list(fill = "grey97", lty = 1, lwd = 1.0, font = 1, cex = 0.68),
    leaf   = list(fill = "#D9E6F2", lty = 1, lwd = 1.7, font = 1, cex = 0.66),
    exit   = list(fill = "white",   lty = 2, lwd = 1.5, font = 3, cex = 0.66))
  rect(x - w/2, y - h/2, x + w/2, y + h/2, col = sty$fill,
       border = INK, lty = sty$lty, lwd = sty$lwd)
  n <- length(txt)
  text(x, y + (n - 1)/2 * gap - (seq_len(n) - 1) * gap,
       txt, cex = sty$cex, font = sty$font, col = INK)
}

dia <- function(x, y, w, h, txt) {
  polygon(c(x - w/2, x, x + w/2, x), c(y, y + h/2, y, y - h/2),
          col = "grey97", border = INK, lwd = 1.0)
  n <- length(txt)
  text(x, y + (n - 1)/2 * 2.7 - (seq_len(n) - 1) * 2.7,
       txt, cex = 0.59, col = INK)
}

## Elbow connector: down out of the parent, across, then down into the child.
elb <- function(x0, y0, x1, y1, lab = NULL, lx = 0) {
  ym <- (y0 + y1) / 2
  segments(x0, y0, x0, ym, col = INK, lwd = 0.9)
  segments(x0, ym, x1, ym, col = INK, lwd = 0.9)
  arrows(x1, ym, x1, y1, length = 0.055, col = INK, lwd = 0.9)
  if (!is.null(lab)) text(x0 + lx, (y0 + ym)/2 + 0.4, lab,
                          cex = 0.60, font = 2, col = INK)
}

## ---- the tree, read downwards ---------------------------------------------
bx(50, 95.5, 46, 7.5, "What do your data look like?", "start")

cxs <- c(12.5, 37.5, 62.5, 87.5)
bx(cxs[1], 78, 23, 12, c("presence-only records", "one point, one photo,", "no idea who looked"), "branch")
bx(cxs[2], 78, 23, 12, c("counts with effort", "checklists, traps,", "timed transects"), "branch")
bx(cxs[3], 78, 23, 12, c("detection /", "non-detection", "one visit per place"), "branch")
bx(cxs[4], 78, 23, 12, c("repeat-visit histories", "same place,", "several times"), "branch")

segments(50, 91.75, 50, 88.5, col = INK, lwd = 0.9)
segments(cxs[1], 88.5, cxs[4], 88.5, col = INK, lwd = 0.9)
for (x in cxs) arrows(x, 88.5, x, 84.0, length = 0.055, col = INK, lwd = 0.9)

## A follow-up question only where an honest exit lives.
dia(cxs[1], 59, 25, 15, c("is the question a", "TREND over years?"))
dia(cxs[4], 59, 25, 15, c("do you need presence", "split from detection?"))
for (x in c(cxs[1], cxs[4])) arrows(x, 72, x, 66.7, length = 0.055, col = INK, lwd = 0.9)

## Leaves. Solid = an article here; dashed = somebody else's tool.
bx( 5.6, 35.5, 10.6, 14.5, c("grid cells +", "list-based", "occupancy", "models"), "exit", 3.2)
bx(20.2, 35.5, 14.4, 14.5, c("joint intensity", "from records", "the point-", "process route"), "leaf", 3.2)
bx(cxs[2], 35.5, 23, 14.5, c("joint intensity", "from records — count", "arm, no recording-", "bias term"), "leaf", 3.2)
bx(cxs[3], 35.5, 23, 14.5, c("binary joint SDM", "with one visit, detection", "cannot be separated", "from occurrence"), "leaf", 3.2)
bx(79.2, 35.5, 12.4, 14.5, c("spOccupancy", "carries the", "detection", "submodel"), "exit", 3.2)
bx(93.2, 35.5, 12.4, 14.5, c("what repeat", "visits add", "to one shared", "intensity"), "leaf", 3.2)

elb(cxs[1], 51.5,  5.6, 42.8, "yes", -8.0)
elb(cxs[1], 51.5, 20.2, 42.8, "no",   6.0)
elb(cxs[4], 51.5, 79.2, 42.8, "yes", -8.0)
elb(cxs[4], 51.5, 93.2, 42.8, "no",   6.0)
elb(cxs[2], 72.0, cxs[2], 42.8)
elb(cxs[3], 72.0, cxs[3], 42.8)

## The band. Only the SOLID leaves feed it; the exits deliberately do not.
for (x in c(20.2, cxs[2], cxs[3], 93.2))
  segments(x, 28.25, x, 25.5, col = INK, lwd = 0.9, lty = 3)
segments(20.2, 25.5, 93.2, 25.5, col = INK, lwd = 0.9, lty = 3)
segments(50, 25.5, 50, 22.5, col = INK, lwd = 1.1)
arrows(50, 24, 50, 21.5, length = 0.06, col = INK, lwd = 1.1)

rect(2, 4.5, 98, 20.5, col = "#D9E6F2", border = INK, lwd = 1.9)
text(50, 16.0, "Have TWO OR MORE of these at once?", cex = 0.78, font = 2, col = INK)
text(50, 11.3, "That is the integrated model: one shared surface, each source keeping its own window.",
     cex = 0.66, col = INK)
text(50,  7.4, "It is this site's centrepiece, and the only route that uses all of your data.",
     cex = 0.66, font = 3, col = INK)

## Legend, parked below everything so it can collide with nothing.
text(50, 1.4, "solid box = an article on this site        dashed box = a better tool elsewhere",
     cex = 0.62, col = "grey30")
par(op)
```

# References

van Strien, A.J., van Swaay, C.A.M. and Termaat, T. (2013). Opportunistic
citizen science data of animal species produce reliable estimates of
distribution trends if analysed with occupancy models. *Journal of Applied
Ecology* **50**, 1450--1458.
