---
title: "Simulation recovery: M3 smoke grid"
tier: 3 # developer validation note, not a first-stop tutorial
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Simulation recovery: M3 smoke grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
m3_smoke_rds <- system.file("extdata", "m3-coverage-grid-smoke.rds",
                            package = "gllvmTMB", mustWork = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = nzchar(m3_smoke_rds) && file.exists(m3_smoke_rds),
  message  = FALSE,
  warning  = FALSE
)
```

> **Developer validation note.** This page records simulation-recovery
> evidence for maintainers and reviewers. It is not a first-stop
> tutorial. Start with [Get started](gllvmTMB.html),
> [Can I trust this fit?](fit-diagnostics.html), and
> [Simulation-based verification](simulation-verification.html) before
> using this page. It is a Tier 3 coverage-triage article: it shows the
> M3 smoke-grid machinery and the under-coverage pattern, but it is not
> evidence that interval coverage has passed. Package-wide interval
> coverage remains incomplete.

This article is for a reader who wants to know **whether the confidence
intervals reported by `gllvmTMB()` actually contain the truth at their nominal
rate**. The standard quantitative-genetic and psychometric tutorials usually
demonstrate convergence and point-estimate recovery; this internal page tracks
the harder coverage question.

Coverage validation is an open M3 target in `gllvmTMB`'s roadmap: empirical
coverage rates at or above 94% for nominal 95% intervals across Gaussian,
binomial, nbinom2, ordinal-probit, and mixed-family fits. The current register
status is **partial**, not covered. The M3.3 production run completed, but did
not clear the gate: only the Gaussian `d = 1` and `d = 3` cells passed, 13 of
15 cells stayed below gate, and 236 of 3000 replicate fits failed. No
production RDS was promoted to `inst/extdata/`.

The full M3 grid is described in
[Design 42](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/42-m3-dgp-grid.md):
5 family cells × 3 latent ranks × R = 200 replicates = 3000 fits,
with profile-likelihood CIs on each free parameter. This article
shows **the M3.3a smoke output** — a 15-cell × 10-rep version of
that grid with **profile-likelihood CIs** on per-trait
diagonal Psi variance $\psi_t$. The M3.3
production grid scales R = 10 to R = 200 (Monte Carlo error
±15 pp → ±3 pp at the 95% level). Treat the smoke output as triage
evidence: it shows which surfaces are hard, not that interval coverage is
validated.

## The target

The DGP simulates a fresh truth
$\boldsymbol\Lambda_{\mathrm{true}} \sim \mathrm{Uniform}(-1.5, 1.5)^{5 \times d}$
and $\boldsymbol\psi_{\mathrm{true}} \sim \mathrm{Gamma}(2, 2)^5$
per replicate, then asks: does the fit's profile-likelihood CI
on each per-trait $\psi_t$ contain the generative truth? $\psi_t$
is the **rotation-invariant target** -- it is the per-trait
diagonal Psi variance, identifiable up to non-rotational sign +
permutation. The shared-Lambda contribution to the full
$\Sigma_{\mathrm{unit}}$ diagonal is rotation-ambiguous on
individual loading entries; the **communality** $c_t^2 =
(\boldsymbol\Lambda\boldsymbol\Lambda^\top)_{tt} / \Sigma_{tt}$
is the corresponding rotation-invariant target for the Lambda
side, deferred to M3.5 (derived-quantity coverage).

Profile CIs are computed via `tmbprofile_wrapper()` warm-started
from the joint MLE — typically ~0.5-1s per CI on a 1.5s fit
(per `R/profile-ci.R:127`), making the production-scale grid
feasible (~12h serial, ~1.5h on 8 cores).

## The grid

Per replicate, the DGP samples a fresh
$\boldsymbol\Lambda_{\mathrm{true}} \sim \mathrm{Uniform}(-1.5, 1.5)^{5 \times d}$
and $\boldsymbol\psi_{\mathrm{true}} \sim \mathrm{Gamma}(2, 2)^5$,
simulates 60 units × 5 traits from that truth, fits `gllvmTMB` with
the matching ordinary `latent()` formula and family, and records
whether the fit's profile-CI on each per-trait $\psi_t$ contains
the generative truth $\boldsymbol\psi_{\mathrm{true}, t}$.
Per-family non-Gaussian responses: binomial uses `rbinom(size = 1)`;
nbinom2 uses `rnbinom(mu, size = phi)` with `phi ~ Gamma(5, 5)`;
ordinal-probit uses 4 categories with quantile-spaced cutpoints.

## Reading the precomputed RDS

```{r read}
artefact <- readRDS(m3_smoke_rds)
artefact$meta[c("label", "gllvmTMB_ver", "n_cells", "n_reps", "elapsed_s")]
```

The `gllvmTMB_ver` field stamps the package version the smoke artefact was
generated under; it records the artefact's provenance and is not necessarily
the version of the installed package rendering this page.

## Per-cell coverage summary

```{r summary}
artefact$summary
```

The `coverage_prof` column is the fraction of (rep × trait) pairs
whose profile-likelihood CI on $\psi_t$ contained the truth.
`passes_94pct_prof` flags cells that meet the M3 audit-1 exit
gate (≥ 94 % nominal at 95 % level).

```{r smoke-honesty, results = "asis"}
n_pass <- sum(artefact$summary$passes_94pct_prof)
n_cells <- nrow(artefact$summary)
fam_avg <- aggregate(coverage_prof ~ family,
                     data = artefact$summary, FUN = mean)
cat(sprintf(
  "**Smoke status**: %d / %d cells pass the 94%% gate at smoke scale (R = 10 reps; Monte Carlo error ~±15 pp).\n\n",
  n_pass, n_cells
))
cat("**Per-family average coverage**:\n\n")
for (i in seq_len(nrow(fam_avg))) {
  cat(sprintf("- %s: %.2f\n", fam_avg$family[i], fam_avg$coverage_prof[i]))
}
```

Interpretation:

- **Gaussian** and **binomial** profile CIs land near nominal at
  smoke scale — the inference machinery appears to work on the
  well-conditioned cases.
- **Ordinal-probit**, **mixed-family**, and especially
  **nbinom2** under-cover — these are real, not placeholder
  artefacts. Likely causes: small-n likelihood non-concavity,
  boundary parameters near zero (where Wald-like profile crossings
  miss the true coverage), and per-trait variance pinning when
  one trait carries little variance signal. M3.4 (boundary
  regimes) and M3.5 (derived-quantity coverage) work on these.

A future recovery pass must:

1. Keep R = 200 or larger so Monte Carlo error is narrow enough for gate
   decisions.
2. Diagnose and reduce replicate-fit failures before any result is promoted.
3. Use target-explicit coverage surfaces, including total `Sigma_unit[tt]`
   where Design 50 requires it.
4. Land a diagnostic report beside the production artefact before promoting
   either the single-family or mixed-family coverage claim.
5. Extend coverage to derived quantities (communality,
   repeatability, phylo signal) — M3.5.

When a later production pass clears its gates, this article can re-render with
promoted coverage estimates. Until then, it is an internal failure-analysis
page.

## Inspecting individual fits

For diagnostic purposes, you can pull a single rep's outputs:

```{r single-rep}
single <- subset(artefact$grid, cell == "gaussian-d2" & rep == 1L)
single[, c("trait_id", "truth_psi", "est_psi",
           "ci_prof_lo", "ci_prof_hi", "covered_prof")]
```

Each row is a trait's $\psi_t$ (per-trait diagonal Psi variance) target.
`truth_psi` is the generative truth for that rep; `est_psi` is
the point estimate; `ci_prof_lo` / `ci_prof_hi` are the 95 %
profile-likelihood CI bounds; `covered_prof` is whether the CI
contained the truth.

## Notes on per-family smoke behaviour

Per-family runtimes and convergence are stable across the smoke
runs; what M3.3a surfaces is the **inference-quality pattern**:

- **Gaussian** (avg 0.95 coverage; ~0.7s per fit): near-nominal at
  smoke scale, as expected for the well-conditioned baseline.
- **Binomial** (avg 0.95; ~0.7-1s): also near-nominal. At smoke
  scale the profile CIs appear to handle the bounded-likelihood
  geometry; binomial has not cleared the production coverage gate.
- **Ordinal-probit** (avg 0.75; ~1-1.5s): under-covers. K-1
  cutpoints per trait + small n = harder identifiability. M3.4
  will diagnose whether single-trait warmup helps.
- **Mixed-family** (avg 0.64; ~0.9-1.8s): under-covers + 30%
  non-convergence at d=2, d=3. Per-row family dispatch tests the
  hardest path. M3.4 priority.
- **nbinom2** (avg 0.38; ~2-3s): worst cell. Two-parameter
  (mean + dispersion) likelihood at small n is genuinely
  unstable. M3.4 + a possible engine-side fix for dispersion
  identifiability are next.

These are real inference patterns, not placeholder artefacts. They should be
treated as failure-analysis evidence until the production grid, target
definition, and replicate-fit failure rate all pass their gates.

## Reproducing the smoke run

The pipeline lives at
[`dev/m3-grid.R`](https://github.com/itchyshin/gllvmTMB/blob/main/dev/m3-grid.R)
(library functions) +
[`dev/precompute-m3-grid.R`](https://github.com/itchyshin/gllvmTMB/blob/main/dev/precompute-m3-grid.R)
(driver). From the repo root:

```bash
# Smoke (Gaussian only, 10 reps × 3 cells): ~18s on a 2025 laptop
Rscript dev/precompute-m3-grid.R

# Smoke across all 5 families (15 cells × 10 reps): ~10-20 min serial
Rscript dev/precompute-m3-grid.R --all-fams

# Full production grid (15 cells × 200 reps): ~hours, M3.3 only
Rscript dev/precompute-m3-grid.R --full
```

Output goes to `dev/precomputed/m3-coverage-grid.rds` (long-format,
150 rows for the default smoke) and `dev/precomputed/m3-coverage-summary.rds`
(per-cell aggregate).

## What this article does NOT yet show

Per Design 42 §6 honest-scope boundaries, the M3 milestone covers:

- **Pipeline machinery (M3.2):** this article and smoke output.
- **Per-family profile CI accuracy (M3.3):** replace Wald
  placeholder with proper inference; expand to all 5 families.
- **Boundary regimes (M3.4):** sparse-Bernoulli at d=3;
  ordinal-probit at d=2 with rare categories.
- **Derived-quantity coverage (M3.5):** communality,
  repeatability, phylogenetic signal.
- **M3 close gate (M3.7):** Ada + Shannon coordination audit.

Single-family and mixed-family interval coverage remain partial. This page
must not be used as release readiness, broad interval calibration, or
scientific coverage evidence.

Out of scope for M3 entirely:

- Model-misspecification coverage (mis-specified rank, omitted
  family). Different validation framework; v0.3.0 scope.
- Asymptotic large-n coverage (n = 600, n = 6000). v0.3.0 scaling
  study.
- Cross-package agreement. Phase 5.5 work.

## See also

- [Design 42](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/42-m3-dgp-grid.md)
  — full M3 grid specification (5 families × 3 dims × 200 reps).
- [`coverage_study()`](../reference/coverage_study.html) —
  per-fit parametric-bootstrap coverage estimator (complementary to
  this external-simulation grid).
- [`confint_inspect()`](../reference/confint_inspect.html) —
  visual diagnosis when coverage falls short.
- [Profile-likelihood CIs](profile-likelihood-ci.html) — the
  inference method M3.3 is trying to validate.
- [Troubleshooting profile](troubleshooting-profile.html) — common
  patterns when profile coverage is unexpectedly low.
- [Animal model](animal-model.html) — quantitative-genetic worked
  example using the same ordinary `latent()` machinery this
  article exercises (point recovery; interval coverage remains open).
