---
title: "Simulation-based verification of gllvmTMB fits"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Simulation-based verification of gllvmTMB fits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

A multivariate latent-variable fit summarises a lot of data into a
small set of statements: *here are the loadings, this is the
correlation matrix, communality is high, phylogenetic signal is
moderate*. Before you build a paper on those statements, you
should verify that the model + inference + extractor pipeline
actually recovers the values you put in. This article walks
through the canonical simulate-fit-check loop with the four
diagnostic functions gllvmTMB ships:

- [`check_identifiability()`](../reference/check_identifiability.html)
  -- Procrustes-aligned simulate-refit identifiability test.
- [`gllvmTMB_check_consistency()`](../reference/gllvmTMB_check_consistency.html)
  -- Laplace-approximation score-centering test.
- [`coverage_study()`](../reference/coverage_study.html) -- empirical
  CI coverage rates against the audit's `>= 94%` exit gate.
- [`confint_inspect()`](../reference/confint_inspect.html) -- visual
  profile-curve verification for a single target.

If you are new to the vocabulary in this article, see
[gllvmTMB vocabulary](gllvm-vocabulary.html) first. If you want
the failure-mode catalogue for profile CIs in particular, see
[Troubleshooting profile-likelihood CIs](troubleshooting-profile.html).

## 1. Why simulate?

Simulation is the only way to verify *the full pipeline*. Unit
tests check that individual functions work; simulation checks that
they compose correctly into the recovery story you intend to tell.

A correctly-implemented gllvmTMB fit on data simulated from a
known DGP should recover (to within sampling noise):

- the loading matrix $\boldsymbol{\Lambda}$ up to rotation;
- the implied trait covariance $\boldsymbol{\Sigma}$ exactly (it's
  rotation-invariant);
- per-trait communality $c_t^2$;
- per-trait phylogenetic signal $H^2$;
- per-trait repeatability $R_t$.

When recovery fails, one of three things is wrong:
**(a)** the DGP doesn't match the fitted model; **(b)** the data
don't constrain some parameter; **(c)** the inference machinery
mis-reports the uncertainty.

## 2. Anatomy of a simulation: DGP -> fit -> check

```{r setup}
library(gllvmTMB)
```

### 2a. Simulate from a known DGP

The package's `simulate_site_trait()` is the canonical DGP for the
worked examples. The truth here is hand-set: three traits with a
shared rank-1 axis (Lambda_B = (0.8, 0.4, -0.3)) plus a
trait-specific unique-variance diagonal (psi_B = (0.3, 0.3, 0.3)).

```{r sim}
set.seed(1)
n_traits <- 3
truth <- list(
  Lambda_B = matrix(c(0.8, 0.4, -0.3), n_traits, 1),
  psi_B    = c(0.3, 0.3, 0.3)
)
sim <- simulate_site_trait(
  n_sites = 40, n_species = 6, n_traits = n_traits,
  mean_species_per_site = 4,
  Lambda_B = truth$Lambda_B, psi_B = truth$psi_B,
  seed = 1
)
df <- sim$data
```

### 2b. Fit the model

```{r fit}
fit <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | site, d = 1) +
          unique(0 + trait | site),
  data = df
)
fit$opt$convergence  # should be 0L
```

### 2c. Compare truth to the fitted Sigma

The implied between-site covariance is the rotation-invariant
biological target. Compare it to the truth:

```{r recover-sigma}
Sigma_truth <- truth$Lambda_B %*% t(truth$Lambda_B) +
               diag(truth$psi_B)
Sigma_fit   <- extract_Sigma(fit, level = "unit", part = "total")$Sigma
data.frame(
  cell  = paste0("[", row(Sigma_truth), ",", col(Sigma_truth), "]"),
  truth = as.numeric(Sigma_truth),
  fit   = as.numeric(Sigma_fit),
  diff  = as.numeric(Sigma_fit - Sigma_truth)
)
```

The diagonal entries (variances) are recovered to within sampling
noise; the off-diagonals (covariances) match the rank-1 structure
encoded in `Lambda_B`. This is the **headline recovery check**:
if `diff` is large on a Tier-1 fixture, something is wrong before
you go any further.

## 3. Common harness mistakes

The verification fails for a *different reason* than the model
fails. The four mistakes that cost the most analysis time:

### 3a. Truth contains terms the fit doesn't (variance absorption)

If the DGP includes a phylogenetic random effect but the fitted
model omits `phylo_*()` terms, the phylogenetic variance gets
absorbed into `Lambda_B` (the fit "captures" what it can with the
machinery you gave it). Symptoms: `Lambda_B` looks bigger than the
truth; communality looks inflated; the phylo-signal extractor
reports zero (because the model has no phy tier).

**Fix**: match the fitted model to the DGP exactly during
verification. After verification passes, you can fit *more
parsimonious* models on real data, but the verification fit
should mirror the DGP's structure.

### 3b. Trait-factor level order silently mis-aligned

Long-format data has a `trait` factor column. R defaults to
alphabetical level ordering. If your DGP `Lambda_B` is indexed in
the order you simulated traits in (e.g. `("length", "mass",
"wing")`) but the data frame's trait factor sorts alphabetically
to `("length", "mass", "wing")` -- ok that order matches -- but
`("Beak.Length", "Beak.Width", "Beak.Depth", "Tarsus.Length")`
sorts to alphabetical `("Beak.Depth", "Beak.Length", "Beak.Width",
"Tarsus.Length")`. Then `Sigma_fit` and `Sigma_truth` have rows in
different orders. Recovery looks broken when only the row labels
are misaligned.

**Fix**: when building the long-format data frame, set
`levels = trait_names` explicitly on the trait factor so the
order matches your `Lambda_B` rows.

### 3c. Insufficient sample size for the rank you fit

A rank-2 latent factor needs more sites than a rank-1. If you fit
`latent(0 + trait | site, d = 2)` on `n_sites = 20`, the second
column of `Lambda_B` will look noisy in the recovery comparison
even when the engine is correct. Use
[`check_identifiability()`](../reference/check_identifiability.html)
to confirm: the spurious factor shows a column of near-zero
Procrustes-aligned residuals across simulate-refit replicates.

**Fix**: fit a smaller rank, get more data, or use
`lambda_constraint` to pin a structural zero.

### 3d. Wrong scale comparison for non-Gaussian traits

When the DGP includes a binomial trait, the "truth" loadings live
on the latent linear-predictor scale. Comparing them to the
fitted `Lambda_B` on the response scale will look like a 1000%
recovery error. The package's
[`extract_correlations(link_residual = "auto")`](../reference/extract_correlations.html)
default adds the family-specific latent-scale residual to the
implied $\boldsymbol\Sigma$ diagonal before computing correlations;
the recovery target should live on the same scale.

**Fix**: do all comparisons on the latent linear-predictor scale.

## 4. The user's verification checklist

Before treating a fit as publication-ready, run all four:

### 4a. `sanity_multi()` -- structural convergence

```{r sanity}
sanity_multi(fit)
```

If anything in `sanity_multi()` returns FALSE -- convergence,
positive-definite Hessian, no large gradient, no implausibly large
SEs -- stop here. The fit has not converged in a useful sense.

### 4b. `check_identifiability()` -- Procrustes recovery

`check_identifiability(fit, sim_reps = 100L)` simulates from the
fit, refits each replica, Procrustes-aligns the loadings, and
reports per-parameter recovery + Hessian rank checks. It catches
the **spurious extra factor** failure mode that no other diagnostic
catches: `pdHess = TRUE`, `sanity_multi()` passes, profile CIs
look tight -- but a column of $\boldsymbol{\Lambda}$ has near-zero
Procrustes-aligned residual magnitude because it's unidentifiable
in direction. See
[`check_identifiability()`](../reference/check_identifiability.html).

(Heavy: `sim_reps = 100` takes a few minutes on a Tier-1 fixture.
For routine checks use `sim_reps = 10`.)

### 4c. `gllvmTMB_check_consistency()` -- Laplace approximation

`gllvmTMB_check_consistency(fit, n_sim = 100L)` wraps
`TMB::checkConsistency()` and tests whether the approximate
marginal score is centred. A non-centred score is a sign that the
Laplace approximation is unreliable for *this* fit. See
[`gllvmTMB_check_consistency()`](../reference/gllvmTMB_check_consistency.html).

```{r check-consistency, eval = FALSE}
gllvmTMB_check_consistency(fit, n_sim = 50L, seed = 1)
```

(Not evaluated inline; the call is slow. Run it locally before
publication-grade reporting.)

### 4d. `coverage_study()` -- empirical CI coverage

`coverage_study(fit, n_reps = 30L)` is the audit's recommended
empirical-coverage check: simulate, refit, count CI hit rate
against the original fit's estimates, flag any rows below the
audit's `>= 94%` exit gate. See
[`coverage_study()`](../reference/coverage_study.html).

```{r coverage, eval = FALSE}
coverage_study(fit, parm = c("sigma_eps", "sd_B[1]", "sd_B[2]", "sd_B[3]"),
               n_reps = 30L, methods = c("wald"), seed = 1,
               progress = FALSE)
```

(Also slow. Run locally.)

## 5. Anatomy of a profile curve: when Wald CIs fail

The four diagnostics above answer "is the fit OK?" at the
overall-model level. For a single parameter,
[`confint_inspect()`](../reference/confint_inspect.html) is the
visual companion: it computes the full profile-likelihood curve
and overlays the MLE, the chi-squared threshold, the profile
bounds (solid green), and the Wald bounds (dashed orange) on the
natural scale.

Profile shape tells you which CI method to trust:

- **Quadratic + symmetric near the MLE**: Wald and profile
  agree. Either method is fine.
- **Asymmetric / skewed**: Wald is wrong; the profile bound is
  right.
- **Flat at the MLE**: weak identifiability; the profile is right
  and Wald hides the flatness. The CI is genuinely wide.
- **Hits the parameter range bound on one side**: boundary
  parameter (variance pinned at 0) or unidentified upper tail.
  Profile reports `+/-Inf`.
- **No crossing**: profile didn't converge. Switch to bootstrap.

```{r confint-inspect, eval = FALSE}
inspect <- confint_inspect(fit, parm = "sigma_eps")
inspect$bounds
inspect$plot
```

The four canonical failure-mode shapes are catalogued in
[`troubleshooting-profile`](troubleshooting-profile.html).

## Summary

A fit has passed verification when:

1. `sanity_multi()` returns clean (convergence, Hessian PD, no
   huge SEs).
2. `check_identifiability()` shows no spurious-factor collapse.
3. `gllvmTMB_check_consistency()` reports `"centred"`.
4. `coverage_study()` shows `passes_94pct = TRUE` on every target
   that matters for the paper.
5. `confint_inspect()` on the headline parameters shows
   well-behaved (`"quadratic"`) profile curves.

If any of these fail, the four common harness mistakes in section
3 are the most likely culprits before the model machinery itself
becomes suspect.

## See also

- [Choose your model](choose-your-model.html) -- pick the right
  model before you verify it.
- [Data-shape flowchart](data-shape-flowchart.html) -- which
  worked example matches your data.
- [Profile-likelihood CIs](profile-likelihood-ci.html) -- the CI
  machinery being audited.
- [Troubleshooting profile-likelihood CIs](troubleshooting-profile.html)
  -- four failure-mode catalogue.
- [Pitfalls](pitfalls.html) -- broader common-gotcha catalogue.
