---
title: "Simulation recovery: validated DGP grid"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Simulation recovery: validated DGP grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
m3_smoke_rds <- system.file("extdata", "m3-coverage-grid-smoke.rds",
                            package = "gllvmTMB", mustWork = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache    = FALSE,
  eval     = nzchar(m3_smoke_rds) && file.exists(m3_smoke_rds),
  message  = FALSE,
  warning  = FALSE
)
```

This article is for a reader who wants to know **whether the
confidence intervals reported by `gllvmTMB()` actually contain the
truth at their nominal rate**. The standard quantitative-genetic and
psychometric tutorials don't usually answer this question — they
demonstrate that fits *converge* and *recover point estimates*, but
not whether the CIs *cover*.

Coverage validation is the M3 milestone in `gllvmTMB`'s roadmap:
empirical coverage rates ≥ 94% (the audit-1 exit gate) at 95%
nominal across Gaussian, binomial, nbinom2, ordinal-probit, and
mixed-family fits.

The full M3 grid is described in
[Design 42](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/42-m3-dgp-grid.md):
5 family cells × 3 latent ranks × R = 200 replicates = 3000 fits,
with profile-likelihood CIs on each free parameter. This article
shows **the M3.2 smoke output** — a 12-cell × 10-rep × Wald-only
sanity-check version of that grid (4 families: Gaussian, binomial,
nbinom2, ordinal-probit, at $d \in \{1, 2, 3\}$ each). Mixed-family
is excluded from smoke pending M3.3's integration of the
`family = list(...)` + `family_var` lookup pattern. The numbers
below are **placeholder Wald** values; the M3.3 production grid
will replace them with profile-likelihood CIs and re-render this
article.

## The grid

Per replicate, the DGP samples a fresh
$\boldsymbol\Lambda_{\mathrm{true}} \sim \mathrm{Uniform}(-1.5, 1.5)^{5 \times d}$
and $\boldsymbol\psi_{\mathrm{true}} \sim \mathrm{Gamma}(2, 2)^5$,
simulates 60 units × 5 traits from that truth, fits `gllvmTMB` with
the matching `latent() + unique()` formula and family, and records
whether the fit's CI on each $\Sigma_{\mathrm{unit}, tt}$ diagonal
contains the generative truth. Per-family non-Gaussian responses:
binomial uses `rbinom(size = 1)`; nbinom2 uses `rnbinom(mu, size = phi)`
with `phi ~ Gamma(5, 5)`; ordinal-probit uses 4 categories with
quantile-spaced cutpoints.

## Reading the precomputed RDS

```{r read}
artefact <- readRDS(m3_smoke_rds)
artefact$meta[c("label", "gllvmTMB_ver", "n_cells", "n_reps", "elapsed_s")]
```

## Per-cell coverage summary

```{r summary}
artefact$summary
```

The `coverage_wald` column is the fraction of (rep × trait) pairs
whose Wald CI contained the truth. The `passes_94pct_wald` flag
checks against the M3 exit gate.

```{r smoke-honesty, results = "asis"}
n_reps_total <- nrow(artefact$grid)
n_pass <- sum(artefact$summary$passes_94pct_wald)
n_cells <- nrow(artefact$summary)
cat(sprintf(
  "**Smoke status**: %d / %d cells pass the 94%% gate at this Wald-placeholder level.\n",
  n_pass, n_cells
))
```

These rates are below the 94% gate because **the Wald CIs are a
placeholder** in the M3.2 pipeline — a rough 20% RSE heuristic that
is intentionally crude. The pipeline is a smoke test of the *plumbing*
(simulate → fit → extract → record), not a validation of inference.
The M3.3 production pass will:

1. Replace the placeholder Wald with proper delta-method Wald and
   profile-likelihood CIs on `extract_Sigma(level = "unit")`
   diagonals.
2. Expand from Gaussian-only to all 5 families (binomial, nbinom2,
   ordinal-probit, mixed-family).
3. Increase `n_reps` from 10 to 200 (Monte Carlo error ±3 pp at 95%
   nominal).
4. Compute the same machinery for derived quantities (communality,
   repeatability) — M3.5.

When M3.3 lands, this article will re-render with the production
numbers. The placeholder columns (`coverage_prof`, `ci_prof_lo`,
`ci_prof_hi`) are filled in then.

## Inspecting individual fits

For diagnostic purposes, you can pull a single rep's outputs:

```{r single-rep}
single <- subset(artefact$grid, cell == "gaussian-d2" & rep == 1L)
single[, c("trait_id", "truth_diag_sigma",
           "ci_wald_lo", "ci_wald_hi", "covered_wald")]
```

## Notes on per-family smoke behaviour

A few patterns visible in the smoke summary above:

- **Gaussian** cells run fastest (~0.5-0.7s per fit) and reach the
  highest placeholder-Wald coverage (54-82%). M3.3 will replace
  Wald with profile and expect coverage near 95% nominal.
- **Binomial** cells run at ~0.7-1s per fit; placeholder coverage
  is low (32-66%) because the 20% RSE heuristic is mis-calibrated
  for Bernoulli inferences. Proper profile CIs in M3.3 should
  recover near-nominal coverage.
- **nbinom2** cells are the slowest (~2-3s per fit) and show
  some non-convergence (45/50 or 40/50 reps converge) due to the
  challenging two-parameter (mean + dispersion) likelihood at the
  small `n = 60 units × 5 traits` regime. M3.3 will diagnose
  whether converged-set coverage is itself near-nominal.
- **Ordinal-probit** cells run at ~1-1.5s per fit; placeholder
  coverage in the 40-50% range. Cutpoint estimation adds K-1 = 3
  extra parameters per trait, so the small-n regime is taxing.
- **Mixed-family** is excluded from smoke — the `family = list(...)`
  + `attr(., 'family_var')` mixed-family API requires a `family_id`
  column in the long-format data frame; M3.3 wires that integration
  per the M1 mixed-family-fixture pattern
  (`inst/extdata/mixed-family-fixture.rds`).

Each row is a trait's $\Sigma_{\mathrm{unit}, tt}$ diagonal. The
`truth_diag_sigma` is the generative truth for that rep; the Wald
CI bounds and `covered_wald` flag are the placeholder smoke-test
results.

## Reproducing the smoke run

The pipeline lives at
[`dev/m3-grid.R`](https://github.com/itchyshin/gllvmTMB/blob/main/dev/m3-grid.R)
(library functions) +
[`dev/precompute-m3-grid.R`](https://github.com/itchyshin/gllvmTMB/blob/main/dev/precompute-m3-grid.R)
(driver). From the repo root:

```bash
# Smoke (Gaussian only, 10 reps × 3 cells): ~18s on a 2025 laptop
Rscript dev/precompute-m3-grid.R

# Smoke across all 5 families (15 cells × 10 reps): ~10-20 min serial
Rscript dev/precompute-m3-grid.R --all-fams

# Full production grid (15 cells × 200 reps): ~hours, M3.3 only
Rscript dev/precompute-m3-grid.R --full
```

Output goes to `dev/precomputed/m3-coverage-grid.rds` (long-format,
150 rows for the default smoke) and `dev/precomputed/m3-coverage-summary.rds`
(per-cell aggregate).

## What this article does NOT yet show

Per Design 42 §6 honest-scope boundaries, the M3 milestone covers:

- ✅ **Pipeline machinery (M3.2)** — this article, smoke output.
- ⚪ **Per-family profile CI accuracy (M3.3)** — replace Wald
  placeholder with proper inference; expand to all 5 families.
- ⚪ **Boundary regimes (M3.4)** — sparse-Bernoulli at d=3;
  ordinal-probit at d=2 with rare categories.
- ⚪ **Derived-quantity coverage (M3.5)** — communality,
  repeatability, phylogenetic signal.
- ⚪ **M3 close gate (M3.7)** — Ada + Shannon coordination audit.

Out of scope for M3 entirely:

- Model-misspecification coverage (mis-specified rank, omitted
  family). Different validation framework; v0.3.0 scope.
- Asymptotic large-n coverage (n = 600, n = 6000). v0.3.0 scaling
  study.
- Cross-package agreement. Phase 5.5 work.

## See also

- [Design 42](https://github.com/itchyshin/gllvmTMB/blob/main/docs/design/42-m3-dgp-grid.md)
  — full M3 grid specification (5 families × 3 dims × 200 reps).
- [`coverage_study()`](../reference/coverage_study.html) —
  per-fit parametric-bootstrap coverage estimator (complementary to
  this external-simulation grid).
- [`confint_inspect()`](../reference/confint_inspect.html) —
  visual diagnosis when coverage falls short.
- [Profile-likelihood CIs](profile-likelihood-ci.html) — the
  inference method M3.3 validates.
- [Troubleshooting profile](troubleshooting-profile.html) — common
  patterns when profile coverage is unexpectedly low.
- [Animal model](animal-model.html) — quantitative-genetic worked
  example using the same `latent() + unique()` machinery this
  article validates.
