---
title: "Multivariate spatial models with an SPDE mesh"
description: "Build a mesh, fit equivalent long- and wide-format Gaussian spatial models, interpret the practical range, and understand the current evidence boundary."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Multivariate spatial models with an SPDE mesh}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 5,
  fig.align = "center",
  cache = FALSE,
  eval = requireNamespace("fmesher", quietly = TRUE),
  message = FALSE,
  warning = FALSE
)
```

Measurements collected at nearby sites can resemble one another even after
the measured environment is in the model. This article answers a practical
question: **how can I fit several continuous traits with spatially structured
residual variation, while keeping the mesh and the trait covariance explicit?**

The first worked model gives each trait an independent spatial field with a
common practical range. It uses `spatial_indep()`: *spatial* identifies the
mesh-based covariance source, and *indep* fixes cross-trait spatial covariance
to zero while estimating a separate field scale for each trait.

The central long-format call is:

```{r quick-start, eval = FALSE}
fit_long <- gllvmTMB(
  value ~ 0 + trait + spatial_indep(0 + trait | site, mesh = mesh),
  data = spatial_long,
  trait = "trait",
  unit = "site",
  mesh = mesh
)
```

> **Evidence boundary.** Mesh construction, projection, finite-element
> matrices, and ingestion by a Gaussian spatial fit are covered (SPA-01).
> Evidence across the complete spatial keyword-by-family surface remains
> partial (FG-13). This article does not support directional anisotropy,
> barriers, spatiotemporal fields, or interval-coverage claims.

## Simulate a spatial teaching dataset

The teaching dataset has 100 sites, three continuous traits, a true practical
range of 0.3 coordinate units, and different spatial field variances across
traits. `simulate_site_trait()` returns stacked long data: each site appears
once per trait.

```{r setup}
library(gllvmTMB)

spatial_sim <- simulate_site_trait(
  n_sites = 100,
  n_species = 1,
  n_traits = 3,
  mean_species_per_site = 1,
  spatial_range = 0.3,
  sigma2_spa = c(0.4, 0.6, 0.3),
  seed = 128
)

spatial_long <- spatial_sim$data
head(spatial_long[c("site", "trait", "value", "lon", "lat")])
```

For this simulation, `lon` and `lat` are normalized equal-distance
coordinates on a unit square. Real longitude and latitude are angular
coordinates, so project them before constructing a distance-based mesh:

```{r project-coordinates, eval = FALSE}
spatial_long <- add_utm_columns(
  spatial_long,
  ll_names = c("longitude", "latitude"),
  utm_names = c("X", "Y"),
  units = "km"
)
```

If sites span several UTM zones or both hemispheres, choose and justify a
study-appropriate projected CRS instead of accepting an automatic zone
silently.

## Build and inspect the mesh

`make_mesh()` constructs the triangular mesh, the observation-to-mesh
projection matrix, and the finite-element matrices used by the SPDE engine.
The `cutoff` is in the same units as the supplied coordinates; increasing it
usually produces a coarser mesh.

```{r build-mesh}
mesh <- make_mesh(
  spatial_long,
  xy_cols = c("lon", "lat"),
  cutoff = 0.07
)
```

The compact summary below confirms that all three SPDE finite-element matrices
are present without printing their sparse entries or every coordinate row.

```{r mesh-summary, echo = FALSE}
mesh_summary <- data.frame(
  component = c(
    "coordinate rows (loc_xy)",
    "mesh vertices",
    "projection matrix (A_st)",
    "SPDE mass matrix (c0)",
    "SPDE stiffness matrix (g1)",
    "SPDE curvature matrix (g2)"
  ),
  size = c(
    paste(dim(mesh$loc_xy), collapse = " x "),
    paste(mesh$mesh$n, "vertices"),
    paste(dim(mesh$A_st), collapse = " x "),
    vapply(
      mesh$spde[c("c0", "g1", "g2")],
      function(x) paste(dim(x), collapse = " x "),
      character(1)
    )
  )
)
knitr::kable(mesh_summary, col.names = c("Component", "Size"))
```

Access `mesh$spde$c0`, `mesh$spde$g1`, or `mesh$spde$g2` directly when you need
to inspect a particular matrix.

```{r plot-mesh}
plot(mesh)
```

Inspect the mesh before fitting. It should cover every sampling location
without a fringe of extremely narrow triangles. A mesh that is too coarse can
erase short-range structure; a needlessly fine mesh adds latent variables and
can make optimization slower.

The projection matrix must follow the rows of the internally stacked data.
The long data already have that order. If your data start wide, make an
equivalent row-major coordinate frame before calling `make_mesh()`:

```{r wide-mesh-data, eval = FALSE}
trait_cols <- c("trait_1", "trait_2", "trait_3")
mesh_data <- spatial_wide[
  rep(seq_len(nrow(spatial_wide)), each = length(trait_cols)),
  c("lon", "lat"),
  drop = FALSE
]
mesh <- make_mesh(mesh_data, c("lon", "lat"), cutoff = 0.07)
```

This explicit alignment protects against the most consequential mesh failure:
a projection matrix whose rows no longer correspond to the fitted response
cells.

## Fit the same model in long and wide form

For trait $t$ at site $i$, the worked model is
$y_{it} = \alpha_t + u_t(s_i) + \varepsilon_{it}$. The fields $u_t$ are
independent across traits. Their shared mesh geometry is represented by

$$
Q(\kappa) = \kappa^4 M_0 + 2\kappa^2 M_1 + M_2,
$$

where the mesh supplies $M_0$, $M_1$, and $M_2$. The parameter $\kappa$
controls the practical range; trait-specific scale parameters control how
strongly each field varies.

```{r fit-long}
fit_long <- gllvmTMB(
  value ~ 0 + trait + spatial_indep(0 + trait | site, mesh = mesh),
  data = spatial_long,
  trait = "trait",
  unit = "site",
  mesh = mesh,
  silent = TRUE
)

fit_long$opt$convergence
```

Some readers store one response column per trait. The compact `traits()`
formula uses `spatial_indep(1 | site)`; the formula expander creates the same
trait-specific intercept and spatial-field design as the long call.

```{r make-wide}
spatial_wide <- reshape(
  spatial_long[c("site", "lon", "lat", "trait", "value")],
  idvar = c("site", "lon", "lat"),
  timevar = "trait",
  direction = "wide"
)
names(spatial_wide) <- sub("value.", "", names(spatial_wide), fixed = TRUE)
row.names(spatial_wide) <- NULL
```

```{r fit-wide}
fit_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~
    1 + spatial_indep(1 | site, mesh = mesh),
  data = spatial_wide,
  unit = "site",
  mesh = mesh,
  silent = TRUE
)

c(
  long_convergence = fit_long$opt$convergence,
  wide_convergence = fit_wide$opt$convergence,
  long_logLik = unname(logLik(fit_long)),
  wide_logLik = unname(logLik(fit_wide))
)
```

The two calls are alternative data interfaces, not different statistical
models. Use long data when trait labels are already rows; use `traits()` when
the responses are columns.

## Interpret the practical range

The current spatial engine assumes isotropy: dependence changes with distance
but not direction. `plot_anisotropy()` therefore displays a circle and reports
the fitted practical range; it does **not** estimate directional anisotropy.

```{r range-summary}
data.frame(
  true_range = spatial_sim$truth$spatial_range,
  fitted_range = sqrt(8) / as.numeric(fit_long$report$kappa[1]),
  assumption = "isotropic (H = I)"
)
```

```{r range-plot}
invisible(plot_anisotropy(fit_long))
```

The fitted value is one point estimate from one simulated dataset. Agreement
with the generating value is a teaching check, not evidence of unbiased range
estimation or calibrated uncertainty across designs.

Run the ordinary fit diagnostics before interpreting spatial parameters:

```{r diagnostics}
invisible(check_gllvmTMB(fit_long))
```

## Choose the spatial covariance mode

All four spatial keywords use the same mesh and SPDE geometry. They answer
different trait-covariance questions.

| Keyword | Spatial covariance question |
|---|---|
| `spatial_scalar()` | Should all traits share one spatial field variance? |
| `spatial_indep()` | Should traits have separate field variances but zero cross-trait spatial covariance? |
| `spatial_latent(..., d = K)` | Can $K$ shared spatial axes describe cross-trait spatial covariance? |
| `spatial_dep()` | Should the model estimate a full cross-trait spatial covariance? |

Start with the simplest mode that answers the biological question. Do not add
`spatial_indep()` to `spatial_latent()` or `spatial_dep()` in the same formula;
those combinations duplicate spatial variance and fail explicitly.

Fit the reduced-rank and full-covariance alternatives by changing only the
spatial term:

```{r correlated-spatial-fits}
fit_latent <- gllvmTMB(
  value ~ 0 + trait +
    spatial_latent(0 + trait | site, d = 2, unique = TRUE, mesh = mesh),
  data = spatial_long,
  trait = "trait",
  unit = "site",
  mesh = mesh,
  silent = TRUE
)

fit_dep <- gllvmTMB(
  value ~ 0 + trait + spatial_dep(0 + trait | site, mesh = mesh),
  data = spatial_long,
  trait = "trait",
  unit = "site",
  mesh = mesh,
  silent = TRUE
)
```

```{r spatial-mode-comparison, echo = FALSE}
spatial_comparison <- data.frame(
  mode = c("indep", "latent (d = 2, unique = TRUE)", "dep"),
  convergence = c(
    fit_long$opt$convergence,
    fit_latent$opt$convergence,
    fit_dep$opt$convergence
  ),
  log_likelihood = c(logLik(fit_long), logLik(fit_latent), logLik(fit_dep)),
  AIC = c(AIC(fit_long), AIC(fit_latent), AIC(fit_dep))
)

knitr::kable(spatial_comparison, digits = 1)
```

Here `d = 2` requests two shared spatial axes, while `unique = TRUE` retains a
trait-specific spatial diagonal. `spatial_dep()` estimates the full
cross-trait spatial covariance. The data were generated with independent
trait fields, so this table demonstrates runnable syntax and relative fit on
one dataset; it is not a recovery or model-selection study. These broader
models remain partial under FG-13/SPA-02, so treat current estimates as
model-based point estimates rather than calibrated spatial-correlation
inference.

## Independence, acknowledgement, and licensing

`gllvmTMB` does not import, link to, or include `sdmTMB`; `sdmTMB` is not a
runtime or build dependency. We thank the sdmTMB authors because their
user-facing spatial interface inspired the original gllvmTMB interface. That
is a courtesy acknowledgement, not a statement that gllvmTMB contains or
adapts sdmTMB source code.

The current helpers in `R/mesh.R`, `R/crs.R`, and `R/plot.R` were independently
authored for gllvmTMB from the published SPDE/GMRF construction and the public
`fmesher` API. They are gllvmTMB code distributed under
[GPL-3](../LICENSE.html). The repository records the detailed code-provenance
boundary in [`inst/COPYRIGHTS`](https://github.com/itchyshin/gllvmTMB/blob/main/inst/COPYRIGHTS).

Users should cite gllvmTMB and TMB through `citation("gllvmTMB")`. The sdmTMB
acknowledgement does not create a separate required citation for using these
independently implemented helpers.

## What to do next

| If your question is... | Continue with... |
|---|---|
| Which spatial covariance mode matches my estimand? | [Formula keyword grid](api-keyword-grid.html) |
| Do species still co-occur after environmental predictors? | [Joint species distribution models](joint-sdm.html) |
| Can I trust this fitted object numerically? | [Fit diagnostics](fit-diagnostics.html) |
| How does the SPDE keyword map to the engine? | [`spatial_indep()` reference](../reference/spatial_indep.html) |

Single-response spatial models belong in
[`sdmTMB`](https://pbs-assess.github.io/sdmTMB/). Directional anisotropy,
barrier meshes, and spatiotemporal fields are not implemented in the current
gllvmTMB spatial helper contract.

## References

Lindgren F, Rue H, Lindström J (2011). An explicit link between Gaussian
fields and Gaussian Markov random fields: the stochastic partial differential
equation approach. *Journal of the Royal Statistical Society: Series B*, 73,
423--498. <https://doi.org/10.1111/j.1467-9868.2011.00777.x>.
