---
title: "Why gllvmTMB: SPDE vs glmmTMB exp() spatial covariance"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Why gllvmTMB: SPDE vs glmmTMB exp() spatial covariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  fig.align = "center",
  cache     = FALSE,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

This article fits the **same** spatial multivariate model two ways and
times each fit. The point is to make the case for the sparse SPDE
engine concrete: when does it beat the dense `glmmTMB::exp()`
covariance path, and by how much?

For the broader case that a spatial random field is the same
statistical object as a phylogenetic random effect -- both
correlated-random-effect models on a distance metric -- see
Mizuno, Williams, Lagisz, Senior & Nakagawa (2026), *A unified
framework for phylogenetic and spatial meta-analysis*, EcoEvoRxiv,
<https://doi.org/10.32942/X2WT1B>. The benchmark below is on the
spatial side.

The model is a long-format multi-trait regression with one
independent Gaussian random field per trait:

* **gllvmTMB**: `value ~ 0 + trait + spatial_unique(0 + trait | coords)`
  -- sdmTMB-style sparse SPDE / GMRF precision matrix
  $Q = \kappa^4 M_0 + 2\kappa^2 M_1 + M_2$ built once on the R side,
  with $n_{\text{mesh}} \times n_{\text{mesh}}$ non-zeros
  (Lindgren, Rue & LindstrĂ¶m 2011).
* **glmmTMB**: `value ~ 0 + trait + exp(pos + 0 | trait)` -- dense
  exponential covariance matrix $V[i, j] = \sigma^2 \exp(-\|p_i - p_j\| / \rho)$,
  one $n \times n$ matrix per trait.

The two engines fit covariance models in the same MatĂ©rn family but
at different smoothnesses (the SPDE here approximates MatĂ©rn
$\nu = 1$; `exp()` is exactly MatĂ©rn $\nu = 0.5$). They are not
numerically identical, but they are directly comparable for the
question this article asks: **how does fit time scale with $n$?**

## Setup

```{r setup}
library(gllvmTMB)
library(glmmTMB)
library(ggplot2)
```

## Benchmark code

This is the function that produces the benchmark numbers. We set
`eval = FALSE` here so the article renders quickly; the actual
numbers shown below were produced on the maintainer's hardware and
are reproduced as a literal data frame.

```{r benchmark-code, eval = FALSE}
bench_one <- function(n_sites, n_traits = 3, seed = 1, reps = 3) {
  sim <- simulate_site_trait(
    n_sites               = n_sites,
    n_species             = 1,
    n_traits              = n_traits,
    mean_species_per_site = 1,
    Lambda_B              = matrix(0.0, n_traits, 1),
    psi_B                 = rep(0.1, n_traits),
    spatial_range         = 0.3,
    sigma2_spa            = rep(0.4, n_traits),
    seed                  = seed
  )
  df   <- sim$data
  mesh <- make_mesh(df, c("lon", "lat"), cutoff = 0.07)
  df$pos <- glmmTMB::numFactor(df$lon, df$lat)

  t_gllvm <- replicate(reps, {
    t0 <- Sys.time()
    gllvmTMB(value ~ 0 + trait + spatial_unique(0 + trait | coords),
             data = df, mesh = mesh, family = gaussian())
    as.numeric(Sys.time() - t0, units = "secs")
  })
  t_glmm <- replicate(reps, {
    t0 <- Sys.time()
    suppressWarnings(glmmTMB::glmmTMB(
      value ~ 0 + trait + exp(pos + 0 | trait),
      data = df, family = gaussian(), REML = FALSE))
    as.numeric(Sys.time() - t0, units = "secs")
  })

  data.frame(n_sites = n_sites, n_obs = nrow(df), n_mesh = mesh$mesh$n,
             gllvmTMB_s = median(t_gllvm), glmmTMB_s = median(t_glmm),
             speedup = median(t_glmm) / median(t_gllvm))
}

set.seed(1)
res <- do.call(rbind, lapply(c(40, 80, 160), bench_one))
```

## Benchmark results (reproduced literally)

```{r load-bench}
res <- data.frame(
  n_sites    = c(40,         80,         160),
  n_obs      = c(120,        240,        480),
  n_mesh     = c(64,         74,         102),
  gllvmTMB_s = c(0.2398331,  0.1216240,  0.1632252),
  glmmTMB_s  = c(0.1872749,  0.5952530,  1.0435259),
  speedup    = c(0.7808552,  4.8942069,  6.3931678)
)
res
```

## Plot: fit time vs `n_sites`

```{r plot, fig.cap = "Median fit time over 3 replications. Both axes log-scale. SPDE engine in blue, glmmTMB::exp() in red."}
res_long <- rbind(
  data.frame(n_sites = res$n_sites, time_s = res$gllvmTMB_s,
             engine  = "gllvmTMB::spatial_unique() -- sparse SPDE (Matern nu=1)"),
  data.frame(n_sites = res$n_sites, time_s = res$glmmTMB_s,
             engine  = "glmmTMB::exp() -- dense exponential (Matern nu=1/2)")
)

ggplot(res_long, aes(n_sites, time_s, colour = engine, shape = engine)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  scale_x_log10(breaks = res$n_sites) +
  scale_y_log10() +
  scale_colour_manual(values = c("#1f78b4", "#e31a1c")) +
  labs(x = "Number of sites", y = "Median fit time (seconds, log)",
       colour = NULL, shape = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.direction = "vertical")
```

## What we found

```{r summary, results = "asis"}
cat(sprintf(
  "* At **n_sites = %d** (%d observations, mesh size %d), `glmmTMB::exp()` is %.2fx faster than `gllvmTMB::spatial_unique()`. The SPDE engine carries fixed mesh-setup overhead that dominates at small n.\n",
  res$n_sites[1], res$n_obs[1], res$n_mesh[1], 1 / res$speedup[1]))
cat(sprintf(
  "* At **n_sites = %d** (%d observations, mesh size %d), `gllvmTMB::spatial_unique()` is %.2fx faster.\n",
  res$n_sites[2], res$n_obs[2], res$n_mesh[2], res$speedup[2]))
cat(sprintf(
  "* At **n_sites = %d** (%d observations, mesh size %d), `gllvmTMB::spatial_unique()` is %.2fx faster.\n",
  res$n_sites[3], res$n_obs[3], res$n_mesh[3], res$speedup[3]))
```

The **crossover** -- the $n$ at which the two engines take the same
time -- sits between `n_sites = 40` and `n_sites = 80` on this
hardware. Below the crossover, `glmmTMB::exp()` wins because its
dense $n \times n$ matrix is small enough that fixed mesh-setup
overhead dominates the SPDE path. Above the crossover the dense-
matrix cost grows faster than the sparse-precision cost and the SPDE
engine pulls ahead -- by **`r round(res$speedup[3], 1)`Ă— at
n_sites = `r res$n_sites[3]`** on this hardware.

## What's actually different

```{r whats-different, eval = FALSE}
# gllvmTMB (sdmTMB style)  : Q_spde = kappa^4 M0 + 2 kappa^2 M1 + M2
#                            sparse, n_mesh x n_mesh, GMRF density
# glmmTMB (exp covariance) : V[i, j] = sigma^2 exp(-||p_i - p_j|| / range)
#                            dense, n_pos x n_pos
```

The sparse-precision approach evaluates `n_mesh` basis functions on
a triangular mesh and projects the field to observation locations
via a sparse $n \times n_{\text{mesh}}$ matrix $A$. The dense-
covariance approach builds an $n \times n$ distance matrix per
trait and Cholesky-factors it. Asymptotic complexity differs
accordingly:

* SPDE / GMRF: $O(n_{\text{mesh}}^{1.5})$ for sparse Cholesky, plus
  one sparse mat-vec per data point.
* Dense `exp()`: $O(n^3)$ Cholesky.

When $n_{\text{mesh}}$ is held roughly constant (controlled by
`cutoff`) and $n$ grows, the SPDE path's per-iteration cost grows
much more slowly.

## Caveats

* The two engines fit MatĂ©rn fields at **different smoothnesses**
  ($\nu = 1$ for SPDE, $\nu = 0.5$ for `exp()`). The point-estimates
  of range and $\sigma$ are not directly comparable. The runtime
  comparison is.
* The benchmark above runs **3 replications per cell** to keep
  vignette build time short. For a publication-grade timing,
  increase `reps` and use `bench::mark()` or
  `microbenchmark::microbenchmark()`.
* Mesh size depends on `cutoff` and on the spatial extent of the
  data, not just on `n_sites`. With finer meshes the SPDE engine
  slows down; with coarser meshes its accuracy degrades. The
  `cutoff = 0.07` used here produces mesh sizes between
  `r res$n_mesh[1]` and `r res$n_mesh[3]`, which is well below
  $n_{\text{sites}}$ at every grid point.

## Reproducibility

The benchmark code above (`bench_one()`) is the canonical timing
function. The numbers in the table are pre-computed values from one
run on the maintainer's hardware. To re-run on your machine, install
`gllvmTMB` and `glmmTMB` and execute the `benchmark-code` chunk
above. Numbers will vary with hardware, BLAS, and TMB version. The
shape of the curves is the invariant claim.

## See also

* The spatial covstruct family: `?spatial_scalar`, `?spatial_unique`,
  `?spatial_indep`, `?spatial_latent`, `?spatial_dep`.
* [Functional biogeography](functional-biogeography.html) -- the
  SPDE term as part of the full stacked-trait decomposition.
* [Stacked-trait GLLVM](stacked-trait-gllvm.html) -- the foundational
  worked example showing all five covariance modes including
  `spatial_unique()` together.
* Anderson SC, Ward EJ, English PA, Barnett LAK, Thorson JT (2025).
  *sdmTMB: an R package for fast, flexible, and user-friendly
  generalized linear mixed effects models with spatial and
  spatiotemporal random fields.* Journal of Statistical Software
  115 (2): 1--46. <https://doi.org/10.18637/jss.v115.i02>
* Lindgren F, Rue H, LindstrĂ¶m J (2011). *An explicit link between
  Gaussian fields and Gaussian Markov random fields: the SPDE
  approach.* Journal of the Royal Statistical Society B
  73 (4): 423--498. <https://doi.org/10.1111/j.1467-9868.2011.00777.x>
