---
title: "Stacked-trait GLLVMs with gllvmTMB"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Stacked-trait GLLVMs with gllvmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 4,
  fig.align  = "center",
  cache      = FALSE,
  eval       = TRUE,
  message    = FALSE,
  warning    = FALSE
)
```

```{r setup}
library(gllvmTMB)
library(ape)
```

This is the **foundational** worked example for the package. It walks
through the canonical full-decomposition `gllvmTMB()` fit -- a
stacked-trait, multivariate generalised linear latent variable model
with fast SPDE spatial fields and (optionally) phylogenetic random
effects.

## The scientific question

A single fit answers a multi-part question that ecology has historically
needed several separate analyses for:

1. **Between-site functional integration**: do trait values *covary*
   across sites? When site A is high in trait 1, is it also high in
   trait 2? (`latent + unique` on `site`)
2. **Within-site functional integration**: when site A's species
   deviate from site A's average, do those deviations covary? (`latent +
   unique` on `site_species`)
3. **Phylogenetic signal**: is some of the species-level variance
   driven by shared ancestry? (`phylo_scalar(species)`)
4. **Residual species effects**: what's left over after accounting for
   phylogeny? (`unique(species)`)
5. **Residual spatial autocorrelation**: do nearby sites' trait values
   look similar after accounting for the predictors and the
   above? (`spatial_unique(coords)`)

The corresponding decomposition is

$$y_{sit} = \alpha_t + \mathbf{x}_s^\top \boldsymbol{\beta}_t
  + r_{st} + u_{st} + e_{sit} + p_{it} + q_{it} + \epsilon_{sit}$$

where:

* $\mu_{st} = \alpha_t + \mathbf{x}_s^\top \boldsymbol{\beta}_t$ -- trait-specific fixed effects;
* $r_{st}$ -- residual spatial autocorrelation among sites for trait $t$;
* $u_{st}$ -- non-spatial between-site axes (global functional integration);
* $e_{sit}$ -- within-site species deviations (local functional integration);
* $p_{it}$ -- phylogenetically structured species effect;
* $q_{it}$ -- non-phylogenetic species effect;
* $\epsilon_{sit}$ -- per-observation residual (Gaussian here).

The reduced-rank decomposition $\Sigma = \boldsymbol{\Lambda}\boldsymbol{\Lambda}^\top + \boldsymbol{\Psi}$
is the stacked-trait GLLVM workhorse (Hui et al. 2017,
*Methods in Ecology and Evolution*; Niku et al. 2017,
*Environmental and Ecological Statistics*) using $\boldsymbol{\Psi}$ for
the diagonal trait-unique-variance matrix (factor-analysis / SEM
convention; Bollen 1989, Mulaik 2010). The SPDE-mesh spatial term
$r_{st}$ is inherited from `sdmTMB` (Anderson et al. 2025,
*Journal of Open Source Software*).

## 1. Simulate a dataset

`simulate_site_trait()` is a flexible simulator that follows the full
decomposition above. We pick a small but realistic example: 4 traits,
60 sites, 12 species, with both global (rank-2) and local (rank-1)
covariance plus a Brownian-motion phylogeny.

```{r sim}
set.seed(42)

# Phylogeny
n_sp  <- 12
tree  <- ape::rcoal(n_sp); tree$tip.label <- paste0("sp", seq_len(n_sp))
Cphy  <- ape::vcv(tree, corr = TRUE)

sim <- simulate_site_trait(
  n_sites               = 60,
  n_species             = n_sp,
  n_traits              = 4,
  mean_species_per_site = 6,
  Lambda_B              = matrix(c(1.0, 0.7, -0.3, 0.5,
                                   0.3, -0.5, 0.8, 0.2),
                                 nrow = 4, ncol = 2),
  psi_B                 = c(0.3, 0.3, 0.3, 0.3),
  Lambda_W              = matrix(c(0.6, 0.3, -0.2, 0.4), nrow = 4, ncol = 1),
  psi_W                 = c(0.4, 0.4, 0.4, 0.4),
  Cphy                  = Cphy,
  sigma2_phy            = c(0.5, 0.5, 0.5, 0.5),
  sigma2_sp             = c(0.2, 0.2, 0.2, 0.2),
  spatial_range         = 0.3,
  sigma2_spa            = c(0.4, 0.4, 0.4, 0.4),
  seed                  = 42
)

df <- sim$data
levels(df$species) <- paste0("sp", seq_len(n_sp))
str(df)
```

## 2. Fit the full model

```{r fit, eval = FALSE}
mesh <- make_mesh(df, c("lon", "lat"), cutoff = 0.07)

fit <- gllvmTMB(
  value ~ 0 + trait + (0 + trait):env_1 + (0 + trait):env_2 +
          spatial_unique(0 + trait | coords, mesh = mesh) +
          latent(0 + trait | site, d = 2) +
          unique(0 + trait | site) +
          latent(0 + trait | site_species, d = 1) +
          unique(0 + trait | site_species) +
          unique(0 + trait | species) +
          phylo_scalar(species, vcv = Cphy),
  data = df
)
```

Each formula term carries a specific role:

- `0 + trait + (0 + trait):env_*` -- trait-specific intercepts and slopes.
- `spatial_unique(0 + trait | coords)` -- fast SPDE spatial term using
  the precomputed `mesh`. Replaces the slow dense
  `glmmTMB::glmmTMB(... + exp(coords + 0 | trait))` /
  `... + gau(...)` paths.
- `latent(0 + trait | site, d = 2)` + `unique(0 + trait | site)` --
  global reduced-rank + diagonal covariance giving
  $\boldsymbol{\Sigma}_{\text{unit}} = \boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^\top + \boldsymbol{\Psi}_B$.
- `latent(0 + trait | site_species, d = 1)` + `unique(0 + trait | site_species)`
  -- local reduced-rank + diagonal covariance giving
  $\boldsymbol{\Sigma}_{\text{unit\_obs}} = \boldsymbol{\Lambda}_W \boldsymbol{\Lambda}_W^\top + \boldsymbol{\Psi}_W$.
- `unique(0 + trait | species)` -- non-phylogenetic species term
  $q_{it}$.
- `phylo_scalar(species, vcv = Cphy)` -- phylogenetic random effect
  $p_{it}$ using an in-keyword `vcv = Cphy` argument.

The `latent + unique` pair is the canonical decomposition mode; see
[the GLLVM vocabulary article](gllvm-vocabulary.html) for the term
glossary and [the API keyword grid](api-keyword-grid.html) for the
full $3\times 5$ keyword grid that organises the formula language.

## 3. Biological summaries

```{r extract, eval = FALSE}
extract_Sigma(fit, level = "unit")$R       # global trait correlation matrix
extract_Sigma(fit, level = "unit_obs")$R   # local trait correlation matrix
extract_ICC_site(fit)                       # site-level ICC per trait
extract_communality(fit, "unit")            # global communality H^2
extract_communality(fit, "unit_obs")        # local communality
ord <- extract_ordination(fit, "unit")
head(ord$scores)                            # site scores on global functional axes
ord$loadings                                # trait loadings
```

Each extractor returns a tidy summary of one biological question. For
the inferential surface (Wald / profile / bootstrap CIs on these
derived quantities) see
[Profile-likelihood CIs](profile-likelihood-ci.html). For the
verification surface (does the fit actually recover what it claims?)
see [Simulation-based verification](simulation-verification.html).

## 4. Two-stage workflow for species-level trait data

When trait values are species-level constants (typical of AVONET-style
trait data: one beak length per species, not multiple replicates per
species per site), the one-shot fit above is **poorly informed about
between-site covariance** -- the data simply don't contain enough
sites-of-the-same-species to estimate global functional integration
directly. The recommended workflow is a **two-stage analysis**:

* **Stage 1**: fit one `gllvmTMB()` model per trait with
  `value ~ 0 + site + phylo_scalar(species, vcv = Cphy)`,
  returning site-level estimates with their sampling variances.
* **Stage 2**: fit the multivariate model on the stage-1 estimates
  using `meta_known_V(value, V = V)` so the stage-1 sampling
  uncertainty propagates into the stage-2 covariance estimates.

Both stages use direct `gllvmTMB()` calls. See
[the corvidae-two-stage worked example](corvidae-two-stage.html) for
the complete pipeline.

## 5. Cross-validation against `glmmTMB`

For the non-spatial covstructs (`latent()`, `unique()`,
`phylo_scalar()`, `meta_known_V()` -- which map internally to
`glmmTMB`'s `rr` / `diag` / `propto` / `equalto`),
`gllvmTMB`'s log-likelihood matches `glmmTMB::glmmTMB()` exactly to
TMB tolerance; see `tests/testthat/test-stage[2-3]-*.R` in the source
for worked examples. The advantage over `glmmTMB` is that the same fit
can include `spatial_unique(0 + trait | coords)` for sparse-precision
spatial fields, which is dramatically faster than `exp()` / `gau()`
on more than a few hundred sites (Anderson et al. 2025).

## References

* Anderson SC, Ward EJ, English PA, Barnett LAK, Thorson JT (2025).
  *sdmTMB: An R package for fast, flexible, and user-friendly
  generalized linear mixed effects models with spatial and
  spatiotemporal random fields.* Journal of Open Source Software,
  10 (109), 7536. <https://doi.org/10.21105/joss.07536>
* Bollen KA (1989). *Structural Equations with Latent Variables.*
  Wiley.
* Hui FKC, Taskinen S, Pledger S, Foster SD, Warton DI (2017).
  *Model-based approaches to unconstrained ordination.* Methods in
  Ecology and Evolution, 6 (4), 399--411.
* Mulaik SA (2010). *Foundations of Factor Analysis*, 2nd ed. CRC
  Press.
* Niku J, Warton DI, Hui FKC, Taskinen S (2017). *Generalized linear
  latent variable models for multivariate count and biomass data in
  ecology.* Journal of Agricultural, Biological, and Environmental
  Statistics, 22 (4), 498--522.
