---
title: "How strongly does a known source connect groups?"
description: "Separate source correlation from trait covariance using fixed attenuation and the admitted replicated Gaussian estimator."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{How strongly does a known source connect groups?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE,comment="#>")
library(gllvmTMB)
```

Suppose we measure six continuous traits in three individuals from each species.
A phylogenetic tree supplies the pattern of relationships between species.
We want to distinguish two questions: how much species vary in their traits,
and how closely correlations between species follow that supplied pattern.

Structured-term `rho` addresses the second question. Fixed values are supported
for one phylogenetic, animal, dense-kernel or spatial trait-intercept covariance block.
The initial estimator requires at least two traits, complete replicated Gaussian
observations, one known source, and an independent observation residual. It does not admit a
competing ordinary covariance on those units. Spatial range remains a separate
estimated parameter; spatial range–rho recovery is not established. Augmented
slopes and source-strength confidence intervals are outside this workflow.
The examples illustrate use; a single simulated dataset does not establish
recovery accuracy.

## What rho changes

Let $K$ be the source covariance after the package's existing source resolution,
and let $D$ contain its diagonal. The covariance used between source levels is

\[
K_\rho=\rho K+(1-\rho)D.
\]

At `rho = 1`, the supplied relationships remain unchanged. At `rho = 0`,
different species are independent at this tier. Interior values reduce the
between-species covariance while preserving each species' marginal variance.
Non-unit diagonals remain non-unit: rho introduces no new normalization.

The trait covariance is separate. With `phylo_dep()` it is an unrestricted
matrix $\Sigma$. With `phylo_latent(..., unique = TRUE)` it is
$\Sigma=\Lambda\Lambda^\top+\Psi$. In both cases the full covariance is
$K_\rho\otimes\Sigma$. The same rho acts on the shared loading component and
the diagonal Psi component. An ordinary random effect remains a separate
component; its variance is not absorbed into rho.

If the scientific question is how phylogenetic and non-phylogenetic variation
differ, keep the phylogenetic source at `rho = 1` and use separate covariance
components. At the species tier, that model has covariance

\[
K\otimes\Sigma_{\mathrm{phy}}+I\otimes\Sigma_{\mathrm{non}}.
\]

The two trait covariance matrices can have different correlations, loadings
and variance patterns. Setting the phylogenetic term's rho to one does not
set the phylogenetic share of total variance to one: the non-phylogenetic
component and observation residuals still contribute variation.

The single-rho model below answers a more restricted question. When $D=I$,
its two species-tier components have trait covariances $\rho\Sigma$ and
$(1-\rho)\Sigma$: they differ in scale but share the same trait correlations.
With non-unit $D$, the independent component also inherits its known
species-specific variance scaling. Estimated rho is therefore an optional
source-strength model, not a substitute for estimating flexible, separate
phylogenetic and non-phylogenetic covariance components.

## Make a replicated example

This example generates the data directly from the covariance equation. The
three individuals per species share a species effect but have independent
Gaussian observation errors. Here `obs` identifies an individual's complete
six-trait vector; `species` identifies the levels connected by the tree.

```{r data}
set.seed(8301)
n_species <- 80L
n_traits <- 6L
tree <- ape::rcoal(n_species)
tree$tip.label <- sprintf("sp%03d",seq_len(n_species))
K <- ape::vcv(tree,corr=TRUE)
K_rho <- .6*K
diag(K_rho) <- diag(K)
loading <- c(.7,.65,-.6,.8,-.55,.75)
psi <- c(.18,.25,.22,.2,.3,.16)
Sigma <- tcrossprod(loading)+diag(psi)
species_effect <- t(chol(K_rho)) %*%
  matrix(rnorm(n_species*n_traits),n_species,n_traits) %*% chol(Sigma)

long <- expand.grid(trait=paste0("t",seq_len(n_traits)),individual=1:3,
                    species=rownames(K))
long$species <- factor(long$species,levels=rownames(K))
long$obs <- interaction(long$species,long$individual,drop=TRUE)
mu <- c(.1,-.2,.25,.05,-.15,.2)
long$value <- mu[as.integer(long$trait)] +
  species_effect[cbind(as.integer(long$species),as.integer(long$trait))] +
  rnorm(nrow(long),sd=.6)
wide <- reshape(long[c("obs","species","trait","value")],
                idvar=c("obs","species"),timevar="trait",direction="wide")
names(wide) <- sub("^value\\.","",names(wide))
```

## Fix the source strength

For a sensitivity analysis, hold rho at a chosen value and estimate the trait
covariance and observation residual. These long and wide calls describe the
same model. `traits()` supplies the trait intercepts in the wide formula.

```{r fixed-long}
set.seed(8401)
fixed_long <- gllvmTMB(
  value ~ 0+trait+phylo_dep(0+trait|species,tree=tree,rho=.6),
  data=long,trait="trait",unit="obs",cluster="species",family=gaussian(),
  control=gllvmTMBcontrol(n_init=1))
```

```{r fixed-wide}
set.seed(8401)
fixed_wide <- gllvmTMB(
  traits(t1,t2,t3,t4,t5,t6) ~ 1+phylo_dep(1|species,tree=tree,rho=.6),
  data=wide,unit="obs",cluster="species",family=gaussian(),
  control=gllvmTMBcontrol(n_init=1))
```

Fixing rho at its generating value is useful for checking this example.
In an application, that value is usually unknown: a sensitivity value should
come from the scientific question, rather than being treated as an estimate.
Omitting rho, or writing `rho = 1`, preserves the existing model.

## Estimate source strength in the admitted Gaussian design

`rho = NULL` requests estimation. It does not remove the Gaussian observation
residual. The estimator starts at rho one-half independently of the generating
value. Replication distinguishes the shared species effect from observation
noise; the source also needs nonzero relationships between modeled species.

```{r estimated-long}
set.seed(8402)
estimated_long <- gllvmTMB(
  value ~ 0+trait+phylo_dep(0+trait|species,tree=tree,rho=NULL),
  data=long,trait="trait",unit="obs",cluster="species",family=gaussian(),
  control=gllvmTMBcontrol(n_init=1))
```

```{r estimated-wide}
set.seed(8402)
estimated_wide <- gllvmTMB(
  traits(t1,t2,t3,t4,t5,t6) ~ 1+phylo_dep(1|species,tree=tree,rho=NULL),
  data=wide,unit="obs",cluster="species",family=gaussian(),
  control=gllvmTMBcontrol(n_init=1))
```

```{r agreement}
c(long=as.numeric(logLik(estimated_long)),
  wide=as.numeric(logLik(estimated_wide)))
```

An estimated latent alternative is initially restricted to rank one and at
least four traits. The matching terms are
`phylo_latent(species, tree=tree, d=1, unique=TRUE, rho=NULL)` in either formula.
`unique=FALSE` requests loadings only. A triangular loading convention fixes a
representation; it does not establish that the loading/Psi decomposition is
well identified. Weak or nearly zero loadings deserve particular caution.

## Read the two kinds of output separately

```{r extraction}
source_result <- extract_Sigma(estimated_long,level="phy",link_residual="none")
round(source_result$Sigma,3)
strength <- source_result$source_strength
strength[c("source","grouping","value","status","resolved_scale")]
strength$diagnostics
```

The generating strength was 0.6; this dataset gives an estimate of
`r round(strength$value,2)`. That difference remains even when the optimizer
converges and no weak-signal diagnostic is triggered. Those diagnostics flag
specific problems; their absence does not certify accurate recovery.

`Sigma` is the covariance among traits at the species tier. `source_strength`
reports rho, its fixed or estimated status, source labels and preserved scale.
It does not construct a large dense covariance merely to print a result.
A value near zero or one needs the boundary diagnostics and fit-health checks;
a small derivative on the logit scale can conceal a substantial slope on the
physical rho scale. The two scores are descriptive checks, not confidence
intervals or proof of a global optimum.

`extract_phylo_signal()` answers a different variance-allocation question.
For these attenuated fits it gives a typed limitation rather than reporting
an allocation that ignores attenuation. Use the covariance and source-strength
output above. Automatic bootstrap refits and rho profiles are also excluded
from this initial workflow; do not interpret an unavailable interval as zero
uncertainty.

## Predict and simulate at known levels

```{r workflow}
known <- predict(estimated_long,newdata=long)
head(known)
simulated <- simulate(estimated_long,nsim=2,seed=8501,condition_on_RE=FALSE)
head(simulated)
```

Known-level predictions include the fitted structured species effects.
Unconditional simulation redraws the whole species covariance, including Psi
when requested, then applies observation replication and residual noise.
This extension does not add prediction for unseen species or ancestral levels.

## Use the same rho for spatial dependence

For `spatial_*()` terms, rho has exactly the same meaning: it attenuates
between-location covariance while preserving each location's marginal variance.
Spatial **range** is a separate estimated parameter, reported through inverse
range `kappa`. Here the source is
$K(\kappa)=A Q(\kappa)^{-1}A^\top$, projected from the mesh to the modeled
locations. Its diagonal is recomputed as range changes; rho adds no normalization.

The following example has four traits and three individuals at each of 40
locations. It generates the whole latent-plus-Psi covariance directly. The
location identifier after the bar must refer to the same coordinates across
traits and individuals. The mesh is built on the long observation rows; those
same projected rows also match the wide call's expansion.

```{r spatial-data}
locations <- expand.grid(x=seq(0,2,length.out=8),ycoord=seq(0,2,length.out=5))
locations$location <- sprintf("s%02d",seq_len(nrow(locations)))
spatial_long <- expand.grid(trait=paste0("t",1:4),individual=1:3,
                            location=locations$location)
spatial_long$location <- factor(spatial_long$location,levels=locations$location)
spatial_long$obs <- interaction(spatial_long$location,spatial_long$individual,drop=TRUE)
spatial_long$x <- locations$x[match(spatial_long$location,locations$location)]
spatial_long$ycoord <- locations$ycoord[match(spatial_long$location,locations$location)]
mesh <- make_mesh(spatial_long,c("x","ycoord"),cutoff=.12)
A <- as.matrix(mesh$A_st[match(locations$location,spatial_long$location),,drop=FALSE])
kappa_truth <- 2
Q <- as.matrix(kappa_truth^4*mesh$spde$c0+
                 2*kappa_truth^2*mesh$spde$g1+mesh$spde$g2)
K_spatial <- A %*% solve(Q,t(A))
K_spatial_rho <- .6*K_spatial
diag(K_spatial_rho) <- diag(K_spatial)
spatial_loading <- c(.9,.7,-.8,.6)/sqrt(mean(diag(K_spatial)))
spatial_psi <- c(.3,.2,.25,.35)/mean(diag(K_spatial))
Sigma_spatial <- tcrossprod(spatial_loading)+diag(spatial_psi)
set.seed(319905)
spatial_effect <- t(chol(K_spatial_rho)) %*%
  matrix(rnorm(160),40,4) %*% chol(Sigma_spatial)
spatial_long$value <- .1*as.integer(spatial_long$trait)+
  spatial_effect[cbind(as.integer(spatial_long$location),as.integer(spatial_long$trait))]+
  rnorm(nrow(spatial_long),sd=.4)
spatial_wide <- reshape(spatial_long[c("obs","location","x","ycoord","trait","value")],
  idvar=c("obs","location","x","ycoord"),timevar="trait",direction="wide")
names(spatial_wide) <- sub("^value\\.","",names(spatial_wide))
```

The generating loading and Psi scales above make the location effects visible
relative to observation noise. They do not change the model's source normalization.

```{r spatial-long}
set.seed(319928)
spatial_fit_long <- gllvmTMB(
  value ~ 0+trait+spatial_latent(0+trait|location,mesh=mesh,d=1,unique=TRUE,rho=NULL),
  data=spatial_long,trait="trait",unit="obs",family=gaussian(),
  control=gllvmTMBcontrol(n_init=1))
```

```{r spatial-wide}
set.seed(319928)
spatial_fit_wide <- gllvmTMB(
  traits(t1,t2,t3,t4) ~ 1+spatial_latent(1|location,mesh=mesh,d=1,unique=TRUE,rho=NULL),
  data=spatial_wide,unit="obs",family=gaussian(),
  control=gllvmTMBcontrol(n_init=1))
```

```{r spatial-output}
c(long=as.numeric(logLik(spatial_fit_long)),wide=as.numeric(logLik(spatial_fit_wide)))
spatial_result <- extract_Sigma(spatial_fit_long,level="spatial",link_residual="none")
spatial_result$source_strength[c("value","status","kappa","resolved_scale")]
spatial_result$source_strength$diagnostics
```

The generating values were rho 0.6 and kappa 2. This dataset estimates rho
`r round(spatial_result$source_strength$value,2)` and kappa
`r round(spatial_result$source_strength$kappa,2)`. Range, attenuation and trait
scale can compete; complete replication separates observation noise but does
not guarantee precise separation of those parameters. The local geometry
diagnostic checks a specific weakness, not global identification or recovery.
The frozen spatial recovery study jointly estimated range and rho in two
geometries, four trait-covariance forms, and two strengths. It classified 14
cells as partial and 2 as blocked; no cell passed its predeclared recovery
criteria. Treat this output as a fitted point estimate and sensitivity tool,
not as evidence of generally reliable spatial range--rho separation.

For a fixed sensitivity value, replace `rho=NULL` with a number in [0,1].
At zero, mesh effects disappear, but range still affects the projected marginal
variances and can be confounded with trait scale. Omitted rho and explicit one
keep the previous spatial model, with range still estimated. Attenuated spatial
models currently predict only at known location identifiers with unchanged
coordinates. Unconditional simulation redraws both the spatial field and its
independent location companion, with the same rho for loadings and Psi.

## When to stop and change the question

With $D=I$, an ordinary unrestricted covariance $T$ on the same units can trade
variance with the structured covariance $S$:

\[
K_\rho\otimes S+I\otimes T
=K_{\rho'}\otimes\frac{\rho}{\rho'}S+
I\otimes\left[T+\left(1-\frac{\rho}{\rho'}\right)S\right],
\]

when the alternative covariances are admissible. Optimizer convergence cannot
resolve that ambiguity. Fix rho for a scientifically justified sensitivity
analysis, or use a design/model with an admitted identification argument.
Do not drop a scientific component or reduce latent rank simply to obtain a
successful fit. Source separation and loading/Psi decomposition are distinct
identification questions.

The animal, kernel and spatial helpers use the same rho convention. Fixed attenuation
retains the existing native Laplace family restrictions. The non-Gaussian
checks establish equivalence to an effective source covariance; they do not
establish recovery of rho outside Gaussian models.
