---
title: "Temporal latent scores for repeated multivariate measurements"
description: "Fit one rank-one AR1 latent process to three repeatedly measured traits, in long or traits(...) wide data."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Temporal latent scores for repeated multivariate measurements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gllvmTMB)
set.seed(2609081)
```

`temporal_latent()` fits one rank-one latent score for every occasion in each
series. The scores follow a stationary AR1 process, so the covariance between
two occasions `h` steps apart is `phi^abs(h) * Lambda %*% t(Lambda)`. The
loadings `Lambda` carry the process amplitude; the score variance is fixed to
one.

This first version admits Gaussian identity-link ML fits with at least three
traits and three consecutive integer occasions per series. It accepts one
temporal intercept block and fixed effects. Forecasts, new-data prediction,
profile and bootstrap intervals, rank selection, other response families,
higher ranks, irregular time, temporal slopes, and combinations with another
random-effect provider are not available yet.

> **Scope boundary.** The likelihood, syntax, score labels, simulation, and
> training-data prediction have focused local checks. The retained bounded
> recovery fixture is still an engineering gate: it does not support a general
> recovery, precision, or interval-coverage claim.

## Ordinary longitudinal observations

Here each `(series, occasion, trait)` cell occurs once. The independent
variation for trait `j` is one total variance, `D_j`; it is not separated into
occasion and measurement pieces.

```{r ordinary-data}
ordinary <- expand.grid(
  series = paste0("series_", 1:3), occasion = 1:3,
  trait = paste0("trait_", 1:3),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
ordinary$value <- rnorm(nrow(ordinary))
```

The long-format call is the canonical form.

```{r ordinary-long}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    temporal_latent(0 + trait | series, time = occasion),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_long)
```

The same observations can be supplied in wide format through the one
`gllvmTMB()` entry point.

```{r ordinary-wide}
ordinary_wide <- reshape(
  ordinary, idvar = c("series", "occasion"), timevar = "trait",
  direction = "wide"
)
names(ordinary_wide) <- sub("value\\.", "", names(ordinary_wide))
fit_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~ 1 +
    temporal_latent(1 | series, time = occasion),
  data = ordinary_wide, unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_wide)$pair_index
```

`getLV()` returns the occasion-level scores as a matrix. Its row names and the
`temporal_index` attribute identify the public series and occasion rather than
the private fitting factor. Score and loading signs are a reporting convention:
the first loading anchors the displayed direction unless it is negligible
relative to the largest loading, in which case that largest trait anchors it.
`extract_temporal()` records this diagnostic; the fitted likelihood is
unchanged by the simultaneous sign change.

```{r ordinary-scores}
scores <- getLV(fit_long)
attr(scores, "temporal_index")
predict(fit_long)
```

## Replicated measurements at each occasion

Add `replicate = measurement` when every trait is measured at least twice at
each occasion. This separates an occasion-shared trait variance `psi_j` from a
common measurement variance. Only the latent scores persist through time.

```{r replicated-data}
replicated <- expand.grid(
  series = paste0("series_", 1:3), occasion = 1:3, measurement = 1:2,
  trait = paste0("trait_", 1:3),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
replicated$value <- rnorm(nrow(replicated))
fit_replicated <- gllvmTMB(
  value ~ 0 + trait + temporal_latent(
    0 + trait | series, time = occasion, replicate = measurement
  ), data = replicated, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_replicated)$variance
```

The wide form keeps the measurement column and uses the same keyword.

```{r replicated-wide}
replicated_wide <- reshape(
  replicated, idvar = c("series", "occasion", "measurement"),
  timevar = "trait", direction = "wide"
)
names(replicated_wide) <- sub("value\\.", "", names(replicated_wide))
fit_replicated_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~ 1 + temporal_latent(
    1 | series, time = occasion, replicate = measurement
  ), data = replicated_wide, unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_replicated_wide)$parameters
```

`simulate()` redraws stationary temporal scores by default. Passing
`condition_on_RE = TRUE` keeps the fitted score modes and redraws only the
independent response variation. `update()` replays the saved public long or
wide call, rebuilding its private temporal index.

```{r simulation-update}
simulated <- simulate(fit_replicated, nsim = 2, seed = 23)
updated_call <- update(fit_replicated_wide, evaluate = FALSE)
```
