---
title: "Temporal covariance for repeated multivariate measurements"
description: "Fit AR1 or OU temporal covariance for three repeatedly measured traits, in long or traits(...) wide data."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Temporal covariance for repeated multivariate measurements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gllvmTMB)
set.seed(2609081)
```

Temporal is a covariance source row with three trait modes:
`temporal_indep()`, `temporal_dep()`, and rank-one `temporal_latent()`.
Each provider has a private `(series, time)` state index; it does not replace
the public `unit` or `unit_obs` labels. For latent mode,
`unique = TRUE` gives
\(K_{time}\otimes(\Lambda\Lambda^T+\Psi_T)\): Psi is correlated through
time, rather than being IID occasion noise.

With `structure = "ar1"`, occasions are ordered integers and gaps are kept:
\(K(t,s)=\phi^{|t-s|}\). With `structure = "ou"`, `time` is elapsed numeric
time and \(K(t,s)=\exp(-\kappa|t-s|)\). This first version admits complete
Gaussian identity-link ML fits with at least three traits and three occasions
per series. Ordinary unit and unit-observation terms can be combined when
their partitions agree and nest. The bounded temporal-only `forecast_temporal()`, `profile_temporal()`,
`bootstrap_temporal()`, and `compare_temporal()` helpers have separate
contracts. A separate replicated Gaussian AR1 `temporal_dep() +
spatial_indep()` cell with one fixed mesh also has direct profile, bootstrap,
and known-series forecast contracts. These narrow results do not establish
calibrated intervals, bootstrap sampling properties, general recovery, or
coverage. Generic forecasts, new-data prediction, generic profile/bootstrap
intervals, automatic rank selection, and temporal combinations with
phylogenetic, animal, other spatial, or dense-kernel sources remain unavailable.

> **Scope boundary.** Gaussian AR1/OU source parsing, dense covariance
> oracles, score labels, simulation, training-data prediction, and update have
> focused local checks. Recovery remains an engineering gate and does not
> support a general recovery, precision, calibration, or interval-coverage
> claim. Temporal is available with ordinary `unit` and nested `unit_obs`
> terms. The first qualified source pair is replicated Gaussian AR1
> `temporal_dep() + spatial_indep()` with one fixed mesh; phylogenetic, animal,
> other spatial, and dense-kernel temporal combinations remain deferred pending
> separate model contracts and evidence. The provider is locally verified only;
> it is not merged, released,
> or cross-platform verified.

## Ordinary longitudinal observations

Here each `(series, occasion, trait)` cell occurs once. The independent
variation for trait `j` is one total variance, `D_j`; it is not separated into
occasion and measurement pieces.

```{r ordinary-data}
ordinary <- expand.grid(
  series = paste0("series_", 1:3), occasion = c(1L, 3L, 7L),
  trait = paste0("trait_", 1:3),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
ordinary$value <- rnorm(nrow(ordinary))
```

The long-format call is the canonical form. Here are the three AR1 cells.

```{r ordinary-long}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    temporal_indep(0 + trait | series, time = occasion),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_long)
fit_dep <- gllvmTMB(
  value ~ 0 + trait + temporal_dep(0 + trait | series, time = occasion),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
fit_latent <- gllvmTMB(
  value ~ 0 + trait + temporal_latent(0 + trait | series, time = occasion,
    d = 1, unique = TRUE),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
```

The same observations can be supplied in wide format through the one
`gllvmTMB()` entry point.

```{r ordinary-wide}
ordinary_wide <- reshape(
  ordinary, idvar = c("series", "occasion"), timevar = "trait",
  direction = "wide"
)
names(ordinary_wide) <- sub("value\\.", "", names(ordinary_wide))
fit_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~ 1 +
    temporal_latent(1 | series, time = occasion, d = 1, unique = TRUE),
  data = ordinary_wide, unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_wide)$pair_index
```

`getLV()` returns the occasion-level scores for `temporal_latent()` as a matrix. Its row names and the
`temporal_index` attribute identify the public series and occasion rather than
the private fitting factor. Score and loading signs are a reporting convention:
the first loading anchors the displayed direction unless it is negligible
relative to the largest loading, in which case that largest trait anchors it.
`extract_temporal()` records this diagnostic; the fitted likelihood is
unchanged by the simultaneous sign change.

```{r ordinary-scores}
scores <- getLV(fit_latent)
attr(scores, "temporal_index")
predict(fit_latent)
```

For elapsed time, use the same grammar with `structure = "ou"`. Translation
of this column does not change the model; rescaling time changes the estimated
decay rate reciprocally.

```{r ordinary-ou}
ordinary$elapsed <- c(0, 0.5, 2)[match(ordinary$occasion, c(1L, 3L, 7L))]
fit_ou <- gllvmTMB(
  value ~ 0 + trait + temporal_dep(0 + trait | series, time = elapsed,
    structure = "ou"),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_ou)$time
```

## Optional repeated rows at each occasion

Use `replicate = measurement` when every trait is measured at least twice at
each occasion. It identifies repeated trait panels while the temporal
covariance remains on the series--occasion state. A `unit_obs` grouping remains
an ordinary, unit-nested random-effect level; it is never silently treated as
a temporal replicate.

```{r replicated-data}
replicated <- expand.grid(
  series = paste0("series_", 1:3), occasion = 1:3, measurement = 1:2,
  trait = paste0("trait_", 1:3),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
replicated$value <- rnorm(nrow(replicated))
fit_replicated <- gllvmTMB(
  value ~ 0 + trait + temporal_latent(
    0 + trait | series, time = occasion, replicate = measurement
  ), data = replicated, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_replicated)$variance
```

The wide form keeps the measurement column and uses the same keyword.

```{r replicated-wide}
replicated_wide <- reshape(
  replicated, idvar = c("series", "occasion", "measurement"),
  timevar = "trait", direction = "wide"
)
names(replicated_wide) <- sub("value\\.", "", names(replicated_wide))
fit_replicated_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~ 1 + temporal_latent(
    1 | series, time = occasion, replicate = measurement
  ), data = replicated_wide, unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_replicated_wide)$parameters
```

## Temporal-only lifecycle helpers

For a fitted temporal-only model, `profile_temporal()` re-optimizes the
remaining parameters at fixed persistence values, `bootstrap_temporal()`
retains every refit outcome, and `compare_temporal()` compares supplied
same-structure candidates by AIC. These are bounded diagnostics, not
calibrated intervals, a hypothesis-testing procedure, or automatic selection.

```{r temporal-lifecycle}
profile_temporal(fit_long, ystep = .25, ytol = 1)
bootstrap_temporal(fit_long, n_boot = 2, seed = 260920L)
compare_temporal(current = fit_long, replay = update(fit_long))
```

`simulate()` redraws stationary temporal scores by default. Passing
`condition_on_RE = TRUE` keeps the fitted score modes and redraws only the
independent response variation. `update()` replays the saved public long or
wide call, rebuilding its private temporal index.

```{r simulation-update}
simulated <- simulate(fit_replicated, nsim = 2, seed = 23)
updated_call <- update(fit_replicated_wide, evaluate = FALSE)
```
