---
title: "Temporal covariance for repeated multivariate measurements"
description: "Fit AR1 or OU temporal covariance for three repeatedly measured traits, in long or traits(...) wide data."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Temporal covariance for repeated multivariate measurements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gllvmTMB)
set.seed(2609081)
```

Temporal is a covariance source row with three trait modes:
`temporal_indep()`, `temporal_dep()`, and rank-one `temporal_latent()`.
Each provider has a private `(series, time)` state index; it does not replace
the public `unit` or `unit_obs` labels. For latent mode,
`unique = TRUE` gives
\(K_{time}\otimes(\Lambda\Lambda^T+\Psi_T)\): Psi is correlated through
time, rather than being IID occasion noise.

With `structure = "ar1"`, occasions are ordered integers and gaps are kept:
\(K(t,s)=\phi^{|t-s|}\). With `structure = "ou"`, `time` is elapsed numeric
time and \(K(t,s)=\exp(-\kappa|t-s|)\). This first version admits complete
Gaussian identity-link ML fits with at least three traits and three occasions
per series. Ordinary unit and unit-observation terms can be combined when
their partitions agree and nest. The bounded temporal-only
`forecast_temporal()`, `profile_temporal()`, `bootstrap_temporal()`, and
`compare_temporal()` helpers have separate contracts. `bootstrap_temporal()`,
`profile_temporal()`, and `compare_temporal()` also support the qualified
replicated-AR1
`temporal_indep() + kernel_indep()` cell. A separate fixed-parameter
future-observation route also admits replicated-AR1
`temporal_dep() + phylo_indep()` with one fixed phylogenetic source; its
conditional standard deviations are not prediction intervals. That same
phylogenetic cell also has a direct persistence profile and an unconditional
bootstrap/refit helper that retains every attempted refit; neither establishes
calibrated intervals or bootstrap sampling properties. A separately qualified
rank-one `temporal_latent(..., unique = FALSE) + animal_indep()` cell also has
that retained bootstrap/refit helper and a known-series future-observation
forecast. Its fitted-parameter conditional standard deviations are not
prediction intervals. Generic forecasts, new-data
prediction, generic profile/bootstrap intervals, automatic rank selection, and
other source-pair helper routes remain unavailable. Other response families,
higher ranks, temporal
slopes, source-by-time interactions, and structured-source combinations beyond
the four narrow AR1 additive pairs and the preliminary OU-kernel pair below remain unavailable.

> **Scope boundary.** Gaussian AR1/OU source parsing, dense covariance
> oracles, score labels, simulation, training-data prediction, and update have
> focused local checks. Recovery remains an engineering gate and does not
> support a general recovery, precision, calibration, or interval-coverage
> claim. Four separately bounded replicated-AR1 additive pairs with one
> diagonal kernel, phylogenetic, animal, or spatial source are implemented;
> their likelihood and lifecycle checks do not give them a common recovery
> status. Kernel has a retained passing local fixture; phylogenetic, animal, and
> spatial retained recovery gates remain partial or failed, and the dependent
> spatial campaign is paused after a retained overrun. One additional replicated
> OU `temporal_indep() + kernel_indep()` cell has independent dense
> likelihood/gradient, time-coordinate, simulation, long/wide, update, direct-profile, bootstrap, and same-structure AIC checks
> on an irregular-time fixture. Its source-pair forecast and recovery gates remain pending, so this article
> does not present it as a worked workflow.

## Ordinary longitudinal observations

Here each `(series, occasion, trait)` cell occurs once. The independent
variation for trait `j` is one total variance, `D_j`; it is not separated into
occasion and measurement pieces.

```{r ordinary-data}
ordinary <- expand.grid(
  series = paste0("series_", 1:3), occasion = c(1L, 3L, 7L),
  trait = paste0("trait_", 1:3),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
ordinary$value <- rnorm(nrow(ordinary))
```

The long-format call is the canonical form. Here are the three AR1 cells.

```{r ordinary-long}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    temporal_indep(0 + trait | series, time = occasion),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_long)
fit_dep <- gllvmTMB(
  value ~ 0 + trait + temporal_dep(0 + trait | series, time = occasion),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
fit_latent <- gllvmTMB(
  value ~ 0 + trait + temporal_latent(0 + trait | series, time = occasion,
    d = 1, unique = TRUE),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
```

The same observations can be supplied in wide format through the one
`gllvmTMB()` entry point.

```{r ordinary-wide}
ordinary_wide <- reshape(
  ordinary, idvar = c("series", "occasion"), timevar = "trait",
  direction = "wide"
)
names(ordinary_wide) <- sub("value\\.", "", names(ordinary_wide))
fit_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~ 1 +
    temporal_latent(1 | series, time = occasion, d = 1, unique = TRUE),
  data = ordinary_wide, unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_wide)$pair_index
```

`getLV()` returns the occasion-level scores for `temporal_latent()` as a matrix. Its row names and the
`temporal_index` attribute identify the public series and occasion rather than
the private fitting factor. Score and loading signs are a reporting convention:
the first loading anchors the displayed direction unless it is negligible
relative to the largest loading, in which case that largest trait anchors it.
`extract_temporal()` records this diagnostic; the fitted likelihood is
unchanged by the simultaneous sign change.

```{r ordinary-scores}
scores <- getLV(fit_latent)
attr(scores, "temporal_index")
predict(fit_latent)
```

For elapsed time, use the same grammar with `structure = "ou"`. Translation
of this column does not change the model; rescaling time changes the estimated
decay rate reciprocally.

```{r ordinary-ou}
ordinary$elapsed <- c(0, 0.5, 2)[match(ordinary$occasion, c(1L, 3L, 7L))]
fit_ou <- gllvmTMB(
  value ~ 0 + trait + temporal_dep(0 + trait | series, time = elapsed,
    structure = "ou"),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_ou)$time
```

## Optional repeated rows at each occasion

Use `replicate = measurement` when every trait is measured at least twice at
each occasion. It identifies repeated trait panels while the temporal
covariance remains on the series--occasion state. A `unit_obs` grouping remains
an ordinary, unit-nested random-effect level; it is never silently treated as
a temporal replicate.

```{r replicated-data}
replicated <- expand.grid(
  series = paste0("series_", 1:3), occasion = 1:3, measurement = 1:2,
  trait = paste0("trait_", 1:3),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
replicated$value <- rnorm(nrow(replicated))
fit_replicated <- gllvmTMB(
  value ~ 0 + trait + temporal_latent(
    0 + trait | series, time = occasion, replicate = measurement
  ), data = replicated, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_replicated)$variance
```

The wide form keeps the measurement column and uses the same keyword.

```{r replicated-wide}
replicated_wide <- reshape(
  replicated, idvar = c("series", "occasion", "measurement"),
  timevar = "trait", direction = "wide"
)
names(replicated_wide) <- sub("value\\.", "", names(replicated_wide))
fit_replicated_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~ 1 + temporal_latent(
    1 | series, time = occasion, replicate = measurement
  ), data = replicated_wide, unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_replicated_wide)$parameters
```

## A qualified kernel-pair refit route

One bounded composed route is available for a replicated AR1
`temporal_indep()` model with one fixed labelled diagonal kernel. The profile
re-optimizes every nuisance parameter, including the kernel variance, at each
fixed persistence value. The bootstrap redraws both sources and refits the
saved public call. `compare_temporal()` accepts only supplied candidates with
the identical labelled kernel and returns their AIC values; it does not perform
a likelihood-ratio test or search candidate models. Its returned profile
endpoints and bootstrap draws are not calibrated intervals.

The same AR1 kernel cell can forecast a **new response measurement** at later
occasions in a known series. Supply each measurement as a complete trait panel.
The forecast conditions on the fitted additive temporal-plus-kernel covariance;
`se.fit` is a fitted-parameter conditional standard deviation, not a
prediction interval.

```{r kernel-pair-lifecycle}
K_series <- diag(3)
dimnames(K_series) <- list(paste0("series_", 1:3), paste0("series_", 1:3))
fit_kernel_pair <- gllvmTMB(
  value ~ 0 + trait +
    temporal_indep(0 + trait | series, time = occasion, replicate = measurement) +
    kernel_indep(series, K = K_series, name = "series_kernel"),
  data = replicated, trait = "trait", unit = "series", cluster = "series",
  family = gaussian(), silent = TRUE
)
profile_temporal(fit_kernel_pair, ystep = .25, ytol = 1)
bootstrap_temporal(fit_kernel_pair, n_boot = 2, seed = 260920L)
compare_temporal(current = fit_kernel_pair, replay = update(fit_kernel_pair))
future_kernel <- expand.grid(
  series = paste0("series_", 1:3), occasion = 5,
  measurement = c("m1", "m2"), trait = paste0("trait_", 1:3),
  KEEP.OUT.ATTRS = FALSE
)
forecast_temporal(fit_kernel_pair, future_kernel, se.fit = TRUE)
```

## A qualified rank-one animal-pair forecast

The rank-one temporal latent cell also has a narrow future-observation route
when it is paired with one fixed labelled `animal_indep()` relationship. The
animal field is static across occasions: it is added to the temporal AR1
covariance rather than multiplied by it. Future rows must retain fitted animal
labels and form complete measurement-by-trait panels. As above, `se.fit` is a
conditional standard deviation at fitted parameters, not a prediction
interval.

```{r animal-pair-forecast}
A_series <- diag(3)
dimnames(A_series) <- list(paste0("series_", 1:3), paste0("series_", 1:3))
fit_animal_pair <- gllvmTMB(
  value ~ 0 + trait +
    temporal_latent(0 + trait | series, time = occasion,
      replicate = measurement, d = 1, unique = FALSE) +
    animal_indep(0 + trait | series, A = A_series),
  data = replicated, trait = "trait", unit = "series", cluster = "series",
  family = gaussian(), silent = TRUE
)
future_animal <- expand.grid(
  series = paste0("series_", 1:3), occasion = 4,
  measurement = 1:2, trait = paste0("trait_", 1:3),
  KEEP.OUT.ATTRS = FALSE
)
forecast_temporal(fit_animal_pair, future_animal, se.fit = TRUE)
```

`simulate()` redraws stationary temporal scores by default. Passing
`condition_on_RE = TRUE` keeps the fitted score modes and redraws only the
independent response variation. `update()` replays the saved public long or
wide call, rebuilding its private temporal index.

```{r simulation-update}
simulated <- simulate(fit_replicated, nsim = 2, seed = 23)
updated_call <- update(fit_replicated_wide, evaluate = FALSE)
```
