---
title: "Temporal covariance for repeated multivariate measurements"
description: "Choose AR1 or OU covariance for repeatedly measured Gaussian traits, then fit a standalone temporal model in long or traits(...) wide data."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Temporal covariance for repeated multivariate measurements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gllvmTMB)
set.seed(2609081)
```

This article is for an applied researcher with several traits measured along an
animal, plot, or transect trajectory. Within each series, the occasions are
temporally correlated: `time` selects the temporal states, whereas `unit` and
`unit_obs` remain ordinary grouping tiers and do not define the temporal clock.
The covariance grammar has five stable-unit sources and a sixth temporal
source for ordered states within a series. This article teaches the standalone
temporal route, with three trait-covariance modes:
`temporal_indep()`, `temporal_dep()`, and rank-one `temporal_latent()`.

First choose the time scale. AR1 is for ordered integer occasions and preserves
their gaps, so \(K(t,s)=\phi^{|t-s|}\). It can describe positive or negative
adjacent-occasion persistence. OU is for numeric elapsed time, with positive
decay rate \(\kappa\) and \(K(t,s)=\exp(-\kappa|t-s|)\). Translation of elapsed
time does not change an OU model; changing its units changes the numerical
decay rate. Next choose the trait covariance carried through that time kernel:

| Question about the three traits | Temporal keyword | Trait covariance |
|---|---|---|
| Do they persist separately? | `temporal_indep()` | a diagonal variance for each trait |
| May every pair persist together? | `temporal_dep()` | an unstructured covariance matrix |
| Is one shared changing axis enough? | `temporal_latent(..., d = 1)` | rank-one \(\Lambda\Lambda^T\) |

With `temporal_latent(unique = TRUE)`, the model adds its trait-diagonal
\(\Psi\) companion, so the temporal covariance is
\(K_{time}\otimes(\Lambda\Lambda^T+\Psi)\). That Psi is correlated through
time; it is not independent occasion noise. Each provider creates a private
`(series, time)` state index and does not replace the public `unit` or
`unit_obs` labels.

This version requires complete Gaussian identity-link data, at least three
traits, and at least three ordered occasions in every series. Ordinary `unit`
and `unit_obs` terms can be combined when their partitions agree and nest. The
bounded temporal-only `forecast_temporal()`, `profile_temporal()`,
`bootstrap_temporal()`, and `compare_temporal()` helpers have separate
contracts.

> **Scope boundary.** The standalone Gaussian AR1/OU routes shown here have
> focused local checks for parsing, dense covariance calculations, score labels,
> simulation, training-data prediction, and update. This is not a general
> recovery, precision, calibration, or interval-coverage claim. Four narrow
> replicated AR1 source combinations are also implemented, with separate
> evidence and restrictions described below. Generic forecasts, new-data
> prediction, generic profile/bootstrap intervals, and automatic model search
> or rank selection remain unavailable.

```{r temporal-concept, echo = FALSE, fig.width = 8.2, fig.height = 4.1, fig.cap = "The correlation panel uses fixed illustrative values, not fitted estimates: AR1 has phi = 0.7 and OU has kappa = -log(0.7). The hierarchy panel shows the standalone temporal provider; the restricted additive source combinations described below are not shown."}
blue <- "#0072B2"
orange <- "#E69F00"
orange_text <- "#8A5A00"
grey <- "#666666"

ar1_points <- data.frame(
  lag = 0:4,
  correlation = 0.7^(0:4)
)
ou_curve <- data.frame(
  lag = seq(0, 4, length.out = 161)
)
ou_curve$correlation <- exp(-(-log(0.7)) * ou_curve$lag)

p_correlation <- ggplot2::ggplot() +
  ggplot2::geom_hline(yintercept = 0, colour = "#D9D9D9", linewidth = 0.4) +
  ggplot2::geom_line(
    data = ou_curve,
    ggplot2::aes(lag, correlation),
    colour = orange, linewidth = 1
  ) +
  ggplot2::geom_point(
    data = ar1_points,
    ggplot2::aes(lag, correlation),
    colour = blue, size = 2.4
  ) +
  ggplot2::annotate(
    "text", x = 2.2, y = 0.54,
    label = "OU: continuous elapsed time\nkappa = -log(0.7)",
    colour = orange_text, hjust = 0, size = 3.05, lineheight = 0.95
  ) +
  ggplot2::annotate(
    "text", x = 2.2, y = 0.29,
    label = "AR1: integer occasions\nphi = 0.7",
    colour = blue, hjust = 0, size = 3.05, lineheight = 0.95
  ) +
  ggplot2::scale_x_continuous(breaks = 0:4, limits = c(0, 4.65)) +
  ggplot2::scale_y_continuous(
    breaks = c(0, 0.25, 0.5, 0.75, 1), limits = c(0, 1.03),
    expand = ggplot2::expansion(mult = c(0, 0.02))
  ) +
  ggplot2::labs(
    title = "Correlation across time",
    x = "Elapsed lag",
    y = "Temporal-process correlation"
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(face = "bold"),
    plot.background = ggplot2::element_rect(fill = "white", colour = NA),
    panel.background = ggplot2::element_rect(fill = "white", colour = NA)
  )

hierarchy_nodes <- data.frame(
  x = c(0.5, 1.9, 3.35, 3.35, 1.25, 2.75),
  y = c(0.76, 0.76, 0.76, 0.47, 0.12, 0.12),
  label = c(
    "Series\nanimal / plot /\ntransect",
    "Occasions / time\nselect temporal\nstates",
    "Complete\ntrait panel",
    "Optional\nreplicate panel",
    "unit",
    "unit_obs"
  ),
  kind = c("temporal", "temporal", "temporal", "optional", "ordinary", "ordinary")
)
hierarchy_links <- data.frame(
  x = c(1.02, 2.37, 3.35),
  xend = c(1.42, 3.02, 3.35),
  y = c(0.76, 0.76, 0.66),
  yend = c(0.76, 0.76, 0.57),
  kind = c("temporal", "temporal", "optional")
)

p_hierarchy <- ggplot2::ggplot() +
  ggplot2::geom_segment(
    data = hierarchy_links,
    ggplot2::aes(x, y, xend = xend, yend = yend, linetype = kind),
    colour = grey, linewidth = 0.55,
    arrow = grid::arrow(length = grid::unit(0.11, "inches"))
  ) +
  ggplot2::geom_label(
    data = hierarchy_nodes,
    ggplot2::aes(x, y, label = label, colour = kind),
    fill = "white", label.size = 0.35, size = 3.05, lineheight = 0.95,
    show.legend = FALSE
  ) +
  ggplot2::annotate(
    "text", x = 2, y = 0.27,
    label = "Separate ordinary grouping tiers (not the temporal clock)",
    colour = grey, size = 3.05
  ) +
  ggplot2::scale_colour_manual(values = c(
    temporal = blue, optional = orange_text, ordinary = grey
  )) +
  ggplot2::scale_linetype_manual(values = c(temporal = "solid", optional = "dashed")) +
  ggplot2::coord_cartesian(xlim = c(0.05, 3.85), ylim = c(0, 1), clip = "off") +
  ggplot2::labs(title = "How observations map to temporal states") +
  ggplot2::theme_void(base_size = 10) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    plot.background = ggplot2::element_rect(fill = "white", colour = NA),
    panel.background = ggplot2::element_rect(fill = "white", colour = NA),
    legend.position = "none",
    plot.margin = ggplot2::margin(8, 8, 8, 8)
  )

grid::grid.newpage()
layout <- grid::grid.layout(1, 2, widths = grid::unit(c(1, 1.2), "null"))
grid::pushViewport(grid::viewport(layout = layout))
print(p_correlation, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1))
print(p_hierarchy, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2))
grid::popViewport()
```

The AR1 points and OU curve above are conceptual only; they show no fitted
estimates and no uncertainty ribbons. With a negative AR1 value, correlations
alternate sign across odd and even integer lags. A positive-decay OU process
has no analogue of that alternating pattern.

## Fit an AR1 model to ordinary longitudinal observations

Here each `(series, occasion, trait)` cell occurs once. Start with AR1 because
the occasions are integers and their unequal gaps, 1 to 2 to 4, matter to the
covariance. The independent-mode variance for each trait is its total temporal
variance; this unreplicated design does not split it into occasion and
measurement components.

```{r ordinary-data}
ordinary <- expand.grid(
  series = paste0("series_", 1:3), occasion = c(1L, 2L, 4L),
  trait = paste0("trait_", 1:3),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
ordinary$value <- rnorm(nrow(ordinary))
```

The long-format call is canonical. The three fits below differ only in their
cross-trait covariance; `extract_temporal()` makes the fitted time index and
mode explicit.

```{r ordinary-long}
fit_long <- gllvmTMB(
  value ~ 0 + trait +
    temporal_indep(0 + trait | series, time = occasion),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
```

Before interpreting the temporal covariance, check the fit. Read and follow
the `action` for every row that is not `PASS`; a converged optimizer alone is
not enough. [Can I trust this fit?](fit-diagnostics.html) explains the
diagnostic workflow.

```{r ordinary-long-check}
health <- check_gllvmTMB(fit_long)
health[health$status != "PASS", c("component", "status", "message", "action")]
```

```{r ordinary-long-comparison}
extract_temporal(fit_long)
fit_dep <- gllvmTMB(
  value ~ 0 + trait + temporal_dep(0 + trait | series, time = occasion),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
fit_latent <- gllvmTMB(
  value ~ 0 + trait + temporal_latent(0 + trait | series, time = occasion,
    d = 1, unique = TRUE),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
```

The same observations can be supplied in wide format through the one
`gllvmTMB()` entry point. `traits(...)` says which response columns are the
three traits; it expands internally to the same trait-intercept structure as
the long call.

```{r ordinary-wide}
ordinary_wide <- reshape(
  ordinary, idvar = c("series", "occasion"), timevar = "trait",
  direction = "wide"
)
names(ordinary_wide) <- sub("value\\.", "", names(ordinary_wide))
fit_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~ 1 +
    temporal_latent(1 | series, time = occasion, d = 1, unique = TRUE),
  data = ordinary_wide, unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_wide)$pair_index
```

`getLV()` returns occasion-level scores for `temporal_latent()` as a matrix.
Its row names and `temporal_index` attribute identify the public series and
occasion, rather than the private fitting factor. Score and loading signs are a
reporting convention: the first loading anchors the displayed direction unless
it is negligible relative to the largest loading, in which case that largest
trait anchors it. A simultaneous sign change leaves the fitted likelihood
unchanged.

```{r ordinary-scores}
scores <- getLV(fit_latent)
attr(scores, "temporal_index")
predict(fit_latent)
```

Read `extract_temporal(fit_latent)$time` as the fitted AR1 persistence (or OU
decay rate), and use `$pair_index` to map each temporal state back to its series
and occasion. For the rank-one model, call
`ordination <- extract_ordination(fit_latent, level = "unit")`, then interpret
`ordination$scores` together with `ordination$loadings`. This paired accessor
applies the same sign orientation to both: the score traces the shared
occasion-level axis, while each loading gives that trait's direction and scale
on the axis. Reversing both signs still describes the same fitted model.

## Use OU when elapsed time, not occasion number, is meaningful

For elapsed time, use the same grammar with `structure = "ou"`. Here 0, 0.5,
and 2 are elapsed time points. Do not replace elapsed time
with consecutive visit numbers: that would fit a different scientific model.

```{r ordinary-ou}
ordinary$elapsed <- c(0, 0.5, 2)[match(ordinary$occasion, c(1L, 2L, 4L))]
fit_ou <- gllvmTMB(
  value ~ 0 + trait + temporal_dep(0 + trait | series, time = elapsed,
    structure = "ou"),
  data = ordinary, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_ou)$time
```

## Add repeated measurements only when they are genuine replicate panels

Use `replicate = measurement` only when every trait is measured at least twice
at each occasion. It identifies repeated complete trait panels while the
temporal covariance remains on the series--occasion state. A `unit_obs`
grouping remains an ordinary, unit-nested random-effect level; it is never
silently treated as a temporal replicate.

```{r replicated-data}
replicated <- expand.grid(
  series = paste0("series_", 1:3), occasion = 1:3, measurement = 1:2,
  trait = paste0("trait_", 1:3),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
replicated$value <- rnorm(nrow(replicated))
fit_replicated <- gllvmTMB(
  value ~ 0 + trait + temporal_latent(
    0 + trait | series, time = occasion, replicate = measurement
  ), data = replicated, trait = "trait", unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_replicated)$variance
```

The wide form keeps the measurement column and uses the same keyword.

```{r replicated-wide}
replicated_wide <- reshape(
  replicated, idvar = c("series", "occasion", "measurement"),
  timevar = "trait", direction = "wide"
)
names(replicated_wide) <- sub("value\\.", "", names(replicated_wide))
fit_replicated_wide <- gllvmTMB(
  traits(trait_1, trait_2, trait_3) ~ 1 + temporal_latent(
    1 | series, time = occasion, replicate = measurement
  ), data = replicated_wide, unit = "series", family = gaussian(), silent = TRUE
)
extract_temporal(fit_replicated_wide)$parameters
```

## What post-fit actions this standalone route supports

The current post-fit helpers are intentionally limited to an unreplicated
Gaussian `temporal_indep()` fit. `profile_temporal()` re-optimizes the temporal
model at fixed persistence or decay values; `bootstrap_temporal()` retains all
refit attempts; and `compare_temporal()` reports AIC without a likelihood-ratio
test. `forecast_temporal()` forecasts a complete future trait panel for an
already observed series. Its `se.fit` is conditional on fitted parameters, not
a calibrated prediction interval.

`simulate()` redraws stationary temporal scores by default. Passing
`condition_on_RE = TRUE` keeps the fitted score modes and redraws only the
independent response variation. `update()` replays the saved public long or
wide call, rebuilding its private temporal index.

```{r simulation-update}
simulated <- simulate(fit_replicated, nsim = 2, seed = 23)
updated_call <- update(fit_replicated_wide, evaluate = FALSE)
```

## Can I add a phylogeny, pedigree, spatial field, or kernel?

Only four narrow additive combinations are implemented: replicated AR1
`temporal_indep()` plus exactly one `kernel_indep()`, `phylo_indep()`,
`animal_indep()`, or `spatial_indep()` term. Both terms describe trait-specific
variances, with zero cross-trait covariance. The model adds their covariance
contributions; it does not fit a source-by-time interaction.

These combinations require complete Gaussian identity-link data, native
TMB/Laplace ML fitting, and at least two complete measurements at every
series--occasion, identified by `replicate = measurement`. The source is fixed
at `rho = 1`. Supply the labelled kernel, `tree =` or `vcv =`, or `pedigree =`,
`A =`, or `Ainv =` inside the corresponding keyword. Use dense `A` for a
relationship matrix and `Ainv` for sparse precision. The spatial route requires
a fixed mesh and spatial contrasts that are not determined solely by temporal
lag; spatial range is still estimated.

The kernel combination met its prespecified recovery criteria in one local
simulated example. The phylogenetic, animal, and spatial combinations have
focused likelihood checks, but their retained recovery studies did not meet the
prespecified recovery criteria. None earns general recovery,
cross-platform, release, precision, calibration, or interval-coverage claims.
OU combinations, temporal `dep`/`latent` combinations, temporal slopes, higher
rank, other families, multiple structured sources, source-by-time interactions,
and spatial forecasting remain unavailable. The temporal-only post-fit helpers
above do not extend to these combinations.

For the standalone examples, inspect the fitted temporal covariance and, for
the latent model, report the labelled scores together with the loadings. Read
[Current limitations](current-limits.html) before interpreting a more complex
model.
