---
title: "What is a site? The unit of analysis, and why you probably do not need a grid"
description: "A count needs a place and an amount of looking. That pair is the unit of analysis — usually an observation event such as a checklist or a station visit, not a grid cell. This page separates the three different things people call a 'cell', shows on toy data that a site's label never enters a prediction, and says when gridding is a reasonable choice and what it costs."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{What is a site? The unit of analysis, and why you probably do not need a grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Before you fit anything, you have to decide what one row of your data *is*.
That decision is called the **unit of analysis**, and for species data it is
the question people get stuck on first: *do I need to put my records into grid
cells?*

The short answer is no. This page explains what a row has to carry instead,
untangles the three unrelated things that get called "cells" in this
literature, and shows the consequences on toy data you can run yourself.
Everything below uses only the installed package, base R, and twelve
simulated sites; the whole page fits in a few seconds of computing.

This page is about the *decision*. For definitions of the package's formula
and model vocabulary — trait, unit, latent variable, loading — see
[Plain-English vocabulary](gllvm-vocabulary.html).

```{r setup, message = FALSE}
library(gllvmTMB)
```

# A count is a number of birds and an amount of looking

Write down a single observation: *three Canada Warblers*. On its own that
number means nothing, because it does not say how hard anyone looked. Three
warblers in five minutes and three warblers in a full morning are different
pieces of evidence about the same place.

So a usable observation needs two things: **where**, and **how much looking**.
The pair is the unit. Everything else in this article follows from that.

An eBird checklist already is such a unit. One outing, one place, 45 minutes,
three warblers — the checklist *is* the counting window, and its duration is
the effort. That is why effort enters a model as `offset(log_effort)`: the
offset says the expected count scales with how long you looked, without
spending a parameter on it. No grid is involved anywhere in that sentence.

A point-count station is the same shape. The protocol defines the window —
ten minutes, 100 m radius, this coordinate — and the protocol is the unit.

A grid cell is a third possibility: "all the birds recorded in this square".
That is the atlas tradition, and it is a perfectly legitimate unit. It is also
entirely optional, and Section [If you do grid, it is a choice with a
price](#if-you-do-grid-it-is-a-choice-with-a-price) says what it costs.

Underneath all three is the same picture. Birds have an expected density that
varies smoothly over the landscape, and any count is *effort × density where
you looked*. When the place you looked is compact — a station, a stationary
checklist — describing that place by its density at a point is an honest
description, not an approximation you have to apologise for.

# Three different things people call a "cell"

These three get conflated constantly, and they are unrelated. Keeping them
apart resolves most of the confusion about gridding.

| | What it is | Do you need it? |
|---|---|---|
| **Observation unit** | What one count means: one place, one amount of looking. Usually an *event* — a checklist, a station visit — not an area. | **Always.** Your data cannot mean anything without it. |
| **Grid cells** | Squares laid over the landscape, used to define units or to pool records. | **Never required.** One optional way of choosing units. |
| **The mesh** | A triangular network the computer uses to represent a smooth surface. | **Required by this package's spatial models** — and nothing to do with counting. |

The mesh is the one that causes trouble, because a picture of it looks
grid-like. It is not a grid, and your data are never snapped to it. It is the
computational device that lets a *continuous* density surface be estimated
from a finite number of numbers: the surface is stored at the mesh vertices,
and each observation's exact coordinate is expressed as a weighted blend of
the nearby vertices. Your coordinates stay exactly where they are.

# What counts as a "site" in your data frame

Whatever unit carries one location and one effort. Here are twelve of them —
twelve checklists, three species, one row per species per checklist.

```{r toy-data}
set.seed(20260818)

n_site <- 12
sites <- data.frame(
  site    = sprintf("checklist_%02d", seq_len(n_site)),
  X       = runif(n_site, 0, 20),          # projected coordinates, in km
  Y       = runif(n_site, 0, 20),
  minutes = sample(c(20, 45, 60, 90), n_site, replace = TRUE)
)
sites$log_effort <- log(sites$minutes / 60)  # effort, on the log scale

species   <- c("warbler", "vireo", "thrush")
abundance <- c(warbler = 0.6, vireo = 0.0, thrush = -1.2)  # thrush is scarce

dat <- do.call(rbind, lapply(species, function(s) transform(
  sites, trait = s,
  value = rpois(n_site, exp(abundance[[s]] + 0.08 * sites$X + sites$log_effort))
)))
dat$trait <- factor(dat$trait)
dat$site  <- factor(dat$site)

str(dat[, c("site", "X", "Y", "log_effort", "trait", "value")])
```

Read that `str()` as the target shape for your own data. `site` labels the
observation unit, `X` and `Y` are its projected coordinates, `log_effort` is
the log of however you measure looking, `trait` is the species, and `value` is
the count.

Two things about the columns are worth saying out loud.

**Project your coordinates before you do anything spatial.** A degree of
longitude and a degree of latitude are different physical distances everywhere
except the equator, so a model built on raw longitude/latitude treats
unequal distances as equal. `add_utm_columns()` will do the conversion.

**The site label is bookkeeping.** It has to be there, and it has to be
distinct per unit, but it carries no meaning of its own — the coordinates and
the effort do all the work. That is worth demonstrating rather than asserting.

# The label really is bookkeeping: a demonstration

Fit a small spatial model to the toy data. The only random structure in it is
the spatial field, which is keyed on the coordinates.

```{r toy-fit}
mesh <- make_mesh(dat, c("X", "Y"), cutoff = 3)

fit <- gllvmTMB(
  value ~ 0 + trait + offset(log_effort) +
    spatial_latent(0 + trait | site, d = 1),
  data = dat, trait = "trait", unit = "site",
  family = poisson(), mesh = mesh, silent = TRUE
)

list(converged = fit$opt$convergence == 0,
     positive_definite_hessian = fit$sd_report$pdHess)
```

Now predict over a grid of new locations twice, changing nothing but the site
label written in the unit column.

```{r label-identity}
newd <- expand.grid(
  X     = seq(2, 18, length.out = 5),
  Y     = seq(2, 18, length.out = 5),
  trait = levels(dat$trait)
)
newd$log_effort <- 0   # predict at one hour of looking
newd$value      <- 0   # placeholder; not used by predict()

lab <- function(d, s) transform(d, site = factor(s, levels = levels(dat$site)))
p1 <- suppressMessages(predict(fit, newdata = lab(newd, "checklist_01"),
                               type = "link"))
p2 <- suppressMessages(predict(fit, newdata = lab(newd, "checklist_09"),
                               type = "link"))

cat("max |difference| between the two labels:", max(abs(p1$est - p2$est)),
    "\nover a prediction range of", round(diff(range(p1$est)), 3), "\n")
```

Exactly zero, across a surface that itself varies by
`r round(diff(range(p1$est)), 2)` on the link scale. The label you write in the
unit column is not consulted. What the model used was the coordinate you asked
about.

**This stops being true the moment your model has a genuine unit-level random
effect** — a `(1 | site)` term, say. Then the label picks out which random
effect to add back, and it matters a great deal. The rule is not "labels never
matter"; it is "labels matter exactly when the model has something keyed on
them".

# The mesh is not your data

The mesh you just built has more vertices than you have sites:

```{r mesh-not-data}
c(sites = nlevels(dat$site), mesh_vertices = mesh$mesh$n)
```

Twelve sites, and the surface is represented at far more locations than that —
including places no one ever visited, and a margin extending beyond the data.
That is the point. The mesh is not a summary of where your data are; it is a
scaffold for the density surface, dense enough to bend where the surface bends
and extended beyond the data so the edges do not distort what is inside.

One practical consequence, which the package will tell you about if you get it
wrong: build the mesh from the same long-format data frame you pass to
`gllvmTMB()`, not from the one-row-per-site table. The projection needs one
row per data row.

# The zeros are data

Notice that the toy table is a **complete crossing**: every site has a row for
every species, whether or not that species was there. A good many of those rows
are zeros, and the scarce species contributes most of them.

```{r zeros}
c(rows = nrow(dat), sites = nlevels(dat$site), species = nlevels(dat$trait),
  zero_rows = sum(dat$value == 0))

table(species = dat$trait, absent = dat$value == 0)
```

Those zeros are not padding. Each one says *we looked here and did not find
this species*, which is genuine evidence about where the species is not. A
joint model of several species leans on them heavily.

This is also where a data source can quietly fail you. In eBird, those zeros
exist only for **complete checklists** — the ones where the observer reported
everything they detected. On an incomplete checklist, a missing species means
nothing at all, and treating it as a zero invents evidence.

A raw presence-only download from GBIF has no zeros to give. That is not a
defect to be patched by writing zeros where records are absent; it is a
different data type with its own route. Fit it with the point-process
quadrature device shown in [Joint ecological intensity from opportunistic
records](gbif-joint-intensity.html), or see the [presence-only
section](isdm-canada-warbler.html#if-your-opportunistic-arm-has-no-counts) of
the Canada Warbler article.

# Counts on both arms are fine

One thing people do unnecessarily when combining an opportunistic source with
a designed survey: binarise. If your structured survey recorded *how many*,
not merely *whether*, keep the counts. Declaring both sources Poisson is
accepted.

```{r isdm-both-poisson}
dat$isdm_source <- factor(rep(c("checklist", "station"),
                              each = n_site / 2, length.out = nrow(dat)))

both <- gllvmTMB(
  value ~ 0 + trait + offset(log_effort),
  data = dat, trait = "trait", unit = "site",
  family = isdm_sources(checklist = poisson(), station = poisson()),
  silent = TRUE
)

c(converged = both$opt$convergence == 0)
```

`isdm_sources()` names each source and its observation law; the data frame
needs a matching `isdm_source` column. Use
`isdm_sources(checklist = poisson(), station = binomial("cloglog"))` when the
structured arm really is detection/non-detection — but do not throw away
abundance you already have.

# When the point approximation blurs

Treating a unit as a point is honest when the unit is compact. It gets less
honest as the unit spreads out. A five-kilometre travelling checklist counted
birds along a route; pinning it to a single coordinate introduces a position
error you have inflicted on yourself.

Two practical responses:

- **Filter.** Restrict to stationary or short-distance checklists. eBird's
  protocol and distance fields make this a one-line subset, and it is the
  cleanest fix.
- **Accept the blur, if it is small relative to how fast the habitat changes.**
  A one-kilometre travel distance is negligible where the habitat gradient
  turns over across fifty kilometres, and serious where it turns over across
  two.

That comparison — positional error against the correlation length of the thing
you are modelling — is the whole question, and it has its own article: [When
one data source knows where it is and the other
does not](isdm-spatial-precision.html) measures what blurred coordinates cost
and gives you a ratio you can compute from your own raster.

# If you do grid, it is a choice with a price

Gridding is legitimate. Sometimes it is forced on you: atlas data arrive as
cells, or a data-sharing agreement releases only coarsened coordinates. But be
clear about what it does. Assigning a record to a cell **displaces it to the
cell centre** — every record moves, by up to half a cell diagonal — and then
pools records that may have had very different effort behind them.

Displacement is positional error by another name, so the cost depends on cell
size relative to the habitat's correlation length — exactly the ratio from the
previous section, and [When one data source knows where it is and the other
does not](isdm-spatial-precision.html) is where that cost is measured.

Two things are worth knowing before you pick a cell size.

First, results can genuinely depend on the grid. Analyses of the *same*
records on different grids can disagree — not through anyone's mistake, but
because a quantity computed over areas depends on how the areas were drawn.
Statisticians call this **change of support**; geographers call it the
**modifiable areal unit problem**. Both names describe the same fact: the
answer belongs to the grid as much as to the landscape.

Second, coarsening does not fix positional error. It is tempting to reason
that a bigger cell swallows a small location error, so a coarse grain is the
safe choice. It is not: Gábor, Jetz, Lu and colleagues tested exactly that
idea, and their title reports the answer — *Positional errors in species
distribution modelling are not overcome by the coarser grains of analysis*
(2022). Coarsening trades one loss for another rather than removing the first.

# Where to go next

- Definitions of the package's model vocabulary:
  [Plain-English vocabulary](gllvm-vocabulary.html).
- A full worked model on species records, with a map at the end:
  [An integrated SDM end to end](isdm-canada-warbler.html).
- Presence-only records on their own:
  [Joint ecological intensity from opportunistic records](gbif-joint-intensity.html).
- What imprecise coordinates cost, and the ratio that decides it:
  [When one data source knows where it is and the other does not](isdm-spatial-precision.html).
- How much structured survey you need alongside opportunistic records:
  [How much survey do you need?](integrated-survey-design.html).
