---
title: "Where does the phylogeny belong?"
description: "Use one phylogeny in two different ways: among sampled plant species or among plant-species response columns."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Where does the phylogeny belong?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggplot2)

theme_article <- theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", colour = "#17324D"),
    plot.title = element_text(face = "bold", colour = "#17324D"),
    plot.caption = element_text(hjust = 0, colour = "#555555"),
    axis.title = element_text(colour = "#17324D")
  )
```

<style>
.table-responsive {
  display: block;
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
</style>

## The decision

Suppose you have a phylogeny for a set of alpine plant species. Before putting
the tree in a model, ask one question: **do its tip labels identify the sampled
species, or the response columns?** The answer changes the model.

<div class="table-responsive">

| Your data question | The tree relates | Use |
|---|---|---|
| Do related species have similar leaf and seed traits? | sampled species | a phylogenetic intercept term |
| Do related species show similar responses to moisture and canopy cover? | species response columns | `phylo_slope(... | trait)` |

</div>

The two examples below use simulated Gaussian data and a fixed tree. They show
which variable the tree connects, not uncertainty calibration. Start with the
row whose tree tips match the variable on the right of the model term.

## Example 1: a tree among sampled species

**Question.** How do I include a tree when each sampled plant species has four
morphology traits, each measured across several populations?

Here the response columns are four measured traits. `species` is a grouping
variable: each long-format row records one trait at one population of one
species. Therefore the tree belongs to `species`, not to `trait`.

<div class="table-responsive">

| Long-data variable | Meaning | Does the tree tip match it? |
|---|---|---|
| `plot_id` | one sampled population | no |
| `trait` | leaf area, leaf mass, seed mass, or stem height | no |
| `species` | the plant species sampled at each population | **yes** |

</div>

```{r species-axis-data, include = FALSE}
library(gllvmTMB)
library(ape)

set.seed(24)
species <- sprintf("plant_%02d", 1:12)
traits <- c("leaf_area", "leaf_mass", "seed_mass", "stem_height")
tree <- rcoal(length(species))
tree$tip.label <- species
A <- vcv.phylo(tree, corr = TRUE)[species, species] + diag(1e-8, length(species))

populations <- expand.grid(
  species = species, population = 1:5, KEEP.OUT.ATTRS = FALSE
)
populations$plot_id <- sprintf("%s_%s", populations$species, populations$population)
populations$elevation <- rep(seq(-1, 1, length.out = 5), each = length(species))
species_by_trait <- t(chol(A)) %*% matrix(rnorm(length(species) * length(traits)),
  nrow = length(species), dimnames = list(species, traits)
)

morphology <- do.call(rbind, lapply(traits, function(trait_name) {
  i <- match(populations$species, species)
  data.frame(
    plot_id = populations$plot_id,
    species = factor(populations$species, levels = species),
    elevation = populations$elevation,
    trait = factor(trait_name, levels = traits),
    value = 0.35 * populations$elevation + species_by_trait[i, trait_name] +
      rnorm(nrow(populations), sd = 0.35)
  )
}))
```

```{r species-axis-fit}
fit_species <- gllvmTMB(
  value ~ 0 + trait + trait:elevation +
    phylo_latent(0 + trait | species, tree = tree, d = 1, unique = TRUE),
  data = morphology, trait = "trait", unit = "plot_id", family = gaussian(),
  silent = TRUE
)
Sigma_species_traits <- extract_Sigma(fit_species, level = "phy", part = "total")$Sigma
round(Sigma_species_traits, 2)
```

`trait:elevation` gives every measured trait its own average elevation effect.
`phylo_latent(... | species)` then models covariance among the **sampled plant
species**. Its extracted matrix describes which morphology traits tend to vary
together across phylogenetically related species: positive diagonal entries are
trait-specific variation, while off-diagonal entries are cross-trait
association. It is not a model of species-specific elevation slopes.

```{r morphology-data, fig.width = 8.4, fig.height = 6.8, fig.cap = "Example 1 in the observed long-data layout. Grey points are five population measurements within each sampled species; amber points are species means. The four panels are the morphology response columns, while the tree belongs to the sampled-species axis.", fig.alt = "Four panels showing leaf area, leaf mass, seed mass, and stem height. Each panel has five pale grey population observations for each of twelve plant species and one amber species mean."}
morphology_plot <- transform(
  morphology,
  trait_label = factor(
    tools::toTitleCase(gsub("_", " ", trait)),
    levels = tools::toTitleCase(gsub("_", " ", traits))
  )
)

ggplot(morphology_plot, aes(x = species, y = value)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "#AAB7C4") +
  geom_point(colour = "#7A8FA0", alpha = 0.55, size = 1.7) +
  stat_summary(
    fun = mean, geom = "point", shape = 21, size = 3.3, stroke = 0.7,
    fill = "#E9A34F", colour = "#173B5E"
  ) +
  facet_wrap(~ trait_label, ncol = 2, scales = "free_y") +
  labs(
    title = "Morphology is observed repeatedly within sampled species",
    subtitle = "Grey points: five populations per species; amber circles: species means",
    x = "Sampled plant species", y = "Simulated morphology value"
  ) +
  theme_article +
  theme(
    plot.title = element_text(face = "bold", colour = "#173B5E"),
    plot.subtitle = element_text(colour = "#4B5563"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold", colour = "#173B5E")
  )
```

```{r morphology-wide, fig.width = 8.4, fig.height = 6.5, fig.cap = "The same Example 1 data after a wide-data summary: each row is a sampled species and each morphology trait is a response column. Cells show standardised species means across the five population measurements; the plot is a data-layout guide, not the fitted model input.", fig.alt = "A twelve by four heatmap. Rows are sampled plant species and columns are leaf area, leaf mass, seed mass, and stem height. Each labelled tile gives a standardised species mean."}
morphology_means <- aggregate(value ~ species + trait, data = morphology, FUN = mean)
morphology_wide <- reshape(
  morphology_means, idvar = "species", timevar = "trait", direction = "wide"
)
morphology_wide <- morphology_wide[, c("species", paste0("value.", traits))]
names(morphology_wide)[-1] <- traits
morphology_standardised <- as.data.frame(scale(morphology_wide[-1]))
morphology_standardised$species <- morphology_wide$species
morphology_wide_plot <- reshape(
  morphology_standardised, varying = traits, v.names = "value",
  timevar = "trait", times = traits, direction = "long"
)
morphology_wide_plot$trait <- factor(
  morphology_wide_plot$trait, levels = traits,
  labels = tools::toTitleCase(gsub("_", " ", traits))
)
morphology_wide_plot$species <- factor(morphology_wide_plot$species, levels = rev(species))
morphology_wide_plot$text_colour <- ifelse(abs(morphology_wide_plot$value) >= 1.45, "white", "#102A43")

ggplot(morphology_wide_plot, aes(x = trait, y = species, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(
    aes(label = sprintf("%.1f", value), colour = text_colour),
    size = 3.1, show.legend = FALSE
  ) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
    name = "Standardised\nspecies mean"
  ) +
  scale_colour_identity() +
  labs(
    title = "Morphology data in wide form",
    subtitle = "Each row is one sampled plant species; the four trait measurements are response columns",
    x = "Response columns: morphology traits", y = "Rows: sampled plant species"
  ) +
  theme_article +
  theme(
    plot.title = element_text(face = "bold", colour = "#173B5E"),
    plot.subtitle = element_text(colour = "#4B5563"),
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )
```

Use this form when the data table contains species as repeated sampling units.
If instead species name the response columns, use the next model.

## Example 2: a tree among species response columns

**Question.** At a set of vegetation plots, how do I model tree-aligned
differences in plant-species moisture and canopy-cover responses?

Now each plot is a sampling unit, and each species is one Gaussian response
column. The same tree tips must therefore match `trait`. The phylogeny relates
the **coefficient deviations across species columns**, while the rows remain
independent plots.

<div class="table-responsive">

| Long-data variable | Meaning | Does the tree tip match it? |
|---|---|---|
| `site_id` | one vegetation plot | no |
| `trait` | the plant species response column | **yes** |
| `moisture`, `canopy` | plot-level predictors | no |

</div>

```{r column-axis-data, include = FALSE}
set.seed(25)
sites <- data.frame(
  site_id = sprintf("site_%02d", 1:50),
  moisture = rnorm(50),
  canopy = rnorm(50)
)
Sigma_predictor_truth <- matrix(
  c(0.35^2, -0.45 * 0.35 * 0.25,
    -0.45 * 0.35 * 0.25, 0.25^2),
  nrow = 2,
  dimnames = list(c("moisture", "canopy"), c("moisture", "canopy"))
)
slope_deviation <- t(chol(A)) %*%
  matrix(rnorm(length(species) * 2), nrow = length(species)) %*%
  chol(Sigma_predictor_truth)
slope_deviation <- sweep(slope_deviation, 2, colMeans(slope_deviation), FUN = "-")
moisture_deviation <- slope_deviation[, "moisture"]
canopy_deviation <- slope_deviation[, "canopy"]

community <- do.call(rbind, lapply(seq_along(species), function(j) {
  data.frame(
    site_id = sites$site_id,
    moisture = sites$moisture,
    canopy = sites$canopy,
    trait = factor(species[[j]], levels = species),
    value = 0.4 + (0.60 + moisture_deviation[[j]]) * sites$moisture +
      (-0.25 + canopy_deviation[[j]]) * sites$canopy + rnorm(nrow(sites), sd = 0.35)
  )
}))
```

```{r column-axis-fit}
fit_columns <- gllvmTMB(
  value ~ 0 + trait + moisture + canopy +
    phylo_slope(moisture + canopy | trait, tree = tree),
  data = community, trait = "trait", unit = "site_id", family = gaussian(),
  silent = TRUE
)
Sigma_predictor <- extract_Sigma(fit_columns, level = "column_slope")$Sigma
round(Sigma_predictor, 2)
```

`phylo_slope()` is a response-column helper, not a fourth covariance mode in
the 5 × 3 grid. The tree supplies the relationship among species columns;
`extract_Sigma(..., level = "column_slope")` returns the two-by-two covariance
of the **moisture and canopy coefficient deviations**. It is neither a
species-by-species covariance matrix nor trait-level `Sigma`.

The two diagonal entries say how much the moisture and canopy responses vary
across species columns. The off-diagonal entry says whether species with a
larger-than-average moisture response also tend to have a larger-than-average
canopy response. It does not describe covariance of the observed responses.
In this simulation, the planted predictor covariance is negative, so the
points in the next figure tend downward.

```{r community-data, fig.width = 8.4, fig.height = 5.8, fig.cap = "Example 2 in long-data form for four selected species. Each point is a site-by-species observation; each line is the planted moisture response at average canopy cover. Here plant species are response columns, so the tree aligns with the `trait` variable.", fig.alt = "A scatterplot of site-by-species observations against soil moisture for four plant species. Pale points surround four directly labelled coloured lines with different slopes."}
focal_species <- c("plant_01", "plant_04", "plant_08", "plant_12")
community_focus <- subset(community, trait %in% focal_species)
community_focus$trait <- factor(as.character(community_focus$trait), levels = focal_species)
species_effects <- data.frame(
  trait = factor(focal_species, levels = focal_species),
  intercept = 0.4,
  moisture = 0.60 + moisture_deviation[match(focal_species, species)],
  canopy = -0.25 + canopy_deviation[match(focal_species, species)]
)
moisture_grid <- seq(min(sites$moisture), max(sites$moisture), length.out = 150)
community_lines <- do.call(rbind, lapply(seq_len(nrow(species_effects)), function(i) {
  data.frame(
    trait = species_effects$trait[i], moisture = moisture_grid,
    value = species_effects$intercept[i] + species_effects$moisture[i] * moisture_grid +
      species_effects$canopy[i] * mean(sites$canopy)
  )
}))
community_ends <- do.call(rbind, lapply(split(community_lines, community_lines$trait), function(x) x[nrow(x), ]))
community_ends$label_y <- community_ends$value + c(0.14, -0.11, 0.14, -0.13)
species_colours <- c(plant_01 = "#0072B2", plant_04 = "#D55E00", plant_08 = "#009E73", plant_12 = "#7A5195")

ggplot(community_focus, aes(x = moisture, y = value, colour = trait)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "#C8CDD2") +
  geom_point(alpha = 0.36, size = 1.6) +
  geom_line(data = community_lines, aes(group = trait), linewidth = 1.05) +
  geom_segment(
    data = community_ends,
    aes(x = max(moisture_grid), xend = max(moisture_grid) + 0.14, y = value, yend = label_y),
    inherit.aes = FALSE, linewidth = 0.45
  ) +
  geom_text(
    data = community_ends,
    aes(x = max(moisture_grid) + 0.18, y = label_y, label = trait, colour = trait),
    inherit.aes = FALSE, hjust = 0, size = 3.7, fontface = "bold", show.legend = FALSE
  ) +
  scale_colour_manual(values = species_colours) +
  coord_cartesian(xlim = c(min(moisture_grid), max(moisture_grid) + 0.75), clip = "off") +
  labs(
    title = "Species are response columns",
    subtitle = "Points are site-by-species observations; lines are planted moisture responses at average canopy cover",
    x = "Soil moisture", y = "Simulated species response"
  ) +
  theme_article +
  theme(
    plot.title = element_text(face = "bold", colour = "#173B5E"),
    plot.subtitle = element_text(colour = "#4B5563"),
    panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
    legend.position = "none", plot.margin = margin(5.5, 105, 5.5, 5.5)
  )
```

```{r community-wide, fig.width = 8.4, fig.height = 7.2, fig.cap = "The same Example 2 layout in wide form. Each row is a site, the first two columns are predictors, and the next four are selected species response columns. Sites are ordered from dry to wet; colours are standardised within columns solely to make the data orientation visible.", fig.alt = "A heatmap with forty-eight sites as rows. Columns are moisture, canopy, and four selected plant species. Moisture changes from blue to red down the rows because sites are ordered from dry to wet."}
community_wide <- reshape(
  community_focus[, c("site_id", "moisture", "canopy", "trait", "value")],
  idvar = c("site_id", "moisture", "canopy"), timevar = "trait", direction = "wide"
)
community_wide <- community_wide[order(community_wide$moisture), ]
community_columns <- c("moisture", "canopy", paste0("value.", focal_species))
community_standardised <- as.data.frame(scale(community_wide[community_columns]))
names(community_standardised) <- c("moisture", "canopy", focal_species)
community_standardised$site_id <- factor(community_wide$site_id, levels = rev(community_wide$site_id))
community_wide_plot <- reshape(
  community_standardised,
  varying = c("moisture", "canopy", focal_species), v.names = "value",
  timevar = "column", times = c("moisture", "canopy", focal_species), direction = "long"
)
community_wide_plot$column <- factor(
  community_wide_plot$column,
  levels = c("moisture", "canopy", focal_species),
  labels = c("moisture\n(predictor)", "canopy\n(predictor)", focal_species)
)

ggplot(community_wide_plot, aes(x = column, y = site_id, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.25) +
  geom_vline(xintercept = 2.5, colour = "#173B5E", linewidth = 0.7) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
    name = "Standardised\nvalue"
  ) +
  labs(
    title = "Community data in wide form",
    subtitle = "Each row is a site (ordered from dry to wet); predictors sit beside four species response columns",
    x = "Columns", y = "Rows: sites"
  ) +
  theme_article +
  theme(
    plot.title = element_text(face = "bold", colour = "#173B5E"),
    plot.subtitle = element_text(colour = "#4B5563"),
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_blank(), axis.ticks.y = element_blank(),
    legend.title = element_text(face = "bold")
  )
```

With several predictors, `|` estimates their full covariance and `||` fixes
their covariance to zero while retaining one variance per predictor. This
response-column route is currently Gaussian and long-format only. Wide
column-slope grammar, non-Gaussian or mixed-family slopes, latent predictor
covariance, simultaneous column-slope sources, and intervals are deliberately
deferred.

## Choose the axis before the helper

The name after the bar is the practical check:

```r
# Tree among sampled species: species is the grouping axis
phylo_latent(0 + trait | species, tree = tree, d = 1, unique = TRUE)

# Tree among response columns: species are the trait levels
phylo_slope(moisture + canopy | trait, tree = tree)
```

If your scientific question is about trait covariance among sampled species,
begin with the first form. If it is about environmental-response coefficients
varying across species columns, begin with the second. For the source ×
trait-covariance lookup, see the [Formula keyword grid](api-keyword-grid.html).
