---
title: "Where does the phylogeny belong?"
description: "Use one phylogeny in two different ways: among sampled plant species or among plant-species response columns."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Where does the phylogeny belong?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggplot2)

theme_article <- theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", colour = "#17324D"),
    plot.title = element_text(face = "bold", colour = "#17324D"),
    plot.caption = element_text(hjust = 0, colour = "#555555"),
    axis.title = element_text(colour = "#17324D")
  )
```

<style>
.table-responsive {
  display: block;
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
</style>

## The decision

Suppose you have a phylogeny for a set of alpine plant species. Before putting
the tree in a model, ask one question: **do its tip labels identify the sampled
species, or the response columns?** The answer changes the model.

<div class="table-responsive">

| Your data question | The tree relates | Use |
|---|---|---|
| Do related species have similar leaf and seed traits? | sampled species | a phylogenetic intercept term |
| Do related species have similar deviations from a C3/C4 latitude gradient? | species response columns | `phylo_slope(... | trait)` |

</div>

The two examples below use simulated Gaussian data and a fixed tree. They show
which variable the tree connects, not uncertainty calibration. Start with the
row whose tree tips match the variable on the right of the model term.

```{r packages, message = FALSE}
library(gllvmTMB)
library(ape)
```

## Example 1: a tree among sampled species

**Question.** How do I include a tree when each sampled plant species has four
morphology traits, each measured across several populations?

Here the response columns are four measured traits. `species` is a grouping
variable: each long-format row records one trait at one population of one
species. Therefore the tree belongs to `species`, not to `trait`.

<div class="table-responsive">

| Long-data variable | Meaning | Does the tree tip match it? |
|---|---|---|
| `plot_id` | one sampled population | no |
| `trait` | leaf area, leaf mass, seed mass, or stem height | no |
| `species` | the plant species sampled at each population | **yes** |

</div>

```{r species-axis-data}
set.seed(24)
species <- sprintf("plant_%02d", 1:12)
traits <- c("leaf_area", "leaf_mass", "seed_mass", "stem_height")
tree <- rcoal(length(species))
tree$tip.label <- species
A <- vcv.phylo(tree, corr = TRUE)[species, species] + diag(1e-8, length(species))

populations <- expand.grid(
  species = species, population = 1:5, KEEP.OUT.ATTRS = FALSE
)
populations$plot_id <- sprintf("%s_%s", populations$species, populations$population)
populations$elevation <- rep(seq(-1, 1, length.out = 5), each = length(species))
species_by_trait <- t(chol(A)) %*% matrix(rnorm(length(species) * length(traits)),
  nrow = length(species), dimnames = list(species, traits)
)

morphology <- do.call(rbind, lapply(traits, function(trait_name) {
  i <- match(populations$species, species)
  data.frame(
    plot_id = populations$plot_id,
    species = factor(populations$species, levels = species),
    elevation = populations$elevation,
    trait = factor(trait_name, levels = traits),
    value = 0.35 * populations$elevation + species_by_trait[i, trait_name] +
      rnorm(nrow(populations), sd = 0.35)
  )
}))
```

```{r species-axis-fit}
fit_species <- gllvmTMB(
  value ~ 0 + trait + trait:elevation +
    phylo_latent(0 + trait | species, tree = tree, d = 1, unique = TRUE),
  data = morphology, trait = "trait", unit = "plot_id", family = gaussian(),
  silent = TRUE
)
Sigma_species_traits <- extract_Sigma(fit_species, level = "phy", part = "total")$Sigma
round(Sigma_species_traits, 2)
```

`trait:elevation` gives every measured trait its own average elevation effect.
`phylo_latent(... | species)` then models covariance among the **sampled plant
species**. Its extracted matrix describes which morphology traits tend to vary
together across phylogenetically related species: positive diagonal entries are
trait-specific variation, while off-diagonal entries are cross-trait
association. It is not a model of species-specific elevation slopes.

```{r morphology-data, echo = FALSE, fig.width = 8.4, fig.height = 6.8, fig.cap = "Example 1 in the observed long-data layout. Grey points are five population measurements within each sampled species; amber points are species means. The four panels are the morphology response columns, while the tree belongs to the sampled-species axis.", fig.alt = "Four panels showing leaf area, leaf mass, seed mass, and stem height. Each panel has five pale grey population observations for each of twelve plant species and one amber species mean."}
morphology_plot <- transform(
  morphology,
  trait_label = factor(
    tools::toTitleCase(gsub("_", " ", trait)),
    levels = tools::toTitleCase(gsub("_", " ", traits))
  )
)

ggplot(morphology_plot, aes(x = species, y = value)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "#AAB7C4") +
  geom_point(colour = "#7A8FA0", alpha = 0.55, size = 1.7) +
  stat_summary(
    fun = mean, geom = "point", shape = 21, size = 3.3, stroke = 0.7,
    fill = "#E9A34F", colour = "#173B5E"
  ) +
  facet_wrap(~ trait_label, ncol = 2, scales = "free_y") +
  labs(
    title = "Morphology is observed repeatedly within sampled species",
    subtitle = "Grey points: five populations per species;\namber circles: species means",
    x = "Sampled plant species", y = "Simulated morphology value"
  ) +
  theme_article +
  theme(
    plot.title = element_text(face = "bold", colour = "#173B5E"),
    plot.subtitle = element_text(colour = "#4B5563"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold", colour = "#173B5E")
  )
```

```{r morphology-wide, echo = FALSE, fig.width = 8.4, fig.height = 6.5, fig.cap = "The same Example 1 data after a wide-data summary: each row is a sampled species and each morphology trait is a response column. Cells show standardised species means across the five population measurements; the plot is a data-layout guide, not the fitted model input.", fig.alt = "A twelve by four heatmap. Rows are sampled plant species and columns are leaf area, leaf mass, seed mass, and stem height. Each labelled tile gives a standardised species mean."}
morphology_means <- aggregate(value ~ species + trait, data = morphology, FUN = mean)
morphology_wide <- reshape(
  morphology_means, idvar = "species", timevar = "trait", direction = "wide"
)
morphology_wide <- morphology_wide[, c("species", paste0("value.", traits))]
names(morphology_wide)[-1] <- traits
morphology_standardised <- as.data.frame(scale(morphology_wide[-1]))
morphology_standardised$species <- morphology_wide$species
morphology_wide_plot <- reshape(
  morphology_standardised, varying = traits, v.names = "value",
  timevar = "trait", times = traits, direction = "long"
)
morphology_wide_plot$trait <- factor(
  morphology_wide_plot$trait, levels = traits,
  labels = tools::toTitleCase(gsub("_", " ", traits))
)
morphology_wide_plot$species <- factor(morphology_wide_plot$species, levels = rev(species))
morphology_wide_plot$text_colour <- ifelse(abs(morphology_wide_plot$value) >= 1.45, "white", "#102A43")

ggplot(morphology_wide_plot, aes(x = trait, y = species, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(
    aes(label = sprintf("%.1f", value), colour = text_colour),
    size = 3.1, show.legend = FALSE
  ) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
    name = "Standardised\nspecies mean"
  ) +
  scale_colour_identity() +
  labs(
    title = "Morphology data in wide form",
    subtitle = "Each row is one sampled plant species;\nthe four trait measurements are response columns",
    x = "Response columns: morphology traits", y = "Rows: sampled plant species"
  ) +
  theme_article +
  theme(
    plot.title = element_text(face = "bold", colour = "#173B5E"),
    plot.subtitle = element_text(colour = "#4B5563"),
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )
```

Use this form when the data table contains species as repeated sampling units.
If instead species name the response columns, use the next model.

## Example 2: a tree among species response columns

**Question.** Across a latitude gradient, how can I estimate different average
responses for C3 and C4 plant species while allowing related species to have
similar species-specific deviations from those averages?

Each site is now a row of the community matrix and each plant species is a
response column. To stay inside the currently tested scope, the response is a
continuous Gaussian intensity: a simulated log-biomass index. Presence–absence
and count responses need the deferred non-Gaussian column-slope extension.

The example separates three different jobs. C3 and C4 are photosynthetic
pathways; contrasting their latitude responses is a simple way to ask whether
two ecological strategies follow different environmental gradients.
`latitude` changes among sites. `pathway` is species-column metadata copied
down the long data for each species. The package calls its response-column
identifier `trait`; here each `trait` level is a plant species. The fixed
`latitude:pathway` term estimates the two pathway mean gradients.
`phylo_slope(latitude | trait)` then lets the latitude coefficient vary among
species columns, with related species receiving correlated deviations from
their pathway mean.

These are different model components. `latitude:pathway` is a **fixed effect**;
`phylo_slope(latitude | trait)` is a **random-effect distribution** for the
remaining species-specific slope deviations. There is no `column_coef()`
helper in the current API. Response-column metadata such as `pathway` enters
the fixed-effects formula directly. `phylo_slope()` uses the supplied tree
covariance directly; it does not estimate a separate IID-versus-phylogenetic
signal mixture.

<div class="table-responsive">

| Long-data variable | Meaning | Does the tree tip match it? |
|---|---|---|
| `site_id` | one vegetation plot | no |
| `trait` | the plant species response column | **yes** |
| `latitude` | a site-level predictor, repeated for every species at that site | no |
| `pathway` | a C3/C4 attribute of each species column | no — it groups tips, but does not identify them |

</div>

```{r column-axis-data}
set.seed(25)
sites <- data.frame(
  site_id = sprintf("site_%02d", 1:50),
  latitude = as.numeric(scale(seq(38, 62, length.out = 50) + rnorm(50, sd = 0.7)))
)
species_info <- data.frame(
  trait = factor(species, levels = species),
  pathway = factor(rep(c("C3", "C4"), each = 6), levels = c("C3", "C4"))
)
pathway_slope <- c(C3 = -0.30, C4 = 0.55)
latitude_deviation <- 0.18 * as.numeric(
  t(chol(A)) %*% rnorm(length(species))
)
species_intercept <- 0.7 + rnorm(length(species), sd = 0.18)

site_scores <- matrix(rnorm(nrow(sites) * 2), ncol = 2)
species_loadings <- matrix(rnorm(length(species) * 2, sd = 0.18), ncol = 2)

community <- do.call(rbind, lapply(seq_along(species), function(j) {
  pathway_j <- as.character(species_info$pathway[[j]])
  data.frame(
    site_id = sites$site_id,
    latitude = sites$latitude,
    trait = factor(species[[j]], levels = species),
    pathway = factor(pathway_j, levels = c("C3", "C4")),
    log_biomass = species_intercept[[j]] +
      (pathway_slope[[pathway_j]] + latitude_deviation[[j]]) * sites$latitude +
      as.numeric(site_scores %*% species_loadings[j, ]) +
      rnorm(nrow(sites), sd = 0.25)
  )
}))

head(species_info, 4)
head(community[c("site_id", "trait", "pathway", "latitude", "log_biomass")], 6)
```

```{r column-axis-fit}
fit_columns <- gllvmTMB(
  log_biomass ~ 0 + trait + latitude:pathway +
    phylo_slope(latitude | trait, tree = tree) +
    latent(0 + trait | site_id, d = 2, unique = FALSE),
  data = community, trait = "trait", unit = "site_id", family = gaussian(),
  silent = TRUE
)
fixed_pathway_slopes <- coef(fit_columns)[
  grepl("^latitude:pathway", names(coef(fit_columns)))
]
pathway_difference <- unname(
  fixed_pathway_slopes[["latitude:pathwayC4"]] -
    fixed_pathway_slopes[["latitude:pathwayC3"]]
)
round(c(fixed_pathway_slopes, `C4 minus C3` = pathway_difference), 2)

Sigma_predictor <- extract_Sigma(fit_columns, level = "column_slope")$Sigma
round(sqrt(Sigma_predictor["latitude", "latitude"]), 2)
```

The first two fixed coefficients are the average latitude slopes for C3 and C4
species. The printed `C4 minus C3` value is their fitted difference in this
simulated sample. That number is descriptive: this article does not calculate
a test or interval for the contrast. The extracted standard deviation describes the residual
species-specific latitude slopes after those pathway means have been fitted.
The tree determines how those deviations covary across species.

`latent(0 + trait | site_id, d = 2, unique = FALSE)` has a separate job: its
two latent variables model remaining association among species within sites.
The explicit `unique = FALSE` leaves independent observation noise to the
Gaussian residual. This term does not turn `pathway` into a random slope and it
does not replace the phylogenetic latitude-slope term. Here it is a nuisance
adjustment rather than a second estimand interpreted in the article.

```{r community-data, echo = FALSE, fig.width = 8.4, fig.height = 5.8, fig.cap = "Example 2 for six selected species. Pale points are observed site-by-species log-biomass values. Thin lines are the planted species-specific latitude responses; thick lines are the C3 and C4 pathway means. The phylogenetic slope term models the deviations of thin lines from their pathway mean.", fig.alt = "Two panels labelled C3 and C4. Each contains site-by-species observations, thin species-specific trend lines, and one thick pathway-average line. The C3 mean decreases with latitude and the C4 mean increases."}
focal_species <- c("plant_01", "plant_03", "plant_05", "plant_07", "plant_09", "plant_11")
community_focus <- subset(community, trait %in% focal_species)
community_focus$trait <- factor(as.character(community_focus$trait), levels = focal_species)
species_effects <- data.frame(
  trait = factor(focal_species, levels = focal_species),
  pathway = species_info$pathway[match(focal_species, species)],
  intercept = species_intercept[match(focal_species, species)],
  latitude_slope = pathway_slope[
    as.character(species_info$pathway[match(focal_species, species)])
  ] + latitude_deviation[match(focal_species, species)]
)
latitude_grid <- seq(min(sites$latitude), max(sites$latitude), length.out = 150)
community_lines <- do.call(rbind, lapply(seq_len(nrow(species_effects)), function(i) {
  data.frame(
    trait = species_effects$trait[i], pathway = species_effects$pathway[i],
    latitude = latitude_grid,
    log_biomass = species_effects$intercept[i] +
      species_effects$latitude_slope[i] * latitude_grid
  )
}))
pathway_lines <- do.call(rbind, lapply(levels(species_info$pathway), function(p) {
  data.frame(
    pathway = factor(p, levels = levels(species_info$pathway)),
    latitude = latitude_grid,
    log_biomass = mean(species_intercept[species_info$pathway == p]) +
      pathway_slope[[p]] * latitude_grid
  )
}))
pathway_colours <- c(C3 = "#0072B2", C4 = "#D55E00")

ggplot(community_focus, aes(x = latitude, y = log_biomass)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "#C8CDD2") +
  geom_point(colour = "#7A8FA0", alpha = 0.30, size = 1.25) +
  geom_line(
    data = community_lines, aes(group = trait),
    colour = "#667788", linewidth = 0.55, alpha = 0.75
  ) +
  geom_line(
    data = pathway_lines, aes(colour = pathway),
    linewidth = 1.6, show.legend = FALSE
  ) +
  facet_wrap(~ pathway, nrow = 1) +
  scale_colour_manual(values = pathway_colours) +
  labs(
    title = "The simulated C3 and C4 gradients differ by design",
    subtitle = "Thin lines: species-specific responses;\nthick lines: pathway means",
    x = "Latitude (centred and scaled)", y = "Simulated log-biomass index"
  ) +
  theme_article +
  theme(
    plot.title = element_text(face = "bold", colour = "#173B5E"),
    plot.subtitle = element_text(colour = "#4B5563"),
    panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
    legend.position = "none"
  )
```

```{r community-wide, echo = FALSE, fig.width = 8.4, fig.height = 7.2, fig.cap = "The same Example 2 data in wide form for eight selected species. Each row is a site, latitude is the row-level predictor, and the remaining columns are species responses. C3/C4 pathway labels belong to species columns, not to sites. Values are standardised within columns only to reveal the data layout.", fig.alt = "A heatmap with sites as rows and nine columns. The first column is latitude; the other eight are plant species labelled C3 or C4. Sites are ordered from low to high latitude."}
wide_species <- c("plant_01", "plant_02", "plant_04", "plant_06", "plant_07", "plant_08", "plant_10", "plant_12")
community_wide_source <- subset(community, trait %in% wide_species)
community_wide <- reshape(
  community_wide_source[, c("site_id", "latitude", "trait", "log_biomass")],
  idvar = c("site_id", "latitude"), timevar = "trait", direction = "wide"
)
community_wide <- community_wide[order(community_wide$latitude), ]
community_columns <- c("latitude", paste0("log_biomass.", wide_species))
community_standardised <- as.data.frame(scale(community_wide[community_columns]))
names(community_standardised) <- c("latitude", wide_species)
community_standardised$site_id <- factor(community_wide$site_id, levels = rev(community_wide$site_id))
community_wide_plot <- reshape(
  community_standardised,
  varying = c("latitude", wide_species), v.names = "value",
  timevar = "column", times = c("latitude", wide_species), direction = "long"
)
wide_pathway <- as.character(species_info$pathway[match(wide_species, species)])
community_wide_plot$column <- factor(
  community_wide_plot$column,
  levels = c("latitude", wide_species),
  labels = c("latitude\n(site predictor)", paste0(wide_species, "\n", wide_pathway))
)

ggplot(community_wide_plot, aes(x = column, y = site_id, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.25) +
  geom_vline(xintercept = 1.5, colour = "#173B5E", linewidth = 0.8) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
    name = "Standardised\nvalue"
  ) +
  labs(
    title = "Sites are rows; plant species are response columns",
    subtitle = "Rows run from low to high latitude;\nC3/C4 is metadata attached to species columns",
    x = "Columns", y = "Rows: sites"
  ) +
  theme_article +
  theme(
    plot.title = element_text(face = "bold", colour = "#173B5E"),
    plot.subtitle = element_text(colour = "#4B5563"),
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_blank(), axis.ticks.y = element_blank(),
    legend.title = element_text(face = "bold")
  )
```

The current helper's coefficient block contains slopes but no phylogenetic
random intercept. `0 + trait` supplies a separate fixed intercept for every
species; it does **not** estimate phylogenetically correlated random intercepts
among species columns. A future intercept-plus-slope coefficient block needs
its own design rather than a silent change to `phylo_slope()`.

With several site-level predictors, `|` estimates their full coefficient
covariance and `||` fixes their covariance to zero while retaining one variance
per predictor. This response-column route is currently Gaussian and
long-format only. Wide column-slope grammar, non-Gaussian or mixed-family
slopes, latent predictor covariance, simultaneous column-slope sources, and
intervals are deliberately deferred.

## Choose the axis before the helper

The name after the bar is the practical check:

```r
# Tree among sampled species: species is the grouping axis
phylo_latent(0 + trait | species, tree = tree, d = 1, unique = TRUE)

# Tree among response columns: latitude varies by site,
# while its coefficient varies by species column
phylo_slope(latitude | trait, tree = tree)
```

If your scientific question is about trait covariance among sampled species,
begin with the first form. If it is about environmental-response coefficients
varying across species columns, begin with the second. A species attribute such
as C3/C4 pathway belongs in a fixed interaction such as `latitude:pathway`, not
inside `phylo_slope()`. Thus the ordinary formula supplies the fixed group
contrast, while `*_slope()` supplies random response-column deviations around
that contrast. For the source × trait-covariance lookup, see the
[Formula keyword grid](api-keyword-grid.html).
