---
title: "Where does the tree go?"
description: "Worked examples for phylogenetic and ordinary covariance on distinct model axes."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Where does the tree go?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

# The tree belongs to the axis it explains

A phylogeny supplies covariance for species labels on a particular model axis.
It does not replace the other covariance sources in the scientific model.
Here species are the **rows** of a morphology matrix, but the **columns** of a
community matrix. Both examples use synthetic Gaussian responses.

**Working draft:** the morphology checks passed, but the community fits have
unresolved convergence warnings. This revision is not ready for publication.

```{r packages}
library(gllvmTMB)
library(ape)
library(ggplot2)
```

<details>
<summary>Simulation helpers: run this code once before the examples</summary>

The code below fixes the trees, predictors, loading matrices and diagonal
variances before fitting. It generates one realization, not a recovery study.

```{r simulation-helpers}
## Synthetic data construction; all seeds and covariance parameters are fixed.

TREE_AXIS_FIXTURE_VERSION <- "tree-axis-latent-v1"
TREE_AXIS_SEEDS <- list(
  morphology_tree = 202608301L,
  morphology_phy = 202608302L,
  morphology_non = 202608303L,
  morphology_elevation = 202608304L,
  community_tree = 202608401L,
  community_latitude = 202608402L,
  community_coef = 202608403L,
  community_site = 202608404L
)

.tree_axis_unit_height <- function(tree) {
  height <- max(ape::node.depth.edgelength(tree))
  if (!is.finite(height) || height <= 0) stop("Tree height must be positive.")
  tree$edge.length <- tree$edge.length / height
  tree
}

.tree_axis_tree_covariance <- function(tree, labels) {
  tip_heights <- ape::node.depth.edgelength(tree)[seq_along(tree$tip.label)]
  if (max(abs(tip_heights - 1)) > 1e-10) {
    stop("The frozen tree must be ultrametric with unit tip height.")
  }
  K <- ape::vcv.phylo(tree, corr = TRUE)[labels, labels, drop = FALSE]
  K_unit_height <- ape::vcv.phylo(tree, corr = FALSE)[labels, labels, drop = FALSE]
  if (max(abs(K - K_unit_height)) > 1e-8) {
    stop("Tree covariance changed scale after unit-height normalization.")
  }
  K
}

.tree_axis_draw_source <- function(K, Lambda, Psi, seed) {
  set.seed(seed)
  n <- nrow(K)
  p <- nrow(Lambda)
  L_K <- t(chol(K))
  shared <- L_K %*% matrix(rnorm(n * ncol(Lambda)), n, ncol(Lambda)) %*%
    t(Lambda)
  unique <- L_K %*% matrix(rnorm(n * p), n, p) %*% chol(Psi)
  list(shared = shared, unique = unique, total = shared + unique)
}

.tree_axis_draw_iid <- function(n, Lambda, Psi, seed) {
  set.seed(seed)
  p <- nrow(Lambda)
  shared <- matrix(rnorm(n * ncol(Lambda)), n, ncol(Lambda)) %*% t(Lambda)
  unique <- matrix(rnorm(n * p), n, p) %*% chol(Psi)
  list(shared = shared, unique = unique, total = shared + unique)
}

.tree_axis_morphology <- function(size = c("target", "canary")) {
  size <- match.arg(size)
  set.seed(TREE_AXIS_SEEDS$morphology_tree)
  n_species <- if (identical(size, "target")) 80L else 20L
  species <- sprintf("morph_%03d", seq_len(n_species))
  traits <- c(
    "leaf_area", "specific_leaf_area", "leaf_dry_matter",
    "stem_height", "seed_mass", "root_depth"
  )
  tree <- .tree_axis_unit_height(ape::rcoal(n_species))
  tree$tip.label <- species
  K <- .tree_axis_tree_covariance(tree, species)

  ## Two deliberately non-collinear trait contrasts.  Their signs label a
  ## synthetic fixture; they do not assert mechanisms in real plants.
  Lambda_phy <- matrix(c(
    0.62, 0.10,
    0.48, 0.34,
   -0.18, 0.58,
    0.71,-0.08,
    0.10, 0.64,
    0.42,-0.38
  ), nrow = length(traits), byrow = TRUE,
  dimnames = list(traits, c("leaf_stature", "reproductive_structure")))
  Lambda_non <- matrix(c(
    0.22, 0.46,
    0.54,-0.16,
    0.63, 0.28,
   -0.15, 0.52,
    0.48,-0.24,
    0.08, 0.57
  ), nrow = length(traits), byrow = TRUE,
  dimnames = list(traits, c("resource_axis", "independent_axis")))
  psi_phy <- c(0.20, 0.16, 0.24, 0.18, 0.22, 0.14)
  psi_non <- c(0.18, 0.25, 0.17, 0.21, 0.15, 0.23)
  Psi_phy <- diag(psi_phy)
  Psi_non <- diag(psi_non)
  dimnames(Psi_phy) <- dimnames(Psi_non) <- list(traits, traits)

  phy <- .tree_axis_draw_source(K, Lambda_phy, Psi_phy, TREE_AXIS_SEEDS$morphology_phy)
  non <- .tree_axis_draw_iid(n_species, Lambda_non, Psi_non,
                              TREE_AXIS_SEEDS$morphology_non)
  set.seed(TREE_AXIS_SEEDS$morphology_elevation)
  elevation <- as.numeric(scale(seq(-1.6, 1.6, length.out = n_species) +
    rnorm(n_species, sd = 0.20)))
  trait_intercept <- c(0.25, 0.05, -0.15, 0.40, -0.25, 0.10)
  trait_elevation <- c(0.20, 0.32, -0.18, 0.38, -0.10, 0.14)
  Y <- sweep(phy$total + non$total, 2L, trait_intercept, "+") +
    tcrossprod(elevation, trait_elevation)
  colnames(Y) <- traits

  wide <- data.frame(species = factor(species, levels = species), elevation = elevation,
                     check.names = FALSE)
  wide[traits] <- as.data.frame(Y, check.names = FALSE)
  long <- data.frame(
    species = factor(rep(species, each = length(traits)), levels = species),
    elevation = rep(elevation, each = length(traits)),
    trait = factor(rep(traits, times = n_species), levels = traits),
    value = as.vector(t(Y))
  )
  list(
    species = species, traits = traits, tree = tree, K = K, long = long, wide = wide,
    truth = list(
      Lambda_phy = Lambda_phy, Psi_phy = Psi_phy,
      Sigma_phy_shared = tcrossprod(Lambda_phy),
      Sigma_phy_total = tcrossprod(Lambda_phy) + Psi_phy,
      Lambda_non = Lambda_non, Psi_non = Psi_non,
      Sigma_non_shared = tcrossprod(Lambda_non),
      Sigma_non_total = tcrossprod(Lambda_non) + Psi_non,
      elevation = trait_elevation
    )
  )
}

.tree_axis_community <- function(size = c("target", "canary")) {
  size <- match.arg(size)
  n_sites <- if (identical(size, "target")) 150L else 50L
  n_species <- if (identical(size, "target")) 50L else 20L
  species <- sprintf("plant_%03d", seq_len(n_species))
  set.seed(TREE_AXIS_SEEDS$community_latitude)
  sites <- data.frame(
    site_id = sprintf("site_%03d", seq_len(n_sites)),
    latitude = as.numeric(scale(seq(38, 62, length.out = n_sites) +
      rnorm(n_sites, sd = 0.7)))
  )
  set.seed(TREE_AXIS_SEEDS$community_tree)
  tree <- .tree_axis_unit_height(ape::rcoal(n_species))
  tree$tip.label <- species
  K <- .tree_axis_tree_covariance(tree, species)
  pathway <- factor(rep(c("C3", "C4"), length.out = n_species), levels = c("C3", "C4"))
  column_data <- data.frame(trait = species, pathway = pathway)

  pathway_intercept <- c(C3 = 0.85, C4 = 0.45)
  pathway_slope <- c(C3 = -0.30, C4 = 0.55)
  rho_truth <- 0.60
  K_rho <- rho_truth * K + (1 - rho_truth) * diag(diag(K))
  coefficient_sd <- c(`(Intercept)` = 0.18, latitude = 0.55)
  coefficient_cor <- -0.25
  Sigma_coef <- outer(coefficient_sd, coefficient_sd) *
    matrix(c(1, coefficient_cor, coefficient_cor, 1), 2L,
           dimnames = list(names(coefficient_sd), names(coefficient_sd)))
  set.seed(TREE_AXIS_SEEDS$community_coef)
  coefficient_draw <- t(chol(K_rho)) %*%
    matrix(rnorm(n_species * 2L), n_species, 2L) %*% chol(Sigma_coef)
  colnames(coefficient_draw) <- names(coefficient_sd)
  rownames(coefficient_draw) <- species

  ## Conditional residual association across species at a site.  There is no
  ## fifth Gaussian noise term: Psi_site is the cell-level independent part.
  Lambda_site <- cbind(
    0.48 * (sin(seq_len(n_species) * 0.22) + 0.20),
    -0.32 * (cos(seq_len(n_species) * 0.31) - 0.15)
  )
  rownames(Lambda_site) <- species
  colnames(Lambda_site) <- c("cooccurrence_axis_1", "cooccurrence_axis_2")
  psi_site <- seq(0.12, 0.28, length.out = n_species)
  Psi_site <- diag(psi_site)
  dimnames(Psi_site) <- list(species, species)
  residual <- .tree_axis_draw_iid(n_sites, Lambda_site, Psi_site,
                                   TREE_AXIS_SEEDS$community_site)$total
  alpha <- pathway_intercept[as.character(pathway)] + coefficient_draw[, "(Intercept)"]
  beta <- pathway_slope[as.character(pathway)] + coefficient_draw[, "latitude"]
  Y <- outer(rep(1, n_sites), alpha) + tcrossprod(sites$latitude, beta) + residual
  colnames(Y) <- species

  wide <- data.frame(site_id = sites$site_id, latitude = sites$latitude,
                     check.names = FALSE)
  wide[species] <- as.data.frame(Y, check.names = FALSE)
  long <- data.frame(
    site_id = factor(rep(sites$site_id, each = n_species), levels = sites$site_id),
    latitude = rep(sites$latitude, each = n_species),
    trait = factor(rep(species, times = n_sites), levels = species),
    pathway = factor(rep(as.character(pathway), times = n_sites), levels = c("C3", "C4")),
    value = as.vector(t(Y))
  )
  list(
    species = species, sites = sites, tree = tree, K = K,
    long = long, wide = wide, column_data = column_data,
    truth = list(
      pathway_intercept = pathway_intercept, pathway_slope = pathway_slope,
      rho = rho_truth, K_rho = K_rho, Sigma_coef = Sigma_coef,
      coefficient_draw = coefficient_draw, Lambda_site = Lambda_site,
      Psi_site = Psi_site, Sigma_site_shared = tcrossprod(Lambda_site),
      Sigma_site_total = tcrossprod(Lambda_site) + Psi_site
    )
  )
}

make_tree_axis_fixture <- function(size = c("target", "canary")) {
  size <- match.arg(size)
  out <- list(
    version = TREE_AXIS_FIXTURE_VERSION,
    morphology = .tree_axis_morphology(size),
    community = .tree_axis_community(size)
  )
  stopifnot(
    identical(levels(out$morphology$long$species), out$morphology$tree$tip.label),
    identical(levels(out$community$long$trait), out$community$tree$tip.label),
    nrow(out$morphology$long) == length(out$morphology$species) * length(out$morphology$traits),
    nrow(out$community$long) == nrow(out$community$sites) * length(out$community$species),
    qr(out$community$truth$Lambda_site)$rank == 2L
  )
  out
}

```

</details>

```{r plot-helper}
corplot <- function(S, title) {
  z <- as.data.frame(as.table(stats::cov2cor(S)))
  names(z) <- c("first", "second", "r")
  p <- ggplot(z, aes(first, second, fill = r)) +
    geom_tile(colour = "white", linewidth = 0.15) +
    scale_fill_gradient2(low = "#b2182b", mid = "white", high = "#2166ac",
                         limits = c(-1, 1)) +
    coord_equal() + labs(title = title, x = NULL, y = NULL, fill = "Correlation") +
    theme_minimal(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  if (nrow(S) <= 12) {
    p <- p + geom_text(aes(label = sprintf("%.2f", r)), size = 2.6)
  } else {
    keep <- rownames(S)[unique(c(1L, seq(10L, nrow(S), by = 10L)))]
    p <- p + scale_x_discrete(breaks = keep) + scale_y_discrete(breaks = keep)
  }
  p
}
```

```{r axis-map, fig.width=8, fig.height=3.5}
axes <- data.frame(
  example = c("1. Species × traits", "2. Sites × species"),
  rows = c("Species rows: tree A", "Site rows: ordinary latent()"),
  columns = c("Trait columns", "Species columns: tree K for coefficients"),
  components = c("phylo_latent() + latent()\nBoth group by species",
                 "phylo_coef() + latent()\nCoefficients by species; residuals by site")
)
ggplot(axes) +
  annotate("rect", xmin = 0, xmax = 3, ymin = 0, ymax = 2,
           fill = "#e4edf2", colour = "#31576b") +
  geom_text(aes(x = 1.5, y = 1, label = components), size = 3.2) +
  geom_text(aes(x = 1.5, y = 2.4, label = columns), size = 3) +
  geom_text(aes(x = -0.4, y = 1, label = rows), angle = 90, size = 3) +
  facet_wrap(~ example) + coord_cartesian(xlim = c(-0.8, 3.3), ylim = c(-0.1, 2.8)) +
  labs(caption = "Schematic response matrices. Each tree follows species labels; the ordinary source remains in the model.") +
  theme_void(base_size = 11) + theme(strip.text = element_text(face = "bold"))
```

# Example 1: six traits, one measurement per species

We simulate 80 species and six standardized traits: leaf area, specific leaf
area, leaf dry matter content, stem height, seed mass, and root depth. Each
species has exactly one multivariate observation: there are no populations and
no within-species variance.

\[
Y=XB+U_{\rm phy}+U_{\rm non}, \qquad
\Sigma_k=\Lambda_k\Lambda_k^\mathsf T+\Psi_k,
\]
\[
\operatorname{Cov}\{\operatorname{vec}(Y^\mathsf T)\}
=A\otimes\Sigma_{\rm phy}+I\otimes\Sigma_{\rm non}.
\]

Both sources are among species. The fixed effects in $XB$ are trait means
and trait-specific effects of standardized species elevation. $A$ is the
unit-diagonal tree covariance; $I$ makes different species independent in the
ordinary source. $\Psi_k$ is a diagonal matrix of positive trait-specific
variances. The vectorization above stacks all traits for each species together.
Two latent dimensions alone do not establish separation of the sources.

```{r morphology-dgp}
fixture <- make_tree_axis_fixture("target")
traits_morph <- fixture$morphology$traits
tree_morph <- fixture$morphology$tree
morphology <- fixture$morphology$long
morphology_wide <- fixture$morphology$wide
truth_morphology <- fixture$morphology$truth
A <- fixture$morphology$K
I <- diag(nrow(A))
source_columns <- cbind(A[lower.tri(A, diag = TRUE)], I[lower.tri(I, diag = TRUE)])
source_columns <- sweep(source_columns, 2, sqrt(colSums(source_columns^2)), "/")
c(source_rank = qr(source_columns)$rank,
  normalized_condition = kappa(source_columns, exact = TRUE))
```

This check distinguishes the two covariance patterns algebraically. It does
not guarantee accurate variance estimates from 80 species.

```{r morphology-long}
fit_morphology <- gllvmTMB(
 value ~ 0+trait+trait:elevation+
   phylo_latent(0+trait|species,tree=tree_morph,d=2,unique=TRUE)+
   latent(0+trait|species,d=2,unique=TRUE),
 data=morphology,trait="trait",unit="species",cluster="species",
 family=gaussian(),silent=TRUE,control=gllvmTMBcontrol(se=FALSE)
)
```

The matching wide-data call uses the same public entry point.

```{r morphology-wide, eval=FALSE}
fit_morphology_wide <- gllvmTMB(
 traits(all_of(traits_morph)) ~ 1+elevation+
   phylo_latent(1|species,tree=tree_morph,d=2,unique=TRUE)+
   latent(1|species,d=2,unique=TRUE),
 data=morphology_wide,unit="species",cluster="species",family=gaussian(),silent=TRUE,control=gllvmTMBcontrol(se=FALSE)
)
```

```{r morphology-figure, fig.width=8, fig.height=6}
Sphy <- extract_Sigma(fit_morphology,level="phy",part="total",link_residual="none")$Sigma
Snon <- extract_Sigma(fit_morphology,level="unit",part="total",link_residual="none")$Sigma
Sps <- extract_Sigma(fit_morphology,level="phy",part="shared",link_residual="none")$Sigma
Sns <- extract_Sigma(fit_morphology,level="unit",part="shared",link_residual="none")$Sigma
pp <- extract_Sigma(fit_morphology,level="phy",part="unique",link_residual="none")$s
pn <- extract_Sigma(fit_morphology,level="unit",part="unique",link_residual="none")$s
parts <- data.frame(trait=rep(traits_morph,4),
 component=rep(c("phylo shared","phylo diagonal","ordinary shared","ordinary diagonal"),each=6),
 variance=c(diag(Sps),pp,diag(Sns),pn))
p1 <- corplot(Sphy,"Fitted phylogenetic trait correlation")
p2 <- corplot(Snon,"Fitted ordinary trait correlation")
parts$trait <- factor(parts$trait, levels = traits_morph)
p3 <- ggplot(parts,aes(trait,variance,fill=component))+geom_col()+
  scale_fill_manual(values = c("phylo shared" = "#156b42", "phylo diagonal" = "#a6dba0",
                              "ordinary shared" = "#70569b", "ordinary diagonal" = "#c5afd9"))+
  labs(title="Fitted four-part variance decomposition",x=NULL,y="Variance")+
  theme_minimal(base_size=10)+theme(axis.text.x=element_text(angle=45,hjust=1))
grid::grid.newpage(); grid::pushViewport(grid::viewport(layout=grid::grid.layout(2,2)))
print(p1,vp=grid::viewport(layout.pos.row=1,layout.pos.col=1))
print(p2,vp=grid::viewport(layout.pos.row=1,layout.pos.col=2))
print(p3,vp=grid::viewport(layout.pos.row=2,layout.pos.col=1:2))
```

```{r morphology-truth-estimate}
truth_estimate <- data.frame(
  trait = traits_morph,
  phylo_total_truth = diag(truth_morphology$Sigma_phy_total),
  phylo_total_fitted = diag(Sphy),
  ordinary_total_truth = diag(truth_morphology$Sigma_non_total),
  ordinary_total_fitted = diag(Snon),
  phylo_communality_truth = diag(truth_morphology$Sigma_phy_shared) / diag(truth_morphology$Sigma_phy_total),
  phylo_communality_fitted = extract_communality(fit_morphology, level = "phy", link_residual = "none"),
  ordinary_communality_truth = diag(truth_morphology$Sigma_non_shared) / diag(truth_morphology$Sigma_non_total),
  ordinary_communality_fitted = extract_communality(fit_morphology, level = "unit", link_residual = "none")
)
knitr::kable(truth_estimate, digits = 3,
  caption = "One synthetic realization: truth and fitted covariance summaries, not recovery evidence.")
```

The table describes one synthetic realization; it is not recovery evidence.
Communality is the shared fraction of variance **within one source**, not that
source's share of total variance. The four-part bars show the latter components
on the original variance scale. The phylogenetic diagonal estimates approach
zero in this realization even though their planted values are positive; the
stable fit does not establish recovery of that decomposition.

The ordinary component is between-species variation not structured by this
tree. It is not automatically plot-level variation, convergence,
microevolution, or measurement error. A near-zero fitted diagonal for a
pre-specified source is reported, not used to alter the frozen model.

# Example 2: species coefficients and residual association at sites

We simulate 150 sites and 50 species. C3 and C4 fixed pathway intercepts are
.85 and .45; their latitude slopes are -.30 and .55. Species have intercept
and slope deviations around those expectations. Both coefficient models retain
a two-dimensional ordinary site residual component. Here $z_i$ denotes
standardized latitude, $C_s$ is species $s$'s pathway, and
$b_s=(b_{0s},b_{1s})^\mathsf T$ contains its coefficient deviations.
The response is a synthetic continuous community measurement, not a count.
The contrasting pathway slopes are design assumptions, not a general claim
about C3 and C4 distributions.

\[
y_{is}=\alpha_{C_s}+\beta_{C_s}z_i+b_{0s}+b_{1s}z_i+e_{is},\qquad
\operatorname{Cov}(e_i)=\Lambda_{\rm site}\Lambda_{\rm site}^{\mathsf T}+\Psi_{\rm site}.
\]

For the phylogenetic coefficient model,
\[
\operatorname{Cov}(b_s,b_t)=(K_\rho)_{st}\Sigma_b,\qquad
K_\rho=\rho K+(1-\rho)\operatorname{diag}\{\operatorname{diag}(K)\}.
\]
We plant $\rho=0.60$ and estimate it with `rho = NULL`. The tree is scaled
exactly as in Example 1, so $K$ has unit diagonal. The coefficient standard
deviations are 0.18 and 0.55, with correlation -0.25. The IID comparator uses
$I$ in place of $K_\rho$, while retaining the same site covariance model.
The coefficient mixture does not describe residual co-occurrence at sites.

```{r community-dgp}
community <- fixture$community$long
community_wide <- fixture$community$wide
column_data <- fixture$community$column_data
tree_columns <- fixture$community$tree
species_comm <- fixture$community$species
truth_community <- fixture$community$truth
```

```{r community-long, eval=FALSE}
fit_columns <- gllvmTMB(
 value ~ 0+pathway+latitude:pathway+column_coef(1+latitude|trait)+
   latent(0+trait|site_id,d=2,unique=TRUE),
 data=community,trait="trait",unit="site_id",family=gaussian(),silent=TRUE,control=gllvmTMBcontrol(se=FALSE)
)
fit_phylo_coef <- gllvmTMB(
 value ~ 0+pathway+latitude:pathway+
   phylo_coef(1+latitude|trait,tree=tree_columns,rho=NULL)+
   latent(0+trait|site_id,d=2,unique=TRUE),
 data=community,trait="trait",unit="site_id",family=gaussian(),silent=TRUE,control=gllvmTMBcontrol(se=FALSE)
)
```

```{r community-wide, eval=FALSE}
fit_columns_wide <- gllvmTMB(
 traits(all_of(species_comm)) ~ 0+pathway+latitude:pathway+
   column_coef(1+latitude|trait)+latent(1|site_id,d=2,unique=TRUE),
 data=community_wide,column_data=column_data,unit="site_id",family=gaussian(),control=gllvmTMBcontrol(se=FALSE)
)
fit_phylo_coef_wide <- gllvmTMB(
 traits(all_of(species_comm)) ~ 0+pathway+latitude:pathway+
   phylo_coef(1+latitude|trait,tree=tree_columns,rho=NULL)+latent(1|site_id,d=2,unique=TRUE),
 data=community_wide,column_data=column_data,unit="site_id",family=gaussian(),control=gllvmTMBcontrol(se=FALSE)
)
```

The target community fits have unresolved convergence warnings. Their formulas
and plotting code are retained in this working draft, but their numerical
results are not rendered or interpreted. The code below uses planted species
curves to illustrate the simulated coefficient deviations. The current public
API provides fitted coefficient covariance, not individual species coefficient
modes; the curves must not be read as estimates of those modes.

```{r community-pathway-coefficients, eval=FALSE, fig.width=8, fig.height=3.6}
fixed <- coef(fit_columns)
curve <- expand.grid(latitude=seq(-2,2,length.out=100), pathway=factor(c("C3","C4"),levels=c("C3","C4")))
X_curve <- model.matrix(~0+pathway+latitude:pathway, data=curve)
curve$mean <- drop(X_curve %*% fixed[colnames(X_curve)])
planted <- expand.grid(latitude = seq(-2, 2, length.out = 100),
                       trait = species_comm, stringsAsFactors = FALSE)
planted$pathway <- column_data$pathway[match(planted$trait, column_data$trait)]
b <- truth_community$coefficient_draw[planted$trait, , drop = FALSE]
planted$mean <- truth_community$pathway_intercept[as.character(planted$pathway)] +
  b[, "(Intercept)"] + planted$latitude *
  (truth_community$pathway_slope[as.character(planted$pathway)] + b[, "latitude"])
p1 <- ggplot() +
  geom_line(data = planted, aes(latitude, mean, group = trait, colour = pathway),
            alpha = 0.18, linewidth = 0.3) +
  geom_line(data = curve, aes(latitude, mean, colour = pathway),
            linewidth = 1.1, linetype = "dashed") +
  labs(title = "Species deviations around pathway expectations",
       x = "Standardized latitude", y = "Synthetic Gaussian mean",
       caption = "Thin: planted species curves. Dashed: fitted IID pathway means.\nOne synthetic realization; species curves are not fitted random effects.") +
  theme_minimal(base_size = 10)
coefphy <- extract_Sigma(fit_phylo_coef,level="column_coef")
p2 <- corplot(coefphy$Sigma,"Fitted phylogenetic coefficient correlation")
grid::grid.newpage(); grid::pushViewport(grid::viewport(layout=grid::grid.layout(1,2)))
print(p1,vp=grid::viewport(layout.pos.row=1,layout.pos.col=1))
print(p2,vp=grid::viewport(layout.pos.row=1,layout.pos.col=2))
```

```{r community-residual-ordination, eval=FALSE, fig.width=8, fig.height=3.6}
Ssite <- extract_Sigma(fit_columns,level="unit",part="total",link_residual="none")$Sigma
p3 <- corplot(Ssite,"Residual correlation among all 50 species")
ord <- extract_ordination(fit_columns,level="unit")
scores <- as.data.frame(ord$scores); names(scores)[1:2] <- c("LV1","LV2")
p4 <- ggplot(scores,aes(LV1,LV2))+geom_point(alpha=.65)+coord_equal()+
 labs(title="Site ordination from ordinary residual fit")+theme_minimal()
grid::grid.newpage(); grid::pushViewport(grid::viewport(layout=grid::grid.layout(1,2)))
print(p3,vp=grid::viewport(layout.pos.row=1,layout.pos.col=1))
print(p4,vp=grid::viewport(layout.pos.row=1,layout.pos.col=2))
```

The residual matrix describes association among species responses at a site,
not correlation between sites. All 50 species are shown in their fixed fixture
order; only selected axis labels are printed for readability.

Residual association can arise from unmeasured environment, sampling, or shared
responses. It is not proof of direct species interactions. Ordination axes can
rotate or change sign and are interpreted within a fit only.

## Gaussian diagonal components and scope

`unique = TRUE` adds a trait-specific diagonal companion to each latent
source. The phylogenetic companion still follows the tree; it is not an
independent nonphylogenetic source. The ordinary companion groups by species
in Example 1 and by site in Example 2. With one value per group and response,
it also absorbs independent cell-level Gaussian variation. Estimating a second
free observation variance would duplicate that diagonal contribution.

The package therefore fixes the separate Gaussian residual standard deviation
to `max(1e-3 * sd(y), 1e-6)` for these fits, as verified during validation.
This small numerical stabilizer is not an estimated biological component or
separately identified measurement error. `control = gllvmTMBcontrol(se = FALSE)`
omits standard-error calculations; it does not itself set that residual scale.

The morphology fit passes the bounded numerical screen. The community
interpretation is withheld until its all-start convergence gate passes. Neither
example makes a recovery, interval, or causal interaction claim.
