---
title: "What does the tree relate?"
description: "Choose whether a phylogeny relates species units or Gaussian response columns before fitting a gradient model."
tier: 1
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What does the tree relate?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.align="center", message=FALSE, warning=FALSE)
```

> **Evidence boundary.** These are simulated Gaussian examples conditional on
> fixed trees. The final section declares source observation models only: it
> does not fit an iSDM or identify abundance, occupancy, or detectability.

## What does the tree relate?

A tree is a relationship matrix, not a button that makes a model
“phylogenetic”. Its tips must match one named axis. In this montane-bird
example the same tree first relates **species units** in a comparative trait
model, then **species response columns** in a community model. Those are
different models, even though both use `tree = tree`.

```{r setup}
library(gllvmTMB); library(ggplot2)
theme_article <- theme_minimal(base_size=12) + theme(panel.grid.minor=element_blank(), panel.grid.major.x=element_blank(), strip.text=element_text(face="bold"), legend.position="bottom")
pal <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7")
bridge_path <- c(file.path("inst","extdata","examples","trait-axis-bridge.rds"), system.file("extdata","examples","trait-axis-bridge.rds",package="gllvmTMB"))
bridge_path <- bridge_path[nzchar(bridge_path) & file.exists(bridge_path)][1]
stopifnot(!is.na(bridge_path)); bridge <- readRDS(bridge_path)
```

```{r front-door, fig.width=10, fig.height=4, fig.cap="Only the first two questions put a tree in the model. A comparative tree joins species units; a community tree joins species response columns. In the third panel, one shared ecology is filtered through source-specific observation models into source-labelled records; no tree is involved.", fig.alt="Three vertical diagrams: tree tips to species units to trait records; tree tips to species response columns to site records; shared ecology to observation models to source-labelled records, without a tree."}
panel_order <- c("Species as units", "Species as columns", "Observation sources")
front <- data.frame(
  panel = factor(rep(panel_order, each = 3), levels = panel_order),
  x = 1, y = rep(c(3, 2, 1), 3),
  label = c("tree tips", "species units", "trait records",
    "tree tips", "species columns", "site records",
    "shared ecology", "observation models", "source-labelled records"),
  kind = c("tree", "data", "data", "tree", "data", "data",
    "data", "source", "source")
)
arrows <- subset(front, y > 1)
ggplot(front, aes(x, y)) +
  geom_segment(data = arrows,
    aes(x = x, y = y - .24, xend = x, yend = y - .76),
    arrow = grid::arrow(length = grid::unit(2, "mm")), colour = "#555555") +
  geom_label(aes(label = label, fill = kind), show.legend = FALSE,
    size = 3.45, label.size = .25) +
  facet_wrap(~ panel, nrow = 1) +
  scale_fill_manual(values = c(tree = "#B8D8E8", data = "#F4D6A0",
    source = "#C9E5C5")) +
  coord_cartesian(xlim = c(.35, 1.65), ylim = c(.65, 3.35), clip = "off") +
  theme_article + theme(axis.text = element_blank(), axis.ticks = element_blank(),
    axis.title = element_blank(), panel.grid = element_blank())
```

## Example A: a PCM tree among species units

Each population record is a row from a species, and each measured trait is a
response column. The fixture has eight measured traits in each of two domains,
`morphology` and `life_history`. Sixteen columns are enough to show the
difference between a domain-average response and variation among the measured
traits without pretending that the columns themselves are phylogenetically
related.

`column_domain:elevation` gives the two domains their mean slopes.
`slope(elevation | trait)` gives measured trait-column deviations around those
means. `phylo_latent()` is instead a **species-axis intercept covariance**
across measured traits. With `d = 2, unique = TRUE`, its trait covariance is
$\Sigma_{\rm trait}=\Lambda\Lambda^\mathsf{T}+\operatorname{diag}(\psi)$:
two shared trait axes plus one trait-specific variance per column. It neither
plants nor estimates species-specific elevation slopes.

The ordinary `slope()` term is used because the variation being described is
among **trait columns**, whereas the tree separately describes covariance among
the **species units**. The single bar, `|`, asks for the full covariance among
the named predictor deviations. `||` is the diagonal alternative: it retains
one variance per predictor but fixes their covariance to zero. Here there is
one predictor, elevation, so `|` and `||` have the same single-predictor
meaning; `|` is retained to make the response-column axis explicit.

$$y_{spt}=\mu_t+\beta_{d(t)}e_{sp}+a_te_{sp}+u_{st}+\epsilon_{spt},\quad
a_t\sim N(0,\sigma_a^2),\quad
\operatorname{vec}(U)\sim N(0,\Sigma_{\rm trait}\otimes A_{\rm tree}).$$

Here $a_t$ is the ordinary slope deviation for measured trait column $t$,
while $U=(u_{st})$ is the species-by-trait intercept matrix. Only $U$ carries
the tree covariance $A_{\rm tree}$.

```{r pcm-fit}
pcm <- bridge$pcm
pcm_data <- pcm$data
tree <- pcm$tree

fit_pcm <- gllvmTMB(
  value_z ~ 0 + trait + column_domain:elevation +
    slope(elevation | trait) +
    phylo_latent(0 + trait | species, tree = tree, d = 2, unique = TRUE),
  data = pcm_data, trait = "trait", unit = "record_id", cluster = "species",
  family = gaussian(), silent = TRUE
)
fit_pcm$fit_health[c("convergence","pd_hessian","converged")]
Sigma_pcm_slope <- extract_Sigma(fit_pcm, level = "column_slope")$Sigma
Sigma_pcm_slope
```

The single entry in `Sigma_pcm_slope` is the fitted variance among measured
trait-column elevation deviations. It is not a phylogenetic variance: the tree
enters separately through `phylo_latent(... | species)`.

```{r domain-trait-slopes, fig.width=9, fig.height=5, fig.cap="Planted domain mean slopes and measured-trait deviations in Example A. Thin direct-labelled lines are the sixteen stored trait-column slopes; bold coloured lines are the two stored domain means. This is a design map, not fitted uncertainty or species trajectories.", fig.alt="Two panels, morphology and life history, each with eight labelled grey trait-column lines around one bold coloured domain-mean elevation line."}
truth_a <- data.frame(trait = names(pcm$truth$trait_slope_deviation),
  deviation = unname(pcm$truth$trait_slope_deviation), stringsAsFactors = FALSE)
truth_a$column_domain <- unname(pcm$metadata$column_domain[truth_a$trait])
truth_a$domain_mean <- unname(pcm$truth$domain_mean_slope[truth_a$column_domain])
truth_a$slope <- truth_a$domain_mean + truth_a$deviation
grid_x <- seq(-1, 1, length.out = 60)
line_data <- merge(truth_a, data.frame(elevation = grid_x))
line_data$value_z <- line_data$slope * line_data$elevation
mean_data <- unique(line_data[c("column_domain", "domain_mean", "elevation")])
mean_data$value_z <- mean_data$domain_mean * mean_data$elevation
mean_label <- subset(mean_data, elevation == max(elevation))
mean_label$label <- paste(tools::toTitleCase(gsub("_", " ", mean_label$column_domain)), "mean")
mean_label$label_x <- 1.38
trait_label <- subset(line_data, elevation == max(elevation))
trait_label$label <- gsub("_", " ", trait_label$trait)
ggplot(line_data, aes(elevation, value_z, group = trait)) +
  geom_hline(yintercept = 0, colour = "#777777", linewidth = .3) +
  geom_line(colour = "#777777", linewidth = .55) +
  geom_line(data = mean_data,
    aes(elevation, value_z, colour = column_domain, group = column_domain),
    inherit.aes = FALSE, linewidth = 1.25) +
  geom_text(data = trait_label, aes(label = label), hjust = -.08,
    colour = "#595959", size = 2.45) +
  geom_text(data = mean_label,
    aes(label_x, value_z, label = label, colour = column_domain),
    inherit.aes = FALSE, hjust = 0, vjust = .35, fontface = "bold",
    size = 2.8) +
  facet_wrap(~ column_domain,
    labeller = as_labeller(c(life_history = "Life history", morphology = "Morphology"))) +
  scale_colour_manual(values = c(morphology = pal[[1]], life_history = pal[[2]]),
    guide = "none") +
  coord_cartesian(xlim = c(-1, 1.75), clip = "off") +
  labs(x = "Elevation (standardised)", y = "Planted linear predictor") +
  theme_article + theme(plot.margin = margin(5.5, 54, 5.5, 5.5))
```

The three contributions stay separate: `column_domain:elevation` supplies the
two fixed domain means, `slope(elevation | trait)` supplies measured-column
deviations, and `phylo_latent(... | species)` supplies species-tree intercept
covariance. The tree does not create a species-specific elevation slope here.

For the covariance interpretation, see [Phylogenetic covariance among traits](phylogenetic-gllvm.html).

## Example B: the same tree among response columns

Sites are now units and species are Gaussian response columns. Fixed guild
effects describe guild-average responses; the tree’s labels must match `trait`,
not `site_id`.

```{r column-fit}
column <- bridge$column
fit_column <- gllvmTMB(value ~ 0 + trait + guild:elevation + guild:forest_cover +
  phylo_slope(elevation + forest_cover | trait, tree=column$tree),
  data=column$data, trait="trait", unit="site_id", family=gaussian(), silent=TRUE)
Sigma_slope <- extract_Sigma(fit_column,level="column_slope")$Sigma
Sigma_slope
```

Read the formula from left to right. `0 + trait` estimates one intercept for
each species response column. `guild:elevation` and `guild:forest_cover` give
guild-average slopes. `phylo_slope()` then adds species-column deviations
around those averages, with the tree relating those species columns.

This predictor-only helper has no random intercept, is Gaussian-only, and
currently has **no wide-data equivalent**. Use the long form so `trait` names
the response-column axis. `Sigma_slope` is covariance between elevation and
forest-cover slope deviations, not a trait-level Sigma. See the [joint SDM article](joint-sdm.html).

Here `phylo_slope()` is appropriate because the tree belongs to the same axis
as the slope deviations: the species response columns. With two predictors,
`|` estimates their covariance in `Sigma_slope`; `||` would keep the same
tree-aligned marginal variances but set their covariance to zero. Neither form
supplies an uncertainty interval in this worked display.

Numerically, this fit estimates elevation and forest-cover deviation variances
of `r round(Sigma_slope["elevation", "elevation"], 2)` and
`r round(Sigma_slope["forest_cover", "forest_cover"], 2)`, with covariance
`r round(Sigma_slope["elevation", "forest_cover"], 2)`. Thus both responses
vary appreciably among species columns, but their fitted association is small
in this simulation. These numbers depend on the predictor scaling; elevation
and forest cover were standardised before fitting.

```{r community-tree-slopes, fig.width=11, fig.height=6, fig.cap="Tree-aligned community slopes. Every point is a planted species-column deviation, aligned to the same tip row as the tree; eight representative tips are directly labelled. The adjacent fitted matrix is covariance between the elevation and forest-cover predictor deviations, not covariance among species.", fig.alt="A phylogenetic tree for 48 montane bird response columns, an aligned point display of each species elevation and forest-cover slope deviation, and a labelled two by two fitted predictor-covariance matrix."}
tree_plot <- ape::reorder.phylo(column$tree, order = "postorder")
edge <- tree_plot$edge
node_x <- ape::node.depth.edgelength(tree_plot)
node_y <- rep(NA_real_, length(node_x))
node_y[seq_len(ape::Ntip(tree_plot))] <- seq_len(ape::Ntip(tree_plot))
children <- split(edge[, 2], edge[, 1])
node_height <- function(node) {
  if (!is.na(node_y[[node]])) return(node_y[[node]])
  child_y <- vapply(children[[as.character(node)]], node_height, numeric(1))
  node_y[[node]] <<- mean(range(child_y))
  node_y[[node]]
}
invisible(node_height(ape::Ntip(tree_plot) + 1L))
horizontal <- data.frame(
  x = node_x[edge[, 1]], xend = node_x[edge[, 2]],
  y = node_y[edge[, 2]], yend = node_y[edge[, 2]]
)
child_rows <- split(edge[, 2], edge[, 1])
vertical <- do.call(rbind, lapply(names(child_rows), function(parent) {
  child_y <- node_y[child_rows[[parent]]]
  data.frame(x = node_x[as.integer(parent)],
    xend = node_x[as.integer(parent)], y = min(child_y), yend = max(child_y))
}))
tip_y <- setNames(node_y[seq_along(tree_plot$tip.label)], tree_plot$tip.label)
labelled_tips <- column$tree$tip.label[unique(round(seq(1, length(tip_y), length.out = 8)))]
tip_labels <- data.frame(trait = labelled_tips, y = tip_y[labelled_tips],
  x = max(node_x[seq_along(tip_y)]))

p_tree <- ggplot() +
  geom_segment(data = horizontal, aes(x, y, xend = xend, yend = yend),
    linewidth = .35, colour = "#3A3A3A") +
  geom_segment(data = vertical, aes(x, y, xend = xend, yend = yend),
    linewidth = .35, colour = "#3A3A3A") +
  geom_text(data = tip_labels, aes(x, y, label = trait), hjust = -.05,
    size = 2.35, colour = "#333333") +
  scale_y_continuous(limits = c(.5, length(tip_y) + .5), expand = c(0, 0)) +
  coord_cartesian(clip = "off") + labs(x = "Phylogenetic distance", y = NULL) +
  theme_article + theme(panel.grid = element_blank(), axis.text.y = element_blank(),
    axis.ticks.y = element_blank(), plot.margin = margin(5.5, 38, 5.5, 5.5))

slope_truth <- rbind(
  data.frame(trait = names(column$truth$elevation_slope_deviation),
    predictor = "Elevation", deviation = unname(column$truth$elevation_slope_deviation)),
  data.frame(trait = names(column$truth$forest_cover_slope_deviation),
    predictor = "Forest cover", deviation = unname(column$truth$forest_cover_slope_deviation))
)
slope_truth$y <- tip_y[slope_truth$trait]
p_slopes <- ggplot(slope_truth, aes(deviation, y, fill = predictor,
    shape = predictor)) +
  geom_vline(xintercept = 0, linewidth = .35, colour = "#777777") +
  geom_point(size = 2.25, colour = "#222222", stroke = .35) +
  scale_fill_manual(values = c(Elevation = pal[[1]], `Forest cover` = pal[[2]])) +
  scale_shape_manual(values = c(Elevation = 21, `Forest cover` = 24)) +
  scale_y_continuous(limits = c(.5, length(tip_y) + .5), expand = c(0, 0)) +
  labs(x = "Planted species-column deviation", y = NULL, fill = NULL, shape = NULL,
    title = "Planted deviations") +
  theme_article + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(), legend.position = "top",
    plot.title = element_text(size = 9.5))

z <- as.data.frame(as.table(Sigma_slope)); names(z) <- c("row", "column", "covariance")
z$row <- factor(gsub("_", " ", as.character(z$row)),
  levels = c("forest cover", "elevation"))
z$column <- factor(gsub("_", " ", as.character(z$column)),
  levels = c("elevation", "forest cover"))
p_cov <- ggplot(z, aes(column, row, fill = covariance)) +
  geom_tile(colour = "white", linewidth = .8) +
  geom_text(aes(label = sprintf("%.2f", covariance)), size = 3.2) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, guide = "none") +
  coord_equal() + labs(x = NULL, y = NULL,
    title = "Fitted predictor\ncovariance") +
  theme_article + theme(panel.grid = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(size = 9.5, lineheight = .95))

grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(
  1, 3, widths = grid::unit(c(1.15, 1.25, .85), "null"))))
print(p_tree, vp = grid::viewport(layout.pos.col = 1))
print(p_slopes, vp = grid::viewport(layout.pos.col = 2))
print(p_cov, vp = grid::viewport(layout.pos.col = 3))
```

## Three sources: no tree axis

Sources select observation processes and are not phylogenetic tips. The
ordinary formulas below use automatic reference coding; do not add a `0 +`
workaround. This supports relative intensity only.

```{r isdm}
source_family <- isdm_sources(
  gbif=isdm_source(poisson(link="log"),observation=~access+popdens),
  inat=isdm_source(poisson(link="log"),observation=~access+popdens),
  survey=isdm_source(poisson(link="log"),observation=~observer+method)
)
```

```{r source-observation-panel, fig.width=10, fig.height=4.5, fig.cap="Three sources record the same simulated ecological relative-intensity curve differently. The dashed black curve is shared ecology; open coloured circles are record-specific expectations after each source's observation process, and solid points are simulated Poisson records. Source formulas change recording, not the tree axis.", fig.alt="Three source panels for GBIF, iNaturalist, and survey data. Each repeats one dashed shared ecological curve, with open coloured record-specific expectations and solid observed count points."}
source_plot <- subset(bridge$isdm$data,
  as.character(trait) == levels(bridge$isdm$data$trait)[1])
source_plot <- source_plot[order(source_plot$isdm_source, source_plot$elevation), ]
source_end <- do.call(rbind, lapply(split(source_plot, source_plot$isdm_source),
  function(x) x[which.max(x$elevation), ]))
source_colours <- c(gbif = pal[[1]], inat = pal[[2]], survey = pal[[3]])

ggplot(source_plot, aes(elevation)) +
  geom_line(aes(y = ecological_intensity, group = 1), colour = "#202020",
    linetype = "22", linewidth = .8) +
  geom_point(aes(y = expected_recorded_intensity, colour = isdm_source),
    shape = 1, stroke = 1, size = 2.4) +
  geom_point(aes(y = value, colour = isdm_source), size = 1.8, alpha = .75) +
  geom_text(data = source_end,
    aes(y = ecological_intensity, label = "shared ecology"),
    hjust = -.06, vjust = 1.3, colour = "#202020", size = 2.55) +
  geom_text(data = source_end,
    aes(y = expected_recorded_intensity, label = "record-specific expectation",
      colour = isdm_source), hjust = -.06, vjust = -.5, size = 2.55) +
  facet_wrap(~ isdm_source) +
  scale_colour_manual(values = source_colours, guide = "none") +
  coord_cartesian(xlim = c(min(source_plot$elevation),
    max(source_plot$elevation) + .55), clip = "off") +
  labs(x = "Elevation (standardised)",
    y = "Observed count or expected recorded intensity") +
  theme_article + theme(plot.margin = margin(5.5, 58, 5.5, 5.5))
```

For the full three-source workflow, see [Integrating three data sources at once](integrated-multi-source-example.html). This article makes no full-iSDM or non-Gaussian multi-predictor column-slope claim.
