---
title: "Where does the phylogeny belong?"
description: "Use one phylogeny in two different ways: among sampled plant species or among plant-species response columns."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Where does the phylogeny belong?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggplot2)

theme_article <- theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", colour = "#17324D"),
    plot.title = element_text(face = "bold", colour = "#17324D"),
    plot.caption = element_text(hjust = 0, colour = "#555555"),
    axis.title = element_text(colour = "#17324D")
  )
```

<style>
.table-responsive {
  display: block;
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
</style>

## The decision

Suppose you have a phylogeny for a set of alpine plant species. Before putting
the tree in a model, ask one question: **do its tip labels identify the sampled
species, or the response columns?** The answer changes the model.

<div class="table-responsive">

| Your data question | The tree relates | Use |
|---|---|---|
| Do related species have similar leaf and seed traits? | sampled species | a phylogenetic intercept term |
| Do related species have similar deviations from a C3/C4 latitude gradient? | species response columns | `phylo_coef(... | trait)` |

</div>

The two examples below use simulated Gaussian data and a fixed tree. They show
which variable the tree connects, not uncertainty calibration. Start with the
row whose tree tips match the variable on the right of the model term.

```{r packages, message = FALSE}
library(gllvmTMB)
library(ape)
```

## Example 1: a tree among sampled species

**Question.** How do I include a tree when each sampled plant species has four
morphology traits, each measured across several populations?

Here the response columns are four measured traits. `species` is a grouping
variable: each long-format row records one trait at one population of one
species. Therefore the tree belongs to `species`, not to `trait`.

<div class="table-responsive">

| Long-data variable | Meaning | Does the tree tip match it? |
|---|---|---|
| `plot_id` | one sampled population | no |
| `trait` | leaf area, leaf mass, seed mass, or stem height | no |
| `species` | the plant species sampled at each population | **yes** |

</div>

```{r species-axis-data}
set.seed(24)
species <- sprintf("plant_%02d", 1:12)
traits <- c("leaf_area", "leaf_mass", "seed_mass", "stem_height")
tree <- rcoal(length(species))
tree$tip.label <- species
K_tree <- vcv.phylo(tree, corr = TRUE)[species, species]
A <- K_tree + diag(1e-8, length(species))

populations <- expand.grid(
  species = species, population = 1:5, KEEP.OUT.ATTRS = FALSE
)
populations$plot_id <- sprintf("%s_%s", populations$species, populations$population)
populations$elevation <- rep(seq(-1, 1, length.out = 5), each = length(species))
species_by_trait <- t(chol(A)) %*% matrix(rnorm(length(species) * length(traits)),
  nrow = length(species), dimnames = list(species, traits)
)

morphology <- do.call(rbind, lapply(traits, function(trait_name) {
  i <- match(populations$species, species)
  data.frame(
    plot_id = populations$plot_id,
    species = factor(populations$species, levels = species),
    elevation = populations$elevation,
    trait = factor(trait_name, levels = traits),
    value = 0.35 * populations$elevation + species_by_trait[i, trait_name] +
      rnorm(nrow(populations), sd = 0.35)
  )
}))
```

```{r species-axis-fit}
fit_species <- gllvmTMB(
  value ~ 0 + trait + trait:elevation +
    phylo_latent(0 + trait | species, tree = tree, d = 1, unique = TRUE),
  data = morphology, trait = "trait", unit = "plot_id", family = gaussian(),
  silent = TRUE
)
Sigma_species_traits <- extract_Sigma(fit_species, level = "phy", part = "total")$Sigma
round(Sigma_species_traits, 2)
```

`trait:elevation` gives every measured trait its own average elevation effect.
`phylo_latent(... | species)` then models covariance among the **sampled plant
species**. Its extracted matrix describes which morphology traits tend to vary
together across phylogenetically related species: positive diagonal entries are
trait-specific variation, while off-diagonal entries are cross-trait
association. It is not a model of species-specific elevation slopes.

```{r morphology-data, echo = FALSE, fig.width = 8.4, fig.height = 6.8, fig.cap = "Example 1 in the observed long-data layout. Grey points are five population measurements within each sampled species; amber points are species means. The four panels are the morphology response columns, while the tree belongs to the sampled-species axis.", fig.alt = "Four panels showing leaf area, leaf mass, seed mass, and stem height. Each panel has five pale grey population observations for each of twelve plant species and one amber species mean."}
morphology_plot <- transform(
  morphology,
  trait_label = factor(
    tools::toTitleCase(gsub("_", " ", trait)),
    levels = tools::toTitleCase(gsub("_", " ", traits))
  )
)

ggplot(morphology_plot, aes(x = species, y = value)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "#AAB7C4") +
  geom_point(colour = "#7A8FA0", alpha = 0.55, size = 1.7) +
  stat_summary(
    fun = mean, geom = "point", shape = 21, size = 3.3, stroke = 0.7,
    fill = "#E9A34F", colour = "#173B5E"
  ) +
  facet_wrap(~ trait_label, ncol = 2, scales = "free_y") +
  labs(
    title = "Morphology is observed repeatedly within sampled species",
    subtitle = "Grey points: five populations per species;\namber circles: species means",
    x = "Sampled plant species", y = "Simulated morphology value"
  ) +
  theme_article +
  theme(
    plot.title = element_text(face = "bold", colour = "#173B5E"),
    plot.subtitle = element_text(colour = "#4B5563"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold", colour = "#173B5E")
  )
```

```{r morphology-wide, echo = FALSE, fig.width = 8.4, fig.height = 6.5, fig.cap = "The same Example 1 data after a wide-data summary: each row is a sampled species and each morphology trait is a response column. Cells show standardised species means across the five population measurements; the plot is a data-layout guide, not the fitted model input.", fig.alt = "A twelve by four heatmap. Rows are sampled plant species and columns are leaf area, leaf mass, seed mass, and stem height. Each labelled tile gives a standardised species mean."}
morphology_means <- aggregate(value ~ species + trait, data = morphology, FUN = mean)
morphology_wide <- reshape(
  morphology_means, idvar = "species", timevar = "trait", direction = "wide"
)
morphology_wide <- morphology_wide[, c("species", paste0("value.", traits))]
names(morphology_wide)[-1] <- traits
morphology_standardised <- as.data.frame(scale(morphology_wide[-1]))
morphology_standardised$species <- morphology_wide$species
morphology_wide_plot <- reshape(
  morphology_standardised, varying = traits, v.names = "value",
  timevar = "trait", times = traits, direction = "long"
)
morphology_wide_plot$trait <- factor(
  morphology_wide_plot$trait, levels = traits,
  labels = tools::toTitleCase(gsub("_", " ", traits))
)
morphology_wide_plot$species <- factor(morphology_wide_plot$species, levels = rev(species))
morphology_wide_plot$text_colour <- ifelse(abs(morphology_wide_plot$value) >= 1.45, "white", "#102A43")

ggplot(morphology_wide_plot, aes(x = trait, y = species, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(
    aes(label = sprintf("%.1f", value), colour = text_colour),
    size = 3.1, show.legend = FALSE
  ) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
    name = "Standardised\nspecies mean"
  ) +
  scale_colour_identity() +
  labs(
    title = "Morphology data in wide form",
    subtitle = "Each row is one sampled plant species;\nthe four trait measurements are response columns",
    x = "Response columns: morphology traits", y = "Rows: sampled plant species"
  ) +
  theme_article +
  theme(
    plot.title = element_text(face = "bold", colour = "#173B5E"),
    plot.subtitle = element_text(colour = "#4B5563"),
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )
```

Use this form when the data table contains species as repeated sampling units.
If instead species name the response columns, use the next model.

## Example 2: a tree among species response columns

**Question.** Do C3 and C4 plants differ in both average biomass and their
average response to latitude, and how much do individual plant species vary
around those pathway averages? Do related species have similar deviations?

Each site is now a row of the community matrix and each plant species is a
response column. To stay inside the currently tested scope, the response is a
continuous Gaussian intensity: a simulated log-biomass index. Presence–absence
and count responses need a future non-Gaussian coefficient extension.

The example separates four different jobs. C3 and C4 are photosynthetic
pathways; contrasting their latitude responses is a simple way to ask whether
two ecological strategies follow different environmental gradients.
`latitude` changes among sites. `pathway` is species-column metadata copied
down the long data for each species. The package calls its response-column
identifier `trait`; here each `trait` level is a plant species. The fixed
`0 + pathway` term estimates separate C3 and C4 mean intercepts, while
`latitude:pathway` estimates separate pathway mean gradients.
`phylo_coef(1 + latitude | trait)` then gives every plant species its own
random intercept and latitude slope around the corresponding pathway means.
Related species receive correlated deviations, and the default `rho = NULL`
estimates how strongly both coefficient deviations follow the supplied tree.
The companion `column_coef()` fit is the IID baseline: it estimates the same
random-intercept and random-slope covariance without a relationship among
plant species.

These are different model components. `pathway` and `latitude:pathway` are
**fixed effects** for the two pathway averages;
`phylo_coef(1 + latitude | trait)` is a **random-effect distribution** for the
remaining species-specific intercept and slope deviations. Response-column
metadata such as `pathway` enters the fixed-effects formula directly: pathway
is observed species information, not a site-level random coefficient. `rho`
changes the correlation among species coefficients while preserving the
marginal scale of the supplied tree covariance.

For plant species (j) in pathway (g(j)), the fitted mean is

\[
E(y_{ij} \mid a_j,b_j) = \{\alpha_{g(j)} + a_j\}
  + \{\beta_{g(j)} + b_j\}\,\mathrm{latitude}_i.
\]

The fixed coefficients \(\alpha_{g(j)}\) and \(\beta_{g(j)}\) are the C3 or C4
pathway averages. The random coefficients \(a_j\) and \(b_j\) are the plant
species deviations around those averages. `column_coef()` treats deviations
from different species as unrelated; `phylo_coef()` gives their coefficient
matrix the phylogenetic row covariance \(K_\rho\).

<div class="table-responsive">

| Long-data variable | Meaning | Does the tree tip match it? |
|---|---|---|
| `site_id` | one vegetation plot | no |
| `trait` | the plant species response column | **yes** |
| `latitude` | a site-level predictor, repeated for every species at that site | no |
| `pathway` | a C3/C4 attribute of each species column | no — it groups tips, but does not identify them |

</div>

```{r column-axis-data}
set.seed(25)
sites <- data.frame(
  site_id = sprintf("site_%02d", 1:50),
  latitude = as.numeric(scale(seq(38, 62, length.out = 50) + rnorm(50, sd = 0.7)))
)
species_info <- data.frame(
  trait = factor(species, levels = species),
  pathway = factor(rep(c("C3", "C4"), each = 6), levels = c("C3", "C4"))
)
pathway_intercept <- c(C3 = 0.85, C4 = 0.45)
pathway_slope <- c(C3 = -0.30, C4 = 0.55)
set.seed(8) # a reproducible interior-rho phylogenetic coefficient draw
rho_truth <- 0.60
A_rho <- rho_truth * K_tree + (1 - rho_truth) * diag(diag(K_tree))
coefficient_sd <- c(intercept = 0.18, latitude = 0.55)
coefficient_cor <- -0.25
Sigma_coef_truth <- outer(coefficient_sd, coefficient_sd) *
  matrix(c(1, coefficient_cor, coefficient_cor, 1), nrow = 2)
coefficient_deviation <- t(chol(A_rho)) %*%
  matrix(rnorm(length(species) * 2), nrow = length(species)) %*%
  chol(Sigma_coef_truth)
colnames(coefficient_deviation) <- c("intercept", "latitude")
species_intercept <- pathway_intercept[as.character(species_info$pathway)] +
  coefficient_deviation[, "intercept"]
species_slope <- pathway_slope[as.character(species_info$pathway)] +
  coefficient_deviation[, "latitude"]

community <- do.call(rbind, lapply(seq_along(species), function(j) {
  pathway_j <- as.character(species_info$pathway[[j]])
  data.frame(
    site_id = sites$site_id,
    latitude = sites$latitude,
    trait = factor(species[[j]], levels = species),
    pathway = factor(pathway_j, levels = c("C3", "C4")),
    log_biomass = species_intercept[[j]] +
      species_slope[[j]] * sites$latitude +
      rnorm(nrow(sites), sd = 0.12)
  )
}))

head(species_info, 4)
head(community[c("site_id", "trait", "pathway", "latitude", "log_biomass")], 6)
```

```{r column-axis-fit}
fit_columns_iid <- gllvmTMB(
  log_biomass ~ 0 + pathway + latitude:pathway +
    column_coef(1 + latitude | trait),
  data = community, trait = "trait", unit = "site_id", family = gaussian(),
  control = gllvmTMBcontrol(se = FALSE), silent = TRUE
)
fit_columns <- gllvmTMB(
  log_biomass ~ 0 + pathway + latitude:pathway +
    phylo_coef(1 + latitude | trait, tree = tree, rho = NULL),
  data = community, trait = "trait", unit = "site_id", family = gaussian(),
  control = gllvmTMBcontrol(se = FALSE), silent = TRUE
)
fixed_pathway_intercepts <- coef(fit_columns)[
  names(coef(fit_columns)) %in% c("pathwayC3", "pathwayC4")
]
fixed_pathway_slopes <- coef(fit_columns)[
  grepl("^pathway.+:latitude$", names(coef(fit_columns)))
]
intercept_difference <- unname(
  fixed_pathway_intercepts[["pathwayC4"]] -
    fixed_pathway_intercepts[["pathwayC3"]]
)
slope_difference <- unname(
  fixed_pathway_slopes[["pathwayC4:latitude"]] -
    fixed_pathway_slopes[["pathwayC3:latitude"]]
)
round(c(
  fixed_pathway_intercepts,
  `C4 minus C3 intercept` = intercept_difference,
  fixed_pathway_slopes,
  `C4 minus C3 slope` = slope_difference
), 2)

coef_iid <- extract_Sigma(fit_columns_iid, level = "column_coef")
coef_phylo <- extract_Sigma(fit_columns, level = "column_coef")
round(c(
  `IID intercept SD` = sqrt(coef_iid$Sigma["(Intercept)", "(Intercept)"]),
  `IID slope SD` = sqrt(coef_iid$Sigma["latitude", "latitude"]),
  `phylogenetic intercept SD` =
    sqrt(coef_phylo$Sigma["(Intercept)", "(Intercept)"]),
  `phylogenetic slope SD` = sqrt(coef_phylo$Sigma["latitude", "latitude"]),
  `intercept-slope correlation` =
    coef_phylo$R["(Intercept)", "latitude"],
  `estimated rho` = coef_phylo$rho
), 2)
```

The first block reports the C3 and C4 average intercepts and latitude slopes.
The two `C4 minus C3` values describe the fitted pathway differences in this
simulated sample; this article does not calculate tests or intervals for those
contrasts. The extracted standard deviations describe the remaining
between-species variation in baseline biomass and latitude response after the
pathway means have been fitted. The intercept-slope correlation asks whether
species with higher-than-expected baseline biomass also tend to have steeper
or shallower latitude responses. The IID baseline treats species deviations
as unrelated. The phylogenetic fit estimates `rho` between zero and one:
values nearer zero remove most between-species correlation, while values
nearer one follow the supplied tree more strongly. These are point estimates;
this article does not provide an interval or test for `rho`, the pathway
contrasts, or the coefficient covariance.
The simulation planted `rho = 0.60`; one finite sample need not return exactly
that value.

The example deliberately fits no second site-level covariance term. That keeps
the model aligned with its question: the only structured random component is
the variation in intercept and latitude coefficients across species columns.
Add other covariance tiers only when the sampling design and scientific question need
them, then re-check whether `rho` remains identifiable.

```{r community-data, echo = FALSE, fig.width = 8.4, fig.height = 5.8, fig.cap = "Example 2 for six selected species. Pale points are observed site-by-species log-biomass values. Thin lines are the planted species-specific latitude responses; thick lines are the C3 and C4 pathway means. The phylogenetic coefficient term models the deviations of thin lines from their pathway mean.", fig.alt = "Two panels labelled C3 and C4. Each contains site-by-species observations, thin species-specific trend lines, and one thick pathway-average line. The C3 mean decreases with latitude and the C4 mean increases."}
focal_species <- c("plant_01", "plant_03", "plant_05", "plant_07", "plant_09", "plant_11")
community_focus <- subset(community, trait %in% focal_species)
community_focus$trait <- factor(as.character(community_focus$trait), levels = focal_species)
species_effects <- data.frame(
  trait = factor(focal_species, levels = focal_species),
  pathway = species_info$pathway[match(focal_species, species)],
  intercept = species_intercept[match(focal_species, species)],
  latitude_slope = species_slope[match(focal_species, species)]
)
latitude_grid <- seq(min(sites$latitude), max(sites$latitude), length.out = 150)
community_lines <- do.call(rbind, lapply(seq_len(nrow(species_effects)), function(i) {
  data.frame(
    trait = species_effects$trait[i], pathway = species_effects$pathway[i],
    latitude = latitude_grid,
    log_biomass = species_effects$intercept[i] +
      species_effects$latitude_slope[i] * latitude_grid
  )
}))
pathway_lines <- do.call(rbind, lapply(levels(species_info$pathway), function(p) {
  data.frame(
    pathway = factor(p, levels = levels(species_info$pathway)),
    latitude = latitude_grid,
    log_biomass = pathway_intercept[[p]] +
      pathway_slope[[p]] * latitude_grid
  )
}))
pathway_colours <- c(C3 = "#0072B2", C4 = "#D55E00")

ggplot(community_focus, aes(x = latitude, y = log_biomass)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "#C8CDD2") +
  geom_point(colour = "#7A8FA0", alpha = 0.30, size = 1.25) +
  geom_line(
    data = community_lines, aes(group = trait),
    colour = "#667788", linewidth = 0.55, alpha = 0.75
  ) +
  geom_line(
    data = pathway_lines, aes(colour = pathway),
    linewidth = 1.6, show.legend = FALSE
  ) +
  facet_wrap(~ pathway, nrow = 1) +
  scale_colour_manual(values = pathway_colours) +
  labs(
    title = "The simulated C3 and C4 gradients differ by design",
    subtitle = "Thin lines: species-specific responses;\nthick lines: pathway means",
    x = "Latitude (centred and scaled)", y = "Simulated log-biomass index"
  ) +
  theme_article +
  theme(
    plot.title = element_text(face = "bold", colour = "#173B5E"),
    plot.subtitle = element_text(colour = "#4B5563"),
    panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
    legend.position = "none"
  )
```

```{r community-wide, echo = FALSE, fig.width = 8.4, fig.height = 7.2, fig.cap = "The same Example 2 data in wide form for eight selected species. Each row is a site, latitude is the row-level predictor, and the remaining columns are species responses. C3/C4 pathway labels belong to species columns, not to sites. Values are standardised within columns only to reveal the data layout.", fig.alt = "A heatmap with sites as rows and nine columns. The first column is latitude; the other eight are plant species labelled C3 or C4. Sites are ordered from low to high latitude."}
wide_species <- c("plant_01", "plant_02", "plant_04", "plant_06", "plant_07", "plant_08", "plant_10", "plant_12")
community_wide_source <- subset(community, trait %in% wide_species)
community_wide <- reshape(
  community_wide_source[, c("site_id", "latitude", "trait", "log_biomass")],
  idvar = c("site_id", "latitude"), timevar = "trait", direction = "wide"
)
community_wide <- community_wide[order(community_wide$latitude), ]
community_columns <- c("latitude", paste0("log_biomass.", wide_species))
community_standardised <- as.data.frame(scale(community_wide[community_columns]))
names(community_standardised) <- c("latitude", wide_species)
community_standardised$site_id <- factor(community_wide$site_id, levels = rev(community_wide$site_id))
community_wide_plot <- reshape(
  community_standardised,
  varying = c("latitude", wide_species), v.names = "value",
  timevar = "column", times = c("latitude", wide_species), direction = "long"
)
wide_pathway <- as.character(species_info$pathway[match(wide_species, species)])
community_wide_plot$column <- factor(
  community_wide_plot$column,
  levels = c("latitude", wide_species),
  labels = c("latitude\n(site predictor)", paste0(wide_species, "\n", wide_pathway))
)

ggplot(community_wide_plot, aes(x = column, y = site_id, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.25) +
  geom_vline(xintercept = 1.5, colour = "#173B5E", linewidth = 0.8) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
    name = "Standardised\nvalue"
  ) +
  labs(
    title = "Sites are rows; plant species are response columns",
    subtitle = "Rows run from low to high latitude;\nC3/C4 is metadata attached to species columns",
    x = "Columns", y = "Rows: sites"
  ) +
  theme_article +
  theme(
    plot.title = element_text(face = "bold", colour = "#173B5E"),
    plot.subtitle = element_text(colour = "#4B5563"),
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_blank(), axis.ticks.y = element_blank(),
    legend.title = element_text(face = "bold")
  )
```

The same hierarchical model can be fitted directly from a wide data frame.
The `pathway` column in `species_info` supplies the two fixed pathway
intercepts and, through `latitude:pathway`, the two fixed pathway slopes.
The coefficient term supplies the plant-species random intercepts and slopes.
It is therefore the same model as the long formula, not a slope-only
approximation.

```{r column-axis-wide-fit}
community_wide_fit <- reshape(
  community[, c("site_id", "latitude", "trait", "log_biomass")],
  idvar = c("site_id", "latitude"), timevar = "trait", direction = "wide"
)
names(community_wide_fit) <- sub("^log_biomass\\.", "", names(community_wide_fit))

fit_columns_wide <- gllvmTMB(
  traits(
    plant_01, plant_02, plant_03, plant_04, plant_05, plant_06,
    plant_07, plant_08, plant_09, plant_10, plant_11, plant_12
  ) ~ 0 + pathway + latitude:pathway +
    phylo_coef(1 + latitude | trait, tree = tree, rho = NULL),
  data = community_wide_fit,
  column_data = species_info,
  unit = "site_id", family = gaussian(),
  control = gllvmTMBcontrol(se = FALSE), silent = TRUE
)
round(extract_Sigma(fit_columns_wide, level = "column_coef")$rho, 2)
```

The fixed intercepts above vary only between C3 and C4, whereas the random
intercepts vary among plant species. Replacing `0 + pathway` with one fixed
intercept per plant species would duplicate the random-intercept space, so the
overlap guard rejects that unidentified model.

With several site-level predictors, `|` estimates their full coefficient
covariance and `||` fixes their covariance to zero while retaining one variance
per predictor. This article demonstrates `column_coef()` and `phylo_coef()` for
Gaussian point models in long and `traits(...)` wide form. The parallel
`animal_coef()` fixed-strength and dense-`kernel_coef()` fixed/estimated-strength
point routes are public but are not worked examples here; the API keyword grid
shows their syntax. Non-Gaussian or mixed-family coefficients, latent
coefficient covariance, simultaneous coefficient sources, spatial coefficient
helpers, and intervals are deliberately deferred. Current `*_slope()` helpers
remain available and warning-free.

One numerical compatibility seam is deliberate. A no-intercept dense-`vcv`
call with `rho = 1` dispatches through the released `phylo_slope()` engine and
therefore uses its `K + 1e-8 I` conditioning. Tree sources use the released
sparse precision; interior or estimated `rho` and intercept-bearing endpoint
fits use the raw covariance-scale mixture. This keeps existing slope fits
exactly unchanged.

## Choose the axis before the helper

The name after the bar is the practical check:

```r
# Tree among sampled species: species is the grouping axis
phylo_latent(0 + trait | species, tree = tree, d = 1, unique = TRUE)

# Tree among response columns: latitude varies by site,
# while its intercept and latitude coefficient vary by species column
phylo_coef(1 + latitude | trait, tree = tree, rho = NULL)
```

If your scientific question is about trait covariance among sampled species,
begin with the first form. If it is about environmental-response coefficients
varying across species columns, begin with the second. A species attribute such
as C3/C4 pathway belongs in a fixed interaction such as `latitude:pathway`, not
inside `phylo_coef()`. Thus the ordinary formula supplies the fixed group
contrast. For this model, pathway metadata enters fixed terms such as
`0 + pathway + latitude:pathway`, while `column_coef()` or `phylo_coef()`
supplies random response-column deviations around those pathway means. For the source ×
trait-covariance lookup, see the
[Formula keyword grid](api-keyword-grid.html).
