---
title: "What does the tree relate?"
description: "Choose the axis of a phylogeny before fitting: species units in a comparative model, Gaussian response columns in a community-gradient model, or observation sources in a compact iSDM declaration."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{What does the tree relate?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  fig.align = "center",
  cache = FALSE,
  message = FALSE,
  warning = FALSE
)
```

> **Evidence boundary.** This is one simulated Gaussian teaching example.
> The comparative and response-column fits use the supplied, fixed trees.
> The iSDM section declares observation models only; it does not fit an iSDM
> or identify abundance, occupancy, or detectability.

## What does the tree relate?

A phylogeny is not a switch that makes a model ``phylogenetic''. It is a
relationship matrix whose rows and columns must match one particular axis of
the data. Before choosing a keyword, ask one concrete question: **which
labels are the tree tips?**

This article follows a simulated montane-bird study through three superficially
similar uses of a tree. First, the tips are the species *units* in a
comparative model: each species has repeated observations along elevation.
Second, the tips are the Gaussian *response columns* in a community-gradient
model: the tree relates the bird columns while sites supply observations.
Finally, three observation sources enter a compact iSDM declaration. They are
not a phylogenetic axis at all.

```{r setup}
library(gllvmTMB)
library(ggplot2)
```

```{r figure-theme, include = FALSE}
theme_article <- theme_minimal(base_size = 12, base_family = "sans") +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title.position = "plot",
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )
bird_palette <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7")
```

## One prepared teaching dataset

The article begins with an installed fixture, rather than asking you to rebuild
the simulation. It contains separate data frames for the three questions and
their fixed trees. The data are simulated to make the axes visible; they are
not evidence about a real bird assemblage.

```{r data}
bridge_path <- system.file(
  "extdata", "examples", "trait-axis-bridge.rds", package = "gllvmTMB"
)
bridge_path <- c(
  file.path("inst", "extdata", "examples", "trait-axis-bridge.rds"),
  file.path("..", "inst", "extdata", "examples", "trait-axis-bridge.rds"),
  file.path("..", "..", "inst", "extdata", "examples", "trait-axis-bridge.rds"),
  bridge_path
)
bridge_path <- bridge_path[nzchar(bridge_path) & file.exists(bridge_path)][1]
stopifnot(!is.na(bridge_path))
bridge <- readRDS(bridge_path)

pcm <- bridge$pcm
column <- bridge$column
isdm <- bridge$isdm
```

```{r axis-map, fig.width=10, fig.height=3.6, fig.cap="The front-door decision: only the first two questions put a tree in the model. In a comparative model the tree joins species units; in a community-gradient model it joins species response columns; source labels instead select observation processes.", fig.alt="A three-panel axis map. Comparative: a species tree is linked to species units, then to repeated elevation records. Community: a species tree is linked to bird response columns, then to site records. iSDM: GBIF, iNaturalist, and survey labels select observation models that feed one ecological process; no tree connects the sources."}
axis_nodes <- data.frame(
  panel = rep(c("Comparative PCM", "Gaussian community", "Three-source iSDM"), each = 3),
  x = rep(c(1, 3, 5), 3),
  y = 2,
  label = c(
    "tree tips", "species units", "repeated\nelevation records",
    "tree tips", "bird response\ncolumns", "site records",
    "source labels", "observation\nmodels", "shared\necology"
  ),
  kind = c("tree", "data", "data", "tree", "data", "data", "source", "source", "data")
)
axis_edges <- data.frame(
  panel = rep(c("Comparative PCM", "Gaussian community", "Three-source iSDM"), each = 2),
  x = c(1.45, 3.45, 1.45, 3.45, 1.45, 3.45),
  xend = c(2.55, 4.55, 2.55, 4.55, 2.55, 4.55),
  y = 2,
  yend = 2
)

ggplot(axis_nodes, aes(x, y)) +
  geom_segment(
    data = axis_edges,
    aes(x = x, y = y, xend = xend, yend = yend),
    inherit.aes = FALSE, linewidth = 0.8, colour = "#4D4D4D",
    arrow = grid::arrow(length = grid::unit(2, "mm"))
  ) +
  geom_label(aes(label = label, fill = kind), colour = "#1A1A1A", size = 3.2,
             label.size = 0.25, show.legend = FALSE) +
  facet_wrap(~ panel) +
  scale_fill_manual(values = c(tree = "#B8D8E8", data = "#F4D6A0", source = "#C9E5C5")) +
  coord_cartesian(xlim = c(0.35, 5.65), ylim = c(1.2, 2.8), clip = "off") +
  labs(x = NULL, y = NULL) +
  theme_article +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())
```

## A PCM tree among species units

Suppose field teams repeatedly measure four morphological and life-history
traits for each
simulated montane bird species across an elevational gradient. Here a tree tip
is a **species unit**. The repeated rows for a species do not become tips: they
are observations of that species at an elevation and in a domain.

The fixed `column_domain:elevation` term gives each domain its own mean
elevation trend. `phylo_indep()` gives each domain a separate phylogenetic
intercept variance. The final term is a different object: it gives every
species one elevation-slope deviation, correlated across species by `tree`.

$$
b_s \sim \mathcal N(0,\, \sigma^2_{\mathrm{slope}} A_{\mathrm{tree}}).
$$

There is one shared phylogenetic elevation-slope variance
\(\sigma^2_{\mathrm{slope}}\), not four domain-specific intercept--slope
covariance blocks. If your scientific question needs those per-domain random
covariances, this compact helper is the wrong model and the formula must be
expanded deliberately.

```{r pcm-fit}
pcm_data <- pcm$data
tree <- pcm$tree

fit_pcm <- gllvmTMB(
  value ~ 0 + trait + column_domain:elevation +
    phylo_indep(0 + trait | species, tree = tree) +
    phylo_slope(elevation | species, tree = tree),
  data = pcm_data,
  trait = "trait",
  unit = "record_id",
  cluster = "species",
  family = gaussian(),
  silent = TRUE
)
fit_pcm$fit_health[c("convergence", "pd_hessian", "converged")]
```

```{r pcm-trajectories, fig.width=10, fig.height=3.7, fig.cap="Simulated montane-bird records by elevation. Each small multiple is one response column; thin lines join repeated records for a species and the heavier line is its column-domain mean. The fitted model adds one phylogenetically structured elevation-slope deviation per species, shared across the four columns.", fig.alt="Four small-multiple line plots, one each for body mass, bill length, clutch size, and laying date. In each plot, faint species trajectories run across elevation and a coloured average trajectory summarizes the corresponding morphology or life-history domain. The horizontal elevation scale is shared."}
pcm_plot_data <- pcm_data
pcm_plot_data$value_z <- ave(
  pcm_plot_data$value, pcm_plot_data$trait,
  FUN = function(x) as.numeric(scale(x))
)
pcm_plot_data$trait_label <- paste(
  gsub("_", " ", pcm_plot_data$trait),
  "—", gsub("_", " ", pcm_plot_data$column_domain)
)

ggplot(pcm_plot_data, aes(elevation, value_z, group = interaction(species, trait))) +
  geom_line(alpha = 0.22, colour = "#5C5C5C", linewidth = 0.45) +
  geom_smooth(aes(colour = column_domain, group = 1), method = "lm", se = FALSE,
              linewidth = 1.1, show.legend = FALSE) +
  facet_wrap(~ trait_label, nrow = 1) +
  scale_colour_manual(values = c(morphology = "#0072B2", life_history = "#D55E00")) +
  labs(x = "Elevation (standardised)", y = "Response (within-column SD)") +
  theme_article
```

The `cluster = "species"` argument identifies the ordinary species-level
grouping in this example; the tree-aware terms also state their own RHS
species axis. Keeping the RHS visible is useful: it tells the next reader
exactly what the tree labels must match.

## A Gaussian tree among community response columns

Now change the question. At each montane site we measure a Gaussian response
for several bird species and record elevation and forest cover. The tree tips
now label the **species response columns**—the levels of `trait`—rather than
the units at which observations were repeatedly made.

The covered response-column model is:

```{r column-fit}
column_data <- column$data
tree <- column$tree

fit_column <- gllvmTMB(
  value ~ 0 + trait +
    phylo_slope(elevation + forest_cover | trait, tree = tree),
  data = column_data,
  trait = "trait",
  unit = "site_id",
  family = gaussian(),
  silent = TRUE
)

Sigma_slope <- extract_Sigma(fit_column, level = "column_slope")$Sigma
Sigma_slope
```

`phylo_slope(elevation + forest_cover | trait, tree = tree)` is a
predictor-only response-column route. It has no random intercept and is
Gaussian only. The returned `Sigma_slope` is the covariance of the two
predictor slopes, not a trait-level `Sigma`: it asks whether elevation and
forest-cover deviations co-vary while the tree relates the bird columns.

```{r community-slopes, fig.width=10, fig.height=4.4, fig.cap="The same simulated community data viewed two ways: species-specific Gaussian response trends across elevation (left) and the fitted predictor-basis covariance returned by `extract_Sigma(fit_column, level = \"column_slope\")$Sigma` (right). The covariance panel is about elevation and forest-cover slopes, not covariance among bird response columns.", fig.alt="Left panel: coloured species-specific Gaussian response lines across elevation, with points. Right panel: a clearly labelled two by two heatmap for elevation and forest-cover slope covariance, using a diverging scale centred at zero and showing each numeric covariance."}
slope_plot_data <- subset(
  column_data,
  trait %in% levels(column_data$trait)[seq(1L, nlevels(column_data$trait), length.out = 12L)]
)
sigma_plot_data <- as.data.frame(as.table(Sigma_slope), stringsAsFactors = FALSE)
names(sigma_plot_data) <- c("predictor_row", "predictor_column", "covariance")
pretty_predictor <- c(elevation = "Elevation", forest_cover = "Forest cover")
sigma_plot_data$predictor_row <- factor(
  pretty_predictor[as.character(sigma_plot_data$predictor_row)],
  levels = rev(unname(pretty_predictor))
)
sigma_plot_data$predictor_column <- factor(
  pretty_predictor[as.character(sigma_plot_data$predictor_column)],
  levels = unname(pretty_predictor)
)

p_slopes <- ggplot(slope_plot_data, aes(elevation, value, colour = guild)) +
  geom_point(alpha = 0.55, size = 1.4) +
  geom_smooth(aes(group = trait), method = "lm", se = FALSE, linewidth = 0.8) +
  scale_colour_manual(values = bird_palette) +
  labs(x = "Elevation (standardised)", y = "Gaussian response", colour = "Guild") +
  theme_article

p_covariance <- ggplot(sigma_plot_data, aes(predictor_column, predictor_row, fill = covariance)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", covariance)), colour = "#1A1A1A", size = 4) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
                       guide = "none") +
  coord_equal() +
  labs(x = NULL, y = NULL) +
  theme_article +
  theme(panel.grid = element_blank(), axis.text = element_text(size = 10))

grid::grid.newpage()
layout <- grid::grid.layout(1, 2, widths = grid::unit(c(1.7, 0.8), "null"))
grid::pushViewport(grid::viewport(layout = layout))
print(p_slopes, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1))
print(p_covariance, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2))
```

The two panels use only base R's grid system, so the covariance display is
never silently omitted. Do not transfer this syntax to non-Gaussian responses or
claim a combined non-Gaussian, multi-predictor slope model: that capability is
not covered here.

## Three sources are an observation axis, not a tree axis

A final dataset records the same simulated bird community through a portal,
an atlas, and a designed survey. This is an observation problem. The source
label says which observation law applies to a row; it is not an evolutionary
relationship and should not be supplied as a tree grouping.

```{r isdm-declare}
isdm_data <- isdm$data

source_family <- isdm_sources(
  gbif = isdm_source(poisson(link = "log"), observation = ~ access + popdens),
  inat = isdm_source(poisson(link = "log"), observation = ~ access + popdens),
  survey = isdm_source(binomial(link = "cloglog"),
                       observation = ~ observer + method)
)
source_family
```

The declaration intentionally uses ordinary R syntax, `~ observer + method`.
It is not a recipe for a full iSDM fit, and the compact source panel below
does not establish abundance, occupancy, or detectability. It simply keeps
the source-specific observation layer separate from the ecological and
phylogenetic questions above.

```{r isdm-sources, fig.width=9, fig.height=3.7, fig.cap="Three simulated observation sources enter separate observation models. GBIF and iNaturalist rows use Poisson-log laws with access and population density; survey rows use a complementary-log-log detection law with observer and method. The source labels are not phylogenetic tips.", fig.alt="A source-flow diagram. Separate GBIF, iNaturalist, and survey observation-model boxes each point to one shared ecological process. A direct annotation says that no tree is defined on sources."}
source_panel <- data.frame(
  source = c("GBIF", "iNaturalist", "Survey"),
  observation = c("GBIF\nPoisson log\naccess + popdens", "iNaturalist\nPoisson log\naccess + popdens", "Survey\nBinomial cloglog\nobserver + method"),
  x = 1.25,
  y = c(2.5, 1.45, 0.4)
)

ggplot(source_panel, aes(x, y)) +
  geom_label(aes(label = observation, fill = source), size = 4, label.size = 0.3,
             colour = "#1A1A1A", show.legend = FALSE) +
  geom_segment(aes(x = 1.95, y = y, xend = 3.25, yend = 1.45),
               arrow = grid::arrow(length = grid::unit(2, "mm")), linewidth = 0.8,
               colour = "#4D4D4D") +
  annotate("label", x = 4.25, y = 1.45, label = "shared ecological\nprocess", fill = "white", size = 4) +
  annotate("text", x = 2.65, y = 3.05, label = "No tree on sources", colour = "#6A3D9A", fontface = "bold") +
  scale_fill_manual(values = c(GBIF = "#B8D8E8", iNaturalist = "#F4D6A0", Survey = "#C9E5C5")) +
  coord_cartesian(xlim = c(0.35, 5.2), ylim = c(-0.1, 3.4), clip = "off") +
  labs(x = NULL, y = NULL) +
  theme_article +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())
```

## Carry this question into the next model

Before fitting a tree-aware term, write down the answer to “what does the tree
relate?” If it relates species units with replicated records, use the
comparative species axis. If it relates Gaussian response columns, use the
predictor-only column-slope route and interpret `column_slope` covariance in
the predictor basis. If a label names an observation source, declare its
observation law instead: a source is not automatically a phylogenetic axis.

These examples cover the two fitted Gaussian routes shown here and the
three-source declaration. They do not cover a full iSDM analysis, tree
uncertainty, abundance or occupancy inference, or non-Gaussian multi-predictor
response-column slopes.
