---
title: "Where does the phylogeny belong?"
description: "Use one phylogeny in two different ways: among sampled plant species or among plant-species response columns."
tier: 1
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Where does the phylogeny belong?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

<style>
.table-responsive {
  display: block;
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
</style>

## The decision

Suppose you have a phylogeny for a set of alpine plant species. Before putting
the tree in a model, ask one question: **do its tip labels identify the sampled
species, or the response columns?** The answer changes the model.

<div class="table-responsive">

| Your data question | The tree relates | Use |
|---|---|---|
| Do related species have similar leaf and seed traits? | sampled species | a phylogenetic intercept term |
| Do related species show similar responses to moisture and canopy cover? | species response columns | `phylo_slope(... | trait)` |

</div>

The two examples below use simulated Gaussian data and a fixed tree. They show
which variable the tree connects, not uncertainty calibration. Start with the
row whose tree tips match the variable on the right of the model term.

## Example 1: a tree among sampled species

**Question.** How do I include a tree when each sampled plant species has four
morphology traits, each measured across several populations?

Here the response columns are four measured traits. `species` is a grouping
variable: each long-format row records one trait at one population of one
species. Therefore the tree belongs to `species`, not to `trait`.

<div class="table-responsive">

| Long-data variable | Meaning | Does the tree tip match it? |
|---|---|---|
| `plot_id` | one sampled population | no |
| `trait` | leaf area, leaf mass, seed mass, or stem height | no |
| `species` | the plant species sampled at each population | **yes** |

</div>

```{r species-axis-data, include = FALSE}
library(gllvmTMB)
library(ape)

set.seed(24)
species <- sprintf("plant_%02d", 1:12)
traits <- c("leaf_area", "leaf_mass", "seed_mass", "stem_height")
tree <- rcoal(length(species))
tree$tip.label <- species
A <- vcv.phylo(tree, corr = TRUE)[species, species] + diag(1e-8, length(species))

populations <- expand.grid(
  species = species, population = 1:5, KEEP.OUT.ATTRS = FALSE
)
populations$plot_id <- sprintf("%s_%s", populations$species, populations$population)
populations$elevation <- rep(seq(-1, 1, length.out = 5), each = length(species))
species_by_trait <- t(chol(A)) %*% matrix(rnorm(length(species) * length(traits)),
  nrow = length(species), dimnames = list(species, traits)
)

morphology <- do.call(rbind, lapply(traits, function(trait_name) {
  i <- match(populations$species, species)
  data.frame(
    plot_id = populations$plot_id,
    species = factor(populations$species, levels = species),
    elevation = populations$elevation,
    trait = factor(trait_name, levels = traits),
    value = 0.35 * populations$elevation + species_by_trait[i, trait_name] +
      rnorm(nrow(populations), sd = 0.35)
  )
}))
```

```{r species-axis-fit}
fit_species <- gllvmTMB(
  value ~ 0 + trait + trait:elevation +
    phylo_latent(0 + trait | species, tree = tree, d = 1, unique = TRUE),
  data = morphology, trait = "trait", unit = "plot_id", family = gaussian(),
  silent = TRUE
)
Sigma_species_traits <- extract_Sigma(fit_species, level = "phy", part = "total")$Sigma
round(Sigma_species_traits, 2)
```

`trait:elevation` gives every measured trait its own average elevation effect.
`phylo_latent(... | species)` then models covariance among the **sampled plant
species**. Its extracted matrix describes which morphology traits tend to vary
together across phylogenetically related species: positive diagonal entries are
trait-specific variation, while off-diagonal entries are cross-trait
association. It is not a model of species-specific elevation slopes.

Use this form when the data table contains species as repeated sampling units.
If instead species name the response columns, use the next model.

## Example 2: a tree among species response columns

**Question.** At a set of vegetation plots, how do I model tree-aligned
differences in plant-species moisture and canopy-cover responses?

Now each plot is a sampling unit, and each species is one Gaussian response
column. The same tree tips must therefore match `trait`. The phylogeny relates
the **coefficient deviations across species columns**, while the rows remain
independent plots.

<div class="table-responsive">

| Long-data variable | Meaning | Does the tree tip match it? |
|---|---|---|
| `site_id` | one vegetation plot | no |
| `trait` | the plant species response column | **yes** |
| `moisture`, `canopy` | plot-level predictors | no |

</div>

```{r column-axis-data, include = FALSE}
set.seed(25)
sites <- data.frame(
  site_id = sprintf("site_%02d", 1:50),
  moisture = rnorm(50),
  canopy = rnorm(50)
)
moisture_deviation <- drop(t(chol(A)) %*% rnorm(length(species), sd = 0.35))
canopy_deviation <- drop(t(chol(A)) %*% rnorm(length(species), sd = 0.25))

community <- do.call(rbind, lapply(seq_along(species), function(j) {
  data.frame(
    site_id = sites$site_id,
    moisture = sites$moisture,
    canopy = sites$canopy,
    trait = factor(species[[j]], levels = species),
    value = 0.4 + (0.60 + moisture_deviation[[j]]) * sites$moisture +
      (-0.25 + canopy_deviation[[j]]) * sites$canopy + rnorm(nrow(sites), sd = 0.35)
  )
}))
```

```{r column-axis-fit}
fit_columns <- gllvmTMB(
  value ~ 0 + trait + moisture + canopy +
    phylo_slope(moisture + canopy | trait, tree = tree),
  data = community, trait = "trait", unit = "site_id", family = gaussian(),
  silent = TRUE
)
Sigma_predictor <- extract_Sigma(fit_columns, level = "column_slope")$Sigma
round(Sigma_predictor, 2)
```

`phylo_slope()` is a response-column helper, not a fourth covariance mode in
the 5 × 3 grid. The tree supplies the relationship among species columns;
`extract_Sigma(..., level = "column_slope")` returns the two-by-two covariance
of the **moisture and canopy coefficient deviations**. It is neither a
species-by-species covariance matrix nor trait-level `Sigma`.

The two diagonal entries say how much the moisture and canopy responses vary
across species columns. The off-diagonal entry says whether species with a
larger-than-average moisture response also tend to have a larger-than-average
canopy response. It does not describe covariance of the observed responses.

With several predictors, `|` estimates their full covariance and `||` fixes
their covariance to zero while retaining one variance per predictor. This
response-column route is currently Gaussian and long-format only. Wide
column-slope grammar, non-Gaussian or mixed-family slopes, latent predictor
covariance, simultaneous column-slope sources, and intervals are deliberately
deferred.

## Choose the axis before the helper

The name after the bar is the practical check:

```r
# Tree among sampled species: species is the grouping axis
phylo_latent(0 + trait | species, tree = tree, d = 1, unique = TRUE)

# Tree among response columns: species are the trait levels
phylo_slope(moisture + canopy | trait, tree = tree)
```

If your scientific question is about trait covariance among sampled species,
begin with the first form. If it is about environmental-response coefficients
varying across species columns, begin with the second. For the source ×
trait-covariance lookup, see the [Formula keyword grid](api-keyword-grid.html).
