---
title: "Joint ecological intensity from opportunistic occurrence records"
output: html_document
vignette: >
  %\VignetteIndexEntry{Joint ecological intensity from opportunistic occurrence records}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- LOCAL-ONLY STAGING DRAFT.  This file is not a package vignette, is not
rendered, and makes no public capability claim.  Its examples are mathematical
sketches until a separately validated public workflow exists. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Why model species jointly when the data are mostly zeros?

Consider a restoration programme that wants to compare six wetland species
across a landscape of reference and restored cells.  Opportunistic records from
a biodiversity portal are plentiful in accessible, popular places and scarce
elsewhere.  A separate single-species model for each taxon can describe six
maps, but it discards a useful biological premise: species can share an
unmeasured ecological gradient while retaining species-specific departures.

This staged article introduces the first part of an integrated joint-species
story: a *relative ecological-intensity* model for opportunistic records.  It
does not estimate abundance, detectability, absolute occurrence probability,
or a spatial field.  It also does not claim that a current released package
interface implements this workflow.  The purpose is to make the estimand and
the source-bias assumption visible before adding survey observations.

# One ecological state for all observation channels

Let (c=1,\ldots,C) index landscape cells and (s=1,\ldots,S) index species.
The ecological linear predictor is

\[
  \eta_{cs}=\alpha_s+x_c^\top\beta_s+\lambda_s z_c+e_{cs}.
\]

Here (x_c) contains measured environmental covariates.  The cell-level
factor (z_c\sim N(0,1)) is shared by all species, and the residual vector
\(e_c=(e_{c1},\ldots,e_{cS})^\top\) has

\[
  e_c\sim N(0,\Psi), \qquad \Psi=\operatorname{diag}(\psi_1^2,\ldots,\psi_S^2).
\]

Consequently the between-cell species covariance is

\[
  \Sigma=\Lambda\Lambda^\top+\Psi,
  \qquad \Lambda=(\lambda_1,\ldots,\lambda_S)^\top.
\]

The rank-one term captures a single unmeasured ecological gradient shared
across species.  The free diagonal \(\Psi\) prevents that factor from being
forced to explain every species-specific deviation.  This is a covariance
model, not a claim that the species interact causally.

# Opportunistic records are a point-process observation channel

Write (N^G_{cs}) for the opportunistic-record count, (a^G_c) for a known
quadrature support or exposure, and (b_c) for a covariate representing
recording bias.  The observation model is

\[
  N^G_{cs}\sim\operatorname{Poisson}(\mu^G_{cs}),\qquad
  \log\mu^G_{cs}=\log a^G_c+\eta_{cs}+b_c\gamma_s.
\]

The \(\gamma_s\) terms belong to the GBIF channel only.  They describe a
species-specific association between records and the named bias covariate;
they do not alter ecological intensity itself.  Identification therefore
depends on a design assumption: the bias covariate must vary separately enough
from the environmental predictors, and the portal channel must contain enough
information to estimate \(\gamma_s\).  That source gate is a property of the
study design, not a numerical convenience.

An observed zero is ordinary under this likelihood.  Low intensity, limited
quadrature support, and weak recording opportunity all make zero records
likely.  A generic zero-inflation term is not the first response to rare
species records because it introduces another process without supplying the
information needed to distinguish it from low intensity.

# A synthetic six-species thought experiment

For a realistic synthetic restoration landscape, let `env` increase from
reference to restored hydrology and let `bias` represent accessibility to
observers.  The following code is intentionally a *data-generating sketch*,
not a user-facing fit recipe.

```{r generate}
set.seed(20260811)
C <- 360L
S <- 6L
env <- seq(-1, 1, length.out = C)
bias <- as.numeric(scale(resid(lm(rnorm(C) ~ env))))
z <- rnorm(C)
Lambda <- c(0.50, -0.35, 0.40, -0.45, 0.30, 0.25)
psi <- c(0.35, 0.30, 0.40, 0.32, 0.38, 0.34)
eta <- outer(env, c(0.55, -0.35, 0.25, 0.45, -0.20, 0.15)) +
  tcrossprod(z, Lambda) +
  matrix(rnorm(C * S, sd = rep(psi, each = C)), C, S)
```

This simulation has a known ecological map and known covariance decomposition.
That lets an analyst ask a stronger question than “did the optimizer converge?”:
do the fitted environmental effects, GBIF-bias effects, ecological maps,
\(\Lambda\Lambda^\top\), and \(\Psi\) recover the values that generated the
data?  The private six-species campaign is designed around precisely those
five checks.  Its present numerical-admission gate is under review, so this
article deliberately does not report a recovery success rate.

# What this first article can and cannot establish

The model is useful when the scientific question concerns relative ecological
intensity and the portal-bias mechanism can be represented by named,
source-specific covariates.  It borrows strength across species without
equating their environmental responses.  It is not yet an integrated
occupancy/detection model, and it has no spatial ecological or spatial bias
field.  The next article adds repeated survey visits specifically to separate
observation detection from ecological occurrence.

# References to verify before promotion

Before this draft becomes a public article, add checked bibliography entries
for point-process and integrated-SDM foundations, GLLVM/JSDM latent-factor
models, and the relevant package interfaces.  It must also gain a runnable
synthetic public workflow only after the private recovery gate passes.
